### 2018-02-12 blueprint.tex
template for the blueprint proof of the forcing theorems

\documentclass[11pt,english]{article}
\usepackage[bottom=2cm, top=3cm, %% Comment if 
  left=4cm, right=4cm]{geometry}             %% using 'ebook' style

\input{header-blueprint}

%%%%%%%%%%%%%%%%%%% Footnotemarks for "Thanks!" %%%%%%%%%%%%%%%%%%
\makeatletter
\renewcommand*{\@fnsymbol}[1]{\ensuremath{\ifcase#1\or *\or \dagger\or \ddagger\or
    \mathsection\or \mathparagraph\or \|\or **\or \dagger\dagger
    \or \ddagger\ddagger \else\@ctrerr\fi}}
\makeatother
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\title{Forcing for Isabelle}
\author{%
  Pedro Sánchez Terraf%
  \thanks{
    Universidad Nacional de C\'ordoba. 
    Facultad de Matem\'atica, Astronom\'{\i}a,  F\'{\i}sica y
    Computaci\'on.}
  \thanks{
    Centro de Investigaci\'on y Estudios de Matem\'atica (CIEM-FaMAF).
    C\'ordoba. Argentina.}
}
%
\maketitle
%
\begin{abstract}
  Basic definitions of the forcing technique and blueprint for an
  Isabelle/ZF proof of the Forcing Theorems.   
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Introduction}
The proofs are mostly based on Chapter IV of
Kunen's~\cite{kunen2011set}.

\section{Definitions}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{alpha}
\bibliography{blueprint}

\bigskip

\begin{small}
  \begin{quote}
    \texttt{sterraf@famaf.unc.edu.ar}

    CIEM --- Facultad de Matem\'atica, Astronom\'{\i}a y F\'{\i}sica 
    (Fa.M.A.F.)

    Universidad Nacional de C\'ordoba --- Ciudad Universitaria

    C\'ordoba 5000. Argentina.
  \end{quote}
\end{small}

\end{document}

%%% Local Variables: 
%%% mode: latex
%%% ispell-local-dictionary: "american"
%%% End:

### 2018-02-12 blueprint.tex
some initial comments concerning implementation of formulas

@@
 \input{header-blueprint}
+\newcommand{\mt}[1]{$\mathtt{#1}$}
+\DeclareUnicodeCharacter{21D2}{\mt\Rightarrow}
+\DeclareUnicodeCharacter{2208}{\mt\in}
+\DeclareUnicodeCharacter{2200}{\mt\forall}
+\DeclareUnicodeCharacter{27F6}{\mt\longrightarrow}
+\DeclareUnicodeCharacter{27E6}{\mt\llbracket}
+\DeclareUnicodeCharacter{27E7}{\mt\rrbracket}
+\DeclareUnicodeCharacter{27F9}{\mt\Longrightarrow}
+
+\renewcommand{\PP}{\mathbb{P}}
+\newcommand{\formula}{\ensuremath{\mathtt{formula}}}
+\newcommand{\tyi}{\mathtt{i}}
+\newcommand{\tyo}{\mathtt{o}}
 
@@
   Basic definitions of the forcing technique and blueprint for an
-  Isabelle/ZF proof of the Forcing Theorems.   
+  Isabelle/ZF proof of the Forcing Theorems, under the assumption of
+  the existence of countable transitive model of $\ZFC$.   
 \end{abstract}
@@
 
-\section{Definitions}
+\section{Forcing notions}
+We will need the following concepts:
+\begin{itemize}
+\item Forcing poset or \emph{notion}: a triple $\lb \PP, \1, \leq\rb$ where $\leq$ is a
+  partial (pre)order on $\PP$ with (a) top element $\1$.
+\item Compatible elements. Antichain in $\PP$. 
+\item Dense subset $D\sbq\PP$ of a forcing poset.
+\item Filter $G\sbq\PP$ in a poset.
+\item $\calD$-generic filter for $\PP$, where $\calD$ is a family of
+  dense subsets of $\PP$.
+\end{itemize}
+
+Most probably, we'll need two or three versions of each of these
+definitions: One internalized as an element of \formula;  a relativized
+one with a class parameter; and finally one without
+relativization. This is done in order to show absolutness and to take
+advantage of the Isabelle/ZF ability to operate with terms of type
+$\tyi$.
+
+As an example, consider $\PP,R \in M$,  and $R\sbq
+\PP\times\PP$. Reflexivity of the relation $R$ is 
+stated as follows.
+\[
+\forall p (p\in \PP \limp \lb p,p\rb \in R).
+\]
+In Isabelle/ZF, we write
+%
+\begin{verbatim}
+definition 
+  reflexivity_abs :: "[i,i] ⇒ o" where
+  "reflexivity_abs(P,r) == ∀p . p∈P ⟶ <p,p>∈r"
+\end{verbatim}
+
+This is different from what would $M$ see from ``within.'' In
+particular, the universal quantifier should range only over elements
+of $M$:
+\[
+\forall p\in M (p\in \PP \limp \lb p,p\rb \in R).
+\]
+In Isabelle/ZF, we will move to a more general setting, by allowing
+that $M$ is a class.
+\begin{verbatim}
+definition  
+  reflexivity_rel :: "[i⇒o,i,i] ⇒ o" where
+  "reflexivity_rel(M,P,r) == ∀p[M].  p∈P ⟶ <p,p>∈r"
+\end{verbatim}
+(Notice: There must be no spaces between \verb|[M]| and the following
+period.) This is the relativized version of
+\verb|reflexivity_abs|.
+
+Finally, the internalized version of reflexivity is as follows:
+\begin{verbatim}
+definition
+  reflexivity_fm :: "[i,i]⇒i" where
+  "reflexivity_fm(x,y) == Forall(Implies(Member(0,succ(x)),
+          Exists(And(pair_fm(1,1,0),Member(0,succ(succ(y)))))))"
+
+lemma reflexivity_type : 
+        "⟦x∈nat ; y∈nat⟧ ⟹ reflexivity_fm(x,y)∈formula"
+  by (simp add:reflexivity_fm_def)
+\end{verbatim}
+
+The last lemma shows that indeed we have defined an element of the set
+\formula. The formula \verb|reflexivity_fm| is an internalization of
+the Isabelle/ZF formula \verb|reflexivity_abs|, not the relativized
+one.
 

### 2018-02-12 blueprint.tex
transitivity => absolutness

@@
 \[
-\forall p (p\in \PP \limp \lb p,p\rb \in R).
+\phi\defi \forall p (p\in \PP \limp \lb p,p\rb \in R).
 \]
@@
 \[
-\forall p\in M (p\in \PP \limp \lb p,p\rb \in R).
+\phi^M\defi\forall p\in M (p\in \PP \limp \lb p,p\rb \in R).
 \]
@@
 
-Finally, the internalized version of reflexivity is as follows:
+Finally, the internalized version $\quine{\phi}$ of reflexivity is as
+follows: 
 \begin{verbatim}
@@
 one.
+
+Under the assumption of transitivity of $M$, the three versions are
+equivalent. This is an \emph{absolutness result}:
+\[
+\phi(p,r) \lsii \phi^M(p,r) \lsii M;[p,r]\models\quine{\phi},
+\]
+when $p,r\in M$, and the list $[p,r]$ is the environment or
+interpretation of free variables.
 

### 2018-02-12 blueprint.tex
section on names

@@
 \input{header-blueprint}
+%%%%%%%%%%%%%%%%%%% % %   Unicodings   % % %%%%%%%%%%%%%%%%%%%%%
 \newcommand{\mt}[1]{$\mathtt{#1}$}
@@
 \newcommand{\tyo}{\mathtt{o}}
+\DeclareMathOperator{\val}{val}
 
@@
 
+\section{Names}
+The following is mostly independent of the previous section, but we
+assume that $\1\in\PP\in M$, $M$ is transitive and $1\in G\sbq\PP$.
+(Note that, in general, $G\notin M$.)
+
+\begin{definition}
+  \begin{enumerate}
+  \item The relation \verb|eee|$(M)$ is the restriction to $M$ of the
+    relational composition 
+    ${\in}\circ{\in}\circ{\in}$. 
+  \item For $\tau\in M$, the \emph{$G$-value} or \emph{interpretation
+    under $G$} of $\tau$ is
+    \[
+    \val(\tau,\PP,G)\defi \{\val(\sigma,\PP,G) : \exists p\in \PP
+    (\lb\sigma,p\rb\in\tau \y p\in G)\}.
+    \]
+  \end{enumerate}
+\end{definition}
+
+The definition of $\val$ proceeds by well founded recursion on
+\verb|eee|$(M)$. Determining  $\val(\tau,\PP,G)$ uses the values
+$\val(\sigma,\PP,G)$ for smaller $\sigma$:
+\[
+\sigma\in\{\sigma\}\in\lb\sigma,p\rb\in\tau.
+\]
+
+\begin{remark}
+  Slightly generalizing the recursive scheme for $\val$ we have:
+  \[
+  F(\tau,\bx)\defi \{F(\sigma,\bx) : \phi(\sigma,\tau,\bx) \y
+  \sigma\mathrel{R}\tau\}, 
+  \]
+  where $R$ is some well-founded relation. Using this scheme the
+  \emph{Mostowski collapse} of $R$ can be defined:
+  \[
+  \mos(\tau)\defi \{\mos(\sigma) : \sigma\mathrel{R}\tau\},
+  \]
+  which is a tool of independent interest.
+\end{remark}
+
+\begin{definition}
+  $\tau$ is a \emph{$\PP$-name} if its elements are of the form
+  $\lb\sigma,p\rb$ where $p\in\PP$ and $\sigma$ is a $\PP$.
+\end{definition}
+The usual way to define a property by well-founded recursion is to
+define the ``characteristic function'' of that property. In this case,
+define by recursion the function $F:M\to\om$ such that $F(x)=1$  if
+$x$ is a $\PP$-name, or $F(x)=0$ otherwise.
+
+\begin{definition}
+  The \emph{extension of $M$ by $G$} is
+  \[
+  \verb|gen_ext|(M,\PP,G) = M[G] \defi \{\val(\tau,\PP,G): \tau\in M\}.
+  \]
+\end{definition}
+The standard definition of $M[G]$ adds the clause that $\tau$ should be
+a $\PP$-name, but this is not necessary.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2018-02-12 blueprint.tex
typos

@@
 \item Forcing poset or \emph{notion}: a triple $\lb \PP, \1, \leq\rb$ where $\leq$ is a
-  partial (pre)order on $\PP$ with (a) top element $\1$.
+  partial (pre)order on $\PP$ with (a distinguished) top element $\1$.
 \item Compatible elements. Antichain in $\PP$. 
@@
 In Isabelle/ZF, we will move to a more general setting, by allowing
-that $M$ is a class.
+ $M$ to be a class.
 \begin{verbatim}

### 2018-02-12 blueprint.tex
definition of "M-generic"

@@
 \DeclareUnicodeCharacter{27F9}{\mt\Longrightarrow}
+\DeclareUnicodeCharacter{03A6}{\mt\Phi}
+%\DeclareUnicodeCharacter{}{\mt}
 
@@
 
+\begin{definition}
+  $G$ is \emph{$M$-generic for $\PP$} if it is $\calD$-generic for
+  $\PP$ for the family $\calD$ of all dense subsets of $\PP$ that live
+  in $M$.
+\end{definition}
+
 \section{Names}
@@
 The standard definition of $M[G]$ adds the clause that $\tau$ should be
-a $\PP$-name, but this is not necessary.
+a $\PP$-name, but this is not necessary. When $G$ is $M$-generic for
+$\PP$, we say that $M[G]$ is a \emph{generic extension}.
+
+\section{The forcing relation}
+In order to prove the independence of $\CH$, we will assume the
+existence of countable transitive model (ctm) $M$ of $\ZFC$. The
+satisfaction relation $M,\mathit{env}\models\Phi$ (where $\Phi$ is a set of
+sentences and $\mathit{env}$ is an interpretation for the free
+variables) is coded in Isabelle/ZF by using \verb|satT|:
+\begin{verbatim}
+definition
+  satT :: "[i,i,i] => o" where
+  "satT(A,Φ,env) == ∀ p ∈ Φ. sats(A,p,env)"
+\end{verbatim}
+The function \verb|satT| takes a subset of \formula{} as its second
+argument, and it is based on the already defined function \verb|sats|.
+
+Given a ctm $M$, and a $M$-generic filter $G\sbq\PP$, the Forcing
+Theorems relate satisfaction of a formula 
+$\phi$ in the generic extension $M[G]$ to the satisfaction of another formula
+$\phi'$ in $M$. The map $\phi\mapsto\phi'$ is defined by recursion on
+the structure of $\phi$.
 

### 2018-02-12 blueprint.tex
intro to forcing. Arriving to hard stuff

@@
   left=4cm, right=4cm]{geometry}             %% using 'ebook' style 
+\usepackage{framed}
 
@@
 \DeclareUnicodeCharacter{03A6}{\mt\Phi}
+\DeclareUnicodeCharacter{03BB}{\mt\lambda}
 %\DeclareUnicodeCharacter{}{\mt}
@@
 \newcommand{\tyo}{\mathtt{o}}
+\newcommand{\forceisa}{\mathtt{forces}}
 \DeclareMathOperator{\val}{val}
@@
 \section{Names}
-The following is mostly independent of the previous section, but we
-assume that $\1\in\PP\in M$, $M$ is transitive and $1\in G\sbq\PP$.
+\begin{framed}
+  The following is mostly independent of the previous section, but 
+  \begin{bfseries}
+    we assume that $\1\in\PP\in M$, $M$ is transitive and $1\in
+    G\sbq\PP$.
+  \end{bfseries}
   (Note that, in general, $G\notin M$.)
+\end{framed}
 
@@
 
-Given a ctm $M$, and a $M$-generic filter $G\sbq\PP$, the Forcing
+\begin{framed}
+  It might be necessary to redefine \verb|sats| and the related
+  function \verb|satisfies| in order to have an independent argument $B$
+  to be the range of values of the environment. We would have, for
+  example, 
+\begin{verbatim}
+consts   satisfies :: "[i,i]=>i"
+primrec 
+  "satisfies(A,B,Member(x,y)) =
+      (λenv ∈ list(B). bool_of_o (nth(x,env) ∈ nth(y,env)))"
+\end{verbatim}
+  \dots
+
+\noindent and then
+\begin{verbatim}
+abbreviation
+  sats :: "[i,i,i] => o" where
+  "sats(A,p,env) == satisfies(A,A,p)`env = 1"
+\end{verbatim}
+\end{framed}
+
+Given a ctm $M$, and an $M$-generic filter $G\sbq\PP$, the Forcing
 Theorems relate satisfaction of a formula 
@@
 $\phi'$ in $M$. The map $\phi\mapsto\phi'$ is defined by recursion on
-the structure of $\phi$.
+the structure of $\phi$. The fact that this map works (in particular,
+in order to show that $M[G]\models\ZFC$), we must use the fact that
+$M\models\ZFC$. Since we can only assert the latter using internalized
+formulas, the aformentioned map will be defined as a function from the
+set \formula{} into itself.
+
+Up to this point, the main reason for working with internalized
+versions is that  it is not possible to do recursion with formulas of
+type $\tyo$ (i.e., the formulas in the first-order logic of
+Isabelle/ZF).
+
+We will now make more precise the definition of the map
+$\phi\mapsto\phi'$ and how it relates satisfaction in $M$ to that in
+$M[G]$. Actually, if the formula $\phi$ has $n$ free variables,
+$\phi'$ will have $n+4$ free variables, where the first four account
+for the forcing notion and a particular condition. If
+$\phi=\phi(x_1,\dots,x_n)$, the standard notation for $\phi'$ is
+\[
+p\forces_{\PP,\leq,\1}^* \phi(\tau_1,\dots,\tau_n).
+\]
+Here, $\PP,\leq,\1,$ and $p\in\PP$ are the extra parameters. For the
+time being, we just consider $\tau_1,\dots,\tau_n$ to be ordinary
+variables. Informally, we call $\forces$ the \emph{forcing relation}.
+Given $\quine\phi\in\formula$, we will write the forcing relation in
+Isabelle/ZF by using  
+\[
+\forceisa(\quine\phi,\PP,\leq,\1)
+\]
+
+The base case (of atomic formulas) is famously the most
+complicated one.
 

### 2018-02-12 blueprint.tex
typo
first insight on why this is so damn difficult

@@
 one with a class parameter; and finally one without
-relativization. This is done in order to show absolutness and to take
+relativization. This is done in order to show absoluteness and to take
 advantage of the Isabelle/ZF ability to operate with terms of type
@@
 Under the assumption of transitivity of $M$, the three versions are
-equivalent. This is an \emph{absolutness result}:
+equivalent. This is an \emph{absoluteness result}:
 \[
@@
 \begin{framed}
-  The following is mostly independent of the previous section, but 
-  \begin{bfseries}
-    we assume that $\1\in\PP\in M$, $M$ is transitive and $1\in
-    G\sbq\PP$.
+  The following is mostly independent of the previous section, but %
+  \begin{bfseries}%
+    we assume that $\1\in\PP\in M$, $M$ is transitive and 
+    $1\in G\sbq\PP$.
   \end{bfseries}
@@
 
-The definition of $\val$ proceeds by well founded recursion on
+The definition of $\val$ proceeds by well-founded recursion on
 \verb|eee|$(M)$. Determining  $\val(\tau,\PP,G)$ uses the values
@@
 \[
-p\forces_{\PP,\leq,\1}^* \phi(\tau_1,\dots,\tau_n).
+p\forces_{\PP,\leq,\1}^* \phi(t_1,\dots,t_n).
 \]
-Here, $\PP,\leq,\1,$ and $p\in\PP$ are the extra parameters. For the
-time being, we just consider $\tau_1,\dots,\tau_n$ to be ordinary
-variables. Informally, we call $\forces$ the \emph{forcing relation}.
+Here, $\PP,\leq,\1,$ and $p\in\PP$ are the extra parameters. 
+%% For the
+%% time being, we just consider $\tau_1,\dots,\tau_n$ to be ordinary
+%% variables. 
+Informally, we call $\forces$ the \emph{forcing relation}.
 Given $\quine\phi\in\formula$, we will write the forcing relation in
@@
 \[
-\forceisa(\quine\phi,\PP,\leq,\1)
+\forceisa(\quine\phi,\PP,\leq,\1,p)
 \]
 
-The base case (of atomic formulas) is famously the most
+The base case (for atomic formulas) is famously the most
 complicated one. 
+\begin{multline}
+  \forceisa(\quine{t_1= t_2},\PP,\leq,\1,p) \defi 
+  \forall s\in\dom(t_1)\cup\dom(t_2)\;\forall q\leq p :\\
+  \forceisa(\quine{s\in t_1},\PP,\leq,\1,q)\lsii 
+  \forceisa(\quine{s\in t_1},\PP,\leq,\1,q),
+\end{multline}
+\begin{multline}
+  \forceisa(\quine{t_1\in t_2},\PP,\leq,\1,p) \defi  \\
+  \text{``}\{q\leq p : \exists \lb s,r\rb\in t_2 (q\leq r \y
+  \forceisa(\quine{t_1 =s},\PP,\leq,\1,q))\} \\
+  \text{is dense below $p$.''}
+\end{multline}
+
+Actually, these cases depend on the theorem of well-founded recursion,
+so the formula $\forceisa(\quine\phi,\PP,\leq,\1,p)$ for atomic $\phi$
+must be obtained from the proof of that theorem. This means that we need
+an internalized version of the theorem.
 

### 2018-02-13 blueprint.tex
first part of the base case of forcing

@@
 The proofs are mostly based on Chapter IV of
-Kunen's~\cite{kunen2011set}.
+Kunen's~\cite{kunen2011set}, and the implementation will make use of
+the work by Paulson~\cite{paulson_2003}.
 
@@
 \end{itemize}
+(Check \verb|ZF/Order.thy| before doing any work.)
 
@@
 definition 
-  reflexivity_abs :: "[i,i] ⇒ o" where
-  "reflexivity_abs(P,r) == ∀p . p∈P ⟶ <p,p>∈r"
+  reflexivity :: "[i,i] ⇒ o" where
+  "reflexivity(P,r) == ∀p . p∈P ⟶ <p,p>∈r"
 \end{verbatim}
@@
 period.) This is the relativized version of
-\verb|reflexivity_abs|.
+\verb|reflexivity|.
 
@@
 \formula. The formula \verb|reflexivity_fm| is an internalization of
-the Isabelle/ZF formula \verb|reflexivity_abs|, not the relativized
+the Isabelle/ZF formula \verb|reflexivity|, not the relativized
 one.
@@
 %% variables. 
-Informally, we call $\forces$ the \emph{forcing relation}.
-Given $\quine\phi\in\formula$, we will write the forcing relation in
+Given $\quine\phi\in\formula$, we will write this  in
 Isabelle/ZF by using  
@@
 \]
+Afterwards, the \emph{forcing relation} $\forces$ is defined by
+relativizing/interpreting $\forces^*$ in a ctm $M$, for fixed
+$\lb\PP,\leq,\1\rb\in M$:
+\[
+p\forces \phi(t_1,\dots,t_n) \defi 
+\bigl(p\forces_{\PP,\leq,\1}^* \phi(t_1,\dots,t_n)\bigr)^M\!.
+\]
 
-The base case (for atomic formulas) is famously the most
-complicated one. 
+\subsection{Forcing for atomic formulas}
+The base case of the definition of the forcing relation, that is, for
+atomic formulas, is (in)famously the most complicated one. 
+The definition adopted by Kunen~\cite[p.~257]{kunen2011set} is the
+following, with two mutually recursive clauses:
 \begin{multline}
@@
 \end{multline}
+Other approaches, like the one in Neeman~\cite{neeman-course}, prefer
+to have a single, more complicated, clause for $t_1=t_2$ using
+well-founded recursion on $t_1$
+and $t_2$ and then define the case $t_1\in t_2$ explicitly.
 
-Actually, these cases depend on the theorem of well-founded recursion,
+In any case, these definitions depend on the theorem of well-founded
+recursion,
 so the formula $\forceisa(\quine\phi,\PP,\leq,\1,p)$ for atomic $\phi$
-must be obtained from the proof of that theorem. This means that we need
-an internalized version of the theorem.
+must be obtained from the proof of that theorem. Below, a defined
+relation $R$ is \emph{set-like} if $R^{-1}\{a\}$ is a set for every $a$.
+%
+\begin{theorem}[{\cite[I.9.11]{kunen2011set}}]
+  Assume that $R$ is a
+  set-like well-founded defined relation on a class $A$, assume that 
+  $\forall  x,s\,\exists!y\,\phi(x,s,y)$ and write $G(x,s)$ for the
+  only $y$ such that $\phi(x,s,y)$. Then there exists another formula 
+   $\psi$ such that $\forall x\,\exists!y\,\psi(x,y)$ and if we write
+  $F(x)$ for the only $y$ satisfying $\psi(x,y)$, we have
+  \[
+  \forall a: F(a) = G(a,F\restr R^{-1}\{a\}).
+  \]
+\end{theorem}
+%
+Observe that since $F$ is a definable function, its restriction to the
+set $R^{-1}\{a\}$ must also be a set by the Replacement Axiom.
+
+There are several steps to apply this theorem to justify the
+definition of forcing for atomic formulas:
+\begin{enumerate}
+\item As mentioned before, $F$ will be the characteristic function of
+  the predicate $\forceisa$.
+  %
+\item The class $A$ is $V\times V = \{\lb x,y\rb : x,y\in V\}$,
+  i.e.\ the class of all ordered pairs. The relation $R$ must be
+  defined.
+  %
+\item Write down explicitly $\psi$ as an element of
+  $\formula$. 
+\end{enumerate}
+%
+Luckily, the definition of $R$ and the formula $\psi$ are spelled out
+in Kunen's.
 

### 2018-02-13 blueprint.tex
atomic formulas seem complete

@@
 \renewcommand{\PP}{\mathbb{P}}
+\renewcommand{\app}{\mathrm{App}}
 \newcommand{\formula}{\ensuremath{\mathtt{formula}}}
@@
 \newcommand{\forceisa}{\mathtt{forces}}
+\newcommand{\equ}{\mathbf{e}}
+\newcommand{\bel}{\mathbf{b}}
+\newcommand{\atr}{\mathit{atr}}
+\newcommand{\reee}{\mathrel{\mathtt{eee}}}
 \DeclareMathOperator{\val}{val}
@@
 following, with two mutually recursive clauses:
-\begin{multline}
+\begin{multline}\label{eq:def-forcing-equality}
   \forceisa(\quine{t_1= t_2},\PP,\leq,\1,p) \defi 
@@
 \end{multline}
-\begin{multline}
+\begin{multline}\label{eq:def-forcing-membership}
   \forceisa(\quine{t_1\in t_2},\PP,\leq,\1,p) \defi  \\
@@
   \[
-  \forall a: F(a) = G(a,F\restr R^{-1}\{a\}).
+  \forall a\in A: F(a) = G(a,F\restr R^{-1}\{a\}).
   \]
@@
   %
-\item The class $A$ is $V\times V = \{\lb x,y\rb : x,y\in V\}$,
-  i.e.\ the class of all ordered pairs. The relation $R$ must be
+\item The class $A$ and the relation $R$ must be
   defined.
@@
 %
-Luckily, the definition of $R$ and the formula $\psi$ are spelled out
-in Kunen's.
+Luckily, the formula $\psi$ is spelled out in Kunen's.
+
+Since the  function $F$ will essentially have 
+the same  arguments  of $\forceisa$, but we will deconstruct the
+formula and pack everything as a tuple. Namely, we will have
+\begin{align}
+  F(\lb\equ,t_1,t_2,\PP,\leq,\1,p\rb) =1 &\iff  
+  \forceisa(\quine{t_1=t_2},\PP,\leq,\1,p) \label{eq:forcing-F-equ}\\
+  F(\lb\bel,s,t_1,\PP,\leq,\1,p\rb) =1 &\iff  
+  \forceisa(\quine{s\in t_1},\PP,\leq,\1,q). \label{eq:forcing-F-bel}
+\end{align}
+Here, $\equ$ and $\bel$ are arbitrary sets that encode which type of
+formula we are considering. We can choose our $A$ to be
+$\{\equ,\bel\}\times V^6$ (i.e., all 7-tuples whose first coordinate
+are  $\equ$ or $\bel$).
+
+Rewriting the clauses (\ref{eq:def-forcing-equality}) and
+(\ref{eq:def-forcing-membership}), we might obtain a first
+approximation to $G$. Notice that its domain is $A\times V$.
+
+$G(\lb\atr,t_1,t_2,\PP,\leq,\1,p\rb,f) =1$ if one of the following hold:
+\begin{itemize}
+\item $\atr=\equ$ and   $\forall s\in\dom(t_1)\cup\dom(t_2)\;\forall q\leq p :
+  f(\lb\bel,s,t_1,\PP,\leq,\1,q\rb) =1  \lsii
+  f(\lb\bel,s,t_2,\PP,\leq,\1,q\rb) =1$; or
+\item $\atr=\bel$ and $\{q\leq p : \exists \lb s,r\rb\in t_2 (q\leq r \y
+  f(\lb\equ,t_1,s,\PP,\leq,\1,q\rb) =1)\}$ is dense below $p$.
+\end{itemize}
+In any other case, $G(\lb\atr,t_1,t_2,\PP,\leq,\1,p\rb,f) =0$. In the
+above, $f(v)=1$ should be interpreted as $\lb v,1\rb\in f$ (i.e., $f$
+must be defined for the argument at hand). From this scheme an element
+of $\formula$ must be produced.
+
+The definition of the relation $R$ is straightforward and follows the
+pattern of \cite[p.~257]{kunen2011set}. For instance, we have that 
+\[
+\lb\bel,s_1,t_1,\PP,\leq,\1,q\rb\mathrel{R}
+\lb\equ,s_2,t_2,\PP,\leq,\1,p\rb
+\]
+if and only if 
+\begin{itemize}
+\item $s_1 \reee  s_2$ or $s_1\reee t_2$; and
+\item $t_1 = s_2$ or $t_1 = t_2$.
+\end{itemize}
+%
+\begin{framed}
+  It seems necessary to prove that the relation $\reee$ is
+  well-founded independently of universe where one is working. We only
+  have a proof that ${\reee}(M)$ is well-founded for each $M$.
+\end{framed}
+
+With these elements we might reproduce the formula $\psi$ that defines
+$F$ as in \cite[p.~49]{kunen2011set}. Let  $\app(d,h)$  be the formula
+that states:
+\begin{enumerate}
+\item \label{item:dom-R-cerrado} $d$ is ``downwards $R$-closed'':
+  $\forall y \in d, 
+  R^{-1}\{y\}\subseteq d$;
+\item \label{item:4} $h$ is a function with $\dom h =d \subseteq A$; and
+\item \label{item:3} for all $y\in d$, $h(y) = G(y,h\restr(R^{-1}\{y\}))$.
+\end{enumerate}
+Then the formula defining $F$ is the following:
+\[
+(x\notin A \y y=0) \o \bigl(x\in A \y \exists d,h\, (\app(d,h)\y
+x\in d \y h(x)=y)\bigr).
+\]
+The well-foundedness of $R$ is used to show that this formula is indeed
+functional, and that it satisfies the required recursive equation
+involving $G$. Once this is all translated to an element of \formula,
+we'll be able to write the base case of the recursive definition of
+$\forceisa$ using Equations~(\ref{eq:forcing-F-equ})
+and~(\ref{eq:forcing-F-bel}).
 

### 2018-02-14 blueprint.tex
typo
inductive step of ⊩ and the Truth Lemma

@@
 \DeclareUnicodeCharacter{2200}{\mt\forall}
+\DeclareUnicodeCharacter{2203}{\mt\exists}
+\DeclareUnicodeCharacter{2227}{\mt\wedge}
 \DeclareUnicodeCharacter{27F6}{\mt\longrightarrow}
@@
   Isabelle/ZF proof of the Forcing Theorems, under the assumption of
-  the existence of countable transitive model of $\ZFC$.   
+  the existence of a countable transitive model of $\ZFC$.   
 \end{abstract}
@@
 \[
-\phi^M\defi\forall p\in M (p\in \PP \limp \lb p,p\rb \in R).
+\phi^M\defi\forall p\in M (p\in \PP \limp \lb p,p\rb^M \in R).
 \]
-In Isabelle/ZF, we will move to a more general setting, by allowing
- $M$ to be a class.
+The use of the relativized term $\lb p,p\rb^M$ implicitly assumes that
+there is an element of $M$ satisfying its defining formula. In
+Isabelle/ZF, we might move to a more general setting, by allowing  $M$
+to be a class. Here, the reference to the ordered pair  is through a
+relational definition.
 \begin{verbatim}
@@
   reflexivity_rel :: "[i⇒o,i,i] ⇒ o" where
-  "reflexivity_rel(M,P,r) == ∀p[M].  p∈P ⟶ <p,p>∈r"
+  "reflexivity_rel(M,P,r) == ∀p[M].  p∈P ⟶ 
+                             (∃x[M]. pair(M,p,p,x) ∧ x∈r)"
 \end{verbatim}
@@
 \end{definition}
-
+This last definition is \emph{not} to be relativized, since for
+nontrivial $\PP$, $G\notin M$. Absoluteness results as the one above
+always assume that the free variables range in the set or class to
+which one is relativizing.
 \section{Names}
@@
 \begin{multline}\label{eq:def-forcing-equality}
-  \forceisa(\quine{t_1= t_2},\PP,\leq,\1,p) \defi 
+  \forceisa({t_1= t_2},\PP,\leq,\1,p) \defi 
   \forall s\in\dom(t_1)\cup\dom(t_2)\;\forall q\leq p :\\
-  \forceisa(\quine{s\in t_1},\PP,\leq,\1,q)\lsii 
-  \forceisa(\quine{s\in t_1},\PP,\leq,\1,q),
+  \forceisa({s\in t_1},\PP,\leq,\1,q)\lsii 
+  \forceisa({s\in t_1},\PP,\leq,\1,q),
 \end{multline}
 \begin{multline}\label{eq:def-forcing-membership}
-  \forceisa(\quine{t_1\in t_2},\PP,\leq,\1,p) \defi  \\
+  \forceisa({t_1\in t_2},\PP,\leq,\1,p) \defi  \\
   \text{``}\{q\leq p : \exists \lb s,r\rb\in t_2 (q\leq r \y
-  \forceisa(\quine{t_1 =s},\PP,\leq,\1,q))\} \\
+  \forceisa({t_1 =s},\PP,\leq,\1,q))\} \\
   \text{is dense below $p$.''}
@@
 
-In any case, these definitions depend on the theorem of well-founded
-recursion,
-so the formula $\forceisa(\quine\phi,\PP,\leq,\1,p)$ for atomic $\phi$
+In any case, these definitions assume that $t_i$ are \emph{names}.
+In fact, the intuitive definition is by recursion on \emph{sentences}
+that are written in the language $\{\in\}$ plus every name as a
+constant; thus they
+depend on the theorem of well-founded recursion,
+so $\forceisa(\phi,\PP,\leq,\1,p)$ for atomic $\phi$
 must be obtained from the proof of that theorem. Below, a defined
@@
 
-Since the  function $F$ will essentially have 
+The  function $F$ will essentially have 
 the same  arguments  of $\forceisa$, but we will deconstruct the
@@
   F(\lb\equ,t_1,t_2,\PP,\leq,\1,p\rb) =1 &\iff  
-  \forceisa(\quine{t_1=t_2},\PP,\leq,\1,p) \label{eq:forcing-F-equ}\\
+  \forceisa(t_1=t_2,\PP,\leq,\1,p) \label{eq:forcing-F-equ}\\
   F(\lb\bel,s,t_1,\PP,\leq,\1,p\rb) =1 &\iff  
-  \forceisa(\quine{s\in t_1},\PP,\leq,\1,q). \label{eq:forcing-F-bel}
+  \forceisa(s\in t_1,\PP,\leq,\1,q). \label{eq:forcing-F-bel}
 \end{align}
@@
   It seems necessary to prove that the relation $\reee$ is
-  well-founded independently of universe where one is working. We only
+  well-founded independently of the universe $M$ where one is
+  working. We only
   have a proof that ${\reee}(M)$ is well-founded for each $M$.
@@
 and~(\ref{eq:forcing-F-bel}).
+
+\subsection{Inductive step}
+%
+The nexts steps in the definition of $\forceisa$ are straightforward
+\cite[IV.2.42]{kunen2011set}. We just highlight some of the cases.
+\begin{itemize}
+\item $\forceisa(\neg\phi,\PP,\leq,\1,p) \defi \neg\exists
+  q\leq p \,(\neg \forceisa(\phi,\PP,\leq,\1,q))$.
+\item
+  $\forceisa(\phi\y\psi,\dots)\defi\forceisa(\phi,\PP,\leq,\1,p)\y\forceisa(\psi,\PP,\leq,\1,p)$.
+\item $\forceisa(\forall x\,\phi(x),\PP,\leq,\1,p)\defi \forall x\,\forceisa(\phi(x),\PP,\leq,\1,p)$. 
+\item
+  $\forceisa(\phi\o\psi,\dots)\defi\text{``}\{q:\forceisa(\phi,\PP,\leq,\1,q)\o\forceisa(\psi,\PP,\leq,\1,q)\}$
+  is dense below $p$''. 
+\end{itemize}
+%
+Actually, we would be done with just the first three.
+
+\section{The fundamental theorems}
+These are stated as two lemmas: \emph{Truth} and
+\emph{Definability}. The second one ``simply'' says that the forcing
+relation is definable. The previous recursive construction amounts to
+that. The Truth Lemma states that the forcing relation indeed relates
+satisfaction in $M[G]$ to that in $M$. 
+\begin{lemma}[Truth Lemma]
+  Let $M$ be a ctm of $\ZFC$ and $\lb\PP,\leq,\1\rb$ a forcing notion
+  in $M$, $p\in\PP$, and $\phi(x_1,\dots,x_n)$ a formula in the
+  language of set 
+  theory with all free variables displayed. Then the
+  following are equivalent, for all $\tau_1,\dots,\tau_n\in M$: 
+  \begin{enumerate}
+  \item     $M[G]\models\phi(\val(\tau_1,G),\dots,\val(\tau_n,G))$ for
+    all filters $G$, $M$-generic for $\PP$, such that $p\in G$.
+  \item $M\models\forceisa(\phi(\tau_1,\dots,\tau_n),\PP,\leq,\1,p)$.
+  \end{enumerate}
+\end{lemma}
+A bit more formally, one should write the parameters to the left of
+$\models$:
+\begin{gather*}
+  M[G],[\val(\tau_1,G),\dots,\val(\tau_n,G)]
+  \models\phi(x_1,\dots,x_n) \\
+  \text{if and only if}\\
+  M,[\tau_1,\dots,\tau_n,\PP,\leq,\1,p] 
+  \models\forceisa(\phi(x_1,\dots,x_n),x_{n+1},x_{n+2},x_{n+3},x_{n+4})
+\end{gather*}
 

### 2018-02-14 blueprint.tex
truth lemma for atomic formulas, statement

@@
 
-\subsection{Forcing for atomic formulas}
+\subsection{Forcing for atomic formulas}\label{sec:forcing-atomic-formulas}
 The base case of the definition of the forcing relation, that is, for
@@
 satisfaction in $M[G]$ to that in $M$. 
-\begin{lemma}[Truth Lemma]
-  Let $M$ be a ctm of $\ZFC$ and $\lb\PP,\leq,\1\rb$ a forcing notion
+\begin{lemma}[Truth Lemma]\label{lem:truth-lemma}
+  Let $M$ be a ctm of $\ZFC$, $\lb\PP,\leq,\1\rb$ a forcing notion
   in $M$, $p\in\PP$, and $\phi(x_1,\dots,x_n)$ a formula in the
@@
 \begin{gather*}
-  M[G],[\val(\tau_1,G),\dots,\val(\tau_n,G)]
+  \forall G : p\in G \limp M[G],[\val(\tau_1,G),\dots,\val(\tau_n,G)]
   \models\phi(x_1,\dots,x_n) \\
@@
 \end{gather*}
+The proof of the Truth Lemma is by recursion in elements of
+$\formula$. The following auxiliary result (adapted from
+\cite[IV.2.43]{kunen2011set}) is also proved by recursion in
+$\formula$.
+\begin{lemma}\label{lem:3-density}
+  Under the same hypothesis of Lemma~\ref{lem:truth-lemma}, we have
+  \begin{enumerate}
+  \item $M\models\forceisa(\phi,\PP,\leq,\1,p)$ and $p_1\leq p$
+    implies $M\models\forceisa(\phi,\PP,\leq,\1,p_1)$.
+  \item $M\models\forceisa(\phi,\dots)$ if and only if
+    $M\models$``$\{p_1\leq p : \forceisa(\phi,\PP,\leq,\1,p_1)\}$ is
+    dense below $p$.''
+  \item $M\models\forceisa(\phi,\dots)$ if and only if 
+    $M\models\neg\exists
+  q\leq p \,(\neg \forceisa(\phi,\PP,\leq,\1,q))$.
+  \end{enumerate}
+\end{lemma}
+
+\subsection{Truth for atomic formulas}
+We restate Lemma~\ref{lem:truth-lemma} in an alternative way, for
+atomic formulas.
+\begin{lemma}
+  Let $M$ be a ctm of $\ZFC$, $\lb\PP,\leq,\1\rb$ a forcing notion
+  in $M$, $p\in\PP$, $G\sbq\PP$ $M$-generic for $\PP$ and $\phi(x_1,x_2)$ an
+  atomic formula. Then the following hold for all
+  $\tau_1,\tau_2\in M$:
+  \begin{enumerate}
+  \item If $p\in G$ and
+    \begin{equation} \label{eq:M-sats-forces-atom}
+      M,[\tau_1,\tau_2,\PP,\leq,\1,p] 
+      \models\forceisa(\phi(x_1,x_2),x_{3},x_{4},x_{5},x_{6})
+    \end{equation}
+    then
+    \begin{equation} \label{eq:MG-sats-atom}
+      M[G],[\val(\tau_1,G),\val(\tau_2,G)]
+      \models\phi(x_1,x_2).
+    \end{equation}
+  \item If (\ref{eq:MG-sats-atom}) holds 
+    then there is some $p\in G$ such that
+    (\ref{eq:M-sats-forces-atom}) holds for that $p$.
+  \end{enumerate}
+\end{lemma}
+This proof consists of two cases for $\phi$ (namely, $x_1\in x_2$ and
+$x_1=x_2$) and it is done by well-founded recursion over $R$ (defined
+in Section~\ref{sec:forcing-atomic-formulas}).
 

### 2018-02-17 blueprint.tex
minor edits: split a Lemma.

@@
 The proof of the Truth Lemma is by recursion in elements of
-$\formula$. The following auxiliary result (adapted from
+$\formula$. The following auxiliary results (adapted from
 \cite[IV.2.43]{kunen2011set}) is also proved by recursion in
 $\formula$.
-\begin{lemma}\label{lem:3-density}
+\begin{lemma}\label{lem:2-density}
   Under the same hypothesis of Lemma~\ref{lem:truth-lemma}, we have
@@
     dense below $p$.''
-  \item $M\models\forceisa(\phi,\dots)$ if and only if 
+  \end{enumerate}
+\end{lemma}
+\begin{lemma}\label{lem:forcing-negation}
+  Under the same hypothesis of Lemma~\ref{lem:truth-lemma}, we have
+  $M\models\forceisa(\phi,\dots)$ if and only if 
   $M\models\neg\exists
   q\leq p \,(\neg \forceisa(\phi,\PP,\leq,\1,q))$.
-  \end{enumerate}
 \end{lemma}
@@
 \subsection{Truth for atomic formulas}
+We start by proving  Lemma~\ref{lem:2-density} for atomic
+formulas.
+\begin{proof}[Proof of  Lemma~\ref{lem:2-density}]
+\end{proof}
+
 We restate Lemma~\ref{lem:truth-lemma} in an alternative way, for

### 2018-02-18 blueprint.tex
formula recursion uses Nand
splitting lemmas

@@
 %
-Actually, we would be done with just the first three.
+Actually, recursion over $\formula$ uses only one propositional
+connective (``nand'') so we need the combination of the first two to
+obtain $\forceisa(\mathtt{Nand(\phi,\psi)},\dots)$:
+\[
+\neg\exists
+  q\leq p :\neg\bigl(\forceisa(\phi,\PP,\leq,\1,q)\y\forceisa(\psi,\PP,\leq,\1,q)\bigr).
+\]
 
@@
 $\formula$.
-\begin{lemma}\label{lem:2-density}
-  Under the same hypothesis of Lemma~\ref{lem:truth-lemma}, we have
-  \begin{enumerate}
-  \item $M\models\forceisa(\phi,\PP,\leq,\1,p)$ and $p_1\leq p$
+\begin{lemma}\label{lem:strengthen} 
+  Assume the same hypothesis of Lemma~\ref{lem:truth-lemma}.
+  $M\models\forceisa(\phi,\PP,\leq,\1,p)$ and $p_1\leq p$
   implies $M\models\forceisa(\phi,\PP,\leq,\1,p_1)$.
-  \item $M\models\forceisa(\phi,\dots)$ if and only if
+\end{lemma}
+\begin{lemma}\label{lem:density}
+  Assume the same hypothesis of Lemma~\ref{lem:truth-lemma}. $M\models\forceisa(\phi,\dots)$ if and only if
     $M\models$``$\{p_1\leq p : \forceisa(\phi,\PP,\leq,\1,p_1)\}$ is
     dense below $p$.''
-  \end{enumerate}
 \end{lemma}
@@
 \subsection{Truth for atomic formulas}
-We start by proving  Lemma~\ref{lem:2-density} for atomic
-formulas.
-\begin{proof}[Proof of  Lemma~\ref{lem:2-density}]
+We start by proving  Lemma~\ref{lem:strengthen}.% and~\ref{lem:density}.
+\begin{proof}[Proof of  Lemma~\ref{lem:strengthen} (atomic formulas)]
+  Let $M$, $\lb\PP,\leq,\1\rb$, $p\in\PP$ as in the hypothesis of
+  Lemma~\ref{lem:truth-lemma}, let $p_1\leq p$ and let $\phi(x_1,x_2)$
+  be atomic. We 
+  have to show that for every    $\tau_1,\tau_2\in M$,
+  \begin{equation} \label{eq:M-phi-atom}
+    M,[\tau_1,\tau_2,\PP,\leq,\1,p] 
+    \models\forceisa(\phi(x_1,x_2),x_{3},x_{4},x_{5},x_{6})
+  \end{equation}
+  then the same holds for the environment
+  $[\tau_1,\tau_2,\PP,\leq,\1,q]$.
 \end{proof}

### 2018-02-20 blueprint.tex
first proof, Strengthening Lemma.

@@
 \end{align}
-Here, $\equ$ and $\bel$ are arbitrary sets that encode which type of
-formula we are considering. We can choose our $A$ to be
+Here, $\equ$ and $\bel$ are arbitrary definable sets that encode which type of
+formula we are considering (for instance, $0$ and $1$, 
+respectively). We can choose our $A$ to be 
 $\{\equ,\bel\}\times V^6$ (i.e., all 7-tuples whose first coordinate
@@
 must be defined for the argument at hand). From this scheme an element
-of $\formula$ must be produced.
+of $\formula$ must be produced. (This is the reason why  $\equ,\bel$
+were chosen to be definable.)
 
@@
 functional, and that it satisfies the required recursive equation
-involving $G$. Once this is all translated to an element of \formula,
+involving $G$. The following intermediate result is needed in the
+sequel.
+\begin{lemma}[Uniqueness of approximations]\label{lem:uniq-approx}
+  Assume $\app(d,h)$ and $\app(d',h')$. Then
+  $h(x)=h'(x)$ for all  $x\in d \cap d'$. In particular,
+  \[
+  h\restr(\{x\}\cup R^{-1}\{x\})=h'\restr(\{x\}\cup R^{-1}\{x\})
+  \]
+  for  all $x\in d \cap d'$. 
+\end{lemma}
+
+Once this is all translated to an element of \formula,
 we'll be able to write the base case of the recursive definition of
@@
 $\formula$. The following auxiliary results (adapted from
-\cite[IV.2.43]{kunen2011set}) is also proved by recursion in
+\cite[IV.2.43]{kunen2011set}) are also proved by recursion in
 $\formula$.
@@
   be atomic. We 
-  have to show that for every    $\tau_1,\tau_2\in M$,
+  have to show that for every    $\tau_1,\tau_2\in M$, if
   \begin{equation} \label{eq:M-phi-atom}
@@
   then the same holds for the environment
-  $[\tau_1,\tau_2,\PP,\leq,\1,q]$.
+  $[\tau_1,\tau_2,\PP,\leq,\1,p_1]$. For this, we will deconstruct the
+  definition of $\forceisa$ in several steps to pin down where the
+  hypothesis of $p_1\leq p$ might be used.
+
+  So, let $\phi$ be $x_1\in x_2$ and assume (\ref{eq:M-phi-atom}). By
+  our definition of $\forceisa$  for atomic formulas, we have
+  $F(\bx)=1$, where
+  $\bx=\lb\equ,\tau_1,\tau_2,\PP,\leq,\1,p\rb$. Expanding the
+  definition of $F$, we obtain that
+  \[
+  (\bx\notin A \y 1=0) \o \bigl(\bx\in A \y \exists d,h\, (\app(d,h)\y
+  \bx\in d \y h(\bx)=1)\bigr)
+  \]
+  holds in $M$. 
+  \begin{framed}
+    By using appropriate absoluteness results we might translate this
+    statement written as an element of $\formula$ to a sentence
+    written in Isabelle/ZF first order logic, making theorems to
+    simplify the statement available, as in the arguments
+    that follow.
+  \end{framed}
+  \noindent Simplifying, we obtain 
+  \begin{equation}\label{eq:1}
+    \exists d,h\, (\app(d,h)\y   \bx\in d \y h(\bx)=1).
+  \end{equation}
+  Take such $d,h\in M$; these will be added to the environment. We
+  will define $d',h'$ that satisfy 
+  $\app(d',h')\y   \by\in d' \y h(\by)=1$ for
+  $\by=\lb\equ,\tau_1,\tau_2,\PP,\leq,\1,p_1\rb$, thereby proving
+  $F(\by)=1$ and hence the thesis.
+
+  By  definition of $\app$ and the fact that $\bx\in d$, we have
+  \[
+  1= h(\bx) = G(\bx,h\restr(R^{-1}\{\bx\})).
+  \]
+  Now the  definition of $G$ for $\bx=\lb\equ,\dots\rb$, yields
+  \[
+  M,[d,h,\tau_1,\tau_2,\PP,\leq,\1,p]\models \forall
+  s\in\dom(x_1)\cup\dom(x_2)\;\forall q\leq x_6 : \Phi
+  \]
+  with $\Phi\defi
+  f(\lb\bel,s,x_1,x_{3},x_{4},x_{5},x_{6}\rb) =1  \lsii
+  f(\lb\bel,s,x_2,x_{3},x_{4},x_{5},x_{6}\rb) =1$ and 
+  $f\defi h\restr(R^{-1}\{\bx\})$.
+  By absoluteness we have
+  \[
+  \forall
+  \sigma\in\dom(\tau_1)\cup\dom(\tau_2)\;\forall q'\leq p :
+  M,[\sigma,q',d,h,\tau_1,\tau_2,\PP,\leq,\1,p]\models  \Phi.
+  \]
+  Since $q'\leq p_1$ implies $q'\leq p$, we have
+  \[
+  \forall
+  \sigma\in\dom(\tau_1)\cup\dom(\tau_2)\;\forall q'\leq p_1 :
+  M,[\sigma,q',d,h,\tau_1,\tau_2,\PP,\leq,\1,p_1]\models  \Phi.
+  \]
+  Using absoluteness again we have
+  \[
+  M,[d,h,\tau_1,\tau_2,\PP,\leq,\1,p_1]\models \forall
+  s\in\dom(x_1)\cup\dom(x_2)\;\forall q\leq p_1 : \Phi.
+  \]
+  Hence we obtain $G(\by,h\restr(R^{-1}\{\bx\}))=1$.
+
+  Now define   $d'\defi d\cup \{\by\}$  and $h'\defi
+  h\cup\{\lb\by,1\rb\}$. Since
+  $R^{-1}\{\bx\} =R^{-1}\{\by\}$ and $\by\in A$, we obtain
+  $\app(d',h')$. This, together with the fact that $h(\by)=1$, shows
+  $F(\by)=1$ and hence (\ref{eq:M-phi-atom}).
+
+  The proof for $\phi= x_1\in x_2$ should be similar.
 \end{proof}
+Lemma~\ref{lem:density} also follows the same pattern, there is a bit
+more of mathematics to it. The same goes for
+Lemma~\ref{lem:forcing-negation}.
 

### 2018-02-25 blueprint.tex
change variable names

@@
 \[
-\phi\defi \forall p (p\in \PP \limp \lb p,p\rb \in R).
+\phi(\PP,R)\defi \forall p (p\in \PP \limp \lb p,p\rb \in R).
 \]
@@
 \[
-\phi(p,r) \lsii \phi^M(p,r) \lsii M;[p,r]\models\quine{\phi},
+\phi(\PP,R) \lsii \phi^M(\PP,R) \lsii M;[\PP,R]\models\quine{\phi},
 \]
-when $p,r\in M$, and the list $[p,r]$ is the environment or
+when $\PP,R\in M$, and the list $[\PP,R]$ is the environment or
 interpretation of free variables.
@@
 \end{align}
-Here, $\equ$ and $\bel$ are arbitrary definable sets that encode which type of
+Here, $\equ$ and $\bel$ are absolute constants that encode which type of
 formula we are considering (for instance, $0$ and $1$, 

### 2018-04-12 blueprint.tex
prueba de separación, primera parte

@@
 \newcommand{\atr}{\mathit{atr}}
+\newcommand{\concat}{\mathbin{@}}
 \newcommand{\reee}{\mathrel{\mathtt{eee}}}
@@
   $\tau$ is a \emph{$\PP$-name} if its elements are of the form
-  $\lb\sigma,p\rb$ where $p\in\PP$ and $\sigma$ is a $\PP$.
+  $\lb\sigma,p\rb$ where $p\in\PP$ and $\sigma$ is a $\PP$-name.
 \end{definition}
@@
 
+\section{The axioms in the generic extension}
+\subsection{The axiom of Separation}
+In this subsection we will show that the Separation Axiom scheme holds
+in the generic extension $M[G]$.
+
+\begin{lemma}
+  Let $G$ be $M$-generic for $\PP$. For every $\phi\in\formula$ with
+  three free variables, the $\phi$-instance of Separation holds in
+  $M[G]$.
+\end{lemma}
+\begin{proof}
+  Let $G$ as in the hypothesis, $\phi(x,z,v)$ be in \formula{} and let
+  $c,w\in M[G]$. Following Kunen \cite[Lemma IV.2.6]{kunen2011set},
+  we have to show that 
+  \[
+  S\defi\{x\in c : M[G]\models \phi(x,c,w)\} \in M[G].
+  \]
+  From this, using absoluteness, we will be able to derive
+  $\phi$-instance of Separation. Let $\pi,\sig\in M$ such that 
+  $\val(\pi)=c$ and $\val(\sig)=w$.
+  We will show that the name
+  \[
+  \tau \defi \{\lb\th,p\rb: \th\in\dom(\pi) \y p\in \PP\y
+  M\models p\forces(\th\in\pi\y\phi(\th,\pi,\sig)\}
+  \]
+  belongs to $M$; for this we will use that an appropriate instance of
+  Separation holds in $M$. Later we will see that $\tau$ names $S$.
+
+  First, we slightly rewrite $\tau$:
+  \[
+  \tau \defi \{w\in\dom(\pi)\times\PP :\exists \th,p (w=\lb\th,p\rb \y 
+  M\models p \forces(\th\in\pi\y\phi(\th,\pi,\sig))\}
+  \]
+
+  The formula giving the relevant instance of Separation in $M$ will be the
+  following:
+  \[
+  \psi \defi \exists \th,p : x=\lb\th,p\rb \y \forceisa(\PP,\leq,\1,p,\th\in\pi\y\phi(\th,\pi,\sig)),
+  \]
+  where $\PP,\leq,\1,\pi,\sig$ are parameters.
+  \begin{framed}
+    I'm changing the order of arguments of $\forceisa$. Also, we
+    will need to show that, in $M$, a version of separation for formulas
+    with  $5$ free variables holds.
+
+    Also, we need to show that $\dom(\pi)\times\PP\in M$.
+  \end{framed}
+  Writing this in de Bruijn notation, gives us something like this:
+  \[
+  \psi \defi  \exists\,\exists\,\verb|pair_fm|(0,1,2) \y
+  \forceisa(4,5,6,7,0\in 1\y\phi(0,1,3)).
+  \]
+  From $M\models \mathtt{Separation}(\psi)$ we have
+  \begin{multline*}
+    \forall u,l,P,s,r\in M \forall d\in M:\\ 
+    M,[d,r,s,P,l,u] \models\exists y \forall x
+    (x\in y \lsii x\in d \y \psi(P,l,u,x,r)).
+  \end{multline*}
+  We instantiate the universal quantifiers using the given parameters
+  and the domain $\dom(\pi)\times\PP$, while changing the free variables in
+  the formula with de Bruijn indices corresponding to the order in the
+  environment.
+  \[
+  M,[\dom(\pi)\times\PP,\pi,\sig,\PP,\leq,\1] \models\exists y \forall x
+  (x\in y \lsii x\in 0 \y \psi(3,4,5,x,1)).
+  \]
+  Extracting the two outermost quantifiers and shifting indices, we get
+  \begin{multline*}
+    \exists y\in M \forall x\in M: \\
+    M,[x,y,\dom(\pi)\times\PP,\pi,\sig,\PP,\leq,\1] \models
+    0\in 1 \lsii 0\in 2 \y \psi(5,6,7,0,3).
+  \end{multline*}
+  Let $\tau\in M$ witness the $\exists y$. So, by absoluteness,
+  \begin{multline*}
+    \forall x\in M: 
+    x\in\tau \lsii  x\in\dom(\pi)\times\PP \y{}\\ 
+  M,[x,\tau,\dom(\pi)\times\PP,\pi,\sig,\PP,\leq,\1] \models
+  \psi(5,6,7,0,3).
+  \end{multline*}
+  Let $\mathit{env}\defi
+  [x,\tau,\dom(\pi)\times\PP,\pi,\sig,\PP,\leq,\1]$. By transitivity
+  of $M$ we can dispense of the bound on $x$, and we get
+  \[
+  \tau = \{x\in\dom(\pi)\times\PP :
+  M, \mathit{env}\models
+  \psi(5,6,7,0,3)\},
+  \]
+  i.e., the relevant name is in $M$.
+
+  Now we turn to the proof that $\val(\tau)=S$.
+
+  Let's first focus into the predicate defining $\tau$ by separation. By expanding $\psi$ we have
+  \[
+   M, \mathit{env}\models
+   \exists \th,p : 0=\lb\th,p\rb \y
+   \forceisa(5,6,7,p,\th\in 3\y\phi(\th,3,4)).
+  \]
+  We proceed as before, using absoluteness and replacing variables by
+  corresponding indices.
+  \[
+  \exists \th,p : x =\lb\th,p\rb \y  M, [\th,p]\concat\mathit{env}\models
+  \forceisa(7,8,9,1,0\in 5\y\phi(0,5,6)).
+  \]
+
+  Considering that only a fraction of elements
+  of the environment  are necessary to interpret
+  $\forceisa(7,8,9,1,0\in\pi\y\phi(0,5,6))$ and by permuting, we can
+  replace
+  \[
+  M, [\th,p,x,\tau,\dom(\pi)\times\PP,\pi,\sig,\PP,\leq,\1]\models
+  \forceisa(7,8,9,1,0\in 5 \y\phi(0,5,6))
+  \]
+  by
+  \[
+  M,[\PP,\leq,\1,p,\th,\pi,\sig]\models
+  \forceisa(0,1,2,3,4\in 5 \y\phi(4,5,6))
+  \]
+  By the Truth Lemma, this is equivalent to
+  \[
+  \forall F\ \PP\text{-generic}, p\in F \limp
+  M[F],[\val(\th),\val(\pi),\val(\sig)]\models
+  0\in1  \y\phi(0,1,2).
+  \]
+
+  Now we calculate $\val(\tau)$.
+  \begin{multline*}
+    \val(\tau) = \\
+    \val\bigl(\{x\in\dom(\pi)\times\PP : \exists \th,p : x =\lb\th,p\rb
+    \y  \\
+    \forall F\ \PP\text{-generic}, p\in F \limp
+    M[F],[\val(\th),\val(\pi),\val(\sig)]\models
+    0\in 1 \y\phi(0,1,2)\}\bigr).
+  \end{multline*}
+  Hence
+  \begin{multline*}
+    \val(\tau) = 
+    \{\val(\th): \th\in\dom(\pi) \y  p \in G 
+    \y \forall F\ \PP\text{-generic}, \\
+   p\in F \longrightarrow
+  M[F],[\val(\th),\val(\pi),\val(\sig)]\models
+  0\in 1 \y\phi(0,1,2)\}.
+  \end{multline*}
+  This set includes
+  \begin{multline*}
+    \{\val(\th): \th\in\dom(\pi) \y  p \in G 
+    \y  \\
+   p\in G \longrightarrow
+   M[G],[\val(\th),\val(\pi),\val(\sig)]\models
+   0\in 1 \y\phi(0,1,2)\},
+  \end{multline*}
+  that is
+  \begin{multline*}
+    \{\val(\th): \th\in\dom(\pi) \y  p \in G 
+    \y 
+   M[G],[\val(\th),\val(\pi),\val(\sig)]\models
+   0\in 1 \y\phi(0,1,2)\}
+  \end{multline*}
+\end{proof}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2018-04-12 blueprint.tex
una inclusión probada. Cambio look de índices de de Bruijn

@@
 \newcommand{\reee}{\mathrel{\mathtt{eee}}}
+\newcommand{\dB}[1]{\mathbf{#1}}
 \DeclareMathOperator{\val}{val}
@@
   \]
-  From this, using absoluteness, we will be able to derive
+  From this, using absoluteness, we will be able to derive the
   $\phi$-instance of Separation. Let $\pi,\sig\in M$ such that 
@@
   \[
-  \psi \defi  \exists\,\exists\,\verb|pair_fm|(0,1,2) \y
-  \forceisa(4,5,6,7,0\in 1\y\phi(0,1,3)).
+  \psi \defi  \exists\,\exists\,\verb|pair_fm|(\dB{0},\dB{1},\dB{2}) \y
+  \forceisa(\dB{4},\dB{5},\dB{6},\dB{7},\dB{0}\in \dB{1}\y\phi(\dB{0},\dB{1},\dB{3})).
   \]
@@
   M,[\dom(\pi)\times\PP,\pi,\sig,\PP,\leq,\1] \models\exists y \forall x
-  (x\in y \lsii x\in 0 \y \psi(3,4,5,x,1)).
+  (x\in y \lsii x\in \dB{0} \y \psi(\dB{3},\dB{4},\dB{5},x,\dB{1})).
   \]
@@
     M,[x,y,\dom(\pi)\times\PP,\pi,\sig,\PP,\leq,\1] \models
-    0\in 1 \lsii 0\in 2 \y \psi(5,6,7,0,3).
+    \dB{0}\in \dB{1} \lsii \dB{0}\in \dB{2} \y \psi(\dB{5},\dB{6},\dB{7},\dB{0},\dB{3}).
   \end{multline*}
@@
   M,[x,\tau,\dom(\pi)\times\PP,\pi,\sig,\PP,\leq,\1] \models
-  \psi(5,6,7,0,3).
+  \psi(\dB{5},\dB{6},\dB{7},\dB{0},\dB{3}).
   \end{multline*}
@@
   M, \mathit{env}\models
-  \psi(5,6,7,0,3)\},
+  \psi(\dB{5},\dB{6},\dB{7},\dB{0},\dB{3})\},
   \]
@@
    M, \mathit{env}\models
-   \exists \th,p : 0=\lb\th,p\rb \y
-   \forceisa(5,6,7,p,\th\in 3\y\phi(\th,3,4)).
+   \exists \th,p : \dB{0}=\lb\th,p\rb \y
+   \forceisa(\dB{5},\dB{6},\dB{7},p,\th\in \dB{3}\y\phi(\th,\dB{3},\dB{4})).
   \]
@@
   \exists \th,p : x =\lb\th,p\rb \y  M, [\th,p]\concat\mathit{env}\models
-  \forceisa(7,8,9,1,0\in 5\y\phi(0,5,6)).
+  \forceisa(\dB{7},\dB{8},\dB{9},\dB{1},\dB{0}\in \dB{5}\y\phi(\dB{0},\dB{5},\dB{6})).
   \]
@@
   of the environment  are necessary to interpret
-  $\forceisa(7,8,9,1,0\in\pi\y\phi(0,5,6))$ and by permuting, we can
+  $\forceisa(\dB{7},\dB{8},\dB{9},\dB{1},\dB{0}\in\pi\y\phi(\dB{0},\dB{5},\dB{6}))$ and by permuting, we can
   replace
@@
   M, [\th,p,x,\tau,\dom(\pi)\times\PP,\pi,\sig,\PP,\leq,\1]\models
-  \forceisa(7,8,9,1,0\in 5 \y\phi(0,5,6))
+  \forceisa(\dB{7},\dB{8},\dB{9},\dB{1},\dB{0}\in \dB{5} \y\phi(\dB{0},\dB{5},\dB{6}))
   \]
@@
   M,[\PP,\leq,\1,p,\th,\pi,\sig]\models
-  \forceisa(0,1,2,3,4\in 5 \y\phi(4,5,6))
+  \forceisa(\dB{0},\dB{1},\dB{2},\dB{3},\dB{4}\in \dB{5} \y\phi(\dB{4},\dB{5},\dB{6}))
   \]
@@
   \forall F\ \PP\text{-generic}, p\in F \limp
-  M[F],[\val(\th),\val(\pi),\val(\sig)]\models
-  0\in1  \y\phi(0,1,2).
+  M[F],[\val(\th,F),\val(\pi,F),\val(\sig,F)]\models
+  \dB{0}\in\dB{1}  \y\phi(\dB{0},\dB{1},\dB{2}).
   \]
 
-  Now we calculate $\val(\tau)$.
+  Now we calculate $\val(\tau,F)$.
   \begin{multline*}
-    \val(\tau) = \\
+    \val(\tau,G) = \\
     \val\bigl(\{x\in\dom(\pi)\times\PP : \exists \th,p : x =\lb\th,p\rb
@@
     \forall F\ \PP\text{-generic}, p\in F \limp
-    M[F],[\val(\th),\val(\pi),\val(\sig)]\models
-    0\in 1 \y\phi(0,1,2)\}\bigr).
+    M[F],[\val(\th,F),\val(\pi,F),\val(\sig,F)]\models
+    \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2})\}\bigr).
   \end{multline*}
@@
   \begin{multline*}
-    \val(\tau) = 
-    \{\val(\th): \th\in\dom(\pi) \y  p \in G 
+    \val(\tau,G) = 
+    \{\val(\th,G): \th\in\dom(\pi) \y  p \in G 
     \y \forall F\ \PP\text{-generic}, \\
    p\in F \longrightarrow
-  M[F],[\val(\th),\val(\pi),\val(\sig)]\models
-  0\in 1 \y\phi(0,1,2)\}.
+  M[F],[\val(\th,F),\val(\pi,F),\val(\sig,F)]\models
+  \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2})\}.
   \end{multline*}
-  This set includes
+  This is included in
   \begin{multline*}
-    \{\val(\th): \th\in\dom(\pi) \y  p \in G 
+    \{\val(\th,G): \th\in\dom(\pi) \y  p \in G 
     \y  \\
    p\in G \longrightarrow
-   M[G],[\val(\th),\val(\pi),\val(\sig)]\models
-   0\in 1 \y\phi(0,1,2)\},
+   M[G],[\val(\th,G),\val(\pi,G),\val(\sig,G)]\models
+   \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2})\},
   \end{multline*}
-  that is
-  \begin{multline*}
-    \{\val(\th): \th\in\dom(\pi) \y  p \in G 
+  and in turn,  replacing $\val(\pi,F),\val(\sig,F)$ by their values,
+  this is included  in 
+  \[ 
+    \{\val(\th,G): \th\in\dom(\pi)
     \y 
-   M[G],[\val(\th),\val(\pi),\val(\sig)]\models
-   0\in 1 \y\phi(0,1,2)\}
-  \end{multline*}
+   M[G],[\val(\th,G),c,w]\models
+   \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2})\}.
+  \]
+  By absoluteness, we have that $\val(\tau,G)$ is
+  \[
+    \{\val(\th,G): \th\in\dom(\pi)
+    \y \val(\th,G)\in c \y 
+   M[G],[\val(\th,G),c,w]\models
+   \phi(\dB{0},\dB{1},\dB{2})\}.
+  \]
+  which equals 
+  \[
+    \{x: x\in c \y 
+    M[G],[x,c,w]\models
+   \phi(\dB{0},\dB{1},\dB{2})\}.
+  \]
+  since for every $x\in c=\val(\pi,G)$ there must
+  exist some $\th\in\dom(\pi)$ with $\val(\th,G)=x$. Therefore
+  $\val(\tau,G)\subseteq S$.
 \end{proof}

### 2018-04-12 blueprint.tex
aparentemente está concluida la prueba de Separación

@@
   Hence
-  \begin{multline*}
+  \begin{multline}\label{eq:val_tau}
     \val(\tau,G) = 
@@
   \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2})\}.
-  \end{multline*}
+  \end{multline}
   This is included in
@@
   $\val(\tau,G)\subseteq S$.
+
+  Now let $x\in S$. Hence there exists $\lb\th,q\rb\in\pi$ such that
+  $q\in G$ and $x=\val(\th,G)$. By the second version of the Truth
+  Lemma, since 
+  \[
+  M[G],[\val(\th,G),\val(\pi,G),\val(\sig,G)]\models
+  \dB{0}\in\dB{1}\y\phi(\dB{0},\dB{1},\dB{2}),
+  \]
+  there must exists $r\in G$ such that 
+  \[
+  M,[\PP,\leq,\1,r,\th,\pi,\sig]\models
+  \forceisa(\dB{0},\dB{1},\dB{2},\dB{3},
+  (\dB{4}\in\dB{5}\y\phi(\dB{4},\dB{5},\dB{6}))).
+  \]
+  Since $G$ is a filter, there is $p\in G$ such that $p\leq q, r$.
+  By Lemma~\ref{lem:strengthen}, we have
+  \[
+  M,[\PP,\leq,\1,p,\th,\pi,\sig]\models
+  \forceisa(\dB{0},\dB{1},\dB{2},\dB{3},
+  (\dB{4}\in\dB{5}\y \phi(\dB{4},\dB{5},\dB{6}))),
+  \]
+  which by the Truth Lemma gives us
+  \[
+  \forall F\ \PP\text{-generic}, \\
+   p\in F \longrightarrow
+  M[F],[\val(\th,F),\val(\pi,F),\val(\sig,F)]\models
+  \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2}).
+  \]
+  Using now (\ref{eq:val_tau}), this shows that $S\sbq \val(\tau,G)$,
+  since $\th\in\dom(\pi)$. 
 \end{proof}

### 2018-06-16 LSFA/first_steps_into_forcing.tex
template for LSFA paper

\documentclass[9pt]{entcs} \usepackage{entcsmacro}
\usepackage{graphicx}
\sloppy

\input{header-draft}
\def\lastname{Gunther, Pagano, S\'anchez Terraf}
\begin{document}
\begin{frontmatter}
  \title{First steps towards a formalization of Forcing}
  \author{Emmanuel Gunther%\thanksref{ALL}
    \thanksref{myemail}}
  \address{FaMAF\\ Universidad Nacional de C\'ordoba\\
    C\'ordoba, Argentina} \author{Miguel Pagano\thanksref{coemail}}
  \address{FaMAF\\Universidad Nacional de C\'ordoba\\
    C\'ordoba, Argentina}
  \author{Pedro S\'anchez Terraf\thanksref{co2email}}
  \address{CIEM-FaMAF\\Universidad Nacional de C\'ordoba\\
    C\'ordoba, Argentina}
 \thanks[myemail]{Email:
    \href{mailto:gunther@famaf.unc.edu.ar} {\texttt{\normalshape
        gunther@famaf.unc.edu.ar}}} \thanks[coemail]{Email:
    \href{mailto:pagano@famaf.unc.edu.ar} {\texttt{\normalshape
        pagano@famaf.unc.edu.ar}}}  \thanks[co2email]{Email:
    \href{mailto:sterraf@famaf.unc.edu.ar} {\texttt{\normalshape
        sterraf@famaf.unc.edu.ar}}} 
 \thanks[ALL]{Supported by Secyt-UNC project 33620180100465CB.} 
\begin{abstract} 
\end{abstract}
\begin{keyword}
Isabelle/ZF, forcing, preorder, Rasiowa-Sikorski lemma, names, generic extension.
\end{keyword}
\end{frontmatter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}\label{sec:introduction}

%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
\subsection{Models of ZFC}

%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
\subsection{Isabelle/ZF}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Forcing posets}\label{sec:forcing-posets}

%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
\subsection{The Rasiowa-Sikorski lemma}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The generic extension}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Names}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Future work}
\begin{itemize}
\item Develop an interface between Paulson's relativization results
  and countable models $M$ of $\ZFC$. This will show
  that every such $M$ is closed under well-founded recursion and, in
  particular, that contains names for each of its
  elements. Consequently, $M\sbq M[G]$.
\item Prove that the basic axioms of $\ZFC$ hold in the generic
  extension. 
\item After prototyping the forcing relation, prove the Axiom Scheme
  of Separation (the first that needs to use the machinery of forcing
  nontrivially).
\end{itemize}

\end{document}

### 2018-06-17 LSFA/first_steps_into_forcing.tex
avanzo lentamente

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Forcing posets}\label{sec:forcing-posets}
+\section{Forcing notions}\label{sec:forcing-posets}
+\textit{Para no confundir, directamente usamos este nombre en lugar de
+  ``forcing posets'' para los preórdenes con un máximo distinguido}
 
@@
 \subsection{The Rasiowa-Sikorski lemma}
+In order to formalize of the proof of the Rasiowa-Sikorski lemma in
+general, a version of the Axiom of Choice ($\AC$) must be proved. This
+version is known as the \emph{Principle of Dependent choices}:
+\begin{quote}
+  ($\DC$) Let $\rr$ be a binary relation on $A$, and $a\in A$. If
+  $\forall x\in A\,  \exists y\in A\, x\rr y$, then there exists
+  $f:\om\to A$ such that $f(0)=a$ and $f(n)\rr f(n+1)$ for all
+  $n\in\om$.
+\end{quote}
+
+A different version of $\DC$ (without the constraint $f(0)=a$, but
+allowing a parameter that allows to prove an equivalence with $\AC$) was
+formalized as part of the Isabelle/ZF library, along with many
+statements that follow from or are equivalent to $\AC$. Since the
+proofs there required a greater generality, it would have been very
+contrived to derived this ``pointed'' version of $\DC$ from that
+development. Instead, we preferred to give a direct proof based in the
+one that appears in Moschovakis \cite{moschovakis1994notes}. Another
+advantage is that this proof depends on the following version of $\AC$
+(existence of choice functions on $\P(X)\sm \{\mty\}$):
+\[
+\exists f:\P(X)\sm \{\mty\}\to X \, \forall A\in \P(X)\sm \{\mty\}\ f(A)
+    \in A.
+\]
+This statement is more standard than the one deemed ``$\AC$'' in the
+Isabelle/ZF development, that involves a formula parameter and thus
+involves a silent use of Replacement.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Names}
 
@@
 \section{The generic extension}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Names}
+Forcing is a technique to extend countable transitive models of
+$\ZFC$. This process is guaranteed to preserve the $\ZFC$
+axioms but allows to fine-tune what other first order properties the
+extension will have. The first, both historically and in importance,
+application of forcing was the proof that $\AC$ and $\CH$ were not
+derivable from $\ZFC$; this was done by Cohen in its seminal work
+\cite{Cohen-CH-PNAS}.
 
@@
 \end{itemize}
+\bibliographystyle{entcs}
+\bibliography{citados,journals&preprints,conferencias}
+\end{document}
 
-\end{document}
+%%% Local Variables:
+%%% mode: latex
+%%% ispell-local-dictionary: "american"
+%%% End:

### 2018-06-18 LSFA/first_steps_into_forcing.tex
estado actual, improlijo

@@
 \[
-\exists f:\P(X)\sm \{\mty\}\to X \, \forall A\in \P(X)\sm \{\mty\}\ f(A)
-    \in A.
+\exists f:\P(X)\sm \{\mty\}\to X \, \forall A\sbq X.\ A\neq \mty \limp
+f(A) \in A.
 \]
 This statement is more standard than the one deemed ``$\AC$'' in the
-Isabelle/ZF development, that involves a formula parameter and thus
-involves a silent use of Replacement.
+Isabelle/ZF development, that involves a silent use of Replacement
+since it is written as an axiom scheme.
 

### 2018-06-18 LSFA/conclusion.tex
División del archivo para dividir el trabajo

--- 2018-06-16 LSFA/first_steps_into_forcing.tex
@@
-\documentclass[9pt]{entcs} \usepackage{entcsmacro}
-\usepackage{graphicx}
-\sloppy
-
-\input{header-draft}
-\def\lastname{Gunther, Pagano, S\'anchez Terraf}
-\begin{document}
-\begin{frontmatter}
-  \title{First steps towards a formalization of Forcing}
-  \author{Emmanuel Gunther%\thanksref{ALL}
-    \thanksref{myemail}}
-  \address{FaMAF\\ Universidad Nacional de C\'ordoba\\
-    C\'ordoba, Argentina} \author{Miguel Pagano\thanksref{coemail}}
-  \address{FaMAF\\Universidad Nacional de C\'ordoba\\
-    C\'ordoba, Argentina}
-  \author{Pedro S\'anchez Terraf\thanksref{co2email}}
-  \address{CIEM-FaMAF\\Universidad Nacional de C\'ordoba\\
-    C\'ordoba, Argentina}
- \thanks[myemail]{Email:
-    \href{mailto:gunther@famaf.unc.edu.ar} {\texttt{\normalshape
-        gunther@famaf.unc.edu.ar}}} \thanks[coemail]{Email:
-    \href{mailto:pagano@famaf.unc.edu.ar} {\texttt{\normalshape
-        pagano@famaf.unc.edu.ar}}}  \thanks[co2email]{Email:
-    \href{mailto:sterraf@famaf.unc.edu.ar} {\texttt{\normalshape
-        sterraf@famaf.unc.edu.ar}}} 
- \thanks[ALL]{Supported by Secyt-UNC project 33620180100465CB.} 
-\begin{abstract} 
-\end{abstract}
-\begin{keyword}
-Isabelle/ZF, forcing, preorder, Rasiowa-Sikorski lemma, names, generic extension.
-\end{keyword}
-\end{frontmatter}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}\label{sec:introduction}
-
-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
-\subsection{Models of ZFC}
-
-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
-\subsection{Isabelle/ZF}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Forcing posets}\label{sec:forcing-posets}
-
-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
-\subsection{The Rasiowa-Sikorski lemma}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{The generic extension}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Names}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Future work}
@@
 
-\end{document}
+%%% Local Variables:
+%%% mode: latex
+%%% ispell-local-dictionary: "american"
+%%% TeX-master: "first_steps_into_forcing"
+%%% End:

### 2018-06-18 LSFA/first_steps_into_forcing.tex
División del archivo para dividir el trabajo

--- 2018-06-16 LSFA/first_steps_into_forcing.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}\label{sec:introduction}
+\input{intro}
 
-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
-\subsection{Models of ZFC}
+\input{posets}
 
-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
-\subsection{Isabelle/ZF}
+\input{rs}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Forcing posets}\label{sec:forcing-posets}
-
-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
-\subsection{The Rasiowa-Sikorski lemma}
+\input{names}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{The generic extension}
+\input{gen-ext}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Names}
+\input{conclusion}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Future work}
-\begin{itemize}
-\item Develop an interface between Paulson's relativization results
-  and countable models $M$ of $\ZFC$. This will show
-  that every such $M$ is closed under well-founded recursion and, in
-  particular, that contains names for each of its
-  elements. Consequently, $M\sbq M[G]$.
-\item Prove that the basic axioms of $\ZFC$ hold in the generic
-  extension. 
-\item After prototyping the forcing relation, prove the Axiom Scheme
-  of Separation (the first that needs to use the machinery of forcing
-  nontrivially).
-\end{itemize}
+\bibliographystyle{entcs}
+\bibliography{first_steps_into_forcing}
+\end{document}
 
-\end{document}
+%%% Local Variables:
+%%% mode: latex
+%%% ispell-local-dictionary: "american"
+%%% End:

### 2018-06-18 LSFA/intro.tex
División del archivo para dividir el trabajo

--- 2018-06-18 LSFA/conclusion.tex
@@
-\section{Future work}
-\begin{itemize}
-\item Develop an interface between Paulson's relativization results
-  and countable models $M$ of $\ZFC$. This will show
-  that every such $M$ is closed under well-founded recursion and, in
-  particular, that contains names for each of its
-  elements. Consequently, $M\sbq M[G]$.
-\item Prove that the basic axioms of $\ZFC$ hold in the generic
-  extension. 
-\item After prototyping the forcing relation, prove the Axiom Scheme
-  of Separation (the first that needs to use the machinery of forcing
-  nontrivially).
-\end{itemize}
+\section{Introduction}\label{sec:introduction}
+
+%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
+\subsection{Models of ZFC}
+
+%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
+\subsection{Isabelle/ZF}
 

### 2018-06-18 LSFA/rs.tex
División del archivo para dividir el trabajo

\subsection{The Rasiowa-Sikorski lemma}
In order to formalize of the proof of the Rasiowa-Sikorski lemma in
general, a version of the Axiom of Choice ($\AC$) must be proved. This
version is known as the \emph{Principle of Dependent choices}:
\begin{quote}
  ($\DC$) Let $\rr$ be a binary relation on $A$, and $a\in A$. If
  $\forall x\in A\,  \exists y\in A\, x\rr y$, then there exists
  $f:\om\to A$ such that $f(0)=a$ and $f(n)\rr f(n+1)$ for all
  $n\in\om$.
\end{quote}

A different version of $\DC$ (without the constraint $f(0)=a$, but
allowing a parameter that allows to prove an equivalence with $\AC$) was
formalized as part of the Isabelle/ZF library, along with many
statements that follow from or are equivalent to $\AC$. Since the
proofs there required a greater generality, it would have been very
contrived to derived this ``pointed'' version of $\DC$ from that
development. Instead, we preferred to give a direct proof based in the
one that appears in Moschovakis \cite{moschovakis1994notes}. Another
advantage is that this proof depends on the following version of $\AC$
(existence of choice functions on $\P(X)\sm \{\mty\}$):
\[
\exists f:\P(X)\sm \{\mty\}\to X \, \forall A\sbq X.\ A\neq \mty \limp
f(A) \in A.
\]
This statement is more standard than the one deemed ``$\AC$'' in the
Isabelle/ZF development, that involves a silent use of Replacement
since it is written as an axiom scheme.

%%% Local Variables:
%%% mode: latex
%%% ispell-local-dictionary: "american"
%%% TeX-master: "first_steps_into_forcing"
%%% End:

### 2018-06-18 LSFA/rs.tex
sequence DC.

@@
 \[
-\exists f:\P(X)\sm \{\mty\}\to X \, \forall A\sbq X.\ A\neq \mty \limp
-f(A) \in A.
+\exists s: \P(A)\sm \{\mty\}\to A \, \forall X\sbq A.\ X\neq \mty \limp
+s(X) \in X.
 \]
@@
 since it is written as an axiom scheme.
+
+The strategy is to define $f$ above by primitive recursion.
+
+\begin{isabellebody}\isamarkuptrue%
+\isanewline
+\isacommand{consts}\isamarkupfalse%
+\ dc{\isacharunderscore}witness\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i\ {\isasymRightarrow}\ i\ {\isasymRightarrow}\ i\ {\isasymRightarrow}\ i\ {\isasymRightarrow}\ i\ {\isasymRightarrow}\ i{\isachardoublequoteclose}\isanewline
+\isacommand{primrec}\isamarkupfalse%
+\isanewline
+\ \ wit{\isadigit{0}}\ \ \ {\isacharcolon}\ {\isachardoublequoteopen}dc{\isacharunderscore}witness{\isacharparenleft}{\isadigit{0}}{\isacharcomma}A{\isacharcomma}a{\isacharcomma}s{\isacharcomma}R{\isacharparenright}\ {\isacharequal}\ a{\isachardoublequoteclose}\isanewline
+\ \ witrec\ {\isacharcolon}{\isachardoublequoteopen}dc{\isacharunderscore}witness{\isacharparenleft}succ{\isacharparenleft}n{\isacharparenright}{\isacharcomma}A{\isacharcomma}a{\isacharcomma}s{\isacharcomma}R{\isacharparenright}\ {\isacharequal}\ s{\isacharbackquote}{\isacharbraceleft}x{\isasymin}A{\isachardot}\ {\isasymlangle}dc{\isacharunderscore}witness{\isacharparenleft}n{\isacharcomma}A{\isacharcomma}a{\isacharcomma}s{\isacharcomma}R{\isacharparenright}{\isacharcomma}x{\isasymrangle}{\isasymin}R\ {\isacharbraceright}{\isachardoublequoteclose}\isanewline
+\end{isabellebody}
+
+The function \verb|dc_witness| has, appart from $A$, $a$ and $R$, a
+function $s$ as a parameter. If this function is a selector for
+$\P(A)\sm \{\mty\}$, the function $f\defi$\verb|dc_witness|$(A,a,s,R)$
+will satify $\DC$.
+
+The proof is mostly routine; after a few lemmas we obtain the
+following theorem:
+
+\begin{isabellebody}
+\isanewline
+\isacommand{theorem}\isamarkupfalse%
+\ pointed{\isacharunderscore}DC\ \ {\isacharcolon}\ {\isachardoublequoteopen}{\isacharparenleft}{\isasymforall}x{\isasymin}A{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}{\isasymin}\ R{\isacharparenright}\ {\isasymLongrightarrow}\isanewline
+\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ {\isasymforall}a{\isasymin}A{\isachardot}\ {\isacharparenleft}{\isasymexists}f\ {\isasymin}\ nat{\isasymrightarrow}A{\isachardot}\ f{\isacharbackquote}{\isadigit{0}}\ {\isacharequal}\ a\ {\isasymand}\ {\isacharparenleft}{\isasymforall}n\ {\isasymin}\ nat{\isachardot}\ {\isasymlangle}f{\isacharbackquote}n{\isacharcomma}f{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isasymrangle}{\isasymin}R{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\end{isabellebody}
+
+We need a further, ``diagonal'' version of $\DC$  to prove
+Rasiowa-Sikorski. That is, if the assumption holds for a sequence of
+relations $S_n$,  then $f(n) \mathrel{S_{n+1}} f(n+1)$.
+
+We first obtain a corollary of $\DC$ changing $A$ for
+$A\times\mathtt{nat}$:
+
+\begin{isabellebody}
+\isanewline
+\isacommand{lemma}\isamarkupfalse%
+\ aux{\isacharunderscore}DC{\isacharunderscore}on{\isacharunderscore}AxNat{\isadigit{2}}\ {\isacharcolon}\ {\isachardoublequoteopen}{\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}{\isasymlangle}y{\isacharcomma}succ{\isacharparenleft}snd{\isacharparenleft}x{\isacharparenright}{\isacharparenright}{\isasymrangle}{\isasymrangle}\ {\isasymin}\ R\ {\isasymLongrightarrow}\isanewline
+\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ {\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ {\isacharbraceleft}{\isasymlangle}a{\isacharcomma}b{\isasymrangle}{\isasymin}R{\isachardot}\ snd{\isacharparenleft}b{\isacharparenright}\ {\isacharequal}\ succ{\isacharparenleft}snd{\isacharparenleft}a{\isacharparenright}{\isacharparenright}{\isacharbraceright}{\isachardoublequoteclose}\isanewline
+\end{isabellebody}
+%
+The following lemma is then proved automatically:
+
+\begin{isabellebody}
+\isanewline
+\isacommand{lemma}\isamarkupfalse%
+\ aux{\isacharunderscore}sequence{\isacharunderscore}DC{\isadigit{2}}\ {\isacharcolon}\ {\isachardoublequoteopen}{\isasymforall}x{\isasymin}A{\isachardot}\ {\isasymforall}n{\isasymin}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ S{\isacharbackquote}n\ {\isasymLongrightarrow}\isanewline
+\ \ \ \ \ \ \ \ {\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}{\isasymlangle}y{\isacharcomma}succ{\isacharparenleft}snd{\isacharparenleft}x{\isacharparenright}{\isacharparenright}{\isasymrangle}{\isasymrangle}\ {\isasymin}\ {\isacharbraceleft}{\isasymlangle}{\isasymlangle}x{\isacharcomma}n{\isasymrangle}{\isacharcomma}{\isasymlangle}y{\isacharcomma}m{\isasymrangle}{\isasymrangle}{\isasymin}{\isacharparenleft}A{\isasymtimes}nat{\isacharparenright}{\isasymtimes}{\isacharparenleft}A{\isasymtimes}nat{\isacharparenright}{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}{\isasymin}S{\isacharbackquote}m\ {\isacharbraceright}{\isachardoublequoteclose}\isanewline
+\ \ %
+\isacommand{by}\isamarkupfalse%
+\ auto%
+\isanewline
+\end{isabellebody}
+%
+And finally, we arrive to $\DC$ for a sequence of relations.
+
+\begin{isabellebody}
+\isanewline
+\isacommand{lemma}\isamarkupfalse%
+\ sequence{\isacharunderscore}DC{\isacharcolon}\ {\isachardoublequoteopen}{\isasymforall}x{\isasymin}A{\isachardot}\ {\isasymforall}n{\isasymin}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ S{\isacharbackquote}n\ {\isasymLongrightarrow}\isanewline
+\ \ \ \ {\isasymforall}a{\isasymin}A{\isachardot}\ {\isacharparenleft}{\isasymexists}f\ {\isasymin}\ nat{\isasymrightarrow}A{\isachardot}\ f{\isacharbackquote}{\isadigit{0}}\ {\isacharequal}\ a\ {\isasymand}\ {\isacharparenleft}{\isasymforall}n\ {\isasymin}\ nat{\isachardot}\ {\isasymlangle}f{\isacharbackquote}n{\isacharcomma}f{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isasymrangle}{\isasymin}S{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\end{isabellebody}
 

### 2018-06-18 LSFA/rs.tex
esquema de RS terminado

@@
-\subsection{The Rasiowa-Sikorski lemma}
-In order to formalize of the proof of the Rasiowa-Sikorski lemma in
+\subsection{A sequence version of $\DC$}
+In order to prove the Rasiowa-Sikorski lemma in
 general, a version of the Axiom of Choice ($\AC$) must be proved. This
@@
 
+\subsection{The Rasiowa-Sikorski lemma}
+In order to state this Lemma, we gather the relevant hypotheses into a locale:
+
+\begin{isabellebody}%
+\isanewline
+\isacommand{locale}\isamarkupfalse%
+\ countable{\isacharunderscore}generic\ {\isacharequal}\ forcing{\isacharunderscore}poset\ {\isacharplus}\isanewline
+\ \ \isakeyword{fixes}\ {\isasymD}\isanewline
+\ \ \isakeyword{assumes}\ countable{\isacharunderscore}subs{\isacharunderscore}of{\isacharunderscore}P{\isacharcolon}\ \ {\isachardoublequoteopen}{\isasymD}\ {\isasymin}\ nat{\isasymrightarrow}Pow{\isacharparenleft}P{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{and}\ \ \ \ \ seq{\isacharunderscore}of{\isacharunderscore}denses{\isacharcolon}\ \ \ \ \ \ \ \ {\isachardoublequoteopen}{\isasymforall}n\ {\isasymin}\ nat{\isachardot}\ dense{\isacharparenleft}{\isasymD}{\isacharbackquote}n{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\end{isabellebody}
+%
+That is, $\calD$ is a sequence of dense subsets of the poset $P$. A
+filter is \emph{$\calD$-generic} if it intersects every dense set in
+the sequence.
+
+\begin{isabellebody}%
+\isanewline
+\isacommand{definition}\isamarkupfalse%
+\ {\isacharparenleft}\isakeyword{in}\ countable{\isacharunderscore}generic{\isacharparenright}\isanewline
+\ \ D{\isacharunderscore}generic\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}o{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
+\ \ {\isachardoublequoteopen}D{\isacharunderscore}generic{\isacharparenleft}G{\isacharparenright}\ {\isacharequal}{\isacharequal}\ filter{\isacharparenleft}G{\isacharparenright}\ {\isasymand}\ {\isacharparenleft}{\isasymforall}n{\isasymin}nat{\isachardot}{\isacharparenleft}{\isasymD}{\isacharbackquote}n{\isacharparenright}{\isasyminter}G{\isasymnoteq}{\isadigit{0}}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\end{isabellebody}
+
+We can now state the Rasiowa-Sikorski Lemma.
+\begin{isabellebody}%
+\isanewline
+\isacommand{theorem}\isamarkupfalse%
+\ {\isacharparenleft}\isakeyword{in}\ countable{\isacharunderscore}generic{\isacharparenright}\ rasiowa{\isacharunderscore}sikorski{\isacharcolon}\isanewline
+\ \ {\isachardoublequoteopen}p{\isasymin}P\ {\isasymLongrightarrow}\ {\isasymexists}G{\isachardot}\ p{\isasymin}G\ {\isasymand}\ D{\isacharunderscore}generic{\isacharparenleft}G{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\end{isabellebody}
+
+The intuitive argument for the result is simple: Once $p_0=p\in P$ is
+fixed, we can recursively choose $p_{n+1}$ such that 
+$p_n \geq p_{n+1}\in \calD_n$, since $\calD_n$ is dense in $P$. Then
+the filter generated by $\{p_n : n\in \om\}$ intersects each set in
+the sequence $\{\calD_n\}_n$. This argument appeals to the sequence
+version of $\DC$; we have to prove first that the relevant relation
+satisfies its hypothesis:
+
+\begin{isabellebody}%
+\isanewline
+\isacommand{lemma}\isamarkupfalse%
+\ {\isacharparenleft}\isakeyword{in}\ countable{\isacharunderscore}generic{\isacharparenright}\ RS{\isacharunderscore}relation{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\isanewline
+\ \ \ \ \ \ \ \ {\isadigit{1}}{\isacharcolon}\ \ {\isachardoublequoteopen}x{\isasymin}P{\isachardoublequoteclose}\isanewline
+\ \ \ \ \ \ \ \ \ \ \ \ \isakeyword{and}\isanewline
+\ \ \ \ \ \ \ \ {\isadigit{2}}{\isacharcolon}\ \ {\isachardoublequoteopen}n{\isasymin}nat{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\isanewline
+\ \ \ \ \ \ \ \ \ \ \ \ {\isachardoublequoteopen}{\isasymexists}y{\isasymin}P{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ {\isacharparenleft}{\isasymlambda}m{\isasymin}nat{\isachardot}\ {\isacharbraceleft}{\isasymlangle}x{\isacharcomma}y{\isasymrangle}{\isasymin}P{\isacharasterisk}P{\isachardot}\ {\isasymlangle}y{\isacharcomma}x{\isasymrangle}{\isasymin}leq\ {\isasymand}\ y{\isasymin}{\isasymD}{\isacharbackquote}{\isacharparenleft}pred{\isacharparenleft}m{\isacharparenright}{\isacharparenright}{\isacharbraceright}{\isacharparenright}{\isacharbackquote}n{\isachardoublequoteclose}\isanewline
+\end{isabellebody}
+%
+Both of the proofs written in the Isar language.
+
 %%% Local Variables:

### 2018-06-18 LSFA/rs.tex

@@
 advantage is that this proof depends on the following version of $\AC$
-(existence of choice functions on $\P(X)\sm \{\mty\}$):
+(existence of choice functions on $\P(A)\sm \{\mty\}$):
 \[
@@
 
-The function \verb|dc_witness| has, appart from $A$, $a$ and $R$, a
+The function \verb|dc_witness| has, apart from $A$, $a$ and $R$, a
 function $s$ as a parameter. If this function is a selector for

### 2018-06-21 LSFA/intro.tex
outline

@@
 \section{Introduction}\label{sec:introduction}
+Set Theory play a double role in Mathematics: It is one of its
+possible foundations, and also an active research area.
+
+As it is widely known, Georg Cantor introduced its main concepts and in
+particular showed the fundamental result that the real line, $\R$  is not
+coordinable with the natural numbers. Soon after this, he posed the
+most important question in the field, written as a conjecture:
+\begin{quote}
+  The \emph{Continuum Hypothesis} ($\CH$). Every uncountable subset of $\R$ is
+  coordinable with $\R$.
+\end{quote}
+\bigskip
+
+\hrule
+\bigskip
+
+The most widely accepted foundation of mathematics is through Set
+Theory. More precisely, using the first-order theory axiomatized by
+Zermelo, Fraenkel, and including the Axiom of Choice ($\AC$) among its
+axioms. This theory is known by the $\ZFC$ acronym.
+\bigskip
+
+\hrule
+\bigskip
+
+The first, both historically and in importance,
+    application of forcing was the proof that $\AC$ and $\CH$ were not
+    derivable from $\ZFC$; this was done by Cohen in its seminal work \cite{Cohen-CH-PNAS}.
 
@@
 \subsection{Models of ZFC}
+\begin{enumerate}
+\item $\CH$ is (Platonistically) undecidable.
+  \begin{enumerate}
+  \item By G\"odel's Second Incompleteness Theorems, we can't show
+    that a model of $\ZFC$ exists.
+  \item G\"odel showed that $\CH$ is relatively consistent with
+    $\ZFC$. ``Easy.''
+  \item Cohen showed that  $\neg\CH$ is relatively consistent with
+    $\ZFC$. Not ``easy.'' Formalization Open.
+  \end{enumerate}
+\item Using $\ZF$, G\"odel defined the constructible universe,
+  $L$. G\"odel showed that $\lb L,\in\rb\models \ZFC+\CH$. Paulson formalized $L$
+  for the consistency of $\AC$.
+\item The main concept for G\"odel's proofs is
+  \emph{relativization}. This is the core of Paulson's development.
+  \begin{enumerate}
+  \item \dots
+  \end{enumerate}
+  %
+\item 
+\item Cohen invented Forcing.
+  \begin{enumerate}
+  \item Countable transitive models $M$. These are enough for consistency
+    proofs by the L\"owenheim-Skolem Theorem, the Reflection
+    Principle, and the Mostowksi Collapse.
+    %
+  \item 
+    %
+  \item Forcing is a technique to extend countable transitive models of
+    $\ZFC$. This process is guaranteed to preserve the $\ZFC$
+    axioms but allows to fine-tune what other first order properties the
+    extension will have. 
+    %
+  \item Given a preorder with top $\lb\PP,\leq,\1\rb$ in $M$, define
+    \emph{filter} $G$, \emph{dense}. In general, there is no  $G\in M$
+    that intersects every dense
+    subset of $\PP$ that lies in $M$ (``generic'').
+  \item Rasiowa-Sikorski states that there is such $G$ for every
+    family of countable dense subsets. Thus, there are generic filters
+    for countable transitive models.
+  \end{enumerate}
+
+\end{enumerate}
 

### 2018-06-21 LSFA/intro.tex
primera parte de la intro

@@
 possible foundations, and also an active research area.
-
+%
 As it is widely known, Georg Cantor introduced its main concepts and in
@@
 \end{quote}
-\bigskip
 
-\hrule
-\bigskip
+The current  axiomatic foundation of Set Theory is through 
+first-order logic and uses the axioms by  Zermelo, Fraenkel, and
+including the Axiom of Choice ($\AC$) among its axioms. This theory is
+known by the $\ZFC$ acronym. G\"odel showed that $\CH$ can't be
+refuted using the $\ZFC$, unless the latter are inconsistent (we say
+that \emph{$\CH$ is relatively 
+  consistent with  $\ZFC$}). For a while, this result left the
+possibility that one might be able show that $\ZFC\models \CH$, but in
+a groundbreaking work \cite{Cohen-CH-PNAS}, Paul Cohen discovered the
+technique of 
+\emph{forcing} and showed that $\neg\CH$ is relatively consistent with
+$\ZFC$. Forcing has since then been used for showing innumerable
+independence results and perform mathematical constructions.
 
-The most widely accepted foundation of mathematics is through Set
-Theory. More precisely, using the first-order theory axiomatized by
-Zermelo, Fraenkel, and including the Axiom of Choice ($\AC$) among its
-axioms. This theory is known by the $\ZFC$ acronym.
-\bigskip
-
-\hrule
-\bigskip
-
-The first, both historically and in importance,
-    application of forcing was the proof that $\AC$ and $\CH$ were not
-    derivable from $\ZFC$; this was done by Cohen in its seminal work \cite{Cohen-CH-PNAS}.
-
-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
-\subsection{Models of ZFC}
 \begin{enumerate}
@@
 \end{enumerate}
-
-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
-\subsection{Isabelle/ZF}
-
-%%% Local Variables:
-%%% mode: latex
-%%% ispell-local-dictionary: "american"
-%%% TeX-master: "first_steps_into_forcing"
-%%% End:

### 2018-06-22 LSFA/rs.tex
body, isanewline -> ""

@@
 
-\begin{isabellebody}\isamarkuptrue%
-\isanewline
+\begin{isabelle}\isamarkuptrue%
 \isacommand{consts}\isamarkupfalse%
@@
 \ \ wit{\isadigit{0}}\ \ \ {\isacharcolon}\ {\isachardoublequoteopen}dc{\isacharunderscore}witness{\isacharparenleft}{\isadigit{0}}{\isacharcomma}A{\isacharcomma}a{\isacharcomma}s{\isacharcomma}R{\isacharparenright}\ {\isacharequal}\ a{\isachardoublequoteclose}\isanewline
-\ \ witrec\ {\isacharcolon}{\isachardoublequoteopen}dc{\isacharunderscore}witness{\isacharparenleft}succ{\isacharparenleft}n{\isacharparenright}{\isacharcomma}A{\isacharcomma}a{\isacharcomma}s{\isacharcomma}R{\isacharparenright}\ {\isacharequal}\ s{\isacharbackquote}{\isacharbraceleft}x{\isasymin}A{\isachardot}\ {\isasymlangle}dc{\isacharunderscore}witness{\isacharparenleft}n{\isacharcomma}A{\isacharcomma}a{\isacharcomma}s{\isacharcomma}R{\isacharparenright}{\isacharcomma}x{\isasymrangle}{\isasymin}R\ {\isacharbraceright}{\isachardoublequoteclose}\isanewline
-\end{isabellebody}
+\ \ witrec\ {\isacharcolon}{\isachardoublequoteopen}dc{\isacharunderscore}witness{\isacharparenleft}succ{\isacharparenleft}n{\isacharparenright}{\isacharcomma}A{\isacharcomma}a{\isacharcomma}s{\isacharcomma}R{\isacharparenright}\ {\isacharequal}\ s{\isacharbackquote}{\isacharbraceleft}x{\isasymin}A{\isachardot}\ {\isasymlangle}dc{\isacharunderscore}witness{\isacharparenleft}n{\isacharcomma}A{\isacharcomma}a{\isacharcomma}s{\isacharcomma}R{\isacharparenright}{\isacharcomma}x{\isasymrangle}{\isasymin}R\ {\isacharbraceright}{\isachardoublequoteclose}
+\end{isabelle}
 
@@
 
-\begin{isabellebody}
-\isanewline
+\begin{isabelle}
 \isacommand{theorem}\isamarkupfalse%
 \ pointed{\isacharunderscore}DC\ \ {\isacharcolon}\ {\isachardoublequoteopen}{\isacharparenleft}{\isasymforall}x{\isasymin}A{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}{\isasymin}\ R{\isacharparenright}\ {\isasymLongrightarrow}\isanewline
-\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ {\isasymforall}a{\isasymin}A{\isachardot}\ {\isacharparenleft}{\isasymexists}f\ {\isasymin}\ nat{\isasymrightarrow}A{\isachardot}\ f{\isacharbackquote}{\isadigit{0}}\ {\isacharequal}\ a\ {\isasymand}\ {\isacharparenleft}{\isasymforall}n\ {\isasymin}\ nat{\isachardot}\ {\isasymlangle}f{\isacharbackquote}n{\isacharcomma}f{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isasymrangle}{\isasymin}R{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
-\end{isabellebody}
+\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ {\isasymforall}a{\isasymin}A{\isachardot}\ {\isacharparenleft}{\isasymexists}f\ {\isasymin}\ nat{\isasymrightarrow}A{\isachardot}\ f{\isacharbackquote}{\isadigit{0}}\ {\isacharequal}\ a\ {\isasymand}\ {\isacharparenleft}{\isasymforall}n\ {\isasymin}\ nat{\isachardot}\ {\isasymlangle}f{\isacharbackquote}n{\isacharcomma}f{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isasymrangle}{\isasymin}R{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
 
@@
 
-\begin{isabellebody}
-\isanewline
+\begin{isabelle}
 \isacommand{lemma}\isamarkupfalse%
 \ aux{\isacharunderscore}DC{\isacharunderscore}on{\isacharunderscore}AxNat{\isadigit{2}}\ {\isacharcolon}\ {\isachardoublequoteopen}{\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}{\isasymlangle}y{\isacharcomma}succ{\isacharparenleft}snd{\isacharparenleft}x{\isacharparenright}{\isacharparenright}{\isasymrangle}{\isasymrangle}\ {\isasymin}\ R\ {\isasymLongrightarrow}\isanewline
-\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ {\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ {\isacharbraceleft}{\isasymlangle}a{\isacharcomma}b{\isasymrangle}{\isasymin}R{\isachardot}\ snd{\isacharparenleft}b{\isacharparenright}\ {\isacharequal}\ succ{\isacharparenleft}snd{\isacharparenleft}a{\isacharparenright}{\isacharparenright}{\isacharbraceright}{\isachardoublequoteclose}\isanewline
-\end{isabellebody}
+\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ {\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ {\isacharbraceleft}{\isasymlangle}a{\isacharcomma}b{\isasymrangle}{\isasymin}R{\isachardot}\ snd{\isacharparenleft}b{\isacharparenright}\ {\isacharequal}\ succ{\isacharparenleft}snd{\isacharparenleft}a{\isacharparenright}{\isacharparenright}{\isacharbraceright}{\isachardoublequoteclose}
+\end{isabelle}
 %
@@
 
-\begin{isabellebody}
-\isanewline
+\begin{isabelle}
 \isacommand{lemma}\isamarkupfalse%
@@
 \ auto%
-\isanewline
-\end{isabellebody}
+\end{isabelle}
 %
@@
 
-\begin{isabellebody}
-\isanewline
+\begin{isabelle}
 \isacommand{lemma}\isamarkupfalse%
 \ sequence{\isacharunderscore}DC{\isacharcolon}\ {\isachardoublequoteopen}{\isasymforall}x{\isasymin}A{\isachardot}\ {\isasymforall}n{\isasymin}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ S{\isacharbackquote}n\ {\isasymLongrightarrow}\isanewline
-\ \ \ \ {\isasymforall}a{\isasymin}A{\isachardot}\ {\isacharparenleft}{\isasymexists}f\ {\isasymin}\ nat{\isasymrightarrow}A{\isachardot}\ f{\isacharbackquote}{\isadigit{0}}\ {\isacharequal}\ a\ {\isasymand}\ {\isacharparenleft}{\isasymforall}n\ {\isasymin}\ nat{\isachardot}\ {\isasymlangle}f{\isacharbackquote}n{\isacharcomma}f{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isasymrangle}{\isasymin}S{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
-\end{isabellebody}
+\ \ \ \ {\isasymforall}a{\isasymin}A{\isachardot}\ {\isacharparenleft}{\isasymexists}f\ {\isasymin}\ nat{\isasymrightarrow}A{\isachardot}\ f{\isacharbackquote}{\isadigit{0}}\ {\isacharequal}\ a\ {\isasymand}\ {\isacharparenleft}{\isasymforall}n\ {\isasymin}\ nat{\isachardot}\ {\isasymlangle}f{\isacharbackquote}n{\isacharcomma}f{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isasymrangle}{\isasymin}S{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
 
@@
 
-\begin{isabellebody}%
-\isanewline
+\begin{isabelle}%
 \isacommand{locale}\isamarkupfalse%
@@
 \ \ \isakeyword{assumes}\ countable{\isacharunderscore}subs{\isacharunderscore}of{\isacharunderscore}P{\isacharcolon}\ \ {\isachardoublequoteopen}{\isasymD}\ {\isasymin}\ nat{\isasymrightarrow}Pow{\isacharparenleft}P{\isacharparenright}{\isachardoublequoteclose}\isanewline
-\ \ \isakeyword{and}\ \ \ \ \ seq{\isacharunderscore}of{\isacharunderscore}denses{\isacharcolon}\ \ \ \ \ \ \ \ {\isachardoublequoteopen}{\isasymforall}n\ {\isasymin}\ nat{\isachardot}\ dense{\isacharparenleft}{\isasymD}{\isacharbackquote}n{\isacharparenright}{\isachardoublequoteclose}\isanewline
-\end{isabellebody}
+\ \ \isakeyword{and}\ \ \ \ \ seq{\isacharunderscore}of{\isacharunderscore}denses{\isacharcolon}\ \ \ \ \ \ \ \ {\isachardoublequoteopen}{\isasymforall}n\ {\isasymin}\ nat{\isachardot}\ dense{\isacharparenleft}{\isasymD}{\isacharbackquote}n{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
 %
@@
 
-\begin{isabellebody}%
-\isanewline
+\begin{isabelle}%
 \isacommand{definition}\isamarkupfalse%
@@
 \ \ D{\isacharunderscore}generic\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}o{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
-\ \ {\isachardoublequoteopen}D{\isacharunderscore}generic{\isacharparenleft}G{\isacharparenright}\ {\isacharequal}{\isacharequal}\ filter{\isacharparenleft}G{\isacharparenright}\ {\isasymand}\ {\isacharparenleft}{\isasymforall}n{\isasymin}nat{\isachardot}{\isacharparenleft}{\isasymD}{\isacharbackquote}n{\isacharparenright}{\isasyminter}G{\isasymnoteq}{\isadigit{0}}{\isacharparenright}{\isachardoublequoteclose}\isanewline
-\end{isabellebody}
+\ \ {\isachardoublequoteopen}D{\isacharunderscore}generic{\isacharparenleft}G{\isacharparenright}\ {\isacharequal}{\isacharequal}\ filter{\isacharparenleft}G{\isacharparenright}\ {\isasymand}\ {\isacharparenleft}{\isasymforall}n{\isasymin}nat{\isachardot}{\isacharparenleft}{\isasymD}{\isacharbackquote}n{\isacharparenright}{\isasyminter}G{\isasymnoteq}{\isadigit{0}}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
 
 We can now state the Rasiowa-Sikorski Lemma.
-\begin{isabellebody}%
-\isanewline
+\begin{isabelle}%
 \isacommand{theorem}\isamarkupfalse%
 \ {\isacharparenleft}\isakeyword{in}\ countable{\isacharunderscore}generic{\isacharparenright}\ rasiowa{\isacharunderscore}sikorski{\isacharcolon}\isanewline
-\ \ {\isachardoublequoteopen}p{\isasymin}P\ {\isasymLongrightarrow}\ {\isasymexists}G{\isachardot}\ p{\isasymin}G\ {\isasymand}\ D{\isacharunderscore}generic{\isacharparenleft}G{\isacharparenright}{\isachardoublequoteclose}\isanewline
-\end{isabellebody}
+\ \ {\isachardoublequoteopen}p{\isasymin}P\ {\isasymLongrightarrow}\ {\isasymexists}G{\isachardot}\ p{\isasymin}G\ {\isasymand}\ D{\isacharunderscore}generic{\isacharparenleft}G{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
 
@@
 
-\begin{isabellebody}%
-\isanewline
+\begin{isabelle}%
 \isacommand{lemma}\isamarkupfalse%
@@
 \ \ \isakeyword{shows}\isanewline
-\ \ \ \ \ \ \ \ \ \ \ \ {\isachardoublequoteopen}{\isasymexists}y{\isasymin}P{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ {\isacharparenleft}{\isasymlambda}m{\isasymin}nat{\isachardot}\ {\isacharbraceleft}{\isasymlangle}x{\isacharcomma}y{\isasymrangle}{\isasymin}P{\isacharasterisk}P{\isachardot}\ {\isasymlangle}y{\isacharcomma}x{\isasymrangle}{\isasymin}leq\ {\isasymand}\ y{\isasymin}{\isasymD}{\isacharbackquote}{\isacharparenleft}pred{\isacharparenleft}m{\isacharparenright}{\isacharparenright}{\isacharbraceright}{\isacharparenright}{\isacharbackquote}n{\isachardoublequoteclose}\isanewline
-\end{isabellebody}
+\ \ \ \ \ \ \ \ \ \ \ \ {\isachardoublequoteopen}{\isasymexists}y{\isasymin}P{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ {\isacharparenleft}{\isasymlambda}m{\isasymin}nat{\isachardot}\ {\isacharbraceleft}{\isasymlangle}x{\isacharcomma}y{\isasymrangle}{\isasymin}P{\isacharasterisk}P{\isachardot}\ {\isasymlangle}y{\isacharcomma}x{\isasymrangle}{\isasymin}leq\ {\isasymand}\ y{\isasymin}{\isasymD}{\isacharbackquote}{\isacharparenleft}pred{\isacharparenleft}m{\isacharparenright}{\isacharparenright}{\isacharbraceright}{\isacharparenright}{\isacharbackquote}n{\isachardoublequoteclose}
+\end{isabelle}
 %

### 2018-06-22 LSFA/isabelle-zf.tex
sección separada
- Primeros comentarios sobre Isabelle/ZF. - Cambiado isabellebody por isabelle en posets.

--- 2018-06-18 LSFA/intro.tex
@@
-\section{Introduction}\label{sec:introduction}
-
-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
-\subsection{Models of ZFC}
-
 %-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
 \subsection{Isabelle/ZF}
+
+Let us introduce briefly Paulson's formalization of ZF
+\cite{paulson2017isabelle} in Isabelle and aspects of his formal proof
+for the relative consistency of the Axiom of Choice
+\cite{paulson_2003}; we will only focus on those aspects that are
+essential to keep this paper self-contained, and refer the interested
+reader to Paulson's articles. Isabelle/ZF is a theory built upon the
+core library FOL of classical first order logic.
+
+In Isabelle/ZF sets are \emph{individuals}, that is terms of type
+\isatt{i} and formulas have type \isatt{o}. As in usual presentations
+of ZF, Isabelle/ZF has only one relation symbol
+\isatt{\ensuremath{\epsilon}} (equality and inclusion are defined
+using membership); it also features one constant \isatt{\isadigit{0}}
+for the empty set and \isatt{inf} for an infinite set and some other
+set constructors, again terms of type \isatt{i}, to ease the
+developments. The axioms, being formulas, are terms of type \isatt{o};
+the foundation axiom, for example, is formalized as:
+\isa{{\isachardoublequoteopen}A\ {\isacharequal}\ {\isadigit{0}}\
+  {\isasymor}%
+  \ {\isacharparenleft}{\isasymexists}x{\isasymin}A{\isachardot}\ %
+  {\isasymforall}y{\isasymin}x{\isachardot}\ %
+  y{\isasymnotin}A{\isacharparenright}%
+  {\isachardoublequoteclose}}. Besides the axioms, Isabelle/ZF also
+introduces several definitions (for example, pairs and sets defined by
+comprehension using separation) and syntactic abbreviations to keep
+the formalization close to the customary manner of doing mathematics.
+Working with the library and extendig it is quite
+straightforward. Since we find a need to combine the replacement and
+separation schemes to speak of the image of a function over some set
+defined by comprehension, we introduced a new term-former:
+\begin{isabelle}
+  \isacommand{definition}\isamarkupfalse%
+  \ SepReplace\ {\isacharcolon}{\isacharcolon}\
+  {\isachardoublequoteopen}{\isacharbrackleft}i{\isacharcomma}\
+  i{\isasymRightarrow}i{\isacharcomma}\ i{\isasymRightarrow}\
+  o{\isacharbrackright}\
+  {\isasymRightarrow}i{\isachardoublequoteclose}\
+  \isakeyword{where}\isanewline \ \
+  {\isachardoublequoteopen}SepReplace{\isacharparenleft}A{\isacharcomma}b{\isacharcomma}Q{\isacharparenright}\
+  {\isacharequal}{\isacharequal}\ {\isacharbraceleft}y\ {\isachardot}\
+  x{\isasymin}A{\isacharcomma}\
+  y{\isacharequal}b{\isacharparenleft}x{\isacharparenright}\
+  {\isasymand}\
+  Q{\isacharparenleft}x{\isacharparenright}{\isacharbraceright}{\isachardoublequoteclose}
+\end{isabelle}
+\noindent It is more convenient to add an abbreviation for a simpler
+writing and a nicer reading, so we use \isa{{\isacharbraceleft}b\
+  {\isachardot}{\isachardot}\ x{\isasymin}A{\isacharcomma}\
+  Q{\isacharbraceright}} for
+\isa{SepReplace{\isacharparenleft}A{\isacharcomma}b{\isacharcomma}Q{\isacharparenright}}. The
+characterization of our new constructor is given by
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ Sep{\isacharunderscore}and{\isacharunderscore}Replace{\isacharcolon}\ {\isachardoublequoteopen}{\isacharbraceleft}b{\isacharparenleft}x{\isacharparenright}\ {\isachardot}{\isachardot}\ x{\isasymin}A{\isacharcomma}\ P{\isacharparenleft}x{\isacharparenright}\ {\isacharbraceright}\ {\isacharequal}\ {\isacharbraceleft}b{\isacharparenleft}x{\isacharparenright}\ {\isachardot}\ x{\isasymin}{\isacharbraceleft}y{\isasymin}A{\isachardot}\ P{\isacharparenleft}y{\isacharparenright}{\isacharbraceright}{\isacharbraceright}{\isachardoublequoteclose}
+\end{isabelle}
+
+% ∀x[c]
+% locales
 

### 2018-06-22 LSFA/intro.tex
esbozo de intro

@@
 first-order logic and uses the axioms by  Zermelo, Fraenkel, and
-including the Axiom of Choice ($\AC$) among its axioms. This theory is
+including the Axiom of Choice ($\AC$) among them. This theory is
 known by the $\ZFC$ acronym. G\"odel showed that $\CH$ can't be
@@
 independence results and perform mathematical constructions.
+
+\subsection{Models of $\ZFC$}
+By G\"odel's Second Incompleteness Theorem, we can't show prove that
+there exists a model of $\ZFC$. More formally, if we assume that
+mathematical proofs can be encoded as theorems of $\ZFC$ and that
+the latter do not lead to contradictions (i.e., $\ZFC$ is
+\emph{consistent}), then we can't prove that there exists a set $M$
+and a binary relation $E$ such that $\lb M,E\rb$ satisfies the $\ZFC$
+axioms.
+
+A relative consistency proof for an axiom $A$ is then obtained by assuming that such a
+$\lb M,E\rb$ exists, and constructing another $\lb M',E'\rb$
+satisfying $\ZFC + A$. We single out a very special kind of models:
+%
+\begin{definition}\label{def:transitive-model}
+  \begin{enumerate}
+  \item A set $M$ (of sets) is \emph{transitive} if for all $x\in M$ and 
+    $y\in x$, we have $y\in M$.
+  \item $\lb M,E\rb$ is a \emph{transitive model} if $M$ is transitive
+    set and  $E$ is the membership relation $\in$ restricted to
+    $M$. It is \emph{countable} if $M$ is coordinable with a subset of
+    $\N$; we then say that $M$ is a \emph{ctm}. 
+  \end{enumerate}
+\end{definition}
+%
+As in the last sentence, one usually uses the underlying set to refer to a transitive
+model, since the relation is fixed.
+
+Despite G\"odel's Second, we can find transitive models for every
+finite fragment of $\ZFC$. More precisely, 
+%
+\begin{theorem}\label{th:ctm-finite-axioms}
+  For each finite subset
+  $\Phi\sbq \ZFC$, the statement \emph{``there exists a countable
+  transitive model of $\Phi$''} is a theorem of $\ZFC$.
+\end{theorem}
+%
+This follows by a
+combination of three results (the Reflection Principle,
+L\"owenheim-Skolem Theorem, and the Mostowksi Collapse). The reader
+can consult the details in \cite{kunen2011set}. Consistency arguments that assume
+the existence of a ctm $M$ of $\ZFC$ can usually be replaced by a
+model as in Theorem~\ref{th:ctm-finite-axioms}, since a first-order
+proof (e.g.\ of contradiction) involves only finitely many axioms.
+
+Assume that $M$ is a ctm of $\ZFC$. In a portrayal convenient for this
+introduction, G\"odel showed that $M$ contains a minimal submodel
+$L^M$ of the same ``height'' (i.e.\ having the same ordinals)
+that satisfies $\ZFC+\CH$. The sets in $L^M$ are called
+\emph{constructible} and are in a sense ``definable.''
+
+It is a primary need to have a means to correlate  (first-order)
+properties satisfied by a model $M$ and those of a 
+submodel $N\sbq M$. As a simple example on this, consider 
+%% $M\defi \{\emptyset,\{\emptyset\},\{\{\emptyset\}\}\}$ and
+%% $N\defi\{\{\emptyset\},\{\{\emptyset\}\}\}$, and let 
+%% $\phi(x)\defi \neg\exists y (y \in x)$. It is clear that 
+%% $M\models \phi(\emptyset)$; but we have
+%% \[
+%% M\not\models \phi(\{\emptyset\}) \quad\text{ but }\quad N\models \phi(\{\emptyset\}).
+%% \]
+%% There is a discrepancy between what $M$ and $N$ \emph{believe} about
+%% their common element $\{\emptyset\}$.
+$M\defi \{a,b,c, \{a,b\},\{a,b,c\}\}$ and
+$N\defi\{a,b,\{a,b,c\}\}$, and let 
+\[
+\phi(x,y,z)\defi \forall w \,( w\in z \lsii w=x \o w=y).
+\]
+%% It is clear that  $M\models \phi(\{a,b\})$; but we have
+Then we have
+\[
+M\not\models \phi(a,b,\{a,b,c\}) \quad\text{ but }\quad N\models \phi(a,b,\{a,b,c\}).
+\]
+There is a discrepancy between  $M$ and $N$ about $\{a,b,c\}$ being ``the
+(unordered) pair of $a$ and $b$.'' We say that $\phi$ holds for
+$a,b,\{a,b,c\}$ \emph{relative} to $N$. It is immediate to see that
+$\phi$ holds  for $x,y,z$ relative to $N$ if and only if
+\[
+\phi^N(x,y,z)\defi \forall w.\ w\in N\limp ( w\in z \lsii w=x \o w=y)
+\] 
+holds. $\phi^N$ is called the \emph{relativization of $\phi$ to
+  $N$}. In general, one can relativized
+
+It can be shown elementarily that if $M$ and $N$ are transitive,
+formula $\phi$ holds for $x,y,z\in N$
+
+It is noteworthy that
+
+ One of the main
+technical tools designed by him
 
@@
 \end{enumerate}
+
+%%% Local Variables:
+%%% mode: latex
+%%% ispell-local-dictionary: "american"
+%%% TeX-master: "first_steps_into_forcing"
+%%% End:

### 2018-06-22 LSFA/intro.tex
línea que demarca el final de lo que tendría sentido.
absolutez

@@
 $\ZFC$. Forcing has since then been used for showing innumerable
-independence results and perform mathematical constructions.
+independence results and to perform mathematical constructions.
 
@@
   \item A set $M$ (of sets) is \emph{transitive} if for all $x\in M$ and 
-    $y\in x$, we have $y\in M$.
+    $y\in x$, we have $y\in M$ (i.e., every element of $M$ is a subset
+    of $M$).
   \item $\lb M,E\rb$ is a \emph{transitive model} if $M$ is transitive
@@
 holds. $\phi^N$ is called the \emph{relativization of $\phi$ to
-  $N$}. In general, one can relativized
+  $N$}. One can generalize this operation of relativization to the
+class of all sets satisfying a first-order predicate $C$ in a
+straightforward way:
+\[
+\phi^C(x,y,z)\defi \forall w.\ C(w)\limp ( w\in z \lsii w=x \o w=y)
+\]
+
+It can be shown elementarily that if $M$ and $N$ are transitive,
+$\phi^N$ holds if and only if $\phi^M$ holds,  for $x,y,z\in N$. We
+say then that $\phi$ is \emph{absolute between $N$ and $M$.}
+
+The concept of relativization and absoluteness are central to task of
+transferring truth of axioms in $M$ to  $L^M$. At this point we 
+point out that although $L^M$ is a subset of $M$. 
+\bigskip
+
+\hrule
+\bigskip
+
+In general, one can relativized
 

### 2018-06-22 LSFA/intro.tex
forcing

@@
 
-The concept of relativization and absoluteness are central to task of
-transferring truth of axioms in $M$ to  $L^M$. At this point we 
-point out that although $L^M$ is a subset of $M$. 
-\bigskip
+The concepts of relativization and absoluteness are central to task of
+transferring truth of axioms in $M$ to  $L^M$.
 
-\hrule
-\bigskip
-
-In general, one can relativized
-
-It can be shown elementarily that if $M$ and $N$ are transitive,
-formula $\phi$ holds for $x,y,z\in N$
-
-It is noteworthy that
-
- One of the main
-technical tools designed by him
-
-\begin{enumerate}
-\item $\CH$ is (Platonistically) undecidable.
-  \begin{enumerate}
-  \item By G\"odel's Second Incompleteness Theorems, we can't show
-    that a model of $\ZFC$ exists.
-  \item G\"odel showed that $\CH$ is relatively consistent with
-    $\ZFC$. ``Easy.''
-  \item Cohen showed that  $\neg\CH$ is relatively consistent with
-    $\ZFC$. Not ``easy.'' Formalization Open.
-  \end{enumerate}
-\item Using $\ZF$, G\"odel defined the constructible universe,
-  $L$. G\"odel showed that $\lb L,\in\rb\models \ZFC+\CH$. Paulson formalized $L$
-  for the consistency of $\AC$.
-\item The main concept for G\"odel's proofs is
-  \emph{relativization}. This is the core of Paulson's development.
-  \begin{enumerate}
-  \item \dots
-  \end{enumerate}
-  %
-\item 
-\item Cohen invented Forcing.
-  \begin{enumerate}
-  \item Countable transitive models $M$. These are enough for consistency
-    proofs by the L\"owenheim-Skolem Theorem, the Reflection
-    Principle, and the Mostowksi Collapse.
-    %
-  \item 
-    %
-  \item Forcing is a technique to extend countable transitive models of
+\subsection{Forcing}
+Forcing is a technique to extend countable transitive models of
 $\ZFC$. This process is guaranteed to preserve the $\ZFC$
 axioms but allows to fine-tune what other first order properties the
-    extension will have. 
-    %
-  \item Given a preorder with top $\lb\PP,\leq,\1\rb$ in $M$, define
-    \emph{filter} $G$, \emph{dense}. In general, there is no  $G\in M$
-    that intersects every dense
-    subset of $\PP$ that lies in $M$ (``generic'').
-  \item Rasiowa-Sikorski states that there is such $G$ for every
-    family of countable dense subsets. Thus, there are generic filters
-    for countable transitive models.
-  \end{enumerate}
+extension will have. Given a ctm $M$ of $\ZFC$, a new set $G$ is
+adjoined to $M$ and under some hypothesis ($G$ being ``generic''),
+there exists a smallest model $M[G]$ of $\ZFC$ that includes $M$ and
+contains $G$.
 
-\end{enumerate}
+\fbox{The next should be spread accross the other sections}
+
+The easiest way to define genericity is by using preorders.
+Given a preorder with top $\lb\PP,\leq,\1\rb$ in $M$, define
+\emph{filter} $G$ ... , \emph{dense} ...  $G$ is an $M$-\emph{generic}
+if it intersects every dense subset of $\PP$ that lies in $M$.
+
+But since $M$ is countable,
+
+The  Rasiowa-Sikorski theorem  states that that for every countable 
+family of dense subsets there is a filter that intersects them
+all. Thus, there are generic filters $G$ for countable transitive models.  In
+general, no such  $G$ belongs to $M$ and therefore the extension
+$M[G]$ is proper.
 

### 2018-06-22 LSFA/names.tex
División del archivo para dividir el trabajo
sección pair-gen-ext y algo de names

--- 2018-06-18 LSFA/intro.tex
@@
-\section{Introduction}\label{sec:introduction}
-
-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
-\subsection{Models of ZFC}
-
-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
-\subsection{Isabelle/ZF}
+\section{Names}
 
@@
 %%% End:
+
+The forcing idea consists of finding a model extending $M$ such that
+it satisfies $ZFC + \neg CH$. This model is called the $\textit{generic extension}$
+$M[G]$, where $G$ is a \emph{$\calD$-generic} filter in $P$.
+
+The elements of $M[G]$ will be constructed by the following function $\val$:
+
+\begin{isabelle}
+\isacommand{definition}\isamarkupfalse%
+\isanewline
+\ \ Hv\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}i{\isasymRightarrow}i{\isasymRightarrow}i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
+\ \ {\isachardoublequoteopen}Hv{\isacharparenleft}G{\isacharcomma}x{\isacharcomma}f{\isacharparenright}\ {\isacharequal}{\isacharequal}\ {\isacharbraceleft}\ f{\isacharbackquote}y\ {\isachardot}{\isachardot}\ y{\isasymin}\ domain{\isacharparenleft}x{\isacharparenright}{\isacharcomma}\ {\isasymexists}p{\isasymin}P{\isachardot}\ {\isacharless}y{\isacharcomma}p{\isachargreater}\ {\isasymin}\ x\ {\isasymand}\ p\ {\isasymin}\ G\ {\isacharbraceright}{\isachardoublequoteclose}\isanewline
+\isanewline
+\isacommand{definition}\isamarkupfalse%
+\isanewline
+\ \ val\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}i{\isasymRightarrow}i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
+\ \ {\isachardoublequoteopen}val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}\ {\isacharequal}{\isacharequal}\ wfrec{\isacharparenleft}edrel{\isacharparenleft}eclose{\isacharparenleft}M{\isacharparenright}{\isacharparenright}{\isacharcomma}\ {\isasymtau}\ {\isacharcomma}Hv{\isacharparenleft}G{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\isanewline
+\end{isabelle}
+
+This definition proceeds by well-founded recursion on relation $edrel(eclose(M))$:
+
+\begin{isabelle}
+\isacommand{definition}\isamarkupfalse%
+\isanewline
+\ \ edrel\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i\ {\isasymRightarrow}\ i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
+\ \ {\isachardoublequoteopen}edrel{\isacharparenleft}A{\isacharparenright}\ {\isacharequal}{\isacharequal}\ {\isacharbraceleft}{\isacharless}x{\isacharcomma}y{\isachargreater}\ {\isasymin}\ A{\isacharasterisk}A\ {\isachardot}\ x\ {\isasymin}\ domain{\isacharparenleft}y{\isacharparenright}{\isacharbraceright}{\isachardoublequoteclose}
+\end{isabelle}
+
+We can characterize the definition of $\val$ with the following lemma:
+
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ def{\isacharunderscore}val{\isacharcolon}\ \ {\isachardoublequoteopen}x{\isasymin}M\ {\isasymLongrightarrow}\ val{\isacharparenleft}G{\isacharcomma}x{\isacharparenright}\ {\isacharequal}\ {\isacharbraceleft}val{\isacharparenleft}G{\isacharcomma}t{\isacharparenright}\ {\isachardot}{\isachardot}\ t{\isasymin}domain{\isacharparenleft}x{\isacharparenright}\ {\isacharcomma}\ {\isasymexists}p{\isasymin}P\ {\isachardot}\ \ {\isasymlangle}t{\isacharcomma}\ p{\isasymrangle}{\isasymin}x\ {\isasymand}\ p\ {\isasymin}\ G\ {\isacharbraceright}{\isachardoublequoteclose}
+\end{isabelle}
+
+Thus, determining $\val(G,\tau)$ uses the values $\val(G,\sigma)$ for smaller $\sigma$ :
+
+    $\langle \sigma , \tau \rangle \in edrel(Memrel(M)) $ iff $\langle \sigma , p \rangle \in \tau$ with $p \in G$.

### 2018-06-23 LSFA/isabelle-zf.tex
Primera explicación de relativización.

@@
 of ZF, Isabelle/ZF has only one relation symbol
-\isatt{\ensuremath{\epsilon}} (equality and inclusion are defined
+\isatt{\ensuremath{\in}} (equality and inclusion are defined
 using membership); it also features one constant \isatt{\isadigit{0}}
@@
 
+We are not discussing how, if possible at all, to express that some
+set satisfies some axiom. For example, we might well be interested in
+proving $\emptyset \models \mathit{foundation}$, but there is no
+sensible way to state this, whithout encoding formulas and axioms in
+ZFC.\footnote{Indeed Paulson has encoded formulas using a very general
+  method to encode recursive definitions. In our development, although
+  not reported in this paper, we use this internalization.} In
+Isabelle/ZF there are, however, relativized versions of the
+axioms. This alternative statement of the axioms are relativized with
+respect to classes, which in Isabelle/ZF corresponds to predicates over
+sets. For example, the relativized axiom of foundation is:
+\begin{isabelle}
+\isacommand{definition}\isamarkupfalse%
+\ foundation{\isacharunderscore}ax\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharparenleft}i{\isacharequal}{\isachargreater}o{\isacharparenright}\ {\isacharequal}{\isachargreater}\ o{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
+\ \ \ \ {\isachardoublequoteopen}foundation{\isacharunderscore}ax{\isacharparenleft}M{\isacharparenright}\ {\isacharequal}{\isacharequal}\isanewline
+\ \ \ \ \ \ \ \ {\isasymforall}x{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ {\isacharparenleft}{\isasymexists}y{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ y{\isasymin}x{\isacharparenright}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\isasymexists}y{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ y{\isasymin}x\ {\isacharampersand}\ {\isachartilde}{\isacharparenleft}{\isasymexists}z{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ z{\isasymin}x\ {\isacharampersand}\ z\ {\isasymin}\ y{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
+
 % ∀x[c]
+\noindent The relativized quantifier
+\isa{{\isasymforall}x{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\
+  P(x)} is a shorthand for \isa{{\isasymforall}x{\isachardot}\ M(x)
+  {\isasymlongrightarrow} P(x)}. We are one step further to be able to
+express that the empty set satisfies foundation; but we must turn the
+set \isa{0 :: i} into a predicate. Isabelle/ZF provides the coercion
+\isa{{\isacharhash}{\isacharhash} :: i ={\isachargreater} (i ={\isachargreater} o)}, thus we can write and prove:
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ emp{\isacharunderscore}foundation\ {\isacharcolon}\ {\isachardoublequoteopen}foundation{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}{\isadigit{0}}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
+
+% Let us ponder if one can state, let alone prove, that some set \isa{M
+%   :: i} satisfies a statement \isa{\ensuremath{\phi} :: o}. Since
+% \isa{o} does not have an inductive definition, we cannot define the
+% satisfaction of \isa{\ensuremath{\phi}}.
+
+% As we have outlined above, relativization is a key concept when
+% proving that axioms satisfied by some model $M$ are also satisfied in
+% another model constructed from $M$. In this paper, we show the complete
+% proof that if a set \isa{M :: i} satisfies the
+
 % locales

### 2018-06-23 LSFA/isabelle-zf.tex
cambios menores y comentarios

@@
 essential to keep this paper self-contained, and refer the interested
-reader to Paulson's articles. Isabelle/ZF is a theory built upon the
-core library FOL of classical first order logic.
+reader to Paulson's articles. 
+%% Isabelle/ZF is a theory built upon the
+%% core library FOL of classical first order logic.
+Isabelle/ZF includes a development of classical first order logic,
+FOL. Both of them are  built upon the core library \emph{Pure}.
 
 In Isabelle/ZF sets are \emph{individuals}, that is terms of type
-\isatt{i} and formulas have type \isatt{o}. As in usual presentations
+\isatt{i} and formulas have type \isatt{o} 
+(akin to a \emph{Bool} type, but at the object level). 
+As in usual presentations
 of ZF, Isabelle/ZF has only one relation symbol
@@
 Working with the library and extendig it is quite
-straightforward. Since we find a need to combine the replacement and
-separation schemes to speak of the image of a function over some set
-defined by comprehension, we introduced a new term-former:
+straightforward. 
+As an example, we introduce  a new term-former which is combination of
+instances of replacement and separation to speak of the image of a function over some set
+defined by comprehension, namely $\{b(x): x\in A\text{ and }Q(x)\}$:
+%% Since we find a need to combine the replacement and
+%% separation schemes to speak of the image of a function over some set
+%% defined by comprehension, we introduced a new term-former:
 \begin{isabelle}
@@
 \end{isabelle}
-\noindent It is more convenient to add an abbreviation for a simpler
-writing and a nicer reading, so we use \isa{{\isacharbraceleft}b\
+\noindent %% It is more convenient to add an abbreviation for a simpler
+%% writing and a nicer reading, so we use
+We are then able to add an abbreviation resembling the usual
+mathematical writing by setting 
+\isa{{\isacharbraceleft}b\
   {\isachardot}{\isachardot}\ x{\isasymin}A{\isacharcomma}\
-  Q{\isacharbraceright}} for
+  Q{\isacharbraceright}} as a notation for
 \isa{SepReplace{\isacharparenleft}A{\isacharcomma}b{\isacharcomma}Q{\isacharparenright}}. The
@@
 proving $\emptyset \models \mathit{foundation}$, but there is no
-sensible way to state this, whithout encoding formulas and axioms in
+sensible way to state this, without encoding formulas and axioms in
 ZFC.\footnote{Indeed Paulson has encoded formulas using a very general
   method to encode recursive definitions. In our development, although
-  not reported in this paper, we use this internalization.} In
+  not reported in this paper, we use this internalization.}
+
+\fbox{ You need this in order to formalize $\models$ as a predicate of
+  type $i\ent i \ent o$. (Don't know how to put that here.)
+}
+
+ In
 Isabelle/ZF there are, however, relativized versions of the

### 2018-06-23 LSFA/intro.tex
linea de más
edicion global

@@
 \subsection{Models of $\ZFC$}
-By G\"odel's Second Incompleteness Theorem, we can't show prove that
+By G\"odel's Second Incompleteness Theorem, we can't  prove that
 there exists a model of $\ZFC$. More formally, if we assume that
@@
   \item $\lb M,E\rb$ is a \emph{transitive model} if $M$ is transitive
-    set and  $E$ is the membership relation $\in$ restricted to
+     and  $E$ is the membership relation $\in$ restricted to
     $M$. It is \emph{countable} if $M$ is coordinable with a subset of
-    $\N$; we then say that $M$ is a \emph{ctm}. 
+    $\N$; we then say that the model  $M$ is a \emph{ctm}. 
   \end{enumerate}
@@
 that satisfies $\ZFC+\CH$. The sets in $L^M$ are called
-\emph{constructible} and are in a sense ``definable.''
+\emph{constructible} and are in a sense ``definable.'' In fact, there
+is a first-order formula $L$ such that $L^M = \{x\in M : M\models
+L(x)\}$.
 
@@
 
-\fbox{The next should be spread accross the other sections}
-
-The easiest way to define genericity is by using preorders.
-Given a preorder with top $\lb\PP,\leq,\1\rb$ in $M$, define
-\emph{filter} $G$ ... , \emph{dense} ...  $G$ is an $M$-\emph{generic}
+The easiest way to define genericity is by using a preorder with top
+$\lb\PP,\leq,\1\rb$ in $M$.   
+In Section~\ref{sec:forcing-notions} we formalize the definitions of
+\emph{dense} subset and  \emph{filter} of  $\PP$, and we say that  $G$
+is an $M$-generic filter
 if it intersects every dense subset of $\PP$ that lies in $M$.
 
-But since $M$ is countable,
-
-The  Rasiowa-Sikorski theorem  states that that for every countable 
-family of dense subsets there is a filter that intersects them
+The  Rasiowa-Sikorski lemma  states that that for every countable 
+family of dense subsets of a preorder there is a filter that intersects them
 all. Thus, there are generic filters $G$ for countable transitive models.  In
 general, no such  $G$ belongs to $M$ and therefore the extension
-$M[G]$ is proper.
+$M[G]$ is proper. We formalize the proof of the Rasiowa-Sikorski lemma
+in Section~\ref{sec:rasiowa-sikorski-lemma}. A preliminary result on
+a version of the Axiom of Choice is formalized in
+Section~\ref{sec:sequence-version-dc}.
+
+Elements of $M[G]$ are coded by the elements of $M$, and are obtained
+from them by applying a function \isatt{val} that uses the generic filter
+as a parameter. In order to show
+that $M[G]$ does  extend $M$, we formalize the function \isatt{check}
+that given an element $x\in M$ provides with a code (a ``name'') for
+it.  This is explained in Section~\ref{sec:names}.
 

### 2018-06-23 LSFA/rs.tex
edicion global

@@
-\subsection{A sequence version of $\DC$}
+\subsection{A sequence version of $\DC$}\label{sec:sequence-version-dc}
 In order to prove the Rasiowa-Sikorski lemma in
@@
 
-The function \verb|dc_witness| has, apart from $A$, $a$ and $R$, a
+The function \isatt{dc{\isacharunderscore}witness} has, apart from $A$, $a$ and $R$, a
 function $s$ as a parameter. If this function is a selector for
-$\P(A)\sm \{\mty\}$, the function $f\defi$\verb|dc_witness|$(A,a,s,R)$
-will satify $\DC$.
+$\P(A)\sm \{\mty\}$, the function $f\defi {}$\isatt{dc{\isacharunderscore}witness}$(A,a,s,R)$
+will satify $\DC$. Notice that $s$ is a term of type \isatt{i} (a
+function construed as a set of pairs) and an expression
+\isatt{s{\isacharbackquote}b} is notation for  \isatt{apply(s,b)},
+where
+\isatt{apply\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i\ {\isasymRightarrow}\ i\ {\isasymRightarrow}\ i{\isachardoublequoteclose}}
+is the operation of function application. We will not make a
+distinction in our prose and simply write \isatt{s}(\isatt{b}) in this case.
 
@@
 
-\subsection{The Rasiowa-Sikorski lemma}
+\subsection{The Rasiowa-Sikorski lemma}\label{sec:rasiowa-sikorski-lemma}
 In order to state this Lemma, we gather the relevant hypotheses into a locale:

### 2018-06-23 LSFA/first_steps_into_forcing.tex
mi base bibliográfica
sólo dejo mi bib porque hay repetidos
macros para markup de Isabelle
sección separada de Isa/ZF
empezando texto de preservación de pares
macros edrel, eclose; abstract; isabelle-zf es Section

@@
 \usepackage{graphicx}
+\usepackage{isabelle,isabellesym}
 \sloppy
@@
 \begin{abstract} 
+  We lay the ground for an Isabelle/ZF formalization of Cohen's technique of
+  \emph{forcing}. We formalize the definition of forcing notions as
+  preorders with top, dense subsets, and generic filters. We formalize
+  a version of the principle of \emph{Dependent Choices} and using it
+  we prove the Rasiowa-Sikorski lemma on existence of generic filters.
+
+  Given a transitive set $M$, we define the generic extension $M[G]$,
+  the canonical names for elements of $M$ and finally show that if $M$
+  satisfy the axiom of pairing, the $M[G]$ also does.
 \end{abstract}
@@
 \input{intro}
+
+\input{isabelle-zf}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\input{pair-gen-ext}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \input{conclusion}
@@
 \bibliographystyle{entcs}
-\bibliography{first_steps_into_forcing}
+\bibliography{citados}
 \end{document}

### 2018-06-23 LSFA/intro.tex
macros edrel, eclose; abstract; isabelle-zf es Section

@@
 independence results and to perform mathematical constructions.
+
+A great part of G\"odel's work on this subject has been formalized in
+Isabelle by Lawrence Paulson. This paper formalizes a first part of
+the machinery of forcing. In the next Subsections we discuss some of
+the set-theoretical  details involved.
 
@@
 The concepts of relativization and absoluteness are central to task of
-transferring truth of axioms in $M$ to  $L^M$.
+transferring truth of axioms in $M$ to  $L^M$, and consitute the
+hardest part of Paulson's development.
 
@@
 $M[G]$ is proper. We formalize the proof of the Rasiowa-Sikorski lemma
-in Section~\ref{sec:rasiowa-sikorski-lemma}. A preliminary result on
+in Section~\ref{sec:rasiowa-sikorski-lemma}. A requisite result on
 a version of the Axiom of Choice is formalized in
@@
 
+A central part of this formalization project involves showing that the
+$\ZFC$ holds in the generic extension. This is most relevant since
+forcing is essentially the only known way to \emph{extend} models of
+$\ZFC$ (while preserving ordinals). The most difficult step to achieve
+this goal is to define the \emph{forcing relation}, that allows to
+connect satisfaction in $M$ to that of $M[G]$; this is needed to show
+that the axiom schemes of Separation and Replacement hold in $M[G]$,
+as well as the Powerset axiom. In this work we tackle the Pairing
+Axiom. This doesn't require the 
+forcing relation, but provides an illustration of the use of names.
+
+We continue with a brief description of the Isabelle framework. 
 %%% Local Variables:

### 2018-06-24 LSFA/intro.tex
dunno if checks are in M
typo
Modificaciones menores y comentarios sobre la intro

--- 2018-06-23 LSFA/intro.tex
@@
 \section{Introduction}\label{sec:introduction}
-Set Theory play a double role in Mathematics: It is one of its
-possible foundations, and also an active research area.
+Set Theory plays a double role in Mathematics: it is one of its
+possible foundations and also an active research area.
 %
@@
 
-The current  axiomatic foundation of Set Theory is through 
-first-order logic and uses the axioms by  Zermelo, Fraenkel, and
-including the Axiom of Choice ($\AC$) among them. This theory is
-known by the $\ZFC$ acronym. G\"odel showed that $\CH$ can't be
-refuted using the $\ZFC$, unless the latter are inconsistent (we say
-that \emph{$\CH$ is relatively 
-  consistent with  $\ZFC$}). For a while, this result left the
-possibility that one might be able show that $\ZFC\models \CH$, but in
-a groundbreaking work \cite{Cohen-CH-PNAS}, Paul Cohen discovered the
-technique of 
-\emph{forcing} and showed that $\neg\CH$ is relatively consistent with
-$\ZFC$. Forcing has since then been used for showing innumerable
-independence results and to perform mathematical constructions.
+The current axiomatic foundation of Set Theory is through first-order
+logic and uses the axioms devised by Zermelo and Fraenkel, including
+the Axiom of Choice ($\AC$) among them. This theory is known by the
+$\ZFC$ acronym. G\"odel \cite{godel-L} showed that $\CH$ cannot be refuted using
+$\ZFC$, unless this theory itself is inconsistent (we say that
+\emph{$\CH$ is relatively consistent with $\ZFC$}). For a while, this
+result left the possibility that one might be able to show
+$\ZFC\models \CH$, but in a groundbreaking work \cite{Cohen-CH-PNAS},
+Paul Cohen discovered the technique of \emph{forcing} and proved that
+$\neg\CH$ is relatively consistent with $\ZFC$. Forcing has been used
+since then for showing innumerable independence results and to perform
+mathematical constructions.
+
+A great part of G\"odel's work on this subject has been formalized in
+Isabelle by Larry Paulson. This paper formalizes a first part of the
+machinery of forcing. In the rest of the introduction we discuss some
+of the set-theoretical details involved and explain briefly Paulson's
+formalization.
 
 \subsection{Models of $\ZFC$}
-By G\"odel's Second Incompleteness Theorem, we can't  prove that
+By G\"odel's Second Incompleteness Theorem, we cannot  prove that
 there exists a model of $\ZFC$. More formally, if we assume that
@@
 the latter do not lead to contradictions (i.e., $\ZFC$ is
-\emph{consistent}), then we can't prove that there exists a set $M$
+\emph{consistent}), then we cannot prove that there exists a set $M$
 and a binary relation $E$ such that $\lb M,E\rb$ satisfies the $\ZFC$
@@
 
-A relative consistency proof for an axiom $A$ is then obtained by assuming that such a
-$\lb M,E\rb$ exists, and constructing another $\lb M',E'\rb$
-satisfying $\ZFC + A$. We single out a very special kind of models:
+A relative consistency proof for an axiom $A$ is then obtained by
+assuming that there exists a model of ZFC, say $\lb M,E\rb$, and
+constructing another model $\lb M',E'\rb$ for $\ZFC + A$. We single
+out a very special kind of models:
 %
@@
 %
-As in the last sentence, one usually uses the underlying set to refer to a transitive
-model, since the relation is fixed.
-
-Despite G\"odel's Second, we can find transitive models for every
-finite fragment of $\ZFC$. More precisely, 
+\noindent As in the last sentence, one usually refer to a transitive
+model by the underlying set because the relation is fixed.
+%% MIGUEL: feísimo este corte. Quizás refrasear la oración anterior
+%% para que entre en un renglón.
+%% MIGUEL: lo que sigue implica que para tener aritmética en ZFC son
+%% necesarios una cantidad infinita de axiomas de ZFC?
+In spite of G\"odel's Second Incompleteness Theorem, one can find
+transitive models for every finite fragment of $\ZFC$. More precisely,
 %
@@
 %
-This follows by a
-combination of three results (the Reflection Principle,
-L\"owenheim-Skolem Theorem, and the Mostowksi Collapse). The reader
-can consult the details in \cite{kunen2011set}. Consistency arguments that assume
-the existence of a ctm $M$ of $\ZFC$ can usually be replaced by a
-model as in Theorem~\ref{th:ctm-finite-axioms}, since a first-order
-proof (e.g.\ of contradiction) involves only finitely many axioms.
+%% MIGUEL: controlar que tiene sentido lo que menciono en la nota al pie!
+%% MIGUEL: Me da la sensación que podemos promover lo de la nota al pie
+%% al cuerpo y luego explicar que por eso es importante el teorema.
+This follows by a combination of the Reflection Principle, the
+L\"owenheim-Skolem Theorem, and the Mostowksi Collapse. The reader can
+consult the details in \cite{kunen2011set}. Consistency arguments that
+assume the existence of a ctm $M$ of $\ZFC$ can usually be replaced by
+a model as in Theorem~\ref{th:ctm-finite-axioms}, since a first-order
+proof (e.g.\ of a contradiction)\footnote{We purportedly mention this
+  example because results about the independence of some axiom $A$
+  with respect to $\ZFC$ can be stated as transforming a proof
+  concluding a contradiction from $\ZFC+A$ to one from $\ZFC$.}
+involves only finitely many axioms.
 
+%% MIGUEL: Qué significa "portrayal convenient"?
+%% MIGUEL: ¿por qué esto es conveniente para esta introducción?
 Assume that $M$ is a ctm of $\ZFC$. In a portrayal convenient for this
@@
 
+%% MIGUEL: acá me parece que hay un salto medio abrupto. Por que
+%% es una necesidad primaria? Lo podés relacionar con lo que venía
+%% del párrafo anterior?
+%% MIGUEL: quizás lo que querés decir acá es para probar que vamos
+%% a construir un modelo N a partir de un modelo M; y para probar
+%% que N satisface un axioma, vamos a utilizar que M lo satisface.
 It is a primary need to have a means to correlate  (first-order)
@@
 say then that $\phi$ is \emph{absolute between $N$ and $M$.}
-
-The concepts of relativization and absoluteness are central to task of
-transferring truth of axioms in $M$ to  $L^M$.
+%% MIGUEL: "truth of axioms in $M$" significa que $M$ satisface los
+%% axiomas?
+The concepts of relativization and absoluteness are central to the
+task of transferring truth of axioms in $M$ to $L^M$, and consitute
+the hardest part of Paulson's development.
 
 \subsection{Forcing}
+%% MIGUEL: no veo la necesidad del adversativo but, pondría and.
+%% MIGUEL: Qué significa "a new set $G$ is adjoined to $M$"?
+%% MIGUEL: este párrafo necesita correcciones.
 Forcing is a technique to extend countable transitive models of
@@
 
-The  Rasiowa-Sikorski lemma  states that that for every countable 
-family of dense subsets of a preorder there is a filter that intersects them
-all. Thus, there are generic filters $G$ for countable transitive models.  In
-general, no such  $G$ belongs to $M$ and therefore the extension
-$M[G]$ is proper. We formalize the proof of the Rasiowa-Sikorski lemma
-in Section~\ref{sec:rasiowa-sikorski-lemma}. A preliminary result on
-a version of the Axiom of Choice is formalized in
+%% MIGUEL: Tenemos que dado un ctm M, existe un filtro genérico G.
+%% ¿Dónde existe? Una primera lectura de ignorante me dice "existe
+%% en M", pero eso me lo desmiente la siguiente oración. ¿Juega algún
+%% rol absolutez? Si es así estaría bien explicitarlo.
+%% MIGUEL: la última oración me resulta un poco difícil de entender.
+The Rasiowa-Sikorski lemma states that, fixed a preorder $\PP$, for
+any countable family of dense subsets of $\PP$ there is a filter
+intersecting all of them. Thus, there are generic filters $G$ for
+countable transitive models; the extension of a ctm $M$ by the filter
+$G$ is denoted by $M[G]$. In general, no such $G$ belongs to $M$ and
+therefore the extension $M[G]$ is proper. We formalize the proof of
+the Rasiowa-Sikorski lemma in
+Section~\ref{sec:rasiowa-sikorski-lemma}. A requisite result on a
+version of the Axiom of Choice is formalized in
 Section~\ref{sec:sequence-version-dc}.
 
-Elements of $M[G]$ are coded by the elements of $M$, and are obtained
-from them by applying a function \isatt{val} that uses the generic filter
-as a parameter. In order to show
-that $M[G]$ does  extend $M$, we formalize the function \isatt{check}
-that given an element $x\in M$ provides with a code (a ``name'') for
-it.  This is explained in Section~\ref{sec:names}.
+%% MIGUEL: redacción alternativa del párrafo siguiente.
+% The elements of $M[G]$ are the decodification of certain elements of
+% $M$, called \emph{names}. The decoding is given by the function
+% $\isa{val}$ which takes the generic filter $G$ as a parameter. To
+% prove that $M$ is contained in $M[G]$ it suffices to give a name for
+% each element of $M$; we define the function \isa{check} which assigns
+% a name for each $x\in M$. Showing that \isa{check(x)\ \isasymin\ M}
+% when \isa{x\ \isasymin\ M} involves some technical issues that will
+% be formalized in a further work. We explain names, \isa{val}, and
+% \isa{check} in Section~\ref{sec:names}.
+
+Elements of $M[G]$ are coded by elements of $M$, and are obtained from
+them by applying a function \isa{val} that uses the generic filter as
+a parameter. In order to show that $M[G]$ does extend $M$, we
+formalize the function \isa{check} that given an element $x\in M$
+provides with a code (a ``name'') for it. Show that this name belongs
+to $M$ involves some technical issues that will be addressed in a
+further work. This is all explained in Section~\ref{sec:names}.
+
+A central part of this formalization project involves showing that
+$\ZFC$ holds in the generic extension. This is most relevant since
+forcing is essentially the only known way to \emph{extend} models of
+$\ZFC$ (while preserving ordinals). The most difficult step to achieve
+this goal is to define the \emph{forcing relation}, that allows to
+connect satisfaction in $M$ to that of $M[G]$; this is needed to show
+that the Powerset axiom and the axiom schemes of Separation and
+Replacement hold in $M[G]$. In this work we tackle the Pairing
+Axiom. This doesn't require the forcing relation, but provides an
+illustration of the use of names.
+
+%We continue with a brief description of the Isabelle framework.
 

### 2018-06-24 LSFA/intro.tex
correcciones de Miguel a la intro

@@
 \section{Introduction}\label{sec:introduction}
-Set Theory plays a double role in Mathematics: it is one of its
+Set Theory plays a double role in Mathematics: It is one of its
 possible foundations and also an active research area.
@@
 %% para que entre en un renglón.
+
 %% MIGUEL: lo que sigue implica que para tener aritmética en ZFC son
@@
 a model as in Theorem~\ref{th:ctm-finite-axioms}, since a first-order
-proof (e.g.\ of a contradiction)\footnote{We purportedly mention this
-  example because results about the independence of some axiom $A$
-  with respect to $\ZFC$ can be stated as transforming a proof
+proof (e.g.\ of a contradiction)%
+%%%%%%%%%%%%%
+\footnote{It is relevant to this point that 
+  both the approaches by G\"odel and Cohen for showing
+  relative independence of an axiom $A$ 
+  can be use to obtain an algorithm transforming a proof
   concluding a contradiction from $\ZFC+A$ to one from $\ZFC$.}
+%%%%%%%%%%%%%
 involves only finitely many axioms.
@@
 is a first-order formula $L$ such that $L^M = \{x\in M : M\models
-L(x)\}$.
+L(x)\}$. To show that  $L^M\models \ZFC+\CH$, one uses that the
+$\ZFC+\CH$ holds in $M$.
 
@@
 %% que N satisface un axioma, vamos a utilizar que M lo satisface.
-It is a primary need to have a means to correlate  (first-order)
+It is therefore a primary need to have a means to correlate  (first-order)
 properties satisfied by a model $M$ and those of a 
@@
 The concepts of relativization and absoluteness are central to the
-task of transferring truth of axioms in $M$ to $L^M$, and consitute
+task of transferring truth of axioms in $M$ to $L^M$, and constitute
 the hardest part of Paulson's development.
@@
 $\ZFC$. This process is guaranteed to preserve the $\ZFC$
-axioms but allows to fine-tune what other first order properties the
-extension will have. Given a ctm $M$ of $\ZFC$, a new set $G$ is
-adjoined to $M$ and under some hypothesis ($G$ being ``generic''),
-there exists a smallest model $M[G]$ of $\ZFC$ that includes $M$ and
-contains $G$.
+axioms while allowing to fine-tune what other first order properties the
+extension will have. Given a ctm $M$  of $\ZFC$ and a set $G$, a new
+ctm  $M[G]$  that includes $M$ and
+contains $G$ is constructed, and under some hypotheses ($G$ being ``generic''),
+$M[G]$ satisfies $\ZFC$.
 
@@
 intersecting all of them. Thus, there are generic filters $G$ for
-countable transitive models; the extension of a ctm $M$ by the filter
-$G$ is denoted by $M[G]$. In general, no such $G$ belongs to $M$ and
+countable transitive models. In general, no such $G$ belongs to $M$ and
 therefore the extension $M[G]$ is proper. We formalize the proof of
@@
 %% MIGUEL: redacción alternativa del párrafo siguiente.
-% The elements of $M[G]$ are the decodification of certain elements of
-% $M$, called \emph{names}. The decoding is given by the function
-% $\isa{val}$ which takes the generic filter $G$ as a parameter. To
-% prove that $M$ is contained in $M[G]$ it suffices to give a name for
-% each element of $M$; we define the function \isa{check} which assigns
-% a name for each $x\in M$. Showing that \isa{check(x)\ \isasymin\ M}
-% when \isa{x\ \isasymin\ M} involves some technical issues that will
-% be formalized in a further work. We explain names, \isa{val}, and
-% \isa{check} in Section~\ref{sec:names}.
-
-Elements of $M[G]$ are coded by elements of $M$, and are obtained from
-them by applying a function \isa{val} that uses the generic filter as
-a parameter. In order to show that $M[G]$ does extend $M$, we
-formalize the function \isa{check} that given an element $x\in M$
-provides with a code (a ``name'') for it. Show that this name belongs
-to $M$ involves some technical issues that will be addressed in a
-further work. This is all explained in Section~\ref{sec:names}.
+Every  $y \in M[G]$ is obtained from an element $\dot y$ of $M$, thus
+elements of $M$ are construed as \emph{names} or codes for elements of
+$M[G]$.
+The decoding is given by the function
+$\val$ which takes the generic filter $G$ as a parameter. To
+prove that $M$ is contained in $M[G]$ it suffices to give a name for
+each element of $M$; we define the function $\chk$ which assigns
+a name for each $x\in M$. Showing that $\chk(x)\in M$
+when $x\in M$ involves some technical issues that will
+be addressed/formalized in a further work. We explain names, $\val$, and
+$\chk$ in Section~\ref{sec:names}.
 

### 2018-06-24 LSFA/names.tex
edicion global
notacion

@@
 \section{Names}
-
-%%% Local Variables:
-%%% mode: latex
-%%% ispell-local-dictionary: "american"
-%%% TeX-master: "first_steps_into_forcing"
-%%% End:
+\label{sec:names}
 
 The forcing idea consists of finding a model extending $M$ such that
-it satisfies $ZFC + \neg CH$. This model is called the $\textit{generic extension}$
+it satisfies $\ZFC + \neg \CH$. This model is called the $\textit{generic extension}$
 $M[G]$, where $G$ is a \emph{$\calD$-generic} filter in $P$.
@@
 
-This definition proceeds by well-founded recursion on relation $edrel(eclose(M))$:
+This definition proceeds by well-founded recursion on relation $\edrel(\eclose(M))$:
 
@@
 
-    $\langle \sigma , \tau \rangle \in edrel(Memrel(M)) $ iff $\langle \sigma , p \rangle \in \tau$ with $p \in G$.
+    $\langle \sigma , \tau \rangle \in \edrel(\Memrel(M)) $ iff $\langle \sigma , p \rangle \in \tau$ with $p \in G$.
+
+%%% Local Variables:
+%%% mode: latex
+%%% ispell-local-dictionary: "american"
+%%% TeX-master: "first_steps_into_forcing"
+%%% End:

### 2018-06-24 LSFA/intro.tex
dejo addressed
Lawrence is back

@@
 A great part of G\"odel's work on this subject has been formalized in
-Isabelle by Larry Paulson. This paper formalizes a first part of the
+Isabelle by Lawrence Paulson. This paper formalizes a first part of the
 machinery of forcing. In the rest of the introduction we discuss some
@@
 when $x\in M$ involves some technical issues that will
-be addressed/formalized in a further work. We explain names, $\val$, and
+be addressed in a further work. We explain names, $\val$, and
 $\chk$ in Section~\ref{sec:names}.

### 2018-06-24 LSFA/names.tex
preliminar

--- 2018-06-22 LSFA/names.tex
@@
 \section{Names}
+\label{sec:names}
+In this section we formalize the function $\val$ that allows to
+construct the elements of the generic extension $M[G]$ from elements
+the ctm $M$ and the generic filter $G$. The definition of $\val$
+can be written succintly as
+%
+\begin{equation}\label{eq:def-val}
+\val(G,\tau)\defi \{\val(G,\sigma) : \exists p\in \PP
+(\lb\sigma,p\rb\in\tau \y p\in G)\}.
+\end{equation}
+%
+This proceeds by well-founded recursion on the relation
+$\mathit{ed}$:
+\[
+x \mathrel{\mathit{ed}} y \iff \exists p : \lb x,p\rb\in y.
+\]
+Recall that in $\ZFC$, an ordered pair $\lb x,y \rb$ is the set
+$\{\{x\},\{x,y\}\}$. Hence this relation is well-founded since $\in$
+is by the Foundation Axiom.
 
-%%% Local Variables:
-%%% mode: latex
-%%% ispell-local-dictionary: "american"
-%%% TeX-master: "first_steps_into_forcing"
-%%% End:
+The  formalization of $\val$ in Isabelle/ZF involves the function
+\isa{wfrec}, which takes as an argument a well-founded relation of
+type \isa{i}, i.e.\ a set of pairs. Therefore we restrict
+$\mathit{ed}$ above to a set.
+%
+\begin{isabelle}
+\isacommand{definition}\isamarkupfalse%
+\isanewline
+\ \ edrel\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i\ {\isasymRightarrow}\ i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
+\ \ {\isachardoublequoteopen}edrel{\isacharparenleft}A{\isacharparenright}\ {\isacharequal}{\isacharequal}\ {\isacharbraceleft}{\isacharless}x{\isacharcomma}y{\isachargreater}\ {\isasymin}\ A{\isacharasterisk}A\ {\isachardot}\ x\ {\isasymin}\ domain{\isacharparenleft}y{\isacharparenright}{\isacharbraceright}{\isachardoublequoteclose}
+\end{isabelle}
+%
+Since \isa{edrel(A)} is a subset of a  well-founded relation (the
+transitive closure of the membership relation restricted to \isa{A}),
+then it is well-founded as well.
 
-The forcing idea consists of finding a model extending $M$ such that
-it satisfies $ZFC + \neg CH$. This model is called the $\textit{generic extension}$
-$M[G]$, where $G$ is a \emph{$\calD$-generic} filter in $P$.
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ edrel{\isacharunderscore}sub{\isacharunderscore}memrel{\isacharcolon}\ {\isachardoublequoteopen}edrel{\isacharparenleft}A{\isacharparenright}\ {\isasymsubseteq}\ trancl{\isacharparenleft}Memrel{\isacharparenleft}eclose{\isacharparenleft}A{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
+\dots
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ wf{\isacharunderscore}edrel\ {\isacharcolon}\ {\isachardoublequoteopen}wf{\isacharparenleft}edrel{\isacharparenleft}A{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+%
+%
+\isacommand{apply}\isamarkupfalse%
+\ {\isacharparenleft}rule{\isacharunderscore}tac\ wf{\isacharunderscore}subset\ {\isacharbrackleft}of\ {\isachardoublequoteopen}trancl{\isacharparenleft}Memrel{\isacharparenleft}eclose{\isacharparenleft}A{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}{\isacharbrackright}{\isacharparenright}\isanewline
+\ \ \isacommand{apply}\isamarkupfalse%
+\ {\isacharparenleft}auto\ simp\ add{\isacharcolon}edrel{\isacharunderscore}sub{\isacharunderscore}memrel\ wf{\isacharunderscore}trancl\ wf{\isacharunderscore}Memrel{\isacharparenright}\isanewline
+\ \ \isacommand{done}\isamarkupfalse%
+%
+\end{isabelle}
 
-The elements of $M[G]$ will be constructed by the following function $\val$:
+We now turn to the definition of $\val$.
 
@@
 \ \ val\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}i{\isasymRightarrow}i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
-\ \ {\isachardoublequoteopen}val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}\ {\isacharequal}{\isacharequal}\ wfrec{\isacharparenleft}edrel{\isacharparenleft}eclose{\isacharparenleft}M{\isacharparenright}{\isacharparenright}{\isacharcomma}\ {\isasymtau}\ {\isacharcomma}Hv{\isacharparenleft}G{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
-\isanewline
+\ \ {\isachardoublequoteopen}val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}\ {\isacharequal}{\isacharequal}\ wfrec{\isacharparenleft}edrel{\isacharparenleft}eclose{\isacharparenleft}M{\isacharparenright}{\isacharparenright}{\isacharcomma}\ {\isasymtau}\ {\isacharcomma}Hv{\isacharparenleft}G{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
 \end{isabelle}
 
-This definition proceeds by well-founded recursion on relation $edrel(eclose(M))$:
-
-\begin{isabelle}
-\isacommand{definition}\isamarkupfalse%
-\isanewline
-\ \ edrel\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i\ {\isasymRightarrow}\ i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
-\ \ {\isachardoublequoteopen}edrel{\isacharparenleft}A{\isacharparenright}\ {\isacharequal}{\isacharequal}\ {\isacharbraceleft}{\isacharless}x{\isacharcomma}y{\isachargreater}\ {\isasymin}\ A{\isacharasterisk}A\ {\isachardot}\ x\ {\isasymin}\ domain{\isacharparenleft}y{\isacharparenright}{\isacharbraceright}{\isachardoublequoteclose}
-\end{isabelle}
-
-We can characterize the definition of $\val$ with the following lemma:
-
+Then we can obtain the recursive expression~(\ref{eq:def-val}) in the
+following lemma:
+%
 \begin{isabelle}
@@
 
-Thus, determining $\val(G,\tau)$ uses the values $\val(G,\sigma)$ for smaller $\sigma$ :
-
-    $\langle \sigma , \tau \rangle \in edrel(Memrel(M)) $ iff $\langle \sigma , p \rangle \in \tau$ with $p \in G$.
+%%% Local Variables:
+%%% mode: latex
+%%% ispell-local-dictionary: "american"
+%%% TeX-master: "first_steps_into_forcing"
+%%% End:

### 2018-06-24 LSFA/isabelle-zf.tex
macros edrel, eclose; abstract; isabelle-zf es Section

@@
 %-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
-\subsection{Isabelle/ZF}
+\section{Isabelle/ZF}
 
@@
 
+Mathematical texts usually start by fixing the context which defines
+both the theory, sometimes one explicitly introduces also the axioms
+of the theory, and syntactical conventions. One might then either
+prove that some structure satisfies the axioms and conclude some
+result about the structure or extend and refine the theory
+itself. \emph{Locales}~\cite{ballarin2010tutorial} are Isabelle's
+facilities to define contexts; there are also operations to
+\emph{extend} a locale with further assumptions (for example, moving
+from lattices to distributive lattices), to \emph{interpret} locales
+(roughly related to signature morphisms of universal algebra), and to
+\emph{instantiate} locales on concrete structures. In our work, we
+use locales to organize the formalization and to make explicit the
+assumptions of the most important results.
 % locales

### 2018-06-24 LSFA/names.tex
presentacion de check

@@
-\section{Names}
+\subsection{Names}
 \label{sec:names}
@@
 construct the elements of the generic extension $M[G]$ from elements
-the ctm $M$ and the generic filter $G$. The definition of $\val$
+of the ctm $M$ and the generic filter $G$. The definition of $\val$
 can be written succintly as
@@
 %
-This proceeds by well-founded recursion on the relation
+This definition proceeds by well-founded recursion on the relation
 $\mathit{ed}$:
@@
 
+%% It is to be noted that this results holds under the assumption that
+%% the name \isa{t} is in $M$.
+
+We now provide names for elements in $M$. That is, for each $x\in M$,
+we define $\chk(x)$ (usually denoted by $\check{x}$ in the literature)
+such that $\val(G,\chk(x))=x$. This will show that $M\sbq M[G]$, with
+a caveat we make explicit in the end of this section.
+
+The definition of $\chk(x)$ is a straightforward $\in$-recursion:
+\begin{equation}
+  \label{eq:def-check}
+  \chk(x)\defi\{\lb\chk(y),\1\rb : y\in x\}
+\end{equation}
+Now the set-relation argument for \isa{wfrec} is the membership
+relation restricted to a set \isa{A}, \isa{Memrel(A)}.
+
+\begin{isabelle}
+\isacommand{definition}\isamarkupfalse%
+\ \isanewline
+\ \ Hcheck\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharbrackleft}i{\isacharcomma}i{\isacharbrackright}\ {\isasymRightarrow}\ i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
+\ \ {\isachardoublequoteopen}Hcheck{\isacharparenleft}z{\isacharcomma}f{\isacharparenright}\ \ {\isacharequal}{\isacharequal}\ {\isacharbraceleft}\ {\isacharless}f{\isacharbackquote}y{\isacharcomma}one{\isachargreater}\ {\isachardot}\ y\ {\isasymin}\ z{\isacharbraceright}{\isachardoublequoteclose}\isanewline
+\isanewline
+\isacommand{definition}\isamarkupfalse%
+\isanewline
+\ \ check\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i\ {\isasymRightarrow}\ i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
+\ \ {\isachardoublequoteopen}check{\isacharparenleft}x{\isacharparenright}\ {\isacharequal}{\isacharequal}\ wfrec{\isacharparenleft}Memrel{\isacharparenleft}eclose{\isacharparenleft}{\isacharbraceleft}x{\isacharbraceright}{\isacharparenright}{\isacharparenright}{\isacharcomma}\ x\ {\isacharcomma}\ Hcheck{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
+Here, \isa{eclose} returns the (downward) $\in$-closure of its
+argument.
+
 %%% Local Variables:

### 2018-06-24 LSFA/names.tex
mi versión casi final de esto

@@
 \end{isabelle}
-
 Then we can obtain the recursive expression~(\ref{eq:def-val}) in the
@@
 
+We can finally define the generic extension of $M$ by $G$, also
+setting up the notation $M[G]$ for it:
+\begin{isabelle}
+\isacommand{definition}\isamarkupfalse%
+\isanewline
+\ \ GenExt\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}i{\isachardoublequoteclose}\ \ \ \ \ {\isacharparenleft}{\isachardoublequoteopen}M{\isacharbrackleft}{\isacharunderscore}{\isacharbrackright}{\isachardoublequoteclose}\ {\isadigit{9}}{\isadigit{0}}{\isacharparenright}\isanewline
+\ \ \isakeyword{where}\ {\isachardoublequoteopen}GenExt{\isacharparenleft}G{\isacharparenright}{\isacharequal}{\isacharequal}\ {\isacharbraceleft}val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}{\isachardot}\ {\isasymtau}\ {\isasymin}\ M{\isacharbraceright}{\isachardoublequoteclose}
+\end{isabelle}
+
 %% It is to be noted that this results holds under the assumption that
@@
 such that $\val(G,\chk(x))=x$. This will show that $M\sbq M[G]$, with
-a caveat we make explicit in the end of this section.
+a caveat we make explicit in the end of this section. As explained in
+the introduction, the fact that $M[G]$ extends $M$ is crucial to show
+that $\ZFC$ holds in the former.
 
@@
 Here, \isa{eclose} returns the (downward) $\in$-closure of its
-argument.
+argument. The main result is stated as follows:
+
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ valcheck\ {\isacharcolon}\ {\isachardoublequoteopen}y\ {\isasymin}\ M\ {\isasymLongrightarrow}\ one\ {\isasymin}\ G\ {\isasymLongrightarrow}\ {\isasymforall}x{\isasymin}M{\isachardot}\ check{\isacharparenleft}x{\isacharparenright}\ {\isasymin}\ M\ {\isasymLongrightarrow}\ val{\isacharparenleft}G{\isacharcomma}check{\isacharparenleft}y{\isacharparenright}{\isacharparenright}\ \ {\isacharequal}\ y{\isachardoublequoteclose}
+\end{isabelle}
+
+On the one hand, the only requirement on the set  \isatt{G} is that it 
+contains the top \isa{one} of the poset \isa{P}. On the other hand, it
+is necessary that all the relevant names are indeed in $M$. It
+requires a serious development  to fulfill this assumption. One of the
+hardest parts of Paulson's formalization of constructibility involves
+showing that models are closed under recursive construction. We will
+eventually formalize the that if $M\models\ZFC$ and the arguments of
+\isa{wfrec} are in $M$, then its value also is. This will require
+to adapt to models $M$ several locales defined in \cite{MR2507047}
+that were intended to be used for the class of constructible sets.
 

### 2018-06-24 LSFA/intro.tex
cambios menores

@@
 %% MIGUEL: la última oración me resulta un poco difícil de entender.
-The Rasiowa-Sikorski lemma states that, fixed a preorder $\PP$, for
+The Rasiowa-Sikorski lemma (RSL) states that, fixed a preorder $\PP$, for
 any countable family of dense subsets of $\PP$ there is a filter
@@
 therefore the extension $M[G]$ is proper. We formalize the proof of
-the Rasiowa-Sikorski lemma in
+RSL in
 Section~\ref{sec:rasiowa-sikorski-lemma}. A requisite result on a
 version of the Axiom of Choice is formalized in
-Section~\ref{sec:sequence-version-dc}.
+Section~\ref{sec:sequence-version-dc}. We then apply RSL to prove the existence of generic
+filters in Section~\ref{sec:generic-filter}.
 
@@
 that the Powerset axiom and the axiom schemes of Separation and
-Replacement hold in $M[G]$. In this work we tackle the Pairing
+Replacement hold in $M[G]$. In
+Section~\ref{sec:pairing-generic-extension}
+ we tackle the Pairing
 Axiom. This doesn't require the forcing relation, but provides an

### 2018-06-24 LSFA/rs.tex
cambios menores

@@
 allowing a parameter that allows to prove an equivalence with $\AC$) was
-formalized as part of the Isabelle/ZF library, along with many
+formalized by Krzysztof Grabczewski as part of the Isabelle/ZF
+library, along with many 
 statements that follow from or are equivalent to $\AC$. Since the

### 2018-06-24 LSFA/conclusion.tex
un poco

@@
-\section{Future work}
+\section{Conclusions and future work}
+There are several technical milestones that have to be traversed in the
+course of a formalization of the theory of forcing. The first and
+obvious one is the bulk of set- and meta-theoretical concepts needed to work
+with. This pushed us, in a sense,  into building on top of Isabelle/ZF,
+since we know of no other development in set theory of such
+depth. In this paper we worked on setting the stage for the work with
+generic extensions; in particular, this involves some purely mathematical
+results, as the Rasiowa-Sikorski lemma.
+
+Other milestones in this formalization project involve the definition
+of the forcing relation, proving the Fundamental Theorem of forcing
+(that relates truth in $M$ to that in $M[G]$), and using it to show
+that $M[G]\models \ZFC$. The theory is extremely modular in the sense
+that the last goal does not depend on the proof of the Fundamental
+Theorem nor on the definition of the forcing relation. In a next work,
+we will obtain the last goal.
+
 \begin{itemize}
+\item Write Isar versions of most proofs in the development.
 \item Develop an interface between Paulson's relativization results

### 2018-06-24 LSFA/isabelle-zf.tex
nueva versión del párrafo de locales

--- 2018-06-23 LSFA/isabelle-zf.tex
@@
 
+Mathematical texts usually start by fixing the context which defines
+parameters and assumptions necessary to develop theorems
+and results. In Isabelle the way of defining contexts is through
+\emph{locales}~\cite{ballarin2010tutorial}.
+Locales can be combined and extended adding more parameters and assuming
+more facts, leading to a new locale. For example a context describing
+lattices can be extended to distributive lattices.
+The way to instantiate a locale is \emph{interpreting} it, which consists
+of giving concrete values to parameters and proving the assumptions.
+In our work, we use locales to organize the formalization and to make
+explicit the assumptions of the most important results.
+
 % locales

### 2018-06-24 LSFA/conclusion.tex
[itemize -> prose] complete

@@
 
-Other milestones in this formalization project involve the definition
-of the forcing relation, proving the Fundamental Theorem of forcing
-(that relates truth in $M$ to that in $M[G]$), and using it to show
-that $M[G]\models \ZFC$. The theory is extremely modular in the sense
+Other milestones in this formalization project
+involve 
+\begin{enumerate}
+\item the definition
+  of the forcing relation, 
+\item proving the Fundamental Theorem of forcing
+  (that relates truth in $M$ to that in $M[G]$), and 
+\item using it to show
+  that $M[G]\models \ZFC$. 
+\end{enumerate}
+The theory is very modular and this is
+witnessed by the fact 
 that the last goal does not depend on the proof of the Fundamental
-Theorem nor on the definition of the forcing relation. In a next work,
-we will obtain the last goal.
+Theorem nor on the definition of the forcing relation. Our next task
+will be to obtain the last goal enumerated.
 
-\begin{itemize}
-\item Write Isar versions of most proofs in the development.
-\item Develop an interface between Paulson's relativization results
-  and countable models $M$ of $\ZFC$. This will show
+To this end, we will develop  an interface between Paulson's
+relativization results and countable models $M$ of $\ZFC$. This will show
 that every such $M$ is closed under well-founded recursion and, in
 particular, that contains names for each of its
-  elements. Consequently, $M\sbq M[G]$.
-\item Prove that the basic axioms of $\ZFC$ hold in the generic
-  extension. 
-\item After prototyping the forcing relation, prove the Axiom Scheme
+elements. Consequently, the proof of  $M\sbq M[G]$ will be
+complete. A landmark will be to prove the Axiom Scheme
 of Separation (the first that needs to use the machinery of forcing
-  nontrivially).
-\end{itemize}
+nontrivially). As a part of the new formalization, we will provide
+Isar versions of the longer applicative proofs presented in this work
 

### 2018-06-24 LSFA/isabelle-zf.tex
compatibilizo relativizacion con la intro

@@
 
-We are not discussing how, if possible at all, to express that some
-set satisfies some axiom. For example, we might well be interested in
-proving $\emptyset \models \mathit{foundation}$, but there is no
-sensible way to state this, without encoding formulas and axioms in
-ZFC.\footnote{Indeed Paulson has encoded formulas using a very general
-  method to encode recursive definitions. In our development, although
-  not reported in this paper, we use this internalization.}
+%% We are not discussing how, if possible at all, to express that some
+%% set satisfies some axiom. For example, we might well be interested in
+%% proving $\emptyset \models \mathit{foundation}$, but there is no
+%% sensible way to state this, without encoding formulas and axioms in
+%% ZFC.\footnote{Indeed Paulson has encoded formulas using a very general
+%%   method to encode recursive definitions. In our development, although
+%%   not reported in this paper, we use this internalization.}
+%% 
+%% \fbox{ You need this in order to formalize $\models$ as a predicate of
+%%   type $i\ent i \ent o$. (Don't know how to put that here.)
+%% }
 
-\fbox{ You need this in order to formalize $\models$ as a predicate of
-  type $i\ent i \ent o$. (Don't know how to put that here.)
-}
-
- In
-Isabelle/ZF there are, however, relativized versions of the
-axioms. This alternative statement of the axioms are relativized with
-respect to classes, which in Isabelle/ZF corresponds to predicates over
-sets. For example, the relativized axiom of foundation is:
+%% In
+%% Isabelle/ZF there are, however, relativized versions of the
+%% axioms. This alternative statement of the axioms are relativized with
+%% respect to classes, which in Isabelle/ZF corresponds to predicates over
+%% sets
+We now discuss relativization in Isabelle/ZF. Relativized versions of the
+ axioms can be found in the formalization of constructibility. For
+ example, the relativized Axiom of Foundation is
 \begin{isabelle}
@@
   P(x)} is a shorthand for \isa{{\isasymforall}x{\isachardot}\ M(x)
-  {\isasymlongrightarrow} P(x)}. We are one step further to be able to
-express that the empty set satisfies foundation; but we must turn the
-set \isa{0 :: i} into a predicate. Isabelle/ZF provides the coercion
-\isa{{\isacharhash}{\isacharhash} :: i ={\isachargreater} (i ={\isachargreater} o)}, thus we can write and prove:
+  {\isasymlongrightarrow} P(x)}. In order to express that a (set)
+satisfies this axiom we use  the coercion
+\isa{{\isacharhash}{\isacharhash} :: i ={\isachargreater} (i
+  ={\isachargreater} o)} provided by Isabelle/ZF. As a trivial example we
+can show that the empty set satisfies Foundation:
 \begin{isabelle}

### 2018-06-24 LSFA/intro.tex
typo

@@
   relative independence of an axiom $A$ 
-  can be use to obtain an algorithm transforming a proof
+  can be used to obtain an algorithm transforming a proof
   concluding a contradiction from $\ZFC+A$ to one from $\ZFC$.}
@@
 
-It can be shown elementarily that if $M$ and $N$ are transitive,
+It can be shown elementally that if $M$ and $N$ are transitive,
 $\phi^N$ holds if and only if $\phi^M$ holds,  for $x,y,z\in N$. We

### 2018-06-24 LSFA/intro.tex
otro typo

@@
   both the approaches by G\"odel and Cohen for showing
-  relative independence of an axiom $A$ 
+  relative consistency of an axiom $A$ 
   can be used to obtain an algorithm transforming a proof

### 2018-06-24 LSFA/isabelle-zf.tex
cambios

@@
 (akin to a \emph{Bool} type, but at the object level).
+
+\fbox{Change all this}
+
 As in usual presentations
-of ZF, Isabelle/ZF has only one relation symbol
+of ZF, Isabelle/ZF is has only one relation symbol
 \isatt{\ensuremath{\in}} (equality and inclusion are defined
-using membership); it also features one constant \isatt{\isadigit{0}}
+using membership);
+
+\fbox{End complaint}
+
+it also features one constant \isatt{\isadigit{0}}
 for the empty set and \isatt{inf} for an infinite set and some other
@@
 the formalization close to the customary manner of doing mathematics.
-Working with the library and extendig it is quite
+Working with the library and extending it is quite
 straightforward. 
@@
 We now discuss relativization in Isabelle/ZF. Relativized versions of the
- axioms can be found in the formalization of constructibility. For
+ axioms can be found in the formalization of constructibility \cite{paulson_2003}. For
  example, the relativized Axiom of Foundation is

### 2018-06-24 LSFA/intro.tex
cambios último momento

@@
 
-It can be shown elementally that if $M$ and $N$ are transitive,
+It can be shown elementarily that if $M$ and $N$ are transitive,
 $\phi^N$ holds if and only if $\phi^M$ holds,  for $x,y,z\in N$. We
@@
 The decoding is given by the function
-$\val$ which takes the generic filter $G$ as a parameter. To
+$\val$, which takes the generic filter $G$ as a parameter. To
 prove that $M$ is contained in $M[G]$ it suffices to give a name for

### 2018-06-24 LSFA/names.tex
cambios último momento

@@
 \isa{wfrec} are in $M$, then its value also is. This will require
-to adapt to models $M$ several locales defined in \cite{MR2507047}
+to adapt to models $M$ several locales defined in \cite{paulson_2003}
 that were intended to be used for the class of constructible sets.

### 2018-06-24 LSFA/isabelle-zf.tex
coercion
saco queja en cuadritos

@@
 (akin to a \emph{Bool} type, but at the object level). 
-
-\fbox{Change all this}
-
-As in usual presentations
-of ZF, Isabelle/ZF is has only one relation symbol
-\isatt{\ensuremath{\in}} (equality and inclusion are defined
-using membership);
-
-\fbox{End complaint}
-
-it also features one constant \isatt{\isadigit{0}}
-for the empty set and \isatt{inf} for an infinite set and some other
-set constructors, again terms of type \isatt{i}, to ease the
-developments. The axioms, being formulas, are terms of type \isatt{o};
-the foundation axiom, for example, is formalized as:
+The  axiomatization 
+of $\ZFC$ in Isabelle/ZF proceeds by postulating  a binary predicate
+\isatt{\ensuremath{\in}} and   several set
+constructors (terms and function with values in \isatt{i}) corresponding to the  empty set
+(the constant \isatt{\isadigit{0}}), powersets, and one further constant
+\isatt{inf} for an infinite set. The axioms, being formulas, are terms
+of type \isatt{o}; 
+the foundation axiom, for example, is formalized as (the universal
+closure of) 
 \isa{{\isachardoublequoteopen}A\ {\isacharequal}\ {\isadigit{0}}\
@@
   P(x)} is a shorthand for \isa{{\isasymforall}x{\isachardot}\ M(x)
-  {\isasymlongrightarrow} P(x)}. In order to express that a (set)
-satisfies this axiom we use  the coercion
+  {\isasymlongrightarrow} P(x)}. In order to express that a (set) model
+satisfies this axiom we use  the ``coercion''
 \isa{{\isacharhash}{\isacharhash} :: i ={\isachargreater} (i
-  ={\isachargreater} o)} provided by Isabelle/ZF. As a trivial example we
+  ={\isachargreater} o)} (that sends a set $A$ to the predicate
+$\lambda x . (x\in A)$) provided by Isabelle/ZF. As a trivial example we
 can show that the empty set satisfies Foundation:

### 2018-06-25 LSFA/rs.tex
corregido global poset->notion
eliminado (in countable_generic) y usando bloque
Comentario sobre def. recursiva

@@
 general, a version of the Axiom of Choice ($\AC$) must be proved. This
-version is known as the \emph{Principle of Dependent choices}:
+version is known as the \emph{Principle of Dependent Choices}:
 \begin{quote}
@@
 proofs there required a greater generality, it would have been very
-contrived to derived this ``pointed'' version of $\DC$ from that
+contrived to derive this ``pointed'' version of $\DC$ from that
 development. Instead, we preferred to give a direct proof based in the
@@
 \[
-\exists s: \P(A)\sm \{\mty\}\to A \, \forall X\sbq A.\ X\neq \mty \limp
+\exists (s: \P(A)\sm \{\mty\}\to A). \, \forall X\sbq A.\ X\neq \mty \limp
 s(X) \in X.
 \]
-This statement is more standard than the one deemed ``$\AC$'' in the
-Isabelle/ZF development, that involves a silent use of Replacement
+This statement is more standard than ``$\AC$'' as stated in the
+Isabelle/ZF development, which involves a silent use of Replacement
 since it is written as an axiom scheme.
 
-The strategy is to define $f$ above by primitive recursion.
+The strategy is to define $f$ above by primitive recursion on the
+natural, which can be done easily thanks to the package for
+definitions by recursion provided by Isabelle/ZF
+\cite{paulson1995set,paulson2000fixedpoint}.\footnote{The package
+  checks that the recursive definition makes sense; for example, it
+  rejects definitions with a missing case.}
 
@@
 \ \ wit{\isadigit{0}}\ \ \ {\isacharcolon}\ {\isachardoublequoteopen}dc{\isacharunderscore}witness{\isacharparenleft}{\isadigit{0}}{\isacharcomma}A{\isacharcomma}a{\isacharcomma}s{\isacharcomma}R{\isacharparenright}\ {\isacharequal}\ a{\isachardoublequoteclose}\isanewline
-\ \ witrec\ {\isacharcolon}{\isachardoublequoteopen}dc{\isacharunderscore}witness{\isacharparenleft}succ{\isacharparenleft}n{\isacharparenright}{\isacharcomma}A{\isacharcomma}a{\isacharcomma}s{\isacharcomma}R{\isacharparenright}\ {\isacharequal}\ s{\isacharbackquote}{\isacharbraceleft}x{\isasymin}A{\isachardot}\ {\isasymlangle}dc{\isacharunderscore}witness{\isacharparenleft}n{\isacharcomma}A{\isacharcomma}a{\isacharcomma}s{\isacharcomma}R{\isacharparenright}{\isacharcomma}x{\isasymrangle}{\isasymin}R\ {\isacharbraceright}{\isachardoublequoteclose}
+\ \ witrec\ {\isacharcolon}\ {\isachardoublequoteopen}dc{\isacharunderscore}witness{\isacharparenleft}succ{\isacharparenleft}n{\isacharparenright}{\isacharcomma}A{\isacharcomma}a{\isacharcomma}s{\isacharcomma}R{\isacharparenright}\ {\isacharequal}\ s{\isacharbackquote}{\isacharbraceleft}x{\isasymin}A{\isachardot}\ {\isasymlangle}dc{\isacharunderscore}witness{\isacharparenleft}n{\isacharcomma}A{\isacharcomma}a{\isacharcomma}s{\isacharcomma}R{\isacharparenright}{\isacharcomma}x{\isasymrangle}{\isasymin}R\ {\isacharbraceright}{\isachardoublequoteclose}
 \end{isabelle}
@@
 function $s$ as a parameter. If this function is a selector for
-$\P(A)\sm \{\mty\}$, the function $f\defi {}$\isatt{dc{\isacharunderscore}witness}$(A,a,s,R)$
+$\P(A)\sm \{\mty\}$, the function $f(n)\defi {}$\isatt{dc{\isacharunderscore}witness}$(n,A,a,s,R)$
 will satify $\DC$. Notice that $s$ is a term of type \isatt{i} (a
@@
 \isacommand{locale}\isamarkupfalse%
-\ countable{\isacharunderscore}generic\ {\isacharequal}\ forcing{\isacharunderscore}poset\ {\isacharplus}\isanewline
+\ countable{\isacharunderscore}generic\ {\isacharequal}\ forcing{\isacharunderscore}notion\ {\isacharplus}\isanewline
 \ \ \isakeyword{fixes}\ {\isasymD}\isanewline
@@
 \isacommand{definition}\isamarkupfalse%
-\ {\isacharparenleft}\isakeyword{in}\ countable{\isacharunderscore}generic{\isacharparenright}\isanewline
+
 \ \ D{\isacharunderscore}generic\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}o{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
@@
 \isacommand{theorem}\isamarkupfalse%
-\ {\isacharparenleft}\isakeyword{in}\ countable{\isacharunderscore}generic{\isacharparenright}\ rasiowa{\isacharunderscore}sikorski{\isacharcolon}\isanewline
+\ rasiowa{\isacharunderscore}sikorski{\isacharcolon}\isanewline
 \ \ {\isachardoublequoteopen}p{\isasymin}P\ {\isasymLongrightarrow}\ {\isasymexists}G{\isachardot}\ p{\isasymin}G\ {\isasymand}\ D{\isacharunderscore}generic{\isacharparenleft}G{\isacharparenright}{\isachardoublequoteclose}
@@
 \isacommand{lemma}\isamarkupfalse%
-\ {\isacharparenleft}\isakeyword{in}\ countable{\isacharunderscore}generic{\isacharparenright}\ RS{\isacharunderscore}relation{\isacharcolon}\isanewline
+\  RS{\isacharunderscore}relation{\isacharcolon}\isanewline
 \ \ \isakeyword{assumes}\isanewline

### 2018-06-25 LSFA/names.tex
Modificaciones menores en names.tex

--- 2018-06-24 LSFA/names.tex
@@
 %
-This definition proceeds by well-founded recursion on the relation
+As with any recursive definition, to justify that the definition makes
+sense (i.e., it defines anything at all) one has to explicit the
+well-founded relation which is used. In this case, we use the relation
 $\mathit{ed}$:
 \[
-x \mathrel{\mathit{ed}} y \iff \exists p : \lb x,p\rb\in y.
+x \mathrel{\mathit{ed}} y \iff \exists p . \lb x,p\rb\in y.
 \]
 Recall that in $\ZFC$, an ordered pair $\lb x,y \rb$ is the set
-$\{\{x\},\{x,y\}\}$. Hence this relation is well-founded since $\in$
-is by the Foundation Axiom.
+$\{\{x\},\{x,y\}\}$. It is trivial to deduce the well-foundedness of
+$\mathit{ed}$ from the fact that $\in$ is well-founded, which follows
+from the Foundation Axiom.
 
@@
 \ wf{\isacharunderscore}edrel\ {\isacharcolon}\ {\isachardoublequoteopen}wf{\isacharparenleft}edrel{\isacharparenleft}A{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
-%
-%
-\isacommand{apply}\isamarkupfalse%
-\ {\isacharparenleft}rule{\isacharunderscore}tac\ wf{\isacharunderscore}subset\ {\isacharbrackleft}of\ {\isachardoublequoteopen}trancl{\isacharparenleft}Memrel{\isacharparenleft}eclose{\isacharparenleft}A{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}{\isacharbrackright}{\isacharparenright}\isanewline
+\ \ \isacommand{apply}\isamarkupfalse%
+\ {\isacharparenleft}rule\ wf{\isacharunderscore}subset\ {\isacharbrackleft}of\ {\isachardoublequoteopen}trancl{\isacharparenleft}Memrel{\isacharparenleft}eclose{\isacharparenleft}A{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}{\isacharbrackright}{\isacharparenright}\isanewline
 \ \ \isacommand{apply}\isamarkupfalse%
@@
 
-We now turn to the definition of $\val$.
+\noindent We now turn to the definition of $\val$. In order to use
+\isa{wfrec} we need to define a functional \isa{Hv}.
 
@@
 \ \ val\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}i{\isasymRightarrow}i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
-\ \ {\isachardoublequoteopen}val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}\ {\isacharequal}{\isacharequal}\ wfrec{\isacharparenleft}edrel{\isacharparenleft}eclose{\isacharparenleft}M{\isacharparenright}{\isacharparenright}{\isacharcomma}\ {\isasymtau}\ {\isacharcomma}Hv{\isacharparenleft}G{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\ \ {\isachardoublequoteopen}val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}\ {\isacharequal}{\isacharequal}\ wfrec{\isacharparenleft}edrel{\isacharparenleft}eclose{\isacharparenleft}M{\isacharparenright}{\isacharparenright}{\isacharcomma}\ {\isasymtau}{\isacharcomma}\ Hv{\isacharparenleft}G{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
 \end{isabelle}
-
-Then we can obtain the recursive expression~(\ref{eq:def-val}) in the
+Then we can recover the recursive expression~(\ref{eq:def-val}) thanks to the
 following lemma:
@@
 \isacommand{lemma}\isamarkupfalse%
-\ def{\isacharunderscore}val{\isacharcolon}\ \ {\isachardoublequoteopen}x{\isasymin}M\ {\isasymLongrightarrow}\ val{\isacharparenleft}G{\isacharcomma}x{\isacharparenright}\ {\isacharequal}\ {\isacharbraceleft}val{\isacharparenleft}G{\isacharcomma}t{\isacharparenright}\ {\isachardot}{\isachardot}\ t{\isasymin}domain{\isacharparenleft}x{\isacharparenright}\ {\isacharcomma}\ {\isasymexists}p{\isasymin}P\ {\isachardot}\ \ {\isasymlangle}t{\isacharcomma}\ p{\isasymrangle}{\isasymin}x\ {\isasymand}\ p\ {\isasymin}\ G\ {\isacharbraceright}{\isachardoublequoteclose}
+\ def{\isacharunderscore}val{\isacharcolon}
+\isanewline
+\ \ \ \ \ \ \ \  {\isachardoublequoteopen}x{\isasymin}M\ {\isasymLongrightarrow}\ val{\isacharparenleft}G{\isacharcomma}x{\isacharparenright}\ {\isacharequal}\ {\isacharbraceleft}val{\isacharparenleft}G{\isacharcomma}t{\isacharparenright}\ {\isachardot}{\isachardot}\ t{\isasymin}domain{\isacharparenleft}x{\isacharparenright}\ {\isacharcomma}\ {\isasymexists}p{\isasymin}P\ {\isachardot}\ \ {\isasymlangle}t{\isacharcomma}\ p{\isasymrangle}{\isasymin}x\ {\isasymand}\ p\ {\isasymin}\ G\ {\isacharbraceright}{\isachardoublequoteclose}
+\end{isabelle}
+
+We can finally define the generic extension of $M$ by $G$, also
+setting up the notation $M[G]$ for it:
+\begin{isabelle}
+\isacommand{definition}\isamarkupfalse%
+\isanewline
+\ \ GenExt\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}i{\isachardoublequoteclose}\ \ \ \ \ {\isacharparenleft}{\isachardoublequoteopen}M{\isacharbrackleft}{\isacharunderscore}{\isacharbrackright}{\isachardoublequoteclose}\ {\isadigit{9}}{\isadigit{0}}{\isacharparenright}\ \isakeyword{where} \isanewline
+\ \ {\isachardoublequoteopen}GenExt{\isacharparenleft}G{\isacharparenright}{\isacharequal}{\isacharequal}\ {\isacharbraceleft}val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}{\isachardot}\ {\isasymtau}\ {\isasymin}\ M{\isacharbraceright}{\isachardoublequoteclose}
 \end{isabelle}
@@
 such that $\val(G,\chk(x))=x$. This will show that $M\sbq M[G]$, with
-a caveat we make explicit in the end of this section.
-
+a caveat we make explicit in the end of this section. As explained in
+the introduction, the fact that $M[G]$ extends $M$ is crucial to show
+that $\ZFC$ holds in the former.
+%
 The definition of $\chk(x)$ is a straightforward $\in$-recursion:
@@
 Here, \isa{eclose} returns the (downward) $\in$-closure of its
-argument.
+argument. The main result is stated as follows:
+
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ valcheck\ {\isacharcolon}\ {\isachardoublequoteopen}y\ {\isasymin}\ M\ {\isasymLongrightarrow}\ one\ {\isasymin}\ G\ {\isasymLongrightarrow}\ {\isasymforall}x{\isasymin}M{\isachardot}\ check{\isacharparenleft}x{\isacharparenright}\ {\isasymin}\ M\ {\isasymLongrightarrow}\ val{\isacharparenleft}G{\isacharcomma}check{\isacharparenleft}y{\isacharparenright}{\isacharparenright}\ \ {\isacharequal}\ y{\isachardoublequoteclose}
+\end{isabelle}
+
+On the one hand, the only requirement on the set \isa{G} is that it
+contains the top element of the poset \isa{P}. On the other hand, it
+is necessary that all the relevant names are indeed in $M$ (i.e.,
+$\chk(x) \in M$ if $x\in M$). It requires a serious development to
+fulfill this assumption. One of the hardest parts of Paulson's
+formalization of constructibility involves showing that models are
+closed under recursive construction. We will eventually formalize that
+if $M\models\ZFC$ and the arguments of \isa{wfrec} are in $M$, then
+its value also is. This will require to adapt to ctm models several
+locales defined in \cite{paulson_2003} that were intended to be used
+for the class of constructible sets.
 

### 2018-06-25 LSFA/names.tex
Tratando de ser más preciso.

@@
 The formalization of $\val$ in Isabelle/ZF involves the function
-\isa{wfrec}, which takes as an argument a well-founded relation of
-type \isa{i}, i.e.\ a set of pairs. Therefore we restrict
-$\mathit{ed}$ above to a set.
+\isa{wfrec},\footnote{Notice that this form of recursive
+  definitions is more general than the one used in the previous
+  section to define \isa{dc{\isacharunderscore}witness}.} which takes as an argument a
+well-founded relation of type \isa{i}, i.e.\ a set of pairs. Therefore
+we restrict $\mathit{ed}$ above to a set.
 %

### 2018-06-25 LSFA/rs.tex
Tratando de ser más preciso.

@@
 The strategy is to define $f$ above by primitive recursion on the
-natural, which can be done easily thanks to the package for
-definitions by recursion provided by Isabelle/ZF
-\cite{paulson1995set,paulson2000fixedpoint}.\footnote{The package
-  checks that the recursive definition makes sense; for example, it
-  rejects definitions with a missing case.}
+naturals, which can be done easily thanks to the package of
+Isabelle/ZF \cite{paulson1995set,paulson2000fixedpoint} for
+definitions by recursion on inductively defined sets.\footnote{The
+  package figures out the inductive set at hand and checks that the
+  recursive definition makes sense; for example, it rejects
+  definitions with a missing case.}
 
@@
 \isacommand{definition}\isamarkupfalse%
-
 \ \ D{\isacharunderscore}generic\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}o{\isachardoublequoteclose}\ \isakeyword{where}\isanewline

### 2018-06-25 LSFA/first_steps_into_forcing.tex
cambio de orden de secciones
saco emph
última leída.

@@
   preorders with top, dense subsets, and generic filters. We formalize
-  a version of the principle of \emph{Dependent Choices} and using it
-  we prove the Rasiowa-Sikorski lemma on existence of generic filters.
+  a version of the principle of Dependent Choices and using it
+  we prove the Rasiowa-Sikorski lemma on the existence of generic filters.
 
-  Given a transitive set $M$, we define the generic extension $M[G]$,
-  the canonical names for elements of $M$ and finally show that if $M$
-  satisfy the axiom of pairing, the $M[G]$ also does.
+  Given a transitive set $M$, we define its generic extension $M[G]$,
+  the canonical names for elements of $M$, and finally show that if $M$
+  satisfy the axiom of pairing, then $M[G]$ also does.
 \end{abstract}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\input{gen-ext}
+
+%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%--%-%-%-%-%-%-%-%
 \input{names}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\input{gen-ext}
 

### 2018-06-25 LSFA/isabelle-zf.tex
última leída.

@@
 %% writing and a nicer reading, so we use
-We are then able to add an abbreviation resembling the usual
-mathematical writing by setting 
-\isa{{\isacharbraceleft}b\
+We are then able to add the abbreviation \isa{{\isacharbraceleft}b\
   {\isachardot}{\isachardot}\ x{\isasymin}A{\isacharcomma}\
@@
 \isa{{\isacharhash}{\isacharhash} :: i ={\isachargreater} (i
-  ={\isachargreater} o)} (that sends a set $A$ to the predicate
+  ={\isachargreater} o)} (that maps a set $A$ to the predicate
 $\lambda x . (x\in A)$) provided by Isabelle/ZF. As a trivial example we
@@
 lattices can be extended to distributive lattices.
-The way to instantiate a locale is \emph{interpreting} it, which consists
+The way to instantiate a locale is by \emph{interpreting} it, which consists
 of giving concrete values to parameters and proving the assumptions.

### 2018-06-25 LSFA/intro.tex
sacándolo a Pagano del camino

@@
 A great part of G\"odel's work on this subject has been formalized in
-Isabelle by Lawrence Paulson. This paper formalizes a first part of the
-machinery of forcing. In the rest of the introduction we discuss some
-of the set-theoretical details involved and explain briefly Paulson's
-formalization.
+Isabelle~\cite{Isabelle} by Lawrence Paulson~\cite{paulson_2003}. This paper formalizes
+a first part of the machinery of forcing. In the rest of the
+introduction we discuss some of the set-theoretical details involved
+and explain briefly Paulson's formalization.
 
@@
 %
-%% MIGUEL: controlar que tiene sentido lo que menciono en la nota al pie!
-%% MIGUEL: Me da la sensación que podemos promover lo de la nota al pie
-%% al cuerpo y luego explicar que por eso es importante el teorema.
-This follows by a combination of the Reflection Principle, the
+\noindent This follows by a combination of the Reflection Principle, the
 L\"owenheim-Skolem Theorem, and the Mostowksi Collapse. The reader can
@@
 
-%% MIGUEL: Qué significa "portrayal convenient"?
-%% MIGUEL: ¿por qué esto es conveniente para esta introducción?
-Assume that $M$ is a ctm of $\ZFC$. In a portrayal convenient for this
-introduction, G\"odel showed that $M$ contains a minimal submodel
-$L^M$ of the same ``height'' (i.e.\ having the same ordinals)
-that satisfies $\ZFC+\CH$. The sets in $L^M$ are called
-\emph{constructible} and are in a sense ``definable.'' In fact, there
-is a first-order formula $L$ such that $L^M = \{x\in M : M\models
-L(x)\}$. To show that  $L^M\models \ZFC+\CH$, one uses that the
-$\ZFC+\CH$ holds in $M$.
+Assume that $M$ is a ctm of $\ZFC$. It is instructive to sketch
+G\"odel's argument of the relative consistency of $\CH$: he showed
+that $M$ contains a minimal submodel $L^M$ of the same ``height''
+(i.e.\ having the same ordinals) that satisfies $\ZFC+\CH$. The sets
+in $L^M$ are called \emph{constructible} and are in a sense
+``definable.'' In fact, there is a first-order formula $L$ such that
+$L^M = \{x\in M : M\models L(x)\}$. To show that
+$L^M\models \ZFC+\CH$, one uses that the $\ZFC+\CH$ holds in $M$.
 
@@
 \[
-\phi(x,y,z)\defi \forall w \,( w\in z \lsii w=x \o w=y).
+\phi(x,y,z)\defi \forall w.\,( w\in z \lsii w=x \o w=y).
 \]
@@
 axioms while allowing to fine-tune what other first order properties the
-extension will have. Given a ctm $M$  of $\ZFC$ and a set $G$, a new
+extension will have. Given a ctm $M$  of $\ZFC$ and a set $G$, one constructs a new
 ctm  $M[G]$  that includes $M$ and
-contains $G$ is constructed, and under some hypotheses ($G$ being ``generic''),
+contains $G$, and proves that under some hypotheses ($G$ being ``generic''),
 $M[G]$ satisfies $\ZFC$.
@@
  we tackle the Pairing
-Axiom. This doesn't require the forcing relation, but provides an
-illustration of the use of names.
+Axiom. This does not require the forcing relation, but provides an
+illustration of the use of names. The development can be downloaded
+from \url{https://cs.famaf.unc.edu.ar/~pedro/forcing/}.
 

### 2018-06-25 LSFA/intro.tex
volvemos por laburo de Manu
Modificación mínima

@@
 
-Assume that $M$ is a ctm of $\ZFC$. It is instructive to sketch
-G\"odel's argument of the relative consistency of $\CH$: he showed
-that $M$ contains a minimal submodel $L^M$ of the same ``height''
-(i.e.\ having the same ordinals) that satisfies $\ZFC+\CH$. The sets
-in $L^M$ are called \emph{constructible} and are in a sense
-``definable.'' In fact, there is a first-order formula $L$ such that
-$L^M = \{x\in M : M\models L(x)\}$. To show that
+It is instructive to sketch G\"odel's argument of the relative
+consistency of $\CH$: Assume that $M$ is a ctm of $\ZFC$, G\"odel
+showed that $M$ contains a minimal submodel $L^M$ of the same
+``height'' (i.e.\ having the same ordinals) that satisfies
+$\ZFC+\CH$. The sets in $L^M$ are called \emph{constructible} and are
+in a sense ``definable.'' In fact, there is a first-order formula $L$
+such that $L^M = \{x\in M : M\models L(x)\}$. To show that
 $L^M\models \ZFC+\CH$, one uses that the $\ZFC+\CH$ holds in $M$.
@@
 illustration of the use of names. The development can be downloaded
-from \url{https://cs.famaf.unc.edu.ar/~pedro/forcing/}.
+from \url{https://cs.famaf.unc.edu.ar/~mpagano/forcing/}.
 

### 2018-07-12 LSFA/arXiv/first_steps_into_forcing.tex
versión para arXiv

--- 2018-06-25 LSFA/first_steps_into_forcing.tex
@@
-\documentclass[9pt]{entcs} \usepackage{entcsmacro}
-\usepackage{graphicx}
+\documentclass{article}
 \usepackage{isabelle,isabellesym}
-\sloppy
+\usepackage{hyperref}
 
 \input{header-draft}
-\def\lastname{Gunther, Pagano, S\'anchez Terraf}
+
 \begin{document}
-\begin{frontmatter}
 \title{First steps towards a formalization of Forcing}
-  \author{Emmanuel Gunther%\thanksref{ALL}
-    \thanksref{myemail}}
-  \address{FaMAF\\ Universidad Nacional de C\'ordoba\\
-    C\'ordoba, Argentina} \author{Miguel Pagano\thanksref{coemail}}
-  \address{FaMAF\\Universidad Nacional de C\'ordoba\\
-    C\'ordoba, Argentina}
-  \author{Pedro S\'anchez Terraf\thanksref{co2email}}
-  \address{CIEM-FaMAF\\Universidad Nacional de C\'ordoba\\
-    C\'ordoba, Argentina}
- \thanks[myemail]{Email:
-    \href{mailto:gunther@famaf.unc.edu.ar} {\texttt{\normalshape
-        gunther@famaf.unc.edu.ar}}} \thanks[coemail]{Email:
-    \href{mailto:pagano@famaf.unc.edu.ar} {\texttt{\normalshape
-        pagano@famaf.unc.edu.ar}}}  \thanks[co2email]{Email:
-    \href{mailto:sterraf@famaf.unc.edu.ar} {\texttt{\normalshape
-        sterraf@famaf.unc.edu.ar}}} 
- \thanks[ALL]{Supported by Secyt-UNC project 33620180100465CB.} 
+\author{Emmanuel Gunther
+  \and 
+  Miguel Pagano
+  \and 
+  Pedro S\'anchez Terraf}
+\maketitle
+
 \begin{abstract} 
@@
 \end{abstract}
-\begin{keyword}
+\keywords{
 Isabelle/ZF, forcing, preorder, Rasiowa-Sikorski lemma, names, generic extension.
-\end{keyword}
-\end{frontmatter}
+}
 
@@
 
-\bibliographystyle{entcs}
-\bibliography{citados}
+\bibliographystyle{plain}
+\bibliography{draft}
 \end{document}

### 2018-07-12 LSFA/arXiv/gen-ext.tex
versión para arXiv

\section{The generic extension}
In the following subsections we show that every preorder in a ctm
admits a generic filter, and develop the machinery of names.

\subsection{The generic filter}
\label{sec:generic-filter}
The following locale gathers the data needed to ensure the 
existence of an $M$-generic filter for a poset \isa{P}.

\begin{isabelle}
\isacommand{locale}\isamarkupfalse%
\ forcing{\isacharunderscore}data\ {\isacharequal}\ forcing{\isacharunderscore}notion\ {\isacharplus}\isanewline
\ \ \isakeyword{fixes}\ M\ enum\isanewline
\ \ \isakeyword{assumes}\ M{\isacharunderscore}countable{\isacharcolon} \ \ {\isachardoublequoteopen}enum{\isasymin}bij{\isacharparenleft}nat{\isacharcomma}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
 \ \ \isakeyword{and}\ P{\isacharunderscore}in{\isacharunderscore}M{\isacharcolon}  \ \ \ {\isachardoublequoteopen}P\ {\isasymin}\ M{\isachardoublequoteclose}\isanewline
 \ \ \isakeyword{and}\ leq{\isacharunderscore}in{\isacharunderscore}M{\isacharcolon}  \ {\isachardoublequoteopen}leq\ {\isasymin}\ M{\isachardoublequoteclose}\isanewline
 \ \ \isakeyword{and}\ trans{\isacharunderscore}M{\isacharcolon}  \ \ {\isachardoublequoteopen}Transset{\isacharparenleft}M{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}

An immediate consequence of the Rasiowa-Sikorski Lemma is the
existence of an $M$-generic filter for a poset \isa{P}.

\begin{isabelle}

\isacommand{lemma}\isamarkupfalse%
\ generic{\isacharunderscore}filter{\isacharunderscore}existence{\isacharcolon}\ \isanewline
\ \ {\isachardoublequoteopen}p{\isasymin}P\ {\isasymLongrightarrow}\ {\isasymexists}G{\isachardot}\ p{\isasymin}G\ {\isasymand}\ M{\isacharunderscore}generic{\isacharparenleft}G{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}

\noindent By defining an appropriate countable sequence of dense subsets of \isa{P},
\begin{isabelle}

\ \ \isacommand{let}\isamarkupfalse%
\isanewline
   \ \ {\isacharquery}D{\isacharequal}{\isachardoublequoteopen}{\isasymlambda}n{\isasymin}nat{\isachardot}\ {\isacharparenleft}if\ {\isacharparenleft}enum{\isacharbackquote}n{\isasymsubseteq}P\ {\isasymand}\ dense{\isacharparenleft}enum{\isacharbackquote}n{\isacharparenright}{\isacharparenright}\ \ then\ enum{\isacharbackquote}n\ else\ P{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
\noindent we can instantiate the locale \isatt{countable{\isacharunderscore}generic}

\begin{isabelle}

\ \ \isacommand{have}\isamarkupfalse%
\ \isanewline
  \ Eq{\isadigit{2}}{\isacharcolon}\ {\isachardoublequoteopen}{\isasymforall}n{\isasymin}nat{\isachardot}\ {\isacharquery}D{\isacharbackquote}n\ {\isasymin}\ Pow{\isacharparenleft}P{\isacharparenright}{\isachardoublequoteclose}\isanewline
 \isacommand{by}\isamarkupfalse%
\ auto\isanewline
\ \ \isacommand{then}\isamarkupfalse%
\ \isacommand{have}\isamarkupfalse%
\isanewline
  \ Eq{\isadigit{3}}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharquery}D{\isacharcolon}nat{\isasymrightarrow}Pow{\isacharparenleft}P{\isacharparenright}{\isachardoublequoteclose}\isanewline
 \isacommand{by}\isamarkupfalse%
\ {\isacharparenleft}rule\ lam{\isacharunderscore}codomain{\isacharparenright}\isanewline
\ \ \isacommand{have}\isamarkupfalse%
\isanewline
  \ Eq{\isadigit{4}}{\isacharcolon}\ {\isachardoublequoteopen}{\isasymforall}n{\isasymin}nat{\isachardot}\ dense{\isacharparenleft}{\isacharquery}D{\isacharbackquote}n{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
\dots
\begin{isabelle}

\ \ \isacommand{from}\isamarkupfalse%
\ Eq{\isadigit{3}}\ \isakeyword{and}\ Eq{\isadigit{4}}\ \isacommand{interpret}\isamarkupfalse%
\ \isanewline
  \ \ cg{\isacharcolon}\ countable{\isacharunderscore}generic\ P\ leq\ uno\ {\isacharquery}D\ \isanewline
 \isacommand{by}\isamarkupfalse%
\ {\isacharparenleft}unfold{\isacharunderscore}locales{\isacharcomma}\ auto{\isacharparenright}
\end{isabelle}
\noindent
and then a  $\calD$-generic filter given by Rasiowa-Sikorski will be $M$-generic by construction.

\begin{isabelle}

\ \ \isacommand{from}\isamarkupfalse%
\ cg{\isachardot}rasiowa{\isacharunderscore}sikorski\ \isakeyword{and}\ Eq{\isadigit{1}}\ \isacommand{obtain}\isamarkupfalse%
\ G\ \isakeyword{where}\ \isanewline
  \ Eq{\isadigit{6}}{\isacharcolon}\ {\isachardoublequoteopen}p{\isasymin}G\ {\isasymand}\ filter{\isacharparenleft}G{\isacharparenright}\ {\isasymand}\ {\isacharparenleft}{\isasymforall}n{\isasymin}nat{\isachardot}{\isacharparenleft}{\isacharquery}D{\isacharbackquote}n{\isacharparenright}{\isasyminter}G{\isasymnoteq}{\isadigit{0}}{\isacharparenright}{\isachardoublequoteclose}\isanewline
 \isacommand{unfolding}\isamarkupfalse%
\ cg{\isachardot}D{\isacharunderscore}generic{\isacharunderscore}def\ \isacommand{by}\isamarkupfalse%
\ blast\isanewline
\ \ \isacommand{then}\isamarkupfalse%
\ \isacommand{have}\isamarkupfalse%
\isanewline
  \ Eq{\isadigit{7}}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharparenleft}{\isasymforall}D{\isasymin}M{\isachardot}\ D{\isasymsubseteq}P\ {\isasymand}\ dense{\isacharparenleft}D{\isacharparenright}{\isasymlongrightarrow}D{\isasyminter}G{\isasymnoteq}{\isadigit{0}}{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}

\noindent We omit the rest of this Isar proof.

%%% Local Variables:
%%% mode: latex
%%% ispell-local-dictionary: "american"
%%% TeX-master: "first_steps_into_forcing"
%%% End:

### 2018-07-12 LSFA/arXiv/pair-gen-ext.tex
versión para arXiv

\section{Pairing in the generic extension}
\label{sec:pairing-generic-extension}
Given a predicate
\isa{C\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}o{\isachardoublequoteclose}},
we can consider the \emph{class} of all sets \isa{x} such that
\isa{C(x)}. In \cite{paulson_2003} a large corpus of set-theoretical
concepts are expressed in relative form.

In the case of the Pairing Axiom the following definition in 
Paulson's formalization captures
what a class \isa{C} believes about a set being the unordered pair of
other two:
\begin{isabelle}
\isacommand{definition}\isamarkupfalse%
\isanewline
\ \ upair\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharbrackleft}i{\isasymRightarrow}o{\isacharcomma}i{\isacharcomma}i{\isacharcomma}i{\isacharbrackright}\ {\isasymRightarrow}\ o{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
 {\isachardoublequoteopen}upair{\isacharparenleft}C{\isacharcomma}a{\isacharcomma}b{\isacharcomma}z{\isacharparenright}\ {\isacharequal}{\isacharequal}\ a\ {\isasymin}\ z\ {\isasymand}\ b\ {\isasymin}\ z\ {\isasymand}\ {\isacharparenleft}{\isasymforall}x{\isacharbrackleft}C{\isacharbrackright}{\isachardot}\ x{\isasymin}z\ {\isasymlongrightarrow}\ x\ {\isacharequal}\ a\ {\isasymor}\ x\ {\isacharequal}\ b{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
%
Then the fact that Pairing axiom holds in \isa{C} is stated as follows.
%
\begin{isabelle}
\isacommand{definition}\isamarkupfalse%
\isanewline
\ \ upair{\isacharunderscore}ax\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharparenleft}i{\isasymRightarrow}o{\isacharparenright}\ {\isasymRightarrow}\ o{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
 {\isachardoublequoteopen}upair{\isacharunderscore}ax{\isacharparenleft}C{\isacharparenright}\ {\isacharequal}{\isacharequal}\ {\isasymforall}x{\isacharbrackleft}C{\isacharbrackright}{\isachardot}\ {\isasymforall}y{\isacharbrackleft}C{\isacharbrackright}{\isachardot}\ {\isasymexists}z{\isacharbrackleft}C{\isacharbrackright}{\isachardot}\ upair{\isacharparenleft}C{\isacharcomma}x{\isacharcomma}y{\isacharcomma}z{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}

In order use the \isa{upair{\isacharunderscore}ax} for a set model, we
use the \isatt{\#\#}  operator.  We state the main result of this
Section in the context \isatt{forcing{\isacharunderscore}data}.  
%
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ pair{\isacharunderscore}preserv\ {\isacharcolon}\ \isanewline
\ \ {\isachardoublequoteopen}one\ {\isasymin}\ G\ {\isasymLongrightarrow}\ upair{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}\ {\isasymLongrightarrow}\ upair{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharbrackleft}G{\isacharbrackright}{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}

Let $x$ and $y$ be elements in $M[G]$. By definition of the generic extension, there exist
elements $\tau$ and $\rho$ in $M$ such that $x = \val(G,\tau)$ and
$y = \val(G,\rho)$.  We need to find an element in $M[G]$ that contains exactly
these elements; for that we should construct a name $\sigma\in M$ such that
$\val(G,\sigma) = \{ \val(G,\tau) , \val(G,\rho) \}$.

The candidate, motivated by the definition of $\chk$,  is
$\sigma = \{\langle \tau , \mathrm{one} \rangle , \langle \rho , \mathrm{one} \rangle \}$. 
Our remaining tasks are to show 
\begin{enumerate}
  \item \label{item:1}$\sigma \in M$, and
  \item \label{item:2} $\val(G,\sigma) = \{ \val(G,\tau) , \val(G,\rho) \}$
\end{enumerate}

By the implementation of pairs  in $\ZFC$, showing (\ref{item:1})
involves using that the
pairing axiom holds in $M$ and the absoluteness of pairing
thanks to $M$ being transitive.

\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ pairs{\isacharunderscore}in{\isacharunderscore}M\ {\isacharcolon}\ \isanewline
\ \ {\isachardoublequoteopen}\ {\isasymlbrakk}\ upair{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}\ {\isacharsemicolon}\ a\ {\isasymin}\ M\ {\isacharsemicolon}\ b\ {\isasymin}\ M\ {\isacharsemicolon}\ c\ {\isasymin}\ M\ {\isacharsemicolon}\ d\ {\isasymin}\ M\ {\isasymrbrakk}\ {\isasymLongrightarrow}\ {\isacharbraceleft}{\isasymlangle}a{\isacharcomma}c{\isasymrangle}{\isacharcomma}{\isasymlangle}b{\isacharcomma}d{\isasymrangle}{\isacharbraceright}\ {\isasymin}\ M{\isachardoublequoteclose}
\end{isabelle}

Item (\ref{item:1}) then follows because \isa{\isasymtau}, \isa{\isasymrho} and
\isa{one} belong to \isa{M} (the last fact holds because \isa{one\isasymin P}, \isa{P\isasymin M} and
\isa{M} is transitive).

\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ sigma{\isacharunderscore}in{\isacharunderscore}M\ {\isacharcolon}\isanewline
\ \ {\isachardoublequoteopen}upair{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}\ {\isasymLongrightarrow}\ one\ {\isasymin}\ G\ {\isasymLongrightarrow}\ {\isasymtau}\ {\isasymin}\ M\ {\isasymLongrightarrow}\ {\isasymrho}\ {\isasymin}\ M\ {\isasymLongrightarrow}\isanewline \ \  \ {\isacharbraceleft}{\isasymlangle}{\isasymtau}{\isacharcomma}one{\isasymrangle}{\isacharcomma}{\isasymlangle}{\isasymrho}{\isacharcomma}one{\isasymrangle}{\isacharbraceright}\ {\isasymin}\ M{\isachardoublequoteclose}

\isacommand{by}\isamarkupfalse%
\ {\isacharparenleft}rule\ pairs{\isacharunderscore}in{\isacharunderscore}M{\isacharcomma}simp{\isacharunderscore}all\ add{\isacharcolon}\ upair{\isacharunderscore}ax{\isacharunderscore}def\ one{\isacharunderscore}in{\isacharunderscore}M{\isacharparenright}%
\end{isabelle}

Under the assumption that \isa{one} belongs to the set \isa{G},
(\ref{item:2}) follows from \isa{def\_val} almost automatically:

\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ valsigma\ {\isacharcolon}\isanewline
\ \ {\isachardoublequoteopen}one\ {\isasymin}\ G\ {\isasymLongrightarrow}\ {\isacharbraceleft}{\isasymlangle}{\isasymtau}{\isacharcomma}one{\isasymrangle}{\isacharcomma}{\isasymlangle}{\isasymrho}{\isacharcomma}one{\isasymrangle}{\isacharbraceright}\ {\isasymin}\ M\ {\isasymLongrightarrow}\isanewline
\ \ \ val{\isacharparenleft}G{\isacharcomma}{\isacharbraceleft}{\isasymlangle}{\isasymtau}{\isacharcomma}one{\isasymrangle}{\isacharcomma}{\isasymlangle}{\isasymrho}{\isacharcomma}one{\isasymrangle}{\isacharbraceright}{\isacharparenright}\ {\isacharequal}\ {\isacharbraceleft}val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}{\isacharcomma}val{\isacharparenleft}G{\isacharcomma}{\isasymrho}{\isacharparenright}{\isacharbraceright}{\isachardoublequoteclose}
\end{isabelle}

This ends the proof of the Pairing Axiom in $M[G]$.

### 2018-07-12 LSFA/arXiv/rs.tex
versión para arXiv

--- 2018-06-25 LSFA/rs.tex
@@
   ($\DC$) Let $\rr$ be a binary relation on $A$, and $a\in A$. If
-  $\forall x\in A\,  \exists y\in A\, x\rr y$, then there exists
+  $\forall x\in A.\,  \exists y\in A.\, x\rr y$, then there exists
   $f:\om\to A$ such that $f(0)=a$ and $f(n)\rr f(n+1)$ for all
@@
 A different version of $\DC$ (without the constraint $f(0)=a$, but
-allowing a parameter that allows to prove an equivalence with $\AC$) was
-formalized by Krzysztof Grabczewski as part of the Isabelle/ZF
-library, along with many 
-statements that follow from or are equivalent to $\AC$. Since the
-proofs there required a greater generality, it would have been very
-contrived to derive this ``pointed'' version of $\DC$ from that
-development. Instead, we preferred to give a direct proof based in the
-one that appears in Moschovakis \cite{moschovakis1994notes}. Another
-advantage is that this proof depends on the following version of $\AC$
-(existence of choice functions on $\P(A)\sm \{\mty\}$):
+allowing a parameter that permits to prove an equivalence with $\AC$)
+was formalized by Krzysztof
+Grabczewski~\cite{DBLP:journals/jar/PaulsonG96} as part of the
+Isabelle/ZF library, along with many statements that follow from or
+are equivalent to $\AC$. Since the proofs there required a greater
+generality, it would have been very contrived to derive this
+``pointed'' version of $\DC$ from that development. Instead, we
+preferred to give a direct proof based on the one in
+Moschovakis' notes \cite{moschovakis1994notes}. Another advantage is that
+this proof depends on the following version of $\AC$ (existence of
+choice functions on $\P(A)\sm \{\mty\}$):
 \[
@@
 \]
-This statement is more standard than ``$\AC$'' as stated in the
+This statement is more standard than $\AC$ as stated in the
 Isabelle/ZF development, which involves a silent use of Replacement
@@
 
-The strategy is to define $f$ above by primitive recursion on the
-natural, which can be done easily thanks to the package for
-definitions by recursion provided by Isabelle/ZF
-\cite{paulson1995set,paulson2000fixedpoint}.\footnote{The package
-  checks that the recursive definition makes sense; for example, it
-  rejects definitions with a missing case.}
+The strategy is to define the function $f$ discussed above by
+primitive recursion on the naturals, which can be done easily thanks
+to the package of Isabelle/ZF
+\cite{paulson1995set,paulson2000fixedpoint} for definitions by
+recursion on inductively defined sets.\footnote{The package figures
+  out the inductive set at hand and checks that the recursive
+  definition makes sense; for example, it rejects definitions with a
+  missing case.}
 
@@
 \ \ wit{\isadigit{0}}\ \ \ {\isacharcolon}\ {\isachardoublequoteopen}dc{\isacharunderscore}witness{\isacharparenleft}{\isadigit{0}}{\isacharcomma}A{\isacharcomma}a{\isacharcomma}s{\isacharcomma}R{\isacharparenright}\ {\isacharequal}\ a{\isachardoublequoteclose}\isanewline
-\ \ witrec\ {\isacharcolon}\ {\isachardoublequoteopen}dc{\isacharunderscore}witness{\isacharparenleft}succ{\isacharparenleft}n{\isacharparenright}{\isacharcomma}A{\isacharcomma}a{\isacharcomma}s{\isacharcomma}R{\isacharparenright}\ {\isacharequal}\ s{\isacharbackquote}{\isacharbraceleft}x{\isasymin}A{\isachardot}\ {\isasymlangle}dc{\isacharunderscore}witness{\isacharparenleft}n{\isacharcomma}A{\isacharcomma}a{\isacharcomma}s{\isacharcomma}R{\isacharparenright}{\isacharcomma}x{\isasymrangle}{\isasymin}R\ {\isacharbraceright}{\isachardoublequoteclose}
+\ \ witrec\ {\isacharcolon}\ {\isachardoublequoteopen}dc{\isacharunderscore}witness{\isacharparenleft}succ{\isacharparenleft}n{\isacharparenright}{\isacharcomma}A{\isacharcomma}a{\isacharcomma}s{\isacharcomma}R{\isacharparenright}\ {\isacharequal}\isanewline \ \ \ \ \ \ \ \ \ \  \ \ s{\isacharbackquote}{\isacharbraceleft}x{\isasymin}A{\isachardot}\ {\isasymlangle}dc{\isacharunderscore}witness{\isacharparenleft}n{\isacharcomma}A{\isacharcomma}a{\isacharcomma}s{\isacharcomma}R{\isacharparenright}{\isacharcomma}x{\isasymrangle}{\isasymin}R\ {\isacharbraceright}{\isachardoublequoteclose}
 \end{isabelle}
@@
 \isatt{apply\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i\ {\isasymRightarrow}\ i\ {\isasymRightarrow}\ i{\isachardoublequoteclose}}
-is the operation of function application. We will not make a
-distinction in our prose and simply write \isatt{s}(\isatt{b}) in this case.
+is the operation of function application.
+% We will not make a distinction in our prose and simply write
+% \isatt{s}(\isatt{b}) in this case.
 
@@
 \ pointed{\isacharunderscore}DC\ \ {\isacharcolon}\ {\isachardoublequoteopen}{\isacharparenleft}{\isasymforall}x{\isasymin}A{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}{\isasymin}\ R{\isacharparenright}\ {\isasymLongrightarrow}\isanewline
-\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ {\isasymforall}a{\isasymin}A{\isachardot}\ {\isacharparenleft}{\isasymexists}f\ {\isasymin}\ nat{\isasymrightarrow}A{\isachardot}\ f{\isacharbackquote}{\isadigit{0}}\ {\isacharequal}\ a\ {\isasymand}\ {\isacharparenleft}{\isasymforall}n\ {\isasymin}\ nat{\isachardot}\ {\isasymlangle}f{\isacharbackquote}n{\isacharcomma}f{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isasymrangle}{\isasymin}R{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+     \ \ \ {\isasymforall}a{\isasymin}A{\isachardot}\ {\isacharparenleft}{\isasymexists}f\ {\isasymin}\ nat{\isasymrightarrow}A{\isachardot}\ f{\isacharbackquote}{\isadigit{0}}\ {\isacharequal}\ a\ {\isasymand}\ {\isacharparenleft}{\isasymforall}n\ {\isasymin}\ nat{\isachardot}\ {\isasymlangle}f{\isacharbackquote}n{\isacharcomma}f{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isasymrangle}{\isasymin}R{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
 \end{isabelle}
@@
 \isacommand{lemma}\isamarkupfalse%
-\ aux{\isacharunderscore}DC{\isacharunderscore}on{\isacharunderscore}AxNat{\isadigit{2}}\ {\isacharcolon}\ {\isachardoublequoteopen}{\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}{\isasymlangle}y{\isacharcomma}succ{\isacharparenleft}snd{\isacharparenleft}x{\isacharparenright}{\isacharparenright}{\isasymrangle}{\isasymrangle}\ {\isasymin}\ R\ {\isasymLongrightarrow}\isanewline
-\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ {\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ {\isacharbraceleft}{\isasymlangle}a{\isacharcomma}b{\isasymrangle}{\isasymin}R{\isachardot}\ snd{\isacharparenleft}b{\isacharparenright}\ {\isacharequal}\ succ{\isacharparenleft}snd{\isacharparenleft}a{\isacharparenright}{\isacharparenright}{\isacharbraceright}{\isachardoublequoteclose}
+\ aux{\isacharunderscore}DC{\isacharunderscore}on{\isacharunderscore}AxNat{\isadigit{2}}\ {\isacharcolon}\isanewline \ \ {\isachardoublequoteopen}{\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}{\isasymlangle}y{\isacharcomma}succ{\isacharparenleft}snd{\isacharparenleft}x{\isacharparenright}{\isacharparenright}{\isasymrangle}{\isasymrangle}\ {\isasymin}\ R\ {\isasymLongrightarrow}\isanewline
+    \ \ {\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ {\isacharbraceleft}{\isasymlangle}a{\isacharcomma}b{\isasymrangle}{\isasymin}R{\isachardot}\ snd{\isacharparenleft}b{\isacharparenright}\ {\isacharequal}\ succ{\isacharparenleft}snd{\isacharparenleft}a{\isacharparenright}{\isacharparenright}{\isacharbraceright}{\isachardoublequoteclose}
 \end{isabelle}
@@
 \ aux{\isacharunderscore}sequence{\isacharunderscore}DC{\isadigit{2}}\ {\isacharcolon}\ {\isachardoublequoteopen}{\isasymforall}x{\isasymin}A{\isachardot}\ {\isasymforall}n{\isasymin}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ S{\isacharbackquote}n\ {\isasymLongrightarrow}\isanewline
-\ \ \ \ \ \ \ \ {\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}{\isasymlangle}y{\isacharcomma}succ{\isacharparenleft}snd{\isacharparenleft}x{\isacharparenright}{\isacharparenright}{\isasymrangle}{\isasymrangle}\ {\isasymin}\ {\isacharbraceleft}{\isasymlangle}{\isasymlangle}x{\isacharcomma}n{\isasymrangle}{\isacharcomma}{\isasymlangle}y{\isacharcomma}m{\isasymrangle}{\isasymrangle}{\isasymin}{\isacharparenleft}A{\isasymtimes}nat{\isacharparenright}{\isasymtimes}{\isacharparenleft}A{\isasymtimes}nat{\isacharparenright}{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}{\isasymin}S{\isacharbackquote}m\ {\isacharbraceright}{\isachardoublequoteclose}\isanewline
+  {\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\isanewline \ \ {\isasymlangle}x{\isacharcomma}{\isasymlangle}y{\isacharcomma}succ{\isacharparenleft}snd{\isacharparenleft}x{\isacharparenright}{\isacharparenright}{\isasymrangle}{\isasymrangle}\ {\isasymin}\ {\isacharbraceleft}{\isasymlangle}{\isasymlangle}x{\isacharcomma}n{\isasymrangle}{\isacharcomma}{\isasymlangle}y{\isacharcomma}m{\isasymrangle}{\isasymrangle}{\isasymin}{\isacharparenleft}A{\isasymtimes}nat{\isacharparenright}{\isasymtimes}{\isacharparenleft}A{\isasymtimes}nat{\isacharparenright}{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}{\isasymin}S{\isacharbackquote}m\ {\isacharbraceright}{\isachardoublequoteclose}\isanewline
 \ \ %
@@
 \ sequence{\isacharunderscore}DC{\isacharcolon}\ {\isachardoublequoteopen}{\isasymforall}x{\isasymin}A{\isachardot}\ {\isasymforall}n{\isasymin}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ S{\isacharbackquote}n\ {\isasymLongrightarrow}\isanewline
-\ \ \ \ {\isasymforall}a{\isasymin}A{\isachardot}\ {\isacharparenleft}{\isasymexists}f\ {\isasymin}\ nat{\isasymrightarrow}A{\isachardot}\ f{\isacharbackquote}{\isadigit{0}}\ {\isacharequal}\ a\ {\isasymand}\ {\isacharparenleft}{\isasymforall}n\ {\isasymin}\ nat{\isachardot}\ {\isasymlangle}f{\isacharbackquote}n{\isacharcomma}f{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isasymrangle}{\isasymin}S{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+ {\isasymforall}a{\isasymin}A{\isachardot}\ {\isacharparenleft}{\isasymexists}f\ {\isasymin}\ nat{\isasymrightarrow}A{\isachardot}\ f{\isacharbackquote}{\isadigit{0}}\ {\isacharequal}\ a\ {\isasymand}\ {\isacharparenleft}{\isasymforall}n\ {\isasymin}\ nat{\isachardot}\ {\isasymlangle}f{\isacharbackquote}n{\isacharcomma}f{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isasymrangle}{\isasymin}S{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
 \end{isabelle}
@@
 \ \ \isakeyword{assumes}\ countable{\isacharunderscore}subs{\isacharunderscore}of{\isacharunderscore}P{\isacharcolon}\ \ {\isachardoublequoteopen}{\isasymD}\ {\isasymin}\ nat{\isasymrightarrow}Pow{\isacharparenleft}P{\isacharparenright}{\isachardoublequoteclose}\isanewline
-\ \ \isakeyword{and}\ \ \ \ \ seq{\isacharunderscore}of{\isacharunderscore}denses{\isacharcolon}\ \ \ \ \ \ \ \ {\isachardoublequoteopen}{\isasymforall}n\ {\isasymin}\ nat{\isachardot}\ dense{\isacharparenleft}{\isasymD}{\isacharbackquote}n{\isacharparenright}{\isachardoublequoteclose}
+\ \ \isakeyword{and} \ seq{\isacharunderscore}of{\isacharunderscore}denses{\isacharcolon}  {\isachardoublequoteopen}{\isasymforall}n\ {\isasymin}\ nat{\isachardot}\ dense{\isacharparenleft}{\isasymD}{\isacharbackquote}n{\isacharparenright}{\isachardoublequoteclose}
 \end{isabelle}
@@
 \ \ \isakeyword{assumes}\isanewline
-\ \ \ \ \ \ \ \ {\isadigit{1}}{\isacharcolon}\ \ {\isachardoublequoteopen}x{\isasymin}P{\isachardoublequoteclose}\isanewline
-\ \ \ \ \ \ \ \ \ \ \ \ \isakeyword{and}\isanewline
-\ \ \ \ \ \ \ \ {\isadigit{2}}{\isacharcolon}\ \ {\isachardoublequoteopen}n{\isasymin}nat{\isachardoublequoteclose}\isanewline
+  {\isadigit{1}}{\isacharcolon}\ \ {\isachardoublequoteopen}x{\isasymin}P{\isachardoublequoteclose}\isanewline
+   \isakeyword{and}\isanewline
+  {\isadigit{2}}{\isacharcolon}\ \ {\isachardoublequoteopen}n{\isasymin}nat{\isachardoublequoteclose}\isanewline
 \ \ \isakeyword{shows}\isanewline
-\ \ \ \ \ \ \ \ \ \ \ \ {\isachardoublequoteopen}{\isasymexists}y{\isasymin}P{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ {\isacharparenleft}{\isasymlambda}m{\isasymin}nat{\isachardot}\ {\isacharbraceleft}{\isasymlangle}x{\isacharcomma}y{\isasymrangle}{\isasymin}P{\isacharasterisk}P{\isachardot}\ {\isasymlangle}y{\isacharcomma}x{\isasymrangle}{\isasymin}leq\ {\isasymand}\ y{\isasymin}{\isasymD}{\isacharbackquote}{\isacharparenleft}pred{\isacharparenleft}m{\isacharparenright}{\isacharparenright}{\isacharbraceright}{\isacharparenright}{\isacharbackquote}n{\isachardoublequoteclose}
+   {\isachardoublequoteopen}{\isasymexists}y{\isasymin}P{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ {\isacharparenleft}{\isasymlambda}m{\isasymin}nat{\isachardot}\ {\isacharbraceleft}{\isasymlangle}x{\isacharcomma}y{\isasymrangle}{\isasymin}P{\isacharasterisk}P{\isachardot}\ {\isasymlangle}y{\isacharcomma}x{\isasymrangle}{\isasymin}leq\ {\isasymand}\ y{\isasymin}{\isasymD}{\isacharbackquote}{\isacharparenleft}pred{\isacharparenleft}m{\isacharparenright}{\isacharparenright}{\isacharbraceright}{\isacharparenright}{\isacharbackquote}n{\isachardoublequoteclose}
 \end{isabelle}
 %
-Both of the proofs written in the Isar language.
+As these two proofs are interesting, we have proved them using the
+Isar proof language~\cite{DBLP:conf/tphol/Wenzel99}.
 

### 2018-07-13 LSFA/arXiv/first_steps_into_forcing.tex
typo

@@
 \usepackage{hyperref}
+\usepackage[numbers]{natbib}
 
@@
   the canonical names for elements of $M$, and finally show that if $M$
-  satisfy the axiom of pairing, then $M[G]$ also does.
+  satisfies the axiom of pairing, then $M[G]$ also does.
 \end{abstract}
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{mi-estilo-else}
 \bibliography{draft}
+%\input{first_steps_into_forcing.bbl}
 \end{document}

### 2018-07-13 LSFA/arXiv/rs.tex
moschovakis BOOK
"AC" is not AC

@@
 preferred to give a direct proof based on the one in
-Moschovakis' notes \cite{moschovakis1994notes}. Another advantage is that
+Moschovakis' textbook \cite{moschovakis1994notes}. Another advantage is that
 this proof depends on the following version of $\AC$ (existence of
@@
 \]
-This statement is more standard than $\AC$ as stated in the
+This statement is more standard than ``$\AC$'' as stated in the
 Isabelle/ZF development, which involves a silent use of Replacement

### 2018-08-01 LSFA/isabelle-zf.tex
Cambios de acuerdo a referees

@@
 Let us introduce briefly Paulson's formalization of ZF
-\cite{paulson2017isabelle} in Isabelle and aspects of his formal proof
-for the relative consistency of the Axiom of Choice
+\cite{paulson2017isabelle} in Isabelle and the main aspects of his
+formal proof for the relative consistency of the Axiom of Choice
 \cite{paulson_2003}; we will only focus on those aspects that are
@@
 
-In Isabelle/ZF sets are \emph{individuals}, that is terms of type
+In Isabelle/ZF sets are \emph{individuals}, i.e.\ terms of type
 \isatt{i} and formulas have type \isatt{o} 
@@
   y{\isasymnotin}A{\isacharparenright}%
-  {\isachardoublequoteclose}}. Besides the axioms, Isabelle/ZF also
-introduces several definitions (for example, pairs and sets defined by
-comprehension using separation) and syntactic abbreviations to keep
-the formalization close to the customary manner of doing mathematics.
-Working with the library and extending it is quite
-straightforward. 
-As an example, we introduce  a new term-former which is combination of
-instances of replacement and separation to speak of the image of a function over some set
-defined by comprehension, namely $\{b(x): x\in A\text{ and }Q(x)\}$:
+  {\isachardoublequoteclose}}. %
+Besides the axioms, Isabelle/ZF also introduces several definitions
+(for example, pairs and sets defined by comprehension using
+separation) and syntactic abbreviations to keep the formalization
+close to the customary manner of doing mathematics.  Working with the
+library and extending it is quite straightforward.  As an example, we
+introduce a new term-former denoting the image of a function
+over a set defined by comprehension, namely
+$\{b(x): x\in A\text{ and }Q(x)\}$:
 %% Since we find a need to combine the replacement and
@@
 \isacommand{lemma}\isamarkupfalse%
-\ Sep{\isacharunderscore}and{\isacharunderscore}Replace{\isacharcolon}\ {\isachardoublequoteopen}{\isacharbraceleft}b{\isacharparenleft}x{\isacharparenright}\ {\isachardot}{\isachardot}\ x{\isasymin}A{\isacharcomma}\ P{\isacharparenleft}x{\isacharparenright}\ {\isacharbraceright}\ {\isacharequal}\ {\isacharbraceleft}b{\isacharparenleft}x{\isacharparenright}\ {\isachardot}\ x{\isasymin}{\isacharbraceleft}y{\isasymin}A{\isachardot}\ P{\isacharparenleft}y{\isacharparenright}{\isacharbraceright}{\isacharbraceright}{\isachardoublequoteclose}
+\ Sep{\isacharunderscore}and{\isacharunderscore}Replace{\isacharcolon}\ {\isachardoublequoteopen}{\isacharbraceleft}b{\isacharparenleft}x{\isacharparenright}\ {\isachardot}{\isachardot}\ x{\isasymin}A{\isacharcomma}\ Q{\isacharparenleft}x{\isacharparenright}\ {\isacharbraceright}\ {\isacharequal}\ {\isacharbraceleft}b{\isacharparenleft}x{\isacharparenright}\ {\isachardot}\ x{\isasymin}{\isacharbraceleft}y{\isasymin}A{\isachardot}\ Q{\isacharparenleft}y{\isacharparenright}{\isacharbraceright}{\isacharbraceright}{\isachardoublequoteclose}
 \end{isabelle}
@@
 
-Mathematical texts usually start by fixing the context which defines
-parameters and assumptions necessary to develop theorems
+Mathematical texts usually start by fixing the context that defines
+parameters and assumptions needed to develop theorems
 and results. In Isabelle the way of defining contexts is through

### 2018-08-10 LSFA/intro.tex
Cambios menores y compromisos.

@@
 %
-\noindent As in the last sentence, one usually refer to a transitive
+\noindent As in the last sentence, one usually refers to a transitive
 model by the underlying set because the relation is fixed.
@@
 It is instructive to sketch G\"odel's argument of the relative
-consistency of $\CH$: Assume that $M$ is a ctm of $\ZFC$, G\"odel
+consistency of $\CH$: Assuming that $M$ is a ctm of $\ZFC$, G\"odel
 showed that $M$ contains a minimal submodel $L^M$ of the same
@@
 such that $L^M = \{x\in M : M\models L(x)\}$. To show that
-$L^M\models \ZFC+\CH$, one uses that the $\ZFC+\CH$ holds in $M$.
+$L^M\models \ZFC+\CH$, one uses the fact that $\ZFC$ holds in $M$.
 
@@
 %% MIGUEL: la última oración me resulta un poco difícil de entender.
-The Rasiowa-Sikorski lemma (RSL) states that, fixed a preorder $\PP$, for
-any countable family of dense subsets of $\PP$ there is a filter
-intersecting all of them. Thus, there are generic filters $G$ for
-countable transitive models. In general, no such $G$ belongs to $M$ and
+The Rasiowa-Sikorski lemma (RSL) states that for any preorder $\PP$
+and any countable family $\{\mathcal{D}_n : n\in\mathbb{N}\}$ of dense
+subsets of $\PP$ there is a filter intersecting every
+$\mathcal{D}_i$. Thus, there are generic filters $G$ for countable
+transitive models. In general, no such $G$ belongs to $M$ and
 therefore the extension $M[G]$ is proper. We formalize the proof of
-RSL in
-Section~\ref{sec:rasiowa-sikorski-lemma}. A requisite result on a
-version of the Axiom of Choice is formalized in
-Section~\ref{sec:sequence-version-dc}. We then apply RSL to prove the existence of generic
-filters in Section~\ref{sec:generic-filter}.
+RSL in Section~\ref{sec:rasiowa-sikorski-lemma}. A requisite result on
+a version of the Axiom of Choice is formalized in
+Section~\ref{sec:sequence-version-dc}. We then apply RSL to prove the
+existence of generic filters in Section~\ref{sec:generic-filter}.
 

### 2018-08-10 LSFA/isabelle-zf.tex
Cambios menores y compromisos.

@@
 In Isabelle/ZF sets are \emph{individuals}, i.e.\ terms of type
-\isatt{i} and formulas have type \isatt{o} 
-(akin to a \emph{Bool} type, but at the object level). 
-The  axiomatization 
-of $\ZFC$ in Isabelle/ZF proceeds by postulating  a binary predicate
-\isatt{\ensuremath{\in}} and   several set
-constructors (terms and function with values in \isatt{i}) corresponding to the  empty set
-(the constant \isatt{\isadigit{0}}), powersets, and one further constant
+\isatt{i} and formulas have type \isatt{o} (akin to a \emph{Bool}
+type, but at the object level).  The axiomatization of $\ZFC$ in
+Isabelle/ZF proceeds by postulating a binary predicate
+\isatt{\ensuremath{\in}} and several set constructors (terms and
+function with values in \isatt{i}) corresponding to the empty set (the
+constant \isatt{\isadigit{0}}), powersets, and one further constant
 \isatt{inf} for an infinite set. The axioms, being formulas, are terms
-of type \isatt{o}; 
-the foundation axiom, for example, is formalized as (the universal
-closure of) 
-\isa{{\isachardoublequoteopen}A\ {\isacharequal}\ {\isadigit{0}}\
-  {\isasymor}%
+of type \isatt{o}; the foundation axiom, for example, is formalized as
+(the universal closure of) \isa{{\isachardoublequoteopen}A\
+  {\isacharequal}\ {\isadigit{0}}\ {\isasymor}%
   \ {\isacharparenleft}{\isasymexists}x{\isasymin}A{\isachardot}\ %
@@
 library and extending it is quite straightforward.  As an example, we
-introduce a new term-former denoting the image of a function
-over a set defined by comprehension, namely
+introduce a new term-former (which is a combination of instances of
+replacement and separation) denoting the image of a function over a
+set defined by comprehension, namely
 $\{b(x): x\in A\text{ and }Q(x)\}$:
@@
 
-Mathematical texts usually start by fixing the context that defines
+Mathematical texts usually start by fixing a context that defines
 parameters and assumptions needed to develop theorems

### 2018-08-10 LSFA/names.tex
última leída.
Cambios menores y compromisos.

@@
 \begin{equation}\label{eq:def-val}
-\val(G,\tau)\defi \{\val(G,\sigma) : \exists p\in \PP
+\val(G,\tau)\defi \{\val(G,\sigma) : \exists p\in \PP.\,
 (\lb\sigma,p\rb\in\tau \y p\in G)\}.
@@
 %
-As with any recursive definition, to justify that the definition makes
-sense (i.e., it defines anything at all) one has to explicit the
-well-founded relation which is used. In this case, we use the relation
-$\mathit{ed}$:
+As with any recursive definition, in order to justify that the
+definition makes sense (i.e., that it defines anything at all) one has
+to make explicit the well-founded relation which is used. In this
+case, we use the relation $\mathit{ed}$:
 \[
@@
 \isacommand{lemma}\isamarkupfalse%
-\ edrel{\isacharunderscore}sub{\isacharunderscore}memrel{\isacharcolon}\ {\isachardoublequoteopen}edrel{\isacharparenleft}A{\isacharparenright}\ {\isasymsubseteq}\ trancl{\isacharparenleft}Memrel{\isacharparenleft}eclose{\isacharparenleft}A{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
-\end{isabelle}
-\dots
-\begin{isabelle}
-\isacommand{lemma}\isamarkupfalse%
 \ wf{\isacharunderscore}edrel\ {\isacharcolon}\ {\isachardoublequoteopen}wf{\isacharparenleft}edrel{\isacharparenleft}A{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
@@
 %
+\end{isabelle}
+
+\noindent We omit the proof of the following lemma used above:
+
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ edrel{\isacharunderscore}sub{\isacharunderscore}memrel{\isacharcolon}\ {\isachardoublequoteopen}edrel{\isacharparenleft}A{\isacharparenright}\ {\isasymsubseteq}\ trancl{\isacharparenleft}Memrel{\isacharparenleft}eclose{\isacharparenleft}A{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
 \end{isabelle}

### 2018-08-13 LSFA/rs.tex
lemma --> corollary (donde correspondía)

--- 2018-06-25 LSFA/rs.tex
@@
 \begin{isabelle}
-\isacommand{lemma}\isamarkupfalse%
-\ aux{\isacharunderscore}DC{\isacharunderscore}on{\isacharunderscore}AxNat{\isadigit{2}}\ {\isacharcolon}\ {\isachardoublequoteopen}{\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}{\isasymlangle}y{\isacharcomma}succ{\isacharparenleft}snd{\isacharparenleft}x{\isacharparenright}{\isacharparenright}{\isasymrangle}{\isasymrangle}\ {\isasymin}\ R\ {\isasymLongrightarrow}\isanewline
-\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ {\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ {\isacharbraceleft}{\isasymlangle}a{\isacharcomma}b{\isasymrangle}{\isasymin}R{\isachardot}\ snd{\isacharparenleft}b{\isacharparenright}\ {\isacharequal}\ succ{\isacharparenleft}snd{\isacharparenleft}a{\isacharparenright}{\isacharparenright}{\isacharbraceright}{\isachardoublequoteclose}
+\isacommand{corollary}\isamarkupfalse%
+\ DC{\isacharunderscore}on{\isacharunderscore}A{\isacharunderscore}x{\isacharunderscore}nat\ {\isacharcolon}\ \isanewline
+\ \ {\isachardoublequoteopen}{\isacharparenleft}{\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}{\isasymlangle}y{\isacharcomma}succ{\isacharparenleft}snd{\isacharparenleft}x{\isacharparenright}{\isacharparenright}{\isasymrangle}{\isasymrangle}\ {\isasymin}\ R{\isacharparenright}\ {\isasymLongrightarrow}\isanewline
+\ \ \ \ {\isasymforall}a{\isasymin}A{\isachardot}\ {\isacharparenleft}{\isasymexists}f\ {\isasymin}\ nat{\isasymrightarrow}A{\isachardot}\ f{\isacharbackquote}{\isadigit{0}}\ {\isacharequal}\ a\ {\isasymand}\ {\isacharparenleft}{\isasymforall}n\ {\isasymin}\ nat{\isachardot}\ {\isasymlangle}{\isasymlangle}f{\isacharbackquote}n{\isacharcomma}n{\isasymrangle}{\isacharcomma}{\isasymlangle}f{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isacharcomma}succ{\isacharparenleft}n{\isacharparenright}{\isasymrangle}{\isasymrangle}{\isasymin}R{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
 \end{isabelle}
+
 %
@@
 \isacommand{lemma}\isamarkupfalse%
-\ aux{\isacharunderscore}sequence{\isacharunderscore}DC{\isadigit{2}}\ {\isacharcolon}\ {\isachardoublequoteopen}{\isasymforall}x{\isasymin}A{\isachardot}\ {\isasymforall}n{\isasymin}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ S{\isacharbackquote}n\ {\isasymLongrightarrow}\isanewline
+\ aux{\isacharunderscore}sequence{\isacharunderscore}DC\ {\isacharcolon}\ {\isachardoublequoteopen}{\isasymforall}x{\isasymin}A{\isachardot}\ {\isasymforall}n{\isasymin}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ S{\isacharbackquote}n\ {\isasymLongrightarrow}\isanewline
 \ \ \ \ \ \ \ \ {\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}{\isasymlangle}y{\isacharcomma}succ{\isacharparenleft}snd{\isacharparenleft}x{\isacharparenright}{\isacharparenright}{\isasymrangle}{\isasymrangle}\ {\isasymin}\ {\isacharbraceleft}{\isasymlangle}{\isasymlangle}x{\isacharcomma}n{\isasymrangle}{\isacharcomma}{\isasymlangle}y{\isacharcomma}m{\isasymrangle}{\isasymrangle}{\isasymin}{\isacharparenleft}A{\isasymtimes}nat{\isacharparenright}{\isasymtimes}{\isacharparenleft}A{\isasymtimes}nat{\isacharparenright}{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}{\isasymin}S{\isacharbackquote}m\ {\isacharbraceright}{\isachardoublequoteclose}\isanewline
@@
 %
-Both of the proofs written in the Isar language.
+As these two proofs are interesting, we have proved them using the
+Isar proof language~\cite{DBLP:conf/tphol/Wenzel99}.
 

### 2018-08-13 LSFA/names.tex
Agregando intros a las secciones 3,4 y 5

@@
 \label{sec:names}
-In this section we formalize the function $\val$ that allows to
+We formalize the function $\val$ that allows to
 construct the elements of the generic extension $M[G]$ from elements

### 2018-08-15 LSFA/rs.tex
DC vs Pointed_DC

@@
-\subsection{A sequence version of $\DC$}\label{sec:sequence-version-dc}
-In order to prove the Rasiowa-Sikorski lemma in
-general, a version of the Axiom of Choice ($\AC$) must be proved. This
-version is known as the \emph{Principle of Dependent Choices}:
+\subsection{A sequence version of Dependent
+  Choices}\label{sec:sequence-version-dc} 
+%% In order to prove the Rasiowa-Sikorski lemma in
+%% general, a consequence of the Axiom of Choice ($\AC$) must be invoked. This
+%% consequence is known as the \emph{Principle of Dependent Choices}:
+The Rasiowa-Sikorski lemma follows naturally from a ``pointed''
+version of   the \emph{Principle of Dependent Choices} which,
+in turn, is a consequence of the Axiom of Choice ($\AC$). It is
+therefore natural to take as a starting point the theory \isatt{AC}
+which adds the latter axiom to the toolkit of Isabelle/ZF.
+
+The statement we are interested in is the following:
 \begin{quote}
-  ($\DC$) Let $\rr$ be a binary relation on $A$, and $a\in A$. If
+  (Pointed $\DC$) Let $\rr$ be a binary relation on $A$, and $a\in A$. If
   $\forall x\in A.\,  \exists y\in A.\, x\rr y$, then there exists
@@
 
-A different version of $\DC$ (without the constraint $f(0)=a$, but
-allowing a parameter that permits to prove an equivalence with $\AC$)
-was formalized by Krzysztof
-Grabczewski~\cite{DBLP:journals/jar/PaulsonG96} as part of the
-Isabelle/ZF library, along with many statements that follow from or
-are equivalent to $\AC$. Since the proofs there required a greater
-generality, it would have been very contrived to derive this
-``pointed'' version of $\DC$ from that development. Instead, we
-preferred to give a direct proof based on the one in
-Moschovakis' notes \cite{moschovakis1994notes}. Another advantage is that
-this proof depends on the following version of $\AC$ (existence of
+Two different versions of $\DC$ (called $\DC0$ and $\DC(\kappa)$) have
+already been formalized  by Krzysztof 
+Grabczewski~\cite{DBLP:journals/jar/PaulsonG96}, as part of a
+study of equivalents to $\AC$ (following Rubin and Rubin
+\cite{rubin1985equivalents}). Nevertheless, those are not convenient
+for our purposes for the reasons we now explain. The axiom $\DC0$
+corresponds essentially to our Pointed $\DC$ but without the
+constraint $f(0)=a$; it is a nice exercise to show that $\DC0$ implies
+Pointed $\DC$, but a formalization would have a moderate length. On
+the other hand, the shortest path from $\AC$ as stated in the
+corresponding Isabelle theory to $\DC0$ using the already formalized
+material involves a complicated detour (130+ proof commands spanning
+various files of the \isatt{ZF-AC} theory), compared to the mere 11 commands from $\AC$ to
+\isatt{AC{\isacharunderscore}func{\isacharunderscore}Pow}. This last
+one is the
+choice principle that we use in our formalization of Pointed $\DC$, and states 
+the existence of
 choice functions on $\P(A)\sm \{\mty\}$):
@@
 \]
-This statement is more standard than $\AC$ as stated in the
-Isabelle/ZF development, which involves a silent use of Replacement
-since it is written as an axiom scheme.
+Another advantage of taking
+\isatt{AC{\isacharunderscore}func{\isacharunderscore}Pow} as a
+starting point is that it does not involve proper classes: The version
+of  $\AC$ in Isabelle/ZF corresponds to an axiom scheme of first order
+logic and as such is not a standard formulation. 
+%% 26 commands to get Pointed $\DC$.
 
-The strategy is to define the function $f$ discussed above by
+The strategy to prove Pointed $\DC$ (following a proof in Moschovakis
+\cite{moschovakis1994notes}) is to define the function $f$ discussed above by
 primitive recursion on the naturals, which can be done easily thanks
@@
 
-The proof is mostly routine; after a few lemmas we obtain the
+The proof is mostly routine; after a few lemmas (having 26 proof
+commands in total) we obtain the
 following theorem:

### 2018-08-15 LSFA/rs.tex
más detalle

@@
 \cite{rubin1985equivalents}). Nevertheless, those are not convenient
-for our purposes for the reasons we now explain. The axiom $\DC0$
+for our purposes for reasons we now explain. The axiom $\DC0$
 corresponds essentially to our Pointed $\DC$ but without the
 constraint $f(0)=a$; it is a nice exercise to show that $\DC0$ implies
-Pointed $\DC$, but a formalization would have a moderate length. On
-the other hand, the shortest path from $\AC$ as stated in the
-corresponding Isabelle theory to $\DC0$ using the already formalized
+Pointed $\DC$, but a formalization would have a moderate
+length. $\DC(\kappa)$ is rather different in nature and it is tailored
+to  obtain another proposition
+equivalent to the axiom of choice (actually,
+$\AC\longleftrightarrow (\forall \kappa
+.\; \mathrm{Card}(\kappa)\longrightarrow \DC(\kappa))$. 
+Finally, the shortest path from $\AC$ %% as stated in the 
+%% corresponding Isabelle theory
+to $\DC0$ using already formalized
 material involves a complicated detour (130+ proof commands spanning
-various files of the \isatt{ZF-AC} theory), compared to the mere 11 commands from $\AC$ to
+various files of the \isatt{ZF-AC} theory and going through the Well
+Ordering Theorem and  $\DC(\omega)$), compared to the mere 11 commands from $\AC$ to
 \isatt{AC{\isacharunderscore}func{\isacharunderscore}Pow}. This last
@@
 
-The proof is mostly routine; after a few lemmas (having 26 proof
+The proof is mostly routine; after a few lemmas (26 proof
 commands in total) we obtain the

### 2018-08-16 LSFA/rs.tex
un parentesis

@@
 The Rasiowa-Sikorski lemma follows naturally from a ``pointed''
-version of   the \emph{Principle of Dependent Choices} which,
+version of   the \emph{Principle of Dependent Choices ($\DC$)} which,
 in turn, is a consequence of the Axiom of Choice ($\AC$). It is
@@
 $\AC\longleftrightarrow (\forall \kappa
-.\; \mathrm{Card}(\kappa)\longrightarrow \DC(\kappa))$. 
+.\; \mathrm{Card}(\kappa)\longrightarrow \DC(\kappa))$). 
 Finally, the shortest path from $\AC$ %% as stated in the 

### 2018-08-17 LSFA/names.tex
Explicación para wfrec y Hv

@@
 of the ctm $M$ and the generic filter $G$. The definition of $\val$
-can be written succintly as
+can be written succintly as a recursive equation
 %
@@
 %
-As with any recursive definition, in order to justify that the
-definition makes sense (i.e., that it defines anything at all) one has
-to make explicit the well-founded relation which is used. In this
-case, we use the relation $\mathit{ed}$:
+The justification that $\val$ is well-defined comes from a general
+result (transfinite recursion on well-founded
+relations~\cite[p. 48]{kunen2011set}). Given a well-founded relation
+$R \subseteq A \times A$ and a functional
+$H : A \times (A \to A) \to A$, the principle asserts the existence of
+a function $F : A \to A$ satisfying
+$F(a) = H(a,F\uparrow\,(R^{-1}(a)))$. This principle is formalized in
+Isabelle/ZF and one can use the operator \isa{wfrec}\footnote{Notice
+  that this form of recursive definitions is more general than the one
+  used in the previous section to define
+  \isa{dc{\isacharunderscore}witness}.} to define functions using
+transfinite recursion. To be precise, \isa{wfrec :: [i, i, [i,i]=>i]
+  => i} is a slight variation, where the first argument is the
+relation, the third is the functional, and the second corresponds to
+the argument of $F$. Notice that the relation and the function argument
+of the functional are internalized as terms of type \isa{i}.
+
+In our case the functional $H$ takes an additional argument for the
+parameter $G$:
+\[
+  H(G,y,f) = \{f(x) : x \in dom(y) \y \exists p\in \PP.\,
+(\lb x,p\rb\in y \y p\in G)\}
+\]
+while the relation is given by:
 \[
@@
 
-The formalization of $\val$ in Isabelle/ZF involves the function
-\isa{wfrec},\footnote{Notice that this form of recursive
-  definitions is more general than the one used in the previous
-  section to define \isa{dc{\isacharunderscore}witness}.} which takes as an argument a
-well-founded relation of type \isa{i}, i.e.\ a set of pairs. Therefore
-we restrict $\mathit{ed}$ above to a set.
+In the formalization of $\mathit{ed}$ is restricted to a set,
+so to have a term of type \isa{i}:
 % 
@@
 
-\noindent We now turn to the definition of $\val$. In order to use
-\isa{wfrec} we need to define a functional \isa{Hv}.
-
+\noindent The formalizaton of the functional $H$ is straightforward and $\val$ is defined using \isa{wfrec}:
+%
 \begin{isabelle}
@@
 \ \ Hv\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}i{\isasymRightarrow}i{\isasymRightarrow}i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
-\ \ {\isachardoublequoteopen}Hv{\isacharparenleft}G{\isacharcomma}x{\isacharcomma}f{\isacharparenright}\ {\isacharequal}{\isacharequal}\ {\isacharbraceleft}\ f{\isacharbackquote}y\ {\isachardot}{\isachardot}\ y{\isasymin}\ domain{\isacharparenleft}x{\isacharparenright}{\isacharcomma}\ {\isasymexists}p{\isasymin}P{\isachardot}\ {\isacharless}y{\isacharcomma}p{\isachargreater}\ {\isasymin}\ x\ {\isasymand}\ p\ {\isasymin}\ G\ {\isacharbraceright}{\isachardoublequoteclose}\isanewline
+\ \ {\isachardoublequoteopen}Hv{\isacharparenleft}G{\isacharcomma}y{\isacharcomma}f{\isacharparenright}\ {\isacharequal}{\isacharequal}\ {\isacharbraceleft}\ f{\isacharbackquote}x\ {\isachardot}{\isachardot}\ x{\isasymin}\ domain{\isacharparenleft}y{\isacharparenright}{\isacharcomma}\ {\isasymexists}p{\isasymin}P{\isachardot}\ {\isacharless}x{\isacharcomma}p{\isachargreater}\ {\isasymin}\ y\ {\isasymand}\ p\ {\isasymin}\ G\ {\isacharbraceright}{\isachardoublequoteclose}\isanewline
 \isanewline

### 2018-08-18 LSFA/related.tex
template subsección trabajo (no) relacionado

--- 2018-06-25 LSFA/intro.tex
@@
-\section{Introduction}\label{sec:introduction}
-Set Theory plays a double role in Mathematics: It is one of its
-possible foundations and also an active research area.
-%
-As it is widely known, Georg Cantor introduced its main concepts and in
-particular showed the fundamental result that the real line, $\R$  is not
-coordinable with the natural numbers. Soon after this, he posed the
-most important question in the field, written as a conjecture:
-\begin{quote}
-  The \emph{Continuum Hypothesis} ($\CH$). Every uncountable subset of $\R$ is
-  coordinable with $\R$.
-\end{quote}
-
-The current axiomatic foundation of Set Theory is through first-order
-logic and uses the axioms devised by Zermelo and Fraenkel, including
-the Axiom of Choice ($\AC$) among them. This theory is known by the
-$\ZFC$ acronym. G\"odel \cite{godel-L} showed that $\CH$ cannot be refuted using
-$\ZFC$, unless this theory itself is inconsistent (we say that
-\emph{$\CH$ is relatively consistent with $\ZFC$}). For a while, this
-result left the possibility that one might be able to show
-$\ZFC\models \CH$, but in a groundbreaking work \cite{Cohen-CH-PNAS},
-Paul Cohen discovered the technique of \emph{forcing} and proved that
-$\neg\CH$ is relatively consistent with $\ZFC$. Forcing has been used
-since then for showing innumerable independence results and to perform
-mathematical constructions.
-
-A great part of G\"odel's work on this subject has been formalized in
-Isabelle~\cite{Isabelle} by Lawrence Paulson~\cite{paulson_2003}. This paper formalizes
-a first part of the machinery of forcing. In the rest of the
-introduction we discuss some of the set-theoretical details involved
-and explain briefly Paulson's formalization.
-
-\subsection{Models of $\ZFC$}
-By G\"odel's Second Incompleteness Theorem, we cannot  prove that
-there exists a model of $\ZFC$. More formally, if we assume that
-mathematical proofs can be encoded as theorems of $\ZFC$ and that
-the latter do not lead to contradictions (i.e., $\ZFC$ is
-\emph{consistent}), then we cannot prove that there exists a set $M$
-and a binary relation $E$ such that $\lb M,E\rb$ satisfies the $\ZFC$
-axioms.
-
-A relative consistency proof for an axiom $A$ is then obtained by
-assuming that there exists a model of ZFC, say $\lb M,E\rb$, and
-constructing another model $\lb M',E'\rb$ for $\ZFC + A$. We single
-out a very special kind of models:
-%
-\begin{definition}\label{def:transitive-model}
-  \begin{enumerate}
-  \item A set $M$ (of sets) is \emph{transitive} if for all $x\in M$ and 
-    $y\in x$, we have $y\in M$ (i.e., every element of $M$ is a subset
-    of $M$).
-  \item $\lb M,E\rb$ is a \emph{transitive model} if $M$ is transitive
-     and  $E$ is the membership relation $\in$ restricted to
-    $M$. It is \emph{countable} if $M$ is coordinable with a subset of
-    $\N$; we then say that the model  $M$ is a \emph{ctm}. 
-  \end{enumerate}
-\end{definition}
-%
-\noindent As in the last sentence, one usually refer to a transitive
-model by the underlying set because the relation is fixed.
-%% MIGUEL: feísimo este corte. Quizás refrasear la oración anterior
-%% para que entre en un renglón.
-
-%% MIGUEL: lo que sigue implica que para tener aritmética en ZFC son
-%% necesarios una cantidad infinita de axiomas de ZFC?
-In spite of G\"odel's Second Incompleteness Theorem, one can find
-transitive models for every finite fragment of $\ZFC$. More precisely,
-%
-\begin{theorem}\label{th:ctm-finite-axioms}
-  For each finite subset
-  $\Phi\sbq \ZFC$, the statement \emph{``there exists a countable
-    transitive model of $\Phi$''} is a theorem of $\ZFC$.
-\end{theorem}
-%
-\noindent This follows by a combination of the Reflection Principle, the
-L\"owenheim-Skolem Theorem, and the Mostowksi Collapse. The reader can
-consult the details in \cite{kunen2011set}. Consistency arguments that
-assume the existence of a ctm $M$ of $\ZFC$ can usually be replaced by
-a model as in Theorem~\ref{th:ctm-finite-axioms}, since a first-order
-proof (e.g.\ of a contradiction)%
-%%%%%%%%%%%%%
-\footnote{It is relevant to this point that 
-  both the approaches by G\"odel and Cohen for showing
-  relative consistency of an axiom $A$ 
-  can be used to obtain an algorithm transforming a proof
-  concluding a contradiction from $\ZFC+A$ to one from $\ZFC$.}
-%%%%%%%%%%%%%
-involves only finitely many axioms.
-
-Assume that $M$ is a ctm of $\ZFC$. It is instructive to sketch
-G\"odel's argument of the relative consistency of $\CH$: he showed
-that $M$ contains a minimal submodel $L^M$ of the same ``height''
-(i.e.\ having the same ordinals) that satisfies $\ZFC+\CH$. The sets
-in $L^M$ are called \emph{constructible} and are in a sense
-``definable.'' In fact, there is a first-order formula $L$ such that
-$L^M = \{x\in M : M\models L(x)\}$. To show that
-$L^M\models \ZFC+\CH$, one uses that the $\ZFC+\CH$ holds in $M$.
-
-%% MIGUEL: acá me parece que hay un salto medio abrupto. Por que
-%% es una necesidad primaria? Lo podés relacionar con lo que venía
-%% del párrafo anterior?
-%% MIGUEL: quizás lo que querés decir acá es para probar que vamos
-%% a construir un modelo N a partir de un modelo M; y para probar
-%% que N satisface un axioma, vamos a utilizar que M lo satisface.
-It is therefore a primary need to have a means to correlate  (first-order)
-properties satisfied by a model $M$ and those of a 
-submodel $N\sbq M$. As a simple example on this, consider 
-%% $M\defi \{\emptyset,\{\emptyset\},\{\{\emptyset\}\}\}$ and
-%% $N\defi\{\{\emptyset\},\{\{\emptyset\}\}\}$, and let 
-%% $\phi(x)\defi \neg\exists y (y \in x)$. It is clear that 
-%% $M\models \phi(\emptyset)$; but we have
-%% \[
-%% M\not\models \phi(\{\emptyset\}) \quad\text{ but }\quad N\models \phi(\{\emptyset\}).
-%% \]
-%% There is a discrepancy between what $M$ and $N$ \emph{believe} about
-%% their common element $\{\emptyset\}$.
-$M\defi \{a,b,c, \{a,b\},\{a,b,c\}\}$ and
-$N\defi\{a,b,\{a,b,c\}\}$, and let 
-\[
-\phi(x,y,z)\defi \forall w.\,( w\in z \lsii w=x \o w=y).
-\]
-%% It is clear that  $M\models \phi(\{a,b\})$; but we have
-Then we have
-\[
-M\not\models \phi(a,b,\{a,b,c\}) \quad\text{ but }\quad N\models \phi(a,b,\{a,b,c\}).
-\]
-There is a discrepancy between  $M$ and $N$ about $\{a,b,c\}$ being ``the
-(unordered) pair of $a$ and $b$.'' We say that $\phi$ holds for
-$a,b,\{a,b,c\}$ \emph{relative} to $N$. It is immediate to see that
-$\phi$ holds  for $x,y,z$ relative to $N$ if and only if
-\[
-\phi^N(x,y,z)\defi \forall w.\ w\in N\limp ( w\in z \lsii w=x \o w=y)
-\] 
-holds. $\phi^N$ is called the \emph{relativization of $\phi$ to
-  $N$}. One can generalize this operation of relativization to the
-class of all sets satisfying a first-order predicate $C$ in a
-straightforward way:
-\[
-\phi^C(x,y,z)\defi \forall w.\ C(w)\limp ( w\in z \lsii w=x \o w=y)
-\]
-
-It can be shown elementarily that if $M$ and $N$ are transitive,
-$\phi^N$ holds if and only if $\phi^M$ holds,  for $x,y,z\in N$. We
-say then that $\phi$ is \emph{absolute between $N$ and $M$.}
-%% MIGUEL: "truth of axioms in $M$" significa que $M$ satisface los
-%% axiomas?
-The concepts of relativization and absoluteness are central to the
-task of transferring truth of axioms in $M$ to $L^M$, and constitute
-the hardest part of Paulson's development.
-
-\subsection{Forcing}
-%% MIGUEL: no veo la necesidad del adversativo but, pondría and.
-%% MIGUEL: Qué significa "a new set $G$ is adjoined to $M$"?
-%% MIGUEL: este párrafo necesita correcciones.
-Forcing is a technique to extend countable transitive models of
-$\ZFC$. This process is guaranteed to preserve the $\ZFC$
-axioms while allowing to fine-tune what other first order properties the
-extension will have. Given a ctm $M$  of $\ZFC$ and a set $G$, one constructs a new
-ctm  $M[G]$  that includes $M$ and
-contains $G$, and proves that under some hypotheses ($G$ being ``generic''),
-$M[G]$ satisfies $\ZFC$.
-
-The easiest way to define genericity is by using a preorder with top
-$\lb\PP,\leq,\1\rb$ in $M$.   
-In Section~\ref{sec:forcing-notions} we formalize the definitions of
-\emph{dense} subset and  \emph{filter} of  $\PP$, and we say that  $G$
-is an $M$-generic filter
-if it intersects every dense subset of $\PP$ that lies in $M$.
-
-%% MIGUEL: Tenemos que dado un ctm M, existe un filtro genérico G.
-%% ¿Dónde existe? Una primera lectura de ignorante me dice "existe
-%% en M", pero eso me lo desmiente la siguiente oración. ¿Juega algún
-%% rol absolutez? Si es así estaría bien explicitarlo.
-%% MIGUEL: la última oración me resulta un poco difícil de entender.
-The Rasiowa-Sikorski lemma (RSL) states that, fixed a preorder $\PP$, for
-any countable family of dense subsets of $\PP$ there is a filter
-intersecting all of them. Thus, there are generic filters $G$ for
-countable transitive models. In general, no such $G$ belongs to $M$ and
-therefore the extension $M[G]$ is proper. We formalize the proof of
-RSL in
-Section~\ref{sec:rasiowa-sikorski-lemma}. A requisite result on a
-version of the Axiom of Choice is formalized in
-Section~\ref{sec:sequence-version-dc}. We then apply RSL to prove the existence of generic
-filters in Section~\ref{sec:generic-filter}.
-
-%% MIGUEL: redacción alternativa del párrafo siguiente.
-Every  $y \in M[G]$ is obtained from an element $\dot y$ of $M$, thus
-elements of $M$ are construed as \emph{names} or codes for elements of
-$M[G]$.
-The decoding is given by the function
-$\val$, which takes the generic filter $G$ as a parameter. To
-prove that $M$ is contained in $M[G]$ it suffices to give a name for
-each element of $M$; we define the function $\chk$ which assigns
-a name for each $x\in M$. Showing that $\chk(x)\in M$
-when $x\in M$ involves some technical issues that will
-be addressed in a further work. We explain names, $\val$, and
-$\chk$ in Section~\ref{sec:names}.
-
-A central part of this formalization project involves showing that
-$\ZFC$ holds in the generic extension. This is most relevant since
-forcing is essentially the only known way to \emph{extend} models of
-$\ZFC$ (while preserving ordinals). The most difficult step to achieve
-this goal is to define the \emph{forcing relation}, that allows to
-connect satisfaction in $M$ to that of $M[G]$; this is needed to show
-that the Powerset axiom and the axiom schemes of Separation and
-Replacement hold in $M[G]$. In
-Section~\ref{sec:pairing-generic-extension}
- we tackle the Pairing
-Axiom. This does not require the forcing relation, but provides an
-illustration of the use of names. The development can be downloaded
-from \url{https://cs.famaf.unc.edu.ar/~pedro/forcing/}.
+\subsection{Related work}\label{sec:related-work}
 

### 2018-08-18 LSFA/related.tex
casi listo

@@
 \subsection{Related work}\label{sec:related-work}
+Before we proceed, we will discuss very succinctly recent related
+formalizations.
 
+There is another major framework in Isabelle based on higher order
+logic, Isabelle/HOL. This framework is very active, and as a 
+consequence more automated tools are available. Isabelle/HOL has 
+basic chapters on set theory. One of those, by Steven Obua, proceeds up to
+well founded relations and provides translations between types in HOL
+(for instance \isatt{nat}) to  sets  (elements of type
+\isatt{ZF}). Another one, by A.~Popescu and D.~Traytel, reaches
+cardinal artihmetic. This is fairly limited for our purposes.
+
+Concerning automation, B.~Zhan has developed a new tool called
+\texttt{auto2} and applied it to untyped set theory
+\cite{10.1007/978-3-319-66107-0_32}. He has redeveloped essentially
+the basic results in Isabelle/ZF, but goes in a different
+direction. Nevertheless, a majority of results in
+Isabelle/ZF are not yet implemented using this tool, and a another
+downside is that proofs using it do not follow the standard Isar
+language.
+
+Finally, there are some forcing-related formalizations in the theory
+of types. Most relevant is the work by K.~Quirin \cite{Quirin}, where
+a sheaf-theoretic initial approach to forcing is implemented in
+Coq. This language is extremely different to the standard approach of
+constructing models of $\ZFC$, and it might be difficult (once the
+forcing machinery is set) to translate results in the literature using
+ctms to this one. In any case, the translation to set theory of what
+Quirin accomplishes is to define a generic extension (where $\CH$
+should fail) and to construct a set $K$ (a candidate counterexample)
+and injections $\N\to K$ and $K\to\R$. But the most important part,
+that is, that there are no surjections
+
+\begin{enumerate}
+\item Jaber et al.\ \cite{jaber:hal-01319066}.
+\end{enumerate}
 %We continue with a brief description of the Isabelle framework.

### 2018-08-18 LSFA/related.tex
final sin revisar, falta Jaber

@@
 
-Finally, there are some forcing-related formalizations in the theory
+There are some forcing-related formalizations in the theory
 of types. Most relevant is the work by K.~Quirin \cite{Quirin}, where
@@
 should fail) and to construct a set $K$ (a candidate counterexample)
-and injections $\N\to K$ and $K\to\R$. But the most important part,
-that is, that there are no surjections
-
-\begin{enumerate}
-\item Jaber et al.\ \cite{jaber:hal-01319066}.
-\end{enumerate}
-%We continue with a brief description of the Isabelle framework.
+and injections $\N\hookrightarrow K$ and $K\hookrightarrow\R$. But the
+most important part, that is, that there are no surjections 
+$\N\twoheadrightarrow K$ and $K\twoheadrightarrow\R$, is left for a future
+work. % Finally,  Jaber et al.\ \cite{jaber:hal-01319066}
 

### 2018-08-18 LSFA/related.tex
mini typo

@@
 direction. Nevertheless, a majority of results in
-Isabelle/ZF are not yet implemented using this tool, and a another
+Isabelle/ZF are not yet implemented using this tool, and  another
 downside is that proofs using it do not follow the standard Isar

### 2018-08-20 LSFA/rs.tex
variable ligada distinta de la libre
Cambios menores de redacción

@@
 already been formalized by Krzysztof
-Grabczewski~\cite{DBLP:journals/jar/PaulsonG96}, as part of a
-study of equivalents to $\AC$ (following Rubin and Rubin
+Grabczewski~\cite{DBLP:journals/jar/PaulsonG96}, as part of a study of
+equivalents of $\AC$ (following Rubin and Rubin
 \cite{rubin1985equivalents}). Nevertheless, those are not convenient
-for our purposes for reasons we now explain. The axiom $\DC0$
-corresponds essentially to our Pointed $\DC$ but without the
-constraint $f(0)=a$; it is a nice exercise to show that $\DC0$ implies
-Pointed $\DC$, but a formalization would have a moderate
-length. $\DC(\kappa)$ is rather different in nature and it is tailored
-to  obtain another proposition
-equivalent to the axiom of choice (actually,
-$\AC\longleftrightarrow (\forall \kappa
-.\; \mathrm{Card}(\kappa)\longrightarrow \DC(\kappa))$). 
-Finally, the shortest path from $\AC$ %% as stated in the 
+for our purposes. In fact, the axiom $\DC0$ corresponds essentially to
+our Pointed $\DC$ but without the constraint $f(0)=a$; it is a nice
+exercise to show that $\DC0$ implies Pointed $\DC$, but a
+formalization would have a moderate length. On the other hand,
+$\DC(\kappa)$ is rather different in nature and it is tailored to
+obtain another proposition equivalent to the axiom of choice
+(actually,
+$\AC\longleftrightarrow (\forall \kappa .\;
+\mathrm{Card}(\kappa)\longrightarrow \DC(\kappa))$).  Finally, the
+shortest path from $\AC$ %% as stated in the
 %% corresponding Isabelle theory
@@
 
-The function \isatt{dc{\isacharunderscore}witness} has, apart from $A$, $a$ and $R$, a
-function $s$ as a parameter. If this function is a selector for
-$\P(A)\sm \{\mty\}$, the function $f(n)\defi {}$\isatt{dc{\isacharunderscore}witness}$(n,A,a,s,R)$
-will satify $\DC$. Notice that $s$ is a term of type \isatt{i} (a
-function construed as a set of pairs) and an expression
+Besides the natural argument and the parameters $A$, $a$, and $R$, the
+function \isatt{dc{\isacharunderscore}witness} has a function $s$ as a
+parameter. If this function is a selector for $\P(A)\sm \{\mty\}$, the
+function
+$f(n)\defi {}$\isatt{dc{\isacharunderscore}witness}$(n,A,a,s,R)$ will
+satify $\DC$. Notice that $s$ is a term of type \isatt{i} (a function
+construed as a set of pairs) and an expression
 \isatt{s{\isacharbackquote}b} is notation for \isatt{apply(s,b)},
-where
-\isatt{apply\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i\ {\isasymRightarrow}\ i\ {\isasymRightarrow}\ i{\isachardoublequoteclose}}
-is the operation of function application.
+where \isatt{apply\ {\isacharcolon}{\isacharcolon}\
+  {\isachardoublequoteopen}i\ {\isasymRightarrow}\ i\
+  {\isasymRightarrow}\ i{\isachardoublequoteclose}} is the operation
+of function application.
 % We will not make a distinction in our prose and simply write
@@
 \ aux{\isacharunderscore}sequence{\isacharunderscore}DC\ {\isacharcolon}\ {\isachardoublequoteopen}{\isasymforall}x{\isasymin}A{\isachardot}\ {\isasymforall}n{\isasymin}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ S{\isacharbackquote}n\ {\isasymLongrightarrow}\isanewline
-\ \ \ \ \ \ \ \ {\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}{\isasymlangle}y{\isacharcomma}succ{\isacharparenleft}snd{\isacharparenleft}x{\isacharparenright}{\isacharparenright}{\isasymrangle}{\isasymrangle}\ {\isasymin}\ {\isacharbraceleft}{\isasymlangle}{\isasymlangle}x{\isacharcomma}n{\isasymrangle}{\isacharcomma}{\isasymlangle}y{\isacharcomma}m{\isasymrangle}{\isasymrangle}{\isasymin}{\isacharparenleft}A{\isasymtimes}nat{\isacharparenright}{\isasymtimes}{\isacharparenleft}A{\isasymtimes}nat{\isacharparenright}{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}{\isasymin}S{\isacharbackquote}m\ {\isacharbraceright}{\isachardoublequoteclose}\isanewline
+ \ \ {\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}{\isasymlangle}y{\isacharcomma}succ{\isacharparenleft}snd{\isacharparenleft}x{\isacharparenright}{\isacharparenright}{\isasymrangle}{\isasymrangle}\ {\isasymin}\ {\isacharbraceleft}{\isasymlangle}{\isasymlangle}w{\isacharcomma}n{\isasymrangle}{\isacharcomma}{\isasymlangle}y{\isacharcomma}m{\isasymrangle}{\isasymrangle}{\isasymin}{\isacharparenleft}A{\isasymtimes}nat{\isacharparenright}{\isasymtimes}{\isacharparenleft}A{\isasymtimes}nat{\isacharparenright}{\isachardot}\ {\isasymlangle}w{\isacharcomma}y{\isasymrangle}{\isasymin}S{\isacharbackquote}m\ {\isacharbraceright}{\isachardoublequoteclose}\isanewline
 \ \ %

### 2018-08-24 blueprint.tex
typos y adecuación con la notación ya implementada

@@
     \[
-    \val(\tau,\PP,G)\defi \{\val(\sigma,\PP,G) : \exists p\in \PP
+    \val(\PP,G,\tau)\defi \{\val(\PP,G,\sigma) : \exists p\in \PP
     (\lb\sigma,p\rb\in\tau \y p\in G)\}.
@@
 The definition of $\val$ proceeds by well-founded recursion on
-\verb|eee|$(M)$. Determining  $\val(\tau,\PP,G)$ uses the values
-$\val(\sigma,\PP,G)$ for smaller $\sigma$:
+\verb|eee|$(M)$. Determining  $\val(\PP,G,\tau)$ uses the values
+$\val(\PP,G,\sigma)$ for smaller $\sigma$:
 \[
@@
   \[
-  \verb|gen_ext|(M,\PP,G) = M[G] \defi \{\val(\tau,\PP,G): \tau\in M\}.
+  \verb|gen_ext|(M,\PP,G) = M[G] \defi \{\val(\PP,G,\tau): \tau\in M\}.
   \]
@@
   \begin{enumerate}
-  \item     $M[G]\models\phi(\val(\tau_1,G),\dots,\val(\tau_n,G))$ for
+  \item     $M[G]\models\phi(\val(G,\tau_1),\dots,\val(G,\tau_n))$ for
     all filters $G$, $M$-generic for $\PP$, such that $p\in G$.
@@
 \begin{gather*}
-  \forall G : p\in G \limp M[G],[\val(\tau_1,G),\dots,\val(\tau_n,G)]
+  \forall G : p\in G \limp M[G],[\val(G,\tau_1),\dots,\val(G,\tau_n)]
   \models\phi(x_1,\dots,x_n) \\
@@
     \begin{equation} \label{eq:MG-sats-atom}
-      M[G],[\val(\tau_1,G),\val(\tau_2,G)]
+      M[G],[\val(G,\tau_1),\val(G,\tau_2)]
       \models\phi(x_1,x_2).
@@
   $\phi$-instance of Separation. Let $\pi,\sig\in M$ such that 
-  $\val(\pi)=c$ and $\val(\sig)=w$.
+  $\val(G,\pi)=c$ and $\val(G,\sig)=w$.
   We will show that the name
@@
   \]
-  By the Truth Lemma, this is equivalent to
+  By the Definition of Forcing, this is equivalent to
   \[
   \forall F\ \PP\text{-generic}, p\in F \limp
-  M[F],[\val(\th,F),\val(\pi,F),\val(\sig,F)]\models
+  M[F],[\val(F,\th),\val(F,\pi),\val(F,\sig)]\models
   \dB{0}\in\dB{1}  \y\phi(\dB{0},\dB{1},\dB{2}).
@@
 
-  Now we calculate $\val(\tau,F)$.
+  Now we calculate $\val(G,\tau)$.
   \begin{multline*}
-    \val(\tau,G) = \\
+    \val(G,\tau) = \\
     \val\bigl(\{x\in\dom(\pi)\times\PP : \exists \th,p : x =\lb\th,p\rb
-    \y  \\
-    \forall F\ \PP\text{-generic}, p\in F \limp
-    M[F],[\val(\th,F),\val(\pi,F),\val(\sig,F)]\models
+    \y \forall F\ \PP\text{-generic}, p\in F \limp \\    
+    M[F],[\val(F,\th),\val(F,\pi),\val(F,\sig)]\models
     \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2})\}\bigr).
@@
   \begin{multline}\label{eq:val_tau}
-    \val(\tau,G) = 
-    \{\val(\th,G): \th\in\dom(\pi) \y  p \in G 
+    \val(G,\tau) = 
+    \{\val(G,\th): \th\in\dom(\pi) \y  p \in G 
     \y \forall F\ \PP\text{-generic}, \\
    p\in F \longrightarrow
-  M[F],[\val(\th,F),\val(\pi,F),\val(\sig,F)]\models
+  M[F],[\val(F,\th),\val(F,\pi),\val(F,\sig)]\models
   \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2})\}.
@@
   \begin{multline*}
-    \{\val(\th,G): \th\in\dom(\pi) \y  p \in G 
+    \{\val(G,\th): \th\in\dom(\pi) \y  p \in G 
     \y  \\
    p\in G \longrightarrow
-   M[G],[\val(\th,G),\val(\pi,G),\val(\sig,G)]\models
+   M[G],[\val(G,\th),\val(G,\pi),\val(G,\sig)]\models
    \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2})\},
   \end{multline*}
-  and in turn,  replacing $\val(\pi,F),\val(\sig,F)$ by their values,
+  and in turn,  replacing $\val(G,\pi),\val(G,\sig)$ by their values,
   this is included  in 
+  \begin{equation}\label{eq:2}
+    \{\val(G,\th): \th\in\dom(\pi)
+    \y 
+   M[G],[\val(G,\th),c,w]\models
+   \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2})\}.
+  \end{equation}
+  By absoluteness, we have that (\ref{eq:2}) is the same as
   \[
-    \{\val(\th,G): \th\in\dom(\pi)
-    \y 
-   M[G],[\val(\th,G),c,w]\models
-   \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2})\}.
-  \]
-  By absoluteness, we have that $\val(\tau,G)$ is
-  \[
-    \{\val(\th,G): \th\in\dom(\pi)
-    \y \val(\th,G)\in c \y 
-   M[G],[\val(\th,G),c,w]\models
+    \{\val(G,\th): \th\in\dom(\pi)
+    \y \val(G,\th)\in c \y 
+   M[G],[\val(G,\th),c,w]\models
    \phi(\dB{0},\dB{1},\dB{2})\}.
@@
   \]
-  since for every $x\in c=\val(\pi,G)$ there must
-  exist some $\th\in\dom(\pi)$ with $\val(\th,G)=x$. Therefore
-  $\val(\tau,G)\subseteq S$.
+  since for every $x\in c=\val(G,\pi)$ there must
+  exist some $\th\in\dom(\pi)$ with $\val(G,\th)=x$. Therefore
+  $\val(G,\tau)\subseteq S$.
 
   Now let $x\in S$. Hence there exists $\lb\th,q\rb\in\pi$ such that
-  $q\in G$ and $x=\val(\th,G)$. By the second version of the Truth
+  $q\in G$ and $x=\val(G,\th)$. By the  Truth
   Lemma, since 
   \[
-  M[G],[\val(\th,G),\val(\pi,G),\val(\sig,G)]\models
+  M[G],[\val(G,\th),\val(G,\pi),\val(G,\sig)]\models
   \dB{0}\in\dB{1}\y\phi(\dB{0},\dB{1},\dB{2}),
   \]
-  there must exists $r\in G$ such that 
+  there must exist $r\in G$ such that 
   \[
@@
    p\in F \longrightarrow
-  M[F],[\val(\th,F),\val(\pi,F),\val(\sig,F)]\models
+  M[F],[\val(F,\th),\val(F,\pi),\val(F,\sig)]\models
   \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2}).
   \]
-  Using now (\ref{eq:val_tau}), this shows that $S\sbq \val(\tau,G)$,
+  Using now (\ref{eq:val_tau}), this shows that $S\sbq \val(G,\tau)$,
   since $\th\in\dom(\pi)$. 

### 2018-11-07 LSFA/ENTCS/conclusion.tex
Copiando los sources

--- 2018-06-24 LSFA/conclusion.tex
@@
 \section{Conclusions and future work}
-There are several technical milestones that have to be traversed in the
-course of a formalization of the theory of forcing. The first and
-obvious one is the bulk of set- and meta-theoretical concepts needed to work
+There are several technical milestones that have to be reached in the
+course of a formalization of the theory of forcing. The first one, and most
+obvious, is the bulk of set- and meta-theoretical concepts needed to work
 with. This pushed us, in a sense,  into building on top of Isabelle/ZF,
 since we know of no other development in set theory of such
-depth. In this paper we worked on setting the stage for the work with
+depth (and breadth). In this paper we worked on setting the stage for the work with
 generic extensions; in particular, this involves some purely mathematical
@@
 Theorem nor on the definition of the forcing relation. Our next task
-will be to obtain the last goal enumerated.
+will be to obtain the last goal in that enumeration.
 
 To this end, we will develop an interface between Paulson's
-relativization results and countable models $M$ of $\ZFC$. This will show
-that every such $M$ is closed under well-founded recursion and, in
+relativization results and countable models of $\ZFC$. This will show
+that every ctm $M$ is closed under well-founded recursion and, in
 particular, that contains names for each of its
@@
 nontrivially). As a part of the new formalization, we will provide
-Isar versions of the longer applicative proofs presented in this work
+Isar versions of the longer applicative proofs presented in this work.
 

### 2018-11-07 LSFA/ENTCS/intro.tex
Copiando los sources

--- 2018-08-10 LSFA/intro.tex
@@
 The Rasiowa-Sikorski lemma (RSL) states that for any preorder $\PP$
-and any countable family $\{\mathcal{D}_n : n\in\mathbb{N}\}$ of dense
+and any countable family $\mathcal{D}_{n\in\mathbb{N}}$ of dense
 subsets of $\PP$ there is a filter intersecting every

### 2018-11-07 LSFA/ENTCS/pair-gen-ext.tex
Copiando los sources

--- 2018-07-12 LSFA/arXiv/pair-gen-ext.tex
@@
 \ \ upair\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharbrackleft}i{\isasymRightarrow}o{\isacharcomma}i{\isacharcomma}i{\isacharcomma}i{\isacharbrackright}\ {\isasymRightarrow}\ o{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
- {\isachardoublequoteopen}upair{\isacharparenleft}C{\isacharcomma}a{\isacharcomma}b{\isacharcomma}z{\isacharparenright}\ {\isacharequal}{\isacharequal}\ a\ {\isasymin}\ z\ {\isasymand}\ b\ {\isasymin}\ z\ {\isasymand}\ {\isacharparenleft}{\isasymforall}x{\isacharbrackleft}C{\isacharbrackright}{\isachardot}\ x{\isasymin}z\ {\isasymlongrightarrow}\ x\ {\isacharequal}\ a\ {\isasymor}\ x\ {\isacharequal}\ b{\isacharparenright}{\isachardoublequoteclose}
+\ \ \ \ {\isachardoublequoteopen}upair{\isacharparenleft}C{\isacharcomma}a{\isacharcomma}b{\isacharcomma}z{\isacharparenright}\ {\isacharequal}{\isacharequal}\ a\ {\isasymin}\ z\ {\isasymand}\ b\ {\isasymin}\ z\ {\isasymand}\ {\isacharparenleft}{\isasymforall}x{\isacharbrackleft}C{\isacharbrackright}{\isachardot}\ x{\isasymin}z\ {\isasymlongrightarrow}\ x\ {\isacharequal}\ a\ {\isasymor}\ x\ {\isacharequal}\ b{\isacharparenright}{\isachardoublequoteclose}
 \end{isabelle}
@@
 \ \ upair{\isacharunderscore}ax\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharparenleft}i{\isasymRightarrow}o{\isacharparenright}\ {\isasymRightarrow}\ o{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
- {\isachardoublequoteopen}upair{\isacharunderscore}ax{\isacharparenleft}C{\isacharparenright}\ {\isacharequal}{\isacharequal}\ {\isasymforall}x{\isacharbrackleft}C{\isacharbrackright}{\isachardot}\ {\isasymforall}y{\isacharbrackleft}C{\isacharbrackright}{\isachardot}\ {\isasymexists}z{\isacharbrackleft}C{\isacharbrackright}{\isachardot}\ upair{\isacharparenleft}C{\isacharcomma}x{\isacharcomma}y{\isacharcomma}z{\isacharparenright}{\isachardoublequoteclose}
+\ \ \ \ {\isachardoublequoteopen}upair{\isacharunderscore}ax{\isacharparenleft}C{\isacharparenright}\ {\isacharequal}{\isacharequal}\ {\isasymforall}x{\isacharbrackleft}C{\isacharbrackright}{\isachardot}\ {\isasymforall}y{\isacharbrackleft}C{\isacharbrackright}{\isachardot}\ {\isasymexists}z{\isacharbrackleft}C{\isacharbrackright}{\isachardot}\ upair{\isacharparenleft}C{\isacharcomma}x{\isacharcomma}y{\isacharcomma}z{\isacharparenright}{\isachardoublequoteclose}
 \end{isabelle}
@@
 use the \isatt{\#\#}  operator.  We state the main result of this
-Section in the context \isatt{forcing{\isacharunderscore}data}.  
+section in the context \isatt{forcing{\isacharunderscore}data}.  
 %
@@
 \ sigma{\isacharunderscore}in{\isacharunderscore}M\ {\isacharcolon}\isanewline
-\ \ {\isachardoublequoteopen}upair{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}\ {\isasymLongrightarrow}\ one\ {\isasymin}\ G\ {\isasymLongrightarrow}\ {\isasymtau}\ {\isasymin}\ M\ {\isasymLongrightarrow}\ {\isasymrho}\ {\isasymin}\ M\ {\isasymLongrightarrow}\isanewline \ \  \ {\isacharbraceleft}{\isasymlangle}{\isasymtau}{\isacharcomma}one{\isasymrangle}{\isacharcomma}{\isasymlangle}{\isasymrho}{\isacharcomma}one{\isasymrangle}{\isacharbraceright}\ {\isasymin}\ M{\isachardoublequoteclose}
+\ \ {\isachardoublequoteopen}upair{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}\ {\isasymLongrightarrow}\ one\ {\isasymin}\ G\ {\isasymLongrightarrow}\ {\isasymtau}\ {\isasymin}\ M\ {\isasymLongrightarrow}\ {\isasymrho}\ {\isasymin}\ M\ {\isasymLongrightarrow}\ {\isacharbraceleft}{\isasymlangle}{\isasymtau}{\isacharcomma}one{\isasymrangle}{\isacharcomma}{\isasymlangle}{\isasymrho}{\isacharcomma}one{\isasymrangle}{\isacharbraceright}\ {\isasymin}\ M{\isachardoublequoteclose}
 

### 2018-11-07 LSFA/ENTCS/rs.tex
Copiando los sources

--- 2018-08-15 LSFA/rs.tex
@@
-\subsection{A sequence version of Dependent
-  Choices}\label{sec:sequence-version-dc} 
-%% In order to prove the Rasiowa-Sikorski lemma in
-%% general, a consequence of the Axiom of Choice ($\AC$) must be invoked. This
-%% consequence is known as the \emph{Principle of Dependent Choices}:
-The Rasiowa-Sikorski lemma follows naturally from a ``pointed''
-version of   the \emph{Principle of Dependent Choices} which,
-in turn, is a consequence of the Axiom of Choice ($\AC$). It is
-therefore natural to take as a starting point the theory \isatt{AC}
-which adds the latter axiom to the toolkit of Isabelle/ZF.
-
-The statement we are interested in is the following:
+\subsection{A sequence version of $\DC$}\label{sec:sequence-version-dc}
+In order to prove the Rasiowa-Sikorski lemma in
+general, a version of the Axiom of Choice ($\AC$) must be proved. This
+version is known as the \emph{Principle of Dependent Choices}:
 \begin{quote}
-  (Pointed $\DC$) Let $\rr$ be a binary relation on $A$, and $a\in A$. If
+  ($\DC$) Let $\rr$ be a binary relation on $A$, and $a\in A$. If
   $\forall x\in A.\,  \exists y\in A.\, x\rr y$, then there exists
@@
 
-Two different versions of $\DC$ (called $\DC0$ and $\DC(\kappa)$) have
-already been formalized  by Krzysztof 
-Grabczewski~\cite{DBLP:journals/jar/PaulsonG96}, as part of a
-study of equivalents to $\AC$ (following Rubin and Rubin
-\cite{rubin1985equivalents}). Nevertheless, those are not convenient
-for our purposes for reasons we now explain. The axiom $\DC0$
-corresponds essentially to our Pointed $\DC$ but without the
-constraint $f(0)=a$; it is a nice exercise to show that $\DC0$ implies
-Pointed $\DC$, but a formalization would have a moderate
-length. $\DC(\kappa)$ is rather different in nature and it is tailored
-to  obtain another proposition
-equivalent to the axiom of choice (actually,
-$\AC\longleftrightarrow (\forall \kappa
-.\; \mathrm{Card}(\kappa)\longrightarrow \DC(\kappa))$. 
-Finally, the shortest path from $\AC$ %% as stated in the 
-%% corresponding Isabelle theory
-to $\DC0$ using already formalized
-material involves a complicated detour (130+ proof commands spanning
-various files of the \isatt{ZF-AC} theory and going through the Well
-Ordering Theorem and  $\DC(\omega)$), compared to the mere 11 commands from $\AC$ to
-\isatt{AC{\isacharunderscore}func{\isacharunderscore}Pow}. This last
-one is the
-choice principle that we use in our formalization of Pointed $\DC$, and states 
-the existence of
+A different version of $\DC$ (without the constraint $f(0)=a$, but
+allowing a parameter that permits to prove an equivalence with $\AC$)
+was formalized by Krzysztof
+Grabczewski~\cite{DBLP:journals/jar/PaulsonG96} as part of the
+Isabelle/ZF library, along with many statements that follow from or
+are equivalent to $\AC$. Since the proofs there required a greater
+generality, it would have been very contrived to derive this
+``pointed'' version of $\DC$ from that development. Instead, we
+preferred to give a direct proof based on the one in
+Moschovakis' notes \cite{moschovakis1994notes}. Another advantage is that
+this proof depends on the following version of $\AC$ (existence of
 choice functions on $\P(A)\sm \{\mty\}$):
@@
 \]
-Another advantage of taking
-\isatt{AC{\isacharunderscore}func{\isacharunderscore}Pow} as a
-starting point is that it does not involve proper classes: The version
-of  $\AC$ in Isabelle/ZF corresponds to an axiom scheme of first order
-logic and as such is not a standard formulation. 
-%% 26 commands to get Pointed $\DC$.
+This statement is more standard than $\AC$ as stated in the
+Isabelle/ZF development, which involves a silent use of Replacement
+since it is written as an axiom scheme.
 
-The strategy to prove Pointed $\DC$ (following a proof in Moschovakis
-\cite{moschovakis1994notes}) is to define the function $f$ discussed above by
+The strategy is to define the function $f$ discussed above by
 primitive recursion on the naturals, which can be done easily thanks
@@
 
-The proof is mostly routine; after a few lemmas (26 proof
-commands in total) we obtain the
+The proof is mostly routine; after a few lemmas we obtain the
 following theorem:
@@
 \begin{isabelle}
-\isacommand{corollary}\isamarkupfalse%
-\ DC{\isacharunderscore}on{\isacharunderscore}A{\isacharunderscore}x{\isacharunderscore}nat\ {\isacharcolon}\ \isanewline
-\ \ {\isachardoublequoteopen}{\isacharparenleft}{\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}{\isasymlangle}y{\isacharcomma}succ{\isacharparenleft}snd{\isacharparenleft}x{\isacharparenright}{\isacharparenright}{\isasymrangle}{\isasymrangle}\ {\isasymin}\ R{\isacharparenright}\ {\isasymLongrightarrow}\isanewline
-\ \ \ \ {\isasymforall}a{\isasymin}A{\isachardot}\ {\isacharparenleft}{\isasymexists}f\ {\isasymin}\ nat{\isasymrightarrow}A{\isachardot}\ f{\isacharbackquote}{\isadigit{0}}\ {\isacharequal}\ a\ {\isasymand}\ {\isacharparenleft}{\isasymforall}n\ {\isasymin}\ nat{\isachardot}\ {\isasymlangle}{\isasymlangle}f{\isacharbackquote}n{\isacharcomma}n{\isasymrangle}{\isacharcomma}{\isasymlangle}f{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isacharcomma}succ{\isacharparenleft}n{\isacharparenright}{\isasymrangle}{\isasymrangle}{\isasymin}R{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\isacommand{lemma}\isamarkupfalse%
+\ aux{\isacharunderscore}DC{\isacharunderscore}on{\isacharunderscore}AxNat{\isadigit{2}}\ {\isacharcolon}\ {\isachardoublequoteopen}{\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}{\isasymlangle}y{\isacharcomma}succ{\isacharparenleft}snd{\isacharparenleft}x{\isacharparenright}{\isacharparenright}{\isasymrangle}{\isasymrangle}\ {\isasymin}\ R\ {\isasymLongrightarrow}\isanewline
+\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ {\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ {\isacharbraceleft}{\isasymlangle}a{\isacharcomma}b{\isasymrangle}{\isasymin}R{\isachardot}\ snd{\isacharparenleft}b{\isacharparenright}\ {\isacharequal}\ succ{\isacharparenleft}snd{\isacharparenleft}a{\isacharparenright}{\isacharparenright}{\isacharbraceright}{\isachardoublequoteclose}
 \end{isabelle}
-
 %
@@
 \isacommand{lemma}\isamarkupfalse%
-\ aux{\isacharunderscore}sequence{\isacharunderscore}DC\ {\isacharcolon}\ {\isachardoublequoteopen}{\isasymforall}x{\isasymin}A{\isachardot}\ {\isasymforall}n{\isasymin}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ S{\isacharbackquote}n\ {\isasymLongrightarrow}\isanewline
+\ aux{\isacharunderscore}sequence{\isacharunderscore}DC{\isadigit{2}}\ {\isacharcolon}\ {\isachardoublequoteopen}{\isasymforall}x{\isasymin}A{\isachardot}\ {\isasymforall}n{\isasymin}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ S{\isacharbackquote}n\ {\isasymLongrightarrow}\isanewline
 \ \ \ \ \ \ \ \ {\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}{\isasymlangle}y{\isacharcomma}succ{\isacharparenleft}snd{\isacharparenleft}x{\isacharparenright}{\isacharparenright}{\isasymrangle}{\isasymrangle}\ {\isasymin}\ {\isacharbraceleft}{\isasymlangle}{\isasymlangle}x{\isacharcomma}n{\isasymrangle}{\isacharcomma}{\isasymlangle}y{\isacharcomma}m{\isasymrangle}{\isasymrangle}{\isasymin}{\isacharparenleft}A{\isasymtimes}nat{\isacharparenright}{\isasymtimes}{\isacharparenleft}A{\isasymtimes}nat{\isacharparenright}{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}{\isasymin}S{\isacharbackquote}m\ {\isacharbraceright}{\isachardoublequoteclose}\isanewline

### 2018-11-16 LSFA/intro.tex
seguimos a Kunen
correcciones viejas de nuevo

@@
 Isabelle~\cite{Isabelle} by Lawrence Paulson~\cite{paulson_2003}. This paper formalizes
-a first part of the machinery of forcing. In the rest of the
+a first part of the machinery of forcing, mostly by following the new
+edition of the classical book on the subject by Kunen \cite{kunen2011set}. In the rest of the
 introduction we discuss some of the set-theoretical details involved
@@
 %% que N satisface un axioma, vamos a utilizar que M lo satisface.
-It is therefore a primary need to have a means to correlate  (first-order)
+It is therefore a primary need to have means to correlate  (first-order)
 properties satisfied by a model $M$ and those of a 

### 2018-11-16 LSFA/intro.tex
traigo todos los archivos de master, donde está la ultima versión enviada

@@
 particular showed the fundamental result that the real line, $\R$  is not
-coordinable with the natural numbers. Soon after this, he posed the
+equipotent to the natural numbers. Soon after this, he posed the
 most important question in the field, written as a conjecture:
@@
   The \emph{Continuum Hypothesis} ($\CH$). Every uncountable subset of $\R$ is
-  coordinable with $\R$.
+  equipotent to $\R$.
 \end{quote}
@@
      and  $E$ is the membership relation $\in$ restricted to
-    $M$. It is \emph{countable} if $M$ is coordinable with a subset of
+    $M$. It is \emph{countable} if $M$ is equipotent to a subset of
     $\N$; we then say that the model  $M$ is a \emph{ctm}. 

### 2018-11-16 LSFA/isabelle-zf.tex
traigo todos los archivos de master, donde está la ultima versión enviada

@@
 \isatt{\ensuremath{\in}} and several set constructors (terms and
-function with values in \isatt{i}) corresponding to the empty set (the
+functions with values in \isatt{i}) corresponding to the empty set (the
 constant \isatt{\isadigit{0}}), powersets, and one further constant

### 2018-11-16 LSFA/related.tex
traigo todos los archivos de master, donde está la ultima versión enviada

@@
 \subsection{Related work}\label{sec:related-work}
-Before we proceed, we will discuss very succinctly recent related
-formalizations.
+Before we proceed, we discuss very succinctly recent formalizations of
+set theory and forcing. The closest formalizations are those based on
+Isabelle. Let us remark that Isabelle allows for different logical
+foundations; in particular, Paulson carried out his formalizations
+on top of Isabelle/FOL which is based on first order logic.
 
@@
 
-There are some forcing-related formalizations in the theory
-of types. Most relevant is the work by K.~Quirin \cite{Quirin}, where
-a sheaf-theoretic initial approach to forcing is implemented in
-Coq. This language is extremely different to the standard approach of
-constructing models of $\ZFC$, and it might be difficult (once the
-forcing machinery is set) to translate results in the literature using
-ctms to this one. In any case, the translation to set theory of what
-Quirin accomplishes is to define a generic extension (where $\CH$
-should fail) and to construct a set $K$ (a candidate counterexample)
-and injections $\N\hookrightarrow K$ and $K\hookrightarrow\R$. But the
-most important part, that is, that there are no surjections 
-$\N\twoheadrightarrow K$ and $K\twoheadrightarrow\R$, is left for a future
+As far as we know, there is little progress on formalizations of
+forcing in type theory. Most relevant is the work by K.~Quirin
+\cite{Quirin}, where a sheaf-theoretic initial approach to forcing is
+implemented in Coq. This language is extremely different to the
+standard approach of constructing models of $\ZFC$, and it might be
+difficult (once the forcing machinery is set) to translate results in
+the literature using ctms to this one. In any case, the translation to
+set theory of what Quirin accomplishes is to define a generic
+extension (where $\CH$ should fail) and to construct a set $K$ (a
+candidate counterexample) and injections $\N\hookrightarrow K$ and
+$K\hookrightarrow\R$. But the most important part, that is, that there
+are no surjections $\N\twoheadrightarrow K$ and
+$K\twoheadrightarrow\R$, is left for a future
 work. % Finally, Jaber et al.\ \cite{jaber:hal-01319066}
+% port forcing to type theory; notice that this is not a formalization
+% of forcing. Instead, they construct new logics from type theory
+% by means of forcing conditions.
 

### 2018-11-16 LSFA/isabelle-zf.tex

@@
 %-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
-\section{Isabelle/ZF}
+\section{Isabelle/ZF}\label{sec:isabellezf}
 
@@
 
-% locales
+We make a final comment about the implementation tools provided by the
+Isabelle framework.
+
+An IDE called \texttt{jEdit} is bundled with the standard Isabelle
+distribution and it offers the user a fair amount of tools in order to
+manage theory files, searching for theorems and concepts spread
+through the source files, and includes tracing utilities for the
+automatic tools. A main feature is a window showing the \emph{proof
+  state}, where the active (sub)goals are shown, along with the already
+obtained results and possibly errors.
+
+Isabelle proofs can be written in two dialects. 
+The most basic one and older (in which most of the original
+work by Paulson is written) follows a procedural approach, where one
+applies several tactics in order to decompose the goal into simpler
+ones and then solving them (with the aid of automation). This approach
+has the drawback that a proof akin to mathematical writing can't be
+read out from the code, since the intermediate steps are only shown in
+the proof state. For this reason, the proof language
+\emph{Isar} was developed, starting with Wenzel's
+work~\cite{DBLP:conf/tphol/Wenzel99}. Isar is mostly declarative, and
+its main purpose is to construct \emph{proof documents} that (in
+principle) can be read and understood without the need of running the
+code.
+
+We started this development by using the procedural approach, but soon
+we realized that for our purposes the Isar language was far more
+appropriate.
 

### 2018-11-16 LSFA/related.tex

@@
 downside is that proofs using it do not follow the standard Isar
-language.
+language (see Section~\ref{sec:isabellezf}).
 

### 2018-11-16 LSFA/isabelle-zf.tex
Pocos cambios

--- 2018-11-16 LSFA/isabelle-zf.tex
@@
 
-% locales
+Let us close this section with a brief comment about the facilities
+provided by the Isabelle framework. The edition is done in an IDE
+called \texttt{jEdit}, which is bundled with the standard Isabelle
+distribution; it offers the user a fair amount of tools in order to
+manage theory files, searching for theorems and concepts spread
+through the source files, and includes tracing utilities for the
+automatic tools. A main feature is a window showing the \emph{proof
+  state}, where the active (sub)goals are shown, along with the
+already obtained results and possibly errors.
+
+Isabelle proofs can be written in two dialects.  The older one, and
+also more basic, follows a procedural approach, where one applies
+several tactics in order to decompose the goal into simpler ones and
+then solving them (with the aid of automation); the original work by
+Paulson used this method. Under this approach proofs are constructed
+bottom-up resulting in proof-scripts that conceal the mathematical
+reasoning behind the proof, since the intermediate steps are only
+shown in the proof state. For this reason, the proof language
+\emph{Isar} was developed, starting with Wenzel's
+work~\cite{DBLP:conf/tphol/Wenzel99}. Isar is mostly declarative, and
+its main purpose is to construct \emph{proof documents} that (in
+principle) can be read and understood without the need of running the
+code.
+
+We started this development using the procedural approach, but soon
+after we realized that for our purposes the Isar language was far more
+appropriate.
 

### 2018-11-17 LSFA/related.tex
formalización de matemática en general

@@
 \subsection{Related work}\label{sec:related-work}
-Before we proceed, we discuss very succinctly recent formalizations of
+Formalization of mathematics serves many purposes. The most obvious
+one is to increase reliability in a result and/or its proof. This has
+been the original motivation that lead Voevodsky to gather many
+researchers around \emph{homotopy type theory} and its formalization in Coq
+\cite{hottbook}; the same applies to the four colour theorem (checked by
+Gonthier \cite{MR2463991}) and the formidable \emph{Flyspeck} project
+\cite{MR3659768} by the team conducted by Hales.
+
+In our particular case, forcing and the set theoretic techniques that
+are being formalized can be regarded as a mature technology and thus
+the main goal is not to increase confidence. Nevertheless, the level
+of detail in a formalization of this sort always provide additional
+information about the inner workings of the theory: It is expected, for
+instance, to have a detailed account of which axioms are necessary to
+define and use forcing. Finally, we support the vision that a growing
+corpus of formalized mathematics can be a useful library for the
+future generations. The question of how to systematize this corpus is
+an ongoing project by Paulson \cite{ALEXANDRIA}.
+
+We will now discuss very succinctly recent formalizations of
 set theory and forcing. The closest formalizations are those based on

### 2018-11-17 LSFA/related.tex
first-order

@@
 foundations; in particular, Paulson carried out his formalizations
-on top of Isabelle/FOL which is based on first order logic.
+on top of Isabelle/FOL which is based on first-order logic.
 

### 2018-11-17 LSFA/rs.tex
first-order

@@
 starting point is that it does not involve proper classes: The version
-of  $\AC$ in Isabelle/ZF corresponds to an axiom scheme of first order
+of  $\AC$ in Isabelle/ZF corresponds to an axiom scheme of first-order
 logic and as such is not a standard formulation. 

### 2018-11-17 LSFA/isabelle-zf.tex
first-order
ronda de correcciones menores

@@
 %% Isabelle/ZF is a theory built upon the
-%% core library FOL of classical first order logic.
-Isabelle/ZF includes a development of classical first order logic,
+%% core library FOL of classical first-order logic.
+Isabelle/ZF includes a development of classical first-order logic,
 FOL. Both of them are  built upon the core library \emph{Pure}.
@@
 \emph{locales}~\cite{ballarin2010tutorial}.
-Locales can be combined and extended adding more parameters and assuming
+Locales can be combined and extended by adding more parameters and assuming
 more facts, leading to a new locale. For example a context describing

### 2018-11-17 LSFA/names.tex
ronda de correcciones menores

@@
 
-In our case the functional $H$ takes an additional argument for the
+In our case the functional is called $Hv$ and takes an additional argument for the
 parameter $G$:
 \[
-  H(G,y,f) = \{f(x) : x \in dom(y) \y \exists p\in \PP.\,
+  \mathit{Hv}(G,y,f) = \{f(x) : x \in dom(y) \y \exists p\in \PP.\,
 (\lb x,p\rb\in y \y p\in G)\}
@@
 
-In the formalization of $\mathit{ed}$ is restricted to a set,
-so to have a term of type \isa{i}:
+In our formalization of this recursion,  the first argument of
+\isa{wfrec} is the term of type  \isa{i} obtained by restricting the
+relation $\mathit{ed}$ to  a set:
 % 
@@
 \end{isabelle}
-
-\noindent We omit the proof of the following lemma used above:
-
+%
+All but one lemma used in the above proof
+(\isa{wf{\isacharunderscore}subset},
+\isa{wf{\isacharunderscore}trancl}, and
+\isa{wf{\isacharunderscore}Memrel}) are already
+present in Isabelle/ZF. The remaining technical result has
+been proved using the Isar language:
+%
 \begin{isabelle}
@@
 
-\noindent The formalizaton of the functional $H$ is straightforward and $\val$ is defined using \isa{wfrec}:
+The formalizaton of the functional $\mathit{Hv}$ is straightforward and $\val$ is defined using \isa{wfrec}:
 %
@@
 \isanewline
-\ \ GenExt\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}i{\isachardoublequoteclose}\ \ \ \ \ {\isacharparenleft}{\isachardoublequoteopen}M{\isacharbrackleft}{\isacharunderscore}{\isacharbrackright}{\isachardoublequoteclose}\ {\isadigit{9}}{\isadigit{0}}{\isacharparenright}\ \isakeyword{where} \isanewline
+\ \ GenExt\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}i{\isachardoublequoteclose}\ \ \ \ \ {\isacharparenleft}{\isachardoublequoteopen}M{\isacharbrackleft}{\isacharunderscore}{\isacharbrackright}{\isachardoublequoteclose}{\isacharparenright}\ \isakeyword{where} \isanewline
 \ \ {\isachardoublequoteopen}GenExt{\isacharparenleft}G{\isacharparenright}{\isacharequal}{\isacharequal}\ {\isacharbraceleft}val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}{\isachardot}\ {\isasymtau}\ {\isasymin}\ M{\isacharbraceright}{\isachardoublequoteclose}

### 2018-11-17 LSFA/rs.tex
ronda de correcciones menores

@@
 the existence of
-choice functions on $\P(A)\sm \{\mty\}$):
+choice functions (``selectors'') on $\P(A)\sm \{\mty\}$):
 \[
@@
 Rasiowa-Sikorski. That is, if the assumption holds for a sequence of
-relations $S_n$,  then $f(n) \mathrel{S_{n+1}} f(n+1)$.
+relations $S_n$,  then $f(n) \mathrel{S_{n+1}} f(n+1)$ for all $n$.
 

### 2018-11-17 LSFA/rs.tex
comentarios sobre pruebas hasta Pointed_DC inclusive

@@
 We first obtain a corollary of $\DC$ changing $A$ for
-$A\times\mathtt{nat}$:
+$A\times\mathtt{nat}$, whose procedural proof takes 16 lines:
 
@@
 %
-And finally, we arrive to $\DC$ for a sequence of relations.
+And after a short proof we arrive to $\DC$ for a sequence of relations:
 
@@
 \ sequence{\isacharunderscore}DC{\isacharcolon}\ {\isachardoublequoteopen}{\isasymforall}x{\isasymin}A{\isachardot}\ {\isasymforall}n{\isasymin}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ S{\isacharbackquote}n\ {\isasymLongrightarrow}\isanewline
-\ \ \ \ {\isasymforall}a{\isasymin}A{\isachardot}\ {\isacharparenleft}{\isasymexists}f\ {\isasymin}\ nat{\isasymrightarrow}A{\isachardot}\ f{\isacharbackquote}{\isadigit{0}}\ {\isacharequal}\ a\ {\isasymand}\ {\isacharparenleft}{\isasymforall}n\ {\isasymin}\ nat{\isachardot}\ {\isasymlangle}f{\isacharbackquote}n{\isacharcomma}f{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isasymrangle}{\isasymin}S{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\ \ \ \ {\isasymforall}a{\isasymin}A{\isachardot}\ {\isacharparenleft}{\isasymexists}f\ {\isasymin}\ nat{\isasymrightarrow}A{\isachardot}\ f{\isacharbackquote}{\isadigit{0}}\ {\isacharequal}\ a\ {\isasymand}\ {\isacharparenleft}{\isasymforall}n\ {\isasymin}\ nat{\isachardot}\ {\isasymlangle}f{\isacharbackquote}n{\isacharcomma}f{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isasymrangle}{\isasymin}S{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+%
+%\isadelimproof
+\ \ %
+%\endisadelimproof
+%
+%\isatagproof
+\isacommand{apply}\isamarkupfalse%
+\ {\isacharparenleft}drule\ aux{\isacharunderscore}sequence{\isacharunderscore}DC{\isacharparenright}\isanewline
+\ \ \isacommand{apply}\isamarkupfalse%
+\ {\isacharparenleft}drule\ DC{\isacharunderscore}on{\isacharunderscore}A{\isacharunderscore}x{\isacharunderscore}nat{\isacharcomma}\ auto{\isacharparenright}\isanewline
+\ \ \isacommand{done}\isamarkupfalse%
+%
 \end{isabelle}

### 2018-11-17 LSFA/cover_letter.tex
descripcion de cambios

\documentclass[12pt,english]{article}
\usepackage{babel}
%
\input{header-draft}
%
%
\begin{document}
\title{First steps towards a formalization of Forcing
  \\ Revisions made to the paper}
\author{E.~Gunther \and  M.~Pagano \and P.~S\'anchez~Terraf
  %% fourth author was partially
  %% supported by CONICET, ANPCyT project PICT 2012-1823, and by
  %% SeCyT-UNC project 05/B284.}
}
\date{}
\maketitle

In first place we want to thank the referee for reading the paper very
carefully and for his detailed and constructive criticism. Below we
detail the main changes that have been made%; also, a diff-PDF is
%provided that shows all the changes in color
.

We have spelled-check the manuscript, corrected punctuation, and took
notice of all the typos indicated in the report.

The Introduction has many changes. A whole new section (``Related
work'') has been added, where the formalization of mathematics in
general and of forcing and set theory in particular are discussed.
We also remark that our  development follows the
presentation by Kunen \cite{kunen2011set}.

%
%
\bibliographystyle{mi-estilo-else}
\bibliography{citados}
\end{document}
%%% Local Variables: 
%%% mode: latex
%%% ispell-local-dictionary: "american"
%%% End:

### 2018-11-17 LSFA/cover_letter.tex
algunas cosas... ver

@@
-\documentclass[12pt,english]{article}
+\documentclass[11pt,english]{article}
 \usepackage{babel}
+\usepackage[numbers]{natbib}
 %
@@
 
-In first place we want to thank the referee for reading the paper very
+In first place we want to thank the referees for reading the paper very
 carefully and for his detailed and constructive criticism. Below we
@@
 general and of forcing and set theory in particular are discussed.
-We also remark that our  development follows the
-presentation by Kunen \cite{kunen2011set}.
+Also, at the end of the subsection on Isabelle we describe the tools
+available for implementation, and a comparison
+between the two dialects available, the basic procedural style and the
+declarative \emph{Isar}. At some point of the Introduction, we also
+remark that our 
+development follows the presentation by Kunen \cite{kunen2011set}.
 
+We added several comments on the implementation details, especially in
+the section on forcing notions; this includes a description of the
+complexity in number of lines and proof commands, and some hints on
+the mechanics of the use Isabelle.
+
+A thorough comparison between the available results on $\AC$ and
+$\DC$, and the need for reimplementing some of those was included in
+the subsection on Dependent Choices.
+
+The section on ``Names'' includes two new applications. One is the
+proof that the generic
+extension $M[G]$ is a transitive set. The second one is even more basic  and
+was missing in the two previous versions of the manuscript: The
+generic filter $G$ belongs to $M[G]$. 
 %
 %
-\bibliographystyle{mi-estilo-else}
+\bibliographystyle{arXiv/mi-estilo-else}
 \bibliography{citados}

### 2018-11-18 LSFA/cover_letter.tex
recursion en Names
responder al Reviewer 3 sobre el esfuerzo y el carácter del trabajo y por qué no agregamos nada de lo que dice la conclusión.

@@
 
-The section on ``Names'' includes two new applications. One is the
+The section on ``Names'' includes a thorough discussion on recursive
+definitions in Isabelle/ZF and two new applications. One is the
 proof that the generic
@@
 generic filter $G$ belongs to $M[G]$.
+
+We take this opportunity to answer a question by Reviewer 3 from the
+first round of corrections. Obtaining the results of this paper per se
+was not very difficult. But as this Reviewer points out, the setting
+up for the project, planning a strategy, and even studying Paulson's
+libraries%
+%%%%%%%%%%%%%%%%%%%%
+\footnote{Even today we find ourselves inadvertently reproving some
+  variant of a lemma in Paulson's development on set theory.}
+%%%%%%%%%%%%%%%%%%%%
+took almost an year of work. With the hindsight of this planning period,
+we would like to point out that most of the results stated in the
+conclusion section involve some serious development; even  a
+publication for each item. It is therefore not feasible to include
+none of them in this revision.
 %

### 2018-11-18 LSFA/hacer_diff.sh
diff desde la versión vieja

@@ -1,0 +1,23 @@
+#!/bin/bash
+
+mkdir diff_dir
+rm -rf diff_dir/*
+mkdir diff_dir/LSFA_old/ diff_dir/LSFA_new/
+git checkout 80e6d1835fc9da4394d0c805d70c3032f6ba297d
+cp * diff_dir/LSFA_old/
+git checkout new_design
+cp * diff_dir/LSFA_new/
+cd diff_dir
+latexdiff --flatten LSFA_old/first_steps_into_forcing.tex LSFA_new/first_steps_into_forcing.tex > LSFA_new/diff.tex
+cd LSFA_new/
+sed -i -- 's/Local Variables://g' diff.tex
+sed -i -- 's/bibliographystyle{entcs}/bibliographystyle{mi-estilo-else}/g' diff.tex
+cp ../../arXiv/mi-estilo-else.bst .
+cp diff.tex diff.tmp
+cat ../../header-of-diff.tex > diff.tex
+tail -n +66 diff.tmp >> diff.tex
+pdflatex diff.tex
+pdflatex diff.tex
+bibtex diff.aux
+pdflatex diff.tex
+pdflatex diff.tex

### 2018-11-19 LSFA/cover_letter.tex
quedó un "his" de otra versión de esta carta.
Miguel haciendo sus cosas tarde

@@
 In first place we want to thank the referees for reading the paper very
-carefully and for his detailed and constructive criticism. Below we
+carefully and for their detailed and constructive criticism. Below we
 detail the main changes that have been made%; also, a diff-PDF is
@@
 The section on ``Names'' includes a thorough discussion on recursive
-definitions in Isabelle/ZF and two new applications. One is the
-proof that the generic
-extension $M[G]$ is a transitive set. The second one is even more basic  and
-was missing in the two previous versions of the manuscript: The
+definitions in Isabelle/ZF and two new applications. One is the proof
+that the generic extension $M[G]$ is a transitive set and that the
 generic filter $G$ belongs to $M[G]$.
@@
 %%%%%%%%%%%%%%%%%%%%
-took almost an year of work. With the hindsight of this planning period,
-we would like to point out that most of the results stated in the
-conclusion section involve some serious development; even  a
-publication for each item. It is therefore not feasible to include
-none of them in this revision.
+took almost an year of work. With the hindsight of this planning
+period, we would like to point out that most of the results stated in
+the conclusion section involve some serious development; and it might
+even result in a publication for each item. It is therefore not
+feasible to include any of them in this revision.
 %

### 2018-11-19 LSFA/names.tex
Miguel haciendo sus cosas tarde

@@
 \end{isabelle}
+It is conventional in Isabelle/ZF to define introduction and
+destruction rules for definitions like \isa{GenExt}; in our case, it is
+enough to know $x\in M$ in order to know $\val(G,x) \in M[G]$:
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ GenExtI{\isacharcolon}\ \ {\isachardoublequoteopen}x\ {\isasymin}\ M\ {\isasymLongrightarrow}\ val{\isacharparenleft}G{\isacharcomma}x{\isacharparenright}\ {\isasymin}\ M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}
+\end{isabelle}
+\noindent The destruction rule corresponding to the generic extension says
+that any $x \in M[G]$ comes from some $\tau \in M$ via $\val$. 
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ GenExtD{\isacharcolon}\ \ {\isachardoublequoteopen}x\ {\isasymin}\ M{\isacharbrackleft}G{\isacharbrackright}\ {\isasymLongrightarrow}\ {\isasymexists}{\isasymtau}{\isasymin}M{\isachardot}\ x\ {\isacharequal}\ val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
 
@@
 Here, \isa{eclose} returns the (downward) $\in$-closure of its
-argument. The main result is stated as follows:
+argument. The main lemmas about $\val$ and $\chk$ require some
+instances of replacement for $M$; we set up a locale to assemble these
+assumptions:
+\begin{isabelle}
+\isacommand{locale}\isamarkupfalse%
+\ M{\isacharunderscore}extra{\isacharunderscore}assms\ {\isacharequal}\ forcing{\isacharunderscore}data\ {\isacharplus}\isanewline
+\ \ \isakeyword{assumes}\ check{\isacharunderscore}in{\isacharunderscore}M\ {\isacharcolon}\ \ \ \ \ \  {\isachardoublequoteopen}{\isasymAnd}x{\isachardot}\ x\ {\isasymin}\ M\ {\isasymLongrightarrow}\ check{\isacharparenleft}x{\isacharparenright}\ {\isasymin}\ M{\isachardoublequoteclose}\isanewline
+\ \ \ \ \ \ \isakeyword{and}\ sats{\isacharunderscore}upair{\isacharunderscore}ax{\isacharcolon}\ \ \ \ {\isachardoublequoteopen}upair{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \ \ \ \ \isakeyword{and}\ repl{\isacharunderscore}check{\isacharunderscore}pair\ {\isacharcolon}\ {\isachardoublequoteopen}strong{\isacharunderscore}replacement{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharcomma}{\isasymlambda}p\ y{\isachardot}\ y\ {\isacharequal}{\isacharless}check{\isacharparenleft}p{\isacharparenright}{\isacharcomma}p{\isachargreater}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
+
+The first assumption asserts that all the relevant names are indeed in
+$M$ (i.e., $\chk(x) \in M$ if $x\in M$) and it is needed to prove that
+$\val(G,\chk(x))=x$. It will take a serious effort to fulfill this
+assumption: One of the hardest parts of Paulson's formalization of
+constructibility involves showing that models are closed under
+recursive construction. We will eventually formalize that if
+$M\models\ZFC$ and the arguments of \isa{wfrec} are in $M$, then its
+value also is. This will require to adapt to ctm models several
+locales defined in \cite{paulson_2003} that were intended to be used
+for the class of constructible sets. Notice that the only requirement
+on the set \isa{G} is that it contains the top element of the poset
+\isa{P}.
 
@@
 \isacommand{lemma}\isamarkupfalse%
-\ valcheck\ {\isacharcolon}\ {\isachardoublequoteopen}y\ {\isasymin}\ M\ {\isasymLongrightarrow}\ one\ {\isasymin}\ G\ {\isasymLongrightarrow}\ {\isasymforall}x{\isasymin}M{\isachardot}\ check{\isacharparenleft}x{\isacharparenright}\ {\isasymin}\ M\ {\isasymLongrightarrow}\ val{\isacharparenleft}G{\isacharcomma}check{\isacharparenleft}y{\isacharparenright}{\isacharparenright}\ \ {\isacharequal}\ y{\isachardoublequoteclose}
+\ valcheck\ {\isacharcolon}\ \isanewline
+\ \ \isakeyword{assumes}\ {\isachardoublequoteopen}one\ {\isasymin}\ G{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\ {\isachardoublequoteopen}y\ {\isasymin}\ M\ {\isasymLongrightarrow}\ val{\isacharparenleft}G{\isacharcomma}check{\isacharparenleft}y{\isacharparenright}{\isacharparenright}\ \ {\isacharequal}\ y{\isachardoublequoteclose}
 \end{isabelle}
 
-On the one hand, the only requirement on the set \isa{G} is that it
-contains the top element of the poset \isa{P}. On the other hand, it
-is necessary that all the relevant names are indeed in $M$ (i.e.,
-$\chk(x) \in M$ if $x\in M$). It requires a serious development to
-fulfill this assumption. One of the hardest parts of Paulson's
-formalization of constructibility involves showing that models are
-closed under recursive construction. We will eventually formalize that
-if $M\models\ZFC$ and the arguments of \isa{wfrec} are in $M$, then
-its value also is. This will require to adapt to ctm models several
-locales defined in \cite{paulson_2003} that were intended to be used
-for the class of constructible sets.
+\subsection{Basic results about the generic extension}
+
+We turn now to prove that $M[G]$ is transitive and $G\in
+M[G]$. Showing that $M[G]$ is transitive amounts to prove $y \in M[G]$
+for any $x \in M[G]$ and $y \in x$. 
+% \begin{isabelle}
+% \isacommand{definition}\isamarkupfalse%
+% \isanewline
+% \ \ Transset\ \ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isacharequal}{\isachargreater}o{\isachardoublequoteclose}\ \ \isakeyword{where}\isanewline
+% \ \ \ \ {\isachardoublequoteopen}Transset{\isacharparenleft}i{\isacharparenright}\ {\isacharequal}{\isacharequal}\ {\isasymforall}x{\isasymin}i{\isachardot}\ x{\isacharless}{\isacharequal}i{\isachardoublequoteclose}
+% \end{isabelle}
+
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ trans{\isacharunderscore}Gen{\isacharunderscore}Ext{\isacharprime}\ {\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\ \ {\isachardoublequoteopen}x\ {\isasymin}\ M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}\ \ \isakeyword{and}\ \ {\isachardoublequoteopen}y\ {\isasymin}\ x{\isachardoublequoteclose}\ \isanewline
+\ \ \isakeyword{shows}\ \ \ \ {\isachardoublequoteopen}y\ {\isasymin}\ M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}
+\end{isabelle}
+The proof of this lemma is straightforward because from $x \in M[G]$
+we can obtain $\tau \in M$ such that $x = \val(G,\tau)$. Notice also
+that using the characterization of $\val$ given by
+\isa{def{\isacharunderscore}val} we can extract some
+$\theta \in \dom(\tau)$ such that $y =\val(G,\theta)$; to conclude
+$\val(G,\theta) \in M[G]$ it is enough to prove $\theta \in M$, which
+follows from the transitivity of $M$.
+
+In contrast, the proof that $G\in M[G]$ is more demanding. In fact, we
+set $\dot{G} = \{ \langle \check{p},p\rangle \,|\, p \in P \}$ as a
+putative name for $G$. Proving that $\dot{G}$ is in fact a name for
+$G$ requires to prove that $\dot{G} \in M$, using an instance of
+replacement for $M$ (\noindent namely that given by the assumption
+\isa{repl{\isacharunderscore}check{\isacharunderscore}pair}), and
+then proving that $\val(G,\dot{G})=G$.
+
+\begin{isabelle}
+\isacommand{definition}\isamarkupfalse%
+\isanewline
+\ \ G{\isacharunderscore}dot\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
+\ \ {\isachardoublequoteopen}G{\isacharunderscore}dot\ {\isacharequal}{\isacharequal}\ {\isacharbraceleft}{\isacharless}check{\isacharparenleft}p{\isacharparenright}{\isacharcomma}p{\isachargreater}\ {\isachardot}\ p{\isasymin}P{\isacharbraceright}{\isachardoublequoteclose}
+\end{isabelle}
+
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ G{\isacharunderscore}dot{\isacharunderscore}in{\isacharunderscore}M\ {\isacharcolon} {\isachardoublequoteopen}G{\isacharunderscore}dot\ {\isasymin}\ M{\isachardoublequoteclose}
+\isanewline\isanewline
+\isacommand{lemma}\isamarkupfalse%
+\ val{\isacharunderscore}G{\isacharunderscore}dot\ {\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\ {\isachardoublequoteopen}G\ {\isasymsubseteq}\ P{\isachardoublequoteclose}\ \ \isakeyword{and}\ \ 
+{\isachardoublequoteopen}one\ {\isasymin}\ G{\isachardoublequoteclose}\ \isanewline
+\ \ \isakeyword{shows}\ {\isachardoublequoteopen}val{\isacharparenleft}G{\isacharcomma}G{\isacharunderscore}dot{\isacharparenright}\ {\isacharequal}\ G{\isachardoublequoteclose}
+\end{isabelle}
 

### 2018-11-19 LSFA/related.tex
Miguel haciendo sus cosas tarde

@@
 \subsection{Related work}\label{sec:related-work}
-Formalization of mathematics serves many purposes. The most obvious
-one is to increase reliability in a result and/or its proof. This has
-been the original motivation that lead Voevodsky to gather many
-researchers around \emph{homotopy type theory} and its formalization in Coq
-\cite{hottbook}; the same applies to the four colour theorem (checked by
-Gonthier \cite{MR2463991}) and the formidable \emph{Flyspeck} project
-\cite{MR3659768} by the team conducted by Hales.
+Formalization of mathematics serves many purposes
+\cite{simpson-theorem-proving-math}. The most obvious one is to
+increase reliability in a result and/or its proof. This has been the
+original motivation that lead Voevodsky to gather many researchers
+around \emph{homotopy type theory} and its formalization in Coq
+\cite{hottbook}; the same applies to the four colour theorem (checked
+by Gonthier \cite{MR2463991}) and the formidable \emph{Flyspeck}
+project \cite{MR3659768} by the team conducted by Hales.
 

### 2018-11-19 LSFA/hacer_diff.sh

@@ -3,7 +3,7 @@
 mkdir diff_dir
 rm -rf diff_dir/*
 mkdir diff_dir/LSFA_old/ diff_dir/LSFA_new/
-git checkout 80e6d1835fc9da4394d0c805d70c3032f6ba297d
+git archive 80e6d1835fc9da4394d0c805d70c3032f6ba297d | tar -x -C diff_dir/LSFA_old
 cp * diff_dir/LSFA_old/
 git checkout new_design
 cp * diff_dir/LSFA_new/

### 2018-11-19 LSFA/hacer_diff.sh
quedaba un checkout, y otro comando viejo

@@ -4,8 +4,8 @@ mkdir diff_dir
 rm -rf diff_dir/*
 mkdir diff_dir/LSFA_old/ diff_dir/LSFA_new/
 git archive 80e6d1835fc9da4394d0c805d70c3032f6ba297d | tar -x -C diff_dir/LSFA_old
-cp * diff_dir/LSFA_old/
-git checkout new_design
+# cp * diff_dir/LSFA_old/
+# git checkout new_design
 cp * diff_dir/LSFA_new/
 cd diff_dir
 latexdiff --flatten LSFA_old/first_steps_into_forcing.tex LSFA_new/first_steps_into_forcing.tex > LSFA_new/diff.tex

### 2018-11-19 LSFA/cover_letter.tex
mandamos el diff
basic -> fundamental

@@
 carefully and for their detailed and constructive criticism. Below we
-detail the main changes that have been made%; also, a diff-PDF is
-%provided that shows all the changes in color
-.
+detail the main changes that have been made; also, a diff-PDF is
+provided that shows all the changes in color.
 
@@
 The section on ``Names'' includes a thorough discussion on recursive
-definitions in Isabelle/ZF and two new applications. One is the proof
-that the generic extension $M[G]$ is a transitive set and that the
-generic filter $G$ belongs to $M[G]$.
+definitions in Isabelle/ZF and two new applications. The first is the proof
+that the generic extension $M[G]$ is a transitive set; and the second
+is that thegeneric filter $G$ belongs to $M[G]$, a fundamental result
+that was missing from the previous two versions.
 

### 2018-11-19 LSFA/cover_letter.tex
spell check

@@
 that the generic extension $M[G]$ is a transitive set; and the second
-is that thegeneric filter $G$ belongs to $M[G]$, a fundamental result
+is that the generic filter $G$ belongs to $M[G]$, a fundamental result
 that was missing from the previous two versions.

### 2018-11-19 LSFA/isabelle-zf.tex
procedural es top-down (estaba al revés)

@@
 Paulson used this method. Under this approach proofs are constructed
-bottom-up resulting in proof-scripts that conceal the mathematical
+top-down resulting in proof-scripts that conceal the mathematical
 reasoning behind the proof, since the intermediate steps are only

### 2018-11-19 LSFA/related.tex
spell check

@@
 around \emph{homotopy type theory} and its formalization in Coq
-\cite{hottbook}; the same applies to the four colour theorem (checked
+\cite{hottbook}; the same applies to the four color theorem (checked
 by Gonthier \cite{MR2463991}) and the formidable \emph{Flyspeck}
@@
 \isatt{ZF}). Another one, by A.~Popescu and D.~Traytel, reaches
-cardinal artihmetic. This is fairly limited for our purposes.
+cardinal arithmetic. This is fairly limited for our purposes.
 

### 2018-11-19 LSFA/names.tex
spell check

@@
 of the ctm $M$ and the generic filter $G$. The definition of $\val$
-can be written succintly as a recursive equation
+can be written succinctly as a recursive equation
 %
@@
 
-The formalizaton of the functional $\mathit{Hv}$ is straightforward and $\val$ is defined using \isa{wfrec}:
+The formalization of the functional $\mathit{Hv}$ is straightforward and $\val$ is defined using \isa{wfrec}:
 %

### 2018-11-19 LSFA/hacer_diff.sh
cambió una línea en algún lugar (????)

@@ -15,7 +15,7 @@ sed -i -- 's/bibliographystyle{entcs}/bibliographystyle{mi-estilo-else}/g' diff.
 cp ../../arXiv/mi-estilo-else.bst .
 cp diff.tex diff.tmp
 cat ../../header-of-diff.tex > diff.tex
-tail -n +66 diff.tmp >> diff.tex
+tail -n +67 diff.tmp >> diff.tex
 pdflatex diff.tex
 pdflatex diff.tex
 bibtex diff.aux

### 2018-11-27 src/document/root.tex
root para generación de tex

\documentclass[11pt,a4paper]{article}
\usepackage{isabelle,isabellesym}

% further packages required for unusual symbols (see also
% isabellesym.sty), use only when needed

%\usepackage{amssymb}
  %for \<leadsto>, \<box>, \<diamond>, \<sqsupset>, \<mho>, \<Join>,
  %\<lhd>, \<lesssim>, \<greatersim>, \<lessapprox>, \<greaterapprox>,
  %\<triangleq>, \<yen>, \<lozenge>

%\usepackage{eurosym}
  %for \<euro>

%\usepackage[only,bigsqcap]{stmaryrd}
  %for \<Sqinter>

%\usepackage{eufrak}
  %for \<AA> ... \<ZZ>, \<aa> ... \<zz> (also included in amssymb)

%\usepackage{textcomp}
  %for \<onequarter>, \<onehalf>, \<threequarters>, \<degree>, \<cent>,
  %\<currency>

% this should be the last package used
\usepackage{pdfsetup}

% urls in roman style, theory text in math-similar italics
\urlstyle{rm}
\isabellestyle{it}

% for uniform font size
%\renewcommand{\isastyle}{\isastyleminor}

\begin{document}

\title{src}
\author{By miguel}
\maketitle

\tableofcontents

% sane default for proof documents
\parindent 0pt\parskip 0.5ex

% generated text of all theories
\input{session}

% optional bibliography
%\bibliographystyle{abbrv}
%\bibliography{root}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2018-11-27 LSFA/arXiv/intro.tex
versión para arXiv
saco comentarios
version actualizada de arXiv

--- 2018-06-25 LSFA/intro.tex
@@
 %
-\noindent As in the last sentence, one usually refer to a transitive
+\noindent As in the last sentence, one usually refers to a transitive
 model by the underlying set because the relation is fixed.
-%% MIGUEL: feísimo este corte. Quizás refrasear la oración anterior
-%% para que entre en un renglón.
+%
+%
 
-%% MIGUEL: lo que sigue implica que para tener aritmética en ZFC son
-%% necesarios una cantidad infinita de axiomas de ZFC?
+%
+%
 In spite of G\"odel's Second Incompleteness Theorem, one can find
@@
 proof (e.g.\ of a contradiction)%
-%%%%%%%%%%%%%
+%
 \footnote{It is relevant to this point that 
@@
   concluding a contradiction from $\ZFC+A$ to one from $\ZFC$.}
-%%%%%%%%%%%%%
+%
 involves only finitely many axioms.
@@
 such that $L^M = \{x\in M : M\models L(x)\}$. To show that
-$L^M\models \ZFC+\CH$, one uses that the $\ZFC+\CH$ holds in $M$.
+$L^M\models \ZFC+\CH$, one uses the fact that $\ZFC$ holds in $M$.
 
-%% MIGUEL: acá me parece que hay un salto medio abrupto. Por que
-%% es una necesidad primaria? Lo podés relacionar con lo que venía
-%% del párrafo anterior?
-%% MIGUEL: quizás lo que querés decir acá es para probar que vamos
-%% a construir un modelo N a partir de un modelo M; y para probar
-%% que N satisface un axioma, vamos a utilizar que M lo satisface.
+%
+%
+%
+%
+%
+%
 It is therefore a primary need to have a means to correlate  (first-order)
@@
 submodel $N\sbq M$. As a simple example on this, consider 
-%% $M\defi \{\emptyset,\{\emptyset\},\{\{\emptyset\}\}\}$ and
-%% $N\defi\{\{\emptyset\},\{\{\emptyset\}\}\}$, and let 
-%% $\phi(x)\defi \neg\exists y (y \in x)$. It is clear that 
-%% $M\models \phi(\emptyset)$; but we have
-%% \[
-%% M\not\models \phi(\{\emptyset\}) \quad\text{ but }\quad N\models \phi(\{\emptyset\}).
-%% \]
-%% There is a discrepancy between what $M$ and $N$ \emph{believe} about
-%% their common element $\{\emptyset\}$.
+%
+%
+%
+%
+%
+%
+%
+%
+%
 $M\defi \{a,b,c, \{a,b\},\{a,b,c\}\}$ and
@@
 \]
-%% It is clear that  $M\models \phi(\{a,b\})$; but we have
+%
 Then we have
@@
 say then that $\phi$ is \emph{absolute between $N$ and $M$.}
-%% MIGUEL: "truth of axioms in $M$" significa que $M$ satisface los
-%% axiomas?
+%
+%
 The concepts of relativization and absoluteness are central to the
@@
 \subsection{Forcing}
-%% MIGUEL: no veo la necesidad del adversativo but, pondría and.
-%% MIGUEL: Qué significa "a new set $G$ is adjoined to $M$"?
-%% MIGUEL: este párrafo necesita correcciones.
+%
+%
+%
 Forcing is a technique to extend countable transitive models of
 $\ZFC$. This process is guaranteed to preserve the $\ZFC$
-axioms while allowing to fine-tune what other first order properties the
+axioms while allowing to fine-tune what other first-order properties the
 extension will have. Given a ctm $M$  of $\ZFC$ and a set $G$, one constructs a new
@@
 
-%% MIGUEL: Tenemos que dado un ctm M, existe un filtro genérico G.
-%% ¿Dónde existe? Una primera lectura de ignorante me dice "existe
-%% en M", pero eso me lo desmiente la siguiente oración. ¿Juega algún
-%% rol absolutez? Si es así estaría bien explicitarlo.
-%% MIGUEL: la última oración me resulta un poco difícil de entender.
-The Rasiowa-Sikorski lemma (RSL) states that, fixed a preorder $\PP$, for
-any countable family of dense subsets of $\PP$ there is a filter
-intersecting all of them. Thus, there are generic filters $G$ for
-countable transitive models. In general, no such $G$ belongs to $M$ and
+%
+%
+%
+%
+%
+The Rasiowa-Sikorski lemma (RSL) states that for any preorder $\PP$
+and any countable family $\{\mathcal{D}_n : n\in\mathbb{N}\}$ of dense
+subsets of $\PP$ there is a filter intersecting every
+$\mathcal{D}_i$. Thus, there are generic filters $G$ for countable
+transitive models. In general, no such $G$ belongs to $M$ and
 therefore the extension $M[G]$ is proper. We formalize the proof of
-RSL in
-Section~\ref{sec:rasiowa-sikorski-lemma}. A requisite result on a
-version of the Axiom of Choice is formalized in
-Section~\ref{sec:sequence-version-dc}. We then apply RSL to prove the existence of generic
-filters in Section~\ref{sec:generic-filter}.
+RSL in Section~\ref{sec:rasiowa-sikorski-lemma}. A requisite result on
+a version of the Axiom of Choice is formalized in
+Section~\ref{sec:sequence-version-dc}. We then apply RSL to prove the
+existence of generic filters in Section~\ref{sec:generic-filter}.
 
-%% MIGUEL: redacción alternativa del párrafo siguiente.
+%
 Every  $y \in M[G]$ is obtained from an element $\dot y$ of $M$, thus
@@
 
-%We continue with a brief description of the Isabelle framework.
+%
 
-%%% Local Variables:
-%%% mode: latex
-%%% ispell-local-dictionary: "american"
-%%% TeX-master: "first_steps_into_forcing"
-%%% End:
+%
+%
+%
+%
+%

### 2018-11-27 LSFA/arXiv/names.tex
versión para arXiv
version actualizada de arXiv

--- 2018-06-25 LSFA/names.tex
@@
 \label{sec:names}
-In this section we formalize the function $\val$ that allows to
+We formalize the function $\val$ that allows to
 construct the elements of the generic extension $M[G]$ from elements
 of the ctm $M$ and the generic filter $G$. The definition of $\val$
-can be written succintly as
+can be written succinctly as a recursive equation
 %
 \begin{equation}\label{eq:def-val}
-\val(G,\tau)\defi \{\val(G,\sigma) : \exists p\in \PP
+\val(G,\tau)\defi \{\val(G,\sigma) : \exists p\in \PP.\,
 (\lb\sigma,p\rb\in\tau \y p\in G)\}.
@@
 %
-As with any recursive definition, to justify that the definition makes
-sense (i.e., it defines anything at all) one has to explicit the
-well-founded relation which is used. In this case, we use the relation
-$\mathit{ed}$:
+The justification that $\val$ is well-defined comes from a general
+result (transfinite recursion on well-founded
+relations~\cite[p. 48]{kunen2011set}). Given a well-founded relation
+$R \subseteq A \times A$ and a functional
+$H : A \times (A \to A) \to A$, the principle asserts the existence of
+a function $F : A \to A$ satisfying
+$F(a) = H(a,F\uparrow\,(R^{-1}(a)))$. This principle is formalized in
+Isabelle/ZF and one can use the operator \isa{wfrec}\footnote{Notice
+  that this form of recursive definitions is more general than the one
+  used in the previous section to define
+  \isa{dc{\isacharunderscore}witness}.} to define functions using
+transfinite recursion. To be precise, \isa{wfrec :: [i, i, [i,i]=>i]
+  => i} is a slight variation, where the first argument is the
+relation, the third is the functional, and the second corresponds to
+the argument of $F$. Notice that the relation and the function argument
+of the functional are internalized as terms of type \isa{i}.
+
+In our case the functional is called $Hv$ and takes an additional argument for the
+parameter $G$:
+\[
+  \mathit{Hv}(G,y,f) = \{f(x) : x \in dom(y) \y \exists p\in \PP.\,
+(\lb x,p\rb\in y \y p\in G)\}
+\]
+while the relation is given by:
 \[
@@
 
-The  formalization of $\val$ in Isabelle/ZF involves the function
-\isa{wfrec}, which takes as an argument a well-founded relation of
-type \isa{i}, i.e.\ a set of pairs. Therefore we restrict
-$\mathit{ed}$ above to a set.
+In our formalization of this recursion,  the first argument of
+\isa{wfrec} is the term of type  \isa{i} obtained by restricting the
+relation $\mathit{ed}$ to  a set:
 % 
@@
 \end{isabelle}
+%
+All but one lemma used in the above proof
+(\isa{wf{\isacharunderscore}subset},
+\isa{wf{\isacharunderscore}trancl}, 
+\isa{wf{\isacharunderscore}Memrel}) are already
+present in Isabelle/ZF. The remaining technical result has
+been proved using the Isar language:
+%
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ edrel{\isacharunderscore}sub{\isacharunderscore}memrel{\isacharcolon}\ {\isachardoublequoteopen}edrel{\isacharparenleft}A{\isacharparenright}\ {\isasymsubseteq}\ trancl{\isacharparenleft}Memrel{\isacharparenleft}eclose{\isacharparenleft}A{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
 
-\noindent We now turn to the definition of $\val$. In order to use
-\isa{wfrec} we need to define a functional \isa{Hv}.
-
+The formalization of the functional $\mathit{Hv}$ is straightforward and $\val$ is defined using \isa{wfrec}:
+%
 \begin{isabelle}
@@
 \isanewline
-\ \ \ \ \ \ \ \  {\isachardoublequoteopen}x{\isasymin}M\ {\isasymLongrightarrow}\ val{\isacharparenleft}G{\isacharcomma}x{\isacharparenright}\ {\isacharequal}\ {\isacharbraceleft}val{\isacharparenleft}G{\isacharcomma}t{\isacharparenright}\ {\isachardot}{\isachardot}\ t{\isasymin}domain{\isacharparenleft}x{\isacharparenright}\ {\isacharcomma}\ {\isasymexists}p{\isasymin}P\ {\isachardot}\ \ {\isasymlangle}t{\isacharcomma}\ p{\isasymrangle}{\isasymin}x\ {\isasymand}\ p\ {\isasymin}\ G\ {\isacharbraceright}{\isachardoublequoteclose}
+   {\isachardoublequoteopen}x{\isasymin}M\ {\isasymLongrightarrow}\ val{\isacharparenleft}G{\isacharcomma}x{\isacharparenright}\ {\isacharequal}\ {\isacharbraceleft}val{\isacharparenleft}G{\isacharcomma}t{\isacharparenright}\ {\isachardot}{\isachardot}\ t{\isasymin}domain{\isacharparenleft}x{\isacharparenright}{\isacharcomma}\ {\isasymexists}p{\isasymin}P\ {\isachardot}\ {\isasymlangle}t{\isacharcomma}\ p{\isasymrangle}{\isasymin}x\ {\isasymand}\ p{\isasymin}G{\isacharbraceright}{\isachardoublequoteclose}
 \end{isabelle}
@@
 \isanewline
-\ \ GenExt\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}i{\isachardoublequoteclose}\ \ \ \ \ {\isacharparenleft}{\isachardoublequoteopen}M{\isacharbrackleft}{\isacharunderscore}{\isacharbrackright}{\isachardoublequoteclose}\ {\isadigit{9}}{\isadigit{0}}{\isacharparenright}\ \isakeyword{where} \isanewline
+\ \ GenExt\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}i{\isachardoublequoteclose}\ {\isacharparenleft}{\isachardoublequoteopen}M{\isacharbrackleft}{\isacharunderscore}{\isacharbrackright}{\isachardoublequoteclose}{\isacharparenright}\ \isakeyword{where} \isanewline
 \ \ {\isachardoublequoteopen}GenExt{\isacharparenleft}G{\isacharparenright}{\isacharequal}{\isacharequal}\ {\isacharbraceleft}val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}{\isachardot}\ {\isasymtau}\ {\isasymin}\ M{\isacharbraceright}{\isachardoublequoteclose}
+\end{isabelle}
+It is conventional in Isabelle/ZF to define introduction and
+destruction rules for definitions like \isa{GenExt}; in our case, it is
+enough to know $x\in M$ in order to know $\val(G,x) \in M[G]$:
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ GenExtI{\isacharcolon}\ \ {\isachardoublequoteopen}x\ {\isasymin}\ M\ {\isasymLongrightarrow}\ val{\isacharparenleft}G{\isacharcomma}x{\isacharparenright}\ {\isasymin}\ M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}
+\end{isabelle}
+\noindent The destruction rule corresponding to the generic extension says
+that any $x \in M[G]$ comes from some $\tau \in M$ via $\val$. 
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ GenExtD{\isacharcolon}\ \ {\isachardoublequoteopen}x\ {\isasymin}\ M{\isacharbrackleft}G{\isacharbrackright}\ {\isasymLongrightarrow}\ {\isasymexists}{\isasymtau}{\isasymin}M{\isachardot}\ x\ {\isacharequal}\ val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}{\isachardoublequoteclose}
 \end{isabelle}
@@
 Here, \isa{eclose} returns the (downward) $\in$-closure of its
-argument. The main result is stated as follows:
+argument. The main lemmas about $\val$ and $\chk$ require some
+instances of replacement for $M$; we set up a locale to assemble these
+assumptions:
+\begin{isabelle}
+\isacommand{locale}\isamarkupfalse%
+\ M{\isacharunderscore}extra{\isacharunderscore}assms\ {\isacharequal}\ forcing{\isacharunderscore}data\ {\isacharplus}\isanewline
+\ \ \isakeyword{assumes}\ check{\isacharunderscore}in{\isacharunderscore}M\ {\isacharcolon}\ \ \  {\isachardoublequoteopen}{\isasymAnd}x{\isachardot}\ x\ {\isasymin}\ M\ {\isasymLongrightarrow}\ check{\isacharparenleft}x{\isacharparenright}\ {\isasymin}\ M{\isachardoublequoteclose}\isanewline
+\ \ \ \isakeyword{and}\ sats{\isacharunderscore}upair{\isacharunderscore}ax{\isacharcolon}\ \ \ \ {\isachardoublequoteopen}upair{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \ \isakeyword{and}\ repl{\isacharunderscore}check{\isacharunderscore}pair{\isacharcolon}\ {\isachardoublequoteopen}strong{\isacharunderscore}replacement{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharcomma}{\isasymlambda}p\ y{\isachardot}\ y\ {\isacharequal}{\isacharless}check{\isacharparenleft}p{\isacharparenright}{\isacharcomma}p{\isachargreater}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
+
+The first assumption asserts that all the relevant names are indeed in
+$M$ (i.e., $\chk(x) \in M$ if $x\in M$) and it is needed to prove that
+$\val(G,\chk(x))=x$. It will take a serious effort to fulfill this
+assumption: One of the hardest parts of Paulson's formalization of
+constructibility involves showing that models are closed under
+recursive construction. We will eventually formalize that if
+$M\models\ZFC$ and the arguments of \isa{wfrec} are in $M$, then its
+value also is. This will require to adapt to ctm models several
+locales defined in \cite{paulson_2003} that were intended to be used
+for the class of constructible sets. Notice that the only requirement
+on the set \isa{G} is that it contains the top element of the poset
+\isa{P}.
 

### 2018-11-27 LSFA/arXiv/pair-gen-ext.tex
version actualizada de arXiv

@@
 \label{sec:pairing-generic-extension}
-Given a predicate
-\isa{C\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}o{\isachardoublequoteclose}},
-we can consider the \emph{class} of all sets \isa{x} such that
-\isa{C(x)}. In \cite{paulson_2003} a large corpus of set-theoretical
-concepts are expressed in relative form.
 
-In the case of the Pairing Axiom the following definition in 
-Paulson's formalization captures
-what a class \isa{C} believes about a set being the unordered pair of
-other two:
+In this section we show that the generic extension satisfies the
+pairing axiom; the purpose of this section is to show how to prove
+that $M[G]$ models one of the axioms of $\ZFC$, assuming that $M$
+satisfies $\ZFC$.\footnote{The proof that $M[G]$ satisfies pairing
+  only needs that $M$ satisfies pairing.} In the locale
+\isa{M{\isacharunderscore}extra{\isacharunderscore}assms} we stated
+the assumption \isa{sats{\isacharunderscore}upair{\isacharunderscore}ax} which captures that
+$M$ satisfies pairing. We use \emph{relativized} versions of the axioms
+in order to express satisfaction.
+
+%% MIGUEL: versión vieja
+% In the previous section we defined the generic extension $M[G]$; we
+% have yet to prove that $M[G]$ is a model of $\ZFC$. This is equivalent
+% to show that each axiom relativized with respect to $M[G]$ holds. In
+% this section we show that the generic extension satisfies the pairing
+% axiom, after explaining relativization in Isabelle/ZF.
+
+As we have already mentioned, in Paulson's library, the relativized
+versions of the $\ZFC$ axioms are defined for classes (which are
+defined as predicates over sets). The definition
+\isa{upair{\isacharunderscore}ax} corresponds to the Pairing Axiom:
+
 \begin{isabelle}
@@
 %
-Then the fact that Pairing axiom holds in \isa{C} is stated as follows.
+%% Then the fact that Pairing axiom holds in \isa{C} is stated as follows.
 %
@@
 
-In order use the \isa{upair{\isacharunderscore}ax} for a set model, we
-use the \isatt{\#\#}  operator.  We state the main result of this
-Section in the context \isatt{forcing{\isacharunderscore}data}.  
+% Given a predicate
+% \isa{C\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}o{\isachardoublequoteclose}},
+% we can consider the \emph{class} of all sets \isa{x} such that
+% \isa{C(x)}. In \cite{paulson_2003} a large corpus of set-theoretical
+% concepts are expressed in relative form.
+
+% In the case of the Pairing Axiom the following definition in 
+% Paulson's formalization captures
+% what a class \isa{C} believes about a set being the unordered pair of
+% other two:
+
+We state the main result of this section in the context
+\isa{M{\isacharunderscore}extra{\isacharunderscore}assms}.
 %
@@
 \isacommand{lemma}\isamarkupfalse%
-\ pair{\isacharunderscore}preserv\ {\isacharcolon}\ \isanewline
-\ \ {\isachardoublequoteopen}one\ {\isasymin}\ G\ {\isasymLongrightarrow}\ upair{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}\ {\isasymLongrightarrow}\ upair{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharbrackleft}G{\isacharbrackright}{\isacharparenright}{\isachardoublequoteclose}
+\ pairing{\isacharunderscore}axiom\ {\isacharcolon}\ \isanewline
+\ \ {\isachardoublequoteopen}one\ {\isasymin}\ G\ {\isasymLongrightarrow}\ upair{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharbrackleft}G{\isacharbrackright}{\isacharparenright}{\isachardoublequoteclose}
 \end{isabelle}
@@
 \ pairs{\isacharunderscore}in{\isacharunderscore}M\ {\isacharcolon}\ \isanewline
-\ \ {\isachardoublequoteopen}\ {\isasymlbrakk}\ upair{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}\ {\isacharsemicolon}\ a\ {\isasymin}\ M\ {\isacharsemicolon}\ b\ {\isasymin}\ M\ {\isacharsemicolon}\ c\ {\isasymin}\ M\ {\isacharsemicolon}\ d\ {\isasymin}\ M\ {\isasymrbrakk}\ {\isasymLongrightarrow}\ {\isacharbraceleft}{\isasymlangle}a{\isacharcomma}c{\isasymrangle}{\isacharcomma}{\isasymlangle}b{\isacharcomma}d{\isasymrangle}{\isacharbraceright}\ {\isasymin}\ M{\isachardoublequoteclose}
+\ \ {\isachardoublequoteopen}\ {\isasymlbrakk}\ a\ {\isasymin}\ M\ {\isacharsemicolon}\ b\ {\isasymin}\ M\ {\isacharsemicolon}\ c\ {\isasymin}\ M\ {\isacharsemicolon}\ d\ {\isasymin}\ M\ {\isasymrbrakk}\ {\isasymLongrightarrow}\ {\isacharbraceleft}{\isasymlangle}a{\isacharcomma}c{\isasymrangle}{\isacharcomma}{\isasymlangle}b{\isacharcomma}d{\isasymrangle}{\isacharbraceright}\ {\isasymin}\ M{\isachardoublequoteclose}
 \end{isabelle}
@@
 \ sigma{\isacharunderscore}in{\isacharunderscore}M\ {\isacharcolon}\isanewline
-\ \ {\isachardoublequoteopen}upair{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}\ {\isasymLongrightarrow}\ one\ {\isasymin}\ G\ {\isasymLongrightarrow}\ {\isasymtau}\ {\isasymin}\ M\ {\isasymLongrightarrow}\ {\isasymrho}\ {\isasymin}\ M\ {\isasymLongrightarrow}\isanewline \ \  \ {\isacharbraceleft}{\isasymlangle}{\isasymtau}{\isacharcomma}one{\isasymrangle}{\isacharcomma}{\isasymlangle}{\isasymrho}{\isacharcomma}one{\isasymrangle}{\isacharbraceright}\ {\isasymin}\ M{\isachardoublequoteclose}
+\ \ {\isachardoublequoteopen} one\ {\isasymin}\ G\ {\isasymLongrightarrow}\ {\isasymtau}\ {\isasymin}\ M\ {\isasymLongrightarrow}\ {\isasymrho}\ {\isasymin}\ M\ {\isasymLongrightarrow}\ {\isacharbraceleft}{\isasymlangle}{\isasymtau}{\isacharcomma}one{\isasymrangle}{\isacharcomma}{\isasymlangle}{\isasymrho}{\isacharcomma}one{\isasymrangle}{\isacharbraceright}\ {\isasymin}\ M{\isachardoublequoteclose}
 
@@
 \end{isabelle}
-
-This ends the proof of the Pairing Axiom in $M[G]$.

### 2018-11-27 LSFA/related.tex
typo

@@
 the main goal is not to increase confidence. Nevertheless, the level
-of detail in a formalization of this sort always provide additional
+of detail in a formalization of this sort always provides additional
 information about the inner workings of the theory: It is expected, for

### 2018-11-27 src/fix_links.sh
arregla links a teorías de base (Main, ZF-Constructible, etc)

@@ -1,0 +1,9 @@
+#!/bin/bash
+old_dir=$PWD
+html_dir="output/html/Unsorted/separation/"
+
+cd $html_dir
+sed -i -- 's/a href="..\/..\/ZF\/ZF\/Main.html/a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF\/ZF.html/g' *.html
+sed -i -- 's/a href="..\/..\/ZF\/ZF\//a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF\//g' *.html
+sed -i -- 's/a href="..\/..\/ZF\/ZF-Constructible\//a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF-Constructible\//g' *.html
+cd $old_dir

### 2018-11-30 LICS/conclusion.tex
modularizando el manuscrito

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
\section{Conclusion}

\section*{Acknowledgment}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "Separation_In_MG"
%%% End:

### 2018-11-30 LICS/forcing.tex
modularizando el manuscrito

--- 2018-04-12 blueprint.tex
@@
-\documentclass[11pt,english]{article}
-\usepackage[bottom=2cm, top=3cm, %% Comment if 
-  left=4cm, right=4cm]{geometry}             %% using 'ebook' style 
-\usepackage{framed}
-
-\input{header-blueprint}
-%%%%%%%%%%%%%%%%%%% % %   Unicodings   % % %%%%%%%%%%%%%%%%%%%%%
-\newcommand{\mt}[1]{$\mathtt{#1}$}
-\DeclareUnicodeCharacter{21D2}{\mt\Rightarrow}
-\DeclareUnicodeCharacter{2208}{\mt\in}
-\DeclareUnicodeCharacter{2200}{\mt\forall}
-\DeclareUnicodeCharacter{2203}{\mt\exists}
-\DeclareUnicodeCharacter{2227}{\mt\wedge}
-\DeclareUnicodeCharacter{27F6}{\mt\longrightarrow}
-\DeclareUnicodeCharacter{27E6}{\mt\llbracket}
-\DeclareUnicodeCharacter{27E7}{\mt\rrbracket}
-\DeclareUnicodeCharacter{27F9}{\mt\Longrightarrow}
-\DeclareUnicodeCharacter{03A6}{\mt\Phi}
-\DeclareUnicodeCharacter{03BB}{\mt\lambda}
-%\DeclareUnicodeCharacter{}{\mt}
-
-\renewcommand{\PP}{\mathbb{P}}
-\renewcommand{\app}{\mathrm{App}}
-\newcommand{\formula}{\ensuremath{\mathtt{formula}}}
-\newcommand{\tyi}{\mathtt{i}}
-\newcommand{\tyo}{\mathtt{o}}
-\newcommand{\forceisa}{\mathtt{forces}}
-\newcommand{\equ}{\mathbf{e}}
-\newcommand{\bel}{\mathbf{b}}
-\newcommand{\atr}{\mathit{atr}}
-\newcommand{\concat}{\mathbin{@}}
-\newcommand{\reee}{\mathrel{\mathtt{eee}}}
-\newcommand{\dB}[1]{\mathbf{#1}}
-\DeclareMathOperator{\val}{val}
-
-%%%%%%%%%%%%%%%%%%% Footnotemarks for "Thanks!" %%%%%%%%%%%%%%%%%%
-\makeatletter
-\renewcommand*{\@fnsymbol}[1]{\ensuremath{\ifcase#1\or *\or \dagger\or \ddagger\or
-    \mathsection\or \mathparagraph\or \|\or **\or \dagger\dagger
-    \or \ddagger\ddagger \else\@ctrerr\fi}}
-\makeatother
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{document}
-\title{Forcing for Isabelle}
-\author{%
-  Pedro Sánchez Terraf%
-  \thanks{
-    Universidad Nacional de C\'ordoba. 
-    Facultad de Matem\'atica, Astronom\'{\i}a,  F\'{\i}sica y
-    Computaci\'on.}
-  \thanks{
-    Centro de Investigaci\'on y Estudios de Matem\'atica (CIEM-FaMAF).
-    C\'ordoba. Argentina.}
-}
-%
-\maketitle
-%
-\begin{abstract}
-  Basic definitions of the forcing technique and blueprint for an
-  Isabelle/ZF proof of the Forcing Theorems, under the assumption of
-  the existence of a countable transitive model of $\ZFC$.   
-\end{abstract}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-\section{Introduction}
-The proofs are mostly based on Chapter IV of
-Kunen's~\cite{kunen2011set}, and the implementation will make use of
-the work by Paulson~\cite{paulson_2003}.
-
-\section{Forcing notions}
-We will need the following concepts:
-\begin{itemize}
-\item Forcing poset or \emph{notion}: a triple $\lb \PP, \1, \leq\rb$ where $\leq$ is a
-  partial (pre)order on $\PP$ with (a distinguished) top element $\1$.
-\item Compatible elements. Antichain in $\PP$. 
-\item Dense subset $D\sbq\PP$ of a forcing poset.
-\item Filter $G\sbq\PP$ in a poset.
-\item $\calD$-generic filter for $\PP$, where $\calD$ is a family of
-  dense subsets of $\PP$.
-\end{itemize}
-(Check \verb|ZF/Order.thy| before doing any work.)
-
-Most probably, we'll need two or three versions of each of these
-definitions: One internalized as an element of \formula;  a relativized
-one with a class parameter; and finally one without
-relativization. This is done in order to show absoluteness and to take
-advantage of the Isabelle/ZF ability to operate with terms of type
-$\tyi$.
-
-As an example, consider $\PP,R \in M$,  and $R\sbq
-\PP\times\PP$. Reflexivity of the relation $R$ is 
-stated as follows.
-\[
-\phi(\PP,R)\defi \forall p (p\in \PP \limp \lb p,p\rb \in R).
-\]
-In Isabelle/ZF, we write
-%
-\begin{verbatim}
-definition 
-  reflexivity :: "[i,i] ⇒ o" where
-  "reflexivity(P,r) == ∀p . p∈P ⟶ <p,p>∈r"
-\end{verbatim}
-
-This is different from what would $M$ see from ``within.'' In
-particular, the universal quantifier should range only over elements
-of $M$:
-\[
-\phi^M\defi\forall p\in M (p\in \PP \limp \lb p,p\rb^M \in R).
-\]
-The use of the relativized term $\lb p,p\rb^M$ implicitly assumes that
-there is an element of $M$ satisfying its defining formula. In
-Isabelle/ZF, we might move to a more general setting, by allowing  $M$
-to be a class. Here, the reference to the ordered pair  is through a
-relational definition.
-\begin{verbatim}
-definition  
-  reflexivity_rel :: "[i⇒o,i,i] ⇒ o" where
-  "reflexivity_rel(M,P,r) == ∀p[M].  p∈P ⟶ 
-                             (∃x[M]. pair(M,p,p,x) ∧ x∈r)"
-\end{verbatim}
-(Notice: There must be no spaces between \verb|[M]| and the following
-period.) This is the relativized version of
-\verb|reflexivity|.
-
-Finally, the internalized version $\quine{\phi}$ of reflexivity is as
-follows: 
-\begin{verbatim}
-definition
-  reflexivity_fm :: "[i,i]⇒i" where
-  "reflexivity_fm(x,y) == Forall(Implies(Member(0,succ(x)),
-          Exists(And(pair_fm(1,1,0),Member(0,succ(succ(y)))))))"
-
-lemma reflexivity_type : 
-        "⟦x∈nat ; y∈nat⟧ ⟹ reflexivity_fm(x,y)∈formula"
-  by (simp add:reflexivity_fm_def)
-\end{verbatim}
-
-The last lemma shows that indeed we have defined an element of the set
-\formula. The formula \verb|reflexivity_fm| is an internalization of
-the Isabelle/ZF formula \verb|reflexivity|, not the relativized
-one.
-
-Under the assumption of transitivity of $M$, the three versions are
-equivalent. This is an \emph{absoluteness result}:
-\[
-\phi(\PP,R) \lsii \phi^M(\PP,R) \lsii M;[\PP,R]\models\quine{\phi},
-\]
-when $\PP,R\in M$, and the list $[\PP,R]$ is the environment or
-interpretation of free variables.
-
-\begin{definition}
-  $G$ is \emph{$M$-generic for $\PP$} if it is $\calD$-generic for
-  $\PP$ for the family $\calD$ of all dense subsets of $\PP$ that live
-  in $M$.
-\end{definition}
-This last definition is \emph{not} to be relativized, since for
-nontrivial $\PP$, $G\notin M$. Absoluteness results as the one above
-always assume that the free variables range in the set or class to
-which one is relativizing.
-\section{Names}
-\begin{framed}
-  The following is mostly independent of the previous section, but %
-  \begin{bfseries}%
-    we assume that $\1\in\PP\in M$, $M$ is transitive and 
-    $1\in G\sbq\PP$.
-  \end{bfseries}
-  (Note that, in general, $G\notin M$.)
-\end{framed}
-
-\begin{definition}
-  \begin{enumerate}
-  \item The relation \verb|eee|$(M)$ is the restriction to $M$ of the
-    relational composition 
-    ${\in}\circ{\in}\circ{\in}$. 
-  \item For $\tau\in M$, the \emph{$G$-value} or \emph{interpretation
-    under $G$} of $\tau$ is
-    \[
-    \val(\tau,\PP,G)\defi \{\val(\sigma,\PP,G) : \exists p\in \PP
-    (\lb\sigma,p\rb\in\tau \y p\in G)\}.
-    \]
-  \end{enumerate}
-\end{definition}
-
-The definition of $\val$ proceeds by well-founded recursion on
-\verb|eee|$(M)$. Determining  $\val(\tau,\PP,G)$ uses the values
-$\val(\sigma,\PP,G)$ for smaller $\sigma$:
-\[
-\sigma\in\{\sigma\}\in\lb\sigma,p\rb\in\tau.
-\]
-
-\begin{remark}
-  Slightly generalizing the recursive scheme for $\val$ we have:
-  \[
-  F(\tau,\bx)\defi \{F(\sigma,\bx) : \phi(\sigma,\tau,\bx) \y
-  \sigma\mathrel{R}\tau\}, 
-  \]
-  where $R$ is some well-founded relation. Using this scheme the
-  \emph{Mostowski collapse} of $R$ can be defined:
-  \[
-  \mos(\tau)\defi \{\mos(\sigma) : \sigma\mathrel{R}\tau\},
-  \]
-  which is a tool of independent interest.
-\end{remark}
-
-\begin{definition}
-  $\tau$ is a \emph{$\PP$-name} if its elements are of the form
-  $\lb\sigma,p\rb$ where $p\in\PP$ and $\sigma$ is a $\PP$-name.
-\end{definition}
-The usual way to define a property by well-founded recursion is to
-define the ``characteristic function'' of that property. In this case,
-define by recursion the function $F:M\to\om$ such that $F(x)=1$  if
-$x$ is a $\PP$-name, or $F(x)=0$ otherwise.
-
-\begin{definition}
-  The \emph{extension of $M$ by $G$} is
-  \[
-  \verb|gen_ext|(M,\PP,G) = M[G] \defi \{\val(\tau,\PP,G): \tau\in M\}.
-  \]
-\end{definition}
-The standard definition of $M[G]$ adds the clause that $\tau$ should be
-a $\PP$-name, but this is not necessary. When $G$ is $M$-generic for
-$\PP$, we say that $M[G]$ is a \emph{generic extension}.
-
-\section{The forcing relation}
-In order to prove the independence of $\CH$, we will assume the
-existence of countable transitive model (ctm) $M$ of $\ZFC$. The
-satisfaction relation $M,\mathit{env}\models\Phi$ (where $\Phi$ is a set of
-sentences and $\mathit{env}$ is an interpretation for the free
-variables) is coded in Isabelle/ZF by using \verb|satT|:
-\begin{verbatim}
-definition
-  satT :: "[i,i,i] => o" where
-  "satT(A,Φ,env) == ∀ p ∈ Φ. sats(A,p,env)"
-\end{verbatim}
-The function \verb|satT| takes a subset of \formula{} as its second
-argument, and it is based on the already defined function \verb|sats|.
-
-\begin{framed}
-  It might be necessary to redefine \verb|sats| and the related
-  function \verb|satisfies| in order to have an independent argument $B$
-  to be the range of values of the environment. We would have, for
-  example, 
-\begin{verbatim}
-consts   satisfies :: "[i,i]=>i"
-primrec 
-  "satisfies(A,B,Member(x,y)) =
-      (λenv ∈ list(B). bool_of_o (nth(x,env) ∈ nth(y,env)))"
-\end{verbatim}
-  \dots
-
-\noindent and then
-\begin{verbatim}
-abbreviation
-  sats :: "[i,i,i] => o" where
-  "sats(A,p,env) == satisfies(A,A,p)`env = 1"
-\end{verbatim}
-\end{framed}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
+\section{Forcing}
 Given a ctm $M$, and an $M$-generic filter $G\sbq\PP$, the Forcing
 Theorems relate satisfaction of a formula 
-$\phi$ in the generic extension $M[G]$ to the satisfaction of another formula
-$\phi'$ in $M$. The map $\phi\mapsto\phi'$ is defined by recursion on
+$\phi$ in the generic extension $M[G]$ to the satisfaction of another
+formula $\phi'$ in $M$. The map $\phi\mapsto\phi'$ is defined by
+recursion on 
 the structure of $\phi$. The fact that this map works (in particular,
@@
 
-\subsection{Forcing for atomic formulas}\label{sec:forcing-atomic-formulas}
-The base case of the definition of the forcing relation, that is, for
-atomic formulas, is (in)famously the most complicated one. 
-The definition adopted by Kunen~\cite[p.~257]{kunen2011set} is the
-following, with two mutually recursive clauses:
-\begin{multline}\label{eq:def-forcing-equality}
-  \forceisa({t_1= t_2},\PP,\leq,\1,p) \defi 
-  \forall s\in\dom(t_1)\cup\dom(t_2)\;\forall q\leq p :\\
-  \forceisa({s\in t_1},\PP,\leq,\1,q)\lsii 
-  \forceisa({s\in t_1},\PP,\leq,\1,q),
-\end{multline}
-\begin{multline}\label{eq:def-forcing-membership}
-  \forceisa({t_1\in t_2},\PP,\leq,\1,p) \defi  \\
-  \text{``}\{q\leq p : \exists \lb s,r\rb\in t_2 (q\leq r \y
-  \forceisa({t_1 =s},\PP,\leq,\1,q))\} \\
-  \text{is dense below $p$.''}
-\end{multline}
-Other approaches, like the one in Neeman~\cite{neeman-course}, prefer
-to have a single, more complicated, clause for $t_1=t_2$ using
-well-founded recursion on $t_1$
-and $t_2$ and then define the case $t_1\in t_2$ explicitly.
-
-In any case, these definitions assume that $t_i$ are \emph{names}.
-In fact, the intuitive definition is by recursion on \emph{sentences}
-that are written in the language $\{\in\}$ plus every name as a
-constant; thus they
-depend on the theorem of well-founded recursion,
-so $\forceisa(\phi,\PP,\leq,\1,p)$ for atomic $\phi$
-must be obtained from the proof of that theorem. Below, a defined
-relation $R$ is \emph{set-like} if $R^{-1}\{a\}$ is a set for every $a$.
-%
-\begin{theorem}[{\cite[I.9.11]{kunen2011set}}]
-  Assume that $R$ is a
-  set-like well-founded defined relation on a class $A$, assume that 
-  $\forall  x,s\,\exists!y\,\phi(x,s,y)$ and write $G(x,s)$ for the
-  only $y$ such that $\phi(x,s,y)$. Then there exists another formula 
-   $\psi$ such that $\forall x\,\exists!y\,\psi(x,y)$ and if we write
-  $F(x)$ for the only $y$ satisfying $\psi(x,y)$, we have
-  \[
-  \forall a\in A: F(a) = G(a,F\restr R^{-1}\{a\}).
-  \]
-\end{theorem}
-%
-Observe that since $F$ is a definable function, its restriction to the
-set $R^{-1}\{a\}$ must also be a set by the Replacement Axiom.
-
-There are several steps to apply this theorem to justify the
-definition of forcing for atomic formulas:
-\begin{enumerate}
-\item As mentioned before, $F$ will be the characteristic function of
-  the predicate $\forceisa$.
-  %
-\item The class $A$ and the relation $R$ must be
-  defined.
-  %
-\item Write down explicitly $\psi$ as an element of
-  $\formula$. 
-\end{enumerate}
-%
-Luckily, the formula $\psi$ is spelled out in Kunen's.
-
-The  function $F$ will essentially have 
-the same  arguments  of $\forceisa$, but we will deconstruct the
-formula and pack everything as a tuple. Namely, we will have
-\begin{align}
-  F(\lb\equ,t_1,t_2,\PP,\leq,\1,p\rb) =1 &\iff  
-  \forceisa(t_1=t_2,\PP,\leq,\1,p) \label{eq:forcing-F-equ}\\
-  F(\lb\bel,s,t_1,\PP,\leq,\1,p\rb) =1 &\iff  
-  \forceisa(s\in t_1,\PP,\leq,\1,q). \label{eq:forcing-F-bel}
-\end{align}
-Here, $\equ$ and $\bel$ are absolute constants that encode which type of
-formula we are considering (for instance, $0$ and $1$, 
-respectively). We can choose our $A$ to be 
-$\{\equ,\bel\}\times V^6$ (i.e., all 7-tuples whose first coordinate
-are  $\equ$ or $\bel$).
-
-Rewriting the clauses (\ref{eq:def-forcing-equality}) and
-(\ref{eq:def-forcing-membership}), we might obtain a first
-approximation to $G$. Notice that its domain is $A\times V$.
-
-$G(\lb\atr,t_1,t_2,\PP,\leq,\1,p\rb,f) =1$ if one of the following hold:
-\begin{itemize}
-\item $\atr=\equ$ and   $\forall s\in\dom(t_1)\cup\dom(t_2)\;\forall q\leq p :
-  f(\lb\bel,s,t_1,\PP,\leq,\1,q\rb) =1  \lsii
-  f(\lb\bel,s,t_2,\PP,\leq,\1,q\rb) =1$; or
-\item $\atr=\bel$ and $\{q\leq p : \exists \lb s,r\rb\in t_2 (q\leq r \y
-  f(\lb\equ,t_1,s,\PP,\leq,\1,q\rb) =1)\}$ is dense below $p$.
-\end{itemize}
-In any other case, $G(\lb\atr,t_1,t_2,\PP,\leq,\1,p\rb,f) =0$. In the
-above, $f(v)=1$ should be interpreted as $\lb v,1\rb\in f$ (i.e., $f$
-must be defined for the argument at hand). From this scheme an element
-of $\formula$ must be produced. (This is the reason why  $\equ,\bel$
-were chosen to be definable.)
-
-The definition of the relation $R$ is straightforward and follows the
-pattern of \cite[p.~257]{kunen2011set}. For instance, we have that 
-\[
-\lb\bel,s_1,t_1,\PP,\leq,\1,q\rb\mathrel{R}
-\lb\equ,s_2,t_2,\PP,\leq,\1,p\rb
-\]
-if and only if 
-\begin{itemize}
-\item $s_1 \reee  s_2$ or $s_1\reee t_2$; and
-\item $t_1 = s_2$ or $t_1 = t_2$.
-\end{itemize}
-%
-\begin{framed}
-  It seems necessary to prove that the relation $\reee$ is
-  well-founded independently of the universe $M$ where one is
-  working. We only
-  have a proof that ${\reee}(M)$ is well-founded for each $M$.
-\end{framed}
-
-With these elements we might reproduce the formula $\psi$ that defines
-$F$ as in \cite[p.~49]{kunen2011set}. Let  $\app(d,h)$  be the formula
-that states:
-\begin{enumerate}
-\item \label{item:dom-R-cerrado} $d$ is ``downwards $R$-closed'':
-  $\forall y \in d, 
-  R^{-1}\{y\}\subseteq d$;
-\item \label{item:4} $h$ is a function with $\dom h =d \subseteq A$; and
-\item \label{item:3} for all $y\in d$, $h(y) = G(y,h\restr(R^{-1}\{y\}))$.
-\end{enumerate}
-Then the formula defining $F$ is the following:
-\[
-(x\notin A \y y=0) \o \bigl(x\in A \y \exists d,h\, (\app(d,h)\y
-x\in d \y h(x)=y)\bigr).
-\]
-The well-foundedness of $R$ is used to show that this formula is indeed
-functional, and that it satisfies the required recursive equation
-involving $G$. The following intermediate result is needed in the
-sequel.
-\begin{lemma}[Uniqueness of approximations]\label{lem:uniq-approx}
-  Assume $\app(d,h)$ and $\app(d',h')$. Then
-  $h(x)=h'(x)$ for all  $x\in d \cap d'$. In particular,
-  \[
-  h\restr(\{x\}\cup R^{-1}\{x\})=h'\restr(\{x\}\cup R^{-1}\{x\})
-  \]
-  for  all $x\in d \cap d'$. 
-\end{lemma}
-
-Once this is all translated to an element of \formula,
-we'll be able to write the base case of the recursive definition of
-$\forceisa$ using Equations~(\ref{eq:forcing-F-equ})
-and~(\ref{eq:forcing-F-bel}).
-
-\subsection{Inductive step}
-%
-The nexts steps in the definition of $\forceisa$ are straightforward
-\cite[IV.2.42]{kunen2011set}. We just highlight some of the cases.
-\begin{itemize}
-\item $\forceisa(\neg\phi,\PP,\leq,\1,p) \defi \neg\exists
-  q\leq p \,(\neg \forceisa(\phi,\PP,\leq,\1,q))$.
-\item
-  $\forceisa(\phi\y\psi,\dots)\defi\forceisa(\phi,\PP,\leq,\1,p)\y\forceisa(\psi,\PP,\leq,\1,p)$.
-\item $\forceisa(\forall x\,\phi(x),\PP,\leq,\1,p)\defi \forall x\,\forceisa(\phi(x),\PP,\leq,\1,p)$. 
-\item
-  $\forceisa(\phi\o\psi,\dots)\defi\text{``}\{q:\forceisa(\phi,\PP,\leq,\1,q)\o\forceisa(\psi,\PP,\leq,\1,q)\}$
-  is dense below $p$''. 
-\end{itemize}
-%
-Actually, recursion over $\formula$ uses only one propositional
-connective (``nand'') so we need the combination of the first two to
-obtain $\forceisa(\mathtt{Nand(\phi,\psi)},\dots)$:
-\[
-\neg\exists
-  q\leq p :\neg\bigl(\forceisa(\phi,\PP,\leq,\1,q)\y\forceisa(\psi,\PP,\leq,\1,q)\bigr).
-\]
-
-\section{The fundamental theorems}
-These are stated as two lemmas: \emph{Truth} and
-\emph{Definability}. The second one ``simply'' says that the forcing
-relation is definable. The previous recursive construction amounts to
-that. The Truth Lemma states that the forcing relation indeed relates
-satisfaction in $M[G]$ to that in $M$. 
-\begin{lemma}[Truth Lemma]\label{lem:truth-lemma}
-  Let $M$ be a ctm of $\ZFC$, $\lb\PP,\leq,\1\rb$ a forcing notion
-  in $M$, $p\in\PP$, and $\phi(x_1,\dots,x_n)$ a formula in the
-  language of set 
-  theory with all free variables displayed. Then the
-  following are equivalent, for all $\tau_1,\dots,\tau_n\in M$: 
-  \begin{enumerate}
-  \item     $M[G]\models\phi(\val(\tau_1,G),\dots,\val(\tau_n,G))$ for
-    all filters $G$, $M$-generic for $\PP$, such that $p\in G$.
-  \item $M\models\forceisa(\phi(\tau_1,\dots,\tau_n),\PP,\leq,\1,p)$.
-  \end{enumerate}
-\end{lemma}
-A bit more formally, one should write the parameters to the left of
-$\models$:
-\begin{gather*}
-  \forall G : p\in G \limp M[G],[\val(\tau_1,G),\dots,\val(\tau_n,G)]
-  \models\phi(x_1,\dots,x_n) \\
-  \text{if and only if}\\
-  M,[\tau_1,\dots,\tau_n,\PP,\leq,\1,p] 
-  \models\forceisa(\phi(x_1,\dots,x_n),x_{n+1},x_{n+2},x_{n+3},x_{n+4})
-\end{gather*}
-The proof of the Truth Lemma is by recursion in elements of
-$\formula$. The following auxiliary results (adapted from
-\cite[IV.2.43]{kunen2011set}) are also proved by recursion in
-$\formula$.
-\begin{lemma}\label{lem:strengthen} 
-  Assume the same hypothesis of Lemma~\ref{lem:truth-lemma}.
-  $M\models\forceisa(\phi,\PP,\leq,\1,p)$ and $p_1\leq p$
-  implies $M\models\forceisa(\phi,\PP,\leq,\1,p_1)$.
-\end{lemma}
-\begin{lemma}\label{lem:density}
-  Assume the same hypothesis of Lemma~\ref{lem:truth-lemma}. $M\models\forceisa(\phi,\dots)$ if and only if
-    $M\models$``$\{p_1\leq p : \forceisa(\phi,\PP,\leq,\1,p_1)\}$ is
-    dense below $p$.''
-\end{lemma}
-\begin{lemma}\label{lem:forcing-negation}
-  Under the same hypothesis of Lemma~\ref{lem:truth-lemma}, we have
-  $M\models\forceisa(\phi,\dots)$ if and only if 
-  $M\models\neg\exists
-  q\leq p \,(\neg \forceisa(\phi,\PP,\leq,\1,q))$.
-\end{lemma}
-
-\subsection{Truth for atomic formulas}
-We start by proving  Lemma~\ref{lem:strengthen}.% and~\ref{lem:density}.
-\begin{proof}[Proof of  Lemma~\ref{lem:strengthen} (atomic formulas)]
-  Let $M$, $\lb\PP,\leq,\1\rb$, $p\in\PP$ as in the hypothesis of
-  Lemma~\ref{lem:truth-lemma}, let $p_1\leq p$ and let $\phi(x_1,x_2)$
-  be atomic. We 
-  have to show that for every    $\tau_1,\tau_2\in M$, if
-  \begin{equation} \label{eq:M-phi-atom}
-    M,[\tau_1,\tau_2,\PP,\leq,\1,p] 
-    \models\forceisa(\phi(x_1,x_2),x_{3},x_{4},x_{5},x_{6})
-  \end{equation}
-  then the same holds for the environment
-  $[\tau_1,\tau_2,\PP,\leq,\1,p_1]$. For this, we will deconstruct the
-  definition of $\forceisa$ in several steps to pin down where the
-  hypothesis of $p_1\leq p$ might be used.
-
-  So, let $\phi$ be $x_1\in x_2$ and assume (\ref{eq:M-phi-atom}). By
-  our definition of $\forceisa$  for atomic formulas, we have
-  $F(\bx)=1$, where
-  $\bx=\lb\equ,\tau_1,\tau_2,\PP,\leq,\1,p\rb$. Expanding the
-  definition of $F$, we obtain that
-  \[
-  (\bx\notin A \y 1=0) \o \bigl(\bx\in A \y \exists d,h\, (\app(d,h)\y
-  \bx\in d \y h(\bx)=1)\bigr)
-  \]
-  holds in $M$. 
-  \begin{framed}
-    By using appropriate absoluteness results we might translate this
-    statement written as an element of $\formula$ to a sentence
-    written in Isabelle/ZF first order logic, making theorems to
-    simplify the statement available, as in the arguments
-    that follow.
-  \end{framed}
-  \noindent Simplifying, we obtain 
-  \begin{equation}\label{eq:1}
-    \exists d,h\, (\app(d,h)\y   \bx\in d \y h(\bx)=1).
-  \end{equation}
-  Take such $d,h\in M$; these will be added to the environment. We
-  will define $d',h'$ that satisfy 
-  $\app(d',h')\y   \by\in d' \y h(\by)=1$ for
-  $\by=\lb\equ,\tau_1,\tau_2,\PP,\leq,\1,p_1\rb$, thereby proving
-  $F(\by)=1$ and hence the thesis.
-
-  By  definition of $\app$ and the fact that $\bx\in d$, we have
-  \[
-  1= h(\bx) = G(\bx,h\restr(R^{-1}\{\bx\})).
-  \]
-  Now the  definition of $G$ for $\bx=\lb\equ,\dots\rb$, yields
-  \[
-  M,[d,h,\tau_1,\tau_2,\PP,\leq,\1,p]\models \forall
-  s\in\dom(x_1)\cup\dom(x_2)\;\forall q\leq x_6 : \Phi
-  \]
-  with $\Phi\defi
-  f(\lb\bel,s,x_1,x_{3},x_{4},x_{5},x_{6}\rb) =1  \lsii
-  f(\lb\bel,s,x_2,x_{3},x_{4},x_{5},x_{6}\rb) =1$ and 
-  $f\defi h\restr(R^{-1}\{\bx\})$.
-  By absoluteness we have
-  \[
-  \forall
-  \sigma\in\dom(\tau_1)\cup\dom(\tau_2)\;\forall q'\leq p :
-  M,[\sigma,q',d,h,\tau_1,\tau_2,\PP,\leq,\1,p]\models  \Phi.
-  \]
-  Since $q'\leq p_1$ implies $q'\leq p$, we have
-  \[
-  \forall
-  \sigma\in\dom(\tau_1)\cup\dom(\tau_2)\;\forall q'\leq p_1 :
-  M,[\sigma,q',d,h,\tau_1,\tau_2,\PP,\leq,\1,p_1]\models  \Phi.
-  \]
-  Using absoluteness again we have
-  \[
-  M,[d,h,\tau_1,\tau_2,\PP,\leq,\1,p_1]\models \forall
-  s\in\dom(x_1)\cup\dom(x_2)\;\forall q\leq p_1 : \Phi.
-  \]
-  Hence we obtain $G(\by,h\restr(R^{-1}\{\bx\}))=1$.
-
-  Now define   $d'\defi d\cup \{\by\}$  and $h'\defi
-  h\cup\{\lb\by,1\rb\}$. Since
-  $R^{-1}\{\bx\} =R^{-1}\{\by\}$ and $\by\in A$, we obtain
-  $\app(d',h')$. This, together with the fact that $h(\by)=1$, shows
-  $F(\by)=1$ and hence (\ref{eq:M-phi-atom}).
-
-  The proof for $\phi= x_1\in x_2$ should be similar.
-\end{proof}
-Lemma~\ref{lem:density} also follows the same pattern, there is a bit
-more of mathematics to it. The same goes for
-Lemma~\ref{lem:forcing-negation}.
-
-We restate Lemma~\ref{lem:truth-lemma} in an alternative way, for
-atomic formulas.
-\begin{lemma}
-  Let $M$ be a ctm of $\ZFC$, $\lb\PP,\leq,\1\rb$ a forcing notion
-  in $M$, $p\in\PP$, $G\sbq\PP$ $M$-generic for $\PP$ and $\phi(x_1,x_2)$ an
-  atomic formula. Then the following hold for all
-  $\tau_1,\tau_2\in M$:
-  \begin{enumerate}
-  \item If $p\in G$ and
-    \begin{equation} \label{eq:M-sats-forces-atom}
-      M,[\tau_1,\tau_2,\PP,\leq,\1,p] 
-      \models\forceisa(\phi(x_1,x_2),x_{3},x_{4},x_{5},x_{6})
-    \end{equation}
-    then
-    \begin{equation} \label{eq:MG-sats-atom}
-      M[G],[\val(\tau_1,G),\val(\tau_2,G)]
-      \models\phi(x_1,x_2).
-    \end{equation}
-  \item If (\ref{eq:MG-sats-atom}) holds 
-    then there is some $p\in G$ such that
-    (\ref{eq:M-sats-forces-atom}) holds for that $p$.
-  \end{enumerate}
-\end{lemma}
-This proof consists of two cases for $\phi$ (namely, $x_1\in x_2$ and
-$x_1=x_2$) and it is done by well-founded recursion over $R$ (defined
-in Section~\ref{sec:forcing-atomic-formulas}).
-
-\section{The axioms in the generic extension}
-\subsection{The axiom of Separation}
-In this subsection we will show that the Separation Axiom scheme holds
-in the generic extension $M[G]$.
-
-\begin{lemma}
-  Let $G$ be $M$-generic for $\PP$. For every $\phi\in\formula$ with
-  three free variables, the $\phi$-instance of Separation holds in
-  $M[G]$.
-\end{lemma}
-\begin{proof}
-  Let $G$ as in the hypothesis, $\phi(x,z,v)$ be in \formula{} and let
-  $c,w\in M[G]$. Following Kunen \cite[Lemma IV.2.6]{kunen2011set},
-  we have to show that 
-  \[
-  S\defi\{x\in c : M[G]\models \phi(x,c,w)\} \in M[G].
-  \]
-  From this, using absoluteness, we will be able to derive the
-  $\phi$-instance of Separation. Let $\pi,\sig\in M$ such that 
-  $\val(\pi)=c$ and $\val(\sig)=w$.
-  We will show that the name
-  \[
-  \tau \defi \{\lb\th,p\rb: \th\in\dom(\pi) \y p\in \PP\y
-  M\models p\forces(\th\in\pi\y\phi(\th,\pi,\sig)\}
-  \]
-  belongs to $M$; for this we will use that an appropriate instance of
-  Separation holds in $M$. Later we will see that $\tau$ names $S$.
-
-  First, we slightly rewrite $\tau$:
-  \[
-  \tau \defi \{w\in\dom(\pi)\times\PP :\exists \th,p (w=\lb\th,p\rb \y 
-  M\models p \forces(\th\in\pi\y\phi(\th,\pi,\sig))\}
-  \]
-
-  The formula giving the relevant instance of Separation in $M$ will be the
-  following:
-  \[
-  \psi \defi \exists \th,p : x=\lb\th,p\rb \y \forceisa(\PP,\leq,\1,p,\th\in\pi\y\phi(\th,\pi,\sig)),
-  \]
-  where $\PP,\leq,\1,\pi,\sig$ are parameters.
-  \begin{framed}
-    I'm changing the order of arguments of $\forceisa$. Also, we
-    will need to show that, in $M$, a version of separation for formulas
-    with  $5$ free variables holds.
-
-    Also, we need to show that $\dom(\pi)\times\PP\in M$.
-  \end{framed}
-  Writing this in de Bruijn notation, gives us something like this:
-  \[
-  \psi \defi  \exists\,\exists\,\verb|pair_fm|(\dB{0},\dB{1},\dB{2}) \y
-  \forceisa(\dB{4},\dB{5},\dB{6},\dB{7},\dB{0}\in \dB{1}\y\phi(\dB{0},\dB{1},\dB{3})).
-  \]
-  From $M\models \mathtt{Separation}(\psi)$ we have
-  \begin{multline*}
-    \forall u,l,P,s,r\in M \forall d\in M:\\ 
-    M,[d,r,s,P,l,u] \models\exists y \forall x
-    (x\in y \lsii x\in d \y \psi(P,l,u,x,r)).
-  \end{multline*}
-  We instantiate the universal quantifiers using the given parameters
-  and the domain $\dom(\pi)\times\PP$, while changing the free variables in
-  the formula with de Bruijn indices corresponding to the order in the
-  environment.
-  \[
-  M,[\dom(\pi)\times\PP,\pi,\sig,\PP,\leq,\1] \models\exists y \forall x
-  (x\in y \lsii x\in \dB{0} \y \psi(\dB{3},\dB{4},\dB{5},x,\dB{1})).
-  \]
-  Extracting the two outermost quantifiers and shifting indices, we get
-  \begin{multline*}
-    \exists y\in M \forall x\in M: \\
-    M,[x,y,\dom(\pi)\times\PP,\pi,\sig,\PP,\leq,\1] \models
-    \dB{0}\in \dB{1} \lsii \dB{0}\in \dB{2} \y \psi(\dB{5},\dB{6},\dB{7},\dB{0},\dB{3}).
-  \end{multline*}
-  Let $\tau\in M$ witness the $\exists y$. So, by absoluteness,
-  \begin{multline*}
-    \forall x\in M: 
-    x\in\tau \lsii  x\in\dom(\pi)\times\PP \y{}\\ 
-  M,[x,\tau,\dom(\pi)\times\PP,\pi,\sig,\PP,\leq,\1] \models
-  \psi(\dB{5},\dB{6},\dB{7},\dB{0},\dB{3}).
-  \end{multline*}
-  Let $\mathit{env}\defi
-  [x,\tau,\dom(\pi)\times\PP,\pi,\sig,\PP,\leq,\1]$. By transitivity
-  of $M$ we can dispense of the bound on $x$, and we get
-  \[
-  \tau = \{x\in\dom(\pi)\times\PP :
-  M, \mathit{env}\models
-  \psi(\dB{5},\dB{6},\dB{7},\dB{0},\dB{3})\},
-  \]
-  i.e., the relevant name is in $M$.
-
-  Now we turn to the proof that $\val(\tau)=S$.
-
-  Let's first focus into the predicate defining $\tau$ by separation. By expanding $\psi$ we have
-  \[
-   M, \mathit{env}\models
-   \exists \th,p : \dB{0}=\lb\th,p\rb \y
-   \forceisa(\dB{5},\dB{6},\dB{7},p,\th\in \dB{3}\y\phi(\th,\dB{3},\dB{4})).
-  \]
-  We proceed as before, using absoluteness and replacing variables by
-  corresponding indices.
-  \[
-  \exists \th,p : x =\lb\th,p\rb \y  M, [\th,p]\concat\mathit{env}\models
-  \forceisa(\dB{7},\dB{8},\dB{9},\dB{1},\dB{0}\in \dB{5}\y\phi(\dB{0},\dB{5},\dB{6})).
-  \]
-
-  Considering that only a fraction of elements
-  of the environment  are necessary to interpret
-  $\forceisa(\dB{7},\dB{8},\dB{9},\dB{1},\dB{0}\in\pi\y\phi(\dB{0},\dB{5},\dB{6}))$ and by permuting, we can
-  replace
-  \[
-  M, [\th,p,x,\tau,\dom(\pi)\times\PP,\pi,\sig,\PP,\leq,\1]\models
-  \forceisa(\dB{7},\dB{8},\dB{9},\dB{1},\dB{0}\in \dB{5} \y\phi(\dB{0},\dB{5},\dB{6}))
-  \]
-  by
-  \[
-  M,[\PP,\leq,\1,p,\th,\pi,\sig]\models
-  \forceisa(\dB{0},\dB{1},\dB{2},\dB{3},\dB{4}\in \dB{5} \y\phi(\dB{4},\dB{5},\dB{6}))
-  \]
-  By the Truth Lemma, this is equivalent to
-  \[
-  \forall F\ \PP\text{-generic}, p\in F \limp
-  M[F],[\val(\th,F),\val(\pi,F),\val(\sig,F)]\models
-  \dB{0}\in\dB{1}  \y\phi(\dB{0},\dB{1},\dB{2}).
-  \]
-
-  Now we calculate $\val(\tau,F)$.
-  \begin{multline*}
-    \val(\tau,G) = \\
-    \val\bigl(\{x\in\dom(\pi)\times\PP : \exists \th,p : x =\lb\th,p\rb
-    \y  \\
-    \forall F\ \PP\text{-generic}, p\in F \limp
-    M[F],[\val(\th,F),\val(\pi,F),\val(\sig,F)]\models
-    \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2})\}\bigr).
-  \end{multline*}
-  Hence
-  \begin{multline*}
-    \val(\tau,G) = 
-    \{\val(\th,G): \th\in\dom(\pi) \y  p \in G 
-    \y \forall F\ \PP\text{-generic}, \\
-   p\in F \longrightarrow
-  M[F],[\val(\th,F),\val(\pi,F),\val(\sig,F)]\models
-  \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2})\}.
-  \end{multline*}
-  This is included in
-  \begin{multline*}
-    \{\val(\th,G): \th\in\dom(\pi) \y  p \in G 
-    \y  \\
-   p\in G \longrightarrow
-   M[G],[\val(\th,G),\val(\pi,G),\val(\sig,G)]\models
-   \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2})\},
-  \end{multline*}
-  and in turn,  replacing $\val(\pi,F),\val(\sig,F)$ by their values,
-  this is included  in 
-  \[ 
-    \{\val(\th,G): \th\in\dom(\pi)
-    \y 
-   M[G],[\val(\th,G),c,w]\models
-   \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2})\}.
-  \]
-  By absoluteness, we have that $\val(\tau,G)$ is
-  \[
-    \{\val(\th,G): \th\in\dom(\pi)
-    \y \val(\th,G)\in c \y 
-   M[G],[\val(\th,G),c,w]\models
-   \phi(\dB{0},\dB{1},\dB{2})\}.
-  \]
-  which equals 
-  \[
-    \{x: x\in c \y 
-    M[G],[x,c,w]\models
-   \phi(\dB{0},\dB{1},\dB{2})\}.
-  \]
-  since for every $x\in c=\val(\pi,G)$ there must
-  exist some $\th\in\dom(\pi)$ with $\val(\th,G)=x$. Therefore
-  $\val(\tau,G)\subseteq S$.
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliographystyle{alpha}
-\bibliography{blueprint}
-
-\bigskip
-
-\begin{small}
-  \begin{quote}
-    \texttt{sterraf@famaf.unc.edu.ar}
-
-    CIEM --- Facultad de Matem\'atica, Astronom\'{\i}a y F\'{\i}sica 
-    (Fa.M.A.F.)
-
-    Universidad Nacional de C\'ordoba --- Ciudad Universitaria
-
-    C\'ordoba 5000. Argentina.
-  \end{quote}
-\end{small}
-
-\end{document}
-
 %%% Local Variables: 
 %%% mode: latex
-%%% ispell-local-dictionary: "american"
+%%% TeX-master: "Separation_In_MG"
 %%% End:

### 2018-11-30 LICS/intro.tex
modularizando el manuscrito

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
\section{Introduction}
% no \IEEEPARstart
We follow \cite{kunen2011set}. An introduction to the issues discussed
in this paper can be found in \cite{2018arXiv180705174G}. An
introduction to forcing can be found at \cite{chow-beginner-forcing},
and the book \cite{weaver2014forcing} contains a thorough treament
minimizing the technicalities.

An enumeration of achievements to be presented:
\begin{enumerate}
\item Discussion of the Isabelle view of life.
  \begin{enumerate}
  \item Meta (Pure) vs Object (FOL+ZF) theory.
  \item Object theory is presented axiomatically; in particular,
    first-order ``formulas'' are not such (i.e., the \emph{initial}
    algebra of formulas is \textbf{not} being specified). Thus, we are
    not able to perform recursion on terms of type \tyo.
  \end{enumerate}
\item Discussion of locales.
  \begin{enumerate}
  \item How they help presenting complex results.
  \item Their key role in stating the Fundamental Theorems (without
    defining \isatt{forces}).
  \item \isatt{forces} is defined for elements of \formula.
  \end{enumerate}
\item Enhancements in the recursion machinery.
  \begin{enumerate}
  \item Impact on names.
  \end{enumerate}
\item Hacking of \isatt{ZF-Constructible}.
  \begin{enumerate}
  \item Elimination of the \tyo-schematic Replacement Axiom from
    locales.
  \item Replacing \isatt{M(nat)} by \isatt{M(0)} (merely non
    emptiness) from Relative (and other places?).
  \end{enumerate}
\item Inteface between first-order properties of $M$ and those of
  $\#\#M$.
  \begin{enumerate}
  \item Enumerate properties and statistics.
  \end{enumerate}
\item Details of the proof of Separation.
\end{enumerate}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "Separation_In_MG"
%%% End:

### 2018-12-01 LICS/proof-of-separation.tex
primeros pasos de una explicación de separación

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
\section{Proof of Separation}

This proof can be found in the file \verb|Separation_Axiom.thy| of the
development, which we proceed to discuss.

The key technical result is the following:
\begin{isabelle}
  \isacommand{lemma}\isamarkupfalse%
  \ Collect{\isacharunderscore}sats{\isacharunderscore}in{\isacharunderscore}MG\ {\isacharcolon}\isanewline
  \ \ \isakeyword{assumes}\isanewline
  \ \ \ \ {\isachardoublequoteopen}{\isasympi}\ {\isasymin}\ M{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymsigma}\ {\isasymin}\ M{\isachardoublequoteclose}\ {\isachardoublequoteopen}val{\isacharparenleft}G{\isacharcomma}\ {\isasympi}{\isacharparenright}\ {\isacharequal}\ c{\isachardoublequoteclose}\ {\isachardoublequoteopen}val{\isacharparenleft}G{\isacharcomma}\ {\isasymsigma}{\isacharparenright}\ {\isacharequal}\ w{\isachardoublequoteclose}\isanewline
  \ \ \ \ {\isachardoublequoteopen}{\isasymphi}\ {\isasymin}\ formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}arity{\isacharparenleft}{\isasymphi}{\isacharparenright}\ {\isasymle}\ {\isadigit{2}}{\isachardoublequoteclose}\isanewline
  \ \ \isakeyword{shows}\ \ \ \ \isanewline
  \ \ \ \ {\isachardoublequoteopen}{\isacharbraceleft}x{\isasymin}c{\isachardot}\ sats{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}\ {\isasymphi}{\isacharcomma}\ {\isacharbrackleft}x{\isacharcomma}\ w{\isacharbrackright}{\isacharparenright}{\isacharbraceright}{\isasymin}\ M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}
\end{isabelle}
%
From this, using absoluteness, we will be able to derive the
$\phi$-instance of Separation.

To show that   
\[
S\defi\{x\in c : M[G]\models \phi(x,w)\} \in M[G],
\]
it is enough to provide a name $n\in M$ for this set.

The candidate name is
\[
n \defi \{u \in\dom(\pi)\times\PP :M,[u,\PP,\leq,\1,\sig,\pi]\models \psi\}
\]
where
\[
\psi \defi \exists \th\, p.\ x_0=\lb\th,p\rb \y 
   \forceisa(\th\in x_5\y\phi(\th,x_4)).
\]
The fact that $n\in M$ follows by an application of a six-variable
instance of Separation in $M$ (lemma \isatt{six{\isacharunderscore}sep{\isacharunderscore}aux}).

Almost a third part of the proof involves the syntactic handling of
internalized formulas and permutation of variables. The more
substantive portion concerns proving that actually $\val(G,n)=S$.

Let's first focus into the predicate 
\[
M,[u,\PP,\leq,\1,\sig,\pi]\models \psi
\]
defining $n$ by separation. By definition of the satisfaction
relation and permuting variables, we have it is equivalent to the fact
that there exist $\th,p\in M$ with   $u=\lb\th,p\rb$  and 
\[
M,[\PP,\leq,\1,p,\th,\sig,\pi]\models \forceisa(x_4\in
x_6\y\phi(x_4,x_5)). 
\]
% (Note that the variable $x_7$ is not used.)
This, in turn is equivalent by the Definition of Forcing to: For all $M$-generic
filters $F$ such that $p\in F$, 
\[
M,[\val(G,\th),\val(G,\sig),\val(G,\pi)]\models x_4\in
x_6\y\phi(x_4,x_5). 
\]

%% The proof of
%% \isatt{Collect{\isacharunderscore}sats{\isacharunderscore}in{\isacharunderscore}MG}
%% has three parts:
%% \begin{enumerate}
%% \item Definition of the name $n$;
%% \item Proving that 
%% \item 
%% \end{enumerate}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "Separation_In_MG"
%%% ispell-local-dictionary: "american"
%%% End:

### 2018-12-02 LICS/forcing.tex
ispell y otras cosas
trayendo cosas de blueprint sin ningún reparo

@@
 
+\subsection{The fundamental theorems}
+These are stated as two lemmas: \emph{Truth} and
+\emph{Definability}. The second one ``simply'' says that the forcing
+relation is definable. The previous recursive construction amounts to
+that. The Truth Lemma states that the forcing relation indeed relates
+satisfaction in $M[G]$ to that in $M$. 
+\begin{lemma}[Truth Lemma]\label{lem:truth-lemma}
+  Let $M$ be a ctm of $\ZFC$, $\lb\PP,\leq,\1\rb$ a forcing notion
+  in $M$, $p\in\PP$, and $\phi(x_1,\dots,x_n)$ a formula in the
+  language of set 
+  theory with all free variables displayed. Then the
+  following are equivalent, for all $\tau_1,\dots,\tau_n\in M$: 
+  \begin{enumerate}
+  \item $M[G],[\val(G,\tau_0),\dots,\val(G,\tau_n)]
+  \models\phi(x_0,\dots,x_n)$ for 
+    all filters $G$, $M$-generic for $\PP$, such that $p\in G$.
+  \item $  M,[\PP,\leq,\1,p,\tau_0,\dots,\tau_n] 
+  \models\forceisa(\phi(x_4,\dots,x_{n+4}))$.
+  \end{enumerate}
+\end{lemma}
+The proof of the Truth Lemma is by recursion in elements of
+$\formula$. The following auxiliary results (adapted from
+\cite[IV.2.43]{kunen2011set}) are also proved by recursion in
+$\formula$.
+\begin{lemma}[Strenghtening]\label{lem:strengthen} 
+  Assume the same hypothesis of Lemma~\ref{lem:truth-lemma}.
+  $M, [\PP,\leq,\1,p]\models\forceisa(\phi)$ and $p_1\leq p$
+  implies $M, [\PP,\leq,\1,p_1]\models\forceisa(\phi)$.
+\end{lemma}
+\begin{lemma}[Density]\label{lem:density}
+  Assume the same hypothesis of
+  Lemma~\ref{lem:truth-lemma}. $M\models\forceisa(\phi,\dots)$ 
+  if and only if 
+  $M\models$``$\{p_1\leq p : \forceisa(\phi,\PP,\leq,\1,p_1)\}$ is
+  dense below $p$.''
+\end{lemma}
+
 %%% Local Variables: 
@@
 %%% TeX-master: "Separation_In_MG"
+%%% ispell-local-dictionary: "american"
 %%% End:

### 2018-12-02 LICS/proof-of-separation.tex
trayendo cosas de blueprint sin ningún reparo

@@
 \[
-S\defi\{x\in c : M[G]\models \phi(x,w)\} \in M[G],
+S\defi\{x\in c : M[G],[x,w]\models \phi(x_0,x_1)\} \in M[G],
 \]
@@
 Let's first focus into the predicate 
-\[
+\begin{equation}\label{eq:1}
 M,[u,\PP,\leq,\1,\sig,\pi]\models \psi
-\]
+\end{equation}
 defining $n$ by separation. By definition of the satisfaction
-relation and permuting variables, we have it is equivalent to the fact
+relation and %% permuting variables
+absoluteness, we have that it is equivalent to the fact
 that there exist $\th,p\in M$ with   $u=\lb\th,p\rb$  and 
@@
 % (Note that the variable $x_7$ is not used.)
-This, in turn is equivalent by the Definition of Forcing to: For all $M$-generic
-filters $F$ such that $p\in F$, 
+This, in turn, is equivalent by the Definition of Forcing to: \emph{For all $M$-generic
+filters $F$ such that $p\in F$,} 
+\begin{equation}\label{eq:2}
+M[F],[\val(F,\th),\val(F,\sig),\val(F,\pi)]\models x_0\in
+x_2\y\phi(x_0,x_1). 
+\end{equation}
+We can instantiate this statement with $G$ and obtain
 \[
-M,[\val(G,\th),\val(G,\sig),\val(G,\pi)]\models x_4\in
-x_6\y\phi(x_4,x_5). 
+p\in G \impl M[G],[\val(G,\th),w,c]\models x_0\in
+x_2\y\phi(x_0,x_1). 
+\] 
+Let $Q(\th,p)$ be the last statement. We have just seen that
+(\ref{eq:1}) implies 
+\[
+\exists \th,p\in M.\ u=\lb\th,p\rb \y Q(\th,p).
+\]
+Hence $n$ is included in 
+\[
+m\defi \{u \in\dom(\pi)\times\PP : \exists \th,p\in M.\ u=\lb\th,p\rb
+\y Q(\th,p)\}. 
 \]
 
-%% The proof of
-%% \isatt{Collect{\isacharunderscore}sats{\isacharunderscore}in{\isacharunderscore}MG}
-%% has three parts:
-%% \begin{enumerate}
-%% \item Definition of the name $n$;
-%% \item Proving that 
-%% \item 
-%% \end{enumerate}
+It can be seen that
+\begin{lemma}
+  $\val(G,m) = S$.
+\end{lemma}
+\noindent And by monotonicity of $\val$ we obtain
+\begin{lemma}
+  $\val(G,n)\sbq \val(G,m)$.
+\end{lemma}
+To complete the proof, it is therefore enough to show that
+$S\sbq \val(G,n)$. For this, let $x\in S$. Hence there exists
+$\lb\th,q\rb\in\pi$ such that 
+$q\in G$ and $x=\val(G,\th)$.
+
+On the other hand, since 
+\[
+M[G],[\val(G,\th),\val(G,\sig),\val(G,\pi)]\models
+ x_0\in x_2\y\phi( x_0, x_1),
+\]
+by the  Truth Lemma there must exist $r\in G$ such that
+\[
+M,[\PP,\leq,\1,r,\th,\sig,\pi]\models
+\forceisa(x_4\in x_6\y\phi( x_4, x_5)).
+\]
+Since $G$ is a filter, there is $p\in G$ such that $p\leq q, r$.
+By Strenghtening, we have
+\[
+M,[\PP,\leq,\1,p,\th,\sig,\pi]\models
+\forceisa(x_4\in x_6\y \phi( x_4, x_5)),
+\]
+which by the Truth Lemma gives us: \emph{for all $M$-generic $F$,
+  $p\in F$ implies} 
+\[
+M[F],[\val(F,\th),\val(F,\sig),\val(F,\pi)]\models
+ x_0\in  x_2 \y\phi( x_0, x_1).
+\]
+Note this is the same as (\ref{eq:2}). In fact, we have that
+\begin{lemma}
+  $\val(G,n)$ equals
+  \[
+  \{\val(G,\th): \th\in\dom(\pi) \y  p \in G 
+  \y \text{(\ref{eq:2}) holds}\}.
+  \]
+\end{lemma}
+This shows that $S\sbq \val(G,n)$, since $\th\in\dom(\pi)$ by construction.
 

### 2018-12-02 LICS/forcing.tex
mejoro un poco los enunciados

@@
 \emph{Definability}. The second one ``simply'' says that the forcing
-relation is definable. The previous recursive construction amounts to
-that. The Truth Lemma states that the forcing relation indeed relates
+relation is definable. The Truth Lemma states that the forcing
+relation indeed relates 
 satisfaction in $M[G]$ to that in $M$. 
@@
   Let $M$ be a ctm of $\ZFC$, $\lb\PP,\leq,\1\rb$ a forcing notion
-  in $M$, $p\in\PP$, and $\phi(x_1,\dots,x_n)$ a formula in the
+  in $M$, $p\in\PP$, and $\phi(x_0,\dots,x_n)$ a formula in the
   language of set 
   theory with all free variables displayed. Then the
-  following are equivalent, for all $\tau_1,\dots,\tau_n\in M$: 
+  following are equivalent, for all $\tau_0,\dots,\tau_n\in M$: 
   \begin{enumerate}
@@
   Assume the same hypothesis of Lemma~\ref{lem:truth-lemma}.
-  $M, [\PP,\leq,\1,p]\models\forceisa(\phi)$ and $p_1\leq p$
-  implies $M, [\PP,\leq,\1,p_1]\models\forceisa(\phi)$.
+  $M, [\PP,\leq,\1,p,\dots]\models\forceisa(\phi)$ and $p_1\leq p$
+  implies $M, [\PP,\leq,\1,p_1,\dots]\models\forceisa(\phi)$.
 \end{lemma}
@@
   Assume the same hypothesis of
-  Lemma~\ref{lem:truth-lemma}. $M\models\forceisa(\phi,\dots)$ 
+  Lemma~\ref{lem:truth-lemma}. $M,[\dots]\models\forceisa(\phi)$ 
   if and only if 
-  $M\models$``$\{p_1\leq p : \forceisa(\phi,\PP,\leq,\1,p_1)\}$ is
+  $M,[\dots]\models$``$\{p_1\leq p : \forceisa(\phi,\PP,\leq,\1,p_1)\}$ is
   dense below $p$.''

### 2018-12-02 LICS/forcing.tex
mejoro un poco y pongo convenciones de notación para variables

@@
 $\phi$ in the generic extension $M[G]$ to the satisfaction of another
-formula $\phi'$ in $M$. The map $\phi\mapsto\phi'$ is defined by
+formula $\forceisa(\phi)$ in $M$. The map $\forceisa$ is defined by
 recursion on 
@@
 We will now make more precise the definition of the map
-$\phi\mapsto\phi'$ and how it relates satisfaction in $M$ to that in
+$\forceisa$ and how it relates satisfaction in $M$ to that in
 $M[G]$. Actually, if the formula $\phi$ has $n$ free variables,
-$\phi'$ will have $n+4$ free variables, where the first four account
-for the forcing notion and a particular condition. If
-$\phi=\phi(x_1,\dots,x_n)$, the standard notation for $\phi'$ is
+$\forceisa(\phi)$ will have $n+4$ free variables, where the first four account
+for the forcing notion and a particular condition.
+
+In referring to formulas with free variables and their interpretation,
+we use the following conventions. If  $M$ is a set, $a_0,\dots,a_n$
+are elements of $M$, and $\phi$ has its free variables
+among $x_0,\dots,x_n$, we write
 \[
-p\forces_{\PP,\leq,\1}^* \phi(t_1,\dots,t_n).
+M,[a_0,\dots,a_n] \models \phi(x_0,\dots,x_n)
 \]
-Here, $\PP,\leq,\1,$ and $p\in\PP$ are the extra parameters. 
-%% For the
-%% time being, we just consider $\tau_1,\dots,\tau_n$ to be ordinary
-%% variables. 
-Given $\quine\phi\in\formula$, we will write this  in
-Isabelle/ZF by using  
+to denote that $\phi$ is satisfied by $M$ when $x_i$ is interpreted
+to be $a_i$ ($i=0,\dots,n$). We call the list $[a_0,\dots,a_n]$ the
+\emph{environment}. In the case of a formula of the form
+$\forceisa(\phi)$, we'll make an abuse of notation and indicate the
+variables inside the argument of $\forceisa$. As an example, take the
+formula $\phi\defi x_1\in x_0$. Then
 \[
-\forceisa(\quine\phi,\PP,\leq,\1,p)
+M,[a,b] \models x_1\in x_0
 \]
-Afterwards, the \emph{forcing relation} $\forces$ is defined by
+will hold whenever $b\in a$; and instead of writing $\forceisa(\phi)$
+we will write $\forceisa(x_5\in x_4)$, as in
+\[
+M,[\PP,\leq,\1,p,\tau,\rho] \models \forceisa(x_5\in x_4).
+\]
+
+If
+$\phi=\phi(x_0,\dots,x_n)$, the standard notation for $\forceisa(\phi)$ is
+\[
+p\forces_{\PP,\leq,\1}^* \phi(x_0,\dots,x_n).
+\]
+Here, the extra parameters are $\PP,\leq,\1,$ and $p\in\PP$, and the
+first three are usually omitted. %(since the forcing notion is fixed
+% throughout the discussion).  
+Afterwards, the \emph{forcing relation}
+$\forces$ is defined by 
 relativizing/interpreting $\forces^*$ in a ctm $M$, for fixed
-$\lb\PP,\leq,\1\rb\in M$:
+$\lb\PP,\leq,\1\rb\in M$: $p\forces \phi(\tau_0,\dots,\tau_n)$ holds
+if and only if
 \[ 
-p\forces \phi(t_1,\dots,t_n) \defi 
-\bigl(p\forces_{\PP,\leq,\1}^* \phi(t_1,\dots,t_n)\bigr)^M\!.
+M,[\PP,\leq,\1,p,\tau_0,\dots,\tau_n]\models \forceisa(x_4,\dots,x_{n+4}).
 \]

### 2018-12-02 LICS/forcing.tex
edición general, y sumo el hacking

@@
 Afterwards, the \emph{forcing relation}
-$\forces$ is defined by 
-relativizing/interpreting $\forces^*$ in a ctm $M$, for fixed
+$\forces$ can be obtained by 
+interpreting $\forces^*$ in a ctm $M$, for fixed
 $\lb\PP,\leq,\1\rb\in M$: $p\forces \phi(\tau_0,\dots,\tau_n)$ holds
 if and only if
-\[ 
+\begin{equation}\label{eq:3}
 M,[\PP,\leq,\1,p,\tau_0,\dots,\tau_n]\models \forceisa(x_4,\dots,x_{n+4}).
-\]
+\end{equation}
 
 \subsection{The fundamental theorems}
-These are stated as two lemmas: \emph{Truth} and
-\emph{Definability}. The second one ``simply'' says that the forcing
-relation is definable. The Truth Lemma states that the forcing
-relation indeed relates 
-satisfaction in $M[G]$ to that in $M$. 
-\begin{lemma}[Truth Lemma]\label{lem:truth-lemma}
+These are stated in \cite{kunen2011set} as two lemmas: \emph{Truth} and
+\emph{Definability}.  Modern treatments of the theory of forcing start
+by defining the 
+forcing relation semantically and later it is proved  that the
+characterization given by (\ref{eq:3}) indeed holds, and hence the
+forcing relation is \emph{definable}. In our case, definability merely
+stands for a typing condition:
+\begin{isabelle}
+{\isachardoublequoteopen}{\isasymphi}{\isasymin}formula\ {\isasymLongrightarrow}\ forces{\isacharparenleft}{\isasymphi}{\isacharparenright}\ {\isasymin}\ formula{\isachardoublequoteclose}
+\end{isabelle}
+
+Then the definition of the forcing relation is stated as a
+\begin{lemma}[Definition of Forcing]\label{lem:definition-of-forcing}
   Let $M$ be a ctm of $\ZFC$, $\lb\PP,\leq,\1\rb$ a forcing notion
@@
   \begin{enumerate}
+  \item $M,[\PP,\leq,\1,p,\tau_0,\dots,\tau_n] 
+  \models\forceisa(\phi(x_4,\dots,x_{n+4}))$.
+  \item For all $M$-generic filters $G$ such that $p\in G$,
+    $M[G],[\val(G,\tau_0),\dots,\val(G,\tau_n)] \models\phi(x_0,\dots,x_n)$.
+  \end{enumerate}
+\end{lemma}
+
+The Truth Lemma states that the forcing
+relation indeed relates 
+satisfaction in $M[G]$ to that in $M$. 
+\begin{lemma}[Truth Lemma]\label{lem:truth-lemma}
+  Assume the same hypothesis of
+  Lemma~\ref{lem:definition-of-forcing}. Then the
+  following are equivalent, for all $\tau_0,\dots,\tau_n\in M$, and
+  $M$-generic $G$: 
+  \begin{enumerate}
   \item $M[G],[\val(G,\tau_0),\dots,\val(G,\tau_n)]
-  \models\phi(x_0,\dots,x_n)$ for 
-    all filters $G$, $M$-generic for $\PP$, such that $p\in G$.
-  \item $  M,[\PP,\leq,\1,p,\tau_0,\dots,\tau_n] 
+  \models\phi(x_0,\dots,x_n)$.
+  \item  There exists $p\in G$ such that $M,[\PP,\leq,\1,p,\tau_0,\dots,\tau_n] 
   \models\forceisa(\phi(x_4,\dots,x_{n+4}))$.
@@
 $\formula$.
-\begin{lemma}[Strenghtening]\label{lem:strengthen} 
-  Assume the same hypothesis of Lemma~\ref{lem:truth-lemma}.
+\begin{lemma}[Strengthening]\label{lem:strengthen} 
+  Assume the same hypothesis of Lemma~\ref{lem:definition-of-forcing}.
   $M, [\PP,\leq,\1,p,\dots]\models\forceisa(\phi)$ and $p_1\leq p$
@@
   Assume the same hypothesis of
-  Lemma~\ref{lem:truth-lemma}. $M,[\dots]\models\forceisa(\phi)$ 
+  Lemma~\ref{lem:definition-of-forcing}. $M,[\dots]\models\forceisa(\phi)$ 
   if and only if 

### 2018-12-02 LICS/hacking.tex
edición general, y sumo el hacking

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
\section{Hacking of \isatt{ZF-Constructible}}

There are several lemmas that were declared, in later developments, as
introduction/simplification rules (notably, the rule
\isatt{equalityI}). They raised a warning, and we have 
eliminated some of them, but in some cases we had to keep them because
the proof works because it is insisted that the rule is \emph{safe}.

The most notable changes, located in the theory \texttt{Relative}, are
the following:
\begin{enumerate}
\item\label{item:1} We eliminated the requirement that the relative Axiom of Replacement
  is satisfied by a class $M$ to be in the locale \isatt{M\_trivial}. 
\item\label{item:2} We moved the requirement of the Powerset Axiom to \isatt{M\_basic}. 
\item\label{item:3} We replaced the need that the set of natural numbers is in $M$ by the
  milder hypothesis that $M(0)$.
\end{enumerate}

As a consequence of Item~\ref{item:1}, the lemma
\isatt{strong\_replacementI} is no longer valid and was commented
out.

We moved the requirement $M(\mathtt{nat})$ to the locale
\isatt{M\_trancl}, where it is needed for the first time. Some results,
for instance \isatt{rtran\_closure\_mem\_iff} and 
\isatt{iterates\_imp\_wfrec\_replacement} had to be moved inside that
locale.

The proof, for instance, that the constructible universe $L$ satisfies
the new locale \isatt{M\_trivial\_no\_repl} hold with minor
modifications.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "Separation_In_MG"
%%% ispell-local-dictionary: "american"
%%% End:

### 2018-12-04 LICS/generic-extensions.tex
cambio esa horrible font tt y agrego secciones

--- 2018-12-02 LICS/hacking.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
-\section{Hacking of \isatt{ZF-Constructible}}
+\section{Generic extensions}
 
-There are several lemmas that were declared, in later developments, as
-introduction/simplification rules (notably, the rule
-\isatt{equalityI}). They raised a warning, and we have 
-eliminated some of them, but in some cases we had to keep them because
-the proof works because it is insisted that the rule is \emph{safe}.
-
-The most notable changes, located in the theory \texttt{Relative}, are
-the following:
-\begin{enumerate}
-\item\label{item:1} We eliminated the requirement that the relative Axiom of Replacement
-  is satisfied by a class $M$ to be in the locale \isatt{M\_trivial}. 
-\item\label{item:2} We moved the requirement of the Powerset Axiom to \isatt{M\_basic}. 
-\item\label{item:3} We replaced the need that the set of natural numbers is in $M$ by the
-  milder hypothesis that $M(0)$.
-\end{enumerate}
-
-As a consequence of Item~\ref{item:1}, the lemma
-\isatt{strong\_replacementI} is no longer valid and was commented
-out.
-
-We moved the requirement $M(\mathtt{nat})$ to the locale
-\isatt{M\_trancl}, where it is needed for the first time. Some results,
-for instance \isatt{rtran\_closure\_mem\_iff} and 
-\isatt{iterates\_imp\_wfrec\_replacement} had to be moved inside that
-locale.
-
-The proof, for instance, that the constructible universe $L$ satisfies
-the new locale \isatt{M\_trivial\_no\_repl} hold with minor
-modifications.
+We will swiftly review some definitions in order to reach the concept
+of \emph{generic extension}.
 

### 2018-12-05 LICS/isabelle.tex
algo de isabelle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
\section{Isabelle}
Isabelle provides a meta-language called \emph{Pure} that consists of
a fragment of higher order logic. The meta-Boolean type is called
\isatt{prop}. Meta connectives
\isatt{\isasymLongrightarrow} and \isatt{\&\&\&} fulfill the role of
implication and conjunction, and the meta-binder \isatt{\isasymAnd}
corresponds to universal quantification.

On top of \emph{Pure}, theories/object logics can be defined, with
their own types, connectives and rules. Rules can be written  using
meta-implication: ``$P$, $Q$, and $R$ yield $S$'' can be written
\[
P \ \isatt{\isasymLongrightarrow}\ Q\ \isatt{\isasymLongrightarrow}\ R\ \isatt{\isasymLongrightarrow}\ S
\]
(as usual,  \isatt{\isasymLongrightarrow} asociates to the right), and
syntactic sugar is provided to curry the previous as follows:
\[
\isasymlbrakk P; Q; R \isasymrbrakk \ \isatt{\isasymLongrightarrow}\ S.
\]

We work in the object theory \emph{Isabelle/ZF}. Two types are defined
in this theory: \tyo, the object-Booleans, and \tyi,
sets. It must be insisted that the types are defined axiomatically, not
recursively. That is, although there are constants and functions that
generate elements of both types, neither of them are 
\emph{initial}, in the sense that they are not the least types
obtained by combining the given constants and operations. This will
have concrete consequences in our strategy to approach the development.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "Separation_In_MG"
%%% ispell-local-dictionary: "american"
%%% End:

### 2018-12-06 LICS/generic-extensions.tex
palabras sobre extensiones

@@
 We will swiftly review some definitions in order to reach the concept
-of \emph{generic extension}.
+of \emph{generic extension}. Given a ctm $M$ of $\ZFC$, a forcing
+notion $\lb\PP,\leq,\1\rb$, and 
+$G\in\P(\PP)\sm M$, a new set $M[G]$ is defined. Each element $a\in M[G]$ is
+determined by an element $\dot a$ of $M$. Actually, the structure of
+each $\dot a$ is used to construct $a$. They are related by a
+map $\val$ that takes $G$ a parameter:
+\[
+\val(G,\dot a) = a.
+\] 
+Then the extension is defined by the image of the map $\val(G,\cdot)$:
+\[
+M[G] \defi \{\val(G,\tau): \tau\in M\}.
+\]
+Metatheoretically, it is straightforward to see that $M[G]$ is a
+transitive set that satisfies the axioms of Pairing and Union and
+includes $M\cup\{G\}$. Nevertheless there are no a priori reasons for
+$M[G]$ to satisfy the rest of the $\ZFC$ 
+axioms. The original insight by Cohen was to define the notion of
+\emph{genericity} for a filter $G\sbq\PP$ and to prove that whenever
+$G$ is generic, $M[G]$ will satisfy $\ZFC$.
+
+The Separation Axiom scheme is the first that requires the notion of
+genericity and the use of the forcing machinery, which we review in
+the next section.
 

### 2018-12-06 LICS/forcing.tex
minor corrections

@@
 recursion on 
-the structure of $\phi$. The fact that this map works (in particular,
-in order to show that $M[G]\models\ZFC$), we must use the fact that
-$M\models\ZFC$. Since we can only assert the latter using internalized
-formulas, the aformentioned map will be defined as a function from the
-set \formula{} into itself.
-
-Up to this point, the main reason for working with internalized
-versions is that  it is not possible to do recursion with formulas of
-type $\tyo$ (i.e., the formulas in the first-order logic of
-Isabelle/ZF).
+the structure of $\phi$. In order to state the properties of this map
+in sufficient generality to prove that  $M[G]$ satisfies $\ZFC$, we work with
+internalized formulas, because it is not possible to carry inductive
+arguments over \tyo. These internalized versions are the elements of
+the recursively constructed set \formula{} (of type \tyi).
 
@@
 If
-$\phi=\phi(x_0,\dots,x_n)$, the standard notation for $\forceisa(\phi)$ is
+$\phi=\phi(x_0,\dots,x_n)$, the notation used by Kunen
+\cite{kunen2011set,kunen1980} for $\forceisa(\phi)$ is 
 \[
@@
 \begin{isabelle}
-{\isachardoublequoteopen}{\isasymphi}{\isasymin}formula\ {\isasymLongrightarrow}\ forces{\isacharparenleft}{\isasymphi}{\isacharparenright}\ {\isasymin}\ formula{\isachardoublequoteclose}
+{\isachardoublequoteopen}{\isasymphi}\ {\isasymin}\ formula\ {\isasymLongrightarrow}\ forces{\isacharparenleft}{\isasymphi}{\isacharparenright}\ {\isasymin}\ formula{\isachardoublequoteclose}
 \end{isabelle}

### 2018-12-06 LICS/forcing.tex
edits variados

@@
 \section{Forcing}
+\label{sec:forcing}
+
 Given a ctm $M$, and an $M$-generic filter $G\sbq\PP$, the Forcing

### 2018-12-06 LICS/generic-extensions.tex
digo filtros de entrada
edits variados

@@
 of \emph{generic extension}. Given a ctm $M$ of $\ZFC$, a forcing
-notion $\lb\PP,\leq,\1\rb$, and 
-$G\in\P(\PP)\sm M$, a new set $M[G]$ is defined. Each element $a\in M[G]$ is
+notion $\lb\PP,\leq,\1\rb$, and a filter
+$G\sbq\PP$, a new set $M[G]$ is defined. Each element $a\in M[G]$ is
 determined by an element $\dot a$ of $M$. Actually, the structure of
@@
 genericity and the use of the forcing machinery, which we review in
-the next section.
+the Section~\ref{sec:forcing}.
 

### 2018-12-06 LICS/renaming.tex
Empezando renaming

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
\section{Renaming}

In the course of our work we need to reason about renaming of formulas
and its effect on their satisfiability. Since internalized formulas
are implemented using de Bruijn indices, thus the arity of a formula
$\phi$ gives the least set containing all the free variables in
$\phi$. Following \citet{fiore-abssyn}, one can understand the arity
of a formula as the context of the free variables and renamings are,
consequently, mappings changing the context.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "Separation_In_MG"
%%% ispell-local-dictionary: "american"
%%% End:

### 2018-12-07 LSFA/intro.tex
first-order
acabo de ver un typo

@@
 A relative consistency proof for an axiom $A$ is then obtained by
-assuming that there exists a model of ZFC, say $\lb M,E\rb$, and
+assuming that there exists a model of $\ZFC$, say $\lb M,E\rb$, and
 constructing another model $\lb M',E'\rb$ for $\ZFC + A$. We single
@@
 $\ZFC$. This process is guaranteed to preserve the $\ZFC$
-axioms while allowing to fine-tune what other first order properties the
+axioms while allowing to fine-tune what other first-order properties the
 extension will have. Given a ctm $M$  of $\ZFC$ and a set $G$, one constructs a new

### 2018-12-07 LICS/zfc-axioms.tex
cambio esa horrible font tt y agrego secciones
Modelos de ZFC

--- 2018-12-04 LICS/generic-extensions.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
-\section{Generic extensions}
+\section{Axioms of set theory}
 
-We will swiftly review some definitions in order to reach the concept
-of \emph{generic extension}.
+\medskip
+\fbox{\em \dots\ discussion of axioms \dots }
+\medskip
 
+An equivalent formulation of the Foundation Axiom states that the
+universe of sets can be decomposed in a transfinite hierarchy of
+sets. 
+\begin{theorem}
+  Let $V_{\al}\defi\union\{\P(V_\be) : \be<\al\}$ for each ordinal
+  $\al$. Then each $V_\al$ is a set and 
+  $\forall x. \exists\al .\ \Ord(\al) \y x\in V_\al$.  
+\end{theorem}
+
+\medskip
+\fbox{\em \dots\ discussion of axioms \dots }
+\medskip
+
+Models of the theory $\ZFC$ consist of a pair $\lb M,E\rb$ where $M$
+is a set and $E$ is a binary relation on $M$. Forcing is a technique
+to extend very special kind of models, where $M$ is a countable
+transitive set (i.e., every element of $M$ is a subset of $M$) and
+$E$ is the membership relation $\in$ restricted to $M$. In this case
+we simply refer to $M$ as a \emph{countable transitive model} or
+\emph{ctm}.
+
+The existence of a ctm of $\ZFC$ can be proved from the existence of a
+model  $\lb N,E\rb$ such that the relation $E$ is well founded. The
+L\"owenheim-Skolem 
+Theorem ensures that there is an countable elementary submodel 
+$\lb N',E\restr N'\rb\preccurlyeq  \lb N,E\rb$ which must also be
+well founded; then the 
+Mostowski collapsing function \cite[Def.~I.9.31]{kunen2011set} sends $\lb
+N',E\restr N'\rb$ 
+isomorphically to an $\lb M,\in\rb$ where $M$ is transitive.
+
+By G\"odel's Second Incompleteness Theorem we cannot  prove that
+there exists a model of $\ZFC$  (assuming that our base theory is not
+stronger than $\ZFC$ to start with), but for consistency proofs,
+usually a model of only a finite subset of $\ZFC$ suffices. Hence the
+following meta-theoretic result by Montague applies:
+%
+\begin{theorem}[Reflection Principle]\label{th:reflection-principle}
+  For every finite $\Phi\sbq\ZFC$, $\ZFC$ proves: ``There exists
+    unboundedly many $\al$ such that $V_\al\models \Phi$.''
+\end{theorem}
+%
+Since the sets $V_\al$ are well founded, we can repeat the above
+argument to obtain a ctm of $\Phi$.
+%
 %%% Local Variables: 

### 2018-12-07 LICS/renaming.tex
Explicación conceptual de renaming

@@
 \section{Renaming}
+\newcommand{\renaming}[2]{#1[#2]}
+\newcommand{\inFm}[2]{\mathsf{Member}(#1,#2)}
+\newcommand{\eqFm}[2]{\mathsf{Equal}(#1,#2)}
+\newcommand{\nandFm}[2]{\mathsf{Nand}(#1,#2)}
+\newcommand{\forallFm}[1]{\mathsf{Forall(#1)}}
 
 In the course of our work we need to reason about renaming of formulas
-and its effect on their satisfiability. Since internalized formulas
-are implemented using de Bruijn indices, thus the arity of a formula
-$\phi$ gives the least set containing all the free variables in
-$\phi$. Following \citet{fiore-abssyn}, one can understand the arity
-of a formula as the context of the free variables and renamings are,
-consequently, mappings changing the context.
+and its effect on their satisfiability. Internalized formulas are
+implemented using de Bruijn indices for variables and the arity of a
+formula $\phi$ gives the least natural number containing all the free
+variables in $\phi$. Following \citet{fiore-abssyn}, one can
+understand the arity of a formula as the context of the free
+variables; notice that the arity of $\forallFm{\phi}$ is the
+predecessor of the arity of $\phi$. Renamings are, consequently,
+mappings between finite sets; since we can think of $\mathsf{succ}(n)$
+as the co-product $1+n$, then given a renaming $f \colon n \to m$, the
+unique morphism $\mathsf{id}_1+f \colon 1+n \to 1+m$ is used to
+rename free variables in a quantified formula:
+
+\begin{definition}[Renaming]
+  Let $\phi$ be a formula of arity $n$ and let $f \colon n \to m$, the
+  renaming of $\phi$ by $f$, denoted $\renaming{\phi}{f}$, is defined
+  by recursion on $\phi$:
+  \begin{gather*}
+    \renaming{\inFm{i}{j}}{f} = \inFm{f\,i}{f\,j}\\
+    \renaming{\eqFm{i}{j}}{f} = \eqFm{f\,i}{f\,j}\\
+    \renaming{\nandFm{\phi}{\psi}}{f} = \nandFm{\renaming{\phi}{f}}{\renaming{\psi}{f}}\\
+    \renaming{\forallFm{\phi}}{f} = \forallFm{\renaming{\phi}{\mathsf{id}+f}}
+  \end{gather*}
+\end{definition}
+
+Environments can be thought, then, as mappings from a finite set to
+the model $M$; the action of renaming on environments is simply given
+by composition.
+
+\begin{lemma}
+  Let $\phi$ be a formula of arity $n$, $f \colon n \to m$ be a
+  renaming, and let $\rho \colon n \to M$ and $\rho' \colon m \to M$
+  be environments. If for all $i \in n$, $\rho\,i = \rho'(f\,i)$, then
+  $M,\rho\models \phi$ is equivalent to
+  $M,\rho' \models \renaming{\phi}{f}$.
+\end{lemma}
 

### 2018-12-07 LICS/renaming.tex
Más texto en renaming.

@@
 \section{Renaming}
-\newcommand{\renaming}[2]{#1[#2]}
-\newcommand{\inFm}[2]{\mathsf{Member}(#1,#2)}
-\newcommand{\eqFm}[2]{\mathsf{Equal}(#1,#2)}
-\newcommand{\nandFm}[2]{\mathsf{Nand}(#1,#2)}
-\newcommand{\forallFm}[1]{\mathsf{Forall(#1)}}
+\newcommand{\renaming}[2]{(#1)[#2]}
+\newcommand{\inFm}[2]{#1 \in #2}
+\newcommand{\eqFm}[2]{#1 = #2}
+\newcommand{\negFm}[1]{\neg #1}
+\newcommand{\andFm}[2]{#1 \wedge #2}
+\newcommand{\forallFm}[1]{\forall #1}
+
+\newcommand{\inIFm}[2]{\mathsf{Member}(#1,#2)}
+\newcommand{\eqIFm}[2]{\mathsf{Equal}(#1,#2)}
+\newcommand{\nandIFm}[2]{\mathsf{Nand}(#1,#2)}
+\newcommand{\forallIFm}[1]{\mathsf{Forall(#1)}}
 
@@
 as the co-product $1+n$, then given a renaming $f \colon n \to m$, the
-unique morphism $\mathsf{id}_1+f \colon 1+n \to 1+m$ is used to
-rename free variables in a quantified formula:
+unique morphism $\mathsf{id}_1+f \colon 1+n \to 1+m$ is used to rename
+free variables in a quantified formula.
 
@@
     \renaming{\eqFm{i}{j}}{f} = \eqFm{f\,i}{f\,j}\\
-    \renaming{\nandFm{\phi}{\psi}}{f} = \nandFm{\renaming{\phi}{f}}{\renaming{\psi}{f}}\\
+    \renaming{\negFm{\phi}}{f} = \negFm{\renaming{\phi}{f}}\\
+    \renaming{\andFm{\phi}{\psi}}{f} = \andFm{\renaming{\phi}{f}}{\renaming{\psi}{f}}\\
     \renaming{\forallFm{\phi}}{f} = \forallFm{\renaming{\phi}{\mathsf{id}+f}}
@@
 the model $M$; the action of renaming on environments is simply given
-by composition.
-
+by composition. An easy proof connects satisfaction with renamings.
 \begin{lemma}
+  \label{lem:renaming}
   Let $\phi$ be a formula of arity $n$, $f \colon n \to m$ be a
   renaming, and let $\rho \colon n \to M$ and $\rho' \colon m \to M$
-  be environments. If for all $i \in n$, $\rho\,i = \rho'(f\,i)$, then
-  $M,\rho\models \phi$ is equivalent to
-  $M,\rho' \models \renaming{\phi}{f}$.
+  be environments. If for all $i \in n$,
+  $\rho\,i = (\rho' \circ f)\,i)$, then $M,\rho\models \phi$ is
+  equivalent to $M,\rho' \models \renaming{\phi}{f}$.
 \end{lemma}
+
+An important resource in Isabelle/ZF is the facility for defining
+inductive sets \cite{paulson2000fixedpoint,paulson1995set} together
+with a principle for defining functions by structural recursion.
+Internalized formulas are a prime example of this, so we define
+a function \isa{ren} that associates to each formula an internalized
+function that can be later applied to suitable arguments.
+\begin{isabelle}
+  \ \isacommand{consts}\isamarkupfalse%
+\ \ \ ren\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isacharequal}{\isachargreater}i{\isachardoublequoteclose}\isanewline
+\ \isacommand{primrec}\isamarkupfalse%
+\isanewline
+\ {\isachardoublequoteopen}ren{\isacharparenleft}Member{\isacharparenleft}x{\isacharcomma}y{\isacharparenright}{\isacharparenright}\ {\isacharequal}\isanewline
+\ \ {\isacharparenleft}{\isasymlambda}n{\isasymin}nat{\isachardot}{\isasymlambda}f\ {\isasymin}\ bij{\isacharparenleft}n{\isacharcomma}n{\isacharparenright}{\isachardot}\ Member\ {\isacharparenleft}f{\isacharbackquote}x{\isacharcomma}\ f{\isacharbackquote}y{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ {\isachardoublequoteopen}ren{\isacharparenleft}Equal{\isacharparenleft}x{\isacharcomma}y{\isacharparenright}{\isacharparenright}\ {\isacharequal}\isanewline
+\ \ {\isacharparenleft}{\isasymlambda}n{\isasymin}nat{\isachardot}\ {\isasymlambda}f\ {\isasymin}\ bij{\isacharparenleft}n{\isacharcomma}n{\isacharparenright}{\isachardot}\ Equal\ {\isacharparenleft}f{\isacharbackquote}x{\isacharcomma}\ f{\isacharbackquote}y{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \isanewline
+\ {\isachardoublequoteopen}ren{\isacharparenleft}Nand{\isacharparenleft}p{\isacharcomma}q{\isacharparenright}{\isacharparenright}\ {\isacharequal}\isanewline
+\ \ {\isacharparenleft}{\isasymlambda}n{\isasymin}nat{\isachardot}\ {\isasymlambda}f\ {\isasymin}\ bij{\isacharparenleft}n{\isacharcomma}n{\isacharparenright}{\isachardot}\isanewline
+\ \ \ \ \ \ \ \ Nand\ {\isacharparenleft}ren{\isacharparenleft}p{\isacharparenright}{\isacharbackquote}n{\isacharbackquote}f{\isacharcomma}\ ren{\isacharparenleft}q{\isacharparenright}{\isacharbackquote}n{\isacharbackquote}f{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \isanewline
+\ {\isachardoublequoteopen}ren{\isacharparenleft}Forall{\isacharparenleft}p{\isacharparenright}{\isacharparenright}\ {\isacharequal}\isanewline
+\ \ {\isacharparenleft}{\isasymlambda}n{\isasymin}nat{\isachardot}\ {\isasymlambda}f\ {\isasymin}\ bij{\isacharparenleft}n{\isacharcomma}n{\isacharparenright}{\isachardot}\isanewline
+\ \ \ \ \ \ \ \ Forall\ {\isacharparenleft}ren{\isacharparenleft}p{\isacharparenright}\ {\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isacharbackquote}\ sum{\isacharunderscore}id{\isacharparenleft}n{\isacharcomma}f{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
+
+Notice that our implementation of renaming is more restricted as we
+work with bijections; \isa{sum{\isacharunderscore}id} corresponds to
+the co-product morfism $\mathsf{id}_{1}+f \colon 1 + n \to 1 +
+n$. This definition of renaming exposes some inconveniences of working
+in the untyped realm of set theory, for example to use \isa{ren} we
+will need to prove that the renaming is a bijection. Another
+inconvenience is that environments are not finite mappings but lists
+(yet another inductively defined set); given an environment and a
+bijection, the following definition returns the composition
+$\rho \circ f$ as a list:
+\begin{isabelle}
+  \isacommand{definition}\isamarkupfalse%
+\ perm{\isacharunderscore}list\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharbrackleft}i{\isacharcomma}i{\isacharbrackright}\ {\isasymRightarrow}\ i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
+\ {\isachardoublequoteopen}perm{\isacharunderscore}list{\isacharparenleft}f{\isacharcomma}\isasymrho{\isacharparenright}\ {\isacharequal}{\isacharequal}tab{\isacharparenleft}nth{\isacharunderscore}i{\isacharparenleft}\isasymrho{\isacharparenright}\ O\ f{\isacharcomma}length{\isacharparenleft}\isasymrho{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
+In the previous definition \isa{tab(g,n)} creates a list tabulating the
+image of the function \isa{g} on the elements of \isa{n}. 
+% \begin{isabelle}
+% \isacommand{lemma}\isamarkupfalse%
+% \ perm{\isacharunderscore}list{\isacharunderscore}tc\ \ {\isacharcolon}\ {\isachardoublequoteopen}\ \isasymrho\ {\isasymin}\ list{\isacharparenleft}A{\isacharparenright}\ {\isasymLongrightarrow}\ \isanewline
+% \ \ \ \ f{\isasymin}\ bij{\isacharparenleft}length{\isacharparenleft}\isasymrho{\isacharparenright}{\isacharcomma}length{\isacharparenleft}\isasymrho{\isacharparenright}{\isacharparenright}\ {\isasymLongrightarrow}\ \isanewline
+% \ \ \ \ perm{\isacharunderscore}list{\isacharparenleft}f{\isacharcomma}\isasymrho{\isacharparenright}\ {\isasymin}\ list{\isacharparenleft}A{\isacharparenright}{\isachardoublequoteclose}
+% \end{isabelle}
+The result connecting renaming with satisfiability, as in
+\ref{lem:renaming}, is the following:
+\fbox{Miguel: esto hay que acomodarlo o no mostrarlo!}
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ renSat\ {\isacharcolon}\ \isanewline
+\ \ \isakeyword{fixes}\ {\isachardoublequoteopen}{\isasymphi}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{assumes}\ {\isachardoublequoteopen}{\isasymphi}\ {\isasymin}\ formula{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\ \ {\isachardoublequoteopen}{\isasymAnd}\ {\isasymrho}\ f\ {\isachardot}\ {\isasymlbrakk}\ {\isasymrho}\ {\isasymin}\ list{\isacharparenleft}M{\isacharparenright}\ {\isacharsemicolon}\ f\ {\isasymin}\ bij{\isacharparenleft}length{\isacharparenleft}{\isasymrho}{\isacharparenright}{\isacharcomma}length{\isacharparenleft}{\isasymrho}{\isacharparenright}{\isacharparenright}\ {\isacharsemicolon}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}\ {\isasymle}\ length{\isacharparenleft}{\isasymrho}{\isacharparenright}\ {\isasymrbrakk}\ {\isasymLongrightarrow}\isanewline
+\ \ \ \ \ \ \ \ \ sats{\isacharparenleft}M{\isacharcomma}{\isasymphi}{\isacharcomma}{\isasymrho}{\isacharparenright}\ {\isasymlongleftrightarrow}\ sats{\isacharparenleft}M{\isacharcomma}rename{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharbackquote}length{\isacharparenleft}{\isasymrho}{\isacharparenright}{\isacharbackquote}converse{\isacharparenleft}f{\isacharparenright}{\isacharcomma}{\isasymrho}{\isasymcirc}f{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
 

### 2018-12-11 LICS/renaming.tex

--- 2018-12-07 LICS/renaming.tex
@@
 the model $M$; the action of renaming on environments is simply given
-by composition.
-
+by composition. An easy proof connects satisfaction with renamings.
 \begin{lemma}
+  \label{lem:renaming}
   Let $\phi$ be a formula of arity $n$, $f \colon n \to m$ be a
   renaming, and let $\rho \colon n \to M$ and $\rho' \colon m \to M$
-  be environments. If for all $i \in n$, $\rho\,i = \rho'(f\,i)$, then
-  $M,\rho\models \phi$ is equivalent to
-  $M,\rho' \models \renaming{\phi}{f}$.
+  be environments. If for all $i \in n$,
+  $\rho\,i = (\rho' \circ f)\,i)$, then $M,\rho\models \phi$ is
+  equivalent to $M,\rho' \models \renaming{\phi}{f}$.
 \end{lemma}
+
+An important resource in Isabelle/ZF is the facility for defining
+inductive sets \cite{paulson2000fixedpoint,paulson1995set} together
+with a principle for defining functions by structural recursion.
+Internalized formulas are a prime example of this, so we define
+a function \isa{ren} that associates to each formula an internalized
+function that can be later applied to suitable arguments.
+\begin{isabelle}
+\isamarkuptrue%
+\isacommand{consts}\isamarkupfalse%
+\ \ \ ren\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isacharequal}{\isachargreater}i{\isachardoublequoteclose}\isanewline
+\isacommand{primrec}\isamarkupfalse%
+\isanewline
+\ \ {\isachardoublequoteopen}ren{\isacharparenleft}Member{\isacharparenleft}x{\isacharcomma}y{\isacharparenright}{\isacharparenright}\ {\isacharequal}\isanewline
+\ \ \ \ \ \ {\isacharparenleft}{\isasymlambda}\ n\ {\isasymin}\ nat\ {\isachardot}\ {\isasymlambda}\ m\ {\isasymin}\ nat{\isachardot}\ {\isasymlambda}f\ {\isasymin}\ n\ {\isasymrightarrow}\ m{\isachardot}\ Member\ {\isacharparenleft}f{\isacharbackquote}x{\isacharcomma}\ f{\isacharbackquote}y{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isanewline
+\ \ {\isachardoublequoteopen}ren{\isacharparenleft}Equal{\isacharparenleft}x{\isacharcomma}y{\isacharparenright}{\isacharparenright}\ {\isacharequal}\isanewline
+\ \ \ \ \ \ {\isacharparenleft}{\isasymlambda}\ n\ {\isasymin}\ nat\ {\isachardot}\ {\isasymlambda}\ m\ {\isasymin}\ nat{\isachardot}\ {\isasymlambda}f\ {\isasymin}\ n\ {\isasymrightarrow}\ m{\isachardot}\ Equal\ {\isacharparenleft}f{\isacharbackquote}x{\isacharcomma}\ f{\isacharbackquote}y{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isanewline
+\ \ {\isachardoublequoteopen}ren{\isacharparenleft}Nand{\isacharparenleft}p{\isacharcomma}q{\isacharparenright}{\isacharparenright}\ {\isacharequal}\isanewline
+\ \ \ \ \ \ {\isacharparenleft}{\isasymlambda}\ n\ {\isasymin}\ nat\ {\isachardot}\ {\isasymlambda}\ m\ {\isasymin}\ nat{\isachardot}\ {\isasymlambda}f\ {\isasymin}\ n\ {\isasymrightarrow}\ m{\isachardot}\ Nand\ {\isacharparenleft}ren{\isacharparenleft}p{\isacharparenright}{\isacharbackquote}n{\isacharbackquote}m{\isacharbackquote}f{\isacharcomma}\ ren{\isacharparenleft}q{\isacharparenright}{\isacharbackquote}n{\isacharbackquote}m{\isacharbackquote}f{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isanewline
+\ \ {\isachardoublequoteopen}ren{\isacharparenleft}Forall{\isacharparenleft}p{\isacharparenright}{\isacharparenright}\ {\isacharequal}\isanewline
+\ \ \ \ \ \ {\isacharparenleft}{\isasymlambda}\ n\ {\isasymin}\ nat\ {\isachardot}\ {\isasymlambda}\ m\ {\isasymin}\ nat{\isachardot}\ {\isasymlambda}f\ {\isasymin}\ n\ {\isasymrightarrow}\ m{\isachardot}\ Forall\ {\isacharparenleft}ren{\isacharparenleft}p{\isacharparenright}{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isacharbackquote}succ{\isacharparenleft}m{\isacharparenright}{\isacharbackquote}sum{\isacharunderscore}id{\isacharparenleft}n{\isacharcomma}f{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
+
+In the last equation, \isa{sum{\isacharunderscore}id} corresponds to
+the co-product morfism $\mathsf{id}_{1}+f \colon 1 + n \to 1 +
+n$. This definition of renaming exposes some inconveniences of working
+in the untyped realm of set theory, for example to use \isa{ren} we
+will need to prove that the renaming is a function. Since environments
+are not finite mappings but lists (yet another inductively defined
+set), the main hypothesis of Lem.~\ref{lem:renaming} is expressed with
+the \isa{nth} function for indexing elements of lists:
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ renSat\ {\isacharcolon}\ \isanewline
+\ \ \isakeyword{fixes}\ {\isasymphi}\isanewline
+\ \ \isakeyword{assumes}\ {\isachardoublequoteopen}{\isasymphi}\ {\isasymin}\ formula{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\ \ {\isachardoublequoteopen}{\isasymAnd}\ n\ m\ {\isasymrho}\ {\isasymrho}{\isacharprime}\ f\ {\isachardot}\ \isanewline
+\ \ {\isasymlbrakk}n{\isasymin}nat\ {\isacharsemicolon}\ m{\isasymin}nat\ {\isacharsemicolon}\ f\ {\isasymin}\ n{\isasymrightarrow}m\ {\isacharsemicolon}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}\ {\isasymle}\ n\ {\isacharsemicolon}\isanewline
+\ \ \ \ \ {\isasymrho}\ {\isasymin}\ list{\isacharparenleft}M{\isacharparenright}\ {\isacharsemicolon}\ {\isasymrho}{\isacharprime}\ {\isasymin}\ list{\isacharparenleft}M{\isacharparenright}\ {\isacharsemicolon}\ \isanewline
+\ \ \ {\isasymAnd}\ i\ {\isachardot}\ i{\isacharless}n\ {\isasymLongrightarrow}\ nth{\isacharparenleft}i{\isacharcomma}{\isasymrho}{\isacharparenright}\ {\isacharequal}\ nth{\isacharparenleft}f{\isacharbackquote}i{\isacharcomma}{\isasymrho}{\isacharprime}{\isacharparenright}\ {\isasymrbrakk}\ {\isasymLongrightarrow}\isanewline
+\ \ sats{\isacharparenleft}M{\isacharcomma}{\isasymphi}{\isacharcomma}{\isasymrho}{\isacharparenright}\ {\isasymlongleftrightarrow}\ sats{\isacharparenleft}M{\isacharcomma}ren{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharbackquote}n{\isacharbackquote}m{\isacharbackquote}f{\isacharcomma}{\isasymrho}{\isacharprime}{\isacharparenright}{\isachardoublequoteclose}\end{isabelle}
 

### 2018-12-11 src/fix_links.sh
cambio nombre de sesión

@@ -1,6 +1,6 @@
 #!/bin/bash
 old_dir=$PWD
-html_dir="output/html/Unsorted/separation/"
+html_dir="output/html/Unsorted/Separation-Generic/"
 
 cd $html_dir
 sed -i -- 's/a href="..\/..\/ZF\/ZF\/Main.html/a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF\/ZF.html/g' *.html

### 2018-12-12 LICS/isabelle.tex
edits variados
ediciones varias

@@
 \section{Isabelle}
+
+%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-
+\subsection{Logics}
+
 Isabelle provides a meta-language called \emph{Pure} that consists of
 a fragment of higher order logic. The meta-Boolean type is called
-\isatt{prop}. Meta connectives
+\isatt{prop}. Meta-connectives
 \isatt{\isasymLongrightarrow} and \isatt{\&\&\&} fulfill the role of
@@
 \]
+A further example is given by induction on the natural numbers
+\isatt{nat},
+\[
+\isasymlbrakk P(0);\ (\textstyle\isasymAnd
+x.\ P(x)\ \isasymLongrightarrow\ P(\isatt{succ}(x))) \isasymrbrakk
+\ \isasymLongrightarrow\ P(n), 
+\]
+where we are omitting the ``typing'' assumtions on $n$ and $x$.
 
@@
 
+%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-
+\subsection{Locales}
+
+Locales \cite{ballarin2010tutorial} provide a neat facility to
+encapsulate a context (fixed objects and assumptions on them) that is
+to be used in proving several theorems, as in usual mathematical
+practice.
+
+In this paper, locales have a further use. The \emph{Fundamental
+  Theorems of Forcing} we use talk about a specific map $\forceisa$
+from formulas to formulas. The definition of $\forceisa$ is involved
+and we will not dwell on this now; but applications of those theorems
+do not require to know how it is defined. Therefore, we black-box it
+and package everything in a locale that assumes that there is such a
+map that satisfies the Fundamental Theorems.
+
 %%% Local Variables: 

### 2018-12-12 LICS/zfc-axioms.tex
ediciones varias

@@
 
+The axioms of Zermelo and Fraenkel with Choice ($\ZFC$) are a
+countably infinite list of first-order sentences.
 \medskip
@@
 \medskip
+
+We chose a presentation of the $\ZFC$ axioms that would be most
+compatible with the development of relativization by Paulson:
+\begin{isabelle}
+\isacommand{locale}\isamarkupfalse%
+\ M{\isacharunderscore}ZF\ {\isacharequal}\ \isanewline
+\ \isakeyword{fixes}\ M\ \isanewline
+\ \isakeyword{assumes}\ \isanewline
+\ \ \ \ \ \ upair{\isacharunderscore}ax{\isacharcolon}\ \ \ \ {\isachardoublequoteopen}upair{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{and}\ Union{\isacharunderscore}ax{\isacharcolon}\ \ \ \ {\isachardoublequoteopen}Union{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{and}\ power{\isacharunderscore}ax{\isacharcolon}\ \ \ \ {\isachardoublequoteopen}power{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{and}\ extensionality{\isacharcolon}\ \ \ {\isachardoublequoteopen}extensionality{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{and}\ foundation{\isacharunderscore}ax{\isacharcolon}\ \ \ \ {\isachardoublequoteopen}foundation{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{and}\ infinity{\isacharunderscore}ax{\isacharcolon}\ {\isachardoublequoteopen}infinity{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{and}\ separation{\isacharunderscore}ax{\isacharcolon}\isanewline
+\ \ {\isachardoublequoteopen}{\isasymlbrakk}\ {\isasymphi}\ {\isasymin}\ formula\ {\isacharsemicolon}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharequal}{\isadigit{1}}\ {\isasymor}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharequal}{\isadigit{2}}\ {\isasymrbrakk}\isanewline\ \ \ \ {\isasymLongrightarrow}\isanewline
+\ \ {\isacharparenleft}{\isasymforall}a{\isasymin}M{\isachardot}\ separation{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharcomma}{\isasymlambda}x{\isachardot}\ sats{\isacharparenleft}M{\isacharcomma}{\isasymphi}{\isacharcomma}{\isacharbrackleft}x{\isacharcomma}a{\isacharbrackright}{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{and}\ replacement{\isacharunderscore}ax{\isacharcolon}\isanewline 
+\ \ {\isachardoublequoteopen}{\isasymlbrakk}\ {\isasymphi}\ {\isasymin}\ formula{\isacharsemicolon}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharequal}{\isadigit{2}}\ {\isasymor}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharequal}{\isadigit{3}}\ {\isasymrbrakk}\isanewline
+\ \ \ \ {\isasymLongrightarrow}\isanewline
+\ \ {\isacharparenleft}{\isasymforall}a{\isasymin}M{\isachardot}\ strong{\isacharunderscore}replacement{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharcomma}\isanewline
+\ \ \ \ {\isasymlambda}x\ y{\isachardot}\ sats{\isacharparenleft}M{\isacharcomma}{\isasymphi}{\isacharcomma}{\isacharbrackleft}x{\isacharcomma}y{\isacharcomma}a{\isacharbrackright}{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
 
@@
 N',E\restr N'\rb$ 
-isomorphically to an $\lb M,\in\rb$ where $M$ is transitive.
+isomorphically to some $\lb M,\in\rb$ with  $M$ transitive.
 

### 2018-12-12 LICS/zfc-axioms.tex
ediciones varias al paper

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
-\section{Axioms of set theory}
+\section{Models and axioms of set theory}
 
-The axioms of Zermelo and Fraenkel with Choice ($\ZFC$) are a
-countably infinite list of first-order sentences.
-\medskip
-\fbox{\em \dots\ discussion of axioms \dots }
-\medskip
+The axioms of Zermelo and Fraenkel ($\ZF$) form a
+countably infinite list of first-order sentences in a language
+consisting of an only binary relation symbol $\in$. These include the
+axioms of Pairing, Union, Powerset, Foundation, Infinity, and two
+axiom-schemes collectively referred to as Axiom of Separation:
+\emph{For every $A$, $a_1,\dots,a_n$,  and  a formula
+  $\psi(x_0,x_1,\dots,x_n)$, there exists $\{x\in  A:
+  \psi(a,a_1,\dots,a_n)\}$}, 
+and Axiom of Replacement: \emph{For every $A$, $a_1,\dots,a_n$,  and
+  a formula   $\psi(x,z,x_1\dots,x_n)$, if 
+  $\forall x.\exists!z.\psi(x,z,x_1,\dots,x_n)$,  there exists 
+  $\{b : \exists a\in A. \psi(a,b,a_1,\dots,a_n)\}$}.
+An excellent introduction to the axioms and the motivation behind them
+can be found in Shoenfield \cite{MR3727410}.
 
-We chose a presentation of the $\ZFC$ axioms that would be most
-compatible with the development of relativization by Paulson:
-\begin{isabelle}
-\isacommand{locale}\isamarkupfalse%
-\ M{\isacharunderscore}ZF\ {\isacharequal}\ \isanewline
-\ \isakeyword{fixes}\ M\ \isanewline
-\ \isakeyword{assumes}\ \isanewline
-\ \ \ \ \ \ upair{\isacharunderscore}ax{\isacharcolon}\ \ \ \ {\isachardoublequoteopen}upair{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
-\ \ \isakeyword{and}\ Union{\isacharunderscore}ax{\isacharcolon}\ \ \ \ {\isachardoublequoteopen}Union{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
-\ \ \isakeyword{and}\ power{\isacharunderscore}ax{\isacharcolon}\ \ \ \ {\isachardoublequoteopen}power{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
-\ \ \isakeyword{and}\ extensionality{\isacharcolon}\ \ \ {\isachardoublequoteopen}extensionality{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
-\ \ \isakeyword{and}\ foundation{\isacharunderscore}ax{\isacharcolon}\ \ \ \ {\isachardoublequoteopen}foundation{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
-\ \ \isakeyword{and}\ infinity{\isacharunderscore}ax{\isacharcolon}\ {\isachardoublequoteopen}infinity{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
-\ \ \isakeyword{and}\ separation{\isacharunderscore}ax{\isacharcolon}\isanewline
-\ \ {\isachardoublequoteopen}{\isasymlbrakk}\ {\isasymphi}\ {\isasymin}\ formula\ {\isacharsemicolon}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharequal}{\isadigit{1}}\ {\isasymor}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharequal}{\isadigit{2}}\ {\isasymrbrakk}\isanewline\ \ \ \ {\isasymLongrightarrow}\isanewline
-\ \ {\isacharparenleft}{\isasymforall}a{\isasymin}M{\isachardot}\ separation{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharcomma}{\isasymlambda}x{\isachardot}\ sats{\isacharparenleft}M{\isacharcomma}{\isasymphi}{\isacharcomma}{\isacharbrackleft}x{\isacharcomma}a{\isacharbrackright}{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
-\ \ \isakeyword{and}\ replacement{\isacharunderscore}ax{\isacharcolon}\isanewline 
-\ \ {\isachardoublequoteopen}{\isasymlbrakk}\ {\isasymphi}\ {\isasymin}\ formula{\isacharsemicolon}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharequal}{\isadigit{2}}\ {\isasymor}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharequal}{\isadigit{3}}\ {\isasymrbrakk}\isanewline
-\ \ \ \ {\isasymLongrightarrow}\isanewline
-\ \ {\isacharparenleft}{\isasymforall}a{\isasymin}M{\isachardot}\ strong{\isacharunderscore}replacement{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharcomma}\isanewline
-\ \ \ \ {\isasymlambda}x\ y{\isachardot}\ sats{\isacharparenleft}M{\isacharcomma}{\isasymphi}{\isacharcomma}{\isacharbrackleft}x{\isacharcomma}y{\isacharcomma}a{\isacharbrackright}{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
-\end{isabelle}
-
-An equivalent formulation of the Foundation Axiom states that the
-universe of sets can be decomposed in a transfinite hierarchy of
-sets. 
-\begin{theorem}
-  Let $V_{\al}\defi\union\{\P(V_\be) : \be<\al\}$ for each ordinal
-  $\al$. Then each $V_\al$ is a set and 
-  $\forall x. \exists\al .\ \Ord(\al) \y x\in V_\al$.  
-\end{theorem}
-
-\medskip
-\fbox{\em \dots\ discussion of axioms \dots }
-\medskip
-
-Models of the theory $\ZFC$ consist of a pair $\lb M,E\rb$ where $M$
+Models of the theory $\ZF$ consist of a pair $\lb M,E\rb$ where $M$
 is a set and $E$ is a binary relation on $M$. Forcing is a technique
@@
 
-The existence of a ctm of $\ZFC$ can be proved from the existence of a
+The existence of a ctm of $\ZF$ can be proved from the existence of a
 model  $\lb N,E\rb$ such that the relation $E$ is well founded. The
@@
 By G\"odel's Second Incompleteness Theorem we cannot  prove that
-there exists a model of $\ZFC$  (assuming that our base theory is not
-stronger than $\ZFC$ to start with), but for consistency proofs,
-usually a model of only a finite subset of $\ZFC$ suffices. Hence the
-following meta-theoretic result by Montague applies:
+there exists a model of $\ZF$  (assuming that our base theory is not
+stronger than $\ZF$ to start with), but for consistency proofs,
+usually a model of only a finite subset of $\ZF$ suffices. Recall that
+an equivalent formulation of the Foundation Axiom states that the 
+universe of sets can be decomposed in a transfinite hierarchy of
+sets. 
+\begin{theorem}
+  Let $V_{\al}\defi\union\{\P(V_\be) : \be<\al\}$ for each ordinal
+  $\al$. Then each $V_\al$ is a set and 
+  $\forall x. \exists\al .\ \Ord(\al) \y x\in V_\al$.  
+\end{theorem}
+%
+Then the following meta-theoretic result by Montague applies:
 %
 \begin{theorem}[Reflection Principle]\label{th:reflection-principle}
-  For every finite $\Phi\sbq\ZFC$, $\ZFC$ proves: ``There exists
+  For every finite $\Phi\sbq\ZF$, $\ZF$ proves: ``There exists
     unboundedly many $\al$ such that $V_\al\models \Phi$.''
@@
 argument to obtain a ctm of $\Phi$.
+
+In this stage of our implementation, we chose a presentation of the
+$\ZF$ axioms that would be most
+compatible with the development of relativization by Paulson. For
+instance, the predicate
+\isatt{upair{\isacharunderscore}ax{\isacharcolon}{\isacharcolon}{\isachardoublequoteopen}{\isacharparenleft}i{\isacharequal}{\isachargreater}o{\isacharparenright}{\isacharequal}{\isachargreater}o{\isachardoublequoteclose}}
+takes a ``class'' (unary predicate) $C$ as an argument and states that
+$C$ satisfies the Pairing Axiom.
+\begin{isabelle}
+upair{\isacharunderscore}ax{\isacharparenleft}C{\isacharparenright}{\isacharequal}{\isacharequal}{\isasymforall}x{\isacharbrackleft}C{\isacharbrackright}{\isachardot}{\isasymforall}y{\isacharbrackleft}C{\isacharbrackright}{\isachardot}{\isasymexists}z{\isacharbrackleft}C{\isacharbrackright}{\isachardot}\ upair{\isacharparenleft}C{\isacharcomma}x{\isacharcomma}y{\isacharcomma}z{\isacharparenright}
+\end{isabelle}
+Here, $\forall x[C]. \phi$ stands for
+$\forall x. C(x) \longrightarrow \phi$, \emph{relative}
+quantification. All of the development of relativization by Paulson is
+written for class arguments, so we set up a locale fixing a set $M$
+and using the class $\#\#M\defi \lambda x. \ x\in M$ as the argument. 
+\begin{isabelle}
+\isacommand{locale}\isamarkupfalse%
+\ M{\isacharunderscore}ZF\ {\isacharequal}\ \isanewline
+\ \isakeyword{fixes}\ M\ \isanewline
+\ \isakeyword{assumes}\ \isanewline
+\ \ \ \ \ \ upair{\isacharunderscore}ax{\isacharcolon}\ \ \ \ {\isachardoublequoteopen}upair{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
+%% \ \ \isakeyword{and}\ Union{\isacharunderscore}ax{\isacharcolon}\ \ \ \ {\isachardoublequoteopen}Union{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
+%% \ \ \isakeyword{and}\ power{\isacharunderscore}ax{\isacharcolon}\ \ \ \ {\isachardoublequoteopen}power{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
+%% \ \ \isakeyword{and}\ extensionality{\isacharcolon}\ \ \ {\isachardoublequoteopen}extensionality{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
+%% \ \ \isakeyword{and}\ foundation{\isacharunderscore}ax{\isacharcolon}\ \ \ \ {\isachardoublequoteopen}foundation{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
+%% \ \ \isakeyword{and}\ infinity{\isacharunderscore}ax{\isacharcolon}\ {\isachardoublequoteopen}infinity{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{and}\ \dots \isanewline
+\ \ \isakeyword{and}\ separation{\isacharunderscore}ax{\isacharcolon}\isanewline
+\ \ {\isachardoublequoteopen}{\isasymlbrakk}\ {\isasymphi}\ {\isasymin}\ formula\ {\isacharsemicolon}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharequal}{\isadigit{1}}\ {\isasymor}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharequal}{\isadigit{2}}\ {\isasymrbrakk}\isanewline\ \ \ \ {\isasymLongrightarrow}\isanewline
+\ \ {\isacharparenleft}{\isasymforall}a{\isasymin}M{\isachardot}\ separation{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharcomma}{\isasymlambda}x{\isachardot}\ sats{\isacharparenleft}M{\isacharcomma}{\isasymphi}{\isacharcomma}{\isacharbrackleft}x{\isacharcomma}a{\isacharbrackright}{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{and}\ replacement{\isacharunderscore}ax{\isacharcolon}\isanewline 
+\ \ {\isachardoublequoteopen}{\isasymlbrakk}\ {\isasymphi}\ {\isasymin}\ formula{\isacharsemicolon}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharequal}{\isadigit{2}}\ {\isasymor}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharequal}{\isadigit{3}}\ {\isasymrbrakk}\isanewline
+\ \ \ \ {\isasymLongrightarrow}\isanewline
+\ \ {\isacharparenleft}{\isasymforall}a{\isasymin}M{\isachardot}\ strong{\isacharunderscore}replacement{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharcomma}\isanewline
+\ \ \ \ {\isasymlambda}x\ y{\isachardot}\ sats{\isacharparenleft}M{\isacharcomma}{\isasymphi}{\isacharcomma}{\isacharbrackleft}x{\isacharcomma}y{\isacharcomma}a{\isacharbrackright}{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
+The rest of the axioms are also
+included. We single out Separation and Replacement: these are written
+for formulas with at most one extra parameter (meaning $n\leq 1$ in the
+above $\phi$). Thanks to Pairing, these
+versions are equivalent to the usual formulations. We are only able to
+prove that the extension satisfies Separation for any particular
+number of parameters, but not in general. This is a consequence that
+induction on terms of typo \tyo{} is not available.
+
 %

### 2018-12-12 src/fix_links.sh
Nombre de la sesión

@@ -1,6 +1,6 @@
 #!/bin/bash
 old_dir=$PWD
-html_dir="output/html/Unsorted/Separation-Generic/"
+html_dir="output/html/Unsorted/Separation/"
 
 cd $html_dir
 sed -i -- 's/a href="..\/..\/ZF\/ZF\/Main.html/a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF\/ZF.html/g' *.html

### 2018-12-13 LICS/generic-extensions.tex

@@
 
-The Separation Axiom scheme is the first that requires the notion of
+The Separation Axiom  is the first that requires the notion of
 genericity and the use of the forcing machinery, which we review in

### 2018-12-13 LICS/isabelle.tex
ediciones varias al paper

@@
 \]
-A further example is given by induction on the natural numbers
+One further example is given by induction on the natural numbers
 \isatt{nat},
@@
 do not require to know how it is defined. Therefore, we black-box it
-and package everything in a locale that assumes that there is such a
+and pack everything in a locale that assumes that there is such a
 map that satisfies the Fundamental Theorems.

### 2018-12-13 LICS/zfc-axioms.tex

@@
 \emph{For every $A$, $a_1,\dots,a_n$,  and  a formula
-  $\psi(x_0,x_1,\dots,x_n)$, there exists $\{x\in  A:
+  $\psi(x_0,x_1,\dots,x_n)$, there exists $\{a\in  A:
   \psi(a,a_1,\dots,a_n)\}$}, 
@@
 
-Models of the theory $\ZF$ consist of a pair $\lb M,E\rb$ where $M$
+A model of the theory $\ZF$ consists of a pair $\lb M,E\rb$ where $M$
 is a set and $E$ is a binary relation on $M$. Forcing is a technique
@@
 quantification. All of the development of relativization by Paulson is
-written for class arguments, so we set up a locale fixing a set $M$
+written for a class model, so we set up a locale fixing a set $M$
 and using the class $\#\#M\defi \lambda x. \ x\in M$ as the argument. 
@@
 for formulas with at most one extra parameter (meaning $n\leq 1$ in the
-above $\phi$). Thanks to Pairing, these
+above $\psi$). Thanks to Pairing, these
 versions are equivalent to the usual formulations. We are only able to
@@
 number of parameters, but not in general. This is a consequence that
-induction on terms of typo \tyo{} is not available.
+induction on terms of type \tyo{} is not available.
 

### 2018-12-13 LICS/easy-axioms.tex
Más ToDos en el texto

--- 2018-12-07 LICS/renaming.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
-\section{Renaming}
-\newcommand{\renaming}[2]{#1[#2]}
-\newcommand{\inFm}[2]{\mathsf{Member}(#1,#2)}
-\newcommand{\eqFm}[2]{\mathsf{Equal}(#1,#2)}
-\newcommand{\nandFm}[2]{\mathsf{Nand}(#1,#2)}
-\newcommand{\forallFm}[1]{\mathsf{Forall(#1)}}
+\section{Foundation, Union, Infinity}
+\label{sec:easy-axioms}
 
-In the course of our work we need to reason about renaming of formulas
-and its effect on their satisfiability. Internalized formulas are
-implemented using de Bruijn indices for variables and the arity of a
-formula $\phi$ gives the least natural number containing all the free
-variables in $\phi$. Following \citet{fiore-abssyn}, one can
-understand the arity of a formula as the context of the free
-variables; notice that the arity of $\forallFm{\phi}$ is the
-predecessor of the arity of $\phi$. Renamings are, consequently,
-mappings between finite sets; since we can think of $\mathsf{succ}(n)$
-as the co-product $1+n$, then given a renaming $f \colon n \to m$, the
-unique morphism $\mathsf{id}_1+f \colon 1+n \to 1+m$ is used to
-rename free variables in a quantified formula:
+\fbox{Hablar de \texttt{M\_trivial}}
 
-\begin{definition}[Renaming]
-  Let $\phi$ be a formula of arity $n$ and let $f \colon n \to m$, the
-  renaming of $\phi$ by $f$, denoted $\renaming{\phi}{f}$, is defined
-  by recursion on $\phi$:
-  \begin{gather*}
-    \renaming{\inFm{i}{j}}{f} = \inFm{f\,i}{f\,j}\\
-    \renaming{\eqFm{i}{j}}{f} = \eqFm{f\,i}{f\,j}\\
-    \renaming{\nandFm{\phi}{\psi}}{f} = \nandFm{\renaming{\phi}{f}}{\renaming{\psi}{f}}\\
-    \renaming{\forallFm{\phi}}{f} = \forallFm{\renaming{\phi}{\mathsf{id}+f}}
-  \end{gather*}
-\end{definition}
-
-Environments can be thought, then, as mappings from a finite set to
-the model $M$; the action of renaming on environments is simply given
-by composition.
-
-\begin{lemma}
-  Let $\phi$ be a formula of arity $n$, $f \colon n \to m$ be a
-  renaming, and let $\rho \colon n \to M$ and $\rho' \colon m \to M$
-  be environments. If for all $i \in n$, $\rho\,i = \rho'(f\,i)$, then
-  $M,\rho\models \phi$ is equivalent to
-  $M,\rho' \models \renaming{\phi}{f}$.
-\end{lemma}
+\fbox{Falta check(x) para Infinity}
 

### 2018-12-14 LICS/recursion.tex
Agregando cosas de recursión

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
\subsection*{Value of names}

The map $\val$ used in the definition of the generic extension is
characterized by the recursive equation
\begin{equation}
  \label{eq:val}
  \val(G,\tau) = \{val(G,\sigma) :\exists p \in\PP .%
  \lb\sigma,p\rb \in \tau \wedge p \in G \}
\end{equation}

As is well-known, the principle of transfinite recursion on
well-founded relations~\cite[p. 48]{kunen2011set}) allows us to define
a recursive function $F \colon A\to A$ by choosing a well-founded
relation $R \subseteq A\times A$ and a functional
$H\colon A\times (A \to A) \to A$ satisfying
$F(a)=H(a,F\!\upharpoonright\!(R^{-1}(a)))$. \citet{paulson1995set}
made this principle available in Isabelle/ZF via the the operator
\isa{wfrec}. The formalization of the functional $\mathit{Hv}$ is
straightforward:
%
\begin{isabelle}
\isacommand{definition}\isamarkupfalse%
\isanewline
\ \ Hv\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}i{\isasymRightarrow}i{\isasymRightarrow}i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
\ \ {\isachardoublequoteopen}Hv{\isacharparenleft}G{\isacharcomma}y{\isacharcomma}f{\isacharparenright}\ {\isacharequal}{\isacharequal}\ {\isacharbraceleft}\ f{\isacharbackquote}x\ {\isachardot}{\isachardot}\ x{\isasymin}\ domain{\isacharparenleft}y{\isacharparenright}{\isacharcomma}\ {\isasymexists}p{\isasymin}P{\isachardot}\ {\isacharless}x{\isacharcomma}p{\isachargreater}\ {\isasymin}\ y\ {\isasymand}\ p\ {\isasymin}\ G\ {\isacharbraceright}{\isachardoublequoteclose}
\end{isabelle}
In the references \cite{kunen2011set,weaver2014forcing} $\val$ is
applied only to \emph{names}, that are certain elements of $M$
characterized by a recursively defined predicate. The well-founded
relation used to justify Equation~\eqref{eq:val} is
\[ x \mathrel{\mathit{ed}} y \iff \exists p . \lb x,p\rb\in y. \] In
order to use \isa{wfrec} the relation should be expressed as a set, so
we took the restriction of $\mathit{ed}$ to the whole universe
$M$; i.e. $\mathit{ed}\cap M\times M$.  It was inconvenient to work
with that definition because it imposed an obligation showing that
some set is in $M$. The remedy was to restrict $\mathit{ed}$ to the
transitive closure of the actual parameter:
\begin{isabelle}
\isacommand{definition}\isamarkupfalse%
\isanewline
\ val\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}i{\isasymRightarrow}i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
\ {\isachardoublequoteopen}val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}{\isacharequal}{\isacharequal}\ wfrec{\isacharparenleft}edrel{\isacharparenleft}eclose{\isacharparenleft}{\isacharbraceleft}{\isasymtau}{\isacharbraceright}{\isacharparenright}{\isacharparenright}{\isacharcomma}{\isasymtau}{\isacharcomma}Hv{\isacharparenleft}G{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
When computing the value of a recursive function on some argument $a$,
one can restrict the relation to some ambient set $A$ if $a$ and
all of its predecessesors are also in $A$.
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ wfrec{\isacharunderscore}restr\ {\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ {\isachardoublequoteopen}relation{\isacharparenleft}r{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}wf{\isacharparenleft}r{\isacharparenright}{\isachardoublequoteclose}\ \isanewline
\ \ \isakeyword{shows}\ \ {\isachardoublequoteopen}a{\isasymin}A\ {\isasymLongrightarrow}\ {\isacharparenleft}r{\isacharcircum}{\isacharplus}{\isacharparenright}{\isacharminus}{\isacharbackquote}{\isacharbackquote}{\isacharbraceleft}a{\isacharbraceright}\ {\isasymsubseteq}\ A\ {\isasymLongrightarrow}\ \isanewline
\ \ \ \ \ \ \ \ \ \ wfrec{\isacharparenleft}r{\isacharcomma}a{\isacharcomma}H{\isacharparenright}\ {\isacharequal}\ wfrec{\isacharparenleft}r{\isasyminter}A{\isasymtimes}A{\isacharcomma}a{\isacharcomma}H{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
Using this lemma we can prove Equation~\eqref{eq:val}:
\begin{isabelle}
  \isacommand{lemma}\isamarkupfalse%
  \ def{\isacharunderscore}val{\isacharcolon}\isanewline
  \ {\isachardoublequoteopen}val{\isacharparenleft}G{\isacharcomma}x{\isacharparenright}\ {\isacharequal}\ {\isacharbraceleft}val{\isacharparenleft}G{\isacharcomma}t{\isacharparenright}\ {\isachardot}{\isachardot}\ t{\isasymin}domain{\isacharparenleft}x{\isacharparenright}\ {\isacharcomma}\isanewline
\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ {\isasymexists}p{\isasymin}P{\isachardot}\ {\isacharless}t{\isacharcomma}p{\isachargreater}{\isasymin}x\ {\isasymand}\ p{\isasymin}G\ {\isacharbraceright}{\isachardoublequoteclose}
\end{isabelle}
\fbox{Pedro: ¿esto sirve cuando la relación
  es una clase pero set-like on $A$?}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "Separation_In_MG"
%%% ispell-local-dictionary: "american"
%%% End:

### 2018-12-14 LICS/isabelle.tex
Descripción corta de locales (copiado de Whatsapp)

@@
 
+\begin{description}
+\item[\texttt{forcing\_notion}] preorden con top
+\item[\texttt{countable\_generic}] lo anterior con una familia contable de densos.
+\item[\texttt{M\_ZF}] axiomas.
+\item[\texttt{forcing\_data}]: lo anterior contable transitivo y una notion.
+\item[\texttt{forcing\_thms}]: eso.
+\item[\texttt{G\_generic}]: G es genérico.
+\item[\texttt{M\_extra\_assms}]: check in M e instancia de reemplazo para G.
+\item[\texttt{G\_generic\_extra}]: los dos anteriores (no sé si sigue estando)
+\end{description}
+
 %%% Local Variables: 

### 2018-12-14 LICS/easy-axioms.tex
precisando comentarios
Descripción corta de locales (copiado de Whatsapp)

@@
 
-\fbox{Hablar de \texttt{M\_trivial}}
+It is straightforward to show that the generic extension $M[G]$
+satisfies foundation, union, and infinity. The easiest one is
+foundation which does not depend on the satisfaction by $M$.
 
-\fbox{Falta check(x) para Infinity}
+\fbox{Hablar de \texttt{M\_trivial} y del uso de interfaz para instanciar}
+
+\fbox{Falta \texttt{check\_in\_M} para completar Infinity}
 

### 2018-12-15 LICS/generic-extensions.tex
ediciones varias

@@
 Metatheoretically, it is straightforward to see that $M[G]$ is a
-transitive set that satisfies the axioms of Pairing and Union and
+transitive set that satisfies the some of axioms of $\ZF$ (see
+Section~\ref{sec:easy-axioms}) and 
 includes $M\cup\{G\}$. Nevertheless there are no a priori reasons for

### 2018-12-15 LICS/proof-of-separation.tex
edición general, y sumo el hacking
ediciones varias

@@
 \] 
-Let $Q(\th,p)$ be the last statement. We have just seen that
+Let $Q(\th,p)$ be the last displayed statement. We have just seen that
 (\ref{eq:1}) implies 
@@
 \end{lemma}
-\noindent And by monotonicity of $\val$ we obtain
-\begin{lemma}
+Also by monotonicity of $\val$ we obtain 
   $\val(G,n)\sbq \val(G,m)$.
-\end{lemma}
 To complete the proof, it is therefore enough to show that
@@
 Since $G$ is a filter, there is $p\in G$ such that $p\leq q, r$.
-By Strenghtening, we have
+By Strengthening, we have
 \[

### 2018-12-17 LICS/forcing.tex
ediciones varias
Más ToDos en el texto

@@
 recursion on 
-the structure of $\phi$. In order to state the properties of this map
+the structure of $\phi$.
+
+\fbox{\em Compatibilizar lo que sigue con Renaming}
+
+In order to state the properties of this map
 in sufficient generality to prove that  $M[G]$ satisfies $\ZFC$, we work with
@@
 
+%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-
 \subsection{The fundamental theorems}
@@
 characterization given by (\ref{eq:3}) indeed holds, and hence the
-forcing relation is \emph{definable}. In our case, definability merely
-stands for a typing condition:
-\begin{isabelle}
-{\isachardoublequoteopen}{\isasymphi}\ {\isasymin}\ formula\ {\isasymLongrightarrow}\ forces{\isacharparenleft}{\isasymphi}{\isacharparenright}\ {\isasymin}\ formula{\isachardoublequoteclose}
-\end{isabelle}
+forcing relation is \emph{definable}.
 
@@
 
+The locale \isatt{forcing\_thms} includes all these results as assumptions on the
+mapping $\forceisa$, plus a typing condition  and its
+effect on arities:
+\begin{isabelle}
+\isacommand{locale}\isamarkupfalse%
+\ forcing{\isacharunderscore}thms\ {\isacharequal}\ forcing{\isacharunderscore}data\ {\isacharplus}\isanewline
+\ \isakeyword{fixes}\ forces\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i\ {\isasymRightarrow}\ i{\isachardoublequoteclose}\isanewline
+\ \isakeyword{assumes}\
+definition{\isacharunderscore}of{\isacharunderscore}forces{\isacharcolon}\isanewline 
+\ \ {\isachardoublequoteopen}p{\isasymin}P\ {\isasymLongrightarrow}\ {\isasymphi}{\isasymin}formula\ {\isasymLongrightarrow}\ env{\isasymin}list{\isacharparenleft}M{\isacharparenright}\ {\isasymLongrightarrow}\isanewline
+\ \ \ \  sats{\isacharparenleft}M{\isacharcomma}forces{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharcomma}\ {\isacharbrackleft}P{\isacharcomma}leq{\isacharcomma}one{\isacharcomma}p{\isacharbrackright}\ {\isacharat}\ env{\isacharparenright}\ {\isasymlongleftrightarrow}\isanewline
+\ \ \ \ \ \
+{\isacharparenleft}{\isasymforall}G{\isachardot}{\isacharparenleft}M{\isacharunderscore}generic{\isacharparenleft}G{\isacharparenright}{\isasymand}\
+p{\isasymin}G{\isacharparenright}\ {\isasymlongrightarrow}\isanewline 
+\ \ \ \ \ \ sats{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}{\isasymphi}{\isacharcomma}map{\isacharparenleft}val{\isacharparenleft}G{\isacharparenright}{\isacharcomma}env{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{and}\ \ definability{\isacharbrackleft}TC{\isacharbrackright}{\isacharcolon}\ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula\ {\isasymLongrightarrow}\isanewline
+\ \ \ \ \ \ \ \ \  forces{\isacharparenleft}{\isasymphi}{\isacharparenright}\ {\isasymin}\ formula{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{and}\ \ \ arity{\isacharunderscore}forces{\isacharcolon}\ \ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula\ {\isasymLongrightarrow}\isanewline
+\ \ \ \ \ \ \
+arity{\isacharparenleft}forces{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharparenright}\
+{\isacharequal}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}\ {\isacharhash}{\isacharplus}\ {\isadigit{4}}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{and}\ \dots
+\end{isabelle}
+
+\fbox{poner acá lo de formal abstracts y el locale \texttt{forcing\_thms}}
 %%% Local Variables: 

### 2018-12-17 LICS/hacking.tex

@@
 The proof, for instance, that the constructible universe $L$ satisfies
-the new locale \isatt{M\_trivial\_no\_repl} hold with minor
-modifications.
+the modified locale \isatt{M\_trivial} holds with minor
+modifications. Nevertheless, in order to have a neater presentation,
+we have stripped off several sections concerning $L$ from the theories
+\isatt{L\_axioms} and \isatt{Internalize}, and we merged them to form
+the new file  \isatt{Internalizations}. Actually we believe that the
+\isatt{ZF-Constructible} session  would benefit from some changes in
+the way some results are organized accross theory files; for
+intance, by cataloging all or most of the internalized formulas in one
+file.
 

### 2018-12-17 LICS/recursion.tex

@@
 \end{isabelle}
+
+\medskip
+\noindent\fbox{Para que el resultado anterior tenga utilidad hay\ \dots\
+}
+\fbox{\dots\ que pedir que la relaci\'on \texttt{"i=>i=>o"} sea set-like}
+\medskip
+
 Using this lemma we can prove Equation~\eqref{eq:val}:
@@
 \end{isabelle}
-\fbox{Pedro: ¿esto sirve cuando la relación
-  es una clase pero set-like on $A$?}
+and the monotonicity of $\val$ follows automatically after a
+substitution.
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ val{\isacharunderscore}mono{\isacharcolon}\ {\isachardoublequoteopen}x{\isasymsubseteq}y\ {\isasymLongrightarrow}\ val{\isacharparenleft}G{\isacharcomma}x{\isacharparenright}\ {\isasymsubseteq}\ val{\isacharparenleft}G{\isacharcomma}y{\isacharparenright}{\isachardoublequoteclose}\isanewline
+%
+\ \ \isacommand{by}\isamarkupfalse%
+\ {\isacharparenleft}subst\ {\isacharparenleft}{\isadigit{1}}\ {\isadigit{2}}{\isacharparenright}\ def{\isacharunderscore}val{\isacharcomma}\ force{\isacharparenright}%
+\end{isabelle}
 

### 2018-12-17 LICS/zfc-axioms.tex

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
-\section{Models and axioms of set theory}
+\section{Axioms and models of set theory}
 
@@
 consisting of an only binary relation symbol $\in$. These include the
-axioms of Pairing, Union, Powerset, Foundation, Infinity, and two
+axioms of Extensionality, Pairing, Union, Powerset, Foundation, Infinity, and two
 axiom-schemes collectively referred to as Axiom of Separation:
@@
 The rest of the axioms are also
-included. We single out Separation and Replacement: these are written
+included. We single out Separation and Replacement: These are written
 for formulas with at most one extra parameter (meaning $n\leq 1$ in the
@@
 versions are equivalent to the usual formulations. We are only able to
-prove that the extension satisfies Separation for any particular
+prove that the generic extension satisfies Separation for any particular
 number of parameters, but not in general. This is a consequence that

### 2018-12-18 LICS/easy-axioms.tex
Primer texto para foundation y extensionality

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
-\section{Foundation, Union, Infinity}
+\section{Extensionality, Foundation, Union, Infinity}
 \label{sec:easy-axioms}
+\newcommand{\forallRel}[2]{\forall #1\kern -1pt[#2]}
+It is straightforward to show that the generic extension $M[G]$
+satisfies extensionality, foundation, union, and infinity. % It is
+% well-known that extensionality is satisfied by any
+% transitive class.\footnote{We plan to define a new hierarchy of
+%   locales for ZF where transitive classes will be at the base.}
+To say that $A$ satisfies some axiom, say extensionality, means
+that the relativized version of extensionality is true in $A$:
+\[
+\forallRel{x}{A}. \forallRel{y}{A}. (\forallRel{w}{A} . w \in x \leftrightarrow w\in y) \rightarrow x = y
+\]
 
-It is straightforward to show that the generic extension $M[G]$
-satisfies foundation, union, and infinity. The easiest one is
-foundation which does not depend on the satisfaction by $M$.
+Extensionality for $M[G]$ is straightforward because $M[G]$ is
+transitive. %\footnote{We generalized slightly some results by Paulson:
+%  in fact, several basic lemmas are valid for any transitive class
+%  (for instance, absoluteness of Union). We plan to define a new
+%  hierarchy of locales for ZF where transitive classes will be at the
+%  base.}
+In fact, for $x,y\in M[G]$, we can prove
+$\forall w. w\in x \leftrightarrow w\in y$ (notice that the binding
+$w$ is not relativized), because $M[G]$ is transitive; therefore by
+extensionality we conclude $x=y$.
 
-\fbox{Hablar de \texttt{M\_trivial} y del uso de interfaz para instanciar}
+Foundation for $M[G]$ does not depend on $M[G]$ being transitive: in
+this case we take $x\in M[G]$ and prove that there is an
+$\in$\kern -1pt-minimal element, relativized to $M[G]$, in $x$. Using
+foundation for $x\cap M[G]$ we get a minimal $y$ (which is not
+relativized to $M[G]$), so it is still minimal when relativized to
+$M[G]$.
+
+The Union axiom asserts that if $x$ is a set, then there exists
+another set (the union of $x$) containing all the elements in each
+element of $x$. The relativized version of Union asks to give a name
+$\pi_a$ for each $a\in M[G]$ and proving $\val(G,\pi)=\bigcup a$.
+
+\fbox{Hablar de \texttt{M\_trivial} y del uso de interfaz para
+  instanciar}
 

### 2018-12-18 LICS/forcing.tex
un poco de fabstracts

@@
 
-\fbox{poner acá lo de formal abstracts y el locale \texttt{forcing\_thms}}
+The presentation of the Fundamental Theorems of Forcing in a locale
+can be regarded as a \emph{formal abstract} as in the project
+envisioned by Hales \cite{fabstracts}, where  statements of
+mathematical theorems proven in the literature are posed in a language
+that is both human- and computer-readable. The point is to take
+particular care so that, v.g., there are no missing hypothesis, so it
+is possible to take this statement as firm ground on which start a
+formalization of a proof.
+
 %%% Local Variables: 

### 2018-12-19 LICS/easy-axioms.tex

@@
 \label{sec:easy-axioms}
-\newcommand{\forallRel}[2]{\forall #1\kern -1pt[#2]}
+\newcommand{\quantRel}[3]{#1 #2\kern -1pt[#3]}
+\newcommand{\forallRel}[2]{\quantRel{\forall}{#1}{#2}}
+\newcommand{\existsRel}[2]{\quantRel{\exists}{#1}{#2}}
+
 It is straightforward to show that the generic extension $M[G]$
-satisfies extensionality, foundation, union, and infinity. % It is
-% well-known that extensionality is satisfied by any
-% transitive class.\footnote{We plan to define a new hierarchy of
-%   locales for ZF where transitive classes will be at the base.}
-To say that $A$ satisfies some axiom, say extensionality, means
-that the relativized version of extensionality is true in $A$:
-\[
-\forallRel{x}{A}. \forallRel{y}{A}. (\forallRel{w}{A} . w \in x \leftrightarrow w\in y) \rightarrow x = y
-\]
+satisfies extensionality and foundation. Showing that $M[G]$ is closed
+under Union depends on $G$ being a filter. On the other hand, infinity
+takes some more effort.
+
+% To say that $A$ satisfies some axiom, say extensionality, means
+% that the relativized version of extensionality is true in $A$:
+% \[
+% \forallRel{x}{A}. \forallRel{y}{A}. (\forallRel{w}{A} . w \in x \leftrightarrow w\in y) \rightarrow x = y
+% \]
 
@@
 
-The Union axiom asserts that if $x$ is a set, then there exists
+The Union Axiom asserts that if $x$ is a set, then there exists
 another set (the union of $x$) containing all the elements in each
 element of $x$. The relativized version of Union asks to give a name
-$\pi_a$ for each $a\in M[G]$ and proving $\val(G,\pi)=\bigcup a$.
+$\pi_a$ for each $a\in M[G]$ and proving $\val(G,\pi_a)=\bigcup a$.
+Let $\tau$ be the name for $a$, i.e.\ $a=\val(G,\tau)$; 
+\citet{kunen2011set} gives $\pi_a$ in terms of $\tau$:
+\begin{align*}
+  \pi_a = \{\langle\theta,p \rangle &\in \dom(\bigcup(\dom(\tau))) \times \PP : \\
+&\exists \langle\sigma,q\rangle  \in \tau .
+ \exists r \in \PP . \langle \theta,r\rangle \in \sigma \wedge
+    p\leqslant r \wedge p \leqslant q \}
+\end{align*}
+Our formal definition is slightly different in order to ease the proof of
+$\pi_a \in M$.  Since $\pi_a$ it is defined using Separation, one
+needs to internalize $\pi_a$ as a formula
+\isa{union{\isacharunderscore}name{\isacharunderscore}fm}. The
+equation $\val(G,\pi_a)=\bigcup a$ is proved by showing the mutual
+inclusion; in both cases one uses that $G$ is a filter.
+\begin{isabelle}
+  \isacommand{lemma}\isamarkupfalse%
+\ Union{\isacharunderscore}MG{\isacharunderscore}Eq\ {\isacharcolon}\ \isanewline
+\ \ \isakeyword{assumes}\ {\isachardoublequoteopen}a\ {\isasymin}\ M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}\ \isakeyword{and}\ {\isachardoublequoteopen}a\ {\isacharequal}\ val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}{\isachardoublequoteclose}\ \isakeyword{and}\isanewline
+\ \ \ \ \ \ \ \ \ \ {\isachardoublequoteopen}filter{\isacharparenleft}G{\isacharparenright}{\isachardoublequoteclose}\ \isakeyword{and}\ {\isachardoublequoteopen}{\isasymtau}\ {\isasymin}\ M{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\ {\isachardoublequoteopen}{\isasymUnion}\ a\ {\isacharequal}\ val{\isacharparenleft}G{\isacharcomma}Union{\isacharunderscore}name{\isacharparenleft}{\isasymtau}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
+Since Union is absolute for any transitive class\footnote{This
+  generalizes a result proved by Paulson.}
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ Union{\isacharunderscore}abs{\isacharunderscore}trans\ {\isacharcolon}\ \isanewline
+\ \ \isakeyword{assumes}\ {\isachardoublequoteopen}Transset{\isacharparenleft}Q{\isacharparenright}{\isachardoublequoteclose}\ \isakeyword{and}\ {\isachardoublequoteopen}a\ {\isasymin}\ Q{\isachardoublequoteclose}\ \isakeyword{and}\isanewline
+\ \ \ \ \ \ \ \ \ \ {\isachardoublequoteopen}z\ {\isasymin}\ Q{\isachardoublequoteclose}\ \isakeyword{and}\ {\isachardoublequoteopen}{\isasymUnion}\ a\ {\isacharequal}\ z{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\ {\isachardoublequoteopen}big{\isacharunderscore}union{\isacharparenleft}{\isacharhash}{\isacharhash}Q{\isacharcomma}a{\isacharcomma}z{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
+\noindent we may conclude that $M[G]$ is closed under Union:
+\begin{isabelle}
+  \isacommand{lemma}\isamarkupfalse%
+  \ union{\isacharunderscore}in{\isacharunderscore}MG\ {\isacharcolon}\ \isanewline
+  \ \ \isakeyword{assumes}\ {\isachardoublequoteopen}filter{\isacharparenleft}G{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\ {\isachardoublequoteopen}Union{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharbrackleft}G{\isacharbrackright}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
 

### 2018-12-19 LICS/easy-axioms.tex
Entry for Formal abstracts

@@
 It is straightforward to show that the generic extension $M[G]$
-satisfies extensionality and foundation. Showing that $M[G]$ is closed
+satisfies extensionality and foundation. Showing that it is closed
 under Union depends on $G$ being a filter. On the other hand, infinity
@@
 element of $x$. The relativized version of Union asks to give a name
-$\pi_a$ for each $a\in M[G]$ and proving $\val(G,\pi_a)=\bigcup a$.
+$\pi_a$ for each $a\in M[G]$ and proving $\val(G,\pi_a)=\union a$.
 Let $\tau$ be the name for $a$, i.e.\ $a=\val(G,\tau)$; 
@@
 \begin{align*}
-  \pi_a = \{\langle\theta,p \rangle &\in \dom(\bigcup(\dom(\tau))) \times \PP : \\
+  \pi_a = \{\langle\theta,p \rangle &\in \dom(\union(\dom(\tau))) \times \PP : \\
 &\exists \langle\sigma,q\rangle  \in \tau .
@@
 \isa{union{\isacharunderscore}name{\isacharunderscore}fm}. The
-equation $\val(G,\pi_a)=\bigcup a$ is proved by showing the mutual
+equation $\val(G,\pi_a)=\union a$ is proved by showing the mutual
 inclusion; in both cases one uses that $G$ is a filter.

### 2018-12-19 LICS/forcing.tex
Entry for Formal abstracts

--- 2018-12-17 LICS/forcing.tex
@@
 
-\fbox{poner acá lo de formal abstracts y el locale \texttt{forcing\_thms}}
+The presentation of the Fundamental Theorems of Forcing in a locale
+can be regarded as a \emph{formal abstract} as in the project
+envisioned by Hales \cite{hales-fabstracts}, where  statements of
+mathematical theorems proven in the literature are posed in a language
+that is both human- and computer-readable. The point is to take
+particular care so that, v.g., there are no missing hypothesis, so it
+is possible to take this statement as firm ground on which start a
+formalization of a proof.
+
 %%% Local Variables: 

### 2018-12-19 LICS/zfc-axioms.tex
ed. varias, esp. Hacking y comentario para Miguel

@@
 \section{Axioms and models of set theory}
+\label{sec:axioms-models-set-theory}
 
@@
 $\ZF$ axioms that would be most
-compatible with the development of relativization by Paulson. For
+compatible with the development by Paulson. For
 instance, the predicate
@@
 for formulas with at most one extra parameter (meaning $n\leq 1$ in the
-above $\psi$). Thanks to Pairing, these
+above $\psi$), where \isatt{sats\,::\,"[i,i,i]=>o"} is the satisfaction
+relation between a set model with an environment and an internalized
+formula. Thanks to Pairing, these 
 versions are equivalent to the usual formulations. We are only able to
@@
 
+It is also possible define a predicate that states that a set
+satisfies a (possibly infinite) set of formulas, and then to state
+that ``$M$ satisfies $\ZF$'' in a standard way. With the
+aforementioned restriction on parameters, it can be shown that this
+statement is equivalent to the set of assumptions of the locale
+\isatt{M\_ZF}.
 %

### 2018-12-19 LICS/easy-axioms.tex
ed. varias, esp. Hacking y comentario para Miguel

@@
 \end{isabelle}
-Since Union is absolute for any transitive class\footnote{This
-  generalizes a result proved by Paulson.}
+
+Since Union is absolute for any transitive class
+\medskip
+
+\fbox{No generaliza a Paulson porque \'el prueba el sii.}
+
+\fbox{Y capaz su prueba de dos l\'{\i}neas tambi\'en lo hace}
+
+\fbox{ac\'a (como Fundaci\'on), as\'{\i} que cuidado.}
 \begin{isabelle}

### 2018-12-19 LICS/hacking.tex
ed. varias, esp. Hacking y comentario para Miguel

@@
 
-There are several lemmas that were declared, in later developments, as
-introduction/simplification rules (notably, the rule
-\isatt{equalityI}). They raised a warning, and we have 
-eliminated some of them, but in some cases we had to keep them because
-the proof works because it is insisted that the rule is \emph{safe}.
+In \cite{paulson_2003}, Paulson presented his formalization of the
+relative consistency of the Axiom of Choice. This development is
+included inside the Isabelle distribution with session name
+\isatt{ZF-Constructible}. The main technical devices, invented by
+G\"odel for this purpose, are \emph{relativization} and
+\emph{absoluteness}. In a nutshell, to relativize a formula $\phi$ to
+a class $C$, it is enough to restrict its quantifiers to $C$. The
+example of \isatt{upair\_ax} in
+Section~\ref{sec:axioms-models-set-theory}, the relativized version of
+the Pairing Axiom, is extracted from \texttt{Relative}, one of the
+core theories of \isatt{ZF-Constructible}. On the other hand, $\phi$
+is \emph{absolute} for $C$ if it is equivalent to its relativization,
+meaning that the statement made $\phi$ coincides with what $C$
+``believes'' $\phi$ is saying. Paulson shows that under certain
+hypothesis  on a class $M$ (condensed in the locale \isatt{M\_trivial}), a plethora of
+absoluteness and closure results can be proved about $M$.
+
+The development of forcing, and the study of ctms in general, takes
+absoluteness as a starting point. We were not able to work with
+\isatt{ZF-Constructible} right out-of-the-box. The main reason is that
+we can't expect state the ``class version'' of Replacement for a
+\emph{set} $M$ by
+using first-order formulas, since predicates \isatt{P::"i=>o"} can't
+be proved to be only the definable ones. Therefore, we had to make
+some modifications to the various sets of hypothesis in several
+locales to make the results available as tools for the present and
+future developments.
+
+%% There are several lemmas that were declared, in later developments, as
+%% introduction/simplification rules (notably, the rule
+%% \isatt{equalityI}). They raised a warning, and we have 
+%% eliminated some of them, but in some cases we had to keep them because
+%% the proof works because it is insisted that the rule is \emph{safe}.
 
@@
 \item\label{item:3} We replaced the need that the set of natural numbers is in $M$ by the
-  milder hypothesis that $M(0)$.
+  milder hypothesis that $M(0)$. Actually, most results should follow
+  by only assuming that $M$ is nonempty.
 \end{enumerate}
@@
 \isatt{L\_axioms} and \isatt{Internalize}, and we merged them to form
-the new file  \isatt{Internalizations}. Actually we believe that the
+the new file  \isatt{Internalizations}.
+
+\medskip
+\fbox{>>>Poner lo que sigue en las conclusiones???}
+\medskip
+
+We believe that the
 \isatt{ZF-Constructible} session  would benefit from some changes in
@@
 intance, by cataloging all or most of the internalized formulas in one
-file.
+file. Another, most basic example would be to start out with an even
+more locale that only assumes $M$ to be a nonempty transitive class,
+as many absoluteness results follow from these hypothesis.  
+It is our desire to advocate a future work to a thorough
+revision of the development of constructibility to maximize modularity.
 

### 2018-12-19 LICS/easy-axioms.tex
correcciones a primeros párrafos

@@
 
-Extensionality for $M[G]$ is straightforward because $M[G]$ is
-transitive. %\footnote{We generalized slightly some results by Paulson:
+For Extensionality in $M[G]$, the assumption 
+$\forall w[M[G]]. w\in x \leftrightarrow w\in y$ yields 
+$\forall w. w\in x \leftrightarrow w\in y$ by transitivity of $M[G]$. %\footnote{We generalized slightly some results by Paulson:
 %  in fact, several basic lemmas are valid for any transitive class
@@
 %  base.}
-In fact, for $x,y\in M[G]$, we can prove
-$\forall w. w\in x \leftrightarrow w\in y$ (notice that the binding
-$w$ is not relativized), because $M[G]$ is transitive; therefore by
-extensionality we conclude $x=y$.
+Therefore, by (ambient) Extensionality we conclude $x=y$.
 
 Foundation for $M[G]$ does not depend on $M[G]$ being transitive: in
-this case we take $x\in M[G]$ and prove that there is an
-$\in$\kern -1pt-minimal element, relativized to $M[G]$, in $x$. Using
-foundation for $x\cap M[G]$ we get a minimal $y$ (which is not
-relativized to $M[G]$), so it is still minimal when relativized to
-$M[G]$.
+this case we take $x\in M[G]$ and prove, relativized to $M[G]$,  that there is an
+$\in$\kern -1pt-minimal element in $x$. Instantiating the global Foundation
+Axiom for $x\cap M[G]$ we get a minimal $y$, so it is still minimal
+when considered relative to $M[G]$.
+
+It is noteworthy that the proofs in the Isar dialect of Isabelle
+strictly follow the argumentation of the two previous paragraphs.
 

### 2018-12-19 LICS/proof-of-separation.tex
line numbers, justo hasta donde la prueba viene igual

@@
 This proof can be found in the file \verb|Separation_Axiom.thy| of the
-development, which we proceed to discuss.
+development, which we proceed to discuss. The order chose to implement
+the proof sought to minimize the crossreference of facts;  it is
+not entirely appropriate for a text version, so we depart from it in
+this presentation.
 
@@
   \ \ \isakeyword{assumes}\isanewline
-  \ \ \ \ {\isachardoublequoteopen}{\isasympi}\ {\isasymin}\ M{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymsigma}\ {\isasymin}\ M{\isachardoublequoteclose}\ {\isachardoublequoteopen}val{\isacharparenleft}G{\isacharcomma}\ {\isasympi}{\isacharparenright}\ {\isacharequal}\ c{\isachardoublequoteclose}\ {\isachardoublequoteopen}val{\isacharparenleft}G{\isacharcomma}\ {\isasymsigma}{\isacharparenright}\ {\isacharequal}\ w{\isachardoublequoteclose}\isanewline
+  \ \ \ \ {\isachardoublequoteopen}{\isasympi}\ {\isasymin}\ M{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymsigma}\ {\isasymin}\ M{\isachardoublequoteclose}\ {\isachardoublequoteopen}val{\isacharparenleft}G{\isacharcomma}\ {\isasympi}{\isacharparenright}\ {\isacharequal}\ c{\isachardoublequoteclose}\isanewline
+  \ \ \ \  {\isachardoublequoteopen}val{\isacharparenleft}G{\isacharcomma}\ {\isasymsigma}{\isacharparenright}\ {\isacharequal}\ w{\isachardoublequoteclose}\isanewline
   \ \ \ \ {\isachardoublequoteopen}{\isasymphi}\ {\isasymin}\ formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}arity{\isacharparenleft}{\isasymphi}{\isacharparenright}\ {\isasymle}\ {\isadigit{2}}{\isachardoublequoteclose}\isanewline
@@
 The candidate name is
-\[
+\begin{equation}\label{eq:4}
 n \defi \{u \in\dom(\pi)\times\PP :M,[u,\PP,\leq,\1,\sig,\pi]\models \psi\}
-\]
+\end{equation}
 where
@@
 \]
-The fact that $n\in M$ follows by an application of a six-variable
-instance of Separation in $M$ (lemma \isatt{six{\isacharunderscore}sep{\isacharunderscore}aux}).
+The fact that $n\in M$ follows (lines 216--220 of the source file) by
+an application of a six-variable instance of Separation in $M$ (lemma
+\isatt{six{\isacharunderscore}sep{\isacharunderscore}aux}). We note in
+passing that it is possible to  abbreviate expressions in Isabelle by
+the use of the  \textbf{let} statement or the \textbf{is} qualifier,
+and metavariables (whose
+identifiers start with a question mark). In this way, the definition in
+(\ref{eq:4}) appears in the sources as letting \isatt{?n} to be that
+set (lines 208--211).
 
@@
 relation and %% permuting variables
-absoluteness, we have that it is equivalent to the fact
+absoluteness, we have (lines 92--98) that it is equivalent to the fact
 that there exist $\th,p\in M$ with   $u=\lb\th,p\rb$  and 
@@
 \end{equation}
-We can instantiate this statement with $G$ and obtain
+(lines 99--185). We can instantiate this statement with $G$ and obtain
+(lines 186--206)
 \[
@@
 \]
-Hence $n$ is included in 
+Hence (lines 207-212) $n$ is included in 
 \[
@@
 
-It can be seen that
+It can be seen (lines 221--274) that
 \begin{lemma}
@@
 Also by monotonicity of $\val$ we obtain 
-  $\val(G,n)\sbq \val(G,m)$.
+  $\val(G,n)\sbq \val(G,m)$ (lines 213--215).
 To complete the proof, it is therefore enough to show that
@@
 \]
-which by the Truth Lemma gives us: \emph{for all $M$-generic $F$,
+which by the Definition of Forcing gives us: \emph{for all $M$-generic $F$,
   $p\in F$ implies} 

### 2018-12-19 LICS/forcing.tex
line numbers, justo hasta donde la prueba viene igual

@@
 $\forceisa(\phi)$ will have $n+4$ free variables, where the first four account
-for the forcing notion and a particular condition.
+for the forcing notion and a particular element of it.
 
@@
 that is both human- and computer-readable. The point is to take
-particular care so that, v.g., there are no missing hypothesis, so it
+particular care so that, v.g., there are no missing hypotheses, so it
 is possible to take this statement as firm ground on which start a

### 2018-12-20 LICS/proof-of-separation.tex
separation line by line

@@
 \begin{lemma}
-  $\val(G,m) = S$.
+  $\val(G,m) = \{x\in c : M[G],[x,w,c]\models \phi(x_0,x_1)\}$.
 \end{lemma}
-Also by monotonicity of $\val$ we obtain 
+The right-hand side is trivially equal to $S$, but as a consequence of
+the definition of 
+\isatt{separation\_ax}, the result contains an extra $c$ in the
+environment.
+
+Also, by monotonicity of $\val$ we obtain 
   $\val(G,n)\sbq \val(G,m)$ (lines 213--215).
-To complete the proof, it is therefore enough to show that
-$S\sbq \val(G,n)$. For this, let $x\in S$. Hence there exists
-$\lb\th,q\rb\in\pi$ such that 
-$q\in G$ and $x=\val(G,\th)$.
+To complete the proof, it is therefore enough to show the other
+inclusion (starting at line 275).
+%$\val(G,m)\sbq \val(G,n)$ 
+For this, let $x\in \val(G,m) = S$ and then $x\in c$. Hence there exists
+$\lb\th,q\rb\in\pi$ such that $q\in G$ and $x=\val(G,\th)$.
 
-On the other hand, since 
+On the other hand, since (line 297)
 \[
@@
 \]
-which by the Definition of Forcing gives us: \emph{for all $M$-generic $F$,
+which by the Definition of Forcing gives us (lines 315--318): \emph{for all $M$-generic $F$,
   $p\in F$ implies} 
@@
 \]
-Note this is the same as (\ref{eq:2}). In fact, we have that
-\begin{lemma}
-  $\val(G,n)$ equals
-  \[
-  \{\val(G,\th): \th\in\dom(\pi) \y  p \in G 
-  \y \text{(\ref{eq:2}) holds}\}.
-  \]
-\end{lemma}
-This shows that $S\sbq \val(G,n)$, since $\th\in\dom(\pi)$ by construction.
+Note this is the same as (\ref{eq:2}). Hence, tracing the equivalence
+up to (\ref{eq:1}), we can show that $x=\val(G,\th)\in \val(G,n)$
+(lines 319--337) and we are done.
 

### 2018-12-20 LICS/easy-axioms.tex
Explicación de Infinity. Nota para introducir check y valcheck.

@@
 \begin{align*}
-  \pi_a = \{\langle\theta,p \rangle &\in \dom(\union(\dom(\tau))) \times \PP : \\
-&\exists \langle\sigma,q\rangle  \in \tau .
- \exists r \in \PP . \langle \theta,r\rangle \in \sigma \wedge
+  \pi_a = \{\langle\theta,p \rangle :  %&\in \dom(\union(\dom(\tau))) \times \PP : \\
+\exists \langle\sigma,q\rangle  \in \tau .
+ \exists r . \langle \theta,r\rangle \in \sigma \wedge
     p\leqslant r \wedge p \leqslant q \}
@@
 \end{isabelle}
-
-Since Union is absolute for any transitive class
-\medskip
-
-\fbox{No generaliza a Paulson porque \'el prueba el sii.}
-
-\fbox{Y capaz su prueba de dos l\'{\i}neas tambi\'en lo hace}
-
-\fbox{ac\'a (como Fundaci\'on), as\'{\i} que cuidado.}
-\begin{isabelle}
-\isacommand{lemma}\isamarkupfalse%
-\ Union{\isacharunderscore}abs{\isacharunderscore}trans\ {\isacharcolon}\ \isanewline
-\ \ \isakeyword{assumes}\ {\isachardoublequoteopen}Transset{\isacharparenleft}Q{\isacharparenright}{\isachardoublequoteclose}\ \isakeyword{and}\ {\isachardoublequoteopen}a\ {\isasymin}\ Q{\isachardoublequoteclose}\ \isakeyword{and}\isanewline
-\ \ \ \ \ \ \ \ \ \ {\isachardoublequoteopen}z\ {\isasymin}\ Q{\isachardoublequoteclose}\ \isakeyword{and}\ {\isachardoublequoteopen}{\isasymUnion}\ a\ {\isacharequal}\ z{\isachardoublequoteclose}\isanewline
-\ \ \isakeyword{shows}\ {\isachardoublequoteopen}big{\isacharunderscore}union{\isacharparenleft}{\isacharhash}{\isacharhash}Q{\isacharcomma}a{\isacharcomma}z{\isacharparenright}{\isachardoublequoteclose}
-\end{isabelle}
-\noindent we may conclude that $M[G]$ is closed under Union:
+Since Union is absolute for any transitive class we may conclude that
+$M[G]$ is closed under Union:
 \begin{isabelle}
@@
 
-\fbox{Hablar de \texttt{M\_trivial} y del uso de interfaz para
-  instanciar}
-
-\fbox{Falta \texttt{check\_in\_M} para completar Infinity}
+The proof of Infinity for $M[G]$ takes advantage of some absoluteness
+results proved in the locale \isa{M{\isacharunderscore}trivial}. Since
+we have already proved that $M[G]$ is transitive, $\emptyset\in M[G]$
+assuming $G$ being generic, and also that it satisfies Pairing and
+Union, we can instantiate \isa{M{\isacharunderscore}trivial}:
+\begin{isabelle}
+\isacommand{sublocale}\isamarkupfalse%
+\ G{\isacharunderscore}generic\ {\isasymsubseteq}\ M{\isacharunderscore}trivial{\isachardoublequoteopen}{\isacharhash}{\isacharhash}M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}
+\end{isabelle}
+We assume that $M$ satisfies Infinity,\footnote{This is equivalent to
+  say that there is an $I\in M$ having an empty set relativized to $M$
+  and closed under successor, again relativized to $M$.} therefore we
+obtain $I \in M$ such that $\emptyset\in I$ and, $x \in I$ implies
+$\mathit{succ}(x)\in I$; this follows by absoluteness of empty and
+succesor for $M$. Using one further assumption, namely that
+$\chk(x) \in M$ if $x\in M$, we deduce $\val(G,\chk(I)) = I \in M[G]$.
+Now we can use absoluteness of emptyness and successor, this time for
+$M[G]$, to conclude that $M[G]$ satisfies Infinity.
+\begin{isabelle}
+  \isacommand{lemma}\isamarkupfalse%
+\ infinty{\isacharunderscore}in{\isacharunderscore}MG\ {\isacharcolon}\ {\isachardoublequoteopen}infinity{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharbrackleft}G{\isacharbrackright}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
+\noindent\fbox{En Generic Extensions decimos que es fácil ver que $M[G]$}
+\fbox{ ... incluye a $M \cup {G}$, por qué no sacar $I\in M[G]$ de ahí?}
 

### 2018-12-20 LICS/easy-axioms.tex
mainly prueba de separación, labels por sección

@@
 \end{isabelle}
-We assume that $M$ satisfies Infinity,\footnote{This is equivalent to
-  say that there is an $I\in M$ having an empty set relativized to $M$
-  and closed under successor, again relativized to $M$.} therefore we
+We assume that $M$ satisfies Infinity, i.e., that Infinity
+relativized to $M$ holds; therefore we
 obtain $I \in M$ such that $\emptyset\in I$ and, $x \in I$ implies
-$\mathit{succ}(x)\in I$; this follows by absoluteness of empty and
-succesor for $M$. Using one further assumption, namely that
+$\mathit{succ}(x)\in I$  by absoluteness of empty and
+successor for $M$. Using one further assumption, namely that
 $\chk(x) \in M$ if $x\in M$, we deduce $\val(G,\chk(I)) = I \in M[G]$.
@@
 \fbox{ ... incluye a $M \cup {G}$, por qué no sacar $I\in M[G]$ de ahí?}
+\medskip
 
+\fbox{Puede ser, pero eso est\'a dentro de  un locale}
+
+\fbox{que asume más cosas (una instancia de Repl)}
 %%% Local Variables: 

### 2018-12-20 LICS/forcing.tex
mainly prueba de separación, labels por sección

@@
 recursion on 
-the structure of $\phi$.
+the structure of $\phi$. It is to be noted that the base case (viz.,
+for atomic $\phi$) contains all the complexity, and indeed it can be
+said that writing down this definition is half
+a way the path to set up the forcing machinery.
+\medskip
 
-\fbox{\em Compatibilizar lo que sigue con Renaming}
+\fbox{Compatibilizar lo que sigue con Renaming}
+\medskip
 
@@
 
-We will now make more precise the definition of the map
+We will now make more precise the properties of the map
 $\forceisa$ and how it relates satisfaction in $M$ to that in
@@
 \subsection{The fundamental theorems}
-These are stated in \cite{kunen2011set} as two lemmas: \emph{Truth} and
-\emph{Definability}.  Modern treatments of the theory of forcing start
+\label{sec:fundamental-theorems}
+Modern treatments of the theory of forcing start
 by defining the 
@@
 
-The Truth Lemma states that the forcing
+The \emph{Truth Lemma} states that the forcing
 relation indeed relates 
@@
 \end{lemma}
-The proof of the Truth Lemma is by recursion in elements of
-$\formula$. The following auxiliary results (adapted from
-\cite[IV.2.43]{kunen2011set}) are also proved by recursion in
-$\formula$.
+The previous two results combined are the ones usually called the
+\emph{fundamental theorems}.
+
+The following auxiliary results (adapted from
+\cite[IV.2.43]{kunen2011set}) are also handy in forcing arguments.
 \begin{lemma}[Strengthening]\label{lem:strengthen} 
@@
 \end{lemma}
+%
+All these results are proved by recursion in
+$\formula$.
 

### 2018-12-20 LICS/hacking.tex
otro archivo con cambios
mainly prueba de separación, labels por sección

@@
 \section{Hacking of \isatt{ZF-Constructible}}
-
+\label{sec:hack-constructible}
 In \cite{paulson_2003}, Paulson presented his formalization of the
@@
 We moved the requirement $M(\mathtt{nat})$ to the locale
-\isatt{M\_trancl}, where it is needed for the first time. Some results,
+\isatt{M\_trancl} (inside the theory \isatt{WF\_absolute}), where it is needed for the first time. Some results,
 for instance \isatt{rtran\_closure\_mem\_iff} and 

### 2018-12-20 LICS/isabelle.tex
mainly prueba de separación, labels por sección

@@
 \section{Isabelle}
-
+\label{sec:isabelle}
 %-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-
 \subsection{Logics}
-
+\label{sec:logics}
 Isabelle provides a meta-language called \emph{Pure} that consists of
@@
 \]
-(as usual,  \isatt{\isasymLongrightarrow} asociates to the right), and
-syntactic sugar is provided to curry the previous as follows:
+(as usual,  \isatt{\isasymLongrightarrow} associates to the right), and
+syntactic sugar is provided to curry the previous rule as follows:
 \[
@@
 \subsection{Locales}
-
+\label{sec:locales}
 Locales \cite{ballarin2010tutorial} provide a neat facility to
@@
 \item[\texttt{forcing\_thms}]: eso.
-\item[\texttt{G\_generic}]: G es genérico.
+\item[\texttt{G\_generic}]: lo anterior y G es genérico.
 \item[\texttt{M\_extra\_assms}]: check in M e instancia de reemplazo para G.

### 2018-12-20 LICS/proof-of-separation.tex
mainly prueba de separación, labels por sección

@@
 \section{Proof of Separation}
+\label{sec:proof-separation}
 
 This proof can be found in the file \verb|Separation_Axiom.thy| of the
-development, which we proceed to discuss. The order chose to implement
-the proof sought to minimize the crossreference of facts;  it is
+development, which we proceed to discuss in detail. The order chosen to
+implement
+the proof sought to minimize the cross-reference of facts;  it is
 not entirely appropriate for a text version, so we depart from it in
-this presentation.
+this presentation. Nevertheless, we will refer to each specific block
+of the code by line numer for ease of reference.
 
@@
 
-It can be seen (lines 221--274) that
-\begin{lemma}
-  $\val(G,m) = \{x\in c : M[G],[x,w,c]\models \phi(x_0,x_1)\}$.
-\end{lemma}
+A result from the theory \isatt{Names} allows to calculate the value
+of sets such as $m$ that frequently are used in forcing arguments: 
+$\val(G,\{x\in A\times \PP.\ Q(x)\})$ equals
+\[
+\{\val(G,t) : t\in A , \exists
+  p\in \PP .\ Q(\lb t,p\rb) \y p \in G\}.
+\]
+This is then used to show  (lines 221--274 of \isatt{Separation\_Axiom}) that 
+\[
+  \val(G,m) = \{x\in c : M[G],[x,w,c]\models \phi(x_0,x_1)\}.
+\]
 The right-hand side is trivially equal to $S$, but as a consequence of
@@
 up to (\ref{eq:1}), we can show that $x=\val(G,\th)\in \val(G,n)$
-(lines 319--337) and we are done.
+(lines 319--337), finishing the main lemma.
+
+The last 20 lines of the theory show, using absoluteness, the two
+instances of Separation for $M[G]$:
+
+\begin{isabelle}
+\isacommand{theorem}\isamarkupfalse%
+\ separation{\isacharunderscore}in{\isacharunderscore}MG{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\ \isanewline
+\ \ \ \ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\ \isakeyword{and}\ {\isachardoublequoteopen}arity{\isacharparenleft}{\isasymphi}{\isacharparenright}\ {\isacharequal}\ {\isadigit{1}}\ {\isasymor}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharequal}{\isadigit{2}}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\ \ \isanewline
+\ \ \ \ {\isachardoublequoteopen}{\isacharparenleft}{\isasymforall}a{\isasymin}{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharparenright}{\isachardot}\ separation{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}{\isasymlambda}x{\isachardot}\ sats{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}{\isasymphi}{\isacharcomma}{\isacharbrackleft}x{\isacharcomma}a{\isacharbrackright}{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
 

### 2018-12-20 LICS/intro.tex
ispell y otras cosas
parrafo incompleto

@@
 % no \IEEEPARstart
+Zermelo-Fraenkel Set Theory $\ZF$) has a prominent place among formal
+theories. The reason for this is that it formalizes many intuitive
+properties of the notion of set. As such, it can be used as a
+foundation for mathematics and thus it has been
+thoroughly studied.
+
+One of the main initial concerns was to decide if this foundation
+completely axiomatized all of mathematics. As it was shown by G\"odel
+in 1931, no  such axiomatization (therefore, including artihmetic) can
+be complete. A refinement of this result, is G\"odel's Second
+Incompleteness Theorem that says that no consistent formal system that
+includes arithmetic can prove its own consistency. As a matter of
+fact, many problem within the area of Set Theory, not directly related
+to consistency or logic, have been proved to be undecided by the
+theory. Two techniques have been developed to do this. One is G\"odel
+theory of \emph{inner models}, which started with his model $L$ of
+\emph{Axiom of Constructibility} and was used to proved the relative
+consistency of the Axiom of Choice and the Generalized Continuum
+Hypothesis with $\ZF$.  he has been formalized by Lawrence
+Paulson\cite{paulson_2003} in the proof assistant \emph{Isabelle},
+
 We follow \cite{kunen2011set}. An introduction to the issues discussed

### 2018-12-20 LICS/intro.tex
una parte de la intro. Faltan algunas cosas

@@
 to consistency or logic, have been proved to be undecided by the
-theory. Two techniques have been developed to do this. One is G\"odel
+theory. Two techniques have been developed to do this. One is Kurt G\"odel's
 theory of \emph{inner models}, which started with his model $L$ of
-\emph{Axiom of Constructibility} and was used to proved the relative
+\emph{Axiom of Constructibility} \cite{godel-L} and was used to proved the relative
 consistency of the Axiom of Choice and the Generalized Continuum
-Hypothesis with $\ZF$.  he has been formalized by Lawrence
-Paulson\cite{paulson_2003} in the proof assistant \emph{Isabelle},
+Hypothesis with $\ZF$. The second one is the technique of
+\emph{forcing}, which was developed 30 years later by Paul
+J. Cohen~\cite{Cohen-CH-PNAS}.
+
+The constructible universe  has been formalized by Lawrence
+Paulson \cite{paulson_2003} in the proof assistant \emph{Isabelle}, and
+the present paper is part of a project of formalizing the whole of the
+forcing machinery, continuing the work by Paulson. The starting point
+of this project is our paper \cite{2018arXiv180705174G}, to which we
+refer the reader for further details on basic matters.
 
@@
 
+We briefly describe the contents of each
+section. Section~\ref{sec:isabelle} contains the bare minimun
+requirements to understand the (meta)logics used in Isabelle. Next, an
+overview of the model theory of set theory is presented in
+Section~\ref{sec:axioms-models-set-theory}. There is an ``internal''
+representation of first-order formulas as sets, implemented by
+Paulson; Section~\ref{sec:renaming} discusses syntactical
+transformations of the former, mainly permutation of variables. 
+In Section~\ref{sec:generic-extensions} the generic extensions are
+succintly reviewed and how the treatment of well founded recursion in
+Isabelle was enhanced. We take care of the ``easy axioms'' in
+Section~\ref{sec:easy-axioms}; these are the ones that
+do not depend on the forcing theorems. We describe the latter in
+Section~\ref{sec:forcing}. We adapted the  work by Paulson to our
+needs, and this is described in
+Section~\ref{sec:hack-constructible}. We present the proof
+of the Separation Axiom Scheme in Section~\ref{sec:proof-separation},
+which follows closely its implementation. A plan for future work and
+some immediate conclusions are offered in
+Section~\ref{sec:conclusions-future-work}.
+
 An enumeration of achievements to be presented:
 \begin{enumerate}
-\item Discussion of the Isabelle view of life.
-  \begin{enumerate}
-  \item Meta (Pure) vs Object (FOL+ZF) theory.
-  \item Object theory is presented axiomatically; in particular,
-    first-order ``formulas'' are not such (i.e., the \emph{initial}
-    algebra of formulas is \textbf{not} being specified). Thus, we are
-    not able to perform recursion on terms of type \tyo.
-  \end{enumerate}
 \item Discussion of locales.
@@
   \end{enumerate}
-\item Hacking of \isatt{ZF-Constructible}.
-  \begin{enumerate}
-  \item Elimination of the \tyo-schematic Replacement Axiom from
-    locales.
-  \item Replacing \isatt{M(nat)} by \isatt{M(0)} (merely non
-    emptiness) from Relative (and other places?).
-  \end{enumerate}
 \item Inteface between first-order properties of $M$ and those of
   $\#\#M$.
-  \begin{enumerate}
-  \item Enumerate properties and statistics.
-  \end{enumerate}
-\item Details of the proof of Separation.
 \end{enumerate}

### 2018-12-20 LICS/intro.tex
más intro y estética

@@
 % no \IEEEPARstart
-Zermelo-Fraenkel Set Theory $\ZF$) has a prominent place among formal
+Zermelo-Fraenkel Set Theory ($\ZF$) has a prominent place among formal
 theories. The reason for this is that it formalizes many intuitive
@@
 in 1931, no  such axiomatization (therefore, including artihmetic) can
-be complete. A refinement of this result, is G\"odel's Second
-Incompleteness Theorem that says that no consistent formal system that
+be complete. A refinement of this result, G\"odel's Second
+Incompleteness Theorem, says that no consistent formal system that
 includes arithmetic can prove its own consistency. As a matter of
-fact, many problem within the area of Set Theory, not directly related
+fact, many problems within the area of Set Theory, not directly related
 to consistency or logic, have been proved to be undecided by the
@@
 \emph{forcing}, which was developed 30 years later by Paul
-J. Cohen~\cite{Cohen-CH-PNAS}.
+J. Cohen~\cite{Cohen-CH-PNAS}. In essence, forcing is the only known
+way of \emph{extending} models of $\ZF$.
 
@@
 forcing machinery, continuing the work by Paulson. The starting point
-of this project is our paper \cite{2018arXiv180705174G}, to which we
+of this project is our previous paper 
+\cite{2018arXiv180705174G}, to which we 
 refer the reader for further details on basic matters.
 
-We follow \cite{kunen2011set}. An introduction to the issues discussed
-in this paper can be found in \cite{2018arXiv180705174G}. An
-introduction to forcing can be found at \cite{chow-beginner-forcing},
-and the book \cite{weaver2014forcing} contains a thorough treament
-minimizing the technicalities.
+In a gross simplification, there are two aspects to a formalization
+project like this one: thematic and programmatic. The first concerns
+the handling of all the theoretical concepts and results in the
+subject, while the second involves the practical issues of the
+implementation and design. In the case of forcing, the main intricacy
+lies in the first aspect. In this sense, following a sensible
+presentation of the material is key.  The authoritative reference 
+on the subject during the last 30 years has been Kunen's classical
+\cite{kunen1980}. In our
+formalizaton we have followed a recent rewrite \cite{kunen2011set}
+of that  textbook,which presents the material in the same sharp 
+style but offering a lot of details. In some sense this project
+wouldn't exist without this book. The reader interested in an
+introduction to the subject can check
+\cite{chow-beginner-forcing}; also, the book \cite{weaver2014forcing}
+contains a thorough treament minimizing the technicalities.
 

### 2018-12-20 LICS/forcing.tex
ediciones mínimas

@@
 In order to state the properties of this map
-in sufficient generality to prove that  $M[G]$ satisfies $\ZFC$, we work with
+in sufficient generality to prove that  $M[G]$ satisfies $\ZF$, we work with
 internalized formulas, because it is not possible to carry inductive
@@
 \begin{lemma}[Definition of Forcing]\label{lem:definition-of-forcing}
-  Let $M$ be a ctm of $\ZFC$, $\lb\PP,\leq,\1\rb$ a forcing notion
+  Let $M$ be a ctm of $\ZF$, $\lb\PP,\leq,\1\rb$ a forcing notion
   in $M$, $p\in\PP$, and $\phi(x_0,\dots,x_n)$ a formula in the

### 2018-12-20 LICS/hacking.tex
ediciones mínimas

@@
 ``believes'' $\phi$ is saying. Paulson shows that under certain
-hypothesis  on a class $M$ (condensed in the locale \isatt{M\_trivial}), a plethora of
+hypotheses  on a class $M$ (condensed in the locale \isatt{M\_trivial}), a plethora of
 absoluteness and closure results can be proved about $M$.
@@
 be proved to be only the definable ones. Therefore, we had to make
-some modifications to the various sets of hypothesis in several
+some modifications to the various sets of hypotheses in several
 locales to make the results available as tools for the present and
@@
 more locale that only assumes $M$ to be a nonempty transitive class,
-as many absoluteness results follow from these hypothesis.  
+as many absoluteness results follow from these hypotheses.  
 It is our desire to advocate a future work to a thorough

### 2018-12-20 LICS/generic-extensions.tex
mainly prueba de separación, labels por sección
ediciones mínimas

@@
 \section{Generic extensions}
-
+\label{sec:generic-extensions}
 We will swiftly review some definitions in order to reach the concept
-of \emph{generic extension}. Given a ctm $M$ of $\ZFC$, a forcing
+of \emph{generic extension}. Given a ctm $M$ of $\ZF$, a forcing
 notion $\lb\PP,\leq,\1\rb$, and a filter
@@
 includes $M\cup\{G\}$. Nevertheless there are no a priori reasons for
-$M[G]$ to satisfy the rest of the $\ZFC$ 
+$M[G]$ to satisfy the rest of the $\ZF$ 
 axioms. The original insight by Cohen was to define the notion of
 \emph{genericity} for a filter $G\sbq\PP$ and to prove that whenever
-$G$ is generic, $M[G]$ will satisfy $\ZFC$.
+$G$ is generic, $M[G]$ will satisfy $\ZF$.
 

### 2018-12-21 LICS/easy-axioms.tex
Ligeras variaciones en easy-axioms.

@@
 \label{sec:easy-axioms}
-\newcommand{\quantRel}[3]{#1 #2\kern -1pt[#3]}
-\newcommand{\forallRel}[2]{\quantRel{\forall}{#1}{#2}}
-\newcommand{\existsRel}[2]{\quantRel{\exists}{#1}{#2}}
 
-It is straightforward to show that the generic extension $M[G]$
-satisfies extensionality and foundation. Showing that it is closed
-under Union depends on $G$ being a filter. On the other hand, infinity
-takes some more effort.
-
-% To say that $A$ satisfies some axiom, say extensionality, means
-% that the relativized version of extensionality is true in $A$:
-% \[
-% \forallRel{x}{A}. \forallRel{y}{A}. (\forallRel{w}{A} . w \in x \leftrightarrow w\in y) \rightarrow x = y
-% \]
+In our first presentation of this project \cite{2018arXiv180705174G},
+we proved tha $M[G]$ satisfies Pairing. It is straightforward to show
+that the generic extension $M[G]$ satisfies extensionality and
+foundation. Showing that it is closed under Union depends on $G$ being
+a filter. Infinity is also easy, but it depends in one further
+assumption.
 
@@
 \end{align*}
-Our formal definition is slightly different in order to ease the proof of
-$\pi_a \in M$.  Since $\pi_a$ it is defined using Separation, one
-needs to internalize $\pi_a$ as a formula
-\isa{union{\isacharunderscore}name{\isacharunderscore}fm}. The
-equation $\val(G,\pi_a)=\union a$ is proved by showing the mutual
-inclusion; in both cases one uses that $G$ is a filter.
+Our formal definition is slightly different in order to ease the proof
+of $\pi_a \in M$; as it is defined using Separation, so one needs to
+define the domain of separation and also internalize the predicate as
+a formula
+\isa{union{\isacharunderscore}name{\isacharunderscore}fm}. Instead of
+working directly with the internalized formula, we define a predicate
+\isa{Unioin{\isacharunderscore}name{\isacharunderscore}body} and prove the equivalence between
+\isa{sats(M,union{\isacharunderscore}name{\isacharunderscore}fm,[P,leq,\isasymtau,x])}
+and
+\isa{Union{\isacharunderscore}Name{\isacharunderscore}body(P,leq,\isasymtau,x)}. The
+definition of $\pi_a$ in our formalization is:
+\begin{isabelle}
+\isacommand{definition}\isamarkupfalse%
+\ Union{\isacharunderscore}name\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i\ {\isasymRightarrow}\ i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
+\ \ {\isachardoublequoteopen}Union{\isacharunderscore}name{\isacharparenleft}{\isasymtau}{\isacharparenright}\ {\isacharequal}{\isacharequal}\ \isanewline
+\ \ \ \ {\isacharbraceleft}u\ {\isasymin}\ domain{\isacharparenleft}{\isasymUnion}{\isacharparenleft}domain{\isacharparenleft}{\isasymtau}{\isacharparenright}{\isacharparenright}{\isacharparenright}\ {\isasymtimes}\ P\ {\isachardot}\isanewline
+\ \ \ \  \ \ \ \ Union{\isacharunderscore}name{\isacharunderscore}body{\isacharparenleft}P{\isacharcomma}leq{\isacharcomma}{\isasymtau}{\isacharcomma}u{\isacharparenright}{\isacharbraceright}{\isachardoublequoteclose}
+\end{isabelle}
+
+Once we know $\pi_a \in M$, the equation $\val(G,\pi_a)=\union a$ is
+proved by showing the mutual inclusion; in both cases one uses that
+$G$ is a filter.
 \begin{isabelle}
@@
 The proof of Infinity for $M[G]$ takes advantage of some absoluteness
-results proved in the locale \isa{M{\isacharunderscore}trivial}. Since
-we have already proved that $M[G]$ is transitive, $\emptyset\in M[G]$
-assuming $G$ being generic, and also that it satisfies Pairing and
-Union, we can instantiate \isa{M{\isacharunderscore}trivial}:
+results proved in the locale \isa{M{\isacharunderscore}trivial}; this
+proof is easy because we assume $\chk(x) \in M$ whenever $x\in
+M$. Since we have already proved that $M[G]$ is transitive,
+$\emptyset\in M[G]$ assuming $G$ being generic, and also that it
+satisfies Pairing and Union, we can instantiate
+\isa{M{\isacharunderscore}trivial}:
 \begin{isabelle}
@@
 \end{isabelle}
-We assume that $M$ satisfies Infinity, i.e., that Infinity
-relativized to $M$ holds; therefore we
-obtain $I \in M$ such that $\emptyset\in I$ and, $x \in I$ implies
-$\mathit{succ}(x)\in I$  by absoluteness of empty and
-successor for $M$. Using one further assumption, namely that
-$\chk(x) \in M$ if $x\in M$, we deduce $\val(G,\chk(I)) = I \in M[G]$.
-Now we can use absoluteness of emptyness and successor, this time for
-$M[G]$, to conclude that $M[G]$ satisfies Infinity.
-\begin{isabelle}
-  \isacommand{lemma}\isamarkupfalse%
-\ infinty{\isacharunderscore}in{\isacharunderscore}MG\ {\isacharcolon}\ {\isachardoublequoteopen}infinity{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharbrackleft}G{\isacharbrackright}{\isacharparenright}{\isachardoublequoteclose}
-\end{isabelle}
-\noindent\fbox{En Generic Extensions decimos que es fácil ver que $M[G]$}
-\fbox{ ... incluye a $M \cup {G}$, por qué no sacar $I\in M[G]$ de ahí?}
-\medskip
+We assume that $M$ satisfies Infinity, i.e., that Infinity relativized
+to $M$ holds; therefore we obtain $I \in M$ such that $\emptyset\in I$
+and, $x \in I$ implies $\mathit{succ}(x)\in I$ by absoluteness of
+empty and successor for $M$. Using the assumption that $M$ is closed
+under $\chk$, we deduce $\val(G,\chk(I)) = I \in M[G]$.  Now we can
+use absoluteness of emptyness and successor, this time for $M[G]$, to
+conclude that $M[G]$ satisfies Infinity.
+% \begin{isabelle}
+%   \isacommand{lemma}\isamarkupfalse%
+% \ infinty{\isacharunderscore}in{\isacharunderscore}MG\ {\isacharcolon}\ {\isachardoublequoteopen}infinity{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharbrackleft}G{\isacharbrackright}{\isacharparenright}{\isachardoublequoteclose}
+% \end{isabelle}
 
-\fbox{Puede ser, pero eso est\'a dentro de  un locale}
-
-\fbox{que asume más cosas (una instancia de Repl)}
 %%% Local Variables: 

### 2018-12-21 LICS/recursion.tex
Explicación de Infinity. Nota para introducir check y valcheck.
Recursión y nombres

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
-\subsection*{Value of names}
+\subsection*{Recursion and Values of Names}
 
@@
 \end{isabelle}
-When computing the value of a recursive function on some argument $a$,
-one can restrict the relation to some ambient set $A$ if $a$ and
-all of its predecessesors are also in $A$.
+A key result, albeit intuitive and rather easy, to make this
+definition work is that when computing the value of a recursive
+function on some argument $a$, one can restrict the relation to some
+ambient set if it includes $a$ and all of its predecessesors.
 \begin{isabelle}
@@
 
-\medskip
-\noindent\fbox{Para que el resultado anterior tenga utilidad hay\ \dots\
-}
-\fbox{\dots\ que pedir que la relaci\'on \texttt{"i=>i=>o"} sea set-like}
-\medskip
-
-Using this lemma we can prove Equation~\eqref{eq:val}:
+This lemma gives rise to some useful results about values of
+names. First, we characterize $\val$ as in Equation~(\ref{eq:val}):
 \begin{isabelle}
@@
 \end{isabelle}
+More interestingly we can give a neat equation for values of
+names defined by Separation, say $B = \{x\in A\times \PP.\ Q(x)\}$,
+then
+\begin{equation}
+\val(G,B) = \{\val(G,t) : t\in A , \exists p\in \PP \cap G.\ Q(\lb t,p\rb) \} \label{eq:val-name-sep}
+\end{equation}
+
+\fbox{Creo que deberíamos poner acá \isa{check} y \isa{check\_in\_M} }
 

### 2018-12-21 LICS/isabelle.tex
comentarios sobre Isar

@@
 obtained by combining the given constants and operations. This will
-have concrete consequences in our strategy to approach the development.
+have concrete consequences in our strategy to approach the
+development.
+
+\medskip
+\fbox{Portar a \textbf{Isabelle2018} antes de enviar.}
+
+\fbox{Hacer benchmark para compararlos.}
+
+\medksip
+
+The source code is written for the 2018 version of Isabelle (with
+minor modifications, it can be run in Isabelle2016-1). Most of it is
+presented in the (nowadays standard) declarative flavour called
+\emph{Isar} \cite{DBLP:conf/tphol/Wenzel99}, where intermediate
+statements in the course 
+of a proof are explicitly stated, interspersed with automatic 
+tactics handling more trivial steps. The goal is that the resulting
+text, a \emph{proof document}, can be understood without the need of
+running it.
 

### 2018-12-21 LICS/generic-extensions.tex

@@
 $G\sbq\PP$, a new set $M[G]$ is defined. Each element $a\in M[G]$ is
-determined by an element $\dot a$ of $M$. Actually, the structure of
+determined by its \emph{name} $\dot a$ of $M$. Actually, the structure of
 each $\dot a$ is used to construct $a$. They are related by a
@@
 Metatheoretically, it is straightforward to see that $M[G]$ is a
-transitive set that satisfies the some of axioms of $\ZF$ (see
-Section~\ref{sec:easy-axioms}) and 
-includes $M\cup\{G\}$. Nevertheless there are no a priori reasons for
-$M[G]$ to satisfy the rest of the $\ZF$ 
-axioms. The original insight by Cohen was to define the notion of
-\emph{genericity} for a filter $G\sbq\PP$ and to prove that whenever
-$G$ is generic, $M[G]$ will satisfy $\ZF$.
+transitive set that satisfies some axioms of $\ZF$ (see
+Section~\ref{sec:easy-axioms}) and includes $M\cup\{G\}$. Nevertheless
+there is no a priori reason for $M[G]$ to satisfy the rest of the
+$\ZF$ axioms. The original insight by Cohen was to define the notion
+of \emph{genericity} for a filter $G\sbq\PP$ and to prove that
+whenever $G$ is generic, $M[G]$ will satisfy $\ZF$. Remember that a
+filter is generic if it intersects all the dense sets in $M$; in
+\cite{2018arXiv180705174G} we formalized Rassiowa-Sikorski lemma which
+proves the existence of generic filters for ctms.
 

### 2018-12-21 LICS/proof-of-separation.tex
comentario para Miguel
más intro y estética
ediciones mínimas

@@
 this presentation. Nevertheless, we will refer to each specific block
-of the code by line numer for ease of reference.
+of code by line number for ease of reference.
 
@@
 
-A result from the theory \isatt{Names} allows to calculate the value
-of sets such as $m$ that frequently are used in forcing arguments: 
-$\val(G,\{x\in A\times \PP.\ Q(x)\})$ equals
-\[
-\{\val(G,t) : t\in A , \exists
-  p\in \PP .\ Q(\lb t,p\rb) \y p \in G\}.
-\]
-This is then used to show  (lines 221--274 of \isatt{Separation\_Axiom}) that 
+Since $m$ is a name defined using Separation, we may use
+(\ref{eq:val-name-sep}) to show (lines 221--274 of
+\isatt{Separation\_Axiom})
 \[
@@
 \ \ \isakeyword{assumes}\ \isanewline
-\ \ \ \ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\ \isakeyword{and}\ {\isachardoublequoteopen}arity{\isacharparenleft}{\isasymphi}{\isacharparenright}\ {\isacharequal}\ {\isadigit{1}}\ {\isasymor}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharequal}{\isadigit{2}}{\isachardoublequoteclose}\isanewline
+\ \ \ \ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\ \isakeyword{and}\isanewline
+\ \ \ \  {\isachardoublequoteopen}arity{\isacharparenleft}{\isasymphi}{\isacharparenright}\ {\isacharequal}\ {\isadigit{1}}\ {\isasymor}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharequal}{\isadigit{2}}{\isachardoublequoteclose}\isanewline
 \ \ \isakeyword{shows}\ \ \isanewline
-\ \ \ \ {\isachardoublequoteopen}{\isacharparenleft}{\isasymforall}a{\isasymin}{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharparenright}{\isachardot}\ separation{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}{\isasymlambda}x{\isachardot}\ sats{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}{\isasymphi}{\isacharcomma}{\isacharbrackleft}x{\isacharcomma}a{\isacharbrackright}{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\ \ \ \ {\isachardoublequoteopen}{\isasymforall}a{\isasymin}{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharparenright}{\isachardot}\isanewline 
+ \ \ \ \ \  separation{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}{\isasymlambda}x{\isachardot}sats{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}{\isasymphi}{\isacharcomma}{\isacharbrackleft}x{\isacharcomma}a{\isacharbrackright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
 \end{isabelle}

### 2018-12-21 LICS/recursion.tex
Hablando un poco de check

@@
 
-\fbox{Creo que deberíamos poner acá \isa{check} y \isa{check\_in\_M} }
+We close our discussion of names and their values by making explicit
+the names for elements in $M$, once more we refer to
+\cite{2018arXiv180705174G} for our formalization. The definition of
+$\chk(x)$ is a straightforward $\in$-recursion:
+\begin{equation*}
+  \label{eq:def-check}
+  \chk(x)\defi\{\lb\chk(y),\1\rb : y\in x\}
+\end{equation*}
+It can be easily proved by $\in$-induction that $\val(G,\chk(x))=x$.
+To conclude $M\subseteq M[G]$ one also needs to have $\chk(x) \in M$;
+this result requires the internalization of recursively defined
+functions. We indicate explicitly where we use this assumption.
 

### 2018-12-21 LICS/easy-axioms.tex
Mencionando dot G y explicitando la dificultad de tener check_in_M

@@
 results proved in the locale \isa{M{\isacharunderscore}trivial}; this
-proof is easy because we assume $\chk(x) \in M$ whenever $x\in
-M$. Since we have already proved that $M[G]$ is transitive,
-$\emptyset\in M[G]$ assuming $G$ being generic, and also that it
-satisfies Pairing and Union, we can instantiate
-\isa{M{\isacharunderscore}trivial}:
+proof is easy because we work in the context of the locale
+\isa{M{\isacharunderscore}extra{\isacharunderscore}assms} which states
+the assumption $\chk(x) \in M$ whenever $x\in M$. Since we have
+already proved that $M[G]$ is transitive, $\emptyset\in M[G]$ assuming
+$G$ being generic, and also that it satisfies Pairing and Union, we
+can instantiate \isa{M{\isacharunderscore}trivial}:
 \begin{isabelle}

### 2018-12-21 LICS/recursion.tex
Mencionando dot G y explicitando la dificultad de tener check_in_M

@@
 It can be easily proved by $\in$-induction that $\val(G,\chk(x))=x$.
-To conclude $M\subseteq M[G]$ one also needs to have $\chk(x) \in M$;
-this result requires the internalization of recursively defined
-functions. We indicate explicitly where we use this assumption.
+But to conclude $M\subseteq M[G]$ one also needs to have
+$\chk(x) \in M$; this result requires the internalization of
+recursively defined functions. This is also needd to prove
+$G \in M[G]$; let us define
+$\dot G= \{\lb \chk(p),p\rb : p \in \PP \}$, it is easy to prove
+$\val(G,\dot G) = G$. Proving $\dot G\in M$ involves
+knowing $\chk(x) \in M$ and using one instance of Replacement.
+
+Paulson proved absoluteness results for definitions by recursion and
+one of our next goals is to instantiate the appropiate locale
+\isa{M{\isacharunderscore}eclose} which is the last layer of a pile of
+locales; it will take us some time to prove that any ctm satisfies the
+assumptions involved in those locales. In the meantime, we declare a
+locale \isa{M{\isacharunderscore}extra{\isacharunderscore}assms}
+assembling both assumptions.
 

### 2018-12-21 LICS/intro.tex

@@
 
-An enumeration of achievements to be presented:
-\begin{enumerate}
-\item Discussion of locales.
-  \begin{enumerate}
-  \item How they help presenting complex results.
-  \item Their key role in stating the Fundamental Theorems (without
-    defining \isatt{forces}).
-  \item \isatt{forces} is defined for elements of \formula.
-  \end{enumerate}
-\item Enhancements in the recursion machinery.
-  \begin{enumerate}
-  \item Impact on names.
-  \end{enumerate}
-\item Inteface between first-order properties of $M$ and those of
-  $\#\#M$.
-\end{enumerate}
-
 %%% Local Variables: 

### 2018-12-21 LICS/recursion.tex
Fixed typo.
Hablando de interfaz (por hacer)

@@
 $\chk(x) \in M$; this result requires the internalization of
-recursively defined functions. This is also needd to prove
+recursively defined functions. This is also needed to prove
 $G \in M[G]$; let us define
 $\dot G= \{\lb \chk(p),p\rb : p \in \PP \}$, it is easy to prove
-$\val(G,\dot G) = G$. Proving $\dot G\in M$ involves
-knowing $\chk(x) \in M$ and using one instance of Replacement.
+$\val(G,\dot G) = G$. Proving $\dot G\in M$ involves knowing
+$\chk(x) \in M$ and using one instance of Replacement.
 
@@
 \isa{M{\isacharunderscore}eclose} which is the last layer of a pile of
-locales; it will take us some time to prove that any ctm satisfies the
-assumptions involved in those locales. In the meantime, we declare a
-locale \isa{M{\isacharunderscore}extra{\isacharunderscore}assms}
-assembling both assumptions.
+locales. It will take us some time to prove that any ctm satisfies the
+assumptions involved in those locales; as we mentioned, Paulson work
+is mostly done at the meta-level, i.e. the assumptions are instances
+of Separation and Replacement where the predicates and functions are
+Isabelle functions of type \isa{i{\isasymRightarrow}i} and
+\isa{i{\isasymRightarrow}o}, respectively. In contrast, we assume that
+$M$ is a model of $ZF$, therefore to deduce that $M$ satisfies a
+Separation instance, we have to define an internalized formula whose
+satisfaction is equivalent to the meta-level predicate (cf. the
+interface described in Section~\ref{sec:axioms-models-set-theory} and
+also the concrete example given in the proof of Union).
+
+In the meantime, we declare a locale
+\isa{M{\isacharunderscore}extra{\isacharunderscore}assms} assembling
+both assumptions ($M$ closed under $\chk$ and the instance of
+Replacement); in this paper we explicitly mention where we use them.
 

### 2018-12-22 LICS/conclusion.tex
ayuda memoria en conclusión
mainly prueba de separación, labels por sección
cosas minimas y cometario en conlusion

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
-\section{Conclusion}
+\section{Conclusions and future work}
+\label{sec:conclusions-future-work}
 
+The proof of the Powerset Axiom does not present difficulties.
+
+The proof of the Replacement Axiom, both in Kunen and in
+other presentations \cite{neeman-course} require a relativized proof
+of the Reflection Principle.
+
+\medskip
+\fbox{Decir que Separaci\'on con 6 variables en $M$}
+
+\fbox{implica Separaci\'on con a lo sumo 2 en $M[G]$}
 \section*{Acknowledgment}

### 2018-12-22 LICS/recursion.tex
cosas minimas y cometario en conlusion

@@
 We close our discussion of names and their values by making explicit
-the names for elements in $M$, once more we refer to
+the names for elements in $M$; once more, we refer to
 \cite{2018arXiv180705174G} for our formalization. The definition of
@@
 Paulson proved absoluteness results for definitions by recursion and
-one of our next goals is to instantiate the appropiate locale
+one of our next goals is to instantiate at $\#\#M$ the appropriate
+locale 
 \isa{M{\isacharunderscore}eclose} which is the last layer of a pile of
-locales. It will take us some time to prove that any ctm satisfies the
-assumptions involved in those locales; as we mentioned, Paulson work
+locales. It will take us some time to prove that any ctm of $\ZF$ 
+satisfies the
+assumptions involved in those locales; as we mentioned, Paulson's work
 is mostly done at the meta-level, i.e. the assumptions are instances

### 2018-12-24 LICS/recursion.tex
meta-level cambiado por external

@@
 assumptions involved in those locales; as we mentioned, Paulson's work
-is mostly done at the meta-level, i.e. the assumptions are instances
+is mostly done externally, i.e. the assumptions are instances
 of Separation and Replacement where the predicates and functions are
@@
 Separation instance, we have to define an internalized formula whose
-satisfaction is equivalent to the meta-level predicate (cf. the
+satisfaction is equivalent to the external predicate (cf. the
 interface described in Section~\ref{sec:axioms-models-set-theory} and

### 2018-12-26 LICS/hacking.tex
jingle bells, jingle bells

@@
 
-\medskip
-\fbox{>>>Poner lo que sigue en las conclusiones???}
-\medskip
-
-We believe that the
-\isatt{ZF-Constructible} session  would benefit from some changes in
-the way some results are organized accross theory files; for
-intance, by cataloging all or most of the internalized formulas in one
-file. Another, most basic example would be to start out with an even
-more locale that only assumes $M$ to be a nonempty transitive class,
-as many absoluteness results follow from these hypotheses.  
-It is our desire to advocate a future work to a thorough
-revision of the development of constructibility to maximize modularity.
-
 %%% Local Variables: 

### 2018-12-26 LICS/recursion.tex
jingle bells, jingle bells

@@
 \isa{i{\isasymRightarrow}o}, respectively. In contrast, we assume that
-$M$ is a model of $ZF$, therefore to deduce that $M$ satisfies a
+$M$ is a model of $\ZF$, therefore to deduce that $M$ satisfies a
 Separation instance, we have to define an internalized formula whose

### 2018-12-26 LICS/zfc-axioms.tex
jingle bells, jingle bells

@@
 we simply refer to $M$ as a \emph{countable transitive model} or
-\emph{ctm}.
-
-The existence of a ctm of $\ZF$ can be proved from the existence of a
-model  $\lb N,E\rb$ such that the relation $E$ is well founded. The
-L\"owenheim-Skolem 
+\emph{ctm}. The following result shows how to obtain ctms from weaker
+hypotheses. 
+%
+\begin{lemma}\label{lem:wf-model-implies-ctm}
+  If there exists  a
+  model  $\lb N,E\rb$  of $\ZF$ such that the relation $E$ is well
+  founded, then there exists a countable transitive one.
+\end{lemma}
+\begin{proof}
+  (Sketch) The L\"owenheim-Skolem 
   Theorem ensures that there is an countable elementary submodel 
@@
   isomorphically to some $\lb M,\in\rb$ with  $M$ transitive.
-
-By G\"odel's Second Incompleteness Theorem we cannot  prove that
-there exists a model of $\ZF$  (assuming that our base theory is not
-stronger than $\ZF$ to start with), but for consistency proofs,
-usually a model of only a finite subset of $\ZF$ suffices. Recall that
-an equivalent formulation of the Foundation Axiom states that the 
-universe of sets can be decomposed in a transfinite hierarchy of
-sets. 
-\begin{theorem}
-  Let $V_{\al}\defi\union\{\P(V_\be) : \be<\al\}$ for each ordinal
-  $\al$. Then each $V_\al$ is a set and 
-  $\forall x. \exists\al .\ \Ord(\al) \y x\in V_\al$.  
-\end{theorem}
-%
-Then the following meta-theoretic result by Montague applies:
-%
-\begin{theorem}[Reflection Principle]\label{th:reflection-principle}
-  For every finite $\Phi\sbq\ZF$, $\ZF$ proves: ``There exists
-    unboundedly many $\al$ such that $V_\al\models \Phi$.''
-\end{theorem}
-%
-Since the sets $V_\al$ are well founded, we can repeat the above
-argument to obtain a ctm of $\Phi$.
+\end{proof}
 

### 2018-12-26 LICS/conclusion.tex
jingle bells, jingle bells

@@
 The proof of the Powerset Axiom does not present difficulties.
+
+By G\"odel's Second Incompleteness Theorem we cannot  prove that
+there exists a model of $\ZF$  (assuming that our base theory is not
+stronger than $\ZF$ to start with), but for consistency proofs,
+usually a model of only a finite subset of $\ZF$ suffices. Recall that
+an equivalent formulation of the Foundation Axiom states that the 
+universe of sets can be decomposed in a transfinite hierarchy of
+sets. 
+\begin{theorem}
+  Let $V_{\al}\defi\union\{\P(V_\be) : \be<\al\}$ for each ordinal
+  $\al$. Then each $V_\al$ is a set and 
+  $\forall x. \exists\al .\ \Ord(\al) \y x\in V_\al$.  
+\end{theorem}
+%
+Then the following meta-theoretic result by Montague applies:
+%
+\begin{theorem}[Reflection Principle]\label{th:reflection-principle}
+  For every finite $\Phi\sbq\ZF$, $\ZF$ proves: ``There exists
+    unboundedly many $\al$ such that $V_\al\models \Phi$.''
+\end{theorem}
+%
+Since the sets $V_\al$ are well founded, we can repeat the
+proof of Lemma~\ref{lem:wf-model-implies-ctm} to obtain a ctm of
+$\Phi$.
 
@@
 \fbox{implica Separaci\'on con a lo sumo 2 en $M[G]$}
+\medskip
+
+We believe that the
+\isatt{ZF-Constructible} session  would benefit from some changes in
+the way some results are organized accross theory files; for
+intance, by cataloging all or most of the internalized formulas in one
+file. Another, most basic example would be to start out with an even
+more locale that only assumes $M$ to be a nonempty transitive class,
+as many absoluteness results follow from these hypotheses.  
+It is our desire to advocate a future work to a thorough
+revision of the development of constructibility to maximize modularity.
+
 \section*{Acknowledgment}

### 2018-12-28 LICS/intro_m.tex
Introducción alternativa.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
\section{Introduction}
% no \IEEEPARstart
Zermelo-Fraenkel Set Theory ($\ZF$) has a prominent place among formal
theories; in particular, it provides a foundation for mathematics and
most of the toolkit used everyday by the computer scientist has also
Set Theory at its base (although much of it can be based in other
formalisms, such as Category Theory) \fbox{citation needed,
  Aczel?}. In this time of mechanization of
mathematics~\cite{avigad2018mechanization}, it seems natural to ask
for a mechanization of the most salient results of Set Theory.
\fbox{Y entra de lleno en la intersección de lógica  y cs}

After G\"odel's Incompleteness Theorems, we cannot expect to have a
formal proof of the consistency of Set Theory in $ZF$. Besides its own
consistency, there are other results which are undecided by $ZF$: the
undecidability of Continuum Hypothesis lead to the development of
techniques for independence proofs. First G\"odel introduced the
theory of \emph{inner models}, which gives rise to his model $L$ of
the \emph{Axiom of Constructibility} \cite{godel-L} and proved the
relative consistency of the Axiom of Choice and the Generalized
Continuum Hypothesis with $\ZF$. Thirty years later Paul
J. Cohen~\cite{Cohen-CH-PNAS} devised the technique of \emph{forcing},
which is the only known way of \emph{extending} models of $\ZF$; in
particular, it can be used to prove the relative consistency of the
negation of the Continuum Hypothesis.

In this work we address a substantial part of formalizing the proof
that given a model $M$ of $\ZF$, any \emph{generic extension} $M[G]$
given by forcing also satisfies $\ZF$. As remarked by
\citet[][p.250]{kunen2011set} \enquote{[...] in verifying that $M[G]$
  is a model for set theory, the hardest axiom to verify is
  Comprehension.}  The most important achievement of this paper is the
mechanization in Isabelle of this result; en route to this, we also
formalized the satisfaction by $M[G]$ of Extensionality, Foundation,
Union, Infinity.

Our development benefited from the extraordinary work done by Lawrence
Paulson \cite{paulson_2003} on the formalization of G\"odel's
constructible universe in the proof assistant \emph{Isabelle}. In the
spirit of the \emph{formal abstract} project~\cite{hales-fabstracts},
we explicit the technical assumptions of forcing that are involved in
the proof of Separation for $M[G]$. It seems that this work is
\emph{Es un lindo test a la capacidad de implementación formal de
  matemática con la tecnología actual}.

\fbox{Acá viene una oración introduciendo lo que sigue.}

In a gross simplification, there are two aspects to a formalization
project like this one: thematic and programmatic. The first concerns
the handling of all the theoretical concepts and results in the
subject, while the second involves the practical issues of the
implementation and design. In the case of forcing, the main intricacy
lies in the first aspect. In this sense, following a sensible
presentation of the material is key.  The authoritative reference 
on the subject during the last 30 years has been Kunen's classical
\cite{kunen1980}. In our
formalizaton we have followed a recent rewrite \cite{kunen2011set}
of that  textbook, which presents the material in the same sharp 
style but offering a lot of details. In some sense this project
wouldn't exist without this book. As alternative, introductory
resources, the  interested reader can check
\cite{chow-beginner-forcing}; also, the book \cite{weaver2014forcing}
contains a thorough treament minimizing the technicalities.

\fbox{El párrafo que sigue lo sacaría}

In this work we address a substantial part the problem of formalizing
the proof that given a model of $\ZF$, its extensions by forcing, 
\emph{generic extensions},
satisfy the same axioms. In particular, the first axiom that
genuinely calls for a use of the forcing machinery is that of
\emph{Separation} (\emph{Aussonderungsaxiom}), otherwise known as
``Axiom of (Restricted) Comprehension''  or ``Specification.'' In
\cite{2018arXiv180705174G} it was shown that  generic extensions of
models of $\ZF$ 
satisfy the Axiom of Pairing, and in this work we show that also
Extensionality, Foundation, Union, Infinity (under extra assumptions),
and Separation go through. The
Axiom of Powerset is not treated in the present paper but it poses no
extra difficulties; we discuss the Axiom of Replacement in the
conclusions.

The proof that these axioms hold in generic extension is independent
of the \emph{proofs} of the ``fundamental'' theorems of forcing. In
particular, one of the most complex parts of those proofs is the
definition of the \emph{forcing relation}. It turns out that the
particular (``implementation'') details of this  definition  does not
appear in the statement of the fundamental theorems, and our
formalization works inside a \emph{locale} (set of assumptions) that
includes the statement of these theorems. The formalization of the
fundamental theorems of forcing roughly comprises little less than a
half of our full project.

We briefly describe the contents of each
section. Section~\ref{sec:isabelle} contains the bare minimun
requirements to understand the (meta)logics used in Isabelle. Next, an
overview of the model theory of set theory is presented in
Section~\ref{sec:axioms-models-set-theory}. There is an ``internal''
representation of first-order formulas as sets, implemented by
Paulson; Section~\ref{sec:renaming} discusses syntactical
transformations of the former, mainly permutation of variables. 
In Section~\ref{sec:generic-extensions} the generic extensions are
succintly reviewed and how the treatment of well founded recursion in
Isabelle was enhanced. We take care of the ``easy axioms'' in
Section~\ref{sec:easy-axioms}; these are the ones that
do not depend on the forcing theorems. We describe the latter in
Section~\ref{sec:forcing}. We adapted the  work by Paulson to our
needs, and this is described in
Section~\ref{sec:hack-constructible}. We present the proof
of the Separation Axiom Scheme in Section~\ref{sec:proof-separation},
which follows closely its implementation. A plan for future work and
some immediate conclusions are offered in
Section~\ref{sec:conclusions-future-work}.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "Separation_In_MG"
%%% ispell-local-dictionary: "american"
%%% End:

### 2018-12-28 LICS/conclusion.tex
empezando a escribir la conclu

@@
 
-The proof of the Powerset Axiom does not present difficulties.
+The proof that the axioms of $\ZF$ hold in generic extensions is 
+the first important landmark in the path of formalization of the
+theory of forcing. The task of proving the Fundamental Theorems can be
+considered a part of that landmark, but from a workflow point of view
+this can be done independently, and one of our design decisions was to
+postpone that task.  Next, various forcing notions must be developed
+in order to obtain extensions satisfying diverse properties. This
+yields relative consistency results.
 
-By G\"odel's Second Incompleteness Theorem we cannot  prove that
-there exists a model of $\ZF$  (assuming that our base theory is not
-stronger than $\ZF$ to start with), but for consistency proofs,
-usually a model of only a finite subset of $\ZF$ suffices. Recall that
+We first discuss the axioms we are now missing. The proof of the
+Powerset Axiom does not present difficulties, and it is one of the
+first goals in the immediate future. This is also the case with the
+Axiom of Infinity; it can be proved in a direct but cumbersome way. We
+preferred to wait to have a full-fledged inteface between ctms and the
+locales providing recursive constructions from Paulson's
+\isatt{Constructibility} session. Then the current proof will hold
+with no extra assumptions. The same goes for the results $M\sbq M[G]$
+and $G\in M[G]$.
+
+The Replacement Axiom, however, requires some more work to be
+done. Both in Kunen and in other presentations \cite{neeman-course}
+it requires a relativized version (i.e., showing that it holds for $M$) of
+the \emph{Reflection Principle}. In order to state this meta-theoretic
+result by Montague, recall that
 an equivalent formulation of the Foundation Axiom states that the 
@@
 \end{theorem}
-%
-Then the following meta-theoretic result by Montague applies:
-%
 \begin{theorem}[Reflection Principle]\label{th:reflection-principle}
@@
 \end{theorem}
-%
-Since the sets $V_\al$ are well founded, we can repeat the
-proof of Lemma~\ref{lem:wf-model-implies-ctm} to obtain a ctm of
-$\Phi$.
 
-The proof of the Replacement Axiom, both in Kunen and in
-other presentations \cite{neeman-course} require a relativized proof
-of the Reflection Principle.
+This result has also other applications; in particular, it dispenses
+the need of ctms of the whole of $\ZF$ for consistency proofs (by
+repeating  the proof of Lemma~\ref{lem:wf-model-implies-ctm}, now
+starting with a  $V_\al$ satisfying $\Phi$).
 

### 2018-12-28 LICS/intro_m.tex
Intro de Miguel

@@
 consistency, there are other results which are undecided by $ZF$: the
-undecidability of Continuum Hypothesis lead to the development of
+undecidability of Continuum Hypothesis leads to the development of
 techniques for independence proofs. First G\"odel introduced the
@@
   Comprehension.}  The most important achievement of this paper is the
-mechanization in Isabelle of this result; en route to this, we also
-formalized the satisfaction by $M[G]$ of Extensionality, Foundation,
-Union, Infinity.
+mechanization in Isabelle of a substantial part of this result; en
+route to this, we also formalized the satisfaction by $M[G]$ of
+Extensionality, Foundation, Union, and Infinity.
 
@@
 Paulson \cite{paulson_2003} on the formalization of G\"odel's
-constructible universe in the proof assistant \emph{Isabelle}. In the
-spirit of the \emph{formal abstract} project~\cite{hales-fabstracts},
-we explicit the technical assumptions of forcing that are involved in
-the proof of Separation for $M[G]$. It seems that this work is
-\emph{Es un lindo test a la capacidad de implementación formal de
-  matemática con la tecnología actual}.
+constructible universe in the proof assistant \emph{Isabelle}. The
+ultimate goal of our project is the formalization of forcing to
+complete the mechanization of the independence of the Continuum
+Hypothesis. We think that this project constitutes an interesting case
+which stresses how feasible is to formally implement mathematics that
+involve several leves of reasoning.
 
-\fbox{Acá viene una oración introduciendo lo que sigue.}
+The \emph{Formal Abstract} project~\cite{hales-fabstracts} proposes
+that the mechanization of mathematics can advance by formalizing some
+main result under the explicit assumption of the material
+(definitions, propositions, lemmas) on which it is based. In line with
+this advice, we profited that the proof that those axioms hold in
+generic extension is independent of the \emph{proofs} of the
+``fundamental'' theorems of forcing. Let us remark that the definition
+of the forces relation is, by itself, quite demanding; the
+formalization of it and of the fundamental theorems of forcing roughly
+comprises little less than a half of our full project.
+
+\fbox{Parece mucho comienzo sólo para introducir a Kunen}
+\fbox{¿lo puedo achurar un poco?}
 
@@
 contains a thorough treament minimizing the technicalities.
-
-\fbox{El párrafo que sigue lo sacaría}
-
-In this work we address a substantial part the problem of formalizing
-the proof that given a model of $\ZF$, its extensions by forcing, 
-\emph{generic extensions},
-satisfy the same axioms. In particular, the first axiom that
-genuinely calls for a use of the forcing machinery is that of
-\emph{Separation} (\emph{Aussonderungsaxiom}), otherwise known as
-``Axiom of (Restricted) Comprehension''  or ``Specification.'' In
-\cite{2018arXiv180705174G} it was shown that  generic extensions of
-models of $\ZF$ 
-satisfy the Axiom of Pairing, and in this work we show that also
-Extensionality, Foundation, Union, Infinity (under extra assumptions),
-and Separation go through. The
-Axiom of Powerset is not treated in the present paper but it poses no
-extra difficulties; we discuss the Axiom of Replacement in the
-conclusions.
-
-The proof that these axioms hold in generic extension is independent
-of the \emph{proofs} of the ``fundamental'' theorems of forcing. In
-particular, one of the most complex parts of those proofs is the
-definition of the \emph{forcing relation}. It turns out that the
-particular (``implementation'') details of this  definition  does not
-appear in the statement of the fundamental theorems, and our
-formalization works inside a \emph{locale} (set of assumptions) that
-includes the statement of these theorems. The formalization of the
-fundamental theorems of forcing roughly comprises little less than a
-half of our full project.
 

### 2018-12-28 LICS/intro_m.tex
trabajos relacionados

@@
 theories; in particular, it provides a foundation for mathematics and
-most of the toolkit used everyday by the computer scientist has also
-Set Theory at its base (although much of it can be based in other
-formalisms, such as Category Theory) \fbox{citation needed,
-  Aczel?}. In this time of mechanization of
-mathematics~\cite{avigad2018mechanization}, it seems natural to ask
-for a mechanization of the most salient results of Set Theory.
-\fbox{Y entra de lleno en la intersección de lógica  y cs}
+most of the formal toolkit used everyday by the computer scientist has
+also Set Theory at its base (cf.~\cite{paulson1995set}). In this time
+of mechanization of mathematics~\cite{avigad2018mechanization}, it
+seems natural to ask for a mechanization of the most salient results
+of Set Theory.
 
@@
 comprises little less than a half of our full project.
+
+It might be a little surprising the lack of formalizations of forcing
+and generic extensions. As far as we know, the development of
+\citet{JFR6232} in homotopy type theory for constructing generic
+extensions in a sheaf-theoretic setting is the unique mechanization of
+forcing. This contrast with the fruitful use of forcing techniques to
+extend the Curry-Howard isomorphism to classical axioms
+\cite{Miquel:2011:FPT:2058525.2059614,lmcs:1070}. Moreover, the
+combination of forcing with intuitionistic type theory
+\cite{coquand2010note,coquand2012computational} gives rise to an
+algorithm to obtain witnesses of the continuity of definable
+functionals.
+
+% moreover forcing translations \cite{jaber:hal-01319066}.
 

### 2018-12-28 LICS/intro_m.tex
cosas menores

@@
 After G\"odel's Incompleteness Theorems, we cannot expect to have a
-formal proof of the consistency of Set Theory in $ZF$. Besides its own
-consistency, there are other results which are undecided by $ZF$: the
+formal proof of the consistency of Set Theory in $\ZF$. Besides its own
+consistency, there are other results which are undecided by $\ZF$: the
 undecidability of Continuum Hypothesis leads to the development of

### 2018-12-28 LICS/intro_m.tex
cosas menores y nuevas referencias

@@
   Comprehension.}  The most important achievement of this paper is the
-mechanization in Isabelle of a substantial part of this result; en
-route to this, we also formalized the satisfaction by $M[G]$ of
+mechanization in Isabelle a considerable part of this result; en route
+to this, we also formalized the satisfaction by $M[G]$ of
 Extensionality, Foundation, Union, and Infinity.
 
-Our development benefited from the extraordinary work done by Lawrence
+Our development benefited from the remarkable work done by Lawrence
 Paulson \cite{paulson_2003} on the formalization of G\"odel's
@@
 which stresses how feasible is to formally implement mathematics that
-involve several leves of reasoning.
+involve several levels of reasoning.
 
@@
 combination of forcing with intuitionistic type theory
-\cite{coquand2010note,coquand2012computational} gives rise to an
+\cite{coquand2010note} gives rise both to positive results (an
 algorithm to obtain witnesses of the continuity of definable
-functionals.
+functionals \cite{coquand2012computational}) and also negative
+(the independence of Markov's principle \cite{lmcs:3859}).
 

### 2018-12-28 LICS/intro_m.tex
Más trabajo relacionado. Menos pegados a fabstracts.

@@
 The \emph{Formal Abstract} project~\cite{hales-fabstracts} proposes
-that the mechanization of mathematics can advance by formalizing some
-main result under the explicit assumption of the material
-(definitions, propositions, lemmas) on which it is based. In line with
-this advice, we profited that the proof that those axioms hold in
-generic extension is independent of the \emph{proofs} of the
-``fundamental'' theorems of forcing. Let us remark that the definition
+that the formalization of mathematics by writing the statements of
+results and the material upon which they are based (definitions,
+propositions, lemmas), but ommiting the proofs. We took a less drastic
+position: since the proof that those axioms hold in generic extension
+is independent of the \emph{proofs} of the ``fundamental'' theorems of
+forcing, we assumed these results. Let us remark that the definition
 of the forces relation is, by itself, quite demanding; the
@@
 combination of forcing with intuitionistic type theory
-\cite{coquand2010note} gives rise both to positive results (an
-algorithm to obtain witnesses of the continuity of definable
-functionals \cite{coquand2012computational}) and also negative
-(the independence of Markov's principle \cite{lmcs:3859}).
-
-% moreover forcing translations \cite{jaber:hal-01319066}.
+\cite{Coquand:2009:FTT:1807662.1807665,coquand2010note} gives rise
+both to positive results (an algorithm to obtain witnesses of the
+continuity of definable functionals \cite{coquand2012computational})
+and also negative (the independence of Markov's principle
+\cite{lmcs:3859}). In the same strand of forcing from the point of
+view of proof theory \cite{avigad_2004} are the conservative
+extensions of CoC with forcing conditions
+\cite{jaber:hal-01319066,Jaber:2012:ETT:2358958.2359524}.
 

### 2018-12-28 LICS/intro_m.tex
Más cambios pavos en la intro

@@
 to this, we also formalized the satisfaction by $M[G]$ of
-Extensionality, Foundation, Union, and Infinity.
+Extensionality, Foundation, Union, and Infinity. % We have already
+% proved Pairing in the first report of our project
+% \cite{2018arXiv180705174G}.
 
@@
 
-\fbox{Parece mucho comienzo sólo para introducir a Kunen}
-\fbox{¿lo puedo achurar un poco?}
+% \fbox{Parece mucho comienzo sólo para introducir a Kunen}
+% \fbox{¿lo puedo achurar un poco?}
 
-In a gross simplification, there are two aspects to a formalization
-project like this one: thematic and programmatic. The first concerns
-the handling of all the theoretical concepts and results in the
-subject, while the second involves the practical issues of the
-implementation and design. In the case of forcing, the main intricacy
-lies in the first aspect. In this sense, following a sensible
-presentation of the material is key.  The authoritative reference 
-on the subject during the last 30 years has been Kunen's classical
-\cite{kunen1980}. In our
-formalizaton we have followed a recent rewrite \cite{kunen2011set}
-of that  textbook, which presents the material in the same sharp 
-style but offering a lot of details. In some sense this project
-wouldn't exist without this book. As alternative, introductory
-resources, the  interested reader can check
-\cite{chow-beginner-forcing}; also, the book \cite{weaver2014forcing}
-contains a thorough treament minimizing the technicalities.
+% In a gross simplification, there are two aspects to a formalization
+% project like this one: thematic and programmatic. The first concerns
+% the handling of all the theoretical concepts and results in the
+% subject, while the second involves the practical issues of the
+% implementation and design. In the case of forcing, the main intricacy
+% lies in the first aspect. In this sense, following a sensible
+% presentation of the material is key.  The authoritative reference 
+% on the subject during the last 30 years has been Kunen's classical
+% \cite{kunen1980}. In our
+% formalizaton we have followed a recent rewrite \cite{kunen2011set}
+% of that  textbook, which presents the material in the same sharp 
+% style but offering a lot of details. In some sense this project
+% wouldn't exist without this book. As alternative, introductory
+% resources, the  interested reader can check
+% \cite{chow-beginner-forcing}; also, the book \cite{weaver2014forcing}
+% contains a thorough treament minimizing the technicalities.
+
+In pursuing the proof of preservation on generic extensions we
+extended Paulson's library with:
+\begin{inlinelist}
+\item renaming of variables, which with little effort can be extended
+  to substitutions;
+\item an improvement on definitions by recursion on well-founded
+  relations; 
+\item a better hierarchy of locales; and
+\item a new choice principle and a version of Rassiowa-Sikorski which
+  ensure the existence of generic filters for countable and transitive
+  models of $\ZF$; these points were already communicated in the
+  first report \cite{2018arXiv180705174G}.
+\end{inlinelist}
 

### 2018-12-29 LICS/conclusion.tex
parrafos sueltos

@@
 We first discuss the axioms we are now missing. The proof of the
-Powerset Axiom does not present difficulties, and it is one of the
+Powerset Axiom does not present new difficulties but  has to
+take care of absoluteness issues; it is one of the
 first goals in the immediate future. This is also the case with the
@@
 \end{theorem}
+%% This result has also other applications; in particular, it dispenses
+%% the need of ctms of the whole of $\ZF$ for consistency proofs (by
+%% repeating  the proof of Lemma~\ref{lem:wf-model-implies-ctm}, now
+%% starting with a  $V_\al$ satisfying $\Phi$).
 
-This result has also other applications; in particular, it dispenses
-the need of ctms of the whole of $\ZF$ for consistency proofs (by
-repeating  the proof of Lemma~\ref{lem:wf-model-implies-ctm}, now
-starting with a  $V_\al$ satisfying $\Phi$).
+This statement is straightforwardly equivalent to that with $\Phi$
+consisting of a single formula. The schematic nature of this result
+hints at a proof by induction on formulas, and hence it must be shown
+internally.
 
-\medskip
-\fbox{Decir que Separaci\'on con 6 variables en $M$}
+This  is an appropriate spot to insist that the internal/external
+dichotomy has been a powerful agent in the shaping of our
+project. It traces back to the definition of the formula-transformer
+$\forceisa$, which must be done by recursion, and this requirement
+spills indirectly to the proof of the Separation Axiom (despite this
+is not by induction). Now the need of using internalized formulas
+reappears with the need of having a general version of the Reflection
+Principle (because the $\Phi$ there depends on the instance of Replacement
+being proved).
 
-\fbox{implica Separaci\'on con a lo sumo 2 en $M[G]$}
-\medskip
+A secondary goal of this project is to assess which assumptions on the
+original ctm $M$ are needed to develop the forcing machinery. Up to this
+point we have some anecdotal data; for example, to show that a
+2-variable instance of Separation holds in $M[G]$, one needs to use a
+6-variable instance in $M$.
 

### 2018-12-29 LICS/intro_m.tex
una pasada a la intro_m

@@
 most of the formal toolkit used everyday by the computer scientist has
-also Set Theory at its base (cf.~\cite{paulson1995set}). In this time
+also some form of set theory at its base
+(cf.~\cite{paulson1995set}). Considering the current trend 
 of mechanization of mathematics~\cite{avigad2018mechanization}, it
@@
 
+The results we are interested in originally arose in connection to
+(relative) consistency proofs.
 After G\"odel's Incompleteness Theorems, we cannot expect to have a
 formal proof of the consistency of Set Theory in $\ZF$. Besides its own
-consistency, there are other results which are undecided by $\ZF$: the
-undecidability of Continuum Hypothesis leads to the development of
-techniques for independence proofs. First G\"odel introduced the
-theory of \emph{inner models}, which gives rise to his model $L$ of
+consistency, there are other statements which are undecided by $\ZF$.
+Perhaps the most prominent  is the
+\emph{Continuum Hypothesis}, which lead to the development of powerful
+techniques for independence proofs. First, G\"odel inaugurated the
+theory of \emph{inner models} by introducing his model $L$ of
 the \emph{Axiom of Constructibility} \cite{godel-L} and proved the
 relative consistency of the Axiom of Choice and the Generalized
-Continuum Hypothesis with $\ZF$. Thirty years later Paul
-J. Cohen~\cite{Cohen-CH-PNAS} devised the technique of \emph{forcing},
-which is the only known way of \emph{extending} models of $\ZF$; in
-particular, it can be used to prove the relative consistency of the
+Continuum Hypothesis with $\ZF$. Thirty years later, Paul
+J.~Cohen~\cite{Cohen-CH-PNAS} devised the technique of \emph{forcing},
+which is the only known way of \emph{extending} models of $\ZF$; this
+was used to prove the relative consistency of the 
 negation of the Continuum Hypothesis.
@@
   is a model for set theory, the hardest axiom to verify is
-  Comprehension.}  The most important achievement of this paper is the
-mechanization in Isabelle a considerable part of this result; en route
+  [Separation].}
+
+\medskip
+\fbox{(sigue p\'arrafo) Versi\'on de Pedro}
+\medskip
+
+The most important achievement of this paper is the
+mechanization in the proof assistant \emph{Isabelle} of a proof of the
+Axiom of Separation in 
+generic extensions by using the ``fundamental'' theorems of forcing.
+En route
 to this, we also formalized the satisfaction by $M[G]$ of
@@
 
+\medskip
+\fbox{Versi\'on de Miguel}
+\medskip
+
+The most important achievement of this paper is the
+mechanization in Isabelle of a considerable part of this result; en route
+to this, we also formalized the satisfaction by $M[G]$ of
+Extensionality, Foundation, Union, and Infinity. % We have already
+% proved Pairing in the first report of our project
+% \cite{2018arXiv180705174G}.
+
+\medskip
+\fbox{Fin p\'arrafo}
+\medskip
+
 Our development benefited from the remarkable work done by Lawrence
 Paulson \cite{paulson_2003} on the formalization of G\"odel's
-constructible universe in the proof assistant \emph{Isabelle}. The
-ultimate goal of our project is the formalization of forcing to
-complete the mechanization of the independence of the Continuum
+constructible universe in Isabelle. The
+ultimate goal of our project is the formalization of the forcing
+techniques needed to show the independence of the Continuum
 Hypothesis. We think that this project constitutes an interesting case
 which stresses how feasible is to formally implement mathematics that
-involve several levels of reasoning.
+involve several levels of reasoning. \fbox{No entiendo la ultima oraci\'on}
 
-The \emph{Formal Abstract} project~\cite{hales-fabstracts} proposes
-that the formalization of mathematics by writing the statements of
+The \emph{Formal Abstracts} project~\cite{hales-fabstracts} proposes
+the formalization of complex pieces of mathematics by writing the
+statements of 
 results and the material upon which they are based (definitions,
-propositions, lemmas), but ommiting the proofs. We took a less drastic
-position: since the proof that those axioms hold in generic extension
-is independent of the \emph{proofs} of the ``fundamental'' theorems of
-forcing, we assumed these results. Let us remark that the definition
-of the forces relation is, by itself, quite demanding; the
+propositions, lemmas), but ommiting the proofs. In this work we
+partially adhere to this vision to delineate our formalization
+strategy:
+since the proofs that the  axioms hold in generic extensions
+is independent of the \emph{proofs} of the fundamental theorems of
+forcing, we assumed these results for the time being. Let us remark
+that those theorems depend on the definition of a function $\forceisa$
+from formulas to formulas which is, by itself, quite demanding; the
 formalization of it and of the fundamental theorems of forcing roughly
-comprises little less than a half of our full project.
+comprises barely less than a half of our full project.
 
@@
   relations; 
-\item a better hierarchy of locales; and
-\item a new choice principle and a version of Rassiowa-Sikorski which
-  ensure the existence of generic filters for countable and transitive
-  models of $\ZF$; these points were already communicated in the
+\item enhancements in the hierarchy of locales; and
+\item a variant of the  principle of dependent choices and a version
+  of Rassiowa-Sikorski, which 
+  ensures the existence of generic filters for countable and transitive
+  models of $\ZF$;
+\end{inlinelist} 
+the last item was already communicated in the
   first report \cite{2018arXiv180705174G}.
-\end{inlinelist}
 

### 2018-12-29 LICS/intro_m.tex
sacando un "preservation"
internalized

@@
 
-In pursuing the proof of preservation on generic extensions we
+In pursuing the proof of Separation on generic extensions we
 extended Paulson's library with:
 \begin{inlinelist}
-\item renaming of variables, which with little effort can be extended
+\item renaming of variables for \emph{internalized} formulas, which
+  with little effort can be extended 
   to substitutions;

### 2018-12-29 LICS/conclusion.tex
un poco más de conclusión

@@
 postpone that task.  Next, various forcing notions must be developed
-in order to obtain extensions satisfying diverse properties. This
-yields relative consistency results.
+in order to obtain extensions satisfying diverse properties (v.g., the
+Continuum Hypothesis). This yields relative consistency results.
 
 We first discuss the axioms we are now missing. The proof of the
-Powerset Axiom does not present new difficulties but  has to
+Powerset Axiom does not present new difficulties but one  has to
 take care of absoluteness issues; it is one of the
@@
 universe of sets can be decomposed in a transfinite hierarchy of
-sets. 
+sets: 
 \begin{theorem}
@@
 $\forceisa$, which must be done by recursion, and this requirement
-spills indirectly to the proof of the Separation Axiom (despite this
+spills indirectly to the proof of the Separation Axiom (despite the
+latter
 is not by induction). Now the need of using internalized formulas
@@
 2-variable instance of Separation holds in $M[G]$, one needs to use a
-6-variable instance in $M$.
+6-variable instance in $M$. In the future, we expect to condense in a
+new locale a finite list of (instances of) axioms on $M$ enough to
+perform forcing constructions; this list will likely include all the
+instances of Separation and Replacement in $M$ that are needed to
+satisfy the requirements of the locales in the
+\isatt{ZF-Constructible} session.
 
-We believe that the
-\isatt{ZF-Constructible} session  would benefit from some changes in
+A few words are in order in relation to \isatt{ZF-Constructible}.
+We believe that this  session  would benefit from some changes in
 the way some results are organized accross theory files; for
@@
 as many absoluteness results follow from these hypotheses.
+As Paulson comments in the sources, it would have been better to
+minimize the use of the Powerset Axiom in locales and proofs. There
+are  useful natural models that satisfy a fraction of $\ZF$ not including
+this particular axiom, and to ensure a broader applicability, it would
+be convenient to have  absoluteness results not assuming it.
 It is our desire to advocate a future work to a thorough
-revision of the development of constructibility to maximize modularity.
+revision of the development of constructibility to address these needs
+and to maximize modularity.
 

### 2018-12-29 LICS/isabelle.tex
Mencionando dot G y explicitando la dificultad de tener check_in_M
jingle bells, jingle bells
eds varias

@@
 Isabelle provides a meta-language called \emph{Pure} that consists of
-a fragment of higher order logic. The meta-Boolean type is called
+a fragment of higher order logic, where \isatt{\isasymRightarrow} is
+the function-space arrow. The meta-Boolean type is called
 \isatt{prop}. Meta-connectives
@@
 have concrete consequences in our strategy to approach the
-development.
+development. From the beginning, we had to resort to
+\emph{internalized} formulas, i.e.\ elements of type $\tyi$ that
+encode first-order formulas with a binary relation symbol, and the
+satisfaction predicate \isatt{sats\,::\,"i\isasymRightarrow i\isasymRightarrow i\isasymRightarrow o"}  between a set
+model with an environment and an internalized formula (where the
+relation symbol is interpreted as membership). The set 
+\isatt{formula::"\tyi"}
+ of internalized
+formulas is defined by recursion and hence it is possible to perform
+inductive arguments using them. In this sense, the object-logic level
+is further divided into \emph{internal} and \emph{external}
+sublevels.
 
@@
 
-\fbox{Hacer benchmark para compararlos.}
-
-\medksip
+\medskip
 

### 2018-12-29 LICS/forcing.tex
eds varias

@@
 the structure of $\phi$. It is to be noted that the base case (viz.,
-for atomic $\phi$) contains all the complexity, and indeed it can be
-said that writing down this definition is half
-a way the path to set up the forcing machinery.
-\medskip
-
-\fbox{Compatibilizar lo que sigue con Renaming}
-\medskip
-
+for atomic $\phi$) contains all the complexity; the case for
+connectives and quantifiers is then straightforward.
+%% and indeed it can be
+%% said that writing down this definition is half
+%% a way the path to set up the forcing machinery.
 In order to state the properties of this map
@@
 internalized formulas, because it is not possible to carry inductive
-arguments over \tyo. These internalized versions are the elements of
-the recursively constructed set \formula{} (of type \tyi).
+arguments over \tyo
 

### 2018-12-29 LICS/forcing.tex
typos

@@
 internalized formulas, because it is not possible to carry inductive
-arguments over \tyo
+arguments over \tyo.
 

### 2018-12-29 LICS/intro_m.tex
typos

@@
 \item a variant of the  principle of dependent choices and a version
-  of Rassiowa-Sikorski, which 
+  of Rasiowa-Sikorski, which 
   ensures the existence of generic filters for countable and transitive

### 2018-12-29 LICS/intro_m.tex
eds varias

@@
 % \cite{2018arXiv180705174G}.
+We have  followed \cite{kunen2011set} for this purpose.
 
@@
 strategy:
-since the proofs that the  axioms hold in generic extensions
+Since the proofs that the  axioms hold in generic extensions
 is independent of the \emph{proofs} of the fundamental theorems of
@@
 from formulas to formulas which is, by itself, quite demanding; the
-formalization of it and of the fundamental theorems of forcing roughly
+formalization of it and of the fundamental theorems of forcing % roughly
 comprises barely less than a half of our full project.

### 2018-12-29 LICS/isabelle.tex
comento este comentario

@@
 do not require to know how it is defined. Therefore, we black-box it
-and pack everything in a locale that assumes that there is such a
+and pack everything in a locale called \texttt{forcing\_thms} that
+assumes that there is such a 
 map that satisfies the Fundamental Theorems.
 
-\begin{description}
-\item[\texttt{forcing\_notion}] preorden con top
-\item[\texttt{countable\_generic}] lo anterior con una familia contable de densos.
-\item[\texttt{M\_ZF}] axiomas.
-\item[\texttt{forcing\_data}]: lo anterior contable transitivo y una notion.
-\item[\texttt{forcing\_thms}]: eso.
-\item[\texttt{G\_generic}]: lo anterior y G es genérico.
-\item[\texttt{M\_extra\_assms}]: check in M e instancia de reemplazo para G.
-\item[\texttt{G\_generic\_extra}]: los dos anteriores (no sé si sigue estando)
-\end{description}
+%% \begin{description}
+%% \item[\texttt{forcing\_notion}] preorden con top
+%% \item[\texttt{countable\_generic}] lo anterior con una familia contable de densos.
+%% \item[\texttt{M\_ZF}] axiomas.
+%% \item[\texttt{forcing\_data}]: lo anterior contable transitivo y una notion.
+%% \item[\texttt{forcing\_thms}]: eso.
+%% \item[\texttt{G\_generic}]: lo anterior y G es genérico.
+%% \item[\texttt{M\_extra\_assms}]: check in M e instancia de reemplazo para G.
+%% \item[\texttt{G\_generic\_extra}]: los dos anteriores (no sé si sigue estando)
+%% \end{description}
 

### 2018-12-30 LICS/intro_m.tex
frase sobre testear la tecnologia de prueba actual... No muy convencido

@@
   [Separation].}  
-
-\medskip
-\fbox{(sigue p\'arrafo) Versi\'on de Pedro}
-\medskip
-
 The most important achievement of this paper is the
@@
 
-\medskip
-\fbox{Versi\'on de Miguel}
-\medskip
-
-The most important achievement of this paper is the
-mechanization in Isabelle of a considerable part of this result; en route
-to this, we also formalized the satisfaction by $M[G]$ of
-Extensionality, Foundation, Union, and Infinity. % We have already
-% proved Pairing in the first report of our project
-% \cite{2018arXiv180705174G}.
-
-\medskip
-\fbox{Fin p\'arrafo}
-\medskip
-
 Our development benefited from the remarkable work done by Lawrence
@@
 techniques needed to show the independence of the Continuum
-Hypothesis. We think that this project constitutes an interesting case
-which stresses how feasible is to formally implement mathematics that
-involve several levels of reasoning. \fbox{No entiendo la ultima oraci\'on}
+Hypothesis. 
+We think that this project constitutes an interesting test-case
+for the current technology of formalization of mathematics, in
+particular for the need of handling several layers of reasoning.
 

### 2018-12-30 LICS/isabelle.tex
typo
internalzd attribd to Paulson

@@
 recursively. That is, although there are constants and functions that
-generate elements of both types, neither of them are 
+generate elements of both types, neither of them is 
 \emph{initial}, in the sense that they are not the least types
@@
 development. From the beginning, we had to resort to
-\emph{internalized} formulas, i.e.\ elements of type $\tyi$ that
+\emph{internalized} formulas \cite[Sect.~6]{paulson_2003}, i.e.\ elements of type $\tyi$ that
 encode first-order formulas with a binary relation symbol, and the

### 2018-12-30 LICS/recursion.tex
section name
eds. varias. esp. recursion

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
-\subsection*{Recursion and Values of Names}
+\subsection*{Recursion and values of names}
 
@@
 
-As is well-known, the principle of transfinite recursion on
-well-founded relations~\cite[p. 48]{kunen2011set}) allows us to define
+As is well-known, the principle of  recursion on
+well-founded relations \cite[p.~48]{kunen2011set} allows us to define
 a recursive function $F \colon A\to A$ by choosing a well-founded
@@
 made this principle available in Isabelle/ZF via the the operator
-\isa{wfrec}. The formalization of the functional $\mathit{Hv}$ is
-straightforward:
+\isa{wfrec}. The formalization of the corresponding functional
+$\mathit{Hv}$ for $\val$ is straightforward:
 %
@@
 order to use \isa{wfrec} the relation should be expressed as a set, so
-we took the restriction of $\mathit{ed}$ to the whole universe
-$M$; i.e. $\mathit{ed}\cap M\times M$.  It was inconvenient to work
-with that definition because it imposed an obligation showing that
-some set is in $M$. The remedy was to restrict $\mathit{ed}$ to the
+in \cite{2018arXiv180705174G} we originally took the restriction of
+$\mathit{ed}$ to the whole universe 
+$M$; i.e. $\mathit{ed}\cap M\times M$.  Although this decision was
+adequate for that work, we now required more flexibility (for
+instance, in order to apply $\val$ to arguments that we can't assume
+that are in $M$, see Eq.~(\ref{eq:val-of-m}) below).
+
+The remedy is to restrict $\mathit{ed}$ to the
 transitive closure of the actual parameter:
@@
 \end{isabelle}
-A key result, albeit intuitive and rather easy, to make this
-definition work is that when computing the value of a recursive
+
+In order to show that this definition satisfies~(\ref{eq:val}) we had
+to supplement the existing recursion tools with a key, albeit
+intuitive, result stating that when computing the value of a recursive 
 function on some argument $a$, one can restrict the relation to some
-ambient set if it includes $a$ and all of its predecessesors.
+ambient set if it includes $a$ and all of its predecessors.
 \begin{isabelle}
@@
 \end{isabelle}
-
-This lemma gives rise to some useful results about values of
-names. First, we characterize $\val$ as in Equation~(\ref{eq:val}):
+As a consequence, we are able to formalize Equation~(\ref{eq:val}) as follows:
 \begin{isabelle}
@@
 interface described in Section~\ref{sec:axioms-models-set-theory} and
-also the concrete example given in the proof of Union).
+also the concrete example given in the proof of Union below).
 
@@
 \isa{M{\isacharunderscore}extra{\isacharunderscore}assms} assembling
-both assumptions ($M$ closed under $\chk$ and the instance of
+both assumptions ($M$ being closed under $\chk$ and the instance of
 Replacement); in this paper we explicitly mention where we use them.

### 2019-01-05 LICS/proof-of-separation.tex
eds. varias. esp. recursion
comiendo de powerset en paper

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
-\section{Proof of Separation}
+\section{Separation and Powerset}
 \label{sec:proof-separation}
+We proceed to describe in detail the main goal of this paper, the formalization
+of the proof of the Separation Axiom. Afterwards, we sketch  the
+implementation of the Powerset Axiom.
 
-This proof can be found in the file \verb|Separation_Axiom.thy| of the
-development, which we proceed to discuss in detail. The order chosen to
+This proof of Separation can be found in the file
+\verb|Separation_Axiom.thy|. The order chosen to
 implement
@@
 \isatt{Separation\_Axiom})
-\[
+\begin{equation}
+  \label{eq:val-of-m}
   \val(G,m) = \{x\in c : M[G],[x,w,c]\models \phi(x_0,x_1)\}.
-\]
+\end{equation}
 The right-hand side is trivially equal to $S$, but as a consequence of
@@
 
+We now turn to the Powerset Axiom. We followed the proof of
+\cite[IV.2.27]{kunen2011set}, and actually the main technical result,
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ Pow{\isacharunderscore}inter{\isacharunderscore}MG{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\isanewline
+\ \ \ \ {\isachardoublequoteopen}a{\isasymin}M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\isanewline
+\ \ \ \ {\isachardoublequoteopen}Pow{\isacharparenleft}a{\isacharparenright}\ {\isasyminter}\ M{\isacharbrackleft}G{\isacharbrackright}\ {\isasymin}\ M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}
+\end{isabelle}
+keeps most of the structure of the printed proof; this ``skeleton'' of the
+argument takes around 120 (short) lines. There are approximately 30
+more lines of bureaucracy in the proof of the last lemma. We preserved the
+names of variables used in the textbook, with the occasional question
+mark that distinguishes meta-variables.
+
+Two more
+absoluteness results concerning powersets were needed: These are
+refinements of results located in \isatt{Relative} that assume that
+$y$ below is in the class $M$.
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ {\isacharparenleft}\isakeyword{in}\ M{\isacharunderscore}trivial{\isacharparenright}\ powerset{\isacharunderscore}subset{\isacharunderscore}Pow{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\ \isanewline
+\ \ \ \ {\isachardoublequoteopen}powerset{\isacharparenleft}M{\isacharcomma}x{\isacharcomma}y{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymAnd}z{\isachardot}\ z{\isasymin}y\ {\isasymLongrightarrow}\ M{\isacharparenleft}z{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\ \isanewline
+\ \ \ \ {\isachardoublequoteopen}y\ {\isasymsubseteq}\ Pow{\isacharparenleft}x{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ {\isacharparenleft}\isakeyword{in}\ M{\isacharunderscore}trivial{\isacharparenright}\ powerset{\isacharunderscore}abs{\isacharcolon}\ \isanewline
+\ \ \isakeyword{assumes}\isanewline
+\ \ \ \ {\isachardoublequoteopen}M{\isacharparenleft}x{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymAnd}z{\isachardot}\ z{\isasymin}y\ {\isasymLongrightarrow}\ M{\isacharparenleft}z{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\isanewline
+\ \ {\isachardoublequoteopen}powerset{\isacharparenleft}M{\isacharcomma}x{\isacharcomma}y{\isacharparenright}\ {\isasymlongleftrightarrow}\ y\ {\isacharequal}\ {\isacharbraceleft}a{\isasymin}Pow{\isacharparenleft}x{\isacharparenright}\ {\isachardot}\ M{\isacharparenleft}a{\isacharparenright}{\isacharbraceright}{\isachardoublequoteclose}
+\end{isabelle}
+
 %%% Local Variables: 

### 2019-01-05 LICS/proof-of-separation.tex
breve reporte sobre powerset

@@
 We now turn to the Powerset Axiom. We followed the proof of
-\cite[IV.2.27]{kunen2011set}, and actually the main technical result,
+\cite[IV.2.27]{kunen2011set}, to which we refer the reader for further
+details, trying to  preserve the
+names of variables used in the textbook (with the occasional question
+mark that distinguishes meta-variables). Actually, the main technical result
 \begin{isabelle}
@@
 argument takes around 120 (short) lines. There are approximately 30
-more lines of bureaucracy in the proof of the last lemma. We preserved the
-names of variables used in the textbook, with the occasional question
-mark that distinguishes meta-variables.
+more lines of bureaucracy in the proof of the last lemma.
 
 Two more
-absoluteness results concerning powersets were needed: These are
-refinements of results located in \isatt{Relative} that assume that
+absoluteness lemmas concerning powersets were needed: These are
+refinements of results (\isatt{powerset\_Pow} and
+\isatt{powerset\_imp\_subset\_Pow}) located in the theory \isatt{Relative} which assume that
 $y$ below is in the class $M$.
@@
 
+Of the rest of the theory file \verb|Powerset_Axiom.thy|, a
+considerable fraction is taken by the proof of a closure property of
+the ctm $M$, that involves renaming of an internalized formula; also,
+the handling of the projections \isatt{fst} and \isatt{snd} must be
+done internally. 
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ sats{\isacharunderscore}fst{\isacharunderscore}snd{\isacharunderscore}in{\isacharunderscore}M{\isacharcolon}\isanewline
+\ \isakeyword{assumes}\ \isanewline
+\  \ {\isachardoublequoteopen}A{\isasymin}M{\isachardoublequoteclose}\ {\isachardoublequoteopen}B{\isasymin}M{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymphi}\ {\isasymin}\ formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}p{\isasymin}M{\isachardoublequoteclose}\ {\isachardoublequoteopen}l{\isasymin}M{\isachardoublequoteclose}\isanewline
+\ \  {\isachardoublequoteopen}o{\isasymin}M{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymchi}{\isasymin}M{\isachardoublequoteclose}\ {\isachardoublequoteopen}arity{\isacharparenleft}{\isasymphi}{\isacharparenright}\ {\isasymle}\ {\isadigit{6}}{\isachardoublequoteclose}\isanewline
+\  \isakeyword{shows}\isanewline
+\  \ {\isachardoublequoteopen}{\isacharbraceleft}sq\ {\isasymin}\ A{\isasymtimes}B\ {\isachardot}\isanewline
+\ \ \ \ \ sats{\isacharparenleft}M{\isacharcomma}{\isasymphi}{\isacharcomma}{\isacharbrackleft}p{\isacharcomma}l{\isacharcomma}o{\isacharcomma}snd{\isacharparenleft}sq{\isacharparenright}{\isacharcomma}fst{\isacharparenleft}sq{\isacharparenright}{\isacharcomma}{\isasymchi}{\isacharbrackright}{\isacharparenright}{\isacharbraceright}\ {\isasymin}\ M{\isachardoublequoteclose}\ \isanewline
+\end{isabelle}
+
 %%% Local Variables: 

### 2019-01-05 LICS/separation_arxiv.tex
versión para arXiv

\documentclass{article}
\usepackage{isabelle,isabellesym}
\renewcommand{\ttdefault}{cmtt}
\usepackage{csquotes}
\usepackage{enumitem}

\newlist{inlinelist}{enumerate*}{1}
\setlist*[inlinelist,1]{%
  label=(\roman*),
}
\usepackage{hyperref}
\usepackage[numbers]{natbib}
\input{header-draft}
\makeatletter
\def\foottext{\gdef\@thefnmark{}\@footnotetext}
\makeatother
\newcommand{\keywords}[1]{\foottext{\emph{Keywords:} #1}}
\newcommand{\ack}[1]{\par\bigskip \noindent \emph{Acknowledgment:} #1}

\begin{document}
\title{Separation in Generic Extensions for Isabelle}
\author{Emmanuel Gunther
  \and 
  Miguel Pagano
  \and 
  Pedro S\'anchez Terraf}
\maketitle

\begin{abstract}
  We mechanize, in the proof assistant
  Isabelle, a proof of the
  axiom-scheme of Separation in 
  generic extensions of models of set theory  
  by using the fundamental theorems of forcing.
  We also formalize the satisfaction of the axioms of
  Extensionality, Foundation, Union, and Powerset. The axiom of
  Infinity is likewise treated, under additional assumptions on the ground
  model.
  We also  extend Paulson's library on constructibility  with
  renaming of variables for internalized formulas, an improvement on
  definitions by recursion on well-founded  relations and sharpening
  of the hypotheses in his development of relativization and
  absoluteness.
\end{abstract}
\keywords{
Isabelle/ZF, forcing, names, generic extension, constructibility.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{intro_m}

\input{isabelle}

\input{zfc-axioms}

\input{renaming}

\input{generic-extensions}

\input{recursion}

% \input{prior-results}

\input{hacking}

\input{easy-axioms}

\input{forcing}

\input{proof-of-separation}

\input{conclusion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliographystyle{mi-estilo-else}
\bibliography{../LSFA/citados}
%\input{separation_arxiv.bbl}
\end{document}

%%% Local Variables:
%%% mode: latex
%%% ispell-local-dictionary: "american"
%%% End:

### 2019-01-05 LICS/forcing.tex
eds varias
ediciones varias, most sugeridas por Miguel

@@
 \label{sec:forcing}
+
+For the most part, we follow Kunen \cite{kunen2011set}. As
+an alternative, introductory 
+resource, the  interested reader can check
+\cite{chow-beginner-forcing}; the book \cite{weaver2014forcing}
+contains a thorough treament minimizing the technicalities.
 
@@
 to denote that $\phi$ is satisfied by $M$ when $x_i$ is interpreted
-to be $a_i$ ($i=0,\dots,n$). We call the list $[a_0,\dots,a_n]$ the
+as $a_i$ ($i=0,\dots,n$). We call the list $[a_0,\dots,a_n]$ the
 \emph{environment}. In the case of a formula of the form
@@
 \end{lemma}
-The previous two results combined are the ones usually called the
+\noindent The previous two results combined are the ones usually called the
 \emph{fundamental theorems}.
@@
 particular care so that, v.g., there are no missing hypotheses, so it
-is possible to take this statement as firm ground on which start a
+is possible to take this statement as firm ground on which to start a
 formalization of a proof.

### 2019-01-05 LICS/hacking.tex
eds varias
ediciones varias, most sugeridas por Miguel

@@
 relative consistency of the Axiom of Choice. This development is
-included inside the Isabelle distribution with session name
+included inside the Isabelle distribution under the session 
 \isatt{ZF-Constructible}. The main technical devices, invented by
@@
 \isatt{ZF-Constructible} right out-of-the-box. The main reason is that
-we can't expect state the ``class version'' of Replacement for a
+we can't expect to state the ``class version'' of Replacement for a
 \emph{set} $M$ by
-using first-order formulas, since predicates \isatt{P::"i=>o"} can't
+using first-order formulas, since predicates
+\isatt{P::"i\isasymRightarrow o"} can't
 be proved to be only the definable ones. Therefore, we had to make
-some modifications to the various sets of hypotheses in several
+some modifications in several
 locales to make the results available as tools for the present and
@@
 
-The most notable changes, located in the theory \texttt{Relative}, are
+The most notable changes, located in the theories \texttt{Relative}
+and \isatt{WF\_absolute}, are
 the following:
 \begin{enumerate}
-\item\label{item:1} We eliminated the requirement that the relative Axiom of Replacement
-  is satisfied by a class $M$ to be in the locale \isatt{M\_trivial}. 
-\item\label{item:2} We moved the requirement of the Powerset Axiom to \isatt{M\_basic}. 
-\item\label{item:3} We replaced the need that the set of natural numbers is in $M$ by the
-  milder hypothesis that $M(0)$. Actually, most results should follow
-  by only assuming that $M$ is nonempty.
-\end{enumerate}
-
-As a consequence of Item~\ref{item:1}, the lemma
+\item\label{item:1} The locale \isatt{M\_trivial}
+  does not assume that the underlying class $M$ satisfies the relative
+  Axiom of replacement.  As a consequence, the lemma
   \isatt{strong\_replacementI} is no longer valid and was commented
   out.
-
-We moved the requirement $M(\mathtt{nat})$ to the locale
-\isatt{M\_trancl} (inside the theory \isatt{WF\_absolute}), where it is needed for the first time. Some results,
+\item\label{item:2}  Originally the Powerset Axiom was assumed by the
+  locale \isatt{M\_trivial},   we moved this requirement to \isatt{M\_basic}. 
+\item\label{item:3} We replaced the need that the set of natural
+  numbers is in $M$ by the   milder hypothesis that $M(0)$. Actually,
+  most results should follow 
+  by only assuming that $M$ is nonempty.
+\item We moved the requirement $M(\mathtt{nat})$ to the locale
+  \isatt{M\_trancl}, where it is needed for the first time. Some results,
   for instance \isatt{rtran\_closure\_mem\_iff} and 
@@
   locale. 
+\end{enumerate}
+Because of these changes, some theory files from the
+\isatt{ZF-Constructible} session have been included among ours.
 

### 2019-01-05 LICS/abstract.tex
abstract

--- 2019-01-05 LICS/separation_arxiv.tex
@@
-\documentclass{article}
-\usepackage{isabelle,isabellesym}
-\renewcommand{\ttdefault}{cmtt}
-\usepackage{csquotes}
-\usepackage{enumitem}
-
-\newlist{inlinelist}{enumerate*}{1}
-\setlist*[inlinelist,1]{%
-  label=(\roman*),
-}
-\usepackage{hyperref}
-\usepackage[numbers]{natbib}
-\input{header-draft}
-\makeatletter
-\def\foottext{\gdef\@thefnmark{}\@footnotetext}
-\makeatother
-\newcommand{\keywords}[1]{\foottext{\emph{Keywords:} #1}}
-\newcommand{\ack}[1]{\par\bigskip \noindent \emph{Acknowledgment:} #1}
-
-\begin{document}
-\title{Separation in Generic Extensions for Isabelle}
-\author{Emmanuel Gunther
-  \and 
-  Miguel Pagano
-  \and 
-  Pedro S\'anchez Terraf}
-\maketitle
-
 \begin{abstract}
@@
   model.
-  We also  extend Paulson's library on constructibility  with
-  renaming of variables for internalized formulas, an improvement on
-  definitions by recursion on well-founded  relations and sharpening
+  In order to achieve these goals, we extended Paulson's library on
+  constructibility  with 
+  renaming of variables for internalized formulas, improved
+  results on 
+  definitions by recursion on well-founded  relations, 
+  and sharpened
   of the hypotheses in his development of relativization and
@@
 \end{abstract}
-\keywords{
-Isabelle/ZF, forcing, names, generic extension, constructibility.
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\input{intro_m}
-
-\input{isabelle}
-
-\input{zfc-axioms}
-
-\input{renaming}
-
-\input{generic-extensions}
-
-\input{recursion}
-
-% \input{prior-results}
-
-\input{hacking}
-
-\input{easy-axioms}
-
-\input{forcing}
-
-\input{proof-of-separation}
-
-\input{conclusion}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\bibliographystyle{mi-estilo-else}
-\bibliography{../LSFA/citados}
-%\input{separation_arxiv.bbl}
-\end{document}
 
@@
 %%% mode: latex
+%%% TeX-master: "Separation_In_MG"
 %%% ispell-local-dictionary: "american"

### 2019-01-05 LICS/separation_arxiv.tex
colores de links, PDF metadata
abstract

@@
 \renewcommand{\ttdefault}{cmtt}
+\usepackage{xcolor}
 \usepackage{csquotes}
@@
 
+\hypersetup{
+  pdftitle={Separation in Generic Extensions for Isabelle},
+  pdfsubject={Computer Science},
+  pdfkeywords={Isabelle/ZF, forcing, names, generic extension, constructibility},
+  colorlinks,
+  linkcolor={blue!40!black},
+  citecolor={blue!40!black},
+  urlcolor={blue!40!black}
+}
+
 \begin{document}
@@
 
-\begin{abstract}
-  We mechanize, in the proof assistant
-  Isabelle, a proof of the
-  axiom-scheme of Separation in 
-  generic extensions of models of set theory  
-  by using the fundamental theorems of forcing.
-  We also formalize the satisfaction of the axioms of
-  Extensionality, Foundation, Union, and Powerset. The axiom of
-  Infinity is likewise treated, under additional assumptions on the ground
-  model.
-  We also  extend Paulson's library on constructibility  with
-  renaming of variables for internalized formulas, an improvement on
-  definitions by recursion on well-founded  relations and sharpening
-  of the hypotheses in his development of relativization and
-  absoluteness.
-\end{abstract}
+\input{abstract}
 \keywords{

### 2019-01-05 LICS/separation_arxiv.tex
titulo e intro

@@
 \hypersetup{
-  pdftitle={Separation in Generic Extensions for Isabelle},
+  pdftitle={Mechanization of Separation in Generic Extensions},
   pdfsubject={Computer Science},
@@
 \begin{document}
-\title{Separation in Generic Extensions for Isabelle}
+\title{Mechanization of Separation in Generic Extensions}
 \author{Emmanuel Gunther

### 2019-01-05 LICS/intro_m.tex
titulo e intro

@@
 Zermelo-Fraenkel Set Theory ($\ZF$) has a prominent place among formal
-theories; in particular, it provides a foundation for mathematics and
-most of the formal toolkit used everyday by the computer scientist has
-also some form of set theory at its base
-(cf.~\cite{paulson1995set}). Considering the current trend 
+theories. The reason for this is that it formalizes many intuitive
+properties of the notion of set. As such, it can be used as a
+foundation for mathematics and thus it has been
+thoroughly studied. 
+Considering the current trend 
 of mechanization of mathematics~\cite{avigad2018mechanization}, it
@@
 The results we are interested in originally arose in connection to
-(relative) consistency proofs.
-After G\"odel's Incompleteness Theorems, we cannot expect to have a
-formal proof of the consistency of Set Theory in $\ZF$. Besides its own
-consistency, there are other statements which are undecided by $\ZF$.
-Perhaps the most prominent  is the
-\emph{Continuum Hypothesis}, which lead to the development of powerful
+\emph{relative consistency} proofs in set theory; that is, showing
+that if $\ZF$ is 
+consistent, the addition of a new axiom $A$ won't make the system
+inconsistent; this is as much as we can expect to obtain, since
+G\"odel's Incompleteness Theorems precludes a  
+formal proof of the consistency of set theory in $\ZF$, unless the
+latter is indeed inconsistent. There are statements $A$ which are
+\emph{undecided} by $\ZF$, in the sense that both $A$ and $\neg A$ are
+consistent relative to $\ZF$; perhaps the most prominent  is the
+\emph{Continuum Hypothesis}, which led to the development of powerful
 techniques for independence proofs. First, G\"odel inaugurated the
@@
 relative consistency of the Axiom of Choice and the Generalized
-Continuum Hypothesis with $\ZF$. Thirty years later, Paul
+Continuum Hypothesis with $\ZF$. More than twenty years later, Paul
 J.~Cohen~\cite{Cohen-CH-PNAS} devised the technique of \emph{forcing},
@@
 that given a model $M$ of $\ZF$, any \emph{generic extension} $M[G]$
-given by forcing also satisfies $\ZF$. As remarked by
+obtained by forcing also satisfies $\ZF$. As remarked by
 \citet[][p.250]{kunen2011set} \enquote{[...] in verifying that $M[G]$
@@
 to this, we also formalized the satisfaction by $M[G]$ of
-Extensionality, Foundation, Union, and Infinity. % We have already
+Extensionality, Foundation, and Union. As a consequence of Separation
+we were able to formalize the proof of the Powerset Axiom; finally,
+the Axiom of Infinity was proved under extra assumptions. % We have already
 % proved Pairing in the first report of our project
 % \cite{2018arXiv180705174G}.
-We have  followed \cite{kunen2011set} for this purpose.
+The theoretical support for this work has been 
+the fine textbook by Kunen \cite{kunen2011set} and our development
+benefited from the remarkable work done by Lawrence 
+Paulson \cite{paulson_2003} on the formalization of G\"odel's
+constructible universe in Isabelle.
 
-Our development benefited from the remarkable work done by Lawrence
-Paulson \cite{paulson_2003} on the formalization of G\"odel's
-constructible universe in Isabelle. The
+The
 ultimate goal of our project is the formalization of the forcing
@@
 Since the proofs that the  axioms hold in generic extensions
-is independent of the \emph{proofs} of the fundamental theorems of
-forcing, we assumed these results for the time being. Let us remark
+are independent of the \emph{proofs} of the fundamental theorems of
+forcing, we assumed the latter for the time being. Let us remark
 that those theorems depend on the definition of a function $\forceisa$

### 2019-01-05 LICS/intro_m.tex
"comments on the proof of the Powerset Axiom."

@@
 of the Separation Axiom Scheme in Section~\ref{sec:proof-separation},
-which follows closely its implementation. A plan for future work and
+which follows closely its implementation, and some comments on the
+proof of the Powerset Axiom. A plan for future work and
 some immediate conclusions are offered in

### 2019-01-05 LICS/conclusion_m.tex
Conclusión modificada

--- 2018-12-29 LICS/conclusion.tex
@@
 
-The proof that the axioms of $\ZF$ hold in generic extensions is 
-the first important landmark in the path of formalization of the
-theory of forcing. The task of proving the Fundamental Theorems can be
-considered a part of that landmark, but from a workflow point of view
-this can be done independently, and one of our design decisions was to
-postpone that task.  Next, various forcing notions must be developed
-in order to obtain extensions satisfying diverse properties (v.g., the
-Continuum Hypothesis). This yields relative consistency results.
+% Recordamos el objetivo del proyecto a "medio-término"
+% (auto-ironía). Podemos luego mencionar la distinción entre thematic y
+% programmatic (describiendo esta también). Luego mostramos ejemplos de
+% decisiones en ambos aspectos: Thematic: renaming(?), locales para
+% forcing (algo más?); Programmatic: trabajar en o, interfaz entre
+% modelo (satisfacción de fórmulas internalizadas) y meta-satisfacción
+% de axiomas relativizados en o. Indicando por qué esas decisiones nos
+% parecieron adecuadas.
 
-We first discuss the axioms we are now missing. The proof of the
-Powerset Axiom does not present new difficulties but one  has to
-take care of absoluteness issues; it is one of the
-first goals in the immediate future. This is also the case with the
-Axiom of Infinity; it can be proved in a direct but cumbersome way. We
-preferred to wait to have a full-fledged inteface between ctms and the
-locales providing recursive constructions from Paulson's
-\isatt{Constructibility} session. Then the current proof will hold
-with no extra assumptions. The same goes for the results $M\sbq M[G]$
-and $G\in M[G]$.
+% Luego señalar los logros del paper poniéndolos en perspectiva del
+% proyecto, destacando que sólo nos queda un axioma para tener M[G]|=
+% ZF. (Alguien quizás se podría preguntar por qué no está C, si es así
+% señalarlo). Explayarse sobre la prueba de replacement.
 
-The Replacement Axiom, however, requires some more work to be
-done. Both in Kunen and in other presentations \cite{neeman-course}
-it requires a relativized version (i.e., showing that it holds for $M$) of
-the \emph{Reflection Principle}. In order to state this meta-theoretic
-result by Montague, recall that
-an equivalent formulation of the Foundation Axiom states that the 
-universe of sets can be decomposed in a transfinite hierarchy of
-sets: 
+The ultimate goal of our project is a complete mechanization of
+forcing allowing for further developments (formalization of the
+relative consistency of CH) and even as a research tool for the
+working set-theorist. In the current paper we reported a first major
+milestone towards that goal; viz. a formal proof in Isabelle/ZF of the
+satisfaction by generic extensions of most of the ZFC axioms, except
+for the axiom-scheme of Replacement.
+
+We cannot overstate the importance of following the detailed
+presentation of forcing, even in its sharp style, given by
+\citet{kunen2011set}. In fact, it helped us to delineate the
+\emph{thematic} aspects of our formalization; i.e.~the handling of all
+the theoretical concepts and results in the subject and it informed
+the structure of locales organizing our development.
+
+% In a gross simplification, there are two aspects to a formalization
+% project like this one: thematic and programmatic. The first concerns
+% the handling of all the theoretical concepts and results in the
+% subject, while the second involves the practical issues of the
+% implementation and design. In the case of forcing, the main intricacy
+% lies in the first aspect. In this sense, following a sensible
+% presentation of the material is key.  The authoritative reference 
+% on the subject during the last 30 years has been Kunen's classical
+% \cite{kunen1980}, and  we have followed a recent rewrite
+% \cite{kunen2011set} 
+% of that  textbook, which presents the material in the same sharp 
+% style but offering a lot of details. In some sense this project
+% wouldn't exist without this book.
+
+The proof of Separation depends on the Fundamental Theorems of
+forcing; in the spirit of the Formal Abstract project we have only
+stated those theorems in our current development, because from a
+engineering point of view this can be done independently. The task of
+proving the Fundamental Theorems is another landmark of our project,
+after completing the formalization that generic extensions satisfy
+the axiom scheme of Replacement.
+
+% Next, various forcing notions must be developed in order to
+% obtain extensions satisfying diverse properties (v.g., the Continuum
+% Hypothesis). This yields relative consistency results.
+
+% We first discuss the axioms we are now missing. We preferred to wait
+% to have a full-fledged interface between ctms and the locales
+% providing recursive constructions from Paulson's
+% \isatt{Constructibility} session. Then the current proof will hold
+% with no extra assumptions. The same goes for the results $M\sbq M[G]$
+% and $G\in M[G]$.
+
+Let us discuss in more detail the proof of the Replacement Axiom. Both
+in Kunen and in other presentations (e.g. \cite{neeman-course}) it
+requires a relativized version (i.e., showing that it holds for $M$)
+of the \emph{Reflection Principle}. In order to state this
+meta-theoretic result by Montague, recall that an equivalent
+formulation of the Foundation Axiom states that the universe of sets
+can be decomposed in a transfinite hierarchy of sets:
 \begin{theorem}
@@
 This is an appropriate spot to insist that the internal/external
-dichotomy has been a powerful agent in the shaping of our
-project. It traces back to the definition of the formula-transformer
-$\forceisa$, which must be done by recursion, and this requirement
-spills indirectly to the proof of the Separation Axiom (despite the
-latter
-is not by induction). Now the need of using internalized formulas
+dichotomy has been a powerful agent in the shaping of our project. It
+traces back to the definition of the formula-transformer $\forceisa$,
+which must be done by recursion, and this requirement spills
+indirectly to the proof of the Separation Axiom (despite the latter is
+not by induction). Now the need of using internalized formulas
 reappears with the need of having a general version of the Reflection
-Principle (because the $\Phi$ there depends on the instance of Replacement
-being proved).
+Principle (because the $\Phi$ there depends on the instance of
+Replacement being proved).
 
-A secondary goal of this project is to assess which assumptions on the
-original ctm $M$ are needed to develop the forcing machinery. Up to this
-point we have some anecdotal data; for example, to show that a
+A secondary outcome of this project is to assess which assumptions on
+the ground model $M$ are needed to develop the forcing machinery. Up
+to this point we have some anecdotal data; for example, to show that a
 2-variable instance of Separation holds in $M[G]$, one needs to use a

### 2019-01-05 LICS/isabelle.tex
Referencias a papers sobre Isabelle. Cambio en la explicación de Isar.

@@
 \label{sec:logics}
-Isabelle provides a meta-language called \emph{Pure} that consists of
-a fragment of higher order logic, where \isatt{\isasymRightarrow} is
-the function-space arrow. The meta-Boolean type is called
-\isatt{prop}. Meta-connectives
+Isabelle \cite{Isabelle,DBLP:books/sp/Paulson94} provides a
+meta-language called \emph{Pure} that consists of a fragment of higher
+order logic, where \isatt{\isasymRightarrow} is the function-space
+arrow. The meta-Boolean type is called \isatt{prop}. Meta-connectives
 \isatt{\isasymLongrightarrow} and \isatt{\&\&\&} fulfill the role of
@@
 \emph{Isar} \cite{DBLP:conf/tphol/Wenzel99}, where intermediate
-statements in the course 
-of a proof are explicitly stated, interspersed with automatic 
-tactics handling more trivial steps. The goal is that the resulting
-text, a \emph{proof document}, can be understood without the need of
-running it.
+statements in the course of a proof are explicitly stated,
+interspersed with automatic tactics handling more trivial steps. The
+goal is that the resulting text, a \emph{proof document}, can be
+understood without the need of re-playing it viewing the proof state
+at each point.
 

### 2019-01-05 LICS/conclusion_m.tex
Profundizando un poco en la dicotomía internalizado/externalizado.

@@
 
-This is an appropriate spot to insist that the internal/external
-dichotomy has been a powerful agent in the shaping of our project. It
-traces back to the definition of the formula-transformer $\forceisa$,
-which must be done by recursion, and this requirement spills
-indirectly to the proof of the Separation Axiom (despite the latter is
-not by induction). Now the need of using internalized formulas
-reappears with the need of having a general version of the Reflection
-Principle (because the $\Phi$ there depends on the instance of
-Replacement being proved).
+This is an appropriate point to insist that the internal/external
+dichotomy has been a powerful agent in the shaping of our project.
+This tension was also pondered by Paulson in his formalization of
+Gödel's constructible universe \cite{MR2051585}; after choosing a
+shallow embedding of $\ZF$, every argument proved by induction on
+formulas (or functions defined by recursion) should be done using
+internalized formulas. Working on top of Paulson's library, we
+prototyped the formula-transformer $\forceisa$ is defined for
+internalized formulas and this affects indirectly the proof of the
+Separation Axiom (despite the latter is not by induction). The need of
+using internalized formulas reappears with the need of a general
+version of the Reflection Principle (because the $\Phi$ there depends
+on the instance of Replacement being proved).
 

### 2019-01-06 LICS/renaming.tex
mainly prueba de separación, labels por sección
renSat --> sats_iff_sats_ren
concretando el coproducto (y horrible ese guión)

@@
 \section{Renaming}
+\label{sec:renaming}
 \newcommand{\renaming}[2]{(#1)[#2]}
@@
 mappings between finite sets; since we can think of $\mathsf{succ}(n)$
-as the co-product $1+n$, then given a renaming $f \colon n \to m$, the
+as the coproduct $1+n = \{0\} \cup \{1,\dots,n\}$, then given a
+renaming $f \colon n \to m$, the 
 unique morphism $\mathsf{id}_1+f \colon 1+n \to 1+m$ is used to rename
@@
 In the last equation, \isa{sum{\isacharunderscore}id} corresponds to
-the co-product morfism $\mathsf{id}_{1}+f \colon 1 + n \to 1 +
+the coproduct morphism $\mathsf{id}_{1}+f \colon 1 + n \to 1 +
 n$. This definition of renaming exposes some inconveniences of working
@@
 \isacommand{lemma}\isamarkupfalse%
-\ renSat\ {\isacharcolon}\ \isanewline
+\ sats{\isacharunderscore}iff{\isacharunderscore}sats{\isacharunderscore}ren\ {\isacharcolon}\ \isanewline
 \ \ \isakeyword{fixes}\ {\isasymphi}\isanewline

### 2019-01-06 LICS/conclusion_m.tex
comentarios y cosas menores

@@
 forcing allowing for further developments (formalization of the
-relative consistency of CH) and even as a research tool for the
-working set-theorist. In the current paper we reported a first major
+relative consistency of $\CH$) and even as a research tool for the
+working set-theorist. \fbox{un poquito jugado lo anterior, lo vemos} In the current paper we reported a first major
 milestone towards that goal; viz. a formal proof in Isabelle/ZF of the
-satisfaction by generic extensions of most of the ZFC axioms, except
+satisfaction by generic extensions of most of the $\ZF$ axioms, except
 for the axiom-scheme of Replacement.
@@
 
+\medskip
+\fbox{Repite intro el p\'arrafo  que sigue}
+
+\medskip
+
 The proof of Separation depends on the Fundamental Theorems of
-forcing; in the spirit of the Formal Abstract project we have only
+Forcing; in the spirit of the Formal Abstracts project we have only
 stated those theorems in our current development, because from a
@@
 This tension was also pondered by Paulson in his formalization of
-Gödel's constructible universe \cite{MR2051585}; after choosing a
+G\"odel's constructible universe \cite{MR2051585}; after choosing a
 shallow embedding of $\ZF$, every argument proved by induction on
@@
 internalized formulas. Working on top of Paulson's library, we
-prototyped the formula-transformer $\forceisa$ is defined for
-internalized formulas and this affects indirectly the proof of the
+prototyped the formula-transformer $\forceisa$, which is defined for
+internalized formulas, and this affects indirectly the proof of the
 Separation Axiom (despite the latter is not by induction). The need of
@@
 file. Another, most basic example would be to start out with an even
-more locale that only assumes $M$ to be a nonempty transitive class,
+more basic locale that only assumes $M$ to be a nonempty transitive class,
 as many absoluteness results follow from these hypotheses.

### 2019-01-06 LICS/appendix.tex
Apéndice

\appendix[A short overview of our development]
In this appendix we succintly describe the contents of each file. In
the next page we include a dependency graph of our formalization. The
theories on a grayish background are directly from Paulson; we
highlight with light-blue/cyan those of Paulson that we modified. We
have developed from scratch the rest, in white.

\begin{description}
\item[\texttt{Nat\_Miscelanea}]Miscelaneous results for naturals, mostly
  needed for renamings.
\item[\texttt{Renaming}] Renaming of internalized formulas, see
  Section \ref{sec:renaming}.
\item[\texttt{Pointed\_DC}] A new choice principle.
\item[\texttt{Recursion\_Thms}] Enhancements about recursively defined
  functions.
\item[\texttt{Forcing\_Notions}] Definition of Posets with maximal
  element, filters, dense sets. Proof of Rasiowa-Sikorski.
\item[\texttt{Forcing\_Data}] Definition of the locales:
  \begin{inlinelist}
  \item \texttt{M\_ZF} satisfaction of axioms; and 
  \item \texttt{Forcing\_Data} extending the previous one with
    \texttt{Forcing\_notion}, transitivity, and being countable.
  \end{inlinelist}
\item[\texttt{Interface}] Instantiation of locales \texttt{M\_trivial}
  and \texttt{M\_basic} for every instance of \texttt{Forcing\_Data}.
\item[\texttt{Names}] Definitions of $\chk$, $\val$, and the generic
  extension. Varios results about them.
\item[\texttt{Forcing\_Theorems}] Specification of fundamental
  theorems of forcing, see Section~\ref{sec:forcing}.
\item[\texttt{*\_Axiom}] Proof of the satisfaction of the
  corresponding axiom in the generic extension.
\end{description}

\includepdf{session_graph_colored.pdf}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "Separation_In_MG"
%%% ispell-local-dictionary: "american"
%%% End:

### 2019-01-06 LICS/conclusion_m.tex
Conclusión, primera versión (casi final).

@@
 \label{sec:conclusions-future-work}
-
-% Recordamos el objetivo del proyecto a "medio-término"
-% (auto-ironía). Podemos luego mencionar la distinción entre thematic y
-% programmatic (describiendo esta también). Luego mostramos ejemplos de
-% decisiones en ambos aspectos: Thematic: renaming(?), locales para
-% forcing (algo más?); Programmatic: trabajar en o, interfaz entre
-% modelo (satisfacción de fórmulas internalizadas) y meta-satisfacción
-% de axiomas relativizados en o. Indicando por qué esas decisiones nos
-% parecieron adecuadas.
-
-% Luego señalar los logros del paper poniéndolos en perspectiva del
-% proyecto, destacando que sólo nos queda un axioma para tener M[G]|=
-% ZF. (Alguien quizás se podría preguntar por qué no está C, si es así
-% señalarlo). Explayarse sobre la prueba de replacement.
 
@@
 
-\medskip
-\fbox{Repite intro el p\'arrafo  que sigue}
-
-\medskip
-
-The proof of Separation depends on the Fundamental Theorems of
-Forcing; in the spirit of the Formal Abstracts project we have only
-stated those theorems in our current development, because from a
-engineering point of view this can be done independently. The task of
-proving the Fundamental Theorems is another landmark of our project,
-after completing the formalization that generic extensions satisfy
-the axiom scheme of Replacement.
+% Repite intro el p\'arrafo  que sigue
+%% The proof of Separation depends on the Fundamental Theorems of
+%% Forcing; in the spirit of the Formal Abstracts project we have only
+%% stated those theorems in our current development, because from a
+%% engineering point of view this can be done independently. The task of
+%% proving the Fundamental Theorems is another landmark of our project,
+%% after completing the formalization that generic extensions satisfy
+%% the axiom scheme of Replacement.
 
@@
 
-This statement is straightforwardly equivalent to that with $\Phi$
-consisting of a single formula. The schematic nature of this result
+% This statement is straightforwardly equivalent to that with $\Phi$
+% consisting of a single formula. 
+It is obvious that we can take the conjunction of $\Phi$ and state
+Theorem~\ref{th:reflection-principle} for a single formula, say $\phi$.
+The schematic nature of this result
 hints at a proof by induction on formulas, and hence it must be shown
@@
 This tension was also pondered by Paulson in his formalization of
-G\"odel's constructible universe \cite{MR2051585}; after choosing a
+G\"odel's constructible universe \cite{paulson_2003}; after choosing a
 shallow embedding of $\ZF$, every argument proved by induction on
@@
 internalized formulas, and this affects indirectly the proof of the
-Separation Axiom (despite the latter is not by induction). The need of
-using internalized formulas reappears with the need of a general
-version of the Reflection Principle (because the $\Phi$ there depends
-on the instance of Replacement being proved).
+Separation Axiom (despite the latter is not by induction). The proof
+of Replacement also calls for internalized formulas, because one needs
+a general version of the Reflection Principle (since the formula
+$\phi$ involved depends on the specific instance of Replacement being
+proved).
 
-A secondary outcome of this project is to assess which assumptions on
-the ground model $M$ are needed to develop the forcing machinery. Up
-to this point we have some anecdotal data; for example, to show that a
-2-variable instance of Separation holds in $M[G]$, one needs to use a
-6-variable instance in $M$. In the future, we expect to condense in a
-new locale a finite list of (instances of) axioms on $M$ enough to
-perform forcing constructions; this list will likely include all the
-instances of Separation and Replacement in $M$ that are needed to
-satisfy the requirements of the locales in the
+A secondary, more prosaic, outcome of this project is to precisely
+assess which assumptions on the ground model $M$ are needed to develop
+the forcing machinery. The obvious are transitivity and $M$ being
+countable (but keep in mind Lemma~\ref{lem:wf-model-implies-ctm}); the
+first because many absoluteness results follows from this, the latter
+for the existence of generic filters.  \fbox{Se me hace que la
+  anécdota empieza con forces} \fbox{que agrega cuatro variables} A
+more anecdotal one is that to show that any instance of Separation
+with at most two parameters holds in $M[G]$, one needs to assume every
+instance with six parameters in $M$ \fbox{Chequear los usos de any y
+  every}. The purpose of identifying those assumptions is to assemble
+in a locale the specific (instances of) axioms that should satisfy the
+ground model in order to perform forcing constructions; this list will
+likely include all the instances of Separation and Replacement that
+are needed to satisfy the requirements of the locales in the
 \isatt{ZF-Constructible} session.
 
-A few words are in order in relation to \isatt{ZF-Constructible}.
-We believe that this  session  would benefit from some changes in
-the way some results are organized accross theory files; for
-intance, by cataloging all or most of the internalized formulas in one
-file. Another, most basic example would be to start out with an even
-more basic locale that only assumes $M$ to be a nonempty transitive class,
-as many absoluteness results follow from these hypotheses.
-As Paulson comments in the sources, it would have been better to
-minimize the use of the Powerset Axiom in locales and proofs. There
-are  useful natural models that satisfy a fraction of $\ZF$ not including
-this particular axiom, and to ensure a broader applicability, it would
-be convenient to have  absoluteness results not assuming it.
-It is our desire to advocate a future work to a thorough
-revision of the development of constructibility to address these needs
-and to maximize modularity.
+We have already commented on our hacking of \isatt{ZF-Constructible}
+to maximize its modularity and thus the re-usability in other
+formalizations. We think it would be desirable to organize it
+somewhat differently: a trivial change is to catalogue in one file all
+the internalized formulas \fbox{No tiene la desventaja de que puede
+  llegar a ser necesario} \fbox{ agregar más en un futuro.}. A more
+conceptual modification would be to start out with an even more basic
+locale that only assumes $M$ to be a non-empty transitive class, as
+many absoluteness results follow from this hypothesis \fbox{Estaba en
+  plural hipótesis, por qué?}. Furthermore, as Paulson comments in the
+sources, it would have been better to minimize the use of the Powerset
+Axiom in locales and proofs. There are useful natural models that
+satisfy a sub-theory of $\ZF$ not including Powersete, and to ensure a
+broader applicability, it would be convenient to have absoluteness
+results not assuming it. Once we gained more confidence that these
+ideas, we plan to contribute back to the official distribution of
+Isabelle/ZF with a thorough revision of the development of
+constructibility.
 
 \section*{Acknowledgment}
+
+The second author is grateful to the others two. They are very nice persons.
 

### 2019-01-06 LICS/zfc-axioms.tex
eds varias
Enumeración de los esquemas

@@
 axioms of Extensionality, Pairing, Union, Powerset, Foundation, Infinity, and two
-axiom-schemes collectively referred to as Axiom of Separation:
+axiom-schemes collectively referred as
+\begin{inlinelist}[label=(\emph{\alph*})]
+\item Axiom of Separation:
 \emph{For every $A$, $a_1,\dots,a_n$,  and  a formula
@@
   \psi(a,a_1,\dots,a_n)\}$}, 
-and Axiom of Replacement: \emph{For every $A$, $a_1,\dots,a_n$,  and
+and \item Axiom of Replacement: \emph{For every $A$, $a_1,\dots,a_n$,  and
   a formula   $\psi(x,z,x_1\dots,x_n)$, if 
@@
   $\{b : \exists a\in A. \psi(a,b,a_1,\dots,a_n)\}$}.
+\end{inlinelist}
 An excellent introduction to the axioms and the motivation behind them
@@
 for formulas with at most one extra parameter (meaning $n\leq 1$ in the
-above $\psi$), where \isatt{sats\,::\,"[i,i,i]=>o"} is the satisfaction
-relation between a set model with an environment and an internalized
-formula. Thanks to Pairing, these 
+above $\psi$). Thanks to Pairing, these 
 versions are equivalent to the usual formulations. We are only able to

### 2019-01-06 LICS/conclusion_m.tex
concl revisada, apéndice en ambos formatos

@@
 forcing allowing for further developments (formalization of the
-relative consistency of $\CH$) and even as a research tool for the
-working set-theorist. \fbox{un poquito jugado lo anterior, lo vemos} In the current paper we reported a first major
+relative consistency of $\CH$), with the long-term hope that 
+working set-theorists will adopt these formal tools as an aid to their
+research. In the current paper we reported a first major
 milestone towards that goal; viz. a formal proof in Isabelle/ZF of the
@@
 
-We cannot overstate the importance of following the detailed
-presentation of forcing, even in its sharp style, given by
+We cannot overstate the importance of following the sharp and detailed 
+presentation of forcing given by
 \citet{kunen2011set}. In fact, it helped us to delineate the
@@
 
-% We first discuss the axioms we are now missing. We preferred to wait
-% to have a full-fledged interface between ctms and the locales
-% providing recursive constructions from Paulson's
-% \isatt{Constructibility} session. Then the current proof will hold
-% with no extra assumptions. The same goes for the results $M\sbq M[G]$
-% and $G\in M[G]$.
+Two axioms have not been treated in full. Infinity was proved under
+two extra assumptions on the model; when we develop a full-fledged
+interface between ctms of $\ZF$  and the locales 
+providing recursive constructions from Paulson's
+\isatt{ZF-Constructible} session, the same current proof will hold
+with no extra assumptions. The same goes for the results $M\sbq M[G]$
+and $G\in M[G]$.
 
-Let us discuss in more detail the proof of the Replacement Axiom. Both
+The Replacement Axiom, however, requires some more work to be
+done. Both
 in Kunen and in other presentations (e.g. \cite{neeman-course}) it
@@
 first because many absoluteness results follows from this, the latter
-for the existence of generic filters.  \fbox{Se me hace que la
-  anécdota empieza con forces} \fbox{que agrega cuatro variables} A
-more anecdotal one is that to show that any instance of Separation
-with at most two parameters holds in $M[G]$, one needs to assume every
-instance with six parameters in $M$ \fbox{Chequear los usos de any y
-  every}. The purpose of identifying those assumptions is to assemble
+for the existence of generic filters. A
+more anecdotal one is that to show that an instance of Separation
+with at most two parameters holds in $M[G]$, one needs to assume a
+particular  six-parameter
+instance in $M$ (four extra parameters can be directly blamed on
+$\forceisa$).  The purpose of identifying those assumptions is to assemble
 in a locale the specific (instances of) axioms that should satisfy the
@@
 formalizations. We think it would be desirable to organize it
-somewhat differently: a trivial change is to catalogue in one file all
-the internalized formulas \fbox{No tiene la desventaja de que puede
-  llegar a ser necesario} \fbox{ agregar más en un futuro.}. A more
+somewhat differently: a trivial change is to catalog in one file all
+the internalized formulas. A more
 conceptual modification would be to start out with an even more basic
 locale that only assumes $M$ to be a non-empty transitive class, as
-many absoluteness results follow from this hypothesis \fbox{Estaba en
-  plural hipótesis, por qué?}. Furthermore, as Paulson comments in the
+many absoluteness results follow from this hypothesis. Furthermore, as
+Paulson comments in the 
 sources, it would have been better to minimize the use of the Powerset
 Axiom in locales and proofs. There are useful natural models that
-satisfy a sub-theory of $\ZF$ not including Powersete, and to ensure a
+satisfy a sub-theory of $\ZF$ not including Powerset, and to ensure a
 broader applicability, it would be convenient to have absoluteness
-results not assuming it. Once we gained more confidence that these
-ideas, we plan to contribute back to the official distribution of
-Isabelle/ZF with a thorough revision of the development of
+results not assuming it. We plan to contribute back to the official
+distribution of Isabelle/ZF with a thorough revision of the development of
 constructibility.
 
-\section*{Acknowledgment}
-
-The second author is grateful to the others two. They are very nice persons.
+%% \section*{Acknowledgment}
+%%
 

### 2019-01-06 LICS/appendix.tex
typo en apéndice
concl revisada, apéndice en ambos formatos

@@
-\appendix[A short overview of our development]
-In this appendix we succintly describe the contents of each file. In
-the next page we include a dependency graph of our formalization. The
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Appendix
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+In this appendix we succintly describe the contents of each file. We
+include in 
+Figure~\ref{fig:deps}  a dependency graph of our formalization. The
 theories on a grayish background are directly from Paulson; we
-highlight with light-blue/cyan those of Paulson that we modified. We
+highlight with blue/cyan those of Paulson that we modified. We
 have developed from scratch the rest, in white.
@@
   Section \ref{sec:renaming}.
-\item[\texttt{Pointed\_DC}] A new choice principle.
+\item[\texttt{Pointed\_DC}] A pointed version of the Principle of
+  Dependent Choices.
 \item[\texttt{Recursion\_Thms}] Enhancements about recursively defined
@@
   \item \texttt{M\_ZF} satisfaction of axioms; and 
-  \item \texttt{Forcing\_Data} extending the previous one with
-    \texttt{Forcing\_notion}, transitivity, and being countable.
+  \item \texttt{forcing\_data} extending the previous one with
+    \texttt{forcing\_notion}, transitivity, and being countable.
   \end{inlinelist}
@@
 \end{description}
-
-\includepdf{session_graph_colored.pdf}
+%
+\begin{figure*}[!b]
+  \begin{center}
+    \includegraphics[height=14.8cm]{session_graph_colored.pdf}
+  \end{center}
+  \caption{Dependency graph of the \isatt{Separation} session.}
+  \label{fig:deps}
+\end{figure*}
 

### 2019-01-06 LICS/forcing.tex
Cambios en renaming; notación para satisfacción en renaming. Ligera adaptación de forcing por eso.

@@
 
-In referring to formulas with free variables and their interpretation,
-we use the following conventions. If  $M$ is a set, $a_0,\dots,a_n$
-are elements of $M$, and $\phi$ has its free variables
-among $x_0,\dots,x_n$, we write
-\[
-M,[a_0,\dots,a_n] \models \phi(x_0,\dots,x_n)
-\]
-to denote that $\phi$ is satisfied by $M$ when $x_i$ is interpreted
-as $a_i$ ($i=0,\dots,n$). We call the list $[a_0,\dots,a_n]$ the
-\emph{environment}. In the case of a formula of the form
-$\forceisa(\phi)$, we'll make an abuse of notation and indicate the
-variables inside the argument of $\forceisa$. As an example, take the
-formula $\phi\defi x_1\in x_0$. Then
+We write $\phi(x_0,\dots,x_n)$ to indicate that the free variables of
+$\phi$ are among $\{x_0,\dots,x_n\}$. In the case of a formula of the
+form $\forceisa(\phi)$, we will make an abuse of notation and indicate
+the variables inside the argument of $\forceisa$. As an example, take
+the formula $\phi\defi x_1\in x_0$. Then
 \[

### 2019-01-06 LICS/renaming.tex
faltaba un subíndice parece
Cambios en renaming; notación para satisfacción en renaming. Ligera adaptación de forcing por eso.

@@
     \renaming{\andFm{\phi}{\psi}}{f} = \andFm{\renaming{\phi}{f}}{\renaming{\psi}{f}}\\
-    \renaming{\forallFm{\phi}}{f} = \forallFm{\renaming{\phi}{\mathsf{id}+f}}
+    \renaming{\forallFm{\phi}}{f} = \forallFm{\renaming{\phi}{\mathsf{id}_1+f}}
   \end{gather*}
@@
 
-Environments can be thought, then, as mappings from a finite set to
-the model $M$; the action of renaming on environments is simply given
-by composition. An easy proof connects satisfaction with renamings.
+As usual, if $M$ is a set, $a_0,\dots,a_{n-1}$ are elements of $M$, and
+$\phi$ is a formula of arity $n$, we write
+\[
+M,[a_0,\dots,a_{n-1}] \models \phi
+\]
+to denote that $\phi$ is satisfied by $M$ when $i$ is interpreted
+as $a_i$ ($i=0,\dots,n-1$). We call the list $[a_0,\dots,a_{n-1}]$ the
+\emph{environment}.
+
+The action of renaming on environments re-indexes the variables. An
+easy proof connects satisfaction with renamings.
 \begin{lemma}
@@
   Let $\phi$ be a formula of arity $n$, $f \colon n \to m$ be a
-  renaming, and let $\rho \colon n \to M$ and $\rho' \colon m \to M$
-  be environments. If for all $i \in n$,
-  $\rho\,i = (\rho' \circ f)\,i)$, then $M,\rho\models \phi$ is
-  equivalent to $M,\rho' \models \renaming{\phi}{f}$.
+  renaming, and let $\rho=[a_1,\ldots,a_n]$ and
+  $\rho'=[b_1,\ldots,b_m]$ be environments of length $n$ and $m$,
+  respectively. If for all $i \in n$, $a_i = b_{j}$ where $j=f\,i$,
+  then $M,\rho\models \phi$ is equivalent to
+  $M,\rho' \models \renaming{\phi}{f}$.
 \end{lemma}
@@
 a function \isa{ren} that associates to each formula an internalized
-function that can be later applied to suitable arguments.
+function that can be later applied to suitable arguments. Notice the
+Paulson used \isa{Nand} because it is more economical.
 \begin{isabelle}
@@
 the coproduct morphism $\mathsf{id}_{1}+f \colon 1 + n \to 1 +
-n$. This definition of renaming exposes some inconveniences of working
-in the untyped realm of set theory, for example to use \isa{ren} we
-will need to prove that the renaming is a function. Since environments
-are not finite mappings but lists (yet another inductively defined
-set), the main hypothesis of Lem.~\ref{lem:renaming} is expressed with
-the \isa{nth} function for indexing elements of lists:
+n$. Since the schema for recursively defined functions does not allow
+parameters, we are forced to return a function of three arguments
+(\isa{n,m,f}). This also exposes some inconveniences of working in the
+untyped realm of set theory; for example to use \isa{ren} we will need
+to prove that the renaming is a function. Besides some auxiliary
+results (the application of renaming to suitable arguments yields a
+formula), the main result corresponding to Lemma~\ref{lem:renaming}
+is:
 \begin{isabelle}
@@
 
+% All our uses of this lemma involve concrete renamings on small
+% numbers, but we also tested it with more abstract renamings for
+% arbitrary numbers. We think that it should be possible, and clearer,
+% to express all the current renamings in the theory \isa{Formula} using
+% our approach.
+
 %%% Local Variables: 

### 2019-01-06 LICS/renaming.tex
comentario del final a medias

@@
 
-% All our uses of this lemma involve concrete renamings on small
-% numbers, but we also tested it with more abstract renamings for
-% arbitrary numbers. We think that it should be possible, and clearer,
-% to express all the current renamings in the theory \isa{Formula} using
-% our approach.
+All our uses of this lemma involve concrete renamings on small
+numbers, but we also tested it with more abstract one for
+arbitrary numbers. All the renamings of the first kind follow the
+same pattern and, more importantly, share equal proofs. We would like
+to develop some \texttt{ML} tools in order to automatize this.
+%% We think that it should be possible, and clearer,
+%% to express all the current renamings in the theory \isa{Formula} using
+%% our approach.
 

### 2019-01-06 LICS/renaming.tex
typo
overflows y concl_m en arXiv

@@
 \isacommand{consts}\isamarkupfalse%
-\ \ \ ren\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isacharequal}{\isachargreater}i{\isachardoublequoteclose}\isanewline
+\ ren\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isacharequal}{\isachargreater}i{\isachardoublequoteclose}\isanewline
 \isacommand{primrec}\isamarkupfalse%
 \isanewline
-\ \ {\isachardoublequoteopen}ren{\isacharparenleft}Member{\isacharparenleft}x{\isacharcomma}y{\isacharparenright}{\isacharparenright}\ {\isacharequal}\isanewline
-\ \ \ \ \ \ {\isacharparenleft}{\isasymlambda}\ n\ {\isasymin}\ nat\ {\isachardot}\ {\isasymlambda}\ m\ {\isasymin}\ nat{\isachardot}\ {\isasymlambda}f\ {\isasymin}\ n\ {\isasymrightarrow}\ m{\isachardot}\ Member\ {\isacharparenleft}f{\isacharbackquote}x{\isacharcomma}\ f{\isacharbackquote}y{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
-\ \ \isanewline
-\ \ {\isachardoublequoteopen}ren{\isacharparenleft}Equal{\isacharparenleft}x{\isacharcomma}y{\isacharparenright}{\isacharparenright}\ {\isacharequal}\isanewline
-\ \ \ \ \ \ {\isacharparenleft}{\isasymlambda}\ n\ {\isasymin}\ nat\ {\isachardot}\ {\isasymlambda}\ m\ {\isasymin}\ nat{\isachardot}\ {\isasymlambda}f\ {\isasymin}\ n\ {\isasymrightarrow}\ m{\isachardot}\ Equal\ {\isacharparenleft}f{\isacharbackquote}x{\isacharcomma}\ f{\isacharbackquote}y{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
-\ \ \isanewline
-\ \ {\isachardoublequoteopen}ren{\isacharparenleft}Nand{\isacharparenleft}p{\isacharcomma}q{\isacharparenright}{\isacharparenright}\ {\isacharequal}\isanewline
-\ \ \ \ \ \ {\isacharparenleft}{\isasymlambda}\ n\ {\isasymin}\ nat\ {\isachardot}\ {\isasymlambda}\ m\ {\isasymin}\ nat{\isachardot}\ {\isasymlambda}f\ {\isasymin}\ n\ {\isasymrightarrow}\ m{\isachardot}\ Nand\ {\isacharparenleft}ren{\isacharparenleft}p{\isacharparenright}{\isacharbackquote}n{\isacharbackquote}m{\isacharbackquote}f{\isacharcomma}\ ren{\isacharparenleft}q{\isacharparenright}{\isacharbackquote}n{\isacharbackquote}m{\isacharbackquote}f{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
-\ \ \isanewline
-\ \ {\isachardoublequoteopen}ren{\isacharparenleft}Forall{\isacharparenleft}p{\isacharparenright}{\isacharparenright}\ {\isacharequal}\isanewline
-\ \ \ \ \ \ {\isacharparenleft}{\isasymlambda}\ n\ {\isasymin}\ nat\ {\isachardot}\ {\isasymlambda}\ m\ {\isasymin}\ nat{\isachardot}\ {\isasymlambda}f\ {\isasymin}\ n\ {\isasymrightarrow}\ m{\isachardot}\ Forall\ {\isacharparenleft}ren{\isacharparenleft}p{\isacharparenright}{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isacharbackquote}succ{\isacharparenleft}m{\isacharparenright}{\isacharbackquote}sum{\isacharunderscore}id{\isacharparenleft}n{\isacharcomma}f{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\ {\isachardoublequoteopen}ren{\isacharparenleft}Member{\isacharparenleft}x{\isacharcomma}y{\isacharparenright}{\isacharparenright}\ {\isacharequal}\isanewline
+\ \ {\isacharparenleft}{\isasymlambda}\ n\ {\isasymin}\ nat\ {\isachardot}\ {\isasymlambda}\ m\ {\isasymin}\ nat{\isachardot}\ {\isasymlambda}f\ {\isasymin}\ n\ {\isasymrightarrow}\ m{\isachardot}\ Member\ {\isacharparenleft}f{\isacharbackquote}x{\isacharcomma}\ f{\isacharbackquote}y{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \isanewline
+\ {\isachardoublequoteopen}ren{\isacharparenleft}Equal{\isacharparenleft}x{\isacharcomma}y{\isacharparenright}{\isacharparenright}\ {\isacharequal}\isanewline
+\ \ {\isacharparenleft}{\isasymlambda}\ n\ {\isasymin}\ nat\ {\isachardot}\ {\isasymlambda}\ m\ {\isasymin}\ nat{\isachardot}\ {\isasymlambda}f\ {\isasymin}\ n\ {\isasymrightarrow}\ m{\isachardot}\ Equal\ {\isacharparenleft}f{\isacharbackquote}x{\isacharcomma}\ f{\isacharbackquote}y{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \isanewline
+\ {\isachardoublequoteopen}ren{\isacharparenleft}Nand{\isacharparenleft}p{\isacharcomma}q{\isacharparenright}{\isacharparenright}\ {\isacharequal}\isanewline
+\ \ {\isacharparenleft}{\isasymlambda}\ n\ {\isasymin}\ nat\ {\isachardot}\ {\isasymlambda}\ m\ {\isasymin}\ nat{\isachardot}\ {\isasymlambda}f\ {\isasymin}\ n\ {\isasymrightarrow}\ m{\isachardot}\ \isanewline
+\ \ \ Nand\ {\isacharparenleft}ren{\isacharparenleft}p{\isacharparenright}{\isacharbackquote}n{\isacharbackquote}m{\isacharbackquote}f{\isacharcomma}\ ren{\isacharparenleft}q{\isacharparenright}{\isacharbackquote}n{\isacharbackquote}m{\isacharbackquote}f{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \isanewline
+\ {\isachardoublequoteopen}ren{\isacharparenleft}Forall{\isacharparenleft}p{\isacharparenright}{\isacharparenright}\ {\isacharequal}\isanewline
+\ \  {\isacharparenleft}{\isasymlambda}\ n\ {\isasymin}\ nat\ {\isachardot}\ {\isasymlambda}\ m\ {\isasymin}\ nat{\isachardot}\ {\isasymlambda}f\ {\isasymin}\ n\ {\isasymrightarrow}\ m{\isachardot}\ \isanewline
+\ \ \ Forall\ {\isacharparenleft}ren{\isacharparenleft}p{\isacharparenright}{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isacharbackquote}succ{\isacharparenleft}m{\isacharparenright}{\isacharbackquote}sum{\isacharunderscore}id{\isacharparenleft}n{\isacharcomma}f{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
 \end{isabelle}
@@
 All our uses of this lemma involve concrete renamings on small
-numbers, but we also tested it with more abstract one for
-arbitrary numbers. All the renamings of the first kind follow the
-same pattern and, more importantly, share equal proofs. We would like
-to develop some \texttt{ML} tools in order to automatize this.
+numbers, but we also tested it with more abstract ones for arbitrary
+numbers. All the renamings of the first kind follow the same pattern
+and, more importantly, share equal proofs. We would like to develop
+some \texttt{ML} tools in order to automatize this.
 %% We think that it should be possible, and clearer,

### 2019-01-06 LICS/easy-axioms.tex
promesa cumplida sobre pairing
comentarios y cosas menores
overflow

@@
 In our first presentation of this project \cite{2018arXiv180705174G},
-we proved tha $M[G]$ satisfies Pairing. It is straightforward to show
-that the generic extension $M[G]$ satisfies extensionality and
-foundation. Showing that it is closed under Union depends on $G$ being
-a filter. Infinity is also easy, but it depends in one further
-assumption.
+we proved that $M[G]$ satisfies Pairing; now we have redone this proof
+in Isar. It is straightforward to show that the generic extension
+$M[G]$ satisfies extensionality and foundation. Showing that it is
+closed under Union depends on $G$ being a filter. Infinity is also
+easy, but it depends in one further assumption.
 
@@
 \isa{Unioin{\isacharunderscore}name{\isacharunderscore}body} and prove the equivalence between
+\begin{center}
 \isa{sats(M,union{\isacharunderscore}name{\isacharunderscore}fm,[P,leq,\isasymtau,x])}
+\end{center}
 and

### 2019-01-07 LICS/conclusion_m.tex
forcing_thms

@@
 milestone towards that goal; viz. a formal proof in Isabelle/ZF of the
-satisfaction by generic extensions of most of the $\ZF$ axioms, except
-for the axiom-scheme of Replacement.
+satisfaction by generic extensions of most of the $\ZF$ axioms.%% , except
+%% for the axiom-scheme of Replacement.
 
@@
 the theoretical concepts and results in the subject and it informed
-the structure of locales organizing our development.
+the structure of locales organizing our development. This had an
+impact, in particular, in the formal statement of the Fundamental
+Theorems. We consider that the writing of the \isatt{forcing\_thms}
+locale, though only taking a few lines of code, is the second most
+important achievement of this work, since there is no obvious
+reference from which to translate this directly.
 

### 2019-01-07 LICS/conclusion_m.tex
otra vuelta más a forcing_thms
kaiser & barras en concl_m. absolutez en powerset

@@
 important achievement of this work, since there is no obvious
-reference from which to translate this directly.
+reference from which to translate this directly. The accomplishment
+of the formalizations of Separation and Powerset are, in a sense,
+certificates that the locale of the Fundamental Theorems was set
+correctly.
 
@@
 $\phi$ involved depends on the specific instance of Replacement being
-proved).
+proved). Other formalizations, v.g.\ that of Barras
+\cite{barras-habilit} and Kaiser \cite{kaiser-set-theory-coq}, take
+advantage of a better articulation between the metatheory (Coq in both
+cases) and the set theory. An alternative road would be to redevelop
+absoluteness 
+results in a more structured metatheory that already includes a
+recursively defined type of first order formulas. Needless to say, this
+comprises an extensive re-engineering.
 

### 2019-01-07 LICS/forcing.tex
compilación diferenciada arXiv/IEEE
uuups... density lemma mal

@@
   \item For all $M$-generic filters $G$ such that $p\in G$,
-    $M[G],[\val(G,\tau_0),\dots,\val(G,\tau_n)] \models\phi(x_0,\dots,x_n)$.
+    \ifIEEE
+    $M[G],[\val(G,\tau_0),\dots,\val(G,\tau_n)]
+    \models\phi(x_0,\dots,x_n)$.
+    \fi
+    \ifarXiv
+    \[
+    M[G],[\val(G,\tau_0),\dots,\val(G,\tau_n)]
+    \models\phi(x_0,\dots,x_n).
+    \]
+    \fi
   \end{enumerate}
@@
   \models\phi(x_0,\dots,x_n)$.
-  \item  There exists $p\in G$ such that $M,[\PP,\leq,\1,p,\tau_0,\dots,\tau_n] 
+  \item  There exists $p\in G$ such that 
+    \ifIEEE 
+    $M,[\PP,\leq,\1,p,\tau_0,\dots,\tau_n] 
     \models\forceisa(\phi(x_4,\dots,x_{n+4}))$.
+    \fi
+    \ifarXiv 
+    \[ 
+    M,[\PP,\leq,\1,p,\tau_0,\dots,\tau_n] 
+    \models\forceisa(\phi(x_4,\dots,x_{n+4})).
+    \]
+    \fi
   \end{enumerate}
@@
   Assume the same hypothesis of
-  Lemma~\ref{lem:definition-of-forcing}. $M,[\dots]\models\forceisa(\phi)$ 
+  Lemma~\ref{lem:definition-of-forcing}. $M,[\PP,\leq,\1,p,\dots]\models\forceisa(\phi)$   
   if and only if 
-  $M,[\dots]\models$``$\{p_1\leq p : \forceisa(\phi,\PP,\leq,\1,p_1)\}$ is
-  dense below $p$.''
+  \[
+  \{p_1\in \PP : M,[\PP,\leq,\1,p_1,\dots]\models
+  \forceisa(\phi)\}
+  \]
+  is dense below $p$.
 \end{lemma}

### 2019-01-08 LICS/proof-of-separation.tex
ediciones varias, most sugeridas por Miguel
kaiser & barras en concl_m. absolutez en powerset
redacción

@@
 \cite[IV.2.27]{kunen2011set}, to which we refer the reader for further
-details, trying to  preserve the
-names of variables used in the textbook (with the occasional question
-mark that distinguishes meta-variables). Actually, the main technical result
+details. Actually, the main technical result,
 \begin{isabelle}
@@
 keeps most of the structure of the printed proof; this ``skeleton'' of the
-argument takes around 120 (short) lines. There are approximately 30
+argument takes around 120 (short) lines, where we tried to  preserve the
+names of variables used in the textbook (with the occasional question
+mark that distinguishes meta-variables). There are approximately 30
 more lines of bureaucracy in the proof of the last lemma.
@@
 refinements of results (\isatt{powerset\_Pow} and
-\isatt{powerset\_imp\_subset\_Pow}) located in the theory \isatt{Relative} which assume that
-$y$ below is in the class $M$.
+\isatt{powerset\_imp\_subset\_Pow}) located in the theory
+\isatt{Relative} 
+% which assumed that $y$ below is in the class $M$.
+where we weakened the assumption ``$y\in M$'' (\isatt{M(y)}) to 
+``$y\sbq  M$'' (second assumption below).  
 \begin{isabelle}

### 2019-01-09 LICS/hacking.tex
guión en non-empty
varias observaciones de Manu

@@
 is \emph{absolute} for $C$ if it is equivalent to its relativization,
-meaning that the statement made $\phi$ coincides with what $C$
+meaning that the statement made by $\phi$ coincides with what $C$
 ``believes'' $\phi$ is saying. Paulson shows that under certain
@@
   most results should follow 
-  by only assuming that $M$ is nonempty.
+  by only assuming that $M$ is non-empty.
 \item We moved the requirement $M(\mathtt{nat})$ to the locale

### 2019-01-09 LICS/isabelle.tex
varias observaciones de Manu

@@
 in this theory: \tyo, the object-Booleans, and \tyi,
-sets. It must be insisted that the types are defined axiomatically, not
-recursively. That is, although there are constants and functions that
-generate elements of both types, neither of them is 
-\emph{initial}, in the sense that they are not the least types
-obtained by combining the given constants and operations. This will
+sets. 
+%% It must be insisted that the types are defined axiomatically, not
+%% recursively. That is, although there are constants and functions that
+%% generate elements of both types, neither of them is 
+%% \emph{initial}, in the sense that they are not the least types
+%% obtained by combining the given constants and operations. 
+It must be observed that predicates (functions with arguments of
+type \tyi{} with values in \tyo) do not correspond to first-order formulas;
+in particular, those are not recursively constructed.
+This will
 have concrete consequences in our strategy to approach the
@@
 to be used in proving several theorems, as in usual mathematical
-practice.
+practice. Furthermore, locales can be organized in hierarchies.
 

### 2019-01-09 LICS/proof-of-separation.tex
varias observaciones de Manu

@@
 passing that it is possible to  abbreviate expressions in Isabelle by
-the use of the  \textbf{let} statement or the \textbf{is} qualifier,
+the use of \textbf{let} statements or  \textbf{is} qualifiers,
 and metavariables (whose

### 2019-01-09 LICS/conclusion_m.tex
Paulson tambien reflexiona
Neeman sobre reflexión. No me convence dónde lo puse

@@
 The Replacement Axiom, however, requires some more work to be
-done. Both
-in Kunen and in other presentations (e.g. \cite{neeman-course}) it
+done. In Kunen it
 requires a relativized version (i.e., showing that it holds for $M$)
@@
 formulation of the Foundation Axiom states that the universe of sets
-can be decomposed in a transfinite hierarchy of sets:
+can be decomposed in a transfinite, cumulative hierarchy of sets:
 \begin{theorem}
@@
 hints at a proof by induction on formulas, and hence it must be shown
-internally.
+internally. It is to be noted that Paulson
+\cite{DBLP:conf/cade/Paulson02} also formalized the
+Reflection principle in Isabelle/ZF, but it is not clear if the
+relativized version follows directly from it. (It may be possible to
+sidestep Reflection, since in Neeman \cite{neeman-course}, only the
+relativization of the cumulative hierarchy is needed; nevertheless, it
+is a nontrivial task.)
 

### 2019-01-09 LICS/abstract.tex
redacción

@@
   definitions by recursion on well-founded  relations, 
-  and sharpened
-  of the hypotheses in his development of relativization and
+  and sharpened hypotheses in his development of relativization and
   absoluteness.

### 2019-01-09 LICS/appendix.tex
otra vez apéndice en ambos formatos
aireando descripción
minor

@@
 
-\begin{description}
+\begin{description}[itemsep=1.5pt]
 \item[\texttt{Nat\_Miscelanea}]Miscelaneous results for naturals, mostly
@@
 \item[\texttt{Recursion\_Thms}] Enhancements about recursively defined
+%\newpage
   functions.
 \item[\texttt{Forcing\_Notions}] Definition of Posets with maximal
-  element, filters, dense sets. Proof of Rasiowa-Sikorski.
+  element, filters, dense sets. Proof of the Rasiowa-Sikorski Lemma.
 \item[\texttt{Forcing\_Data}] Definition of the locales:
@@
 \item[\texttt{Names}] Definitions of $\chk$, $\val$, and the generic
-  extension. Varios results about them.
+  extension. Various results about them.
 \item[\texttt{Forcing\_Theorems}] Specification of fundamental
@@
   \begin{center}
-    \includegraphics[height=14.8cm]{session_graph_colored.pdf}
+    \includegraphics[height=14cm]{session_graph_colored.pdf}
   \end{center}

### 2019-01-10 LICS/fix_biblio.sh
facilidades para envio de arXiv

@@ -1,0 +1,4 @@
+#!/bin/bash
+sed -i -- s@\bibliographystyle{mi-estilo-else}@@g arXiv_tmp/separation_arxiv.tex
+sed -i -- s@\bibliography{../LSFA/citados}@\input{separation_arxiv.bbl}@g arXiv_tmp/separation_arxiv.tex
+

### 2019-01-10 LICS/kill_comments.sh

@@ -1,0 +1,2 @@
+#!/bin/bash
+for x in $*; do ./stripcomments.pl $x > arXiv_tmp/$x; done

### 2019-01-10 LICS/isabelle.tex
saco advertencia gracias a Miguel
barras out y link a web de Miguel

@@
 
-\medskip
-\fbox{Portar a \textbf{Isabelle2018} antes de enviar.}
-
-\medskip
-
-The source code is written for the 2018 version of Isabelle (with
+The source code is written for the 2018 version of Isabelle and can be downloaded
+from 
+\ifIEEE
+\begin{center}
+\texttt{https://cs.famaf.unc.edu.ar/\~{}mpagano/forcing/}
+\end{center}
+\fi
+\ifarXiv
+\url{https://cs.famaf.unc.edu.ar/~mpagano/forcing/}
+\fi
+(with
 minor modifications, it can be run in Isabelle2016-1). Most of it is

### 2019-01-10 LICS/conclusion_m.tex
barras out y link a web de Miguel
ispell... LSFA cayó en la volteada

@@
 \begin{theorem}[Reflection Principle]\label{th:reflection-principle}
-  For every finite $\Phi\sbq\ZF$, $\ZF$ proves: ``There exists
+  For every finite $\Phi\sbq\ZF$, $\ZF$ proves: ``There exist
   unboundedly many $\al$ such that $V_\al\models \Phi$.''
@@
 $\phi$ involved depends on the specific instance of Replacement being
-proved). Other formalizations, v.g.\ that of Barras
-\cite{barras-habilit} and Kaiser \cite{kaiser-set-theory-coq}, take
-advantage of a better articulation between the metatheory (Coq in both
-cases) and the set theory. An alternative road would be to redevelop
+proved). 
+%% Other formalizations, v.g.\ that of Barras
+%% \cite{barras-habilit} and Kaiser \cite{kaiser-set-theory-coq}, take
+%% advantage of a better articulation between the metatheory (Coq in both
+%% cases) and the set theory.
+
+An alternative road to internalization would be to redevelop
 absoluteness 

### 2019-01-10 LICS/recursion.tex
Cambio pavo por espacios
spacing
le saco * para uniformizar estilo
ispell... LSFA cayó en la volteada

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
-\subsection*{Recursion and values of names}
+\subsection{Recursion and values of names}
 
@@
 \ \ Hv\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}i{\isasymRightarrow}i{\isasymRightarrow}i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
-\ \ {\isachardoublequoteopen}Hv{\isacharparenleft}G{\isacharcomma}y{\isacharcomma}f{\isacharparenright}\ {\isacharequal}{\isacharequal}\ {\isacharbraceleft}\ f{\isacharbackquote}x\ {\isachardot}{\isachardot}\ x{\isasymin}\ domain{\isacharparenleft}y{\isacharparenright}{\isacharcomma}\ {\isasymexists}p{\isasymin}P{\isachardot}\ {\isacharless}x{\isacharcomma}p{\isachargreater}\ {\isasymin}\ y\ {\isasymand}\ p\ {\isasymin}\ G\ {\isacharbraceright}{\isachardoublequoteclose}
+\ \ {\isachardoublequoteopen}Hv{\isacharparenleft}G{\isacharcomma}y{\isacharcomma}f{\isacharparenright}\ {\isacharequal}{\isacharequal}\ {\isacharbraceleft}f{\isacharbackquote}x\ {\isachardot}{\isachardot}\ x{\isasymin}domain{\isacharparenleft}y{\isacharparenright}{\isacharcomma}\ {\isasymexists}p{\isasymin}P{\isachardot}\ {\isacharless}x{\isacharcomma}p{\isachargreater}\ {\isasymin}\ y\ {\isasymand}\ p\ {\isasymin}\ G\ {\isacharbraceright}{\isachardoublequoteclose}
 \end{isabelle}
@@
 \end{equation*}
-It can be easily proved by $\in$-induction that $\val(G,\chk(x))=x$.
+An easy $\in$-induction shows $\val(G,\chk(x))=x$.
 But to conclude $M\subseteq M[G]$ one also needs to have
@@
 assumptions involved in those locales; as we mentioned, Paulson's work
-is mostly done externally, i.e. the assumptions are instances
+is mostly done \emph{externally}, i.e. the assumptions are instances
 of Separation and Replacement where the predicates and functions are

### 2019-01-10 LICS/appendix.tex
ispell... LSFA cayó en la volteada

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-In this appendix we succintly describe the contents of each file. We
+In this appendix we succinctly describe the contents of each file. We
 include in 
@@
 \begin{description}[itemsep=1.5pt]
-\item[\texttt{Nat\_Miscelanea}]Miscelaneous results for naturals, mostly
+\item[\texttt{Nat\_Miscelanea}]Miscellaneous results for naturals, mostly
   needed for renamings.

### 2019-01-10 LICS/isabelle.tex

@@
 goal is that the resulting text, a \emph{proof document}, can be
-understood without the need of re-playing it viewing the proof state
+understood without the need of replaying it and viewing the proof state
 at each point.

### 2019-01-10 LICS/renaming.tex
fixing typo the -> that

@@
 a function \isa{ren} that associates to each formula an internalized
-function that can be later applied to suitable arguments. Notice the
+function that can be later applied to suitable arguments. Notice that
 Paulson used \isa{Nand} because it is more economical.
@@
 to prove that the renaming is a function. Besides some auxiliary
-results (the application of renaming to suitable arguments yields a
-formula), the main result corresponding to Lemma~\ref{lem:renaming}
-is:
+results (for example that the application of renaming to suitable
+arguments yields a formula), the main result corresponding to
+Lemma~\ref{lem:renaming} is:
 \begin{isabelle}

### 2019-01-13 LICS/hacer_diff.sh
version 2.0 del diff

@@ -1,0 +1,34 @@
+#!/bin/bash
+DIFF_DIR=diff_dir
+PREFIX=LICS
+OLD_DIR=$DIFF_DIR/$PREFIX"_old"
+NEW_DIR=$DIFF_DIR/$PREFIX"_new"
+OLDSHA=73ed3222f1edb01dd260f7ea7f9bfe4acb785fa5
+MAIN=Separation_In_MG
+
+mkdir $DIFF_DIR
+rm -rf $DIFF_DIR/*
+mkdir $OLD_DIR $NEW_DIR
+git archive $OLDSHA | tar -x -C $OLD_DIR
+cp * $NEW_DIR
+sed -i -- 's@\\appendix\[A short overview of our development\]@@g' $OLD_DIR/$MAIN.tex
+sed -i -- 's@\\appendix\[A short overview of our development\]@@g' $NEW_DIR/$MAIN.tex
+sed -i -- 's@\\input{appendix}@@g' $OLD_DIR/$MAIN.tex
+sed -i -- 's@\\input{appendix}@@g' $NEW_DIR/$MAIN.tex
+sed -i -- 's@\\newpage\n\\onecolumn@@g' $OLD_DIR/$MAIN.tex
+sed -i -- 's@\\newpage\n\\onecolumn@@g' $NEW_DIR/$MAIN.tex
+cd $DIFF_DIR
+latexdiff --flatten $PREFIX"_old"/$MAIN.tex $PREFIX"_new"/$MAIN.tex > $PREFIX"_new"/diff.tex
+cd $PREFIX"_new"/
+sed -i -- 's/Local Variables://g' diff.tex
+sed -i -- 's/bibliographystyle{entcs}/bibliographystyle{mi-estilo-else}/g' diff.tex
+sed -i -- 's@bibliography{../LSFA/citados}@bibliography{../../../LSFA/citados}@g' diff.tex
+# cp ../../mi-estilo-else.bst .
+# cp diff.tex diff.tmp
+# cat ../../header-of-diff.tex > diff.tex
+# tail -n +67 diff.tmp >> diff.tex
+pdflatex \\nonstopmode\\input diff.tex
+pdflatex \\nonstopmode\\input diff.tex
+bibtex diff.aux
+pdflatex \\nonstopmode\\input diff.tex
+pdflatex \\nonstopmode\\input diff.tex

### 2019-01-13 LICS/hacer_diff.sh
diff 2.1

@@ -6,17 +6,25 @@ NEW_DIR=$DIFF_DIR/$PREFIX"_new"
 OLDSHA=73ed3222f1edb01dd260f7ea7f9bfe4acb785fa5
 MAIN=Separation_In_MG
 
+function clear_appendix {
+    sed -i -- 's@\\appendix\[A short overview of our development\]@@g' $1
+    sed -i -- 's@\\input{appendix}@@g' $1
+    sed -i -- 's@\\newpage\n\\onecolumn@@g' $1
+}
+
 mkdir $DIFF_DIR
 rm -rf $DIFF_DIR/*
 mkdir $OLD_DIR $NEW_DIR
 git archive $OLDSHA | tar -x -C $OLD_DIR
 cp * $NEW_DIR
-sed -i -- 's@\\appendix\[A short overview of our development\]@@g' $OLD_DIR/$MAIN.tex
-sed -i -- 's@\\appendix\[A short overview of our development\]@@g' $NEW_DIR/$MAIN.tex
-sed -i -- 's@\\input{appendix}@@g' $OLD_DIR/$MAIN.tex
-sed -i -- 's@\\input{appendix}@@g' $NEW_DIR/$MAIN.tex
-sed -i -- 's@\\newpage\n\\onecolumn@@g' $OLD_DIR/$MAIN.tex
-sed -i -- 's@\\newpage\n\\onecolumn@@g' $NEW_DIR/$MAIN.tex
+clear_appendix $OLD_DIR/$MAIN.tex
+clear_appendix $NEW_DIR/$MAIN.tex
+# sed -i -- 's@\\appendix\[A short overview of our development\]@@g' $OLD_DIR/$MAIN.tex
+# sed -i -- 's@\\appendix\[A short overview of our development\]@@g' $NEW_DIR/$MAIN.tex
+# sed -i -- 's@\\input{appendix}@@g' $OLD_DIR/$MAIN.tex
+# sed -i -- 's@\\input{appendix}@@g' $NEW_DIR/$MAIN.tex
+# sed -i -- 's@\\newpage\n\\onecolumn@@g' $OLD_DIR/$MAIN.tex
+# sed -i -- 's@\\newpage\n\\onecolumn@@g' $NEW_DIR/$MAIN.tex
 cd $DIFF_DIR
 latexdiff --flatten $PREFIX"_old"/$MAIN.tex $PREFIX"_new"/$MAIN.tex > $PREFIX"_new"/diff.tex
 cd $PREFIX"_new"/

### 2019-01-13 LICS/hacer_diff.sh
sacando código al pedo. Dejo comentado el SHA posta

@@ -3,7 +3,10 @@ DIFF_DIR=diff_dir
 PREFIX=LICS
 OLD_DIR=$DIFF_DIR/$PREFIX"_old"
 NEW_DIR=$DIFF_DIR/$PREFIX"_new"
+## SHA for testing purposes
 OLDSHA=73ed3222f1edb01dd260f7ea7f9bfe4acb785fa5
+## SHA of submitted version (the real thing)
+# OLDSHA=457792cabb9db838fe26a23a9b70d735423a45c0
 MAIN=Separation_In_MG
 
 function clear_appendix {
@@ -19,22 +22,11 @@ git archive $OLDSHA | tar -x -C $OLD_DIR
 cp * $NEW_DIR
 clear_appendix $OLD_DIR/$MAIN.tex
 clear_appendix $NEW_DIR/$MAIN.tex
-# sed -i -- 's@\\appendix\[A short overview of our development\]@@g' $OLD_DIR/$MAIN.tex
-# sed -i -- 's@\\appendix\[A short overview of our development\]@@g' $NEW_DIR/$MAIN.tex
-# sed -i -- 's@\\input{appendix}@@g' $OLD_DIR/$MAIN.tex
-# sed -i -- 's@\\input{appendix}@@g' $NEW_DIR/$MAIN.tex
-# sed -i -- 's@\\newpage\n\\onecolumn@@g' $OLD_DIR/$MAIN.tex
-# sed -i -- 's@\\newpage\n\\onecolumn@@g' $NEW_DIR/$MAIN.tex
 cd $DIFF_DIR
 latexdiff --flatten $PREFIX"_old"/$MAIN.tex $PREFIX"_new"/$MAIN.tex > $PREFIX"_new"/diff.tex
 cd $PREFIX"_new"/
 sed -i -- 's/Local Variables://g' diff.tex
-sed -i -- 's/bibliographystyle{entcs}/bibliographystyle{mi-estilo-else}/g' diff.tex
 sed -i -- 's@bibliography{../LSFA/citados}@bibliography{../../../LSFA/citados}@g' diff.tex
-# cp ../../mi-estilo-else.bst .
-# cp diff.tex diff.tmp
-# cat ../../header-of-diff.tex > diff.tex
-# tail -n +67 diff.tmp >> diff.tex
 pdflatex \\nonstopmode\\input diff.tex
 pdflatex \\nonstopmode\\input diff.tex
 bibtex diff.aux

### 2019-01-17 LSFA/names.tex
Atendiendo las sugerencias de referees para post-proceedings.

@@
 \label{sec:names}
-We formalize the function $\val$ that allows to
+We formalize the function $\val$ that allows us to
 construct the elements of the generic extension $M[G]$ from elements

### 2019-01-17 LSFA/related.tex
Atendiendo las sugerencias de referees para post-proceedings.

@@
 instance, to have a detailed account of which axioms are necessary to
-define and use forcing. Finally, we support the vision that a growing
+define and use forcing. Moreover, we support the vision that a growing
 corpus of formalized mathematics can be a useful library for the
@@
 logic, Isabelle/HOL. This framework is very active, and as a
-consequence more automated tools are available. Isabelle/HOL has 
-basic chapters on set theory. One of those, by Steven Obua, proceeds up to
-well founded relations and provides translations between types in HOL
-(for instance \isatt{nat}) to  sets  (elements of type
-\isatt{ZF}). Another one, by A.~Popescu and D.~Traytel, reaches
+consequence more automated tools are available. Isabelle/HOL has basic
+chapters on set theory. One of those, by Steven
+Obua~\cite{DBLP:conf/ictac/Obua06}, proceeds up to well founded
+relations and provides translations between types in HOL (for instance
+\isatt{nat}) to sets (elements of type \isatt{ZF}). Another one, by
+A.~Popescu and D.~Traytel~\cite{DBLP:conf/itp/Blanchette0T14}, reaches
 cardinal arithmetic. This is fairly limited for our purposes.

### 2019-01-17 LSFA/rs.tex
comentario que quedaba raro dado el nuevo contexto.
Atendiendo las sugerencias de referees para post-proceedings.

@@
 
-We need a further, ``diagonal'' version of $\DC$  to prove
+We need a further ``diagonal'' version of $\DC$  to prove
 Rasiowa-Sikorski. That is, if the assumption holds for a sequence of
@@
 %
-As these two proofs are interesting, we have proved them using the
-Isar proof language~\cite{DBLP:conf/tphol/Wenzel99}.
+These two proofs have been implemented  using the
+Isar proof language.
 

### 2019-03-07 LICS/rebuttal.txt
feedback con comentarios para empezar

> ----------------------- REVIEW 1 ---------------------
> PAPER: 165
> TITLE: Mechanization of Separation in Generic Extensions
> AUTHORS: Emmanuel Gunther, Miguel Pagano and Pedro Sánchez Terraf
>
> Overall evaluation: -2 (reject)
>
> ----------- Overall evaluation -----------
> ========== Summary =============
>
>
> The paper "Mechanization of Separation in Generic Extensions" presents a formalization in Isabelle/ZF of Cohen's forcing. To this end, they build on Paulson's ZF theory in Isabelle, which they partly modify. A prequel of this work was already presented at LSFA'18 by the same authors. In this paper, they focus on the proof that the extended forcing model M[G] satisfies ZF axioms, and especially the Separation axiom which turns out to be more problematic. After a general introduction to forcing and existing works (almost none)

No entiendo bien. ¿Está confirmando que no hay trabajos en esto? ¿O sólo repitiendo lo que decimos nosotros?
> on its formalization, the paper essentially consists in a survey of the project and of the techniques used in the formalization, before a particular focus on the proof of validity for Separation and Powerset.
>
>
>
> ========== Review =============
>
> While I am convinced of the interest of formalizations and aware of their potential difficulty, I find myself bothered with the nature of this paper. Indeed, it mainly describes the work of the author in order to formalize the theory of Cohen's forcing, and as such consists of an alternation between Isabelle code and its description. Again, I am sincerely in favor of works like this one, but insofar as it does not bring any thing new from a theoretical/logical point of view, I am afraid that it falls out of LICS scope. Conferences on the topic (e.g., ITP) would have been, I believe, much more adapted for presenting such a work. In particular, I cannot see how it can be of interest for people that are not interested in formalization: most of the paper is devoted to technicalities in implementing the formalization. For instance, these (subjectively chosen) extracts highlight the fact that, in my opinion, the paper is already more directed to the specific community around formalized mathematics and proof assistants than to the broad audience of LICS:
> - "Since the schema for recursively defined functions does not allow parameters, we are forced to return a function of three arguments (n,m,f )."
> - "Nevertheless, in order to have a neater presentation, we have stripped off several sections concerning L from the theories L_axioms and Internalize, and we merged them to form the new file Internalizations."
> - "We note in passing that it is possible to abbreviate expressions in Isabelle by the use of let statements or is qualifiers, and metavariables (whose identifiers start with a question mark)."

66% de responsabilidad mía :'D
>
> Moreover, I would have expected at least a bit of discussion on the benefits/drawbacks of choosing Isabelle/ZF over other possible frameworks (Coq, Mizar, Isabelle/HOL, HOL Light, ... and the corresponding underlying logics).
Insistir que el overhead era tremendo si no usábamos Isabelle/ZF.
>
> Last, I find the paper really hard to read without being familiar with Isabelle/ZF and/or the specific project. As I said, the paper including a lot of code lines, it requires to process a lot of notations/declarations (for instance f'x or sats(M,φ,[a,b,c])), and sometimes complicate the readings. In particular, the whole renaming process makes some notations more computer- than human-readable. For instance, in p8, the reader has to remember that x_0 ∈ x_2 is transformed into x_4 ∈ x_6 when it is inside the forces(_) function, because the latter have 4 more free variables and thus imposes a name translations.

Otra culpa mía... Pero bueno, hay que ver si esta cosa aislada es tan relevante (estoy pensando en ITP o donde carajo lo mandemos).
>
> To wrap it up, despite my personal interest for forcing and formalization of mathematics (and hereby for the present work), I cannot recommend the acceptation of this paper at LICS. Nonetheless, I strongly advise the authors to improve the presentation and to consider applying to a conference or journal more specifically focused on formalized mathematics.

Dale... Me parece que nos bocha pero con onda.
>
>
> ========= Comments ============
>
> Here follows a list of specific comments, typos and so on.
> Globally, I would suggest to consider doing a trade-off between sticking to Isabelle code and making the paper more readable. In particular, I would take a step back from the formalization and try to avoid all the part describing the code without further informations (e.g., "in the file Bli there is 30 lines of this").

Me siento nuevamente aludido... Igual creo que en una conferencia especializada capaz sí sea apropiado.
> Also, make sure that no notation is used without having been previously introduced in the text (e.g., ed, edrel, etc.).
Esto es importante
>
> *p.1*
>
> - "strategy: Since" -> since
> - "the fruitful use of forcing techniques to extend the Curry-Howard isomorphism to classical axioms" -> this claim is misleading. The two papers you cite show how to express forcing within Krivine realizability, a framework which is already classical. Therefore, in these papers, forcing techniques are *not* used to extend CH to classical logic (which has to do with control operators, not forcing). They rather unveil the computational content of forcing within Krivine realizability, just like Jaber et al. show how to define it within type theory [14,15].
>
Llamada para el Dr. Pagano...
>
> *p2*
>
> - "a version of Rasiowa-Sikorski" -> ...'s theorem?
> - "typing assumtions" -> assumptions
> - sats:: "i ⇒ i ⇒ i ⇒ o" -> without exploring the code, I find it hard to understand the type of sats
>
>
> *p3*
>
> - ∈ restricted to M -> this suggests that it is defined outside of M, right? Isn't this strange?

Hay que decir que ∈ es el mismo para todos los modelos (el externo).
> - "an countable" -> a countable
> - "in this stage" -> at this stage?
> - relative quantification -> relativized? (it seems more natural to me and you speak of "relativization" all the time afterwards. Also, I would parenthesize ∀x.(C(x) → φ) to make sure it is read correctly
> - induction on terms of type o is not available -> is this related to the chosen framework (Isabelle/ZF) or is it more general?

Aclarar
> - "It is also possible define" -> to define
> - "it is also possible", "it can be shown" -> if you did it, say it as such, otherwise I am not sure that this is informative or useful (but I might be missing something)

Decir que lo hicimos pero no es parte de esta submission
> - think of succ(n) as .. -> I have been a bit confused by your explanation, because I didn't understand at first the relation of succ with the rest of the paragraph and I was thinking that you wanted the reader to think to the function succ itself, not as a renaming...
> - block starting at "consts ren" -> I am not sure of the utility of giving the four examples in the paper...

¿Ejemplos? Es la definición.
>
>
>
> *p4*
>
> - develop ML tools -> good idea indeed, seems boring to do manually otherwise
> - a preorder with top -> is top 1 in your tuple?

¿Medio boludo esto o me parece mal?
> - its name \dot a of M -> \dot a in M (of \dot a \in M)
> - "As is well-known ..." -> come a bit out of nowhere
> - "x ed y ⇐⇒ ..." -> what is ed? Isn't it a definition rather than an equivalence?
> - edrel, edclose -> we don't know what they are?
> - (r^+)-''{a} -> is the reader expected to be able to parse this?

Jajaja, ta bien.
>
>
> *p5*
>
> - 2) Originally,... 4) ... -> again, why is it informative for external reader? (by this, I mean not an Isabelle/ZF developer)
> - Nervetheless, in order to ... -> again, it looks more like a description of your technical process than like something useful for the reader.

Qué orsai estamos, che.
>
>
> *p6*
>
> - as it is defined using Separation, so one needsx -> remove "so"
> - Union is absolute for any transitive class -> is it well-known ?

YES!
>
>
> *p7*
>
> - "We note in passing that it is possible to abbreviate expression in Isabelle..." -> I don't want to rude, but I really don't see how this could be a relevant information in a LICS paper...

:'(
>
> *p8*
>
> - as said above, the use of variable names x_0, x_1, etc that are automatically translated in forces makes it hard to read...
> - "two instances" -> which are?
> - using absoluteness -> the absoluteness of?
> - in the last block of code, the reader is left guessing what p,l,o,χ are
>
> *p9*
>
> - you should give a reference for Theorem 8 & 9
>
> ----------------------- REVIEW 2 ---------------------
> PAPER: 165
> TITLE: Mechanization of Separation in Generic Extensions
> AUTHORS: Emmanuel Gunther, Miguel Pagano and Pedro Sánchez Terraf
>
> Overall evaluation: -2 (reject)
>
> ----------- Overall evaluation -----------
> Summary:
> This paper presents a formalization in Isabelle of some axioms of Set theory of a generic extension of a model of Set theory using forcing. The main axiom, according to the authors, that has been shown to hold in a generic extension is Separation. The paper starts by presenting very briefly Isabelle, axioms of Set theory, mentioning the representation of first-order formulas as sets done by Paulson and
> discuss their presentation of renaming of formulas. Then the authors explain briefly the generic extension of a model of Set theory using forcing, some modifications they have done on the development of Paulson of “ZF-Constructible”. The authors explain how to prove the “easy axioms” first. Then they present the fundamental lemmas of forcing which are given as axioms in the formalization. After that, they show that Separation holds in the generic extension providing the fundamental lemmas holds. Finally, they conclude by discussing how the main missing axiom, i.e. Replacement, could be proven.
>
> Evaluation:
> Overall, I can’t see more in this paper than a formalization exercise, following the library developed by Paulson and the textbook of Kunen. The reader does not get any insight on forcing in Set theory, and very few on how to formalize theorems on Set theoretic tools in Isabelle. For those reasons,
> I suggest to the authors to submit their work to a more appropriate venue such as CPP or ITP, and focus more on the insight of the formalization.

Sí, dale, vamos a ITP... Basta...
> I think this paper is very hard to follow for a general audience that does not know about Cohen forcing and/or Isabelle. The paper interleaves short explanations on Set theory and forcing with technical remarks on the Isabelle formalization in a rather unstructured way, which makes it even harder to read.
>
> Some Remarks:
>
> - “We think that this project constitutes an interesting test-case for the current technology of  formalization of mathematics, in particular for the need of handling several layers of reasoning. “
> Not really, it is only formalization of set theory, which is a very little part of what is usually  considered when one say formalizing mathematics. For instance, the work of the Mathematical Component  Library in Coq is more in this flavor of multiple layers of definitions/reasoning. - “It might be a little surprising the lack of formalizations of forcing and generic extensions”
> It is indeed disappointing, it seems that the author have only partially formalize that given a good forcing translation,  the induced generic model is indeed a model of Set theory. But nothing is done on concrete translations which may be the hard part in Isabelle.
>
> Questions to the authors:
> “We are only able to prove that the generic extension satisfies Separation for any particular number of parameters, but not in general.”
> This looks like a strong limitation, can you comment on this.

Hay que decir que no (de hecho con 2 parámetros estamos haciendo todo), pero tendría que sentarme un rato a confirmar de nuevo esto.
>
> ----------------------- REVIEW 3 ---------------------
> PAPER: 165
> TITLE: Mechanization of Separation in Generic Extensions
> AUTHORS: Emmanuel Gunther, Miguel Pagano and Pedro Sánchez Terraf
>
> Overall evaluation: -2 (reject)
>
> ----------- Overall evaluation -----------
> Summary:
>
> This paper presents a formalization of the separation axiom from set theory using the forcing library that has already been built in Isabelle/ZF by Paulson. The main contribution of the paper is a formalization of parts of Kunen's textbook in Isabelle/ZF.
>
> Evaluation:
>
> The paper is decently written. As someone whose poison of choice is Coq, I nevertheless found the paper fairly easy to follow since it doesn't use any complicated Isabelle-specific machinery.
>
> Despite this, I regret to say that I cannot recommend acceptance at LICS.
>
> My biggest complaint is that this work does not seem to provide the kind of fundamental contributions to logic that LICS is looking for. There is no new mathematics here, and this isn't even the first formalization of separation.

WTF!? Acá hay que preguntar cuál es la otra.
>
> I think a conference like CPP or ITP is a much better fit for this kind of paper that is more of a documentation of a formalization project than a description of new results.
>
> I am not giving a more thorough review because I think the program committees of these other conferences would be better equipped to critique a formalization artifact.

O sea, no tenés la más puta idea.
>
> ----------------------- REVIEW 4 ---------------------
> PAPER: 165
> TITLE: Mechanization of Separation in Generic Extensions
> AUTHORS: Emmanuel Gunther, Miguel Pagano and Pedro Sánchez Terraf
>
> Overall evaluation: -1 (weak reject)

Oh, dadivoso réferi...
>
> ----------- Overall evaluation -----------
> This paper takes place in the long term goal of formalizing the method
> of forcing *in set theory*, such as introduced by Cohen in the sixties
> to prove the relative consistency of the negation of the continuum
> hypothesis. (As opposed to more recent works cited by the authors, in
> which the method of forcing has been adapted to various
> type-theoretical frameworks, for instance to study its computational
> contents via the proofs-as-programs correspondence)
>
> Reusing parts of an earlier formalization (by Paulson) of Gödel's
> method of constructible sets in the proof-assistant Isabelle, the
> authors started a formalization of Cohen's method of forcing (in the
> same proof-assistant) based on Kunen's presentation of forcing in his
> famous textbook on set theory. In this paper, the authors present the
> formalization of the soundness proof for the Separation scheme and for
> the Powerset axiom. (The treatment of the Extensionality, Regularity,
> Pairing, Union and Infinity axioms has been already published in an
> earlier paper by the same authors, whereas the Replacement scheme is
> still to be treated.)

Acá la pifió bárbaro. Después repite esto que todos los axiomas fáciles están en otro paper. ¿Cómo llegó a tener esa impresión?
>
> Together with the authors, I consider that the achievement of their
> long term goal (which is to formalize the *set-theoretic* method of
> forcing) would constitute a milestone in the formalization of
> mathematics. (After all, Cohen was awarded the Fields medal for this
> beautiful piece of mathematics.) Moreover, the paper is technically
> sound and the methodology followed by the authors is clearly
> explained.

Lástima que este réferi sea el/la único/a que lo vea...
>
> Nevertheless, and despite its real interest, the paper suffers from
> important weaknesses which I will now address.
>
> The more important point is that the "increment" presented in this
> paper is (IMO) too little for being published in a top conference
> such as LICS. Indeed, the paper only deals with the soundness of
> Separation and Powerset in generic extensions. The formalization
> of the soundness of Extensionality, Regularity, Pairing, Union and
> Infinity has been published before,

Otra vez esta pelotudez
> and the soundness of Replacement
> has not been formalized yet. More important: the authors' work relies
> on some fundamental properties of forcing (relating the notion of
> satisfaction in the generic extension with the notion of satisfaction
> in the ground model) that have not been proven yet. (In their formal
> development, these properties are only axiomatized.)
>
> The other point deals with the formalization choices of the authors,
> which are never really discussed. It seems to me that the authors
> blindly followed Kunen's presentation of forcing (based on countable
> transitive models and the Rasiowa-Sikorski lemma), without comparing
> it with other presentations that would be (IMO) *much more suitable*
> to a formalization of forcing in a proof-assistant, especially in the
> perspective of deriving relative consistency proofs from this work.
> (Which is proof-theoretically impossible from Kunen's approach, as I
> explain below in my detailed comments.)

Raro esto. Kunen tiene una sección específica sobre cómo se puede hacer esto.
> Moreover, the choice of the
> authors to follow Kunen's approach induces many complications in their
> formalization (for instance: the use of internalized formulas) that
> could have been avoided by taking another approach (for instance,
> the one presented by Jech in his textbook on Set theory). Perhaps
> these complications explain why the authors had to split the
> presentation of their formalization in several papers.
>
> For all theses reasons - and despite the merits of the authors' work -
> I cannot recommend this paper for publication in the LICS conference.
> Nevertheless, I strongly encourage the authors to pursue their
> (valuable) effort, but also to consider other formalization choices
> that would be (IMO) more suited to their long-term goal, which is to
> get relative consistency proofs.

Acá estaría bueno delimitar más los goles. Efectivamente, los móviles iniciales son las pruebas de consistencia relativa, pero lo que estamos haciendo está más cerca del desarrollo de la maquinaria matemática del forcing que de la correspondiente metamatemática de las pruebas de consistencia. En ese sentido, capaz habría que aclarar más este punto en próximas versiones del paper.
>
>
> Detailed comments and suggestions for improvement
> -------------------------------------------------
>
> p. 2, last paragraph of the 1st column. "It must be observed that
> predicates (...) do not correspond to first-order formulas". Here,
> you should be more precise, to help the reader. As far as I
> understand, you mean: "It must be observed that Isabelle higher-order
> predicates (...) do not correspond to ZF first-order formulas."
>
> p. 2, second column: internalized formulas. Here you should justify
> why you use internalized formulas, that is: formulas as sets.
>
> IMO, it would be more natural to inductively define an Isabelle
> type "f" of first-order formulas (or codes of first-order formulas),
> together with a decoding function that would map each code of a
> first-order formula to a real Isabelle formula. So that the image
> of this map would capture exactly the class you are interested in.
> (At least, this would be the standard way of doing in Coq or Agda.)

Bueno, se ve que no le quedó claro que esto no se puede hacer. Hay que insistir.
> Moreover, such a definition would reflect the fact that, in model
> theory, formulas are traditionally considered as external objects:
> they live in the ambient theory, outside all models.
>
> On the other hand, I agree that "formulas as sets" are useful to
> define the notion of constructible set, and the corresponding
> constructible universe. But the point is that in forcing, we
> (normally) never need such internalized formulas! The theory
> of forcing is independent from the theory of constructible sets,
> and this should appear in a formalization. (Although both theories
> can be nicely combined in particular situations.)

Acá otro pifie, pero es por algo que no aclaramos nosotros: nuestra presentación no depende de constructibilidad per se, sino de relativización y absolutez, que es de hecho un prerrequisito para ambas teorías.
>
> So why did you consider internalized formulas, and not external
> formulas, inductively defined "outside" as explained above?
> I understand you want to reuse parts of the (previous) formalization
> of the constructible universe, but this is (scientifically) a very
> weak argument. IMO, considering internalized formulas for defining
> forcing is just introducing unnecessary complication.

Bueno, hacer todo desde cero es una complicación innecesaria para nosotros (!).
>
> Moreover, considering internalized formulas (that is: formulas as
> sets) raises another problem, which is that there might be more
> internalized formulas than first-order formulas (such as seen
> outside). Indeed, your axiomatized type "i" of ZF sets is nothing
> but a (hypothetical) model of ZF inside Isabelle. But such a model
> can be non-standard, in the sense that it might contain more natural
> numbers (and thus more formulas-as-sets) than the Isabelle type of
> natural numbers. Note that this problem is harmless in practice, but
> it is worth mentioning it.

Todo está pasando dentro de "i", así que no es relevante esta observación. Hay que discutirlo un poco más.
>
> p. 2, last paragraph of 2nd column. "Forcing is a technique to extend
> very special kind of models...". Actually, not. There are several
> ways to construct forcing models of ZF, and the one you consider in
> your paper (taken from Kunen's book) is only *one* of them.
>
> The method you consider is based on countable transitive models
> (ctms) of ZF, whose main advantage is that they allow the use of
> the Rasiowa-Sikorski lemma to construct the generic filter.
> Basically, a ctm of ZF captures three properties:
>
> (1) M is a set equipped with a binary relation (to interpret
> membership) that is well-founded _from the point of view of the
> outer set-theoretic universe_ (the type "i"). This property
> implies that M fulfills the foundation axiom, but the converse
> is not true. So basically, this property says that the notion
> of well-foundedness is the same inside M and outside M (that is:
> in the outer model "i").
>
> (2) M is transitive, which means that the binary relation interpreting
> membership is membership itself (in the sense of "i"). This point
> is actually just a technical facility, to avoid distinguishing two
> membership predicates.

No sólo eso. Se está olvidando de los resultados de absolutez.
>
> (3) M is countable (from the point of view of the universe "i").
> This point is crucial to prove the Rasiowa-Sikorski lemma.
>
> As you mention it in Lemma 1, the properties (2) and (3) are
> proof-theoretically free, since once you have (1), you can easily
> enforce (2) and (3) by combining a Mostowski collapse with the
> (descending) Löwenheim-Skolem theorem.
>
> But the point is that (1) is actually very strong, and more
> important: it is *not known* to be equivalent to the consistency
> of ZF. By this, I mean that it might be the case that ZF has
> no transitive model while being consistent.

Tranquilate un poco... Esto también es irrelevante, porque sólo necesitamos un ctm de un fragmento suficientemente grande de ZFC (cosa que el Teorema de Reflexión nos da).
>
> This simple argument shows that transitive models of ZF, although
> they are very convenient from the point of view of model theory,
> *cannot* be used to prove relative consistency results about forcing
> (in the proof-theoretic sense), since there is *no known way* to
> turn an arbitrary model of ZF (such as given by the completeness
> theorem of 1st-order logic) into a transitive model.

Ídem... Sólo necesitamos un fragmento finito de esto, como lo explica Kunen en el libro nuevo.
>
> This is why other approaches have been developed in the literature.
> For instance, the construction of a Boolean-valued model of names
> inside the ground model M, and then a quotient by a suitable
> ultrafilter (see Jech's book for instance). The main advantage of
> this approach is that it can be applied to _any model_ of ZF. (In
> this case, the generic filter is directly constructed from the
> corresponding name, without the Rasiowa-Sikorski lemma.)
>
> However, the ground model M and the corresponding generic extension
> M[G] are in general *not* transitive models. But from a proof-
> theoretic point of view, this property is definitely useless!
>
> So, since you are interested in relative consistency results (as
> mentioned in the introduction of the paper), why didn't you consider
> this alternative approach, that would work with _any_ model of ZF?
> (Moreover, it seems to me that the job of formalization would have
> been easier.) IMO, the paper is dramatically missing a discussion
> about this important foundational issue (that has also a dramatic
> impact on your formalization choices).

Me gusta más el enfoque de los ctm para trabajar en general (y off-the-record, es el que más entiendo de todos), y además tratamos de seguir lo más posible una sola referencia (i.e. Kunen) para darle cohesión al trabajo. La discusión sobre la conveniencia de usar el enfoque de los modelos a valores booleanos es más interesante, pero creo que otra vez nos pone en el brete de tener que reimplementar varias cosas.
>
> Another advantage of constructing the generic extension as a quotient
> of the Boolean-valued model of names (defined as a class within M)
> would be the following. In your work, you need to distinguish three
> models:
> - the Isabelle type i of sets (the ambient model);
> - the ctm M, that is an object of type i (the ground model);
> - the generic extension M[G].
> By considering the approach I suggest, you would only need two models:
> - the type i, which would serve as a ground model M as well;
> - the generic extension M[G], that would be defined as a quotient of
> the class of names (within i = M).
>
> p. 3, Proof of Lemma 1. "an countable..." -> "a countable".
>
> p. 3, second last par. of column 1. "We are only able to prove that
> the generic extension satisfies Separation for any particular number
> of parameters, but not in general." What do you mean precisely?
> At which level lives the quantification in "any particular number of
> parameters"? A quantification over the set N (inside M and "i")?
> Or a quantification over the type of natural numbers in Isabelle?
> Please clarify this point.
Aclarar esto.
>
> p. 3, next par. "possible define" -> "possible to define".
>
> p. 4, last remark before Sect. V. "All our uses... (etc.)".
> What do you mean precisely? I am completely confused.

Tener en cuenta quizá para la próxima versión
>
> p. 4, first paragraph of Sect. V. "be its _name_ in M".
> Here you should precise what you mean here by a name.
> Do you consider a special class of names inside M?
> Or do you consider all elements of M as names?
> (Note that both options are possible.) From the discussion of
> Sect. V. A., it seems that you consider the second option, but
> you should clarify it.
>
> p. 5, second last par. before Sect. VI. "Paulson proved... (etc.)"
> This discussion is really hard to follow, due to the several levels
> that are involved.
>
> p. 6, column 2: Sect. VIII. In this section, you say that the
> fundamental properties of forcing are axiomatized (in a locale)
> rather than proved. IMO, these properties should have been proven
> first (before proving the soundness of the axioms of ZF), since they
> are at the core of the method of forcing.

A esto lo mando a una revista, no a una conferencia, GIL.
>
> p. 9, column 1, 3rd paragraph: "The Replacement Axiom [...] requires
> more work to be done. In Kunen it requires a relativized version
> [...] of the Reflection Principle." IMO, this requirement is another
> drawback of following Kunen's approach. By taking the other method I
> suggest, you would definitely not need the Reflection Principle to
> prove the soundness of Replacement.

Tendría que ver esto, no estoy seguro ya mismo. Este tipo entiende, sí, pero hay gente que entiende bien un enfoque y no el otro (bueno, a mí me pasa un poco pero al revés de este muchacho).

### 2019-03-07 LICS/rebuttal.txt
limpiando un poco

@@
-> ----------------------- REVIEW 1 ---------------------
-> PAPER: 165
-> TITLE: Mechanization of Separation in Generic Extensions
-> AUTHORS: Emmanuel Gunther, Miguel Pagano and Pedro Sánchez Terraf
->
-> Overall evaluation: -2 (reject)
->
-> ----------- Overall evaluation -----------
-> ========== Summary =============
->
->
-> The paper "Mechanization of Separation in Generic Extensions" presents a formalization in Isabelle/ZF of Cohen's forcing.
-To this end, they build on Paulson's ZF theory in Isabelle, which they partly modify.
-A prequel of this work was already presented at LSFA'18 by the same authors.
-In this paper, they focus on the proof that the extended forcing model M[G] satisfies ZF axioms, and especially the Separation axiom which turns out to be more problematic.
-After a general introduction to forcing and existing works (almost none)
+First and foremost, we would like to thank the referees for their
+careful reading and detailed criticism. We are indeed greatly
+benefited by their evaluation.
 
-No entiendo bien. ¿Está confirmando que no hay trabajos en esto? ¿O sólo repitiendo lo que decimos nosotros?
-> on its formalization, the paper essentially consists in a survey of the project and of the techniques used in the formalization, before a particular focus on the proof of validity for Separation and Powerset.
->
->
->
-> ========== Review =============
->
-> While I am convinced of the interest of formalizations and aware of their potential difficulty, I find myself bothered with the nature of this paper.
-Indeed, it mainly describes the work of the author in order to formalize the theory of Cohen's forcing, and as such consists of an alternation between Isabelle code and its description.
-Again, I am sincerely in favor of works like this one, but insofar as it does not bring any thing new from a theoretical/logical point of view, I am afraid that it falls out of LICS scope.
-Conferences on the topic (e.g., ITP) would have been, I believe, much more adapted for presenting such a work.
-In particular, I cannot see how it can be of interest for people that are not interested in formalization: most of the paper is devoted to technicalities in implementing the formalization.
-For instance, these (subjectively chosen) extracts highlight the fact that, in my opinion, the paper is already more directed to the specific community around formalized mathematics and proof assistants than to the broad audience of LICS:
-> - "Since the schema for recursively defined functions does not allow parameters, we are forced to return a function of three arguments (n,m,f )."
-> - "Nevertheless, in order to have a neater presentation, we have stripped off several sections concerning L from the theories L_axioms and Internalize, and we merged them to form the new file Internalizations."
-> - "We note in passing that it is possible to abbreviate expressions in Isabelle by the use of let statements or is qualifiers, and metavariables (whose identifiers start with a question mark)."
+----------------------- REVIEW 1 ---------------------
 
-66% de responsabilidad mía :'D
->
-> Moreover, I would have expected at least a bit of discussion on the benefits/drawbacks of choosing Isabelle/ZF over other possible frameworks (Coq, Mizar, Isabelle/HOL, HOL Light, ...
-and the corresponding underlying logics).
+> Moreover, I would have expected at least a bit of discussion on the
+  benefits/drawbacks of choosing Isabelle/ZF over other possible
+  frameworks (Coq, Mizar, Isabelle/HOL, HOL Light, ... and the
+  corresponding underlying logics).
+
 Insistir que el overhead era tremendo si no usábamos Isabelle/ZF.
->
-> Last, I find the paper really hard to read without being familiar with Isabelle/ZF and/or the specific project.
-As I said, the paper including a lot of code lines, it requires to process a lot of notations/declarations (for instance f'x or sats(M,φ,[a,b,c])), and sometimes complicate the readings.
-In particular, the whole renaming process makes some notations more computer- than human-readable.
-For instance, in p8, the reader has to remember that x_0 ∈ x_2 is transformed into x_4 ∈ x_6 when it is inside the forces(_) function, because the latter have 4 more free variables and thus imposes a name translations.
 
-Otra culpa mía... Pero bueno, hay que ver si esta cosa aislada es tan relevante (estoy pensando en ITP o donde carajo lo mandemos).
->
-> To wrap it up, despite my personal interest for forcing and formalization of mathematics (and hereby for the present work), I cannot recommend the acceptation of this paper at LICS.
-Nonetheless, I strongly advise the authors to improve the presentation and to consider applying to a conference or journal more specifically focused on formalized mathematics.
+> - "the fruitful use of forcing techniques to extend the Curry-Howard
+>isomorphism to classical axioms" -> this claim is misleading. The two
+>papers you cite show how to express forcing within Krivine
+>realizability, a framework which is already classical. Therefore, in
+>these papers, forcing techniques are *not* used to extend CH to
+>classical logic (which has to do with control operators, not
+>forcing). They rather unveil the computational content of forcing
+>within Krivine realizability, just like Jaber et al. show how to
+>define it within type theory [14,15].  Llamada para el Dr. Pagano...
 
-Dale... Me parece que nos bocha pero con onda.
->
->
-> ========= Comments ============
->
-> Here follows a list of specific comments, typos and so on.
-> Globally, I would suggest to consider doing a trade-off between sticking to Isabelle code and making the paper more readable.
-In particular, I would take a step back from the formalization and try to avoid all the part describing the code without further informations (e.g., "in the file Bli there is 30 lines of this").
-
-Me siento nuevamente aludido... Igual creo que en una conferencia especializada capaz sí sea apropiado.
-> Also, make sure that no notation is used without having been previously introduced in the text (e.g., ed, edrel, etc.).
-Esto es importante
->
-> *p.1*
->
-> - "strategy: Since" -> since
-> - "the fruitful use of forcing techniques to extend the Curry-Howard isomorphism to classical axioms" -> this claim is misleading.
-The two papers you cite show how to express forcing within Krivine realizability, a framework which is already classical.
-Therefore, in these papers, forcing techniques are *not* used to extend CH to classical logic (which has to do with control operators, not forcing).
-They rather unveil the computational content of forcing within Krivine realizability, just like Jaber et al.
-show how to define it within type theory [14,15].
->
-Llamada para el Dr. Pagano...
->
-> *p2*
->
-> - "a version of Rasiowa-Sikorski" -> ...'s theorem?
-> - "typing assumtions" -> assumptions
-> - sats:: "i ⇒ i ⇒ i ⇒ o" -> without exploring the code, I find it hard to understand the type of sats
->
->
-> *p3*
->
-> - ∈ restricted to M -> this suggests that it is defined outside of M, right? Isn't this strange?
+> - ∈ restricted to M -> this suggests that it is defined outside of
+    M, right? Isn't this strange?
 
 Hay que decir que ∈ es el mismo para todos los modelos (el externo).
-> - "an countable" -> a countable
-> - "in this stage" -> at this stage?
-> - relative quantification -> relativized?
-(it seems more natural to me and you speak of "relativization" all the time afterwards.
-Also, I would parenthesize ∀x.(C(x) → φ) to make sure it is read correctly
-> - induction on terms of type o is not available -> is this related to the chosen framework (Isabelle/ZF) or is it more general?
+
+> - induction on terms of type o is not available -> is this related
+    to the chosen framework (Isabelle/ZF) or is it more general?
 
 Aclarar
-> - "It is also possible define" -> to define
-> - "it is also possible", "it can be shown" -> if you did it, say it as such, otherwise I am not sure that this is informative or useful (but I might be missing something)
 
-Decir que lo hicimos pero no es parte de esta submission
-> - think of succ(n) as ..
--> I have been a bit confused by your explanation, because I didn't understand at first the relation of succ with the rest of the paragraph and I was thinking that you wanted the reader to think to the function succ itself, not as a renaming...
-> - block starting at "consts ren" -> I am not sure of the utility of giving the four examples in the paper...
+> - "it is also possible", "it can be shown" -> if you did it, say it
+    as such, otherwise I am not sure that this is informative or
+    useful (but I might be missing something)
 
-¿Ejemplos? Es la definición.
->
->
->
-> *p4*
->
-> - develop ML tools -> good idea indeed, seems boring to do manually otherwise
-> - a preorder with top -> is top 1 in your tuple?
-
-¿Medio boludo esto o me parece mal?
-> - its name \dot a of M -> \dot a in M (of \dot a \in M)
-> - "As is well-known ..." -> come a bit out of nowhere
-> - "x ed y ⇐⇒ ..." -> what is ed? Isn't it a definition rather than an equivalence?
-> - edrel, edclose -> we don't know what they are?
-> - (r^+)-''{a} -> is the reader expected to be able to parse this?
-
-Jajaja, ta bien.
->
->
-> *p5*
->
-> - 2) Originally,... 4) ... -> again, why is it informative for external reader? (by this, I mean not an Isabelle/ZF developer)
-> - Nervetheless, in order to ... -> again, it looks more like a description of your technical process than like something useful for the reader.
-
-Qué orsai estamos, che.
->
->
-> *p6*
->
-> - as it is defined using Separation, so one needsx -> remove "so"
-> - Union is absolute for any transitive class -> is it well-known ?
+Decir que lo hicimos pero no es parte de esta submission > - Union is
+absolute for any transitive class -> is it well-known ?
 
 YES!
->
->
-> *p7*
->
-> - "We note in passing that it is possible to abbreviate expression in Isabelle..." -> I don't want to rude, but I really don't see how this could be a relevant information in a LICS paper...
 
-:'(
->
-> *p8*
->
-> - as said above, the use of variable names x_0, x_1, etc that are automatically translated in forces makes it hard to read...
-> - "two instances" -> which are?
-> - using absoluteness -> the absoluteness of?
-> - in the last block of code, the reader is left guessing what p,l,o,χ are
->
-> *p9*
->
-> - you should give a reference for Theorem 8 & 9
->
-> ----------------------- REVIEW 2 ---------------------
-> PAPER: 165
-> TITLE: Mechanization of Separation in Generic Extensions
-> AUTHORS: Emmanuel Gunther, Miguel Pagano and Pedro Sánchez Terraf
->
-> Overall evaluation: -2 (reject)
->
-> ----------- Overall evaluation -----------
-> Summary:
-> This paper presents a formalization in Isabelle of some axioms of Set theory of a generic extension of a model of Set theory using forcing.
-The main axiom, according to the authors, that has been shown to hold in a generic extension is Separation.
-The paper starts by presenting very briefly Isabelle, axioms of Set theory, mentioning the representation of first-order formulas as sets done by Paulson and
-> discuss their presentation of renaming of formulas.
-Then the authors explain briefly the generic extension of a model of Set theory using forcing, some modifications they have done on the development of Paulson of “ZF-Constructible”.
-The authors explain how to prove the “easy axioms” first.
-Then they present the fundamental lemmas of forcing which are given as axioms in the formalization.
-After that, they show that Separation holds in the generic extension providing the fundamental lemmas holds.
-Finally, they conclude by discussing how the main missing axiom, i.e.
-Replacement, could be proven.
->
-> Evaluation:
-> Overall, I can’t see more in this paper than a formalization exercise, following the library developed by Paulson and the textbook of Kunen.
-The reader does not get any insight on forcing in Set theory, and very few on how to formalize theorems on Set theoretic tools in Isabelle.
-For those reasons,
-> I suggest to the authors to submit their work to a more appropriate venue such as CPP or ITP, and focus more on the insight of the formalization.
+----------------------- REVIEW 2 ---------------------
 
-Sí, dale, vamos a ITP... Basta...
-> I think this paper is very hard to follow for a general audience that does not know about Cohen forcing and/or Isabelle.
-The paper interleaves short explanations on Set theory and forcing with technical remarks on the Isabelle formalization in a rather unstructured way, which makes it even harder to read.
->
-> Some Remarks:
->
-> - “We think that this project constitutes an interesting test-case for the current technology of  formalization of mathematics, in particular for the need of handling several layers of reasoning. “
-> Not really, it is only formalization of set theory, which is a very little part of what is usually  considered when one say formalizing mathematics.
-For instance, the work of the Mathematical Component  Library in Coq is more in this flavor of multiple layers of definitions/reasoning.
-- “It might be a little surprising the lack of formalizations of forcing and generic extensions”
-> It is indeed disappointing, it seems that the author have only partially formalize that given a good forcing translation,  the induced generic model is indeed a model of Set theory.
-But nothing is done on concrete translations which may be the hard part in Isabelle.
->
-> Questions to the authors:
-> “We are only able to prove that the generic extension satisfies Separation for any particular number of parameters, but not in general.”
-> This looks like a strong limitation, can you comment on this.
+> Summary: This paper presents a formalization in Isabelle of some
+> axioms of Set theory of a generic extension of a model of Set theory
+> using forcing. The main axiom, according to the authors, that has
+> been shown to hold in a generic extension is Separation. The paper
+> starts by presenting very briefly Isabelle, axioms of Set theory,
+> mentioning the representation of first-order formulas as sets done
+> by Paulson and discuss their presentation of renaming of
+> formulas. Then the authors explain briefly the generic extension of
+> a model of Set theory using forcing, some modifications they have
+> done on the development of Paulson of “ZF-Constructible”. The
+> authors explain how to prove the “easy axioms” first. Then they
+> present the fundamental lemmas of forcing which are given as axioms
+> in the formalization. After that, they show that Separation holds in
+> the generic extension providing the fundamental lemmas
+> holds. Finally, they conclude by discussing how the main missing
+> axiom, i.e. Replacement, could be proven.
 
-Hay que decir que no (de hecho con 2 parámetros estamos haciendo todo), pero tendría que sentarme un rato a confirmar de nuevo esto.
->
-> ----------------------- REVIEW 3 ---------------------
-> PAPER: 165
-> TITLE: Mechanization of Separation in Generic Extensions
-> AUTHORS: Emmanuel Gunther, Miguel Pagano and Pedro Sánchez Terraf
->
-> Overall evaluation: -2 (reject)
->
-> ----------- Overall evaluation -----------
-> Summary:
->
-> This paper presents a formalization of the separation axiom from set theory using the forcing library that has already been built in Isabelle/ZF by Paulson.
-The main contribution of the paper is a formalization of parts of Kunen's textbook in Isabelle/ZF.
->
-> Evaluation:
->
-> The paper is decently written. As someone whose poison of choice is Coq, I nevertheless found the paper fairly easy to follow since it doesn't use any complicated Isabelle-specific machinery.
->
-> Despite this, I regret to say that I cannot recommend acceptance at LICS.
->
-> My biggest complaint is that this work does not seem to provide the kind of fundamental contributions to logic that LICS is looking for.
-There is no new mathematics here, and this isn't even the first formalization of separation.
+> Questions to the authors: “We are only able to prove that the
+> generic extension satisfies Separation for any particular number of
+> parameters, but not in general.”  This looks like a strong
+> limitation, can you comment on this.
+
+Hay que decir que no (de hecho con 2 parámetros estamos haciendo
+todo), pero tendría que sentarme un rato a confirmar de nuevo esto.
+
+----------------------- REVIEW 3 ---------------------
+
+> My biggest complaint is that this work does not seem to provide the
+  kind of fundamental contributions to logic that LICS is looking
+  for. There is no new mathematics here, and this isn't even the first
+  formalization of separation.
 
 WTF!? Acá hay que preguntar cuál es la otra.
->
-> I think a conference like CPP or ITP is a much better fit for this kind of paper that is more of a documentation of a formalization project than a description of new results.
->
-> I am not giving a more thorough review because I think the program committees of these other conferences would be better equipped to critique a formalization artifact.
 
-O sea, no tenés la más puta idea.
->
-> ----------------------- REVIEW 4 ---------------------
-> PAPER: 165
-> TITLE: Mechanization of Separation in Generic Extensions
-> AUTHORS: Emmanuel Gunther, Miguel Pagano and Pedro Sánchez Terraf
->
-> Overall evaluation: -1 (weak reject)
+----------------------- REVIEW 4 ---------------------
 
-Oh, dadivoso réferi...
->
-> ----------- Overall evaluation -----------
-> This paper takes place in the long term goal of formalizing the method
-> of forcing *in set theory*, such as introduced by Cohen in the sixties
-> to prove the relative consistency of the negation of the continuum
-> hypothesis. (As opposed to more recent works cited by the authors, in
-> which the method of forcing has been adapted to various
-> type-theoretical frameworks, for instance to study its computational
-> contents via the proofs-as-programs correspondence)
->
-> Reusing parts of an earlier formalization (by Paulson) of Gödel's
-> method of constructible sets in the proof-assistant Isabelle, the
-> authors started a formalization of Cohen's method of forcing (in the
-> same proof-assistant) based on Kunen's presentation of forcing in his
-> famous textbook on set theory. In this paper, the authors present the
-> formalization of the soundness proof for the Separation scheme and for
-> the Powerset axiom. (The treatment of the Extensionality, Regularity,
-> Pairing, Union and Infinity axioms has been already published in an
-> earlier paper by the same authors, whereas the Replacement scheme is
-> still to be treated.)
-
-Acá la pifió bárbaro. Después repite esto que todos los axiomas fáciles están en otro paper. ¿Cómo llegó a tener esa impresión?
->
-> Together with the authors, I consider that the achievement of their
-> long term goal (which is to formalize the *set-theoretic* method of
-> forcing) would constitute a milestone in the formalization of
-> mathematics. (After all, Cohen was awarded the Fields medal for this
-> beautiful piece of mathematics.) Moreover, the paper is technically
-> sound and the methodology followed by the authors is clearly
-> explained.
-
-Lástima que este réferi sea el/la único/a que lo vea...
->
-> Nevertheless, and despite its real interest, the paper suffers from
-> important weaknesses which I will now address.
->
 > The more important point is that the "increment" presented in this
@@
 > such as LICS. Indeed, the paper only deals with the soundness of
-> Separation and Powerset in generic extensions. The formalization
-> of the soundness of Extensionality, Regularity, Pairing, Union and
+> Separation and Powerset in generic extensions. The formalization of
+> the soundness of Extensionality, Regularity, Pairing, Union and
 > Infinity has been published before,
 
-Otra vez esta pelotudez
-> and the soundness of Replacement
-> has not been formalized yet. More important: the authors' work relies
-> on some fundamental properties of forcing (relating the notion of
-> satisfaction in the generic extension with the notion of satisfaction
-> in the ground model) that have not been proven yet. (In their formal
-> development, these properties are only axiomatized.)
->
+Acá la pifió bárbaro. Después repite esto que todos los axiomas
+fáciles están en otro paper. ¿Cómo llegó a tener esa impresión?
+
 > The other point deals with the formalization choices of the authors,
@@
 > it with other presentations that would be (IMO) *much more suitable*
-> to a formalization of forcing in a proof-assistant, especially in the
-> perspective of deriving relative consistency proofs from this work.
-> (Which is proof-theoretically impossible from Kunen's approach, as I
-> explain below in my detailed comments.)
+> to a formalization of forcing in a proof-assistant, especially in
+> the perspective of deriving relative consistency proofs from this
+> work.  (Which is proof-theoretically impossible from Kunen's
+> approach, as I explain below in my detailed comments.)
 
-Raro esto. Kunen tiene una sección específica sobre cómo se puede hacer esto.
-> Moreover, the choice of the
-> authors to follow Kunen's approach induces many complications in their
-> formalization (for instance: the use of internalized formulas) that
-> could have been avoided by taking another approach (for instance,
-> the one presented by Jech in his textbook on Set theory). Perhaps
-> these complications explain why the authors had to split the
-> presentation of their formalization in several papers.
->
-> For all theses reasons - and despite the merits of the authors' work -
-> I cannot recommend this paper for publication in the LICS conference.
-> Nevertheless, I strongly encourage the authors to pursue their
-> (valuable) effort, but also to consider other formalization choices
-> that would be (IMO) more suited to their long-term goal, which is to
-> get relative consistency proofs.
+Raro esto. Kunen tiene una sección específica sobre cómo se puede
+hacer esto.
 
-Acá estaría bueno delimitar más los goles.
-Efectivamente, los móviles iniciales son las pruebas de consistencia relativa, pero lo que estamos haciendo está más cerca del desarrollo de la maquinaria matemática del forcing que de la correspondiente metamatemática de las pruebas de consistencia.
-En ese sentido, capaz habría que aclarar más este punto en próximas versiones del paper.
->
->
-> Detailed comments and suggestions for improvement
-> -------------------------------------------------
->
-> p. 2, last paragraph of the 1st column. "It must be observed that
-> predicates (...) do not correspond to first-order formulas". Here,
-> you should be more precise, to help the reader. As far as I
-> understand, you mean: "It must be observed that Isabelle higher-order
-> predicates (...) do not correspond to ZF first-order formulas."
->
+> Moreover, the choice of the authors to follow Kunen's approach
+> induces many complications in their formalization (for instance: the
+> use of internalized formulas) that could have been avoided by taking
+> another approach (for instance, the one presented by Jech in his
+> textbook on Set theory). Perhaps these complications explain why the
+> authors had to split the presentation of their formalization in
+> several papers.
+
 > p. 2, second column: internalized formulas. Here you should justify
-> why you use internalized formulas, that is: formulas as sets.
->
-> IMO, it would be more natural to inductively define an Isabelle
-> type "f" of first-order formulas (or codes of first-order formulas),
-> together with a decoding function that would map each code of a
-> first-order formula to a real Isabelle formula. So that the image
-> of this map would capture exactly the class you are interested in.
-> (At least, this would be the standard way of doing in Coq or Agda.)
+> why you use internalized formulas, that is: formulas as sets.  IMO,
+> it would be more natural to inductively define an Isabelle type "f"
+> of first-order formulas (or codes of first-order formulas), together
+> with a decoding function that would map each code of a first-order
+> formula to a real Isabelle formula. So that the image of this map
+> would capture exactly the class you are interested in.  (At least,
+> this would be the standard way of doing in Coq or Agda.)
 
-Bueno, se ve que no le quedó claro que esto no se puede hacer. Hay que insistir.
+Bueno, se ve que no le quedó claro que esto no se puede hacer. Hay que
+insistir.
+
 > Moreover, such a definition would reflect the fact that, in model
@@
 > they live in the ambient theory, outside all models.
->
+
 > On the other hand, I agree that "formulas as sets" are useful to
@@
 > constructible universe. But the point is that in forcing, we
-> (normally) never need such internalized formulas! The theory
-> of forcing is independent from the theory of constructible sets,
-> and this should appear in a formalization. (Although both theories
-> can be nicely combined in particular situations.)
+> (normally) never need such internalized formulas! The theory of
+> forcing is independent from the theory of constructible sets, and
+> this should appear in a formalization. (Although both theories can
+> be nicely combined in particular situations.)
 
-Acá otro pifie, pero es por algo que no aclaramos nosotros: nuestra presentación no depende de constructibilidad per se, sino de relativización y absolutez, que es de hecho un prerrequisito para ambas teorías.
->
+Acá otro pifie, pero es por algo que no aclaramos nosotros: nuestra
+presentación no depende de constructibilidad per se, sino de
+relativización y absolutez, que es de hecho un prerrequisito para
+ambas teorías.
+
 > So why did you consider internalized formulas, and not external
-> formulas, inductively defined "outside" as explained above?
-> I understand you want to reuse parts of the (previous) formalization
+> formulas, inductively defined "outside" as explained above?  I
+> understand you want to reuse parts of the (previous) formalization
 > of the constructible universe, but this is (scientifically) a very
@@
 
-Bueno, hacer todo desde cero es una complicación innecesaria para nosotros (!).
->
+Bueno, hacer todo desde cero es una complicación innecesaria para
+nosotros (!).
+
 > Moreover, considering internalized formulas (that is: formulas as
@@
 
-Todo está pasando dentro de "i", así que no es relevante esta observación. Hay que discutirlo un poco más.
->
-> p. 2, last paragraph of 2nd column. "Forcing is a technique to extend
-> very special kind of models...". Actually, not. There are several
-> ways to construct forcing models of ZF, and the one you consider in
-> your paper (taken from Kunen's book) is only *one* of them.
->
-> The method you consider is based on countable transitive models
-> (ctms) of ZF, whose main advantage is that they allow the use of
-> the Rasiowa-Sikorski lemma to construct the generic filter.
-> Basically, a ctm of ZF captures three properties:
->
-> (1) M is a set equipped with a binary relation (to interpret
-> membership) that is well-founded _from the point of view of the
-> outer set-theoretic universe_ (the type "i"). This property
-> implies that M fulfills the foundation axiom, but the converse
-> is not true. So basically, this property says that the notion
-> of well-foundedness is the same inside M and outside M (that is:
-> in the outer model "i").
->
-> (2) M is transitive, which means that the binary relation interpreting
-> membership is membership itself (in the sense of "i"). This point
-> is actually just a technical facility, to avoid distinguishing two
-> membership predicates.
+Todo está pasando dentro de "i", así que no es relevante esta
+observación. Hay que discutirlo un poco más.  > > p. 2, last paragraph
+of 2nd column. "Forcing is a technique to extend > very special kind
+of models...". Actually, not. There are several > ways to construct
+forcing models of ZF, and the one you consider in > your paper (taken
+from Kunen's book) is only *one* of them.  > > The method you consider
+is based on countable transitive models > (ctms) of ZF, whose main
+advantage is that they allow the use of > the Rasiowa-Sikorski lemma
+to construct the generic filter.  > Basically, a ctm of ZF captures
+three properties: > > (1) M is a set equipped with a binary relation
+(to interpret > membership) that is well-founded _from the point of
+view of the > outer set-theoretic universe_ (the type "i"). This
+property > implies that M fulfills the foundation axiom, but the
+converse > is not true. So basically, this property says that the
+notion > of well-foundedness is the same inside M and outside M (that
+is: > in the outer model "i").  > > (2) M is transitive, which means
+that the binary relation interpreting > membership is membership
+itself (in the sense of "i"). This point > is actually just a
+technical facility, to avoid distinguishing two > membership
+predicates.
 
-No sólo eso. Se está olvidando de los resultados de absolutez.
->
-> (3) M is countable (from the point of view of the universe "i").
-> This point is crucial to prove the Rasiowa-Sikorski lemma.
->
-> As you mention it in Lemma 1, the properties (2) and (3) are
-> proof-theoretically free, since once you have (1), you can easily
-> enforce (2) and (3) by combining a Mostowski collapse with the
-> (descending) Löwenheim-Skolem theorem.
->
-> But the point is that (1) is actually very strong, and more
-> important: it is *not known* to be equivalent to the consistency
-> of ZF. By this, I mean that it might be the case that ZF has
-> no transitive model while being consistent.
+No sólo eso. Se está olvidando de los resultados de absolutez.  > >
+(3) M is countable (from the point of view of the universe "i").  >
+This point is crucial to prove the Rasiowa-Sikorski lemma.  > > As you
+mention it in Lemma 1, the properties (2) and (3) are >
+proof-theoretically free, since once you have (1), you can easily >
+enforce (2) and (3) by combining a Mostowski collapse with the >
+(descending) Löwenheim-Skolem theorem.  > > But the point is that (1)
+is actually very strong, and more > important: it is *not known* to be
+equivalent to the consistency > of ZF. By this, I mean that it might
+be the case that ZF has > no transitive model while being consistent.
 
-Tranquilate un poco... Esto también es irrelevante, porque sólo necesitamos un ctm de un fragmento suficientemente grande de ZFC (cosa que el Teorema de Reflexión nos da).
->
+Tranquilate un poco... Esto también es irrelevante, porque sólo
+necesitamos un ctm de un fragmento suficientemente grande de ZFC (cosa
+que el Teorema de Reflexión nos da).
+
 > This simple argument shows that transitive models of ZF, although
@@
 
-Ídem... Sólo necesitamos un fragmento finito de esto, como lo explica Kunen en el libro nuevo.
->
-> This is why other approaches have been developed in the literature.
-> For instance, the construction of a Boolean-valued model of names
-> inside the ground model M, and then a quotient by a suitable
-> ultrafilter (see Jech's book for instance). The main advantage of
-> this approach is that it can be applied to _any model_ of ZF. (In
-> this case, the generic filter is directly constructed from the
-> corresponding name, without the Rasiowa-Sikorski lemma.)
->
-> However, the ground model M and the corresponding generic extension
-> M[G] are in general *not* transitive models. But from a proof-
-> theoretic point of view, this property is definitely useless!
->
-> So, since you are interested in relative consistency results (as
-> mentioned in the introduction of the paper), why didn't you consider
-> this alternative approach, that would work with _any_ model of ZF?
-> (Moreover, it seems to me that the job of formalization would have
-> been easier.) IMO, the paper is dramatically missing a discussion
-> about this important foundational issue (that has also a dramatic
-> impact on your formalization choices).
+Ídem... Sólo necesitamos un fragmento finito de esto, como lo explica
+Kunen en el libro nuevo.  > > This is why other approaches have been
+developed in the literature.  > For instance, the construction of a
+Boolean-valued model of names > inside the ground model M, and then a
+quotient by a suitable > ultrafilter (see Jech's book for
+instance). The main advantage of > this approach is that it can be
+applied to _any model_ of ZF. (In > this case, the generic filter is
+directly constructed from the > corresponding name, without the
+Rasiowa-Sikorski lemma.)  > > However, the ground model M and the
+corresponding generic extension > M[G] are in general *not* transitive
+models. But from a proof- > theoretic point of view, this property is
+definitely useless!  > > So, since you are interested in relative
+consistency results (as > mentioned in the introduction of the paper),
+why didn't you consider > this alternative approach, that would work
+with _any_ model of ZF?  > (Moreover, it seems to me that the job of
+formalization would have > been easier.) IMO, the paper is
+dramatically missing a discussion > about this important foundational
+issue (that has also a dramatic > impact on your formalization
+choices).
 
-Me gusta más el enfoque de los ctm para trabajar en general (y off-the-record, es el que más entiendo de todos), y además tratamos de seguir lo más posible una sola referencia (i.e.
-Kunen) para darle cohesión al trabajo.
-La discusión sobre la conveniencia de usar el enfoque de los modelos a valores booleanos es más interesante, pero creo que otra vez nos pone en el brete de tener que reimplementar varias cosas.
->
-> Another advantage of constructing the generic extension as a quotient
-> of the Boolean-valued model of names (defined as a class within M)
-> would be the following. In your work, you need to distinguish three
-> models:
-> - the Isabelle type i of sets (the ambient model);
-> - the ctm M, that is an object of type i (the ground model);
-> - the generic extension M[G].
-> By considering the approach I suggest, you would only need two models:
-> - the type i, which would serve as a ground model M as well;
-> - the generic extension M[G], that would be defined as a quotient of
-> the class of names (within i = M).
->
-> p. 3, Proof of Lemma 1. "an countable..." -> "a countable".
->
+Me gusta más el enfoque de los ctm para trabajar en general (y
+off-the-record, es el que más entiendo de todos), y además tratamos de
+seguir lo más posible una sola referencia (i.e. Kunen) para darle
+cohesión al trabajo. La discusión sobre la conveniencia de usar el
+enfoque de los modelos a valores booleanos es más interesante, pero
+creo que otra vez nos pone en el brete de tener que reimplementar
+varias cosas.
+
+> Another advantage of constructing the generic extension as a
+> quotient of the Boolean-valued model of names (defined as a class
+> within M) would be the following. In your work, you need to
+> distinguish three models: - the Isabelle type i of sets (the ambient
+> model); - the ctm M, that is an object of type i (the ground model);
+> - the generic extension M[G].  By considering the approach I
+> suggest, you would only need two models: - the type i, which would
+> serve as a ground model M as well; - the generic extension M[G],
+> that would be defined as a quotient of the class of names (within i
+> = M).
+
 > p. 3, second last par. of column 1. "We are only able to prove that
 > the generic extension satisfies Separation for any particular number
-> of parameters, but not in general." What do you mean precisely?
-> At which level lives the quantification in "any particular number of
-> parameters"? A quantification over the set N (inside M and "i")?
-> Or a quantification over the type of natural numbers in Isabelle?
+> of parameters, but not in general." What do you mean precisely?  At
+> which level lives the quantification in "any particular number of
+> parameters"? A quantification over the set N (inside M and "i")?  Or
+> a quantification over the type of natural numbers in Isabelle?
 > Please clarify this point.
+
 Aclarar esto.
->
-> p. 3, next par. "possible define" -> "possible to define".
->
-> p. 4, last remark before Sect. V. "All our uses... (etc.)".
-> What do you mean precisely? I am completely confused.
 
-Tener en cuenta quizá para la próxima versión
->
-> p. 4, first paragraph of Sect. V. "be its _name_ in M".
-> Here you should precise what you mean here by a name.
-> Do you consider a special class of names inside M?
-> Or do you consider all elements of M as names?
-> (Note that both options are possible.) From the discussion of
-> Sect. V. A., it seems that you consider the second option, but
-> you should clarify it.
->
-> p. 5, second last par. before Sect. VI. "Paulson proved... (etc.)"
-> This discussion is really hard to follow, due to the several levels
-> that are involved.
->
-> p. 6, column 2: Sect. VIII. In this section, you say that the
-> fundamental properties of forcing are axiomatized (in a locale)
-> rather than proved. IMO, these properties should have been proven
-> first (before proving the soundness of the axioms of ZF), since they
-> are at the core of the method of forcing.
-
-A esto lo mando a una revista, no a una conferencia, GIL.
->
 > p. 9, column 1, 3rd paragraph: "The Replacement Axiom [...] requires
@@
 
-Tendría que ver esto, no estoy seguro ya mismo. Este tipo entiende, sí, pero hay gente que entiende bien un enfoque y no el otro (bueno, a mí me pasa un poco pero al revés de este muchacho).
+Tendría que ver esto, no estoy seguro ya mismo. Este tipo entiende,
+sí, pero hay gente que entiende bien un enfoque y no el otro (bueno, a
+mí me pasa un poco pero al revés de este muchacho).

### 2019-03-07 LICS/rebuttal.txt
versión limpia, pero incompleta y no revisada

@@
 
------------------------ REVIEW 1 ---------------------
+We first address some issues that appeared in more than one
+review. Our choice of Isabelle/ZF as framework was simply based on the
+need to decrease the overhead of the formalization process. Indeed,
+the main drawback of this choice is that we cannot perform inductive
+arguments on "external" formulas (i.e., Isabelle terms of type
+o). This is the reason why we are using internalized formulas.
 
-> Moreover, I would have expected at least a bit of discussion on the
-  benefits/drawbacks of choosing Isabelle/ZF over other possible
-  frameworks (Coq, Mizar, Isabelle/HOL, HOL Light, ... and the
-  corresponding underlying logics).
-
-Insistir que el overhead era tremendo si no usábamos Isabelle/ZF.
-
-> - "the fruitful use of forcing techniques to extend the Curry-Howard
->isomorphism to classical axioms" -> this claim is misleading. The two
->papers you cite show how to express forcing within Krivine
->realizability, a framework which is already classical. Therefore, in
->these papers, forcing techniques are *not* used to extend CH to
->classical logic (which has to do with control operators, not
->forcing). They rather unveil the computational content of forcing
->within Krivine realizability, just like Jaber et al. show how to
->define it within type theory [14,15].  Llamada para el Dr. Pagano...
+Questions by REVIEWER 1
+=======================
 
 > - ∈ restricted to M -> this suggests that it is defined outside of
-    M, right? Isn't this strange?
+>   M, right? Isn't this strange?
 
-Hay que decir que ∈ es el mismo para todos los modelos (el externo).
-
-> - induction on terms of type o is not available -> is this related
-    to the chosen framework (Isabelle/ZF) or is it more general?
-
-Aclarar
+Not really, since ∈ is defined in the "ambient" Isabelle type i, and
+one can restrict it to any set.
 
 > - "it is also possible", "it can be shown" -> if you did it, say it
-    as such, otherwise I am not sure that this is informative or
-    useful (but I might be missing something)
+>    as such, otherwise I am not sure that this is informative or
+>    useful (but I might be missing something)
 
-Decir que lo hicimos pero no es parte de esta submission > - Union is
-absolute for any transitive class -> is it well-known ?
+We actually did, but that wasn't part of this submission. We are sorry
+for the imprecision
 
-YES!
+> - Union is absolute for any transitive class - is it well-known ?
 
------------------------ REVIEW 2 ---------------------
+Yes, indeed.
 
-> Summary: This paper presents a formalization in Isabelle of some
-> axioms of Set theory of a generic extension of a model of Set theory
-> using forcing. The main axiom, according to the authors, that has
-> been shown to hold in a generic extension is Separation. The paper
-> starts by presenting very briefly Isabelle, axioms of Set theory,
-> mentioning the representation of first-order formulas as sets done
-> by Paulson and discuss their presentation of renaming of
-> formulas. Then the authors explain briefly the generic extension of
-> a model of Set theory using forcing, some modifications they have
-> done on the development of Paulson of “ZF-Constructible”. The
-> authors explain how to prove the “easy axioms” first. Then they
-> present the fundamental lemmas of forcing which are given as axioms
-> in the formalization. After that, they show that Separation holds in
-> the generic extension providing the fundamental lemmas
-> holds. Finally, they conclude by discussing how the main missing
-> axiom, i.e. Replacement, could be proven.
+Questions by REVIEWER 2
+=======================
 
@@
 
-Hay que decir que no (de hecho con 2 parámetros estamos haciendo
-todo), pero tendría que sentarme un rato a confirmar de nuevo esto.
+Actually, it is well known [Schindler] that there are axiomatizations
+of ZF without any parameters. We have chosen an axiomatization using
+at most one parameter in both Separation and Replacement. In this
+sense, we can prove that M[G] satisfies the "same amount of
+Separation" that M is assumed to satisfy.
 
------------------------ REVIEW 3 ---------------------
+Questions by REVIEWER 3
+=======================
 
-> My biggest complaint is that this work does not seem to provide the
-  kind of fundamental contributions to logic that LICS is looking
-  for. There is no new mathematics here, and this isn't even the first
-  formalization of separation.
+There are no questions by the third Reviewer, but this Reviewer states
+that ours is not the first formalization of separation. We are
+extremely eager to know about this, would they provide a reference for
+such a strong claim?
 
-WTF!? Acá hay que preguntar cuál es la otra.
-
------------------------ REVIEW 4 ---------------------
+Questions by REVIEWER 4
+=======================
 
@@
 
-Acá la pifió bárbaro. Después repite esto que todos los axiomas
-fáciles están en otro paper. ¿Cómo llegó a tener esa impresión?
-
-> The other point deals with the formalization choices of the authors,
-> which are never really discussed. It seems to me that the authors
-> blindly followed Kunen's presentation of forcing (based on countable
-> transitive models and the Rasiowa-Sikorski lemma), without comparing
-> it with other presentations that would be (IMO) *much more suitable*
-> to a formalization of forcing in a proof-assistant, especially in
-> the perspective of deriving relative consistency proofs from this
-> work.  (Which is proof-theoretically impossible from Kunen's
-> approach, as I explain below in my detailed comments.)
-
-Raro esto. Kunen tiene una sección específica sobre cómo se puede
-hacer esto.
-
-> Moreover, the choice of the authors to follow Kunen's approach
-> induces many complications in their formalization (for instance: the
-> use of internalized formulas) that could have been avoided by taking
-> another approach (for instance, the one presented by Jech in his
-> textbook on Set theory). Perhaps these complications explain why the
-> authors had to split the presentation of their formalization in
-> several papers.
-
-> p. 2, second column: internalized formulas. Here you should justify
-> why you use internalized formulas, that is: formulas as sets.  IMO,
-> it would be more natural to inductively define an Isabelle type "f"
-> of first-order formulas (or codes of first-order formulas), together
-> with a decoding function that would map each code of a first-order
-> formula to a real Isabelle formula. So that the image of this map
-> would capture exactly the class you are interested in.  (At least,
-> this would be the standard way of doing in Coq or Agda.)
-
-Bueno, se ve que no le quedó claro que esto no se puede hacer. Hay que
-insistir.
-
-> Moreover, such a definition would reflect the fact that, in model
-> theory, formulas are traditionally considered as external objects:
-> they live in the ambient theory, outside all models.
+There is a mistake in this assertion. The only axiom that has been
+published before (at least by us) was Pairing. Is Reviewer 4 referring
+to some work not known to us?
 
@@
 > forcing is independent from the theory of constructible sets, and
-> this should appear in a formalization. (Although both theories can
-> be nicely combined in particular situations.)
+> this should appear in a formalization.
 
-Acá otro pifie, pero es por algo que no aclaramos nosotros: nuestra
-presentación no depende de constructibilidad per se, sino de
-relativización y absolutez, que es de hecho un prerrequisito para
-ambas teorías.
-
-> So why did you consider internalized formulas, and not external
-> formulas, inductively defined "outside" as explained above?  I
-> understand you want to reuse parts of the (previous) formalization
-> of the constructible universe, but this is (scientifically) a very
-> weak argument. IMO, considering internalized formulas for defining
-> forcing is just introducing unnecessary complication.
-
-Bueno, hacer todo desde cero es una complicación innecesaria para
-nosotros (!).
-
-> Moreover, considering internalized formulas (that is: formulas as
-> sets) raises another problem, which is that there might be more
-> internalized formulas than first-order formulas (such as seen
-> outside). Indeed, your axiomatized type "i" of ZF sets is nothing
-> but a (hypothetical) model of ZF inside Isabelle. But such a model
-> can be non-standard, in the sense that it might contain more natural
-> numbers (and thus more formulas-as-sets) than the Isabelle type of
-> natural numbers. Note that this problem is harmless in practice, but
-> it is worth mentioning it.
-
-Todo está pasando dentro de "i", así que no es relevante esta
-observación. Hay que discutirlo un poco más.  > > p. 2, last paragraph
-of 2nd column. "Forcing is a technique to extend > very special kind
-of models...". Actually, not. There are several > ways to construct
-forcing models of ZF, and the one you consider in > your paper (taken
-from Kunen's book) is only *one* of them.  > > The method you consider
-is based on countable transitive models > (ctms) of ZF, whose main
-advantage is that they allow the use of > the Rasiowa-Sikorski lemma
-to construct the generic filter.  > Basically, a ctm of ZF captures
-three properties: > > (1) M is a set equipped with a binary relation
-(to interpret > membership) that is well-founded _from the point of
-view of the > outer set-theoretic universe_ (the type "i"). This
-property > implies that M fulfills the foundation axiom, but the
-converse > is not true. So basically, this property says that the
-notion > of well-foundedness is the same inside M and outside M (that
-is: > in the outer model "i").  > > (2) M is transitive, which means
-that the binary relation interpreting > membership is membership
-itself (in the sense of "i"). This point > is actually just a
-technical facility, to avoid distinguishing two > membership
-predicates.
-
-No sólo eso. Se está olvidando de los resultados de absolutez.  > >
-(3) M is countable (from the point of view of the universe "i").  >
-This point is crucial to prove the Rasiowa-Sikorski lemma.  > > As you
-mention it in Lemma 1, the properties (2) and (3) are >
-proof-theoretically free, since once you have (1), you can easily >
-enforce (2) and (3) by combining a Mostowski collapse with the >
-(descending) Löwenheim-Skolem theorem.  > > But the point is that (1)
-is actually very strong, and more > important: it is *not known* to be
-equivalent to the consistency > of ZF. By this, I mean that it might
-be the case that ZF has > no transitive model while being consistent.
-
-Tranquilate un poco... Esto también es irrelevante, porque sólo
-necesitamos un ctm de un fragmento suficientemente grande de ZFC (cosa
-que el Teorema de Reflexión nos da).
+We do need to ponder the suggestion by this Reviewer of defining a new
+Isabelle type "f" for formulas, and we thank this
+insight. Nevertheless, we are *not* using the theory of constructible
+sets for our formalization, only the absoluteness results (which are
+prerrequisites for both theories).
 
@@
 
-Ídem... Sólo necesitamos un fragmento finito de esto, como lo explica
-Kunen en el libro nuevo.  > > This is why other approaches have been
-developed in the literature.  > For instance, the construction of a
-Boolean-valued model of names > inside the ground model M, and then a
-quotient by a suitable > ultrafilter (see Jech's book for
-instance). The main advantage of > this approach is that it can be
-applied to _any model_ of ZF. (In > this case, the generic filter is
-directly constructed from the > corresponding name, without the
-Rasiowa-Sikorski lemma.)  > > However, the ground model M and the
-corresponding generic extension > M[G] are in general *not* transitive
-models. But from a proof- > theoretic point of view, this property is
-definitely useless!  > > So, since you are interested in relative
-consistency results (as > mentioned in the introduction of the paper),
-why didn't you consider > this alternative approach, that would work
-with _any_ model of ZF?  > (Moreover, it seems to me that the job of
-formalization would have > been easier.) IMO, the paper is
-dramatically missing a discussion > about this important foundational
-issue (that has also a dramatic > impact on your formalization
-choices).
-
-Me gusta más el enfoque de los ctm para trabajar en general (y
-off-the-record, es el que más entiendo de todos), y además tratamos de
-seguir lo más posible una sola referencia (i.e. Kunen) para darle
-cohesión al trabajo. La discusión sobre la conveniencia de usar el
-enfoque de los modelos a valores booleanos es más interesante, pero
-creo que otra vez nos pone en el brete de tener que reimplementar
-varias cosas.
-
-> Another advantage of constructing the generic extension as a
-> quotient of the Boolean-valued model of names (defined as a class
-> within M) would be the following. In your work, you need to
-> distinguish three models: - the Isabelle type i of sets (the ambient
-> model); - the ctm M, that is an object of type i (the ground model);
-> - the generic extension M[G].  By considering the approach I
-> suggest, you would only need two models: - the type i, which would
-> serve as a ground model M as well; - the generic extension M[G],
-> that would be defined as a quotient of the class of names (within i
-> = M).
+Though it is a fact that Con(ZF) does not imply the existence of ctms
+of ZF, finitary consistency proofs can be obtained through the ctm
+approach, since one only needs that the ctm M satisfy only a finite
+fragment of ZF at a time (and such an M can be obtained by a standard
+reflection argument). For a reference, check [Kunen, IV.5].
 
@@
 
-Aclarar esto.
-
-> p. 9, column 1, 3rd paragraph: "The Replacement Axiom [...] requires
-> more work to be done. In Kunen it requires a relativized version
-> [...] of the Reflection Principle." IMO, this requirement is another
-> drawback of following Kunen's approach. By taking the other method I
-> suggest, you would definitely not need the Reflection Principle to
-> prove the soundness of Replacement.
-
-Tendría que ver esto, no estoy seguro ya mismo. Este tipo entiende,
-sí, pero hay gente que entiende bien un enfoque y no el otro (bueno, a
-mí me pasa un poco pero al revés de este muchacho).
+Isabelle/ZF does not define a type of natural numbers (we just have i
+and o). "In general" would mean an universal quantification over the
+set nat::i.

### 2019-03-07 LICS/rebuttal.txt
References

@@
 o). This is the reason why we are using internalized formulas.
+
+There is are some questions in common on parameters of Separation, but
+those are answered below.
 
@@
 
-Actually, it is well known [Schindler] that there are axiomatizations
-of ZF without any parameters. We have chosen an axiomatization using
-at most one parameter in both Separation and Replacement. In this
-sense, we can prove that M[G] satisfies the "same amount of
-Separation" that M is assumed to satisfy.
+Actually, it is well known [Levy, SchiSchli] that there are
+axiomatizations of ZF without any parameters. We have chosen an
+axiomatization using at most one parameter in both Separation and
+Replacement. In this sense, we can prove that M[G] satisfies the "same
+amount of Separation" that M is assumed to satisfy.
 
@@
 set nat::i.
+
+References
+==========
+
+    [Kunen] K. Kunen, Set Theory, 2nd ed., ser. Studies in Logic.
+    	    College Publications, 2011, revised edition, 2013.
+
+     [Levy] Azriel Levy, Parameters in the comprehension axiom schemas
+     	    of set theory, in Proceedings of the Tarski symposium,
+     	    Proceedings of Symposia in Pure Mathematics, vol. 25, American
+     	    Mathematical Society, Providence, RI, pp. 309–324.
+
+[SchiSchli] Ralf Schindler and Philipp Schlicht, ZFC without parameters (A note on a question of Kai Wehmeier) http://wwwmath.uni-muenster.de/logik/Personen/rds/ZFC_without_parameters.pdf

### 2019-03-07 LICS/rebuttal.txt
bibstyle
typo (ispell)

@@
 sets for our formalization, only the absoluteness results (which are
-prerrequisites for both theories).
+prerequisites for both theories).
 
@@
 
-[SchiSchli] Ralf Schindler and Philipp Schlicht, ZFC without parameters (A note on a question of Kai Wehmeier) http://wwwmath.uni-muenster.de/logik/Personen/rds/ZFC_without_parameters.pdf
+[SchiSchli] Ralf Schindler and Philipp Schlicht, ZFC without
+	    parameters (A note on a question of Kai Wehmeier)
+	    http://wwwmath.uni-muenster.de/logik/Personen/rds/ZFC_without_parameters.pdf

### 2019-03-07 LICS/rebuttal.txt
cambios menores
Pasada de Miguel por respuesta
Cambios en las respuestas al 3 y al 4.

@@
 
+> Summary:
+>
+> This paper presents a formalization of the separation axiom from set
+> theory using the forcing library that has already been built in
+> Isabelle/ZF by Paulson.
+
+This is a mistake. Forcing was not formalized by Paulson.
+
 There are no questions by the third Reviewer, but this Reviewer states
-that ours is not the first formalization of separation. We are
-extremely eager to know about this, would they provide a reference for
-such a strong claim?
+that ours is not the firsgt formalization of separation. This strong
+claim is quite misleading: does the Reviewer refer to some
+formalization of the Axiom of Separation of some set theory? It's hard
+to imagine how does such a formalization look like. We are extremely
+eager to know about this, would they provide a reference for such a
+strong claim?
 
@@
 There is a mistake in this assertion. The only axiom that has been
-published before (at least by us) was Pairing. Is Reviewer 4 referring
-to some work not known to us?
+published before (at least by us) was Pairing.
 
@@
 
-We do need to ponder the suggestion by this Reviewer of defining a new
-Isabelle type "f" for formulas, and we thank this
-insight. Nevertheless, we are *not* using the theory of constructible
-sets for our formalization, only the absoluteness results (which are
+We agree that internalized formulas are not needed per se; however in
+Isabelle/ZF internalized formulas are the only representation of
+formulas that are syntactically manipulable. We welcome the suggestion
+of extending Isabelle/ZF with a new inductive type "f" for formulas,
+this would, however, change the meta-theory (in contrast with Agda, Coq,
+or even Isabelle/HOL there is no support in Isabelle/ZF for introducing
+inductive types).
+
+Nevertheless, we are *not* using the theory of constructible sets for
+our formalization, only the absoluteness results (which are
 prerequisites for both theories).

### 2019-03-07 LICS/rebuttal.txt
comentario Reviewer 3 cambiado y al principio

@@
 those are answered below.
+
+Finally, we were a bit preoccupied by a comment made in one of the
+reviews. Reviewer 3 states that ours is not the first formalization of
+separation. We are extremely eager to know about this, would they
+provide a reference for such a strong claim?
 
@@
 This is a mistake. Forcing was not formalized by Paulson.
-
-There are no questions by the third Reviewer, but this Reviewer states
-that ours is not the firsgt formalization of separation. This strong
-claim is quite misleading: does the Reviewer refer to some
-formalization of the Axiom of Separation of some set theory? It's hard
-to imagine how does such a formalization look like. We are extremely
-eager to know about this, would they provide a reference for such a
-strong claim?
 

### 2019-03-08 LICS/rebuttal.txt
comentario extra de separación
confirmo la dificultad de implementar "f"

@@
 
-There is are some questions in common on parameters of Separation, but
+There are some questions in common on parameters of Separation, but
 those are answered below.
 
-Finally, we were a bit preoccupied by a comment made in one of the
-reviews. Reviewer 3 states that ours is not the first formalization of
-separation. We are extremely eager to know about this, would they
-provide a reference for such a strong claim?
+Finally, a comment in one of the reviews surprised us. Reviewer 3
+states that ours is not the first formalization of separation. We are
+extremely eager to know about this, would they provide a reference for
+such a strong claim?
 
@@
 
-Not really, since ∈ is defined in the "ambient" Isabelle type i, and
+Not really, since ∈ is defined over the "ambient" Isabelle type i, and
 one can restrict it to any set.
@@
 formulas that are syntactically manipulable. We welcome the suggestion
-of extending Isabelle/ZF with a new inductive type "f" for formulas,
-this would, however, change the meta-theory (in contrast with Agda, Coq,
-or even Isabelle/HOL there is no support in Isabelle/ZF for introducing
-inductive types).
+of extending Isabelle/ZF with a new inductive type "f" for formulas;
+this would, however, require a change in the metatheory. In contrast
+with Agda, Coq, or even Isabelle/HOL, there is no support in
+Isabelle/ZF for introducing inductive types, save recursively
+defined "sets" (elements of i).
 
@@
 and o). "In general" would mean an universal quantification over the
-set nat::i.
+set nat::i. For our purposes, showing Separation for at most one
+parameter (that is, for at most two free variables) is enough.
 

### 2019-03-08 LICS/rebuttal.txt
ejemplo de recursively defined

@@
 Isabelle/ZF for introducing inductive types, save recursively
-defined "sets" (elements of i).
+defined "sets" (elements of i, like nat).
 

### 2019-03-19 ITP/Separation_In_MG.tex
Primeros pasos hacia ITP

\documentclass[a4paper,UKenglish,cleveref, autoref]{lipics-v2019}
%This is a template for producing LIPIcs articles. 
%See lipics-manual.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling cleveref support, use "autoref"

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\bibliographystyle{plainurl}% the mandatory bibstyle

\title{Mechanization of Separation in Generic Extensions} %TODO Please add

%\titlerunning{}%optional, please use if title is longer than one line

\author{Emmanuel Gunther}{FAMAF, Universidad Nacional de Córdoba, [optional: Address], Argentina}{gunther@famaf.unc.edu.ar}{}{Partially supported by a CONICET scholarship}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional

\author{Miguel Pagano}{FAMAF, Universidad Nacional de Córdoba, Argentina}{pagano@famaf.unc.edu.ar}{}{}%{[orcid]}{[funding]}
\author{Pedro Sánchez Terraf}{CIEM-FAMAF, Universidad Nacional de Córdoba, Argentina}{sterraf@famaf.unc.edu.ar}{}{}%{[orcid]}{[funding]}

\authorrunning{E. Gunther and M. Pagano and P. Sánchez Terraf}%TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Emmanuel Gunther and Miguel Pagano and Pedro Sánchez Terraf}%TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\ccsdesc[100]{General and reference~General literature}
\ccsdesc[100]{General and reference}%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

\keywords{Isabelle/ZF, forcing, names, generic extension, constructibility}%TODO mandatory; please add comma-separated list of keywords

\category{}%optional, e.g. invited paper

\relatedversion{A full version of the paper is available at \url{https://arxiv.org/abs/1901.03313}}%optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversion{A full version of the paper is available at \url{...}.}

\supplement{The whole Isabelle/ZF development is available at \url{https://arxiv.org/abs/1901.03313}}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

\acknowledgements{We would like to thanks to anonymous referees of a
  previous version of this paper that helped us to enhance the
  presentation.}%optional

%\nolinenumbers %uncomment to disable line numbering

%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)
\usepackage{isabelle,isabellesym}
\usepackage{csquotes}
\input{header-draft}

\begin{document}

\maketitle
\input{abstract}

\input{intro}

\input{isabelle}

\input{zfc-axioms}

\input{renaming}

\input{generic-extensions}

\input{recursion}

% \input{prior-results}

\input{hacking}

\input{easy-axioms}

\input{forcing}

\input{proof-of-separation}

\input{conclusion}

\bibliography{../LSFA/citados}

\newpage
\onecolumn
\appendix[A short overview of our development]
\input{appendix}
% that's all folks
\end{document}

%%% Local Variables:
%%% mode: latex
%%% ispell-local-dictionary: "american"
%%% End:

### 2019-03-19 ITP/conclusion.tex
Primeros pasos hacia ITP

--- 2019-01-10 LICS/conclusion_m.tex
@@
 presentation of forcing given by
-\citet{kunen2011set}. In fact, it helped us to delineate the
+Kunen \cite{kunen2011set}. In fact, it helped us to delineate the
 \emph{thematic} aspects of our formalization; i.e.~the handling of all
@@
 
-Two axioms have not been treated in full. Infinity was proved under
-two extra assumptions on the model; when we develop a full-fledged
-interface between ctms of $\ZF$  and the locales 
-providing recursive constructions from Paulson's
+Replacement is the only one axiom that we have not yet addressed (see
+below) and Infinity was proved under two extra assumptions on the
+ground model; when we develop a full-fledged interface between ctms of
+$\ZF$ and the locales providing recursive constructions from Paulson's
 \isatt{ZF-Constructible} session, the same current proof will hold
@@
 It is obvious that we can take the conjunction of $\Phi$ and state
-Theorem~\ref{th:reflection-principle} for a single formula, say $\phi$.
-The schematic nature of this result
-hints at a proof by induction on formulas, and hence it must be shown
-internally. It is to be noted that Paulson
-\cite{DBLP:conf/cade/Paulson02} also formalized the
-Reflection principle in Isabelle/ZF, but it is not clear if the
-relativized version follows directly from it. (It may be possible to
-sidestep Reflection, since in Neeman \cite{neeman-course}, only the
-relativization of the cumulative hierarchy is needed; nevertheless, it
-is a nontrivial task.)
+Theorem~\ref{th:reflection-principle} for a single formula, say
+$\phi$.  The schematic nature of this result hints at a proof by
+induction on formulas, and hence it must be shown internally. It is to
+be noted that Paulson \cite{DBLP:conf/cade/Paulson02} also formalized
+the Reflection principle in Isabelle/ZF, but it is not clear if the
+relativized version follows directly from it. Neeman
+\cite{neeman-course} has an alternative proof for Replacement that
+might avoid Reflection, for only the relativization of the
+cumulative hierarchy is needed; it is, nevertheless, a non-trivial
+task.
 
@@
 An alternative road to internalization would be to redevelop
-absoluteness 
-results in a more structured metatheory that already includes a
-recursively defined type of first order formulas. Needless to say, this
-comprises an extensive re-engineering.
+absoluteness results in a more structured metatheory that already
+includes a recursively defined type of first order formulas. Needless
+to say, this change to a deeper-embedding comprises an extensive
+re-engineering.
 

### 2019-03-19 ITP/easy-axioms.tex
Primeros pasos hacia ITP

--- 2019-01-06 LICS/easy-axioms.tex
@@
 Let $\tau$ be the name for $a$, i.e.\ $a=\val(G,\tau)$; 
-\citet{kunen2011set} gives $\pi_a$ in terms of $\tau$:
+Kunen \cite{kunen2011set} gives $\pi_a$ in terms of $\tau$:
 \begin{align*}
@@
 working directly with the internalized formula, we define a predicate
-\isa{Unioin{\isacharunderscore}name{\isacharunderscore}body} and prove the equivalence between
+\isa{Union{\isacharunderscore}name{\isacharunderscore}body} and prove the equivalence between
 \begin{center}
@@
 under $\chk$, we deduce $\val(G,\chk(I)) = I \in M[G]$.  Now we can
-use absoluteness of emptyness and successor, this time for $M[G]$, to
+use absoluteness of emptiness and successor, this time for $M[G]$, to
 conclude that $M[G]$ satisfies Infinity.

### 2019-03-19 ITP/intro.tex
Primeros pasos hacia ITP

--- 2019-01-05 LICS/intro_m.tex
@@
 that given a model $M$ of $\ZF$, any \emph{generic extension} $M[G]$
-obtained by forcing also satisfies $\ZF$. As remarked by
-\citet[][p.250]{kunen2011set} \enquote{[...] in verifying that $M[G]$
-  is a model for set theory, the hardest axiom to verify is
-  [Separation].}  
-The most important achievement of this paper is the
+obtained by forcing also satisfies $\ZF$. As remarked by Kunen
+\cite[p.250]{kunen2011set} \enquote{[...] in verifying that $M[G]$ is
+  a model for set theory, the hardest axiom to verify is
+  [Separation].}  The most important achievement of this paper is the
 mechanization in the proof assistant \emph{Isabelle} of a proof of the
-Axiom of Separation in 
-generic extensions by using the ``fundamental'' theorems of forcing.
-En route
-to this, we also formalized the satisfaction by $M[G]$ of
-Extensionality, Foundation, and Union. As a consequence of Separation
-we were able to formalize the proof of the Powerset Axiom; finally,
-the Axiom of Infinity was proved under extra assumptions. % We have already
+Axiom of Separation in generic extensions by using the ``fundamental''
+theorems of forcing.  En route to this, we also formalized the
+satisfaction by $M[G]$ of Extensionality, Foundation, and Union. As a
+consequence of Separation we were able to formalize the proof of the
+Powerset Axiom; finally, the Axiom of Infinity was proved under extra
+assumptions. % We have already
 % proved Pairing in the first report of our project
@@
 results and the material upon which they are based (definitions,
-propositions, lemmas), but ommiting the proofs. In this work we
+propositions, lemmas), but omitting the proofs. In this work we
 partially adhere to this vision to delineate our formalization
@@
 and generic extensions. As far as we know, the development of
-\citet{JFR6232} in homotopy type theory for constructing generic
+Quirin and Tabareau \cite{JFR6232} in homotopy type theory for constructing generic
 extensions in a sheaf-theoretic setting is the unique mechanization of
@@
 extended Paulson's library with:
-\begin{inlinelist}
+\begin{enumerate}
 \item renaming of variables for \emph{internalized} formulas, which
@@
   models of $\ZF$;
-\end{inlinelist} 
+\end{enumerate}
 the last item was already communicated in the
@@
 We briefly describe the contents of each
-section. Section~\ref{sec:isabelle} contains the bare minimun
+section. Section~\ref{sec:isabelle} contains the bare minimum
 requirements to understand the (meta)logics used in Isabelle. Next, an
@@
 In Section~\ref{sec:generic-extensions} the generic extensions are
-succintly reviewed and how the treatment of well founded recursion in
+succinctly reviewed and how the treatment of well founded recursion in
 Isabelle was enhanced. We take care of the ``easy axioms'' in

### 2019-03-19 ITP/prior-results.tex
Primeros pasos hacia ITP

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
\section{Prior results}

Paulson's work provides an appropriate starting point for our
developments.

There are other formalizations of set theory (Isabelle/HOL, Mizar, Automath).

To the best of our knowledge, only preliminary approaches to the
formalization of forcing exist; v.g. \cite{Quirin}, which has already been
discussed in our previous work \cite{2018arXiv180705174G}.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "Separation_In_MG"
%%% ispell-local-dictionary: "american"
%%% End:

### 2019-03-19 ITP/renaming.tex
Primeros pasos hacia ITP

--- 2019-01-10 LICS/renaming.tex
@@
 formula $\phi$ gives the least natural number containing all the free
-variables in $\phi$. Following \citet{fiore-abssyn}, one can
+variables in $\phi$. Following Fiore et al. \cite{fiore-abssyn}, one can
 understand the arity of a formula as the context of the free

### 2019-03-19 ITP/zfc-axioms.tex
Primeros pasos hacia ITP

--- 2018-12-26 LICS/zfc-axioms.tex
@@
 axioms of Extensionality, Pairing, Union, Powerset, Foundation, Infinity, and two
-axiom-schemes collectively referred to as Axiom of Separation:
-\emph{For every $A$, $a_1,\dots,a_n$,  and  a formula
+axiom-schemes collectively referred as
+(a) \emph{For every $A$, $a_1,\dots,a_n$,  and  a formula
   $\psi(x_0,x_1,\dots,x_n)$, there exists $\{a\in  A:
   \psi(a,a_1,\dots,a_n)\}$}, 
-and Axiom of Replacement: \emph{For every $A$, $a_1,\dots,a_n$,  and
+and (b) Axiom of Replacement: \emph{For every $A$, $a_1,\dots,a_n$,  and
   a formula   $\psi(x,z,x_1\dots,x_n)$, if 
@@
 A model of the theory $\ZF$ consists of a pair $\lb M,E\rb$ where $M$
-is a set and $E$ is a binary relation on $M$. Forcing is a technique
+is a set and $E$ is a binary relation on $M$ satisfying the
+axioms. Forcing is a technique 
 to extend very special kind of models, where $M$ is a countable
@@
 \end{isabelle}
-The rest of the axioms are also
-included. We single out Separation and Replacement: These are written
-for formulas with at most one extra parameter (meaning $n\leq 1$ in the
-above $\psi$), where \isatt{sats\,::\,"[i,i,i]=>o"} is the satisfaction
-relation between a set model with an environment and an internalized
-formula. Thanks to Pairing, these 
-versions are equivalent to the usual formulations. We are only able to
-prove that the generic extension satisfies Separation for any particular
-number of parameters, but not in general. This is a consequence that
-induction on terms of type \tyo{} is not available.
+The rest of the axioms are also included. We single out Separation and
+Replacement: These are written for formulas with at most one extra
+parameter (meaning $n\leq 1$ in the formula $\psi$ of the
+axiom-schemes above). Thanks to Pairing, these versions are equivalent
+to the usual formulations. We are only able to prove that the generic
+extension satisfies Separation for any particular number of
+parameters, but not in general. This is a consequence that induction
+on terms of type \tyo{} is not available.
 

### 2019-03-26 ITP/generic-extensions.tex
Primeros avances en la corrección

--- 2018-12-21 LICS/generic-extensions.tex
@@
 We will swiftly review some definitions in order to reach the concept
-of \emph{generic extension}. Given a ctm $M$ of $\ZF$, a forcing
-notion $\lb\PP,\leq,\1\rb$, and a filter
-$G\sbq\PP$, a new set $M[G]$ is defined. Each element $a\in M[G]$ is
-determined by its \emph{name} $\dot a$ of $M$. Actually, the structure of
+of \emph{generic extension}. As first preliminary definitions, a \emph{forcing
+notion} $\lb\PP,\leq,\1\rb$ is simply a preorder with top, and a \emph{filter}
+$G\sbq\PP$ is an increasing subset which is downwards
+compatible. Given a ctm $M$ of $\ZF$, a forcing
+notion in $M$, and a filter $G$, a new set $M[G]$ is defined. Each
+element $a\in M[G]$ is 
+determined by its \emph{name} $\dot a$ in $M$. Actually, the structure of
 each $\dot a$ is used to construct $a$. They are related by a
-map $\val$ that takes $G$ a parameter:
+map $\val$ that takes $G$ as a parameter:
 \[
@@
 Section~\ref{sec:easy-axioms}) and includes $M\cup\{G\}$. Nevertheless
-there is no a priori reason for $M[G]$ to satisfy the rest of the
-$\ZF$ axioms. The original insight by Cohen was to define the notion
+there is no a priori reason for $M[G]$ to satisfy either Separation, Powerset
+or Replacement. The original insight by Cohen was to define the notion
 of \emph{genericity} for a filter $G\sbq\PP$ and to prove that
@@
 filter is generic if it intersects all the dense sets in $M$; in
-\cite{2018arXiv180705174G} we formalized Rassiowa-Sikorski lemma which
+\cite{2018arXiv180705174G} we formalized the Rasiowa-Sikorski lemma which
 proves the existence of generic filters for ctms.

### 2019-03-26 ITP/intro.tex
Primeros avances en la corrección

@@
 partially adhere to this vision to delineate our formalization
-strategy:
-Since the proofs that the  axioms hold in generic extensions
+strategy: since the proofs that the  axioms hold in generic extensions
 are independent of the \emph{proofs} of the fundamental theorems of
@@
 \item a variant of the  principle of dependent choices and a version
-  of Rasiowa-Sikorski, which 
+  of Rasiowa-Sikorski's theorem, which 
   ensures the existence of generic filters for countable and transitive

### 2019-03-26 ITP/isabelle.tex
Primeros avances en la corrección

--- 2019-01-10 LICS/isabelle.tex
@@
 \]
-where we are omitting the ``typing'' assumtions on $n$ and $x$.
+where we are omitting the ``typing'' assumptions on $n$ and $x$.
 
@@
  of internalized
-formulas is defined by recursion and hence it is possible to perform
-inductive arguments using them. In this sense, the object-logic level
+formulas is inductively defined and hence it is possible to perform
+inductive arguments on them, such as the definition of the renaming map. In this sense, the object-logic level
 is further divided into \emph{internal} and \emph{external}

### 2019-03-26 ITP/renaming.tex
Primeros avances en la corrección

@@
 formula $\phi$ gives the least natural number containing all the free
-variables in $\phi$. Following Fiore et al. \cite{fiore-abssyn}, one can
-understand the arity of a formula as the context of the free
+variables in $\phi$. Following Fiore et al. \cite{fiore-abssyn}, one
+can understand the arity of a formula as the context of the free
 variables; notice that the arity of $\forallFm{\phi}$ is the
-predecessor of the arity of $\phi$. Renamings are, consequently,
-mappings between finite sets; since we can think of $\mathsf{succ}(n)$
-as the coproduct $1+n = \{0\} \cup \{1,\dots,n\}$, then given a
-renaming $f \colon n \to m$, the 
-unique morphism $\mathsf{id}_1+f \colon 1+n \to 1+m$ is used to rename
-free variables in a quantified formula.
+predecessor of the arity of $\phi$. In order to understand renamings,
+it is helpful to think of $\mathsf{succ}(n)$ as the coproduct
+$1+n = \{0\} \cup \{1,\dots,n\}$; given a renaming $f \colon n \to m$,
+the unique morphism $\mathsf{id}_1+f \colon 1+n \to 1+m$ is used to
+rename free variables in a quantified formula.
 
@@
 \ {\isachardoublequoteopen}ren{\isacharparenleft}Member{\isacharparenleft}x{\isacharcomma}y{\isacharparenright}{\isacharparenright}\ {\isacharequal}\isanewline
-\ \ {\isacharparenleft}{\isasymlambda}\ n\ {\isasymin}\ nat\ {\isachardot}\ {\isasymlambda}\ m\ {\isasymin}\ nat{\isachardot}\ {\isasymlambda}f\ {\isasymin}\ n\ {\isasymrightarrow}\ m{\isachardot}\ Member\ {\isacharparenleft}f{\isacharbackquote}x{\isacharcomma}\ f{\isacharbackquote}y{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ {\isacharparenleft}{\isasymlambda}n\ {\isasymin}\ nat\ {\isachardot}\ {\isasymlambda}\ m\ {\isasymin}\ nat{\isachardot}\ {\isasymlambda}f\ {\isasymin}\ n\ {\isasymrightarrow}\ m{\isachardot}\ Member\ {\isacharparenleft}f{\isacharbackquote}x{\isacharcomma}\ f{\isacharbackquote}y{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
 \ \isanewline
 \ {\isachardoublequoteopen}ren{\isacharparenleft}Equal{\isacharparenleft}x{\isacharcomma}y{\isacharparenright}{\isacharparenright}\ {\isacharequal}\isanewline
-\ \ {\isacharparenleft}{\isasymlambda}\ n\ {\isasymin}\ nat\ {\isachardot}\ {\isasymlambda}\ m\ {\isasymin}\ nat{\isachardot}\ {\isasymlambda}f\ {\isasymin}\ n\ {\isasymrightarrow}\ m{\isachardot}\ Equal\ {\isacharparenleft}f{\isacharbackquote}x{\isacharcomma}\ f{\isacharbackquote}y{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ {\isacharparenleft}{\isasymlambda}n\ {\isasymin}\ nat\ {\isachardot}\ {\isasymlambda}\ m\ {\isasymin}\ nat{\isachardot}\ {\isasymlambda}f\ {\isasymin}\ n\ {\isasymrightarrow}\ m{\isachardot}\ Equal\ {\isacharparenleft}f{\isacharbackquote}x{\isacharcomma}\ f{\isacharbackquote}y{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
 \ \isanewline
 \ {\isachardoublequoteopen}ren{\isacharparenleft}Nand{\isacharparenleft}p{\isacharcomma}q{\isacharparenright}{\isacharparenright}\ {\isacharequal}\isanewline
-\ \ {\isacharparenleft}{\isasymlambda}\ n\ {\isasymin}\ nat\ {\isachardot}\ {\isasymlambda}\ m\ {\isasymin}\ nat{\isachardot}\ {\isasymlambda}f\ {\isasymin}\ n\ {\isasymrightarrow}\ m{\isachardot}\ \isanewline
-\ \ \ Nand\ {\isacharparenleft}ren{\isacharparenleft}p{\isacharparenright}{\isacharbackquote}n{\isacharbackquote}m{\isacharbackquote}f{\isacharcomma}\ ren{\isacharparenleft}q{\isacharparenright}{\isacharbackquote}n{\isacharbackquote}m{\isacharbackquote}f{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ {\isacharparenleft}{\isasymlambda}n\ {\isasymin}\ nat\ {\isachardot}\ {\isasymlambda}\ m\ {\isasymin}\ nat{\isachardot}\ {\isasymlambda}f\ {\isasymin}\ n\ {\isasymrightarrow}\ m{\isachardot}\ 
+Nand\ {\isacharparenleft}ren{\isacharparenleft}p{\isacharparenright}{\isacharbackquote}n{\isacharbackquote}m{\isacharbackquote}f{\isacharcomma}\ ren{\isacharparenleft}q{\isacharparenright}{\isacharbackquote}n{\isacharbackquote}m{\isacharbackquote}f{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
 \ \isanewline
 \ {\isachardoublequoteopen}ren{\isacharparenleft}Forall{\isacharparenleft}p{\isacharparenright}{\isacharparenright}\ {\isacharequal}\isanewline
-\ \  {\isacharparenleft}{\isasymlambda}\ n\ {\isasymin}\ nat\ {\isachardot}\ {\isasymlambda}\ m\ {\isasymin}\ nat{\isachardot}\ {\isasymlambda}f\ {\isasymin}\ n\ {\isasymrightarrow}\ m{\isachardot}\ \isanewline
+\ \  {\isacharparenleft}{\isasymlambda}n\ {\isasymin}\ nat\ {\isachardot}\ {\isasymlambda}\ m\ {\isasymin}\ nat{\isachardot}\ {\isasymlambda}f\ {\isasymin}\ n\ {\isasymrightarrow}\ m{\isachardot}\ \isanewline
 \ \ \ Forall\ {\isacharparenleft}ren{\isacharparenleft}p{\isacharparenright}{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isacharbackquote}succ{\isacharparenleft}m{\isacharparenright}{\isacharbackquote}sum{\isacharunderscore}id{\isacharparenleft}n{\isacharcomma}f{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}

### 2019-03-26 ITP/zfc-axioms.tex
Primeros avances en la corrección

@@
   (Sketch) The L\"owenheim-Skolem 
-  Theorem ensures that there is an countable elementary submodel 
+  Theorem ensures that there is a countable elementary submodel 
   $\lb N',E\restr N'\rb\preccurlyeq  \lb N,E\rb$ which must also be
@@
 
-In this stage of our implementation, we chose a presentation of the
+At this stage of our implementation, we chose a presentation of the
 $\ZF$ axioms that would be most
@@
 
-It is also possible define a predicate that states that a set
+It is also possible to define a predicate that states that a set
 satisfies a (possibly infinite) set of formulas, and then to state
 that ``$M$ satisfies $\ZF$'' in a standard way. With the
-aforementioned restriction on parameters, it can be shown that this
+aforementioned restriction on parameters, we have shown that this
 statement is equivalent to the set of assumptions of the locale

### 2019-03-27 ITP/conclusion.tex
Terminando con todas las observaciones del review 1(?)

@@
 can be decomposed in a transfinite, cumulative hierarchy of sets:
-\begin{theorem}
+\begin{theorem}[{\cite[Thms.~I.9.23 and I.9.30]{kunen2011set}}]
   Let $V_{\al}\defi\union\{\P(V_\be) : \be<\al\}$ for each ordinal
@@
 \end{theorem}
-\begin{theorem}[Reflection Principle]\label{th:reflection-principle}
+\begin{theorem}[Reflection Principle, {\cite[Thm.~II.5.3]{kunen2011set}}]\label{th:reflection-principle}
   For every finite $\Phi\sbq\ZF$, $\ZF$ proves: ``There exist

### 2019-03-27 ITP/easy-axioms.tex
Terminando con todas las observaciones del review 1(?)

@@
 Our formal definition is slightly different in order to ease the proof
-of $\pi_a \in M$; as it is defined using Separation, so one needs to
+of $\pi_a \in M$; as it is defined using Separation, one needs to
 define the domain of separation and also internalize the predicate as
@@
 \end{isabelle}
-Since Union is absolute for any transitive class we may conclude that
-$M[G]$ is closed under Union:
+Since Union is absolute for any transitive class (an easy exercise)
+we may conclude that $M[G]$ is closed under Union:
 \begin{isabelle}

### 2019-03-27 ITP/generic-extensions.tex
Terminando con todas las observaciones del review 1(?)

@@
 of \emph{generic extension}. As first preliminary definitions, a \emph{forcing
-notion} $\lb\PP,\leq,\1\rb$ is simply a preorder with top, and a \emph{filter}
+notion} $\lb\PP,\leq,\1\rb$ is simply a preorder with top ($\1$), and a \emph{filter}
 $G\sbq\PP$ is an increasing subset which is downwards

### 2019-03-27 ITP/hacking.tex
Terminando con todas las observaciones del review 1(?)

--- 2019-01-09 LICS/hacking.tex
@@
 \item\label{item:2} Originally the Powerset Axiom was assumed by the
-  locale \isatt{M\_trivial},   we moved this requirement to \isatt{M\_basic}. 
+  locale \isatt{M\_trivial} but it was not needed to prove any result
+  of that locale, so we moved this requirement to \isatt{M\_basic};
+  thus turning \isatt{M\_trivial} more usable.
 \item\label{item:3} We replaced the need that the set of natural
@@
 Because of these changes, some theory files from the
-\isatt{ZF-Constructible} session have been included among ours.
-
-The proof, for instance, that the constructible universe $L$ satisfies
-the modified locale \isatt{M\_trivial} holds with minor
+\isatt{ZF-Constructible} session have been included among ours. The
+proof, for instance, that the constructible universe $L$ satisfies the
+modified locale \isatt{M\_trivial} holds with minor
 modifications. Nevertheless, in order to have a neater presentation,
@@
 \isatt{L\_axioms} and \isatt{Internalize}, and we merged them to form
-the new file  \isatt{Internalizations}.
+the new file \isatt{Internalizations}. We discuss in the conclusion further
+modifications of the constructible session.
 

### 2019-03-27 ITP/proof-of-separation.tex
Terminando con todas las observaciones del review 1(?)

--- 2019-01-09 LICS/proof-of-separation.tex
@@
 \]
-it is enough to provide a name $n\in M$ for this set.
-
-The candidate name is
+it is enough to provide a name $n\in M$ for this set. The candidate name is
 \begin{equation}\label{eq:4}
@@
 
-The last 20 lines of the theory show, using absoluteness, the two
-instances of Separation for $M[G]$:
+The last 20 lines of the theory show, using absoluteness of
+Separation, every instance of Separation for $M[G]$ for formulas with
+at most two parameters:
 
@@
 \ \ \isakeyword{shows}\ \ \isanewline
-\ \ \ \ {\isachardoublequoteopen}{\isasymforall}a{\isasymin}{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharparenright}{\isachardot}\isanewline 
- \ \ \ \ \  separation{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}{\isasymlambda}x{\isachardot}sats{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}{\isasymphi}{\isacharcomma}{\isacharbrackleft}x{\isacharcomma}a{\isacharbrackright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\ \ \ \ {\isachardoublequoteopen}{\isasymforall}a{\isasymin}{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharparenright}{\isachardot} \ \  separation{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}{\isasymlambda}x{\isachardot}sats{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}{\isasymphi}{\isacharcomma}{\isacharbrackleft}x{\isacharcomma}a{\isacharbrackright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
 \end{isabelle}
@@
 considerable fraction is taken by the proof of a closure property of
-the ctm $M$, that involves renaming of an internalized formula; also,
-the handling of the projections \isatt{fst} and \isatt{snd} must be
-done internally. 
-\begin{isabelle}
-\isacommand{lemma}\isamarkupfalse%
-\ sats{\isacharunderscore}fst{\isacharunderscore}snd{\isacharunderscore}in{\isacharunderscore}M{\isacharcolon}\isanewline
-\ \isakeyword{assumes}\ \isanewline
-\  \ {\isachardoublequoteopen}A{\isasymin}M{\isachardoublequoteclose}\ {\isachardoublequoteopen}B{\isasymin}M{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymphi}\ {\isasymin}\ formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}p{\isasymin}M{\isachardoublequoteclose}\ {\isachardoublequoteopen}l{\isasymin}M{\isachardoublequoteclose}\isanewline
-\ \  {\isachardoublequoteopen}o{\isasymin}M{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymchi}{\isasymin}M{\isachardoublequoteclose}\ {\isachardoublequoteopen}arity{\isacharparenleft}{\isasymphi}{\isacharparenright}\ {\isasymle}\ {\isadigit{6}}{\isachardoublequoteclose}\isanewline
-\  \isakeyword{shows}\isanewline
-\  \ {\isachardoublequoteopen}{\isacharbraceleft}sq\ {\isasymin}\ A{\isasymtimes}B\ {\isachardot}\isanewline
-\ \ \ \ \ sats{\isacharparenleft}M{\isacharcomma}{\isasymphi}{\isacharcomma}{\isacharbrackleft}p{\isacharcomma}l{\isacharcomma}o{\isacharcomma}snd{\isacharparenleft}sq{\isacharparenright}{\isacharcomma}fst{\isacharparenleft}sq{\isacharparenright}{\isacharcomma}{\isasymchi}{\isacharbrackright}{\isacharparenright}{\isacharbraceright}\ {\isasymin}\ M{\isachardoublequoteclose}\ \isanewline
-\end{isabelle}
+the ctm $M$, that involves renaming of an internalized formula
+involving a rearrangement of the components of pairs via the
+projections \isatt{fst} and \isatt{snd}.
+% \begin{isabelle}
+% \isacommand{lemma}\isamarkupfalse%
+% \ sats{\isacharunderscore}fst{\isacharunderscore}snd{\isacharunderscore}in{\isacharunderscore}M{\isacharcolon}\isanewline
+% \ \isakeyword{assumes}\ \isanewline
+% \  \ {\isachardoublequoteopen}A{\isasymin}M{\isachardoublequoteclose}\ {\isachardoublequoteopen}B{\isasymin}M{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymphi}\ {\isasymin}\ formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}p{\isasymin}M{\isachardoublequoteclose}\ {\isachardoublequoteopen}l{\isasymin}M{\isachardoublequoteclose}\isanewline
+% \ \  {\isachardoublequoteopen}o{\isasymin}M{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymchi}{\isasymin}M{\isachardoublequoteclose}\ {\isachardoublequoteopen}arity{\isacharparenleft}{\isasymphi}{\isacharparenright}\ {\isasymle}\ {\isadigit{6}}{\isachardoublequoteclose}\isanewline
+% \  \isakeyword{shows}\isanewline
+% \  \ {\isachardoublequoteopen}{\isacharbraceleft}sq\ {\isasymin}\ A{\isasymtimes}B\ {\isachardot}\ \ sats{\isacharparenleft}M{\isacharcomma}{\isasymphi}{\isacharcomma}{\isacharbrackleft}p{\isacharcomma}l{\isacharcomma}o{\isacharcomma}snd{\isacharparenleft}sq{\isacharparenright}{\isacharcomma}fst{\isacharparenleft}sq{\isacharparenright}{\isacharcomma}{\isasymchi}{\isacharbrackright}{\isacharparenright}{\isacharbraceright}\ {\isasymin}\ M{\isachardoublequoteclose}
+% \end{isabelle}
 

### 2019-04-01 ITP/Separation_In_MG.tex
Cambios de acuerdo al review 4 de LICS

@@
   previous version of this paper that helped us to enhance the
-  presentation.}%optional
+  presentation. We are grateful to Alexandre Miquel for his comments
+  on forcing and Boolean-valued models.}%optional
 
@@
 \input{appendix}
-% that's all folks
+
 \end{document}

### 2019-04-01 ITP/conclusion.tex
Cambios de acuerdo al review 4 de LICS

@@
 first because many absoluteness results follows from this, the latter
-for the existence of generic filters. A
-more anecdotal one is that to show that an instance of Separation
-with at most two parameters holds in $M[G]$, one needs to assume a
-particular  six-parameter
-instance in $M$ (four extra parameters can be directly blamed on
-$\forceisa$).  The purpose of identifying those assumptions is to assemble
+for the existence of generic filters. % A
+% more anecdotal one is that to show that an instance of Separation
+% with at most two parameters holds in $M[G]$, one needs to assume a
+% particular  six-parameter
+% instance in $M$ (four extra parameters can be directly blamed on
+% $\forceisa$).
+The purpose of identifying those assumptions is to assemble
 in a locale the specific (instances of) axioms that should satisfy the
@@
 
+Recently Alexandre Miquel suggested us to look into Boolean-valued
+models as an alternative method for extending models of ZF. This is a
+very interesting option to consider, since that would work for any
+model (and not just transitivive ones); some of the others issues (the
+most salient being the use of internalized formulas) are consequences
+of working in Isabelle/ZF and would be inevitable.
+
 %% \section*{Acknowledgment}

### 2019-04-01 ITP/easy-axioms.tex
Cambios de acuerdo al review 4 de LICS

@@
 
-In our first presentation of this project \cite{2018arXiv180705174G},
-we proved that $M[G]$ satisfies Pairing; now we have redone this proof
-in Isar. It is straightforward to show that the generic extension
-$M[G]$ satisfies extensionality and foundation. Showing that it is
-closed under Union depends on $G$ being a filter. Infinity is also
-easy, but it depends in one further assumption.
+In this section we briefly comment the proofs showing that the generic
+extension $M[G]$ satisfies extensionality and foundation. Showing that
+it is closed under Union depends on $G$ being a filter. Infinity is
+also easy, but it depends in one further assumption.
 

### 2019-04-01 ITP/intro.tex
Cambios de acuerdo al review 4 de LICS

@@
 forcing. This contrast with the fruitful use of forcing techniques to
-extend the Curry-Howard isomorphism to classical axioms
+extend the Curry-Howard isomorphism to classical axioms in order to
+get programs for proofs using those axioms
 \cite{Miquel:2011:FPT:2058525.2059614,lmcs:1070}. Moreover, the
@@
 \end{enumerate}
-the last item was already communicated in the
-  first report \cite{2018arXiv180705174G}.
+the last item was already communicated in the first report
+\cite{2018arXiv180705174G} of our project (where we also proved that
+$M[G]$ satisfies Pairing; now we have redone this proof in Isar).
 

### 2019-04-01 ITP/isabelle.tex
Terminando con todas las observaciones del review 1(?)
Cambios de acuerdo al review 4 de LICS

@@
 %% obtained by combining the given constants and operations. 
-It must be observed that predicates (functions with arguments of
-type \tyi{} with values in \tyo) do not correspond to first-order formulas;
-in particular, those are not recursively constructed.
-This will
-have concrete consequences in our strategy to approach the
+It must be observed that predicates over \tyi{} do not correspond to
+first-order formulas; in particular, those are not recursively
+constructed and there is no induction principle associated with them.
+This will have concrete consequences in our strategy to approach the
 development. From the beginning, we had to resort to
-\emph{internalized} formulas \cite[Sect.~6]{paulson_2003}, i.e.\ elements of type $\tyi$ that
-encode first-order formulas with a binary relation symbol, and the
-satisfaction predicate \isatt{sats\,::\,"i\isasymRightarrow i\isasymRightarrow i\isasymRightarrow o"}  between a set
-model with an environment and an internalized formula (where the
-relation symbol is interpreted as membership). The set 
-\isatt{formula::"\tyi"}
- of internalized
-formulas is inductively defined and hence it is possible to perform
-inductive arguments on them, such as the definition of the renaming map. In this sense, the object-logic level
-is further divided into \emph{internal} and \emph{external}
-sublevels.
+\emph{internalized} formulas \cite[Sect.~6]{paulson_2003}, i.e.\
+elements of type $\tyi$ that encode first-order formulas with a binary
+relation symbol, and the satisfaction predicate
+\isatt{sats\,::\,"i\isasymRightarrow i\isasymRightarrow
+  i\isasymRightarrow o"} between a set model with an environment and
+an internalized formula (where the relation symbol is interpreted as
+membership). The set \isatt{formula::"\tyi"} of internalized formulas
+is inductively defined and hence it is possible to perform inductive
+arguments on them, such as the definition of the renaming map. In this
+sense, the object-logic level is further divided into \emph{internal}
+and \emph{external} sublevels. We remark that Isabelle/ZF lacks any
+mechanism to define inductively defined types at the same level of
+\tyi{} or \tyo.
 

### 2019-04-01 ITP/Separation_In_MG.tex
IEEE
link a sources
typo cerca de nombre Manu

@@
 
-\author{Emmanuel Gunther}{FAMAF, Universidad Nacional de Córdoba, [optional: Address], Argentina}{gunther@famaf.unc.edu.ar}{}{Partially supported by a CONICET scholarship}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
+\author{Emmanuel Gunther}{FAMAF, Universidad Nacional de Córdoba, Argentina}{gunther@famaf.unc.edu.ar}{}{Partially supported by a CONICET scholarship}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
 Please provide at least the name of the affiliation and the country.
@@
 
-\supplement{The whole Isabelle/ZF development is available at \url{https://arxiv.org/abs/1901.03313}}%optional, e.g.
+\supplement{The whole Isabelle/ZF development is available at \url{https://cs.famaf.unc.edu.ar/\~{}mpagano/forcing/}}%optional, e.g.
 related research data, source code, ...
@@
 \input{header-draft}
+\IEEEtrue
 
@@
 \onecolumn
-\appendix[A short overview of our development]
+\appendix
+\section{A short overview of our development}
 \input{appendix}

### 2019-10-15 ITP/recursion.tex
Primeros pasos hacia ITP

--- 2019-01-10 LICS/recursion.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
-\subsection{Recursion and values of names}
+\subsection*{Recursion and values of names}
 
@@
 
-As is well-known, the principle of  recursion on
-well-founded relations \cite[p.~48]{kunen2011set} allows us to define
-a recursive function $F \colon A\to A$ by choosing a well-founded
-relation $R \subseteq A\times A$ and a functional
-$H\colon A\times (A \to A) \to A$ satisfying
-$F(a)=H(a,F\!\upharpoonright\!(R^{-1}(a)))$. \citet{paulson1995set}
-made this principle available in Isabelle/ZF via the the operator
-\isa{wfrec}. The formalization of the corresponding functional
-$\mathit{Hv}$ for $\val$ is straightforward:
+Recursive definitions such as this are justified by the principle of
+recursion on well-founded relations \cite[p.~48]{kunen2011set}; this
+principle allows us to define a recursive function $F \colon A\to A$
+by choosing a well-founded relation $R \subseteq A\times A$ and a
+functional $H\colon A\times (A \to A) \to A$ satisfying
+$F(a)=H(a,F\!\upharpoonright\!(R^{-1}(a)))$. Paulson
+\cite{paulson1995set} made this principle available in Isabelle/ZF via
+the the operator \isa{wfrec}. The formalization of the corresponding
+functional $\mathit{Hv}$ for $\val$ is straightforward:
 %
@@
 characterized by a recursively defined predicate. The well-founded
-relation used to justify Equation~\eqref{eq:val} is
+relation used to justify Equation~\eqref{eq:val} is $\mathrel{\mathit{ed}}$
+defined by
 \[ x \mathrel{\mathit{ed}} y \iff \exists p . \lb x,p\rb\in y. \] In
@@
 instance, in order to apply $\val$ to arguments that we can't assume
-that are in $M$, see Eq.~(\ref{eq:val-of-m}) below).
-
-The remedy is to restrict $\mathit{ed}$ to the
-transitive closure of the actual parameter:
+that are in $M$, see Eq.~(\ref{eq:val-of-m}) below); so we define
 \begin{isabelle}
@@
 \isanewline
-\ val\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}i{\isasymRightarrow}i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
+\ edrel\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}i{\isachardoublequoteclose}\ \isakeyword{where}
+\ {\isachardoublequoteopen}edrel{\isacharparenleft}A{\isacharparenright}{\isacharequal}{\isacharequal}\ \{<x,y> . <x,y> {\isasymin} A{\isasymtimes}A {\isasymand} ed(x,y) \}
+\end{isabelle}
+
+The remedy is to restrict $\mathit{ed}$ to the
+transitive closure (\isatt{eclose}) of the actual parameter:
+\begin{isabelle}
+\isacommand{definition}\isamarkupfalse%
+\isanewline
+\ val\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}i{\isasymRightarrow}i{\isachardoublequoteclose}\ \isakeyword{where}
 \ {\isachardoublequoteopen}val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}{\isacharequal}{\isacharequal}\ wfrec{\isacharparenleft}edrel{\isacharparenleft}eclose{\isacharparenleft}{\isacharbraceleft}{\isasymtau}{\isacharbraceright}{\isacharparenright}{\isacharparenright}{\isacharcomma}{\isasymtau}{\isacharcomma}Hv{\isacharparenleft}G{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
@@
 function on some argument $a$, one can restrict the relation to some
-ambient set if it includes $a$ and all of its predecessors.
+ambient set if it includes $a$ and all of its predecessors (i.e. every
+element in the image of the transitive closure of the relation on $a$,
+\isatt{{\isacharparenleft}r{\isacharcircum}{\isacharplus}{\isacharparenright}{\isacharminus}{\isacharbackquote}{\isacharbackquote}{\isacharbraceleft}a{\isacharbraceright}}).
 \begin{isabelle}
@@
   \ def{\isacharunderscore}val{\isacharcolon}\isanewline
-  \ {\isachardoublequoteopen}val{\isacharparenleft}G{\isacharcomma}x{\isacharparenright}\ {\isacharequal}\ {\isacharbraceleft}val{\isacharparenleft}G{\isacharcomma}t{\isacharparenright}\ {\isachardot}{\isachardot}\ t{\isasymin}domain{\isacharparenleft}x{\isacharparenright}\ {\isacharcomma}\isanewline
-\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ {\isasymexists}p{\isasymin}P{\isachardot}\ {\isacharless}t{\isacharcomma}p{\isachargreater}{\isasymin}x\ {\isasymand}\ p{\isasymin}G\ {\isacharbraceright}{\isachardoublequoteclose}
+  \ {\isachardoublequoteopen}val{\isacharparenleft}G{\isacharcomma}x{\isacharparenright}\ {\isacharequal}\ {\isacharbraceleft}val{\isacharparenleft}G{\isacharcomma}t{\isacharparenright}\ {\isachardot}{\isachardot}\ t{\isasymin}domain{\isacharparenleft}x{\isacharparenright}\ {\isacharcomma}\ {\isasymexists}p{\isasymin}P{\isachardot}\ {\isacharless}t{\isacharcomma}p{\isachargreater}{\isasymin}x\ {\isasymand}\ p{\isasymin}G\ {\isacharbraceright}{\isachardoublequoteclose}
 \end{isabelle}

### 2019-10-15 LSFA/arXiv/gen-ext.tex

@@
 \section{The generic extension}
-In the following subsections we show that every preorder in a ctm
-admits a generic filter, and develop the machinery of names.
+
+Cohen's technique of forcing consists of constructing new models of
+$\ZFC$ by adding a \emph{generic} subset $G$ of the forcing notion $P$
+(a preorder with top). Given a model $M$ of $\ZFC$, the extension with
+the generic subset $G$ is called \emph{the generic extension} of $M$,
+denoted $M[G]$.  In this section we introduce all the necessary
+concepts and results for defining $M[G]$; namely, we show, using
+Rasiowa-Sikorski, that every preorder in a ctm admits a generic filter
+and also develop the machinery of names. As an application of the
+latter, we prove some basic
+results about the generic extension.
 
@@
 \ \isanewline
-  \ \ cg{\isacharcolon}\ countable{\isacharunderscore}generic\ P\ leq\ uno\ {\isacharquery}D\ \isanewline
+  \ \ cg{\isacharcolon}\ countable{\isacharunderscore}generic\ P\ leq\ one\ {\isacharquery}D\ \isanewline
  \isacommand{by}\isamarkupfalse%
@@
 \end{isabelle}
-\noindent
+%
 and then a  $\calD$-generic filter given by Rasiowa-Sikorski will be $M$-generic by construction.

### 2019-10-15 LSFA/arXiv/isabelle-zf.tex
versión para arXiv
saco comentarios

--- 2018-06-25 LSFA/isabelle-zf.tex
@@
 %-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
-\section{Isabelle/ZF}
+\section{Isabelle/ZF}\label{sec:isabellezf}
 
 Let us introduce briefly Paulson's formalization of ZF
-\cite{paulson2017isabelle} in Isabelle and aspects of his formal proof
-for the relative consistency of the Axiom of Choice
+\cite{paulson2017isabelle} in Isabelle and the main aspects of his
+formal proof for the relative consistency of the Axiom of Choice
 \cite{paulson_2003}; we will only focus on those aspects that are
@@
 %% Isabelle/ZF is a theory built upon the
-%% core library FOL of classical first order logic.
-Isabelle/ZF includes a development of classical first order logic,
+%% core library FOL of classical first-order logic.
+Isabelle/ZF includes a development of classical first-order logic,
 FOL. Both of them are  built upon the core library \emph{Pure}.
 
-In Isabelle/ZF sets are \emph{individuals}, that is terms of type
-\isatt{i} and formulas have type \isatt{o} 
-(akin to a \emph{Bool} type, but at the object level). 
-The  axiomatization 
-of $\ZFC$ in Isabelle/ZF proceeds by postulating  a binary predicate
-\isatt{\ensuremath{\in}} and   several set
-constructors (terms and function with values in \isatt{i}) corresponding to the  empty set
-(the constant \isatt{\isadigit{0}}), powersets, and one further constant
+In Isabelle/ZF sets are \emph{individuals}, i.e.\ terms of type
+\isatt{i} and formulas have type \isatt{o} (akin to a \emph{Bool}
+type, but at the object level).  The axiomatization of $\ZFC$ in
+Isabelle/ZF proceeds by postulating a binary predicate
+\isatt{\ensuremath{\in}} and several set constructors (terms and
+functions with values in \isatt{i}) corresponding to the empty set (the
+constant \isatt{\isadigit{0}}), powersets, and one further constant
 \isatt{inf} for an infinite set. The axioms, being formulas, are terms
-of type \isatt{o}; 
-the foundation axiom, for example, is formalized as (the universal
-closure of) 
-\isa{{\isachardoublequoteopen}A\ {\isacharequal}\ {\isadigit{0}}\
-  {\isasymor}%
+of type \isatt{o}; the foundation axiom, for example, is formalized as
+(the universal closure of) \isa{{\isachardoublequoteopen}A\
+  {\isacharequal}\ {\isadigit{0}}\ {\isasymor}%
   \ {\isacharparenleft}{\isasymexists}x{\isasymin}A{\isachardot}\ %
@@
   y{\isasymnotin}A{\isacharparenright}%
-  {\isachardoublequoteclose}}. Besides the axioms, Isabelle/ZF also
-introduces several definitions (for example, pairs and sets defined by
-comprehension using separation) and syntactic abbreviations to keep
-the formalization close to the customary manner of doing mathematics.
-Working with the library and extending it is quite
-straightforward. 
-As an example, we introduce  a new term-former which is combination of
-instances of replacement and separation to speak of the image of a function over some set
-defined by comprehension, namely $\{b(x): x\in A\text{ and }Q(x)\}$:
-%% Since we find a need to combine the replacement and
-%% separation schemes to speak of the image of a function over some set
-%% defined by comprehension, we introduced a new term-former:
+  {\isachardoublequoteclose}}. %
+Besides the axioms, Isabelle/ZF also introduces several definitions
+(for example, pairs and sets defined by comprehension using
+separation) and syntactic abbreviations to keep the formalization
+close to the customary manner of doing mathematics.  Working with the
+library and extending it is quite straightforward.  As an example, we
+introduce a new term-former (which is a combination of instances of
+replacement and separation) denoting the image of a function over a
+set defined by comprehension, namely
+$\{b(x): x\in A\text{ and }Q(x)\}$:
+%
+%
+%
 \begin{isabelle}
@@
 \end{isabelle}
-\noindent %% It is more convenient to add an abbreviation for a simpler
-%% writing and a nicer reading, so we use
+\noindent %
+%
 We are then able to add the abbreviation \isa{{\isacharbraceleft}b\
@@
 \isacommand{lemma}\isamarkupfalse%
-\ Sep{\isacharunderscore}and{\isacharunderscore}Replace{\isacharcolon}\ {\isachardoublequoteopen}{\isacharbraceleft}b{\isacharparenleft}x{\isacharparenright}\ {\isachardot}{\isachardot}\ x{\isasymin}A{\isacharcomma}\ P{\isacharparenleft}x{\isacharparenright}\ {\isacharbraceright}\ {\isacharequal}\ {\isacharbraceleft}b{\isacharparenleft}x{\isacharparenright}\ {\isachardot}\ x{\isasymin}{\isacharbraceleft}y{\isasymin}A{\isachardot}\ P{\isacharparenleft}y{\isacharparenright}{\isacharbraceright}{\isacharbraceright}{\isachardoublequoteclose}
+\ Sep{\isacharunderscore}and{\isacharunderscore}Replace{\isacharcolon}\ {\isachardoublequoteopen}{\isacharbraceleft}b{\isacharparenleft}x{\isacharparenright}\ {\isachardot}{\isachardot}\ x{\isasymin}A{\isacharcomma}\ Q{\isacharparenleft}x{\isacharparenright}\ {\isacharbraceright}\ {\isacharequal}\ {\isacharbraceleft}b{\isacharparenleft}x{\isacharparenright}\ {\isachardot}\ x{\isasymin}{\isacharbraceleft}y{\isasymin}A{\isachardot}\ Q{\isacharparenleft}y{\isacharparenright}{\isacharbraceright}{\isacharbraceright}{\isachardoublequoteclose}
 \end{isabelle}
 
-%% We are not discussing how, if possible at all, to express that some
-%% set satisfies some axiom. For example, we might well be interested in
-%% proving $\emptyset \models \mathit{foundation}$, but there is no
-%% sensible way to state this, without encoding formulas and axioms in
-%% ZFC.\footnote{Indeed Paulson has encoded formulas using a very general
-%%   method to encode recursive definitions. In our development, although
-%%   not reported in this paper, we use this internalization.}
-%% 
-%% \fbox{ You need this in order to formalize $\models$ as a predicate of
-%%   type $i\ent i \ent o$. (Don't know how to put that here.)
-%% }
+%
+%
+%
+%
+%
+%
+%
+%
+%
+%
+%
 
-%% In
-%% Isabelle/ZF there are, however, relativized versions of the
-%% axioms. This alternative statement of the axioms are relativized with
-%% respect to classes, which in Isabelle/ZF corresponds to predicates over
-%% sets
+%
+%
+%
+%
+%
 We now discuss relativization in Isabelle/ZF. Relativized versions of the
@@
 \ foundation{\isacharunderscore}ax\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharparenleft}i{\isacharequal}{\isachargreater}o{\isacharparenright}\ {\isacharequal}{\isachargreater}\ o{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
-\ \ \ \ {\isachardoublequoteopen}foundation{\isacharunderscore}ax{\isacharparenleft}M{\isacharparenright}\ {\isacharequal}{\isacharequal}\isanewline
-\ \ \ \ \ \ \ \ {\isasymforall}x{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ {\isacharparenleft}{\isasymexists}y{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ y{\isasymin}x{\isacharparenright}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\isasymexists}y{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ y{\isasymin}x\ {\isacharampersand}\ {\isachartilde}{\isacharparenleft}{\isasymexists}z{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ z{\isasymin}x\ {\isacharampersand}\ z\ {\isasymin}\ y{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+ {\isachardoublequoteopen}foundation{\isacharunderscore}ax{\isacharparenleft}M{\isacharparenright}\ {\isacharequal}{\isacharequal}\isanewline
+  {\isasymforall}x{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ {\isacharparenleft}{\isasymexists}y{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ y{\isasymin}x{\isacharparenright}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\isasymexists}y{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ y{\isasymin}x\ {\isacharampersand}\ {\isachartilde}{\isacharparenleft}{\isasymexists}z{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ z{\isasymin}x\ {\isacharampersand}\ z\ {\isasymin}\ y{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
 \end{isabelle}
 
-% ∀x[c]
+%
 \noindent The relativized quantifier
@@
 
-% Let us ponder if one can state, let alone prove, that some set \isa{M
-%   :: i} satisfies a statement \isa{\ensuremath{\phi} :: o}. Since
-% \isa{o} does not have an inductive definition, we cannot define the
-% satisfaction of \isa{\ensuremath{\phi}}.
+%
+%
+%
+%
 
-% As we have outlined above, relativization is a key concept when
-% proving that axioms satisfied by some model $M$ are also satisfied in
-% another model constructed from $M$. In this paper, we show the complete
-% proof that if a set \isa{M :: i} satisfies the
+%
+%
+%
+%
 
-Mathematical texts usually start by fixing the context which defines
-parameters and assumptions necessary to develop theorems
+Mathematical texts usually start by fixing a context that defines
+parameters and assumptions needed to develop theorems
 and results. In Isabelle the way of defining contexts is through
 \emph{locales}~\cite{ballarin2010tutorial}.
-Locales can be combined and extended adding more parameters and assuming
+Locales can be combined and extended by adding more parameters and assuming
 more facts, leading to a new locale. For example a context describing
@@
 
-% locales
+Let us close this section with a brief comment about the facilities
+provided by the Isabelle framework. The edition is done in an IDE
+called \texttt{jEdit}, which is bundled with the standard Isabelle
+distribution; it offers the user a fair amount of tools in order to
+manage theory files, searching for theorems and concepts spread
+through the source files, and includes tracing utilities for the
+automatic tools. A main feature is a window showing the \emph{proof
+  state}, where the active (sub)goals are shown, along with the
+already obtained results and possibly errors.
+
+Isabelle proofs can be written in two dialects.  The older one, and
+also more basic, follows a procedural approach, where one applies
+several tactics in order to decompose the goal into simpler ones and
+then solving them (with the aid of automation); the original work by
+Paulson used this method. Under this approach proofs are constructed
+top-down resulting in proof-scripts that conceal the mathematical
+reasoning behind the proof, since the intermediate steps are only
+shown in the proof state. For this reason, the proof language
+\emph{Isar} was developed, starting with Wenzel's
+work~\cite{DBLP:conf/tphol/Wenzel99}. Isar is mostly declarative, and
+its main purpose is to construct \emph{proof documents} that (in
+principle) can be read and understood without the need of running the
+code.
+
+We started this development using the procedural approach, but soon
+after we realized that for our purposes the Isar language was far more
+appropriate.
 

### 2019-10-15 LSFA/conclusion.tex
punto final!!

@@
 Other milestones in this formalization project
-involve 
+involve:
 \begin{enumerate}
@@
 \end{enumerate}
-The theory is very modular and this is
+The theory is modular and this is
 witnessed by the fact 
-that the last goal does not depend on the proof of the Fundamental
+that the last goal neither depends on the proof of the Fundamental
 Theorem nor on the definition of the forcing relation. Our next task
-will be to obtain the last goal enumerated.
+will be to obtain the last goal in that enumeration.
 
 To this end, we will develop an interface between Paulson's
-relativization results and countable models $M$ of $\ZFC$. This will show
-that every such $M$ is closed under well-founded recursion and, in
+relativization results and countable models of $\ZFC$. This will show
+that every ctm $M$ is closed under well-founded recursion and, in
 particular, that contains names for each of its
@@
 nontrivially). As a part of the new formalization, we will provide
-Isar versions of the longer applicative proofs presented in this work
+Isar versions of the longer applicative proofs presented in this work.
 
+\ack{We'd like to thank the anonymous referees for reading the paper
+  carefully and for their detailed and constructive criticism.}
 %%% Local Variables:

### 2019-11-01 ITP/generic-extensions.tex

--- 2019-03-26 ITP/generic-extensions.tex
@@
 We will swiftly review some definitions in order to reach the concept
-of \emph{generic extension}. As first preliminary definitions, a \emph{forcing
-notion} $\lb\PP,\leq,\1\rb$ is simply a preorder with top, and a \emph{filter}
-$G\sbq\PP$ is an increasing subset which is downwards
-compatible. Given a ctm $M$ of $\ZF$, a forcing
+of \emph{generic extension}. As first preliminary definitions, a
+\emph{forcing notion} $\lb\PP,\leq,\1\rb$ is simply a preorder with
+top ($\1$), and a \emph{filter} $G\sbq\PP$ is an increasing subset
+which is downwards compatible. Given a ctm $M$ of $\ZF$, a forcing
 notion in $M$, and a filter $G$, a new set $M[G]$ is defined. Each
-element $a\in M[G]$ is 
-determined by its \emph{name} $\dot a$ in $M$. Actually, the structure of
-each $\dot a$ is used to construct $a$. They are related by a
-map $\val$ that takes $G$ as a parameter:
+element $a\in M[G]$ is determined by its \emph{name} $\dot a$ in $M$;
+sometimes names are elements of $M$ that satisfy a recursively defined
+predicate, we prefer to assume that any element of $M$ can be a
+name. Actually, the structure of each $\dot a$ is used to construct
+$a$. They are related by a map $\val$ that takes $G$ as a parameter:
 \[

### 2019-11-01 ITP/renaming.tex

@@
 
-All our uses of this lemma involve concrete renamings on small
-numbers, but we also tested it with more abstract ones for arbitrary
-numbers. All the renamings of the first kind follow the same pattern
-and, more importantly, share equal proofs. We would like to develop
+The use of this lemma involves some repetitive tasks (mainly proving
+that the renaming is in fact a function). We would like to develop
 some \texttt{ML} tools in order to automatize this.

### 2019-11-01 LICS/conclusion.tex
typo

@@
 \label{sec:conclusions-future-work}
+
+In a gross simplification, there are two aspects to a formalization
+project like this one: thematic and programmatic. The first concerns
+the handling of all the theoretical concepts and results in the
+subject, while the second involves the practical issues of the
+implementation and design. In the case of forcing, the main intricacy
+lies in the first aspect. In this sense, following a sensible
+presentation of the material is key.  The authoritative reference 
+on the subject during the last 30 years has been Kunen's classical
+\cite{kunen1980}, and  we have followed a recent rewrite
+\cite{kunen2011set} 
+of that  textbook, which presents the material in the same sharp 
+style but offering a lot of details. In some sense this project
+wouldn't exist without this book.
 
@@
 Axiom of Infinity; it can be proved in a direct but cumbersome way. We
-preferred to wait to have a full-fledged inteface between ctms and the
+preferred to wait to have a full-fledged interface between ctms and the
 locales providing recursive constructions from Paulson's

### 2019-11-01 LICS/generic-extensions.tex

@@
 We will swiftly review some definitions in order to reach the concept
-of \emph{generic extension}. Given a ctm $M$ of $\ZF$, a forcing
-notion $\lb\PP,\leq,\1\rb$, and a filter
-$G\sbq\PP$, a new set $M[G]$ is defined. Each element $a\in M[G]$ is
+of \emph{generic extension}. As first preliminary definitions, a \emph{forcing
+notion} $\lb\PP,\leq,\1\rb$ is simply a preorder with top, and a \emph{filter}
+$G\sbq\PP$ is an increasing subset which is downwards
+compatible. Given a ctm $M$ of $\ZF$, a forcing
+notion in $M$, and a filter $G$, a new set $M[G]$ is defined. Each
+element $a\in M[G]$ is 
 determined by its \emph{name} $\dot a$ of $M$. Actually, the structure of
 each $\dot a$ is used to construct $a$. They are related by a
-map $\val$ that takes $G$ a parameter:
+map $\val$ that takes $G$ as a parameter:
 \[

### 2019-11-01 LICS/intro.tex

@@
 formalizaton we have followed a recent rewrite \cite{kunen2011set}
-of that  textbook,which presents the material in the same sharp 
+of that  textbook, which presents the material in the same sharp 
 style but offering a lot of details. In some sense this project
-wouldn't exist without this book. The reader interested in an
-introduction to the subject can check
+wouldn't exist without this book. As alternative, introductory
+resources, the  interested reader can check
 \cite{chow-beginner-forcing}; also, the book \cite{weaver2014forcing}
 contains a thorough treament minimizing the technicalities.
+
+In this work we address a substantial part the problem of formalizing
+the proof that given a model of $\ZF$, its extensions by forcing, 
+\emph{generic extensions},
+satisfy the same axioms. In particular, the first axiom that
+genuinely calls for a use of the forcing machinery is that of
+\emph{Separation} (\emph{Aussonderungsaxiom}), otherwise known as
+``Axiom of (Restricted) Comprehension''  or ``Specification.'' In
+\cite{2018arXiv180705174G} it was shown that  generic extensions of
+models of $\ZF$ 
+satisfy the Axiom of Pairing, and in this work we show that also
+Extensionality, Foundation, Union, Infinity (under extra assumptions),
+and Separation go through. The
+Axiom of Powerset is not treated in the present paper but it poses no
+extra difficulties; we discuss the Axiom of Replacement in the
+conclusions.
+
+The proof that these axioms hold in generic extension is independent
+of the \emph{proofs} of the ``fundamental'' theorems of forcing. In
+particular, one of the most complex parts of those proofs is the
+definition of the \emph{forcing relation}. It turns out that the
+particular (``implementation'') details of this  definition  does not
+appear in the statement of the fundamental theorems, and our
+formalization works inside a \emph{locale} (set of assumptions) that
+includes the statement of these theorems. The formalization of the
+fundamental theorems of forcing roughly comprises little less than a
+half of our full project.
 

### 2019-11-01 LICS/zfc-axioms.tex
varias observaciones de Manu

@@
 \end{isabelle}
-The rest of the axioms are also
-included. We single out Separation and Replacement: These are written
-for formulas with at most one extra parameter (meaning $n\leq 1$ in the
-above $\psi$). Thanks to Pairing, these 
-versions are equivalent to the usual formulations. We are only able to
-prove that the generic extension satisfies Separation for any particular
-number of parameters, but not in general. This is a consequence that
-induction on terms of type \tyo{} is not available.
+The rest of the axioms are also included. We single out Separation and
+Replacement: These are written for formulas with at most one extra
+parameter (meaning $n\leq 1$ in the formula $\psi$ of the
+axiom-schemes above). Thanks to Pairing, these versions are equivalent
+to the usual formulations. We are only able to prove that the generic
+extension satisfies Separation for any particular number of
+parameters, but not in general. This is a consequence that induction
+on terms of type \tyo{} is not available.
 

### 2019-11-01 LSFA/arXiv/rs.tex

--- 2018-07-12 LSFA/arXiv/rs.tex
@@
-\subsection{A sequence version of $\DC$}\label{sec:sequence-version-dc}
-In order to prove the Rasiowa-Sikorski lemma in
-general, a version of the Axiom of Choice ($\AC$) must be proved. This
-version is known as the \emph{Principle of Dependent Choices}:
+\subsection{A sequence version of Dependent
+  Choices}\label{sec:sequence-version-dc} 
+%% In order to prove the Rasiowa-Sikorski lemma in
+%% general, a consequence of the Axiom of Choice ($\AC$) must be invoked. This
+%% consequence is known as the \emph{Principle of Dependent Choices}:
+The Rasiowa-Sikorski lemma follows naturally from a ``pointed''
+version of   the \emph{Principle of Dependent Choices ($\DC$)} which,
+in turn, is a consequence of the Axiom of Choice ($\AC$). It is
+therefore natural to take as a starting point the theory \isatt{AC}
+which adds the latter axiom to the toolkit of Isabelle/ZF.
+
+The statement we are interested in is the following:
 \begin{quote}
-  ($\DC$) Let $\rr$ be a binary relation on $A$, and $a\in A$. If
+  (Pointed $\DC$) Let $\rr$ be a binary relation on $A$, and $a\in A$. If
   $\forall x\in A.\,  \exists y\in A.\, x\rr y$, then there exists
@@
 
-A different version of $\DC$ (without the constraint $f(0)=a$, but
-allowing a parameter that permits to prove an equivalence with $\AC$)
-was formalized by Krzysztof
-Grabczewski~\cite{DBLP:journals/jar/PaulsonG96} as part of the
-Isabelle/ZF library, along with many statements that follow from or
-are equivalent to $\AC$. Since the proofs there required a greater
-generality, it would have been very contrived to derive this
-``pointed'' version of $\DC$ from that development. Instead, we
-preferred to give a direct proof based on the one in
-Moschovakis' notes \cite{moschovakis1994notes}. Another advantage is that
-this proof depends on the following version of $\AC$ (existence of
-choice functions on $\P(A)\sm \{\mty\}$):
+Two different versions of $\DC$ (called $\DC0$ and $\DC(\kappa)$) have
+already been formalized by Krzysztof
+Grabczewski~\cite{DBLP:journals/jar/PaulsonG96}, as part of a study of
+equivalents of $\AC$ (following Rubin and Rubin
+\cite{rubin1985equivalents}). Nevertheless, those are not convenient
+for our purposes. In fact, the axiom $\DC0$ corresponds essentially to
+our Pointed $\DC$ but without the constraint $f(0)=a$; it is a nice
+exercise to show that $\DC0$ implies Pointed $\DC$, but a
+formalization would have a moderate length. On the other hand,
+$\DC(\kappa)$ is rather different in nature and it is tailored to
+obtain another proposition equivalent to the axiom of choice
+(actually,
+$\AC\longleftrightarrow (\forall \kappa .\;
+\mathrm{Card}(\kappa)\longrightarrow \DC(\kappa))$).  Finally, the
+shortest path from $\AC$ %% as stated in the
+%% corresponding Isabelle theory
+to $\DC0$ using already formalized
+material involves a complicated detour (130+ proof commands spanning
+various files of the \isatt{ZF-AC} theory and going through the Well
+Ordering Theorem and  $\DC(\omega)$), compared to the mere 11 commands from $\AC$ to
+\isatt{AC{\isacharunderscore}func{\isacharunderscore}Pow}. This last
+one is the
+choice principle that we use in our formalization of Pointed $\DC$, and states 
+the existence of
+choice functions (``selectors'') on $\P(A)\sm \{\mty\}$):
 \[
@@
 \]
-This statement is more standard than $\AC$ as stated in the
-Isabelle/ZF development, which involves a silent use of Replacement
-since it is written as an axiom scheme.
+Another advantage of taking
+\isatt{AC{\isacharunderscore}func{\isacharunderscore}Pow} as a
+starting point is that it does not involve proper classes: The version
+of  $\AC$ in Isabelle/ZF corresponds to an axiom scheme of first-order
+logic and as such is not a standard formulation. 
+%% 26 commands to get Pointed $\DC$.
 
-The strategy is to define the function $f$ discussed above by
+The strategy to prove Pointed $\DC$ (following a proof in Moschovakis
+\cite{moschovakis1994notes}) is to define the function $f$ discussed above by
 primitive recursion on the naturals, which can be done easily thanks
@@
 
-The proof is mostly routine; after a few lemmas we obtain the
+The proof is mostly routine; after a few lemmas (26 proof
+commands in total) we obtain the
 following theorem:
@@
 Rasiowa-Sikorski. That is, if the assumption holds for a sequence of
-relations $S_n$,  then $f(n) \mathrel{S_{n+1}} f(n+1)$.
+relations $S_n$,  then $f(n) \mathrel{S_{n+1}} f(n+1)$ for all $n$.
 
 We first obtain a corollary of $\DC$ changing $A$ for
-$A\times\mathtt{nat}$:
+$A\times\mathtt{nat}$, whose procedural proof takes 16 lines:
 
 \begin{isabelle}
-\isacommand{lemma}\isamarkupfalse%
-\ aux{\isacharunderscore}DC{\isacharunderscore}on{\isacharunderscore}AxNat{\isadigit{2}}\ {\isacharcolon}\isanewline \ \ {\isachardoublequoteopen}{\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}{\isasymlangle}y{\isacharcomma}succ{\isacharparenleft}snd{\isacharparenleft}x{\isacharparenright}{\isacharparenright}{\isasymrangle}{\isasymrangle}\ {\isasymin}\ R\ {\isasymLongrightarrow}\isanewline
-    \ \ {\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ {\isacharbraceleft}{\isasymlangle}a{\isacharcomma}b{\isasymrangle}{\isasymin}R{\isachardot}\ snd{\isacharparenleft}b{\isacharparenright}\ {\isacharequal}\ succ{\isacharparenleft}snd{\isacharparenleft}a{\isacharparenright}{\isacharparenright}{\isacharbraceright}{\isachardoublequoteclose}
+\isacommand{corollary}\isamarkupfalse%
+\ DC{\isacharunderscore}on{\isacharunderscore}A{\isacharunderscore}x{\isacharunderscore}nat\ {\isacharcolon}\ \isanewline
+\ \ {\isachardoublequoteopen}{\isacharparenleft}{\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}{\isasymlangle}y{\isacharcomma}succ{\isacharparenleft}snd{\isacharparenleft}x{\isacharparenright}{\isacharparenright}{\isasymrangle}{\isasymrangle}\ {\isasymin}\ R{\isacharparenright}\ {\isasymLongrightarrow}\isanewline
+\ \ \ \ {\isasymforall}a{\isasymin}A{\isachardot}\ {\isacharparenleft}{\isasymexists}f\ {\isasymin}\ nat{\isasymrightarrow}A{\isachardot}\ f{\isacharbackquote}{\isadigit{0}}\ {\isacharequal}\ a\ {\isasymand}\isanewline
+\ \ \ \ \ \ \ \ \ \ \ \ {\isacharparenleft}{\isasymforall}n\ {\isasymin}\ nat{\isachardot}\ {\isasymlangle}{\isasymlangle}f{\isacharbackquote}n{\isacharcomma}n{\isasymrangle}{\isacharcomma}{\isasymlangle}f{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isacharcomma}succ{\isacharparenleft}n{\isacharparenright}{\isasymrangle}{\isasymrangle}{\isasymin}R{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
 \end{isabelle}
+
 %
@@
 \isacommand{lemma}\isamarkupfalse%
-\ aux{\isacharunderscore}sequence{\isacharunderscore}DC{\isadigit{2}}\ {\isacharcolon}\ {\isachardoublequoteopen}{\isasymforall}x{\isasymin}A{\isachardot}\ {\isasymforall}n{\isasymin}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ S{\isacharbackquote}n\ {\isasymLongrightarrow}\isanewline
-  {\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\isanewline \ \ {\isasymlangle}x{\isacharcomma}{\isasymlangle}y{\isacharcomma}succ{\isacharparenleft}snd{\isacharparenleft}x{\isacharparenright}{\isacharparenright}{\isasymrangle}{\isasymrangle}\ {\isasymin}\ {\isacharbraceleft}{\isasymlangle}{\isasymlangle}x{\isacharcomma}n{\isasymrangle}{\isacharcomma}{\isasymlangle}y{\isacharcomma}m{\isasymrangle}{\isasymrangle}{\isasymin}{\isacharparenleft}A{\isasymtimes}nat{\isacharparenright}{\isasymtimes}{\isacharparenleft}A{\isasymtimes}nat{\isacharparenright}{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}{\isasymin}S{\isacharbackquote}m\ {\isacharbraceright}{\isachardoublequoteclose}\isanewline
+\ aux{\isacharunderscore}sequence{\isacharunderscore}DC\ {\isacharcolon}\ {\isachardoublequoteopen}{\isasymforall}x{\isasymin}A{\isachardot}\ {\isasymforall}n{\isasymin}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ S{\isacharbackquote}n\ {\isasymLongrightarrow}\isanewline
+ \ \ {\isasymforall}x{\isasymin}A{\isasymtimes}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\isanewline
+ \ \ \ {\isasymlangle}x{\isacharcomma}{\isasymlangle}y{\isacharcomma}succ{\isacharparenleft}snd{\isacharparenleft}x{\isacharparenright}{\isacharparenright}{\isasymrangle}{\isasymrangle}\ {\isasymin}\ {\isacharbraceleft}{\isasymlangle}{\isasymlangle}w{\isacharcomma}n{\isasymrangle}{\isacharcomma}{\isasymlangle}y{\isacharcomma}m{\isasymrangle}{\isasymrangle}{\isasymin}{\isacharparenleft}A{\isasymtimes}nat{\isacharparenright}{\isasymtimes}{\isacharparenleft}A{\isasymtimes}nat{\isacharparenright}{\isachardot}\ {\isasymlangle}w{\isacharcomma}y{\isasymrangle}{\isasymin}S{\isacharbackquote}m\ {\isacharbraceright}{\isachardoublequoteclose}\isanewline
 \ \ %
@@
 \ sequence{\isacharunderscore}DC{\isacharcolon}\ {\isachardoublequoteopen}{\isasymforall}x{\isasymin}A{\isachardot}\ {\isasymforall}n{\isasymin}nat{\isachardot}\ {\isasymexists}y{\isasymin}A{\isachardot}\ {\isasymlangle}x{\isacharcomma}y{\isasymrangle}\ {\isasymin}\ S{\isacharbackquote}n\ {\isasymLongrightarrow}\isanewline
- {\isasymforall}a{\isasymin}A{\isachardot}\ {\isacharparenleft}{\isasymexists}f\ {\isasymin}\ nat{\isasymrightarrow}A{\isachardot}\ f{\isacharbackquote}{\isadigit{0}}\ {\isacharequal}\ a\ {\isasymand}\ {\isacharparenleft}{\isasymforall}n\ {\isasymin}\ nat{\isachardot}\ {\isasymlangle}f{\isacharbackquote}n{\isacharcomma}f{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isasymrangle}{\isasymin}S{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\ \ \ \ {\isasymforall}a{\isasymin}A{\isachardot}\ {\isacharparenleft}{\isasymexists}f\ {\isasymin}\ nat{\isasymrightarrow}A{\isachardot}\ f{\isacharbackquote}{\isadigit{0}}\ {\isacharequal}\ a\ {\isasymand}\isanewline
+\ \ \ \ \ \ \  {\isacharparenleft}{\isasymforall}n\ {\isasymin}\ nat{\isachardot}\ {\isasymlangle}f{\isacharbackquote}n{\isacharcomma}f{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isasymrangle}{\isasymin}S{\isacharbackquote}succ{\isacharparenleft}n{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+%
+%\isadelimproof
+\ \ %
+%\endisadelimproof
+%
+%\isatagproof
+\isacommand{apply}\isamarkupfalse%
+\ {\isacharparenleft}drule\ aux{\isacharunderscore}sequence{\isacharunderscore}DC{\isacharparenright}\isanewline
+\ \ \isacommand{apply}\isamarkupfalse%
+\ {\isacharparenleft}drule\ DC{\isacharunderscore}on{\isacharunderscore}A{\isacharunderscore}x{\isacharunderscore}nat{\isacharcomma}\ auto{\isacharparenright}\isanewline
+\ \ \isacommand{done}\isamarkupfalse%
+%
 \end{isabelle}

### 2019-11-01 LSFA/talk/talk.tex

\documentclass{beamer}
\usetheme{metropolis}           % Use metropolis theme
\usepackage[spanish]{babel}
\usepackage{amsmath,mathtools,array}
\usepackage{amssymb} 
\usepackage{amsthm} 
\usepackage{tikz}
\usepackage{microtype}
% \usepackage{../paper/paper}

\title{First steps towards a formalization of Forcing}
\author{Emmanuel Gunther\and \emph{Miguel Pagano}\and Pedro Sánchez Terraf}
\date{LSFA - Fortaleza, 26 September 2018}
\input{../header-draft}
\usepackage{../isabelle}
\usepackage{../isabellesym}
\begin{document}

\maketitle
%% Outline
%

\begin{frame}{Zermelo-Fraenkel Set Theory}
  \begin{block}{}
    Set theory arose at the early twentieth century as one
    possible foundation for mathematics.
  \end{block}
  \begin{block}{}
    More formally, set theory is a first order theory ($ZFC$) with no
    constants and one relational symbol $\in$. % It has some axioms
    % (extensionality, pairing, foundation, union, powerset, infinity)
    % and two axiom schemes (separation and replacement). This theory
    % is known by $ZF$; while $ZFC$ is $ZF$ plus the axiom of Choice.
  \end{block}
  \begin{block}{}
    To say that $ZFC$ is a possible foundation for mathematics means
    that any mathematical theorem, say $T$, can be deduced from $ZFC$:
    \[ ZFC \vdash T \]
  \end{block}
\end{frame}
\begin{frame}{Cantor's Question and Gödel's and Cohen's answers}
  \begin{block}{}
    Cantor posed the \emph{Continuum Hypothesis} ($\CH$):
    \begin{quote}
      Every uncountable subset of $\R$ is equipotent with $\R$.
    \end{quote}
  \end{block}
  \begin{block}{}
    Gödel showed that $\CH$ is \emph{relatively consistent} with $ZFC$.
  \end{block}
  \begin{block}{}
    Later, Cohen showed that also the negation of $\CH$ is relatively
    consistent with $ZFC$. For this he invented the technique of
    \emph{forcing}.
  \end{block}
\end{frame}

\begin{frame}{Some preliminaries}
  \begin{block}{}
    To say that $T$ is \emph{relatively consistent} with $ZFC$ can be
    understood in two ways:
    \begin{enumerate}
    \item Construct a model $M'$ of $ZFC+T$ from a model $M$ for
      $ZFC$.
    \item Alternatively, deduce a contradiction in $ZFC$ from a
      contradiction in $ZFC+T$.% (Interestingly both proofs give
      % rise to algorithms for this)
    \end{enumerate}
  \end{block}

  \begin{block}{}
    Cohen's proofs construct a model $ZFC+\neg\CH$ assuming a
    countable and transitive model (ctm).
  \end{block}
  \begin{itemize}
  \item  $M$ is a \emph{transitive} model of $ZFC$ if $x \in M$ implies
    $x\subset M$.
  \item $M$ is \emph{countable} if it is equipotent with $\N$.
  \end{itemize}

  % \begin{block}{}
  %   Notice that it is fine to ask for a ctm because at the end we
  %   want to transforms contradictions 
  % \end{block}
\end{frame}

\begin{frame}{Forcing}
  \begin{block}{}
    \begin{itemize}
    \item Given a ctm $M$ and a \emph{generic} set $G$ \emph{for} $M$,
      one constructs a new ctm $M[G]$.
    \item One defines a poset $(P,\leq)$, where $P \in M$. The
      existence of a generic filter $G$ follows from Rasiowa-Sikorski.
    \item $M[G] = \{\val(G,x)\, |\, x \in M \}$ 
    \item Depending on the definition of $P$, $M[G]$ will satisfy
      $\CH$ or $\neg\CH$.
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}{Our goal: formalize forcing}
  \begin{block}{Questions?}
    \begin{itemize}
    \item Why forcing? Has not been formalized yet. % (Paulson formalized Gödel's result)
    \item Related work. Paulson formalized Gödel's proof. %is the largest development and
%      we can build on top of his work.
    \item Which proof assistant? Because of the previous answer:
      Isabelle/ZF.
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}{Isabelle/ZF}
  \begin{block}{Axiomatization of ZF in Isabelle/FOL}
    \begin{itemize}
    \item Sets are terms of type \isatt{i} and formulas have type
      \isatt{o}. % None of them are recursive data-types.
    \item Axioms of $ZFC$ are postulated as axioms of Isabelle:
      \isa{extension: A\ =\ B\ \isasymlongleftrightarrow\ A\ \isasymsubseteq\ B\ \isasymand\ B \isasymsubseteq\ A}
    \item A theorem of $ZFC$ is a \emph{lemma} of Isabelle/ZF:
           \includegraphics[scale=1.3]{eqI.png}
     \item In order to speak of formulas and satisfaction
           (models), Paulson coded formulas (and much more) as sets.
         \item He also defined relativized (to a class) versions of the axioms\\
\hspace{-0.5cm}   \isa{upair\_ax(M) == \isasymforall x[M]. \isasymforall  y[M].\isasymexists z[M]. upair(M,x,y,z)}
     \end{itemize}
\end{block}
\end{frame}

\begin{frame}{First steps towards Forcings}
  \begin{block}{Our strategy}
    \begin{enumerate}
    \item Be as modular as possible, using \emph{locales}.
    \item Define interfaces to deal with models of $ZFC$.
    \item Exploit the modularity of forcing to divide the work.
    \end{enumerate}
  \end{block}
\end{frame}

\begin{frame}{What have we done?}
  \begin{block}{Existence of a generic filter}
  \begin{itemize}
  \item Proved a principle of dependent choice.
  \item Using that principle, proved Rasiowa-Sikorski.
  \item If $M$ is a ctm and $P\in M$ is a poset, then
    there exists a generic filter for $M$.
  \end{itemize}
\end{block}
\begin{block}{Definition of $M[G]$}
  \begin{itemize}
  \item Defined \emph{names} for $x \in M$ and \emph{$\val$}.
  \item Defined $M[G]$, assuming $M$ ctm and $G$ generic for $M$.
  \item Proved $M \subseteq M[G]$.% (using $\check{x}$).
\end{itemize}
\end{block}
\end{frame}

\begin{frame}{$M[G] \models ZFC$}
  \begin{block}{Axiom by axiom}
    \begin{itemize}
    \item One of the hardest part is transfering truth on $M$ to
      truth on $M[G]$.
    % \item The most obvious, say for pairing, would be
    %   \isa{sats(M,upair\_fm,[])} implies \isa{sats(M[G],upair\_fm,[])}.
    \item We proved \isa{upair\_ax(\#\#M)} implies \isa{upair\_ax(\#\#M[G])}
    \item $M[G]\models \mathit{separation}$ (almost).
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}{Future work: The next steps}
  \begin{block}{Soon}
    \begin{itemize}
    \item $x \in M$ implies $\check{x}\in M$.
    \item Define the poset we are interested in ($Add(\omega,\kappa)$).
    \item Prove that $M[G]$ satisfies union, foundation, infinity.
    \end{itemize}
  \end{block}
  \begin{block}{Not so soon}
    \begin{itemize}
    \item Define the forcing relation ($\forces$).
    \item Prove the fundamental theorem of forcing.
    \item $M[G]\models ZFC$.
    \item If $P$ is ccc, then cardinals are preserved in $M[G]$.
    \item Prove that for any generic $G$ for $Add(\omega,\aleph_2)$,
      then $M[G]\models ZFC+\neg\CH$.
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}{Questions?}
  \begin{center}
    \begin{block}{}
      \large{Thank you!}
    \end{block}
  \end{center}
\end{frame}

\end{document}

### 2019-12-10 IJCAR2020/forcing_in_isabelle_zf.tex
comienzo del paper que hará historia

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
% \renewcommand\UrlFont{\color{blue}\rmfamily}

\begin{document}
%
\title{Formalization of Forcing in Isabelle/ZF%
  \thanks{Supported by Secyt-UNC project 33620180100465CB and Conicet.}%
}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Emmanuel Gunther\inst{1} \and
Miguel Pagano\inst{1} \and \\
Pedro Sánchez Terraf\inst{1,2}\orcidID{0000-0003-3928-6942}}
%
\authorrunning{E. Gunther et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Universidad Nacional de C\'ordoba. 
  \\  Facultad de Matem\'atica, Astronom\'{\i}a,  F\'{\i}sica y
    Computaci\'on. \\
\email{\{gunther,pagano,sterraf\}@famaf.unc.edu.ar}
\and
    Centro de Investigaci\'on y Estudios de Matem\'atica (CIEM-FaMAF),
    Conicet. C\'ordoba. Argentina.
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
A better paper than that of Han \& van Doorn \cite{han_et_al:LIPIcs:2019:11074}
\keywords{Isabelle/ZF \and forcing \and names \and generic extension \and constructibility.}
\end{abstract}
%
%
%

\input{intro}

%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
\bibliographystyle{splncs04}
\bibliography{forcing_in_isabelle_zf}
\end{document}

### 2019-12-10 IJCAR2020/intro.tex

--- 2019-03-19 ITP/intro.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
 \section{Introduction}
-% no \IEEEPARstart
-Zermelo-Fraenkel Set Theory ($\ZF$) has a prominent place among formal
-theories. The reason for this is that it formalizes many intuitive
-properties of the notion of set. As such, it can be used as a
-foundation for mathematics and thus it has been
-thoroughly studied. 
-Considering the current trend 
-of mechanization of mathematics~\cite{avigad2018mechanization}, it
-seems natural to ask for a mechanization of the most salient results
-of Set Theory.
-
-The results we are interested in originally arose in connection to
-\emph{relative consistency} proofs in set theory; that is, showing
-that if $\ZF$ is 
-consistent, the addition of a new axiom $A$ won't make the system
-inconsistent; this is as much as we can expect to obtain, since
-G\"odel's Incompleteness Theorems precludes a  
-formal proof of the consistency of set theory in $\ZF$, unless the
-latter is indeed inconsistent. There are statements $A$ which are
-\emph{undecided} by $\ZF$, in the sense that both $A$ and $\neg A$ are
-consistent relative to $\ZF$; perhaps the most prominent  is the
-\emph{Continuum Hypothesis}, which led to the development of powerful
-techniques for independence proofs. First, G\"odel inaugurated the
-theory of \emph{inner models} by introducing his model $L$ of
-the \emph{Axiom of Constructibility} \cite{godel-L} and proved the
-relative consistency of the Axiom of Choice and the Generalized
-Continuum Hypothesis with $\ZF$. More than twenty years later, Paul
-J.~Cohen~\cite{Cohen-CH-PNAS} devised the technique of \emph{forcing},
-which is the only known way of \emph{extending} models of $\ZF$; this
-was used to prove the relative consistency of the 
-negation of the Continuum Hypothesis.
-
-In this work we address a substantial part of formalizing the proof
-that given a model $M$ of $\ZF$, any \emph{generic extension} $M[G]$
-obtained by forcing also satisfies $\ZF$. As remarked by Kunen
-\cite[p.250]{kunen2011set} \enquote{[...] in verifying that $M[G]$ is
-  a model for set theory, the hardest axiom to verify is
-  [Separation].}  The most important achievement of this paper is the
-mechanization in the proof assistant \emph{Isabelle} of a proof of the
-Axiom of Separation in generic extensions by using the ``fundamental''
-theorems of forcing.  En route to this, we also formalized the
-satisfaction by $M[G]$ of Extensionality, Foundation, and Union. As a
-consequence of Separation we were able to formalize the proof of the
-Powerset Axiom; finally, the Axiom of Infinity was proved under extra
-assumptions. % We have already
-% proved Pairing in the first report of our project
-% \cite{2018arXiv180705174G}.
-The theoretical support for this work has been 
-the fine textbook by Kunen \cite{kunen2011set} and our development
-benefited from the remarkable work done by Lawrence 
-Paulson \cite{paulson_2003} on the formalization of G\"odel's
-constructible universe in Isabelle.
-
-The
-ultimate goal of our project is the formalization of the forcing
-techniques needed to show the independence of the Continuum
-Hypothesis. 
-We think that this project constitutes an interesting test-case
-for the current technology of formalization of mathematics, in
-particular for the need of handling several layers of reasoning.
-
-The \emph{Formal Abstracts} project~\cite{hales-fabstracts} proposes
-the formalization of complex pieces of mathematics by writing the
-statements of 
-results and the material upon which they are based (definitions,
-propositions, lemmas), but omitting the proofs. In this work we
-partially adhere to this vision to delineate our formalization
-strategy:
-Since the proofs that the  axioms hold in generic extensions
-are independent of the \emph{proofs} of the fundamental theorems of
-forcing, we assumed the latter for the time being. Let us remark
-that those theorems depend on the definition of a function $\forceisa$
-from formulas to formulas which is, by itself, quite demanding; the
-formalization of it and of the fundamental theorems of forcing % roughly
-comprises barely less than a half of our full project.
-
-It might be a little surprising the lack of formalizations of forcing
-and generic extensions. As far as we know, the development of
-Quirin and Tabareau \cite{JFR6232} in homotopy type theory for constructing generic
-extensions in a sheaf-theoretic setting is the unique mechanization of
-forcing. This contrast with the fruitful use of forcing techniques to
-extend the Curry-Howard isomorphism to classical axioms
-\cite{Miquel:2011:FPT:2058525.2059614,lmcs:1070}. Moreover, the
-combination of forcing with intuitionistic type theory
-\cite{Coquand:2009:FTT:1807662.1807665,coquand2010note} gives rise
-both to positive results (an algorithm to obtain witnesses of the
-continuity of definable functionals \cite{coquand2012computational})
-and also negative (the independence of Markov's principle
-\cite{lmcs:3859}). In the same strand of forcing from the point of
-view of proof theory \cite{avigad_2004} are the conservative
-extensions of CoC with forcing conditions
-\cite{jaber:hal-01319066,Jaber:2012:ETT:2358958.2359524}.
-
-% \fbox{Parece mucho comienzo sólo para introducir a Kunen}
-% \fbox{¿lo puedo achurar un poco?}
-
-% In a gross simplification, there are two aspects to a formalization
-% project like this one: thematic and programmatic. The first concerns
-% the handling of all the theoretical concepts and results in the
-% subject, while the second involves the practical issues of the
-% implementation and design. In the case of forcing, the main intricacy
-% lies in the first aspect. In this sense, following a sensible
-% presentation of the material is key.  The authoritative reference 
-% on the subject during the last 30 years has been Kunen's classical
-% \cite{kunen1980}. In our
-% formalizaton we have followed a recent rewrite \cite{kunen2011set}
-% of that  textbook, which presents the material in the same sharp 
-% style but offering a lot of details. In some sense this project
-% wouldn't exist without this book. As alternative, introductory
-% resources, the  interested reader can check
-% \cite{chow-beginner-forcing}; also, the book \cite{weaver2014forcing}
-% contains a thorough treament minimizing the technicalities.
-
-In pursuing the proof of Separation on generic extensions we
-extended Paulson's library with:
-\begin{enumerate}
-\item renaming of variables for \emph{internalized} formulas, which
-  with little effort can be extended 
-  to substitutions;
-\item an improvement on definitions by recursion on well-founded
-  relations; 
-\item enhancements in the hierarchy of locales; and
-\item a variant of the  principle of dependent choices and a version
-  of Rasiowa-Sikorski, which 
-  ensures the existence of generic filters for countable and transitive
-  models of $\ZF$;
-\end{enumerate}
-the last item was already communicated in the
-  first report \cite{2018arXiv180705174G}.
-
-We briefly describe the contents of each
-section. Section~\ref{sec:isabelle} contains the bare minimum
-requirements to understand the (meta)logics used in Isabelle. Next, an
-overview of the model theory of set theory is presented in
-Section~\ref{sec:axioms-models-set-theory}. There is an ``internal''
-representation of first-order formulas as sets, implemented by
-Paulson; Section~\ref{sec:renaming} discusses syntactical
-transformations of the former, mainly permutation of variables. 
-In Section~\ref{sec:generic-extensions} the generic extensions are
-succinctly reviewed and how the treatment of well founded recursion in
-Isabelle was enhanced. We take care of the ``easy axioms'' in
-Section~\ref{sec:easy-axioms}; these are the ones that
-do not depend on the forcing theorems. We describe the latter in
-Section~\ref{sec:forcing}. We adapted the  work by Paulson to our
-needs, and this is described in
-Section~\ref{sec:hack-constructible}. We present the proof
-of the Separation Axiom Scheme in Section~\ref{sec:proof-separation},
-which follows closely its implementation, and some comments on the
-proof of the Powerset Axiom. A plan for future work and
-some immediate conclusions are offered in
-Section~\ref{sec:conclusions-future-work}.
 
@@
 %%% mode: latex
-%%% TeX-master: "Separation_In_MG"
+%%% TeX-master: "forcing_in_isabelle_zf"
 %%% ispell-local-dictionary: "american"

### 2019-12-14 IJCAR2020/absoluteness.tex
pedacitos del paper

\section{Relativization and absoluteness}

The concepts of relativization and absoluteness (due to Gödel, in his
proof of the relative consistency of $\AC$ \cite{godel-L}) 
are both
prerrequisites and powerful tools in working with transitive
models. The \emph{relativization} $\phi^C(\bx)$ of a set-theoretic
definition
$\phi$ (of relation such
as ``$x$ is a subset of $y$'' or of a function like $y=\P(x)$) to
a class $C$ is obtained by taking its (first-order) definition and
restricting all of its quantifiers to $C$.

\centerline{[example: $\sbq$]}

The new formula $\phi^C(\bx)$ corresponds to what is obtained by defining
the concept ``inside'' $C$. It turns out that many concepts mean the
same after relativization to a nonempty transitive class $C$; formally
\[
\forall\bx.\ C(\bx) \longrightarrow (\phi^C(\bx) \longleftrightarrow
\phi^C(\bx))
\]
When this is the case, we say that the relation defined by $\phi$ is
\emph{absolute for transitive models}. (The absoluteness of functions
moreover requires that
the relativized definition also behaves functionally over $C$.) As
examples, the relation of inclusion $\subseteq$ ---and actually, any
relation defined by a formula using only bounded quantifiers 
$(\forall x\in y)$ and $(\exists x\in y)$--- is absolute for
transitive models. On the contrary, this is not the case with the powerset
operation.

A benefit of the work with transitive models is that [one concept of
  ordinal accross the universe / $M$ / $M[G]$]

A part of this project is to refactorize Paulson's formalization
\cite{paulson_2003} of Gödel's \cite{godel-L}. The main objective is
to maximize applicability of the relativization machinery by adjusting
the hypothesis of a greater part of it early development. Paulson's
architecture had only in mind the consistency of $\ZFC$, but, for
instance, in order to apply it in the development of forcing, too much
is assumed at the beginning; more seriously, some assumptions can't be
regarded as ``first-order'' (v.g. the Replacement Scheme).

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "forcing_in_isabelle_zf"
%%% ispell-local-dictionary: "american"
%%% End:

### 2019-12-14 IJCAR2020/conclusion.tex
pedacitos del paper

--- 2019-12-14 IJCAR2020/absoluteness.tex
@@
-\section{Relativization and absoluteness}
-
-The concepts of relativization and absoluteness (due to Gödel, in his
-proof of the relative consistency of $\AC$ \cite{godel-L}) 
-are both
-prerrequisites and powerful tools in working with transitive
-models. The \emph{relativization} $\phi^C(\bx)$ of a set-theoretic
-definition
-$\phi$ (of relation such
-as ``$x$ is a subset of $y$'' or of a function like $y=\P(x)$) to
-a class $C$ is obtained by taking its (first-order) definition and
-restricting all of its quantifiers to $C$.
-
-\centerline{[example: $\sbq$]}
-
-The new formula $\phi^C(\bx)$ corresponds to what is obtained by defining
-the concept ``inside'' $C$. It turns out that many concepts mean the
-same after relativization to a nonempty transitive class $C$; formally
-\[
-\forall\bx.\ C(\bx) \longrightarrow (\phi^C(\bx) \longleftrightarrow
-\phi^C(\bx))
-\]
-When this is the case, we say that the relation defined by $\phi$ is
-\emph{absolute for transitive models}. (The absoluteness of functions
-moreover requires that
-the relativized definition also behaves functionally over $C$.) As
-examples, the relation of inclusion $\subseteq$ ---and actually, any
-relation defined by a formula using only bounded quantifiers 
-$(\forall x\in y)$ and $(\exists x\in y)$--- is absolute for
-transitive models. On the contrary, this is not the case with the powerset
-operation.
-
-A benefit of the work with transitive models is that [one concept of
-  ordinal accross the universe / $M$ / $M[G]$]
-
-A part of this project is to refactorize Paulson's formalization
-\cite{paulson_2003} of Gödel's \cite{godel-L}. The main objective is
-to maximize applicability of the relativization machinery by adjusting
-the hypothesis of a greater part of it early development. Paulson's
-architecture had only in mind the consistency of $\ZFC$, but, for
-instance, in order to apply it in the development of forcing, too much
-is assumed at the beginning; more seriously, some assumptions can't be
-regarded as ``first-order'' (v.g. the Replacement Scheme).
+\section{Conclusion}
 

### 2019-12-14 IJCAR2020/isabelle_metatheory.tex
pedacitos del paper

\section{Isabelle and (meta)theories}

\begin{enumerate}
\item Description of \emph{Pure} (meta-theory).
\item Isabelle/ZF + $\exists M$ ctm of $\ZF(C)$ (theory). 
\item $\formula$ (inner theory).
\item \textbf{Theorem}: $\exists N$ ctm of $\ZF(C)+\neg\CH$.
\end{enumerate}

in comparison to the ``proof theory approach'':

\begin{enumerate}
\item Primitive recursive arithmetic (meta-theory).
\item \textbf{Theorem}: Con($\ZF(C)$) $\implies$ Con($\ZF(C)+\neg\CH$).
\end{enumerate}

and then to the ``type-theoretic approach'':

\begin{enumerate}
\item CiC (meta-theory).
\item CiC (theory, $\geq \ZFC + \exists \kappa$ inaccessible).
\item \textbf{Theorem}: Con($\ZF(C)+\neg\CH$).
\end{enumerate}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "forcing_in_isabelle_zf"
%%% ispell-local-dictionary: "american"
%%% End:

### 2019-12-16 IJCAR2020/absoluteness.tex
pequeñas ediciones al paper

@@
 
-\centerline{[example: $\sbq$]}
+\[
+x \sbq^C y \equiv \forall z.\ C(z) \longrightarrow (z\in x
+\longrightarrow z\in y)
+\]
 
@@
 examples, the relation of inclusion $\subseteq$ ---and actually, any
-relation defined by a formula using only bounded quantifiers 
+relation defined by a formula (equivalent to one) using only bounded
+quantifiers 
 $(\forall x\in y)$ and $(\exists x\in y)$--- is absolute for
@@
 
+The version of \isatt{ZF-Constructible} we present weakens the
+assumptions of [many / a dozen / several] absoluteness theorems to that
+of a nonempty transitive class.
+
 %%% Local Variables: 

### 2019-12-16 IJCAR2020/forcing_in_isabelle_zf.tex
headers compilando
pedacitos del paper
pequeñas ediciones al paper

@@
 \usepackage[utf8]{inputenc}
+\usepackage{isabelle,isabellesym}
+\input{header-draft}
 \usepackage{graphicx}
@@
 %
+\hypersetup{
+  colorlinks,
+  urlcolor={blue},
+  linkcolor={blue!50!black},
+  citecolor={blue!50!black},
+}
 % If you use the hyperref package, please uncomment the following line
 % to display URLs in blue roman font according to Springer's eBook style:
-% \renewcommand\UrlFont{\color{blue}\rmfamily}
+\renewcommand\UrlFont{\color{blue}\rmfamily}
 
@@
 %
-\authorrunning{E. Gunther et al.}
+\authorrunning{E.~Gunther, M.~Pagano, P.~Sánchez Terraf}
+%\authorrunning{E. Gunther et al.}
 % First names are abbreviated in the running head.
@@
 \begin{abstract}
-A better paper than that of Han \& van Doorn \cite{han_et_al:LIPIcs:2019:11074}
+We formalize the development of forcing in the set theory framework of
+Isabelle/ZF. Under the assumption of the existence of countable
+transitive model of $\ZF$, we construct a generic extension and show
+that the latter also satisfies $\ZF$.
 \keywords{Isabelle/ZF \and forcing \and names \and generic extension \and constructibility.}
@@
 %
-
 \input{intro}
 
+\input{isabelle_metatheory}
+
+\input{absoluteness}
+
+\input{conclusion}
 %

### 2019-12-19 IJCAR2020/intro.tex
enumeración de (algunos) avances por sobre Separation

@@
 
+\subsection{Related work}
+
+Han \& van Doorn \cite{han_et_al:LIPIcs:2019:11074}.
+
+Updates over the (unpublished) technical report \cite{2019arXiv190103313G}.
+
+\begin{enumerate}
+\item The development is now independent of $\AC$: We modularized the
+  Rasiowa-Sikorski lemma in such a way that a version for countable
+  posets does not require choice, while we have the full result in an
+  independent file.
+  %
+\item We lifted the restriction on arities on the axiom schemes of
+  Separation and Replacement, streamlining various proofs as an
+  immediate benefit and avoiding the need for tupling. 
+  %
+\item We fulfilled the promised formal abstract for the Forcing
+  Theorems almost to the letter, with some slight modification
+  regarding arities of formulas.
+  %
+\item We implemented an automatic ML machinery of (bounded) renaming
+  of formulas.
+\item We finished the inteface between our development and
+  \isatt{ZF-Constructible} which enables us to do well-founded
+  recursion inside transitive models of an appropriate finite fragment
+  of $\ZF$. As a consequence, we have an unconditional proof of
+  the Axiom of Infinity and a long debt: the fact that the generic
+  filter $G$ belongs to the extension $M[G]$ and $M\sbq M[G]$.
+\end{enumerate}
 %%% Local Variables: 

### 2019-12-28 IJCAR2020/absoluteness.tex
enumeración de (algunos) avances por sobre Separation
ToDo list, typos, nuevos archivos

@@
 definition
-$\phi$ (of relation such
+$\phi$ (of a relation such
 as ``$x$ is a subset of $y$'' or of a function like $y=\P(x)$) to
@@
 
+[New absoluteness results on powersets (/ universes?)]
+
 %%% Local Variables: 

### 2019-12-28 IJCAR2020/conclusion.tex
ToDo list, typos, nuevos archivos

@@
 \section{Conclusion}
+
+TODO:
+\begin{enumerate}
+\item Implement \emph{Basic Set Theory (BST)} by Kunen in
+  Constructible: the use of alternatively Replacement or Powerset to
+  prove basic absoluteness and closure resuls.
+\item Enhance the automatization of formulas
+\item Develop the forcing notions to obtain the independence of $\CH$,
+  along with the prerrequisite combinatorial results (v.g.\ the
+  $\Delta$-system lemma).
+\end{enumerate}
 

### 2019-12-28 IJCAR2020/forces_definition.tex
nuevas secciones
minimos avances en escritura

--- 2019-12-14 IJCAR2020/isabelle_metatheory.tex
@@
-\section{Isabelle and (meta)theories}
+\section{The definition of $\forceisa$}
 
-\begin{enumerate}
-\item Description of \emph{Pure} (meta-theory).
-\item Isabelle/ZF + $\exists M$ ctm of $\ZF(C)$ (theory). 
-\item $\formula$ (inner theory).
-\item \textbf{Theorem}: $\exists N$ ctm of $\ZF(C)+\neg\CH$.
-\end{enumerate}
+The core of the development is showing the definability of the
+relation of forcing. As explained in
+\cite[Sect.~8]{2019arXiv190103313G}, this comprises the 
+definition of a function $\forceisa$ that maps the set of internal
+formulas into itself.
 
-in comparison to the ``proof theory approach'':
+The definition of $\forceisa$ proceeds by recursion
+over the set $\formula$ and its base case, that is, for
+atomic formulas, is (in)famously the most complicated one. Actually,
+newcomers can be puzzled by the fact that forcing for atomic
+formulas is also defined by (mutual) recursion: to know if $\tau_1\in\tau_2$ is
+forced, one must check if $\tau_1=\sigma$ is forced for $\sigma$
+moving in the transitive closure of $\tau_2$. To disentangle this, one
+must realize that this last recursion must be described sintactically:
+the definition of $\forceisa(\phi)$ for atomic $\phi$ is then an
+internal definition of the alleged recursion on names.
 
-\begin{enumerate}
-\item Primitive recursive arithmetic (meta-theory).
-\item \textbf{Theorem}: Con($\ZF(C)$) $\implies$ Con($\ZF(C)+\neg\CH$).
-\end{enumerate}
+Our definition, in the lines of Kunen~\cite[p.~257]{kunen2011set}, is the
+following, with two mutually recursive clauses:
+\begin{multline}\label{eq:def-forcing-equality}
+  \forceseq (p,t_1,t_2) \defi 
+  \forall s\in\dom(t_1)\cup\dom(t_2)\;\forall q\pleq p :\\
+  \forcesmem(q,s,t_1)\lsii 
+  \forcesmem(q,s,t_2),
+\end{multline}
+\begin{multline}\label{eq:def-forcing-membership}
+  \forcesmem(p,t_1,t_2) \defi  \forall v\pleq p \limp\\
+      \exists q\pleq v. \ \exists s.\ \exists r\in P .\ \lb s,r\rb \in
+      t_2 \land q \pleq r \land \forceseq(q,t_1,s)
+\end{multline}
+Note that the definition of $\forcesmem$ states that the set 
+\[
+\{q\pleq p : \exists \lb s,r\rb\in t_2 . \ q\pleq r \land \forceseq(q,t_1,s)\}
+\]
+is dense below $p$.
 
-and then to the ``type-theoretic approach'':
+It was not straightforward to use the recursion machinery of
+Isabelle/ZF to define $\forceseq$ and $\forcesmem$. For this, we
+defined a relation $\frecR$ on 4-tuples of elements of $M$, proved an
+induction principle for this relation,
 
-\begin{enumerate}
-\item CiC (meta-theory).
-\item CiC (theory, $\geq \ZFC + \exists \kappa$ inaccessible).
-\item \textbf{Theorem}: Con($\ZF(C)+\neg\CH$).
-\end{enumerate}
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ forces{\isacharunderscore}induction{\isacharunderscore}with{\isacharunderscore}conds{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\isanewline
+\ \ \ \ {\isachardoublequoteopen}{\isasymAnd}{\isasymtau}\ {\isasymtheta}\ p{\isachardot}\ p\ {\isasymin}\ P\ {\isasymLongrightarrow}\ {\isasymlbrakk}{\isasymAnd}q\ {\isasymsigma}{\isachardot}\ {\isasymlbrakk}q{\isasymin}P\ {\isacharsemicolon}\ {\isasymsigma}{\isasymin}domain{\isacharparenleft}{\isasymtheta}{\isacharparenright}{\isasymrbrakk}\ {\isasymLongrightarrow}\ Q{\isacharparenleft}q{\isacharcomma}{\isasymtau}{\isacharcomma}{\isasymsigma}{\isacharparenright}{\isasymrbrakk}\ {\isasymLongrightarrow}\ R{\isacharparenleft}p{\isacharcomma}{\isasymtau}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \ \ {\isachardoublequoteopen}{\isasymAnd}{\isasymtau}\ {\isasymtheta}\ p{\isachardot}\ p\ {\isasymin}\ P\ {\isasymLongrightarrow}\ {\isasymlbrakk}{\isasymAnd}q\ {\isasymsigma}{\isachardot}\ {\isasymlbrakk}q{\isasymin}P\ {\isacharsemicolon}\ {\isasymsigma}{\isasymin}domain{\isacharparenleft}{\isasymtau}{\isacharparenright}\ {\isasymunion}\ domain{\isacharparenleft}{\isasymtheta}{\isacharparenright}{\isasymrbrakk}\ {\isasymLongrightarrow}\ R{\isacharparenleft}q{\isacharcomma}{\isasymsigma}{\isacharcomma}{\isasymtau}{\isacharparenright}\ {\isasymand}\ R{\isacharparenleft}q{\isacharcomma}{\isasymsigma}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isasymrbrakk}\ {\isasymLongrightarrow}\ Q{\isacharparenleft}p{\isacharcomma}{\isasymtau}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \ \ {\isachardoublequoteopen}p\ {\isasymin}\ P{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\isanewline
+\ \ \ \ {\isachardoublequoteopen}Q{\isacharparenleft}p{\isacharcomma}{\isasymtau}{\isacharcomma}{\isasymtheta}{\isacharparenright}\ {\isasymand}\ R{\isacharparenleft}p{\isacharcomma}{\isasymtau}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
+%
+and 
+obtained both functions as cases of a another one, 
+$\forcesat$, using a single recursion on $\frecR$. Then we obtained 
+(\ref{eq:def-forcing-equality}) and (\ref{eq:def-forcing-membership})
+as our corollaries \isatt{def\_forces\_eq} and
+\isatt{def\_forces\_mem}.
+
+Other approaches, like the one in Neeman~\cite{neeman-course}, prefer
+to have a single, more complicated, clause for $t_1=t_2$ using
+well-founded recursion on $t_1$
+and $t_2$ and then define the case $t_1\in t_2$ explicitly.
 

### 2019-12-29 IJCAR2020/forces_definition.tex
avances en la definición de forces en paper, nueva sección

--- 2019-12-14 IJCAR2020/isabelle_metatheory.tex
@@
-\section{Isabelle and (meta)theories}
+\section{The definition of $\forceisa$}
 
-\begin{enumerate}
-\item Description of \emph{Pure} (meta-theory).
-\item Isabelle/ZF + $\exists M$ ctm of $\ZF(C)$ (theory). 
-\item $\formula$ (inner theory).
-\item \textbf{Theorem}: $\exists N$ ctm of $\ZF(C)+\neg\CH$.
-\end{enumerate}
+The core of the development is showing the definability of the
+relation of forcing. As explained in
+\cite[Sect.~8]{2019arXiv190103313G}, this comprises the 
+definition of a function $\forceisa$ that maps the set of internal
+formulas into itself.
 
-in comparison to the ``proof theory approach'':
+The definition of $\forceisa$ proceeds by recursion
+over the set $\formula$ and its base case, that is, for
+atomic formulas, is (in)famously the most complicated one. Actually,
+newcomers can be puzzled by the fact that forcing for atomic
+formulas is also defined by (mutual) recursion: to know if $\tau_1\in\tau_2$ is
+forced, one must check if $\tau_1=\sigma$ is forced for $\sigma$
+moving in the transitive closure of $\tau_2$. To disentangle this, one
+must realize that this last recursion must be described sintactically:
+the definition of $\forceisa(\phi)$ for atomic $\phi$ is then an
+internal definition of the alleged recursion on names.
 
-\begin{enumerate}
-\item Primitive recursive arithmetic (meta-theory).
-\item \textbf{Theorem}: Con($\ZF(C)$) $\implies$ Con($\ZF(C)+\neg\CH$).
-\end{enumerate}
+Our aim was to follow Kunen~\cite[p.~257]{kunen2011set}, where the
+following mutual recursion is given:
+\begin{multline}\label{eq:def-forcing-equality}
+  \forceseq (p,t_1,t_2) \defi 
+  \forall s\in\dom(t_1)\cup\dom(t_2).\ \forall q\pleq p .\\
+  \forcesmem(q,s,t_1)\lsii 
+  \forcesmem(q,s,t_2),
+\end{multline}
+\begin{multline}\label{eq:def-forcing-membership}
+  \forcesmem(p,t_1,t_2) \defi  \forall v\pleq p. \ \exists q\pleq v. \\
+  \exists s.\ \exists r\in \PP .\ \lb s,r\rb \in
+      t_2 \land q \pleq r \land \forceseq(q,t_1,s)
+\end{multline}
+Note that the definition of $\forcesmem$ states that the set 
+\[
+\{q\pleq p : \exists \lb s,r\rb\in t_2 . \ q\pleq r \land \forceseq(q,t_1,s)\}
+\]
+is dense below $p$.
 
-and then to the ``type-theoretic approach'':
+It was not straightforward to use the recursion machinery of
+Isabelle/ZF to define $\forceseq$ and $\forcesmem$. For this, we
+defined a relation $\frecR$ on 4-tuples of elements of $M$, proved
+that it is wellfounded ---indeed, we also proved an
+induction principle for this relation:
+%
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ forces{\isacharunderscore}induction{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\isanewline
+\ \ \ \ {\isachardoublequoteopen}{\isasymAnd}{\isasymtau}\ {\isasymtheta}{\isachardot}\ {\isasymlbrakk}{\isasymAnd}{\isasymsigma}{\isachardot}\ {\isasymsigma}{\isasymin}domain{\isacharparenleft}{\isasymtheta}{\isacharparenright}\ {\isasymLongrightarrow}\ Q{\isacharparenleft}{\isasymtau}{\isacharcomma}{\isasymsigma}{\isacharparenright}{\isasymrbrakk}\ {\isasymLongrightarrow}\ R{\isacharparenleft}{\isasymtau}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \ \ {\isachardoublequoteopen}{\isasymAnd}{\isasymtau}\ {\isasymtheta}{\isachardot}\ {\isasymlbrakk}{\isasymAnd}{\isasymsigma}{\isachardot}\ {\isasymsigma}{\isasymin}domain{\isacharparenleft}{\isasymtau}{\isacharparenright}\ {\isasymunion}\ domain{\isacharparenleft}{\isasymtheta}{\isacharparenright}\ {\isasymLongrightarrow}\ R{\isacharparenleft}{\isasymsigma}{\isacharcomma}{\isasymtau}{\isacharparenright}\ {\isasymand}\ R{\isacharparenleft}{\isasymsigma}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isasymrbrakk}\isanewline
+\ \ \ \ \ \  {\isasymLongrightarrow}\ Q{\isacharparenleft}{\isasymtau}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\isanewline
+\ \ \ \ {\isachardoublequoteopen}Q{\isacharparenleft}{\isasymtau}{\isacharcomma}{\isasymtheta}{\isacharparenright}\ {\isasymand}\ R{\isacharparenleft}{\isasymtau}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
+%
+and 
+obtained both functions as cases of a another one, 
+$\forcesat$, using a single recursion on $\frecR$. Then we obtained 
+(\ref{eq:def-forcing-equality}) and (\ref{eq:def-forcing-membership})
+as our corollaries \isatt{def\_forces\_eq} and
+\isatt{def\_forces\_mem}.
 
-\begin{enumerate}
-\item CiC (meta-theory).
-\item CiC (theory, $\geq \ZFC + \exists \kappa$ inaccessible).
-\item \textbf{Theorem}: Con($\ZF(C)+\neg\CH$).
-\end{enumerate}
+Other approaches, like the one in Neeman~\cite{neeman-course} (and
+Kunen's older \cite{kunen1980}), prefer
+to have a single, more complicated, definition by simple recursion for
+$\forceseq$ and then define $\forcesmem$ explicitly. On hindsight,
+this might have been a little simpler to do, but we preferred to be as
+faithful as possible concerning this point to the text.
+
+Once $\forcesat$ and its relativized version $\isatt{is\_forces\_at}$
+were defined, we proceeded to show absoluteness and provided internal
+definitions for the recursion on names using results in
+\isatt{ZF-Constructible}. This finished the atomic case of the
+formula-transformer $\forceisa$.
+
+\begin{framed}
+  In order to prove absoluteness of functions defined by wellfounded
+  recursion, we need to show that we can replace the relevant
+  wellfounded relations by their transitive closures. (EXPAND HERE OR
+  SOMEWHERE ELSE).
+\end{framed}
 

### 2019-12-30 src/fix_links.sh
generacion del documento

@@ -1,6 +1,6 @@
 #!/bin/bash
 old_dir=$PWD
-html_dir="output/html/Unsorted/Separation/"
+html_dir="output/html/Unsorted/Forcing/"
 
 cd $html_dir
 sed -i -- 's/a href="..\/..\/ZF\/ZF\/Main.html/a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF\/ZF.html/g' *.html

### 2019-12-30 IJCAR2020/forcing_theorems.tex
avances en la definición de forces en paper, nueva sección
los dos tipos de inducción en Forcing_Thms

--- 2019-12-14 IJCAR2020/isabelle_metatheory.tex
@@
-\section{Isabelle and (meta)theories}
+\section{The forcing theorems}
 
-\begin{enumerate}
-\item Description of \emph{Pure} (meta-theory).
-\item Isabelle/ZF + $\exists M$ ctm of $\ZF(C)$ (theory). 
-\item $\formula$ (inner theory).
-\item \textbf{Theorem}: $\exists N$ ctm of $\ZF(C)+\neg\CH$.
-\end{enumerate}
+After the definition of $\forceisa$ is complete, the proof of the
+Fundamental Theorems of forcing is comparatively straightforward, and
+we were able to follow Kunen very closely. The more involved points of
+this part of the development were those where we needed to proved that
+various (dense) subsets of $\PP$ were in $M$; for this, we had to
+recourse to several absoluteness ad-hoc lemmas.
 
-in comparison to the ``proof theory approach'':
+As with the previous section, the proofs here have two different
+flavours: The ones for the atomic formulas proceed by using the
+\isatt{forces\_induction} principle, and then an induction on
+$\formula$ wraps the former with the remaining (inductive) cases.
 
-\begin{enumerate}
-\item Primitive recursive arithmetic (meta-theory).
-\item \textbf{Theorem}: Con($\ZF(C)$) $\implies$ Con($\ZF(C)+\neg\CH$).
-\end{enumerate}
+As an example of the first class, we can take a look at our
+formalization of \cite[Lem.~IV.2.40(a)]{kunen2011set}:
 
-and then to the ``type-theoretic approach'':
+\begin{isabelle}
+  \isacommand{lemma}\isamarkupfalse%
+  \ IV{\isadigit{2}}{\isadigit{4}}{\isadigit{0}}a{\isacharcolon}\isanewline
+  \ \ \isakeyword{assumes}\isanewline
+  \ \ \ \ {\isachardoublequoteopen}M{\isacharunderscore}generic{\isacharparenleft}G{\isacharparenright}{\isachardoublequoteclose}\isanewline
+  \ \ \isakeyword{shows}\ \isanewline
+  \ \ \ \ {\isachardoublequoteopen}{\isacharparenleft}{\isasymtau}{\isasymin}M{\isasymlongrightarrow}{\isasymtheta}{\isasymin}M{\isasymlongrightarrow}{\isacharparenleft}{\isasymforall}p{\isasymin}G{\isachardot}forces{\isacharunderscore}eq{\isacharparenleft}p{\isacharcomma}{\isasymtau}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isasymlongrightarrow}val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}{\isacharequal}val{\isacharparenleft}G{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isasymand}\isanewline
+  \ \ \ \ \ {\isacharparenleft}{\isasymtau}{\isasymin}M{\isasymlongrightarrow}{\isasymtheta}{\isasymin}M{\isasymlongrightarrow}{\isacharparenleft}{\isasymforall}p{\isasymin}G{\isachardot}forces{\isacharunderscore}mem{\isacharparenleft}p{\isacharcomma}{\isasymtau}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isasymlongrightarrow}val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}{\isasymin}val{\isacharparenleft}G{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
+%
+Its proof starts by an introduction of \isatt{forces\_induction}, and
+the two inductive cases for each atomic type were handled before as
+separate lemmas (\isatt{IV240a\_mem} and \isatt{IV240a\_eq}). We
+illustrate with the statement of the latter.
+%
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ IV{\isadigit{2}}{\isadigit{4}}{\isadigit{0}}a{\isacharunderscore}eq{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\isanewline
+\ \ \ \ {\isachardoublequoteopen}M{\isacharunderscore}generic{\isacharparenleft}G{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}p{\isasymin}G{\isachardoublequoteclose}\ {\isachardoublequoteopen}forces{\isacharunderscore}eq{\isacharparenleft}p{\isacharcomma}{\isasymtau}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \ \ \isakeyword{and}\isanewline
+\ \ \ \ IH{\isacharcolon}{\isachardoublequoteopen}{\isasymAnd}q\ {\isasymsigma}{\isachardot}\ q{\isasymin}P\ {\isasymLongrightarrow}\ q{\isasymin}G\ {\isasymLongrightarrow}\ {\isasymsigma}{\isasymin}domain{\isacharparenleft}{\isasymtau}{\isacharparenright}\ {\isasymunion}\ domain{\isacharparenleft}{\isasymtheta}{\isacharparenright}\ {\isasymLongrightarrow}\ \isanewline
+\ \ \ \ \ \ \ \ {\isacharparenleft}forces{\isacharunderscore}mem{\isacharparenleft}q{\isacharcomma}{\isasymsigma}{\isacharcomma}{\isasymtau}{\isacharparenright}\ {\isasymlongrightarrow}\ val{\isacharparenleft}G{\isacharcomma}{\isasymsigma}{\isacharparenright}\ {\isasymin}\ val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}{\isacharparenright}\ {\isasymand}\isanewline
+\ \ \ \ \ \ \ \ {\isacharparenleft}forces{\isacharunderscore}mem{\isacharparenleft}q{\isacharcomma}{\isasymsigma}{\isacharcomma}{\isasymtheta}{\isacharparenright}\ {\isasymlongrightarrow}\ val{\isacharparenleft}G{\isacharcomma}{\isasymsigma}{\isacharparenright}\ {\isasymin}\ val{\isacharparenleft}G{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\isanewline
+\ \ \ \ {\isachardoublequoteopen}val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}\ {\isacharequal}\ val{\isacharparenleft}G{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
 
-\begin{enumerate}
-\item CiC (meta-theory).
-\item CiC (theory, $\geq \ZFC + \exists \kappa$ inaccessible).
-\item \textbf{Theorem}: Con($\ZF(C)+\neg\CH$).
-\end{enumerate}
+As an example of the second kind of induction (on formulas), we choose the
+following relatively simple result:
+
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ strengthening{\isacharunderscore}lemma{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\ \isanewline
+\ \ \ \ {\isachardoublequoteopen}p{\isasymin}P{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}r{\isasymin}P{\isachardoublequoteclose}\ {\isachardoublequoteopen}r{\isasympreceq}p{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\isanewline
+\ \ \ \ {\isachardoublequoteopen}{\isasymAnd}env{\isachardot}\ env{\isasymin}list{\isacharparenleft}M{\isacharparenright}\ {\isasymLongrightarrow}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isasymle}length{\isacharparenleft}env{\isacharparenright}\ {\isasymLongrightarrow}\ p\ {\isasymtturnstile}\ {\isasymphi}\ env\isanewline
+\ \ \ \ \ {\isasymLongrightarrow}\ r\ {\isasymtturnstile}\ {\isasymphi}\ env{\isachardoublequoteclose}\isanewline
+%
+%
+\isacommand{using}\isamarkupfalse%
+\ assms{\isacharparenleft}{\isadigit{2}}{\isacharparenright}
+\end{isabelle}
+%
+The proof is divided in the 4 cases of definition of an element of $\formula$,
+%
+\begin{isabelle}
+\isacommand{proof}\isamarkupfalse%
+\ {\isacharparenleft}induct{\isacharparenright}\isanewline
+\ \ \isacommand{case}\isamarkupfalse%
+\ {\isacharparenleft}Member\ n\ m{\isacharparenright}\isanewline
+\ \ \isacommand{then}\isamarkupfalse%
+\isanewline
+\ \ \dots
+\isanewline
+\ \ \isacommand{show}\isamarkupfalse%
+\ {\isacharquery}case\ \isanewline
+\ \ \ \ \isacommand{using}\isamarkupfalse%
+\ Forces{\isacharunderscore}Member{\isacharbrackleft}of\ {\isacharunderscore}\ {\isachardoublequoteopen}nth{\isacharparenleft}n{\isacharcomma}env{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}nth{\isacharparenleft}m{\isacharcomma}env{\isacharparenright}{\isachardoublequoteclose}\ env\ n\ m{\isacharbrackright}\isanewline
+\ \ \ \ \ \ strengthening{\isacharunderscore}mem{\isacharbrackleft}of\ p\ r\ {\isachardoublequoteopen}nth{\isacharparenleft}n{\isacharcomma}env{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}nth{\isacharparenleft}m{\isacharcomma}env{\isacharparenright}{\isachardoublequoteclose}{\isacharbrackright}\ \isacommand{by}\isamarkupfalse%
+\ simp
+\end{isabelle}
+%
+where the final step depends on previously proved
+\isatt{strengthening{\isacharunderscore}mem} and the characterization of
+$\forceisa$ for membership (\isatt{Forces{\isacharunderscore}Member}):
+\isatt{{\isacharparenleft}p\ {\isasymtturnstile}\ Member{\isacharparenleft}n{\isacharcomma}m{\isacharparenright}\ env{\isacharparenright}\ {\isasymlongleftrightarrow}\ forces{\isacharunderscore}mem{\isacharparenleft}p{\isacharcomma}t{\isadigit{1}}{\isacharcomma}t{\isadigit{2}}{\isacharparenright}},
+where \isatt{t{\isadigit{1}}} and \isatt{t{\isadigit{1}}} are the
+\isatt{n}th resp.\ \isatt{m}th elements of \isatt{env}.
+
+The case of equality is entirely analogous, and the \isatt{Nand} and
+\isatt{Forall} cases are handled very simply.
+%
+\begin{isabelle}
+\isacommand{next}\isamarkupfalse%
+\isanewline
+\ \ \isacommand{case}\isamarkupfalse%
+\ {\isacharparenleft}Equal\ n\ m{\isacharparenright}\isanewline
+\ \ \dots\isanewline
+\isacommand{next}\isamarkupfalse%
+\isanewline
+\ \ \isacommand{case}\isamarkupfalse%
+\ {\isacharparenleft}Nand\ {\isasymphi}\ {\isasympsi}{\isacharparenright}\isanewline
+\ \ \isacommand{with}\isamarkupfalse%
+\ assms\isanewline
+\ \ \isacommand{show}\isamarkupfalse%
+\ {\isacharquery}case\ \isanewline
+\ \ \ \ \isacommand{using}\isamarkupfalse%
+\ Forces{\isacharunderscore}Nand\ Transset{\isacharunderscore}intf{\isacharbrackleft}OF\ trans{\isacharunderscore}M\ {\isacharunderscore}\ P{\isacharunderscore}in{\isacharunderscore}M{\isacharbrackright}\ pair{\isacharunderscore}in{\isacharunderscore}M{\isacharunderscore}iff\isanewline
+\ \ \ \ \ \ Transset{\isacharunderscore}intf{\isacharbrackleft}OF\ trans{\isacharunderscore}M\ {\isacharunderscore}\ leq{\isacharunderscore}in{\isacharunderscore}M{\isacharbrackright}\ leq{\isacharunderscore}transD\ \isacommand{by}\isamarkupfalse%
+\ auto\isanewline
+\isacommand{next}\isamarkupfalse%
+\isanewline
+\ \ \isacommand{case}\isamarkupfalse%
+\ {\isacharparenleft}Forall\ {\isasymphi}{\isacharparenright}\isanewline
+\ \ \isacommand{with}\isamarkupfalse%
+\ assms\isanewline
+\ \ \isacommand{have}\isamarkupfalse%
+\ {\isachardoublequoteopen}p\ {\isasymtturnstile}\ {\isasymphi}\ {\isacharparenleft}{\isacharbrackleft}x{\isacharbrackright}\ {\isacharat}\ env{\isacharparenright}{\isachardoublequoteclose}\ \isakeyword{if}\ {\isachardoublequoteopen}x{\isasymin}M{\isachardoublequoteclose}\ \isakeyword{for}\ x\isanewline
+\ \ \ \ \isacommand{using}\isamarkupfalse%
+\ that\ Forces{\isacharunderscore}Forall\ \isacommand{by}\isamarkupfalse%
+\ simp\isanewline
+\ \ \isacommand{with}\isamarkupfalse%
+\ \underline{Forall}\ \isanewline
+\ \ \isacommand{have}\isamarkupfalse%
+\ {\isachardoublequoteopen}r\ {\isasymtturnstile}\ {\isasymphi}\ {\isacharparenleft}{\isacharbrackleft}x{\isacharbrackright}\ {\isacharat}\ env{\isacharparenright}{\isachardoublequoteclose}\ \isakeyword{if}\ {\isachardoublequoteopen}x{\isasymin}M{\isachardoublequoteclose}\ \isakeyword{for}\ x\isanewline
+\ \ \ \ \isacommand{using}\isamarkupfalse%
+\ that\ pred{\isacharunderscore}le{\isadigit{2}}\ \isacommand{by}\isamarkupfalse%
+\ {\isacharparenleft}simp{\isacharparenright}\isanewline
+\ \ \isacommand{with}\isamarkupfalse%
+\ assms\ \underline{Forall}\isanewline
+\ \ \isacommand{show}\isamarkupfalse%
+\ {\isacharquery}case\ \isanewline
+\ \ \ \ \isacommand{using}\isamarkupfalse%
+\ Forces{\isacharunderscore}Forall\ \isacommand{by}\isamarkupfalse%
+\ simp\isanewline
+\isacommand{qed}\isamarkupfalse
+\end{isabelle}
+%
+In these two proofs, analogous results to
+\isatt{Forces{\isacharunderscore}Member} for \isatt{Nand} and
+\isatt{Forall} are used. It can be noted that the inductive hypothesis
+gets used in the last case (underlined here as
+\isatt{\underline{Forall}}), but not in the case for \isatt{Nand}.
 

### 2019-12-31 IJCAR2020/forcing_theorems.tex
avances en escritura

@@
 After the definition of $\forceisa$ is complete, the proof of the
-Fundamental Theorems of forcing is comparatively straightforward, and
+Fundamental Theorems of Forcing is comparatively straightforward, and
 we were able to follow Kunen very closely. The more involved points of
@@
 
-As with the previous section, the proofs here have two different
+The first results concern characterizations of the forcing
+relation. Two of them are \isatt{Forces{\isacharunderscore}Member}:
+\begin{center}
+  \isatt{{\isacharparenleft}p\ {\isasymtturnstile}\ Member{\isacharparenleft}n{\isacharcomma}m{\isacharparenright}\ env{\isacharparenright}\ {\isasymlongleftrightarrow}\ forces{\isacharunderscore}mem{\isacharparenleft}p{\isacharcomma}t{\isadigit{1}}{\isacharcomma}t{\isadigit{2}}{\isacharparenright}},
+\end{center}
+where \isatt{t{\isadigit{1}}} and \isatt{t{\isadigit{1}}} are the
+\isatt{n}th resp.\ \isatt{m}th elements of \isatt{env}, and  \isatt{Forces{\isacharunderscore}Forall}:
+\begin{center}
+  \isatt{{\isacharparenleft}p\ {\isasymtturnstile}\ Forall{\isacharparenleft}{\isasymphi}{\isacharparenright}\ env{\isacharparenright}\ {\isasymlongleftrightarrow}\ {\isacharparenleft}{\isasymforall}x{\isasymin}M{\isachardot}\ {\isacharparenleft}p\ {\isasymtturnstile}\ {\isasymphi}\ {\isacharparenleft}{\isacharbrackleft}x{\isacharbrackright}\ {\isacharat}\ env{\isacharparenright}{\isacharparenright}{\isacharparenright}}.
+\end{center}
+These two, along with  \isatt{Forces{\isacharunderscore}Equal} and
+\isatt{Forces{\isacharunderscore}Nand}, appear in Kunen as the
+inductive definition of the forcing relation \cite[Def.~IV.2.42]{kunen2011set}.
+
+As with the previous section, the proofs of the forcing theorems have two different
 flavours: The ones for the atomic formulas proceed by using the
-\isatt{forces\_induction} principle, and then an induction on
-$\formula$ wraps the former with the remaining (inductive) cases.
+principle of 
+\isatt{forces\_induction}, and then an induction on
+$\formula$ wraps the former with the remaining cases (\isatt{Nand} and \isatt{Forall}).
 
@@
 %
-Its proof starts by an introduction of \isatt{forces\_induction}, and
-the two inductive cases for each atomic type were handled before as
+Its proof starts by an introduction of \isatt{forces\_induction};
+the  inductive cases for each atomic type were handled before as
 separate lemmas (\isatt{IV240a\_mem} and \isatt{IV240a\_eq}). We
@@
 \isatt{strengthening{\isacharunderscore}mem} and the characterization of
-$\forceisa$ for membership (\isatt{Forces{\isacharunderscore}Member}):
-\isatt{{\isacharparenleft}p\ {\isasymtturnstile}\ Member{\isacharparenleft}n{\isacharcomma}m{\isacharparenright}\ env{\isacharparenright}\ {\isasymlongleftrightarrow}\ forces{\isacharunderscore}mem{\isacharparenleft}p{\isacharcomma}t{\isadigit{1}}{\isacharcomma}t{\isadigit{2}}{\isacharparenright}},
-where \isatt{t{\isadigit{1}}} and \isatt{t{\isadigit{1}}} are the
-\isatt{n}th resp.\ \isatt{m}th elements of \isatt{env}.
+$\forceisa$ for membership
 
@@
 %
-In these two proofs, analogous results to
-\isatt{Forces{\isacharunderscore}Member} for \isatt{Nand} and
-\isatt{Forall} are used. It can be noted that the inductive hypothesis
+It can be noted that the inductive hypothesis
 gets used in the last case (underlined here as

### 2019-12-31 IJCAR2020/proper_extension.tex
avances en escritura

\section{Example of proper extension}

Even when the axioms of $\ZFC$ are proved in the generic extension,
one can't claim that the magic of forcing has taken place unless one
is able to provide some \emph{proper} extension with the \emph{same
ordinals}. After all, one is assuming from starters a model $M$ of $\ZFC$,
and in some trivial cases $M[G]$ might end up to be exactly $M$; this
is where \emph{proper} enters the stage. But, for instance, in the
presence of large cardinals, a model $M'\supsetneq M$ might be an
end-extension of $M$ ---this is were we ask the two models to have the
same ordinals, the same \emph{height}.

Two short theory files contains the relevant
results. \texttt{Ordinals\_In\_MG.thy} shows, using the closure of $M$
under ranks, that $M$ and $M[G]$ share the same ordinals (actually,
ranks of elements of $M[G]$ are bounded by the ranks of their names in
$M$):
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ rank{\isacharunderscore}val{\isacharcolon}\ {\isachardoublequoteopen}rank{\isacharparenleft}val{\isacharparenleft}G{\isacharcomma}x{\isacharparenright}{\isacharparenright}\ {\isasymle}\ rank{\isacharparenleft}x{\isacharparenright}{\isachardoublequoteclose}\isanewline
\isacommand{lemma}\isamarkupfalse%
\ Ord{\isacharunderscore}MG{\isacharunderscore}iff{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ {\isachardoublequoteopen}Ord{\isacharparenleft}{\isasymalpha}{\isacharparenright}{\isachardoublequoteclose}\ \isanewline
\ \ \isakeyword{shows}\ {\isachardoublequoteopen}{\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}
\end{isabelle}

\texttt{Non\_Constructible.thy} contains the first example of a poset
that interprets the locale
\isatt{forcing{\isacharunderscore}notion}. It is the set of all binary
lists with the partial order given by  ``\isatt{xs} is an extension of
\isatt{ys}.''

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "forcing_in_isabelle_zf"
%%% ispell-local-dictionary: "american"
%%% End:

### 2019-12-31 IJCAR2020/replacement.tex
nuevas secciones
avances en escritura

--- 2019-12-14 IJCAR2020/isabelle_metatheory.tex
@@
-\section{Isabelle and (meta)theories}
+\section{The axiom of replacement}
 
-\begin{enumerate}
-\item Description of \emph{Pure} (meta-theory).
-\item Isabelle/ZF + $\exists M$ ctm of $\ZF(C)$ (theory). 
-\item $\formula$ (inner theory).
-\item \textbf{Theorem}: $\exists N$ ctm of $\ZF(C)+\neg\CH$.
-\end{enumerate}
+The proof of the Replacement Axiom scheme in $M[G]$ in Kunen uses the
+Reflection Principle relativized to $M$. We took an alternative
+pathway, following Neeman \cite{neeman-course}. In his course notes,
+he uses the relativization of the cumulative hierarchy of sets.
 
-in comparison to the ``proof theory approach'':
+The
+family of all sets of rank less than $\alpha$ is called
+\isatt{Vset}$(\alpha)$ in Isabelle/ZF. We showed, in the theory file
+\texttt{Relative\_Univ.thy}
+ the following
+relativization and closure results concerning this function, for a
+class $M$ satisfying the locale \isatt{M{\isacharunderscore}eclose}
+plus the Powerset Axiom and four instances of replacement.
+%
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ Vset{\isacharunderscore}abs{\isacharcolon}\ {\isachardoublequoteopen}{\isasymlbrakk}\ M{\isacharparenleft}i{\isacharparenright}{\isacharsemicolon}\ M{\isacharparenleft}V{\isacharparenright}{\isacharsemicolon}\ Ord{\isacharparenleft}i{\isacharparenright}\ {\isasymrbrakk}\ {\isasymLongrightarrow}\ is{\isacharunderscore}Vset{\isacharparenleft}M{\isacharcomma}i{\isacharcomma}V{\isacharparenright}\ {\isasymlongleftrightarrow}\ V\ {\isacharequal}\ {\isacharbraceleft}x{\isasymin}Vset{\isacharparenleft}i{\isacharparenright}{\isachardot}\ M{\isacharparenleft}x{\isacharparenright}{\isacharbraceright}{\isachardoublequoteclose}
+\end{isabelle}
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ Vset{\isacharunderscore}closed{\isacharcolon}\ {\isachardoublequoteopen}{\isasymlbrakk}\ M{\isacharparenleft}i{\isacharparenright}{\isacharsemicolon}\ Ord{\isacharparenleft}i{\isacharparenright}\ {\isasymrbrakk}\ {\isasymLongrightarrow}\ M{\isacharparenleft}{\isacharbraceleft}x{\isasymin}Vset{\isacharparenleft}i{\isacharparenright}{\isachardot}\ M{\isacharparenleft}x{\isacharparenright}{\isacharbraceright}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
+We also have the basic result
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ M{\isacharunderscore}into{\isacharunderscore}Vset{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\ {\isachardoublequoteopen}M{\isacharparenleft}a{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\ {\isachardoublequoteopen}{\isasymexists}i{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ {\isasymexists}V{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ ordinal{\isacharparenleft}M{\isacharcomma}i{\isacharparenright}\ {\isasymand}\ is{\isacharunderscore}Vfrom{\isacharparenleft}M{\isacharcomma}{\isadigit{0}}{\isacharcomma}i{\isacharcomma}V{\isacharparenright}\ {\isasymand}\ a{\isasymin}V{\isachardoublequoteclose}
+\end{isabelle}
+stating that $M$ is included in 
+$\union\{\isatt{Vset}(\alpha) : \alpha\in M\}$ (it's actually equal).
 
-\begin{enumerate}
-\item Primitive recursive arithmetic (meta-theory).
-\item \textbf{Theorem}: Con($\ZF(C)$) $\implies$ Con($\ZF(C)+\neg\CH$).
-\end{enumerate}
+For the proof of the Replacement Axiom, we assume that $\phi$ is
+functional in its first two variables when interpreted in $M[G]$ and
+the first ranges over the domain \isatt{c}. Then we show that the collection of
+all values of the second variable when the first ranges over
+\isatt{c} belongs to $M[G]$:
+%
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ Replace{\isacharunderscore}sats{\isacharunderscore}in{\isacharunderscore}MG{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\isanewline
+\ \ \ \ {\isachardoublequoteopen}c{\isasymin}M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}env\ {\isasymin}\ list{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \ \ {\isachardoublequoteopen}{\isasymphi}\ {\isasymin}\ formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}arity{\isacharparenleft}{\isasymphi}{\isacharparenright}\ {\isasymle}\ {\isadigit{2}}\ {\isacharhash}{\isacharplus}\ length{\isacharparenleft}env{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \ \ {\isachardoublequoteopen}univalent{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}\ c{\isacharcomma}\ {\isasymlambda}x\ v{\isachardot}\ sats{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}\ {\isasymphi}{\isacharcomma}\ {\isacharbrackleft}x{\isacharcomma}\ v{\isacharbrackright}\ {\isacharat}\ env{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\isanewline
+\ \ \ \ {\isachardoublequoteopen}{\isacharbraceleft}v{\isachardot}\ x{\isasymin}c{\isacharcomma}\ v{\isasymin}M{\isacharbrackleft}G{\isacharbrackright}\ {\isasymand}\ sats{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}\ {\isasymphi}{\isacharcomma}\ {\isacharbrackleft}x{\isacharcomma}v{\isacharbrackright}\ {\isacharat}\ env{\isacharparenright}{\isacharbraceright}\ {\isasymin}\ M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}
+\end{isabelle}
+%
+From this, the satisfaction of the Replacement Axiom in $M[G]$ follows
+very easily.
 
-and then to the ``type-theoretic approach'':
+The proof of the previous lemma, following Neeman, proceeds as usual
+by turning an argument concerning elements of $M[G]$ to one involving
+names lying in $M$, and connecting both worlds by using the forcing
+theorems. In the case at hand, by functionality of $\phi$ we know that
+for every $x\in c\cap M[G]$ there exists exactly one $v\in M[G]$ such
+that
+\isatt{sats{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}{\isasymphi}{\isacharcomma}{\isacharbrackleft}x{\isacharcomma}v{\isacharbrackright}{\isacharat}env{\isacharparenright}}.
+Now,
+given a name $\pi'\in M$ for $c$, every name of an element of $c$
+belongs to $\pi\defi \dom(\pi')\times \PP$, which is easily seen to be
+in $M$. We'll use $\pi$ to be the domain in an application of the
+Replacement Axiom in $M$. But now, obviously, we have lost
+functionality since there are many names $\dot v\in M$ for a fixed $v$
+in $M[G]$. To solve this issue, for each $\rho p \defi\lb\rho,p\rb\in \pi$ calculate the
+minimum rank of some $\tau\in M$ such that 
+$p\forces \phi(\rho,\tau,\dots)$ if there is one, or $0$ otherwise. By
+Replacement in $M$, we can show that the supremum \isatt{?sup} of these ordinals
+belongs to $M$ and we can construct a \isatt{bigname} $\defi$ 
+\isatt{{\isacharbraceleft}x{\isasymin}Vset{\isacharparenleft}{\isacharquery}sup{\isacharparenright}{\isachardot}\ x\ {\isasymin}\ M{\isacharbraceright}\ {\isasymtimes}\ {\isacharbraceleft}one{\isacharbraceright}}
+whose interpretation by (any generic) $G$ will include all possible elements
+as $v$ above.
 
-\begin{enumerate}
-\item CiC (meta-theory).
-\item CiC (theory, $\geq \ZFC + \exists \kappa$ inaccessible).
-\item \textbf{Theorem}: Con($\ZF(C)+\neg\CH$).
-\end{enumerate}
+The previous calculation required some absoluteness and closure
+results regarding the minimum ordinal binder, \isatt{Least}$(Q)$, also
+denoted $\mu x. Q(x)$, that can be found in the theory file
+\texttt{Least.thy}.
 

### 2020-01-01 IJCAR2020/proper_extension.tex
redondeando lo de extensión propia

@@
 
+To prove this results, we found it useful to formalize induction over
+the relation \isatt{ed}$(x,y) \defi x\in\dom(y)$, which is key
+to arguments involving names.
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ ed{\isacharunderscore}induction{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\ {\isachardoublequoteopen}{\isasymAnd}x{\isachardot}\ {\isasymlbrakk}{\isasymAnd}y{\isachardot}\ \ ed{\isacharparenleft}y{\isacharcomma}x{\isacharparenright}\ {\isasymLongrightarrow}\ Q{\isacharparenleft}y{\isacharparenright}\ {\isasymrbrakk}\ {\isasymLongrightarrow}\ Q{\isacharparenleft}x{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\ {\isachardoublequoteopen}Q{\isacharparenleft}a{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
+
 \texttt{Non\_Constructible.thy} contains the first example of a poset
@@
 lists with the partial order given by  ``\isatt{xs} is an extension of
-\isatt{ys}.''
+\isatt{ys}.'' The sufficient condition for a proper extension is that
+the forcing poset is \emph{separative}: every element has two
+incompatible extensions.
+
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ incompatible{\isacharunderscore}extensions{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\ {\isachardoublequoteopen}p{\isasymin}list{\isacharparenleft}{\isadigit{2}}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\ {\isachardoublequoteopen}{\isacharparenleft}p\ {\isacharat}\ {\isacharbrackleft}{\isadigit{0}}{\isacharbrackright}{\isacharparenright}\ {\isasymbottom}\ {\isacharparenleft}p\ {\isacharat}\ {\isacharbrackleft}{\isadigit{1}}{\isacharbrackright}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
+
+It is to be noted that both proof-documents comprise mere 99 and 79
+(resp.) lines.
 

### 2020-01-01 IJCAR2020/replacement_choice.tex
Choice, y ediciones menores

--- 2019-12-31 IJCAR2020/replacement.tex
@@
-\section{The axiom of replacement}
+\section{The axioms of replacement and choice}
+
+Apart from the obvious modifications arising from the change in the
+statement of the axiom schemes, the proofs that all axioms apart from
+those in the title hold in $M[G]$ are obtained exactly as in
+\cite{2019arXiv190103313G}.
+
+\subsection{Replacement}
 
@@
 
+\subsection{Choice}
+A first important observation is that the proof of $\AC$ in $M[G]$
+only requires the assumption that $M$ satisfies (a finite fragment of)
+$\ZFC$. There is no need to invoke Choice in the metatheory.
+
+Although our previous version of the developement used $\AC$, that was
+only needed to show the Rasiowa-Sikorski Lemma (RSL) for
+arbitrary posets. We have modularized the proof of the latter
+and now the version for countable posets that we use to show the
+existence of generic filters
+does not require Choice (as it is well known). We also bundled the
+full RSL along with our implementation of the principle of dependent
+choices in an independent branch of the dependency graph, which is the
+only place where the theory \texttt{ZF.AC} is invoked.
+
+Our statement of the axiom of choice is the one preferred for
+arguments involving transitive classes satisfying $\ZF$:
+%
+\begin{center}
+\isatt{{\isasymforall}x{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ {\isasymexists}a{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ {\isasymexists}f{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ ordinal{\isacharparenleft}M{\isacharcomma}a{\isacharparenright}\ {\isasymand}\ surjection{\isacharparenleft}M{\isacharcomma}a{\isacharcomma}x{\isacharcomma}f{\isacharparenright}}
+\end{center}
+%
+The Isabelle simplifier is able to show automatically that this
+statement is equivalent to the next one, in which the real notions of
+ordinal and surjection appear:
+%
+\begin{center}
+\isatt{{\isasymforall}x{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ {\isasymexists}a{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ {\isasymexists}f{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ Ord{\isacharparenleft}a{\isacharparenright}\ {\isasymand}\ f\ {\isasymin}\ surj{\isacharparenleft}a{\isacharcomma}x{\isacharparenright}}
+\end{center}
+
+As with the forcing axioms, the proof of $\AC$ in $M[G]$ follows the pattern of Kunen
+\cite[IV.2.27]{kunen2011set} and is rather
+straightforward; the only complicated technical point being to show
+that the relevant name belongs to $M$. We assume that \isatt{a}${}\neq\emptyset$
+belongs to $M[G]$ and has a name $\tau\in M$. By $\AC$ in $M$, there
+is a surjection \isatt{s} from an ordinal $\alpha\in M$ ($\subseteq M[G]$) onto
+$\dom(\tau)$. Now
+%
+\begin{center}
+\isatt{{\isacharbraceleft}opair{\isacharunderscore}name{\isacharparenleft}check{\isacharparenleft}{\isasymbeta}{\isacharparenright}{\isacharcomma}s{\isacharbackquote}{\isasymbeta}{\isacharparenright}{\isachardot}\ {\isasymbeta}{\isasymin}{\isasymalpha}{\isacharbraceright}\ {\isasymtimes}\ {\isacharbraceleft}one{\isacharbraceright}}
+\end{center}
+%
+is a name for a function \isatt{f} with domain $\alpha$ such that \isatt{a}
+is included in its range, and where
+\isatt{opair{\isacharunderscore}name}$(\sig,\rho)$ is a name for the
+ordered pair $\lb\val(G,\sig),\val(G,\rho)\rb$. From this, $\AC$ in
+$M[G]$ follows easily.
+
 %%% Local Variables: 

### 2020-01-01 IJCAR2020/forces_definition.tex
los dos tipos de inducción en Forcing_Thms
typos y labels para cada sección

@@
 \section{The definition of $\forceisa$}
+\label{sec:definition-forces}
 
@@
 (\ref{eq:def-forcing-equality}) and (\ref{eq:def-forcing-membership})
-as our corollaries \isatt{def\_forces\_eq} and
-\isatt{def\_forces\_mem}.
+as our corollaries \isatt{def{\isacharunderscore}forces{\isacharunderscore}eq} and
+\isatt{def{\isacharunderscore}forces{\isacharunderscore}mem}.
 
@@
 this might have been a little simpler to do, but we preferred to be as
-faithful as possible concerning this point to the text.
+faithful to the text as possible concerning this point.
 
-Once $\forcesat$ and its relativized version $\isatt{is\_forces\_at}$
+Once $\forcesat$ and its relativized version $\isatt{is{\isacharunderscore}forces{\isacharunderscore}at}$
 were defined, we proceeded to show absoluteness and provided internal

### 2020-01-01 IJCAR2020/forcing_theorems.tex
typos y labels para cada sección

@@
 \section{The forcing theorems}
+\label{sec:forcing-theorems}
 
@@
 principle of 
-\isatt{forces\_induction}, and then an induction on
+\isatt{forces{\isacharunderscore}induction}, and then an induction on
 $\formula$ wraps the former with the remaining cases (\isatt{Nand} and \isatt{Forall}).
@@
 %
-Its proof starts by an introduction of \isatt{forces\_induction};
+Its proof starts by an introduction of \isatt{forces{\isacharunderscore}induction};
 the  inductive cases for each atomic type were handled before as
-separate lemmas (\isatt{IV240a\_mem} and \isatt{IV240a\_eq}). We
+separate lemmas (\isatt{IV240a{\isacharunderscore}mem} and \isatt{IV240a{\isacharunderscore}eq}). We
 illustrate with the statement of the latter.

### 2020-01-01 IJCAR2020/forcing_theorems.tex
los teoremas de forcing faltantes

@@
 
+The main results in this section are the Density and Truth Lemmas,
+which we reproduce next.
+%
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ density{\isacharunderscore}lemma{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\isanewline
+\ \ \ \ {\isachardoublequoteopen}p{\isasymin}P{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}env{\isasymin}list{\isacharparenleft}M{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isasymle}length{\isacharparenleft}env{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\isanewline
+\ \ \ \ {\isachardoublequoteopen}{\isacharparenleft}p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}\ {\isasymlongleftrightarrow}\ dense{\isacharunderscore}below{\isacharparenleft}{\isacharbraceleft}q{\isasymin}P{\isachardot}\ {\isacharparenleft}q\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}{\isacharbraceright}{\isacharcomma}p{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ truth{\isacharunderscore}lemma{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\ \isanewline
+\ \ \ \ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}M{\isacharunderscore}generic{\isacharparenleft}G{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\ \isanewline
+\ \ \ \ \ {\isachardoublequoteopen}{\isasymAnd}env{\isachardot}\ env{\isasymin}list{\isacharparenleft}M{\isacharparenright}\ {\isasymLongrightarrow}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isasymle}length{\isacharparenleft}env{\isacharparenright}\ {\isasymLongrightarrow}\ \isanewline
+\ \ \ \ \ \ {\isacharparenleft}{\isasymexists}p{\isasymin}G{\isachardot}\ {\isacharparenleft}p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}{\isacharparenright}\ {\isasymlongleftrightarrow}\ sats{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}{\isasymphi}{\isacharcomma}map{\isacharparenleft}val{\isacharparenleft}G{\isacharparenright}{\isacharcomma}env{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
+%
+From these results, the semantical characterization of the forcing
+relation (the ``definition of $\forces$''
+\cite[IV.2.22]{kunen2011set}) follows easily (49 lines):
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ definition{\isacharunderscore}of{\isacharunderscore}forces{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\isanewline
+\ \ \ \ {\isachardoublequoteopen}p{\isasymin}P{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}env{\isasymin}list{\isacharparenleft}M{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isasymle}length{\isacharparenleft}env{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\isanewline
+\ \ \ \ {\isachardoublequoteopen}{\isacharparenleft}p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}\ {\isasymlongleftrightarrow}\isanewline
+\ \ \ \ \ {\isacharparenleft}{\isasymforall}G{\isachardot}{\isacharparenleft}M{\isacharunderscore}generic{\isacharparenleft}G{\isacharparenright}{\isasymand}\ p{\isasymin}G{\isacharparenright}{\isasymlongrightarrow}sats{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}{\isasymphi}{\isacharcomma}map{\isacharparenleft}val{\isacharparenleft}G{\isacharparenright}{\isacharcomma}env{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
 %%% Local Variables: 

### 2020-01-02 src/link-forcing.sh
archivos para el crosslinking de html

@@ -1,0 +1,57 @@
+#!/bin/bash
+old_dir=$PWD
+html_dir="output/html/Unsorted/Forcing/"
+
+function link_item {
+    #
+    # Usage:
+    # 
+    #   link_item HTML ITEMLIST SUFFIX
+    for z in `cat $2`
+    do 
+	l=`echo $z | cut -d"." -f2`$3
+	lprime=`echo $l | sed -e "s/&/\\\\\\&/g"`
+	t=`echo $z | cut -d"." -f1`
+	sed -i -e "s/>$l</><a class=\"pst-lnk\" href=\"$t.html#$t.$lprime\">$lprime<\/a></g" $1
+    done
+}
+
+function partial_job {
+    #
+    # Usage:
+    #
+    #   partial_job INITIALS ITEMLIST SUFFIX
+    for x in [$1]*.html
+    do 
+	link_item $x $2 $3
+    done
+}
+
+function full_job {
+    #
+    # Usage:
+    #
+    #   full_job ITEMLIST SUFFIX
+    partial_job CDE $1 $2 &
+    partial_job F $1 $2 &
+    partial_job ILN $1 $2 &
+    partial_job OPQ $1 $2 &
+    partial_job RSTUVW $1 $2 
+}
+
+cd $html_dir
+
+## Script de indizado y linkeo de definiciones completo
+for x in *.html; do y=`basename -s".html" $x`; sed -i -e "h;s/.*definition<\/span><\/span><span> <\/span><span>\([^<>\/]*\)<\/span>.*/$y.\1/w definiciones_$y.sn.txt" -e "g;s/<span \(class=\"command\">definition<\/span><\/span><span> <\/span><span>\)\([^<>\/]*\)<\/span><span>/<span id=\"$y.\2\"\1\2<\/span><span>/" $x; sed -i -e "N;h;s/.*<span class=\"command\">definition<\/span><\/span><span>\n<\/span><span>[ ]\+<\/span><span>\([^<>\/]*\)<\/span><span>.*/$y.\1/w definiciones_$y.cn.txt" -e "g;s/<span \(class=\"command\">definition<\/span><\/span><span>\n<\/span><span>[ ]\+<\/span><span>\)\([^<>\/]*\)<\/span><span>/<span id=\"$y.\2\"\1\2<\/span><span>/;P;D" $x ; done; cat definiciones_*txt > definiciones.txt ; rm -f definiciones_*txt
[…]

### 2020-01-02 src/fix_links.sh
Makefile y aledaños

@@ -3,7 +3,7 @@ old_dir=$PWD
 html_dir="output/html/Unsorted/Forcing/"
 
 cd $html_dir
-sed -i -- 's/a href="..\/..\/ZF\/ZF\/Main.html/a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF\/ZF.html/g' *.html
+sed -i -- 's/a href="..\/..\/ZF\/ZF\/ZF.html/a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF\/ZF.html/g' *.html
 sed -i -- 's/a href="..\/..\/ZF\/ZF\//a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF\//g' *.html
 sed -i -- 's/a href="..\/..\/ZF\/ZF-Constructible\//a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF-Constructible\//g' *.html
 cd $old_dir

### 2020-01-02 src/link-forcing.sh
arreglo problema de blank space

@@ -42,13 +42,13 @@ function full_job {
 cd $html_dir
 
 ## Script de indizado y linkeo de definiciones completo
-for x in *.html; do y=`basename -s".html" $x`; sed -i -e "h;s/.*definition<\/span><\/span><span> <\/span><span>\([^<>\/]*\)<\/span>.*/$y.\1/w definiciones_$y.sn.txt" -e "g;s/<span \(class=\"command\">definition<\/span><\/span><span> <\/span><span>\)\([^<>\/]*\)<\/span><span>/<span id=\"$y.\2\"\1\2<\/span><span>/" $x; sed -i -e "N;h;s/.*<span class=\"command\">definition<\/span><\/span><span>\n<\/span><span>[ ]\+<\/span><span>\([^<>\/]*\)<\/span><span>.*/$y.\1/w definiciones_$y.cn.txt" -e "g;s/<span \(class=\"command\">definition<\/span><\/span><span>\n<\/span><span>[ ]\+<\/span><span>\)\([^<>\/]*\)<\/span><span>/<span id=\"$y.\2\"\1\2<\/span><span>/;P;D" $x ; done; cat definiciones_*txt > definiciones.txt ; rm -f definiciones_*txt
+for x in *.html; do y=`basename -s".html" $x`; sed -i -e "h;s/.*definition<\/span><\/span><span>[ ]\+<\/span><span>\([^<>\/]*\)<\/span>.*/$y.\1/w definiciones_$y.sn.txt" -e "g;s/<span \(class=\"command\">definition<\/span><\/span><span> <\/span><span>\)\([^<>\/]*\)<\/span><span>/<span id=\"$y.\2\"\1\2<\/span><span>/" $x; sed -i -e "N;h;s/.*<span class=\"command\">definition<\/span><\/span><span>\n<\/span><span>[ ]\+<\/span><span>\([^<>\/]*\)<\/span><span>.*/$y.\1/w definiciones_$y.cn.txt" -e "g;s/<span \(class=\"command\">definition<\/span><\/span><span>\n<\/span><span>[ ]\+<\/span><span>\)\([^<>\/]*\)<\/span><span>/<span id=\"$y.\2\"\1\2<\/span><span>/;P;D" $x ; done; cat definiciones_*txt > definiciones.txt ; rm -f definiciones_*txt
 
[…]

### 2020-01-02 src/link-forcing.sh
linkeo de definiciones corregido

@@ -9,10 +9,10 @@ function link_item {
     #   link_item HTML ITEMLIST SUFFIX
     for z in `cat $2`
     do 
-	l=`echo $z | cut -d"." -f2`$3
+	l=`echo $z | cut -d"." -f2`
 	lprime=`echo $l | sed -e "s/&/\\\\\\&/g"`
 	t=`echo $z | cut -d"." -f1`
-	sed -i -e "s/>$l</><a class=\"pst-lnk\" href=\"$t.html#$t.$lprime\">$lprime<\/a></g" $1
+	sed -i -e "s/>$l$3</><a class=\"pst-lnk\" href=\"$t.html#$t.$lprime\">$lprime$3<\/a></g" $1
     done
 }
 

### 2020-01-03 IJCAR2020/conclusion.tex
avances en escritura
typos y labels para cada sección
los teoremas de forcing faltantes
algo en la conclusión

--- 2019-12-14 IJCAR2020/conclusion.tex
@@
-\section{Conclusion}
+\section{Conclusion and future work}
+\label{sec:conclusion}
+
+We consider that the formalization of the definition of $\forceisa$
+and its recursive characterization of forcing for atomic formulas is
+turning point
+in our project; the reason for this is that all further
+developments will not involve such a daunting metamathematical
+component. Even the proofs of the Fundamental Theorems of Forcing
+turned out to follow rather smoothly after this initial setup was
+ready, the only complicated affair being to show that various dense sets belong
+to $M$. 
+\begin{framed}
+  Actually, this is a point to taken care of: the machinery of
+  absoluteness and relativization must be developed for every new
+  concept that is introduced.
+\end{framed}
+
+In the course of obtaining internal formulas for the atomic case of
+forcing, a fruitful discussion
+concerning complementary perspectives on the role of proof assistants
+took place. An earlier approach relied more heavily in formula
+synthesis, thus making the Isabelle simplifier an indispensable main
+character. Following this line was quickier from the coding point of
+view since few new primitives were introduced and thus fewer lemmas
+concerning absoluteness and arities. On the downside, processing was a
+bit slower, the formulas synthesized were gigantic and the process on
+a whole as more error-prone. In fact, this approach was unsuccessful
+and we opted for a more detailed engineering, defining all
+intermediate steps. So the load on the assistant, in this part of the
+development, balanced from code-production to code-verification.
+
+The next task in our path is pretty clear: To develop the forcing
+notions to obtain the independence of $\CH$ 
+along with the prerrequisite combinatorial results, v.g.\ the
+$\Delta$-system lemma. A development of cofinality is under way in a
+joint work with E.~Pacheco Rodríguez, which is needed for a general
+statement of the latter.
+
+There are other debts, as well. In a second
+release of \isatt{ZF-Constructible}, we intend to conform it to the
+lines of \emph{Basic Set Theory (BST)} \cite[I.3.1]{kunen2011set} in
+which elementary results have proofs using alternatively Powerset or
+Replacement. The interest in this arises because many natural set
+models
+(rank-initial segments of the universe or the family $H(\kappa)$ of
+sets hereditarily of cardinality less than $\kappa$) satisfy one of
+those axioms and not the other. There are also still some older or
+less significant proofs are in tactical (\textbf{apply}) format; we
+hope we'll find the time to translate them to Isar. Finally, the
+automation of formula synthesis is on an early stage of
+development;  finishing that module will make writing our proofs of closure
+under various operations faster.
+\bigskip
+
+COMMENTS:
+\begin{enumerate}
+\item No merit in ``not formalizing'' the Reflection Principle.
+\end{enumerate}
 

### 2020-01-03 IJCAR2020/forcing_theorems.tex

@@
 \end{center}
-These two, along with  \isatt{Forces{\isacharunderscore}Equal} and
+Equivalent statements, along with the ones corresponding to \isatt{Forces{\isacharunderscore}Equal} and
 \isatt{Forces{\isacharunderscore}Nand}, appear in Kunen as the
@@
 %
-where the final step depends on previously proved
+(7 lines omitted) where the final step depends on the previously
+proved lemma
 \isatt{strengthening{\isacharunderscore}mem} and the characterization of
-$\forceisa$ for membership
+$\forceisa$ for membership.
 

### 2020-01-04 IJCAR2020/absoluteness.tex
typo
typos y labels para cada sección
mención de los axiomas

@@
-\section{Relativization and absoluteness}
+\section{Relativization,  absoluteness, and the axioms}
+\label{sec:relat-absol}
 
@@
 prerrequisites and powerful tools in working with transitive
-models. The \emph{relativization} $\phi^C(\bx)$ of a set-theoretic
+models. A \emph{class} is simply a predicate $C(x)$ with at least one
+free variable $x$.
+The \emph{relativization} $\phi^C(\bx)$ of a set-theoretic
 definition
@@
 The new formula $\phi^C(\bx)$ corresponds to what is obtained by defining
-the concept ``inside'' $C$. It turns out that many concepts mean the
+the concept ``inside'' $C$. In fact, for a class corresponding to a
+set $c$ (i.e.\ $C(x) \defi x \in c$), the relativization $\phi^C$ of a 
+sentence $\phi$ is equivalent to the satisfaction of $\phi$ in the
+first-order model $\lb c, \in\rb$.
+
+It turns out that many concepts mean the
 same after relativization to a nonempty transitive class $C$; formally
@@
 \forall\bx.\ C(\bx) \longrightarrow (\phi^C(\bx) \longleftrightarrow
-\phi^C(\bx))
+\phi(\bx))
 \]
@@
 of a nonempty transitive class.
+%
+\begin{framed}
+  New absoluteness results on powersets --- for \isatt{Least} and
+  \isatt{Vset} see Section~\ref{sec:axioms-replacement-choice}
+\end{framed}
 
-[New absoluteness results on powersets (/ universes?)]
+Apart from the axiom schemes, the rest of the $\ZFC$ 
+axioms are stated as predicates on classes (that is,
+of type \isatt{(i{\isasymRightarrow}o){\isasymRightarrow}o}); some
+of them were already defined by Paulson, and this formulation allows a
+better interaction with \isatt{ZF-Constructible}. 
+The axioms of Pairing, Union, Foundation,
+Extensionality, and Infinity are relativizations of the respective
+traditional 
+first-order sentences to the class argument. For the Axiom of Choice
+we selected a version best suited for the work with transitive
+models: the relativization of a sentence stating that for every $x$
+there is surjection from an ordinal onto $x$. Finally, Separation and
+Replacement were treated separately to effectively obtain first-order
+versions. It is to be noted that predicates in Isabelle/ZF are akin to
+second order variables and thus do not correspond to first-order
+formulas; in particular, there is no induction principle for functions
+of type \isatt{i{\isasymRightarrow}o}. For that reason, Separation and
+Replacement predicate \emph{on the satisfaction} of a formula $\phi$.
+With respect to our previous \cite{2019arXiv190103313G}, we lifted the
+arity restriction for the parameter $\phi$ on these schemes,
+streamlining various proofs as an immediate benefit and avoiding the
+need for tupling.
 

### 2020-01-04 IJCAR2020/absoluteness.tex
fuera explicación línea por línea (podría quedar en un apéndice?)

@@
 
-Apart from the axiom schemes, the rest of the $\ZFC$ 
+Apart from the axiom schemes, the $\ZFC$ 
 axioms are stated as predicates on classes (that is,

### 2020-01-04 IJCAR2020/forcing_theorems.tex
sacando cantidades de líneas
fuera explicación línea por línea (podría quedar en un apéndice?)

@@
 
-As an example of the second kind of induction (on formulas), we choose the
-following relatively simple result:
-
-\begin{isabelle}
-\isacommand{lemma}\isamarkupfalse%
-\ strengthening{\isacharunderscore}lemma{\isacharcolon}\isanewline
-\ \ \isakeyword{assumes}\ \isanewline
-\ \ \ \ {\isachardoublequoteopen}p{\isasymin}P{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}r{\isasymin}P{\isachardoublequoteclose}\ {\isachardoublequoteopen}r{\isasympreceq}p{\isachardoublequoteclose}\isanewline
-\ \ \isakeyword{shows}\isanewline
-\ \ \ \ {\isachardoublequoteopen}{\isasymAnd}env{\isachardot}\ env{\isasymin}list{\isacharparenleft}M{\isacharparenright}\ {\isasymLongrightarrow}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isasymle}length{\isacharparenleft}env{\isacharparenright}\ {\isasymLongrightarrow}\ p\ {\isasymtturnstile}\ {\isasymphi}\ env\isanewline
-\ \ \ \ \ {\isasymLongrightarrow}\ r\ {\isasymtturnstile}\ {\isasymphi}\ env{\isachardoublequoteclose}\isanewline
-%
-%
-\isacommand{using}\isamarkupfalse%
-\ assms{\isacharparenleft}{\isadigit{2}}{\isacharparenright}
-\end{isabelle}
-%
-The proof is divided in the 4 cases of definition of an element of $\formula$,
-%
-\begin{isabelle}
-\isacommand{proof}\isamarkupfalse%
-\ {\isacharparenleft}induct{\isacharparenright}\isanewline
-\ \ \isacommand{case}\isamarkupfalse%
-\ {\isacharparenleft}Member\ n\ m{\isacharparenright}\isanewline
-\ \ \isacommand{then}\isamarkupfalse%
-\isanewline
-\ \ \dots
-\isanewline
-\ \ \isacommand{show}\isamarkupfalse%
-\ {\isacharquery}case\ \isanewline
-\ \ \ \ \isacommand{using}\isamarkupfalse%
-\ Forces{\isacharunderscore}Member{\isacharbrackleft}of\ {\isacharunderscore}\ {\isachardoublequoteopen}nth{\isacharparenleft}n{\isacharcomma}env{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}nth{\isacharparenleft}m{\isacharcomma}env{\isacharparenright}{\isachardoublequoteclose}\ env\ n\ m{\isacharbrackright}\isanewline
-\ \ \ \ \ \ strengthening{\isacharunderscore}mem{\isacharbrackleft}of\ p\ r\ {\isachardoublequoteopen}nth{\isacharparenleft}n{\isacharcomma}env{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}nth{\isacharparenleft}m{\isacharcomma}env{\isacharparenright}{\isachardoublequoteclose}{\isacharbrackright}\ \isacommand{by}\isamarkupfalse%
-\ simp
-\end{isabelle}
-%
-(7 lines omitted) where the final step depends on the previously
-proved lemma
-\isatt{strengthening{\isacharunderscore}mem} and the characterization of
-$\forceisa$ for membership.
-
-The case of equality is entirely analogous, and the \isatt{Nand} and
-\isatt{Forall} cases are handled very simply.
-%
-\begin{isabelle}
-\isacommand{next}\isamarkupfalse%
-\isanewline
-\ \ \isacommand{case}\isamarkupfalse%
-\ {\isacharparenleft}Equal\ n\ m{\isacharparenright}\isanewline
-\ \ \dots\isanewline
-\isacommand{next}\isamarkupfalse%
-\isanewline
-\ \ \isacommand{case}\isamarkupfalse%
-\ {\isacharparenleft}Nand\ {\isasymphi}\ {\isasympsi}{\isacharparenright}\isanewline
-\ \ \isacommand{with}\isamarkupfalse%
-\ assms\isanewline
-\ \ \isacommand{show}\isamarkupfalse%
-\ {\isacharquery}case\ \isanewline
-\ \ \ \ \isacommand{using}\isamarkupfalse%
-\ Forces{\isacharunderscore}Nand\ Transset{\isacharunderscore}intf{\isacharbrackleft}OF\ trans{\isacharunderscore}M\ {\isacharunderscore}\ P{\isacharunderscore}in{\isacharunderscore}M{\isacharbrackright}\ pair{\isacharunderscore}in{\isacharunderscore}M{\isacharunderscore}iff\isanewline
-\ \ \ \ \ \ Transset{\isacharunderscore}intf{\isacharbrackleft}OF\ trans{\isacharunderscore}M\ {\isacharunderscore}\ leq{\isacharunderscore}in{\isacharunderscore}M{\isacharbrackright}\ leq{\isacharunderscore}transD\ \isacommand{by}\isamarkupfalse%
-\ auto\isanewline
-\isacommand{next}\isamarkupfalse%
-\isanewline
-\ \ \isacommand{case}\isamarkupfalse%
-\ {\isacharparenleft}Forall\ {\isasymphi}{\isacharparenright}\isanewline
-\ \ \isacommand{with}\isamarkupfalse%
-\ assms\isanewline
-\ \ \isacommand{have}\isamarkupfalse%
-\ {\isachardoublequoteopen}p\ {\isasymtturnstile}\ {\isasymphi}\ {\isacharparenleft}{\isacharbrackleft}x{\isacharbrackright}\ {\isacharat}\ env{\isacharparenright}{\isachardoublequoteclose}\ \isakeyword{if}\ {\isachardoublequoteopen}x{\isasymin}M{\isachardoublequoteclose}\ \isakeyword{for}\ x\isanewline
-\ \ \ \ \isacommand{using}\isamarkupfalse%
-\ that\ Forces{\isacharunderscore}Forall\ \isacommand{by}\isamarkupfalse%
-\ simp\isanewline
-\ \ \isacommand{with}\isamarkupfalse%
-\ \underline{Forall}\ \isanewline
-\ \ \isacommand{have}\isamarkupfalse%
-\ {\isachardoublequoteopen}r\ {\isasymtturnstile}\ {\isasymphi}\ {\isacharparenleft}{\isacharbrackleft}x{\isacharbrackright}\ {\isacharat}\ env{\isacharparenright}{\isachardoublequoteclose}\ \isakeyword{if}\ {\isachardoublequoteopen}x{\isasymin}M{\isachardoublequoteclose}\ \isakeyword{for}\ x\isanewline
-\ \ \ \ \isacommand{using}\isamarkupfalse%
-\ that\ pred{\isacharunderscore}le{\isadigit{2}}\ \isacommand{by}\isamarkupfalse%
-\ {\isacharparenleft}simp{\isacharparenright}\isanewline
-\ \ \isacommand{with}\isamarkupfalse%
-\ assms\ \underline{Forall}\isanewline
-\ \ \isacommand{show}\isamarkupfalse%
-\ {\isacharquery}case\ \isanewline
-\ \ \ \ \isacommand{using}\isamarkupfalse%
-\ Forces{\isacharunderscore}Forall\ \isacommand{by}\isamarkupfalse%
-\ simp\isanewline
-\isacommand{qed}\isamarkupfalse
-\end{isabelle}
-%
-It can be noted that the inductive hypothesis
-gets used in the last case (underlined here as
-\isatt{\underline{Forall}}), but not in the case for \isatt{Nand}.
-
-The main results in this section are the Density and Truth Lemmas,
+Examples of proofs  using the second kind of induction include
+the basic \isatt{strengthening{\isacharunderscore}lemma} and the main
+results in this section, the lemmas of  Density  (actually, its nontrivial
+direction
+\isatt{dense{\isacharunderscore}below{\isacharunderscore}imp{\isacharunderscore}forces})
+and Truth, 
 which we reproduce next.
@@
 relation (the ``definition of $\forces$''
-\cite[IV.2.22]{kunen2011set}) follows easily (49 lines):
+\cite[IV.2.22]{kunen2011set}) follows easily:
 \begin{isabelle}
@@
 \end{isabelle}
+
 %%% Local Variables: 

### 2020-01-04 IJCAR2020/proper_extension.tex
Choice, y ediciones menores
typos y labels para cada sección
sacando cantidades de líneas
fuera explicación línea por línea (podría quedar en un apéndice?)

@@
 \section{Example of proper extension}
+\label{sec:example-proper-extension}
 
@@
 
-Two short theory files contains the relevant
+Two short theory files contain the relevant
 results. \texttt{Ordinals\_In\_MG.thy} shows, using the closure of $M$
@@
 
-\texttt{Non\_Constructible.thy} contains the first example of a poset
+\texttt{Non\_Constructible.thy} contains our first example of a poset
 that interprets the locale
@@
 
-It is to be noted that both proof-documents comprise mere 99 and 79
-(resp.) lines.
-
 %%% Local Variables: 

### 2020-01-04 IJCAR2020/isabelle_metatheory.tex
pequeñas ediciones al paper
typos y labels para cada sección
tres párrafos de metateoría

@@
 \section{Isabelle and (meta)theories}
+\label{sec:isabelle-metatheories}
 
-\begin{enumerate}
-\item Description of \emph{Pure} (meta-theory).
-\item Isabelle/ZF + $\exists M$ ctm of $\ZF(C)$ (theory). 
-\item $\formula$ (inner theory).
-\item \textbf{Theorem}: $\exists N$ ctm of $\ZF(C)+\neg\CH$.
-\end{enumerate}
+Isabelle \cite{Isabelle,DBLP:books/sp/Paulson94} is a general proof
+assistant based on fragment of higher-order logic called
+\emph{Pure}. The logical primitives of this meta-level are
+\isatt{\isasymLongrightarrow}, \isatt{\&\&\&}, and \isatt{\isasymAnd}
+(implication, conjunction, and universal
+quantification, resp.), which operate on the meta-booleans
+\isatt{prop}. The results presented in this work are theorems of a
+version of $\ZF$ set theory called \emph{Isabelle/ZF}, which is one of the
+``object logics'' that can be defined on top of Pure (which is then
+used as a language to define rules). Isabelle/ZF defines types
+\isatt{i} and \isatt{o} for sets and booleans, resp., and the $\ZF$
+axioms are written down as terms of type \isatt{o}.
 
-in comparison to the ``proof theory approach'':
+More specifically, our results work under the hypothesis of
+the existence of a countable transitive model (ctm) of $\ZFC$.% 
+%%
+\footnote{The countability is only used to prove the existence of
+  generic filters and can be thus replaced in favor of this
+  hypothesis.} 
+%%
+This hypothesis follows, for instance, from the existence of an
+inaccessible cardinal. As such, our framework is weaker than those
+found usually in type theories with universes, but allows us to work
+``platonistically''--- assuming we are in a universe of sets (namely,
+\isatt{i}) an performing constructions there.
 
-\begin{enumerate}
-\item Primitive recursive arithmetic (meta-theory).
-\item \textbf{Theorem}: Con($\ZF(C)$) $\implies$ Con($\ZF(C)+\neg\CH$).
-\end{enumerate}
-
-and then to the ``type-theoretic approach'':
-
-\begin{enumerate}
-\item CiC (meta-theory).
-\item CiC (theory, $\geq \ZFC + \exists \kappa$ inaccessible).
-\item \textbf{Theorem}: Con($\ZF(C)+\neg\CH$).
-\end{enumerate}
+On the downside, our approach is not able to provide us with finitary
+consistency proofs. It is well known that, for example, the
+implication $\Con(\ZF) \implies \Con(\ZFC+\neg\CH)$ can be proved in
+\emph{primitive recursive arithmetic (PRA)}. To achieve this, however,
+it would have implied to work focusing on the proof mechanisms
+and distracting us from our main goal, that is, formalize the ctm
+approach currently used by many mathematicians.
 

### 2020-01-04 IJCAR2020/conclusion.tex
ediciones minimas al paper

@@
 those axioms and not the other. There are also still some older or
-less significant proofs are in tactical (\textbf{apply}) format; we
+less significant proofs written in tactical (\textbf{apply}) format; we
 hope we'll find the time to translate them to Isar. Finally, the

### 2020-01-04 IJCAR2020/replacement_choice.tex
typos y labels para cada sección
algo en la conclusión
ediciones minimas al paper

@@
 \section{The axioms of replacement and choice}
+\label{sec:axioms-replacement-choice}
 
@@
 functional in its first two variables when interpreted in $M[G]$ and
-the first ranges over the domain \isatt{c}. Then we show that the collection of
-all values of the second variable when the first ranges over
-\isatt{c} belongs to $M[G]$:
+the first ranges over the domain \isatt{c}${}\in M[G]$. Then we show
+that the collection of
+all values of the second variable, when the first ranges over
+\isatt{c}, belongs to $M[G]$:
 %
@@
 functionality since there are many names $\dot v\in M$ for a fixed $v$
-in $M[G]$. To solve this issue, for each $\rho p \defi\lb\rho,p\rb\in \pi$ calculate the
+in $M[G]$. To solve this issue, for each $\rho p \defi\lb\rho,p\rb\in
+\pi$ we calculate the
 minimum rank of some $\tau\in M$ such that 
@@
 Replacement in $M$, we can show that the supremum \isatt{?sup} of these ordinals
-belongs to $M$ and we can construct a \isatt{bigname} $\defi$ 
+belongs to $M$ and we can construct a \isatt{?bigname} $\defi$ 
 \isatt{{\isacharbraceleft}x{\isasymin}Vset{\isacharparenleft}{\isacharquery}sup{\isacharparenright}{\isachardot}\ x\ {\isasymin}\ M{\isacharbraceright}\ {\isasymtimes}\ {\isacharbraceleft}one{\isacharbraceright}}

### 2020-01-04 IJCAR2020/isabelle_metatheory.tex

@@
 \isatt{prop}. The results presented in this work are theorems of a
-version of $\ZF$ set theory called \emph{Isabelle/ZF}, which is one of the
+version of $\ZF$ set theory (without the Axiom of Choice, $\AC$) 
+called \emph{Isabelle/ZF}, which is one of the
+
 ``object logics'' that can be defined on top of Pure (which is then

### 2020-01-04 IJCAR2020/proper_extension.tex

@@
 
-To prove this results, we found it useful to formalize induction over
+To prove these results, we found it useful to formalize induction over
 the relation \isatt{ed}$(x,y) \defi x\in\dom(y)$, which is key

### 2020-01-05 IJCAR2020/absoluteness.tex
resumen de lo que hemos estado haciendo, en intro

@@
 as ``$x$ is a subset of $y$'' or of a function like $y=\P(x)$) to
-a class $C$ is obtained by taking its (first-order) definition and
-restricting all of its quantifiers to $C$.
+a class $C$ is obtained by restricting all of its quantifiers to $C$.
 

### 2020-01-05 IJCAR2020/forcing_theorems.tex
resumen de lo que hemos estado haciendo, en intro

@@
 various (dense) subsets of $\PP$ were in $M$; for this, we had to
-recourse to several absoluteness ad-hoc lemmas.
+recourse to several ad-hoc absoluteness lemmas.
 
@@
 
+The present statement of the Fundamental Theorems is almost exactly
+the same of those of \cite{2019arXiv190103313G}, with the only
+modification being the bound on arities and a missing typing
+constraint. This implied only minor adjustments in the proofs of the
+satisfaction of axioms.
+
 %%% Local Variables: 

### 2020-01-05 IJCAR2020/intro.tex
typos y labels para cada sección
mención de los axiomas
resumen de lo que hemos estado haciendo, en intro

--- 2019-12-10 IJCAR2020/intro.tex
@@
 \section{Introduction}
+\label{sec:introduction}
+
+The present work reports on the third stage of our project of
+formalizing the theory of forcing and its applications as presented in
+one of the more
+important references on the subject, Kunen's Set Theory
+\cite{kunen2011set} (a rewrite of the classical \cite{kunen1980}).
+
+We
+work using the  implementation of Zermelo-Fraenkel ($\ZF$)
+set theory \emph{Isabelle/ZF} by Paulson and Grabczewski \cite{DBLP:journals/jar/PaulsonG96}. In
+an early paper \cite{2018arXiv180705174G}, we set up the first
+elements, defining forcing notions, names, generic extensions, and
+showing the existence of generic filters via the Rasiowa-Sikorski
+lemma (RSL). Our second (unpublished) technical report
+\cite{2019arXiv190103313G} advanced by presenting the first accurate
+\emph{formal abstract} of the Fundamental Theorems of Forcing, and
+using them to show that that the $\ZF$ axioms apart from Replacement
+and Infinity hold in all generic extensions.
+
+This paper contains the proof of Fundamental Theorems and complete
+proofs of the Axioms of Infinity, Replacement, and Choice in all
+generic extensions. In particular, we were able to fulfill the
+promised formal abstract for the Forcing 
+Theorems almost to the letter. A requirement for Infinity and the
+absoluteness of forcing for atomic formulas, we finished the inteface
+between our development and
+Paulson's constructibility library \cite{paulson_2003} which enables
+us to do well-founded
+recursion inside transitive models of an appropriate finite fragment
+of $\ZF$. As a by-product, we honored a long debt: the fact that the
+generic filter $G$ belongs to the extension $M[G]$ and $M\sbq M[G]$.
+Finally, our development is now independent of $\AC$: We modularized
+RSL in such a way that a version for countable 
+posets does not require choice.
+
+\begin{framed}
+  We implemented an automatic ML machinery of (bounded) renaming
+  of formulas.
+\end{framed}
+
+\subsection{Related work}
+To the best of our knowledge, all of the previous works in
+formalization of the method 
+of forcing has been done in different variants of type theory. The
+most important is the recent one by 
+Han \& van Doorn
+\cite{han_et_al:LIPIcs:2019:11074,DBLP:conf/cpp/HanD20}, which includes
+a formalization of the independence of $\CH$ by means
+the Boolean-valued approach to forcing, using the Lean
+proof assistant \cite{DBLP:conf/cade/MouraKADR15}.
 

### 2020-01-05 IJCAR2020/isabelle_metatheory.tex
resumen de lo que hemos estado haciendo, en intro

--- 2020-01-04 IJCAR2020/isabelle_metatheory.tex
@@
 \isatt{prop}. The results presented in this work are theorems of a
-version of $\ZF$ set theory called \emph{Isabelle/ZF}, which is one of the
+version of $\ZF$ set theory (without the Axiom of Choice, $\AC$) 
+called \emph{Isabelle/ZF}, which is one of the
 ``object logics'' that can be defined on top of Pure (which is then
@@
 
+\begin{framed}
+  Internalized formulas, MIGHT BE HERE OR IN THE NEXT SECTION.
+\end{framed}
+
 %%% Local Variables: 

### 2020-01-05 IJCAR2020/proper_extension.tex
resumen de lo que hemos estado haciendo, en intro

@@
 Two short theory files contain the relevant
-results. \texttt{Ordinals\_In\_MG.thy} shows, using the closure of $M$
+results. \verb|Ordinals_In_MG.thy| shows, using the closure of $M$
 under ranks, that $M$ and $M[G]$ share the same ordinals (actually,
@@
 
-\texttt{Non\_Constructible.thy} contains our first example of a poset
+\verb|Non_Constructible.thy| contains our first example of a poset
 that interprets the locale

### 2020-01-05 IJCAR2020/replacement_choice.tex
resumen de lo que hemos estado haciendo, en intro

@@
 \isatt{Vset}$(\alpha)$ in Isabelle/ZF. We showed, in the theory file
-\texttt{Relative\_Univ.thy}
+\verb|Relative_Univ.thy|
  the following
@@
 denoted $\mu x. Q(x)$, that can be found in the theory file
-\texttt{Least.thy}.
+\verb|Least.thy|.
 
@@
 
-Our statement of the axiom of choice is the one preferred for
+Our statement of the Axiom of Choice is the one preferred for
 arguments involving transitive classes satisfying $\ZF$:

### 2020-01-06 IJCAR2020/forces_definition.tex
haciendo (demasiado?) énfasis en ctms

@@
 moving in the transitive closure of $\tau_2$. To disentangle this, one
-must realize that this last recursion must be described sintactically:
+must realize that this last recursion must be described syntactically:
 the definition of $\forceisa(\phi)$ for atomic $\phi$ is then an

### 2020-01-06 IJCAR2020/forcing_in_isabelle_zf.tex
ToDo list, typos, nuevos archivos
avances en la definición de forces en paper, nueva sección
avances en escritura
Choice, y ediciones menores
algo en la conclusión
apendice con prueba levemente explicada
haciendo (demasiado?) énfasis en ctms

@@
 Miguel Pagano\inst{1} \and \\
-Pedro Sánchez Terraf\inst{1,2}\orcidID{0000-0003-3928-6942}}
+Pedro Sánchez Terraf\inst{1,2}%\orcidID{0000-0003-3928-6942}
+}
 %
@@
 \begin{abstract}
-We formalize the development of forcing in the set theory framework of
+We formalize the theory of forcing in the set theory framework of
 Isabelle/ZF. Under the assumption of the existence of countable
-transitive model of $\ZF$, we construct a generic extension and show
-that the latter also satisfies $\ZF$.
+transitive model of $\ZFC$, we construct a proper generic extension and show
+that the latter also satisfies $\ZFC$. In doing so, we remodularized
+Paulson's ZF-Constructibility library.
 \keywords{Isabelle/ZF \and forcing \and names \and generic extension \and constructibility.}
@@
 
+\input{forces_definition}
+
+\input{forcing_theorems}
+
+\input{proper_extension}
+
+\input{replacement_choice}
+
 \input{conclusion}
@@
 \bibliography{forcing_in_isabelle_zf}
+
+\appendix
+
+\input{sample_proof}
+
 \end{document}

### 2020-01-06 IJCAR2020/intro.tex
apendice con prueba levemente explicada
haciendo (demasiado?) énfasis en ctms

@@
 an early paper \cite{2018arXiv180705174G}, we set up the first
-elements, defining forcing notions, names, generic extensions, and
+elements of the countable transitive model (ctm) approach, defining
+forcing notions, names, generic extensions, and
 showing the existence of generic filters via the Rasiowa-Sikorski
@@
 of $\ZF$. As a by-product, we honored a long debt: the fact that the
-generic filter $G$ belongs to the extension $M[G]$ and $M\sbq M[G]$.
+generic filter $G$ belongs to the extension $M[G]$ and 
+$M\sbq M[G]$. We also include an example of forcing notion for which
+the last inclusion is proper.
 Finally, our development is now independent of $\AC$: We modularized
@@
 formalization of the method 
-of forcing has been done in different variants of type theory. The
+of forcing have been done in different variants of type theory, and
+none of them uses the ctm approach. The
 most important is the recent one by 
-Han \& van Doorn
+Han and van Doorn
 \cite{han_et_al:LIPIcs:2019:11074,DBLP:conf/cpp/HanD20}, which includes

### 2020-01-08 IJCAR2020/forcing_in_isabelle_zf.tex
inicio de absolutez de concatenacion de listas

@@
 We formalize the theory of forcing in the set theory framework of
-Isabelle/ZF. Under the assumption of the existence of countable
+Isabelle/ZF. Under the assumption of the existence of a countable
 transitive model of $\ZFC$, we construct a proper generic extension and show
 that the latter also satisfies $\ZFC$. In doing so, we remodularized
-Paulson's ZF-Constructibility library.
+Paulson's \verb|ZF-Constructibility| library.
 \keywords{Isabelle/ZF \and forcing \and names \and generic extension \and constructibility.}

### 2020-01-17 IJCAR2020/proper_extension.tex
adecuando al codigo nuevo

@@
 
-\verb|Non_Constructible.thy| contains our first example of a poset
+\verb|Succession_Poset.thy| contains our first example of a poset
 that interprets the locale
-\isatt{forcing{\isacharunderscore}notion}. It is the set of all binary
-lists with the partial order given by  ``\isatt{xs} is an extension of
-\isatt{ys}.'' The sufficient condition for a proper extension is that
+\isatt{forcing{\isacharunderscore}notion}, essentially the notion for
+adding one Cohen real. It is the set $2^{<\om}$ of all finite binary
+sequences partially  ordered by reverse inclusion.
+The sufficient condition for a proper extension is that
 the forcing poset is \emph{separative}: every element has two
-incompatible extensions.
+incompatible (\isatt{{\isasymbottom}f}) extensions. Here,
+\isatt{fun{\isacharunderscore}upd{\isacharparenleft}f{\isacharcomma}x{\isacharparenright}}
+adds \isatt{x} to the end of the sequence.
 
@@
 \isacommand{lemma}\isamarkupfalse%
-\ incompatible{\isacharunderscore}extensions{\isacharcolon}\isanewline
-\ \ \isakeyword{assumes}\ {\isachardoublequoteopen}p{\isasymin}list{\isacharparenleft}{\isadigit{2}}{\isacharparenright}{\isachardoublequoteclose}\isanewline
-\ \ \isakeyword{shows}\ {\isachardoublequoteopen}{\isacharparenleft}p\ {\isacharat}\ {\isacharbrackleft}{\isadigit{0}}{\isacharbrackright}{\isacharparenright}\ {\isasymbottom}\ {\isacharparenleft}p\ {\isacharat}\ {\isacharbrackleft}{\isadigit{1}}{\isacharbrackright}{\isacharparenright}{\isachardoublequoteclose}
+\ seqspace{\isacharunderscore}separative{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\ {\isachardoublequoteopen}f{\isasymin}{\isadigit{2}}{\isacharcircum}{\isacharless}{\isasymomega}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\ {\isachardoublequoteopen}fun{\isacharunderscore}upd{\isacharparenleft}f{\isacharcomma}{\isadigit{0}}{\isacharparenright}\ {\isasymbottom}f\ fun{\isacharunderscore}upd{\isacharparenleft}f{\isacharcomma}{\isadigit{1}}{\isacharparenright}{\isachardoublequoteclose}
 \end{isabelle}
+
+We prove in the theory file \verb|Proper_Extension.thy| that, in
+general, every separative forcing notion gives rise to a proper
+extension.
 

### 2020-01-21 IJCAR2020/intro.tex

@@
 
+In the course of our work we find it useful to develop Isar methods to
+automate repetitive tasks. Part of the interface with Paulson's
+library consisted in constructing formulas for each relativized
+concept; Isar already has a method (\texttt{schematic\_goal}) which can
+synthesize terms for unbound schematic variables in theorems. The
+synthesized term, however, is not available outside the theorem; we
+introduced a method that creates a definition from a schematic
+goal. The second method is concerned with renaming of formulas: we
+improved our small library of bounded renamings with a method that
+given the source and target environments figures out the renaming
+function and produces the relevant lemmas about it.
+
 \begin{framed}
-  We implemented an automatic ML machinery of (bounded) renaming
-  of formulas.
+  TODO: Comentar en algún lado las limitaciones del método de
+  definiciones sintetizadas.
+
+  TODO: Explicar mejor que las fórmulas que necesitamos definir son un
+  ejemplo exacto de lo que se puede hacer con schematic\_goal.
 \end{framed}

### 2020-01-21 IJCAR2020/forces_definition.tex
Párrafo pelotudo sobre recursión.

@@
 
+\paragraph*{Technical details}
+Paulson proved absoluteness of functions defined by wellfounded
+recursion over a transitive relation. Some of our definitions by
+recursion (\emph{check} and \emph{forces}) do not fit in that scheme.
+One can replace the relation $R$ for its transitive closure $R^+$ in the
+recursive definition because one can prove, in general, that
+$F\!\upharpoonright\!(R^{-1}(x))(y) =
+F\!\upharpoonright\!({R^+}^{-1}(x))(y)$ whenever $(x,y) \in R$.
+
 \begin{framed}
-  In order to prove absoluteness of functions defined by wellfounded
-  recursion, we need to show that we can replace the relevant
-  wellfounded relations by their transitive closures. (EXPAND HERE OR
-  SOMEWHERE ELSE).
+  TODO: ¿Qué más decimos?
+
+  TODO: por qué no tenemos ese resultado general que mencionamos?
 \end{framed}

### 2020-01-21 IJCAR2020/proper_extension.tex
cambiando notacion del poset de sucesiones

@@
 the forcing poset is \emph{separative}: every element has two
-incompatible (\isatt{{\isasymbottom}f}) extensions. Here,
-\isatt{fun{\isacharunderscore}upd{\isacharparenleft}f{\isacharcomma}x{\isacharparenright}}
-adds \isatt{x} to the end of the sequence.
+incompatible (\isatt{{\isasymbottom}s}) extensions. Here,
+\isatt{seq{\isacharunderscore}upd{\isacharparenleft}f{\isacharcomma}x{\isacharparenright}}
+adds \isatt{x} to the end of the sequence \isatt{f}.
 
@@
 \ \ \isakeyword{assumes}\ {\isachardoublequoteopen}f{\isasymin}{\isadigit{2}}{\isacharcircum}{\isacharless}{\isasymomega}{\isachardoublequoteclose}\isanewline
-\ \ \isakeyword{shows}\ {\isachardoublequoteopen}fun{\isacharunderscore}upd{\isacharparenleft}f{\isacharcomma}{\isadigit{0}}{\isacharparenright}\ {\isasymbottom}f\ fun{\isacharunderscore}upd{\isacharparenleft}f{\isacharcomma}{\isadigit{1}}{\isacharparenright}{\isachardoublequoteclose}
+\ \ \isakeyword{shows}\ {\isachardoublequoteopen}seq{\isacharunderscore}upd{\isacharparenleft}f{\isacharcomma}{\isadigit{0}}{\isacharparenright}\ {\isasymbottom}s\ seq{\isacharunderscore}upd{\isacharparenleft}f{\isacharcomma}{\isadigit{1}}{\isacharparenright}{\isachardoublequoteclose}
 \end{isabelle}

### 2020-01-22 IJCAR2020/intro.tex
ediciones menores

@@
 Theorems almost to the letter. A requirement for Infinity and the
-absoluteness of forcing for atomic formulas, we finished the inteface
+absoluteness of forcing for atomic formulas, we finished the interface
 between our development and

### 2020-01-22 IJCAR2020/isabelle_metatheory.tex
haciendo (demasiado?) énfasis en ctms
ediciones menores

@@
 More specifically, our results work under the hypothesis of
-the existence of a countable transitive model (ctm) of $\ZFC$.% 
+the existence of a ctm of $\ZFC$.% 
 %%
@@
 ``platonistically''--- assuming we are in a universe of sets (namely,
-\isatt{i}) an performing constructions there.
+\isatt{i}) an performing constructions there. We may state the main
+theorem of our formalization:
+\begin{isabelle}
+\isacommand{theorem}\isamarkupfalse%
+\ extensions{\isacharunderscore}of{\isacharunderscore}ctms{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\ \isanewline
+\ \ \ \ {\isachardoublequoteopen}M\ {\isasymapprox}\ nat{\isachardoublequoteclose}\ {\isachardoublequoteopen}Transset{\isacharparenleft}M{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ ZF{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\ \isanewline
+\ \ \ \ {\isachardoublequoteopen}{\isasymexists}N{\isachardot}\ \isanewline
+\ \ \ \ \ \ M\ {\isasymsubseteq}\ N\ {\isasymand}\ N\ {\isasymapprox}\ nat\ {\isasymand}\ Transset{\isacharparenleft}N{\isacharparenright}\ {\isasymand}\ N\ {\isasymTurnstile}\ ZF\ {\isasymand}\ M{\isasymnoteq}N\ {\isasymand}\isanewline
+\ \ \ \ \ \ {\isacharparenleft}{\isasymforall}{\isasymalpha}{\isachardot}\ Ord{\isacharparenleft}{\isasymalpha}{\isacharparenright}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\isacharparenright}\ {\isasymand}\isanewline
+\ \ \ \ \ \ {\isacharparenleft}M{\isacharcomma}\ {\isacharbrackleft}{\isacharbrackright}{\isasymTurnstile}\ AC\ {\isasymlongrightarrow}\ N\ {\isasymTurnstile}\ ZFC{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
+Here, \isatt{\isasymapprox} stands for equipotence, \isatt{nat} is the
+set of natural numbers, and the predicate 
+\isatt{Transset} indicates transitivity.
 

### 2020-01-22 IJCAR2020/absoluteness.tex
ediciones menores al paper

@@
 The version of \isatt{ZF-Constructible} we present weakens the
-assumptions of [many / a dozen / several] absoluteness theorems to that
-of a nonempty transitive class.
-%
-\begin{framed}
-  New absoluteness results on powersets --- for \isatt{Least} and
-  \isatt{Vset} see Section~\ref{sec:axioms-replacement-choice}
-\end{framed}
+assumptions of many absoluteness theorems to that
+of a nonempty transitive class; also there are some stronger results
+such as the relativization of powersets.
 

### 2020-01-22 IJCAR2020/intro.tex
ediciones menores al paper

@@
 
-In the course of our work we find it useful to develop Isar methods to
+In the course of our work we found it useful to develop Isar methods to
 automate repetitive tasks. Part of the interface with Paulson's
 library consisted in constructing formulas for each relativized
-concept; Isar already has a method (\texttt{schematic\_goal}) which can
+concept; and actually Isabelle's Simplifier can
 synthesize terms for unbound schematic variables in theorems. The

### 2020-01-23 IJCAR2020/absoluteness.tex
necesidad de internalizar, obtención axiomas en primer orden, movi main_theorem

@@
 
-A benefit of the work with transitive models is that [one concept of
-  ordinal accross the universe / $M$ / $M[G]$]
+A benefit of the work with transitive models is that many 
+concepts concepts (pairs, unions, and fundamentally ordinals) are
+uniform across the universe \isatt{i}, a ctm (of an adequate fragment of
+$\ZF$) $M$ and any of its extensions $M[G]$
 
@@
 
+Once this class versions of the axioms are set up, it is not hard to
+apply our synthesis method to obtain their internal, first-order
+counterparts.
+\begin{framed}
+  Some code for axiom synthesis.
+\end{framed}
+
 %%% Local Variables: 

### 2020-01-23 IJCAR2020/conclusion.tex
resumen de lo que hemos estado haciendo, en intro
necesidad de internalizar, obtención axiomas en primer orden, movi main_theorem

@@
 took place. An earlier approach relied more heavily in formula
-synthesis, thus making the Isabelle simplifier an indispensable main
+synthesis, thus making the Simplifier an indispensable main
 character. Following this line was quickier from the coding point of
@@
 under various operations faster.
-\bigskip
-
-COMMENTS:
-\begin{enumerate}
-\item No merit in ``not formalizing'' the Reflection Principle.
-\end{enumerate}
 

### 2020-01-23 IJCAR2020/isabelle_metatheory.tex
ediciones menores al paper
necesidad de internalizar, obtención axiomas en primer orden, movi main_theorem

--- 2020-01-05 IJCAR2020/isabelle_metatheory.tex
@@
 assistant based on fragment of higher-order logic called
-\emph{Pure}. The logical primitives of this meta-level are
-\isatt{\isasymLongrightarrow}, \isatt{\&\&\&}, and \isatt{\isasymAnd}
-(implication, conjunction, and universal
-quantification, resp.), which operate on the meta-booleans
-\isatt{prop}. The results presented in this work are theorems of a
+\emph{Pure}. %% The logical primitives of this meta-level are
+%% \isatt{\isasymLongrightarrow}, \isatt{\&\&\&}, and \isatt{\isasymAnd}
+%% (implication, conjunction, and universal
+%% quantification, resp.), which operate on the meta-booleans
+%% \isatt{prop}. 
+The results presented in this work are theorems of a
 version of $\ZF$ set theory (without the Axiom of Choice, $\AC$) 
@@
 More specifically, our results work under the hypothesis of
-the existence of a countable transitive model (ctm) of $\ZFC$.% 
+the existence of a ctm of $\ZFC$.% 
 %%
-\footnote{The countability is only used to prove the existence of
+\footnote{By Gödel's Second incompleteness theorem, one must assume at
+  least the existence of some model of $\ZF$. 
+  The countability is only used to prove the existence of
   generic filters and can be thus replaced in favor of this
@@
 ``platonistically''--- assuming we are in a universe of sets (namely,
-\isatt{i}) an performing constructions there.
+\isatt{i}) and performing constructions there.
 
@@
 
-\begin{framed}
-  Internalized formulas, MIGHT BE HERE OR IN THE NEXT SECTION.
-\end{framed}
+It should be noted that Pure is a very weak framework and has no
+induction/recursion capabilities. So the only way to define functions
+by recursion is inside the object logic. (This works the same for
+Isabelle/HOL.) For this reason, to define the relation of forcing, we
+needed to resort to \emph{internalized} first-order formulas: they
+form a recursively defined set \isatt{formula}. For example, the
+predicate of safisfaction
+\isatt{sats::i{\isasymRightarrow}i{\isasymRightarrow}i{\isasymRightarrow}o}
+(noted $M,\mathit{ms}\models\phi$ for a set $M$, for a set $M$,
+$\mathit{ms}\in\isatt{list}(M)$ and $\phi\in\formula$
+%% (noted \isatt{M, ms \isasymTurnstile{ }\isasymphi} for a set \isatt{M},
+%% \isatt{ms {\isasymin} list(M)} and \isatt{{\isasymphi} {\isasymin} formula})
+had already been defined by recursion in \cite{paulson_2003}. We added
+the corresponding notion $M\models \Phi$ for satisfaction of a set of
+sentences.
 

### 2020-01-23 IJCAR2020/replacement_choice.tex
necesidad de internalizar, obtención axiomas en primer orden, movi main_theorem

@@
 %
-The Isabelle simplifier is able to show automatically that this
+The Simplifier is able to show automatically that this
 statement is equivalent to the next one, in which the real notions of
@@
 
+\subsection{The main theorem}
+With all these elements in place, we are able to transcript the main
+theorem of our formalization:
+\begin{isabelle}
+\isacommand{theorem}\isamarkupfalse%
+\ extensions{\isacharunderscore}of{\isacharunderscore}ctms{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\ \isanewline
+\ \ \ \ {\isachardoublequoteopen}M\ {\isasymapprox}\ nat{\isachardoublequoteclose}\ {\isachardoublequoteopen}Transset{\isacharparenleft}M{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ ZF{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\ \isanewline
+\ \ \ \ {\isachardoublequoteopen}{\isasymexists}N{\isachardot}\ \isanewline
+\ \ \ \ \ \ M\ {\isasymsubseteq}\ N\ {\isasymand}\ N\ {\isasymapprox}\ nat\ {\isasymand}\ Transset{\isacharparenleft}N{\isacharparenright}\ {\isasymand}\ N\ {\isasymTurnstile}\ ZF\ {\isasymand}\ M{\isasymnoteq}N\ {\isasymand}\isanewline
+\ \ \ \ \ \ {\isacharparenleft}{\isasymforall}{\isasymalpha}{\isachardot}\ Ord{\isacharparenleft}{\isasymalpha}{\isacharparenright}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\isacharparenright}\ {\isasymand}\isanewline
+\ \ \ \ \ \ {\isacharparenleft}M{\isacharcomma}\ {\isacharbrackleft}{\isacharbrackright}{\isasymTurnstile}\ AC\ {\isasymlongrightarrow}\ N\ {\isasymTurnstile}\ ZFC{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
+Here, \isatt{\isasymapprox} stands for equipotence, \isatt{nat} is the
+set of natural numbers, and the predicate 
+\isatt{Transset} indicates transitivity.
+
 %%% Local Variables: 

### 2020-01-23 IJCAR2020/forcing_theorems.tex
abreviación de sats en algunos teos

@@
 \ \ \ \ \ {\isachardoublequoteopen}{\isasymAnd}env{\isachardot}\ env{\isasymin}list{\isacharparenleft}M{\isacharparenright}\ {\isasymLongrightarrow}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isasymle}length{\isacharparenleft}env{\isacharparenright}\ {\isasymLongrightarrow}\ \isanewline
-\ \ \ \ \ \ {\isacharparenleft}{\isasymexists}p{\isasymin}G{\isachardot}\ {\isacharparenleft}p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}{\isacharparenright}\ {\isasymlongleftrightarrow}\ sats{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}{\isasymphi}{\isacharcomma}map{\isacharparenleft}val{\isacharparenleft}G{\isacharparenright}{\isacharcomma}env{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\ \ \ \ \ \ {\isacharparenleft}{\isasymexists}p{\isasymin}G{\isachardot}\ {\isacharparenleft}p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}{\isacharparenright}\ \ {\isasymlongleftrightarrow}\ \ M{\isacharbrackleft}G{\isacharbrackright}\ {\isacharcomma}\ map{\isacharparenleft}val{\isacharparenleft}G{\isacharparenright}{\isacharcomma}env{\isacharparenright}\ {\isasymTurnstile}\ {\isasymphi}{\isachardoublequoteclose}
 \end{isabelle}
@@
 \ \ \ \ {\isachardoublequoteopen}{\isacharparenleft}p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}\ {\isasymlongleftrightarrow}\isanewline
-\ \ \ \ \ {\isacharparenleft}{\isasymforall}G{\isachardot}{\isacharparenleft}M{\isacharunderscore}generic{\isacharparenleft}G{\isacharparenright}{\isasymand}\ p{\isasymin}G{\isacharparenright}{\isasymlongrightarrow}sats{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}{\isasymphi}{\isacharcomma}map{\isacharparenleft}val{\isacharparenleft}G{\isacharparenright}{\isacharcomma}env{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\ \ \ \ \ {\isacharparenleft}{\isasymforall}G{\isachardot}{\isacharparenleft}M{\isacharunderscore}generic{\isacharparenleft}G{\isacharparenright}{\isasymand}\ p{\isasymin}G{\isacharparenright}{\isasymlongrightarrow}\ M{\isacharbrackleft}G{\isacharbrackright}\ {\isacharcomma}\ map{\isacharparenleft}val{\isacharparenleft}G{\isacharparenright}{\isacharcomma}env{\isacharparenright}\ {\isasymTurnstile}\ {\isasymphi}{\isacharparenright}{\isachardoublequoteclose}
 \end{isabelle}

### 2020-01-23 IJCAR2020/replacement_choice.tex
abreviación de sats en algunos teos
menor: mudo comentarios de Pure, explico más main theorem

@@
 \ \ \ \ {\isachardoublequoteopen}{\isasymphi}\ {\isasymin}\ formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}arity{\isacharparenleft}{\isasymphi}{\isacharparenright}\ {\isasymle}\ {\isadigit{2}}\ {\isacharhash}{\isacharplus}\ length{\isacharparenleft}env{\isacharparenright}{\isachardoublequoteclose}\isanewline
-\ \ \ \ {\isachardoublequoteopen}univalent{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}\ c{\isacharcomma}\ {\isasymlambda}x\ v{\isachardot}\ sats{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}\ {\isasymphi}{\isacharcomma}\ {\isacharbrackleft}x{\isacharcomma}\ v{\isacharbrackright}\ {\isacharat}\ env{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \ \ {\isachardoublequoteopen}univalent{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}\ c{\isacharcomma}\ {\isasymlambda}x\ v{\isachardot}\ {\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}\ {\isacharbrackleft}x{\isacharcomma}v{\isacharbrackright}{\isacharat}env\ {\isasymTurnstile}\ {\isasymphi}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
 \ \ \isakeyword{shows}\isanewline
-\ \ \ \ {\isachardoublequoteopen}{\isacharbraceleft}v{\isachardot}\ x{\isasymin}c{\isacharcomma}\ v{\isasymin}M{\isacharbrackleft}G{\isacharbrackright}\ {\isasymand}\ sats{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}\ {\isasymphi}{\isacharcomma}\ {\isacharbrackleft}x{\isacharcomma}v{\isacharbrackright}\ {\isacharat}\ env{\isacharparenright}{\isacharbraceright}\ {\isasymin}\ M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}
+\ \ \ \ {\isachardoublequoteopen}{\isacharbraceleft}v{\isachardot}\ x{\isasymin}c{\isacharcomma}\ v{\isasymin}M{\isacharbrackleft}G{\isacharbrackright}\ {\isasymand}\ {\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}\ {\isacharbrackleft}x{\isacharcomma}v{\isacharbrackright}{\isacharat}env\ {\isasymTurnstile}\ {\isasymphi}{\isacharparenright}{\isacharbraceright}\ {\isasymin}\ M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}
 \end{isabelle}
@@
 that
-\isatt{sats{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}{\isasymphi}{\isacharcomma}{\isacharbrackleft}x{\isacharcomma}v{\isacharbrackright}{\isacharat}env{\isacharparenright}}.
-Now,
+$M[G], [x,v]\mathbin{@} \mathit{env} \models \phi$. Now,
 given a name $\pi'\in M$ for $c$, every name of an element of $c$
@@
 belongs to $M$ and we can construct a \isatt{?bigname} $\defi$ 
-\isatt{{\isacharbraceleft}x{\isasymin}Vset{\isacharparenleft}{\isacharquery}sup{\isacharparenright}{\isachardot}\ x\ {\isasymin}\ M{\isacharbraceright}\ {\isasymtimes}\ {\isacharbraceleft}one{\isacharbraceright}}
+\isatt{{\isacharbraceleft}x{\isasymin}Vset{\isacharparenleft}{\isacharquery}sup{\isacharparenright}{\isachardot}\ x\ {\isasymin}\
+}$M$\isatt{{\isacharbraceright}\ {\isasymtimes}\ {\isacharbraceleft}one{\isacharbraceright}}
 whose interpretation by (any generic) $G$ will include all possible elements
@@
 set of natural numbers, and the predicate 
-\isatt{Transset} indicates transitivity.
+\isatt{Transset} indicates transitivity; and as usual, \isatt{AC}
+denotes the Axiom of Choice, and \isatt{ZF} and \isatt{ZFC} the
+corresponding subsets of \isatt{formula}.
 

### 2020-01-23 IJCAR2020/forces_definition.tex
menor: mudo comentarios de Pure, explico más main theorem
dos explicaciones de forcing y teoremas

@@
 definition of a function $\forceisa$ that maps the set of internal
-formulas into itself.
+formulas into itself. It is the very reason of applicability of
+forcing that the satisfaction of the formula $\forceisa(\phi)$ in a
+model controls the satisfaction of $\phi$ in all of its generic
+extensions.
 
@@
 \ \ \isakeyword{assumes}\isanewline
-\ \ \ \ {\isachardoublequoteopen}{\isasymAnd}{\isasymtau}\ {\isasymtheta}{\isachardot}\ {\isasymlbrakk}{\isasymAnd}{\isasymsigma}{\isachardot}\ {\isasymsigma}{\isasymin}domain{\isacharparenleft}{\isasymtheta}{\isacharparenright}\ {\isasymLongrightarrow}\ Q{\isacharparenleft}{\isasymtau}{\isacharcomma}{\isasymsigma}{\isacharparenright}{\isasymrbrakk}\ {\isasymLongrightarrow}\ R{\isacharparenleft}{\isasymtau}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\ \ \ \ {\isachardoublequoteopen}{\isasymAnd}{\isasymtau}\ {\isasymtheta}{\isachardot}\ {\isasymlbrakk}{\isasymAnd}{\isasymsigma}{\isachardot}\ {\isasymsigma}{\isasymin}domain{\isacharparenleft}{\isasymtheta}{\isacharparenright}\ {\isasymLongrightarrow}\ Q{\isacharparenleft}{\isasymtau}{\isacharcomma}{\isasymsigma}{\isacharparenright}{\isasymrbrakk}\ {\isasymLongrightarrow}\ R{\isacharparenleft}{\isasymtau}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isachardoublequoteclose}\footnotemark\isanewline
 \ \ \ \ {\isachardoublequoteopen}{\isasymAnd}{\isasymtau}\ {\isasymtheta}{\isachardot}\ {\isasymlbrakk}{\isasymAnd}{\isasymsigma}{\isachardot}\ {\isasymsigma}{\isasymin}domain{\isacharparenleft}{\isasymtau}{\isacharparenright}\ {\isasymunion}\ domain{\isacharparenleft}{\isasymtheta}{\isacharparenright}\ {\isasymLongrightarrow}\ R{\isacharparenleft}{\isasymsigma}{\isacharcomma}{\isasymtau}{\isacharparenright}\ {\isasymand}\ R{\isacharparenleft}{\isasymsigma}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isasymrbrakk}\isanewline
@@
 \end{isabelle}
+\footnotetext{The logical primitives of \emph{Pure} are
+\isatt{\isasymLongrightarrow}, \isatt{\&\&\&}, and \isatt{\isasymAnd}
+(implication, conjunction, and universal
+quantification, resp.), which operate on the meta-booleans
+\isatt{prop}.}
 %

### 2020-01-23 IJCAR2020/forcing_theorems.tex
dos explicaciones de forcing y teoremas

@@
 As an example of the first class, we can take a look at our
-formalization of \cite[Lem.~IV.2.40(a)]{kunen2011set}:
+formalization of \cite[Lem.~IV.2.40(a)]{kunen2011set}. Note that the
+\emph{context} of the lemma includes the assumption of \isatt{P} being
+a forcing notion, and the predicate of being $M$-generic is relative
+to \isatt{P}:
 
@@
 \ \ \ \ \ {\isachardoublequoteopen}{\isasymAnd}env{\isachardot}\ env{\isasymin}list{\isacharparenleft}M{\isacharparenright}\ {\isasymLongrightarrow}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isasymle}length{\isacharparenleft}env{\isacharparenright}\ {\isasymLongrightarrow}\ \isanewline
-\ \ \ \ \ \ {\isacharparenleft}{\isasymexists}p{\isasymin}G{\isachardot}\ {\isacharparenleft}p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}{\isacharparenright}\ \ {\isasymlongleftrightarrow}\ \ M{\isacharbrackleft}G{\isacharbrackright}\ {\isacharcomma}\ map{\isacharparenleft}val{\isacharparenleft}G{\isacharparenright}{\isacharcomma}env{\isacharparenright}\ {\isasymTurnstile}\ {\isasymphi}{\isachardoublequoteclose}
+\ \ \ \ \ \ {\isacharparenleft}{\isasymexists}p{\isasymin}G{\isachardot}\ {\isacharparenleft}p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}{\isacharparenright}\ \ {\isasymlongleftrightarrow}\ \ M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}\ map{\isacharparenleft}val{\isacharparenleft}G{\isacharparenright}{\isacharcomma}env{\isacharparenright}\ {\isasymTurnstile}\ {\isasymphi}{\isachardoublequoteclose}
 \end{isabelle}
@@
 \ \ \ \ {\isachardoublequoteopen}{\isacharparenleft}p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}\ {\isasymlongleftrightarrow}\isanewline
-\ \ \ \ \ {\isacharparenleft}{\isasymforall}G{\isachardot}{\isacharparenleft}M{\isacharunderscore}generic{\isacharparenleft}G{\isacharparenright}{\isasymand}\ p{\isasymin}G{\isacharparenright}{\isasymlongrightarrow}\ M{\isacharbrackleft}G{\isacharbrackright}\ {\isacharcomma}\ map{\isacharparenleft}val{\isacharparenleft}G{\isacharparenright}{\isacharcomma}env{\isacharparenright}\ {\isasymTurnstile}\ {\isasymphi}{\isacharparenright}{\isachardoublequoteclose}
+\ \ \ \ \ {\isacharparenleft}{\isasymforall}G{\isachardot}\ M{\isacharunderscore}generic{\isacharparenleft}G{\isacharparenright}{\isasymand}\ p{\isasymin}G\ {\isasymlongrightarrow}\ M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}\ map{\isacharparenleft}val{\isacharparenleft}G{\isacharparenright}{\isacharcomma}env{\isacharparenright}\ {\isasymTurnstile}\ {\isasymphi}{\isacharparenright}{\isachardoublequoteclose}
 \end{isabelle}

### 2020-01-24 src/fix_links.sh
make actualizado

@@ -1,6 +1,6 @@
 #!/bin/bash
 old_dir=$PWD
-html_dir="output/html/Unsorted/Forcing/"
+html_dir="output/html/ZF/Forcing/"
 
 cd $html_dir
 sed -i -- 's/a href="..\/..\/ZF\/ZF\/ZF.html/a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF\/ZF.html/g' *.html

### 2020-01-24 src/link-forcing.sh

@@ -1,6 +1,19 @@
 #!/bin/bash
 old_dir=$PWD
-html_dir="output/html/Unsorted/Forcing/"
+html_dir="output/html/ZF/Forcing/"
+list1=Forces_Definition.html
+list2="Forcing_Theorems.html Arities.html"
+list3="Names.html Interface.html"
+list4="Forcing_Main.html FrecR.html Nat_Miscellanea.html"
+list5="Forcing_Notions.html Relative_Univ.html Forcing_Data.html\
+ Renaming.html Internalizations.html"
+list6="Replacement_Axiom.html Succession_Poset.html Recursion_Thms.html\
+ Separation_Rename.html Choice_Axiom.html Least.html Pointed_DC.html\
+ Renaming_Auto.html Union_Axiom.html"
+list7="Ordinals_In_MG.html Powerset_Axiom.html Separation_Axiom.html\
+ Foundation_Axiom.html Pairing_Axiom.html\
+ Proper_Extension.html Rasiowa_Sikorski.html Extensionality_Axiom.html\
+ Infinity_Axiom.html Synthetic_Definition.html"
 
 function link_item {
     #
@@ -20,8 +33,8 @@ function partial_job {
     #
     # Usage:
     #
-    #   partial_job INITIALS ITEMLIST SUFFIX
-    for x in [$1]*.html
+    #   partial_job HTMLS ITEMLIST SUFFIX
+    for x in $1
     do 
 	link_item $x $2 $3
     done
@@ -32,26 +45,45 @@ function full_job {
     # Usage:
     #
     #   full_job ITEMLIST SUFFIX
-    partial_job CDE $1 $2 &
-    partial_job F $1 $2 &
-    partial_job ILN $1 $2 &
-    partial_job OPQ $1 $2 &
-    partial_job RSTUVW $1 $2 
+    partial_job "$list1" $1 $2 &
+    partial_job "$list2" $1 $2 &
+    partial_job "$list3" $1 $2 &
+    partial_job "$list4" $1 $2 &
+    partial_job "$list5" $1 $2 &
+    partial_job "$list6" $1 $2 &
+    partial_job "$list7" $1 $2
 }
 
+# echo Back-up of html directory
+# cp -R $html_dir ~/tmp/
+
+echo Changing to html directory
 cd $html_dir
 
[…]

### 2020-01-24 src/fix_links.sh
arreglo linkeo a ZF y Constructible

@@ -3,7 +3,7 @@ old_dir=$PWD
 html_dir="output/html/ZF/Forcing/"
 
 cd $html_dir
-sed -i -- 's/a href="..\/..\/ZF\/ZF\/ZF.html/a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF\/ZF.html/g' *.html
-sed -i -- 's/a href="..\/..\/ZF\/ZF\//a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF\//g' *.html
-sed -i -- 's/a href="..\/..\/ZF\/ZF-Constructible\//a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF-Constructible\//g' *.html
+sed -i -- 's/a href="..\/ZF\/ZF.html/a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF\/ZF.html/g' *.html
+sed -i -- 's/a href="..\/ZF\/ZF\//a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF\//g' *.html
+#sed -i -- 's/a href="..\/ZF\/ZF-Constructible\//a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF-Constructible\//g' *.html
 cd $old_dir

### 2020-01-24 src/link-forcing.sh
mejor distribución de archivos en tar

@@ -4,7 +4,8 @@ html_dir="output/html/ZF/Forcing/"
 list1=Forces_Definition.html
 list2="Forcing_Theorems.html Arities.html"
 list3="Names.html Interface.html"
-list4="Forcing_Main.html FrecR.html Nat_Miscellanea.html"
+list4="Forcing_Main.html FrecR.html Nat_Miscellanea.html\
+ Internal_ZFC_Axioms.html"
 list5="Forcing_Notions.html Relative_Univ.html Forcing_Data.html\
  Renaming.html Internalizations.html"
 list6="Replacement_Axiom.html Succession_Poset.html Recursion_Thms.html\

### 2020-01-24 IJCAR2020/absoluteness.tex
Arreglos muy pavos. Comentario sobre definiciones sintetizadas.

@@
 apply our synthesis method to obtain their internal, first-order
-counterparts.
+counterparts.\footnote{It should be noted that our method for
+  synthetic definition requires that the schematic goal had a
+  particular format where the only unbound schematic variable occurs
+  as an (open) formula whose satisfaction is equivalent to some
+  relativized predicate.}
 \begin{framed}

### 2020-01-24 IJCAR2020/conclusion.tex
Arreglos muy pavos. Comentario sobre definiciones sintetizadas.

@@
 and its recursive characterization of forcing for atomic formulas is
-turning point
+a turning point
 in our project; the reason for this is that all further

### 2020-01-25 IJCAR2020/absoluteness.tex
código de síntesis y saco ese apéndice de mierda

--- 2020-01-23 IJCAR2020/absoluteness.tex
@@
 apply our synthesis method to obtain their internal, first-order
-counterparts.
-\begin{framed}
-  Some code for axiom synthesis.
-\end{framed}
+counterparts.\footnote{It should be noted that our method for
+  synthetic definition requires that the schematic goal had a
+  particular format where the only unbound schematic variable occurs
+  as an (open) formula whose satisfaction is equivalent to some
+  relativized predicate.}
+For the case of the Union Axiom, the statement for classes is the
+following
+\begin{isabelle}
+upair{\isacharunderscore}ax{\isacharparenleft}C{\isacharparenright}{\isacharequal}{\isacharequal}{\isasymforall}x{\isacharbrackleft}C{\isacharbrackright}{\isachardot}{\isasymforall}y{\isacharbrackleft}C{\isacharbrackright}{\isachardot}{\isasymexists}z{\isacharbrackleft}C{\isacharbrackright}{\isachardot}\ upair{\isacharparenleft}C{\isacharcomma}x{\isacharcomma}y{\isacharcomma}z{\isacharparenright}
+\end{isabelle}
+where \isatt{upair} says that \isatt{z} is the unordered pair of
+\isatt{x} and \isatt{y}, relative to \isatt{C}.
+
+The following schematic lemma synthesizes its internal version,
+\begin{isabelle}
+\isacommand{schematic{\isacharunderscore}goal}\isamarkupfalse%
+\ ZF{\isacharunderscore}pairing{\isacharunderscore}auto{\isacharcolon}\isanewline
+\ \ \ \ {\isachardoublequoteopen}upair{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}A{\isacharparenright}\ {\isasymlongleftrightarrow}\ {\isacharparenleft}A{\isacharcomma}\ {\isacharbrackleft}{\isacharbrackright}\ {\isasymTurnstile}\ {\isacharquery}zfpair{\isacharparenright}{\isachardoublequoteclose}\isanewline
+\isacommand{unfolding}\isamarkupfalse%
+\ upair{\isacharunderscore}ax{\isacharunderscore}def\ \isanewline
+\ \ \isacommand{by}\isamarkupfalse%
+\ {\isacharparenleft}{\isacharparenleft}rule\ sep{\isacharunderscore}rules\ {\isacharbar}\ simp{\isacharparenright}{\isacharplus}{\isacharparenright}
+\end{isabelle}
+(the actual formula obtained is
+\isatt{Forall(Forall(Exists(upair{\isacharunderscore}fm(2,1,0))))}) and
+our \isatt{synthesize} method introduces a new term
+\isatt{ZF{\isacharunderscore}pairing{\isacharunderscore}fm} for it.
+\begin{isabelle}
+\isacommand{synthesize}\isamarkupfalse%
+\ {\isachardoublequoteopen}ZF{\isacharunderscore}pairing{\isacharunderscore}fm{\isachardoublequoteclose}\ \isakeyword{from{\isacharunderscore}schematic}\ {\isachardoublequoteopen}ZF{\isacharunderscore}pairing{\isacharunderscore}auto{\isachardoublequoteclose}%
+\end{isabelle}
 

### 2020-01-25 IJCAR2020/forcing_theorems.tex
digo que existen los locales

@@
 formalization of \cite[Lem.~IV.2.40(a)]{kunen2011set}. Note that the
-\emph{context} of the lemma includes the assumption of \isatt{P} being
-a forcing notion, and the predicate of being $M$-generic is relative
-to \isatt{P}:
+context (a ``locale'' in Isabelle terminology, namely \isatt{forcing{\isacharunderscore}data}) of the lemma 
+includes the assumption of \isatt{P} being
+a forcing notion, and the predicate of being $M$-generic is defined in
+terms of \isatt{P}:
 

### 2020-01-25 IJCAR2020/conclusion.tex
pongo recuadro inline con un poco más

@@
 to $M$. 
-\begin{framed}
-  Actually, this is a point to taken care of: the machinery of
-  absoluteness and relativization must be developed for every new
-  concept that is introduced.
-\end{framed}
+%
+Actually, this is a point to be taken care of: For every new
+concept that is introduced, some lemmas concerning 
+relativization and closure must be proved to be able to synthesize its
+internal definition. Further automation must be developed for this
+purpose.
 

### 2020-01-25 IJCAR2020/intro.tex
Arreglos muy pavos. Comentario sobre definiciones sintetizadas.
web resources

@@
 
-\begin{framed}
-  TODO: Comentar en algún lado las limitaciones del método de
-  definiciones sintetizadas.
-
-  TODO: Explicar mejor que las fórmulas que necesitamos definir son un
-  ejemplo exacto de lo que se puede hacer con schematic\_goal.
-\end{framed}
+The source code of our formalization, written for the 2019 version of
+Isabelle, can be browsed and downloaded at
+\url{https://cs.famaf.unc.edu.ar/~pedro/forcing/}
 

### 2020-01-25 IJCAR2020/absoluteness.tex
ediciones menores

@@
 
-The version of \isatt{ZF-Constructible} we present weakens the
+The version we present of constructibility library,
+\isatt{ZF-Constructible-Trans}, weakens the
 assumptions of many absoluteness theorems to that
@@
 Apart from the axiom schemes, the $\ZFC$ 
-axioms are stated as predicates on classes (that is,
+axioms are initially stated as predicates on classes (that is,
 of type \isatt{(i{\isasymRightarrow}o){\isasymRightarrow}o}); some
@@
 Replacement were treated separately to effectively obtain first-order
-versions. It is to be noted that predicates in Isabelle/ZF are akin to
-second order variables and thus do not correspond to first-order
-formulas; in particular, there is no induction principle for functions
-of type \isatt{i{\isasymRightarrow}o}. For that reason, Separation and
+versions afterwards. It is to be noted that predicates in Isabelle/ZF
+are akin to second order variables and thus do not correspond to first-order
+formulas.
+%% in particular, there is no induction principle for functions
+%% of type \isatt{i{\isasymRightarrow}o}. 
+For that reason, Separation and
 Replacement predicate \emph{on the satisfaction} of a formula $\phi$.
@@
 
-The following schematic lemma synthesizes its internal version,
+The following schematic lemma synthesizes its internal version,%
+\footnote{The use of such schematic goals and the orginal definition
+  of the collection of lemmas \isatt{sep{\isacharunderscore}rules} are due to Paulson.}
 \begin{isabelle}

### 2020-01-25 IJCAR2020/forces_definition.tex
Arreglos muy pavos. Comentario sobre definiciones sintetizadas.
ediciones menores

@@
 formulas is also defined by (mutual) recursion: to know if $\tau_1\in\tau_2$ is
-forced, one must check if $\tau_1=\sigma$ is forced for $\sigma$
+forced by $p$ (notation: $\forcesmem(p,\tau_1,\tau_2)$), one must check if $\tau_1=\sigma$ is forced for $\sigma$
 moving in the transitive closure of $\tau_2$. To disentangle this, one
@@
 
-\begin{framed}
-  TODO: ¿Qué más decimos?
-
-  TODO: por qué no tenemos ese resultado general que mencionamos?
-\end{framed}
-
 %%% Local Variables: 

### 2020-01-25 IJCAR2020/forcing_theorems.tex
ediciones menores

--- 2020-01-23 IJCAR2020/forcing_theorems.tex
@@
 formalization of \cite[Lem.~IV.2.40(a)]{kunen2011set}. Note that the
-\emph{context} of the lemma includes the assumption of \isatt{P} being
-a forcing notion, and the predicate of being $M$-generic is relative
-to \isatt{P}:
+context (a ``locale,'' in Isabelle terminology, namely \isatt{forcing{\isacharunderscore}data}) of the lemma 
+includes the assumption of \isatt{P} being
+a forcing notion, and the predicate of being $M$-generic is defined in
+terms of \isatt{P}:
 

### 2020-01-25 IJCAR2020/absoluteness.tex
Por acá pasó Miguel.

@@
 When this is the case, we say that the relation defined by $\phi$ is
-\emph{absolute for transitive models}. (The absoluteness of functions
-moreover requires that
-the relativized definition also behaves functionally over $C$.) As
-examples, the relation of inclusion $\subseteq$ ---and actually, any
-relation defined by a formula (equivalent to one) using only bounded
-quantifiers 
-$(\forall x\in y)$ and $(\exists x\in y)$--- is absolute for
-transitive models. On the contrary, this is not the case with the powerset
-operation.
+\emph{absolute for transitive models}.\footnote{Absoluteness of
+  functions also requires the relativized definition to behave
+  functionally over $C$.} As examples, the relation of inclusion
+$\subseteq$ ---and actually, any relation defined by a formula
+(equivalent to one) using only bounded quantifiers $(\forall x\in y)$
+and $(\exists x\in y)$--- is absolute for transitive models. On the
+contrary, this is not the case with the powerset operation.
 
-A benefit of the work with transitive models is that many 
-concepts concepts (pairs, unions, and fundamentally ordinals) are
-uniform across the universe \isatt{i}, a ctm (of an adequate fragment of
-$\ZF$) $M$ and any of its extensions $M[G]$
+% If a concept is absolute, then one can reason ``externally'' instead
+% of ``inside'' the model; thus, one has at hand any lemma about ``the''
+% concept and not what the model believes of the concept.
+A benefit of working with transitive models
+is that many concepts (pairs, unions, and fundamentally ordinals) are
+uniform across the universe \isatt{i}, a ctm (of an adequate fragment
+of $\ZF$) $M$ and any of its extensions $M[G]$.
 
-A part of this project is to refactorize Paulson's formalization
-\cite{paulson_2003} of Gödel's \cite{godel-L}. The main objective is
-to maximize applicability of the relativization machinery by adjusting
-the hypothesis of a greater part of it early development. Paulson's
-architecture had only in mind the consistency of $\ZFC$, but, for
-instance, in order to apply it in the development of forcing, too much
-is assumed at the beginning; more seriously, some assumptions can't be
-regarded as ``first-order'' (v.g. the Replacement Scheme).
+Paulson's formalization \cite{paulson_2003} of Gödel's \cite{godel-L}
+already contains absoluteness results which were crucial to our
+project; we realized however that they could be refactorized to be
+more useful. The main objective is to maximize applicability of
+the relativization machinery by adjusting the hypothesis of a greater
+part of it early development. Paulson's architecture had only in mind
+the consistency of $\ZFC$, but, for instance, in order to apply it in
+the development of forcing, too much is assumed at the beginning; more
+seriously, some assumptions cannot be regarded as ``first-order''
+(v.g. the Replacement Scheme).
 
-The version we present of constructibility library,
+In our version of the constructibility library,
 \isatt{ZF-Constructible-Trans}, weakens the
 assumptions of many absoluteness theorems to that
-of a nonempty transitive class; also there are some stronger results
+of a nonempty transitive class; we also include some stronger results
 such as the relativization of powersets.
@@
 %% of type \isatt{i{\isasymRightarrow}o}. 
-For that reason, Separation and
-Replacement predicate \emph{on the satisfaction} of a formula $\phi$.
-With respect to our previous \cite{2019arXiv190103313G}, we lifted the
-arity restriction for the parameter $\phi$ on these schemes,
-streamlining various proofs as an immediate benefit and avoiding the
-need for tupling.
+For that reason, Separation and Replacement predicate \emph{on the
+  satisfaction} of a formula $\phi$.  We improved their specification,
+with respect to our previous \cite{2019arXiv190103313G}, by lifting
+the arity restriction for the parameter $\phi$; consequently we
+get rid of tupling and thus various proofs are slicker.
 
-Once this class versions of the axioms are set up, it is not hard to
+A benefit of having class versions of the axioms is that we can
 apply our synthesis method to obtain their internal, first-order
@@
   relativized predicate.}
-For the case of the Union Axiom, the statement for classes is the
+For the case of the Unordered Pair Axiom, the statement for classes is the
 following

### 2020-01-25 IJCAR2020/conclusion.tex
Por acá pasó Miguel.

@@
 
-There are other debts, as well. In a second
-release of \isatt{ZF-Constructible}, we intend to conform it to the
-lines of \emph{Basic Set Theory (BST)} \cite[I.3.1]{kunen2011set} in
-which elementary results have proofs using alternatively Powerset or
-Replacement. The interest in this arises because many natural set
-models
-(rank-initial segments of the universe or the family $H(\kappa)$ of
-sets hereditarily of cardinality less than $\kappa$) satisfy one of
-those axioms and not the other. There are also still some older or
-less significant proofs written in tactical (\textbf{apply}) format; we
-hope we'll find the time to translate them to Isar. Finally, the
-automation of formula synthesis is on an early stage of
-development;  finishing that module will make writing our proofs of closure
-under various operations faster.
+In a second release of \isatt{ZF-Constructible-Trans}, we intend to conform
+it to the lines of \emph{Basic Set Theory (BST)}
+\cite[I.3.1]{kunen2011set} in which elementary results have proofs
+using alternatively Powerset or Replacement. The interest in this
+arises because many natural set models (rank-initial segments of the
+universe or the family $H(\kappa)$ of sets hereditarily of cardinality
+less than $\kappa$) satisfy one of those axioms and not the
+other. There are also still some older or less significant proofs
+written in tactical (\textbf{apply}) format; we hope we will find the
+time to translate them to Isar. Finally, the automation of formula
+synthesis is on an early stage of development; finishing that module
+will make writing our proofs of closure under various operations
+faster.
+
+It is our intention to contribute back our improvements over
+\isatt{ZF-Constructible} to the official Isabelle distribution: We
+expect that furnishing it with more automatic tools might turn it more
+usable to other researchers; making it a feasible alternative to the
+interesting Flypitch project of Han and van Doorn.
 

### 2020-01-25 IJCAR2020/forces_definition.tex
Por acá pasó Miguel.

@@
 defined a relation $\frecR$ on 4-tuples of elements of $M$, proved
-that it is wellfounded ---indeed, we also proved an
+that it is wellfounded and, more important, we also proved an
 induction principle for this relation:
@@
 
-\paragraph*{Technical details}
-Paulson proved absoluteness of functions defined by wellfounded
-recursion over a transitive relation. Some of our definitions by
-recursion (\emph{check} and \emph{forces}) do not fit in that scheme.
-One can replace the relation $R$ for its transitive closure $R^+$ in the
+Another improvement over our previous report are a couple of new
+technical results concerning recursive definitions: Paulson proved
+absoluteness of functions defined by wellfounded recursion over a
+transitive relation. Some of our definitions by recursion
+(\emph{check} and \emph{forces}) do not fit in that scheme.  One can
+replace the relation $R$ for its transitive closure $R^+$ in the
 recursive definition because one can prove, in general, that

### 2020-01-25 IJCAR2020/intro.tex
Por acá pasó Miguel.

@@
 generic filter $G$ belongs to the extension $M[G]$ and 
-$M\sbq M[G]$. We also include an example of forcing notion for which
+$M\sbq M[G]$. In order to take full advantage of the constructibility library
+we enhanced it by weakening the assumption of many results and also
+extended it with stronger results. % 
+We also include an example of forcing notion for which
 the last inclusion is proper. 
@@
 
+\subsection*{Outline}
+\label{sec:outline}
+
+%% Esto está como subsection para que no parezca como parte de Related
+%% Work (que tranquilamente podría ir a las conclusiones).
+We assume some familiarity with Isabelle and some terminology of set
+theory. The current paper is organized as follows. In
+Sect.~\ref{sec:isabelle-metatheories} we comment briefly on the
+metha-theoretical implications of using Isabelle/ZF. In
+Sect.~\ref{sec:relat-absol} we explain the use of relativized concepts
+and its importance for the ctm approach. The next sections cover the
+core of this report: in Sect.~\ref{sec:definition-forces} we introduce
+the definition of the formula transformer $\forceisa$ and reasoning
+principles about it; in Sect.~\ref{sec:forcing-theorems} we present
+the proofs of the fundamental theorems of forcing. We show in
+Sect.~\ref{sec:example-proper-extension} a concrete poset that leads
+to a proper extension of the ground model. In
+Sect.~\ref{sec:axioms-replacement-choice} we complete the proof that
+every axiom and axiom scheme of ZFC is valid in any generic extension.
+We close the paper by noting the next steps in our project and drawing
+conclusions from this formalization.
+
 %%% Local Variables: 

### 2020-01-25 IJCAR2020/proper_extension.tex
menor: mudo comentarios de Pure, explico más main theorem
Por acá pasó Miguel.

@@
 Even when the axioms of $\ZFC$ are proved in the generic extension,
-one can't claim that the magic of forcing has taken place unless one
+one cannot claim that the magic of forcing has taken place unless one
 is able to provide some \emph{proper} extension with the \emph{same
@@
 
-Two short theory files contain the relevant
+Three theory files contain the relevant
 results. \verb|Ordinals_In_MG.thy| shows, using the closure of $M$

### 2020-01-25 IJCAR2020/replacement_choice.tex
Por acá pasó Miguel.

@@
 
-Apart from the obvious modifications arising from the change in the
-statement of the axiom schemes, the proofs that all axioms apart from
-those in the title hold in $M[G]$ are obtained exactly as in
-\cite{2019arXiv190103313G}.
+In \cite{2019arXiv190103313G} we proved that any generic extension
+preserve the satisfaction of all the axioms and the separtion scheme
+(we adapted those proofs to the current statement of the axiom
+schemes). Our proofs that Replacement and choice hold in every generic
+extension depend on further relativized concepts and closure properties.
 
@@
 belongs to $\pi\defi \dom(\pi')\times \PP$, which is easily seen to be
-in $M$. We'll use $\pi$ to be the domain in an application of the
+in $M$. We will use $\pi$ to be the domain in an application of the
 Replacement Axiom in $M$. But now, obviously, we have lost

### 2020-01-25 IJCAR2020/absoluteness.tex
reediciones (faltan dos observaciones by Miguel)

@@
 
-% If a concept is absolute, then one can reason ``externally'' instead
-% of ``inside'' the model; thus, one has at hand any lemma about ``the''
-% concept and not what the model believes of the concept.
 A benefit of working with transitive models
@@
 of $\ZF$) $M$ and any of its extensions $M[G]$.
+For that reason, then one can reason
+``externally'' about absolute concepts, instead
+of ``inside'' the model; thus, one has at hand any already proved
+lemma about the real concept.
 
@@
 seriously, some assumptions cannot be regarded as ``first-order''
-(v.g. the Replacement Scheme).
-
-In our version of the constructibility library,
+(v.g. the Replacement Scheme) because of the occurrence of predicate
+variables.
+The version we present of the constructibility library,
 \isatt{ZF-Constructible-Trans}, weakens the
@@
 the arity restriction for the parameter $\phi$; consequently we
-get rid of tupling and thus various proofs are slicker.
+get rid of tupling and thus various proofs are now slicker.
 
@@
   relativized predicate.}
-For the case of the Unordered Pair Axiom, the statement for classes is the
+For the case of the Pairing Axiom, the statement for classes is the
 following

### 2020-01-25 IJCAR2020/conclusion.tex
reediciones (faltan dos observaciones by Miguel)

@@
 will make writing our proofs of closure under various operations
-faster.
-
-It is our intention to contribute back our improvements over
-\isatt{ZF-Constructible} to the official Isabelle distribution: We
-expect that furnishing it with more automatic tools might turn it more
-usable to other researchers; making it a feasible alternative to the
-interesting Flypitch project of Han and van Doorn.
+faster, and also turn the set theory libraries  more
+usable to other researchers.
 

### 2020-01-25 IJCAR2020/intro.tex
reediciones (faltan dos observaciones by Miguel)

@@
 generic filter $G$ belongs to the extension $M[G]$ and 
-$M\sbq M[G]$. In order to take full advantage of the constructibility library
+$M\sbq M[G]$. 
+In order to take full advantage of the constructibility library
 we enhanced it by weakening the assumption of many results and also
 extended it with stronger results. % 
-We also include an example of forcing notion for which
-the last inclusion is proper. 
 Finally, our development is now independent of $\AC$: We modularized
@@
 
-\subsection*{Outline}
+\subsection{Outline}
 \label{sec:outline}
@@
 Sect.~\ref{sec:isabelle-metatheories} we comment briefly on the
-metha-theoretical implications of using Isabelle/ZF. In
+meta-theoretical implications of using Isabelle/ZF. In
 Sect.~\ref{sec:relat-absol} we explain the use of relativized concepts
 and its importance for the ctm approach. The next sections cover the
-core of this report: in Sect.~\ref{sec:definition-forces} we introduce
+core of this report: In Sect.~\ref{sec:definition-forces} we introduce
 the definition of the formula transformer $\forceisa$ and reasoning

### 2020-01-25 IJCAR2020/proper_extension.tex
reediciones (faltan dos observaciones by Miguel)

@@
 \end{isabelle}
-
 To prove these results, we found it useful to formalize induction over
@@
 to arguments involving names.
-\begin{isabelle}
-\isacommand{lemma}\isamarkupfalse%
-\ ed{\isacharunderscore}induction{\isacharcolon}\isanewline
-\ \ \isakeyword{assumes}\ {\isachardoublequoteopen}{\isasymAnd}x{\isachardot}\ {\isasymlbrakk}{\isasymAnd}y{\isachardot}\ \ ed{\isacharparenleft}y{\isacharcomma}x{\isacharparenright}\ {\isasymLongrightarrow}\ Q{\isacharparenleft}y{\isacharparenright}\ {\isasymrbrakk}\ {\isasymLongrightarrow}\ Q{\isacharparenleft}x{\isacharparenright}{\isachardoublequoteclose}\isanewline
-\ \ \isakeyword{shows}\ {\isachardoublequoteopen}Q{\isacharparenleft}a{\isacharparenright}{\isachardoublequoteclose}
-\end{isabelle}
 

### 2020-01-25 IJCAR2020/replacement_choice.tex
reediciones (faltan dos observaciones by Miguel)

@@
 In \cite{2019arXiv190103313G} we proved that any generic extension
-preserve the satisfaction of all the axioms and the separtion scheme
+preserves the satisfaction of almost all the axioms, including the separation scheme
 (we adapted those proofs to the current statement of the axiom

### 2020-01-25 IJCAR2020/forces_definition.tex
reediciones (faltan dos observaciones by Miguel)
dos observaciones by Miguel (sólo apliqué una)

@@
 formulas into itself. It is the very reason of applicability of
-forcing that the satisfaction of the formula $\forceisa(\phi)$ in a
-model controls the satisfaction of $\phi$ in all of its generic
-extensions.
+forcing that the satisfaction of a first-order formula  $\phi$ in all of the generic
+extensions of a ctm $M$ can be ``controlled'' in a definable way from
+$M$ (viz., by  satisfaction the of the formula $\forceisa(\phi)$).
 
@@
 \end{multline}
-Note that the definition of $\forcesmem$ states that the set 
+Note that the definition of $\forcesmem$ is equivalent to require 
+ the set 
 \[
@@
 \]
-is dense below $p$.
+to be dense below $p$.
 

### 2020-01-25 IJCAR2020/absoluteness.tex
spelling

@@
 are both
-prerrequisites and powerful tools in working with transitive
+prerequisites and powerful tools in working with transitive
 models. A \emph{class} is simply a predicate $C(x)$ with at least one
@@
 The following schematic lemma synthesizes its internal version,%
-\footnote{The use of such schematic goals and the orginal definition
+\footnote{The use of such schematic goals and the original definition
   of the collection of lemmas \isatt{sep{\isacharunderscore}rules} are due to Paulson.}

### 2020-01-25 IJCAR2020/conclusion.tex
spelling

@@
 notions to obtain the independence of $\CH$ 
-along with the prerrequisite combinatorial results, v.g.\ the
+along with the prerequisite combinatorial results, v.g.\ the
 $\Delta$-system lemma. A development of cofinality is under way in a
@@
 arises because many natural set models (rank-initial segments of the
-universe or the family $H(\kappa)$ of sets hereditarily of cardinality
-less than $\kappa$) satisfy one of those axioms and not the
+universe or the family $H(\kappa)$ of sets  of cardinality
+less than $\kappa$ hereditarily) satisfy one of those axioms and not the
 other. There are also still some older or less significant proofs

### 2020-01-25 IJCAR2020/forces_definition.tex
spelling

@@
 defined a relation $\frecR$ on 4-tuples of elements of $M$, proved
-that it is wellfounded and, more important, we also proved an
+that it is well-founded and, more important, we also proved an
 induction principle for this relation:
@@
 technical results concerning recursive definitions: Paulson proved
-absoluteness of functions defined by wellfounded recursion over a
+absoluteness of functions defined by well-founded recursion over a
 transitive relation. Some of our definitions by recursion

### 2020-01-25 IJCAR2020/forcing_theorems.tex
reediciones (faltan dos observaciones by Miguel)
spelling

@@
 As with the previous section, the proofs of the forcing theorems have two different
-flavours: The ones for the atomic formulas proceed by using the
+flavors: The ones for the atomic formulas proceed by using the
 principle of 
@@
 and Truth, 
-which we reproduce next.
+which we state next.
 %

### 2020-01-25 IJCAR2020/isabelle_metatheory.tex
menor: mudo comentarios de Pure, explico más main theorem
Arreglos muy pavos. Comentario sobre definiciones sintetizadas.
reediciones (faltan dos observaciones by Miguel)
spelling

@@
 assistant based on fragment of higher-order logic called
-\emph{Pure}. %% The logical primitives of this meta-level are
-%% \isatt{\isasymLongrightarrow}, \isatt{\&\&\&}, and \isatt{\isasymAnd}
-%% (implication, conjunction, and universal
-%% quantification, resp.), which operate on the meta-booleans
-%% \isatt{prop}. 
+\emph{Pure}. 
 The results presented in this work are theorems of a
@@
 form a recursively defined set \isatt{formula}. For example, the
-predicate of safisfaction
+predicate of satisfaction
 \isatt{sats::i{\isasymRightarrow}i{\isasymRightarrow}i{\isasymRightarrow}o}
-(noted $M,\mathit{ms}\models\phi$ for a set $M$, for a set $M$,
-$\mathit{ms}\in\isatt{list}(M)$ and $\phi\in\formula$
+(written $M,\mathit{ms}\models\phi$ for a set $M$,
+$\mathit{ms}\in\isatt{list}(M)$ and $\phi\in\formula$)
 %% (noted \isatt{M, ms \isasymTurnstile{ }\isasymphi} for a set \isatt{M},
 %% \isatt{ms {\isasymin} list(M)} and \isatt{{\isasymphi} {\isasymin} formula})
-had already been defined by recursion in \cite{paulson_2003}. We added
-the corresponding notion $M\models \Phi$ for satisfaction of a set of
-sentences.
+had already been defined by recursion in \cite{paulson_2003}.
 

### 2020-01-25 IJCAR2020/replacement_choice.tex
spelling

@@
 
-Although our previous version of the developement used $\AC$, that was
+Although our previous version of the development used $\AC$, that was
 only needed to show the Rasiowa-Sikorski Lemma (RSL) for
@@
 \end{isabelle}
-Here, \isatt{\isasymapprox} stands for equipotence, \isatt{nat} is the
+Here, \isatt{\isasymapprox} stands for equipotency, \isatt{nat} is the
 set of natural numbers, and the predicate 

### 2020-01-27 IJCAR2020/absoluteness.tex
Cambios muy pavos.

@@
 apply our synthesis method to obtain their internal, first-order
-counterparts.\footnote{It should be noted that our method for
-  synthetic definition requires that the schematic goal had a
-  particular format where the only unbound schematic variable occurs
-  as an (open) formula whose satisfaction is equivalent to some
-  relativized predicate.}
+counterparts.% \footnote{It should be noted that our method for
+  % synthetic definition requires that the schematic goal had a
+  % particular format where the only unbound schematic variable occurs
+  % as an (open) formula whose satisfaction is equivalent to some
+  % relativized predicate.}
 For the case of the Pairing Axiom, the statement for classes is the

### 2020-01-27 IJCAR2020/forces_definition.tex
Cambios muy pavos.
Definición pavísima de forces.
Introducción de ||-

@@
 extensions of a ctm $M$ can be ``controlled'' in a definable way from
-$M$ (viz., by  satisfaction the of the formula $\forceisa(\phi)$).
+$M$ (viz., by  satisfaction of the formula $\forceisa(\phi)$).
+
+In fact, given a forcing notion (ie. a poset with a top element) Kunen
+defines meta-theoretically the forcing relation $p \Vdash \phi$ as the
+satisfaction by $M$ of $\forceisa(\phi)$ in an environment containing
+the forcing notion and $p$; next he shows that this relation is
+definable. To make sense of this in our formalization, we started with
+an internalized relation and then proved the equivalence of that
+relation with the semantic version of the forcing relation.  The
+latter relation is captured by the following abbreviation, after
+defining $\forceisa$:
+\begin{isabelle}
+\ \ {\isachardoublequoteopen}p\ {\isasymtturnstile}\ {\isasymphi}\ env\ {\isasymequiv}\ sats{\isacharparenleft}M{\isacharcomma}forces{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharcomma}\ {\isacharbrackleft}p{\isacharcomma}P{\isacharcomma}leq{\isacharcomma}one{\isacharbrackright}\ {\isacharat}\ env{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
 
@@
 
-Once $\forcesat$ and its relativized version $\isatt{is{\isacharunderscore}forces{\isacharunderscore}at}$
-were defined, we proceeded to show absoluteness and provided internal
+Once $\forcesat$ and its relativized version
+$\isatt{is{\isacharunderscore}forces{\isacharunderscore}at}$ were
+defined, we proceeded to show absoluteness and provided internal
 definitions for the recursion on names using results in
 \isatt{ZF-Constructible}. This finished the atomic case of the
-formula-transformer $\forceisa$.
+formula-transformer $\forceisa$. The characterization of $\forceisa$
+for negated and universal quantified formulas is given by the
+following lemmas, respectively:
+%
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ sats{\isacharunderscore}forces{\isacharunderscore}Neg{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\isanewline
+\ \ \ \ {\isachardoublequoteopen}p{\isasymin}P{\isachardoublequoteclose}\ {\isachardoublequoteopen}env\ {\isasymin}\ list{\isacharparenleft}M{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\isanewline
+\ \ \ \ {\isachardoublequoteopen}M{\isacharcomma}\ {\isacharbrackleft}p{\isacharcomma}P{\isacharcomma}leq{\isacharcomma}one{\isacharbrackright}\ {\isacharat}\ env\ {\isasymTurnstile}\ forces{\isacharparenleft}Neg{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharparenright}\ \ \ {\isasymlongleftrightarrow}\ \isanewline
+\ \ \ \ \ {\isasymnot}{\isacharparenleft}{\isasymexists}q{\isasymin}M{\isachardot}\ q{\isasymin}P\ {\isasymand}\ is{\isacharunderscore}leq{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharcomma}leq{\isacharcomma}q{\isacharcomma}p{\isacharparenright}\ {\isasymand}\ \isanewline
+\ \ \ \ \ \ \ \ \ \ M{\isacharcomma}\ {\isacharbrackleft}q{\isacharcomma}P{\isacharcomma}leq{\isacharcomma}one{\isacharbrackright}{\isacharat}env\ {\isasymTurnstile}\ forces{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}\isanewline
 
-Another improvement over our previous report are a couple of new
-technical results concerning recursive definitions: Paulson proved
-absoluteness of functions defined by well-founded recursion over a
-transitive relation. Some of our definitions by recursion
-(\emph{check} and \emph{forces}) do not fit in that scheme.  One can
-replace the relation $R$ for its transitive closure $R^+$ in the
-recursive definition because one can prove, in general, that
+\isacommand{lemma}\isamarkupfalse%
+\ sats{\isacharunderscore}forces{\isacharunderscore}Forall{\isacharcolon}\isanewline
+\ \ \isakeyword{assumes}\isanewline
+\ \ \ \ {\isachardoublequoteopen}p{\isasymin}P{\isachardoublequoteclose}\ {\isachardoublequoteopen}env\ {\isasymin}\ list{\isacharparenleft}M{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\isanewline
+\ \ \ \ {\isachardoublequoteopen}M{\isacharcomma}{\isacharbrackleft}p{\isacharcomma}P{\isacharcomma}leq{\isacharcomma}one{\isacharbrackright}\ {\isacharat}\ env\ {\isasymTurnstile}\ forces{\isacharparenleft}Forall{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharparenright}\ {\isasymlongleftrightarrow}\ \isanewline
+\ \ \ \ \ {\isacharparenleft}{\isasymforall}x{\isasymin}M{\isachardot}\ \ \ M{\isacharcomma}\ {\isacharbrackleft}p{\isacharcomma}P{\isacharcomma}leq{\isacharcomma}one{\isacharcomma}x{\isacharbrackright}\ {\isacharat}\ env\ {\isasymTurnstile}\ forces{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+\end{isabelle}
+
+Let us note in passing another improvement over our previous report:
+we made a couple of new technical results concerning recursive
+definitions. Paulson proved absoluteness of functions defined by
+well-founded recursion over a transitive relation. Some of our
+definitions by recursion (\emph{check} and \emph{forces}) do not fit
+in that scheme.  One can replace the relation $R$ for its transitive
+closure $R^+$ in the recursive definition because one can prove, in
+general, that
 $F\!\upharpoonright\!(R^{-1}(x))(y) =

### 2020-01-27 IJCAR2020/replacement_choice.tex
Sacando overflows

@@
 \isacommand{lemma}\isamarkupfalse%
-\ Vset{\isacharunderscore}abs{\isacharcolon}\ {\isachardoublequoteopen}{\isasymlbrakk}\ M{\isacharparenleft}i{\isacharparenright}{\isacharsemicolon}\ M{\isacharparenleft}V{\isacharparenright}{\isacharsemicolon}\ Ord{\isacharparenleft}i{\isacharparenright}\ {\isasymrbrakk}\ {\isasymLongrightarrow}\ is{\isacharunderscore}Vset{\isacharparenleft}M{\isacharcomma}i{\isacharcomma}V{\isacharparenright}\ {\isasymlongleftrightarrow}\ V\ {\isacharequal}\ {\isacharbraceleft}x{\isasymin}Vset{\isacharparenleft}i{\isacharparenright}{\isachardot}\ M{\isacharparenleft}x{\isacharparenright}{\isacharbraceright}{\isachardoublequoteclose}
+\ Vset{\isacharunderscore}abs{\isacharcolon}\ {\isachardoublequoteopen}{\isasymlbrakk}\ M{\isacharparenleft}i{\isacharparenright}{\isacharsemicolon}\ M{\isacharparenleft}V{\isacharparenright}{\isacharsemicolon}\ Ord{\isacharparenleft}i{\isacharparenright}\ {\isasymrbrakk}\ {\isasymLongrightarrow}\ \isanewline
+\ \ \ \ \ \ \ \  \ \  \ \ \ \ \ \ \ \ \ \ is{\isacharunderscore}Vset{\isacharparenleft}M{\isacharcomma}i{\isacharcomma}V{\isacharparenright}\ {\isasymlongleftrightarrow}\ V\ {\isacharequal}\ {\isacharbraceleft}x{\isasymin}Vset{\isacharparenleft}i{\isacharparenright}{\isachardot}\ M{\isacharparenleft}x{\isacharparenright}{\isacharbraceright}{\isachardoublequoteclose}
 \end{isabelle}

### 2020-03-02 IJCAR2020/intro.tex

@@
 
-\subsection{Related work}
+\paragraph{Related work}
 To the best of our knowledge, all of the previous works in
@@
 
-\subsection{Outline}
+\paragraph{Outline}
 \label{sec:outline}

### 2020-03-02 IJCAR2020/sample_proof.tex

\section{A sample proof}

We would like to illustrate the point of the simplicity of proofs
after the definition of $\forceisa$ is ready to use. We consider the following:

\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ strengthening{\isacharunderscore}lemma{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ \isanewline
\ \ \ \ {\isachardoublequoteopen}p{\isasymin}P{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}r{\isasymin}P{\isachardoublequoteclose}\ {\isachardoublequoteopen}r{\isasympreceq}p{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isasymAnd}env{\isachardot}\ env{\isasymin}list{\isacharparenleft}M{\isacharparenright}\ {\isasymLongrightarrow}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isasymle}length{\isacharparenleft}env{\isacharparenright}\ {\isasymLongrightarrow}\ p\ {\isasymtturnstile}\ {\isasymphi}\ env\isanewline 
\ \ \ \ \ {\isasymLongrightarrow}\ r\ {\isasymtturnstile}\ {\isasymphi}\ env{\isachardoublequoteclose}\isanewline
%
%
\isacommand{using}\isamarkupfalse%
\ assms{\isacharparenleft}{\isadigit{2}}{\isacharparenright}
\end{isabelle}
%
The proof is divided in the 4 cases of definition of an element of $\formula$,
%
\begin{isabelle}
\isacommand{proof}\isamarkupfalse%
\ {\isacharparenleft}induct{\isacharparenright}\isanewline
\ \ \isacommand{case}\isamarkupfalse%
\ {\isacharparenleft}Member\ n\ m{\isacharparenright}\isanewline
\ \ \isacommand{then}\isamarkupfalse%
\isanewline
\ \ \isacommand{have}\ {\isachardoublequoteopen}n<length(env){\isachardoublequoteclose} {\isachardoublequoteopen}m<length(env){\isachardoublequoteclose}\isanewline
\ \ \ \  using arities{\isacharunderscore}at{\isacharunderscore}aux by simp{\isacharunderscore}all\isanewline
\ \ \isacommand{moreover}\isanewline
\ \ \isacommand{assume} {\isachardoublequoteopen}env{\isasymin}list(M){\isachardoublequoteclose}\isanewline
\ \ \isacommand{moreover}\isanewline
\ \ \isacommand{note} assms Member\isanewline
\ \ \isacommand{show}\isamarkupfalse%
\ {\isacharquery}case\ \isanewline
\ \ \ \ \isacommand{using}\isamarkupfalse%
\ Forces{\isacharunderscore}Member{\isacharbrackleft}of\ {\isacharunderscore}\ {\isachardoublequoteopen}nth{\isacharparenleft}n{\isacharcomma}env{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}nth{\isacharparenleft}m{\isacharcomma}env{\isacharparenright}{\isachardoublequoteclose}\ env\ n\ m{\isacharbrackright}\isanewline
\ \ \ \ \ \ strengthening{\isacharunderscore}mem{\isacharbrackleft}of\ p\ r\ {\isachardoublequoteopen}nth{\isacharparenleft}n{\isacharcomma}env{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}nth{\isacharparenleft}m{\isacharcomma}env{\isacharparenright}{\isachardoublequoteclose}{\isacharbrackright}\ \isacommand{by}\isamarkupfalse%
\ simp
\end{isabelle}
%
where the final step depends on the previously
proved lemma
\isatt{strengthening{\isacharunderscore}mem} and the characterization of
$\forceisa$ for membership.

The case of equality is entirely analogous, and the \isatt{Nand} and
\isatt{Forall} cases are handled very simply.
%
\begin{isabelle}
\isacommand{next}\isamarkupfalse%
\isanewline
\ \ \isacommand{case}\isamarkupfalse%
\ {\isacharparenleft}Equal\ n\ m{\isacharparenright}\isanewline
\ \ \dots\isanewline
\isacommand{next}\isamarkupfalse%
\isanewline
\ \ \isacommand{case}\isamarkupfalse%
\ {\isacharparenleft}Nand\ {\isasymphi}\ {\isasympsi}{\isacharparenright}\isanewline
\ \ \isacommand{with}\isamarkupfalse%
\ assms\isanewline
\ \ \isacommand{show}\isamarkupfalse%
\ {\isacharquery}case\ \isanewline
\ \ \ \ \isacommand{using}\isamarkupfalse%
\ Forces{\isacharunderscore}Nand\ Transset{\isacharunderscore}intf{\isacharbrackleft}OF\ trans{\isacharunderscore}M\ {\isacharunderscore}\ P{\isacharunderscore}in{\isacharunderscore}M{\isacharbrackright}\ pair{\isacharunderscore}in{\isacharunderscore}M{\isacharunderscore}iff\isanewline
\ \ \ \ \ \ Transset{\isacharunderscore}intf{\isacharbrackleft}OF\ trans{\isacharunderscore}M\ {\isacharunderscore}\ leq{\isacharunderscore}in{\isacharunderscore}M{\isacharbrackright}\ leq{\isacharunderscore}transD\ \isacommand{by}\isamarkupfalse%
\ auto\isanewline
\isacommand{next}\isamarkupfalse%
\isanewline
\ \ \isacommand{case}\isamarkupfalse%
\ {\isacharparenleft}Forall\ {\isasymphi}{\isacharparenright}\isanewline
\ \ \isacommand{with}\isamarkupfalse%
\ assms\isanewline
\ \ \isacommand{have}\isamarkupfalse%
\ {\isachardoublequoteopen}p\ {\isasymtturnstile}\ {\isasymphi}\ {\isacharparenleft}{\isacharbrackleft}x{\isacharbrackright}\ {\isacharat}\ env{\isacharparenright}{\isachardoublequoteclose}\ \isakeyword{if}\ {\isachardoublequoteopen}x{\isasymin}M{\isachardoublequoteclose}\ \isakeyword{for}\ x\isanewline
\ \ \ \ \isacommand{using}\isamarkupfalse%
\ that\ Forces{\isacharunderscore}Forall\ \isacommand{by}\isamarkupfalse%
\ simp\isanewline
\ \ \isacommand{with}\isamarkupfalse%
\ \underline{Forall}\ \isanewline
\ \ \isacommand{have}\isamarkupfalse%
\ {\isachardoublequoteopen}r\ {\isasymtturnstile}\ {\isasymphi}\ {\isacharparenleft}{\isacharbrackleft}x{\isacharbrackright}\ {\isacharat}\ env{\isacharparenright}{\isachardoublequoteclose}\ \isakeyword{if}\ {\isachardoublequoteopen}x{\isasymin}M{\isachardoublequoteclose}\ \isakeyword{for}\ x\isanewline
\ \ \ \ \isacommand{using}\isamarkupfalse%
\ that\ pred{\isacharunderscore}le{\isadigit{2}}\ \isacommand{by}\isamarkupfalse%
\ {\isacharparenleft}simp{\isacharparenright}\isanewline
\ \ \isacommand{with}\isamarkupfalse%
\ assms\ \underline{Forall}\isanewline
\ \ \isacommand{show}\isamarkupfalse%
\ {\isacharquery}case\ \isanewline
\ \ \ \ \isacommand{using}\isamarkupfalse%
\ Forces{\isacharunderscore}Forall\ \isacommand{by}\isamarkupfalse%
\ simp\isanewline
\isacommand{qed}\isamarkupfalse
\end{isabelle}
%
It can be noted that the inductive hypothesis
gets used in the last case (underlined here as
\isatt{\underline{Forall}}), but not in the case for \isatt{Nand}.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "forcing_in_isabelle_zf"
%%% ispell-local-dictionary: "american"
%%% End:

### 2020-03-02 src/fix_links.sh
usando ahora versiones parametricas en ../scripts

@@ -1,9 +1,0 @@
-#!/bin/bash
-old_dir=$PWD
-html_dir="output/html/ZF/Forcing/"
-
-cd $html_dir
-sed -i -- 's/a href="..\/ZF\/ZF.html/a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF\/ZF.html/g' *.html
-sed -i -- 's/a href="..\/ZF\/ZF\//a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF\//g' *.html
-#sed -i -- 's/a href="..\/ZF\/ZF-Constructible\//a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF-Constructible\//g' *.html
-cd $old_dir

### 2020-03-02 src/link-forcing.sh

@@ -1,90 +1,0 @@
-#!/bin/bash
-old_dir=$PWD
-html_dir="output/html/ZF/Forcing/"
-list1=Forces_Definition.html
-list2="Forcing_Theorems.html Arities.html"
-list3="Names.html Interface.html"
-list4="Forcing_Main.html FrecR.html Nat_Miscellanea.html\
- Internal_ZFC_Axioms.html"
-list5="Forcing_Notions.html Relative_Univ.html Forcing_Data.html\
- Renaming.html Internalizations.html"
-list6="Replacement_Axiom.html Succession_Poset.html Recursion_Thms.html\
- Separation_Rename.html Choice_Axiom.html Least.html Pointed_DC.html\
- Renaming_Auto.html Union_Axiom.html"
-list7="Ordinals_In_MG.html Powerset_Axiom.html Separation_Axiom.html\
- Foundation_Axiom.html Pairing_Axiom.html\
- Proper_Extension.html Rasiowa_Sikorski.html Extensionality_Axiom.html\
- Infinity_Axiom.html Synthetic_Definition.html"
-
-function link_item {
-    #
-    # Usage:
-    # 
-    #   link_item HTML ITEMLIST SUFFIX
-    for z in `cat $2`
-    do 
-	l=`echo $z | cut -d"." -f2`
-	lprime=`echo $l | sed -e "s/&/\\\\\\&/g"`
-	t=`echo $z | cut -d"." -f1`
-	sed -i -e "s/>$l$3</><a class=\"pst-lnk\" href=\"$t.html#$t.$lprime\">$lprime$3<\/a></g" $1
-    done
-}
-
-function partial_job {
-    #
-    # Usage:
-    #
-    #   partial_job HTMLS ITEMLIST SUFFIX
-    for x in $1
-    do 
-	link_item $x $2 $3
-    done
-}
-
-function full_job {
-    #
-    # Usage:
-    #
-    #   full_job ITEMLIST SUFFIX
-    partial_job "$list1" $1 $2 &
-    partial_job "$list2" $1 $2 &
-    partial_job "$list3" $1 $2 &
-    partial_job "$list4" $1 $2 &
-    partial_job "$list5" $1 $2 &
-    partial_job "$list6" $1 $2 &
-    partial_job "$list7" $1 $2
-}
-
-# echo Back-up of html directory
-# cp -R $html_dir ~/tmp/
-
-echo Changing to html directory
[…]

### 2020-03-03 scripts/fix_links.sh
reemplazo para tener links en el micrositio

@@ -6,4 +6,5 @@ cd $html_dir
 sed -i -- 's/a href="..\/ZF\/ZF.html/a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF\/ZF.html/g' *.html
 sed -i -- 's/a href="..\/ZF\/ZF\//a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF\//g' *.html
 sed -i -- 's/a href="..\/ZF\/ZF-Constructible\//a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF-Constructible\//g' *.html
+sed -i -- 's|a href="../ZF-Constructible-Trans/|a href="https://cs.famaf.unc.edu.ar/~pedro/forcing/ZF-Constructible-Trans/|g' *.html
 cd $old_dir

### 2020-03-03 scripts/linking.sh
parto un script de linkeo en dos, actualizo Makefiles

@@ -1,0 +1,66 @@
+#!/bin/bash
+session=$1 # link to this ---
+src_dir=$2 # where are the lists of items of $session
+href=$3    # where to link
+
+function link_item {
+    #
+    # Usage:
+    # 
+    #   link_item HTML ITEMLIST SUFFIX
+    for z in `cat $2`
+    do 
+	l=`echo $z | cut -d"." -f2`
+	lprime=`echo $l | sed -e "s/&/\\\\\\&/g"`
+	t=`echo $z | cut -d"." -f1`
+	sed -i -e "s|>$l$3<|><a class=\"pst-lnk\" href=\"$href/$t.html#$t.$lprime\">$lprime$3</a><|g" $1
+    done
+}
+
+function partial_job {
+    #
+    # Usage:
+    #
+    #   partial_job HTMLS ITEMLIST SUFFIX
+    for x in $1
+    do 
+	link_item $x $2 $3
+    done
+}
+
+function full_job {
+    #
+    # Usage:
+    #
+    #   full_job ITEMLIST SUFFIX LISTS
+    for (( x=3; $#-$x ; x=$x+1))
+    do 
+	partial_job "${!x}" $1 $2 &
+    done
+    partial_job "${!#}" $1 $2
+}
+
+echo ${@:4} | grep "|"  -q && \
+{ 
+    echo Error: a filename contains a pipe \("|"\).
+    echo Aborted
+    exit 1
+}
+cat $src_dir/lemas_$session.txt $src_dir/lemaslemmas_$session.txt\
+   $src_dir/definiciones_$session.txt | grep "|"  -q && \
+{ 
+    echo Error: an item contains a pipe \("|"\).
+    echo Aborted
+    exit 1
+}
+
+echo -n Crosslinking "lemma" items...
+full_job $src_dir/lemas_$session.txt "" ${@:4}
+echo  Done
+echo -n Crosslinking "lemmas" items...
+full_job $src_dir/lemaslemmas_$session.txt "" ${@:4}
+echo  Done
+echo -n Crosslinking "definition" items...
+full_job $src_dir/definiciones_$session.txt _def ${@:4}
+echo  Done
+echo Finished

### 2020-03-09 IJCAR2020/rebuttal.txt
primera version de rebuttal

We would like to thank the referees for their
careful reading and detailed criticism.

> ----------------------- REVIEW 1 ---------------------
> 
> The paper's main limitation is its treatment of related work. On
> general principles, the main discussion of related work should be near
> the end of the paper, where it could be compared with the work just
> presented. There, it would be appropriate to contrast what was done in
> Paulson's work on constructibility with the current work on
> forcing. More to the point, the recent work by Han and van Doorn on
> forcing in Lean deserves more discussion.

We will gladly incorporate this changes into a revised version.

> ----------------------- REVIEW 2 ---------------------
> 
> In a single
> publication, Han and van Doorn using forcing to show the independence
> of the continuum hypothesis, whereas the authors of this paper are on
> their third publication, and still seem to be at least one paper away
> from showing any independence result.

We believe that the previous assertions are not entirely accurate. As
it is clear from our introduction, our second manuscript on the
subject is not published (and it won't be) in any venue other than the
arXiv. (As a side note, Han & van Doorn have two papers detailing
the full account of their formalization.)

> The main result in this paper is
> simply that forcing extensions satisfy replacement and choice, which
> seem to be fairly straightforward.

In some sense, this is a misstatement of our achievements. The "Main
theorem" of our paper is just the icing on the cake, since there is no
other way to prove this result other than formalizing the machinery of
forcing, which is the real achievement of this paper. Han & van Doorn
[CPP 2020] hint that the formalization of forcing in set theoretical
foundations is practically impossible.

> 
> 
> ----------------------- REVIEW 3 ---------------------
> This work presents a formalization of forcing in Isabelle/ZF and the 
> proof of a first lemma about the forcing relation.

We would like to point out that we proved the fundamental theorems of
forcing, which are the main tools to work with the forcing relation.

### 2020-03-09 IJCAR2020/rebuttal-miguel.txt
La versión de Miguel.

We would like to thank the referees for their
careful reading and detailed criticism.

> ----------------------- REVIEW 1 ---------------------
>
> The paper's main limitation is its treatment of related work. On
> general principles, the main discussion of related work should be near
> the end of the paper, where it could be compared with the work just
> presented. There, it would be appropriate to contrast what was done in
> Paulson's work on constructibility with the current work on
> forcing. More to the point, the recent work by Han and van Doorn on
> forcing in Lean deserves more discussion.

We will gladly incorporate this changes into a revised version.

> ----------------------- REVIEW 2 ---------------------
>
> In a single publication, Han and van Doorn using forcing to show the
> independence of the continuum hypothesis, whereas the authors of
> this paper are on their third publication, and still seem to be at
> least one paper away from showing any independence result.

We knew of Han and van Doorn when we were somewhat advanced in our
project and decided to continue because of the differences: a
different meta-theory and a different approach to forcing. We think
that this paper is interesting per se (we can agree that the arXiv
paper is a bit premature) as it establish a ground on which do forcing
arguments.

> The main result in this paper is simply that forcing extensions
> satisfy replacement and choice, which seem to be fairly
> straightforward.
> ----------------------- REVIEW 3 ---------------------
> This work presents a formalization of forcing in Isabelle/ZF and the
> proof of a first lemma about the forcing relation.

We would like to point that the main result may seem straightforward
thanks to the more substantial contributin of the paper: the
definition of the forces function and the proof of the forcing
theorems. Han & van Doorn [CPP 2020] hint that the formalization of
forcing using the ctm and in a set theoretical foundations is more
intricate than via the Boolean-valued models on type-theory.

### 2020-03-09 IJCAR2020/rebuttal-miguel.txt
La versión de Miguel.

@@
 that this paper is interesting per se (we can agree that the arXiv
-paper is a bit premature) as it establish a ground on which do forcing
-arguments.
+paper is a bit premature) as it establish a ground on which to do
+forcing arguments.
 

### 2020-03-12 IJCAR2020/rebuttal.txt
version final de rebuttal

@@
 
-We will gladly incorporate this changes into a revised version.
+We will gladly incorporate these changes and discussion on the
+questions posed into a revised version.
 
@@
 > their third publication, and still seem to be at least one paper away
-> from showing any independence result.
+> from showing any independence result. In particular, CH will require
+> some careful cardinal combinatorics.
 
-We believe that the previous assertions are not entirely accurate. As
-it is clear from our introduction, our second manuscript on the
-subject is not published (and it won't be) in any venue other than the
-arXiv. (As a side note, Han & van Doorn have two papers detailing
-the full account of their formalization.)
+We knew of Han and van Doorn when we were somewhat advanced in our
+project and decided to continue because of the differences:
+
+ 1) a weaker meta-theory (Han & van Doorn [CPP 2020] hint that the
+ formalization of forcing in set theoretical foundations is practically
+ impossible); and
+
+ 2) an approach to forcing (ctm) used in a important fraction of the
+ literature.
+
+We agree that the remaining work towards the independence of CH will
+require one more paper. Notice that, in fact, Han & van Doorn have two
+papers (ITP2019, CPP2020) detailing the full account of their
+formalization. (As a side note, our second manuscript on the subject
+is not published.)
 
@@
 
-In some sense, this is a misstatement of our achievements. The "Main
-theorem" of our paper is just the icing on the cake, since there is no
-other way to prove this result other than formalizing the machinery of
-forcing, which is the real achievement of this paper. Han & van Doorn
-[CPP 2020] hint that the formalization of forcing in set theoretical
-foundations is practically impossible.
+This is a misstatement of our achievements. Our "Main Theorem" may
+seem straightforward thanks to the more substantial contribution of
+the paper: the definition of the forcing relation and the proof of the
+forcing theorems. The Main Theorem is a succinct way to demonstrate
+the forcing machinery, since there is no other way to produce
+extensions of ctms.
 
-> 
-> 
 > ----------------------- REVIEW 3 ---------------------

### 2020-03-13 IJCAR2020/rebuttal-miguel.txt

@@
 We would like to point that the main result may seem straightforward
-thanks to the more substantial contributin of the paper: the
+thanks to the more substantial contribution of the paper: the
 definition of the forces function and the proof of the forcing

### 2020-03-13 IJCAR2020/absoluteness.tex

@@
 
-Apart from the axiom schemes, the $\ZFC$ 
-axioms are initially stated as predicates on classes (that is,
-of type \isatt{(i{\isasymRightarrow}o){\isasymRightarrow}o}); some
-of them were already defined by Paulson, and this formulation allows a
-better interaction with \isatt{ZF-Constructible}. 
-The axioms of Pairing, Union, Foundation,
-Extensionality, and Infinity are relativizations of the respective
-traditional 
-first-order sentences to the class argument. For the Axiom of Choice
-we selected a version best suited for the work with transitive
-models: the relativization of a sentence stating that for every $x$
-there is surjection from an ordinal onto $x$. Finally, Separation and
-Replacement were treated separately to effectively obtain first-order
-versions afterwards. It is to be noted that predicates in Isabelle/ZF
-are akin to second order variables and thus do not correspond to first-order
-formulas.
+Apart from the axiom schemes, the $\ZFC$ axioms are initially stated
+as predicates on classes (that is, of type
+\isatt{(i{\isasymRightarrow}o){\isasymRightarrow}o}); this formulation
+allows a better interaction with \isatt{ZF-Constructible}.  The axioms
+of Pairing, Union, Foundation, Extensionality, and Infinity are
+relativizations of the respective traditional first-order sentences to
+the class argument. For the Axiom of Choice we selected a version best
+suited for the work with transitive models: the relativization of a
+sentence stating that for every $x$ there is surjection from an
+ordinal onto $x$. Finally, Separation and Replacement were treated
+separately to effectively obtain first-order versions afterwards. It
+is to be noted that predicates in Isabelle/ZF are akin to second order
+variables and thus do not correspond to first-order formulas.
 %% in particular, there is no induction principle for functions

### 2020-03-13 IJCAR2020/forces_definition.tex

@@
 
-In fact, given a forcing notion (ie. a poset with a top element) Kunen
-defines meta-theoretically the forcing relation $p \Vdash \phi$ as the
-satisfaction by $M$ of $\forceisa(\phi)$ in an environment containing
-the forcing notion and $p$; next he shows that this relation is
-definable. To make sense of this in our formalization, we started with
-an internalized relation and then proved the equivalence of that
-relation with the semantic version of the forcing relation.  The
-latter relation is captured by the following abbreviation, after
-defining $\forceisa$:
+In fact, given a forcing notion $\PP$ (i.e. a preorder with a top element)
+in a ctm $M$,
+Kunen defines the \emph{forcing relation} model-theoretically 
+by considering all extensions $M[G]$ with $G$ generic for $\PP$.
+Then two fundamental results are proved, the Truth Lemma and the
+Definability Lemma; but the proof of the first one is based on the
+formula that witnesses Definability. To make sense of this in our 
+formalization, we started with the internalized relation and then
+proved that it is equivalent to the semantic version 
+(``\isatt{definition{\isacharunderscore}of{\isacharunderscore}forces},'' in
+the next section).
+For that reason, the usual notation of the forcing relation 
+$p \Vdash \phi\ \mathit{env}$ (for $\mathit{env}$ a list of
+``names''), abbreviates in our code the
+satisfaction by $M$ of $\forceisa(\phi)$:
 \begin{isabelle}
-\ \ {\isachardoublequoteopen}p\ {\isasymtturnstile}\ {\isasymphi}\ env\ {\isasymequiv}\ sats{\isacharparenleft}M{\isacharcomma}forces{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharcomma}\ {\isacharbrackleft}p{\isacharcomma}P{\isacharcomma}leq{\isacharcomma}one{\isacharbrackright}\ {\isacharat}\ env{\isacharparenright}{\isachardoublequoteclose}
+\ \ {\isachardoublequoteopen}p\ {\isasymtturnstile}\ {\isasymphi}\ env\ \ \ {\isasymequiv}\ \ \ M{\isacharcomma}\ {\isacharparenleft}{\isacharbrackleft}p{\isacharcomma}P{\isacharcomma}leq{\isacharcomma}one{\isacharbrackright}\ {\isacharat}\ env{\isacharparenright}
+    {\isasymTurnstile}\ forces{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isachardoublequoteclose}
 \end{isabelle}

### 2020-03-24 IJCAR2020/related.tex
nueva sección

--- 2020-03-02 IJCAR2020/intro.tex
@@
-\section{Introduction}
-\label{sec:introduction}
+\section{Related work}
+\label{sec:related-work}
 
-The present work reports on the third stage of our project of
-formalizing the theory of forcing and its applications as presented in
-one of the more
-important references on the subject, Kunen's Set Theory
-\cite{kunen2011set} (a rewrite of the classical \cite{kunen1980}).
-
-We
-work using the  implementation of Zermelo-Fraenkel ($\ZF$)
-set theory \emph{Isabelle/ZF} by Paulson and Grabczewski \cite{DBLP:journals/jar/PaulsonG96}. In
-an early paper \cite{2018arXiv180705174G}, we set up the first
-elements of the countable transitive model (ctm) approach, defining
-forcing notions, names, generic extensions, and
-showing the existence of generic filters via the Rasiowa-Sikorski
-lemma (RSL). Our second (unpublished) technical report
-\cite{2019arXiv190103313G} advanced by presenting the first accurate
-\emph{formal abstract} of the Fundamental Theorems of Forcing, and
-using them to show that that the $\ZF$ axioms apart from Replacement
-and Infinity hold in all generic extensions.
-
-This paper contains the proof of Fundamental Theorems and complete
-proofs of the Axioms of Infinity, Replacement, and Choice in all
-generic extensions. In particular, we were able to fulfill the
-promised formal abstract for the Forcing 
-Theorems almost to the letter. A requirement for Infinity and the
-absoluteness of forcing for atomic formulas, we finished the interface
-between our development and
-Paulson's constructibility library \cite{paulson_2003} which enables
-us to do well-founded
-recursion inside transitive models of an appropriate finite fragment
-of $\ZF$. As a by-product, we honored a long debt: the fact that the
-generic filter $G$ belongs to the extension $M[G]$ and 
-$M\sbq M[G]$. 
-In order to take full advantage of the constructibility library
-we enhanced it by weakening the assumption of many results and also
-extended it with stronger results. % 
-Finally, our development is now independent of $\AC$: We modularized
-RSL in such a way that a version for countable 
-posets does not require choice.
-
-In the course of our work we found it useful to develop Isar methods to
-automate repetitive tasks. Part of the interface with Paulson's
-library consisted in constructing formulas for each relativized
-concept; and actually Isabelle's Simplifier can
-synthesize terms for unbound schematic variables in theorems. The
-synthesized term, however, is not available outside the theorem; we
-introduced a method that creates a definition from a schematic
-goal. The second method is concerned with renaming of formulas: we
-improved our small library of bounded renamings with a method that
-given the source and target environments figures out the renaming
-function and produces the relevant lemmas about it.
-
-The source code of our formalization, written for the 2019 version of
-Isabelle, can be browsed and downloaded at
-\url{https://cs.famaf.unc.edu.ar/~pedro/forcing/}
-
-\paragraph{Related work}
 To the best of our knowledge, all of the previous works in
@@
 
-\paragraph{Outline}
-\label{sec:outline}
-
-%% Esto está como subsection para que no parezca como parte de Related
-%% Work (que tranquilamente podría ir a las conclusiones).
-We assume some familiarity with Isabelle and some terminology of set
-theory. The current paper is organized as follows. In
-Sect.~\ref{sec:isabelle-metatheories} we comment briefly on the
-meta-theoretical implications of using Isabelle/ZF. In
-Sect.~\ref{sec:relat-absol} we explain the use of relativized concepts
-and its importance for the ctm approach. The next sections cover the
-core of this report: In Sect.~\ref{sec:definition-forces} we introduce
-the definition of the formula transformer $\forceisa$ and reasoning
-principles about it; in Sect.~\ref{sec:forcing-theorems} we present
-the proofs of the fundamental theorems of forcing. We show in
-Sect.~\ref{sec:example-proper-extension} a concrete poset that leads
-to a proper extension of the ground model. In
-Sect.~\ref{sec:axioms-replacement-choice} we complete the proof that
-every axiom and axiom scheme of ZFC is valid in any generic extension.
-We close the paper by noting the next steps in our project and drawing
-conclusions from this formalization.
-
 %%% Local Variables: 

### 2020-03-24 IJCAR2020/intro.tex
nueva sección, 2

@@
 
-\paragraph{Related work}
-To the best of our knowledge, all of the previous works in
-formalization of the method 
-of forcing have been done in different variants of type theory, and
-none of them uses the ctm approach. The
-most important is the recent one by 
-Han and van Doorn
-\cite{han_et_al:LIPIcs:2019:11074,DBLP:conf/cpp/HanD20}, which includes
-a formalization of the independence of $\CH$ by means
-the Boolean-valued approach to forcing, using the Lean
-proof assistant \cite{DBLP:conf/cade/MouraKADR15}.
-
-\paragraph{Outline}
-\label{sec:outline}
-
-%% Esto está como subsection para que no parezca como parte de Related
-%% Work (que tranquilamente podría ir a las conclusiones).
+%% \paragraph{Outline}
+%% \label{sec:outline}
+%% %% Esto está como subsection para que no parezca como parte de Related
+%% %% Work (que tranquilamente podría ir a las conclusiones).
 We assume some familiarity with Isabelle and some terminology of set
@@
 Sect.~\ref{sec:axioms-replacement-choice} we complete the proof that
-every axiom and axiom scheme of ZFC is valid in any generic extension.
-We close the paper by noting the next steps in our project and drawing
-conclusions from this formalization.
+every axiom and axiom scheme of ZFC is valid in any generic
+extension. Sect.~\ref{sec:related-work} briefly discusses related
+works and we close the paper by noting the next steps in our project
+and drawing conclusions from this formalization.
 

### 2020-04-03 IJCAR2020/related.tex

@@
 \label{sec:related-work}
+
+\textbf{Reviewer's comments}
+{\it
+  \begin{itemize}
+  \item There, it would be appropriate to contrast what was done in
+    Paulson's work on constructibility with the current work on forcing.
+  \item More to the point, the recent work by Han and van Doorn on
+    forcing in Lean deserves more discussion.  They have gone further
+    than the current authors, having proved the independence of the
+    continuum hypothesis.  They prefer Boolean-valued models as being
+    more direct in use than the authors' countable transitive models.
+    \begin{itemize}
+    \item Readers will want to know whether the type-theoretic approach
+      is better/worse/just different than using Isabelle/ZF, and
+    \item are there any benefits to the ctm approach?
+    \item Is the type-theory encoding of ZF really accurate?
+    \item How about comparing proofs of equivalent statements in the two
+      approaches for length and readability?
+    \end{itemize}
+  \end{itemize}
+}
 
@@
 
+\begin{itemize}
+\item The consistency strength of Lean seems to require infinitely
+  many inaccessibles. More precisely, Carneiro proved the  consistency
+  of the CiC foundations used by Lean from $\ZFC$ plus the assumption
+  of the existence of countably infinitely many inaccessible cardinals
+  (as claimed by Hales \cite{hales-math-defs-formally} at the Big
+  Proof 2019 workshop).%
+  \footnote{We do not have a direct reference to Carneiro's proof.}
+
+  On the other hand, Isabelle's \emph{Pure} is based on
+  ``intuitionistic higher order logic.'' In Paulson
+  \cite{Paulson1989} it is proved that \emph{Pure} is sound for
+  intuitionistic first order logic, thus it does not add any strength
+  to it. On top of this, the axiomatization of Isabelle/ZF results in
+  a system equiconsistent with $\ZFC$. Our running assumption, that of
+  the existence of a countable transitive model, is considerably
+  weaker (directly and consistency-wise) than the existence of a
+  single inaccesible cardinal. In that sense, directly obtain
+  unprovability results in first order logic, the meta theoretic
+  machinery used to obtain them is far heavier than the one we use to
+  operate model-theoretically.
+  %
+\item We may discuss in finer detail the shape of the axioms of
+  Isabelle/ZF. It is perhaps more correct to say it is an
+  notational variant of NBG set theory, because the schemes of
+  Replacement and Separation feature higher order (free) variables
+  playing the role of formula variables. It can't be proved that the
+  axioms thus written correspond to first order sentences. This is the
+  reason that our relativized versions only apply to set models, where
+  we can restrict the formula variables to predicates that actually
+  come from first order variables. In that sense, the axioms of the
+  locale \isatt{M{\isacharunderscore}ZF} correspond faithfully to the
+  $ZF$ axioms.
+\item \fbox{\bf take care of repetitions} In our opinion, one of the
+  main benefits of using transitive models is that many fundamental
+  notions are absolute and thus the many statements can be interpreted
+  transparently. It also provides a very concrete way to understand
+  generic objects: as sets that (in the non trivial case) are provably
+  not in the original model; this dispells any mystical feel around
+  this concept (contrary to the case when the ground model is the
+  universe of all sets). In addition, two-valued semantics is
+  closer to our intuition ($\leftarrow$ revise).
+\end{itemize}
 %%% Local Variables: 

### 2020-04-12 IJCAR2020/proper_extension.tex
mas precisión sobre consistencia

@@
 presence of large cardinals, a model $M'\supsetneq M$ might be an
-end-extension of $M$ ---this is were we ask the two models to have the
+end-extension of $M$ ---this is where we ask the two models to have the
 same ordinals, the same \emph{height}.

### 2020-04-12 IJCAR2020/related.tex
mas precisión sobre consistencia

@@
 \begin{itemize}
-\item The consistency strength of Lean seems to require infinitely
-  many inaccessibles. More precisely, Carneiro proved the  consistency
-  of the CiC foundations used by Lean from $\ZFC$ plus the assumption
-  of the existence of countably infinitely many inaccessible cardinals
-  (as claimed by Hales \cite{hales-math-defs-formally} at the Big
-  Proof 2019 workshop).%
-  \footnote{We do not have a direct reference to Carneiro's proof.}
+\item The consistency strength of Lean requires infinitely
+  many inaccessibles. More precisely, let Lean$_n$ be the theory of
+  CiC foundations of Lean restricted to $n$ type universes.  Carneiro
+  \cite{carneiro-ms-thesis}, proved the consistency of Lean$_n$ from
+  $\ZFC$ plus the existence of $n$ inaccessible
+  cardinals. It is also reported in \cite{carneiro-ms-thesis} that
+  Werner's results in \cite{10.5555/645869.668660} can be adapted to
+  show that Lean$_{n+2}$ proves the consistency of the latter theory.
 

### 2020-04-15 IJCAR2020/absoluteness.tex
refs y typo

@@
 the relativization machinery by adjusting the hypothesis of a greater
-part of it early development. Paulson's architecture had only in mind
+part of its earlier development. Paulson's architecture had only in mind
 the consistency of $\ZFC$, but, for instance, in order to apply it in

### 2020-04-15 IJCAR2020/absoluteness.tex
Various fixes suggested by the first referee and some other fixes.

@@
 
-Paulson's formalization \cite{paulson_2003} of Gödel's \cite{godel-L}
-already contains absoluteness results which were crucial to our
-project; we realized however that they could be refactorized to be
-more useful. The main objective is to maximize applicability of
-the relativization machinery by adjusting the hypothesis of a greater
-part of its earlier development. Paulson's architecture had only in mind
-the consistency of $\ZFC$, but, for instance, in order to apply it in
-the development of forcing, too much is assumed at the beginning; more
-seriously, some assumptions cannot be regarded as ``first-order''
-(v.g. the Replacement Scheme) because of the occurrence of predicate
-variables.
+Paulson's formalization \cite{paulson_2003} of the relative
+consistency of AC by Gödel \cite{godel-L} already contains
+absoluteness results which were crucial to our project; we realized
+however that they could be refactored to be more useful. The main
+objective is to maximize applicability of the relativization machinery
+by adjusting the hypothesis of a greater part of its earlier
+development. Paulson's architecture had only in mind the consistency
+of $\ZFC$, but, for instance, in order to apply it in the development
+of forcing, too much is assumed at the beginning; more seriously, some
+assumptions cannot be regarded as ``first-order'' (v.g. the
+Replacement Scheme) because of the occurrence of predicate variables.
 The version we present of the constructibility library,
-\isatt{ZF-Constructible-Trans}, weakens the
-assumptions of many absoluteness theorems to that
-of a nonempty transitive class; we also include some stronger results
-such as the relativization of powersets.
+\isatt{ZF-Constructible-Trans}, weakens the assumptions of many
+absoluteness theorems to that of a nonempty transitive class; we also
+include some stronger results such as the relativization of powersets.
 
@@
   satisfaction} of a formula $\phi$.  We improved their specification,
-with respect to our previous \cite{2019arXiv190103313G}, by lifting
+with respect to our previous report~\cite{2019arXiv190103313G}, by lifting
 the arity restriction for the parameter $\phi$; consequently we
@@
 apply our synthesis method to obtain their internal, first-order
-counterparts.% \footnote{It should be noted that our method for
+counterparts. % \footnote{It should be noted that our method for
   % synthetic definition requires that the schematic goal had a

### 2020-04-15 IJCAR2020/conclusion.tex
Various fixes suggested by the first referee and some other fixes.

@@
 
-In a second release of \isatt{ZF-Constructible-Trans}, we intend to conform
-it to the lines of \emph{Basic Set Theory (BST)}
-\cite[I.3.1]{kunen2011set} in which elementary results have proofs
-using alternatively Powerset or Replacement. The interest in this
-arises because many natural set models (rank-initial segments of the
-universe or the family $H(\kappa)$ of sets  of cardinality
-less than $\kappa$ hereditarily) satisfy one of those axioms and not the
+In a second release of \isatt{ZF-Constructible-Trans}, we intend to
+conform it to the lines of \emph{Basic Set Theory (BST)} proposed by
+Kunen~\cite[I.3.1]{kunen2011set} in which elementary results have
+proofs using alternatively Powerset or Replacement. The interest in
+this arises because many natural set models (rank-initial segments of
+the universe or the family $H(\kappa)$ of sets of cardinality less
+than $\kappa$ hereditarily) satisfy one of those axioms and not the
 other. There are also still some older or less significant proofs
@@
 will make writing our proofs of closure under various operations
-faster, and also turn the set theory libraries  more
-usable to other researchers.
+faster, and also turn the set theory libraries more usable to other
+researchers.
 

### 2020-04-15 IJCAR2020/forces_definition.tex
Various fixes suggested by the first referee and some other fixes.

@@
 The core of the development is showing the definability of the
-relation of forcing. As explained in
-\cite[Sect.~8]{2019arXiv190103313G}, this comprises the 
+relation of forcing. As we explained in our previous
+report~\cite[Sect.~8]{2019arXiv190103313G}, this comprises the
 definition of a function $\forceisa$ that maps the set of internal
 formulas into itself. It is the very reason of applicability of
-forcing that the satisfaction of a first-order formula  $\phi$ in all of the generic
-extensions of a ctm $M$ can be ``controlled'' in a definable way from
-$M$ (viz., by  satisfaction of the formula $\forceisa(\phi)$).
+forcing that the satisfaction of a first-order formula $\phi$ in all
+of the generic extensions of a ctm $M$ can be ``controlled'' in a
+definable way from $M$ (viz., by satisfaction of the formula
+$\forceisa(\phi)$).
 
@@
 
-Our aim was to follow Kunen~\cite[p.~257]{kunen2011set}, where the
-following mutual recursion is given:
+Our aim was to follow the definition proposed by Kunen
+in~\cite[p.~257]{kunen2011set}, where the following mutual recursion
+is given:
 \begin{multline}\label{eq:def-forcing-equality}
@@
 Other approaches, like the one in Neeman~\cite{neeman-course} (and
-Kunen's older \cite{kunen1980}), prefer
+Kunen's older book \cite{kunen1980}), prefer
 to have a single, more complicated, definition by simple recursion for

### 2020-04-15 IJCAR2020/forcing_theorems.tex
mas precisión sobre consistencia
Various fixes suggested by the first referee and some other fixes.

@@
   \ \ \isakeyword{shows}\ \isanewline
-  \ \ \ \ {\isachardoublequoteopen}{\isacharparenleft}{\isasymtau}{\isasymin}M{\isasymlongrightarrow}{\isasymtheta}{\isasymin}M{\isasymlongrightarrow}{\isacharparenleft}{\isasymforall}p{\isasymin}G{\isachardot}forces{\isacharunderscore}eq{\isacharparenleft}p{\isacharcomma}{\isasymtau}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isasymlongrightarrow}val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}{\isacharequal}val{\isacharparenleft}G{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isasymand}\isanewline
-  \ \ \ \ \ {\isacharparenleft}{\isasymtau}{\isasymin}M{\isasymlongrightarrow}{\isasymtheta}{\isasymin}M{\isasymlongrightarrow}{\isacharparenleft}{\isasymforall}p{\isasymin}G{\isachardot}forces{\isacharunderscore}mem{\isacharparenleft}p{\isacharcomma}{\isasymtau}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isasymlongrightarrow}val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}{\isasymin}val{\isacharparenleft}G{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
+  \ \ {\isachardoublequoteopen}{\isacharparenleft}{\isasymtau}{\isasymin}M{\isasymlongrightarrow}{\isasymtheta}{\isasymin}M{\isasymlongrightarrow}{\isacharparenleft}{\isasymforall}p{\isasymin}G{\isachardot}forces{\isacharunderscore}eq{\isacharparenleft}p{\isacharcomma}{\isasymtau}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isasymlongrightarrow}val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}{\isacharequal}val{\isacharparenleft}G{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isacharparenright}{\isacharparenright}\isanewline
+  \ \  {\isasymand}\isanewline
+  \ \ {\isacharparenleft}{\isasymtau}{\isasymin}M{\isasymlongrightarrow}{\isasymtheta}{\isasymin}M{\isasymlongrightarrow}{\isacharparenleft}{\isasymforall}p{\isasymin}G{\isachardot}forces{\isacharunderscore}mem{\isacharparenleft}p{\isacharcomma}{\isasymtau}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isasymlongrightarrow}val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}{\isasymin}val{\isacharparenleft}G{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
 \end{isabelle}
@@
 From these results, the semantical characterization of the forcing
-relation (the ``definition of $\forces$''
+relation (the ``definition of $\forces$'' in 
 \cite[IV.2.22]{kunen2011set}) follows easily:
@@
 \isacommand{lemma}\isamarkupfalse%
-\ definition{\isacharunderscore}of{\isacharunderscore}forces{\isacharcolon}\isanewline
+\ definition{\isacharunderscore}of{\isacharunderscore}forcing{\isacharcolon}\isanewline
 \ \ \isakeyword{assumes}\isanewline
@@
 The present statement of the Fundamental Theorems is almost exactly
-the same of those of \cite{2019arXiv190103313G}, with the only
-modification being the bound on arities and a missing typing
-constraint. This implied only minor adjustments in the proofs of the
-satisfaction of axioms.
+the same of those in our previous report \cite{2019arXiv190103313G},
+with the only modification being the bound on arities and a missing
+typing constraint. This implied only minor adjustments in the proofs
+of the satisfaction of axioms.
 

### 2020-04-15 IJCAR2020/intro.tex
Various fixes suggested by the first referee and some other fixes.

@@
 formalizing the theory of forcing and its applications as presented in
-one of the more
-important references on the subject, Kunen's Set Theory
-\cite{kunen2011set} (a rewrite of the classical \cite{kunen1980}).
+one of the more important references on the subject, Kunen's Set
+Theory \cite{kunen2011set} (a rewrite of the classical book~\cite{kunen1980}).
 
@@
 showing the existence of generic filters via the Rasiowa-Sikorski
-lemma (RSL). Our second (unpublished) technical report
-\cite{2019arXiv190103313G} advanced by presenting the first accurate
+lemma (RSL). In our second (unpublished) technical report
+\cite{2019arXiv190103313G} we advanced by presenting the first accurate
 \emph{formal abstract} of the Fundamental Theorems of Forcing, and
@@
 recursion inside transitive models of an appropriate finite fragment
-of $\ZF$. As a by-product, we honored a long debt: the fact that the
+of $\ZF$. As a by-product, we finally met two long-standing goals: the fact that the
 generic filter $G$ belongs to the extension $M[G]$ and 

### 2020-04-15 IJCAR2020/replacement_choice.tex
typo matemático (faltó relativizar)
Various fixes suggested by the first referee and some other fixes.

@@
 
-In \cite{2019arXiv190103313G} we proved that any generic extension
-preserves the satisfaction of almost all the axioms, including the separation scheme
-(we adapted those proofs to the current statement of the axiom
-schemes). Our proofs that Replacement and choice hold in every generic
-extension depend on further relativized concepts and closure properties.
+In our report~\cite{2019arXiv190103313G} we proved that any generic
+extension preserves the satisfaction of almost all the axioms,
+including the separation scheme (we adapted those proofs to the
+current statement of the axiom schemes). Our proofs that Replacement
+and choice hold in every generic extension depend on further
+relativized concepts and closure properties.
 
@@
 stating that $M$ is included in 
-$\union\{\isatt{Vset}(\alpha) : \alpha\in M\}$ (it's actually equal).
+$\union\{\isatt{Vset}^M(\alpha) : \alpha\in M\}$ (actually they are equal).
 

### 2020-04-15 IJCAR2020/absoluteness.tex
Finishing with the typos and grammar comments by the first referee.

@@
 %% of type \isatt{i{\isasymRightarrow}o}. 
-For that reason, Separation and Replacement predicate \emph{on the
-  satisfaction} of a formula $\phi$.  We improved their specification,
-with respect to our previous report~\cite{2019arXiv190103313G}, by lifting
-the arity restriction for the parameter $\phi$; consequently we
-get rid of tupling and thus various proofs are now slicker.
+For that reason, Separation and Replacement are defined in terms of
+\emph{the satisfaction} of an internalized formula $\phi$.  We improved their
+specification, with respect to our previous
+report~\cite{2019arXiv190103313G}, by lifting the arity restriction
+for the parameter $\phi$; consequently we get rid of tupling and thus
+various proofs are now slicker.
 
@@
 \end{isabelle}
-(the actual formula obtained is
-\isatt{Forall(Forall(Exists(upair{\isacharunderscore}fm(2,1,0))))}) and
+and
 our \isatt{synthesize} method introduces a new term
-\isatt{ZF{\isacharunderscore}pairing{\isacharunderscore}fm} for it.
+\isatt{ZF{\isacharunderscore}pairing{\isacharunderscore}fm} for it:
 \begin{isabelle}
@@
 \end{isabelle}
+the actual formula obtained is
+\isatt{Forall(Forall(Exists(upair{\isacharunderscore}fm(2,1,0))))}.
 

### 2020-04-15 IJCAR2020/forcing_theorems.tex
Finishing with the typos and grammar comments by the first referee.

@@
 we were able to follow Kunen very closely. The more involved points of
-this part of the development were those where we needed to proved that
-various (dense) subsets of $\PP$ were in $M$; for this, we had to
-recourse to several ad-hoc absoluteness lemmas.
+this part of the development were those where we needed to prove that
+various (dense) subsets of $\PP$ were in $M$; for this, we have
+resorted to several ad-hoc absoluteness lemmas.
 

### 2020-04-15 IJCAR2020/proper_extension.tex
Finishing with the typos and grammar comments by the first referee.

@@
 is able to provide some \emph{proper} extension with the \emph{same
-ordinals}. After all, one is assuming from starters a model $M$ of $\ZFC$,
+ordinals}. After all, one is assuming from the start a model $M$ of $\ZFC$,
 and in some trivial cases $M[G]$ might end up to be exactly $M$; this

### 2020-04-15 IJCAR2020/related.tex
versión sin arriesgar comparación de código (aún)

--- 2020-03-24 IJCAR2020/related.tex
@@
 
-To the best of our knowledge, all of the previous works in
-formalization of the method 
-of forcing have been done in different variants of type theory, and
+%% \textbf{Reviewer's comments}
+%% {\it
+%%   \begin{itemize}
+%%   \item There, it would be appropriate to contrast what was done in
+%%     Paulson's work on constructibility with the current work on forcing.
+%%   \item More to the point, the recent work by Han and van Doorn on
+%%     forcing in Lean deserves more discussion.  They have gone further
+%%     than the current authors, having proved the independence of the
+%%     continuum hypothesis.  They prefer Boolean-valued models as being
+%%     more direct in use than the authors' countable transitive models.
+%%     \begin{itemize}
+%%     \item Readers will want to know whether the type-theoretic approach
+%%       is better/worse/just different than using Isabelle/ZF, and
+%%     \item are there any benefits to the ctm approach?
+%%     \item Is the type-theory encoding of ZF really accurate?
+%%     \item How about comparing proofs of equivalent statements in the two
+%%       approaches for length and readability?
+%%     \end{itemize}
+%%   \end{itemize}
+%% }
+
+There are various formalizations of Zermelo-Fraenkel set theory in
+proof assistants (v.g.\ Mizar, Metamath, and recently Lean
+\cite{DBLP:conf/cade/MouraKADR15}) that proceed to different levels of
+sophistication. Isabelle/ZF can be regarded as a notational variant of
+NGB set theory \cite[Sect.~II.10]{kunen2011set}, because the schemes
+of Replacement and Separation feature higher order (free) variables
+playing the role of formula variables. It can't be proved that the
+axioms thus written correspond to first order sentences. This is the
+reason that our relativized versions only apply to set models, where
+we can restrict those variables to predicates that actually come
+from first order formulas. In that sense, the axioms of the locale
+\isatt{M{\isacharunderscore}ZF} correspond more faithfully to the
+$\ZF$ axioms.
+
+Traditional expositions of the method of forcing
+\cite{kunen2011set,Jech_Millennium} are preceded by a study of
+relativization and absoluteness. For this reason, it was a natural
+choice at the beginning of this project to build on top of Paulson's
+formalization of constructibility on Isabelle/ZF, and that was a one
+of the main early reasons to work on that logic instead of, e.g., HOL
+---there are other reasons, that are discussed below. In any case, our
+development of forcing does not depend on constructibility
+itself (in contrast to Cohen's
+original presentation, in which ground models are initial segments of the
+constructible hierarchy).
+
+A natural question is whether Isabelle/HOL (with a far more solid
+framework to work with given its infrastructure and automation) would
+have been a better choice than Isabelle/ZF. In fact,
+there are two developments of Zermelo-Fraenkel set theory available on
+it: \isatt{HOLZF} by Obua \cite{DBLP:conf/ictac/Obua06} and
+\isatt{ZFC{\isacharunderscore}in{\isacharunderscore}HOL} by Paulson
+\cite{ZFC_in_HOL-AFP}. But these (logically equivalent) frameworks are
+higher in consistency strength than Isabelle/ZF. To elaborate on this,
+both ZF and HOL are axiomatized on top of Isabelle's metalogic
+\emph{Pure}, which is a version of ``intuitionistic higher order
+logic.'' In  \cite{Paulson1989} Paulson proves that \emph{Pure}
+is sound for intuitionistic first order logic, thus it does not add
+any strength to it. On top of this, the axiomatization of Isabelle/ZF
+results in a system equiconsistent with $\ZFC$. On the other hand,
+showing the consistency of \isatt{HOLZF} (and thus
+\isatt{ZFC{\isacharunderscore}in{\isacharunderscore}HOL}) requires
+assuming the consistency of $\ZFC$ plus the existence of an
+inaccesible cardinal \cite[Sect.~3]{DBLP:conf/ictac/Obua06}. We note,
+in contrast, that our extra running assumption of the existence of a
+countable transitive model is considerably weaker (directly and
+consistency-wise) than the existence of an inaccessible cardinal.
+
+Concerning the formalization of the method 
+of forcing, all of the previous works (to the best of our knowledge)
+have been done interpreting set theory into type theory, and
 none of them uses the ctm approach. The
-most important is the recent one by 
+most important is the recent \emph{Flypitch} project  by 
 Han and van Doorn
@@
 the Boolean-valued approach to forcing, using the Lean
-proof assistant \cite{DBLP:conf/cade/MouraKADR15}.
+proof assistant. The Flypitch
+formalization is largely orthogonal to ours, and this
+precludes a direct comparison of code. But we can
+highlight some conceptual differences.
+
+A first observation concerns consistency strength. The consistency of Lean requires infinitely many
+inaccessibles. More precisely, let Lean$_n$ be the theory of CiC
+foundations of Lean restricted to $n$ type universes.  Carneiro
+\cite{carneiro-ms-thesis}, proved the consistency of Lean$_n$ from
+$\ZFC$ plus the existence of $n$ inaccessible cardinals. It is also
+reported in \cite{carneiro-ms-thesis} that Werner's results in
+\cite{10.5555/645869.668660} can be adapted to show that Lean$_{n+2}$
+proves the consistency of the latter theory.
+In that sense, although Flypitch includes proofs of
+unprovability results in first order logic, the meta-theoretic
+machinery used to obtain them is far heavier than the one we use to
+operate model-theoretically.
+
+As a matter of taste, one of the main benefits of using transitive
+models is that many fundamental notions are absolute and thus most of
+the concepts and statements can be interpreted transparently, as we
+have noted before. It
+also provides a very concrete way to understand generic objects: as
+sets that (in the non trivial case) are provably not in the original
+model; this dispells any mystical feel around this concept (contrary
+to the case when the ground model is the universe of all sets). In
+addition, two-valued semantics is closer to our intuition.
 

### 2020-04-16 IJCAR2020/conclusion.tex
Finishing with the typos and grammar comments by the first referee.
intento de version final, con oracion en conclusion y footnote nueva

@@
 bit slower, the formulas synthesized were gigantic and the process on
-a whole as more error-prone. In fact, this approach was unsuccessful
+a whole was more error-prone. In fact, this approach was unsuccessful
 and we opted for a more detailed engineering, defining all
@@
 joint work with E.~Pacheco Rodríguez, which is needed for a general
-statement of the latter.
+statement of the latter. Once these developments are finished, we will
+be able to give a more thorough comparison between our project and
+the \emph{Flypitch} approach using Boolean-valued models.
 

### 2020-04-16 IJCAR2020/related.tex
intento de version final, con oracion en conclusion y footnote nueva

@@
 reason that our relativized versions only apply to set models, where
-we can restrict those variables to predicates that actually come
+we restrict those variables to predicates that actually come
 from first order formulas. In that sense, the axioms of the locale
@@
 precludes a direct comparison of code. But we can
-highlight some conceptual differences.
+highlight some conceptual differences between our development and the
+corresponding fraction of Flypitch.
 
-A first observation concerns consistency strength. The consistency of Lean requires infinitely many
-inaccessibles. More precisely, let Lean$_n$ be the theory of CiC
-foundations of Lean restricted to $n$ type universes.  Carneiro
-\cite{carneiro-ms-thesis}, proved the consistency of Lean$_n$ from
-$\ZFC$ plus the existence of $n$ inaccessible cardinals. It is also
-reported in \cite{carneiro-ms-thesis} that Werner's results in
+A first observation concerns consistency strength. The consistency of
+Lean requires infinitely many inaccessibles. More precisely, let
+Lean$_n$ be the theory of CiC foundations of Lean restricted to $n$
+type universes.  Carneiro \cite{carneiro-ms-thesis}, proved the
+consistency of Lean$_n$ from $\ZFC$ plus the existence of $n$
+inaccessible cardinals. It is also reported in
+\cite{carneiro-ms-thesis} that Werner's results in
 \cite{10.5555/645869.668660} can be adapted to show that Lean$_{n+2}$
-proves the consistency of the latter theory.
-In that sense, although Flypitch includes proofs of
-unprovability results in first order logic, the meta-theoretic
-machinery used to obtain them is far heavier than the one we use to
-operate model-theoretically.
+proves the consistency of the latter theory.  In that sense, although
+Flypitch includes proofs of unprovability results in first order
+logic, the meta-theoretic machinery used to obtain them is far heavier
+than the one we use to operate model-theoretically.
 
-As a matter of taste, one of the main benefits of using transitive
+In second place, a formalization of forcing with general partial
+orders, generic filters and  ctms has ---in our opinion--- the added value
+that this approach is used in an important (perhaps the greatest)
+fraction of the literature, both in exposition and in research
+articles and monographs. In verifying a piece of mature mathematics as the
+present one, representing the actual practice seems paramount to us.
+
+Finally, as a matter of taste, one of the main benefits of using transitive
 models is that many fundamental notions are absolute and thus most of

### 2020-04-16 IJCAR2020/related.tex
Minor corrections and phrasing.

@@
 of Replacement and Separation feature higher order (free) variables
-playing the role of formula variables. It can't be proved that the
-axioms thus written correspond to first order sentences. This is the
-reason that our relativized versions only apply to set models, where
+playing the role of formula variables. It cannot be proved that the
+axioms thus written correspond to first order sentences. For this
+reason, our relativized versions only apply to set models, where
 we restrict those variables to predicates that actually come
@@
 choice at the beginning of this project to build on top of Paulson's
-formalization of constructibility on Isabelle/ZF, and that was a one
+formalization of constructibility on Isabelle/ZF, and that was one
 of the main early reasons to work on that logic instead of, e.g., HOL
----there are other reasons, that are discussed below. In any case, our
+---below we discuss other reasons. In any case, our
 development of forcing does not depend on constructibility
@@
 \cite{han_et_al:LIPIcs:2019:11074,DBLP:conf/cpp/HanD20}, which includes
-a formalization of the independence of $\CH$ by means
+a formalization of the independence of $\CH$ by means of
 the Boolean-valued approach to forcing, using the Lean
@@
 Lean$_n$ be the theory of CiC foundations of Lean restricted to $n$
-type universes.  Carneiro \cite{carneiro-ms-thesis}, proved the
-consistency of Lean$_n$ from $\ZFC$ plus the existence of $n$
-inaccessible cardinals. It is also reported in
-\cite{carneiro-ms-thesis} that Werner's results in
+type universes.  Carneiro proved in his MSc thesis~\cite{carneiro-ms-thesis} the consistency of Lean$_n$ from $\ZFC$ plus
+the existence of $n$ inaccessible cardinals. It is also reported in
+Carneiro's thesis that Werner's results in
 \cite{10.5555/645869.668660} can be adapted to show that Lean$_{n+2}$

### 2020-04-17 IJCAR2020/related.tex
Removed redundancy about their approach.

@@
 
-Concerning the formalization of the method 
-of forcing, all of the previous works (to the best of our knowledge)
-have been done interpreting set theory into type theory, and
-none of them uses the ctm approach. The
-most important is the recent \emph{Flypitch} project  by 
-Han and van Doorn
-\cite{han_et_al:LIPIcs:2019:11074,DBLP:conf/cpp/HanD20}, which includes
-a formalization of the independence of $\CH$ by means of
-the Boolean-valued approach to forcing, using the Lean
-proof assistant. The Flypitch
-formalization is largely orthogonal to ours, and this
-precludes a direct comparison of code. But we can
-highlight some conceptual differences between our development and the
+Concerning the formalization of the method of forcing, to the best of
+our knowledge there is only another one that deals with forcing for
+set theory: the recent \emph{Flypitch} project by Han and van Doorn
+\cite{han_et_al:LIPIcs:2019:11074,DBLP:conf/cpp/HanD20}, which
+includes a formalization of the independence of CH by means of the
+Boolean-valued approach to forcing, using the Lean proof
+assistant. The Flypitch formalization is largely orthogonal to ours
+(it is based on Boolean-valued models, which are interpreted into
+type theory through a variant of the Aczel encoding of set theory),
+and this precludes a direct comparison of code. But we can highlight
+some conceptual differences between our development and the
 corresponding fraction of Flypitch.

### 2020-04-17 IJCAR2020/related.tex
frase rara e ispell

@@
 assuming the consistency of $\ZFC$ plus the existence of an
-inaccesible cardinal \cite[Sect.~3]{DBLP:conf/ictac/Obua06}. We note,
+inaccessible cardinal \cite[Sect.~3]{DBLP:conf/ictac/Obua06}. We note,
 in contrast, that our extra running assumption of the existence of a
@@
 Concerning the formalization of the method of forcing, to the best of
-our knowledge there is only another one that deals with forcing for
+our knowledge there is only one other that deals with forcing for
 set theory: the recent \emph{Flypitch} project by Han and van Doorn
@@
 sets that (in the non trivial case) are provably not in the original
-model; this dispells any mystical feel around this concept (contrary
+model; this dispels any mystical feel around this concept (contrary
 to the case when the ground model is the universe of all sets). In

### 2020-04-17 IJCAR2020/forces_definition.tex
Cambios de minúsculas por mayúsculas. Remoción de archivo no usado.

@@
 (implication, conjunction, and universal
-quantification, resp.), which operate on the meta-booleans
+quantification, resp.), which operate on the meta-Booleans
 \isatt{prop}.}

### 2020-04-17 IJCAR2020/isabelle_metatheory.tex
Various fixes suggested by the first referee and some other fixes.
Cambios de minúsculas por mayúsculas. Remoción de archivo no usado.

@@
 used as a language to define rules). Isabelle/ZF defines types
-\isatt{i} and \isatt{o} for sets and booleans, resp., and the $\ZF$
+\isatt{i} and \isatt{o} for sets and Booleans, resp., and the $\ZF$
 axioms are written down as terms of type \isatt{o}.
@@
 found usually in type theories with universes, but allows us to work
-``platonistically''--- assuming we are in a universe of sets (namely,
+``Platonistically''--- assuming we are in a universe of sets (namely,
 \isatt{i}) and performing constructions there.
@@
 %% \isatt{ms {\isasymin} list(M)} and \isatt{{\isasymphi} {\isasymin} formula})
-had already been defined by recursion in \cite{paulson_2003}.
+had already been defined by recursion in Paulson~\cite{paulson_2003}.
 

### 2020-04-18 IJCAR2020/forces_definition.tex
3 typos, esencialmente

@@
 proved that it is equivalent to the semantic version 
-(``\isatt{definition{\isacharunderscore}of{\isacharunderscore}forces},'' in
+(``\isatt{definition{\isacharunderscore}of{\isacharunderscore}forcing},'' in
 the next section).
@@
 $F\!\upharpoonright\!(R^{-1}(x))(y) =
-F\!\upharpoonright\!({R^+}^{-1}(x))(y)$ whenever $(x,y) \in R$.
+F\!\upharpoonright\!\bigl((R^+)^{-1}(x)\bigr)(y)$ whenever $(x,y) \in R$.
 

### 2020-04-18 IJCAR2020/related.tex
3 typos, esencialmente

@@
 \cite{han_et_al:LIPIcs:2019:11074,DBLP:conf/cpp/HanD20}, which
-includes a formalization of the independence of CH by means of the
-Boolean-valued approach to forcing, using the Lean proof
+includes a formalization of the independence of CH using the Lean proof
 assistant. The Flypitch formalization is largely orthogonal to ours

### 2020-04-25 src/document/root.tex
cambio de título del documento
simbolos nuevos
arreglados guiones bajos
arreglo algunos símbolos en documento y paper (por ejemplo, ||-)
abstracts & references

@@
 \usepackage{isabelle,isabellesym}
+\usepackage[numbers]{natbib}
 
@@
 
-%\usepackage{amssymb}
+\usepackage{amssymb}
   %for \<leadsto>, \<box>, \<diamond>, \<sqsupset>, \<mho>, \<Join>,
@@
 
+\renewcommand{\isacharunderscorekeyword}{\mbox{\_}}
+\renewcommand{\isacharunderscore}{\mbox{\_}}
+\renewcommand{\isasymtturnstile}{\isamath{\Vdash}}
+\renewcommand{\isacharminus}{-}
+\newcommand{\axiomas}[1]{\mathit{#1}}
+\newcommand{\ZFC}{\axiomas{ZFC}}
+
 \begin{document}
 
-\title{src}
-\author{By miguel}
+\title{Formalization of Forcing in Isabelle/ZF}
+\author{Emmanuel Gunther
+  \and 
+  Miguel Pagano
+  \and 
+  Pedro S\'anchez Terraf
+}
 \maketitle
+
+\begin{abstract}
+We formalize the theory of forcing in the set theory framework of
+Isabelle/ZF. Under the assumption of the existence of a countable
+transitive model of $\ZFC$, we construct a proper generic extension and show
+that the latter also satisfies $\ZFC$. In doing so, we remodularized
+Paulson's \verb|ZF-Constructibility| library.
+\end{abstract}
 
@@
 
+\section{Introduction}
+This development is described in more detail in our papers
+\cite{2018arXiv180705174G} (LSFA 2018), \cite{2019arXiv190103313G},
+and \cite{2020arXiv200109715G} (IJCAR 2020).
+
 % generated text of all theories
@@
 % optional bibliography
-%\bibliographystyle{abbrv}
-%\bibliography{root}
+\bibliographystyle{root}
+\bibliography{root}
 

### 2020-05-04 src/document/root.tex
no reference to ZF-Constructible in abstract
Release notes in root.

@@
 transitive model of $\ZFC$, we construct a proper generic extension and show
-that the latter also satisfies $\ZFC$. In doing so, we remodularized
-Paulson's \verb|ZF-Constructibility| library.
+that the latter also satisfies $\ZFC$.
 \end{abstract}
@@
 \section{Introduction}
-This development is described in more detail in our papers
+We formalize the theory of forcing. We work on top of the Isabelle/ZF
+framework developed by \citet{DBLP:journals/jar/PaulsonG96}. Our
+mechanization is described in more detail in our papers
 \cite{2018arXiv180705174G} (LSFA 2018), \cite{2019arXiv190103313G},
 and \cite{2020arXiv200109715G} (IJCAR 2020).
+
+\subsection*{Release notes}
+\label{sec:release-notes}
+
+We have improved several aspects of our development before submitting
+it to the AFP:
+\begin{enumerate}
+\item Our session \isatt{Forcing} depends on the new release of
+  \isatt{ZF-Constructible}.
+\item We streamlined the commands for synthesizing renames and formulas.
+\item The command that synthesizes formulas produces the lemmas for
+  them (the synthesized term is a formula and the equivalence between
+  the satisfaction of the synthesized term and the relativized term).
+\item Consistently use of structured proofs using Isar (except for one
+  coming from a schematic goal command).
+\end{enumerate}
 

### 2020-06-08 IJCAR2020/slides/forcing_IJCAR2020_slides.tex
template for IJCAR slides

\documentclass[english]{beamer}
%\usepackage{beamerarticle}
\usepackage{graphicx}
\usepackage{babel}
\usepackage{tikz}
\usetikzlibrary{arrows,shapes}
\usepackage[all]{xy}
\usepackage[invisible]{arxiv-beamer-theme}

\usepackage[utf8]{inputenc}
%%%
%%% Natbib must be here!! (after Beamer stuff)
\usepackage[numbers]{natbib}
\input{header-slides}
\usepackage{isabelle,isabellesym}

\CompileMatrices

\title[Forcing in Isabelle/ZF]{Formalization of Forcing in Isabelle/ZF}

\author[E.~Gunther, M.~Pagano, PST]{E.~Gunther \qquad M.~Pagano \qquad P.~Sánchez Terraf\thanks{Supported by
    CONICET and SeCyT-UNC} 
}
\institute[UNC]{CIEM-FaMAF --- Universidad Nacional de Córdoba}

\date[IJCAR 2020]{International Joint Conference on Automated
  Reasoning\\ Paris (Virtual),\ \  2020 / 07 / 02}

\subject{Logic in Computer Science}

\pgfdeclareimage[height=3ex]{UNC-logo}{unc_400_Blanco_Negro}
\logo{\pgfuseimage{UNC-logo}}

%% Para usar overlays con tikz
\tikzset{onslide/.code args={<#1>#2}{%
    \only<#1>{\pgfkeysalso{#2}} % \pgfkeysalso doesn't change the path
}}

\begin{document}
\begin{frame}[plain]
  \titlepage
  \begin{center}
    \insertlogo
  \end{center}
\end{frame}
%
\begin{frame}
  \frametitle{Summary}
  \tableofcontents
  \transwipe
\end{frame}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{}

%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
\subsection{}

\begin{frame}
  \frametitle{}
\end{frame}

%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
\begin{frame}
  \begin{shadowblock}{}
    \begin{center}
      {\Huge Thank you!}
    \end{center}
  \end{shadowblock}
\end{frame}

\end{document}
%%% Local Variables: 
%%% mode: latex
%%% ispell-local-dictionary: "american"
%%% End:

### 2020-06-10 IJCAR2020/slides/forcing_IJCAR2020_slides.tex
first draft of some structure for slides

@@
 %%% Natbib must be here!! (after Beamer stuff)
-\usepackage[numbers]{natbib}
+\usepackage[authoryear,square]{natbib}
 \input{header-slides}
 \usepackage{isabelle,isabellesym}
+\newcommand{\uscore}{\isacharunderscore}
 
@@
 \begin{document}
+%\nobibliography*
+
 \begin{frame}[plain]
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{}
+\section{Introduction}
+
+%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
+\subsection{Why Isabelle/ZF?}
+
+\begin{frame}
+  \frametitle{Why Isabelle/ZF?}
+\end{frame}
+
+\begin{frame}
+  \frametitle{}
+
+\end{frame}
+
+%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
+\subsection{Paulson's \texttt{ZF-Constructible}}
+
+\begin{frame}
+  \frametitle{Paulson's \texttt{ZF-Constructible}}
+\end{frame}
+
+%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
+\subsection{The ctm approach to forcing}
+
+\begin{frame}
+  \frametitle{The ctm approach to forcing, I}
+\end{frame}
+
+\begin{frame}
+  \frametitle{The ctm approach to forcing, II}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Other approaches}
+  \begin{itemize}
+  \item<+-> \emph{Set theory in HOL}
+    \begin{itemize}
+    \item \isatt{HOLZF} \citep{DBLP:conf/ictac/Obua06}
+    \item \isatt{ZFC{\uscore}in{\uscore}HOL}
+      \citep{ZFC_in_HOL-AFP}
+    \end{itemize}
+  \item<+-> \alert{Lean}: Full formalization of the Boolean-valued approach to
+    forcing and the independence of $\CH$ \citep{DBLP:conf/cpp/HanD20}.
+  \end{itemize}
+  %
+  \begin{block}{A word on consistency}<+->
+    \begin{description}
+      \setlength{\labelwidth}{12em}
+      \setlength{\labelsep}{2em}
+      \setlength{\itemindent}{6em}
+    \item[Isabelle/ZF + ctm]  (far) less than $\ZF$ +  one
+      inaccessible.
+    \item[\isatt{HOLZF},
+      \isatt{ZFC{\uscore}in{\uscore}HOL}]
+      approximately $\ZF$ +  one   inaccessible. 
+    \item[Lean (CiC)] $\ZF$ + $\omega$ inaccessibles. 
+    \end{description}
+  \end{block}
+\end{frame}
+
+%% \begin{frame}
+%%   \frametitle{A word on consistency}
+%% \end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{The development}
+
+%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
+\subsection{What did we accomplished?}
+
+\begin{frame}
+  \frametitle{What did we accomplished?}
+  %
+  \begin{shadowblock}{}
+    \begin{enumerate}
+    \item We adapted \texttt{ZF-Constructible} to obtain
+      absoluteness results for nonempty transitive classes ($\leftarrow$
+      already in the 2020 release of Isabelle).
+    \item We formalized the forcing relation and proved the Fundamental
+      Theorems.
+    \item We showed that generic extensions of ctms of $\ZF$ are also
+      ctms of $\ZF$ (respectively, adding $\AC$).
+    \item We provided the forcing notion that adds a Cohen real,
+      therefore proving the existence of a nontrivial extension.
+    \end{enumerate}
+  \end{shadowblock}
+\end{frame}
+
+%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
+\subsection{A map}
+
+\begin{frame}
+  \frametitle{A map to the development}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Locale structure} 
+             {  
+               \renewcommand{\arraystretch}{1.5}               \begin{tabular}{rcl}
+                 \texttt{forcing{\uscore}notion} & = & preorder $\PP$ with top. \\
+                 %    \texttt{M{\uscore}ZF} & = & set model of the $\ZF$ axioms \\ 
+                 \texttt{M{\uscore}ZF{\uscore}trans} & = & set model $M$ of the $\ZF$
+                 axioms \alert{+}  $M$ transitive \\ 
+                 \texttt{M{\uscore}ctm} & = &  \texttt{M{\uscore}ZF{\uscore}trans} \alert{+}
+                 $M$ countable \\
+                 \texttt{forcing{\uscore}data} & =  & \texttt{M{\uscore}ctm} \alert{+}
+                 \texttt{forcing{\uscore}notion} in $M$\\
+                 \texttt{separative{\uscore}notion} & = &
+                 \texttt{forcing{\uscore}notion} \alert{+} $\PP$ separative \\
+                 \texttt{M{\uscore}ctm{\uscore}separative} & = &
+                 \texttt{forcing{\uscore}data} \alert{+} \texttt{separative{\uscore}notion}
+               \end{tabular} 
+             }
+\end{frame}
 
@@
   \frametitle{}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{}
+%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
+\subsection{}
+
+\begin{frame}
+  \frametitle{}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Looking forward}
+
+\begin{frame}
+  \frametitle{Looking forward}
+  \begin{block}{Formalizing math}
+    \begin{itemize}
+    \item<+-> Cofinality, K\H{o}nig's Theorem, Shanin's $\Delta$-system Lemma.
+    \item<+-> Forcing notion for adding $\kappa$ Cohen reals.
+    \item<+-> Theorems on preservation of cardinals.
+    \end{itemize}
+  \end{block}
+  %
+  \begin{block}{Technical aids}<+->
+    \begin{enumerate}
+    \item<.-> Automatic relativization of concepts.
+    \item<+-> Relative functions.
+    \item<+->
+    \end{enumerate}
+  \end{block}
 \end{frame}
@@
 
+\begin{frame}
+  \frametitle{References}
+  \bibliographystyle{mi-estilo-else}
+  \bibliography{../forcing_in_isabelle_zf}
+\end{frame}
+
 \end{document}

### 2020-06-12 IJCAR2020/slides/forcing_IJCAR2020_slides.tex
code discussed in slides

@@
   \frametitle{Why Isabelle/ZF?}
+  \begin{block}{Pros}
+    \begin{itemize}
+    \item Most advanced set theory formalized (around 2017).
+    \item Closer to the mathematical practice of forcing.
+    \item Structured proof language Isar
+      \citep{DBLP:conf/tphol/Wenzel99}.
+    \item Weak metatheory.
+    \end{itemize}
+  \end{block}
+  %
+  \pause
+  %
+  \begin{block}{Cons}
+    \begin{itemize}
+    \item A fraction of automation of Isabelle (\texttt{sledgehammer},
+      etc).
+    \item Untyped, and too weak a metatheory.
+    \end{itemize}
+  \end{block}
 \end{frame}
@@
 \begin{frame}
-  \frametitle{}
-
+  \frametitle{Isabelle/ZF \citep{DBLP:journals/jar/PaulsonG96}}
+  \begin{itemize}
+  \item<+-> An object logic of Isabelle axiomatized over the
+    intuitionistic fragment of HOL \emph{Pure}. 
+  \item<+-> It postulates \alert<3->{two types}: \isatt{i}
+    (sets) and \isatt{o} (booleans). \visible<3->{\alert{Not
+      inductively defined!}}
+    \pause
+  \item<+-> More stuff.
+  \end{itemize}
 \end{frame}
@@
   \frametitle{Paulson's \texttt{ZF-Constructible}}
+
 \end{frame}
@@
   \frametitle{The ctm approach to forcing, I}
+  \begin{block}{Countable transitive model (ctm) of $\ZF$}
+    $\lb M, E\rb \models \ZF$ where
+    \begin{itemize}
+    \item $M$ is countable.
+    \item $E \defi {\in}\restriction M$.
+    \item $x\in y \in M \implies x\in M$.
+    \end{itemize}
+  \end{block}
+  %
+  \pause%
+  %
+  \textbf{Note}. If $\lb N, R\rb \models \ZF$ with $R$ wellfounded, then there
+  exists a ctm $M$ of $\ZF$.
+  %
+  \pause
+  %
+  \begin{block}{Forcing notions and generics}
+    \begin{itemize}
+    \item $\lb \PP, {\preceq} ,\1\rb$ preorder with top.
+    \item A filter $G\sbq \PP$ is \textbf{$M$-generic} if it meets
+      every dense $D\sbq \PP$ in $M$.
+    \end{itemize}
+  \end{block}
 \end{frame}
@@
   \frametitle{The ctm approach to forcing, II}
+  Given $G\sbq \PP$, we can adjoin it to $M$ to form the
+  \textbf{generic extension} $M[G]$.
+  \pause
+  Every $a\in M[G]$ is coded by
+  some $\punto{a}\in M$ through the definable (in $M$) function $\val$:
+  \[
+  M[G]\defi \{ \val(G,\punto{a}) : \punto{a}\in M\}
+  \]
+  \pause
+  \emph{Fundamentally}, \textbf{truth} in $M[G]$ is coded in $M$ by the
+  function $\forceisa$.
+  \pause
+  \begin{theorem}[Cohen, 1963]
+    \begin{enumerate}
+    \item $\bigl(\forall G\visible<5->{\alert{\ni p}}.\quad M[G], [\val(G,\punto{a})]
+      \models \phi\bigr)
+      \iff
+      M,
+      [\only<-4>{\1}\alert{\only<5->{p\hspace{0.2ex}}},\PP,\preceq,\punto{a}]\models
+      \forceisa(\phi)$.
+    \item $\forceisa$ is definable over $M$.\phantom{$\punto{a}).\models \phi\bigr)
+      \iff M,$}\visible<6->{%
+      $\overset{\searrow}{}$ \ 
+      $\alert{p} \forces^M_{\PP,\preceq} \phi(\punto{a})$
+      \  $\overset{\swarrow}{}$}
+      \pause% 
+      \pause% 
+      \pause%       
+    \item For $env\in\mathsf{list}(M)$,
+      \[
+      M[G], \mathrm{map}(\val(G),env) \models \phi)
+      \iff
+      \exists p\in G.\ M, [p,\preceq,\PP] \mathbin{{+}\!\!\!\!{+}} env \models
+      \forceisa(\phi).
+      \]
+    \end{enumerate}
+  \end{theorem} 
+\end{frame}
+
+\begin{frame}
+  \frametitle{The ctm approach to forcing, III}
+  By choosing the forcing notion $\lb \PP, {\preceq} ,\1\rb$
+  appropriately one can tune the first order properties of  $M[G]$ (for
+  any generic $G$).
+  \pause
+
 \end{frame}
@@
   \begin{itemize}
-  \item<+-> \emph{Set theory in HOL}
+  \item<+-> Set theory over Isabelle/HOL:
     \begin{itemize}
@@
   %
-  \begin{block}{A word on consistency}<+->
+  \begin{block}{A word on consistency strength}<+->
     \begin{description}
@@
       approximately $\ZF$ +  one   inaccessible. 
-    \item[Lean (CiC)] $\ZF$ + $\omega$ inaccessibles. 
+    \item[Lean (CiC)] $\ZF$ + $\omega$ inaccessibles \citep{carneiro-ms-thesis}. 
     \end{description}
@@
 %-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
-\subsection{What did we accomplished?}
+\subsection{What did we accomplish?}
 
 \begin{frame}
-  \frametitle{What did we accomplished?}
+  \frametitle{What did we accomplish?}
   %
@@
     \begin{enumerate}
-    \item We adapted \texttt{ZF-Constructible} to obtain
+    \item<+-> We adapted \texttt{ZF-Constructible} to obtain
       absoluteness results for nonempty transitive classes ($\leftarrow$
-      already in the 2020 release of Isabelle).
-    \item We formalized the forcing relation and proved the Fundamental
+      \alert{in Isabelle2020}).
+      %
+      \pause
+
+      Around 40 absoluteness/closure lemmas now hold using weaker
+      hypotheses on the class $M$ (most of them, just that $M$ is
+      transitive and nonempty). 
+      %
+      \pause
+    \item<+-> We formalized the formula transformer $\forceisa$ and
+      hence the forcing relation $\forces$, and proved the Fundamental
       Theorems.
-    \item We showed that generic extensions of ctms of $\ZF$ are also
+    \item<+-> We showed that generic extensions of ctms of $\ZF$ are also
       ctms of $\ZF$ (respectively, adding $\AC$).
-    \item We provided the forcing notion that adds a Cohen real,
+    \item<+-> We provided the forcing notion that adds a Cohen real,
       therefore proving the existence of a nontrivial extension.
@@
   \end{shadowblock}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Some details}
+  \begin{enumerate}
+  \item<+-> Modifications to \texttt{ZF-Constructible}.
+
+  \item<+-> We formalized the formula transformer $\forceisa$ and
+    hence the forcing relation $\forces$, and proved the Fundamental
+    Theorems.
+  \item<+-> We showed that generic extensions of ctms of $\ZF$ are also
+    ctms of $\ZF$ (respectively, adding $\AC$).
+  \item<+-> We provided the forcing notion that adds a Cohen real,
+    therefore proving the existence of a nontrivial extension.
+  \end{enumerate}
 \end{frame}
@@
 \begin{frame}
-  \frametitle{A map to the development}
+  \begin{center}
+    \only<01>{\includegraphics[scale=0.23]{kunen_powerset_01.png} \\ \includegraphics[scale=0.26]{isa_converted_01.png}}%
+    \only<02>{\includegraphics[scale=0.23]{kunen_powerset_02.png} \\ \includegraphics[scale=0.26]{isa_converted_02.png}}%
+    \only<03>{\includegraphics[scale=0.23]{kunen_powerset_03.png} \\ \includegraphics[scale=0.26]{isa_converted_03.png}}%
+    \only<04>{\includegraphics[scale=0.23]{kunen_powerset_04.png} \\ \includegraphics[scale=0.26]{isa_converted_04.png}}%
+    \only<05>{\includegraphics[scale=0.23]{kunen_powerset_05.png} \\ \includegraphics[scale=0.26]{isa_converted_05.png}}%
+    \only<06>{\includegraphics[scale=0.23]{kunen_powerset_06.png} \\ \includegraphics[scale=0.26]{isa_converted_06.png}}%
+    \only<07>{\includegraphics[scale=0.23]{kunen_powerset_07.png} \\ \includegraphics[scale=0.26]{isa_converted_07.png}}%
+    \only<08>{\includegraphics[scale=0.23]{kunen_powerset_08.png} \\ \includegraphics[scale=0.26]{isa_converted_08.png}}%
+    \only<09>{\includegraphics[scale=0.23]{kunen_powerset_09.png} \\ \includegraphics[scale=0.26]{isa_converted_09.png}}%
+    \only<10>{\includegraphics[scale=0.23]{kunen_powerset_10.png} \\ \includegraphics[scale=0.26]{isa_converted_10.png}}%
+    \only<11>{\includegraphics[scale=0.23]{kunen_powerset_11.png} \\ \includegraphics[scale=0.26]{isa_converted_11.png}}%
+    \only<12>{\includegraphics[scale=0.23]{kunen_powerset_12.png} \\ \includegraphics[scale=0.26]{isa_converted_12.png}}%
+    \only<13>{\includegraphics[scale=0.23]{kunen_powerset_13.png} \\ \includegraphics[scale=0.26]{isa_converted_13.png}}%
+    \only<14>{\includegraphics[scale=0.23]{kunen_powerset_14.png} \\ \includegraphics[scale=0.26]{isa_converted_14.png}}%
+    \only<15>{\includegraphics[scale=0.23]{kunen_powerset_15.png} \\ \includegraphics[scale=0.26]{isa_converted_15.png}}%
+    \only<16>{\includegraphics[scale=0.23]{kunen_powerset_16.png} \\ \includegraphics[scale=0.26]{isa_converted_16.png}}%
+    \only<17>{\includegraphics[scale=0.23]{kunen_powerset_17.png} \\ \includegraphics[scale=0.26]{isa_converted_17.png}}%
+    \only<18>{\includegraphics[scale=0.23]{kunen_powerset_18.png} \\ \includegraphics[scale=0.26]{isa_converted_18.png}}%
+    \only<19>{\includegraphics[scale=0.23]{kunen_powerset_19.png} \\ \includegraphics[scale=0.26]{isa_converted_19.png}}%
+    \only<20>{\includegraphics[scale=0.23]{kunen_powerset_20.png} \\ \includegraphics[scale=0.26]{isa_converted_20.png}}%
+  \end{center}
 \end{frame}
@@
   \begin{block}{Technical aids}<+->
-    \begin{enumerate}
-    \item<.-> Automatic relativization of concepts.
-    \item<+-> Relative functions.
+    \begin{itemize}
+    \item<.-> Automatic relativization and proof of absoluteness of concepts.
+    \item<+-> ``Relative functions'' (e.g., $\P^M$, $\card{\cdot}^M$, $\cf^M$).
     \item<+-> 
-    \end{enumerate}
+    \end{itemize}
   \end{block}
@@
 \end{document}
+
+%% - Más bola a la metateoría (ZF).
+%% 
+%% * es weak ---> falta de inducción/recursión, y el objeto
+%% central (forces) se define así.
+%% * too strong ---> not possible to have finitary consistency proofs
+%% 
+%% - ctm approach: bondades 
+%% 
+%% * absolutez de ordinal, etc
+%% * Modelos a dos valores.
+%% 
+%% - no definir locales, hablar menos de la arquitectura modular
+%% (no tanto del codigo en si)
+%% 
+%% - Pruebas mostrar:
+%% 
+%% density_mem ?
+%% powerset?
+%% choice?
+
 %%% Local Variables: 

### 2020-06-13 IJCAR2020/slides/forcing_IJCAR2020_slides.tex
minimal changes to the slides; draft mode for pictures

@@
     \item Most advanced set theory formalized (around 2017).
-    \item Closer to the mathematical practice of forcing.
+      % \item Closer to the mathematical practice of forcing.
     \item Structured proof language Isar
@@
   \frametitle{Paulson's \texttt{ZF-Constructible}}
-
+  Let $C :: {}$ \isatt{i} $\Rightarrow$ \isatt{o} be a predicate on sets.
+  \pause
+  \begin{block}{Relativization of $\sbq$}
+    $x \sbq^C y \equiv \forall z.\ C(z) \longrightarrow (z\in x
+    \longrightarrow z\in y)$.
+  \end{block}
+  \pause
+  \begin{block}{$\sbq$ is absolute}
+    $\forall x\ y.\ C(x) \land C(y) \longrightarrow (x \sbq^C y
+    \longleftrightarrow x\sbq y)$.
+  \end{block}
+  \pause 
+  \texttt{ZF-Constructible} develops relativized version of the axioms
+  and proves that many concepts are absolute for \textbf{transitive
+    classes} $C$
+  \[
+  x\in y \land C(y)  \longrightarrow C(x).
+  \]
+  satisfying additional axioms.
 \end{frame}
@@
   \]
-  \pause
   \emph{Fundamentally}, \textbf{truth} in $M[G]$ is coded in $M$ by the
@@
   \pause
-  \begin{theorem}[Cohen, 1963]
+  \begin{theorem}[\citep{Cohen-CH-PNAS}]
     \begin{enumerate}
-    \item $\bigl(\forall G\visible<5->{\alert{\ni p}}.\quad M[G], [\val(G,\punto{a})]
+    \item $\bigl(\forall G\visible<4->{\alert{\ni p}}.\quad M[G], [\val(G,\punto{a})]
       \models \phi\bigr)
@@
       M,
-      [\only<-4>{\1}\alert{\only<5->{p\hspace{0.2ex}}},\PP,\preceq,\punto{a}]\models
+      [\only<-3>{\1}\alert{\only<4->{p\hspace{0.2ex}}},\PP,\preceq,\punto{a}]\models
       \forceisa(\phi)$.
     \item $\forceisa$ is definable over $M$.\phantom{$\punto{a}).\models \phi\bigr)
-      \iff M,$}\visible<6->{%
+      \iff M,$}\visible<5->{%
       $\overset{\searrow}{}$ \ 
@@
       \[
-      M[G], \mathrm{map}(\val(G),env) \models \phi)
+      M[G], \mathrm{map}(\val(G),env) \models \phi
       \iff
@@
   \pause
-
+  %
+  \begin{theorem}[\citep{Cohen-CH-PNAS}]
+    If $\PP$ is the set of finite partial binary functions with domain
+    included in $\aleph_2$, $M[G]$ satisfies the negation of the
+    \textbf{Continuum Hypothesis}:
+    \[
+    M[G] \models 2^{\aleph_0} >\aleph_1
+    \]
+  \end{theorem}
 \end{frame}
@@
     \begin{enumerate}
-    \item<+-> We adapted \texttt{ZF-Constructible} to obtain
+    \item<1-> We adapted \texttt{ZF-Constructible} to obtain
       absoluteness results for nonempty transitive classes ($\leftarrow$
       \alert{in Isabelle2020}).
-      %
-      \pause
 
+      \visible<2->{
         Around 40 absoluteness/closure lemmas now hold using weaker
-      hypotheses on the class $M$ (most of them, just that $M$ is
+        hypotheses on the class $C$ (most of them, just that $C$ is
         transitive and nonempty).
-      %
-      \pause
-    \item<+-> We formalized the formula transformer $\forceisa$ and
+      }
+    \item<3-> We formalized the formula transformer $\forceisa$ and
       hence the forcing relation $\forces$, and proved the Fundamental
       Theorems.
-    \item<+-> We showed that generic extensions of ctms of $\ZF$ are also
+    \item<4-> We showed that generic extensions of ctms of $\ZF$ are also
       ctms of $\ZF$ (respectively, adding $\AC$).
-    \item<+-> We provided the forcing notion that adds a Cohen real,
+    \item<5-> We provided the forcing notion that adds a Cohen real,
       therefore proving the existence of a nontrivial extension.
@@
   \frametitle{Some details}
-  \begin{enumerate}
-  \item<+-> Modifications to \texttt{ZF-Constructible}.
-
+  \begin{enumerate}\setcounter{enumi}{1}
   \item<+-> We formalized the formula transformer $\forceisa$ and
@@
     Theorems.
+    \begin{itemize}
+    \item<+-> 
+    \item<+-> 
+    \end{itemize}
   \item<+-> We showed that generic extensions of ctms of $\ZF$ are also
     ctms of $\ZF$ (respectively, adding $\AC$).
-  \item<+-> We provided the forcing notion that adds a Cohen real,
-    therefore proving the existence of a nontrivial extension.
+    \begin{itemize}
+    \item<+-> We can see a bit of the code, compared to the actual math.
+    \end{itemize}
   \end{enumerate}
@@
 %-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
-\subsection{A map}
+\subsection{Math vs Code}
 
@@
   \begin{center}
-    \only<01>{\includegraphics[scale=0.23]{kunen_powerset_01.png} \\ \includegraphics[scale=0.26]{isa_converted_01.png}}%
-    \only<02>{\includegraphics[scale=0.23]{kunen_powerset_02.png} \\ \includegraphics[scale=0.26]{isa_converted_02.png}}%
-    \only<03>{\includegraphics[scale=0.23]{kunen_powerset_03.png} \\ \includegraphics[scale=0.26]{isa_converted_03.png}}%
-    \only<04>{\includegraphics[scale=0.23]{kunen_powerset_04.png} \\ \includegraphics[scale=0.26]{isa_converted_04.png}}%
-    \only<05>{\includegraphics[scale=0.23]{kunen_powerset_05.png} \\ \includegraphics[scale=0.26]{isa_converted_05.png}}%
-    \only<06>{\includegraphics[scale=0.23]{kunen_powerset_06.png} \\ \includegraphics[scale=0.26]{isa_converted_06.png}}%
-    \only<07>{\includegraphics[scale=0.23]{kunen_powerset_07.png} \\ \includegraphics[scale=0.26]{isa_converted_07.png}}%
-    \only<08>{\includegraphics[scale=0.23]{kunen_powerset_08.png} \\ \includegraphics[scale=0.26]{isa_converted_08.png}}%
-    \only<09>{\includegraphics[scale=0.23]{kunen_powerset_09.png} \\ \includegraphics[scale=0.26]{isa_converted_09.png}}%
-    \only<10>{\includegraphics[scale=0.23]{kunen_powerset_10.png} \\ \includegraphics[scale=0.26]{isa_converted_10.png}}%
-    \only<11>{\includegraphics[scale=0.23]{kunen_powerset_11.png} \\ \includegraphics[scale=0.26]{isa_converted_11.png}}%
-    \only<12>{\includegraphics[scale=0.23]{kunen_powerset_12.png} \\ \includegraphics[scale=0.26]{isa_converted_12.png}}%
-    \only<13>{\includegraphics[scale=0.23]{kunen_powerset_13.png} \\ \includegraphics[scale=0.26]{isa_converted_13.png}}%
-    \only<14>{\includegraphics[scale=0.23]{kunen_powerset_14.png} \\ \includegraphics[scale=0.26]{isa_converted_14.png}}%
-    \only<15>{\includegraphics[scale=0.23]{kunen_powerset_15.png} \\ \includegraphics[scale=0.26]{isa_converted_15.png}}%
-    \only<16>{\includegraphics[scale=0.23]{kunen_powerset_16.png} \\ \includegraphics[scale=0.26]{isa_converted_16.png}}%
-    \only<17>{\includegraphics[scale=0.23]{kunen_powerset_17.png} \\ \includegraphics[scale=0.26]{isa_converted_17.png}}%
-    \only<18>{\includegraphics[scale=0.23]{kunen_powerset_18.png} \\ \includegraphics[scale=0.26]{isa_converted_18.png}}%
-    \only<19>{\includegraphics[scale=0.23]{kunen_powerset_19.png} \\ \includegraphics[scale=0.26]{isa_converted_19.png}}%
-    \only<20>{\includegraphics[scale=0.23]{kunen_powerset_20.png} \\ \includegraphics[scale=0.26]{isa_converted_20.png}}%
+    \only<01>{\includegraphics[draft,scale=0.23]{kunen_powerset_01.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_01.png}}%
+    \only<02>{\includegraphics[draft,scale=0.23]{kunen_powerset_02.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_02.png}}%
+    \only<03>{\includegraphics[draft,scale=0.23]{kunen_powerset_03.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_03.png}}%
+    \only<04>{\includegraphics[draft,scale=0.23]{kunen_powerset_04.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_04.png}}%
+    \only<05>{\includegraphics[draft,scale=0.23]{kunen_powerset_05.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_05.png}}%
+    \only<06>{\includegraphics[draft,scale=0.23]{kunen_powerset_06.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_06.png}}%
+    \only<07>{\includegraphics[draft,scale=0.23]{kunen_powerset_07.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_07.png}}%
+    \only<08>{\includegraphics[draft,scale=0.23]{kunen_powerset_08.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_08.png}}%
+    \only<09>{\includegraphics[draft,scale=0.23]{kunen_powerset_09.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_09.png}}%
+    \only<10>{\includegraphics[draft,scale=0.23]{kunen_powerset_10.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_10.png}}%
+    \only<11>{\includegraphics[draft,scale=0.23]{kunen_powerset_11.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_11.png}}%
+    \only<12>{\includegraphics[draft,scale=0.23]{kunen_powerset_12.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_12.png}}%
+    \only<13>{\includegraphics[draft,scale=0.23]{kunen_powerset_13.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_13.png}}%
+    \only<14>{\includegraphics[draft,scale=0.23]{kunen_powerset_14.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_14.png}}%
+    \only<15>{\includegraphics[draft,scale=0.23]{kunen_powerset_15.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_15.png}}%
+    \only<16>{\includegraphics[draft,scale=0.23]{kunen_powerset_16.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_16.png}}%
+    \only<17>{\includegraphics[draft,scale=0.23]{kunen_powerset_17.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_17.png}}%
+    \only<18>{\includegraphics[draft,scale=0.23]{kunen_powerset_18.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_18.png}}%
+    \only<19>{\includegraphics[draft,scale=0.23]{kunen_powerset_19.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_19.png}}%
+    \only<20>{\includegraphics[draft,scale=0.23]{kunen_powerset_20.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_20.png}}%
   \end{center}
@@
 
-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
-\subsection{}
-
-\begin{frame}
-  \frametitle{}
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{}
-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
-\subsection{}
-
-\begin{frame}
-  \frametitle{}
-\end{frame}
+%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% \section{}
+%% %-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
+%% \subsection{}
+%% 
+%% \begin{frame}
+%%   \frametitle{}
+%% \end{frame}
 
@@
     \item<+-> ``Relative functions'' (e.g., $\P^M$, $\card{\cdot}^M$, $\cf^M$).
-    \item<+-> 
     \end{itemize}

### 2020-06-13 IJCAR2020/slides/forcing_IJCAR2020_slides.tex
benefits of ctm approach, moving things to the end

--- 2020-06-12 IJCAR2020/slides/forcing_IJCAR2020_slides.tex
@@
       M,
-      [\only<-4>{\1}\alert{\only<5->{p\hspace{0.2ex}}},\PP,\preceq,\punto{a}]\models
+      [\only<-3>{\1}\alert{\only<4->{p\hspace{0.2ex}}},\PP,\preceq,\punto{a}]\models
       \forceisa(\phi)$.
     \item $\forceisa$ is definable over $M$.\phantom{$\punto{a}).\models \phi\bigr)
-      \iff M,$}\visible<6->{%
+      \iff M,$}\visible<4->{%
       $\overset{\searrow}{}$ \ 
@@
       \pause% 
-      \pause%       
     \item For $env\in\mathsf{list}(M)$,
       \[
-      M[G], \mathrm{map}(\val(G),env) \models \phi)
+      M[G], \mathrm{map}(\val(G),env) \models \phi
       \iff
@@
   \pause
-
+  %
+  \begin{theorem}[\citep{Cohen-CH-PNAS}]
+    If $\PP$ is the set of finite partial binary functions with domain
+    included in $\aleph_2$, $M[G]$ satisfies the negation of the
+    \textbf{Continuum Hypothesis}:
+    \[
+    M[G] \models 2^{\aleph_0} >\aleph_1.
+    \]
+  \end{theorem}
+  \pause
+  %
+  In the ctm approach:
+  \begin{enumerate}
+  \item<+-> Countability ensures that generics exist (by
+    Rasiowa-Sikorski Lemma).
+    %  
+  \item<+-> Absoluteness provide seamless
+    treatment of many concepts.
+    \begin{center}
+      $\alpha$ is an ordinal \ $\iff$ \  $M\models \alpha$ is
+      an ordinal \  $\iff$ \  $M[G]\models \alpha$ is an ordinal
+    \end{center}
+    %
+  \item<+-> Both $M$ and $M[G]$ are standard (two-valued) models.
+  \end{enumerate}
 \end{frame}
@@
     Theorems.
+    \begin{itemize}
+    \item<+-> 
+    \item<+-> 
+    \end{itemize}
   \item<+-> We showed that generic extensions of ctms of $\ZF$ are also
     ctms of $\ZF$ (respectively, adding $\AC$).
-  \item<+-> We provided the forcing notion that adds a Cohen real,
-    therefore proving the existence of a nontrivial extension.
+    \begin{itemize}
+    \item<+-> We may compare some of the code with the actual
+      math \citep{kunen2011set}.
+    \end{itemize}
   \end{enumerate}
@@
 %-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
-\subsection{A map}
+\subsection{Math vs Code}
 
-\begin{frame}
+\begin{frame}<1-12>[label=code-review]
   \begin{center}
@@
 
+%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% \section{}
+%% %-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
+%% \subsection{}
+%% 
+%% \begin{frame}
+%%   \frametitle{}
+%% \end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Looking forward}
+
 \begin{frame}
-  \frametitle{Locale structure} 
+  \frametitle{Looking forward}
+  \begin{block}{Formalizing math}
+    \begin{itemize}
+    \item<+-> Cofinality, K\H{o}nig's Theorem, Shanin's $\Delta$-system Lemma.
+    \item<+-> Forcing notion for adding $\kappa$ Cohen reals.
+    \item<+-> Theorems on preservation of cardinals.
+    \end{itemize}
+  \end{block}
+  %
+  \begin{block}{Technical aids}<+->
+    \begin{itemize}
+    \item<.-> Automatic relativization and proof of absoluteness of concepts.
+    \item<+-> ``Relative functions'' (e.g., $\P^M$, $\card{\cdot}^M$, $\cf^M$).
+    \end{itemize}
+  \end{block}
+\end{frame}
+
+%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
+\begin{frame}
+  \begin{shadowblock}{}
+    \begin{center}
+      {\Huge Thank you!}
+    \end{center}
+  \end{shadowblock}
+\end{frame}
+
+\begin{frame}
+  \frametitle{References}
+  \bibliographystyle{mi-estilo-else}
+  \bibliography{../forcing_in_isabelle_zf}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Extra: Locale structure} 
              {  
@@
 
-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
-\subsection{}
-
 \begin{frame}
-  \frametitle{}
+  \frametitle{Extra: More code review} 
+  We only show the second inclusion $c\sbq \vartheta_G =
+  \val(G,\vartheta)$ (the first one is proved in the
+  course of the 24 folded lines).
 \end{frame}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{}
-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
-\subsection{}
-
-\begin{frame}
-  \frametitle{}
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Looking forward}
-
-\begin{frame}
-  \frametitle{Looking forward}
-  \begin{block}{Formalizing math}
-    \begin{itemize}
-    \item<+-> Cofinality, K\H{o}nig's Theorem, Shanin's $\Delta$-system Lemma.
-    \item<+-> Forcing notion for adding $\kappa$ Cohen reals.
-    \item<+-> Theorems on preservation of cardinals.
-    \end{itemize}
-  \end{block}
-  %
-  \begin{block}{Technical aids}<+->
-    \begin{itemize}
-    \item<.-> Automatic relativization and proof of absoluteness of concepts.
-    \item<+-> ``Relative functions'' (e.g., $\P^M$, $\card{\cdot}^M$, $\cf^M$).
-    \item<+-> 
-    \end{itemize}
-  \end{block}
-\end{frame}
-
-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
-\begin{frame}
-  \begin{shadowblock}{}
-    \begin{center}
-      {\Huge Thank you!}
-    \end{center}
-  \end{shadowblock}
-\end{frame}
-
-\begin{frame}
-  \frametitle{References}
-  \bibliographystyle{mi-estilo-else}
-  \bibliography{../forcing_in_isabelle_zf}
-\end{frame}
+\againframe<12-20>{code-review}
 

### 2020-06-13 IJCAR2020/slides/forcing_IJCAR2020_slides.tex
I keep playing with the code review

@@
 \begin{frame}
-  \frametitle{Paulson's \texttt{ZF-Constructible}}
+  \frametitle{\texttt{ZF-Constructible} \citep{paulson_2003}}
   Let $C :: {}$ \isatt{i} $\Rightarrow$ \isatt{o} be a predicate on sets.
@@
 \begin{frame}<1-12>[label=code-review]
-  \begin{center}
-    \only<01>{\includegraphics[scale=0.23]{kunen_powerset_01.png} \\ \includegraphics[scale=0.26]{isa_converted_01.png}}%
-    \only<02>{\includegraphics[scale=0.23]{kunen_powerset_02.png} \\ \includegraphics[scale=0.26]{isa_converted_02.png}}%
-    \only<03>{\includegraphics[scale=0.23]{kunen_powerset_03.png} \\ \includegraphics[scale=0.26]{isa_converted_03.png}}%
-    \only<04>{\includegraphics[scale=0.23]{kunen_powerset_04.png} \\ \includegraphics[scale=0.26]{isa_converted_04.png}}%
-    \only<05>{\includegraphics[scale=0.23]{kunen_powerset_05.png} \\ \includegraphics[scale=0.26]{isa_converted_05.png}}%
-    \only<06>{\includegraphics[scale=0.23]{kunen_powerset_06.png} \\ \includegraphics[scale=0.26]{isa_converted_06.png}}%
-    \only<07>{\includegraphics[scale=0.23]{kunen_powerset_07.png} \\ \includegraphics[scale=0.26]{isa_converted_07.png}}%
-    \only<08>{\includegraphics[scale=0.23]{kunen_powerset_08.png} \\ \includegraphics[scale=0.26]{isa_converted_08.png}}%
-    \only<09>{\includegraphics[scale=0.23]{kunen_powerset_09.png} \\ \includegraphics[scale=0.26]{isa_converted_09.png}}%
-    \only<10>{\includegraphics[scale=0.23]{kunen_powerset_10.png} \\ \includegraphics[scale=0.26]{isa_converted_10.png}}%
-    \only<11>{\includegraphics[scale=0.23]{kunen_powerset_11.png} \\ \includegraphics[scale=0.26]{isa_converted_11.png}}%
-    \only<12>{\includegraphics[scale=0.23]{kunen_powerset_12.png} \\ \includegraphics[scale=0.26]{isa_converted_12.png}}%
-    \only<13>{\includegraphics[scale=0.23]{kunen_powerset_13.png} \\ \includegraphics[scale=0.26]{isa_converted_13.png}}%
-    \only<14>{\includegraphics[scale=0.23]{kunen_powerset_14.png} \\ \includegraphics[scale=0.26]{isa_converted_14.png}}%
-    \only<15>{\includegraphics[scale=0.23]{kunen_powerset_15.png} \\ \includegraphics[scale=0.26]{isa_converted_15.png}}%
-    \only<16>{\includegraphics[scale=0.23]{kunen_powerset_16.png} \\ \includegraphics[scale=0.26]{isa_converted_16.png}}%
-    \only<17>{\includegraphics[scale=0.23]{kunen_powerset_17.png} \\ \includegraphics[scale=0.26]{isa_converted_17.png}}%
-    \only<18>{\includegraphics[scale=0.23]{kunen_powerset_18.png} \\ \includegraphics[scale=0.26]{isa_converted_18.png}}%
-    \only<19>{\includegraphics[scale=0.23]{kunen_powerset_19.png} \\ \includegraphics[scale=0.26]{isa_converted_19.png}}%
-    \only<20>{\includegraphics[scale=0.23]{kunen_powerset_20.png} \\ \includegraphics[scale=0.26]{isa_converted_20.png}}%
+  \begin{center}\setlength{\fboxsep}{0pt}
+    \only<01>{\includegraphics[scale=0.23]{kunen_powerset_01.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_01.png}}}%
+    \only<02>{\includegraphics[scale=0.23]{kunen_powerset_02.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_02.png}}}%
+    \only<03>{\includegraphics[scale=0.23]{kunen_powerset_03.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_03.png}}}%
+    \only<04>{\includegraphics[scale=0.23]{kunen_powerset_04.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_04.png}}}%
+    \only<05>{\includegraphics[scale=0.23]{kunen_powerset_05.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_05.png}}}%
+    \only<06>{\includegraphics[scale=0.23]{kunen_powerset_06.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_06.png}}}%
+    \only<07>{\includegraphics[scale=0.23]{kunen_powerset_07.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_07.png}}}%
+    \vspace{3.00115pt}\only<08>{\includegraphics[scale=0.23]{kunen_powerset_08.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_08.png}}}%
+    \only<09>{\includegraphics[scale=0.23]{kunen_powerset_09.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_09.png}}}%
+    \only<10>{\includegraphics[scale=0.23]{kunen_powerset_10.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_10.png}}}%
+    \only<11>{\includegraphics[scale=0.23]{kunen_powerset_11.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_11.png}}}%
+    \only<12>{\includegraphics[scale=0.23]{kunen_powerset_12.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_12.png}}}%
+    \only<13>{\includegraphics[scale=0.23]{kunen_powerset_13.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_13.png}}}%
+    \only<14>{\includegraphics[scale=0.23]{kunen_powerset_14.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_14.png}}}%
+    \only<15>{\includegraphics[scale=0.23]{kunen_powerset_15.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_15.png}}}%
+    \only<16>{\includegraphics[scale=0.23]{kunen_powerset_16.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_16.png}}}%
+    \only<17>{\includegraphics[scale=0.23]{kunen_powerset_17.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_17.png}}}%
+    \only<18>{\includegraphics[scale=0.23]{kunen_powerset_18.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_18.png}}}%
+    \only<19>{\includegraphics[scale=0.23]{kunen_powerset_19.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_19.png}}}%
+    \only<20>{\includegraphics[scale=0.23]{kunen_powerset_20.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_20.png}}}%
   \end{center}

### 2020-06-14 IJCAR2020/slides/forcing_IJCAR2020_slides.tex
first re-engineering of slides

--- 2020-06-13 IJCAR2020/slides/forcing_IJCAR2020_slides.tex
@@
       \citep{DBLP:conf/tphol/Wenzel99}.
-    \item Weak metatheory.
+    \item Low in consistency strength.
     \end{itemize}
@@
   \item<+-> An object logic of Isabelle axiomatized over the
-    intuitionistic fragment of HOL \emph{Pure}. 
+    intuitionistic fragment  \emph{Pure} of higher order logic.
   \item<+-> It postulates \alert<3->{two types}: \isatt{i}
@@
 
-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
-\subsection{Paulson's \texttt{ZF-Constructible}}
-
-\begin{frame}
-  \frametitle{Paulson's \texttt{ZF-Constructible}}
-  Let $C :: {}$ \isatt{i} $\Rightarrow$ \isatt{o} be a predicate on sets.
-  \pause
-  \begin{block}{Relativization of $\sbq$}
-    $x \sbq^C y \equiv \forall z.\ C(z) \longrightarrow (z\in x
-    \longrightarrow z\in y)$.
-  \end{block}
-  \pause
-  \begin{block}{$\sbq$ is absolute}
-    $\forall x\ y.\ C(x) \land C(y) \longrightarrow (x \sbq^C y
-    \longleftrightarrow x\sbq y)$.
-  \end{block}
-  \pause 
-  \texttt{ZF-Constructible} develops relativized version of the axioms
-  and proves that many concepts are absolute for \textbf{transitive
-    classes} $C$
-  \[
-  x\in y \land C(y)  \longrightarrow C(x).
-  \]
-  satisfying additional axioms.
-\end{frame}
+%% \begin{frame}
+%%   \frametitle{\texttt{ZF-Constructible} \citep{paulson_2003}}
+%%   Let $C :: {}$ \isatt{i} $\Rightarrow$ \isatt{o} be a predicate on sets.
+%%   \pause
+%%   \begin{block}{Relativization of $\sbq$}
+%%     $x \sbq^C y \equiv \forall z.\ C(z) \longrightarrow (z\in x
+%%     \longrightarrow z\in y)$.
+%%   \end{block}
+%%   \pause
+%%   \begin{block}{$\sbq$ is absolute}
+%%     $\forall x\ y.\ C(x) \land C(y) \longrightarrow (x \sbq^C y
+%%     \longleftrightarrow x\sbq y)$.
+%%   \end{block}
+%%   \pause 
+%%   \texttt{ZF-Constructible} develops relativized version of the axioms
+%%   and proves that many concepts are absolute for \textbf{transitive
+%%     classes} $C$
+%%   \[
+%%   x\in y \land C(y)  \longrightarrow C(x).
+%%   \]
+%%   satisfying additional axioms.
+%% \end{frame}
 
@@
 \begin{frame}
-  \frametitle{The ctm approach to forcing, I}
+  \frametitle{The ctm approach to forcing, 1/4}
   \begin{block}{Countable transitive model (ctm) of $\ZF$}
@@
     \begin{itemize}
-    \item $M$ is countable.
-    \item $E \defi {\in}\restriction M$.
-    \item $x\in y \in M \implies x\in M$.
+    \item \alert<3>{$M$ is \emph{standard}:\quad $E \defi {\in}\restriction M$.}
+    \item $M$ is countable, and \emph{transitive}:\quad $x\in y \in M \implies x\in M$.
     \end{itemize}
@@
   exists a ctm $M$ of $\ZF$.
+  \pause
+
+  It makes sense to compare, for $x,y\in M$:
+  \[
+  x \sbq y
+  \begin{overlayarea}{4.5em}{2ex}
+    \ \ \ \only<3-4>{\ \  and}\only<5>{\alert{$\iff$}}
+  \end{overlayarea}
+  M \models x\sbq y
+  \]
+  \pause
+  The right hand side can be written as a term of type \isatt{o}:
+  $\forall z.\ z\in M \longrightarrow (z\in x \longrightarrow z\in
+  y)$,
+  the \textbf{relativization} of $\sbq$ to $M$. 
+  \pause
+
+  In this case, we know that  $\sbq$  is \textbf{absolute}.
   %
-  \pause
-  %
-  \begin{block}{Forcing notions and generics}
-    \begin{itemize}
-    \item $\lb \PP, {\preceq} ,\1\rb$ preorder with top.
-    \item A filter $G\sbq \PP$ is \textbf{$M$-generic} if it meets
-      every dense $D\sbq \PP$ in $M$.
-    \end{itemize}
-  \end{block}
+  %% \begin{block}{Forcing notions and generics}
+  %%   \begin{itemize}
+  %%   \item $\lb \PP, {\preceq} ,\1\rb$ preorder with top.
+  %%   \item A filter $G\sbq \PP$ is \textbf{$M$-generic} if it meets
+  %%     every dense $D\sbq \PP$ in $M$.
+  %%   \end{itemize}
+  %% \end{block}
 \end{frame}
@@
 \begin{frame}
-  \frametitle{The ctm approach to forcing, II}
-  Given $G\sbq \PP$, we can adjoin it to $M$ to form the
+  \frametitle{The ctm approach to forcing, 2/4}
+  Let $\lb \PP, {\preceq} ,\1\rb \in M$ be a \emph{forcing notion} (a preorder with top). Given 
+  an \emph{$M$-generic} filter $G\sbq \PP$, 
+  we can adjoin it to $M$ to form the
   \textbf{generic extension} $M[G]$.
   \pause
+
   Every $a\in M[G]$ is coded by
-  some $\punto{a}\in M$ through the definable (in $M$) function $\val$:
+  some $\punto{a}\in M$ through the $M$-definable function $\val$:
   \[
@@
 \begin{frame}
-  \frametitle{The ctm approach to forcing, III}
-  By choosing the forcing notion $\lb \PP, {\preceq} ,\1\rb$
-  appropriately one can tune the first order properties of  $M[G]$ (for
-  any generic $G$).
-  \pause
-  %
-  \begin{theorem}[\citep{Cohen-CH-PNAS}]
-    If $\PP$ is the set of finite partial binary functions with domain
-    included in $\aleph_2$, $M[G]$ satisfies the negation of the
-    \textbf{Continuum Hypothesis}:
-    \[
-    M[G] \models 2^{\aleph_0} >\aleph_1.
-    \]
-  \end{theorem}
-  \pause
-  %
-  In the ctm approach:
+  \frametitle{The ctm approach to forcing, 3/4}
+  \begin{block}{Benefits of using ctms}
     \begin{enumerate}
@@
     \item<+-> Both $M$ and $M[G]$ are standard (two-valued) models.
+      %
+    \item<+-> Ctms are
+      used in an important (perhaps the greatest) fraction of the literature.
     \end{enumerate}
+  \end{block}
 \end{frame}
@@
 \begin{frame}
-  \frametitle{Other approaches}
+  \frametitle{The ctm approach to forcing, 4/4}
+  By choosing  $\lb \PP, {\preceq} ,\1\rb$
+  appropriately one can tune the first order properties of  $M[G]$ (for
+  any generic $G$).
+  \pause
+  %
+  \begin{theorem}[\citep{Cohen-CH-PNAS}]
+    If $\PP$ is the set of finite partial binary functions with domain
+    included in $\aleph_2$, $M[G]$ satisfies the negation of the
+    \textbf{Continuum Hypothesis} ($\CH$):
+    \[
+    M[G] \models 2^{\aleph_0} >\aleph_1.
+    \]
+  \end{theorem}
+  \pause
+
+  Formalizing the independence of $\CH$ from the axioms of $\ZFC$
+  using ctms is one of the main goals of our project.
+\end{frame}
+
+%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
+\subsection{Other approaches}
+\begin{frame}
+  \frametitle{Other approaches to set theory and forcing}
   \begin{itemize}
+  \item<+-> \alert{Lean}: Full formalization of the Boolean-valued approach to
+    forcing and the independence of $\CH$ \citep{DBLP:conf/cpp/HanD20}.
   \item<+-> Set theory over Isabelle/HOL:
@@
     \end{itemize}
-  \item<+-> \alert{Lean}: Full formalization of the Boolean-valued approach to
-    forcing and the independence of $\CH$ \citep{DBLP:conf/cpp/HanD20}.
   \end{itemize}
@@
     \begin{itemize}
-    \item<+-> 
-    \item<+-> 
+    \item<+-> $\forceisa$ is defined by induction on internalized
+      formulas.
+    \item<+-> The base cases  $\forceisa(x\in y)$ and  $\forceisa(x =
+      y)$ are  defined by wellfounded
+      recursion, which \alert{must also be internalized}.
     \end{itemize}
@@
 \begin{frame}<1-12>[label=code-review]
-  \begin{center}
-    \only<01>{\includegraphics[scale=0.23]{kunen_powerset_01.png} \\ \includegraphics[scale=0.26]{isa_converted_01.png}}%
-    \only<02>{\includegraphics[scale=0.23]{kunen_powerset_02.png} \\ \includegraphics[scale=0.26]{isa_converted_02.png}}%
-    \only<03>{\includegraphics[scale=0.23]{kunen_powerset_03.png} \\ \includegraphics[scale=0.26]{isa_converted_03.png}}%
-    \only<04>{\includegraphics[scale=0.23]{kunen_powerset_04.png} \\ \includegraphics[scale=0.26]{isa_converted_04.png}}%
-    \only<05>{\includegraphics[scale=0.23]{kunen_powerset_05.png} \\ \includegraphics[scale=0.26]{isa_converted_05.png}}%
-    \only<06>{\includegraphics[scale=0.23]{kunen_powerset_06.png} \\ \includegraphics[scale=0.26]{isa_converted_06.png}}%
-    \only<07>{\includegraphics[scale=0.23]{kunen_powerset_07.png} \\ \includegraphics[scale=0.26]{isa_converted_07.png}}%
-    \only<08>{\includegraphics[scale=0.23]{kunen_powerset_08.png} \\ \includegraphics[scale=0.26]{isa_converted_08.png}}%
-    \only<09>{\includegraphics[scale=0.23]{kunen_powerset_09.png} \\ \includegraphics[scale=0.26]{isa_converted_09.png}}%
-    \only<10>{\includegraphics[scale=0.23]{kunen_powerset_10.png} \\ \includegraphics[scale=0.26]{isa_converted_10.png}}%
-    \only<11>{\includegraphics[scale=0.23]{kunen_powerset_11.png} \\ \includegraphics[scale=0.26]{isa_converted_11.png}}%
-    \only<12>{\includegraphics[scale=0.23]{kunen_powerset_12.png} \\ \includegraphics[scale=0.26]{isa_converted_12.png}}%
-    \only<13>{\includegraphics[scale=0.23]{kunen_powerset_13.png} \\ \includegraphics[scale=0.26]{isa_converted_13.png}}%
-    \only<14>{\includegraphics[scale=0.23]{kunen_powerset_14.png} \\ \includegraphics[scale=0.26]{isa_converted_14.png}}%
-    \only<15>{\includegraphics[scale=0.23]{kunen_powerset_15.png} \\ \includegraphics[scale=0.26]{isa_converted_15.png}}%
-    \only<16>{\includegraphics[scale=0.23]{kunen_powerset_16.png} \\ \includegraphics[scale=0.26]{isa_converted_16.png}}%
-    \only<17>{\includegraphics[scale=0.23]{kunen_powerset_17.png} \\ \includegraphics[scale=0.26]{isa_converted_17.png}}%
-    \only<18>{\includegraphics[scale=0.23]{kunen_powerset_18.png} \\ \includegraphics[scale=0.26]{isa_converted_18.png}}%
-    \only<19>{\includegraphics[scale=0.23]{kunen_powerset_19.png} \\ \includegraphics[scale=0.26]{isa_converted_19.png}}%
-    \only<20>{\includegraphics[scale=0.23]{kunen_powerset_20.png} \\ \includegraphics[scale=0.26]{isa_converted_20.png}}%
+  \begin{center}\setlength{\fboxsep}{0pt}
+    \only<01>{\includegraphics[draft,scale=0.23]{kunen_powerset_01.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_01.png}}}%
+    \only<02>{\includegraphics[draft,scale=0.23]{kunen_powerset_02.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_02.png}}}%
+    \only<03>{\includegraphics[draft,scale=0.23]{kunen_powerset_03.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_03.png}}}%
+    \only<04>{\includegraphics[draft,scale=0.23]{kunen_powerset_04.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_04.png}}}%
+    \only<05>{\includegraphics[draft,scale=0.23]{kunen_powerset_05.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_05.png}}}%
+    \only<06>{\includegraphics[draft,scale=0.23]{kunen_powerset_06.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_06.png}}}%
+    \only<07>{\includegraphics[draft,scale=0.23]{kunen_powerset_07.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_07.png}}}%
+    \vspace{3.00115pt}\only<08>{\includegraphics[draft,scale=0.23]{kunen_powerset_08.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_08.png}}}%
+    \only<09>{\includegraphics[draft,scale=0.23]{kunen_powerset_09.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_09.png}}}%
+    \only<10>{\includegraphics[draft,scale=0.23]{kunen_powerset_10.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_10.png}}}%
+    \only<11>{\includegraphics[draft,scale=0.23]{kunen_powerset_11.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_11.png}}}%
+    \only<12>{\includegraphics[draft,scale=0.23]{kunen_powerset_12.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_12.png}}}%
+    \only<13>{\includegraphics[draft,scale=0.23]{kunen_powerset_13.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_13.png}}}%
+    \only<14>{\includegraphics[draft,scale=0.23]{kunen_powerset_14.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_14.png}}}%
+    \only<15>{\includegraphics[draft,scale=0.23]{kunen_powerset_15.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_15.png}}}%
+    \only<16>{\includegraphics[draft,scale=0.23]{kunen_powerset_16.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_16.png}}}%
+    \only<17>{\includegraphics[draft,scale=0.23]{kunen_powerset_17.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_17.png}}}%
+    \only<18>{\includegraphics[draft,scale=0.23]{kunen_powerset_18.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_18.png}}}%
+    \only<19>{\includegraphics[draft,scale=0.23]{kunen_powerset_19.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_19.png}}}%
+    \only<20>{\includegraphics[draft,scale=0.23]{kunen_powerset_20.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_20.png}}}%
   \end{center}
@@
 \begin{frame}
-  \frametitle{Extra: Locale structure} 
+  \frametitle{Extra: Locale structure involving set models} 
              {  
@@
                  \texttt{forcing{\uscore}data} & =  & \texttt{M{\uscore}ctm} \alert{+}
-                 \texttt{forcing{\uscore}notion} in $M$\\
+                 \texttt{forcing{\uscore}notion} $\PP\in M$\\
                  \texttt{separative{\uscore}notion} & = &
@@
                  \texttt{M{\uscore}ctm{\uscore}separative} & = &
-                 \texttt{forcing{\uscore}data} \alert{+} \texttt{separative{\uscore}notion}
+                 \texttt{forcing{\uscore}data} \alert{+}
+                 \texttt{separative{\uscore}notion} \\
+                 \texttt{G{\uscore}generic} & = & \texttt{forcing{\uscore}data} \alert{+} $G$ is $M$-generic
                \end{tabular} 
@@
 
-\againframe<12-20>{code-review}
+\againframe<13-20>{code-review}
 
@@
 %%
-%% - ctm approach: bondades 
-%% 
-%% * absolutez de ordinal, etc
-%% * Modelos a dos valores.
-%% 
-%% - no definir locales, hablar menos de la arquitectura modular
-%% (no tanto del codigo en si)
-%% 
-%% - Pruebas mostrar:
-%% 
-%% density_mem ?
-%% powerset?
-%% choice?
 

### 2020-06-17 IJCAR2020/slides/forcing_IJCAR2020_slides.tex
full account of ZF-Constructible, more details on achievements

@@
   \pause
-  The right hand side can be written as a term of type \isatt{o}:
-  $\forall z.\ z\in M \longrightarrow (z\in x \longrightarrow z\in
+  The rhs can be written as a term of type \isatt{o}:
+
+  {} \hfill   $\forall z.\ z\in M \longrightarrow (z\in x \longrightarrow z\in
   y)$,
-  the \textbf{relativization} of $\sbq$ to $M$. 
+  \hfill the \textbf{relativization} $\sbq^M$ of $\sbq$ to $M$. 
   \pause
 
-  In this case, we know that  $\sbq$  is \textbf{absolute}.
+  In this case, we know that  $\sbq$  is \textbf{absolute for
+    transitive models}.
   %
@@
   \]
-  \emph{Fundamentally}, \textbf{truth} in $M[G]$ is coded in $M$ by the
+  Fundamentally, \textbf{truth} in $M[G]$ is coded in $M$ by the
   function $\forceisa$.
@@
       %  
-    \item<+-> Absoluteness provide seamless
+    \item<+-> Absoluteness provides seamless
       treatment of many concepts.
@@
     If $\PP$ is the set of finite partial binary functions with domain
-    included in $\aleph_2$, $M[G]$ satisfies the negation of the
+    included in $\aleph_2^M$, $M[G]$ satisfies the negation of the
     \textbf{Continuum Hypothesis} ($\CH$):
@@
     \begin{enumerate}
-    \item<1-> We adapted \texttt{ZF-Constructible} to obtain
-      absoluteness results for nonempty transitive classes ($\leftarrow$
-      \alert{in Isabelle2020}).
-
-      \visible<2->{
-        Around 40 absoluteness/closure lemmas now hold using weaker
-        hypotheses on the class $C$ (most of them, just that $C$ is
-        transitive and nonempty).
-      }
-    \item<3-> We formalized the formula transformer $\forceisa$ and
+    \item<1-> We adapted the \texttt{ZF-Constructible} library
+      \citep{paulson_2003} to obtain
+      absoluteness results for nonempty transitive classes ($\to$
+      \alert{Isabelle2020}).
+    \item<2-> We formalized the formula transformer $\forceisa$ and
       hence the forcing relation $\forces$, and proved the Fundamental
       Theorems.
-    \item<4-> We showed that generic extensions of ctms of $\ZF$ are also
+    \item<3-> We showed that generic extensions of ctms of $\ZF$ are also
       ctms of $\ZF$ (respectively, adding $\AC$).
-    \item<5-> We provided the forcing notion that adds a Cohen real,
+    \item<4-> We provided the forcing notion that adds a Cohen real,
       therefore proving the existence of a nontrivial extension.
@@
   \frametitle{Some details}
+  \begin{enumerate}%\setcounter{enumi}{1}
+  \item We adapted \texttt{ZF-Constructible} to obtain
+    sharper absoluteness results.
+  \end{enumerate}
+
+  This library contains a major development of relativization and
+  absoluteness for \emph{classes} $C :: {}$ \isatt{i} $\Rightarrow$
+  \isatt{o}, including the definition of the set \isatt{formula} ::
+  \isatt{i} of internal formulas and model theoretic satisfaction. 
+  \pause
+  \begin{block}{Relativization and synthesis discipline}
+    \begin{enumerate}[I]
+    \item<+-> $p=\{x,y\} :: {}$\isatt{i}
+      \hfill
+      (original term).
+    \item<+-> \isatt{upair}$(C,x,y,p) :: {}$\isatt{o}
+      \hfill
+      (relativization, fully
+      relational). 
+    \item<+-> \isatt{upair{\uscore}fm}$(0,1,2) :: {}$\isatt{i} 
+      \hfill
+      (synthesized member of \isatt{formula}).
+    \end{enumerate}
+  \end{block}
+  \begin{alertblock}{}<+->
+    Around 40 absoluteness/closure lemmas now hold using weaker
+    hypotheses on the class $C$ (most of them, just that $C$ is
+    transitive and nonempty).
+  \end{alertblock}
+\end{frame}
+
+\begin{frame}{Some details}
   \begin{enumerate}\setcounter{enumi}{1}
@@
     Theorems.
-    \begin{itemize}
+  \end{enumerate}
+  \begin{itemize}[I]
   \item<+-> $\forceisa$ is defined by induction on internalized
@@
   \end{itemize}
-  \item<+-> We showed that generic extensions of ctms of $\ZF$ are also
-    ctms of $\ZF$ (respectively, adding $\AC$).
-    \begin{itemize}
-    \item<+-> We may compare some of the code with the actual
+\end{frame}
+
+\begin{frame}{Some details}
+  \begin{enumerate}\setcounter{enumi}{2}
+  \item For any ctm $M$ of $\ZF$ and $M$-generic $G$, $M[G] \models
+    \ZF$ (respectively, adding $\AC$).
+  \end{enumerate}
+  \pause
+  Actually, the modularity of the theory of forcing allowed us to do
+  this \alert{before} we write down the definition of $\forceisa$.
+  \pause
+
+  We may compare some of the code with the actual
   math \citep{kunen2011set}.
-    \end{itemize}
-  \end{enumerate}
 \end{frame}

### 2020-06-17 IJCAR2020/slides/forcing_IJCAR2020_slides.tex
almost final version of slides

@@
       etc).
-    \item Untyped, and too weak a metatheory.
+    \item ``Untyped'', and too weak a metatheory.
     \end{itemize}
@@
   \begin{itemize}
-  \item<+-> An object logic of Isabelle axiomatized over the
+  \item<.-> An object logic of Isabelle axiomatized over the
     intuitionistic fragment  \emph{Pure} of higher order logic.
-  \item<+-> It postulates \alert<3->{two types}: \isatt{i}
+  \item<+-> It postulates two types: \isatt{i}
     (sets) and \isatt{o} (booleans). \visible<3->{\alert{Not
       inductively defined!}}
-    \pause
-  \item<+-> More stuff.
+  \item<+-> The Replacement and Separation axiom schemes feature free
+    high order variables.
+  \item<+-> Induction/recursion is \emph{internal} to the theory (it works as a
+    layer on top of set-theoretical proofs of well-foundedness (of
+    $\N$, of $\Ord$, etc).
   \end{itemize}
 \end{frame}
-
-%% \begin{frame}
-%%   \frametitle{\texttt{ZF-Constructible} \citep{paulson_2003}}
-%%   Let $C :: {}$ \isatt{i} $\Rightarrow$ \isatt{o} be a predicate on sets.
-%%   \pause
-%%   \begin{block}{Relativization of $\sbq$}
-%%     $x \sbq^C y \equiv \forall z.\ C(z) \longrightarrow (z\in x
-%%     \longrightarrow z\in y)$.
-%%   \end{block}
-%%   \pause
-%%   \begin{block}{$\sbq$ is absolute}
-%%     $\forall x\ y.\ C(x) \land C(y) \longrightarrow (x \sbq^C y
-%%     \longleftrightarrow x\sbq y)$.
-%%   \end{block}
-%%   \pause 
-%%   \texttt{ZF-Constructible} develops relativized version of the axioms
-%%   and proves that many concepts are absolute for \textbf{transitive
-%%     classes} $C$
-%%   \[
-%%   x\in y \land C(y)  \longrightarrow C(x).
-%%   \]
-%%   satisfying additional axioms.
-%% \end{frame}
 
@@
   %
-  \textbf{Note}. If $\lb N, R\rb \models \ZF$ with $R$ wellfounded, then there
+  \textbf{Note}. If $\lb N, R\rb \models \ZF$ with $R$ well-founded, then there
   exists a ctm $M$ of $\ZF$.
@@
     transitive models}.
-  %
-  %% \begin{block}{Forcing notions and generics}
-  %%   \begin{itemize}
-  %%   \item $\lb \PP, {\preceq} ,\1\rb$ preorder with top.
-  %%   \item A filter $G\sbq \PP$ is \textbf{$M$-generic} if it meets
-  %%     every dense $D\sbq \PP$ in $M$.
-  %%   \end{itemize}
-  %% \end{block}
 \end{frame}
@@
   Every $a\in M[G]$ is coded by
-  some $\punto{a}\in M$ through the $M$-definable function $\val$:
+  some $\punto{a}\in M$ through the function $\val$:
   \[
@@
     \begin{enumerate}
-    \item $\bigl(\forall G\visible<4->{\alert{\ni p}}.\quad M[G], [\val(G,\punto{a})]
-      \models \phi\bigr)
-      \iff
-      M,
-      [\only<-3>{\1}\alert{\only<4->{p\hspace{0.2ex}}},\PP,\preceq,\punto{a}]\models
-      \forceisa(\phi)$.
-    \item $\forceisa$ is definable over $M$.\phantom{$\punto{a}).\models \phi\bigr)
-      \iff M,$}\visible<4->{%
-      $\overset{\searrow}{}$ \ 
-      $\alert{p} \forces^M_{\PP,\preceq} \phi(\punto{a})$
-      \  $\overset{\swarrow}{}$}
-      \pause% 
-      \pause% 
-    \item For $env\in\mathsf{list}(M)$,
+    \item $\forceisa$ is definable over $M$.
+      %
+    \item For every $\phi$, $M$-generic $G$, and $\punto{a}\in M$,
       \[
-      M[G], \mathrm{map}(\val(G),env) \models \phi
-      \iff
-      \exists p\in G.\ M, [p,\preceq,\PP] \mathbin{{+}\!\!\!\!{+}} env \models
+      M[G], [\val(G,\punto{a})] \models \phi
+      \quad      \iff  \quad 
+      \exists p\in G.\ \; M, [p,\preceq,\PP, \punto{a}]  \models
       \forceisa(\phi).
       \]
+      \pause
+      \rule{21em}{0pt}
+      \raisebox{1.6ex}[0ex]{$\overset{\searrow}{}$ \ \ 
+        $p \alert{\forces^M_{\PP,\preceq}} \phi(\punto{a})$
+        \  \ $\overset{\swarrow}{}$}
     \end{enumerate}
@@
   \begin{enumerate}\setcounter{enumi}{1}
-  \item<+-> We formalized the formula transformer $\forceisa$ and
+  \item We formalized the formula transformer $\forceisa$ and
     hence the forcing relation $\forces$, and proved the Fundamental
@@
   \end{enumerate}
-  \begin{itemize}[I]
-  \item<+-> $\forceisa$ is defined by induction on internalized
-    formulas.
-  \item<+-> The base cases  $\forceisa(x\in y)$ and  $\forceisa(x =
-    y)$ are  defined by wellfounded
-    recursion, which \alert{must also be internalized}.
-  \end{itemize}
+  %
+  \pause
+  %
+  \begin{block}{$\forceisa$ is defined by recursion on  formulas}
+    This is the main reason we work with the set of internalized
+    formulas, and that we require legit first-order expressions for
+    the axiom schemes (Separation and Replacement).
+
+    The base cases  $\forceisa(x\in y)$ and  $\forceisa(x =
+    y)$ are  defined by \alert{internal} well-founded
+    recursion.
+  \end{block}
+  %
+  \pause
+  \begin{alertblock}{}
+    We enhanced the recursion results of Isabelle/ZF as well as
+    the relevant preservation results in ZF-Constructible, thus
+    showing that forcing is absolute for atomic formulas.
+  \end{alertblock}
 \end{frame}
@@
   \begin{center}\setlength{\fboxsep}{0pt}
-    \only<01>{\includegraphics[draft,scale=0.23]{kunen_powerset_01.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_01.png}}}%
-    \only<02>{\includegraphics[draft,scale=0.23]{kunen_powerset_02.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_02.png}}}%
-    \only<03>{\includegraphics[draft,scale=0.23]{kunen_powerset_03.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_03.png}}}%
-    \only<04>{\includegraphics[draft,scale=0.23]{kunen_powerset_04.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_04.png}}}%
-    \only<05>{\includegraphics[draft,scale=0.23]{kunen_powerset_05.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_05.png}}}%
-    \only<06>{\includegraphics[draft,scale=0.23]{kunen_powerset_06.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_06.png}}}%
-    \only<07>{\includegraphics[draft,scale=0.23]{kunen_powerset_07.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_07.png}}}%
-    \vspace{3.00115pt}\only<08>{\includegraphics[draft,scale=0.23]{kunen_powerset_08.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_08.png}}}%
-    \only<09>{\includegraphics[draft,scale=0.23]{kunen_powerset_09.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_09.png}}}%
-    \only<10>{\includegraphics[draft,scale=0.23]{kunen_powerset_10.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_10.png}}}%
-    \only<11>{\includegraphics[draft,scale=0.23]{kunen_powerset_11.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_11.png}}}%
-    \only<12>{\includegraphics[draft,scale=0.23]{kunen_powerset_12.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_12.png}}}%
-    \only<13>{\includegraphics[draft,scale=0.23]{kunen_powerset_13.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_13.png}}}%
-    \only<14>{\includegraphics[draft,scale=0.23]{kunen_powerset_14.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_14.png}}}%
-    \only<15>{\includegraphics[draft,scale=0.23]{kunen_powerset_15.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_15.png}}}%
-    \only<16>{\includegraphics[draft,scale=0.23]{kunen_powerset_16.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_16.png}}}%
-    \only<17>{\includegraphics[draft,scale=0.23]{kunen_powerset_17.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_17.png}}}%
-    \only<18>{\includegraphics[draft,scale=0.23]{kunen_powerset_18.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_18.png}}}%
-    \only<19>{\includegraphics[draft,scale=0.23]{kunen_powerset_19.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_19.png}}}%
-    \only<20>{\includegraphics[draft,scale=0.23]{kunen_powerset_20.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_20.png}}}%
+    \only<01>{\includegraphics[scale=0.23]{kunen_powerset_01.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_01.png}}}%
+    \only<02>{\includegraphics[scale=0.23]{kunen_powerset_02.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_02.png}}}%
+    \only<03>{\includegraphics[scale=0.23]{kunen_powerset_03.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_03.png}}}%
+    \only<04>{\includegraphics[scale=0.23]{kunen_powerset_04.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_04.png}}}%
+    \only<05>{\includegraphics[scale=0.23]{kunen_powerset_05.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_05.png}}}%
+    \only<06>{\includegraphics[scale=0.23]{kunen_powerset_06.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_06.png}}}%
+    \only<07>{\includegraphics[scale=0.23]{kunen_powerset_07.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_07.png}}}%
+    \vspace{3.00115pt}\only<08>{\includegraphics[scale=0.23]{kunen_powerset_08.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_08.png}}}%
+    \only<09>{\includegraphics[scale=0.23]{kunen_powerset_09.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_09.png}}}%
+    \only<10>{\includegraphics[scale=0.23]{kunen_powerset_10.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_10.png}}}%
+    \only<11>{\includegraphics[scale=0.23]{kunen_powerset_11.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_11.png}}}%
+    \only<12>{\includegraphics[scale=0.23]{kunen_powerset_12.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_12.png}}}%
+    \only<13>{\includegraphics[scale=0.23]{kunen_powerset_13.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_13.png}}}%
+    \only<14>{\includegraphics[scale=0.23]{kunen_powerset_14.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_14.png}}}%
+    \only<15>{\includegraphics[scale=0.23]{kunen_powerset_15.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_15.png}}}%
+    \only<16>{\includegraphics[scale=0.23]{kunen_powerset_16.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_16.png}}}%
+    \only<17>{\includegraphics[scale=0.23]{kunen_powerset_17.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_17.png}}}%
+    \only<18>{\includegraphics[scale=0.23]{kunen_powerset_18.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_18.png}}}%
+    \only<19>{\includegraphics[scale=0.23]{kunen_powerset_19.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_19.png}}}%
+    \only<20>{\includegraphics[scale=0.23]{kunen_powerset_20.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_20.png}}}%
   \end{center}
@@
     \begin{itemize}
-    \item<+-> Cofinality, K\H{o}nig's Theorem, Shanin's $\Delta$-system Lemma.
-    \item<+-> Forcing notion for adding $\kappa$ Cohen reals.
-    \item<+-> Theorems on preservation of cardinals.
+    \item Cofinality, K\H{o}nig's Theorem, Shanin's $\Delta$-system Lemma.
+    \item Forcing notion for adding $\kappa$ Cohen reals.
+    \item Theorems on preservation of cardinals.
     \end{itemize}
   \end{block}
-  %
-  \begin{block}{Technical aids}<+->
+  \pause
+  \begin{block}{Technical aids}
     \begin{itemize}
-    \item<.-> Automatic relativization and proof of absoluteness of concepts.
-    \item<+-> ``Relative functions'' (e.g., $\P^M$, $\card{\cdot}^M$, $\cf^M$).
+    \item Automatic relativization and proof of absoluteness of concepts.
+    \item ``Relative functions'' (e.g., $\P^M$, $\card{\cdot}^M$, $\cf^M$).
     \end{itemize}

### 2020-06-17 IJCAR2020/slides/forcing_IJCAR2020_slides.tex
highliting in slides, script almost complete
slides: minor. script: ZF-Constructible

@@
 \begin{document}
-%\nobibliography*
 
@@
     \item Most advanced set theory formalized (around 2017).
-      % \item Closer to the mathematical practice of forcing.
     \item Structured proof language Isar
       \citep{DBLP:conf/tphol/Wenzel99}.
-    \item Low in consistency strength.
+    \item Comparatively low in consistency strength.
     \end{itemize}
@@
 
-%% \begin{frame}
-%%   \frametitle{A word on consistency}
-%% \end{frame}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
   \frametitle{Some details}
+  \begin{alertblock}{}
     \begin{enumerate}%\setcounter{enumi}{1}
@@
     \end{enumerate}
+  \end{alertblock}
 
-  This library contains a major development of relativization and
+  This library originally contains a major development of relativization and
   absoluteness for \emph{classes} $C :: {}$ \isatt{i} $\Rightarrow$
@@
 \begin{frame}{Some details}
+  \begin{alertblock}{}
     \begin{enumerate}\setcounter{enumi}{1}
@@
     \end{enumerate}
+  \end{alertblock}
   %
@@
   %
-  \begin{block}{$\forceisa$ is defined by recursion on  formulas}
+  \begin{block}{$\;\forceisa$ is defined by recursion on  formulas}
     This is the main reason we work with the set of internalized
@@
 \begin{frame}{Some details}
+  \begin{alertblock}{}
     \begin{enumerate}\setcounter{enumi}{2}
@@
     \end{enumerate}
+  \end{alertblock}
   \pause
@@
 \end{frame}
-
-%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% \section{}
-%% %-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
-%% \subsection{}
-%% 
-%% \begin{frame}
-%%   \frametitle{}
-%% \end{frame}
 
@@
                  \texttt{forcing{\uscore}notion} & = & preorder $\PP$ with top. \\
-                 %    \texttt{M{\uscore}ZF} & = & set model of the $\ZF$ axioms \\ 
                  \texttt{M{\uscore}ZF{\uscore}trans} & = & set model $M$ of the $\ZF$
@@
 
-%% - Más bola a la metateoría (ZF).
-%% 
-%% * es weak ---> falta de inducción/recursión, y el objeto
-%% central (forces) se define así.
-%% * too strong ---> not possible to have finitary consistency proofs
-%%
-
 %%% Local Variables: 

### 2020-06-26 IJCAR2020/slides/forcing_IJCAR2020_slides.tex
submitted version of talk. Extra slide in code review
minor adjustments to talk

@@
   \item<+-> Induction/recursion is \emph{internal} to the theory (it works as a
-    layer on top of set-theoretical proofs of well-foundedness (of
+    layer on top of set-theoretical proofs of well-foundedness--- of
     $\N$, of $\Ord$, etc).
@@
 
-\begin{frame}<1-12>[label=code-review]
+\begin{frame}<1-10>[label=code-review]
   \begin{center}\setlength{\fboxsep}{0pt}
@@
 \end{frame}
+
+\begin{frame}
+  \begin{center}\setlength{\fboxsep}{0pt}
+    \vspace{3.00115pt}\includegraphics[scale=0.23]{kunen_powerset_10.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_10_bis.png}}%
+  \end{center}
+\end{frame}
+
+\againframe<11-12>{code-review}
 

### 2020-11-27 Delta_System_Lemma/document/root.tex
renombrando "Edwin"
document preparation for Cofinality and Delta System Lemma

--- 2018-11-27 src/document/root.tex
@@
 
+\input{header-delta-system}
+
 % this should be the last package used
@@
 
+\renewcommand{\isacharunderscorekeyword}{\mbox{\_}}
+\renewcommand{\isacharunderscore}{\mbox{\_}}
+\renewcommand{\isasymtturnstile}{\isamath{\Vdash}}
+\renewcommand{\isacharminus}{-}
+
 \begin{document}
 
-\title{src}
-\author{By miguel}
+\title{Cofinality and the Delta System Lemma}
+\author{
+  Pedro S\'anchez Terraf\thanks{Universidad Nacional de C\'ordoba. 
+    Facultad de Matem\'atica, Astronom\'{\i}a,  F\'{\i}sica y
+    Computaci\'on.}
+  \thanks{%
+    Centro de Investigaci\'on y Estudios de Matem\'atica
+    (CIEM-FaMAF), Conicet. C\'ordoba. Argentina.
+    Supported by Secyt-UNC project 33620180100465CB.}
+}
 \maketitle
+
+\begin{abstract}
+  We formalize the basic results on cofinality of linearly ordered
+  sets and ordinals and \v{S}anin's Lemma for uncountable families of
+  finite sets. We work in the set theory framework of
+  Isabelle/ZF, using the Axiom of Choice as needed.
+\end{abstract}
 

### 2020-11-30 src/document/root.tex
Fixed latex compilation

@@
 \usepackage[numbers]{natbib}
+\usepackage{babel}
 
@@
 \title{Formalization of Forcing in Isabelle/ZF}
-\author{Emmanuel Gunther
+\author{Emmanuel Gunther\thanks{Universidad Nacional de C\'ordoba. 
+    Facultad de Matem\'atica, Astronom\'{\i}a,  F\'{\i}sica y
+    Computaci\'on.}
   \and
-  Miguel Pagano
+  Miguel Pagano\footnotemark[1]
   \and
-  Pedro S\'anchez Terraf
+  Pedro S\'anchez Terraf\footnotemark[1] \thanks{Centro de Investigaci\'on y Estudios de Matem\'atica
+    (CIEM-FaMAF), Conicet. C\'ordoba. Argentina.
+    Supported by Secyt-UNC project 33620180100465CB.}
 }
@@
 
+A cross-linked HTML version of the development can be found at
+\url{https://cs.famaf.unc.edu.ar/~pedro/forcing/}.
+
 % generated text of all theories

### 2020-12-26 Delta_System_Lemma/document/root.tex
better document sub/super-scripts
introduction text to DSL

@@
 \usepackage[numbers]{natbib}
+
+\usepackage{relsize}
+\DeclareRobustCommand{\isactrlbsub}{\emph\bgroup\math{}\sb\bgroup\mbox\bgroup\isaspacing\itshape\smaller}
+\DeclareRobustCommand{\isactrlesub}{\egroup\egroup\endmath\egroup}
+\DeclareRobustCommand{\isactrlbsup}{\emph\bgroup\math{}\sp\bgroup\mbox\bgroup\isaspacing\itshape\smaller}
+\DeclareRobustCommand{\isactrlesup}{\egroup\egroup\endmath\egroup}
 
@@
 % urls in roman style, theory text in math-similar italics
-\urlstyle{rm}
+\urlstyle{tt}
 \isabellestyle{it}
@@
 
+\section{Introduction}
+
+The session we present gathers very basic results built on the set
+theory formalization of Isabelle/ZF
+\cite{DBLP:journals/jar/PaulsonG96}. In a sense, some of the material
+formalized here corresponds to a natural continuation of that
+work. This is even clearer after perusing Section~\ref{sec:zf-lib},
+where notions like cardinal exponentiation are first defined, together
+with various lemmas that do not depend on the Axiom of Choice ($\AC$);
+the same holds for the basic theory of cofinality of ordinals, which
+is developed in Section~\ref{sec:cofinality}. In
+Section~\ref{sec:cardinal-lib}, (un)countability is defined
+and several results proved, now using $\AC$ freely; the latter is also
+needed to prove König's Theorem on cofinality of cardinal
+exponentiation. Finally, the simplest infinitary version of the Delta
+System Lemma (DSL, also known as the ``Sunflower Lemma'') due to  \v{S}anin
+is proved in Section~\ref{sec:dsl}.
+
+A greater part of this development was motivated by a joint project on
+the formalization of the ctm approach to
+forcing~\cite{2020arXiv200109715G} by Gunther, Pagano, Steinberg, and
+the author. Indeed, most of the results presented here are required
+for the development of forcing; in particular, DSL is used in the
+proof that Cohen posets satisfy \emph{countable chain condition}, which
+in turn is used to characterize cardinals in generic extensions
+\cite[Chap.~IV]{kunen2011set}. As it turns out, the material as
+formalized presently will not be part of the forcing formalization,
+since for that goal we need relativized versions of both the
+concepts and the proofs.
+
 % generated text of all theories

### 2021-05-28 IJCAR2020/slides/forcing_IJCAR2020_slides.tex

@@
   \begin{theorem}[\citep{Cohen-CH-PNAS}]
-    \begin{enumerate}
-    \item $\forceisa$ is definable over $M$.
-      %
-    \item For every $\phi$, $M$-generic $G$, and $\punto{a}\in M$,
+    There exists a formula-transformer  $\forceisa$
+    such that for every $\phi$, $M$-generic $G$, and $\punto{a}\in M$,
       \[
@@
         \  \ $\overset{\swarrow}{}$}
-    \end{enumerate}
   \end{theorem} 
@@
     \item<+-> Ctms are
-      used in an important (perhaps the greatest) fraction of the literature.
+      used in an important fraction of the literature.
     \end{enumerate}
@@
   \end{shadowblock}
+  \transduration{2}
 \end{frame}
@@
   \bibliography{../forcing_in_isabelle_zf}
+  \transduration{5}
 \end{frame}

### 2021-12-04 src/document/root.tex
better rendering of sub/superindexes at document
filling the gaps in Definitions_Main
minor edits for document; hiding FIXME and TODO from view

@@
 \usepackage{babel}
+
+\usepackage{relsize}
+\DeclareRobustCommand{\isactrlbsub}{\emph\bgroup\math{}\sb\bgroup\mbox\bgroup\isaspacing\itshape\smaller}
+\DeclareRobustCommand{\isactrlesub}{\egroup\egroup\endmath\egroup}
+\DeclareRobustCommand{\isactrlbsup}{\emph\bgroup\math{}\sp\bgroup\mbox\bgroup\isaspacing\itshape\smaller}
+\DeclareRobustCommand{\isactrlesup}{\egroup\egroup\endmath\egroup}
 
@@
 \newcommand{\ZFC}{\axiomas{ZFC}}
+\newcommand{\AC}{\axiomas{AC}}
+\newcommand{\CH}{\axiomas{CH}}
 
@@
     Supported by Secyt-UNC project 33620180100465CB.}
+  \and
+  Mat\'{\i}as Steinberg\footnotemark[1]
 }
@@
 \begin{abstract}
-  We formalize the theory of forcing in the set theory framework of
+  We redeveloped our formalization of forcing in the set theory framework of
   Isabelle/ZF. Under the assumption of the existence of a countable
-  transitive model of $\ZFC$, we construct a proper generic extension and show
-  that the latter also satisfies $\ZFC$.
+  transitive model of ZFC, we construct proper generic extensions 
+  that satisfy the Continuum Hypothesis and its negation.
 \end{abstract}
@@
 
-We have improved several aspects of our development before submitting
-it to the AFP:
-\begin{enumerate}
-\item Our session \isatt{Forcing} depends on the new release of
-  \isatt{ZF-Constructible}.
-\item We streamlined the commands for synthesizing renames and formulas.
-\item The command that synthesizes formulas produces the lemmas for
-  them (the synthesized term is a formula and the equivalence between
-  the satisfaction of the synthesized term and the relativized term).
-\item Consistently use of structured proofs using Isar (except for one
-  coming from a schematic goal command).
-\end{enumerate}
-
-A cross-linked HTML version of the development can be found at
+Previous versions of this development can be found at
 \url{https://cs.famaf.unc.edu.ar/~pedro/forcing/}.

### 2022-02-25 Transitive_Models/document/root.tex
Split of the development in two sessions.

--- 2021-12-04 src/document/root.tex
@@
 %\renewcommand{\isastyle}{\isastyleminor}
-
+\newcommand{\forces}{\Vdash}
+\newcommand{\dom}{\mathsf{dom}}
 \renewcommand{\isacharunderscorekeyword}{\mbox{\_}}
@@
 \newcommand{\ZFC}{\axiomas{ZFC}}
+\newcommand{\ZF}{\axiomas{ZF}}
 \newcommand{\AC}{\axiomas{AC}}
@@
 
-\title{Formalization of Forcing in Isabelle/ZF}
+\title{Transitive Models of ZF}
 \author{Emmanuel Gunther\thanks{Universidad Nacional de C\'ordoba. 
@@
 \section{Introduction}
-We formalize the theory of forcing. We work on top of the Isabelle/ZF
-framework developed by \citet{DBLP:journals/jar/PaulsonG96}. Our
-mechanization is described in more detail in our papers
-\cite{2018arXiv180705174G} (LSFA 2018), \cite{2019arXiv190103313G},
-and \cite{2020arXiv200109715G} (IJCAR 2020).
-
-\subsection*{Release notes}
-\label{sec:release-notes}
-
-Previous versions of this development can be found at
-\url{https://cs.famaf.unc.edu.ar/~pedro/forcing/}.
 
@@
 \input{session}
-
-% optional bibliography
-\bibliographystyle{root}
-\bibliography{root}
 

### 2022-02-26 Transitive_Models/document/root.tex
Correcting some section titles. Minor edition in abstract of document

@@
 \begin{abstract}
-  We redeveloped our formalization of forcing in the set theory framework of
-  Isabelle/ZF. Under the assumption of the existence of a countable
-  transitive model of ZFC, we construct proper generic extensions 
-  that satisfy the Continuum Hypothesis and its negation.
+  We extend the ZF-Constructibility session, by relativizing several
+  theories of Isabelle/ZF to a transitive class. This work is a
+  prerrequisite to our formalization of the independence of the
+  Continuum Hypothesis.
 \end{abstract}

### 2022-02-26 Transitive_Models/document/root.tex
Use of countability highlighted. Merging abstracts and titles
Brief description of how porting affected theories

@@
 \renewcommand{\isacharminus}{-}
+\newcommand{\session}[1]{\textit{#1}}
+\newcommand{\theory}[1]{\texttt{#1}}
 \newcommand{\axiomas}[1]{\mathit{#1}}
@@
 
-\title{Transitive Models of ZF}
+\title{Transitive Models of Fragments of ZF}
 \author{Emmanuel Gunther\thanks{Universidad Nacional de C\'ordoba. 
@@
 \begin{abstract}
-  We extend the ZF-Constructibility session, by relativizing several
-  theories of Isabelle/ZF to a transitive class. This work is a
+  We extend the ZF-Constructibility library by relativizing theories
+  of the Isabelle/ZF and Delta System Lemma sessions to a transitive
+  class. We also relativize Paulson's work on Aleph and our former
+  treatment of the Axiom of Dependent Choices. This work is a
   prerrequisite to our formalization of the independence of the
@@
 
+As it was explained in \cite[Sect.~3]{2020arXiv200109715G} and elsewhere,
+relativization of concepts is a key tool to obtain results in forcing.
+
+In this session, we cast some theories in relative form, in a
+way that they now refer to a fixed class $M$ as the universe of
+discourse. Whenever it was possible, we tried to minimize the changes
+to the structure and proof scripts. For this reason, some
+comments of the original text as well as outdated \textbf{apply}
+commands appear profusely in the following theories.
+
+A repeated pattern that appears is that the relativized result can be
+proved \emph{mutatis mutandis}, with remaining proof obligations that
+the objects constructed actually belong to the model $M$. Another
+aspect was that the management of higher order constructs always posed
+some extra problems, already noted by Paulson \cite[Sect.~7.3]{MR2051585}.
+
+We proceed to enumerate the theories that were “ported” to relative
+form, briefly commenting on each of them. Below, we refer to the
+original theories as the \emph{source} and, correspondingly, call
+\emph{target} the relativized version. We omit the \theory{.thy}
+suffixes.
+
+\begin{enumerate}
+\item From \session{ZF}:
+  \begin{enumerate}
+  \item \theory{Univ}. Here we decided to relativize only the term
+    \isatt{Vfrom} that constructs the cumulative hierarchy up to some
+    ordinal length and starting from an arbitrary set.
+  \item \theory{Cardinal}. There are two targets for this source,
+    \theory{Least} and \theory{Cardinal\_Relative}. Both require some
+    fair amount of preparation, trying to take advantage of absolute
+    concepts. It is not straightforward to compare source and targets
+    in a line-by-line fashion at this point.
+  \item \theory{CardinalArith}. The hardest part was to formalize the
+    cardinal successor function. We also disregarded the part treating
+    finite cardinals since it is an absolute concept. Apart from that,
+    the relative version closely parallels the source.
+  \item \theory{Cardinal\_AC}. After some boilerplate, porting was
+    rather straightforward, excepting cardinal arithmetic involving
+    the higher-order union operator.
+  \end{enumerate}
+\item From \session{ZF-Constructible}:
+  \begin{enumerate}
+  \item \theory{Normal}. The target here is \theory{Aleph\_Relative}
+    since that is the only concept that we ported. Instead of porting
+    all the machinery of normal functions (since it involved
+    higher-order variables), we particularized the results for the
+    Aleph function. We also used an alternative definition of the
+    latter that worked better with our relativization discipline.
+  \end{enumerate}
+\item From \session{Delta\_System\_Lemma}:
+  \begin{enumerate}
+  \item \theory{ZF\_Library}. The target includes a big section of
+    auxiliary lemmas and commands that aid the relativization. We
+    needed to make explicit the witnesses (mainly functions) in some of the 
+    existential results proved in the source, since only in that way
+    we would be able to show that they belonged to the model.
+  \item \theory{Cardinal\_Library}. Porting was relatively
+    straightforward; most of the extra work laid in adjusting locale
+    assumptions to obtain an appropriate context to state and prove
+    the theorems.
+  \item \theory{Delta\_System}. Same comments as in the case of
+    \theory{Cardinal\_Library} apply here.
+  \end{enumerate}
+\item From \session{Forcing}:
+  \begin{enumerate}
+  \item \theory{Pointed\_DC}. This case was similart to
+    \theory{Cardinal\_AC} above, alhough a bit of care was needed to
+    handle the recursive construction. Also, a fraction of the theory
+    \theory{AC} from \session{ZF} was ported here as it was a
+    prerrequisite. A complete relativization of \theory{AC} would be
+    desirable but still missing.
+  \end{enumerate}
+\end{enumerate}
 % generated text of all theories
 \input{session}
+
+\bibliographystyle{root}
+\bibliography{root}
 
@@
 %%% TeX-master: t
+%%% ispell-local-dictionary: "american"
 %%% End:

### 2022-02-28 scripts/fix_links.sh
updating Makefiles and associated script

@@ -3,8 +3,7 @@ old_dir=$PWD
 html_dir="$1"
 
 cd $html_dir
-sed -i -- 's/a href="..\/ZF\/ZF.html/a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF\/ZF.html/g' *.html
-sed -i -- 's/a href="..\/ZF\/ZF\//a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF\//g' *.html
-sed -i -- 's/a href="..\/ZF\/ZF-Constructible\//a href="https:\/\/www.cl.cam.ac.uk\/research\/hvg\/Isabelle\/dist\/library\/ZF\/ZF-Constructible\//g' *.html
-sed -i -- 's|a href="../ZF-Constructible-Trans/|a href="https://cs.famaf.unc.edu.ar/~pedro/forcing/ZF-Constructible-Trans/|g' *.html
+sed -i -- 's|a href="../ZF/|a href="https://www.cl.cam.ac.uk/research/hvg/Isabelle/dist/library/ZF/ZF/|g' *.html
+sed -i -- 's|a href="../ZF-Constructible/|a href="https://www.cl.cam.ac.uk/research/hvg/Isabelle/dist/library/ZF/ZF-Constructible/|g' *.html
+sed -i -- 's|a href="../../AFP/Delta_System_Lemma/|a href="https://www.isa-afp.org/browser_info/current/AFP/Delta_System_Lemma/|g' *.html
 cd $old_dir

### 2022-03-03 Transitive_Models/document/root.tex
Some documentation

@@
 
-As it was explained in \cite[Sect.~3]{2020arXiv200109715G} and elsewhere,
-relativization of concepts is a key tool to obtain results in forcing.
+Relativization of concepts is a key tool to obtain results in forcing,
+as it is explained in \cite[Sect.~3]{2020arXiv200109715G} and elsewhere.
 
@@
 discourse. Whenever it was possible, we tried to minimize the changes
-to the structure and proof scripts. For this reason, some
-comments of the original text as well as outdated \textbf{apply}
-commands appear profusely in the following theories.
+to the structure and proof scripts of the absolute concepts. For
+this reason, some comments of the original text as well as
+outdated \textbf{apply} commands appear profusely in the following
+theories.
 
@@
 
-We proceed to enumerate the theories that were “ported” to relative
+In the theory \theory{Lambda\_Replacement} we introduce a new locale assuming
+two instances of separation and twelve instances of ``lambda replacements''
+(ie, replacement of the form $\lambda x y. y=\langle x, f(x) \rangle$)
+that allow for having some form of compositionality of further instances
+of separations and replacements.
+
+We proceed to enumerate the theories that were ``ported'' to relative
 form, briefly commenting on each of them. Below, we refer to the
@@
 % generated text of all theories
+
 \input{session}

### 2022-03-04 src/document/root.tex
Names and Choice_Axiom according to Fm_Definitions.
I've changed some definitions in Names making them more
general; we should decide if are worth-keeping. We should
understand better why the satisfaction for is_check_fm
cannot be in Fm_Definitions.

I've added some text and comments in Fm_Definitions.
Editing deps of Forces_Definition. Missing LaTeX macro
A bit more of (sub)section structure
Comments on Defs_Main and Demonstrations (now in the session). TODOs and several FIXMEs deleted/addressed/transformed. New way to run (see README). Repeated code deleted.

@@
 \renewcommand{\isacharminus}{-}
+\newcommand{\session}[1]{\textit{#1}}
+\newcommand{\theory}[1]{\texttt{#1}}
 \newcommand{\axiomas}[1]{\mathit{#1}}
 \newcommand{\ZFC}{\axiomas{ZFC}}
+\newcommand{\ZF}{\axiomas{ZF}}
 \newcommand{\AC}{\axiomas{AC}}
 \newcommand{\CH}{\axiomas{CH}}
+\newcommand{\calV}{\mathcal{V}}
 
@@
 
-\title{Formalization of Forcing in Isabelle/ZF}
+\title{The Independence of the Continuum Hypothesis in Isabelle/ZF}
 \author{Emmanuel Gunther\thanks{Universidad Nacional de C\'ordoba. 
@@
 
+The main entry point of the present session is the theory
+\theory{Defitions\_Main.thy}, in which a path from fundamental set
+theoretic concepts formalized in Isabelle reaching our main theorems
+is expounded. Cross-references to major milestones are provided there.
+
+In order to provide evidence for the correctness of several of our relativized
+definitions, we needed to assume the Axiom of Choice ($\AC$) during the
+aforementioned theory. Nevertheless, the whole of our development is
+independent of $\AC$, and the theory \theory{CH.thy} already provides
+all of our results and does not import that axiom.
+
 \subsection*{Release notes}

### 2022-03-04 scripts/fix_links.sh
pattern-match all AFP entries

@@ -5,5 +5,5 @@ html_dir="$1"
 cd $html_dir
 sed -i -- 's|a href="../ZF/|a href="https://www.cl.cam.ac.uk/research/hvg/Isabelle/dist/library/ZF/ZF/|g' *.html
 sed -i -- 's|a href="../ZF-Constructible/|a href="https://www.cl.cam.ac.uk/research/hvg/Isabelle/dist/library/ZF/ZF-Constructible/|g' *.html
-sed -i -- 's|a href="../../AFP/Delta_System_Lemma/|a href="https://www.isa-afp.org/browser_info/current/AFP/Delta_System_Lemma/|g' *.html
+sed -i -- 's|a href="../../AFP/|a href="https://www.isa-afp.org/browser_info/current/AFP/|g' *.html
 cd $old_dir

### 2022-03-06 src/document/root.tex
Typos, hyphening

@@
-\documentclass[11pt,a4paper]{article}
+\documentclass[11pt,a4paper,english]{article}
 \usepackage{isabelle,isabellesym}
@@
   Isabelle/ZF. Under the assumption of the existence of a countable
-  transitive model of ZFC, we construct proper generic extensions 
+  transitive model of $\ZFC$, we construct proper generic extensions 
   that satisfy the Continuum Hypothesis and its negation.
@@
 The main entry point of the present session is the theory
-\theory{Defitions\_Main.thy}, in which a path from fundamental set
+\theory{Definitions\_Main.thy}, in which a path from fundamental set
 theoretic concepts formalized in Isabelle reaching our main theorems

### 2022-03-26 JSL/independence_ch_isabelle.tex
Random draft for JSL paper

--- 2019-12-16 IJCAR2020/forcing_in_isabelle_zf.tex
@@
 \usepackage[utf8]{inputenc}
-\usepackage{isabelle,isabellesym}
-\input{header-draft}
+\usepackage{isabelle_indepCH,isabellesym_indepCH}
+\input{header-indepCH}
 \usepackage{graphicx}
@@
 % to display URLs in blue roman font according to Springer's eBook style:
-\renewcommand\UrlFont{\color{blue}\rmfamily}
+% \renewcommand\UrlFont{\color{blue}\rmfamily}
 
@@
 %
-\title{Formalization of Forcing in Isabelle/ZF%
+\title{Some lessons after the formalization of the ctm approach to forcing%
   \thanks{Supported by Secyt-UNC project 33620180100465CB and Conicet.}%
@@
 %
-%\titlerunning{Abbreviated paper title}
+\titlerunning{Lessons after formalizing ctm forcing}
 % If the paper title is too long for the running head, you can set
@@
 Miguel Pagano\inst{1} \and \\
-Pedro Sánchez Terraf\inst{1,2}\orcidID{0000-0003-3928-6942}}
+Pedro Sánchez Terraf\inst{1,2}%\orcidID{0000-0003-3928-6942}
+\and
+Matías Steinberg\inst{1}
+}
 %
-\authorrunning{E.~Gunther, M.~Pagano, P.~Sánchez Terraf}
-%\authorrunning{E. Gunther et al.}
+\authorrunning{E.~Gunther, M.~Pagano, P.~Sánchez Terraf, M.~Steinberg}
 % First names are abbreviated in the running head.
@@
   \\  Facultad de Matem\'atica, Astronom\'{\i}a,  F\'{\i}sica y
-    Computaci\'on. \\
-\email{\{gunther,pagano,sterraf\}@famaf.unc.edu.ar}
+  Computaci\'on.
   \and
     Centro de Investigaci\'on y Estudios de Matem\'atica (CIEM-FaMAF),
-    Conicet. C\'ordoba. Argentina.
+    Conicet. C\'ordoba. Argentina. \\
+    \email{\{gunther,pagano,sterraf\}@famaf.unc.edu.ar\\
+        matias.steinberg@mi.unc.edu.ar}
 }
@@
 \begin{abstract}
-We formalize the development of forcing in the set theory framework of
-Isabelle/ZF. Under the assumption of the existence of countable
-transitive model of $\ZF$, we construct a generic extension and show
-that the latter also satisfies $\ZF$.
-\keywords{Isabelle/ZF \and forcing \and names \and generic extension \and constructibility.}
+  We'll discuss some highlights of our computer-verified
+  proof of the construction, given a countable transitive set model $M$
+  of $\ZFC$, of generic extensions  satisfying $\ZFC+\neg\CH$ and $\ZFC+\CH$.
+  In particular,
+  we isolated a set $\Delta$ of 39 instances
+  of Axiom of Replacement and a function $F$
+  such that such that for any finite fragment $\Phi\sbq\ZFC$,
+  $F(\Phi)\sbq\ZFC$ is also finite and if
+  $M\models F(\Phi) + \Delta$ then $M[G]\models \Phi + \neg \CH$.
+  We also obtained the formulas yielded by the Forcing Definability Theorem
+  explicitly.
+
+  To achieve this, we worked in the proof assistant \emph{Isabelle},
+  basing our development on the theory Isabelle/ZF by L.~Paulson and
+  others.
+
+  %% The vantage point of the talk will be that of a mathematician but
+  %% elements from the computer science perspective will be
+  %% present. Perhaps some myths regarding what can effectively be done
+  %% using  proof assistants/checkers will
+  %% be dispelled.
+
+  We'll also compare our formalization with the recent one by Jesse
+  M.~Han and Floris van Doorn in the proof assistant \emph{Lean}.
+
+  \keywords{forcing \and Isabelle/ZF \and countable transitive models
+    \and absoluteness \and generic extension \and constructibility.}
 \end{abstract}
@@
 \input{intro}
+%% 
+%% \input{isabelle_metatheory}
+%% 
+%% \input{absoluteness}
+%% 
+%% \input{forces_definition}
+%% 
+%% \input{forcing_theorems}
+%% 
+%% \input{proper_extension}
+%% 
+%% \input{replacement_choice}
+%% 
+\input{random}
 
-\input{isabelle_metatheory}
+\input{related}
 
-\input{absoluteness}
-
-\input{conclusion}
+\input{lessons}
+%%
+%% \input{conclusion}
 %
@@
 %
-\bibliographystyle{splncs04}
-\bibliography{forcing_in_isabelle_zf}
+%\bibliographystyle{splncs04}
+\bibliographystyle{mi-estilo-else}
+\bibliography{independence_ch_isabelle,clam2021}
+
 \end{document}
+
+%%% Local Variables: 
+%%% mode: latex
+%%% ispell-local-dictionary: "american"
+%%% End:

### 2022-03-26 JSL/intro.tex
Random draft for JSL paper

\section{Introduction}
\label{sec:introduction}

\subsection{Formalized mathematics}
There is a growing trend in the use of computers to assist in the
\emph{formal verification} of mathematics. Note, for instance, the forthcoming
\href{https://web.archive.org/web/20211220044218/https://icm2022.org/plenary-lectures#Kevin-Buzzard}{ICM2022
  plenary lecture} by Kevin Buzzard.

Several venues devoted to the communication of formalized mathematics;
today, mainly CS conferences: ITP, IJCAR, CPP, CICM, and others.

Many proof assistants, each with a different flavour and underlying
foundations: Naproche (Koepke, mainly FOL + ATP), Mizar,
Isabelle/HOL, Lean, \dots, Coq, Agda,\dots.

These assistants provide several dialects, among which we single out:
\begin{enumerate}
\item Procedural: Useful for exploration/research.
\item Declarative: Only one that can be read by humans!
\end{enumerate}

Before we start an in-depth discussion, a point should be made clear:
A formalized proof is not the same as an \emph{automatic proof}. The
reader surely understands the aside from trivial results, no current
technology allows us to state a reasonable (and correct) theorem
statement and expect to obtain a proof after hitting ``Enter'', at
least not after a humanly feasible wait. On the other hand, it is
quite possible that the same reader has some mental image that
formalizing a proof requires making each application of Modus Ponens
explicit.

The fact is that proof assistants are designed for the human prover to
be able to decompose a statement to be proved into smaller subgoals
which can actually be fed to some automatic tool. The balance between
what these tools are able to handle is not  easily appreciated by
intuition: Sometimes, ``trivial'' steps are not solved by them, which
can result in obvious frustration; but they would quickly solve some
goals that do not look like a ``mere calculation.''

The same happens to the extent of mathematics formalizable (or even,
which has \emph{already} been formalized): The community seems not to
be fully aware of the potential of the technology. A recent discussion
at the FOM mailing list shows this state of affairs within our fellow
logicians. Major formalization projects include the Four Color Theorem
\cite{MR2463991}, the odd order theorem
\cite{10.1007/978-3-642-39634-2_14}, and the proof the Kepler's
Conjecture \cite{MR3659768}. There is a vast mathematical corpus at
the Archive of Formal Proofs based on Isabelle, and formalizations of
brand new and complex objects like the definition of perfectoid spaces
and the Liquid Tensor Project on Lean.

\subsection{Our achievements}
We formalized a model-theoretic rendition of forcing.
\begin{itemize}
\item Proper extensions of ctms of $\ZF$.
\item The construction of ctms of $\ZFC + \neg\CH$ and of $\ZFC + \CH$ given one of
  $\ZFC$.
\end{itemize}

In order to do so, we extended the treatment of relativization of
set-theoretical concepts. 
\begin{itemize}
\item We redesigned Isabelle/ZF results on non-absolute concepts to work
  relative to a class.
\item We provided missing results (from basic ones involving cardinal
  successors, countable sets, \dots)
\end{itemize}
No meta-theoretic issues (consistency, FOL calculi, etc) were formalized.

One added value that is obtained from the present formalization is
that we identified which instances of  Replacement are
needed to set the forcing machinery up (Separation is a bit more
complicated).
%% A verification that no choice  is needed at the level of the
%%  object logic.
The eagerness to obtain this level of detail might be a consequence of
“an unnatural tendency to investigate, for the most part, trivial
minutiae of the formalism”, as it was put by Cohen
\cite{zbMATH02012060}, but we were also drove by curiosity.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "independence_ch_isabelle"
%%% ispell-local-dictionary: "american"
%%% End:

### 2022-03-26 JSL/lessons.tex
Random draft for JSL paper

\section{Lessons}\label{sec:lessons}

\subsection{Ask about your project}
\url{https://mathoverflow.net/q/265435/66044}

Disclaimer: It may be useless.

\subsection{Read about what is done}
\begin{enumerate}
\item Related work at that moment.
\item Precise enumeration of what \session{ZF-Constructible} had.
\end{enumerate}

\subsection{Read the fine print}
\begin{enumerate}
\item Foundations of your proof assistant.
\end{enumerate}

\subsection{Plan before formalizing}
\begin{enumerate}
\item We should had better used predicates for the forcing posets'
  order relations (the way DSL is written), and go for class forcing
  (with $\PP$ a definable subset of $M$). The latter change seems to be
  easy, but the former doesn't.

  “Sometimes it felt like sculpting on marble: The fear of having
  carved too deep and hence needing to change the whole stone!”
\end{enumerate}

\subsection{In case of doubt, formalize it}
\begin{enumerate}
\item Example: the abstract of the CLAM2021 talk.
\end{enumerate}

\subsection{Control your bureaucracy. Automate early}
\begin{enumerate}
\item Bureaucracy vs ML programming.
\item The “math” was already formalized on 22 November 2020.
  We finished the last goal on 22 August 2021.
  (Update: 20 November 2021 \& 28 November 2021, for CH)
\item Missing: automation of closure of models under operations.
\item Missing: basic arithmetic for dealing with arities.
\end{enumerate}

\subsection{Beware of scale factors}
\begin{enumerate}
\item It is extremely misleading when automatic tools (\isatt{simp}, \isatt{auto}, etc)
  stop working just because of the sheer size of the goal. Oftentimes,
  in math, we disregard scale issues but they must always be taken
  into account in CS.
\item Example: $\forceisa(0\in 1)$ is expandable,
  $\forceisa(\neg\neg  0\in 1)$ is not.
\item Example: Synthesis of $\forceisa$; could have been fully synthesized,
  but that was dirty “strategy”.
\item The know-how of computer scientists on this kind of engineering is
  very important
\end{enumerate}

\subsection{You might have formalized it, and still be wrong}
\begin{enumerate}
\item Example: restriction of relations.
\item Pollack, “Pollack consistency” by Wiedijk. Opacity of automated
  proofs.
\item Plot twist: You can be right without knowing. Intuition may drive proofs
  even if we are not working on what we believe we are.
\end{enumerate}

\subsection{Beware of the “Code fever”}
\begin{enumerate}
\item “We know that doing math is fun---formalization is like DRUGS”
\item Feeling of accomplishment after seeing your writings
  validated beyond reasonable doubt.
\item One easily forgets about the “Power of the Board.”
\end{enumerate}

\subsection{The Devil's on the shortcuts}
\begin{enumerate}
\item
  Our proofs of the "definition of forces" (and many
  consequences) and of the lemma for "forcing a value" of function
  depend on the countability of the ground model. 
\item
  Density arguments (look for "TODO", "general versions").
\end{enumerate}

\subsection{Document your project}
\begin{enumerate}
\item \theory{Definitions\_Main}, thanks to Vidnyánszky.
\end{enumerate}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "independence_ch_isabelle"
%%% ispell-local-dictionary: "american"
%%% End:

### 2022-03-26 JSL/random.tex
Random draft for JSL paper

\subsection{Random comments}

\begin{enumerate}
\item Satisfaction of having one's material fully formalized
  (v.g. cofinality).
\item The concept of absoluteness (\verb|_abs|) splitted into two components
  during the formalization: \verb|is_foo_iff| and \verb|foo_rel_abs| (this last one
  being the concept of absoluteness from  “mathematical practice”) and
  the first one involves a purely relational presentation, introduced
  by Paulson.
\item
  In order to extract information concerning the proof, the need for a
  proof assistant with a more computational flavour (Agda, Coq)
  arises! As a second thought, obviously.
\item  \textbf{Pros and cons of an untyped environment}: Closer to the
  set-theoretic point of view/prone to the most stupid errors.
\item \textbf{“Pros” and cons of lack of automation}: we had to be extremely
  detailed, which provides explicit information / the formalization
  work was pain in the neck and much lower.
\item In order to extract information concerning the proof, the need for a
  proof assistant with a more computational flavour (Agda, Coq)
  arises! As a second thought, obviously.
\item Promotion of formalization by Buzzard.
\end{enumerate}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "independence_ch_isabelle"
%%% ispell-local-dictionary: "american"
%%% End:

### 2022-03-26 JSL/related.tex
Random draft for JSL paper

\section{Related work}
\label{sec:related-work}

\paragraph*{Question} How do they overcome Tarski's undefinability of truth?

In some sense, their presentation is mathematically more "mature",
but the type-theoretic machinery might be a bit alien to most
set-theorists.

Quoting Flypitch-CPP,
\begin{quote}
  The types of cardinals and ordinals in mathlib, which
  are defined as equivalence classes of (well-ordered) types, live one
  universe level higher than the types used to construct them
\end{quote}

\paragraph*{Questions} How do they handle recursions? Are these all
\emph{external}? Does this has any impact on their (non?)ability of
identifying replacement instances used by their approach?

They seem to require $\AC$ in their meta-theory.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "independence_ch_isabelle"
%%% ispell-local-dictionary: "american"
%%% End:

### 2022-03-26 JSL/intro.tex
Typos and hyphening

@@
 A formalized proof is not the same as an \emph{automatic proof}. The
-reader surely understands the aside from trivial results, no current
+reader surely understands that, aside from trivial results, no current
 technology allows us to state a reasonable (and correct) theorem

### 2022-03-26 JSL/intro.tex
Minor edits

@@
 be able to decompose a statement to be proved into smaller subgoals
-which can actually be fed to some automatic tool. The balance between
+which can actually be fed into some automatic tool. The balance between
 what these tools are able to handle is not  easily appreciated by
@@
 can result in obvious frustration; but they would quickly solve some
-goals that do not look like a ``mere calculation.''
+goals that do not look like a ``mere computation.''
 
@@
 at the FOM mailing list shows this state of affairs within our fellow
-logicians. Major formalization projects include the Four Color Theorem
-\cite{MR2463991}, the odd order theorem
+logicians. It is thus convenient to recall
+some major formalization projects, such as the Four Color Theorem
+\cite{MR2463991}, the Odd Order Theorem
 \cite{10.1007/978-3-642-39634-2_14}, and the proof the Kepler's

### 2022-03-31 JSL/forcing.tex
Starting the section on forcing. Liquid Tensor Experiment refs

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
\section{Forcing}
\label{sec:forcing}

Prerequisite: a set $\formula$ of codes for first order formulas, and
the (recursive) definition of the Tarski satisfaction relation in $\ZF$.

Let $\lb \PP, {\preceq} ,\1\rb \in M$ be a forcing notion. Given $G\sbq \PP$, we have
$M[G]\defi \{ \val(\PP,G,\punto{a}) : \punto{a}\in M \}$.

The following form of the Forcing Theorems  is the one
that we formalized.
\begin{theorem}
  There exists a function  $\forceisa: \formula \to  \formula$
  such that for every
  $\phi$ and $\punto{a}_0,\dots,\punto{a}_n\in M$,
  \begin{enumerate}
  \item (Truth Lemma) for every $M$-generic $G$,
    \[
      M[G], [\val(\PP,G,\punto{a}_0),\dots,\val(\PP,G,\punto{a}_n)]
      \models \phi
    \]
    is equivalent to 
    \[
      \exists p\in G.\ \; M, [p,\PP,\preceq,\1, \punto{a}_0,\dots,\punto{a}_n]  \models
      \forceisa(\phi).\]

    We use the notation $p \forces
    \phi\ [\punto{a}_0,\dots,\punto{a}_n]$ for this last assertion.
  \item (Density Lemma) $p \forces \phi\ [\punto{a}_0,\dots,\punto{a}_n]$
    if and only if 
    $\{q\in \PP :  q \forces \phi\ [\punto{a}_0,\dots,\punto{a}_n]\}$
    is dense below $p$.
  \end{enumerate}
\end{theorem}

Note: $\forceisa$ was formalized as a class-function (i.e., of type
$\tyi\Rightarrow\tyi$), to take advantage of the Isabelle's
conveniences for primitive recursion.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "independence_ch_isabelle"
%%% ispell-local-dictionary: "american"
%%% End:

### 2022-03-31 JSL/intro.tex
Starting the section on forcing. Liquid Tensor Experiment refs

@@
 brand new and complex objects like the definition of perfectoid spaces
-and the Liquid Tensor Project on Lean.
+and the Liquid Tensor Project \cite{LTE2020,LTE2021} on Lean.
 

### 2022-04-05 JSL/lessons.tex
Starting the section on forcing. Liquid Tensor Experiment refs
Trimming list of lessons

@@
 
-\subsection{Ask about your project}
+\subsection{Plan before formalizing}
+\begin{enumerate}
+\item Ask about your project
   \url{https://mathoverflow.net/q/265435/66044}
@@
   Disclaimer: It may be useless.
+\item Antecedents: Precise enumeration of what \session{ZF-Constructible} had.
 
-\subsection{Read about what is done}
-\begin{enumerate}
-\item Related work at that moment.
-\item Precise enumeration of what \session{ZF-Constructible} had.
-\end{enumerate}
-
-\subsection{Read the fine print}
-\begin{enumerate}
-\item Foundations of your proof assistant.
-\end{enumerate}
-
-\subsection{Plan before formalizing}
-\begin{enumerate}
 \item We should had better used predicates for the forcing posets'
@@
   carved too deep and hence needing to change the whole stone!”
+
+  Note: This is in direct contradiction with “code fever”
+  (\ref{sec:beware-code-fever} below).
 \end{enumerate}
 
-\subsection{In case of doubt, formalize it}
+\subsection{Read the fine print}
 \begin{enumerate}
-\item Example: the abstract of the CLAM2021 talk.
+\item Foundations of your proof assistant.
 \end{enumerate}
@@
 
-\subsection{Beware of the “Code fever”}
+\subsection{Beware of the “Code fever”}\label{sec:beware-code-fever}
 \begin{enumerate}

### 2022-04-27 JSL/details.tex
Preliminary analysis of instances; a few words on proof assistants

\section{Details of the formalization}

\subsection{35 replacement instances to rule them all}
We isolated 35 instances of Replacement that are needed to force
$\CH$/$\neg\CH$. The first 14 instances are needed to set up
cardinal arithmetic in $M$.

\begin{itemize}
\item 7 instances for iteration of operations.
  \begin{itemize}
  \item Fixpoints: lists
    \isatt{list{\uscore}repl1{\uscore}intf{\uscore}fm},
    \isatt{list{\uscore}repl2{\uscore}intf{\uscore}fm}; formulas
    \isatt{formula{\uscore}repl1{\uscore}intf{\uscore}fm},
    \isatt{formula{\uscore}repl1{\uscore}intf{\uscore}fm}.
  \item Transitive closure \isatt{eclose{\uscore}repl1{\uscore}intf{\uscore}fm},
    \isatt{eclose{\uscore}repl2{\uscore}intf{\uscore}fm} (iteration of $\union$).
  \item $n$th element of a list \isatt{tl{\uscore}repl{\uscore}intf{\uscore}fm} (iteration of the tail operation; fake?).
  \end{itemize}

\item 4 instances for definitions by well-founded recursion.
  \begin{itemize}
  \item Ordertypes \isatt{wfrec{\uscore}replacement{\uscore}order{\uscore}pred{\uscore}fm}.
  \item Rank \isatt{wfrec{\uscore}rank{\uscore}fm}.
  \item Cumulative hierarchy \isatt{trans{\uscore}repl{\uscore}HVFrom{\uscore}fm} (rank initial segments).
  \item Aleph \isatt{replacement{\uscore}HAleph{\uscore}wfrec{\uscore}repl{\uscore}body{\uscore}fm}.
  \end{itemize}

\item 3 other replacements.
  \begin{itemize}
  \item Definition of ordertypes \isatt{replacement{\uscore}is{\uscore}order{\uscore}body{\uscore}fm},
    \isatt{}.
  \item Definition of cardinal successor \isatt{replacement{\uscore}is{\uscore}jump{\uscore}cardinal{\uscore}body{\uscore}fm}.
  \end{itemize}
\end{itemize}

We also need a one extra replacement instance $\psi$ on $M$ for each
$\phi$ of the
previous ones to have them in $M[G]$.
\[
  \psi(x,\alpha,y_1,\dots,y_n) \defi \quine{\alpha = \min \bigl\{
    \beta \mid \exists\tau\in V_\beta.\  \mathit{snd}(x) \forces
    \phi\ [\mathit{fst}(x),\tau,y_1,\dots,y_n]\bigr\}}
\]
That makes 28 instances up to now. For the setup of forcing, we
require the following 7 instances (the first two needed for the $\Delta$-system Lemma):

\begin{itemize}
%\verb|replacement_ax2(6)| \verb|replacement_transrec_apply_image_body_fm| seems fake!
\item Recursive construction of sets using a choice function
  \isatt{replacement{\uscore}is{\uscore}trans{\uscore}apply{\uscore}image{\uscore}fm}.
% \item Replacement through  $x\mapsto \check{x}$ \isatt{check{\uscore}replacement{\uscore}fm}.
\item $x\mapsto |x|^M$
  \isatt{Lambda{\uscore}in{\uscore}M{\uscore}fm(is{\uscore}cardinal{\uscore}fm(0,1),0)}.
\item 2 instances for definitions by well-founded recursion \isatt{wfrec{\uscore}Hcheck{\uscore}fm} (check
  names) and \isatt{wfrec{\uscore}Hfrc{\uscore}at{\uscore}fm} (forcing for atomic formulas).
\item Replacement through $x\mapsto \lb\check{x},x\rb$
  \isatt{G{\uscore}dot{\uscore}in{\uscore}M{\uscore}fm} (for the
  definition of $\punto{G}$).
% \item Name of choice functions in $M[G]$ \isatt{repl{\uscore}opname{\uscore}check{\uscore}fm}.
\item 2 instances for the slalom needed for cardinal preservation
  by ccc forcing.
\end{itemize}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "independence_ch_isabelle"
%%% ispell-local-dictionary: "american"
%%% End:

### 2022-04-27 JSL/proof_assists_isabelle.tex
Preliminary analysis of instances; a few words on proof assistants

\section{Proof assistants and Isabelle/ZF}
\label{sec:proof-assist-isabelle}

Proof assistants provide diverse aids for the task
of formalizing a piece of mathematics. They are usually implemented
using a typed programming language; rigor is enforced by defining a
type of ``theorems'', whose members can only be constructed using
operations stipulated in a small \emph{kernel} which encodes the
underlying foundation of the assistant. Further developments interact
with the type of theorems only through the kernel, and thus the latter
is the only ``trusted'' part of the assistant's code.

Several of the more established assistants (HOL Light, Coq, Isabelle)
are programmed in some variant of the ML language (which was
originally designed for this purpose); the newer Lean, on the other
hand, was originally conceived as a standalone functional programming
language with all the features of a standard assistant.

\subsection{Four layers}

In the case of Isabelle, Standard ML is the deeper layer of the
assistant, in which the kernel is written.

%% Isabelle \cite{Isabelle,DBLP:books/sp/Paulson94} is a general proof
%% assistant based on fragment of higher-order logic called
%% \emph{Pure}. 
%% The results presented in this work are theorems of a
%% version of $\ZF$ set theory (without the Axiom of Choice, $\AC$) 
%% called \emph{Isabelle/ZF}, which is one of the
%% ``object logics'' that can be defined on top of Pure (which is then
%% used as a language to define rules). Isabelle/ZF defines types
%% \isatt{i} and \isatt{o} for sets and Booleans, resp., and the $\ZF$
%% axioms are written down as terms of type \isatt{o}.
%% 
%% It should be noted that Pure is a very weak framework and has no
%% induction/recursion capabilities. So the only way to define functions
%% by recursion is inside the object logic. (This works the same for
%% Isabelle/HOL.) For this reason, to define the relation of forcing, we
%% needed to resort to \emph{internalized} first-order formulas: they
%% form a recursively defined set \isatt{formula}. For example, the
%% predicate of satisfaction
%% \isatt{sats::i{\isasymRightarrow}i{\isasymRightarrow}i{\isasymRightarrow}o}
%% (written $M,\mathit{ms}\models\phi$ for a set $M$,
%% $\mathit{ms}\in\isatt{list}(M)$ and $\phi\in\formula$)
%% had already been defined by recursion in Paulson~\cite{paulson_2003}.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "independence_ch_isabelle"
%%% ispell-local-dictionary: "american"
%%% End:

### 2022-04-27 JSL/details.tex
2 instances supposedly out

@@
 
-\subsection{35 replacement instances to rule them all}
-We isolated 35 instances of Replacement that are needed to force
+\subsection{33 replacement instances to rule them all}
+We isolated 33 instances of Replacement that are needed to force
 $\CH$/$\neg\CH$. The first 14 instances are needed to set up
@@
   \begin{itemize}
-  \item Ordertypes \isatt{wfrec{\uscore}replacement{\uscore}order{\uscore}pred{\uscore}fm}.
+  \item Ordertypes
+    \isatt{wfrec{\uscore}replacement{\uscore}order{\uscore}pred{\uscore}fm}.
   \item Rank \isatt{wfrec{\uscore}rank{\uscore}fm}.
@@
 
-\item 3 other replacements.
+\item 2 instances for definition of ordertypes.
   \begin{itemize}
-  \item Definition of ordertypes \isatt{replacement{\uscore}is{\uscore}order{\uscore}body{\uscore}fm},
-    \isatt{}.
-  \item Definition of cardinal successor \isatt{replacement{\uscore}is{\uscore}jump{\uscore}cardinal{\uscore}body{\uscore}fm}.
+  \item \isatt{replacement{\uscore}is{\uscore}order{\uscore}body{\uscore}fm}.
+  \item \isatt{replacement{\uscore}is{\uscore}order{\uscore}eq{\uscore}map{\uscore}fm}.
+  \end{itemize}
+\item 1 instance for definition of cardinal successor.
+  \begin{itemize}
+  \item
+    \isatt{replacement{\uscore}is{\uscore}jump{\uscore}cardinal{\uscore}body{\uscore}fm}.
   \end{itemize}
@@
 That makes 28 instances up to now. For the setup of forcing, we
-require the following 7 instances (the first two needed for the $\Delta$-system Lemma):
+require the following 5 instances (the first two needed for the $\Delta$-system Lemma):
 
@@
 % \item Name of choice functions in $M[G]$ \isatt{repl{\uscore}opname{\uscore}check{\uscore}fm}.
-\item 2 instances for the slalom needed for cardinal preservation
-  by ccc forcing.
+% \item 2 instances for the slalom needed for cardinal preservation by ccc forcing.
 \end{itemize}

### 2022-05-04 JSL/lessons.tex
Preliminary analysis of instances; a few words on proof assistants
types and sorts in Isabelle's M

@@
   (\ref{sec:beware-code-fever} below).
-\end{enumerate}
-
-\subsection{Read the fine print}
-\begin{enumerate}
-\item Foundations of your proof assistant.
+\item Read the fine print: Foundations of your proof assistant.
 \end{enumerate}
@@
 \item Feeling of accomplishment after seeing your writings
-  validated beyond reasonable doubt.
+  validated beyond reasonable doubt (v.g. cofinality).
+
 \item One easily forgets about the “Power of the Board.”
@@
 \item
-  Our proofs of the "definition of forces" (and many
-  consequences) and of the lemma for "forcing a value" of function
+  Our proofs of the “definition of forces” (and many
+  consequences) and of the lemma for “forcing a value” of function
   depend on the countability of the ground model. 
 \item
-  Density arguments (look for "TODO", "general versions").
+  Density arguments (look for “TODO”, “general versions”).
 \end{enumerate}

### 2022-05-04 JSL/proof_assists_isabelle.tex
types and sorts in Isabelle's M

@@
 are programmed in some variant of the ML language (which was
-originally designed for this purpose); the newer Lean, on the other
+designed for this purpose); the newer Lean, on the other
 hand, was originally conceived as a standalone functional programming
@@
 assistant, in which the kernel is written.
+
+The logical foundation of Isabelle is an intuitionistic fragment of
+higher-order logic (or simple type theory) called $\Meta$;
+original version was described in \cite{Paulson1989}, and the addition
+of  sorts appears in \cite{Nipkow-LF-91}.
+
+The only predefined type is $\prop$ (“propositions”); new base types
+can be postulated when defining objects logics. Types of higher order can be
+assembled using the function space constructor $\fun$.
+
+The type of propositions $\prop$ equipped with a binary operation
+$\implies$ (“meta-implication”) that is used to represent the object
+logic rules. As an example, the axiomatization of first-order logic
+postulates a type $\tyo$ of booleans, and Modus Ponens
+% https://isabelle.in.tum.de/dist/library/FOL/FOL/IFOL.html#IFOL.mp|axiom
+is written as
+\[
+  [P\limp Q] \implies ([P] \implies [Q]).
+\]
+The square brackets represent an injection from $\tyo$ into
+$\prop$. % ($[P]$ can be read as “$P$ holds”)
+A consequence of this representation is that every
+formula of the object logic appears as atomic for $\Meta$.
+
+%% Quantification is handled in the meta-level using a functional $\ALL$
+%% with polymorphic type $(\alpha \fun \prop) \fun \prop$. 
+Types in Isabelle are organized into \emph{classes} and \emph{sorts};
+for ease of exposition, we will omit the former.  The axiomatization
+of first-order logic postulates a sort $\type{term}$ (of
+“individuals”, or elements of a first-order universe of discourse) and
+stipulates that every further type variable $\alpha$ must be of that
+sort. In particular, Isabelle/ZF only postulates one new type $\tyi$
+(“sets”) of sort $\type{term}$. Hence, from the type of the universal
+quantifier functional $\forall :: (\alpha \fun \tyo) \fun \tyo$ it
+follows that it may only be applied to predicates with a variable of
+type $\tyi$. This ensures that the object logic is effectively
+first-order.
+
+Paulson  \cite{Paulson1989} carried out a proof that the encoding
+$\Meta_{\mathrm{IFOL}}$ of
+intuitionistic first-order logic without equality in the original $\Meta$ is
+conservative (there is a correspondence between provable $\phi$ in
+IFOL and provable $[\phi]$ in $\Meta_{\mathrm{IFOL}}$) by putting
+$\Meta_{\mathrm{IFOL}}$ proofs in \emph{expanded normal form}
+\cite{MR0387024}. Passing to classical logic does not presents
+difficulties, but the treatment of meta-equality differs between the
+original and the present incarnation of $\Meta$; details are
+exhaustively expounded in the recent formalization by Nipkow and
+Roßkof \cite{10.1007/978-3-030-79876-5_6}.
 

### 2022-05-04 JSL/proof_assists_isabelle.tex
Isabelle/ZF, Constructible, and internalized formulas

@@
 
-\subsection{Four layers}
+In the case of Isabelle, Standard ML is the first of the four layers
+in which we work in this assistant. Both the kernel and automation
+of proofs is coded in ML, sometimes as a substitute for induction on
+formulas, as the next section explains.
 
-In the case of Isabelle, Standard ML is the deeper layer of the
-assistant, in which the kernel is written.
+\subsection{Isabelle metalogic $\Meta$}
 
-The logical foundation of Isabelle is an intuitionistic fragment of
-higher-order logic (or simple type theory) called $\Meta$;
-original version was described in \cite{Paulson1989}, and the addition
-of  sorts appears in \cite{Nipkow-LF-91}.
+The second layer of Isabelle is an
+intuitionistic fragment of higher-order logic (or simple type theory)
+called $\Meta$; its original version was described in \cite{Paulson1989},
+and the addition of “sorts” appears in \cite{Nipkow-LF-91}.
 
@@
 \]
-The square brackets represent an injection from $\tyo$ into
-$\prop$. % ($[P]$ can be read as “$P$ holds”)
-A consequence of this representation is that every
-formula of the object logic appears as atomic for $\Meta$.
+The square brackets (which are omitted in Isabelle theories) represent
+an injection from $\tyo$ into $\prop$. % ($[P]$ can be read as “$P$
+% holds”)
+A consequence of this representation is that every formula of
+the object logic appears as atomic for $\Meta$.
 
@@
 for ease of exposition, we will omit the former.  The axiomatization
-of first-order logic postulates a sort $\type{term}$ (of
-“individuals”, or elements of a first-order universe of discourse) and
+of first-order logic postulates a sort $\{\type{term}\}$ (of
+“individuals,” or elements of a first-order universe of discourse) and
 stipulates that every further type variable $\alpha$ must be of that
 sort. In particular, Isabelle/ZF only postulates one new type $\tyi$
-(“sets”) of sort $\type{term}$. Hence, from the type of the universal
+(“sets”) of sort $\{\type{term}\}$. Hence, from the type of the universal
 quantifier functional $\forall :: (\alpha \fun \tyo) \fun \tyo$ it
@@
 $\Meta_{\mathrm{IFOL}}$ of
-intuitionistic first-order logic without equality in the original $\Meta$ is
+intuitionistic first-order logic IFOL without equality  in the original $\Meta$ is
 conservative (there is a correspondence between provable $\phi$ in
@@
 $\Meta_{\mathrm{IFOL}}$ proofs in \emph{expanded normal form}
-\cite{MR0387024}. Passing to classical logic does not presents
+\cite{MR0387024}. Passing to classical logic does not present
 difficulties, but the treatment of meta-equality differs between the
-original and the present incarnation of $\Meta$; details are
-exhaustively expounded in the recent formalization by Nipkow and
-Roßkof \cite{10.1007/978-3-030-79876-5_6}.
+original and the present incarnation of $\Meta$; details for the
+latter are exhaustively expounded in the recent formalization by
+Nipkow and Roßkof \cite{10.1007/978-3-030-79876-5_6}.
 
-%% Isabelle \cite{Isabelle,DBLP:books/sp/Paulson94} is a general proof
-%% assistant based on fragment of higher-order logic called
-%% \emph{Pure}. 
-%% The results presented in this work are theorems of a
-%% version of $\ZF$ set theory (without the Axiom of Choice, $\AC$) 
-%% called \emph{Isabelle/ZF}, which is one of the
-%% ``object logics'' that can be defined on top of Pure (which is then
-%% used as a language to define rules). Isabelle/ZF defines types
-%% \isatt{i} and \isatt{o} for sets and Booleans, resp., and the $\ZF$
-%% axioms are written down as terms of type \isatt{o}.
-%% 
-%% It should be noted that Pure is a very weak framework and has no
-%% induction/recursion capabilities. So the only way to define functions
-%% by recursion is inside the object logic. (This works the same for
-%% Isabelle/HOL.) For this reason, to define the relation of forcing, we
-%% needed to resort to \emph{internalized} first-order formulas: they
-%% form a recursively defined set \isatt{formula}. For example, the
-%% predicate of satisfaction
-%% \isatt{sats::i{\isasymRightarrow}i{\isasymRightarrow}i{\isasymRightarrow}o}
-%% (written $M,\mathit{ms}\models\phi$ for a set $M$,
-%% $\mathit{ms}\in\isatt{list}(M)$ and $\phi\in\formula$)
-%% had already been defined by recursion in Paulson~\cite{paulson_2003}.
+The meta-logic $\Meta$ is rather weak; it has no induction/recursion
+principles. Types are not inductively presented and, in particular, it
+is not possible to perform proofs by induction statements about
+object-logic formulas (which are construed as terms of type $\tyi \fun
+\dots \fun \tyi \fun \tyo$). Two ways to overcome this limitation are:
+\begin{enumerate}
+\item
+  to construct the
+  proof of each instance of the statement by hand or by programming on
+  ML; or 
+\item
+  to encode formulas as sets and prove an internal version statement
+  using induction of $\ZF$.
+\end{enumerate}
+
+For recursive definitions, only the second option is available, and
+that is the way the Definition of Forcing is implemented in our
+formalization.
+
+\subsection{Isabelle/ZF}
+
+For the most part, the development of set theory in Isabelle is
+carried out using its ZF object logic
+\cite{DBLP:journals/jar/PaulsonG96}. Apart from the type and sort
+declarations detailed above, it features a finite axiomatization,
+% https://isabelle.in.tum.de/dist/library/FOL/FOL/IFOL.html#ZF_Base.mem|const
+with a predicate for membership, constants for the empty set and an
+infinite set, and functions $\isatt{Pow}::\tyi\fun\tyi$,
+$\union::\tyi\fun\tyi$, and $\isatt{PrimReplace} :: \tyi \fun (\tyi
+\fun \tyi \fun\tyo)\fun \tyi$ (for Replacement). The Axiom of
+Replacement
+% https://isabelle.in.tum.de/dist/library/FOL/FOL/IFOL.html#ZF_Base.replacement|axiom
+has a free predicate variable $P$: % $P ::\tyi\fun\tyi\fun\tyo$:
+\begin{multline*}
+  (\forall x \in A .\ \forall y\, z.\ P(x, y) \wedge P(x, z)
+  \longrightarrow y=z) \implies \\
+  b \in \isatt{PrimReplace}(A, P)
+  \longleftrightarrow(\exists x \in A .\ P(x, b)) 
+\end{multline*}
+The restrictions on sorts described above ensure that it is not
+possible that higher-order quantification gets used in $P$. The
+statement of $\AC$ also has a higher-order free variable, but we only
+use $\AC$ for demonstration purposes.
+
+Isabelle/ZF reaches Hessenberg's $|A|\cdot|A| = |A|$. Our decision
+(during 2017) to
+use this assistant was triggered by its constructibility
+library \citep{paulson_2003},
+% https://isabelle.in.tum.de/dist/library/ZF/ZF-Constructible/
+which contains the development of $L$, the proof that it satisfies
+$\AC$, and a version of the Reflection Principle. The latter was
+actually encoded as a series of instructions to Isabelle automatic
+proof tools that would prove each particular instance of reflection:
+This is an example of what was said at the end of the previous
+section.
+
+The development of relativization and absoluteness for classes $C::
+\isatt{i} \fun \isatt{o}$ follows the same pattern. Each particular
+concept was manually written in a relational form and relativized.
+Here, the contrast between the usual way one regards $\ZF$ as
+first-order theory in the language $\{\in \}$ and the mathematical
+practice of freely using defined concepts comes to the
+forefront. Assistants have refined mechanisms to cope with defined
+concepts (which also make their foundations a more complicated
+than plain first-order logic), and this is the only that  nontrivial
+mathematics can be formalized. But when studying relative interpretations, one
+usually assumes a spartan syntax and defines relativization by
+induction of formulas of the more succinct language. The approach
+taken in \session{ZF-Constructible} is to consider relativizations of
+the formulas (of type $\tyo$) that define each concept. For instance,
+in the case of unions, we find a relativization
+$\isatt{big{\uscore}union}:: (\isatt{i} \fun \isatt{o}) \fun \isatt{i}
+\fun \isatt{i} \fun \isatt{o}$ of the statement
+“$\union A = z$”:
+\[
+ \isatt{big{\uscore}union}(M,A,Z) \equiv \forall x[M].\ x \in z
+ \longleftrightarrow (\exists y[M].\ y \in A \land x \in y)
+\]
+where $\forall x[M]\dots$ stands for $\forall x.\ M(x)\limp \dots$,
+etc. The need to work with \emph{relational} presentations of defined
+concepts stems from the fact that the model-theoretic definition of
+$L$ requires working with set models and satisfaction, which is
+defined for codes of formulas.
+
+\subsection{Internalized formulas}
+
+\session{ZF-Constructible} defines the set $\formula$ of codes for
+first-order formulas. These, alongside with lists, are instances of
+Isabelle/ZF treatment of inductively defined datatypes; induction
+and recursion theorems for them are proved automatically.
+
+To avoid problems related to the binding of variables, de Bruijn
+indices are used instead. The satisfaction predicate
+$\isatt{sats}::\tyi\fun\tyi\fun\tyi\fun\tyo$ takes as arguments a set $M$, a list
+$\mathit{ms}\in\isatt{list}(M)$ for the assignment of free indices,
+and $\phi\in\formula$, and it is written
+$M,\mathit{ms}\models\phi$ in our formalization. This completes the
+description of the fourth and last formal layer of the development.
+
+Internalized formulas for most (but not all) of the relational
+concepts is obtained by guiding the automatic tactics. Actually,
+great many of the concepts in \session{ZF-Constructible} where
+internalized by hand, as is the case for union,
+%% \begin{multline*}
+%%   \isatt{big{\uscore}union{\uscore}fm}(A,z) \equiv \\
+%%   ((\cdot\forall\cdot\cdot0 \in succ(z)\cdot \longleftrightarrow
+%%   (\cdot\exists\cdot\cdot0 \in succ(succ(A))\cdot \land \cdot1 \in
+%%   0\cdot\cdot\cdot)\cdot\cdot)
+%% \end{multline*}
+for which we have the following satisfaction lemma:
+\begin{multline*}
+  x \in \omega \implies y \in \omega \implies \mathit{env} \in \isatt{list}(A)
+  \implies \\
+  (A, \mathit{env} \models \isatt{big{\uscore}union{\uscore}fm}(x,y)%\cdot\union x \isatt{ is } y\cdot
+  ) \longleftrightarrow  \isatt{big{\uscore}union}(\isatt{\#\#} A, \isatt{nth}(x,
+  \mathit{env}), \isatt{nth}(y, \mathit{env}))
+\end{multline*}
+Note that $x$ and $y$ above are de Bruijn indices,
+$\isatt{nth}(x,\mathit{env})$ is the $x$th element of $\mathit{env}$
+and $\isatt{\#\#}A::\tyi\fun \tyo$ is the class corresponding to the
+set $A::\tyi$.
 

### 2022-05-04 JSL/non-absolute.tex
New discipline

\section{Relative versions of non-absolute concepts}
\label{sec:relat-vers-non-absol}

The treatment of relativization/internalization described in the
previous section was enough for the Paulson's treatment of
constructibility. This is the case because essentially all the
concepts in the way of proving the consistency of $\AC$ are
absolute, and treatment of relational versions and relativized notions
could be minimized after proving the relevant absoluteness results:
For example, the lemma \isatt{Union{\uscore}abs},
\[
  M(A) \implies M(z) \implies \isatt{big{\uscore}union}(M, A, z) \longleftrightarrow z = \union
  A
\]
proved under the assumption that $M$ is transitive nonempty. Working in this
way with powersets, cardinalities, and the like would be
unfeasible.

In order to cope with this, we added the missing step from the
literature consisting of relative versions of the various non-absolute
functions and limited automatic facilities that define, state, and
prove the needed concepts. For instance, the 
$\isatt{cardinal}::\tyi \fun \tyi$ function is defined in
\session{Isabelle/ZF}, and the commands
\begin{isabelle}
  \isacommand{relativize}\isamarkupfalse%
  \ \isakeyword{functional}\ {\isachardoublequoteopen}cardinal{\isachardoublequoteclose}\ {\isachardoublequoteopen}cardinal{\isacharunderscore}{\kern0pt}rel{\isachardoublequoteclose}\ \isakeyword{external}\isanewline
  \isacommand{relationalize}\isamarkupfalse%
  \ {\isachardoublequoteopen}cardinal{\isacharunderscore}{\kern0pt}rel{\isachardoublequoteclose}\ {\isachardoublequoteopen}is{\isacharunderscore}{\kern0pt}cardinal{\isachardoublequoteclose}\isanewline
  \isacommand{synthesize}\isamarkupfalse%
  \ {\isachardoublequoteopen}is{\isacharunderscore}{\kern0pt}cardinal{\isachardoublequoteclose}\ \isakeyword{from{\isacharunderscore}{\kern0pt}definition}\ \isakeyword{assuming}\ {\isachardoublequoteopen}nonempty{\isachardoublequoteclose}%
\end{isabelle}
define the relative cardinal function
$\isatt{cardinal{\uscore}rel}::(\tyi \fun \tyo) \fun \tyi \fun\tyi$,
its relational version $\isatt{is{\uscore}cardinal}$ that encodes the
statement $|x|^M = z$, the
internalized formula \isatt{is{\uscore}cardinal{\uscore}fm} whose
satisfaction by a set is equivalent to the relational version, and the
proof of the previous statements.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "independence_ch_isabelle"
%%% ispell-local-dictionary: "american"
%%% End:

### 2022-05-04 JSL/proof_assists_isabelle.tex
New discipline

@@
 Internalized formulas for most (but not all) of the relational
-concepts is obtained by guiding the automatic tactics. Actually,
+concepts can be obtained by guiding the automatic tactics. Actually,
 great many of the concepts in \session{ZF-Constructible} where
-internalized by hand, as is the case for union,
+internalized by hand; this is the case for union,
 %% \begin{multline*}
@@
 for which we have the following satisfaction lemma:
+% https://isabelle.in.tum.de/dist/library/ZF/ZF-Constructible/L_axioms.html#L_axioms.sats_big_union_fm|fact
 \begin{multline*}

### 2022-05-06 JSL/forcing.tex
New discipline
More on relativization. Extensions to Isabelle/ZF. Few words on forcing

@@
 \label{sec:forcing}
-
-Prerequisite: a set $\formula$ of codes for first order formulas, and
-the (recursive) definition of the Tarski satisfaction relation in $\ZF$.
-
 Let $\lb \PP, {\preceq} ,\1\rb \in M$ be a forcing notion. Given $G\sbq \PP$, we have
@@
 \begin{theorem}
-  There exists a function  $\forceisa: \formula \to  \formula$
+  There exists a function  $\forceisa:: \tyi \fun  \tyi$
   such that for every
-  $\phi$ and $\punto{a}_0,\dots,\punto{a}_n\in M$,
+  $\phi\in\formula$ and $\punto{a}_0,\dots,\punto{a}_n\in M$,
   \begin{enumerate}
+  \item (Definability) $\forceisa(\phi)\in\formula$;
   \item (Truth Lemma) for every $M$-generic $G$,
@@
 
-Note: $\forceisa$ was formalized as a class-function (i.e., of type
-$\tyi\Rightarrow\tyi$), to take advantage of the Isabelle's
-conveniences for primitive recursion.
+We followed the new Kunen's book to define
+$\forceisa$.  Forcing for atomic formulas is described as a mutual
+recursion
+%% \begin{multline*}
+%%   \forceseq (p,t_1,t_2) \defi 
+%%   \forall s\in\dom(t_1)\cup\dom(t_2).\ \forall q\pleq p .\\
+%%   \forcesmem(q,s,t_1)\lsii \forcesmem(q,s,t_2)
+%% \end{multline*}
+%% \begin{multline*}
+%%   \forcesmem(p,t_1,t_2) \defi  \forall v\pleq p. \ \exists q\pleq v.\\  
+%%   \exists s.\ \exists r\in \PP .\ \lb s,r\rb \in  t_2 \land q
+%%   \pleq r \land \forceseq(q,t_1,s)
+%% \end{multline*}
+but then \cite[p.~257]{kunen2011set} it is cast as a single
+recursively defined function $\frcat$ over the wellfounded relation
+$\isatt{frecR}$ on tuples $\lb \mathit{ft},t_1,t_2,p\rb$ (where
+$\mathit{ft}\in\{0,1\}$ indicates the type of the atomic formula being
+forced). Forcing for general formulas is defined by recursion on the
+datatype $\formula$. Details on the implementation and proofs of the
+Forcing Theorems have been spelled out in our
+\cite{2020arXiv200109715G}.
+
+It is to be noted that application of the Forcing theorems do not
+require any extra Replacement instances on $M$.
 

### 2022-05-06 JSL/non-absolute.tex
More on relativization. Extensions to Isabelle/ZF. Few words on forcing

@@
 \]
-proved under the assumption that $M$ is transitive nonempty. Working in this
+proved under the assumption that $M$ is transitive nonempty. Working
+in this relational 
 way with powersets, cardinalities, and the like would be
-unfeasible.
+unfeasible. As such, cardinal arithmetic was not put in relative form
+in \session{ZF-Constructible}.
 
+\subsection{Tools for relativization}
+\label{sec:tools-relativization}
 In order to cope with this, we added the missing step from the
 literature consisting of relative versions of the various non-absolute
-functions and limited automatic facilities that define, state, and
-prove the needed concepts. For instance, the 
+functions and programmed in ML limited automatic facilities that
+define the needed concepts, and  state and
+prove the requisite lemmas. For instance, the 
 $\isatt{cardinal}::\tyi \fun \tyi$ function is defined in
@@
 define the relative cardinal function
-$\isatt{cardinal{\uscore}rel}::(\tyi \fun \tyo) \fun \tyi \fun\tyi$,
-its relational version $\isatt{is{\uscore}cardinal}$ that encodes the
-statement $|x|^M = z$, the
+$\isatt{cardinal{\uscore}rel}::(\tyi \fun \tyo) \fun \tyi \fun\tyi$
+(denoted  $|\cdot|^M$, as expected),
+its relational version $\isatt{is{\uscore}cardinal}$, the
 internalized formula \isatt{is{\uscore}cardinal{\uscore}fm} whose
-satisfaction by a set is equivalent to the relational version, and the
-proof of the previous statements.
+satisfaction by a set is equivalent to the relational version, and
+prove the previous statement (analogous to (\ref{eq:sats_big_union_fm})).
+The proof that $\isatt{is{\uscore}cardinal}(M,x,z)$  encodes the
+statement $|x|^M = z$ must still be done by hand, since the definition
+of $\isatt{cardinal{\uscore}rel}$ already involves some tacit
+absoluteness result (“\textit{the least $\ka \in \Ord$ such that \dots}” instead
+of “\textit{the least $\ka \in \Ord^M$ such that \dots}”, and the like).
+
+\subsection{Extension of Isabelle/ZF}
+\label{sec:extension-isabellezf}
+We extended \cite{Delta_System_Lemma-AFP} the material formalized in
+Isabelle, from basic results involving function spaces and the
+definition of cardinal exponentiation, to a treatment of cofinality
+and the Delta System Lemma for $\omega_1$-families. We also included a
+concise treatment of the axiom of Dependent Choices $\DC$ and the
+general version of Rasiowa-Sikorski Lemma \cite{2018arXiv180705174G}
+and a choiceless one for countable preorders.
+
+This material was subsequently put in relative form in our formal
+development on transitive class models \cite{Transitive_Models-AFP}
+using as an aid the tools from
+Section~\ref{sec:tools-relativization}. We also relativized many
+original theories appearing in \session{ZF}, including the
+fundamentals of cardinal arithmetic, the cumulative hierarchy, and the
+definition of the $\ale{}$ function.
 

### 2022-05-06 JSL/proof_assists_isabelle.tex
More on relativization. Extensions to Isabelle/ZF. Few words on forcing

@@
 \subsection{Isabelle metalogic $\Meta$}
+\label{sec:isabelle-metalogic-meta}
 
@@
 The type of propositions $\prop$ equipped with a binary operation
-$\implies$ (“meta-implication”) that is used to represent the object
+$\implies$ (“meta-implication”) and a universal “meta-quantifier”
+$\ALL$, that are used to represent the object
 logic rules. As an example, the axiomatization of first-order logic
@@
 \[
-  [P\limp Q] \implies ([P] \implies [Q]).
+  \ALL P\,Q.\ \ [P\limp Q] \implies ([P] \implies [Q]).
 \]
@@
 A consequence of this representation is that every formula of
-the object logic appears as atomic for $\Meta$.
+the object logic appears atomic to $\Meta$.
 
@@
 (“sets”) of sort $\{\type{term}\}$. Hence, from the type of the universal
-quantifier functional $\forall :: (\alpha \fun \tyo) \fun \tyo$ it
+quantifier functional $\forall :: (\alpha \fun \tyo) \fun \tyo$, it
 follows that it may only be applied to predicates with a variable of
@@
 \subsection{Isabelle/ZF}
+\label{sec:isabellezf}
 
@@
 proof tools that would prove each particular instance of reflection:
-This is an example of what was said at the end of the previous
-section.
+This is an example of what was said at the end of Section~\ref{sec:isabelle-metalogic-meta}.
 
@@
 \subsection{Internalized formulas}
+\label{sec:internalized-formulas}
 
@@
 first-order formulas. These, alongside with lists, are instances of
-Isabelle/ZF treatment of inductively defined datatypes; induction
-and recursion theorems for them are proved automatically.
+Isabelle/ZF treatment of inductively defined (internal) datatypes; induction
+and recursion theorems for them are proved automatically (this is in
+constrast to general well-founded recursion, for which one has to work
+with the fundamental recursor $\isatt{wfrec}$).
 
 To avoid problems related to the binding of variables, de Bruijn
-indices are used instead. The satisfaction predicate
+indices \cite{MR0321704} are used instead. The satisfaction predicate
 $\isatt{sats}::\tyi\fun\tyi\fun\tyi\fun\tyo$ takes as arguments a set $M$, a list
-$\mathit{ms}\in\isatt{list}(M)$ for the assignment of free indices,
+$\mathit{env}\in\isatt{list}(M)$ for the assignment of free indices,
 and $\phi\in\formula$, and it is written
-$M,\mathit{ms}\models\phi$ in our formalization. This completes the
+$M,\mathit{env}\models\phi$ in our formalization. This completes the
 description of the fourth and last formal layer of the development.
@@
 % https://isabelle.in.tum.de/dist/library/ZF/ZF-Constructible/L_axioms.html#L_axioms.sats_big_union_fm|fact
-\begin{multline*}
+\begin{multline}\label{eq:sats_big_union_fm}
   x \in \omega \implies y \in \omega \implies \mathit{env} \in \isatt{list}(A)
   \implies \\
-  (A, \mathit{env} \models \isatt{big{\uscore}union{\uscore}fm}(x,y)%\cdot\union x \isatt{ is } y\cdot
-  ) \longleftrightarrow  \isatt{big{\uscore}union}(\isatt{\#\#} A, \isatt{nth}(x,
+  A, \mathit{env} \models \isatt{big{\uscore}union{\uscore}fm}(x,y)%\cdot\union x \isatt{ is } y\cdot
+  \longleftrightarrow \\
+  \isatt{big{\uscore}union}(\isatt{\#\#} A, \isatt{nth}(x,
   \mathit{env}), \isatt{nth}(y, \mathit{env}))
-\end{multline*}
+\end{multline}
 Note that $x$ and $y$ above are de Bruijn indices,

### 2022-05-08 JSL/proof_assists_isabelle.tex
Minor edits

@@
 
-The type of propositions $\prop$ equipped with a binary operation
+The type of propositions $\prop$ is equipped with a binary operation
 $\implies$ (“meta-implication”) and a universal “meta-quantifier”
@@
 principles. Types are not inductively presented and, in particular, it
-is not possible to perform proofs by induction statements about
+is not possible to prove by induction statements about
 object-logic formulas (which are construed as terms of type $\tyi \fun

### 2022-05-08 JSL/random.tex
Starting the section on forcing. Liquid Tensor Experiment refs
Preliminary analysis of instances; a few words on proof assistants
types and sorts in Isabelle's M
Minor edits

@@
-\subsection{Random comments}
+\section{Random comments}
 
 \begin{enumerate}
-\item Satisfaction of having one's material fully formalized
-  (v.g. cofinality).
-\item The concept of absoluteness (\verb|_abs|) splitted into two components
-  during the formalization: \verb|is_foo_iff| and \verb|foo_rel_abs| (this last one
-  being the concept of absoluteness from  “mathematical practice”) and
-  the first one involves a purely relational presentation, introduced
-  by Paulson.
-\item
-  In order to extract information concerning the proof, the need for a
-  proof assistant with a more computational flavour (Agda, Coq)
-  arises! As a second thought, obviously.
 \item  \textbf{Pros and cons of an untyped environment}: Closer to the
@@
   arises! As a second thought, obviously.
-\item Promotion of formalization by Buzzard.
 \end{enumerate}

### 2022-05-10 JSL/details.tex
types and sorts in Isabelle's M
A bit more precise on instances

@@
 $\CH$/$\neg\CH$. The first 14 instances are needed to set up
-cardinal arithmetic in $M$.
+cardinal arithmetic in $M$. Many of these were already present in
+relational form in the \session{ZF-Constructible} library.
 
 \begin{itemize}
-\item 7 instances for iteration of operations.
+\item 7 instances for iteration of operations. For each construct,
+  Paulson used one replacement instance in order to have absoluteness,
+  and a second one to obtain closure.
   \begin{itemize}
-  \item Fixpoints: lists
+  \item
+    Lists
     \isatt{list{\uscore}repl1{\uscore}intf{\uscore}fm},
-    \isatt{list{\uscore}repl2{\uscore}intf{\uscore}fm}; formulas
+    \isatt{list{\uscore}repl2{\uscore}intf{\uscore}fm}.
+  \item
+    Formulas
     \isatt{formula{\uscore}repl1{\uscore}intf{\uscore}fm},
     \isatt{formula{\uscore}repl1{\uscore}intf{\uscore}fm}.
-  \item Transitive closure \isatt{eclose{\uscore}repl1{\uscore}intf{\uscore}fm},
-    \isatt{eclose{\uscore}repl2{\uscore}intf{\uscore}fm} (iteration of $\union$).
-  \item $n$th element of a list \isatt{tl{\uscore}repl{\uscore}intf{\uscore}fm} (iteration of the tail operation; fake?).
+  \item
+    Transitive closure (iteration of $\union$)
+    \isatt{eclose{\uscore}repl1{\uscore}intf{\uscore}fm},
+    \isatt{eclose{\uscore}repl2{\uscore}intf{\uscore}fm}.
+  \item
+    $n$th element of a list (iteration of the tail operation)
+    \isatt{tl{\uscore}repl{\uscore}intf{\uscore}fm}. By transitivity,
+    we do not need an instance for closure; the only instance was required
+    for absoluteness.
   \end{itemize}
-
-\item 4 instances for definitions by well-founded recursion.
-  \begin{itemize}
-  \item Ordertypes
-    \isatt{wfrec{\uscore}replacement{\uscore}order{\uscore}pred{\uscore}fm}.
-  \item Rank \isatt{wfrec{\uscore}rank{\uscore}fm}.
-  \item Cumulative hierarchy \isatt{trans{\uscore}repl{\uscore}HVFrom{\uscore}fm} (rank initial segments).
-  \item Aleph \isatt{replacement{\uscore}HAleph{\uscore}wfrec{\uscore}repl{\uscore}body{\uscore}fm}.
-  \end{itemize}
-
-\item 2 instances for definition of ordertypes.
+\item
+  2 instances for definition of ordertypes.
   \begin{itemize}
@@
     \isatt{replacement{\uscore}is{\uscore}jump{\uscore}cardinal{\uscore}body{\uscore}fm}.
+  \end{itemize}
+\item
+  4 instances for definitions by well-founded recursion.
+  \begin{itemize}
+  \item Ordertypes
+    \isatt{wfrec{\uscore}replacement{\uscore}order{\uscore}pred{\uscore}fm}.
+  \item Rank \isatt{wfrec{\uscore}rank{\uscore}fm}.
+  \item Cumulative hierarchy (rank initial segments) \isatt{trans{\uscore}repl{\uscore}HVFrom{\uscore}fm}.
+  \item Aleph \isatt{replacement{\uscore}HAleph{\uscore}wfrec{\uscore}repl{\uscore}body{\uscore}fm}.
   \end{itemize}
@@
 \]
+In our development, the mapping $\phi\mapsto\psi$ is given by the
+$\isatt{ground{\uscore}repl{\uscore}fm}$ function.
+
 That makes 28 instances up to now. For the setup of forcing, we
-require the following 5 instances (the first two needed for the $\Delta$-system Lemma):
+require the following 5 instances (the first two are  needed for the $\Delta$-system Lemma):
 

### 2022-05-10 JSL/details.tex
Statement of the axioms, locales

@@
 \subsection{33 replacement instances to rule them all}
-We isolated 33 instances of Replacement that are needed to force
+\label{sec:33-repl-inst-rule}
+
+We isolated 33 instances of Replacement that are sufficient to force
 $\CH$/$\neg\CH$. The first 14 instances are needed to set up
@@
   \item
-    Transitive closure (iteration of $\union$)
+    Transitive closure %(iteration of $\union$)
     \isatt{eclose{\uscore}repl1{\uscore}intf{\uscore}fm},
@@
 \]
-In our development, the mapping $\phi\mapsto\psi$ is given by the
-$\isatt{ground{\uscore}repl{\uscore}fm}$ function.
+In our development, the mapping $\phi\mapsto\psi$ defined above is given by the
+$\isatt{ground{\uscore}repl{\uscore}fm}$ function, and all ground replacement
+instances appear in the locale \isatt{M{\uscore}ZF4}. These are expressed using
+the \isatt{ground{\uscore}replacement{\uscore}assm(M,\isasymphi)} predicate
+obtained by replacing $\phi$ by
+$\isatt{ground{\uscore}repl{\uscore}fm}(\phi)$ in Eq.~(\ref{eq:replacement_assm_def}).
 
@@
   \isatt{Lambda{\uscore}in{\uscore}M{\uscore}fm(is{\uscore}cardinal{\uscore}fm(0,1),0)}.
-\item 2 instances for definitions by well-founded recursion \isatt{wfrec{\uscore}Hcheck{\uscore}fm} (check
-  names) and \isatt{wfrec{\uscore}Hfrc{\uscore}at{\uscore}fm} (forcing for atomic formulas).
-\item Replacement through $x\mapsto \lb\check{x},x\rb$
-  \isatt{G{\uscore}dot{\uscore}in{\uscore}M{\uscore}fm} (for the
-  definition of $\punto{G}$).
+\item 2 instances for definitions by well-founded recursion.
+  \begin{itemize}
+  \item
+    Check names
+    \isatt{wfrec{\uscore}Hcheck{\uscore}fm}.
+  \item Forcing for atomic formulas
+    \isatt{wfrec{\uscore}Hfrc{\uscore}at{\uscore}fm}.
+  \end{itemize}
+\item Replacement through $x\mapsto \lb\check{x},x\rb$ (for the
+  definition of $\punto{G}$)
+  \isatt{G{\uscore}dot{\uscore}in{\uscore}M{\uscore}fm}.
 % \item Name of choice functions in $M[G]$ \isatt{repl{\uscore}opname{\uscore}check{\uscore}fm}.
@@
 \end{itemize}
+
+It is to be noted that application of the Forcing Theorems do not
+require any extra Replacement instances on $M$ (independently of the
+formula $\phi$ for which they are invoked). This is not the case for
+Separation, at least from our formalization: More instances are needed
+as the complexity of $\phi$ grows. One point where this is apparent is
+in the proof of Theorem~\ref{th:forcing-thms}(\ref{item:truth-lemma}),
+that appears as the \isatt{truth{\uscore}lemma} in our development; it
+depends on \isatt{truth{\uscore}lemma'} and
+\isatt{truth{\uscore}lemma{\uscore}Neg} which explicitly invoke
+\isatt{separation{\uscore}ax}.
 

### 2022-05-10 JSL/forcing.tex
Statement of the axioms, locales

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
-\section{Forcing}
+\section{Set models and forcing}
 \label{sec:forcing}
-Let $\lb \PP, {\preceq} ,\1\rb \in M$ be a forcing notion. Given $G\sbq \PP$, we have
+
+\subsection{The $\ZFC$ axioms as locales}
+The description of set models of fragments of $\ZFC$ was performed
+using Isabelle contexts (\emph{locales}) that fix a variable $M::\tyi$
+and pack assumptions stating that $\lb M, \in\rb$ satisfy some
+axioms. For example, the locale \isatt{M{\uscore}Z{\uscore}basic}
+states that Zermelo set theory holds in $M$. The Union Axiom (\isatt{Union{\uscore}ax}), for
+instance, is defined as follows:
+\[
+\forall x[\isatt{\#\#}M].\ \exists z[\isatt{\#\#}M].\ \isatt{big{\uscore}union}(\isatt{\#\#}M,x,z)
+\]
+%% % Same a above
+%% \begin{isabelle}%
+%% {\isasymforall}x{\isacharbrackleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymexists}z{\isacharbrackleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ big{\isacharunderscore}{\kern0pt}union{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}M{\isacharcomma}{\kern0pt}\ x{\isacharcomma}{\kern0pt}\ z{\isacharparenright}{\kern0pt}%
+%% \end{isabelle}%
+using relativized, relational versions of the axioms of Isabelle/ZF
+since the interaction with \session{ZF-Constructible} was smoother
+that way and, as it was mentioned in Section~\ref{sec:isabellezf},
+this third layer of the formalization has more tools at our
+disposal. Later, an equivalent statement “in the codes” (our fourth
+layer) is obtained,
+\[
+  \isatt{Union{\uscore}ax}(\isatt{\#\#}M) \lsii A, [\,]\models \isatt{{\isasymcdot}Union\ Ax{\isasymcdot}}
+\]
+%% % Same a above
+%% \begin{isabelle}
+%% Union{\isacharunderscore}{\kern0pt}ax{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}A{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\ A{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}Union\ Ax{\isasymcdot}\isasep
+%% \end{isabelle}
+where \isatt{{\isasymcdot}Union\ Ax{\isasymcdot}} is the $\formula$ code for the
+Union axiom. For the axiom schemes, \session{ZF-Constructible} defines
+the expressions
+\[
+  \isatt{separation}(N,Q)
+  \text{ and }
+  \isatt{strong{\uscore}replacement}(N,R)
+\]
+that consume a class $N$ and predicates $Q::\tyi\fun\tyo$ and $R::\tyi\fun\tyi\fun\tyo$, and state that $N$
+satisfies the $Q$-instance of Separation and the $R$-instance of
+Replacement, respectively. In the statement of the Separation Axiom in
+\isatt{M{\uscore}Z{\uscore}basic} and
+the many replacement instances, predicates $Q$ and $R$ are actually
+defined as the of formulas of the appropriate arity; we can only show
+that this form of the axioms hold in generic extensions.
+
+Further locales gather the assumption of transitivity of $M$ and
+particular replacement instances expressed by means of the
+\isatt{replacement{\uscore}assm(M,\isasymphi)} predicate:
+\begin{multline}\label{eq:replacement_assm_def}
+\varphi \in \formula  \limp \mathit{env} \in \isatt{list}(M) \limp \isatt{arity}(\varphi) \leq 2+ \isatt{length}(\mathit{env}) \limp \\
+ \isatt{strong{\uscore}replacement}(\isatt{\#\#} M, \lambda x\, y.\ (M, [x,y]
+\mathbin{@} \mathit{env}  \models \varphi))
+\end{multline}
+%% % Same as above
+%% \begin{isabelle}
+%% {\isasymphi}\;{\isasymin}\;formula\ {\isasymlongrightarrow}\ env\;{\isasymin}\;list{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow} arity{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}\;{\isasymle}\;{\isadigit{2}}\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ length{\isacharparenleft}{\kern0pt}env{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\isanewline
+%% \ \ \ \ strong{\isacharunderscore}{\kern0pt}replacement{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}M{\isacharcomma}{\kern0pt}{\isasymlambda}x\ y{\isachardot}{\kern0pt}\ {\isacharparenleft}{\kern0pt}M\ {\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}x{\isacharcomma}{\kern0pt}y{\isacharbrackright}{\kern0pt}{\isacharat}{\kern0pt}env\ {\isasymTurnstile}\ {\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}
+%% \end{isabelle}
+Some of those instances, which were assumed in the locales of
+\session{ZF-Constructible} and we included for simplicity of design,
+are actually “fake” instances in that they can be obtained from
+Powerset. An example of this is the sole instance of the
+\isatt{M{\uscore}basic} locale; we managed to eliminate it but in
+doing so we had to reprove many lemmas from
+\session{ZF-Constructible} in the weakened context. Our locales
+\isatt{M{\uscore}ZF1} through \isatt{M{\uscore}ZF4} list a total of 86
+replacement instances, from which we identified 53 fake ones (see
+Section~\ref{sec:33-repl-inst-rule}).
+
+\subsection{The fundamental theorems}
+Now let $\lb \PP, {\preceq} ,\1\rb \in M$ be a forcing notion. Given $G\sbq \PP$, we have
 $M[G]\defi \{ \val(\PP,G,\punto{a}) : \punto{a}\in M \}$.
@@
 that we formalized.
-\begin{theorem}
+\begin{theorem}\label{th:forcing-thms}
   There exists a function  $\forceisa:: \tyi \fun  \tyi$
@@
   \begin{enumerate}
-  \item (Definability) $\forceisa(\phi)\in\formula$;
-  \item (Truth Lemma) for every $M$-generic $G$,
+  \item\label{item:definability} (Definability)
+    $\forceisa(\phi)\in\formula$,
+  \end{enumerate}
+  and if
+  “$p \forces \phi\ [\punto{a}_0,\dots,\punto{a}_n]$”
+  denotes
+  “$M, [p,\PP,\preceq,\1, \punto{a}_0,\dots,\punto{a}_n]  \models
+  \forceisa(\phi)$”, we have:
+  \begin{enumerate}
+    \setcounter{enumi}{1}
+  \item\label{item:truth-lemma} (Truth Lemma) for every $M$-generic $G$,
     \[
-      M[G], [\val(\PP,G,\punto{a}_0),\dots,\val(\PP,G,\punto{a}_n)]
-      \models \phi
+      \exists p\in G.\ \; p \forces \phi\ [\punto{a}_0,\dots,\punto{a}_n]
     \]
@@
     \[
-      \exists p\in G.\ \; M, [p,\PP,\preceq,\1, \punto{a}_0,\dots,\punto{a}_n]  \models
-      \forceisa(\phi).\]
-
-    We use the notation $p \forces
-    \phi\ [\punto{a}_0,\dots,\punto{a}_n]$ for this last assertion.
-  \item (Density Lemma) $p \forces \phi\ [\punto{a}_0,\dots,\punto{a}_n]$
+      M[G], [\val(\PP,G,\punto{a}_0),\dots,\val(\PP,G,\punto{a}_n)]
+      \models \phi.
+    \]
+  \item \label{item:density-lemma} (Density Lemma) $p \forces \phi\ [\punto{a}_0,\dots,\punto{a}_n]$
     if and only if 
@@
 \end{theorem}
+The statements appear as three separated items in
+\theory{Forcing{\uscore}Theorems.thy} in the \session{Independence\_CH} session,
+and benefit from the $\isatt{map}$ function that applies a function to
+each element of a list. For instance, the Truth Lemma is stated as
+follows:
+\begin{isabelle}
+\isacommand{lemma}\isamarkupfalse%
+\ truth{\isacharunderscore}{\kern0pt}lemma{\isacharcolon}{\kern0pt}\isanewline
+\ \ \isakeyword{assumes}\isanewline
+\ \ \ \ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}M{\isacharunderscore}{\kern0pt}generic{\isacharparenleft}{\kern0pt}G{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\isanewline
+\ \ \ \ {\isachardoublequoteopen}env{\isasymin}list{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\ {\isachardoublequoteopen}arity{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymle}length{\isacharparenleft}{\kern0pt}env{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\isanewline
+\ \ \ \ {\isachardoublequoteopen}{\isacharparenleft}{\kern0pt}{\isasymexists}p{\isasymin}G{\isachardot}{\kern0pt}\ p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}{\kern0pt}\ \ \ {\isasymlongleftrightarrow}\ \ \ M{\isacharbrackleft}{\kern0pt}G{\isacharbrackright}{\kern0pt}{\isacharcomma}{\kern0pt}\ map{\isacharparenleft}{\kern0pt}val{\isacharparenleft}{\kern0pt}P{\isacharcomma}{\kern0pt}G{\isacharparenright}{\kern0pt}{\isacharcomma}{\kern0pt}env{\isacharparenright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymphi}{\isachardoublequoteclose}
+\end{isabelle}
+where the $\forces$ notation had already been set up in the theory
+\theory{Forces{\uscore}Definition} as follows:
+\begin{isabelle}
+\isacommand{abbreviation}\isamarkupfalse%
+\ Forces\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}{\isacharbrackleft}{\kern0pt}i{\isacharcomma}{\kern0pt}\ i{\isacharcomma}{\kern0pt}\ i{\isacharbrackright}{\kern0pt}\ {\isasymRightarrow}\ o{\isachardoublequoteclose}\ \ {\isacharparenleft}{\kern0pt}{\isachardoublequoteopen}{\isacharunderscore}{\kern0pt}\ {\isasymtturnstile}\ {\isacharunderscore}{\kern0pt}\ {\isacharunderscore}{\kern0pt}{\isachardoublequoteclose}\ {\isacharbrackleft}{\kern0pt}{\isadigit{3}}{\isadigit{6}}{\isacharcomma}{\kern0pt}{\isadigit{3}}{\isadigit{6}}{\isacharcomma}{\kern0pt}{\isadigit{3}}{\isadigit{6}}{\isacharbrackright}{\kern0pt}\ {\isadigit{6}}{\isadigit{0}}{\isacharparenright}{\kern0pt}\ \isakeyword{where}\isanewline
+\ \ {\isachardoublequoteopen}p\ {\isasymtturnstile}\ {\isasymphi}\ env\ \ \ {\isasymequiv}\ \ \ M{\isacharcomma}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isacharbrackleft}{\kern0pt}p{\isacharcomma}{\kern0pt}P{\isacharcomma}{\kern0pt}leq{\isacharcomma}{\kern0pt}{\isasymone}{\isacharbrackright}{\kern0pt}\ {\isacharat}{\kern0pt}\ env{\isacharparenright}{\kern0pt}\ {\isasymTurnstile}\ forces{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\isanewline
+\end{isabelle}
 
-We followed the new Kunen's book to define
+We followed Kunen's new \textit{Set Theory} \cite{kunen2011set} to define
 $\forceisa$.  Forcing for atomic formulas is described as a mutual
@@
 
-It is to be noted that application of the Forcing theorems do not
-require any extra Replacement instances on $M$.
-
 %%% Local Variables: 

### 2022-05-10 JSL/proof_assists_isabelle.tex
Statement of the axioms, locales

@@
 \cite{MR0387024}. Passing to classical logic does not present
-difficulties, but the treatment of meta-equality differs between the
+difficulties, but the addition of meta-equality must be taken care of.
+Even more so, since the treatment of equality differs between the
 original and the present incarnation of $\Meta$; details for the
@@
 carried out using its ZF object logic
-\cite{DBLP:journals/jar/PaulsonG96}. Apart from the type and sort
+\cite{DBLP:journals/jar/PaulsonG96}, which is the third logical layer
+of the formalization and the most versatile one, since 
+Isabelle's native automation is available at this level. Apart from
+the type and sort
 declarations detailed above, it features a finite axiomatization,
@@
 use this assistant was triggered by its constructibility
-library \citep{paulson_2003},
+library, \session{ZF-Constructible} \citep{paulson_2003},
 % https://isabelle.in.tum.de/dist/library/ZF/ZF-Constructible/

### 2022-05-11 Independence_CH/forcing.tex
Target journal now unspecified

--- 2022-05-10 JSL/forcing.tex
@@
 particular replacement instances expressed by means of the
-\isatt{replacement{\uscore}assm(M,\isasymphi)} predicate:
+\isatt{replacement{\uscore}assm(M,\isasymphi)} predicate ($@$ denotes
+list concatenation):
 \begin{multline}\label{eq:replacement_assm_def}
@@
 \isatt{M{\uscore}ZF1} through \isatt{M{\uscore}ZF4} list a total of 86
-replacement instances, from which we identified 53 fake ones (see
+replacement instances, from which there are at least 53 fake ones (see
 Section~\ref{sec:33-repl-inst-rule}).
@@
   \end{enumerate}
-  and if
+  where the 
+  arity of $\forceisa(\phi)$ is at most $\isatt{arity}(\phi) + 4$; and if
   “$p \forces \phi\ [\punto{a}_0,\dots,\punto{a}_n]$”
@@
 forced). Forcing for general formulas is defined by recursion on the
-datatype $\formula$. Details on the implementation and proofs of the
+datatype $\formula$. Technical details on the implementation and proofs of the
 Forcing Theorems have been spelled out in our

### 2022-05-11 Independence_CH/intro.tex
Target journal now unspecified

--- 2022-03-31 JSL/intro.tex
@@
 “an unnatural tendency to investigate, for the most part, trivial
-minutiae of the formalism”, as it was put by Cohen
+minutiae of the formalism,” as it was put by Cohen
 \cite{zbMATH02012060}, but we were also drove by curiosity.

### 2022-05-11 Independence_CH/non-absolute.tex
Target journal now unspecified

--- 2022-05-06 JSL/non-absolute.tex
@@
 The treatment of relativization/internalization described in the
-previous section was enough for the Paulson's treatment of
+previous sections was enough for the Paulson's treatment of
 constructibility. This is the case because essentially all the
@@
 of $\isatt{cardinal{\uscore}rel}$ already involves some tacit
-absoluteness result (“\textit{the least $\ka \in \Ord$ such that \dots}” instead
-of “\textit{the least $\ka \in \Ord^M$ such that \dots}”, and the like).
+absoluteness results (“\textit{the least $z \in \Ord$ such that $z
+  \approx^M x$}” instead
+of “\textit{the least $z \in \Ord^M$ such that $z
+  \approx^M x$}”, and the like).
 

### 2022-05-11 Independence_CH/proof_assists_isabelle.tex
Target journal now unspecified

--- 2022-05-10 JSL/proof_assists_isabelle.tex
@@
 $\isatt{sats}::\tyi\fun\tyi\fun\tyi\fun\tyo$ takes as arguments a set $M$, a list
-$\mathit{env}\in\isatt{list}(M)$ for the assignment of free indices,
+$\mathit{env}\in\isatt{list}(M)$ for the assignment of free indices
+(which are counted by the \isatt{arity} function),
 and $\phi\in\formula$, and it is written

### 2022-05-11 Independence_CH/related.tex
Target journal now unspecified

--- 2022-03-26 JSL/related.tex
@@
 
+\subsection{Random comments}
+
+\begin{enumerate}
+\item  \textbf{Pros and cons of an untyped environment}: Closer to the
+  set-theoretic point of view/prone to the most stupid errors.
+\item \textbf{“Pros” and cons of lack of automation}: we had to be extremely
+  detailed, which provides explicit information / the formalization
+  work was pain in the neck and much lower.
+\item In order to extract information concerning the proof, the need for a
+  proof assistant with a more computational flavour (Agda, Coq)
+  arises! As a second thought, obviously.
+\end{enumerate}
+
 %%% Local Variables: 

### 2022-05-11 Independence_CH/details.tex
Sample formal proof

\section{Details of the formalization}

\subsection{33 replacement instances to rule them all}
\label{sec:33-repl-inst-rule}

We isolated 33 instances of Replacement that are sufficient to force
$\CH$/$\neg\CH$. The first 14 instances are needed to set up
cardinal arithmetic in $M$. Many of these were already present in
relational form in the \session{ZF-Constructible} library.

\begin{itemize}
\item 7 instances for iteration of operations. For each construct,
  Paulson used one replacement instance in order to have absoluteness,
  and a second one to obtain closure.
  \begin{itemize}
  \item
    Lists
    \isatt{list{\uscore}repl1{\uscore}intf{\uscore}fm},
    \isatt{list{\uscore}repl2{\uscore}intf{\uscore}fm}.
  \item
    Formulas
    \isatt{formula{\uscore}repl1{\uscore}intf{\uscore}fm},
    \isatt{formula{\uscore}repl1{\uscore}intf{\uscore}fm}.
  \item
    Transitive closure %(iteration of $\union$)
    \isatt{eclose{\uscore}repl1{\uscore}intf{\uscore}fm},
    \isatt{eclose{\uscore}repl2{\uscore}intf{\uscore}fm}.
  \item
    $n$th element of a list (iteration of the tail operation)
    \isatt{tl{\uscore}repl{\uscore}intf{\uscore}fm}. By transitivity,
    we do not need an instance for closure; the only instance was required
    for absoluteness.
  \end{itemize}
\item
  2 instances for definition of ordertypes.
  \begin{itemize}
  \item \isatt{replacement{\uscore}is{\uscore}order{\uscore}body{\uscore}fm}.
  \item \isatt{replacement{\uscore}is{\uscore}order{\uscore}eq{\uscore}map{\uscore}fm}.
  \end{itemize}
\item 1 instance for definition of cardinal successor.
  \begin{itemize}
  \item
    \isatt{replacement{\uscore}is{\uscore}jump{\uscore}cardinal{\uscore}body{\uscore}fm}.
  \end{itemize}
\item
  4 instances for definitions by well-founded recursion.
  \begin{itemize}
  \item Ordertypes
    \isatt{wfrec{\uscore}replacement{\uscore}order{\uscore}pred{\uscore}fm}.
  \item Rank \isatt{wfrec{\uscore}rank{\uscore}fm}.
  \item Cumulative hierarchy (rank initial segments) \isatt{trans{\uscore}repl{\uscore}HVFrom{\uscore}fm}.
  \item Aleph \isatt{replacement{\uscore}HAleph{\uscore}wfrec{\uscore}repl{\uscore}body{\uscore}fm}.
  \end{itemize}
\end{itemize}

We also need a one extra replacement instance $\psi$ on $M$ for each
$\phi$ of the
previous ones to have them in $M[G]$.
\[
  \psi(x,\alpha,y_1,\dots,y_n) \defi \quine{\alpha = \min \bigl\{
    \beta \mid \exists\tau\in V_\beta.\  \mathit{snd}(x) \forces
    \phi\ [\mathit{fst}(x),\tau,y_1,\dots,y_n]\bigr\}}
\]
In our development, the mapping $\phi\mapsto\psi$ defined above is given by the
$\isatt{ground{\uscore}repl{\uscore}fm}$ function, and all ground replacement
instances appear in the locale \isatt{M{\uscore}ZF4}. These are expressed using
the \isatt{ground{\uscore}replacement{\uscore}assm(M,\isasymphi)} predicate
obtained by replacing $\phi$ by
$\isatt{ground{\uscore}repl{\uscore}fm}(\phi)$ in Eq.~(\ref{eq:replacement_assm_def}).

That makes 28 instances up to now. For the setup of forcing, we
require the following 5 instances (the first two are  needed for the $\Delta$-system Lemma):

\begin{itemize}
%\verb|replacement_ax2(6)| \verb|replacement_transrec_apply_image_body_fm| seems fake!
\item Recursive construction of sets using a choice function
  \isatt{replacement{\uscore}is{\uscore}trans{\uscore}apply{\uscore}image{\uscore}fm}.
% \item Replacement through  $x\mapsto \check{x}$ \isatt{check{\uscore}replacement{\uscore}fm}.
\item $x\mapsto |x|^M$
  \isatt{Lambda{\uscore}in{\uscore}M{\uscore}fm(is{\uscore}cardinal{\uscore}fm(0,1),0)}.
\item 2 instances for definitions by well-founded recursion.
  \begin{itemize}
  \item
    Check names
    \isatt{wfrec{\uscore}Hcheck{\uscore}fm}.
  \item Forcing for atomic formulas
    \isatt{wfrec{\uscore}Hfrc{\uscore}at{\uscore}fm}.
  \end{itemize}
\item Replacement through $x\mapsto \lb\check{x},x\rb$ (for the
  definition of $\punto{G}$)
  \isatt{G{\uscore}dot{\uscore}in{\uscore}M{\uscore}fm}.
% \item Name of choice functions in $M[G]$ \isatt{repl{\uscore}opname{\uscore}check{\uscore}fm}.
% \item 2 instances for the slalom needed for cardinal preservation by ccc forcing.
\end{itemize}

It is to be noted that application of the Forcing Theorems do not
require any extra Replacement instances on $M$ (independently of the
formula $\phi$ for which they are invoked). This is not the case for
Separation, at least from our formalization: More instances are needed
as the complexity of $\phi$ grows. One point where this is apparent is
in the proof of Theorem~\ref{th:forcing-thms}(\ref{item:truth-lemma}),
that appears as the \isatt{truth{\uscore}lemma} in our development; it
depends on \isatt{truth{\uscore}lemma'} and
\isatt{truth{\uscore}lemma{\uscore}Neg} which explicitly invoke
\isatt{separation{\uscore}ax}.

\subsection{A sample formal proof}
\label{sec:sample-formal-proof}

We present a fragment of the formal version of Kunen's proof that the
Powerset Axiom holds in a generic extension. We quote the relevant
paragraph of \cite[Thm.~IV.2.27]{kunen2011set}:
\begin{quote}
  For Power Set (similarly to Union above), it is sufficient to prove
  that whenever $a \in M[G]$, there is a $b \in M[G]$ such that
  $\mathcal{P}(a) \cap M[G] \subseteq b$. Fix $\tau \in
  M^{\mathbb{P}}$ such that $\tau_{G}=a$. Let
  $Q=(\mathcal{P}(\operatorname{dom}(\tau) \times
  \mathbb{P}))^{M}$. This is the set of all names $\vartheta \in
  M^{\mathcal{P}}$ such that $\operatorname{dom}(\vartheta) \subseteq
  \operatorname{dom}(\tau)$. Let $\pi=Q \times\{\1\}$ and let
  $b=\pi_{G}=$ $\left\{\vartheta_{G}: \vartheta \in Q\right\}$. Now,
  consider any $c \in \mathcal{P}(a) \cap M[G]$; we need to show that
  $c \in b$. Fix $\varkappa \in M^{\mathbb{P}}$ such that
  $\varkappa_{G}=c$, and let $\vartheta=\{\langle\sigma, p\rangle:
  \sigma \in \operatorname{dom}(\tau) \wedge p \Vdash \sigma \in
  \varkappa\}$; $\vartheta \in M$ by the Definability Lemma. Since
  $\vartheta \in Q$, we are done if we can show that
  $\vartheta_{G}=c$.
\end{quote}
The assumption $a\in M[G]$ appears in the lemma statement, and the
goal involving $b$ in the first sentence will appear below (page \pageref{goal-on-b}); formalized
material necessarily tends to be much more linear than usual prose. In
what follows, we
will intersperse the relevant passages of the proof.
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ Pow{\isacharunderscore}{\kern0pt}inter{\isacharunderscore}{\kern0pt}MG{\isacharcolon}{\kern0pt}\isanewline
\ \ \isakeyword{assumes}\ {\isachardoublequoteopen}a{\isasymin}M{\isacharbrackleft}{\kern0pt}G{\isacharbrackright}{\kern0pt}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\ {\isachardoublequoteopen}Pow{\isacharparenleft}{\kern0pt}a{\isacharparenright}{\kern0pt}\ {\isasyminter}\ M{\isacharbrackleft}{\kern0pt}G{\isacharbrackright}{\kern0pt}\ {\isasymin}\ M{\isacharbrackleft}{\kern0pt}G{\isacharbrackright}{\kern0pt}{\isachardoublequoteclose}\isanewline
%
\isacommand{proof}\isamarkupfalse%
\ {\isacharminus}{\kern0pt}
\end{isabelle}
\textit{Fix $\tau \in  M^{\mathbb{P}}$ such that $\tau_{G}=a$.}
\begin{isabelle}
\ \ \isacommand{from}\isamarkupfalse%
\ assms\isanewline
\ \ \isacommand{obtain}\isamarkupfalse%
\ {\isasymtau}\ \isakeyword{where}\ {\isachardoublequoteopen}{\isasymtau}\ {\isasymin}\ M{\isachardoublequoteclose}\ {\isachardoublequoteopen}val{\isacharparenleft}{\kern0pt}P{\isacharcomma}{\kern0pt}G{\isacharcomma}{\kern0pt}\ {\isasymtau}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ a{\isachardoublequoteclose}\isanewline
\ \ \ \ \isacommand{using}\isamarkupfalse%
\ GenExtD\ \isacommand{by}\isamarkupfalse%
\ auto
\end{isabelle}
\textit{Let
  $Q=(\mathcal{P}(\operatorname{dom}(\tau) \times
  \mathbb{P}))^{M}$.}
\begin{isabelle}
\ \ \isacommand{let}\isamarkupfalse%
\ {\isacharquery}{\kern0pt}Q{\isacharequal}{\kern0pt}{\isachardoublequoteopen}Pow{\isacharparenleft}{\kern0pt}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}{\isasymtimes}P{\isacharparenright}{\kern0pt}\ {\isasyminter}\ M{\isachardoublequoteclose}
\end{isabelle}
\textit{This is the set of all names $\vartheta \in
  M^{\mathcal{P}}$\dots}---it is pretty laborious to show that things
are in $M$; it takes 17 more lines of code (not shown below) that
apply further previously proved lemmas.
\begin{isabelle}
\ \ \isacommand{from}\isamarkupfalse%
\ {\isacartoucheopen}{\isasymtau}{\isasymin}M{\isacartoucheclose}\isanewline
\ \ \isacommand{have}\isamarkupfalse%
\ {\isachardoublequoteopen}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}{\isasymtimes}P\ {\isasymin}\ M{\isachardoublequoteclose}\ {\isachardoublequoteopen}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}\ {\isasymin}\ M{\isachardoublequoteclose}\isanewline
\ \ \ \ \isacommand{using}\isamarkupfalse%
\ domain{\isacharunderscore}{\kern0pt}closed\ cartprod{\isacharunderscore}{\kern0pt}closed\ P{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}M\isanewline
\ \ \ \ \isacommand{by}\isamarkupfalse%
\ simp{\isacharunderscore}{\kern0pt}all\isanewline
\ \ \isacommand{then}\isamarkupfalse%
\isanewline
\ \ \isacommand{have}\isamarkupfalse%
\ {\isachardoublequoteopen}{\isacharquery}{\kern0pt}Q\ {\isasymin}\ M{\isachardoublequoteclose}
\end{isabelle}
[\dots]\textit{ Let $\pi=Q \times\{\1\}$ and let
  $b=\pi_{G}=$ $\left\{\vartheta_{G}: \vartheta \in Q\right\}$.}
\begin{isabelle}
\ \ \isacommand{let}\isamarkupfalse%
\ {\isacharquery}{\kern0pt}{\isasympi}{\isacharequal}{\kern0pt}{\isachardoublequoteopen}{\isacharquery}{\kern0pt}Q{\isasymtimes}{\isacharbraceleft}{\kern0pt}{\isasymone}{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline
\ \ \isacommand{let}\isamarkupfalse%
\ {\isacharquery}{\kern0pt}b{\isacharequal}{\kern0pt}{\isachardoublequoteopen}val{\isacharparenleft}{\kern0pt}P{\isacharcomma}{\kern0pt}G{\isacharcomma}{\kern0pt}{\isacharquery}{\kern0pt}{\isasympi}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}%% \isanewline
%% \ \ \isacommand{from}\isamarkupfalse%
%% \ {\isacartoucheopen}{\isacharquery}{\kern0pt}Q{\isasymin}M{\isacartoucheclose}\isanewline
%% \ \ \isacommand{have}\isamarkupfalse%
%% \ {\isachardoublequoteopen}{\isacharquery}{\kern0pt}{\isasympi}{\isasymin}M{\isachardoublequoteclose}\isanewline
%% \ \ \ \ \isacommand{using}\isamarkupfalse%
%% \ one{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}P\ P{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}M\ transitivity\isanewline
%% \ \ \ \ \isacommand{by}\isamarkupfalse%
%% \ {\isacharparenleft}{\kern0pt}simp\ flip{\isacharcolon}{\kern0pt}\ setclass{\isacharunderscore}{\kern0pt}iff{\isacharparenright}{\kern0pt}\isanewline
%% \ \ \isacommand{then}\isamarkupfalse%
%% \isanewline
%% \ \ \isacommand{have}\isamarkupfalse%
%% \ {\isachardoublequoteopen}{\isacharquery}{\kern0pt}b\ {\isasymin}\ M{\isacharbrackleft}{\kern0pt}G{\isacharbrackright}{\kern0pt}{\isachardoublequoteclose}\isanewline
%% \ \ \ \ \isacommand{using}\isamarkupfalse%
%% \ GenExtI\ \isacommand{by}\isamarkupfalse%
%% \ simp
\end{isabelle}
\textit{Now,
  consider any $c \in \mathcal{P}(a) \cap M[G]$; we need to show that
  $c \in b$.}
\begin{isabelle}
  \label{goal-on-b}
\ \ \isacommand{have}\isamarkupfalse%
\ {\isachardoublequoteopen}Pow{\isacharparenleft}{\kern0pt}a{\isacharparenright}{\kern0pt}\ {\isasyminter}\ M{\isacharbrackleft}{\kern0pt}G{\isacharbrackright}{\kern0pt}\ {\isasymsubseteq}\ {\isacharquery}{\kern0pt}b{\isachardoublequoteclose}\isanewline
\ \ \isacommand{proof}\isamarkupfalse%
\isanewline
\ \ \ \ \isacommand{fix}\isamarkupfalse%
\ c\isanewline
\ \ \ \ \isacommand{assume}\isamarkupfalse%
\ {\isachardoublequoteopen}c\ {\isasymin}\ Pow{\isacharparenleft}{\kern0pt}a{\isacharparenright}{\kern0pt}\ {\isasyminter}\ M{\isacharbrackleft}{\kern0pt}G{\isacharbrackright}{\kern0pt}{\isachardoublequoteclose}
\end{isabelle}
\textit{Fix $\varkappa \in M^{\mathbb{P}}$ such that
  $\varkappa_{G}=c$,}
\begin{isabelle}
\ \ \ \ \isacommand{then}\isamarkupfalse%
\isanewline
\ \ \ \ \isacommand{obtain}\isamarkupfalse%
\ {\isasymchi}\ \isakeyword{where}\ {\isachardoublequoteopen}c{\isasymin}M{\isacharbrackleft}{\kern0pt}G{\isacharbrackright}{\kern0pt}{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymchi}\ {\isasymin}\ M{\isachardoublequoteclose}\ {\isachardoublequoteopen}val{\isacharparenleft}{\kern0pt}P{\isacharcomma}{\kern0pt}G{\isacharcomma}{\kern0pt}{\isasymchi}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ c{\isachardoublequoteclose}\isanewline
\ \ \ \ \ \ \isacommand{using}\isamarkupfalse%
\ GenExt{\isacharunderscore}{\kern0pt}iff\ \isacommand{by}\isamarkupfalse%
\ auto
\end{isabelle}
\textit{and let $\vartheta=\{\langle\sigma, p\rangle:
  \sigma \in \operatorname{dom}(\tau) \wedge p \Vdash \sigma \in
  \varkappa\}$;}
\begin{isabelle}
\ \ \ \ \isacommand{let}\isamarkupfalse%
\ {\isacharquery}{\kern0pt}{\isasymtheta}{\isacharequal}{\kern0pt}{\isachardoublequoteopen}{\isacharbraceleft}{\kern0pt}{\isasymlangle}{\isasymsigma}{\isacharcomma}{\kern0pt}p{\isasymrangle}\ {\isasymin}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}{\isasymtimes}P\ {\isachardot}{\kern0pt}\ p\ {\isasymtturnstile}\ {\isasymcdot}{\isadigit{0}}\ {\isasymin}\ {\isadigit{1}}{\isasymcdot}\ {\isacharbrackleft}{\kern0pt}{\isasymsigma}{\isacharcomma}{\kern0pt}{\isasymchi}{\isacharbrackright}{\kern0pt}\ {\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}
\end{isabelle}
\textit{$\vartheta \in M$ by the Definability Lemma.}
\begin{isabelle}
\ \ \ \ \isacommand{have}\isamarkupfalse%
\ {\isachardoublequoteopen}arity{\isacharparenleft}{\kern0pt}forces{\isacharparenleft}{\kern0pt}Member{\isacharparenleft}{\kern0pt}{\isadigit{0}}{\isacharcomma}{\kern0pt}{\isadigit{1}}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ {\isadigit{6}}{\isachardoublequoteclose}\isanewline
\ \ \ \ \ \ \isacommand{using}\isamarkupfalse%
\ arity{\isacharunderscore}{\kern0pt}forces{\isacharunderscore}{\kern0pt}at\ \isacommand{by}\isamarkupfalse%
\ auto\isanewline
\ \ \ \ \isacommand{with}\isamarkupfalse%
\ {\isacartoucheopen}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}\ {\isasymin}\ M{\isacartoucheclose}\ {\isacartoucheopen}{\isasymchi}\ {\isasymin}\ M{\isacartoucheclose}\isanewline
\ \ \ \ \isacommand{have}\isamarkupfalse%
\ {\isachardoublequoteopen}{\isacharquery}{\kern0pt}{\isasymtheta}\ {\isasymin}\ M{\isachardoublequoteclose}\isanewline
\ \ \ \ \ \ \isacommand{using}\isamarkupfalse%
\ P{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}M\ one{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}M\ leq{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}M\ sats{\isacharunderscore}{\kern0pt}fst{\isacharunderscore}{\kern0pt}snd{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}M\isanewline
\ \ \ \ \ \ \isacommand{by}\isamarkupfalse%
\ simp
\end{isabelle}
\textit{Since
  $\vartheta \in Q$,}
\begin{isabelle}
\ \ \ \ \isacommand{then}\isamarkupfalse%
\isanewline
\ \ \ \ \isacommand{have}\isamarkupfalse%
\ {\isachardoublequoteopen}{\isacharquery}{\kern0pt}{\isasymtheta}\ {\isasymin}\ {\isacharquery}{\kern0pt}Q{\isachardoublequoteclose}\ \isacommand{by}\isamarkupfalse%
\ auto\isanewline
\ \ \ \ \isacommand{then}\isamarkupfalse%
\isanewline
\ \ \ \ \isacommand{have}\isamarkupfalse%
\ {\isachardoublequoteopen}val{\isacharparenleft}{\kern0pt}P{\isacharcomma}{\kern0pt}G{\isacharcomma}{\kern0pt}{\isacharquery}{\kern0pt}{\isasymtheta}{\isacharparenright}{\kern0pt}\ {\isasymin}\ {\isacharquery}{\kern0pt}b{\isachardoublequoteclose}\isanewline
\ \ \ \ \ \ \isacommand{using}\isamarkupfalse%
\ one{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}G\ one{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}P\ generic\ val{\isacharunderscore}{\kern0pt}of{\isacharunderscore}{\kern0pt}elem\ {\isacharbrackleft}{\kern0pt}of\ {\isacharquery}{\kern0pt}{\isasymtheta}\ {\isasymone}\ {\isacharquery}{\kern0pt}{\isasympi}\ G{\isacharbrackright}{\kern0pt}\isanewline
\ \ \ \ \ \ \isacommand{by}\isamarkupfalse%
\ auto
\end{isabelle}
\textit{we are done if we can show that
  $\vartheta_{G}=c$.}
\begin{isabelle}
\ \ \ \ \isacommand{have}\isamarkupfalse%
\ {\isachardoublequoteopen}val{\isacharparenleft}{\kern0pt}P{\isacharcomma}{\kern0pt}G{\isacharcomma}{\kern0pt}{\isacharquery}{\kern0pt}{\isasymtheta}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ c{\isachardoublequoteclose}
\end{isabelle}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "independence_ch_isabelle"
%%% ispell-local-dictionary: "american"
%%% End:

### 2022-05-11 Independence_CH/forcing.tex
Sample formal proof

@@
 %% \end{isabelle}
-Some of those instances, which were assumed in the locales of
-\session{ZF-Constructible} and we included for simplicity of design,
+Some of those instances, which we included for simplicity of design,
 are actually “fake” instances in that they can be obtained from
 Powerset. An example of this is the sole instance of the
-\isatt{M{\uscore}basic} locale; we managed to eliminate it but in
+\isatt{M{\uscore}basic} locale, originally from
+\session{ZF-Constructible}; we managed to eliminate it but in
 doing so we had to reprove many lemmas from
-\session{ZF-Constructible} in the weakened context. Our locales
+that session in the weakened context. Our locales
 \isatt{M{\uscore}ZF1} through \isatt{M{\uscore}ZF4} list a total of 86

### 2022-05-25 Independence_CH/details.tex
Articulating some conclusions, first try

@@
 It is to be noted that application of the Forcing Theorems do not
-require any extra Replacement instances on $M$ (independently of the
+require any extra replacement instances on $M$ (independently of the
 formula $\phi$ for which they are invoked). This is not the case for
@@
 \textit{This is the set of all names $\vartheta \in
-  M^{\mathcal{P}}$\dots}---it is pretty laborious to show that things
+  M^{\mathcal{P}}$} [\dots]---it is pretty laborious to show that things
 are in $M$; it takes 17 more lines of code (not shown below) that

### 2022-05-25 Independence_CH/forcing.tex
Articulating some conclusions, first try

@@
 the many replacement instances, predicates $Q$ and $R$ are actually
-defined as the of formulas of the appropriate arity; we can only show
+defined as the satisfaction of formulas of the appropriate arity; we can only show
 that this form of the axioms hold in generic extensions.
@@
 are actually “fake” instances in that they can be obtained from
-Powerset. An example of this is the sole instance of the
+Powerset, or from other replacement instances already available in the
+respective context. An example of this is the sole instance of the
 \isatt{M{\uscore}basic} locale, originally from
@@
 \end{isabelle}
-where the $\forces$ notation had already been set up in the theory
-\theory{Forces{\uscore}Definition} as follows:
+where the $\forces$ notation had already been set up in the
+\theory{Forces{\uscore}Definition} theory as follows:
 \begin{isabelle}
@@
 $\isatt{frecR}$ on tuples $\lb \mathit{ft},t_1,t_2,p\rb$ (where
-$\mathit{ft}\in\{0,1\}$ indicates the type of the atomic formula being
-forced). Forcing for general formulas is defined by recursion on the
+$\mathit{ft}\in\{0,1\}$ indicates whether the atomic formula being
+forced is an equality or a membership, respectively). Forcing for general formulas is defined by recursion on the
 datatype $\formula$. Technical details on the implementation and proofs of the

### 2022-05-25 Independence_CH/lessons.tex
Articulating some conclusions, first try

\section{Some lessons}\label{sec:lessons}

We want to finish this report by gathering some of the conclusions we
reached after the experience of formalizing the basics of forcing in a
proof assistant.

\subsection{Aims of a formalization and planning}

We believe that in every project of formalization of mathematics,
there is a tension between the haste to verify the target results and
the need to obtain a readable, albeit extremely detailed, corpus of
statements and proofs. This tension is mirrored in two differents
purposes of formalization: Developing new mathematics from scratch and
producing verified results en route to this, versus verifying and
documenting material that has already been produced on paper.

Our present project clearly belongs to the second category, so we
prioritized trying to obtain formal proofs that mimicked standard
prose (the highlight being the sample proof in
Section~\ref{sec:sample-formal-proof}). We feel that the Isar language
provided with Isabelle has the right balance between elegance and
efficacy. Another crucial aspect to achieve this goal is the level of
detail of the blueprint for the formalization. We must however confess
that we learned many of the subtleties of Isabelle in the making, and
many engineering decisions were also taken before it was clear the
precise way things would develop in the future.

A similar experience, but on an opposite side of the formalization
spectrum happened to the Liquid Tensor Project as described by Scholze
in \cite{LTE2021}. People involved in the formalization simply pushed
their way to reach the summit, formalizing lemma after lemma. They
actually wrote the blueprint for that formalization \emph{afterwards}
it was complete! From time to time, we were also frenziedly trying to
get the results formalized, going beyond what we had planned.

As a result from this, some design choices that seemed reasonable at
first were proved to be inconvenient. For instance, we should had
better used predicates (of type $\tyi\fun\tyi\fun\tyo$) for the
forcing posets' order relations; this is the way they
are presented in the \session{Delta\_System\_Lemma} session. A similar
problem is that we require the forcing poset to be an element of $M$,
so the present infrastructure does not allow class forcing out of the
box. (The latter change seems to be rather straightforward, but the
former does not.)

Nearly the final stage of the project, we decided to go for the minimal
set of definitions and versions of lemmas that were needed to obtain
our target results. For example, we only proved the Delta System Lemma
for $\aleph_1$-sized families (thus limiting us to the plain ccc) and
showed preservation of sequences by considering countably closed
forcings (in fact, we formalized the bare minimum requirement of being
$(<\omega+1)$-closed). In doing this we went against the tried and
true advice that one should formalize the most general version of the
results available.

\subsection{Bureaucracy and scale factors}

\begin{enumerate}
\item Bureaucracy vs ML programming.
\item The “math” was already formalized on 22 November 2020.
  We finished the last goal on 22 August 2021.
  (Update: 20 November 2021 \& 28 November 2021, for CH)
\item Missing: automation of closure of models under operations.
\item Missing: basic arithmetic for dealing with arities.
\end{enumerate}

\begin{enumerate}
\item It is extremely misleading when automatic tools (\isatt{simp}, \isatt{auto}, etc)
  stop working just because of the sheer size of the goal. Oftentimes,
  in math, we disregard scale issues but they must always be taken
  into account in CS.
\item Example: $\forceisa(0\in 1)$ is expandable,
  $\forceisa(\neg\neg  0\in 1)$ is not.
\item Example: Synthesis of $\forceisa$; could have been fully synthesized,
  but that was dirty “strategy”.
\item The know-how of computer scientists on this kind of engineering is
  very important
\end{enumerate}

\subsection{You might have formalized it, and still be wrong}
\begin{enumerate}
\item Example: restriction of relations.
\item Pollack, “Pollack consistency” by
  Wiedijk. Cf. \theory{Definitions\_Main} (thanks to discussions with
  Vidnyánszky). Opacity of automated proofs. 
\item Plot twist: You can be right without knowing. Intuition may drive proofs
  even if we are not working on what we believe we are.
\end{enumerate}

\subsection{Beware of the “Code fever”}\label{sec:beware-code-fever}
\begin{enumerate}
\item “We know that doing math is fun---formalization is like DRUGS”
\item Feeling of accomplishment after seeing your writings
  validated beyond reasonable doubt (v.g. cofinality).

\item One easily forgets about the “Power of the Board.”
\end{enumerate}

\subsection{The Devil's on the shortcuts}
\begin{enumerate}
\item
  Our proofs of the “definition of forces” (and many
  consequences) and of the lemma for “forcing a value” of function
  depend on the countability of the ground model. 
\item
  Density arguments (look for “TODO”, “general versions”).
\end{enumerate}

\end{enumerate}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "independence_ch_isabelle"
%%% ispell-local-dictionary: "american"
%%% End:

### 2022-05-27 Independence_CH/conclusion.tex
More lessons into slightly-sensical paragraphs

\section{Conclusion}
\label{sec:conclusion}

One of our main goals was to provide a \emph{proof document}, which
can hopefully be read by mathematicians/set-theorists.

Where did we fail on this?
\begin{enumerate}
\item
  The definition of forces is too codish.
\item
  Proving instances of axiom schemes.
\end{enumerate}

The experience of formalizing math:
\begin{enumerate}
\item It is fun, almost addictive.
\item Feeling of accomplishment after seeing your writings
  validated beyond reasonable doubt (v.g. cofinality).
\item On the grim side: It is easy to forget to put perspective on a
  formalization issue and meditate on paper.
\end{enumerate}

\section*{Acknowledgments}
\label{sec:acknowledgments}
We thanks Z.~Vidnyánszky for some discussions where the idea of
building the summary in \theory{Definitions\_Main} took shape.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "independence_ch_isabelle"
%%% ispell-local-dictionary: "american"
%%% End:

### 2022-05-27 Independence_CH/lessons.tex
Removed an extra \end{enumerate}
More lessons into slightly-sensical paragraphs

@@
 set of definitions and versions of lemmas that were needed to obtain
-our target results. For example, we only proved the Delta System Lemma
-for $\aleph_1$-sized families (thus limiting us to the plain ccc) and
-showed preservation of sequences by considering countably closed
-forcings (in fact, we formalized the bare minimum requirement of being
-$(<\omega+1)$-closed). In doing this we went against the tried and
+our target results. For example,
+\begin{itemize}
+\item
+  we only proved the Delta System Lemma for $\aleph_1$-sized families
+  (thus limiting us to the plain ccc);
+\item we showed preservation of sequences by considering countably
+  closed forcings (in fact, we formalized the bare minimum requirement
+  of being $(<\omega+1)$-closed); and
+\item we proved that appropriate subsets were dense only for Cohen posets
+  relevant for the arguments at hand, but could have been done in general.
+\end{itemize}
+In doing this we went against the tried and
 true advice that one should formalize the most general version of the
-results available.
+results available. Another shortcut we took was to
+simplify some proofs by appealing to the countability of the ground
+model; this is the case of
+\isatt{definition{\uscore}of{\uscore}forces} and the result on forcing
+values of a function.
+
+\subsection{How to believe in the formalization}
+This is a rather tricky question, that was addressed by Pollack in his
+\cite{MR1686867}. There is little point to discuss that, after an
+assistant has accepted some input successfully, \emph{some
+mathematics} has been formally verified. What might not be apparent is
+if the claimed theorems are indeed the results that have been
+checked. One key aspect of this is the logical foundation of the
+assistant (Section~\ref{sec:isabelle-metalogic-meta}). But the weakest
+link in the chain is the laying down of definitions building up to the
+concepts needed to state the target results. (Another related aspect,
+concerning the way results are printed and parsed by assistant versus
+there internal meaning, was studied by Wiedijk \cite{zbMATH06319597}.)
+
+We took care of this matter by providing, as an entry point for our
+whole development, the theory \theory{Definitions\_Main} in which a
+path from some the fundamental concepts from Isabelle/ZF reaching to our main theorems
+is expounded. Cross-references to major milestones (which can be
+navigated by using Isabelle) are provided there. It
+corresponds roughly to less than seven journal pages.
+
+Frequently, we formalized material by directly typing the proof we
+new by heart, and in doing so we assumed that some definitions
+accommodated some of our preconceptions.
+It is significant that in a few such occasions, we were doubly
+surprised by the fact that some supposedly trivial lemma would not go
+through, because the definitions addressed something different (think
+of \emph{restriction} of a function to a set versus that of a
+relation), and also that we were able to proved the previous
+results. The takeaway is that intuition may drive proofs
+even if you are not working on what we believe we are.
+
+A final aspect on this topic concerns automated methods. As it was
+mentioned in the Introduction, some proofs can actually be
+\emph{obscured} by automation, simply because one has to take
+complicated opaque computations “on faith.”
 
@@
 
-\begin{enumerate}
-\item Bureaucracy vs ML programming.
-\item The “math” was already formalized on 22 November 2020.
-  We finished the last goal on 22 August 2021.
-  (Update: 20 November 2021 \& 28 November 2021, for CH)
-\item Missing: automation of closure of models under operations.
-\item Missing: basic arithmetic for dealing with arities.
-\end{enumerate}
+It is noteworthy that although the “math” of the construction of a
+model of $\ZFC+\neg\CH$ was already in place by the end of November
+2020, it was only 9 months later that we were able to finish the
+formalization of that result. The missing pieces were essentially
+bureacracy. Some of the material filed under this category comprises:
+\begin{itemize}
+\item calculation of arities of internalized formulas;
+\item proving that certain constructions belong to the relevant
+  models;
+\item (required for the above) showing that particular instances of
+  separation and replacement hold in the ground model;
+\item etc.
+\end{itemize}
 
-\begin{enumerate}
-\item It is extremely misleading when automatic tools (\isatt{simp}, \isatt{auto}, etc)
-  stop working just because of the sheer size of the goal. Oftentimes,
-  in math, we disregard scale issues but they must always be taken
-  into account in CS.
-\item Example: $\forceisa(0\in 1)$ is expandable,
-  $\forceisa(\neg\neg  0\in 1)$ is not.
-\item Example: Synthesis of $\forceisa$; could have been fully synthesized,
-  but that was dirty “strategy”.
-\item The know-how of computer scientists on this kind of engineering is
-  very important
-\end{enumerate}
+Some of those proofs were almost copypasted once and again with minor
+variants; this would usually be relegated to some function in the
+meta-language, but we were unable to do this due to our limitations in
+programming Isabelle/ML\footnote{%
+  On the other hand, our inability to automate proofs of replacement
+  instances paved the way for identifying which were the ones needed for
+  forcing! ($\leftarrow$\textbf{put somewhere, don't know if this is the right
+    place}).}
+and  (an ML style having several quirks and which
+is not very well document).
 
-\subsection{You might have formalized it, and still be wrong}
-\begin{enumerate}
-\item Example: restriction of relations.
-\item Pollack, “Pollack consistency” by
-  Wiedijk. Cf. \theory{Definitions\_Main} (thanks to discussions with
-  Vidnyánszky). Opacity of automated proofs. 
-\item Plot twist: You can be right without knowing. Intuition may drive proofs
-  even if we are not working on what we believe we are.
-\end{enumerate}
+Nevertheless, the experience of the CS members of the team (and
+community) is invaluable for the kind of engineering that is needed in
+a large project like the present one. For example, it is
+(mathematically) misleading when automatic tools (\isatt{simp},
+\isatt{auto}, etc) stop working just because of the sheer size of the
+goal (v.g., the same statement with 7 variables succeed but with 8
+variables doesn't). Scale issues are very easily diregarded in the
+abstract; for example, the formula $\forceisa({\cdot}0\in 1{\cdot})$
+can be explicitly printed by Isabelle (it spans nearly 20k symbols), but
+$\forceisa({\cdot}\neg{\cdot}\neg{\cdot}0\in 1{\cdot}{\cdot}{\cdot})$
+can not.
 
-\subsection{Beware of the “Code fever”}\label{sec:beware-code-fever}
-\begin{enumerate}
-\item “We know that doing math is fun---formalization is like DRUGS”
-\item Feeling of accomplishment after seeing your writings
-  validated beyond reasonable doubt (v.g. cofinality).
-
-\item One easily forgets about the “Power of the Board.”
-\end{enumerate}
-
-\subsection{The Devil's on the shortcuts}
-\begin{enumerate}
-\item
-  Our proofs of the “definition of forces” (and many
-  consequences) and of the lemma for “forcing a value” of function
-  depend on the countability of the ground model. 
-\item
-  Density arguments (look for “TODO”, “general versions”).
-\end{enumerate}
-
-\end{enumerate}
+Another example concerns the very definition of $\forceisa$. As a
+proof of concept, one of us tried to obtain its definition by
+exclusively using formula synthesis, which was supposed to be trivial,
+since it is mathematically. But in fact, some early minor mistake
+rendered the whole effort useless. We then turned to a more
+informed programming discipline, which involved decomposing the
+definition in stages, each of which was checked for correctness, and
+in that way we were able to reach our objective.
 

### 2022-05-28 Independence_CH/proof_assists_isabelle.tex
An alternative start for Sec. 2

@@
 \label{sec:proof-assist-isabelle}
+
+Let us briefly introduce Isabelle in the large landscape of proof
+assistans (or ITPs for interactive theorem provers); we refer to the
+excelent chapter by \citet{DBLP:series/hhl/HarrisonUW14} for a more
+thorough reconstruction of the history of ITPs.
+
+It is expected that an ITP assists the human user while mechanizing
+some piece of mathematics; the interaction varies from system to
+system, but a common interface consists on the system showing the
+current goal and context. The user instructs the ITP to modify the
+goal/context by means of tactics; a forward rule changes the context
+with the intention of having in the new context a hypothesis closer to
+the goal. A backward rule represents a deduction whose conclusion is
+the goal, thus the new goal corresponds to the premises of the rule. A
+proof is completed when the (current) goal is an instance of the
+hypothesis (in the current context).
+
+In that dialog, the user produces a script of tactics that can be
+later reproduced step-by-step by the system (to check, for example,
+that an imported theory is correct) or by the user to understand
+the proof.\footnote{As we will see there are more declaratives
+proof languages that aims to have intelligible proof scripts.}
+
+To have any value at all, the system should be able to say if a tactic
+makes sense (for instance, it must be forbidden to use the backward
+tactic \textit{conjI} when the main connective is a
+disjunction). Moreover one should be able to ascertain the validity of
+the reaction of an ITP (be its rejection or its acceptance of some
+alleged logical step). % Here I would merge the next paragraph.
 

### 2022-05-28 Independence_CH/lessons.tex
Moving bib, typos, example of synthesis triviality

@@
 exclusively using formula synthesis, which was supposed to be trivial,
-since it is mathematically. But in fact, some early minor mistake
+since it is mathematically (as in the case of
+Equation~(\ref{eq:sats_big_union_fm})).
+But in fact, some early minor mistake
 rendered the whole effort useless. We then turned to a more

### 2022-05-28 Independence_CH/proof_assists_isabelle.tex
Moving bib, typos, example of synthesis triviality

@@
 excelent chapter by \citet{DBLP:series/hhl/HarrisonUW14} for a more
-thorough reconstruction of the history of ITPs.
+thorough recollection of the history of ITPs.
 
@@
 the proof.\footnote{As we will see there are more declaratives
-proof languages that aims to have intelligible proof scripts.}
+proof languages that aim to have intelligible proof scripts.}
 
@@
 using a typed programming language; rigor is enforced by defining a
-type of ``theorems'', whose members can only be constructed using
+type of ``theorems,'' whose members can only be constructed using
 operations stipulated in a small \emph{kernel} which encodes the

### 2022-05-28 Independence_CH/lessons.tex
Even more typos

@@
 the need to obtain a readable, albeit extremely detailed, corpus of
-statements and proofs. This tension is mirrored in two differents
+statements and proofs. This tension is mirrored in two different
 purposes of formalization: Developing new mathematics from scratch and
@@
 formalization of that result. The missing pieces were essentially
-bureacracy. Some of the material filed under this category comprises:
+bureaucracy. Some of the material filed under this category comprises:
 \begin{itemize}
@@
 
-Some of those proofs were almost copypasted once and again with minor
+Some of those proofs were almost copy-pasted once and again with minor
 variants; this would usually be relegated to some function in the
@@
 goal (v.g., the same statement with 7 variables succeed but with 8
-variables doesn't). Scale issues are very easily diregarded in the
+variables doesn't). Scale issues are very easily disregarded in the
 abstract; for example, the formula $\forceisa({\cdot}0\in 1{\cdot})$

### 2022-05-28 Independence_CH/conclusion.tex
Ack

@@
 \label{sec:acknowledgments}
-We thanks Z.~Vidnyánszky for some discussions where the idea of
+We thank Z.~Vidnyánszky for some discussions where the idea of
 building the summary in \theory{Definitions\_Main} took shape.
+Mikhail Mandrykin helped us with technical assistance regarding the
+rendering of formulas in the Isabelle/jEdit interface, which improved
+our presentation of this material during the Latin American Congress
+of Mathematicians (CLAM 2021); we make this extensive to the rest of
+the Isabelle community, for its kind support through the Users mailing
+list. We also want to warmly thank Larry Paulson for his work and
+encouragement.  Last, and definitely not least, we are deeply grateful
+to Ken Kunen for his
+inspiring expositions. This formalization is a tiny homage to his
+memory.
 

### 2022-05-29 Independence_CH/intro.tex
bibs, minor edits

@@
 the Archive of Formal Proofs based on Isabelle, and formalizations of
-brand new and complex objects like the definition of perfectoid spaces
-and the Liquid Tensor Project \cite{LTE2020,LTE2021} on Lean.
+brand new and complex objects like the definition of perfectoid spaces \cite{10.1145/3372885.3373830}
+and the Liquid Tensor Experiment \cite{LTE2020,LTE2021} on Lean.
 
 \subsection{Our achievements}
-We formalized a model-theoretic rendition of forcing.
-\begin{itemize}
-\item Proper extensions of ctms of $\ZF$.
-\item The construction of ctms of $\ZFC + \neg\CH$ and of $\ZFC + \CH$ given one of
-  $\ZFC$.
-\end{itemize}
+We formalized a model-theoretic rendition of forcing (Section~\ref{sec:forcing}), showing how to
+construct proper extension of ctms of $\ZF$ (respectively, with
+$\AC$), and we formalized the basic forcing notions required to obtain
+ctms of $\ZFC + \neg\CH$ and of $\ZFC + \CH$. No meta-theoretic issues
+(consistency, FOL calculi, etc) were formalized, so we were merely
+concerned with the mathematics of forcing. Nevertheless, by inspecting
+the foundations underlying our proof assistant Isabelle
+(Section~\ref{sec:isabelle-metalogic-meta}) it can be stated that our
+formalization is closest to a proof in $\ZF$ of the previous
+implications.
 
-In order to do so, we extended the treatment of relativization of
-set-theoretical concepts. 
-\begin{itemize}
-\item We redesigned Isabelle/ZF results on non-absolute concepts to work
-  relative to a class.
-\item We provided missing results (from basic ones involving cardinal
-  successors, countable sets, \dots)
-\end{itemize}
-No meta-theoretic issues (consistency, FOL calculi, etc) were formalized.
+In order to reach our goals, we provided missing results, from basic ones
+involving cardinal successors, countable sets, etc
+(Section~\ref{sec:extension-isabellezf}); we extended the treatment of relativization of
+set-theoretical concepts (Section~\ref{sec:tools-relativization}).
+%% We redesigned Isabelle/ZF results on non-absolute concepts to work
+%% relative to a class.
 
 One added value that is obtained from the present formalization is
-that we identified which instances of  Replacement are
-needed to set the forcing machinery up (Separation is a bit more
-complicated).
+that we identified a handfull of instances of Replacement which are
+sufficient to set the forcing machinery up (Section~\ref{sec:33-repl-inst-rule}).
+%% (Separation is a bit more complicated).
 %% A verification that no choice  is needed at the level of the

### 2022-05-29 Independence_CH/proof_assists_isabelle.tex
bibs, minor edits

@@
 The type of propositions $\prop$ is equipped with a binary operation
-$\implies$ (“meta-implication”) and a universal “meta-quantifier”
+${\implies}$ (“meta-implication”) and a universal “meta-quantifier”
 $\ALL$, that are used to represent the object
@@
 For recursive definitions, only the second option is available, and
-that is the way the Definition of Forcing is implemented in our
-formalization.
+that is the way the definition of the forcing relation is implemented
+in our formalization.
 

### 2022-05-29 Independence_CH/proof_assists_isabelle.tex
One paragraph about Edinburgh LCF.

@@
 excelent chapter by \citet{DBLP:series/hhl/HarrisonUW14} for a more
-thorough recollection of the history of ITPs.
+thorough reconstruction of the history of ITPs.
 
@@
 the goal, thus the new goal corresponds to the premises of the rule. A
-proof is completed when the (current) goal is an instance of the
+proof is completed when the (current) goal is an instance of a
 hypothesis (in the current context).
@@
 the reaction of an ITP (be its rejection or its acceptance of some
-alleged logical step). % Here I would merge the next paragraph.
+alleged logical step). Edinburgh LCF~\cite{Gordon1979-qm} was an
+influential proof assistant where that verification could be achieved
+by checking a small \emph{kernel}; this is the only module that
+constructs proofs in response to user scripts. The rest of the system
+can interact with the kernel via an interface that exposes functions
+guaranteeing the validity of the proofs.
 

### 2022-05-29 Independence_CH/related.tex
Comparison with Flypitch. Moving bibs

--- 2022-03-26 JSL/related.tex
@@
 
-\paragraph*{Question} How do they overcome Tarski's undefinability of truth?
+There is another formalization of forcing in Lean by Han and van
+Doorn, under the name \emph{Flypitch} \cite{han_et_al:LIPIcs:2019:11074,DBLP:conf/cpp/HanD20}. When our project started, we
+were unaware of this initiative, and the same as them, we were deeply
+influenced by Wiedijk's list of 100 theorems \cite{Formalizing100}.
 
-In some sense, their presentation is mathematically more "mature",
-but the type-theoretic machinery might be a bit alien to most
-set-theorists.
+Many aspects make their formalization different from ours. Their
+presentation of the mathematics is somewhat more elegant and cohesive,
+since they go for the Boolean valued approach and this is closer to
+obtain proofs of relative consistency; they actually set up the
+calculus of first-order logic, and \emph{en route} to forcing they
+formalized the basic model theory of Boolean valued models and Gödel's
+Completeness Theorem. They also provided the treatment of the regular
+open algebra, and the general version of the Delta System
+Lemma. Putting this together they readily obtain a proof that
+(*) $\Con(\ZFC)$ implies 
+$\ZFC\nvdash \CH$ \cite{han_et_al:LIPIcs:2019:11074} and after
+formalizing collapse forcing they show  $\ZFC\nvdash \neg\CH$
+\cite{DBLP:conf/cpp/HanD20}.
 
-Quoting Flypitch-CPP,
-\begin{quote}
-  The types of cardinals and ordinals in mathlib, which
-  are defined as equivalence classes of (well-ordered) types, live one
-  universe level higher than the types used to construct them
-\end{quote}
+It should be emphasized, however, that the Flypitch project was
+carried out assuming a rather strong meta-theory. 
+Carneiro \cite{carneiro-ms-thesis} reports that Werner's results in
+\cite{10.5555/645869.668660} can be adapted to show that the base
+logic  of Lean (restricted to $n$
+type universes) proves the consistency of $\ZFC$ plus $n$
+inaccessibles. Han and van Doorn did used universes in their
+implementation; for instance, ordinals are “defined as equivalence
+classes of (well-ordered) types, live one
+universe level higher than the types used to construct them”
+\cite{han_et_al:LIPIcs:2019:11074}. It is not clear to us if they are
+able to avoid strength (even for the proof of the finitary implication
+(*)).
 
-\paragraph*{Questions} How do they handle recursions? Are these all
-\emph{external}? Does this has any impact on their (non?)ability of
-identifying replacement instances used by their approach?
+On a lesser note, in order to prove $\AC$ in the generic extension,
+Flypitch requires choice in the metatheory
+\cite[p.~11]{han_et_al:LIPIcs:2019:11074}. Our formalization works
+entirely in $\ZF$, and only import the Isabelle theory postulating
+$\AC$ for demostration purposes.
 
-They seem to require $\AC$ in their meta-theory.
+We believe that our alternative formalization using the ctm approach
+over Isabelle/ZF might be more appealing to set-theorists, because the
+type-theoretic machinery used in Flypitch might be a bit alien, and
+since absoluteness grant us extra naturality. This last point may also
+be illustrated by the treatment of ordinals; in our formalization, as
+it is expected intuitively, the following are equivalent for every $x$
+in a ctm $M$:
+\begin{itemize}
+\item $\isatt{Ord}(x)$,
+\item $\isatt{ordinal}(\isatt{\#\#}M,x)$ (the relational
+  definition relativized to $M$),
+\item
+  $M,[x] \models {\cdot}0\isatt{ is ordinal}{\cdot}$,
+\item
+  $M[G], [x] \models {\cdot}0\isatt{ is ordinal}{\cdot}$ for every generic $G$,
+\end{itemize}
+where ${\cdot}n\isatt{ is ordinal}{\cdot}$ is the code for the
+appropriate first-order formula  ($0$ is a de
+Bruijn index above!). In contrast, Han and van Doorn require an
+injection from the ordinals of the corresponding type universe into
+their encoding of a model of $\ZF$, and a further necessary injection into the Boolean
+valued model using checks---this last step obviously appears in our
+presentation, but the $\val$ function used to construct will turn
+check-names into the corresponding set in the generic extension, as expected.
+
+This faithfulness to set-theoretical practices does not come for
+free. Recursive constructions and inductive definitions are far easier
+to perform in the Calculus of Inductive Constructions on which Lean
+(and Coq) are based, and in Isabelle/ZF are (still) rather
+cumbersome. Also, a typed discipline provides aid to write succintly
+and many assumptions are satisfied by mere notation.
+
+%% Comment on ZFC_in_HOL
 

### 2022-05-29 Independence_CH/proof_assists_isabelle.tex
Reverting changes accidentally lost by me.

@@
 The type of propositions $\prop$ is equipped with a binary operation
-${\implies}$ (“meta-implication”) and a universal “meta-quantifier”
+$\implies$ (“meta-implication”) and a universal “meta-quantifier”
 $\ALL$, that are used to represent the object
@@
 For recursive definitions, only the second option is available, and
-that is the way the definition of the forcing relation is implemented
-in our formalization.
+that is the way the Definition of Forcing is implemented in our
+formalization.
 

### 2022-05-29 Independence_CH/proof_assists_isabelle.tex
First version of intro for section 2.

--- 2022-05-29 Independence_CH/proof_assists_isabelle.tex
@@
 
-Proof assistants provide diverse aids for the task
-of formalizing a piece of mathematics. They are usually implemented
-using a typed programming language; rigor is enforced by defining a
-type of ``theorems,'' whose members can only be constructed using
-operations stipulated in a small \emph{kernel} which encodes the
-underlying foundation of the assistant. Further developments interact
-with the type of theorems only through the kernel, and thus the latter
-is the only ``trusted'' part of the assistant's code.
+The core of Isabelle, as well as LCF, is based on higher-order logic;
+another heritage of LCF is the possibility for the user to define
+\emph{tactics} to encapsulate common patterns used to solve goals. In
+fact, this is the origin of the ML family of languages: a
+Meta-Language for programming tactics. In contrast with LCF, Isabelle
+was designed~\cite{Paulson1990} as a generic theorem for several
+logics. Isabelle still offers a programmable interface (in ML) that
+allows the user, for instance, to define a tactic solving a statement
+about a natural number $n$ by doing cases on whether $n=0$ or
+$n\neq 0$.  Moreover one can define functions that manipulate terms
+(in the usual sense) or even formulas.
 
-Several of the more established assistants (HOL Light, Coq, Isabelle)
-are programmed in some variant of the ML language (which was
-designed for this purpose); the newer Lean, on the other
-hand, was originally conceived as a standalone functional programming
-language with all the features of a standard assistant.
+In contrast with LCF, some of the more prominent ITPs of today are
+based on some type theory. Both Coq~\cite{coq} and
+lean~\cite{DBLP:conf/cade/Moura021} are based on the Calculus of
+Inductive Constructions
+\cite{DBLP:conf/mfps/PfenningP89,DBLP:journals/iandc/CoquandH88},
+while Agda~\cite{agda} is an extension of Martin-Löf type
+theory~\cite{DBLP:books/daglib/0000395}. The oldest ITP still
+still used today is Mizar~\cite{mizar} is far away both in terms
+of foundations and architecture; it inspired, though, the Isar
+dialect used in nowadays Isabelle.
 
-In the case of Isabelle, Standard ML is the first of the four layers
-in which we work in this assistant. Both the kernel and automation
-of proofs is coded in ML, sometimes as a substitute for induction on
-formulas, as the next section explains.
+% Proof assistants provide diverse aids for the task
+% of formalizing a piece of mathematics. They are usually implemented
+% using a typed programming language; rigor is enforced by defining a
+% type of ``theorems,'' whose members can only be constructed using
+% operations stipulated in a small \emph{kernel} which encodes the
+% underlying foundation of the assistant. Further developments interact
+% with the type of theorems only through the kernel, and thus the latter
+% is the only ``trusted'' part of the assistant's code.
+
+% Several of the more established assistants (HOL Light, Coq, Isabelle)
+% are programmed in some variant of the ML language (which was
+% designed for this purpose); the newer Lean, on the other
+% hand, was originally conceived as a standalone functional programming
+% language with all the features of a standard assistant.
+
+% In the case of Isabelle, Standard ML is the first of the four layers
+% in which we work in this assistant. Both the kernel and automation
+% of proofs is coded in ML, sometimes as a substitute for induction on
+% formulas, as the next section explains.
 

### 2022-05-29 Independence_CH/proof_assists_isabelle.tex
minor changes.

@@
 while Agda~\cite{agda} is an extension of Martin-Löf type
-theory~\cite{DBLP:books/daglib/0000395}. The oldest ITP still
-still used today is Mizar~\cite{mizar} is far away both in terms
-of foundations and architecture; it inspired, though, the Isar
-dialect used in nowadays Isabelle.
+theory~\cite{DBLP:books/daglib/0000395}. Mizar~\cite{mizar} is the
+oldest ITP still used today but is far away both in terms of
+foundations and architecture from Isabelle; Mizar inspired, though, the
+Isar dialect used in nowadays Isabelle.
 

### 2022-05-30 Independence_CH/non-absolute.tex
Minor editing

@@
 (denoted  $|\cdot|^M$, as expected),
-its relational version $\isatt{is{\uscore}cardinal}$, the
+the relational version of the latter $\isatt{is{\uscore}cardinal}$, the
 internalized formula \isatt{is{\uscore}cardinal{\uscore}fm} whose

### 2022-05-30 Independence_CH/proof_assists_isabelle.tex
Minor editing

@@
 is written as
-\[
+\begin{equation}\label{eq:modus-ponens}
   \ALL P\,Q.\ \ [P\limp Q] \implies ([P] \implies [Q]).
-\]
-The square brackets (which are omitted in Isabelle theories) represent
+\end{equation}
+The square brackets (which are omitted in Isabelle theories, as well
+as the outermost universal quantifiers) represent
 an injection from $\tyo$ into $\prop$. % ($[P]$ can be read as “$P$
@@
 $\Meta_{\mathrm{IFOL}}$ proofs in \emph{expanded normal form}
-\cite{MR0387024}. Passing to classical logic does not present
+\cite{MR0387024}; atomicity as stated after
+Equation~(\ref{eq:modus-ponens}) plays a role in this argument. Passing to classical logic does not present
 difficulties, but the addition of meta-equality must be taken care of.
@@
 possible that higher-order quantification gets used in $P$. The
-statement of $\AC$ also has a higher-order free variable, but we only
-use $\AC$ for demonstration purposes.
+statement of $\AC$, also uses a free higher-order variable to denote
+an indexed family of nonempty sets. % but we only
+% use $\AC$ for demonstration purposes.
 
@@
 forefront. Assistants have refined mechanisms to cope with defined
-concepts (which also make their foundations a more complicated
-than plain first-order logic), and this is the only that  nontrivial
+concepts and the introduction of new notation (which also make their foundations more complicated
+than plain first-order logic), and this is the only way that nontrivial
 mathematics can be formalized. But when studying relative interpretations, one
@@
 taken in \session{ZF-Constructible} is to consider relativizations of
-the formulas (of type $\tyo$) that define each concept. For instance,
+the formulas that define each concept. For instance,
 in the case of unions, we find a relativization
+%https://isabelle.in.tum.de/dist/library/ZF/ZF-Constructible/Relative.html#Relative.big_union_def_raw|axiom
 $\isatt{big{\uscore}union}:: (\isatt{i} \fun \isatt{o}) \fun \isatt{i}
@@
 \[
- \isatt{big{\uscore}union}(M,A,Z) \equiv \forall x[M].\ x \in z
+ \isatt{big{\uscore}union}(M,A,z) \equiv \forall x[M].\ x \in z
  \longleftrightarrow (\exists y[M].\ y \in A \land x \in y)
@@
 $L$ requires working with set models and satisfaction, which is
-defined for codes of formulas.
+defined for codes of formulas in the language $\{\in\}$.
 

### 2022-05-30 Independence_CH/related.tex
More minor revisions

@@
 type universes) proves the consistency of $\ZFC$ plus $n$
-inaccessibles. Han and van Doorn did used universes in their
+inaccessibles. Han and van Doorn did use universes in their
 implementation; for instance, ordinals are “defined as equivalence
-classes of (well-ordered) types, live one
+classes of (well-ordered) types, [\dots] one
 universe level higher than the types used to construct them”
@@
 We believe that our alternative formalization using the ctm approach
-over Isabelle/ZF might be more appealing to set-theorists, because the
-type-theoretic machinery used in Flypitch might be a bit alien, and
+over Isabelle/ZF might be more appealing to set-theorists because of the
+type-theoretic machinery used in Flypitch, and % might be a bit alien
 since absoluteness grant us extra naturality. This last point may also
@@
 valued model using checks---this last step obviously appears in our
-presentation, but the $\val$ function used to construct will turn
-check-names into the corresponding set in the generic extension, as expected.
+presentation, but the $\val$ function used to construct $M[G]$ will turn
+check-names into the corresponding argument, as expected.
 
-This faithfulness to set-theoretical practices does not come for
+This faithfulness to set-theoretical practice does not come for
 free. Recursive constructions and inductive definitions are far easier

### 2022-05-30 Independence_CH/lessons.tex
Misprints after second read

@@
 purposes of formalization: Developing new mathematics from scratch and
-producing verified results en route to this, versus verifying and
+producing verified results on the way, versus verifying and
 documenting material that has already been produced on paper.
@@
 A similar experience, but on an opposite side of the formalization
-spectrum happened to the Liquid Tensor Project as described by Scholze
+spectrum happened to the Liquid Tensor Experiment as described by Scholze
 in \cite{LTE2021}. People involved in the formalization simply pushed
@@
 concerning the way results are printed and parsed by assistant versus
-there internal meaning, was studied by Wiedijk \cite{zbMATH06319597}.)
+their internal meaning, was studied by Wiedijk \cite{zbMATH06319597}.)
 
@@
 Frequently, we formalized material by directly typing the proof we
-new by heart, and in doing so we assumed that some definitions
+knew by heart, and in doing so we assumed that some definitions
 accommodated some of our preconceptions.
@@
 of \emph{restriction} of a function to a set versus that of a
-relation), and also that we were able to proved the previous
+relation), and also that we were able to prove the adjacent
 results. The takeaway is that intuition may drive proofs
-even if you are not working on what we believe we are.
+even if you are not working on what you believe we are.
 
@@
     place}).}
-and  (an ML style having several quirks and which
+(an ML style having several quirks and which
 is not very well document).
@@
 variables doesn't). Scale issues are very easily disregarded in the
-abstract; for example, the formula $\forceisa({\cdot}0\in 1{\cdot})$
+abstract but, as an example, the formula $\forceisa({\cdot}0\in 1{\cdot})$
 can be explicitly printed by Isabelle (it spans nearly 20k symbols), but
@@
 
-Another example concerns the very definition of $\forceisa$. As a
+Another point where CS expertise was a prime asset was
+the very definition of $\forceisa$. As a
 proof of concept, one of us tried to obtain its definition by
-exclusively using formula synthesis, which was supposed to be trivial,
+using formula synthesis exclusively, which was supposed to be trivial,
 since it is mathematically (as in the case of

### 2022-06-03 Independence_CH/forcing.tex
bibs, minor edits
Comments for further edits

@@
 that we formalized.
+% Is the existential quantification here confusing?
 \begin{theorem}\label{th:forcing-thms}
@@
 \end{theorem}
+% Describe differences between Kunen's and our approach.
+% We can do everything using "internal forcing" (vs. forcing with
+% formulas in the metatheory).
+
 The statements appear as three separated items in
-\theory{Forcing{\uscore}Theorems.thy} in the \session{Independence\_CH} session,
+\theory{Forcing{\uscore}Theorems.thy} in the
+\session{Independence\_CH} session \cite{Independence_CH-AFP},
 and benefit from the $\isatt{map}$ function that applies a function to

### 2022-06-07 Independence_CH/proof_assists_isabelle.tex
Some re-arrangements concerning internalized formulas.

@@
 $L$ requires working with set models and satisfaction, which is
-defined for codes of formulas in the language $\{\in\}$.
+defined for (codes of) formulas in the language $\{\in\}$.
 
@@
 
-\session{ZF-Constructible} defines the set $\formula$ of codes for
-first-order formulas. These, alongside with lists, are instances of
-Isabelle/ZF treatment of inductively defined (internal) datatypes; induction
-and recursion theorems for them are proved automatically (this is in
-constrast to general well-founded recursion, for which one has to work
-with the fundamental recursor $\isatt{wfrec}$).
+\session{ZF-Constructible} defines the set $\formula$ of first-order
+formulas.\footnote{These, alongside with lists, are instances of
+  Isabelle/ZF treatment of inductively defined (internal) datatypes;
+  induction and recursion theorems for them are proved automatically
+  (this is in constrast to general well-founded recursion, for which
+  one has to work with the fundamental recursor $\isatt{wfrec}$).}  It
+is a minimalistic language only suited for set theory, since it has no
+term constructors and the atomic formulas are $x \in y$ and $x = y$.
+Variables are represented by de Bruijn indices \cite{MR0321704}, so in
+those formulas $x,y \in \omega$; for $\phi\in\formula$ and
+$z\in\omega$, $z$ is free in $\phi$ if it occurs under less than $z+1$
+quantifiers. The \isatt{arity} function on $\phi$ is one more than the
+maximum free index occurring in $\phi$.
 
-To avoid problems related to the binding of variables, de Bruijn
-indices \cite{MR0321704} are used instead. The satisfaction predicate
-$\isatt{sats}::\tyi\fun\tyi\fun\tyi\fun\tyo$ takes as arguments a set $M$, a list
-$\mathit{env}\in\isatt{list}(M)$ for the assignment of free indices
-(which are counted by the \isatt{arity} function),
-and $\phi\in\formula$, and it is written
-$M,\mathit{env}\models\phi$ in our formalization. This completes the
+The satisfaction predicate
+$\isatt{sats}::\tyi\fun\tyi\fun\tyi\fun\tyo$ takes as arguments a set
+$M$, a list $\mathit{env}\in\isatt{list}(M)$ for the assignment of
+free indices, and $\phi\in\formula$, and it is written
+$M,\mathit{env}\models\phi$ in our formalization.\footnote{Notice that
+  $M,\mathit{env}\models\phi$ only is meaningful if
+  $\mathit{length(env)} \geqslant arity(\phi)$.} This completes the
 description of the fourth and last formal layer of the development.
@@
 Internalized formulas for most (but not all) of the relational
-concepts can be obtained by guiding the automatic tactics. Actually,
-great many of the concepts in \session{ZF-Constructible} where
-internalized by hand; this is the case for union,
+concepts can be obtained by guiding the automatic tactics. In
+\session{ZF-Constructible} most of the concepts were internalized by
+hand; this is the case for union,
 %% \begin{multline*}
@@
 \end{multline}
-Note that $x$ and $y$ above are de Bruijn indices,
+Remember that $x$ and $y$ above are de Bruijn indices,
 $\isatt{nth}(x,\mathit{env})$ is the $x$th element of $\mathit{env}$

### 2022-06-07 Independence_CH/related.tex
Comments for further edits, 2nd part
Some points for Isar.

@@
 to perform in the Calculus of Inductive Constructions on which Lean
-(and Coq) are based, and in Isabelle/ZF are (still) rather
-cumbersome. Also, a typed discipline provides aid to write succintly
-and many assumptions are satisfied by mere notation.
-
-%% Comment on ZFC_in_HOL
+(and Coq) are based, and in Isabelle/ZF are rather cumbersome. Also, a
+typed discipline provides aid to write succintly and many assumptions
+are satisfied by mere notation. To be clear, those benefits come from
+doing set theory in a non set-theoretical meta-language. On the other
+hand, Isar proofs, as the one shown in
+Sec.~\ref{sec:sample-formal-proof}, are easier to write and understand
+than the language of tactics of Lean.
+%% Comment on ZFC_in_HOL and other developments using it (mostly by
+%% Paulson: Ramsey theory, (ordinal) partition theorems)
 

### 2022-06-08 Independence_CH/proof_assists_isabelle.tex
Minor edits and questions in boldface

--- 2022-05-30 Independence_CH/proof_assists_isabelle.tex
@@
 called $\Meta$; its original version was described in \cite{Paulson1989},
-and the addition of “sorts” appears in \cite{Nipkow-LF-91}.
+and the addition of “sorts” was reported in \cite{Nipkow-LF-91}.
 
@@
 $L$ requires working with set models and satisfaction, which is
-defined for codes of formulas in the language $\{\in\}$.
+defined for (“codes” of) formulas in the language $\{\in\}$.
 
@@
 
-\session{ZF-Constructible} defines the set $\formula$ of codes for
-first-order formulas. These, alongside with lists, are instances of
-Isabelle/ZF treatment of inductively defined (internal) datatypes; induction
-and recursion theorems for them are proved automatically (this is in
-constrast to general well-founded recursion, for which one has to work
-with the fundamental recursor $\isatt{wfrec}$).
+\session{ZF-Constructible} defines the set $\formula$ of first-order
+formulas internalized as sets.\footnote{These, alongside with lists, are instances of
+  Isabelle/ZF treatment of inductively defined (internal) datatypes;
+  induction and recursion theorems for them are proved automatically
+  (this is in constrast to general well-founded recursion, for which
+  one has to work with the fundamental recursor $\isatt{wfrec}$).}  It
+is a minimalistic language only suited for set theory
+(\textbf{just say they're formulas in language $\{\in\}$?}), since its
+atomic formulas have the form $\cdot x \in y\cdot$ and $\cdot x =
+y\cdot$. (We use dots as a visual aid signaling internalized formulas.)
+Variables are represented by de Bruijn indices \cite{MR0321704}, so in
+those formulas $x,y \in \omega$; for $\phi\in\formula$ and
+$z\in\omega$, $z$ is free in $\phi$ if it occurs under at most $z$
+quantifiers. The \isatt{arity} function on $\phi$ is one more than the
+maximum free index occurring in $\phi$.
 
-To avoid problems related to the binding of variables, de Bruijn
-indices \cite{MR0321704} are used instead. The satisfaction predicate
-$\isatt{sats}::\tyi\fun\tyi\fun\tyi\fun\tyo$ takes as arguments a set $M$, a list
-$\mathit{env}\in\isatt{list}(M)$ for the assignment of free indices
-(which are counted by the \isatt{arity} function),
-and $\phi\in\formula$, and it is written
-$M,\mathit{env}\models\phi$ in our formalization. This completes the
+The satisfaction predicate
+$\isatt{sats}::\tyi\fun\tyi\fun\tyi\fun\tyo$ takes as arguments a set
+$M$, a list $\mathit{env}\in\isatt{list}(M)$ for the assignment of
+free indices, and $\phi\in\formula$, and it is written
+$M,\mathit{env}\models\phi$ in our formalization.\footnote{Notice that
+  $M,\mathit{env}\models\phi$ is only meaningful if
+  $\isatt{length}(\mathit{env}) \geqslant
+\isatt{arity}(\phi)$.\textbf{$\leftarrow$ necessary?}} This completes the
 description of the fourth and last formal layer of the development.
@@
 Internalized formulas for most (but not all) of the relational
-concepts can be obtained by guiding the automatic tactics. Actually,
-great many of the concepts in \session{ZF-Constructible} where
-internalized by hand; this is the case for union,
-%% \begin{multline*}
-%%   \isatt{big{\uscore}union{\uscore}fm}(A,z) \equiv \\
-%%   ((\cdot\forall\cdot\cdot0 \in succ(z)\cdot \longleftrightarrow
-%%   (\cdot\exists\cdot\cdot0 \in succ(succ(A))\cdot \land \cdot1 \in
-%%   0\cdot\cdot\cdot)\cdot\cdot)
-%% \end{multline*}
+concepts can be obtained by guiding the automatic tactics. In
+\session{ZF-Constructible} most of the concepts were internalized by
+hand; this is the case for union,
+\begin{multline*}
+  \isatt{big{\uscore}union{\uscore}fm}(A,z) \equiv \\
+  ((\cdot\forall\cdot\cdot0 \in \mathit{succ}(z)\cdot \longleftrightarrow
+  (\cdot\exists\cdot\cdot0 \in \mathit{succ}(\mathit{succ}(A))\cdot \land \cdot1 \in
+  0\cdot\cdot\cdot)\cdot\cdot)
+\end{multline*}
 for which we have the following satisfaction lemma:
@@
 \end{multline}
-Note that $x$ and $y$ above are de Bruijn indices,
-$\isatt{nth}(x,\mathit{env})$ is the $x$th element of $\mathit{env}$
+Above, $\isatt{nth}(x,\mathit{env})$ is the $x$th element of $\mathit{env}$
 and $\isatt{\#\#}A::\tyi\fun \tyo$ is the class corresponding to the

### 2022-06-08 Independence_CH/proof_assists_isabelle.tex
Spacing the dots... sort of

@@
   \isatt{big{\uscore}union{\uscore}fm}(A,z) \equiv \\
-  ((\cdot\forall\cdot\cdot0 \in \mathit{succ}(z)\cdot \longleftrightarrow
-  (\cdot\exists\cdot\cdot0 \in \mathit{succ}(\mathit{succ}(A))\cdot \land \cdot1 \in
-  0\cdot\cdot\cdot)\cdot\cdot)
+  (({\cdot}\forall{\cdot}\,{\cdot}0 \in \mathit{succ}(z){\cdot} \longleftrightarrow
+  ({\cdot}\exists{\cdot}\,{\cdot}0 \in \mathit{succ}(\mathit{succ}(A)){\cdot} \land {\cdot}1 \in
+  0{\cdot}\,{\cdot}{\cdot}){\cdot}{\cdot})
 \end{multline*}

### 2022-06-10 Independence_CH/details.tex
Eliminating one instance from the list, associating fm to instances

@@
 
-\subsection{33 replacement instances to rule them all}
-\label{sec:33-repl-inst-rule}
+\subsection{32 replacement instances to rule them all}
+\label{sec:repl-instances}
 
-We isolated 33 instances of Replacement that are sufficient to force
+We isolated 32 instances of Replacement that are sufficient to force
 $\CH$/$\neg\CH$. The first 14 instances are needed to set up
@@
   \begin{itemize}
-  \item \isatt{replacement{\uscore}is{\uscore}order{\uscore}body{\uscore}fm}.
   \item \isatt{replacement{\uscore}is{\uscore}order{\uscore}eq{\uscore}map{\uscore}fm}.
+  \item Replacement through $x\mapsto \otype(x)$ \isatt{replacement{\uscore}is{\uscore}order{\uscore}body{\uscore}fm}.
   \end{itemize}
@@
 That makes 28 instances up to now. For the setup of forcing, we
-require the following 5 instances (the first two are  needed for the $\Delta$-system Lemma):
+require the following 4 instances:
 
@@
 %\verb|replacement_ax2(6)| \verb|replacement_transrec_apply_image_body_fm| seems fake!
-\item Recursive construction of sets using a choice function
+\item Recursive construction of sets using a choice function (for the $\Delta$-system Lemma)
   \isatt{replacement{\uscore}is{\uscore}trans{\uscore}apply{\uscore}image{\uscore}fm}.
 % \item Replacement through  $x\mapsto \check{x}$ \isatt{check{\uscore}replacement{\uscore}fm}.
-\item $x\mapsto |x|^M$
-  \isatt{Lambda{\uscore}in{\uscore}M{\uscore}fm(is{\uscore}cardinal{\uscore}fm(0,1),0)}.
+%\item $x\mapsto |x|^M$
+%  \isatt{Lambda{\uscore}in{\uscore}M{\uscore}fm(is{\uscore}cardinal{\uscore}fm(0,1),0)}. % fake since the image is known
 \item 2 instances for definitions by well-founded recursion.

### 2022-06-10 Independence_CH/forcing.tex
Comments for further edits, 2nd part
Eliminating one instance from the list, associating fm to instances

@@
 replacement instances, from which there are at least 53 fake ones (see
-Section~\ref{sec:33-repl-inst-rule}).
+Section~\ref{sec:repl-instances}).
+
+%% Missing: Locales have to be interpreted (part of "big picture"?)
 

### 2022-06-10 Independence_CH/intro.tex
Eliminating one instance from the list, associating fm to instances

@@
 that we identified a handfull of instances of Replacement which are
-sufficient to set the forcing machinery up (Section~\ref{sec:33-repl-inst-rule}).
+sufficient to set the forcing machinery up (Section~\ref{sec:repl-instances}).
 %% (Separation is a bit more complicated).

### 2022-06-11 Independence_CH/details.tex
Minor edits after meeting with Miguel

--- 2022-05-25 Independence_CH/details.tex
@@
 \subsection{33 replacement instances to rule them all}
-\label{sec:33-repl-inst-rule}
+\label{sec:repl-instances}
 
@@
 % \item Replacement through  $x\mapsto \check{x}$ \isatt{check{\uscore}replacement{\uscore}fm}.
-\item $x\mapsto |x|^M$
-  \isatt{Lambda{\uscore}in{\uscore}M{\uscore}fm(is{\uscore}cardinal{\uscore}fm(0,1),0)}.
+%\item $x\mapsto |x|^M$
+%  \isatt{Lambda{\uscore}in{\uscore}M{\uscore}fm(is{\uscore}cardinal{\uscore}fm(0,1),0)}. % fake since the image is known
+\item Absoluteness of the previous construction \isatt{replacement{\uscore}transrec{\uscore}apply{\uscore}image{\uscore}body{\uscore}fm}.
 \item 2 instances for definitions by well-founded recursion.

### 2022-06-11 Independence_CH/lessons.tex
Comments for further edits
Eliminating one instance from the list, associating fm to instances
Minor edits after meeting with Miguel

@@
   of being $(<\omega+1)$-closed); and
-\item we proved that appropriate subsets were dense only for Cohen posets
-  relevant for the arguments at hand, but could have been done in general.
 \end{itemize}
@@
   forcing! ($\leftarrow$\textbf{put somewhere, don't know if this is the right
-    place}).}
-(an ML style having several quirks and which
-is not very well document).
+    place}).}.
 
@@
 in that way we were able to reach our objective.
+% Provide details, what was precisely being synthesized
+% Connect to the relativization tools (that came afterwards).
 

### 2022-06-15 Independence_CH/discipline.tex
First brief description of the discipline.

\appendix

\section{Discipline for relativization}

As we said in Sec.~\ref{sec:relat-vers-non-absol}, in order to force
CH and its negation we depended on having relativized versions of
cardinals, Alephs, etc. It was clear for us that our efforts would be
more efficient if we set up a discipline for relativizing sets (terms
of type $\tyi$) and predicates/relations (terms of type $\tyo$).

Paulson only had, for each set, the relational version. It seemed
clearer to us to have a functional version of the relativized concept.
Going back to our example in \ref{sec:tools-relativization}, for the
concept $\isatt{cardinal}::\tyi \fun \tyi$ we want its relative
version
$\isatt{cardinal{\uscore}rel}::(\tyi \fun \tyo) \fun \tyi \fun\tyi$
and the relational version of the latter
$\isatt{is{\uscore}cardinal}::(\tyi \fun \tyo) \fun \tyi \fun \tyi
\fun \tyo$.

Our first attempt of defining a discipline was inspired by
mathematical considerations: if we might prove that
$\isatt{is{\uscore}cardinal}$ is functional and also prove the
existence of a witness $\isatt{c}$ such that $\isatt{M(c)}$ and
$\isatt{is{\uscore}cardinal(M,x,c)}$ then
$\isatt{cardinal{\uscore}rel(M,x)}$ can be obtained by the operator of
definite descriptions.

Soon we realized that resorting to definite descriptions was needed
only for the most primitive concepts. In fact, once we have a
relativized concept, we can use it to define other relativizations.
For instance, $\isatt{cardinal{\uscore}rel}$ depends on having
relative versions of $\isatt{bij}$. Instead of relationalizing
$\isatt{bij}$ to get $\isatt{is{\uscore}bij}$ and then prove
uniqueness and existence of a witness, we define
$\isatt{bij{\uscore}rel}$ using $\isatt{inj{\uscore}rel}$ and
$\isatt{surj{\uscore}rel}$.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "independence_ch_isabelle"
%%% ispell-local-dictionary: "american"
%%% End:

### 2022-06-30 Independence_CH/lessons.tex
We didn't relativized Cofinality + one typo on DSL

@@
 \item
-  we only proved the Delta System Lemma for $\aleph_1$-sized families
-  (thus limiting us to the plain ccc);
+  we only proved the Delta System Lemma for $\aleph_1$-sized families;
+  thus limiting us to the plain ccc, and avoiding the relativization
+  of the material on cofinalities \cite{Delta_System_Lemma-AFP};
 \item we showed preservation of sequences by considering countably

### 2022-07-21 Independence_CH/details.tex
Signaling fake instances, and updating counts in the manuscript

@@
 
-\subsection{33 replacement instances to rule them all}
+\subsection{36 replacement instances to rule them all}
 \label{sec:repl-instances}
 
-We isolated 33 instances of Replacement that are sufficient to force
-$\CH$/$\neg\CH$. The first 14 instances are needed to set up
+We isolated 36 instances of Replacement that are sufficient to force
+$\CH$/$\neg\CH$. The first 15 instances are needed to set up
 cardinal arithmetic in $M$. Many of these were already present in
@@
 \begin{itemize}
-\item 7 instances for iteration of operations. For each construct,
+\item 8 instances for iteration of operations. For each construct,
   Paulson used one replacement instance in order to have absoluteness,
@@
     for absoluteness.
+  \item Absoluteness of the recursive construction in the proof of the
+    Schröder-Bernstein Theorem \isatt{banach{\uscore}iterates{\uscore}fm}.
   \end{itemize}
@@
 
-That makes 28 instances up to now. For the setup of forcing, we
-require the following 5 instances (the first two are needed for the $\Delta$-system Lemma):
+That makes 30 instances up to now. For the setup of forcing, we
+require the following 6 instances (the first two are needed for the $\Delta$-System Lemma):
 
@@
 \item Absoluteness of the previous construction \isatt{replacement{\uscore}transrec{\uscore}apply{\uscore}image{\uscore}body{\uscore}fm}.
+\item Absoluteness of the recursive construction in the proof of the
+    Dependent Choices \isatt{replacement{\uscore}dcwit{\uscore}repl{\uscore}body{\uscore}fm}.
 \item 2 instances for definitions by well-founded recursion.
@@
   definition of $\punto{G}$)
-  \isatt{G{\uscore}dot{\uscore}in{\uscore}M{\uscore}fm}.
+  \isatt{Lambda{\uscore}in{\uscore}M{\uscore}fm(check{\uscore}fm(2,0,1),1)}.
 % \item Name of choice functions in $M[G]$ \isatt{repl{\uscore}opname{\uscore}check{\uscore}fm}.
@@
 \end{itemize}
+
+The need for some of the instances above arose from the particular
+engineering of the some of the proofs. For instance, the construction
+of the cardinal successor should go through by reusing the replacement
+$x\mapsto \otype(x)$. Nevertheless, we relativized Paulson's
+approach direclty, which had an alternative formulation to obtain a
+simpler proof of Hartogs' Theorem (in particular, that $\kappa^+$ is
+indeed a cardinal).
+% https://isabelle.in.tum.de/dist/library/ZF/ZF/CardinalArith.html#CardinalArith.jump_cardinal_def|fact
 

### 2022-07-24 Independence_CH/details.tex
Updating count and sorting instances in the text

@@
 We isolated 36 instances of Replacement that are sufficient to force
-$\CH$/$\neg\CH$. The first 15 instances are needed to set up
+$\CH$ or $\neg\CH$. The first 15 instances are needed to set up
 cardinal arithmetic in $M$. Many of these were already present in
@@
 That makes 30 instances up to now. For the setup of forcing, we
-require the following 6 instances (the first two are needed for the $\Delta$-System Lemma):
+require the following 5 instances (the first two are needed for the $\Delta$-System Lemma):
 
@@
 \item Absoluteness of the previous construction \isatt{replacement{\uscore}transrec{\uscore}apply{\uscore}image{\uscore}body{\uscore}fm}.
-\item Absoluteness of the recursive construction in the proof of the
-    Dependent Choices \isatt{replacement{\uscore}dcwit{\uscore}repl{\uscore}body{\uscore}fm}.
 \item 2 instances for definitions by well-founded recursion.
@@
   \isatt{Lambda{\uscore}in{\uscore}M{\uscore}fm(check{\uscore}fm(2,0,1),1)}.
-% \item Name of choice functions in $M[G]$ \isatt{repl{\uscore}opname{\uscore}check{\uscore}fm}.
-% \item 2 instances for the slalom needed for cardinal preservation by ccc forcing.
+\end{itemize}
+%
+This is enough to force $\neg\CH$. To force $\CH$, one further instance is needed:
+%
+\begin{itemize}
+\item Absoluteness of the recursive construction in the proof of the
+  Dependent Choices \isatt{replacement{\uscore}dcwit{\uscore}repl{\uscore}body{\uscore}fm}.
 \end{itemize}
@@
   \mathbb{P}))^{M}$. This is the set of all names $\vartheta \in
-  M^{\mathcal{P}}$ such that $\operatorname{dom}(\vartheta) \subseteq
+  M^{\mathbb{P}}$ such that $\operatorname{dom}(\vartheta) \subseteq
   \operatorname{dom}(\tau)$. Let $\pi=Q \times\{\1\}$ and let
@@
 \textit{This is the set of all names $\vartheta \in
-  M^{\mathcal{P}}$} [\dots]---it is pretty laborious to show that things
+  M^{\mathbb{P}}$} [\dots]---it is pretty laborious to show that things
 are in $M$; it takes 17 more lines of code (not shown below) that

### 2022-08-02 Independence_CH/details.tex
Sit-rep on manuscript. Eliminating original comments by Larry and repeated arity thms

@@
 
-\subsection{36 replacement instances to rule them all}
+\subsection{34 replacement instances to rule them all}
 \label{sec:repl-instances}
 
-We isolated 36 instances of Replacement that are sufficient to force
-$\CH$ or $\neg\CH$. The first 15 instances are needed to set up
+We isolated 34 instances of Replacement that are sufficient to force
+$\CH$ or $\neg\CH$. The first 14 instances are needed to set up
 cardinal arithmetic in $M$. Many of these were already present in
@@
 \begin{itemize}
-\item 8 instances for iteration of operations. For each construct,
+\item 6 instances for iteration of operations. For each construct,
   Paulson used one replacement instance in order to have absoluteness,
@@
     \isatt{eclose{\uscore}repl2{\uscore}intf{\uscore}fm}.
+  \end{itemize}
+\item 2 instances for absoluteness of iterative constructions.
+  \begin{itemize}
   \item
@@
     \isatt{tl{\uscore}repl{\uscore}intf{\uscore}fm}. By transitivity,
-    we do not need an instance for closure; the only instance was required
-    for absoluteness.
+    we do not need an instance for closure.
   \item Absoluteness of the recursive construction in the proof of the
@@
 \item
-  2 instances for definition of ordertypes.
+  2 instances for ordertypes.
   \begin{itemize}
-  \item \isatt{replacement{\uscore}is{\uscore}order{\uscore}eq{\uscore}map{\uscore}fm}.
-  \item Replacement through $x\mapsto \otype(x)$ \isatt{replacement{\uscore}is{\uscore}order{\uscore}body{\uscore}fm}.
-  \end{itemize}
-\item 1 instance for definition of cardinal successor.
-  \begin{itemize}
-  \item
-    \isatt{replacement{\uscore}is{\uscore}jump{\uscore}cardinal{\uscore}body{\uscore}fm}.
+  \item Auxiliary for the definition of ordertypes
+    \isatt{replacement{\uscore}is{\uscore}order{\uscore}eq{\uscore}map{\uscore}fm}.
+  \item Replacement through $x\mapsto \otype(x)$, for Hartogs' Theorem
+    \isatt{replacement{\uscore}is{\uscore}order{\uscore}body{\uscore}fm}.
   \end{itemize}
@@
 
-That makes 30 instances up to now. For the setup of forcing, we
-require the following 5 instances (the first two are needed for the $\Delta$-System Lemma):
+That makes 28 instances up to now. For the setup of forcing, we
+require the following 5 instances (the last two are needed for the $\Delta$-System Lemma):
 
 \begin{itemize}
-%\verb|replacement_ax2(6)| \verb|replacement_transrec_apply_image_body_fm| seems fake!
-\item Recursive construction of sets using a choice function
-  \isatt{replacement{\uscore}is{\uscore}trans{\uscore}apply{\uscore}image{\uscore}fm}.
-% \item Replacement through  $x\mapsto \check{x}$ \isatt{check{\uscore}replacement{\uscore}fm}.
-%\item $x\mapsto |x|^M$
-%  \isatt{Lambda{\uscore}in{\uscore}M{\uscore}fm(is{\uscore}cardinal{\uscore}fm(0,1),0)}. % fake since the image is known
-\item Absoluteness of the previous construction \isatt{replacement{\uscore}transrec{\uscore}apply{\uscore}image{\uscore}body{\uscore}fm}.
 \item 2 instances for definitions by well-founded recursion.
@@
   \end{itemize}
-\item Replacement through $x\mapsto \lb\check{x},x\rb$ (for the
+\item Replacement through $x\mapsto \lb x,\check{x}\rb$ (for the
   definition of $\punto{G}$)
   \isatt{Lambda{\uscore}in{\uscore}M{\uscore}fm(check{\uscore}fm(2,0,1),1)}.
+\item Recursive construction of sets using a choice function
+  \isatt{replacement{\uscore}is{\uscore}trans{\uscore}apply{\uscore}image{\uscore}fm}.
+\item Absoluteness of the previous construction \isatt{replacement{\uscore}transrec{\uscore}apply{\uscore}image{\uscore}body{\uscore}fm}.
 \end{itemize}
@@
 The need for some of the instances above arose from the particular
-engineering of the some of the proofs. For instance, the construction
-of the cardinal successor should go through by reusing the replacement
-$x\mapsto \otype(x)$. Nevertheless, we relativized Paulson's
-approach direclty, which had an alternative formulation to obtain a
-simpler proof of Hartogs' Theorem (in particular, that $\kappa^+$ is
-indeed a cardinal).
-% https://isabelle.in.tum.de/dist/library/ZF/ZF/CardinalArith.html#CardinalArith.jump_cardinal_def|fact
+engineering of the some of the proofs. For instance, the first
+instance required for the definition of relative ordertypes arises
+from Paulson's \session{ZF-Constructible}.
+% https://isabelle.in.tum.de/dist/library/ZF/ZF-Constructible/Rank.html#offset_1123..1139
 

### 2022-08-03 Independence_CH/details.tex
Documenting the need for M_datatypes.
Documenting the need for one instance of replacement.

@@
   \end{itemize}
+  These instances together with \isatt{tl{\uscore}repl{\uscore}intf{\uscore}fm}
+  are needed to instantiate the locales \isatt{M{\uscore}datatypes} and
+  \isatt{M{\uscore}eclose}. The former is used in the relative definition of
+  \isatt{Vset}.
 \item 2 instances for absoluteness of iterative constructions.
@@
   \end{itemize}
+  The first one is needed for the interpretation of \isatt{M{\uscore}ordertype}.
 \item

### 2022-08-03 Independence_CH/details.tex
Updating counts once again. Deleting Zermelo's locale assms

--- 2022-08-02 Independence_CH/details.tex
@@
 
-\subsection{34 replacement instances to rule them all}
+\subsection{32 replacement instances to rule them all}
 \label{sec:repl-instances}
 
-We isolated 34 instances of Replacement that are sufficient to force
-$\CH$ or $\neg\CH$. The first 14 instances are needed to set up
+We isolated 32 instances of Replacement that are sufficient to force
+$\CH$ or $\neg\CH$. The first 13 instances are needed to set up
 cardinal arithmetic in $M$. Many of these were already present in
@@
   \end{itemize}
-\item 2 instances for absoluteness of iterative constructions.
-  \begin{itemize}
-  \item
+\item 1 instance for absoluteness of the definition of the
   $n$th element of a list (iteration of the tail operation)
@@
   we do not need an instance for closure.
-  \item Absoluteness of the recursive construction in the proof of the
-    Schröder-Bernstein Theorem \isatt{banach{\uscore}iterates{\uscore}fm}.
 \end{itemize}
+The instances so far
+are needed to interpret locales \isatt{M{\uscore}datatypes} and
+\isatt{M{\uscore}eclose}. The former is used in the relative definition of
+\isatt{Vset}.
+\begin{itemize}
 \item
@@
     \isatt{replacement{\uscore}is{\uscore}order{\uscore}eq{\uscore}map{\uscore}fm}.
+
+    This one is needed for the interpretation of
+    \isatt{M{\uscore}ordertype}.
+
   \item Replacement through $x\mapsto \otype(x)$, for Hartogs' Theorem
@@
 
-That makes 28 instances up to now. For the setup of forcing, we
+That makes 26 instances up to now. For the setup of forcing, we
 require the following 5 instances (the last two are needed for the $\Delta$-System Lemma):
@@
 
-The need for some of the instances above arose from the particular
-engineering of the some of the proofs. For instance, the first
-instance required for the definition of relative ordertypes arises
-from Paulson's \session{ZF-Constructible}.
+The particular choice of some of the instances above arose from
+Paulson's architecture on which we based our development.
+This applies every time
+a locale from \session{ZF-Constructible} has to be
+interpreted (namely \isatt{M{\uscore}datatypes}
+\isatt{M{\uscore}eclose}, and \isatt{M{\uscore}ordertype}).
+%% For instance, the first
+%% instance required for the definition of relative ordertypes arises
+%% from Paulson's \session{ZF-Constructible}.
 % https://isabelle.in.tum.de/dist/library/ZF/ZF-Constructible/Rank.html#offset_1123..1139
+
+On the other hand, as explained in
+Section~\ref{sec:zfc-axioms-as-locales}, we managed to eliminate the
+instance arising from the \isatt{M{\uscore}basic} locale. Similarly,
+we replaced the original proof of the Schröder-Bernstein Theorem by
+Zermelo's one \cite[Exr. x4.27]{moschovakis1994notes}, because the
+former required at least one extra instance
+% (\isatt{banach{\uscore}iterates{\uscore}fm})
+arising from an iteration.
 

### 2022-08-03 Independence_CH/discipline.tex
Updating counts once again. Deleting Zermelo's locale assms

@@
 As we said in Sec.~\ref{sec:relat-vers-non-absol}, in order to force
-CH and its negation we depended on having relativized versions of
+$\CH$ and its negation we depended on having relativized versions of
 cardinals, Alephs, etc. It was clear for us that our efforts would be

### 2022-08-03 Independence_CH/forcing.tex
Updating counts once again. Deleting Zermelo's locale assms

@@
 
-\subsection{The $\ZFC$ axioms as locales}
+\subsection{The $\ZFC$ axioms as locales}\label{sec:zfc-axioms-as-locales}
 The description of set models of fragments of $\ZFC$ was performed
@@
 %% \end{isabelle}
-Some of those instances, which we included for simplicity of design,
-are actually “fake” instances in that they can be obtained from
-Powerset, or from other replacement instances already available in the
-respective context. An example of this is the sole instance of the
-\isatt{M{\uscore}basic} locale, originally from
-\session{ZF-Constructible}; we managed to eliminate it but in
-doing so we had to reprove many lemmas from
-that session in the weakened context. Our locales
-\isatt{M{\uscore}ZF1} through \isatt{M{\uscore}ZF4} list a total of 86
-replacement instances, from which there are at least 53 fake ones (see
-Section~\ref{sec:repl-instances}).
+%
+%% Some of those instances, which we included for simplicity of design,
+%% are actually “fake” instances in that they can be obtained from
+%% Powerset, or from other replacement instances already available in the
+%% respective context. An example of this is the sole instance of the
+%% \isatt{M{\uscore}basic} locale, originally from
+%% \session{ZF-Constructible}; we managed to eliminate it but in
+%% doing so we had to reprove many lemmas from
+%% that session in the weakened context. 
+The locales gathering all 32 replacement instances necessary are named
+\isatt{M{\uscore}ZF1} through \isatt{M{\uscore}ZF4},
+\isatt{M{\uscore}ZF{\uscore}ground},
+\isatt{M{\uscore}ZF{\uscore}ground{\uscore}notCH}, and
+\isatt{M{\uscore}ZF{\uscore}ground{\uscore}CH} (see
+Section~\ref{sec:repl-instances} for details).
 

### 2022-08-03 Independence_CH/lessons.tex
Updating counts once again. Deleting Zermelo's locale assms

@@
 abstract but, as an example, the formula $\forceisa({\cdot}0\in 1{\cdot})$
-can be explicitly printed by Isabelle (it spans nearly 20k symbols), but
+can be explicitly printed by Isabelle2021-1 (it spans nearly 20k symbols), but
 $\forceisa({\cdot}\neg{\cdot}\neg{\cdot}0\in 1{\cdot}{\cdot}{\cdot})$

### 2022-08-08 Transitive_Models/document/root.tex
Comments on Card_[O]UN, Schröder-Bernstein. Update on Lam_Repl, missing section headers, style

@@
   treatment of the Axiom of Dependent Choices. This work is a
-  prerrequisite to our formalization of the independence of the
+  prerequisite to our formalization of the independence of the
   Continuum Hypothesis.
@@
 In the theory \theory{Lambda\_Replacement} we introduce a new locale assuming
-two instances of separation and twelve instances of ``lambda replacements''
-(ie, replacement of the form $\lambda x y. y=\langle x, f(x) \rangle$)
-that allow for having some form of compositionality of further instances
+two instances of separation and six instances of ``lambda replacements''
+(i.e., replacements using definable functions of the form $\lambda x y. y=\langle x, f(x) \rangle$)
+that allow some form of compositionality of further instances
 of separations and replacements.
@@
   \item \theory{CardinalArith}. The hardest part was to formalize the
-    cardinal successor function. We also disregarded the part treating
+    cardinal successor function and ordertypes.
+    We also disregarded the part treating
     finite cardinals since it is an absolute concept. Apart from that,
@@
   \begin{enumerate}
-  \item \theory{Pointed\_DC}. This case was similart to
-    \theory{Cardinal\_AC} above, alhough a bit of care was needed to
+  \item \theory{Pointed\_DC}. This case was similar to
+    \theory{Cardinal\_AC} above, although a bit of care was needed to
     handle the recursive construction. Also, a fraction of the theory
     \theory{AC} from \session{ZF} was ported here as it was a
-    prerrequisite. A complete relativization of \theory{AC} would be
+    prerequisite. A complete relativization of \theory{AC} would be
     desirable but still missing.

### 2022-08-10 Independence_CH/forcing.tex
A few more words on forcing

@@
 \subsection{The fundamental theorems}
-Now let $\lb \PP, {\preceq} ,\1\rb \in M$ be a forcing notion. Given $G\sbq \PP$, we have
+Let $\lb \PP, {\preceq} ,\1\rb \in M$ be a forcing notion. The
+$\val$ interpretation function takes $3$ arguments so that if $G\sbq \PP$, we have
 $M[G]\defi \{ \val(\PP,G,\punto{a}) : \punto{a}\in M \}$.
 
-The following form of the Forcing Theorems  is the one
-that we formalized.
-% Is the existential quantification here confusing?
+The version of the Forcing Theorems that we formalized follows the
+considerations on the $\forces^*$ relation as discussed in Kunen's new
+\emph{Set Theory}
+\cite[p.~257ff]{kunen2011set}. But in contrast to the point made on
+p.~260 of this book, we internalized the recursion to define the forcing relation,
+in that it involves codes for formulas, and the meta-theoretic formula
+transformer $\phi\mapsto\mathit{Forces}_\phi$ is replaced by the
+set-theoretic function $\forceisa$, which was defined by using
+Isabelle/ZF facilities for primitive recursion.
+
+Next we state the fundamental theorems in compact form.
 \begin{theorem}\label{th:forcing-thms}
@@
 \end{theorem}
-% Describe differences between Kunen's and our approach.
-% We can do everything using "internal forcing" (vs. forcing with
-% formulas in the metatheory).
-
-The statements appear as three separated items in
-\theory{Forcing{\uscore}Theorems.thy} in the
+We actually have to define $\forceisa$ before we state the fundamental
+theorems, so the main existential quantifier above does not appear in the
+formalization.
+Moreover, the statements appear as three separated items in
+\theory{Forcing{\uscore}Theorems} of our
 \session{Independence\_CH} session \cite{Independence_CH-AFP},
@@
 \end{isabelle}
-where the $\forces$ notation had already been set up in the
+where the $\forces$ notation (and its precedence) had already been set up in the
 \theory{Forces{\uscore}Definition} theory as follows:
@@
 
-We followed Kunen's new \textit{Set Theory} \cite{kunen2011set} to define
-$\forceisa$.  Forcing for atomic formulas is described as a mutual
+Kunen first describes forcing for atomic formulas using a mutual
 recursion
@@
 but then \cite[p.~257]{kunen2011set} it is cast as a single
-recursively defined function $\frcat$ over the wellfounded relation
-$\isatt{frecR}$ on tuples $\lb \mathit{ft},t_1,t_2,p\rb$ (where
+recursively defined function $F$ over a wellfounded  relation $R$.
+In our formalization, these are called $\frcat$ and 
+$\isatt{frecR}$, respectively, and are defined on tuples $\lb \mathit{ft},t_1,t_2,p\rb$ (where
 $\mathit{ft}\in\{0,1\}$ indicates whether the atomic formula being
-forced is an equality or a membership, respectively). Forcing for general formulas is defined by recursion on the
-datatype $\formula$. Technical details on the implementation and proofs of the
+forced is an equality or a membership, respectively).
+Forcing for general formulas is then defined by recursion on the
+datatype $\formula$ as indicated above. Technical details on the
+implementation and proofs of the
 Forcing Theorems have been spelled out in our

### 2022-08-11 Independence_CH/forcing.tex
Minor edits

@@
 \subsection{The fundamental theorems}
-Let $\lb \PP, {\preceq} ,\1\rb \in M$ be a forcing notion. The
+Let $\lb \PP, {\preceq} ,\1\rb \in M$ be a forcing notion. In order to
+fix the notation, the
 $\val$ interpretation function takes $3$ arguments so that if $G\sbq \PP$, we have
@@
 transformer $\phi\mapsto\mathit{Forces}_\phi$ is replaced by the
-set-theoretic function $\forceisa$, which was defined by using
+set-theoretic class function $\forceisa:: \tyi \fun \tyi$, which was defined by using
 Isabelle/ZF facilities for primitive recursion.
 
-Next we state the fundamental theorems in compact form.
+Next we state this version of the fundamental theorems in a compact
+way.
 \begin{theorem}\label{th:forcing-thms}
-  There exists a function  $\forceisa:: \tyi \fun  \tyi$
+  There exists a function  $\forceisa$
   such that for every
@@
 formalization.
-Moreover, the statements appear as three separated items in
+Moreover, the items in Theorem~\ref{th:forcing-thms} appear as three separated lemmas in
 \theory{Forcing{\uscore}Theorems} of our

### 2022-08-11 Independence_CH/forcing.tex
Comments on locales

@@
 The description of set models of fragments of $\ZFC$ was performed
-using Isabelle contexts (\emph{locales}) that fix a variable $M::\tyi$
+using locales that fix a variable $M::\tyi$
 and pack assumptions stating that $\lb M, \in\rb$ satisfy some
@@
 Section~\ref{sec:repl-instances} for details).
-
-%% Missing: Locales have to be interpreted (part of "big picture"?)
+These assumptions are then invoked to interpret at the class
+$\isatt{\#\#} M$ the relevant locales appearing in
+\session{ZF-Constructible}, and further ones required for the relative
+results from Section~\ref{sec:extension-isabellezf}.
 

### 2022-08-11 Independence_CH/proof_assists_isabelle.tex
Comments on locales

@@
 $L$ requires working with set models and satisfaction, which is
-defined for (“codes” of) formulas in the language $\{\in\}$.
+defined for (“codes” of) formulas in the language $\{\in \}$
+(viz.\ next Section~\ref{sec:internalized-formulas}).
+
+There is one further point concerning the organization of
+\session{ZF-Constructible}. Isabelle provides a very convenient way to
+define “contexts,” called \emph{locales}, in which some variables are
+fixed and assumptions are made. In the case of the constructibility
+library, several locales are defined where the variable $M$ assumed to
+denote a class satisfying certain finite amount of $\ZF$; the weakest
+one, \isatt{M{\uscore}trans} \cite[Sect.~3]{2020arXiv200109715G}, just
+assumes that $M$ is transitive nonempty. Inside such context, many
+absoluteness results are proved. In order to quote those results for a
+particular class $C$, one has to \emph{interpret} the locale at
+$C$, which amounts to prove that $C$ satisfies the assumptions made by
+the context.
 

### 2022-08-11 Independence_CH/related.tex
On ZFC_in_HOL

@@
 than the language of tactics of Lean.
-%% Comment on ZFC_in_HOL and other developments using it (mostly by
-%% Paulson: Ramsey theory, (ordinal) partition theorems)
+
+A sweet spot combining the best of both worlds is to be found on
+developments in Isabelle/HOL based on the AFP entry
+\session{ZFC\_in\_HOL} by Paulson \cite{ZFC_in_HOL-AFP}. There is a
+range of results in combinatorics and other set-theoretical material
+that was swiftly formalized in this setting: Nash-Williams theorem and
+Larson's $\forall k\ \omega^{\omega}\longrightarrow(\omega^\omega,k)$
+\cite{doi:10.1080/10586458.2021.1980464}, Design Theory
+\cite{10.1007/978-3-030-81097-9_1}, and Wetzel's problem
+\cite{2022arXiv220503159P}; this last paper describes in a brief and
+clear way the conveniences of the interaction with Isabelle/HOL
+(which, in this case, is also paid in consistency-strength currency
+\cite[Sect.~3]{DBLP:conf/ictac/Obua06}).
 

### 2022-08-11 Independence_CH/details.tex
Formal proof of Powerset now consistent with code

@@
 \ \ \isacommand{obtain}\isamarkupfalse%
-\ {\isasymtau}\ \isakeyword{where}\ {\isachardoublequoteopen}{\isasymtau}\ {\isasymin}\ M{\isachardoublequoteclose}\ {\isachardoublequoteopen}val{\isacharparenleft}{\kern0pt}P{\isacharcomma}{\kern0pt}G{\isacharcomma}{\kern0pt}\ {\isasymtau}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ a{\isachardoublequoteclose}\isanewline
+\ {\isasymtau}\ \isakeyword{where}\ {\isachardoublequoteopen}{\isasymtau}\ {\isasymin}\ M{\isachardoublequoteclose}\ {\isachardoublequoteopen}val{\isacharparenleft}{\kern0pt}G{\isacharcomma}{\kern0pt}\ {\isasymtau}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ a{\isachardoublequoteclose}\isanewline
 \ \ \ \ \isacommand{using}\isamarkupfalse%
@@
 \ {\isachardoublequoteopen}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}{\isasymtimes}P\ {\isasymin}\ M{\isachardoublequoteclose}\ {\isachardoublequoteopen}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}\ {\isasymin}\ M{\isachardoublequoteclose}\isanewline
-\ \ \ \ \isacommand{using}\isamarkupfalse%
-\ domain{\isacharunderscore}{\kern0pt}closed\ cartprod{\isacharunderscore}{\kern0pt}closed\ P{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}M\isanewline
 \ \ \ \ \isacommand{by}\isamarkupfalse%
@@
 \ \ \isacommand{let}\isamarkupfalse%
-\ {\isacharquery}{\kern0pt}b{\isacharequal}{\kern0pt}{\isachardoublequoteopen}val{\isacharparenleft}{\kern0pt}P{\isacharcomma}{\kern0pt}G{\isacharcomma}{\kern0pt}{\isacharquery}{\kern0pt}{\isasympi}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}%% \isanewline
+\ {\isacharquery}{\kern0pt}b{\isacharequal}{\kern0pt}{\isachardoublequoteopen}val{\isacharparenleft}{\kern0pt}G{\isacharcomma}{\kern0pt}{\isacharquery}{\kern0pt}{\isasympi}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}%% \isanewline
 %% \ \ \isacommand{from}\isamarkupfalse%
@@
 \ \ \ \ \isacommand{obtain}\isamarkupfalse%
-\ {\isasymchi}\ \isakeyword{where}\ {\isachardoublequoteopen}c{\isasymin}M{\isacharbrackleft}{\kern0pt}G{\isacharbrackright}{\kern0pt}{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymchi}\ {\isasymin}\ M{\isachardoublequoteclose}\ {\isachardoublequoteopen}val{\isacharparenleft}{\kern0pt}P{\isacharcomma}{\kern0pt}G{\isacharcomma}{\kern0pt}{\isasymchi}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ c{\isachardoublequoteclose}\isanewline
+\ {\isasymchi}\ \isakeyword{where}\ {\isachardoublequoteopen}c{\isasymin}M{\isacharbrackleft}{\kern0pt}G{\isacharbrackright}{\kern0pt}{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymchi}\ {\isasymin}\ M{\isachardoublequoteclose}\ {\isachardoublequoteopen}val{\isacharparenleft}{\kern0pt}G{\isacharcomma}{\kern0pt}{\isasymchi}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ c{\isachardoublequoteclose}\isanewline
 \ \ \ \ \ \ \isacommand{using}\isamarkupfalse%
@@
 \ \ \ \ \ \ \isacommand{using}\isamarkupfalse%
-\ P{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}M\ one{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}M\ leq{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}M\ sats{\isacharunderscore}{\kern0pt}fst{\isacharunderscore}{\kern0pt}snd{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}M\isanewline
+\ sats{\isacharunderscore}{\kern0pt}fst{\isacharunderscore}{\kern0pt}snd{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}M\isanewline
 \ \ \ \ \ \ \isacommand{by}\isamarkupfalse%
@@
 \ \ \ \ \isacommand{have}\isamarkupfalse%
-\ {\isachardoublequoteopen}val{\isacharparenleft}{\kern0pt}P{\isacharcomma}{\kern0pt}G{\isacharcomma}{\kern0pt}{\isacharquery}{\kern0pt}{\isasymtheta}{\isacharparenright}{\kern0pt}\ {\isasymin}\ {\isacharquery}{\kern0pt}b{\isachardoublequoteclose}\isanewline
+\ {\isachardoublequoteopen}val{\isacharparenleft}{\kern0pt}G{\isacharcomma}{\kern0pt}{\isacharquery}{\kern0pt}{\isasymtheta}{\isacharparenright}{\kern0pt}\ {\isasymin}\ {\isacharquery}{\kern0pt}b{\isachardoublequoteclose}\isanewline
 \ \ \ \ \ \ \isacommand{using}\isamarkupfalse%
-\ one{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}G\ one{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}P\ generic\ val{\isacharunderscore}{\kern0pt}of{\isacharunderscore}{\kern0pt}elem\ {\isacharbrackleft}{\kern0pt}of\ {\isacharquery}{\kern0pt}{\isasymtheta}\ {\isasymone}\ {\isacharquery}{\kern0pt}{\isasympi}\ G{\isacharbrackright}{\kern0pt}\isanewline
+\ one{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}G\ generic\ val{\isacharunderscore}{\kern0pt}of{\isacharunderscore}{\kern0pt}elem\ {\isacharbrackleft}{\kern0pt}of\ {\isacharquery}{\kern0pt}{\isasymtheta}\ {\isasymone}\ {\isacharquery}{\kern0pt}{\isasympi}\ G{\isacharbrackright}{\kern0pt}\isanewline
 \ \ \ \ \ \ \isacommand{by}\isamarkupfalse%
@@
 \ \ \ \ \isacommand{have}\isamarkupfalse%
-\ {\isachardoublequoteopen}val{\isacharparenleft}{\kern0pt}P{\isacharcomma}{\kern0pt}G{\isacharcomma}{\kern0pt}{\isacharquery}{\kern0pt}{\isasymtheta}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ c{\isachardoublequoteclose}
+\ {\isachardoublequoteopen}val{\isacharparenleft}{\kern0pt}G{\isacharcomma}{\kern0pt}{\isacharquery}{\kern0pt}{\isasymtheta}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ c{\isachardoublequoteclose}
 \end{isabelle}

### 2022-08-11 Independence_CH/details.tex
Comment on Naproche

@@
 The assumption $a\in M[G]$ appears in the lemma statement, and the
-goal involving $b$ in the first sentence will appear below (page \pageref{goal-on-b}); formalized
+goal involving $b$ in the first sentence will appear below (signaled
+by “{\small (**)}”); formalized
 material necessarily tends to be much more linear than usual prose. In
@@
 \ \ \isacommand{have}\isamarkupfalse%
-\ {\isachardoublequoteopen}Pow{\isacharparenleft}{\kern0pt}a{\isacharparenright}{\kern0pt}\ {\isasyminter}\ M{\isacharbrackleft}{\kern0pt}G{\isacharbrackright}{\kern0pt}\ {\isasymsubseteq}\ {\isacharquery}{\kern0pt}b{\isachardoublequoteclose}\isanewline
+\ {\isachardoublequoteopen}Pow{\isacharparenleft}{\kern0pt}a{\isacharparenright}{\kern0pt}\ {\isasyminter}\ M{\isacharbrackleft}{\kern0pt}G{\isacharbrackright}{\kern0pt}\ {\isasymsubseteq}\ {\isacharquery}{\kern0pt}b{\isachardoublequoteclose}\hfill
+\mbox{\rm\small(**)}\isanewline
 \ \ \isacommand{proof}\isamarkupfalse%
@@
 
+Undoubtedly, even for this cherry-picked example, the formalization
+looks codish. It is therefore inevitable to compare this to the
+magnificent results obtained by P.~Koepke and his team by using
+Isabelle/Naproche \cite{10.1007/978-3-030-81097-9_2} (notoriously,
+König's Theorem). The trick there consists in presenting the result
+being formalized as restricted first-order problem, and then each
+proof step can be handled by an automatic theorem prover.
+
 %%% Local Variables: 

### 2022-08-11 Independence_CH/related.tex
Correct statement of results of Flypitch

@@
 Lemma. Putting this together they readily obtain a proof that
-(*) $\Con(\ZFC)$ implies 
 $\ZFC\nvdash \CH$ \cite{han_et_al:LIPIcs:2019:11074} and after
 formalizing collapse forcing they show  $\ZFC\nvdash \neg\CH$
-\cite{DBLP:conf/cpp/HanD20}.
+\cite[Sect.~5.6]{DBLP:conf/cpp/HanD20}.
 
@@
 \cite{han_et_al:LIPIcs:2019:11074}. It is not clear to us if they are
-able to avoid strength (even for the proof of the finitary implication
-(*)).
+able to avoid such strength (at least $\Con(\ZFC)$ is readily
+available in order to get either unprovability result).
+% by Gödel's Second).
 

### 2022-08-11 Independence_CH/related.tex
typo

@@
 \cite{2022arXiv220503159P}; this last paper describes in a brief and
-clear way the conveniences of the interaction with Isabelle/HOL
+clear way the convenience of the interaction with Isabelle/HOL
 (which, in this case, is also paid in consistency-strength currency

### 2022-08-11 Independence_CH/related.tex
Now I now where that 's' came from

@@
 type-theoretic machinery used in Flypitch, and % might be a bit alien
-since absoluteness grant us extra naturality. This last point may also
+since absoluteness grants us extra naturality. This last point may also
 be illustrated by the treatment of ordinals; in our formalization, as

### 2022-08-11 Independence_CH/details.tex
Different order of presentation of repl insts, with a bit more of info

@@
 \begin{itemize}
-\item 6 instances for iteration of operations. For each construct,
+\item 6 instances for iteration of operations (between parentheses). For each construct,
   Paulson used one replacement instance in order to have absoluteness,
@@
   \item
-    Lists
-    \isatt{list{\uscore}repl1{\uscore}intf{\uscore}fm},
+    \isatt{list{\uscore}repl1{\uscore}intf{\uscore}fm} and 
     \isatt{list{\uscore}repl2{\uscore}intf{\uscore}fm}.
+
+    For lists on a set $A$ ($X\mapsto \{0 \} \oplus A \times X$, where
+    $D\oplus E$ stands for the disjoint union/coproduct $(\{0 \}\times
+    D) \cup (\{1 \}\times E)$).
+    %
   \item
-    Formulas
-    \isatt{formula{\uscore}repl1{\uscore}intf{\uscore}fm},
+    \isatt{formula{\uscore}repl1{\uscore}intf{\uscore}fm} and
     \isatt{formula{\uscore}repl1{\uscore}intf{\uscore}fm}.
+
+    For (codes of) first order formulas ($X\mapsto (\omega \times
+    \omega \oplus \omega \times \omega) \oplus X \times X \oplus X$).
+    %
   \item
-    Transitive closure %(iteration of $\union$)
-    \isatt{eclose{\uscore}repl1{\uscore}intf{\uscore}fm},
+    \isatt{eclose{\uscore}repl1{\uscore}intf{\uscore}fm} and 
     \isatt{eclose{\uscore}repl2{\uscore}intf{\uscore}fm}.
+
+    For transitive closure ($X\mapsto\union X$).
   \end{itemize}
-\item 1 instance for absoluteness of the definition of the
-  $n$th element of a list (iteration of the tail operation)
-  \isatt{tl{\uscore}repl{\uscore}intf{\uscore}fm}. By transitivity,
-  we do not need an instance for closure.
+\item \isatt{tl{\uscore}repl{\uscore}intf{\uscore}fm}
+
+  Absoluteness of the definition of the
+  $n$th element of a list (iteration of the tail operation).
+  By transitivity, we do not need an instance for closure.
 \end{itemize}
@@
   \begin{itemize}
-  \item Auxiliary for the definition of ordertypes
+  \item
     \isatt{replacement{\uscore}is{\uscore}order{\uscore}eq{\uscore}map{\uscore}fm}.
 
-    This one is needed for the interpretation of
-    \isatt{M{\uscore}ordertype}.
+    Auxiliary instance for the definition of ordertypes, needed for
+    the interpretation of \isatt{M{\uscore}ordertype}.
+    %
+  \item
+    \isatt{replacement{\uscore}is{\uscore}order{\uscore}body{\uscore}fm}.
 
-  \item Replacement through $x\mapsto \otype(x)$, for Hartogs' Theorem
-    \isatt{replacement{\uscore}is{\uscore}order{\uscore}body{\uscore}fm}.
+    Replacement through $x\mapsto \otype(x)$, for Hartogs' Theorem.
   \end{itemize}
@@
   \begin{itemize}
-  \item Ordertypes
-    \isatt{wfrec{\uscore}replacement{\uscore}order{\uscore}pred{\uscore}fm}.
-  \item Rank \isatt{wfrec{\uscore}rank{\uscore}fm}.
-  \item Cumulative hierarchy (rank initial segments) \isatt{trans{\uscore}repl{\uscore}HVFrom{\uscore}fm}.
-  \item Aleph \isatt{replacement{\uscore}HAleph{\uscore}wfrec{\uscore}repl{\uscore}body{\uscore}fm}.
+  \item \isatt{wfrec{\uscore}replacement{\uscore}order{\uscore}pred{\uscore}fm}.
+
+    For ordertypes
+    %
+  \item \isatt{wfrec{\uscore}rank{\uscore}fm}.
+
+    For $\in$-rank.
+    %
+  \item \isatt{trans{\uscore}repl{\uscore}HVFrom{\uscore}fm}.
+
+    For the cumulative hierarchy (rank initial segments).
+    %
+  \item
+    \isatt{replacement{\uscore}HAleph{\uscore}wfrec{\uscore}repl{\uscore}body{\uscore}fm}.
+
+    For Aleph.
   \end{itemize}
@@
 \]
+Here, $\mathit{fst}(\lb a,b\rb) = a$, $\mathit{snd}(\lb a,b\rb) = b$
+(with default value $0$ for non pairs).
 In our development, the mapping $\phi\mapsto\psi$ defined above is given by the
@@
   \begin{itemize}
-  \item
-    Check names
-    \isatt{wfrec{\uscore}Hcheck{\uscore}fm}.
-  \item Forcing for atomic formulas
-    \isatt{wfrec{\uscore}Hfrc{\uscore}at{\uscore}fm}.
-  \end{itemize}
-\item Replacement through $x\mapsto \lb x,\check{x}\rb$ (for the
-  definition of $\punto{G}$)
-  \isatt{Lambda{\uscore}in{\uscore}M{\uscore}fm(check{\uscore}fm(2,0,1),1)}.
-\item Recursive construction of sets using a choice function
-  \isatt{replacement{\uscore}is{\uscore}trans{\uscore}apply{\uscore}image{\uscore}fm}.
-\item Absoluteness of the previous construction \isatt{replacement{\uscore}transrec{\uscore}apply{\uscore}image{\uscore}body{\uscore}fm}.
+  \item \isatt{wfrec{\uscore}Hcheck{\uscore}fm}.
+
+    For check-names.
+    %
+  \item \isatt{wfrec{\uscore}Hfrc{\uscore}at{\uscore}fm}.
+
+    Definition of forcing for atomic formulas.
   \end{itemize}
   %
-This is enough to force $\neg\CH$. To force $\CH$, one further instance is needed:
+\item
+  \isatt{Lambda{\uscore}in{\uscore}M{\uscore}fm(check{\uscore}fm(2,0,1),1)}.
+
+  Replacement through $x\mapsto \lb x,\check{x}\rb$ (for the
+  definition of $\punto{G}$).
+  %
+\item
+  \isatt{replacement{\uscore}is{\uscore}trans{\uscore}apply{\uscore}image{\uscore}fm}.
+
+  Recursive construction of sets using a choice function (as in the
+  construction of a wellorder of $X$ given a choice function on $\Pow(X)$).
+  %
+\item
+  \isatt{replacement{\uscore}transrec{\uscore}apply{\uscore}image{\uscore}body{\uscore}fm}.
+
+  Absoluteness of the previous construction.
+\end{itemize}
+%
+This is enough to force $\neg\CH$. To force $\CH$, we required one further instance:
 %
 \begin{itemize}
-\item Absoluteness of the recursive construction in the proof of the
-  Dependent Choices \isatt{replacement{\uscore}dcwit{\uscore}repl{\uscore}body{\uscore}fm}.
+\item
+  \isatt{replacement{\uscore}dcwit{\uscore}repl{\uscore}body{\uscore}fm}.
+
+  Absoluteness of the recursive construction in the proof of the
+  Dependent Choices from $\AC$.
 \end{itemize}

### 2022-08-11 Independence_CH/independence_ch_isabelle.tex
Target journal now unspecified
One bib, one more entry
Another possible title
typo :facepalm:
First brief description of the discipline.
ITP -> assistant. Updated abstract

--- 2022-03-26 JSL/independence_ch_isabelle.tex
@@
 %
-\title{Some lessons after the formalization of the ctm approach to forcing%
+\title{The formal verification of the ctm approach to forcing%Some lessons after the formalization of the ctm approach to forcing%
   \thanks{Supported by Secyt-UNC project 33620180100465CB and Conicet.}%
@@
 %
-\titlerunning{Lessons after formalizing ctm forcing}
+\titlerunning{Formalization of ctm forcing}%Lessons after formalizing ctm forcing}
 % If the paper title is too long for the running head, you can set
@@
 \begin{abstract}
-  We'll discuss some highlights of our computer-verified
-  proof of the construction, given a countable transitive set model $M$
+  We discuss some highlights of our computer-verified
+  proof of the construction, given a countable transitive set-model $M$
   of $\ZFC$, of generic extensions  satisfying $\ZFC+\neg\CH$ and $\ZFC+\CH$.
-  In particular,
-  we isolated a set $\Delta$ of 39 instances
-  of Axiom of Replacement and a function $F$
-  such that such that for any finite fragment $\Phi\sbq\ZFC$,
-  $F(\Phi)\sbq\ZFC$ is also finite and if
-  $M\models F(\Phi) + \Delta$ then $M[G]\models \Phi + \neg \CH$.
+  Moreover, let $\calR$ be the set of instances of the Axiom of
+  Replacement. We isolated a 32-element subset $\Omega\sbq\calR$ and
+  defined $\calF:\calR\to\calR$
+  such that such that for every finite $\Phi\sbq\calR$
+  we have
+  $M\models Z \cup \calF\dimg\Phi \cup \Delta$ implies $M[G]\models Z
+  \cup \Phi \cup \{ \neg \CH \}$ for any $M$-generic $G$, where $Z$ is
+  Zermelo set theory.
   We also obtained the formulas yielded by the Forcing Definability Theorem
@@
 
-  We'll also compare our formalization with the recent one by Jesse
-  M.~Han and Floris van Doorn in the proof assistant \emph{Lean}.
+  We'll also compare our formalization with the recent one by J.~M.~Han and F.~van Doorn in the proof assistant \emph{Lean}.
 
@@
 \input{intro}
-%% 
-%% \input{isabelle_metatheory}
-%% 
-%% \input{absoluteness}
-%% 
+
+\input{proof_assists_isabelle}
+
+\input{non-absolute}
+
 %% \input{forces_definition}
+
+\input{forcing}
+
+\input{details}
 %% 
@@
 %%
-\input{random}
 
@@
 \input{lessons}
-%%
-%% \input{conclusion}
+
+\input{conclusion}
 %
@@
 \bibliographystyle{mi-estilo-else}
-\bibliography{independence_ch_isabelle,clam2021}
+\bibliography{independence_ch_isabelle}
 
+\input{discipline}
 \end{document}

### 2022-08-11 Independence_CH/intro.tex
ITP -> assistant. Updated abstract

@@
 We formalized a model-theoretic rendition of forcing (Section~\ref{sec:forcing}), showing how to
-construct proper extension of ctms of $\ZF$ (respectively, with
+construct proper extensions of ctms of $\ZF$ (respectively, with
 $\AC$), and we formalized the basic forcing notions required to obtain
@@
 formalization is closest to a proof in $\ZF$ of the previous
-implications.
+constructions.
 

### 2022-08-11 Independence_CH/proof_assists_isabelle.tex
ITP -> assistant. Updated abstract

@@
 
-Let us briefly introduce Isabelle in the large landscape of proof
-assistans (or ITPs for interactive theorem provers); we refer to the
-excelent chapter by \citet{DBLP:series/hhl/HarrisonUW14} for a more
-thorough reconstruction of the history of ITPs.
+Let us briefly introduce Isabelle in the large landscape of \emph{proof
+assistants} (“assitants” for short; also known as “interactive theorem provers”); we refer to the
+excellent chapter by \citet{DBLP:series/hhl/HarrisonUW14} for a more
+thorough reconstruction of the history of assistants.
 
-It is expected that an ITP assists the human user while mechanizing
+It is expected that an assistant aids the human user while mechanizing
 some piece of mathematics; the interaction varies from system to
 system, but a common interface consists on the system showing the
-current goal and context. The user instructs the ITP to modify the
+current goal and context. The user instructs the assistant to modify the
 goal/context by means of tactics; a forward rule changes the context
@@
 disjunction). Moreover one should be able to ascertain the validity of
-the reaction of an ITP (be its rejection or its acceptance of some
+the reaction of an assistant (be its rejection or its acceptance of some
 alleged logical step). Edinburgh LCF~\cite{Gordon1979-qm} was an
@@
 
-In contrast with LCF, some of the more prominent ITPs of today are
+In contrast with LCF, some of the more prominent assistants of today are
 based on some type theory. Both Coq~\cite{coq} and
-lean~\cite{DBLP:conf/cade/Moura021} are based on the Calculus of
+Lean~\cite{DBLP:conf/cade/Moura021} are based on the Calculus of
 Inductive Constructions
@@
 theory~\cite{DBLP:books/daglib/0000395}. Mizar~\cite{mizar} is the
-oldest ITP still used today but is far away both in terms of
+oldest assistant still used today but is far away both in terms of
 foundations and architecture from Isabelle; Mizar inspired, though, the
-Isar dialect used in nowadays Isabelle.
+Isar dialect used in Isabelle nowadays.
 

### 2022-08-11 Independence_CH/related.tex
Minor edits

@@
 \cite{han_et_al:LIPIcs:2019:11074}. It is not clear to us if they are
-able to avoid such strength (at least $\Con(\ZFC)$ is readily
-available in order to get either unprovability result).
+able to avoid such strength (at least $\Con(\ZFC)$ is provable in
+their context, a requisite to get either unprovability result).
 % by Gödel's Second).
@@
 \cite[p.~11]{han_et_al:LIPIcs:2019:11074}. Our formalization works
-entirely in $\ZF$, and only import the Isabelle theory postulating
+entirely in $\ZF$, and only imports the Isabelle theory postulating
 $\AC$ for demostration purposes.

### 2022-08-12 Independence_CH/details.tex
ITP -> assistant. Updated abstract
Minor edits
Appendix on lam_replacements. Macro for locales

@@
 The instances so far
-are needed to interpret locales \isatt{M{\uscore}datatypes} and
-\isatt{M{\uscore}eclose}. The former is used in the relative definition of
-\isatt{Vset}.
+are needed to interpret locales \locale{M{\uscore}datatypes} and
+\locale{M{\uscore}eclose}.
+%% \locale{M{\uscore}datatypes} is used in the relative definition of
+%% \isatt{Vset}.
 \begin{itemize}
@@
     Auxiliary instance for the definition of ordertypes, needed for
-    the interpretation of \isatt{M{\uscore}ordertype}.
+    the interpretation of \locale{M{\uscore}ordertype}.
     %
@@
 $\isatt{ground{\uscore}repl{\uscore}fm}$ function, and all ground replacement
-instances appear in the locale \isatt{M{\uscore}ZF4}. These are expressed using
+instances appear in the locale \locale{M{\uscore}ZF4}. These are expressed using
 the \isatt{ground{\uscore}replacement{\uscore}assm(M,\isasymphi)} predicate
@@
 a locale from \session{ZF-Constructible} has to be
-interpreted (namely \isatt{M{\uscore}datatypes}
-\isatt{M{\uscore}eclose}, and \isatt{M{\uscore}ordertype}).
+interpreted (namely \locale{M{\uscore}datatypes}
+\locale{M{\uscore}eclose}, and \locale{M{\uscore}ordertype}).
 %% For instance, the first
@@
 Section~\ref{sec:zfc-axioms-as-locales}, we managed to eliminate the
-instance arising from the \isatt{M{\uscore}basic} locale. Similarly,
+instance arising from the \locale{M{\uscore}basic} locale. Similarly,
 we replaced the original proof of the Schröder-Bernstein Theorem by
@@
 Undoubtedly, even for this cherry-picked example, the formalization
-looks codish. It is therefore inevitable to compare this to the
+looks a bit codish. It is therefore inevitable to compare this to the
 magnificent results obtained by P.~Koepke and his team by using

### 2022-08-12 Independence_CH/forcing.tex
Minor edits
Appendix on lam_replacements. Macro for locales

@@
 and pack assumptions stating that $\lb M, \in\rb$ satisfy some
-axioms. For example, the locale \isatt{M{\uscore}Z{\uscore}basic}
+axioms. For example, the locale \locale{M{\uscore}Z{\uscore}basic}
 states that Zermelo set theory holds in $M$. The Union Axiom (\isatt{Union{\uscore}ax}), for
@@
 Replacement, respectively. In the statement of the Separation Axiom in
-\isatt{M{\uscore}Z{\uscore}basic} and
+\locale{M{\uscore}Z{\uscore}basic} and
 the many replacement instances, predicates $Q$ and $R$ are actually
@@
 %% respective context. An example of this is the sole instance of the
-%% \isatt{M{\uscore}basic} locale, originally from
+%% \locale{M{\uscore}basic} locale, originally from
 %% \session{ZF-Constructible}; we managed to eliminate it but in
@@
 The locales gathering all 32 replacement instances necessary are named
-\isatt{M{\uscore}ZF1} through \isatt{M{\uscore}ZF4},
-\isatt{M{\uscore}ZF{\uscore}ground},
-\isatt{M{\uscore}ZF{\uscore}ground{\uscore}notCH}, and
-\isatt{M{\uscore}ZF{\uscore}ground{\uscore}CH} (see
+\locale{M{\uscore}ZF1} through \locale{M{\uscore}ZF4},
+\locale{M{\uscore}ZF{\uscore}ground},
+\locale{M{\uscore}ZF{\uscore}ground{\uscore}notCH}, and
+\locale{M{\uscore}ZF{\uscore}ground{\uscore}CH} (see
 Section~\ref{sec:repl-instances} for details).
@@
 formalization.
-Moreover, the items in Theorem~\ref{th:forcing-thms} appear as three separated lemmas in
+Moreover, the items in Theorem~\ref{th:forcing-thms} appear as three
+separated lemmas in the theory
 \theory{Forcing{\uscore}Theorems} of our

### 2022-08-12 Independence_CH/lam_replacement.tex
Appendix on lam_replacements. Macro for locales

\section{Lambda replacements}\label{sec:lambda-replacements}

The development of the locale structure of the project was a dynamical
process. As further properties of closure of the ground $M$ were
required, we gather the relevant instances of Separation and
Replacement into a new locale (always assuming a class model, for
added generality), and proceded to apply them to those closure proofs.

This procedure lead to a steady grow in the number of interpretation
obligations and therefore, of formula synthesis (since the two axiom
schemes were postulated using codes for formulas). That number would
easily surpass the hundred, and the automatic tools at our disposal
for that task were rudimentary (as discussed in
Section~\ref{sec:bureaucracy-scale-factors}).

Faced with this situation, we decided that we needed some sort of
\emph{compositionality} in order to obtain instances new instances
from the ones already proved: Having Replacement for class functions
$F$ and $G$ does not entail immediately replacement under $F\circ G$
(unless you use one further instance of Separation, and the net gain
is zero). The solution was to postulate for the relevant $F$s, instead
of replacement through $x\mapsto F(x)$, a \emph{lambda replacement}
through $x\mapsto \lb x,F(x)\rb$. The name “lambda” corresponds to the
fact that this type of replacement is equivalent to closure under
$(\lambda x\in A.\ F(x)) \defi \{ \lb x,F(x)\rb : x\in A \}$ for every
$A\in M$.

Now, a fixed set of six replacements and one separation (apart from
those in \locale{M{\uscore}basic}, which also assumes the Powerset
Axiom for the class $M$) is sufficient to obtain the lambda
replacement under $x\mapsto \lb x,F(G(x))\rb$ given those for $F$ and
$G$. To obtain compositions with binary class functions $H$, it is
enough to assume the lambda replacement
$x\mapsto \lb x,H(\mathit{fst}(x),\mathit{snd}(x)))\rb$.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "independence_ch_isabelle"
%%% ispell-local-dictionary: "american"
%%% End:

### 2022-08-12 Independence_CH/lessons.tex
Appendix on lam_replacements. Macro for locales

@@
 \subsection{Bureaucracy and scale factors}
+\label{sec:bureaucracy-scale-factors}
 
@@
 \isatt{auto}, etc) stop working just because of the sheer size of the
-goal (v.g., the same statement with 7 variables succeed but with 8
+goal (v.g., the same statement with 7 variables succeeds but with 8
 variables doesn't). Scale issues are very easily disregarded in the

### 2022-08-12 Independence_CH/proof_assists_isabelle.tex
Appendix on lam_replacements. Macro for locales

@@
 denote a class satisfying certain finite amount of $\ZF$; the weakest
-one, \isatt{M{\uscore}trans} \cite[Sect.~3]{2020arXiv200109715G}, just
+one, \locale{M{\uscore}trans} \cite[Sect.~3]{2020arXiv200109715G}, just
 assumes that $M$ is transitive nonempty. Inside such context, many

### 2022-08-12 Independence_CH/cofinality.tex
Optional section on cofinality

\section{Recursions in cofinality}\label{sec:recursions-cofinality}

As we mentioned near the end of
Section~\ref{sec:aims-formalization-planning}, we decided to minimize
the requirements being formalized in order to achieve our immediate
goal. In particular, the treatment of cofinality in the companion
project \cite{Delta_System_Lemma-AFP} was left behind.

We already observed that well-founded, and in particular transfinite,
recursion is not easily dealt with in Isabelle/ZF. Nevertheless, and
mainly as a curiosity, we found out that only one recursive
construction is needed for the development of the basic theory of
cofinality (as in \cite[Sect.~I.13]{kunen2011set}), which is used in
the proof of the following “factorization” lemma:

\begin{lemma}
  Let $\del,\ga\in\Ord$ and assume $f:\del\to\ga$ is cofinal.  There exists
  a strictly increasing $g:\cf(\ga)\to \del$ such that $f\circ g$ is
  strictly increasing and cofinal in $\ga$. Moreover, if $f$ is
  strictly increasing, then $g$ must also be cofinal.
\end{lemma}

It turns out that the rest of the basic results on cofinality (namely,
idempotence of $\cf$, that regular ordinals are cardinals, the
cofinality of Alephs, König's Theorem) follow easily from the previous
Lemma by “algebraic” reasoning only.

We therefore expect that the relativization of these
results be straightforward, when time permits.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "independence_ch_isabelle"
%%% ispell-local-dictionary: "american"
%%% End:

### 2022-08-14 Independence_CH/forcing.tex
Defs Main as an appendix

@@
 
-\subsection{The fundamental theorems}
+%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
+
+\subsection{The fundamental theorems}\label{sec:fundamental-theorems}
 Let $\lb \PP, {\preceq} ,\1\rb \in M$ be a forcing notion. In order to
@@
 
+%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
+\subsection{Models for $\CH$ and its negation}
+\label{sec:models-ch-negation}
+
+The statements of the existence of models of $\ZFC + \neg\CH$ and of
+$\ZFC + \CH$ negation appear in our formalization as follows:
+
+\begin{isabelle}
+\isacommand{corollary}\isamarkupfalse%
+\ ctm{\isacharunderscore}{\kern0pt}ZFC{\isacharunderscore}{\kern0pt}imp{\isacharunderscore}{\kern0pt}ctm{\isacharunderscore}{\kern0pt}not{\isacharunderscore}{\kern0pt}CH{\isacharcolon}{\kern0pt}\isanewline
+\ \ \isakeyword{assumes}\isanewline
+\ \ \ \ {\isachardoublequoteopen}M\ {\isasymapprox}\ {\isasymomega}{\isachardoublequoteclose}\ {\isachardoublequoteopen}Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ ZFC{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\isanewline
+\ \ \ \ {\isachardoublequoteopen}{\isasymexists}N{\isachardot}{\kern0pt}\isanewline
+\ \ \ \ \ \ M\ {\isasymsubseteq}\ N\ {\isasymand}\ N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\ Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\ N\ {\isasymTurnstile}\ ZFC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}{\isasymnot}{\isasymcdot}CH{\isasymcdot}{\isasymcdot}{\isacharbraceright}{\kern0pt}\ {\isasymand}\isanewline
+\ \ \ \ \ \ {\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}
+\end{isabelle}
+
+\begin{isabelle}
+\isacommand{corollary}\isamarkupfalse%
+\ ctm{\isacharunderscore}{\kern0pt}ZFC{\isacharunderscore}{\kern0pt}imp{\isacharunderscore}{\kern0pt}ctm{\isacharunderscore}{\kern0pt}CH{\isacharcolon}{\kern0pt}\isanewline
+\ \ \isakeyword{assumes}\isanewline
+\ \ \ \ {\isachardoublequoteopen}M\ {\isasymapprox}\ {\isasymomega}{\isachardoublequoteclose}\ {\isachardoublequoteopen}Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ ZFC{\isachardoublequoteclose}\isanewline
+\ \ \isakeyword{shows}\isanewline
+\ \ \ \ {\isachardoublequoteopen}{\isasymexists}N{\isachardot}{\kern0pt}\isanewline
+\ \ \ \ \ \ M\ {\isasymsubseteq}\ N\ {\isasymand}\ N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\ Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\ N\ {\isasymTurnstile}\ ZFC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}CH{\isasymcdot}{\isacharbraceright}{\kern0pt}\ {\isasymand}\isanewline
+\ \ \ \ \ \ {\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}
+\end{isabelle}
+
 %%% Local Variables: 

### 2022-08-14 Independence_CH/main_defs.tex
Defs Main as an appendix

\section{Main definitions of the formalization}\label{sec:definitions_main}

This appendix, which appears almost verbatim as
the theory \theory{Definitions\_Main} in \cite{Independence_CH-AFP},
might be considered as the bare minimum reading requisite to
trust that our development indeed formalizes the theory of
forcing.

The reader trusting
all the libraries on which our development is based, might jump
directly to Section~\ref{sec:relative-arith}, which treats relative
cardinal arithmetic as implemented in
\isatt{T{\kern0pt}r{\kern0pt}a{\kern0pt}n{\kern0pt}s{\kern0pt}i{\kern0pt}t{\kern0pt}i{\kern0pt}v{\kern0pt}e{\kern0pt}{\char`\_}{\kern0pt}M{\kern0pt}o{\kern0pt}d{\kern0pt}e{\kern0pt}l{\kern0pt}s{\kern0pt}}. But in case one wants to dive deeper, the
following sections treat some basic concepts of the ZF logic
(Section~\ref{sec:def-main-ZF}) and in the
ZF-Constructible library (Section~\ref{sec:def-main-relative})
on which our definitions are built.

\subsection{ZF\label{sec:def-main-ZF}%
}
For the basic logic ZF we restrict ourselves to just a few
concepts.
%
\begin{isabelle}%
bij{\isacharparenleft}{\kern0pt}A{\isacharcomma}{\kern0pt}\ B{\isacharparenright}{\kern0pt}\ {\isasymequiv}\isanewline
{\isacharbraceleft}{\kern0pt}f\ {\isasymin}\ A\ {\isasymrightarrow}\ B\ {\isachardot}{\kern0pt}\ {\isasymforall}w{\isasymin}A{\isachardot}{\kern0pt}\ {\isasymforall}x{\isasymin}A{\isachardot}{\kern0pt}\ f\ {\isacharbackquote}{\kern0pt}\ w\ {\isacharequal}{\kern0pt}\ f\ {\isacharbackquote}{\kern0pt}\ x\ {\isasymlongrightarrow}\ w\ {\isacharequal}{\kern0pt}\ x{\isacharbraceright}{\kern0pt}\ {\isasyminter}\isanewline
{\isacharbraceleft}{\kern0pt}f\ {\isasymin}\ A\ {\isasymrightarrow}\ B\ {\isachardot}{\kern0pt}\ {\isasymforall}y{\isasymin}B{\isachardot}{\kern0pt}\ {\isasymexists}x{\isasymin}A{\isachardot}{\kern0pt}\ f\ {\isacharbackquote}{\kern0pt}\ x\ {\isacharequal}{\kern0pt}\ y{\isacharbraceright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
A\ {\isasymapprox}\ B\ {\isasymequiv}\ {\isasymexists}f{\isachardot}{\kern0pt}\ f\ {\isasymin}\ bij{\isacharparenleft}{\kern0pt}A{\isacharcomma}{\kern0pt}\ B{\isacharparenright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
Transset{\isacharparenleft}{\kern0pt}i{\isacharparenright}{\kern0pt}\ {\isasymequiv}\ {\isasymforall}x{\isasymin}i{\isachardot}{\kern0pt}\ x\ {\isasymsubseteq}\ i%
\end{isabelle}%
\begin{isabelle}%
Ord{\isacharparenleft}{\kern0pt}i{\isacharparenright}{\kern0pt}\ {\isasymequiv}\ Transset{\isacharparenleft}{\kern0pt}i{\isacharparenright}{\kern0pt}\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymforall}x{\isasymin}i{\isachardot}{\kern0pt}\ Transset{\isacharparenleft}{\kern0pt}x{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
i\ {\isacharless}{\kern0pt}\ j\ {\isasymequiv}\ i\ {\isasymin}\ j\ {\isasymand}\ Ord{\isacharparenleft}{\kern0pt}j{\isacharparenright}{\kern0pt}\isasep\isanewline%
i\ {\isasymle}\ j\ {\isasymlongleftrightarrow}\ i\ {\isacharless}{\kern0pt}\ j\ {\isasymor}\ i\ {\isacharequal}{\kern0pt}\ j\ {\isasymand}\ Ord{\isacharparenleft}{\kern0pt}j{\isacharparenright}{\kern0pt}%
\end{isabelle}%
With the concepts of empty set and successor in place,%

\begin{isabelle}
\isacommand{lemma}
\ empty{\uscore}{\kern0pt}def{\isacharprime}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}{\isasymforall}x{\isachardot}{\kern0pt}\ x\ {\isasymnotin}\ {\isadigit{0}}{\isachardoublequoteclose}%
\isanewline
\isacommand{lemma}
\ succ{\uscore}{\kern0pt}def{\isacharprime}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}succ{\isacharparenleft}{\kern0pt}i{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ i\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}i{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}%
\end{isabelle}
%
we can define the set of natural numbers \isa{{\isasymomega}}. In the
sources, it is  defined as a fixpoint, but here we just write
its characterization as the first limit ordinal.%
\begin{isabelle}%
Ord{\isacharparenleft}{\kern0pt}{\isasymomega}{\isacharparenright}{\kern0pt}\ {\isasymand}\ {\isadigit{0}}\ {\isacharless}{\kern0pt}\ {\isasymomega}\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymforall}y{\isachardot}{\kern0pt}\ y\ {\isacharless}{\kern0pt}\ {\isasymomega}\ {\isasymlongrightarrow}\ succ{\isacharparenleft}{\kern0pt}y{\isacharparenright}{\kern0pt}\ {\isacharless}{\kern0pt}\ {\isasymomega}{\isacharparenright}{\kern0pt}\isasep\isanewline%
Ord{\isacharparenleft}{\kern0pt}i{\isacharparenright}{\kern0pt}\ {\isasymand}\ {\isadigit{0}}\ {\isacharless}{\kern0pt}\ i\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymforall}y{\isachardot}{\kern0pt}\ y\ {\isacharless}{\kern0pt}\ i\ {\isasymlongrightarrow}\ succ{\isacharparenleft}{\kern0pt}y{\isacharparenright}{\kern0pt}\ {\isacharless}{\kern0pt}\ i{\isacharparenright}{\kern0pt}\ {\isasymLongrightarrow}\ {\isasymomega}\ {\isasymle}\ i%
\end{isabelle}%
Then, addition and predecessor on \isa{{\isasymomega}} are inductively
characterized as follows:%
\begin{isabelle}%
m\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ succ{\isacharparenleft}{\kern0pt}n{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ succ{\isacharparenleft}{\kern0pt}m\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ n{\isacharparenright}{\kern0pt}\isasep\isanewline%
m\ {\isasymin}\ {\isasymomega}\ {\isasymLongrightarrow}\ m\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ {\isadigit{0}}\ {\isacharequal}{\kern0pt}\ m\isasep\isanewline%
pred{\isacharparenleft}{\kern0pt}{\isadigit{0}}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ {\isadigit{0}}\isasep\isanewline%
pred{\isacharparenleft}{\kern0pt}succ{\isacharparenleft}{\kern0pt}y{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ y%
\end{isabelle}%
Lists on a set \isa{A} can be characterized by being
recursively generated from the empty list \isa{{\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}} and the
operation \isa{Cons} that adds a new element to the left end;
the induction theorem for them shows that the characterization is
“complete”.%
\begin{isabelle}%
{\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}A{\isacharparenright}{\kern0pt}\isasep\isanewline%
{\isasymlbrakk}a\ {\isasymin}\ A{\isacharsemicolon}{\kern0pt}\ l\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}A{\isacharparenright}{\kern0pt}{\isasymrbrakk}\ {\isasymLongrightarrow}\ Cons{\isacharparenleft}{\kern0pt}a{\isacharcomma}{\kern0pt}\ l{\isacharparenright}{\kern0pt}\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}A{\isacharparenright}{\kern0pt}\isasep\isanewline%
{\isasymlbrakk}x\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}A{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ P{\isacharparenleft}{\kern0pt}{\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ {\isasymAnd}a\ l{\isachardot}{\kern0pt}\ {\isasymlbrakk}a\ {\isasymin}\ A{\isacharsemicolon}{\kern0pt}\ l\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}A{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ P{\isacharparenleft}{\kern0pt}l{\isacharparenright}{\kern0pt}{\isasymrbrakk}\ {\isasymLongrightarrow}\ P{\isacharparenleft}{\kern0pt}Cons{\isacharparenleft}{\kern0pt}a{\isacharcomma}{\kern0pt}\ l{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isasymrbrakk}\isanewline
{\isasymLongrightarrow}\ P{\isacharparenleft}{\kern0pt}x{\isacharparenright}{\kern0pt}%
\end{isabelle}%
Length, concatenation, and \isa{n}th element of lists are
recursively characterized as follows.%
\begin{isabelle}%
length{\isacharparenleft}{\kern0pt}{\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ {\isadigit{0}}\isasep\isanewline%
length{\isacharparenleft}{\kern0pt}Cons{\isacharparenleft}{\kern0pt}a{\isacharcomma}{\kern0pt}\ l{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ succ{\isacharparenleft}{\kern0pt}length{\isacharparenleft}{\kern0pt}l{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\isasep\isanewline%
{\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isacharat}{\kern0pt}\ ys\ {\isacharequal}{\kern0pt}\ ys\isasep\isanewline%
Cons{\isacharparenleft}{\kern0pt}a{\isacharcomma}{\kern0pt}\ l{\isacharparenright}{\kern0pt}\ {\isacharat}{\kern0pt}\ ys\ {\isacharequal}{\kern0pt}\ Cons{\isacharparenleft}{\kern0pt}a{\isacharcomma}{\kern0pt}\ l\ {\isacharat}{\kern0pt}\ ys{\isacharparenright}{\kern0pt}\isasep\isanewline%
nth{\isacharparenleft}{\kern0pt}{\isadigit{0}}{\isacharcomma}{\kern0pt}\ Cons{\isacharparenleft}{\kern0pt}a{\isacharcomma}{\kern0pt}\ l{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ a\isasep\isanewline%
n\ {\isasymin}\ {\isasymomega}\ {\isasymLongrightarrow}\ nth{\isacharparenleft}{\kern0pt}succ{\isacharparenleft}{\kern0pt}n{\isacharparenright}{\kern0pt}{\isacharcomma}{\kern0pt}\ Cons{\isacharparenleft}{\kern0pt}a{\isacharcomma}{\kern0pt}\ l{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ nth{\isacharparenleft}{\kern0pt}n{\isacharcomma}{\kern0pt}\ l{\isacharparenright}{\kern0pt}%
\end{isabelle}%
We have the usual Haskell-like notation for iterated applications
of \isa{Cons}:%
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ Cons{\isacharunderscore}{\kern0pt}app{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}{\isacharbrackleft}{\kern0pt}a{\isacharcomma}{\kern0pt}b{\isacharcomma}{\kern0pt}c{\isacharbrackright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ Cons{\isacharparenleft}{\kern0pt}a{\isacharcomma}{\kern0pt}Cons{\isacharparenleft}{\kern0pt}b{\isacharcomma}{\kern0pt}Cons{\isacharparenleft}{\kern0pt}c{\isacharcomma}{\kern0pt}{\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}%
\end{isabelle}

%
%
Relative quantifiers restrict the range of the bound variable to a
class \isa{M} of type \isa{i\ {\isasymRightarrow}\ o}; that is, a truth-valued function with
set arguments.%
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ {\isachardoublequoteopen}{\isasymforall}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ P{\isacharparenleft}{\kern0pt}x{\isacharparenright}{\kern0pt}\ {\isasymequiv}\ {\isasymforall}x{\isachardot}{\kern0pt}\ M{\isacharparenleft}{\kern0pt}x{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ P{\isacharparenleft}{\kern0pt}x{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\isanewline
\ \ \ \ \ \ {\isachardoublequoteopen}{\isasymexists}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ P{\isacharparenleft}{\kern0pt}x{\isacharparenright}{\kern0pt}\ {\isasymequiv}\ {\isasymexists}x{\isachardot}{\kern0pt}\ M{\isacharparenleft}{\kern0pt}x{\isacharparenright}{\kern0pt}\ {\isasymand}\ P{\isacharparenleft}{\kern0pt}x{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}
\end{isabelle}
%
%
Finally, a set can be viewed (“cast”) as a class using the
following function of type \isa{i\ {\isasymRightarrow}\ i\ {\isasymRightarrow}\ o}.%
\begin{isabelle}%
{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}A{\isacharparenright}{\kern0pt}{\isacharparenleft}{\kern0pt}x{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\ x\ {\isasymin}\ A%
\end{isabelle}%
\subsection{Relative concepts\label{sec:def-main-relative}%
}
A list of relative concepts (mostly from the \session{ZF-Constructible}
    library) follows next.%
\begin{isabelle}%
big{\isacharunderscore}{\kern0pt}union{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ A{\isacharcomma}{\kern0pt}\ z{\isacharparenright}{\kern0pt}\ {\isasymequiv}\ {\isasymforall}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ x\ {\isasymin}\ z\ {\isasymlongleftrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymexists}y{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ y\ {\isasymin}\ A\ {\isasymand}\ x\ {\isasymin}\ y{\isacharparenright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
upair{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ a{\isacharcomma}{\kern0pt}\ b{\isacharcomma}{\kern0pt}\ z{\isacharparenright}{\kern0pt}\ {\isasymequiv}\ a\ {\isasymin}\ z\ {\isasymand}\ b\ {\isasymin}\ z\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymforall}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ x\ {\isasymin}\ z\ {\isasymlongrightarrow}\ x\ {\isacharequal}{\kern0pt}\ a\ {\isasymor}\ x\ {\isacharequal}{\kern0pt}\ b{\isacharparenright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
pair{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ a{\isacharcomma}{\kern0pt}\ b{\isacharcomma}{\kern0pt}\ z{\isacharparenright}{\kern0pt}\ {\isasymequiv}\isanewline
{\isasymexists}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ upair{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ a{\isacharcomma}{\kern0pt}\ a{\isacharcomma}{\kern0pt}\ x{\isacharparenright}{\kern0pt}\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymexists}y{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ upair{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ a{\isacharcomma}{\kern0pt}\ b{\isacharcomma}{\kern0pt}\ y{\isacharparenright}{\kern0pt}\ {\isasymand}\ upair{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ x{\isacharcomma}{\kern0pt}\ y{\isacharcomma}{\kern0pt}\ z{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
successor{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ a{\isacharcomma}{\kern0pt}\ z{\isacharparenright}{\kern0pt}\ {\isasymequiv}\isanewline
{\isasymexists}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ upair{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ a{\isacharcomma}{\kern0pt}\ a{\isacharcomma}{\kern0pt}\ x{\isacharparenright}{\kern0pt}\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymforall}xa{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ xa\ {\isasymin}\ z\ {\isasymlongleftrightarrow}\ xa\ {\isasymin}\ x\ {\isasymor}\ xa\ {\isasymin}\ a{\isacharparenright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
empty{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ z{\isacharparenright}{\kern0pt}\ {\isasymequiv}\ {\isasymforall}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ x\ {\isasymnotin}\ z%
\end{isabelle}%
\begin{isabelle}%
transitive{\isacharunderscore}{\kern0pt}set{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ a{\isacharparenright}{\kern0pt}\ {\isasymequiv}\ {\isasymforall}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ x\ {\isasymin}\ a\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymforall}xa{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ xa\ {\isasymin}\ x\ {\isasymlongrightarrow}\ xa\ {\isasymin}\ a{\isacharparenright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
ordinal{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ a{\isacharparenright}{\kern0pt}\ {\isasymequiv}\isanewline
transitive{\isacharunderscore}{\kern0pt}set{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ a{\isacharparenright}{\kern0pt}\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymforall}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ x\ {\isasymin}\ a\ {\isasymlongrightarrow}\ transitive{\isacharunderscore}{\kern0pt}set{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ x{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
image{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ r{\isacharcomma}{\kern0pt}\ A{\isacharcomma}{\kern0pt}\ z{\isacharparenright}{\kern0pt}\ {\isasymequiv}\isanewline
{\isasymforall}y{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ y\ {\isasymin}\ z\ {\isasymlongleftrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymexists}w{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ w\ {\isasymin}\ r\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymexists}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ x\ {\isasymin}\ A\ {\isasymand}\ pair{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ x{\isacharcomma}{\kern0pt}\ y{\isacharcomma}{\kern0pt}\ w{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
is{\isacharunderscore}{\kern0pt}apply{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ f{\isacharcomma}{\kern0pt}\ x{\isacharcomma}{\kern0pt}\ y{\isacharparenright}{\kern0pt}\ {\isasymequiv}\isanewline
{\isasymexists}xs{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\isanewline
\isaindent{\ \ \ }{\isasymexists}fxs{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ upair{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ x{\isacharcomma}{\kern0pt}\ x{\isacharcomma}{\kern0pt}\ xs{\isacharparenright}{\kern0pt}\ {\isasymand}\ image{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ f{\isacharcomma}{\kern0pt}\ xs{\isacharcomma}{\kern0pt}\ fxs{\isacharparenright}{\kern0pt}\ {\isasymand}\ big{\isacharunderscore}{\kern0pt}union{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ fxs{\isacharcomma}{\kern0pt}\ y{\isacharparenright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
is{\isacharunderscore}{\kern0pt}function{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ r{\isacharparenright}{\kern0pt}\ {\isasymequiv}\isanewline
{\isasymforall}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\isanewline
\isaindent{\ \ \ }{\isasymforall}y{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\isanewline
\isaindent{\ \ \ \ \ \ }{\isasymforall}y{\isacharprime}{\kern0pt}{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\isanewline
\isaindent{\ \ \ \ \ \ \ \ \ }{\isasymforall}p{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\isanewline
\isaindent{\ \ \ \ \ \ \ \ \ \ \ \ }{\isasymforall}p{\isacharprime}{\kern0pt}{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\isanewline
\isaindent{\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ }pair{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ x{\isacharcomma}{\kern0pt}\ y{\isacharcomma}{\kern0pt}\ p{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\isanewline
\isaindent{\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ }pair{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ x{\isacharcomma}{\kern0pt}\ y{\isacharprime}{\kern0pt}{\isacharcomma}{\kern0pt}\ p{\isacharprime}{\kern0pt}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ p\ {\isasymin}\ r\ {\isasymlongrightarrow}\ p{\isacharprime}{\kern0pt}\ {\isasymin}\ r\ {\isasymlongrightarrow}\ y\ {\isacharequal}{\kern0pt}\ y{\isacharprime}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
is{\isacharunderscore}{\kern0pt}relation{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ r{\isacharparenright}{\kern0pt}\ {\isasymequiv}\ {\isasymforall}z{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ z\ {\isasymin}\ r\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymexists}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymexists}y{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ pair{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ x{\isacharcomma}{\kern0pt}\ y{\isacharcomma}{\kern0pt}\ z{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
is{\isacharunderscore}{\kern0pt}domain{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ r{\isacharcomma}{\kern0pt}\ z{\isacharparenright}{\kern0pt}\ {\isasymequiv}\isanewline
{\isasymforall}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ x\ {\isasymin}\ z\ {\isasymlongleftrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymexists}w{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ w\ {\isasymin}\ r\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymexists}y{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ pair{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ x{\isacharcomma}{\kern0pt}\ y{\isacharcomma}{\kern0pt}\ w{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
typed{\isacharunderscore}{\kern0pt}function{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ A{\isacharcomma}{\kern0pt}\ B{\isacharcomma}{\kern0pt}\ r{\isacharparenright}{\kern0pt}\ {\isasymequiv}\isanewline
is{\isacharunderscore}{\kern0pt}function{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ r{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
is{\isacharunderscore}{\kern0pt}relation{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ r{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
is{\isacharunderscore}{\kern0pt}domain{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ r{\isacharcomma}{\kern0pt}\ A{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
{\isacharparenleft}{\kern0pt}{\isasymforall}u{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ u\ {\isasymin}\ r\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymforall}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymforall}y{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ pair{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ x{\isacharcomma}{\kern0pt}\ y{\isacharcomma}{\kern0pt}\ u{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ y\ {\isasymin}\ B{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
is{\isacharunderscore}{\kern0pt}function{\isacharunderscore}{\kern0pt}space{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ A{\isacharcomma}{\kern0pt}\ B{\isacharcomma}{\kern0pt}\ fs{\isacharparenright}{\kern0pt}\ {\isasymequiv}\isanewline
M{\isacharparenleft}{\kern0pt}fs{\isacharparenright}{\kern0pt}\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymforall}f{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ f\ {\isasymin}\ fs\ {\isasymlongleftrightarrow}\ typed{\isacharunderscore}{\kern0pt}function{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ A{\isacharcomma}{\kern0pt}\ B{\isacharcomma}{\kern0pt}\ f{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\isasep\isanewline%
A\ {\isasymrightarrow}\isactrlbsup M\isactrlesup \ B\ {\isasymequiv}\ THE\ d{\isachardot}{\kern0pt}\ is{\isacharunderscore}{\kern0pt}function{\isacharunderscore}{\kern0pt}space{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ A{\isacharcomma}{\kern0pt}\ B{\isacharcomma}{\kern0pt}\ d{\isacharparenright}{\kern0pt}\isasep\isanewline%
surjection{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ A{\isacharcomma}{\kern0pt}\ B{\isacharcomma}{\kern0pt}\ f{\isacharparenright}{\kern0pt}\ {\isasymequiv}\isanewline
typed{\isacharunderscore}{\kern0pt}function{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ A{\isacharcomma}{\kern0pt}\ B{\isacharcomma}{\kern0pt}\ f{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
{\isacharparenleft}{\kern0pt}{\isasymforall}y{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ y\ {\isasymin}\ B\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymexists}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ x\ {\isasymin}\ A\ {\isasymand}\ is{\isacharunderscore}{\kern0pt}apply{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ f{\isacharcomma}{\kern0pt}\ x{\isacharcomma}{\kern0pt}\ y{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}%
\end{isabelle}%

\subsubsection*{Relative version of the $\ZFC$ axioms}
\begin{isabelle}%
extensionality{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}\ {\isasymequiv}\ {\isasymforall}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymforall}y{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isasymforall}z{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ z\ {\isasymin}\ x\ {\isasymlongleftrightarrow}\ z\ {\isasymin}\ y{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ x\ {\isacharequal}{\kern0pt}\ y%
\end{isabelle}%
\begin{isabelle}%
foundation{\isacharunderscore}{\kern0pt}ax{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}\ {\isasymequiv}\isanewline
{\isasymforall}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isasymexists}y{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ y\ {\isasymin}\ x{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymexists}y{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ y\ {\isasymin}\ x\ {\isasymand}\ {\isasymnot}\ {\isacharparenleft}{\kern0pt}{\isasymexists}z{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ z\ {\isasymin}\ x\ {\isasymand}\ z\ {\isasymin}\ y{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
upair{\isacharunderscore}{\kern0pt}ax{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}\ {\isasymequiv}\ {\isasymforall}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymforall}y{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymexists}z{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ upair{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ x{\isacharcomma}{\kern0pt}\ y{\isacharcomma}{\kern0pt}\ z{\isacharparenright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
Union{\isacharunderscore}{\kern0pt}ax{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}\ {\isasymequiv}\ {\isasymforall}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymexists}z{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ big{\isacharunderscore}{\kern0pt}union{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ x{\isacharcomma}{\kern0pt}\ z{\isacharparenright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
power{\isacharunderscore}{\kern0pt}ax{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}\ {\isasymequiv}\ {\isasymforall}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymexists}z{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymforall}xa{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ xa\ {\isasymin}\ z\ {\isasymlongleftrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymforall}xb{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ xb\ {\isasymin}\ xa\ {\isasymlongrightarrow}\ xb\ {\isasymin}\ x{\isacharparenright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
infinity{\isacharunderscore}{\kern0pt}ax{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}\ {\isasymequiv}\isanewline
{\isasymexists}I{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\isanewline
\isaindent{\ \ \ }{\isacharparenleft}{\kern0pt}{\isasymexists}z{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ empty{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ z{\isacharparenright}{\kern0pt}\ {\isasymand}\ z\ {\isasymin}\ I{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
\isaindent{\ \ \ }{\isacharparenleft}{\kern0pt}{\isasymforall}y{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ y\ {\isasymin}\ I\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymexists}sy{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ successor{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ y{\isacharcomma}{\kern0pt}\ sy{\isacharparenright}{\kern0pt}\ {\isasymand}\ sy\ {\isasymin}\ I{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
choice{\isacharunderscore}{\kern0pt}ax{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}\ {\isasymequiv}\ {\isasymforall}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymexists}a{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymexists}f{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ ordinal{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ a{\isacharparenright}{\kern0pt}\ {\isasymand}\ surjection{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ a{\isacharcomma}{\kern0pt}\ x{\isacharcomma}{\kern0pt}\ f{\isacharparenright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
separation{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ P{\isacharparenright}{\kern0pt}\ {\isasymequiv}\ {\isasymforall}z{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymexists}y{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymforall}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ x\ {\isasymin}\ y\ {\isasymlongleftrightarrow}\ x\ {\isasymin}\ z\ {\isasymand}\ P{\isacharparenleft}{\kern0pt}x{\isacharparenright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
univalent{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ A{\isacharcomma}{\kern0pt}\ P{\isacharparenright}{\kern0pt}\ {\isasymequiv}\isanewline
{\isasymforall}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ x\ {\isasymin}\ A\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymforall}y{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymforall}z{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ P{\isacharparenleft}{\kern0pt}x{\isacharcomma}{\kern0pt}\ y{\isacharparenright}{\kern0pt}\ {\isasymand}\ P{\isacharparenleft}{\kern0pt}x{\isacharcomma}{\kern0pt}\ z{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ y\ {\isacharequal}{\kern0pt}\ z{\isacharparenright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
strong{\isacharunderscore}{\kern0pt}replacement{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ P{\isacharparenright}{\kern0pt}\ {\isasymequiv}\isanewline
{\isasymforall}A{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\isanewline
\isaindent{\ \ \ }univalent{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ A{\isacharcomma}{\kern0pt}\ P{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymexists}Y{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymforall}b{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ b\ {\isasymin}\ Y\ {\isasymlongleftrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymexists}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ x\ {\isasymin}\ A\ {\isasymand}\ P{\isacharparenleft}{\kern0pt}x{\isacharcomma}{\kern0pt}\ b{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}%
\end{isabelle}%
\subsubsection*{Internalized formulas}
“Codes” for formulas (as sets) are constructed from natural
numbers using \isa{Member}, \isa{Equal}, \isa{Nand},
and \isa{Forall}.%
\begin{isabelle}%
{\isasymlbrakk}x\ {\isasymin}\ {\isasymomega}{\isacharsemicolon}{\kern0pt}\ y\ {\isasymin}\ {\isasymomega}{\isasymrbrakk}\ {\isasymLongrightarrow}\ {\isasymcdot}x\ {\isasymin}\ y{\isasymcdot}\ {\isasymin}\ formula\isasep\isanewline%
{\isasymlbrakk}x\ {\isasymin}\ {\isasymomega}{\isacharsemicolon}{\kern0pt}\ y\ {\isasymin}\ {\isasymomega}{\isasymrbrakk}\ {\isasymLongrightarrow}\ {\isasymcdot}x\ {\isacharequal}{\kern0pt}\ y{\isasymcdot}\ {\isasymin}\ formula\isasep\isanewline%
{\isasymlbrakk}p\ {\isasymin}\ formula{\isacharsemicolon}{\kern0pt}\ q\ {\isasymin}\ formula{\isasymrbrakk}\ {\isasymLongrightarrow}\ {\isasymcdot}{\isasymnot}{\isacharparenleft}{\kern0pt}p\ {\isasymand}\ q{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isasymin}\ formula\isasep\isanewline%
p\ {\isasymin}\ formula\ {\isasymLongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymcdot}{\isasymforall}p{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymin}\ formula\isasep\isanewline\isanewline%
{\isasymlbrakk}x\ {\isasymin}\ formula{\isacharsemicolon}{\kern0pt}\ {\isasymAnd}x\ y{\isachardot}{\kern0pt}\ {\isasymlbrakk}x\ {\isasymin}\ {\isasymomega}{\isacharsemicolon}{\kern0pt}\ y\ {\isasymin}\ {\isasymomega}{\isasymrbrakk}\ {\isasymLongrightarrow}\ P{\isacharparenleft}{\kern0pt}{\isasymcdot}x\ {\isasymin}\ y{\isasymcdot}{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\isanewline
\isaindent{\ }{\isasymAnd}x\ y{\isachardot}{\kern0pt}\ {\isasymlbrakk}x\ {\isasymin}\ {\isasymomega}{\isacharsemicolon}{\kern0pt}\ y\ {\isasymin}\ {\isasymomega}{\isasymrbrakk}\ {\isasymLongrightarrow}\ P{\isacharparenleft}{\kern0pt}{\isasymcdot}x\ {\isacharequal}{\kern0pt}\ y{\isasymcdot}{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\isanewline
\isaindent{\ }{\isasymAnd}p\ q{\isachardot}{\kern0pt}\ {\isasymlbrakk}p\ {\isasymin}\ formula{\isacharsemicolon}{\kern0pt}\ P{\isacharparenleft}{\kern0pt}p{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ q\ {\isasymin}\ formula{\isacharsemicolon}{\kern0pt}\ P{\isacharparenleft}{\kern0pt}q{\isacharparenright}{\kern0pt}{\isasymrbrakk}\ {\isasymLongrightarrow}\ P{\isacharparenleft}{\kern0pt}{\isasymcdot}{\isasymnot}{\isacharparenleft}{\kern0pt}p\ {\isasymand}\ q{\isacharparenright}{\kern0pt}{\isasymcdot}{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\isanewline
\isaindent{\ }{\isasymAnd}p{\isachardot}{\kern0pt}\ {\isasymlbrakk}p\ {\isasymin}\ formula{\isacharsemicolon}{\kern0pt}\ P{\isacharparenleft}{\kern0pt}p{\isacharparenright}{\kern0pt}{\isasymrbrakk}\ {\isasymLongrightarrow}\ P{\isacharparenleft}{\kern0pt}{\isacharparenleft}{\kern0pt}{\isasymcdot}{\isasymforall}p{\isasymcdot}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isasymrbrakk}\isanewline
{\isasymLongrightarrow}\ P{\isacharparenleft}{\kern0pt}x{\isacharparenright}{\kern0pt}%
\end{isabelle}%
Definitions for the other connectives and the internal existential
quantifier are also provided. For instance, negation:%
\begin{isabelle}%
{\isasymcdot}{\isasymnot}p{\isasymcdot}\ {\isasymequiv}\ {\isasymcdot}{\isasymnot}{\isacharparenleft}{\kern0pt}p\ {\isasymand}\ p{\isacharparenright}{\kern0pt}{\isasymcdot}%
\end{isabelle}%
\begin{isabelle}%
arity{\isacharparenleft}{\kern0pt}{\isasymcdot}x\ {\isasymin}\ y{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ succ{\isacharparenleft}{\kern0pt}x{\isacharparenright}{\kern0pt}\ {\isasymunion}\ succ{\isacharparenleft}{\kern0pt}y{\isacharparenright}{\kern0pt}\isasep\isanewline%
arity{\isacharparenleft}{\kern0pt}{\isasymcdot}x\ {\isacharequal}{\kern0pt}\ y{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ succ{\isacharparenleft}{\kern0pt}x{\isacharparenright}{\kern0pt}\ {\isasymunion}\ succ{\isacharparenleft}{\kern0pt}y{\isacharparenright}{\kern0pt}\isasep\isanewline%
arity{\isacharparenleft}{\kern0pt}{\isasymcdot}{\isasymnot}{\isacharparenleft}{\kern0pt}p\ {\isasymand}\ q{\isacharparenright}{\kern0pt}{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ arity{\isacharparenleft}{\kern0pt}p{\isacharparenright}{\kern0pt}\ {\isasymunion}\ arity{\isacharparenleft}{\kern0pt}q{\isacharparenright}{\kern0pt}\isasep\isanewline%
arity{\isacharparenleft}{\kern0pt}{\isacharparenleft}{\kern0pt}{\isasymcdot}{\isasymforall}p{\isasymcdot}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ pred{\isacharparenleft}{\kern0pt}arity{\isacharparenleft}{\kern0pt}p{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}%
\end{isabelle}%
We have the satisfaction relation between $\in$-models and
    first order formulas (given a “environment” list representing
    the assignment of free variables),%
\begin{isabelle}%
{\isasymlbrakk}nth{\isacharparenleft}{\kern0pt}i{\isacharcomma}{\kern0pt}\ env{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ x{\isacharsemicolon}{\kern0pt}\ nth{\isacharparenleft}{\kern0pt}j{\isacharcomma}{\kern0pt}\ env{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ y{\isacharsemicolon}{\kern0pt}\ env\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}A{\isacharparenright}{\kern0pt}{\isasymrbrakk}\isanewline
{\isasymLongrightarrow}\ x\ {\isasymin}\ y\ {\isasymlongleftrightarrow}\ A{\isacharcomma}{\kern0pt}\ env\ {\isasymTurnstile}\ {\isasymcdot}i\ {\isasymin}\ j{\isasymcdot}\isasep\isanewline\isanewline%
{\isasymlbrakk}nth{\isacharparenleft}{\kern0pt}i{\isacharcomma}{\kern0pt}\ env{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ x{\isacharsemicolon}{\kern0pt}\ nth{\isacharparenleft}{\kern0pt}j{\isacharcomma}{\kern0pt}\ env{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ y{\isacharsemicolon}{\kern0pt}\ env\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}A{\isacharparenright}{\kern0pt}{\isasymrbrakk}\isanewline
{\isasymLongrightarrow}\ x\ {\isacharequal}{\kern0pt}\ y\ {\isasymlongleftrightarrow}\ A{\isacharcomma}{\kern0pt}\ env\ {\isasymTurnstile}\ {\isasymcdot}i\ {\isacharequal}{\kern0pt}\ j{\isasymcdot}\isasep\isanewline\isanewline%
env\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}A{\isacharparenright}{\kern0pt}\ {\isasymLongrightarrow}\ {\isacharparenleft}{\kern0pt}A{\isacharcomma}{\kern0pt}\ env\ {\isasymTurnstile}\ {\isasymcdot}{\isasymnot}{\isacharparenleft}{\kern0pt}p\ {\isasymand}\ q{\isacharparenright}{\kern0pt}{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\ {\isasymnot}\ {\isacharparenleft}{\kern0pt}{\isacharparenleft}{\kern0pt}A{\isacharcomma}{\kern0pt}\ env\ {\isasymTurnstile}\ p{\isacharparenright}{\kern0pt}\ {\isasymand}\ {\isacharparenleft}{\kern0pt}A{\isacharcomma}{\kern0pt}\ env\ {\isasymTurnstile}\ q{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\isasep\isanewline\isanewline%
env\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}A{\isacharparenright}{\kern0pt}\ {\isasymLongrightarrow}\ {\isacharparenleft}{\kern0pt}A{\isacharcomma}{\kern0pt}\ env\ {\isasymTurnstile}\ {\isacharparenleft}{\kern0pt}{\isasymcdot}{\isasymforall}p{\isasymcdot}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymforall}x{\isasymin}A{\isachardot}{\kern0pt}\ A{\isacharcomma}{\kern0pt}\ Cons{\isacharparenleft}{\kern0pt}x{\isacharcomma}{\kern0pt}\ env{\isacharparenright}{\kern0pt}\ {\isasymTurnstile}\ p{\isacharparenright}{\kern0pt}%
\end{isabelle}%
as well as the satisfaction of an arbitrary set of sentences.%
\begin{isabelle}%
A\ {\isasymTurnstile}\ {\isasymPhi}\ {\isasymequiv}\ {\isasymforall}{\isasymphi}{\isasymin}{\isasymPhi}{\isachardot}{\kern0pt}\ A{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymphi}%
\end{isabelle}%
The internalized (viz. as elements of the set \isa{formula})
    version of the axioms follow next.%
\begin{isabelle}%
Union{\isacharunderscore}{\kern0pt}ax{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}A{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\ A{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}Union\ Ax{\isasymcdot}\isasep\isanewline%
power{\isacharunderscore}{\kern0pt}ax{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}A{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\ A{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}Powerset\ Ax{\isasymcdot}\isasep\isanewline%
upair{\isacharunderscore}{\kern0pt}ax{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}A{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\ A{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}Pairing{\isasymcdot}\isasep\isanewline%
foundation{\isacharunderscore}{\kern0pt}ax{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}A{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\ A{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}Foundation{\isasymcdot}\isasep\isanewline%
extensionality{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}A{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\ A{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}Extensionality{\isasymcdot}\isasep\isanewline%
infinity{\isacharunderscore}{\kern0pt}ax{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}A{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\ A{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}Infinity{\isasymcdot}\isasep\isanewline\isanewline%
{\isasymphi}\ {\isasymin}\ formula\ {\isasymLongrightarrow}\isanewline
{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}Separation{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\isanewline
{\isacharparenleft}{\kern0pt}{\isasymforall}env{\isasymin}list{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isachardot}{\kern0pt}\isanewline
\isaindent{{\isacharparenleft}{\kern0pt}\ \ \ }arity{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isadigit{1}}\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ length{\isacharparenleft}{\kern0pt}env{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ separation{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isasymlambda}x{\isachardot}{\kern0pt}\ M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}x{\isacharbrackright}{\kern0pt}\ {\isacharat}{\kern0pt}\ env\ {\isasymTurnstile}\ {\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\isasep\isanewline\isanewline%
{\isasymphi}\ {\isasymin}\ formula\ {\isasymLongrightarrow}\isanewline
{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\isanewline
{\isacharparenleft}{\kern0pt}{\isasymforall}env{\isasymin}list{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isachardot}{\kern0pt}\isanewline
\isaindent{{\isacharparenleft}{\kern0pt}\ \ \ }arity{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isadigit{2}}\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ length{\isacharparenleft}{\kern0pt}env{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\isanewline
\isaindent{{\isacharparenleft}{\kern0pt}\ \ \ }strong{\isacharunderscore}{\kern0pt}replacement{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isasymlambda}x\ y{\isachardot}{\kern0pt}\ M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}x{\isacharcomma}{\kern0pt}\ y{\isacharbrackright}{\kern0pt}\ {\isacharat}{\kern0pt}\ env\ {\isasymTurnstile}\ {\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\isasep\isanewline\isanewline%
choice{\isacharunderscore}{\kern0pt}ax{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}A{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\ A{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}AC{\isasymcdot}%
\end{isabelle}%
\begin{isabelle}%
ZF{\isacharunderscore}{\kern0pt}fin\ {\isasymequiv}\isanewline
{\isacharbraceleft}{\kern0pt}{\isasymcdot}Extensionality{\isasymcdot}{\isacharcomma}{\kern0pt}\ {\isasymcdot}Foundation{\isasymcdot}{\isacharcomma}{\kern0pt}\ {\isasymcdot}Pairing{\isasymcdot}{\isacharcomma}{\kern0pt}\ {\isasymcdot}Union\ Ax{\isasymcdot}{\isacharcomma}{\kern0pt}\ {\isasymcdot}Infinity{\isasymcdot}{\isacharcomma}{\kern0pt}\isanewline
\isaindent{{\isacharbraceleft}{\kern0pt}}{\isasymcdot}Powerset\ Ax{\isasymcdot}{\isacharbraceright}{\kern0pt}\isasep\isanewline\isanewline%
ZF{\isacharunderscore}{\kern0pt}schemes\ {\isasymequiv}\isanewline
{\isacharbraceleft}{\kern0pt}{\isasymcdot}Separation{\isacharparenleft}{\kern0pt}p{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ p\ {\isasymin}\ formula{\isacharbraceright}{\kern0pt}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}p{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ p\ {\isasymin}\ formula{\isacharbraceright}{\kern0pt}\isasep\isanewline\isanewline%
{\isasymcdot}Z{\isasymcdot}\ {\isasymequiv}\ ZF{\isacharunderscore}{\kern0pt}fin\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Separation{\isacharparenleft}{\kern0pt}p{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ p\ {\isasymin}\ formula{\isacharbraceright}{\kern0pt}\isasep\isanewline%
ZC\ {\isasymequiv}\ {\isasymcdot}Z{\isasymcdot}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}AC{\isasymcdot}{\isacharbraceright}{\kern0pt}\isasep\isanewline%
ZF\ {\isasymequiv}\ ZF{\isacharunderscore}{\kern0pt}schemes\ {\isasymunion}\ ZF{\isacharunderscore}{\kern0pt}fin\isasep\isanewline%
ZFC\ {\isasymequiv}\ ZF\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}AC{\isasymcdot}{\isacharbraceright}{\kern0pt}%
\end{isabelle}%

\subsection{Relativization of infinitary arithmetic\label{sec:relative-arith}%
}
In order to state the defining property of the relative
equipotence relation, we work under the assumptions of the
locale \isa{M{\isacharunderscore}{\kern0pt}cardinals}. They comprise a finite set
of instances of Separation and Replacement to prove
closure properties of the transitive class \isa{M}.%
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ {\isacharparenleft}{\kern0pt}\isakeyword{in}\ M{\isacharunderscore}{\kern0pt}cardinals{\isacharparenright}{\kern0pt}\ eqpoll{\isacharunderscore}{\kern0pt}def{\isacharprime}{\kern0pt}{\isacharcolon}{\kern0pt}\isanewline
\ \ \isakeyword{assumes}\ {\isachardoublequoteopen}M{\isacharparenleft}{\kern0pt}A{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\ {\isachardoublequoteopen}M{\isacharparenleft}{\kern0pt}B{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\ \isakeyword{shows}\ {\isachardoublequoteopen}A\ {\isasymapprox}\isactrlbsup M\isactrlesup \ B\ {\isasymlongleftrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymexists}f{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ f\ {\isasymin}\ bij{\isacharparenleft}{\kern0pt}A{\isacharcomma}{\kern0pt}B{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}
\end{isabelle}

%
Below, $\mu$ denotes the minimum operator on the ordinals.%
\begin{isabelle}
  \isacommand{lemma}\isamarkupfalse%
\ cardinalities{\isacharunderscore}{\kern0pt}defs{\isacharcolon}{\kern0pt}\isanewline
\ \ \isakeyword{fixes}\ M{\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}{\isachardoublequoteopen}i{\isasymRightarrow}o{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isacharbar}{\kern0pt}A{\isacharbar}{\kern0pt}\isactrlbsup M\isactrlesup \ {\isasymequiv}\ {\isasymmu}\ i{\isachardot}{\kern0pt}\ M{\isacharparenleft}{\kern0pt}i{\isacharparenright}{\kern0pt}\ {\isasymand}\ i\ {\isasymapprox}\isactrlbsup M\isactrlesup \ A{\isachardoublequoteclose}\isanewline
\ \ \ \ {\isachardoublequoteopen}Card\isactrlbsup M\isactrlesup {\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymequiv}\ {\isasymalpha}\ {\isacharequal}{\kern0pt}\ {\isacharbar}{\kern0pt}{\isasymalpha}{\isacharbar}{\kern0pt}\isactrlbsup M\isactrlesup {\isachardoublequoteclose}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isasymkappa}\isactrlbsup {\isasymup}{\isasymnu}{\isacharcomma}{\kern0pt}M\isactrlesup \ {\isasymequiv}\ {\isacharbar}{\kern0pt}{\isasymnu}\ {\isasymrightarrow}\isactrlbsup M\isactrlesup \ {\isasymkappa}{\isacharbar}{\kern0pt}\isactrlbsup M\isactrlesup {\isachardoublequoteclose}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isacharparenleft}{\kern0pt}{\isasymkappa}\isactrlsup {\isacharplus}{\kern0pt}{\isacharparenright}{\kern0pt}\isactrlbsup M\isactrlesup \ {\isasymequiv}\ {\isasymmu}\ x{\isachardot}{\kern0pt}\ M{\isacharparenleft}{\kern0pt}x{\isacharparenright}{\kern0pt}\ {\isasymand}\ Card\isactrlbsup M\isactrlesup {\isacharparenleft}{\kern0pt}x{\isacharparenright}{\kern0pt}\ {\isasymand}\ {\isasymkappa}\ {\isacharless}{\kern0pt}\ x{\isachardoublequoteclose}
\end{isabelle}
Analogous to the previous Lemma
\isa{eqpoll{\isacharunderscore}{\kern0pt}def{\isacharprime}{\kern0pt}},
the next lemma holds under
the assumptions of the locale \isa{M{\isacharunderscore}{\kern0pt}aleph}. The axiom instances
included are sufficient to state and prove the defining
properties of the relativized \isa{Aleph} function
(in particular, the required ability to perform transfinite recursions).%
\begin{isabelle}%
\isacommand{context}\isamarkupfalse%
\ M{\isacharunderscore}{\kern0pt}aleph\isanewline
\isakeyword{begin}%
\isanewline
\isanewline
{\isasymaleph}\isactrlbsub {\isadigit{0}}\isactrlesub \isactrlbsup M\isactrlesup \ {\isacharequal}{\kern0pt}\ {\isasymomega}\isasep\isanewline%
{\isasymlbrakk}Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ M{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}{\isasymrbrakk}\ {\isasymLongrightarrow}\ {\isasymaleph}\isactrlbsub succ{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\isactrlesub \isactrlbsup M\isactrlesup \ {\isacharequal}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isasymaleph}\isactrlbsub {\isasymalpha}\isactrlesub \isactrlbsup M\isactrlesup \isactrlsup {\isacharplus}{\kern0pt}{\isacharparenright}{\kern0pt}\isactrlbsup M\isactrlesup \isasep\isanewline%
{\isasymlbrakk}Limit{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ M{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}{\isasymrbrakk}\ {\isasymLongrightarrow}\ {\isasymaleph}\isactrlbsub {\isasymalpha}\isactrlesub \isactrlbsup M\isactrlesup \ {\isacharequal}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isasymUnion}j{\isasymin}{\isasymalpha}{\isachardot}{\kern0pt}\ {\isasymaleph}\isactrlbsub j\isactrlesub \isactrlbsup M\isactrlesup {\isacharparenright}{\kern0pt}%
\end{isabelle}%
\isacommand{end}\isamarkupfalse%
\ %
\isamarkupcmt{\isa{M{\isacharunderscore}{\kern0pt}aleph}%
}
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ ContHyp{\isacharunderscore}{\kern0pt}rel{\isacharunderscore}{\kern0pt}def{\isacharprime}{\kern0pt}{\isacharcolon}{\kern0pt}\isanewline
\ \ \isakeyword{fixes}\ N{\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}{\isachardoublequoteopen}i{\isasymRightarrow}o{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ \ \ {\isachardoublequoteopen}CH\isactrlbsup N\isactrlesup \ {\isasymequiv}\ {\isasymaleph}\isactrlbsub {\isadigit{1}}\isactrlesub \isactrlbsup N\isactrlesup \ {\isacharequal}{\kern0pt}\ {\isadigit{2}}\isactrlbsup {\isasymup}{\isasymaleph}\isactrlbsub {\isadigit{0}}\isactrlesub \isactrlbsup N\isactrlesup {\isacharcomma}{\kern0pt}N\isactrlesup {\isachardoublequoteclose}
\end{isabelle}

%
%
Under appropriate hypotheses (this time, from the locale \isa{M{\isacharunderscore}{\kern0pt}ZF{\isacharunderscore}{\kern0pt}library}),
   \isa{CH\isactrlbsup M\isactrlesup } is equivalent to its fully relational version \isa{is{\isacharunderscore}{\kern0pt}ContHyp}.
    As a sanity check, we see that if the transitive class is indeed \isa{{\isasymV}},
    we recover the original $\CH$.%
\begin{isabelle}%
M{\isacharunderscore}{\kern0pt}ZF{\isacharunderscore}{\kern0pt}library{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}\ {\isasymLongrightarrow}\ is{\isacharunderscore}{\kern0pt}ContHyp{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\ CH\isactrlbsup M\isactrlesup \isasep\isanewline%
is{\isacharunderscore}{\kern0pt}ContHyp{\isacharparenleft}{\kern0pt}{\isasymV}{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\ {\isasymaleph}\isactrlbsub {\isadigit{1}}\isactrlesub \ {\isacharequal}{\kern0pt}\ {\isadigit{2}}\isactrlbsup {\isasymup}{\isasymaleph}\isactrlbsub {\isadigit{0}}\isactrlesub \isactrlesup %
\end{isabelle}%
In turn, the fully relational version evaluated on a nonempty
    transitive \isa{A} is equivalent to the satisfaction of the
    first-order formula \isa{{\isasymcdot}CH{\isasymcdot}}.%
\begin{isabelle}%
{\isasymlbrakk}env\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}A{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ {\isadigit{0}}\ {\isasymin}\ A{\isasymrbrakk}\ {\isasymLongrightarrow}\ is{\isacharunderscore}{\kern0pt}ContHyp{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}A{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\ A{\isacharcomma}{\kern0pt}\ env\ {\isasymTurnstile}\ {\isasymcdot}CH{\isasymcdot}%
\end{isabelle}%
\subsection{Forcing \label{sec:def-main-forcing}%
}
Our first milestone was to obtain a proper extension using forcing.
Its original proof didn't required the previous developments involving
below.%
\begin{isabelle}%
{\isasymlbrakk}M\ {\isasymapprox}\ {\isasymomega}{\isacharsemicolon}{\kern0pt}\ Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ M\ {\isasymTurnstile}\ ZF{\isasymrbrakk}\isanewline
{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ M\ {\isasymsubseteq}\ N\ {\isasymand}\isanewline
\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\isanewline
\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }N\ {\isasymTurnstile}\ ZF\ {\isasymand}\isanewline
\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }M\ {\isasymnoteq}\ N\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}AC{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ N\ {\isasymTurnstile}\ ZFC{\isacharparenright}{\kern0pt}%
\end{isabelle}%
We can finally state our main results, namely, the existence of models
for $\ZFC + \CH$ and $\ZFC + \neg\CH$ under the assumption of a ctm of $\ZFC$.%
\begin{isabelle}%
{\isasymlbrakk}M\ {\isasymapprox}\ {\isasymomega}{\isacharsemicolon}{\kern0pt}\ Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ M\ {\isasymTurnstile}\ ZFC{\isasymrbrakk}\isanewline
{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ M\ {\isasymsubseteq}\ N\ {\isasymand}\isanewline
\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\isanewline
\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\ N\ {\isasymTurnstile}\ ZFC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}{\isasymnot}{\isasymcdot}CH{\isasymcdot}{\isasymcdot}{\isacharbraceright}{\kern0pt}\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
{\isasymlbrakk}M\ {\isasymapprox}\ {\isasymomega}{\isacharsemicolon}{\kern0pt}\ Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ M\ {\isasymTurnstile}\ ZFC{\isasymrbrakk}\isanewline
{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ M\ {\isasymsubseteq}\ N\ {\isasymand}\isanewline
\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\isanewline
\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\ N\ {\isasymTurnstile}\ ZFC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}CH{\isasymcdot}{\isacharbraceright}{\kern0pt}\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}%
\end{isabelle}%
These results can be strengthened by enumerating six finite sets of
replacement instances which are sufficient to develop forcing and for
the construction of the aforementioned models: \isa{instances{\isadigit{1}}{\isacharunderscore}{\kern0pt}fms}
through \isa{instances{\isadigit{4}}{\isacharunderscore}{\kern0pt}fms}, \isa{instances{\isacharunderscore}{\kern0pt}ground{\isacharunderscore}{\kern0pt}fms}, and
\isa{instances{\isacharunderscore}{\kern0pt}ground{\isacharunderscore}{\kern0pt}notCH{\isacharunderscore}{\kern0pt}fms},
which are then collected into the $31$-element set \isa{overhead{\isacharunderscore}{\kern0pt}notCH}.
For example, we have:%
\begin{isabelle}%
instances{\isadigit{1}}{\isacharunderscore}{\kern0pt}fms\ {\isasymequiv}\isanewline
{\isacharbraceleft}{\kern0pt}list{\isacharunderscore}{\kern0pt}repl{\isadigit{1}}{\isacharunderscore}{\kern0pt}intf{\isacharunderscore}{\kern0pt}fm{\isacharcomma}{\kern0pt}\ list{\isacharunderscore}{\kern0pt}repl{\isadigit{2}}{\isacharunderscore}{\kern0pt}intf{\isacharunderscore}{\kern0pt}fm{\isacharcomma}{\kern0pt}\ formula{\isacharunderscore}{\kern0pt}repl{\isadigit{1}}{\isacharunderscore}{\kern0pt}intf{\isacharunderscore}{\kern0pt}fm{\isacharcomma}{\kern0pt}\isanewline
\isaindent{{\isacharbraceleft}{\kern0pt}}formula{\isacharunderscore}{\kern0pt}repl{\isadigit{2}}{\isacharunderscore}{\kern0pt}intf{\isacharunderscore}{\kern0pt}fm{\isacharcomma}{\kern0pt}\ eclose{\isacharunderscore}{\kern0pt}repl{\isadigit{1}}{\isacharunderscore}{\kern0pt}intf{\isacharunderscore}{\kern0pt}fm{\isacharcomma}{\kern0pt}\ eclose{\isacharunderscore}{\kern0pt}repl{\isadigit{2}}{\isacharunderscore}{\kern0pt}intf{\isacharunderscore}{\kern0pt}fm{\isacharcomma}{\kern0pt}\isanewline
\isaindent{{\isacharbraceleft}{\kern0pt}}wfrec{\isacharunderscore}{\kern0pt}rank{\isacharunderscore}{\kern0pt}fm{\isacharcomma}{\kern0pt}\ trans{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}HVFrom{\isacharunderscore}{\kern0pt}fm{\isacharcomma}{\kern0pt}\ tl{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}intf{\isacharunderscore}{\kern0pt}fm{\isacharbraceright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
overhead\ {\isasymequiv}\ instances{\isadigit{1}}{\isacharunderscore}{\kern0pt}fms\ {\isasymunion}\ instances{\isadigit{2}}{\isacharunderscore}{\kern0pt}fms\ {\isasymunion}\ instances{\isacharunderscore}{\kern0pt}ground{\isacharunderscore}{\kern0pt}fms\isasep\isanewline\isanewline%
overhead{\isacharunderscore}{\kern0pt}notCH\ {\isasymequiv}\isanewline
overhead\ {\isasymunion}\ instances{\isadigit{3}}{\isacharunderscore}{\kern0pt}fms\ {\isasymunion}\ instances{\isadigit{4}}{\isacharunderscore}{\kern0pt}fms\ {\isasymunion}\ instances{\isacharunderscore}{\kern0pt}ground{\isacharunderscore}{\kern0pt}notCH{\isacharunderscore}{\kern0pt}fms\isasep\isanewline\isanewline%
overhead{\isacharunderscore}{\kern0pt}CH\ {\isasymequiv}\ overhead{\isacharunderscore}{\kern0pt}notCH\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}replacement{\isacharunderscore}{\kern0pt}dcwit{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}body{\isacharunderscore}{\kern0pt}fm{\isacharbraceright}{\kern0pt}%
\end{isabelle}%
One further instance is needed to force $\CH$, with a total count
of $32$ instances:%
\begin{isabelle}%
overhead{\isacharunderscore}{\kern0pt}CH\ {\isasymequiv}\ overhead{\isacharunderscore}{\kern0pt}notCH\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}replacement{\isacharunderscore}{\kern0pt}dcwit{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}body{\isacharunderscore}{\kern0pt}fm{\isacharbraceright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
{\isasymlbrakk}M\ {\isasymapprox}\ {\isasymomega}{\isacharsemicolon}{\kern0pt}\ Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ M\ {\isasymTurnstile}\ {\isasymcdot}Z{\isasymcdot}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}p{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ p\ {\isasymin}\ overhead{\isacharbraceright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\isanewline
\isaindent{\ }{\isasymPhi}\ {\isasymsubseteq}\ formula{\isacharsemicolon}{\kern0pt}\ M\ {\isasymTurnstile}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}ground{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}fm{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}{\isasymrbrakk}\isanewline
{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ M\ {\isasymsubseteq}\ N\ {\isasymand}\isanewline
\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\isanewline
\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }M\ {\isasymnoteq}\ N\ {\isasymand}\isanewline
\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }{\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }{\isacharparenleft}{\kern0pt}{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}AC{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ N{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}AC{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }N\ {\isasymTurnstile}\ {\isasymcdot}Z{\isasymcdot}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
{\isasymlbrakk}M\ {\isasymapprox}\ {\isasymomega}{\isacharsemicolon}{\kern0pt}\ Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ M\ {\isasymTurnstile}\ ZC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}p{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ p\ {\isasymin}\ overhead{\isacharunderscore}{\kern0pt}notCH{\isacharbraceright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\isanewline
\isaindent{\ }{\isasymPhi}\ {\isasymsubseteq}\ formula{\isacharsemicolon}{\kern0pt}\ M\ {\isasymTurnstile}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}ground{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}fm{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}{\isasymrbrakk}\isanewline
{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ M\ {\isasymsubseteq}\ N\ {\isasymand}\isanewline
\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\isanewline
\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }N\ {\isasymTurnstile}\ ZC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}{\isasymnot}{\isasymcdot}CH{\isasymcdot}{\isasymcdot}{\isacharbraceright}{\kern0pt}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}\ {\isasymand}\isanewline
\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }{\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}%
\end{isabelle}%
\begin{isabelle}%
{\isasymlbrakk}M\ {\isasymapprox}\ {\isasymomega}{\isacharsemicolon}{\kern0pt}\ Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ M\ {\isasymTurnstile}\ ZC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}p{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ p\ {\isasymin}\ overhead{\isacharunderscore}{\kern0pt}CH{\isacharbraceright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\isanewline
\isaindent{\ }{\isasymPhi}\ {\isasymsubseteq}\ formula{\isacharsemicolon}{\kern0pt}\ M\ {\isasymTurnstile}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}ground{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}fm{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}{\isasymrbrakk}\isanewline
{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ M\ {\isasymsubseteq}\ N\ {\isasymand}\isanewline
\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\isanewline
\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }N\ {\isasymTurnstile}\ ZC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}CH{\isasymcdot}{\isacharbraceright}{\kern0pt}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}\ {\isasymand}\isanewline
\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }{\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}%
\end{isabelle}%
In the above three statements, the function \isa{ground{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}fm}
takes an element \isa{{\isasymphi}} of \isa{formula} and returns the
replacement instance in the ground model that produces the
\isa{{\isasymphi}}-replacement instance in the generic extension. The next
result is stated in the context \isa{G{\isacharunderscore}{\kern0pt}generic{\isadigit{1}}}, which assumes
the existence of a generic filter.
%
\begin{isabelle}%
\isacommand{context}\isamarkupfalse%
\ G{\isacharunderscore}{\kern0pt}generic{\isadigit{1}}\isanewline
\isakeyword{begin}\isanewline
\isanewline
{\isasymlbrakk}{\isasymphi}\ {\isasymin}\ formula{\isacharsemicolon}{\kern0pt}\ M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}ground{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}fm{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isasymcdot}{\isasymrbrakk}\isanewline
{\isasymLongrightarrow}\ M{\isacharbrackleft}{\kern0pt}G{\isacharbrackright}{\kern0pt}{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}%
\end{isabelle}%
\isacommand{end}\isamarkupfalse%
\ %
\isamarkupcmt{\isa{G{\isacharunderscore}{\kern0pt}generic{\isadigit{1}}}%
}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "independence_ch_isabelle"
%%% ispell-local-dictionary: "american"
%%% End:

### 2022-08-14 Independence_CH/main_defs.tex
Revision of Main Defs on paper

@@
 
-This appendix, which appears almost verbatim as
+This section, which appears almost verbatim as
 the theory \theory{Definitions\_Main} in \cite{Independence_CH-AFP},
@@
 (Section~\ref{sec:def-main-ZF}) and in the
-ZF-Constructible library (Section~\ref{sec:def-main-relative})
+\session{ZF-Constructible} library (Section~\ref{sec:def-main-relative})
 on which our definitions are built.
@@
 m\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ succ{\isacharparenleft}{\kern0pt}n{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ succ{\isacharparenleft}{\kern0pt}m\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ n{\isacharparenright}{\kern0pt}\isasep\isanewline%
-m\ {\isasymin}\ {\isasymomega}\ {\isasymLongrightarrow}\ m\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ {\isadigit{0}}\ {\isacharequal}{\kern0pt}\ m\isasep\isanewline%
+m\ {\isasymin}\ {\isasymomega}\ {\isasymLongrightarrow}\ m\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ {\isadigit{0}}\ {\isacharequal}{\kern0pt}\ m\isasep\isanewline\isanewline%
 pred{\isacharparenleft}{\kern0pt}{\isadigit{0}}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ {\isadigit{0}}\isasep\isanewline%
@@
 {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}A{\isacharparenright}{\kern0pt}\isasep\isanewline%
-{\isasymlbrakk}a\ {\isasymin}\ A{\isacharsemicolon}{\kern0pt}\ l\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}A{\isacharparenright}{\kern0pt}{\isasymrbrakk}\ {\isasymLongrightarrow}\ Cons{\isacharparenleft}{\kern0pt}a{\isacharcomma}{\kern0pt}\ l{\isacharparenright}{\kern0pt}\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}A{\isacharparenright}{\kern0pt}\isasep\isanewline%
-{\isasymlbrakk}x\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}A{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ P{\isacharparenleft}{\kern0pt}{\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ {\isasymAnd}a\ l{\isachardot}{\kern0pt}\ {\isasymlbrakk}a\ {\isasymin}\ A{\isacharsemicolon}{\kern0pt}\ l\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}A{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ P{\isacharparenleft}{\kern0pt}l{\isacharparenright}{\kern0pt}{\isasymrbrakk}\ {\isasymLongrightarrow}\ P{\isacharparenleft}{\kern0pt}Cons{\isacharparenleft}{\kern0pt}a{\isacharcomma}{\kern0pt}\ l{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isasymrbrakk}\isanewline
+{\isasymlbrakk}a\ {\isasymin}\ A{\isacharsemicolon}{\kern0pt}\ l\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}A{\isacharparenright}{\kern0pt}{\isasymrbrakk}\ {\isasymLongrightarrow}\ Cons{\isacharparenleft}{\kern0pt}a{\isacharcomma}{\kern0pt}\ l{\isacharparenright}{\kern0pt}\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}A{\isacharparenright}{\kern0pt}\isasep\isanewline\isanewline%
+{\isasymlbrakk}x\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}A{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ P{\isacharparenleft}{\kern0pt}{\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ {\isasymAnd}a\ l{\isachardot}{\kern0pt}\ {\isasymlbrakk}a\ {\isasymin}\ A{\isacharsemicolon}{\kern0pt}\ l\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}A{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ P{\isacharparenleft}{\kern0pt}l{\isacharparenright}{\kern0pt}{\isasymrbrakk}\ {\isasymLongrightarrow}\isanewline
+\ \ P{\isacharparenleft}{\kern0pt}Cons{\isacharparenleft}{\kern0pt}a{\isacharcomma}{\kern0pt}\ l{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isasymrbrakk}
 {\isasymLongrightarrow}\ P{\isacharparenleft}{\kern0pt}x{\isacharparenright}{\kern0pt}%
@@
 length{\isacharparenleft}{\kern0pt}{\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ {\isadigit{0}}\isasep\isanewline%
-length{\isacharparenleft}{\kern0pt}Cons{\isacharparenleft}{\kern0pt}a{\isacharcomma}{\kern0pt}\ l{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ succ{\isacharparenleft}{\kern0pt}length{\isacharparenleft}{\kern0pt}l{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\isasep\isanewline%
+length{\isacharparenleft}{\kern0pt}Cons{\isacharparenleft}{\kern0pt}a{\isacharcomma}{\kern0pt}\ l{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ succ{\isacharparenleft}{\kern0pt}length{\isacharparenleft}{\kern0pt}l{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\isasep\isanewline\isanewline%
 {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isacharat}{\kern0pt}\ ys\ {\isacharequal}{\kern0pt}\ ys\isasep\isanewline%
-Cons{\isacharparenleft}{\kern0pt}a{\isacharcomma}{\kern0pt}\ l{\isacharparenright}{\kern0pt}\ {\isacharat}{\kern0pt}\ ys\ {\isacharequal}{\kern0pt}\ Cons{\isacharparenleft}{\kern0pt}a{\isacharcomma}{\kern0pt}\ l\ {\isacharat}{\kern0pt}\ ys{\isacharparenright}{\kern0pt}\isasep\isanewline%
+Cons{\isacharparenleft}{\kern0pt}a{\isacharcomma}{\kern0pt}\ l{\isacharparenright}{\kern0pt}\ {\isacharat}{\kern0pt}\ ys\ {\isacharequal}{\kern0pt}\ Cons{\isacharparenleft}{\kern0pt}a{\isacharcomma}{\kern0pt}\ l\ {\isacharat}{\kern0pt}\ ys{\isacharparenright}{\kern0pt}\isasep\isanewline\isanewline%
 nth{\isacharparenleft}{\kern0pt}{\isadigit{0}}{\isacharcomma}{\kern0pt}\ Cons{\isacharparenleft}{\kern0pt}a{\isacharcomma}{\kern0pt}\ l{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ a\isasep\isanewline%
@@
 {\isasymexists}xs{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\isanewline
-\isaindent{\ \ \ }{\isasymexists}fxs{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ upair{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ x{\isacharcomma}{\kern0pt}\ x{\isacharcomma}{\kern0pt}\ xs{\isacharparenright}{\kern0pt}\ {\isasymand}\ image{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ f{\isacharcomma}{\kern0pt}\ xs{\isacharcomma}{\kern0pt}\ fxs{\isacharparenright}{\kern0pt}\ {\isasymand}\ big{\isacharunderscore}{\kern0pt}union{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ fxs{\isacharcomma}{\kern0pt}\ y{\isacharparenright}{\kern0pt}%
+\isaindent{\ \ \ }{\isasymexists}fxs{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ upair{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ x{\isacharcomma}{\kern0pt}\ x{\isacharcomma}{\kern0pt}\ xs{\isacharparenright}{\kern0pt}\ {\isasymand}\ image{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ f{\isacharcomma}{\kern0pt}\ xs{\isacharcomma}{\kern0pt}\ fxs{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
+\isaindent{\ \ \ \ \ }big{\isacharunderscore}{\kern0pt}union{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ fxs{\isacharcomma}{\kern0pt}\ y{\isacharparenright}{\kern0pt}%
 \end{isabelle}%
@@
 is{\isacharunderscore}{\kern0pt}function{\isacharunderscore}{\kern0pt}space{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ A{\isacharcomma}{\kern0pt}\ B{\isacharcomma}{\kern0pt}\ fs{\isacharparenright}{\kern0pt}\ {\isasymequiv}\isanewline
-M{\isacharparenleft}{\kern0pt}fs{\isacharparenright}{\kern0pt}\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymforall}f{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ f\ {\isasymin}\ fs\ {\isasymlongleftrightarrow}\ typed{\isacharunderscore}{\kern0pt}function{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ A{\isacharcomma}{\kern0pt}\ B{\isacharcomma}{\kern0pt}\ f{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\isasep\isanewline%
-A\ {\isasymrightarrow}\isactrlbsup M\isactrlesup \ B\ {\isasymequiv}\ THE\ d{\isachardot}{\kern0pt}\ is{\isacharunderscore}{\kern0pt}function{\isacharunderscore}{\kern0pt}space{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ A{\isacharcomma}{\kern0pt}\ B{\isacharcomma}{\kern0pt}\ d{\isacharparenright}{\kern0pt}\isasep\isanewline%
+M{\isacharparenleft}{\kern0pt}fs{\isacharparenright}{\kern0pt}\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymforall}f{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ f\ {\isasymin}\ fs\ {\isasymlongleftrightarrow}\ typed{\isacharunderscore}{\kern0pt}function{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ A{\isacharcomma}{\kern0pt}\ B{\isacharcomma}{\kern0pt}\ f{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\isasep\isanewline\isanewline%
+A\ {\isasymrightarrow}\isactrlbsup M\isactrlesup \ B\ {\isasymequiv}\ THE\ d{\isachardot}{\kern0pt}\ is{\isacharunderscore}{\kern0pt}function{\isacharunderscore}{\kern0pt}space{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ A{\isacharcomma}{\kern0pt}\ B{\isacharcomma}{\kern0pt}\ d{\isacharparenright}{\kern0pt}\isasep\isanewline\isanewline%
 surjection{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ A{\isacharcomma}{\kern0pt}\ B{\isacharcomma}{\kern0pt}\ f{\isacharparenright}{\kern0pt}\ {\isasymequiv}\isanewline
@@
 \end{isabelle}%
+The \isatt{arity} function strictly bounding the free de Bruijn
+indices of a formula is defined below:
 \begin{isabelle}%
@@
 {\isasymLongrightarrow}\ x\ {\isacharequal}{\kern0pt}\ y\ {\isasymlongleftrightarrow}\ A{\isacharcomma}{\kern0pt}\ env\ {\isasymTurnstile}\ {\isasymcdot}i\ {\isacharequal}{\kern0pt}\ j{\isasymcdot}\isasep\isanewline\isanewline%
-env\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}A{\isacharparenright}{\kern0pt}\ {\isasymLongrightarrow}\ {\isacharparenleft}{\kern0pt}A{\isacharcomma}{\kern0pt}\ env\ {\isasymTurnstile}\ {\isasymcdot}{\isasymnot}{\isacharparenleft}{\kern0pt}p\ {\isasymand}\ q{\isacharparenright}{\kern0pt}{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\ {\isasymnot}\ {\isacharparenleft}{\kern0pt}{\isacharparenleft}{\kern0pt}A{\isacharcomma}{\kern0pt}\ env\ {\isasymTurnstile}\ p{\isacharparenright}{\kern0pt}\ {\isasymand}\ {\isacharparenleft}{\kern0pt}A{\isacharcomma}{\kern0pt}\ env\ {\isasymTurnstile}\ q{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\isasep\isanewline\isanewline%
+env\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}A{\isacharparenright}{\kern0pt}\ {\isasymLongrightarrow}\ {\isacharparenleft}{\kern0pt}A{\isacharcomma}{\kern0pt}\ env\ {\isasymTurnstile}\ {\isasymcdot}{\isasymnot}{\isacharparenleft}{\kern0pt}p\ {\isasymand}\ q{\isacharparenright}{\kern0pt}{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\ {\isasymnot}\ {\isacharparenleft}{\kern0pt}{\isacharparenleft}{\kern0pt}A{\isacharcomma}{\kern0pt}\ env\ {\isasymTurnstile}\ p{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline%
+\ \ {\isacharparenleft}{\kern0pt}A{\isacharcomma}{\kern0pt}\ env\ {\isasymTurnstile}\ q{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\isasep\isanewline\isanewline%
 env\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}A{\isacharparenright}{\kern0pt}\ {\isasymLongrightarrow}\ {\isacharparenleft}{\kern0pt}A{\isacharcomma}{\kern0pt}\ env\ {\isasymTurnstile}\ {\isacharparenleft}{\kern0pt}{\isasymcdot}{\isasymforall}p{\isasymcdot}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymforall}x{\isasymin}A{\isachardot}{\kern0pt}\ A{\isacharcomma}{\kern0pt}\ Cons{\isacharparenleft}{\kern0pt}x{\isacharcomma}{\kern0pt}\ env{\isacharparenright}{\kern0pt}\ {\isasymTurnstile}\ p{\isacharparenright}{\kern0pt}%
@@
 The internalized (viz. as elements of the set \isa{formula})
-    version of the axioms follow next.%
+versions of the axioms are checked next against the relative statements.%
 \begin{isabelle}%
@@
 transitive \isa{A} is equivalent to the satisfaction of the
-    first-order formula \isa{{\isasymcdot}CH{\isasymcdot}}.%
+first-order formula \isa{{\isasymcdot}CH{\isasymcdot}} (since it
+actually is a sentence, it does not depend on \isatt{env}, which
+appears only because the definition of $\models$ requires that argument).%
 \begin{isabelle}%

### 2022-08-14 Independence_CH/main_defs.tex
Change in section ordering. Statement of the main theorems

@@
 \end{isabelle}%
-\subsection{Forcing \label{sec:def-main-forcing}%
-}
-Our first milestone was to obtain a proper extension using forcing.
-Its original proof didn't required the previous developments involving
-below.%
-\begin{isabelle}%
-{\isasymlbrakk}M\ {\isasymapprox}\ {\isasymomega}{\isacharsemicolon}{\kern0pt}\ Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ M\ {\isasymTurnstile}\ ZF{\isasymrbrakk}\isanewline
-{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ M\ {\isasymsubseteq}\ N\ {\isasymand}\isanewline
-\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\isanewline
-\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
-\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }N\ {\isasymTurnstile}\ ZF\ {\isasymand}\isanewline
-\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }M\ {\isasymnoteq}\ N\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}AC{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ N\ {\isasymTurnstile}\ ZFC{\isacharparenright}{\kern0pt}%
-\end{isabelle}%
-We can finally state our main results, namely, the existence of models
-for $\ZFC + \CH$ and $\ZFC + \neg\CH$ under the assumption of a ctm of $\ZFC$.%
-\begin{isabelle}%
-{\isasymlbrakk}M\ {\isasymapprox}\ {\isasymomega}{\isacharsemicolon}{\kern0pt}\ Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ M\ {\isasymTurnstile}\ ZFC{\isasymrbrakk}\isanewline
-{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ M\ {\isasymsubseteq}\ N\ {\isasymand}\isanewline
-\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\isanewline
-\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\ N\ {\isasymTurnstile}\ ZFC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}{\isasymnot}{\isasymcdot}CH{\isasymcdot}{\isasymcdot}{\isacharbraceright}{\kern0pt}\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}%
-\end{isabelle}%
-\begin{isabelle}%
-{\isasymlbrakk}M\ {\isasymapprox}\ {\isasymomega}{\isacharsemicolon}{\kern0pt}\ Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ M\ {\isasymTurnstile}\ ZFC{\isasymrbrakk}\isanewline
-{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ M\ {\isasymsubseteq}\ N\ {\isasymand}\isanewline
-\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\isanewline
-\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\ N\ {\isasymTurnstile}\ ZFC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}CH{\isasymcdot}{\isacharbraceright}{\kern0pt}\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}%
-\end{isabelle}%
-These results can be strengthened by enumerating six finite sets of
-replacement instances which are sufficient to develop forcing and for
-the construction of the aforementioned models: \isa{instances{\isadigit{1}}{\isacharunderscore}{\kern0pt}fms}
-through \isa{instances{\isadigit{4}}{\isacharunderscore}{\kern0pt}fms}, \isa{instances{\isacharunderscore}{\kern0pt}ground{\isacharunderscore}{\kern0pt}fms}, and
-\isa{instances{\isacharunderscore}{\kern0pt}ground{\isacharunderscore}{\kern0pt}notCH{\isacharunderscore}{\kern0pt}fms},
-which are then collected into the $31$-element set \isa{overhead{\isacharunderscore}{\kern0pt}notCH}.
-For example, we have:%
-\begin{isabelle}%
-instances{\isadigit{1}}{\isacharunderscore}{\kern0pt}fms\ {\isasymequiv}\isanewline
-{\isacharbraceleft}{\kern0pt}list{\isacharunderscore}{\kern0pt}repl{\isadigit{1}}{\isacharunderscore}{\kern0pt}intf{\isacharunderscore}{\kern0pt}fm{\isacharcomma}{\kern0pt}\ list{\isacharunderscore}{\kern0pt}repl{\isadigit{2}}{\isacharunderscore}{\kern0pt}intf{\isacharunderscore}{\kern0pt}fm{\isacharcomma}{\kern0pt}\ formula{\isacharunderscore}{\kern0pt}repl{\isadigit{1}}{\isacharunderscore}{\kern0pt}intf{\isacharunderscore}{\kern0pt}fm{\isacharcomma}{\kern0pt}\isanewline
-\isaindent{{\isacharbraceleft}{\kern0pt}}formula{\isacharunderscore}{\kern0pt}repl{\isadigit{2}}{\isacharunderscore}{\kern0pt}intf{\isacharunderscore}{\kern0pt}fm{\isacharcomma}{\kern0pt}\ eclose{\isacharunderscore}{\kern0pt}repl{\isadigit{1}}{\isacharunderscore}{\kern0pt}intf{\isacharunderscore}{\kern0pt}fm{\isacharcomma}{\kern0pt}\ eclose{\isacharunderscore}{\kern0pt}repl{\isadigit{2}}{\isacharunderscore}{\kern0pt}intf{\isacharunderscore}{\kern0pt}fm{\isacharcomma}{\kern0pt}\isanewline
-\isaindent{{\isacharbraceleft}{\kern0pt}}wfrec{\isacharunderscore}{\kern0pt}rank{\isacharunderscore}{\kern0pt}fm{\isacharcomma}{\kern0pt}\ trans{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}HVFrom{\isacharunderscore}{\kern0pt}fm{\isacharcomma}{\kern0pt}\ tl{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}intf{\isacharunderscore}{\kern0pt}fm{\isacharbraceright}{\kern0pt}%
-\end{isabelle}%
-\begin{isabelle}%
-overhead\ {\isasymequiv}\ instances{\isadigit{1}}{\isacharunderscore}{\kern0pt}fms\ {\isasymunion}\ instances{\isadigit{2}}{\isacharunderscore}{\kern0pt}fms\ {\isasymunion}\ instances{\isacharunderscore}{\kern0pt}ground{\isacharunderscore}{\kern0pt}fms\isasep\isanewline\isanewline%
-overhead{\isacharunderscore}{\kern0pt}notCH\ {\isasymequiv}\isanewline
-overhead\ {\isasymunion}\ instances{\isadigit{3}}{\isacharunderscore}{\kern0pt}fms\ {\isasymunion}\ instances{\isadigit{4}}{\isacharunderscore}{\kern0pt}fms\ {\isasymunion}\ instances{\isacharunderscore}{\kern0pt}ground{\isacharunderscore}{\kern0pt}notCH{\isacharunderscore}{\kern0pt}fms\isasep\isanewline\isanewline%
-overhead{\isacharunderscore}{\kern0pt}CH\ {\isasymequiv}\ overhead{\isacharunderscore}{\kern0pt}notCH\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}replacement{\isacharunderscore}{\kern0pt}dcwit{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}body{\isacharunderscore}{\kern0pt}fm{\isacharbraceright}{\kern0pt}%
-\end{isabelle}%
-One further instance is needed to force $\CH$, with a total count
-of $32$ instances:%
-\begin{isabelle}%
-overhead{\isacharunderscore}{\kern0pt}CH\ {\isasymequiv}\ overhead{\isacharunderscore}{\kern0pt}notCH\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}replacement{\isacharunderscore}{\kern0pt}dcwit{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}body{\isacharunderscore}{\kern0pt}fm{\isacharbraceright}{\kern0pt}%
-\end{isabelle}%
-\begin{isabelle}%
-{\isasymlbrakk}M\ {\isasymapprox}\ {\isasymomega}{\isacharsemicolon}{\kern0pt}\ Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ M\ {\isasymTurnstile}\ {\isasymcdot}Z{\isasymcdot}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}p{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ p\ {\isasymin}\ overhead{\isacharbraceright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\isanewline
-\isaindent{\ }{\isasymPhi}\ {\isasymsubseteq}\ formula{\isacharsemicolon}{\kern0pt}\ M\ {\isasymTurnstile}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}ground{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}fm{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}{\isasymrbrakk}\isanewline
-{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ M\ {\isasymsubseteq}\ N\ {\isasymand}\isanewline
-\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\isanewline
-\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
-\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }M\ {\isasymnoteq}\ N\ {\isasymand}\isanewline
-\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }{\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
-\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }{\isacharparenleft}{\kern0pt}{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}AC{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ N{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}AC{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
-\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }N\ {\isasymTurnstile}\ {\isasymcdot}Z{\isasymcdot}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}%
-\end{isabelle}%
-\begin{isabelle}%
-{\isasymlbrakk}M\ {\isasymapprox}\ {\isasymomega}{\isacharsemicolon}{\kern0pt}\ Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ M\ {\isasymTurnstile}\ ZC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}p{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ p\ {\isasymin}\ overhead{\isacharunderscore}{\kern0pt}notCH{\isacharbraceright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\isanewline
-\isaindent{\ }{\isasymPhi}\ {\isasymsubseteq}\ formula{\isacharsemicolon}{\kern0pt}\ M\ {\isasymTurnstile}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}ground{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}fm{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}{\isasymrbrakk}\isanewline
-{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ M\ {\isasymsubseteq}\ N\ {\isasymand}\isanewline
-\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\isanewline
-\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
-\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }N\ {\isasymTurnstile}\ ZC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}{\isasymnot}{\isasymcdot}CH{\isasymcdot}{\isasymcdot}{\isacharbraceright}{\kern0pt}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}\ {\isasymand}\isanewline
-\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }{\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}%
-\end{isabelle}%
-\begin{isabelle}%
-{\isasymlbrakk}M\ {\isasymapprox}\ {\isasymomega}{\isacharsemicolon}{\kern0pt}\ Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ M\ {\isasymTurnstile}\ ZC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}p{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ p\ {\isasymin}\ overhead{\isacharunderscore}{\kern0pt}CH{\isacharbraceright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\isanewline
-\isaindent{\ }{\isasymPhi}\ {\isasymsubseteq}\ formula{\isacharsemicolon}{\kern0pt}\ M\ {\isasymTurnstile}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}ground{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}fm{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}{\isasymrbrakk}\isanewline
-{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ M\ {\isasymsubseteq}\ N\ {\isasymand}\isanewline
-\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\isanewline
-\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
-\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }N\ {\isasymTurnstile}\ ZC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}CH{\isasymcdot}{\isacharbraceright}{\kern0pt}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}\ {\isasymand}\isanewline
-\isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }{\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}%
-\end{isabelle}%
-In the above three statements, the function \isa{ground{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}fm}
-takes an element \isa{{\isasymphi}} of \isa{formula} and returns the
-replacement instance in the ground model that produces the
-\isa{{\isasymphi}}-replacement instance in the generic extension. The next
-result is stated in the context \isa{G{\isacharunderscore}{\kern0pt}generic{\isadigit{1}}}, which assumes
-the existence of a generic filter.
-%
-\begin{isabelle}%
-\isacommand{context}\isamarkupfalse%
-\ G{\isacharunderscore}{\kern0pt}generic{\isadigit{1}}\isanewline
-\isakeyword{begin}\isanewline
-\isanewline
-{\isasymlbrakk}{\isasymphi}\ {\isasymin}\ formula{\isacharsemicolon}{\kern0pt}\ M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}ground{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}fm{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isasymcdot}{\isasymrbrakk}\isanewline
-{\isasymLongrightarrow}\ M{\isacharbrackleft}{\kern0pt}G{\isacharbrackright}{\kern0pt}{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}%
-\end{isabelle}%
-\isacommand{end}\isamarkupfalse%
-\ %
-\isamarkupcmt{\isa{G{\isacharunderscore}{\kern0pt}generic{\isadigit{1}}}%
-}
+%% \subsection{Forcing \label{sec:def-main-forcing}%
+%% }
+%% Our first milestone was to obtain a proper extension using forcing.
+%% Its original proof didn't required the previous developments involving
+%% below.%
+%% \begin{isabelle}%
+%% {\isasymlbrakk}M\ {\isasymapprox}\ {\isasymomega}{\isacharsemicolon}{\kern0pt}\ Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ M\ {\isasymTurnstile}\ ZF{\isasymrbrakk}\isanewline
+%% {\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ M\ {\isasymsubseteq}\ N\ {\isasymand}\isanewline
+%% \isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\isanewline
+%% \isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
+%% \isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }N\ {\isasymTurnstile}\ ZF\ {\isasymand}\isanewline
+%% \isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }M\ {\isasymnoteq}\ N\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}AC{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ N\ {\isasymTurnstile}\ ZFC{\isacharparenright}{\kern0pt}%
+%% \end{isabelle}%
+%% We can finally state our main results, namely, the existence of models
+%% for $\ZFC + \CH$ and $\ZFC + \neg\CH$ under the assumption of a ctm of $\ZFC$.%
+%% \begin{isabelle}%
+%% {\isasymlbrakk}M\ {\isasymapprox}\ {\isasymomega}{\isacharsemicolon}{\kern0pt}\ Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ M\ {\isasymTurnstile}\ ZFC{\isasymrbrakk}\isanewline
+%% {\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ M\ {\isasymsubseteq}\ N\ {\isasymand}\isanewline
+%% \isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\isanewline
+%% \isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\ N\ {\isasymTurnstile}\ ZFC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}{\isasymnot}{\isasymcdot}CH{\isasymcdot}{\isasymcdot}{\isacharbraceright}{\kern0pt}\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}%
+%% \end{isabelle}%
+%% \begin{isabelle}%
+%% {\isasymlbrakk}M\ {\isasymapprox}\ {\isasymomega}{\isacharsemicolon}{\kern0pt}\ Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\ M\ {\isasymTurnstile}\ ZFC{\isasymrbrakk}\isanewline
+%% {\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ M\ {\isasymsubseteq}\ N\ {\isasymand}\isanewline
+%% \isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\isanewline
+%% \isaindent{{\isasymLongrightarrow}\ {\isasymexists}N{\isachardot}{\kern0pt}\ }Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\ N\ {\isasymTurnstile}\ ZFC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}CH{\isasymcdot}{\isacharbraceright}{\kern0pt}\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}%
+%% \end{isabelle}%
+%% In the above three statements, the function \isa{ground{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}fm}
+%% takes an element \isa{{\isasymphi}} of \isa{formula} and returns the
+%% replacement instance in the ground model that produces the
+%% \isa{{\isasymphi}}-replacement instance in the generic extension. The next
+%% result is stated in the context \isa{G{\isacharunderscore}{\kern0pt}generic{\isadigit{1}}}, which assumes
+%% the existence of a generic filter.
+%% %
+%% \begin{isabelle}%
+%% \isacommand{context}\isamarkupfalse%
+%% \ G{\isacharunderscore}{\kern0pt}generic{\isadigit{1}}\isanewline
+%% \isakeyword{begin}\isanewline
+%% \isanewline
+%% {\isasymlbrakk}{\isasymphi}\ {\isasymin}\ formula{\isacharsemicolon}{\kern0pt}\ M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}ground{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}fm{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isasymcdot}{\isasymrbrakk}\isanewline
+%% {\isasymLongrightarrow}\ M{\isacharbrackleft}{\kern0pt}G{\isacharbrackright}{\kern0pt}{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}%
+%% \end{isabelle}%
+%% \isacommand{end}\isamarkupfalse%
+%% \ %
+%% \isamarkupcmt{\isa{G{\isacharunderscore}{\kern0pt}generic{\isadigit{1}}}%
+%% }
 

### 2022-08-14 Independence_CH/main_results.tex
Change in section ordering. Statement of the main theorems

\section{Main achievements of the formalization}
\label{sec:main-achievements}

\subsection{32 replacement instances to rule them all}
\label{sec:repl-instances}

We isolated 32 instances of Replacement that are sufficient to force
$\CH$ or $\neg\CH$, which are enumerated below by the name of the
corresponding internalized first order formula. Many of these were already present in
relational form in the \session{ZF-Constructible} library.

The first 13 instances are needed to set up
cardinal arithmetic in $M$, along some other basic constructions.

\begin{itemize}
\item 6 instances for iteration of operations (between parentheses). For each construct,
  Paulson used one replacement instance in order to have absoluteness,
  and a second one to obtain closure.
  \begin{itemize}
  \item
    \isatt{list{\uscore}repl1{\uscore}intf{\uscore}fm} and 
    \isatt{list{\uscore}repl2{\uscore}intf{\uscore}fm}.

    For lists on a set $A$ ($X\mapsto \{0 \} \oplus A \times X$, where
    $D\oplus E$ stands for the disjoint union/coproduct $(\{0 \}\times
    D) \cup (\{1 \}\times E)$).
    %
  \item
    \isatt{formula{\uscore}repl1{\uscore}intf{\uscore}fm} and
    \isatt{formula{\uscore}repl1{\uscore}intf{\uscore}fm}.

    For (codes of) first order formulas ($X\mapsto (\omega \times
    \omega \oplus \omega \times \omega) \oplus X \times X \oplus X$).
    %
  \item
    \isatt{eclose{\uscore}repl1{\uscore}intf{\uscore}fm} and 
    \isatt{eclose{\uscore}repl2{\uscore}intf{\uscore}fm}.

    For transitive closure ($X\mapsto\union X$).
  \end{itemize}
\item \isatt{tl{\uscore}repl{\uscore}intf{\uscore}fm}.

  Absoluteness of the definition of the
  $n$th element of a list (iteration of the tail operation).
  By transitivity, we do not need an instance for closure.
\end{itemize}
%% The instances so far
%% are needed to interpret locales \locale{M{\uscore}datatypes} and
%% \locale{M{\uscore}eclose}.
%% \locale{M{\uscore}datatypes} is used in the relative definition of
%% \isatt{Vset}.
These formulas, together with the next two, form the set
\isa{instances{\isadigit{1}}{\isacharunderscore}{\kern0pt}fms}.
\begin{itemize}
\item
  4 instances for definitions by well-founded recursion.
  \begin{itemize}
  \item \isatt{wfrec{\uscore}rank{\uscore}fm}.

    For $\in$-rank.
    %
  \item \isatt{trans{\uscore}repl{\uscore}HVFrom{\uscore}fm}.

    For the cumulative hierarchy (rank initial segments).
    %
  \item \isatt{wfrec{\uscore}replacement{\uscore}order{\uscore}pred{\uscore}fm}.

    For ordertypes
    %
  \item
    \isatt{replacement{\uscore}HAleph{\uscore}wfrec{\uscore}repl{\uscore}body{\uscore}fm}.

    For Aleph.
  \end{itemize}
\item
  2 instances for ordertypes.
  \begin{itemize}
  \item
    \isatt{replacement{\uscore}is{\uscore}order{\uscore}eq{\uscore}map{\uscore}fm}.

    Auxiliary instance for the definition of ordertypes, needed for
    the interpretation of \locale{M{\uscore}ordertype}.
    %
  \item
    \isatt{replacement{\uscore}is{\uscore}order{\uscore}body{\uscore}fm}.

    Replacement through $x\mapsto \otype(x)$, for Hartogs' Theorem.
  \end{itemize}
\end{itemize}

We also need a one extra replacement instance $\psi$ on $M$ for each
$\phi$ of the
previous ones to have them in $M[G]$.
\[
  \psi(x,\alpha,y_1,\dots,y_n) \defi \quine{\alpha = \min \bigl\{
    \beta \mid \exists\tau\in V_\beta.\  \mathit{snd}(x) \forces
    \phi\ [\mathit{fst}(x),\tau,y_1,\dots,y_n]\bigr\}}
\]
Here, $\mathit{fst}(\lb a,b\rb) = a$, $\mathit{snd}(\lb a,b\rb) = b$
(with default value $0$ for non pairs).
In our development, the mapping $\phi\mapsto\psi$ defined above is given by the
$\isatt{ground{\uscore}repl{\uscore}fm}$ function, and all ground replacement
instances appear in the locale \locale{M{\uscore}ZF4} and form the set
\isa{instances{\isadigit{4}}{\isacharunderscore}{\kern0pt}fms}. These are expressed using
the \isatt{ground{\uscore}replacement{\uscore}assm(M,\isasymphi)} predicate
obtained by replacing $\phi$ by
$\isatt{ground{\uscore}repl{\uscore}fm}(\phi)$ in Eq.~(\ref{eq:replacement_assm_def}).

That makes 26 instances up to now. For the setup of forcing, we
require the following 3 instances, which form the set \isa{instances{\isacharunderscore}{\kern0pt}ground{\isacharunderscore}{\kern0pt}fms}.
\begin{itemize}
\item 2 instances for definitions by well-founded recursion.
  \begin{itemize}
  \item \isatt{wfrec{\uscore}Hcheck{\uscore}fm}.

    For check-names.
    %
  \item \isatt{wfrec{\uscore}Hfrc{\uscore}at{\uscore}fm}.

    Definition of forcing for atomic formulas.
  \end{itemize}
  %
\item
  \isatt{Lambda{\uscore}in{\uscore}M{\uscore}fm(check{\uscore}fm(2,0,1),1)}.

  Replacement through $x\mapsto \lb x,\check{x}\rb$ (for the
  definition of $\punto{G}$).
  %
\end{itemize}
The next two formulas, which form the set
\isa{instances{\isacharunderscore}{\kern0pt}ground{\isacharunderscore}{\kern0pt}notCH{\isacharunderscore}{\kern0pt}fms},
are needed for the $\Delta$-System Lemma.
\begin{itemize}
\item
  \isatt{replacement{\uscore}is{\uscore}trans{\uscore}apply{\uscore}image{\uscore}fm}.

  Recursive construction of sets using a choice function (as in the
  construction of a wellorder of $X$ given a choice function on $\Pow(X)$).
  %
\item
  \isatt{replacement{\uscore}transrec{\uscore}apply{\uscore}image{\uscore}body{\uscore}fm}.

  Absoluteness of the previous construction.
\end{itemize}
%
The $31$ formulas up to this point are collected into the set
\isa{overhead{\isacharunderscore}{\kern0pt}notCH}, which is enough to
force $\neg\CH$. To force $\CH$, we required one further instance:
%
\begin{itemize}
\item
  \isatt{replacement{\uscore}dcwit{\uscore}repl{\uscore}body{\uscore}fm}.

  Absoluteness of the recursive construction in the proof of the
  Dependent Choices from $\AC$.
\end{itemize}

The particular choice of some of the instances above arose from
Paulson's architecture on which we based our development.
This applies every time
a locale from \session{ZF-Constructible} has to be
interpreted (namely \locale{M{\uscore}datatypes}
\locale{M{\uscore}eclose}, and \locale{M{\uscore}ordertype}).
%% For instance, the first
%% instance required for the definition of relative ordertypes arises
%% from Paulson's \session{ZF-Constructible}.
% https://isabelle.in.tum.de/dist/library/ZF/ZF-Constructible/Rank.html#offset_1123..1139

On the other hand, as explained in
Section~\ref{sec:zfc-axioms-as-locales}, we managed to eliminate the
instance arising from the \locale{M{\uscore}basic} locale. Similarly,
we replaced the original proof of the Schröder-Bernstein Theorem by
Zermelo's one \cite[Exr. x4.27]{moschovakis1994notes}, because the
former required at least one extra instance
% (\isatt{banach{\uscore}iterates{\uscore}fm})
arising from an iteration.

It is to be noted that application of the Forcing Theorems do not
require any extra replacement instances on $M$ (independently of the
formula $\phi$ for which they are invoked). This is not the case for
Separation, at least from our formalization: More instances are needed
as the complexity of $\phi$ grows. One point where this is apparent is
in the proof of Theorem~\ref{th:forcing-thms}(\ref{item:truth-lemma}),
that appears as the \isatt{truth{\uscore}lemma} in our development; it
depends on \isatt{truth{\uscore}lemma'} and
\isatt{truth{\uscore}lemma{\uscore}Neg} which explicitly invoke
\isatt{separation{\uscore}ax}.

%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
\subsection{Models for $\CH$ and its negation}
\label{sec:models-ch-negation}

The statements of the existence of models of $\ZFC + \neg\CH$ and of
$\ZFC + \CH$ negation appear in our formalization as follows:

\begin{isabelle}
\isacommand{corollary}\isamarkupfalse%
\ ctm{\isacharunderscore}{\kern0pt}ZFC{\isacharunderscore}{\kern0pt}imp{\isacharunderscore}{\kern0pt}ctm{\isacharunderscore}{\kern0pt}not{\isacharunderscore}{\kern0pt}CH{\isacharcolon}{\kern0pt}\isanewline
\ \ \isakeyword{assumes}\isanewline
\ \ \ \ {\isachardoublequoteopen}M\ {\isasymapprox}\ {\isasymomega}{\isachardoublequoteclose}\ {\isachardoublequoteopen}Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ ZFC{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isasymexists}N{\isachardot}{\kern0pt}\isanewline
\ \ \ \ \ \ M\ {\isasymsubseteq}\ N\ {\isasymand}\ N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\ Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\ N\ {\isasymTurnstile}\ ZFC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}{\isasymnot}{\isasymcdot}CH{\isasymcdot}{\isasymcdot}{\isacharbraceright}{\kern0pt}\ {\isasymand}\isanewline
\ \ \ \ \ \ {\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}
\end{isabelle}

\begin{isabelle}
\isacommand{corollary}\isamarkupfalse%
\ ctm{\isacharunderscore}{\kern0pt}ZFC{\isacharunderscore}{\kern0pt}imp{\isacharunderscore}{\kern0pt}ctm{\isacharunderscore}{\kern0pt}CH{\isacharcolon}{\kern0pt}\isanewline
\ \ \isakeyword{assumes}\isanewline
\ \ \ \ {\isachardoublequoteopen}M\ {\isasymapprox}\ {\isasymomega}{\isachardoublequoteclose}\ {\isachardoublequoteopen}Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ ZFC{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isasymexists}N{\isachardot}{\kern0pt}\isanewline
\ \ \ \ \ \ M\ {\isasymsubseteq}\ N\ {\isasymand}\ N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\ Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\ N\ {\isasymTurnstile}\ ZFC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}CH{\isasymcdot}{\isacharbraceright}{\kern0pt}\ {\isasymand}\isanewline
\ \ \ \ \ \ {\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}
\end{isabelle}

As the code indicates, these results are obtained as corollaries from
three theorems in which only a subset of the aforementioned
replacement instances are assumed of the ground model.

\begin{isabelle}
\isacommand{theorem}\isamarkupfalse%
\ extensions{\isacharunderscore}{\kern0pt}of{\isacharunderscore}{\kern0pt}ctms{\isacharcolon}{\kern0pt}\isanewline
\ \ \isakeyword{assumes}\isanewline
\ \ \ \ {\isachardoublequoteopen}M\ {\isasymapprox}\ {\isasymomega}{\isachardoublequoteclose}\ {\isachardoublequoteopen}Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\isanewline
\ \ \ \ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ {\isasymcdot}Z{\isasymcdot}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}p{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ p\ {\isasymin}\ overhead{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isasymPhi}\ {\isasymsubseteq}\ formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ {\isacharbraceleft}{\kern0pt}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}ground{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}fm{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isasymexists}N{\isachardot}{\kern0pt}\isanewline
\ \ \ \ \ \ M\ {\isasymsubseteq}\ N\ {\isasymand}\ N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\ Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\ M{\isasymnoteq}N\ {\isasymand}\isanewline
\ \ \ \ \ \ {\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
\ \ \ \ \ \ {\isacharparenleft}{\kern0pt}{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}{\isasymTurnstile}\ {\isasymcdot}AC{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ N{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}AC{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymand}\ N\ {\isasymTurnstile}\ {\isasymcdot}Z{\isasymcdot}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}
\end{isabelle}

Here, the 14-element \isatt{overhead} is enough to construct a proper
extension. It is  the union of
\isa{instances{\isadigit{1}}{\isacharunderscore}{\kern0pt}fms},
\isa{instances{\isacharunderscore}{\kern0pt}ground{\isacharunderscore}{\kern0pt}fms},
and the two instances
\isatt{replacement{\uscore}HAleph{\uscore}wfrec{\uscore}repl{\uscore}body{\uscore}fm}
and
\isatt{replacement{\uscore}is{\uscore}order{\uscore}eq{\uscore}map{\uscore}fm}.

\begin{isabelle}
\isacommand{theorem}\isamarkupfalse%
\ ctm{\isacharunderscore}{\kern0pt}of{\isacharunderscore}{\kern0pt}not{\isacharunderscore}{\kern0pt}CH{\isacharcolon}{\kern0pt}\isanewline
\ \ \isakeyword{assumes}\isanewline
\ \ \ \ {\isachardoublequoteopen}M\ {\isasymapprox}\ {\isasymomega}{\isachardoublequoteclose}\ {\isachardoublequoteopen}Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ ZC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}p{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ p\ {\isasymin}\ overhead{\isacharunderscore}{\kern0pt}notCH{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isasymPhi}\ {\isasymsubseteq}\ formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ {\isacharbraceleft}{\kern0pt}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}ground{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}fm{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isasymexists}N{\isachardot}{\kern0pt}\isanewline
\ \ \ \ \ \ M\ {\isasymsubseteq}\ N\ {\isasymand}\ N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\ Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\ N\ {\isasymTurnstile}\ ZC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}{\isasymnot}{\isasymcdot}CH{\isasymcdot}{\isasymcdot}{\isacharbraceright}{\kern0pt}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}\ {\isasymand}\isanewline
\ \ \ \ \ \ {\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}
\end{isabelle}

\begin{isabelle}
\isacommand{theorem}\isamarkupfalse%
\ ctm{\isacharunderscore}{\kern0pt}of{\isacharunderscore}{\kern0pt}CH{\isacharcolon}{\kern0pt}\isanewline
\ \ \isakeyword{assumes}\isanewline
\ \ \ \ {\isachardoublequoteopen}M\ {\isasymapprox}\ {\isasymomega}{\isachardoublequoteclose}\ {\isachardoublequoteopen}Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\isanewline
\ \ \ \ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ ZC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}p{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ p\ {\isasymin}\ overhead{\isacharunderscore}{\kern0pt}CH{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isasymPhi}\ {\isasymsubseteq}\ formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ {\isacharbraceleft}{\kern0pt}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}ground{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}fm{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isasymexists}N{\isachardot}{\kern0pt}\isanewline
\ \ \ \ \ \ M\ {\isasymsubseteq}\ N\ {\isasymand}\ N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\ Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\ N\ {\isasymTurnstile}\ ZC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}CH{\isasymcdot}{\isacharbraceright}{\kern0pt}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}\ {\isasymand}\isanewline
\ \ \ \ \ \ {\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}
\end{isabelle}

Finally, \isatt{overhead{\isacharunderscore}{\kern0pt}CH} is the union
of \isatt{overhead{\isacharunderscore}{\kern0pt}CH} with the $\DC$
instance \isatt{replacement{\isacharunderscore}{\kern0pt}dcwit{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}body{\isacharunderscore}{\kern0pt}fm}.
%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "independence_ch_isabelle"
%%% ispell-local-dictionary: "american"
%%% End:

### 2022-08-14 Independence_CH/sample_proof.tex
Change in section ordering. Statement of the main theorems

--- 2022-08-11 Independence_CH/details.tex
@@
-\section{Details of the formalization}
-
-\subsection{32 replacement instances to rule them all}
-\label{sec:repl-instances}
-
-We isolated 32 instances of Replacement that are sufficient to force
-$\CH$ or $\neg\CH$. The first 13 instances are needed to set up
-cardinal arithmetic in $M$. Many of these were already present in
-relational form in the \session{ZF-Constructible} library.
-
-\begin{itemize}
-\item 6 instances for iteration of operations. For each construct,
-  Paulson used one replacement instance in order to have absoluteness,
-  and a second one to obtain closure.
-  \begin{itemize}
-  \item
-    Lists
-    \isatt{list{\uscore}repl1{\uscore}intf{\uscore}fm},
-    \isatt{list{\uscore}repl2{\uscore}intf{\uscore}fm}.
-  \item
-    Formulas
-    \isatt{formula{\uscore}repl1{\uscore}intf{\uscore}fm},
-    \isatt{formula{\uscore}repl1{\uscore}intf{\uscore}fm}.
-  \item
-    Transitive closure %(iteration of $\union$)
-    \isatt{eclose{\uscore}repl1{\uscore}intf{\uscore}fm},
-    \isatt{eclose{\uscore}repl2{\uscore}intf{\uscore}fm}.
-  \end{itemize}
-\item 1 instance for absoluteness of the definition of the
-  $n$th element of a list (iteration of the tail operation)
-  \isatt{tl{\uscore}repl{\uscore}intf{\uscore}fm}. By transitivity,
-  we do not need an instance for closure.
-\end{itemize}
-The instances so far
-are needed to interpret locales \isatt{M{\uscore}datatypes} and
-\isatt{M{\uscore}eclose}. The former is used in the relative definition of
-\isatt{Vset}.
-\begin{itemize}
-\item
-  2 instances for ordertypes.
-  \begin{itemize}
-  \item Auxiliary for the definition of ordertypes
-    \isatt{replacement{\uscore}is{\uscore}order{\uscore}eq{\uscore}map{\uscore}fm}.
-
-    This one is needed for the interpretation of
-    \isatt{M{\uscore}ordertype}.
-
-  \item Replacement through $x\mapsto \otype(x)$, for Hartogs' Theorem
-    \isatt{replacement{\uscore}is{\uscore}order{\uscore}body{\uscore}fm}.
-  \end{itemize}
-\item
-  4 instances for definitions by well-founded recursion.
-  \begin{itemize}
-  \item Ordertypes
-    \isatt{wfrec{\uscore}replacement{\uscore}order{\uscore}pred{\uscore}fm}.
-  \item Rank \isatt{wfrec{\uscore}rank{\uscore}fm}.
-  \item Cumulative hierarchy (rank initial segments) \isatt{trans{\uscore}repl{\uscore}HVFrom{\uscore}fm}.
-  \item Aleph \isatt{replacement{\uscore}HAleph{\uscore}wfrec{\uscore}repl{\uscore}body{\uscore}fm}.
-  \end{itemize}
-\end{itemize}
-
-We also need a one extra replacement instance $\psi$ on $M$ for each
-$\phi$ of the
-previous ones to have them in $M[G]$.
-\[
-  \psi(x,\alpha,y_1,\dots,y_n) \defi \quine{\alpha = \min \bigl\{
-    \beta \mid \exists\tau\in V_\beta.\  \mathit{snd}(x) \forces
-    \phi\ [\mathit{fst}(x),\tau,y_1,\dots,y_n]\bigr\}}
-\]
-In our development, the mapping $\phi\mapsto\psi$ defined above is given by the
-$\isatt{ground{\uscore}repl{\uscore}fm}$ function, and all ground replacement
-instances appear in the locale \isatt{M{\uscore}ZF4}. These are expressed using
-the \isatt{ground{\uscore}replacement{\uscore}assm(M,\isasymphi)} predicate
-obtained by replacing $\phi$ by
-$\isatt{ground{\uscore}repl{\uscore}fm}(\phi)$ in Eq.~(\ref{eq:replacement_assm_def}).
-
-That makes 26 instances up to now. For the setup of forcing, we
-require the following 5 instances (the last two are needed for the $\Delta$-System Lemma):
-
-\begin{itemize}
-\item 2 instances for definitions by well-founded recursion.
-  \begin{itemize}
-  \item
-    Check names
-    \isatt{wfrec{\uscore}Hcheck{\uscore}fm}.
-  \item Forcing for atomic formulas
-    \isatt{wfrec{\uscore}Hfrc{\uscore}at{\uscore}fm}.
-  \end{itemize}
-\item Replacement through $x\mapsto \lb x,\check{x}\rb$ (for the
-  definition of $\punto{G}$)
-  \isatt{Lambda{\uscore}in{\uscore}M{\uscore}fm(check{\uscore}fm(2,0,1),1)}.
-\item Recursive construction of sets using a choice function
-  \isatt{replacement{\uscore}is{\uscore}trans{\uscore}apply{\uscore}image{\uscore}fm}.
-\item Absoluteness of the previous construction \isatt{replacement{\uscore}transrec{\uscore}apply{\uscore}image{\uscore}body{\uscore}fm}.
-\end{itemize}
-%
-This is enough to force $\neg\CH$. To force $\CH$, one further instance is needed:
-%
-\begin{itemize}
-\item Absoluteness of the recursive construction in the proof of the
-  Dependent Choices \isatt{replacement{\uscore}dcwit{\uscore}repl{\uscore}body{\uscore}fm}.
-\end{itemize}
-
-The particular choice of some of the instances above arose from
-Paulson's architecture on which we based our development.
-This applies every time
-a locale from \session{ZF-Constructible} has to be
-interpreted (namely \isatt{M{\uscore}datatypes}
-\isatt{M{\uscore}eclose}, and \isatt{M{\uscore}ordertype}).
-%% For instance, the first
-%% instance required for the definition of relative ordertypes arises
-%% from Paulson's \session{ZF-Constructible}.
-% https://isabelle.in.tum.de/dist/library/ZF/ZF-Constructible/Rank.html#offset_1123..1139
-
-On the other hand, as explained in
-Section~\ref{sec:zfc-axioms-as-locales}, we managed to eliminate the
-instance arising from the \isatt{M{\uscore}basic} locale. Similarly,
-we replaced the original proof of the Schröder-Bernstein Theorem by
-Zermelo's one \cite[Exr. x4.27]{moschovakis1994notes}, because the
-former required at least one extra instance
-% (\isatt{banach{\uscore}iterates{\uscore}fm})
-arising from an iteration.
-
-It is to be noted that application of the Forcing Theorems do not
-require any extra replacement instances on $M$ (independently of the
-formula $\phi$ for which they are invoked). This is not the case for
-Separation, at least from our formalization: More instances are needed
-as the complexity of $\phi$ grows. One point where this is apparent is
-in the proof of Theorem~\ref{th:forcing-thms}(\ref{item:truth-lemma}),
-that appears as the \isatt{truth{\uscore}lemma} in our development; it
-depends on \isatt{truth{\uscore}lemma'} and
-\isatt{truth{\uscore}lemma{\uscore}Neg} which explicitly invoke
-\isatt{separation{\uscore}ax}.
-
-\subsection{A sample formal proof}
+\section{A sample formal proof}
 \label{sec:sample-formal-proof}

### 2022-08-14 Independence_CH/main_results.tex
Eliminating M_ZF2 and derivatives (assms now part of ZF3)

@@
 The statements of the existence of models of $\ZFC + \neg\CH$ and of
-$\ZFC + \CH$ negation appear in our formalization as follows:
+$\ZFC + \CH$  appear in our formalization as follows:
 
@@
 
-As the code indicates, these results are obtained as corollaries from
-three theorems in which only a subset of the aforementioned
-replacement instances are assumed of the ground model.
-
+As the code indicates, these results are obtained as corollaries of
+two theorems in which only a subset of the aforementioned
+replacement instances are assumed of the ground model. We begin the
+discussion of these stronger results by
+considering extensions of ctms of fragments of $\ZF$.
 \begin{isabelle}
@@
 \ \ \ \ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ {\isasymcdot}Z{\isasymcdot}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}p{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ p\ {\isasymin}\ overhead{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline
-\ \ \ \ {\isachardoublequoteopen}{\isasymPhi}\ {\isasymsubseteq}\ formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ {\isacharbraceleft}{\kern0pt}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}ground{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}fm{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline
+\ \ \ \ {\isachardoublequoteopen}{\isasymPhi}\ {\isasymsubseteq}\ formula{\isachardoublequoteclose}\isanewline%
+\ \ \ \ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ {\isacharbraceleft}{\kern0pt}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}ground{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}fm{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline
 \ \ \isakeyword{shows}\isanewline
@@
 \ \ \ \ \ \ {\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
-\ \ \ \ \ \ {\isacharparenleft}{\kern0pt}{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}{\isasymTurnstile}\ {\isasymcdot}AC{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ N{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}AC{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymand}\ N\ {\isasymTurnstile}\ {\isasymcdot}Z{\isasymcdot}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}
+\ \ \ \ \ \ {\isacharparenleft}{\kern0pt}{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}{\isasymTurnstile}\ {\isasymcdot}AC{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ N{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}AC{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
+\ \ \ \ \ \ N\ {\isasymTurnstile}\ {\isasymcdot}Z{\isasymcdot}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}
 \end{isabelle}
 
-Here, the 14-element \isatt{overhead} is enough to construct a proper
+Here, the 12-element \isatt{overhead} is enough to construct a proper
 extension. It is  the union of
-\isa{instances{\isadigit{1}}{\isacharunderscore}{\kern0pt}fms},
-\isa{instances{\isacharunderscore}{\kern0pt}ground{\isacharunderscore}{\kern0pt}fms},
-and the two instances
-\isatt{replacement{\uscore}HAleph{\uscore}wfrec{\uscore}repl{\uscore}body{\uscore}fm}
-and
-\isatt{replacement{\uscore}is{\uscore}order{\uscore}eq{\uscore}map{\uscore}fm}.
+\isa{instances{\isadigit{1}}{\isacharunderscore}{\kern0pt}fms} and
+\isa{instances{\isacharunderscore}{\kern0pt}ground{\isacharunderscore}{\kern0pt}fms}.
+Also,
+\isatt{{\isasymcdot}Z{\isasymcdot}} denotes Zermelo set theory. In the
+next theorem, the relevant set of formulas is
+\isatt{overhead{\isacharunderscore}{\kern0pt}notCH}, defined above in
+Section~\ref{sec:repl-instances}, and \isatt{ZC} denotes Zermelo set
+theory plus Choice:
 
@@
 \ \ \isakeyword{assumes}\isanewline
-\ \ \ \ {\isachardoublequoteopen}M\ {\isasymapprox}\ {\isasymomega}{\isachardoublequoteclose}\ {\isachardoublequoteopen}Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ ZC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}p{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ p\ {\isasymin}\ overhead{\isacharunderscore}{\kern0pt}notCH{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline
-\ \ \ \ {\isachardoublequoteopen}{\isasymPhi}\ {\isasymsubseteq}\ formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ {\isacharbraceleft}{\kern0pt}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}ground{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}fm{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline
+\ \ \ \ {\isachardoublequoteopen}M\ {\isasymapprox}\ {\isasymomega}{\isachardoublequoteclose}\ {\isachardoublequoteopen}Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\isanewline
+\ \ \ \ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ ZC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}p{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ p\ {\isasymin}\ overhead{\isacharunderscore}{\kern0pt}notCH{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline
+\ \ \ \ {\isachardoublequoteopen}{\isasymPhi}\ {\isasymsubseteq}\ formula{\isachardoublequoteclose}\isanewline
+\ \ \ \ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ {\isacharbraceleft}{\kern0pt}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}ground{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}fm{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline
 \ \ \isakeyword{shows}\isanewline
 \ \ \ \ {\isachardoublequoteopen}{\isasymexists}N{\isachardot}{\kern0pt}\isanewline
-\ \ \ \ \ \ M\ {\isasymsubseteq}\ N\ {\isasymand}\ N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\ Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\ N\ {\isasymTurnstile}\ ZC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}{\isasymnot}{\isasymcdot}CH{\isasymcdot}{\isasymcdot}{\isacharbraceright}{\kern0pt}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}\ {\isasymand}\isanewline
+\ \ \ \ \ \ M\ {\isasymsubseteq}\ N\ {\isasymand}\ N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\ Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
+\ \ \ \ \ \ N\ {\isasymTurnstile}\ ZC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}{\isasymnot}{\isasymcdot}CH{\isasymcdot}{\isasymcdot}{\isacharbraceright}{\kern0pt}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}\ {\isasymand}\isanewline
 \ \ \ \ \ \ {\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}
@@
 
+Finally, \isatt{overhead{\isacharunderscore}{\kern0pt}CH} is the union
+of \isatt{overhead{\isacharunderscore}{\kern0pt}CH} with the $\DC$
+instance \isatt{replacement{\isacharunderscore}{\kern0pt}dcwit{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}body{\isacharunderscore}{\kern0pt}fm}:
 \begin{isabelle}
@@
 \ \ \ \ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ ZC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}p{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ p\ {\isasymin}\ overhead{\isacharunderscore}{\kern0pt}CH{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline
-\ \ \ \ {\isachardoublequoteopen}{\isasymPhi}\ {\isasymsubseteq}\ formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ {\isacharbraceleft}{\kern0pt}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}ground{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}fm{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline
+\ \ \ \ {\isachardoublequoteopen}{\isasymPhi}\ {\isasymsubseteq}\ formula{\isachardoublequoteclose}\isanewline
+\ \ \ \ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ {\isacharbraceleft}{\kern0pt}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}ground{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}fm{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline
 \ \ \isakeyword{shows}\isanewline
 \ \ \ \ {\isachardoublequoteopen}{\isasymexists}N{\isachardot}{\kern0pt}\isanewline
-\ \ \ \ \ \ M\ {\isasymsubseteq}\ N\ {\isasymand}\ N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\ Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\ N\ {\isasymTurnstile}\ ZC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}CH{\isasymcdot}{\isacharbraceright}{\kern0pt}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}\ {\isasymand}\isanewline
+\ \ \ \ \ \ M\ {\isasymsubseteq}\ N\ {\isasymand}\ N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\ Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
+\ \ \ \ \ \ N\ {\isasymTurnstile}\ ZC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}CH{\isasymcdot}{\isacharbraceright}{\kern0pt}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}\ {\isasymand}\isanewline
 \ \ \ \ \ \ {\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}
@@
 
-Finally, \isatt{overhead{\isacharunderscore}{\kern0pt}CH} is the union
-of \isatt{overhead{\isacharunderscore}{\kern0pt}CH} with the $\DC$
-instance \isatt{replacement{\isacharunderscore}{\kern0pt}dcwit{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}body{\isacharunderscore}{\kern0pt}fm}.
 %%% Local Variables: 

### 2022-08-14 Independence_CH/sample_proof.tex
Eliminating M_ZF2 and derivatives (assms now part of ZF3)

@@
 König's Theorem). The trick there consists in presenting the result
-being formalized as restricted first-order problem, and then each
+being formalized as a restricted first-order problem, and then each
 proof step can be handled by an automatic theorem prover.

### 2022-08-15 Independence_CH/main_results.tex
Shifting the remaining locales

@@
 $\isatt{ground{\uscore}repl{\uscore}fm}$ function, and all ground replacement
-instances appear in the locale \locale{M{\uscore}ZF4} and form the set
-\isa{instances{\isadigit{4}}{\isacharunderscore}{\kern0pt}fms}. These are expressed using
+instances appear in the locale \locale{M{\uscore}ZF3} and form the set
+\isa{instances{\isadigit{3}}{\isacharunderscore}{\kern0pt}fms}. These are expressed using
 the \isatt{ground{\uscore}replacement{\uscore}assm(M,\isasymphi)} predicate
@@
 
-Here, the 12-element \isatt{overhead} is enough to construct a proper
+Here, the 12-element set \isatt{overhead} is enough to construct a proper
 extension. It is  the union of

### 2022-08-15 Independence_CH/forcing.tex
Change in section ordering. Statement of the main theorems
Shifting the remaining locales
Using the replacement_assm function a bit more

--- 2022-08-12 Independence_CH/forcing.tex
@@
 particular replacement instances expressed by means of the
-\isatt{replacement{\uscore}assm(M,\isasymphi)} predicate ($@$ denotes
+\isatt{replacement{\uscore}assm(M,\mathit{env},\isasymphi)} predicate ($@$ denotes
 list concatenation):
@@
 %% \end{isabelle}
+In turn, the \isatt{{\isasymcdot}Replacement{\isasymcdot}} function
+takes a formula code and returns the corresponding replacement
+instance:
+\begin{isabelle}
+{\isasymphi}\ {\isasymin}\ formula\ {\isasymLongrightarrow}\isanewline
+{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymforall}env{\isachardot}{\kern0pt}\ replacement{\isacharunderscore}{\kern0pt}assm{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ env{\isacharcomma}{\kern0pt}\ {\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}
+\end{isabelle}
 %
@@
 %% that session in the weakened context.
+
 The locales gathering all 32 replacement instances necessary are named
-\locale{M{\uscore}ZF1} through \locale{M{\uscore}ZF4},
+\locale{M{\uscore}ZF1} through \locale{M{\uscore}ZF3},
 \locale{M{\uscore}ZF{\uscore}ground},

### 2022-08-15 Independence_CH/main_defs.tex
Using the replacement_assm function a bit more

@@
 {\isacharparenleft}{\kern0pt}{\isasymforall}env{\isasymin}list{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isachardot}{\kern0pt}\isanewline
-\isaindent{{\isacharparenleft}{\kern0pt}\ \ \ }arity{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isadigit{1}}\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ length{\isacharparenleft}{\kern0pt}env{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ separation{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isasymlambda}x{\isachardot}{\kern0pt}\ M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}x{\isacharbrackright}{\kern0pt}\ {\isacharat}{\kern0pt}\ env\ {\isasymTurnstile}\ {\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\isasep\isanewline\isanewline%
+\isaindent{{\isacharparenleft}{\kern0pt}\ \ \ }arity{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isadigit{1}}\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ length{\isacharparenleft}{\kern0pt}env{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\isanewline
+\ \ \ \ separation{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isasymlambda}x{\isachardot}{\kern0pt}\ M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}x{\isacharbrackright}{\kern0pt}\ {\isacharat}{\kern0pt}\ env\ {\isasymTurnstile}\ {\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\isasep\isanewline\isanewline%
 {\isasymphi}\ {\isasymin}\ formula\ {\isasymLongrightarrow}\isanewline
-{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\isanewline
-{\isacharparenleft}{\kern0pt}{\isasymforall}env{\isasymin}list{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isachardot}{\kern0pt}\isanewline
-\isaindent{{\isacharparenleft}{\kern0pt}\ \ \ }arity{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isadigit{2}}\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ length{\isacharparenleft}{\kern0pt}env{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\isanewline
-\isaindent{{\isacharparenleft}{\kern0pt}\ \ \ }strong{\isacharunderscore}{\kern0pt}replacement{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isasymlambda}x\ y{\isachardot}{\kern0pt}\ M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}x{\isacharcomma}{\kern0pt}\ y{\isacharbrackright}{\kern0pt}\ {\isacharat}{\kern0pt}\ env\ {\isasymTurnstile}\ {\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\isasep\isanewline\isanewline%
+{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymforall}env{\isachardot}{\kern0pt}\ replacement{\isacharunderscore}{\kern0pt}assm{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ env{\isacharcomma}{\kern0pt}\ {\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}\isanewline\isanewline%
 choice{\isacharunderscore}{\kern0pt}ax{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}A{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\ A{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}AC{\isasymcdot}%

### 2022-08-15 Independence_CH/main_defs.tex
llbrack ... rrbrack sugar

@@
 the induction theorem for them shows that the characterization is
-“complete”.%
+“complete”. (Mind the
+\isatt{\isasymlbrakk P; Q\isasymrbrakk\ \isasymLongrightarrow\ R}
+abbreviation for
+\isatt{P\ \isasymLongrightarrow\ Q\ \isasymLongrightarrow\ R}.)
+
 \begin{isabelle}%

### 2022-08-16 Independence_CH/lessons.tex
Optional section on cofinality
replacement_assm_def; ref to appendix

@@
 \subsection{Aims of a formalization and planning}
+\label{sec:aims-formalization-planning}
 
@@
 \subsection{How to believe in the formalization}
+\label{sec:how-believe-formalization}
+
 This is a rather tricky question, that was addressed by Pollack in his
@@
 is expounded. Cross-references to major milestones (which can be
-navigated by using Isabelle) are provided there. It
-corresponds roughly to less than seven journal pages.
+navigated by using Isabelle) are provided there. A curated version can
+be found as Appendix~\ref{sec:definitions_main} to this paper.
 
@@
 \begin{itemize}
-\item calculation of arities of internalized formulas;
+\item permutation of indices and calculation of arities of
+  internalized formulas;
 \item proving that certain constructions belong to the relevant

### 2022-08-16 Independence_CH/main_defs.tex
replacement_assm_def; ref to appendix

@@
 \end{isabelle}%
+
+Finally, the axiom sets are defined as follows.
+
 \begin{isabelle}%

### 2022-08-16 Independence_CH/main_results.tex
replacement_assm_def; ref to appendix

@@
 \end{isabelle}
+Both results are proved in $\ZF$ (no choice is used).
 
-As the code indicates, these results are obtained as corollaries of
+As the excerpts indicate, these results are obtained as corollaries of
 two theorems in which only a subset of the aforementioned

### 2022-08-16 Independence_CH/forcing.tex
Small font for code blocks

@@
 particular replacement instances expressed by means of the
-\isatt{replacement{\uscore}assm(M,\mathit{env},\isasymphi)} predicate ($@$ denotes
+$\isatt{replacement{\uscore}assm}(M,\mathit{env},\phi)$ predicate ($@$ denotes
 list concatenation):

### 2022-08-20 Independence_CH/main_results.tex
Updating counts (wrt no_datatype). Yet another way of presenting instances

@@
 
-\subsection{32 replacement instances to rule them all}
+\subsection{22 replacement instances to rule them all}
 \label{sec:repl-instances}
 
-We isolated 32 instances of Replacement that are sufficient to force
+We isolated 22 instances of Replacement that are sufficient to force
 $\CH$ or $\neg\CH$, which are enumerated below by the name of the
@@
 
-The first 13 instances are needed to set up
-cardinal arithmetic in $M$, along some other basic constructions.
+The first 4 instances, collected in the subset
+\isatt{instances1{\uscore}fms} of \formula, consist of basic
+constructions.
 
 \begin{itemize}
-\item 6 instances for iteration of operations (between parentheses). For each construct,
-  Paulson used one replacement instance in order to have absoluteness,
-  and a second one to obtain closure.
+\item 2 instances for transitive closure.
   \begin{itemize}
   \item
-    \isatt{list{\uscore}repl1{\uscore}intf{\uscore}fm} and 
-    \isatt{list{\uscore}repl2{\uscore}intf{\uscore}fm}.
+    \isatt{eclose{\uscore}repl1{\uscore}intf{\uscore}fm}.
 
-    For lists on a set $A$ ($X\mapsto \{0 \} \oplus A \times X$, where
-    $D\oplus E$ stands for the disjoint union/coproduct $(\{0 \}\times
-    D) \cup (\{1 \}\times E)$).
-    %
+    To prove closure under iteration of $X\mapsto\union X$.
   \item
-    \isatt{formula{\uscore}repl1{\uscore}intf{\uscore}fm} and
-    \isatt{formula{\uscore}repl1{\uscore}intf{\uscore}fm}.
-
-    For (codes of) first order formulas ($X\mapsto (\omega \times
-    \omega \oplus \omega \times \omega) \oplus X \times X \oplus X$).
-    %
-  \item
-    \isatt{eclose{\uscore}repl1{\uscore}intf{\uscore}fm} and 
     \isatt{eclose{\uscore}repl2{\uscore}intf{\uscore}fm}.
 
-    For transitive closure ($X\mapsto\union X$).
-  \end{itemize}
-\item \isatt{tl{\uscore}repl{\uscore}intf{\uscore}fm}.
-
-  Absoluteness of the definition of the
-  $n$th element of a list (iteration of the tail operation).
-  By transitivity, we do not need an instance for closure.
+    Auxiliary instance used to show absoluteness.
   \end{itemize}
 %% The instances so far
-%% are needed to interpret locales \locale{M{\uscore}datatypes} and
+%% are needed to interpret locale
 %% \locale{M{\uscore}eclose}.
-%% \locale{M{\uscore}datatypes} is used in the relative definition of
-%% \isatt{Vset}.
-These formulas, together with the next two, form the set
-\isa{instances{\isadigit{1}}{\isacharunderscore}{\kern0pt}fms}.
-\begin{itemize}
-\item
-  4 instances for definitions by well-founded recursion.
-  \begin{itemize}
 \item \isatt{wfrec{\uscore}rank{\uscore}fm}.
@@
   For the cumulative hierarchy (rank initial segments).
+\end{itemize}
+
+%% The last two and next pair have the same syntactic structure, because
+%% they are definitions by well-founded recursion.
+The next 4 instances (gathered in \isatt{instances2{\uscore}fms})
+are needed to set up
+cardinal arithmetic in $M$.
+\begin{itemize}
+\item 2 instances for ordertypes:
+  \begin{itemize}
+  \item
+    \isatt{wfrec{\uscore}replacement{\uscore}order{\uscore}pred{\uscore}fm}.
+
+    Well-founded recursion for the construction of ordertypes.
     %
-  \item \isatt{wfrec{\uscore}replacement{\uscore}order{\uscore}pred{\uscore}fm}.
-
-    For ordertypes
-    %
-  \item
-    \isatt{replacement{\uscore}HAleph{\uscore}wfrec{\uscore}repl{\uscore}body{\uscore}fm}.
-
-    For Aleph.
-  \end{itemize}
-\item
-  2 instances for ordertypes.
-  \begin{itemize}
   \item
@@
 
-    Auxiliary instance for the definition of ordertypes, needed for
-    the interpretation of \locale{M{\uscore}ordertype}.
+    Auxiliary instance for the definition of ordertypes.
+  \end{itemize}
+\item 2 instances for Aleph:
+  \begin{itemize}
     %
@@
     Replacement through $x\mapsto \otype(x)$, for Hartogs' Theorem.
+    %
+  \item
+    \isatt{replacement{\uscore}HAleph{\uscore}wfrec{\uscore}repl{\uscore}body{\uscore}fm}.
+
+    The well-founded recursion to define Aleph.
   \end{itemize}
@@
 instances appear in the locale \locale{M{\uscore}ZF3} and form the set
-\isa{instances{\isadigit{3}}{\isacharunderscore}{\kern0pt}fms}. These are expressed using
+\isatt{instances3{\uscore}fms}. These are expressed using
 the \isatt{ground{\uscore}replacement{\uscore}assm(M,\isasymphi)} predicate
@@
 
-That makes 26 instances up to now. For the setup of forcing, we
-require the following 3 instances, which form the set \isa{instances{\isacharunderscore}{\kern0pt}ground{\isacharunderscore}{\kern0pt}fms}.
-\begin{itemize}
-\item 2 instances for definitions by well-founded recursion.
+That makes 16 instances up to now. For the setup of forcing, we
+require the following 3 instances, which form the set
+\isatt{instances{\uscore}ground{\uscore}fms}.
+
 \begin{itemize}
@@
 
-    For check-names.
+  Well-founded recursion to define check-names.
   %
@@
 
-    Definition of forcing for atomic formulas.
-  \end{itemize}
+  Well-founded recursion for the definition of forcing for atomic formulas.
   %
@@
 The next two formulas, which form the set
-\isa{instances{\isacharunderscore}{\kern0pt}ground{\isacharunderscore}{\kern0pt}notCH{\isacharunderscore}{\kern0pt}fms},
+\isatt{instances{\uscore}ground{\uscore}notCH{\uscore}fms},
 are needed for the $\Delta$-System Lemma.
@@
 %
-The $31$ formulas up to this point are collected into the set
-\isa{overhead{\isacharunderscore}{\kern0pt}notCH}, which is enough to
+The $21$ formulas up to this point are collected into the set
+\isatt{overhead{\uscore}notCH}, which is enough to
 force $\neg\CH$. To force $\CH$, we required one further instance:
@@
 a locale from \session{ZF-Constructible} has to be
-interpreted (namely \locale{M{\uscore}datatypes}
-\locale{M{\uscore}eclose}, and \locale{M{\uscore}ordertype}).
+interpreted (\locale{M{\uscore}eclose} and
+\locale{M{\uscore}ordertype}, respectively, for the “auxiliary” instances).
 %% For instance, the first
@@
 % (\isatt{banach{\uscore}iterates{\uscore}fm})
-arising from an iteration.
+arising from an iteration. We also managed to avoid 10 further
+replacements by restructuring some of original theories in
+\session{ZF-Constructible}, so these modifications are included as
+part of our project.
 
-It is to be noted that application of the Forcing Theorems do not
+It is to be noted that the application of the Forcing Theorems do not
 require any extra replacement instances on $M$ (independently of the
@@
 
-Here, the 12-element set \isatt{overhead} is enough to construct a proper
+Here, the 7-element set \isatt{overhead} is enough to construct a proper
 extension. It is  the union of

### 2022-08-22 Independence_CH/forcing.tex
Updating counts (wrt no_datatype). Yet another way of presenting instances
Revision of manuscript, except Intro and from Related to end

@@
 
-The locales gathering all 32 replacement instances necessary are named
+The locales gathering all 22 replacement instances necessary are named
 \locale{M{\uscore}ZF1} through \locale{M{\uscore}ZF3},
@@
 \subsection{The fundamental theorems}\label{sec:fundamental-theorems}
-Let $\lb \PP, {\preceq} ,\1\rb \in M$ be a forcing notion. In order to
-fix the notation, the
+In order to fix notation, let $\lb \PP, {\preceq} ,\1\rb \in M$ be a forcing notion. The
 $\val$ interpretation function takes $3$ arguments so that if $G\sbq \PP$, we have
@@
 
-Next we state this version of the fundamental theorems in a compact
+Next, we state this version of the fundamental theorems in a compact
 way.

### 2022-08-22 Independence_CH/independence_ch_isabelle.tex
Appendix on lam_replacements. Macro for locales
Optional section on cofinality
Defs Main as an appendix
Change in section ordering. Statement of the main theorems
Revision of manuscript, except Intro and from Related to end

@@
   Moreover, let $\calR$ be the set of instances of the Axiom of
-  Replacement. We isolated a 32-element subset $\Omega\sbq\calR$ and
+  Replacement. We isolated a 22-element subset $\Omega\sbq\calR$ and
   defined $\calF:\calR\to\calR$
-  such that such that for every finite $\Phi\sbq\calR$
+  such that for every finite $\Phi\sbq\calR$
   we have
-  $M\models Z \cup \calF\dimg\Phi \cup \Delta$ implies $M[G]\models Z
-  \cup \Phi \cup \{ \neg \CH \}$ for any $M$-generic $G$, where $Z$ is
-  Zermelo set theory.
+  $M\models \ZC \cup \calF\dimg\Phi \cup \Delta$ implies
+  $M[G]\models \ZC \cup \Phi \cup \{ \neg \CH \}$ for any $M$-generic $G$, where $\ZC$ is
+  Zermelo set theory with Choice.
   We also obtained the formulas yielded by the Forcing Definability Theorem
@@
 
-  We'll also compare our formalization with the recent one by J.~M.~Han and F.~van Doorn in the proof assistant \emph{Lean}.
+  We also compare our formalization with the recent one by J.~M.~Han and F.~van Doorn in the proof assistant \emph{Lean}.
 
   \keywords{forcing \and Isabelle/ZF \and countable transitive models
-    \and absoluteness \and generic extension \and constructibility.}
+    \and absoluteness \and generic extension.}
 \end{abstract}
@@
 
-%% \input{forces_definition}
-
 \input{forcing}
 
-\input{details}
-%% 
-%% \input{forcing_theorems}
-%% 
-%% \input{proper_extension}
-%% 
-%% \input{replacement_choice}
-%%
+\input{sample_proof}
+
+\input{main_results}
 
@@
 
+\appendix
+
+\input{lam_replacement}
+
 \input{discipline}
+
+\input{cofinality}
+
+\input{main_defs}
+
 \end{document}

### 2022-08-22 Independence_CH/intro.tex
Revision of manuscript, except Intro and from Related to end

@@
 \subsection{Our achievements}
-We formalized a model-theoretic rendition of forcing (Section~\ref{sec:forcing}), showing how to
+We formalized a model-theoretic rendition of forcing (Sect.~\ref{sec:forcing}), showing how to
 construct proper extensions of ctms of $\ZF$ (respectively, with
 $\AC$), and we formalized the basic forcing notions required to obtain
-ctms of $\ZFC + \neg\CH$ and of $\ZFC + \CH$. No meta-theoretic issues
-(consistency, FOL calculi, etc) were formalized, so we were merely
+ctms of $\ZFC + \neg\CH$ and of $\ZFC + \CH$ (Sect.~\ref{sec:models-ch-negation}). No meta-theoretic issues
+(consistency, FOL calculi, etc) were formalized, so we were mainly
 concerned with the mathematics of forcing. Nevertheless, by inspecting
@@
 
-In order to reach our goals, we provided missing results, from basic ones
+In order to reach our goals, we provided missing results, starting from basic ones
 involving cardinal successors, countable sets, etc

### 2022-08-22 Independence_CH/main_results.tex
Revision of manuscript, except Intro and from Related to end

@@
 $\phi$ of the
-previous ones to have them in $M[G]$.
+previous ones to have them in $M[G]$:
 \[
@@
 \isatt{instances3{\uscore}fms}. These are expressed using
-the \isatt{ground{\uscore}replacement{\uscore}assm(M,\isasymphi)} predicate
+the $\isatt{ground{\uscore}replacement{\uscore}assm}(M,\mathit{env},\phi)$ predicate
 obtained by replacing $\phi$ by
@@
 
-On the other hand, as explained in
-Section~\ref{sec:zfc-axioms-as-locales}, we managed to eliminate the
-instance arising from the \locale{M{\uscore}basic} locale. Similarly,
-we replaced the original proof of the Schröder-Bernstein Theorem by
-Zermelo's one \cite[Exr. x4.27]{moschovakis1994notes}, because the
-former required at least one extra instance
+On the other hand, we replaced the original proof of the
+Schröder-Bernstein Theorem by Zermelo's one
+\cite[Exr. x4.27]{moschovakis1994notes}, because the former required
+at least one extra instance
 % (\isatt{banach{\uscore}iterates{\uscore}fm})
-arising from an iteration. We also managed to avoid 10 further
+arising from an iteration. We also managed to avoid 12 further
 replacements by restructuring some of original theories in
@@
 
-It is to be noted that the application of the Forcing Theorems do not
-require any extra replacement instances on $M$ (independently of the
-formula $\phi$ for which they are invoked). This is not the case for
-Separation, at least from our formalization: More instances are needed
+It is to be noted that the \emph{application} of the Forcing Theorems \emph{does not
+require any extra replacement}. That is, no further replacement
+instance holding at $M$ is necessary to invoke
+Theorem~\ref{th:forcing-thms}, independently of the
+formula $\phi$. But this seems  not be the case for
+Separation, at least from inspecting our formalization: More instances are needed
 as the complexity of $\phi$ grows. One point where this is apparent is
@@
 depends on \isatt{truth{\uscore}lemma'} and
-\isatt{truth{\uscore}lemma{\uscore}Neg} which explicitly invoke
-\isatt{separation{\uscore}ax}.
+\isatt{truth{\uscore}lemma{\uscore}Neg}, which explicitly invoke
+\isatt{separation{\uscore}ax}. In any case, our intended grounds
+(v.g., the transitive collapse of countable elementary submodels of a
+rank initial segment $V_\alpha$ or an $H(\kappa)$) all satisfy full
+Separation.
 
@@
 \end{isabelle}
-Both results are proved in $\ZF$ (no choice is used).
+where $\approx$ is equipotence, and the predicate \isatt{Transset}
+holds for
+transitive sets. Both results are proved without using Choice.
 

### 2022-08-22 Independence_CH/non-absolute.tex
Revision of manuscript, except Intro and from Related to end

--- 2022-05-11 Independence_CH/non-absolute.tex
@@
 concepts in the way of proving the consistency of $\AC$ are
-absolute, and treatment of relational versions and relativized notions
+absolute, and the treatment of relational versions and relativized notions
 could be minimized after proving the relevant absoluteness results:
@@
 define the needed concepts, and  state and
-prove the requisite lemmas. For instance, the 
-$\isatt{cardinal}::\tyi \fun \tyi$ function is defined in
-\session{Isabelle/ZF}, and the commands
+prove the requisite lemmas. For instance, consider the 
+$\isatt{cardinal}::\tyi \fun \tyi$ function defined in
+\session{Isabelle/ZF}. Then the commands
 \begin{isabelle}
@@
 (denoted  $|\cdot|^M$, as expected),
-its relational version $\isatt{is{\uscore}cardinal}$, the
+the relational version $\isatt{is{\uscore}cardinal}$ of the latter, the
 internalized formula \isatt{is{\uscore}cardinal{\uscore}fm} whose
@@
 Section~\ref{sec:tools-relativization}. We also relativized many
-original theories appearing in \session{ZF}, including the
+original theories appearing in Isabelle/ZF, including the
 fundamentals of cardinal arithmetic, the cumulative hierarchy, and the

### 2022-08-22 Independence_CH/proof_assists_isabelle.tex
Revision of manuscript, except Intro and from Related to end

@@
 
-Let us briefly introduce Isabelle in the large landscape of \emph{proof
-assistants} (“assitants” for short; also known as “interactive theorem provers”); we refer to the
+Let us briefly introduce Isabelle \cite{DBLP:books/sp/Paulson94} in the large landscape of \emph{proof
+assistants} (“assistants” for short; also known as “interactive theorem provers”); we refer to the
 excellent chapter by \citet{DBLP:series/hhl/HarrisonUW14} for a more
@@
 some piece of mathematics; the interaction varies from system to
-system, but a common interface consists on the system showing the
-current goal and context. The user instructs the assistant to modify the
-goal/context by means of tactics; a forward rule changes the context
-with the intention of having in the new context a hypothesis closer to
-the goal. A backward rule represents a deduction whose conclusion is
-the goal, thus the new goal corresponds to the premises of the rule. A
-proof is completed when the (current) goal is an instance of a
-hypothesis (in the current context).
+system, but a common interface consists of a display showing the
+current goal and assumptions. The user instructs the assistant to modify the
+latter by means of \emph{tactics}; a %% forward rule changes the context
+%% with the intention of having in the new context a hypothesis closer to
+%% the goal. A backward rule represents a deduction whose conclusion is
+%% the goal, thus the new goal corresponds to the premises of the rule. A
+proof is completed when the (current) goal is an instance of one of
+the assumptions.
 
@@
 that an imported theory is correct) or by the user to understand
-the proof.\footnote{As we will see there are more declaratives
-proof languages that aim to have intelligible proof scripts.}
+the proof.
+%% \footnote{As we will see there are more declaratives
+%% proof languages that aim to have intelligible proof scripts.}
 
-To have any value at all, the system should be able to say if a tactic
-makes sense (for instance, it must be forbidden to use the backward
-tactic \textit{conjI} when the main connective is a
-disjunction). Moreover one should be able to ascertain the validity of
-the reaction of an assistant (be its rejection or its acceptance of some
-alleged logical step). Edinburgh LCF~\cite{Gordon1979-qm} was an
-influential proof assistant where that verification could be achieved
-by checking a small \emph{kernel}; this is the only module that
-constructs proofs in response to user scripts. The rest of the system
-can interact with the kernel via an interface that exposes functions
-guaranteeing the validity of the proofs.
+To have any value at all, the system should only allow the application
+of sound tactics; in particular, neither the author nor (any part) of
+the system should be able to fake proofs. Edinburgh LCF~\cite{Gordon1979-qm} was an
+influential proof assistant in which the critical code (that
+constructs proofs in response to user scripts or other modules) was reduced to a small
+\emph{kernel}. Hence, by verfying the correctness of the kernel, one
+achieves confidence on the whole system.
+%% The rest of the system
+%% can interact with the kernel via an interface that exposes functions
+%% guaranteeing the validity of the proofs.
 
-The core of Isabelle, as well as LCF, is based on higher-order logic;
-another heritage of LCF is the possibility for the user to define
-\emph{tactics} to encapsulate common patterns used to solve goals. In
-fact, this is the origin of the ML family of languages: a
-Meta-Language for programming tactics. In contrast with LCF, Isabelle
-was designed~\cite{Paulson1990} as a generic theorem for several
-logics. Isabelle still offers a programmable interface (in ML) that
-allows the user, for instance, to define a tactic solving a statement
-about a natural number $n$ by doing cases on whether $n=0$ or
-$n\neq 0$.  Moreover one can define functions that manipulate terms
-(in the usual sense) or even formulas.
-
-In contrast with LCF, some of the more prominent assistants of today are
-based on some type theory. Both Coq~\cite{coq} and
+The metalogic of Isabelle, as well as LCF, is based on higher-order logic.
+%% In contrast with LCF, Isabelle
+%% was designed~\cite{Paulson1990} as a generic theorem for several
+%% logics. Isabelle still offers a programmable interface (in ML) that
+%% allows the user, for instance, to define a tactic solving a statement
+%% about a natural number $n$ by doing cases on whether $n=0$ or
+%% $n\neq 0$.  Moreover one can define functions that manipulate terms
+%% (in the usual sense) or even formulas.
+In contrast, some of the other prominent assistants of today are
+based on some (dependent) type theory. Both Coq~\cite{coq} and
 Lean~\cite{DBLP:conf/cade/Moura021} are based on the Calculus of
@@
 foundations and architecture from Isabelle; Mizar inspired, though, the
-Isar dialect used in Isabelle nowadays.
+Isar \cite{DBLP:conf/tphol/Wenzel99} dialect used in Isabelle nowadays.
 
@@
 
-% In the case of Isabelle, Standard ML is the first of the four layers
-% in which we work in this assistant. Both the kernel and automation
-% of proofs is coded in ML, sometimes as a substitute for induction on
-% formulas, as the next section explains.
+Isabelle also inherited from LCF the possibility for the user to define
+tactics to encapsulate common patterns used to solve goals. In
+fact, this is the origin of the ML family of languages: a
+Meta-Language for programming tactics.
+In the case of Isabelle, \emph{Standard} ML is the first of the four layers
+in which we work in this assistant. Both the kernel and the automation
+of proofs are coded in ML, sometimes as a substitute for induction on
+formulas, as the next section explains.
 
@@
 
-Isabelle/ZF reaches Hessenberg's $|A|\cdot|A| = |A|$. Our decision
+Isabelle/ZF reaches essentially Hessenberg's $|A|\cdot|A| = |A|$. Our decision
 (during 2017) to
@@
 \session{ZF-Constructible} defines the set $\formula$ of first-order 
-formulas internalized as sets.\footnote{These, alongside with lists, are instances of
+formulas in the language $\{ \in \}$, internalized as sets.%
+\footnote{These, alongside with lists, are instances of
 Isabelle/ZF treatment of inductively defined (internal) datatypes;
@@
 (this is in constrast to general well-founded recursion, for which
-  one has to work with the fundamental recursor $\isatt{wfrec}$).}  It
-is a minimalistic language only suited for set theory
-(\textbf{just say they're formulas in language $\{\in\}$?}), since its
+one has to work with the fundamental recursor $\isatt{wfrec}$).}  Its
 atomic formulas have the form $\cdot x \in y\cdot$ and $\cdot x =
@@
 $z\in\omega$, $z$ is free in $\phi$ if it occurs under at most $z$
-quantifiers. The \isatt{arity} function on $\phi$ is one more than the
+quantifiers. The \isatt{arity} function on $\phi$ is one plus the
 maximum free index occurring in $\phi$.
@@
 free indices, and $\phi\in\formula$, and it is written
-$M,\mathit{env}\models\phi$ in our formalization.\footnote{Notice that
-  $M,\mathit{env}\models\phi$ is only meaningful if
-  $\isatt{length}(\mathit{env}) \geqslant
-\isatt{arity}(\phi)$.\textbf{$\leftarrow$ necessary?}} This completes the
+$M,\mathit{env}\models\phi$ in our formalization.
+%% \footnote{Notice that
+%%   $M,\mathit{env}\models\phi$ is only meaningful if
+%%   $\isatt{length}(\mathit{env}) \geqslant
+%% \isatt{arity}(\phi)$.\textbf{$\leftarrow$ necessary?}}
+This completes the
 description of the fourth and last formal layer of the development.
@@
 Internalized formulas for most (but not all) of the relational
-concepts can be obtained by guiding the automatic tactics. In
-\session{ZF-Constructible} most of the concepts were internalized by
+concepts can be obtained by guiding the automatic tactics. But in the
+early development of 
+\session{ZF-Constructible}, most of the concepts were internalized by
 hand; this is the case for union,

### 2022-08-22 Independence_CH/sample_proof.tex
Revision of manuscript, except Intro and from Related to end

@@
 
-We present a fragment of the formal version of Kunen's proof that the
-Powerset Axiom holds in a generic extension. We quote the relevant
-paragraph of \cite[Thm.~IV.2.27]{kunen2011set}:
+We present a fragment of the formal version of the proof that the
+Powerset Axiom holds in a generic extension, which also serves to
+illustrate the Isar dialect of Isabelle.
+
+We quote the relevant
+paragraph of Kunen's \cite[Thm.~IV.2.27]{kunen2011set}:
 \begin{quote}
@@
 magnificent results obtained by P.~Koepke and his team by using
-Isabelle/Naproche \cite{10.1007/978-3-030-81097-9_2} (notoriously,
-König's Theorem). The trick there consists in presenting the result
+Isabelle/Naproche \cite{10.1007/978-3-030-81097-9_2} (particularly,
+the proof König's Theorem). The trick there consists in presenting the result
 being formalized as a restricted first-order problem, and then each

### 2022-08-22 Independence_CH/related.tex
Minor in Related

@@
 to perform in the Calculus of Inductive Constructions on which Lean
-(and Coq) are based, and in Isabelle/ZF are rather cumbersome. Also, a
+is based, and in Isabelle/ZF are rather cumbersome. Also, a
 typed discipline provides aid to write succintly and many assumptions
@@
 clear way the convenience of the interaction with Isabelle/HOL
-(which, in this case, is also paid in consistency-strength currency
+(which is also paid in consistency-strength currency
 \cite[Sect.~3]{DBLP:conf/ictac/Obua06}).

### 2022-08-22 Independence_CH/lessons.tex
Slight revision of Lessons

@@
 A similar experience, but on an opposite side of the formalization
-spectrum happened to the Liquid Tensor Experiment as described by Scholze
+spectrum happened to the Liquid Tensor Experiment \cite{nature-lte} as described by Scholze
 in \cite{LTE2021}. People involved in the formalization simply pushed
@@
   we only proved the Delta System Lemma for $\aleph_1$-sized families;
-  thus limiting us to the plain ccc, and avoiding the relativization
+  thus limiting us to $\kappa=\aleph_1$ case of the $\kappa$-chain condition, and dodging the relativization
   of the material on cofinalities \cite{Delta_System_Lemma-AFP};
@@
   closed forcings (in fact, we formalized the bare minimum requirement
-  of being $(<\omega+1)$-closed); and
+  of being $(<\omega+1)$-closed, that is, under plain $\omega$-sequences).
 \end{itemize}
@@
 \item (required for the above) showing that particular instances of
-  separation and replacement hold in the ground model;
-\item etc.
+  separation and replacement hold in the ground model.
 \end{itemize}
@@
   instances paved the way for identifying which were the ones needed for
-  forcing! ($\leftarrow$\textbf{put somewhere, don't know if this is the right
-    place}).}.
+  forcing!}.
 

### 2022-08-22 Independence_CH/conclusion.tex
Comments for further edits
Added one question in conclusions.
A few more words on forcing
Some words as a conclusion

@@
-\section{Conclusion}
+\section{Future directions}
 \label{sec:conclusion}
 
-One of our main goals was to provide a \emph{proof document}, which
-can hopefully be read by mathematicians/set-theorists.
+There are many possibilities for further work starting from this
+formalization. We will mention just a few.
 
-Where did we fail on this?
-\begin{enumerate}
-\item
-  The definition of forces is too codish.
-\item
-  Proving instances of axiom schemes.
-\end{enumerate}
+Obvious missing pieces would be proving the standard properties of
+general Cohen posets $\Fn_\kappa(I,J)$, and to modify the core
+definitions to allow for class forcing. We would also like to try the
+Boolean valued approach to compare the (un)ease of formalization using
+Isabelle/ZF.
 
-The experience of formalizing math:
-\begin{enumerate}
-\item It is fun, almost addictive.
-\item Feeling of accomplishment after seeing your writings
-  validated beyond reasonable doubt (v.g. cofinality).
-\item On the grim side: It is easy to forget to put perspective on a
-  formalization issue and meditate on paper.
-\end{enumerate}
+Another desirable goal is to construct transitive set models of $\ZFC$
+from a large cardinal. There's some work to be done for that: Even the
+definition of inaccessibles, and of the transitive collapse (for that
+matter), are still missing.
+
+In line with identifying (almost) minimal set of axioms, we would like
+to obtained formalized results concerning which instances of
+Separation are needed to use forcing. A necessary ingredient will
+certainly be an implementation of Gödel operations
+\cite[Thm.~13.4]{Jech_Millennium}.
+
+In our previous landmark \cite{2020arXiv200109715G}, we contributed
+with some modifications to \session{ZF-Constructible}; this is now
+part of the official Isabelle distribution.  We intend ask Isabelle
+maintainers to consider the more modular versions of some of those
+theories that we are presenting in this project.
+
+As final words about our journey, we believe that, as in mathematics
+in general, the experience of working in a formal environment can be
+daunting, but at the same time extremely rewarding: The feeling of
+accomplishment after seeing one's writings validated beyond doubt is
+in some ways comparable that of finding a proof of an important
+lemma. It also allows subtly different ways of reasoning (with their
+own merits and pitfalls---it is easy to forget how easy a proof on
+paper is once you are fully engaged in directing your assistant). We
+hope that at some point these experiencies are shared by our community
+at large.
 

### 2022-08-23 Independence_CH/forcing.tex
Locales used for forcing

@@
 
-Further locales gather the assumption of transitivity of $M$ and
+Further locales gather
 particular replacement instances expressed by means of the
@@
 \subsection{The fundamental theorems}\label{sec:fundamental-theorems}
-In order to fix notation, let $\lb \PP, {\preceq} ,\1\rb \in M$ be a forcing notion. The
-$\val$ interpretation function takes $3$ arguments so that if $G\sbq \PP$, we have
-$M[G]\defi \{ \val(\PP,G,\punto{a}) : \punto{a}\in M \}$.
+At this point, we work inside the locale \isatt{M{\uscore}ctm1} that
+assumes $M$ to be
+countable and transitive, and satisfies some fragment of
+$\ZFC$.
+\footnote{%
+Namely, Zermelo set theory plus the replacement instances included in
+\locale{M{\uscore}ZF1} and \locale{M{\uscore}ZF{\uscore}ground}.
+}%
+conjunction of the sets This is further extended by assuming a forcing notion
+$\lb \PP, {\preceq} ,\1\rb \in M$. The actual implementation reads:
+
+\begin{isabelle}
+\isacommand{locale}\isamarkupfalse%
+\ forcing{\isacharunderscore}{\kern0pt}notion\ {\isacharequal}{\kern0pt}\isanewline
+\ \ \isakeyword{fixes}\ P\ {\isacharparenleft}{\kern0pt}{\isacartoucheopen}{\isasymbbbP}{\isacartoucheclose}{\isacharparenright}{\kern0pt}\ \isakeyword{and}\ leq\ \isakeyword{and}\ one\ {\isacharparenleft}{\kern0pt}{\isacartoucheopen}{\isasymone}{\isacartoucheclose}{\isacharparenright}{\kern0pt}\isanewline
+\ \ \isakeyword{assumes}\ one{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}P{\isacharcolon}{\kern0pt}\ \ \ \ \ \ \ {\isachardoublequoteopen}{\isasymone}\ {\isasymin}\ {\isasymbbbP}{\isachardoublequoteclose}\isanewline
+\ \ \ \ \isakeyword{and}\ leq{\isacharunderscore}{\kern0pt}preord{\isacharcolon}{\kern0pt}\ \ \ \ \ \ \ {\isachardoublequoteopen}preorder{\isacharunderscore}{\kern0pt}on{\isacharparenleft}{\kern0pt}{\isasymbbbP}{\isacharcomma}{\kern0pt}leq{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\isanewline
+\ \ \ \ \isakeyword{and}\ one{\isacharunderscore}{\kern0pt}max{\isacharcolon}{\kern0pt}\ \ \ \ \ \ \ \ \ \ {\isachardoublequoteopen}{\isasymforall}p{\isasymin}{\isasymbbbP}{\isachardot}{\kern0pt}\ {\isasymlangle}p{\isacharcomma}{\kern0pt}{\isasymone}{\isasymrangle}{\isasymin}leq{\isachardoublequoteclose}
+\end{isabelle}
+
+\begin{isabelle}
+\isacommand{locale}\isamarkupfalse%
+\ forcing{\isacharunderscore}{\kern0pt}data{\isadigit{1}}\ {\isacharequal}{\kern0pt}\ forcing{\isacharunderscore}{\kern0pt}notion\ {\isacharplus}{\kern0pt}\ M{\isacharunderscore}{\kern0pt}ctm{\isadigit{1}}\ {\isacharplus}{\kern0pt}\isanewline
+\ \ \isakeyword{assumes}\ P{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}M{\isacharcolon}{\kern0pt}\ \ \ \ \ \ \ \ \ \ \ {\isachardoublequoteopen}{\isasymbbbP}\ {\isasymin}\ M{\isachardoublequoteclose}\isanewline
+\ \ \ \ \isakeyword{and}\ leq{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}M{\isacharcolon}{\kern0pt}\ \ \ \ \ \ \ \ \ {\isachardoublequoteopen}leq\ {\isasymin}\ M{\isachardoublequoteclose}
+\end{isabelle}
+In order to fix notation, the
+$\val$ interpretation function takes as its first argument the filter
+and then the name, so we have
+$M[G]\defi \{ \val(G,\tau) : \tau\in M \}$.
 
@@
   such that for every
-  $\phi\in\formula$ and $\punto{a}_0,\dots,\punto{a}_n\in M$,
+  $\phi\in\formula$ and $\tau_0,\dots,\tau_n\in M$,
   \begin{enumerate}
@@
   arity of $\forceisa(\phi)$ is at most $\isatt{arity}(\phi) + 4$; and if
-  “$p \forces \phi\ [\punto{a}_0,\dots,\punto{a}_n]$”
+  “$p \forces \phi\ [\tau_0,\dots,\tau_n]$”
   denotes
-  “$M, [p,\PP,\preceq,\1, \punto{a}_0,\dots,\punto{a}_n]  \models
+  “$M, [p,\PP,\preceq,\1, \tau_0,\dots,\tau_n]  \models
   \forceisa(\phi)$”, we have:
@@
     \[
-      \exists p\in G.\ \; p \forces \phi\ [\punto{a}_0,\dots,\punto{a}_n]
+      \exists p\in G.\ \; p \forces \phi\ [\tau_0,\dots,\tau_n]
     \]
@@
     \[
-      M[G], [\val(\PP,G,\punto{a}_0),\dots,\val(\PP,G,\punto{a}_n)]
+      M[G], [\val(G,\tau_0),\dots,\val(G,\tau_n)]
       \models \phi.
     \]
-  \item \label{item:density-lemma} (Density Lemma) $p \forces \phi\ [\punto{a}_0,\dots,\punto{a}_n]$
+  \item \label{item:density-lemma} (Density Lemma) $p \forces \phi\ [\tau_0,\dots,\tau_n]$
     if and only if 
-    $\{q\in \PP :  q \forces \phi\ [\punto{a}_0,\dots,\punto{a}_n]\}$
+    $\{q\in \PP :  q \forces \phi\ [\tau_0,\dots,\tau_n]\}$
     is dense below $p$.
@@
 \ \ \isakeyword{assumes}\isanewline
-\ \ \ \ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}M{\isacharunderscore}{\kern0pt}generic{\isacharparenleft}{\kern0pt}G{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\isanewline
+\ \ \ \ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\isanewline
 \ \ \ \ {\isachardoublequoteopen}env{\isasymin}list{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\ {\isachardoublequoteopen}arity{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymle}length{\isacharparenleft}{\kern0pt}env{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\isanewline
 \ \ \isakeyword{shows}\isanewline
-\ \ \ \ {\isachardoublequoteopen}{\isacharparenleft}{\kern0pt}{\isasymexists}p{\isasymin}G{\isachardot}{\kern0pt}\ p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}{\kern0pt}\ \ \ {\isasymlongleftrightarrow}\ \ \ M{\isacharbrackleft}{\kern0pt}G{\isacharbrackright}{\kern0pt}{\isacharcomma}{\kern0pt}\ map{\isacharparenleft}{\kern0pt}val{\isacharparenleft}{\kern0pt}P{\isacharcomma}{\kern0pt}G{\isacharparenright}{\kern0pt}{\isacharcomma}{\kern0pt}env{\isacharparenright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymphi}{\isachardoublequoteclose}
+\ \ \ \ {\isachardoublequoteopen}{\isacharparenleft}{\kern0pt}{\isasymexists}p{\isasymin}G{\isachardot}{\kern0pt}\ p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}{\kern0pt}\ \ \ {\isasymlongleftrightarrow}\ \ \ M{\isacharbrackleft}{\kern0pt}G{\isacharbrackright}{\kern0pt}{\isacharcomma}{\kern0pt}\ map{\isacharparenleft}{\kern0pt}val{\isacharparenleft}{\kern0pt}G{\isacharparenright}{\kern0pt}{\isacharcomma}{\kern0pt}env{\isacharparenright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymphi}{\isachardoublequoteclose}
 \end{isabelle}
@@
 \ Forces\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}{\isacharbrackleft}{\kern0pt}i{\isacharcomma}{\kern0pt}\ i{\isacharcomma}{\kern0pt}\ i{\isacharbrackright}{\kern0pt}\ {\isasymRightarrow}\ o{\isachardoublequoteclose}\ \ {\isacharparenleft}{\kern0pt}{\isachardoublequoteopen}{\isacharunderscore}{\kern0pt}\ {\isasymtturnstile}\ {\isacharunderscore}{\kern0pt}\ {\isacharunderscore}{\kern0pt}{\isachardoublequoteclose}\ {\isacharbrackleft}{\kern0pt}{\isadigit{3}}{\isadigit{6}}{\isacharcomma}{\kern0pt}{\isadigit{3}}{\isadigit{6}}{\isacharcomma}{\kern0pt}{\isadigit{3}}{\isadigit{6}}{\isacharbrackright}{\kern0pt}\ {\isadigit{6}}{\isadigit{0}}{\isacharparenright}{\kern0pt}\ \isakeyword{where}\isanewline
-\ \ {\isachardoublequoteopen}p\ {\isasymtturnstile}\ {\isasymphi}\ env\ \ \ {\isasymequiv}\ \ \ M{\isacharcomma}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isacharbrackleft}{\kern0pt}p{\isacharcomma}{\kern0pt}P{\isacharcomma}{\kern0pt}leq{\isacharcomma}{\kern0pt}{\isasymone}{\isacharbrackright}{\kern0pt}\ {\isacharat}{\kern0pt}\ env{\isacharparenright}{\kern0pt}\ {\isasymTurnstile}\ forces{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\isanewline
+\ \ {\isachardoublequoteopen}p\ {\isasymtturnstile}\ {\isasymphi}\ env\ \ \ {\isasymequiv}\ \ \ M{\isacharcomma}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isacharbrackleft}{\kern0pt}p{\isacharcomma}{\kern0pt}{\isasymbbbP}{\isacharcomma}{\kern0pt}leq{\isacharcomma}{\kern0pt}{\isasymone}{\isacharbrackright}{\kern0pt}\ {\isacharat}{\kern0pt}\ env{\isacharparenright}{\kern0pt}\ {\isasymTurnstile}\ forces{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}
 \end{isabelle}

### 2022-08-23 Independence_CH/sample_proof.tex
Locales used for forcing

@@
 \ \ \isacommand{let}\isamarkupfalse%
-\ {\isacharquery}{\kern0pt}Q{\isacharequal}{\kern0pt}{\isachardoublequoteopen}Pow{\isacharparenleft}{\kern0pt}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}{\isasymtimes}P{\isacharparenright}{\kern0pt}\ {\isasyminter}\ M{\isachardoublequoteclose}
+\ {\isacharquery}{\kern0pt}Q{\isacharequal}{\kern0pt}{\isachardoublequoteopen}Pow{\isacharparenleft}{\kern0pt}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}{\isasymtimes}\isasymbbbP{\isacharparenright}{\kern0pt}\ {\isasyminter}\ M{\isachardoublequoteclose}
 \end{isabelle}
@@
 \ \ \isacommand{have}\isamarkupfalse%
-\ {\isachardoublequoteopen}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}{\isasymtimes}P\ {\isasymin}\ M{\isachardoublequoteclose}\ {\isachardoublequoteopen}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}\ {\isasymin}\ M{\isachardoublequoteclose}\isanewline
+\ {\isachardoublequoteopen}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}{\isasymtimes}\isasymbbbP\ {\isasymin}\ M{\isachardoublequoteclose}\ {\isachardoublequoteopen}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}\ {\isasymin}\ M{\isachardoublequoteclose}\isanewline
 \ \ \ \ \isacommand{by}\isamarkupfalse%
@@
 %% \ \ \ \ \isacommand{using}\isamarkupfalse%
-%% \ one{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}P\ P{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}M\ transitivity\isanewline
+%% \ one{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}\isasymbbbP\ P{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}M\ transitivity\isanewline
 %% \ \ \ \ \isacommand{by}\isamarkupfalse%
@@
 \ \ \ \ \isacommand{let}\isamarkupfalse%
-\ {\isacharquery}{\kern0pt}{\isasymtheta}{\isacharequal}{\kern0pt}{\isachardoublequoteopen}{\isacharbraceleft}{\kern0pt}{\isasymlangle}{\isasymsigma}{\isacharcomma}{\kern0pt}p{\isasymrangle}\ {\isasymin}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}{\isasymtimes}P\ {\isachardot}{\kern0pt}\ p\ {\isasymtturnstile}\ {\isasymcdot}{\isadigit{0}}\ {\isasymin}\ {\isadigit{1}}{\isasymcdot}\ {\isacharbrackleft}{\kern0pt}{\isasymsigma}{\isacharcomma}{\kern0pt}{\isasymchi}{\isacharbrackright}{\kern0pt}\ {\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}
+\ {\isacharquery}{\kern0pt}{\isasymtheta}{\isacharequal}{\kern0pt}{\isachardoublequoteopen}{\isacharbraceleft}{\kern0pt}{\isasymlangle}{\isasymsigma}{\isacharcomma}{\kern0pt}p{\isasymrangle}\ {\isasymin}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}{\isasymtimes}\isasymbbbP\ {\isachardot}{\kern0pt}\ p\ {\isasymtturnstile}\ {\isasymcdot}{\isadigit{0}}\ {\isasymin}\ {\isadigit{1}}{\isasymcdot}\ {\isacharbrackleft}{\kern0pt}{\isasymsigma}{\isacharcomma}{\kern0pt}{\isasymchi}{\isacharbrackright}{\kern0pt}\ {\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}
 \end{isabelle}

### 2022-08-23 Independence_CH/forcing.tex
Typo spotted during an unauthorized perusal by Miguel :rofl:

@@
 countable and transitive, and satisfies some fragment of
-$\ZFC$.
+$\ZFC$%
 \footnote{%
@@
 \locale{M{\uscore}ZF1} and \locale{M{\uscore}ZF{\uscore}ground}.
-}%
-conjunction of the sets This is further extended by assuming a forcing notion
+}.
+This is further extended by assuming a forcing notion
 $\lb \PP, {\preceq} ,\1\rb \in M$. The actual implementation reads:

### 2022-08-26 Independence_CH/main_results.tex
Attempt to clarify the use of axiom schemes in Forcing Theorems

@@
 
-It is to be noted that the \emph{application} of the Forcing Theorems \emph{does not
-require any extra replacement}. That is, no further replacement
-instance holding at $M$ is necessary to invoke
-Theorem~\ref{th:forcing-thms}, independently of the
-formula $\phi$. But this seems  not be the case for
-Separation, at least from inspecting our formalization: More instances are needed
+It is to be noted that the proofs of the Forcing Theorems do not
+require any extra replacement; actually, they only need the 7
+instances appearing in \isatt{instances1{\uscore}fms} and
+\isatt{instances{\uscore}ground{\uscore}fms}.  But this seems not be
+the case for Separation, at least by inspecting our formalization:
+More instances holding in $M$ are needed 
 as the complexity of $\phi$ grows. One point where this is apparent is

### 2022-08-27 Independence_CH/proof_assists_isabelle.tex
Minor typos

@@
 To have any value at all, the system should only allow the application
-of sound tactics; in particular, neither the author nor (any part) of
+of sound tactics; in particular, neither the author nor (any part of)
 the system should be able to fake proofs. Edinburgh LCF~\cite{Gordon1979-qm} was an
@@
 In the case of Isabelle, \emph{Standard} ML is the first of the four layers
-in which we work in this assistant. Both the kernel and the automation
+on which we work in this assistant. Both the kernel and the automation
 of proofs are coded in ML, sometimes as a substitute for induction on
@@
 possible that higher-order quantification gets used in $P$. The
-statement of $\AC$, also uses a free higher-order variable to denote
+statement of $\AC$ also uses a free higher-order variable to denote
 an indexed family of nonempty sets. % but we only

### 2022-08-28 Independence_CH/intro.tex
First try at an articulate introduction

@@
 
+This paper is the culmination of our project on the computerized
+formalization of the undecidability of the Continuum Hypothesis
+($\CH$) from Zermelo-Fraenkel set theory with Choice ($\ZFC$), under the
+assumption of the existence of a countable transitive model (ctm) of
+$\ZFC$. In contrast to our reports of the previous steps towards this
+goal
+\cite{2018arXiv180705174G,2019arXiv190103313G,2020arXiv200109715G}, we
+intend here to present our development to the mathematical logic
+community. For this reason, we start with a general discussion around
+the formalization of mathematics.
+
 \subsection{Formalized mathematics}
-There is a growing trend in the use of computers to assist in the
-\emph{formal verification} of mathematics. Note, for instance, the forthcoming
-\href{https://web.archive.org/web/20211220044218/https://icm2022.org/plenary-lectures#Kevin-Buzzard}{ICM2022
-  plenary lecture} by Kevin Buzzard.
+The use of computers to assist the creation and verification of
+mathematics has seen a steady grow. But the general awareness on the
+matter still seems to be a bit scant (even among mathematicians
+involved in foundations), and the venues devoted to the communication
+of formalized mathematics are, mainly, computer science journals and
+conferences: JAR, ITP, IJCAR, CPP, CICM, and others.
 
-Several venues devoted to the communication of formalized mathematics;
-today, mainly CS conferences: ITP, IJCAR, CPP, CICM, and others.
+Nevertheless, the discussion about the subject in central mathematical
+circles is increasing; there were some hints on the ICM2018 panel on
+“machine-assisted” proofs
+\cite{https://doi.org/10.48550/arxiv.1809.08062} and a lively
+promotion by Kevin Buzzard, during his ICM2022 special plenary lecture
+\cite{2021arXiv211211598B}.
 
-Many proof assistants, each with a different flavour and underlying
-foundations: Naproche (Koepke, mainly FOL + ATP), Mizar,
-Isabelle/HOL, Lean, \dots, Coq, Agda,\dots.
-
-These assistants provide several dialects, among which we single out:
-\begin{enumerate}
-\item Procedural: Useful for exploration/research.
-\item Declarative: Only one that can be read by humans!
-\end{enumerate}
+%% These assistants provide several dialects, among which we single out:
+%% \begin{enumerate}
+%% \item Procedural: Useful for exploration/research.
+%% \item Declarative: Only one that can be read by humans!
+%% \end{enumerate}
 
@@
 reader surely understands that, aside from trivial results, no current
-technology allows us to state a reasonable (and correct) theorem
+technology allows us to state a reasonably complex (and correct) theorem
 statement and expect to obtain a proof after hitting ``Enter'', at
@@
 
-The fact is that proof assistants are designed for the human prover to
+The fact is that \emph{proof assistants} are designed for the human prover to
 be able to decompose a statement to be proved into smaller subgoals
@@
 
-The same happens to the extent of mathematics formalizable (or even,
-which has \emph{already} been formalized): The community seems not to
-be fully aware of the potential of the technology. A recent discussion
-at the FOM mailing list shows this state of affairs within our fellow
-logicians. It is thus convenient to recall
-some major formalization projects, such as the Four Color Theorem
+To appreciate the extent of mathematics formalizable, it is convenient to recall
+some major successful projects, such as the Four Color Theorem
 \cite{MR2463991}, the Odd Order Theorem
@@
 and the Liquid Tensor Experiment \cite{LTE2020,LTE2021} on Lean.
+
+We'll continue our description of proof assistants in
+Section~\ref{sec:proof-assist-isabelle}. We kindly invite the reader
+to supplement the previous comments by reading the apt summary by
+A.~Koutsoukou-Argyraki \cite{angeliki} and the interviews
+therein; some of the experts consulted have also discussed
+in \cite{2022arXiv220704779B} the status of formalized versus standard
+proof in mathematics.
 
@@
 One added value that is obtained from the present formalization is
-that we identified a handfull of instances of Replacement which are
-sufficient to set the forcing machinery up (Section~\ref{sec:repl-instances}).
-%% (Separation is a bit more complicated).
-%% A verification that no choice  is needed at the level of the
-%%  object logic.
+that we identified a handful of instances of Replacement which are
+sufficient to set the forcing machinery up
+(Section~\ref{sec:repl-instances}), on the basis of Zermelo set theory.
 The eagerness to obtain this level of detail might be a consequence of
 “an unnatural tendency to investigate, for the most part, trivial
-minutiae of the formalism,” as it was put by Cohen
-\cite{zbMATH02012060}, but we were also drove by curiosity.
+minutiae of the formalism” on our part, as it was put by Cohen
+\cite{zbMATH02012060}, but we'd rather say that we were driven by
+curiosity.
 

### 2022-08-28 Independence_CH/proof_assists_isabelle.tex
First try at an articulate introduction

@@
 
-Let us briefly introduce Isabelle \cite{DBLP:books/sp/Paulson94} in the large landscape of \emph{proof
-assistants} (“assistants” for short; also known as “interactive theorem provers”); we refer to the
+Let us briefly introduce Isabelle \cite{DBLP:books/sp/Paulson94} in the large landscape of proof
+assistants (“assistants” for short; also known as “interactive theorem provers”); we refer to the
 excellent chapter by \citet{DBLP:series/hhl/HarrisonUW14} for a more
@@
 foundations and architecture from Isabelle; Mizar inspired, though, the
-Isar \cite{DBLP:conf/tphol/Wenzel99} dialect used in Isabelle nowadays.
+Isar \cite{DBLP:conf/tphol/Wenzel99} dialect used in Isabelle
+nowadays, which aims at the production of proof scripts that are
+closer to mathematical texts.%
+\footnote{%
+We recommend the excellent survey \cite{2020arXiv200909541A} by
+J.~Avigad for details about the different logical foundations in which
+assistants are based.}
 
@@
 % is the only ``trusted'' part of the assistant's code.
-
+%
 % Several of the more established assistants (HOL Light, Coq, Isabelle)

### 2022-08-28 Independence_CH/sample_proof.tex
First try at an articulate introduction

@@
 Isabelle/Naproche \cite{10.1007/978-3-030-81097-9_2} (particularly,
-the proof König's Theorem). The trick there consists in presenting the result
+the proof of König's Theorem). The trick there consists in presenting the result
 being formalized as a restricted first-order problem, and then each

### 2022-08-28 Independence_CH/independence_ch_isabelle.tex
Typos in abstract

@@
   Moreover, let $\calR$ be the set of instances of the Axiom of
-  Replacement. We isolated a 22-element subset $\Omega\sbq\calR$ and
+  Replacement. We isolated a 21-element subset $\Omega\sbq\calR$ and
   defined $\calF:\calR\to\calR$
@@
   we have
-  $M\models \ZC \cup \calF\dimg\Phi \cup \Delta$ implies
+  $M\models \ZC \cup \calF\dimg\Phi \cup \Omega$ implies
   $M[G]\models \ZC \cup \Phi \cup \{ \neg \CH \}$ for any $M$-generic $G$, where $\ZC$ is

### 2022-08-29 Independence_CH/intro.tex
Link to AFP-devel

@@
 Conjecture \cite{MR3659768}. There is a vast mathematical corpus at
-the Archive of Formal Proofs based on Isabelle, and formalizations of
+the Archive of Formal Proofs (AFP) based on Isabelle, and formalizations of
 brand new and complex objects like the definition of perfectoid spaces \cite{10.1145/3372885.3373830}
@@
 
+The lastest version of our formalization can be accessed at the
+development site of the AFP, via the following link:
+\begin{center}
+  \url{https://devel.isa-afp.org/entries/Independence_CH.html}
+\end{center}
 %%% Local Variables: 

### 2022-08-29 Independence_CH/intro.tex
First round of corrections with Miguel

@@
 Section~\ref{sec:proof-assist-isabelle}. We kindly invite the reader
-to supplement the previous comments by reading the apt summary by
+to deepen the previous exposition by reading the apt summary by
 A.~Koutsoukou-Argyraki \cite{angeliki} and the interviews
@@
 (Section~\ref{sec:isabelle-metalogic-meta}) it can be stated that our
-formalization is closest to a proof in $\ZF$ of the previous
+formalization is a bona fide proof in $\ZF$ of the previous
 constructions.
 
-In order to reach our goals, we provided missing results, starting from basic ones
-involving cardinal successors, countable sets, etc
+In order to reach our goals, we provided basic results that were
+missing from Isabelle's $\ZF$ library, starting from ones
+involving cardinal successors, countable sets, etc.
 (Section~\ref{sec:extension-isabellezf}); we extended the treatment of relativization of

### 2022-08-29 Independence_CH/non-absolute.tex
First round of corrections with Miguel

@@
 The treatment of relativization/internalization described in the
-previous sections was enough for the Paulson's treatment of
+previous sections was enough for Paulson's treatment of
 constructibility. This is the case because essentially all the
@@
 \]
-proved under the assumption that $M$ is transitive nonempty. Working
-in this relational 
-way with powersets, cardinalities, and the like would be
-unfeasible. As such, cardinal arithmetic was not put in relative form
-in \session{ZF-Constructible}.
+proved under the assumption that $M$ is transitive and nonempty.
 
-\subsection{Tools for relativization}
+Our first attempt to relativize cardinal arithmetic proceeded in the
+same way
+and we rapidly found out that stating and proving statements like $(||A||
+= |A|) ^M$ in a completely relational language was extremely
+cumbersome. This observation lead to the discovery of the discipline
+expounded in the next subsection.
+
+%% Working
+%% in this relational 
+%% way with powersets, cardinalities, and the like would be
+%% unfeasible. As such, cardinal arithmetic was not put in relative form
+%% in \session{ZF-Constructible}.
+
+\subsection{Discipline and tools for relativization}
 \label{sec:tools-relativization}

### 2022-08-29 Independence_CH/proof_assists_isabelle.tex
First round of corrections with Miguel

@@
 system, but a common interface consists of a display showing the
-current goal and assumptions. The user instructs the assistant to modify the
-latter by means of \emph{tactics}; a %% forward rule changes the context
+current goal and assumptions. The user instructs the assistant to
+modify them by means of \emph{tactics}; a %% forward rule changes the context
 %% with the intention of having in the new context a hypothesis closer to
@@
 To have any value at all, the system should only allow the application
-of sound tactics; in particular, neither the author nor (any part of)
-the system should be able to fake proofs. Edinburgh LCF~\cite{Gordon1979-qm} was an
+of sound tactics.
+%% ; neither the author nor (any part of)
+%% the system should be able to counterfeit proofs.
+Edinburgh LCF~\cite{Gordon1979-qm} was an
 influential proof assistant in which the critical code (that
@@
 
-The metalogic of Isabelle, as well as LCF, is based on higher-order logic.
+The metalogic of Isabelle, as well as that of LCF, is based on higher-order logic.
 %% In contrast with LCF, Isabelle
@@
 \footnote{%
-We recommend the excellent survey \cite{2020arXiv200909541A} by
+We recommend the survey \cite{2020arXiv200909541A} by
 J.~Avigad for details about the different logical foundations in which
@@
 In the case of Isabelle, \emph{Standard} ML is the first of the four layers
-on which we work in this assistant. Both the kernel and the automation
+on which we worked in this assistant. Both the kernel and the automation
 of proofs are coded in ML, sometimes as a substitute for induction on
@@
 concept was manually written in a relational form and relativized.
-Here, the contrast between the usual way one regards $\ZF$ as
+Here, the contrast between the usual way one regards $\ZF$ as a
 first-order theory in the language $\{\in \}$ and the mathematical
@@
 usually assumes a spartan syntax and defines relativization by
-induction of formulas of the more succinct language. The approach
+induction on formulas of the more succinct language. The approach
 taken in \session{ZF-Constructible} is to consider relativizations of
@@
 fixed and assumptions are made. In the case of the constructibility
-library, several locales are defined where the variable $M$ assumed to
+library, several locales are defined where the variable $M$ is assumed to
 denote a class satisfying certain finite amount of $\ZF$; the weakest
 one, \locale{M{\uscore}trans} \cite[Sect.~3]{2020arXiv200109715G}, just
-assumes that $M$ is transitive nonempty. Inside such context, many
+assumes that $M$ is transitive and nonempty. Inside such context, many
 absoluteness results are proved. In order to quote those results for a
@@
 \begin{multline}\label{eq:sats_big_union_fm}
-  x \in \omega \implies y \in \omega \implies \mathit{env} \in \isatt{list}(A)
+  A \in \omega \implies z \in \omega \implies \mathit{env} \in \isatt{list}(M)
   \implies \\
-  A, \mathit{env} \models \isatt{big{\uscore}union{\uscore}fm}(x,y)%\cdot\union x \isatt{ is } y\cdot
+  \bigl(M, \mathit{env} \models \isatt{big{\uscore}union{\uscore}fm}(A,z)\bigr)%\cdot\union x \isatt{ is } y\cdot
   \longleftrightarrow \\
-  \isatt{big{\uscore}union}(\isatt{\#\#} A, \isatt{nth}(x,
-  \mathit{env}), \isatt{nth}(y, \mathit{env}))
+  \isatt{big{\uscore}union}(\isatt{\#\#} M, \isatt{nth}(A,
+  \mathit{env}), \isatt{nth}(z, \mathit{env}))
 \end{multline}
 Above, $\isatt{nth}(x,\mathit{env})$ is the $x$th element of $\mathit{env}$
-and $\isatt{\#\#}A::\tyi\fun \tyo$ is the class corresponding to the
-set $A::\tyi$.
+and $\isatt{\#\#}M::\tyi\fun \tyo$ is the class corresponding to the
+set $M::\tyi$.
 

### 2022-08-30 Independence_CH/intro.tex
Minor edits

@@
 Section~\ref{sec:proof-assist-isabelle}. We kindly invite the reader
-to deepen the previous exposition by reading the apt summary by
+to enrich the previous exposition by reading the apt summary by
 A.~Koutsoukou-Argyraki \cite{angeliki} and the interviews
@@
 involving cardinal successors, countable sets, etc.
-(Section~\ref{sec:extension-isabellezf}); we extended the treatment of relativization of
+(Section~\ref{sec:extension-isabellezf}). We also extended the treatment of relativization of
 set-theoretical concepts (Section~\ref{sec:tools-relativization}).

### 2022-09-12 Independence_CH/forcing.tex
First round of corrections with Miguel
Signaling edits after meeting with Miguel. Proposed new section on replacement instances, and new names in appendix

@@
 and pack assumptions stating that $\lb M, \in\rb$ satisfy some
-axioms. For example, the locale \locale{M{\uscore}Z{\uscore}basic}
+axioms.
+\begin{itemize}
+\item First introduce axioms as satisfactions;
+\item Comment on the convenience of relative forms;
+\item We retain satisfactions for the predicates for Separation and Replacement.
+\end{itemize}
+For example, the locale \locale{M{\uscore}Z{\uscore}basic}
 states that Zermelo set theory holds in $M$. The Union Axiom (\isatt{Union{\uscore}ax}), for
@@
 \[
-  \isatt{Union{\uscore}ax}(\isatt{\#\#}M) \lsii A, [\,]\models \isatt{{\isasymcdot}Union\ Ax{\isasymcdot}}
+  \isatt{Union{\uscore}ax}(\isatt{\#\#}M) \lsii M, [\,]\models \isatt{{\isasymcdot}Union\ Ax{\isasymcdot}}
 \]
@@
 \]
-that consume a class $N$ and predicates $Q::\tyi\fun\tyo$ and $R::\tyi\fun\tyi\fun\tyo$, and state that $N$
+%%%%%%%%%%
+\sout{that consume a class $N$ and predicates $Q::\tyi\fun\tyo$ and $R::\tyi\fun\tyi\fun\tyo$, and state that $N$
 satisfies the $Q$-instance of Separation and the $R$-instance of
-Replacement, respectively. In the statement of the Separation Axiom in
+Replacement, respectively.
+In the statement of the Separation Axiom in
 \locale{M{\uscore}Z{\uscore}basic} and
@@
 defined as the satisfaction of formulas of the appropriate arity; we can only show
-that this form of the axioms hold in generic extensions.
-
-Further locales gather
-particular replacement instances expressed by means of the
+that this form of the axioms hold in generic extensions.}
+%%%%%%%%%%
+whose first argument $N$ is a class and their second arguments $Q$ and
+$R$ are higher-order predicates of types $R :: \tyi \fun \tyo$ and $Q ::
+\tyi \fun \tyi \fun \tyo$, resp.
+The Separation axiom is axiomatized like
+\[
+  \text{This}
+\]
+Particular replacement instances expressed by means of the
 $\isatt{replacement{\uscore}assm}(M,\mathit{env},\phi)$ predicate ($@$ denotes
@@
 \end{isabelle}
+
+In \session{ZF-Constructible} some results are proved assuming some instances of
+separation or replacement (mejor si ponemos ejemplo concreto). In contrast,
+our locale \isatt{M{\uscore}Z{\uscore}basic} assumes the full scheme of separation in terms
+of satisfaction of some internalized formula but no instance of replacement. (Mostrar el enunciado de separation\_ax:
+es distinto a Reemplazo y se nombra después)
+
+In order to have a finer control of the instances of replacement needed,
+we postulate further locales with the instances needed to interpret at
+the class\dots
+(tal vez no sea necesario mostrar la definición de .Replacement.)
+
 %
@@
 
-The locales gathering all 22 replacement instances necessary are named
+\sout{The locales gathering all 22 replacement instances necessary are named
 \locale{M{\uscore}ZF1} through \locale{M{\uscore}ZF3},
@@
 \locale{M{\uscore}ZF{\uscore}ground{\uscore}CH} (see
-Section~\ref{sec:repl-instances} for details).
+Section~\ref{sec:repl-instances} for details).}
 These assumptions are then invoked to interpret at the class
@@
 \session{ZF-Constructible}, and further ones required for the relative
-results from Section~\ref{sec:extension-isabellezf}.
+results from Section~\ref{sec:extension-isabellezf}. See
+Section~\ref{sec:repl-instances} for details.
 
@@
 \footnote{%
-Namely, Zermelo set theory plus the replacement instances included in
-\locale{M{\uscore}ZF1} and \locale{M{\uscore}ZF{\uscore}ground}.
+Namely, Zermelo set theory plus the 7 replacement instances included in
+the locales \locale{M{\uscore}ZF1} and \locale{M{\uscore}ZF{\uscore}ground}.
 }.
@@
 \end{isabelle}
-In order to fix notation, the
-$\val$ interpretation function takes as its first argument the filter
-and then the name, so we have
-$M[G]\defi \{ \val(G,\tau) : \tau\in M \}$.
-
 The version of the Forcing Theorems that we formalized follows the
@@
 \emph{Set Theory}
-\cite[p.~257ff]{kunen2011set}. But in contrast to the point made on
-p.~260 of this book, we internalized the recursion to define the forcing relation,
-in that it involves codes for formulas, and the meta-theoretic formula
-transformer $\phi\mapsto\mathit{Forces}_\phi$ is replaced by the
+\cite[p.~257ff]{kunen2011set}.
+We defined forcing for atomic formulas by recursion on names in an
+analogous fashion. But, in contrast to the point made on
+p.~260 of this book, the structural recursion to define the forcing
+relation was replaced by one  involving codes for formulas. Thus, the meta-theoretic formula
+transformer $\phi\mapsto\mathit{Forces}_\phi$ was replaced by a
 set-theoretic class function $\forceisa:: \tyi \fun \tyi$, which was defined by using
@@
 Next, we state this version of the fundamental theorems in a compact
-way.
+way. For any $G\subseteq P$, our notation for the extension of $M$ by
+$G$ is the  customary one: $M[G]\defi
+\{ \val(G,\tau) : \tau\in M \}$, where the interpretation
+$\val(G,\tau)$ is defined by well-founded recursion on
+$\tau$.
+
 \begin{theorem}\label{th:forcing-thms}
-  There exists a function  $\forceisa$
-  such that for every
+  For every
   $\phi\in\formula$ and $\tau_0,\dots,\tau_n\in M$,
@@
 \end{theorem}
-We actually have to define $\forceisa$ before we state the fundamental
-theorems, so the main existential quantifier above does not appear in the
-formalization.
-Moreover, the items in Theorem~\ref{th:forcing-thms} appear as three
+The items in Theorem~\ref{th:forcing-thms} appear as three
 separated lemmas in the theory
 \theory{Forcing{\uscore}Theorems} of our
-\session{Independence\_CH} session \cite{Independence_CH-AFP},
-and benefit from the $\isatt{map}$ function that applies a function to
-each element of a list. For instance, the Truth Lemma is stated as
+\session{Independence\_CH} session \cite{Independence_CH-AFP}.
+For instance, the Truth Lemma is stated as
 follows:
@@
 \end{isabelle}
+%% where the $\isatt{map}$ function that applies a function to
+%% each element of a list and
 where the $\forces$ notation (and its precedence) had already been set up in the
-\theory{Forces{\uscore}Definition} theory as follows:
-\begin{isabelle}
-\isacommand{abbreviation}\isamarkupfalse%
-\ Forces\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}{\isacharbrackleft}{\kern0pt}i{\isacharcomma}{\kern0pt}\ i{\isacharcomma}{\kern0pt}\ i{\isacharbrackright}{\kern0pt}\ {\isasymRightarrow}\ o{\isachardoublequoteclose}\ \ {\isacharparenleft}{\kern0pt}{\isachardoublequoteopen}{\isacharunderscore}{\kern0pt}\ {\isasymtturnstile}\ {\isacharunderscore}{\kern0pt}\ {\isacharunderscore}{\kern0pt}{\isachardoublequoteclose}\ {\isacharbrackleft}{\kern0pt}{\isadigit{3}}{\isadigit{6}}{\isacharcomma}{\kern0pt}{\isadigit{3}}{\isadigit{6}}{\isacharcomma}{\kern0pt}{\isadigit{3}}{\isadigit{6}}{\isacharbrackright}{\kern0pt}\ {\isadigit{6}}{\isadigit{0}}{\isacharparenright}{\kern0pt}\ \isakeyword{where}\isanewline
-\ \ {\isachardoublequoteopen}p\ {\isasymtturnstile}\ {\isasymphi}\ env\ \ \ {\isasymequiv}\ \ \ M{\isacharcomma}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isacharbrackleft}{\kern0pt}p{\isacharcomma}{\kern0pt}{\isasymbbbP}{\isacharcomma}{\kern0pt}leq{\isacharcomma}{\kern0pt}{\isasymone}{\isacharbrackright}{\kern0pt}\ {\isacharat}{\kern0pt}\ env{\isacharparenright}{\kern0pt}\ {\isasymTurnstile}\ forces{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}
-\end{isabelle}
+\theory{Forces{\uscore}Definition} theory.
 
@@
 but then \cite[p.~257]{kunen2011set} it is cast as a single
-recursively defined function $F$ over a wellfounded  relation $R$.
+recursively defined function $F$ over a well-founded  relation $R$.
 In our formalization, these are called $\frcat$ and 

### 2022-09-12 Independence_CH/main_results.tex
Signaling edits after meeting with Miguel. Proposed new section on replacement instances, and new names in appendix

@@
 
-\subsection{22 replacement instances to rule them all}
+\subsection{A sufficient set of replacement instances}
 \label{sec:repl-instances}
@@
 \isatt{instances1{\uscore}fms} of \formula, consist of basic
-constructions.
+constructions:
 
 \begin{itemize}
-\item 2 instances for transitive closure.
-  \begin{itemize}
-  \item
-    \isatt{eclose{\uscore}repl1{\uscore}intf{\uscore}fm}.
-
-    To prove closure under iteration of $X\mapsto\union X$.
-  \item
-    \isatt{eclose{\uscore}repl2{\uscore}intf{\uscore}fm}.
-
-    Auxiliary instance used to show absoluteness.
-  \end{itemize}
-%% The instances so far
-%% are needed to interpret locale
-%% \locale{M{\uscore}eclose}.
-\item \isatt{wfrec{\uscore}rank{\uscore}fm}.
-
-  For $\in$-rank.
+\item 2 instances for transitive closure: one to prove closure under
+  iteration of $X\mapsto\union X$ and an auxiliary one used to show absoluteness.
+\item 1 instance to define $\in$-rank.
   %
-\item \isatt{trans{\uscore}repl{\uscore}HVFrom{\uscore}fm}.
-
-  For the cumulative hierarchy (rank initial segments).
+\item 1 instance to construct the cumulative hierarchy (rank initial segments).
 \end{itemize}
 
-%% The last two and next pair have the same syntactic structure, because
-%% they are definitions by well-founded recursion.
 The next 4 instances (gathered in \isatt{instances2{\uscore}fms})
 are needed to set up
-cardinal arithmetic in $M$.
+cardinal arithmetic in $M$:  and 2 more for the definition
 \begin{itemize}
-\item 2 instances for ordertypes:
-  \begin{itemize}
-  \item
-    \isatt{wfrec{\uscore}replacement{\uscore}order{\uscore}pred{\uscore}fm}.
-
-    Well-founded recursion for the construction of ordertypes.
-    %
-  \item
-    \isatt{replacement{\uscore}is{\uscore}order{\uscore}eq{\uscore}map{\uscore}fm}.
-
-    Auxiliary instance for the definition of ordertypes.
-  \end{itemize}
-\item 2 instances for Aleph:
-  \begin{itemize}
-    %
-  \item
-    \isatt{replacement{\uscore}is{\uscore}order{\uscore}body{\uscore}fm}.
-
-    Replacement through $x\mapsto \otype(x)$, for Hartogs' Theorem.
-    %
-  \item
-    \isatt{replacement{\uscore}HAleph{\uscore}wfrec{\uscore}repl{\uscore}body{\uscore}fm}.
-
-    The well-founded recursion to define Aleph.
-  \end{itemize}
+\item 2 instances for the definition of
+  ordertypes;
+\item 2 instances for Aleph: Replacement through $x\mapsto
+  \otype(x)$ (for Hartogs' Theorem) and the well-founded recursion
+  using it.
 \end{itemize}
@@
 \isatt{instances3{\uscore}fms}. These are expressed using
-the $\isatt{ground{\uscore}replacement{\uscore}assm}(M,\mathit{env},\phi)$ predicate
+the
+\[
+  \isatt{ground{\uscore}replacement{\uscore}assm}(M,\mathit{env},\phi)
+\]
+predicate
 obtained by replacing $\phi$ by
@@
 require the following 3 instances, which form the set
-\isatt{instances{\uscore}ground{\uscore}fms}.
-
+\isatt{instances{\uscore}ground{\uscore}fms}:
+%
 \begin{itemize}
-\item \isatt{wfrec{\uscore}Hcheck{\uscore}fm}.
-
-  Well-founded recursion to define check-names.
+\item Well-founded recursion to define check-names.
   %
-\item \isatt{wfrec{\uscore}Hfrc{\uscore}at{\uscore}fm}.
-
-  Well-founded recursion for the definition of forcing for atomic formulas.
+\item Well-founded recursion for the definition of forcing for atomic formulas.
   %
-\item
-  \isatt{Lambda{\uscore}in{\uscore}M{\uscore}fm(check{\uscore}fm(2,0,1),1)}.
-
-  Replacement through $x\mapsto \lb x,\check{x}\rb$ (for the
+\item Replacement through $x\mapsto \lb x,\check{x}\rb$ (for the
   definition of $\punto{G}$).
@@
 \end{itemize}
-The next two formulas, which form the set
-\isatt{instances{\uscore}ground{\uscore}notCH{\uscore}fms},
-are needed for the $\Delta$-System Lemma.
-\begin{itemize}
-\item
-  \isatt{replacement{\uscore}is{\uscore}trans{\uscore}apply{\uscore}image{\uscore}fm}.
+The proof of the $\Delta$-System Lemma requires two instances which form the set
+\isatt{instances{\uscore}ground{\uscore}notCH{\uscore}fms}, that are
+used for the recursive construction of sets using a choice function (as in the
+construction of a wellorder of $X$ given a choice function on
+$\Pow(X)$), and to show its absoluteness.
 
-  Recursive construction of sets using a choice function (as in the
-  construction of a wellorder of $X$ given a choice function on $\Pow(X)$).
-  %
-\item
-  \isatt{replacement{\uscore}transrec{\uscore}apply{\uscore}image{\uscore}body{\uscore}fm}.
-
-  Absoluteness of the previous construction.
-\end{itemize}
-%
 The $21$ formulas up to this point are collected into the set
 \isatt{overhead{\uscore}notCH}, which is enough to
-force $\neg\CH$. To force $\CH$, we required one further instance:
-%
-\begin{itemize}
-\item
-  \isatt{replacement{\uscore}dcwit{\uscore}repl{\uscore}body{\uscore}fm}.
-
-  Absoluteness of the recursive construction in the proof of the
+force $\neg\CH$. To force $\CH$, we required one further instance for
+the absoluteness of the recursive construction in the proof of
 Dependent Choices from $\AC$.
-\end{itemize}
 

### 2022-09-12 Independence_CH/non-absolute.tex
Signaling edits after meeting with Miguel. Proposed new section on replacement instances, and new names in appendix

@@
 \label{sec:tools-relativization}
-In order to cope with this, we added the missing step from the
+In order to cope with the inconveniences of the completely relational approach, \sout{we added the missing step from the
 literature consisting of relative versions of the various non-absolute
@@
 define the needed concepts, and  state and
-prove the requisite lemmas. For instance, consider the 
+prove the requisite lemmas} we
+provide automatic tools
+to ease the definitions of relativized versions, their fully relational
+partenaire, and the internalized formulas. For instance, consider the 
 $\isatt{cardinal}::\tyi \fun \tyi$ function defined in

### 2022-09-12 Independence_CH/replacement_instances.tex
Signaling edits after meeting with Miguel. Proposed new section on replacement instances, and new names in appendix

--- 2022-08-26 Independence_CH/main_results.tex
@@
-\section{Main achievements of the formalization}
-\label{sec:main-achievements}
-
-\subsection{22 replacement instances to rule them all}
-\label{sec:repl-instances}
+\section{22 replacement instances to rule them all}
+\label{sec:repl-instances-appendix}
 
@@
   \item
-    \isatt{eclose{\uscore}repl1{\uscore}intf{\uscore}fm}.
+    \sout{\isatt{eclose{\uscore}repl1{\uscore}intf{\uscore}fm}.} \isatt{eclose{\uscore}closed{\uscore}fm}.
 
@@
   \item
-    \isatt{eclose{\uscore}repl2{\uscore}intf{\uscore}fm}.
+    \sout{\isatt{eclose{\uscore}repl2{\uscore}intf{\uscore}fm}} \isatt{eclose{\uscore}abs{\uscore}fm}.
 
@@
   %
-\item \isatt{trans{\uscore}repl{\uscore}HVFrom{\uscore}fm}.
+\item \sout{\isatt{trans{\uscore}repl{\uscore}HVFrom{\uscore}fm}.} \isatt{transrec{\uscore}VFrom{\uscore}fm}.
 
@@
   \item
-    \isatt{wfrec{\uscore}replacement{\uscore}order{\uscore}pred{\uscore}fm}.
+    \sout{\isatt{wfrec{\uscore}replacement{\uscore}order{\uscore}pred{\uscore}fm}.}
+    \isatt{wfrec{\uscore}ordertype{\uscore}fm}.
 
@@
   \item
-    \isatt{replacement{\uscore}is{\uscore}order{\uscore}eq{\uscore}map{\uscore}fm}.
+    \sout{\isatt{replacement{\uscore}is{\uscore}order{\uscore}eq{\uscore}map{\uscore}fm}.}
+    \isatt{omap{\uscore}replacement{\uscore}fm}.
 
@@
   \item
-    \isatt{replacement{\uscore}is{\uscore}order{\uscore}body{\uscore}fm}.
+    \sout{\isatt{replacement{\uscore}is{\uscore}order{\uscore}body{\uscore}fm}.}
+    \isatt{otype{\uscore}replacement{\uscore}fm}.
 
@@
   \item
-    \isatt{replacement{\uscore}HAleph{\uscore}wfrec{\uscore}repl{\uscore}body{\uscore}fm}.
+    \sout{\isatt{replacement{\uscore}HAleph{\uscore}wfrec{\uscore}repl{\uscore}body{\uscore}fm}.}
+    \isatt{wfrec{\uscore}Aleph{\uscore}fm}.
 
@@
 \item
-  \isatt{Lambda{\uscore}in{\uscore}M{\uscore}fm(check{\uscore}fm(2,0,1),1)}.
+  \sout{\isatt{Lambda{\uscore}in{\uscore}M{\uscore}fm(check{\uscore}fm(2,0,1),1)}.}
+  \isatt{lam{\uscore}replacement{\uscore}check{\uscore}fm}.
 
@@
 \item
-  \isatt{replacement{\uscore}is{\uscore}trans{\uscore}apply{\uscore}image{\uscore}fm}.
+  \sout{\isatt{replacement{\uscore}is{\uscore}trans{\uscore}apply{\uscore}image{\uscore}fm}.}
+  \isatt{recursive{\uscore}construction{\uscore}fm}.
 
@@
 \item
-  \isatt{replacement{\uscore}transrec{\uscore}apply{\uscore}image{\uscore}body{\uscore}fm}.
+  \sout{\isatt{replacement{\uscore}transrec{\uscore}apply{\uscore}image{\uscore}body{\uscore}fm}.}
+  \isatt{recursive{\uscore}construction{\uscore}abs{\uscore}fm}.
 
@@
 \item
-  \isatt{replacement{\uscore}dcwit{\uscore}repl{\uscore}body{\uscore}fm}.
+  \sout{\isatt{replacement{\uscore}dcwit{\uscore}repl{\uscore}body{\uscore}fm}.}
+  \isatt{dc{\uscore}abs{\uscore}fm}.
 
@@
 Separation.
-
-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
-\subsection{Models for $\CH$ and its negation}
-\label{sec:models-ch-negation}
-
-The statements of the existence of models of $\ZFC + \neg\CH$ and of
-$\ZFC + \CH$  appear in our formalization as follows:
-
-\begin{isabelle}
-\isacommand{corollary}\isamarkupfalse%
-\ ctm{\isacharunderscore}{\kern0pt}ZFC{\isacharunderscore}{\kern0pt}imp{\isacharunderscore}{\kern0pt}ctm{\isacharunderscore}{\kern0pt}not{\isacharunderscore}{\kern0pt}CH{\isacharcolon}{\kern0pt}\isanewline
-\ \ \isakeyword{assumes}\isanewline
-\ \ \ \ {\isachardoublequoteopen}M\ {\isasymapprox}\ {\isasymomega}{\isachardoublequoteclose}\ {\isachardoublequoteopen}Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ ZFC{\isachardoublequoteclose}\isanewline
-\ \ \isakeyword{shows}\isanewline
-\ \ \ \ {\isachardoublequoteopen}{\isasymexists}N{\isachardot}{\kern0pt}\isanewline
-\ \ \ \ \ \ M\ {\isasymsubseteq}\ N\ {\isasymand}\ N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\ Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\ N\ {\isasymTurnstile}\ ZFC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}{\isasymnot}{\isasymcdot}CH{\isasymcdot}{\isasymcdot}{\isacharbraceright}{\kern0pt}\ {\isasymand}\isanewline
-\ \ \ \ \ \ {\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}
-\end{isabelle}
-
-\begin{isabelle}
-\isacommand{corollary}\isamarkupfalse%
-\ ctm{\isacharunderscore}{\kern0pt}ZFC{\isacharunderscore}{\kern0pt}imp{\isacharunderscore}{\kern0pt}ctm{\isacharunderscore}{\kern0pt}CH{\isacharcolon}{\kern0pt}\isanewline
-\ \ \isakeyword{assumes}\isanewline
-\ \ \ \ {\isachardoublequoteopen}M\ {\isasymapprox}\ {\isasymomega}{\isachardoublequoteclose}\ {\isachardoublequoteopen}Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ ZFC{\isachardoublequoteclose}\isanewline
-\ \ \isakeyword{shows}\isanewline
-\ \ \ \ {\isachardoublequoteopen}{\isasymexists}N{\isachardot}{\kern0pt}\isanewline
-\ \ \ \ \ \ M\ {\isasymsubseteq}\ N\ {\isasymand}\ N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\ Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\ N\ {\isasymTurnstile}\ ZFC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}CH{\isasymcdot}{\isacharbraceright}{\kern0pt}\ {\isasymand}\isanewline
-\ \ \ \ \ \ {\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}
-\end{isabelle}
-where $\approx$ is equipotence, and the predicate \isatt{Transset}
-holds for
-transitive sets. Both results are proved without using Choice.
-
-As the excerpts indicate, these results are obtained as corollaries of
-two theorems in which only a subset of the aforementioned
-replacement instances are assumed of the ground model. We begin the
-discussion of these stronger results by
-considering extensions of ctms of fragments of $\ZF$.
-\begin{isabelle}
-\isacommand{theorem}\isamarkupfalse%
-\ extensions{\isacharunderscore}{\kern0pt}of{\isacharunderscore}{\kern0pt}ctms{\isacharcolon}{\kern0pt}\isanewline
-\ \ \isakeyword{assumes}\isanewline
-\ \ \ \ {\isachardoublequoteopen}M\ {\isasymapprox}\ {\isasymomega}{\isachardoublequoteclose}\ {\isachardoublequoteopen}Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\isanewline
-\ \ \ \ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ {\isasymcdot}Z{\isasymcdot}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}p{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ p\ {\isasymin}\ overhead{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline
-\ \ \ \ {\isachardoublequoteopen}{\isasymPhi}\ {\isasymsubseteq}\ formula{\isachardoublequoteclose}\isanewline%
-\ \ \ \ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ {\isacharbraceleft}{\kern0pt}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}ground{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}fm{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline
-\ \ \isakeyword{shows}\isanewline
-\ \ \ \ {\isachardoublequoteopen}{\isasymexists}N{\isachardot}{\kern0pt}\isanewline
-\ \ \ \ \ \ M\ {\isasymsubseteq}\ N\ {\isasymand}\ N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\ Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\ M{\isasymnoteq}N\ {\isasymand}\isanewline
-\ \ \ \ \ \ {\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
-\ \ \ \ \ \ {\isacharparenleft}{\kern0pt}{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}{\isasymTurnstile}\ {\isasymcdot}AC{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ N{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}AC{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
-\ \ \ \ \ \ N\ {\isasymTurnstile}\ {\isasymcdot}Z{\isasymcdot}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}
-\end{isabelle}
-
-Here, the 7-element set \isatt{overhead} is enough to construct a proper
-extension. It is  the union of
-\isa{instances{\isadigit{1}}{\isacharunderscore}{\kern0pt}fms} and
-\isa{instances{\isacharunderscore}{\kern0pt}ground{\isacharunderscore}{\kern0pt}fms}.
-Also,
-\isatt{{\isasymcdot}Z{\isasymcdot}} denotes Zermelo set theory. In the
-next theorem, the relevant set of formulas is
-\isatt{overhead{\isacharunderscore}{\kern0pt}notCH}, defined above in
-Section~\ref{sec:repl-instances}, and \isatt{ZC} denotes Zermelo set
-theory plus Choice:
-
-\begin{isabelle}
-\isacommand{theorem}\isamarkupfalse%
-\ ctm{\isacharunderscore}{\kern0pt}of{\isacharunderscore}{\kern0pt}not{\isacharunderscore}{\kern0pt}CH{\isacharcolon}{\kern0pt}\isanewline
-\ \ \isakeyword{assumes}\isanewline
-\ \ \ \ {\isachardoublequoteopen}M\ {\isasymapprox}\ {\isasymomega}{\isachardoublequoteclose}\ {\isachardoublequoteopen}Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\isanewline
-\ \ \ \ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ ZC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}p{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ p\ {\isasymin}\ overhead{\isacharunderscore}{\kern0pt}notCH{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline
-\ \ \ \ {\isachardoublequoteopen}{\isasymPhi}\ {\isasymsubseteq}\ formula{\isachardoublequoteclose}\isanewline
-\ \ \ \ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ {\isacharbraceleft}{\kern0pt}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}ground{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}fm{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline
-\ \ \isakeyword{shows}\isanewline
-\ \ \ \ {\isachardoublequoteopen}{\isasymexists}N{\isachardot}{\kern0pt}\isanewline
-\ \ \ \ \ \ M\ {\isasymsubseteq}\ N\ {\isasymand}\ N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\ Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
-\ \ \ \ \ \ N\ {\isasymTurnstile}\ ZC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}{\isasymnot}{\isasymcdot}CH{\isasymcdot}{\isasymcdot}{\isacharbraceright}{\kern0pt}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}\ {\isasymand}\isanewline
-\ \ \ \ \ \ {\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}
-\end{isabelle}
-
-Finally, \isatt{overhead{\isacharunderscore}{\kern0pt}CH} is the union
-of \isatt{overhead{\isacharunderscore}{\kern0pt}CH} with the $\DC$
-instance \isatt{replacement{\isacharunderscore}{\kern0pt}dcwit{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}body{\isacharunderscore}{\kern0pt}fm}:
-\begin{isabelle}
-\isacommand{theorem}\isamarkupfalse%
-\ ctm{\isacharunderscore}{\kern0pt}of{\isacharunderscore}{\kern0pt}CH{\isacharcolon}{\kern0pt}\isanewline
-\ \ \isakeyword{assumes}\isanewline
-\ \ \ \ {\isachardoublequoteopen}M\ {\isasymapprox}\ {\isasymomega}{\isachardoublequoteclose}\ {\isachardoublequoteopen}Transset{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\isanewline
-\ \ \ \ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ ZC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}p{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ p\ {\isasymin}\ overhead{\isacharunderscore}{\kern0pt}CH{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline
-\ \ \ \ {\isachardoublequoteopen}{\isasymPhi}\ {\isasymsubseteq}\ formula{\isachardoublequoteclose}\isanewline
-\ \ \ \ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ {\isacharbraceleft}{\kern0pt}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}ground{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}fm{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline
-\ \ \isakeyword{shows}\isanewline
-\ \ \ \ {\isachardoublequoteopen}{\isasymexists}N{\isachardot}{\kern0pt}\isanewline
-\ \ \ \ \ \ M\ {\isasymsubseteq}\ N\ {\isasymand}\ N\ {\isasymapprox}\ {\isasymomega}\ {\isasymand}\ Transset{\isacharparenleft}{\kern0pt}N{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
-\ \ \ \ \ \ N\ {\isasymTurnstile}\ ZC\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}{\isasymcdot}CH{\isasymcdot}{\isacharbraceright}{\kern0pt}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}\ {\isachardot}{\kern0pt}\ {\isasymphi}\ {\isasymin}\ {\isasymPhi}{\isacharbraceright}{\kern0pt}\ {\isasymand}\isanewline
-\ \ \ \ \ \ {\isacharparenleft}{\kern0pt}{\isasymforall}{\isasymalpha}{\isachardot}{\kern0pt}\ Ord{\isacharparenleft}{\kern0pt}{\isasymalpha}{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}
-\end{isabelle}
-
-%%% Local Variables: 
-%%% mode: latex
-%%% TeX-master: "independence_ch_isabelle"
-%%% ispell-local-dictionary: "american"
-%%% End:

### 2022-09-12 Independence_CH/sample_proof.tex
Minor edits
Signaling edits after meeting with Miguel. Proposed new section on replacement instances, and new names in appendix

@@
   consider any $c \in \mathcal{P}(a) \cap M[G]$; we need to show that
-  $c \in b$. Fix $\varkappa \in M^{\mathbb{P}}$ such that
-  $\varkappa_{G}=c$, and let $\vartheta=\{\langle\sigma, p\rangle:
+  $c \in b$. Fix $\chi \in M^{\mathbb{P}}$ such that
+  $\chi_{G}=c$, and let $\vartheta=\{\langle\sigma, p\rangle:
   \sigma \in \operatorname{dom}(\tau) \wedge p \Vdash \sigma \in
-  \varkappa\}$; $\vartheta \in M$ by the Definability Lemma. Since
+  \chi\}$; $\vartheta \in M$ by the Definability Lemma. Since
   $\vartheta \in Q$, we are done if we can show that
@@
   M^{\mathbb{P}}$} [\dots]---it is pretty laborious to show that things
-are in $M$; it takes 17 more lines of code (not shown below) that
-apply further previously proved lemmas.
+are in $M$; we omit 17 lines of code to that effect,
+that also apply previously proved lemmas.
 \begin{isabelle}
@@
 \end{isabelle}
-\textit{Fix $\varkappa \in M^{\mathbb{P}}$ such that
-  $\varkappa_{G}=c$,}
+\textit{Fix $\chi \in M^{\mathbb{P}}$ such that
+  $\chi_{G}=c$,}
 \begin{isabelle}
@@
   \sigma \in \operatorname{dom}(\tau) \wedge p \Vdash \sigma \in
-  \varkappa\}$;}
+  \chi\}$;}
 \begin{isabelle}
@@
 \ \ \ \ \isacommand{have}\isamarkupfalse%
-\ {\isachardoublequoteopen}val{\isacharparenleft}{\kern0pt}G{\isacharcomma}{\kern0pt}{\isacharquery}{\kern0pt}{\isasymtheta}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ c{\isachardoublequoteclose}
+\ {\isachardoublequoteopen}val{\isacharparenleft}{\kern0pt}G{\isacharcomma}{\kern0pt}{\isacharquery}{\kern0pt}{\isasymtheta}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ c{\isachardoublequoteclose}\isanewline
+%
+\ \ \ \ \isacommand{proof}\isamarkupfalse%
+\ {\isacharminus}{\kern0pt}\isanewline
+\ \ \ \ \ \ \dots\isanewline
+%
+\ \ \ \ \isacommand{qed}\isamarkupfalse%
 \end{isabelle}

### 2022-09-12 Independence_CH/main_results.tex
Minor edits

@@
 are needed to set up
-cardinal arithmetic in $M$:  and 2 more for the definition
+cardinal arithmetic in $M$:
 \begin{itemize}
@@
 \end{itemize}
-The proof of the $\Delta$-System Lemma requires two instances which form the set
+The proof of the $\Delta$-System Lemma requires 2 instances which form the set
 \isatt{instances{\uscore}ground{\uscore}notCH{\uscore}fms}, that are

### 2022-09-12 Independence_CH/forcing.tex
Arity of phi and reindexing of env at the fundamental theorem

@@
 Next, we state this version of the fundamental theorems in a compact
-way. For any $G\subseteq P$, our notation for the extension of $M$ by
+way. For any $G\subseteq \PP$, our notation for the extension of $M$ by
 $G$ is the  customary one: $M[G]\defi
@@
   For every
-  $\phi\in\formula$ and $\tau_0,\dots,\tau_n\in M$,
+  $\phi\in\formula$ with $\isatt{arity}(\phi)\leq n$ and $\tau_1,\dots,\tau_n\in M$,
   \begin{enumerate}
@@
   arity of $\forceisa(\phi)$ is at most $\isatt{arity}(\phi) + 4$; and if
-  “$p \forces \phi\ [\tau_0,\dots,\tau_n]$”
+  “$p \forces \phi\ [\tau_1,\dots,\tau_n]$”
   denotes
-  “$M, [p,\PP,\preceq,\1, \tau_0,\dots,\tau_n]  \models
+  “$M, [p,\PP,\preceq,\1, \tau_1,\dots,\tau_n]  \models
   \forceisa(\phi)$”, we have:
@@
     \[
-      \exists p\in G.\ \; p \forces \phi\ [\tau_0,\dots,\tau_n]
+      \exists p\in G.\ \; p \forces \phi\ [\tau_1,\dots,\tau_n]
     \]
@@
     \[
-      M[G], [\val(G,\tau_0),\dots,\val(G,\tau_n)]
+      M[G], [\val(G,\tau_1),\dots,\val(G,\tau_n)]
       \models \phi.
     \]
-  \item \label{item:density-lemma} (Density Lemma) $p \forces \phi\ [\tau_0,\dots,\tau_n]$
+  \item \label{item:density-lemma} (Density Lemma) $p \forces \phi\ [\tau_1,\dots,\tau_n]$
     if and only if 
-    $\{q\in \PP :  q \forces \phi\ [\tau_0,\dots,\tau_n]\}$
+    $\{q\in \PP :  q \forces \phi\ [\tau_1,\dots,\tau_n]\}$
     is dense below $p$.

### 2022-09-12 Independence_CH/intro.tex
Arity of phi and reindexing of env at the fundamental theorem

@@
 Conjecture \cite{MR3659768}. There is a vast mathematical corpus at
-the Archive of Formal Proofs (AFP) based on Isabelle, and formalizations of
+the Archive of Formal Proofs (AFP) based on Isabelle; and formalizations of
 brand new and complex objects like the definition of perfectoid spaces \cite{10.1145/3372885.3373830}
-and the Liquid Tensor Experiment \cite{LTE2020,LTE2021} on Lean.
+and the Liquid Tensor Experiment \cite{LTE2020,LTE2021} have been
+achieved using Lean.
 

### 2022-09-12 Independence_CH/discipline.tex
Appendix on lam_replacements. Macro for locales
Bye isatt, bye slanted

@@
-\appendix
-
 \section{Discipline for relativization}
+\label{sec:discipline-relativization}
 
@@
 Going back to our example in \ref{sec:tools-relativization}, for the
-concept $\isatt{cardinal}::\tyi \fun \tyi$ we want its relative
+concept $\isa{cardinal}::\tyi \fun \tyi$ we want its relative
 version
-$\isatt{cardinal{\uscore}rel}::(\tyi \fun \tyo) \fun \tyi \fun\tyi$
+$\isa{cardinal{\uscore}rel}::(\tyi \fun \tyo) \fun \tyi \fun\tyi$
 and the relational version of the latter
-$\isatt{is{\uscore}cardinal}::(\tyi \fun \tyo) \fun \tyi \fun \tyi
+$\isa{is{\uscore}cardinal}::(\tyi \fun \tyo) \fun \tyi \fun \tyi
 \fun \tyo$.
@@
 mathematical considerations: if we might prove that
-$\isatt{is{\uscore}cardinal}$ is functional and also prove the
-existence of a witness $\isatt{c}$ such that $\isatt{M(c)}$ and
-$\isatt{is{\uscore}cardinal(M,x,c)}$ then
-$\isatt{cardinal{\uscore}rel(M,x)}$ can be obtained by the operator of
+$\isa{is{\uscore}cardinal}$ is functional and also prove the
+existence of a witness $\isa{c}$ such that $\isa{M(c)}$ and
+$\isa{is{\uscore}cardinal(M,x,c)}$ then
+$\isa{cardinal{\uscore}rel(M,x)}$ can be obtained by the operator of
 definite descriptions.
@@
 relativized concept, we can use it to define other relativizations.
-For instance, $\isatt{cardinal{\uscore}rel}$ depends on having
-relative versions of $\isatt{bij}$. Instead of relationalizing
-$\isatt{bij}$ to get $\isatt{is{\uscore}bij}$ and then prove
+For instance, $\isa{cardinal{\uscore}rel}$ depends on having
+relative versions of $\isa{bij}$. Instead of relationalizing
+$\isa{bij}$ to get $\isa{is{\uscore}bij}$ and then prove
 uniqueness and existence of a witness, we define
-$\isatt{bij{\uscore}rel}$ using $\isatt{inj{\uscore}rel}$ and
-$\isatt{surj{\uscore}rel}$.
+$\isa{bij{\uscore}rel}$ using $\isa{inj{\uscore}rel}$ and
+$\isa{surj{\uscore}rel}$.
 

### 2022-09-12 Independence_CH/forcing.tex
Bye isatt, bye slanted

@@
 For example, the locale \locale{M{\uscore}Z{\uscore}basic}
-states that Zermelo set theory holds in $M$. The Union Axiom (\isatt{Union{\uscore}ax}), for
+states that Zermelo set theory holds in $M$. The Union Axiom (\isa{Union{\uscore}ax}), for
 instance, is defined as follows:
 \[
-\forall x[\isatt{\#\#}M].\ \exists z[\isatt{\#\#}M].\ \isatt{big{\uscore}union}(\isatt{\#\#}M,x,z)
+\forall x[\isa{\#\#}M].\ \exists z[\isa{\#\#}M].\ \isa{big{\uscore}union}(\isa{\#\#}M,x,z)
 \]
@@
 \[
-  \isatt{Union{\uscore}ax}(\isatt{\#\#}M) \lsii M, [\,]\models \isatt{{\isasymcdot}Union\ Ax{\isasymcdot}}
+  \isa{Union{\uscore}ax}(\isa{\#\#}M) \lsii M, [\,]\models \isa{{\isasymcdot}Union\ Ax{\isasymcdot}}
 \]
@@
 %% \end{isabelle}
-where \isatt{{\isasymcdot}Union\ Ax{\isasymcdot}} is the $\formula$ code for the
+where \isa{{\isasymcdot}Union\ Ax{\isasymcdot}} is the $\formula$ code for the
 Union axiom. For the axiom schemes, \session{ZF-Constructible} defines
@@
 \[
-  \isatt{separation}(N,Q)
+  \isa{separation}(N,Q)
   \text{ and }
-  \isatt{strong{\uscore}replacement}(N,R)
+  \isa{strong{\uscore}replacement}(N,R)
 \]
@@
 Particular replacement instances expressed by means of the
-$\isatt{replacement{\uscore}assm}(M,\mathit{env},\phi)$ predicate ($@$ denotes
+$\isa{replacement{\uscore}assm}(M,\mathit{env},\phi)$ predicate ($@$ denotes
 list concatenation):
 \begin{multline}\label{eq:replacement_assm_def}
-\varphi \in \formula  \limp \mathit{env} \in \isatt{list}(M) \limp \isatt{arity}(\varphi) \leq 2+ \isatt{length}(\mathit{env}) \limp \\
- \isatt{strong{\uscore}replacement}(\isatt{\#\#} M, \lambda x\, y.\ (M, [x,y]
+\varphi \in \formula  \limp \mathit{env} \in \isa{list}(M) \limp \isa{arity}(\varphi) \leq 2+ \isa{length}(\mathit{env}) \limp \\
+ \isa{strong{\uscore}replacement}(\isa{\#\#} M, \lambda x\, y.\ (M, [x,y]
 \mathbin{@} \mathit{env}  \models \varphi))
@@
 %% \end{isabelle}
-In turn, the \isatt{{\isasymcdot}Replacement{\isasymcdot}} function
+In turn, the \isa{{\isasymcdot}Replacement{\isasymcdot}} function
 takes a formula code and returns the corresponding replacement
@@
 separation or replacement (mejor si ponemos ejemplo concreto). In contrast,
-our locale \isatt{M{\uscore}Z{\uscore}basic} assumes the full scheme of separation in terms
+our locale \isa{M{\uscore}Z{\uscore}basic} assumes the full scheme of separation in terms
 of satisfaction of some internalized formula but no instance of replacement. (Mostrar el enunciado de separation\_ax:
@@
 These assumptions are then invoked to interpret at the class
-$\isatt{\#\#} M$ the relevant locales appearing in
+$\isa{\#\#} M$ the relevant locales appearing in
 \session{ZF-Constructible}, and further ones required for the relative
@@
 \subsection{The fundamental theorems}\label{sec:fundamental-theorems}
-At this point, we work inside the locale \isatt{M{\uscore}ctm1} that
+At this point, we work inside the locale \isa{M{\uscore}ctm1} that
 assumes $M$ to be
@@
   For every
-  $\phi\in\formula$ with $\isatt{arity}(\phi)\leq n$ and $\tau_1,\dots,\tau_n\in M$,
+  $\phi\in\formula$ with $\isa{arity}(\phi)\leq n$ and $\tau_1,\dots,\tau_n\in M$,
   \begin{enumerate}
@@
   where the 
-  arity of $\forceisa(\phi)$ is at most $\isatt{arity}(\phi) + 4$; and if
+  arity of $\forceisa(\phi)$ is at most $\isa{arity}(\phi) + 4$; and if
   “$p \forces \phi\ [\tau_1,\dots,\tau_n]$”
@@
 \end{isabelle}
-%% where the $\isatt{map}$ function that applies a function to
+%% where the $\isa{map}$ function that applies a function to
 %% each element of a list and
@@
 In our formalization, these are called $\frcat$ and 
-$\isatt{frecR}$, respectively, and are defined on tuples $\lb \mathit{ft},t_1,t_2,p\rb$ (where
+$\isa{frecR}$, respectively, and are defined on tuples $\lb \mathit{ft},t_1,t_2,p\rb$ (where
 $\mathit{ft}\in\{0,1\}$ indicates whether the atomic formula being

### 2022-09-12 Independence_CH/lessons.tex
Signaling edits after meeting with Miguel. Proposed new section on replacement instances, and new names in appendix
Bye isatt, bye slanted

@@
 box. (The latter change seems to be rather straightforward, but the
-former does not.)
+former does not.)\\
+\fbox{Note that this is a consequence of following
+  Kunen's blueprint for the formalization}
 
@@
 model; this is the case of
-\isatt{definition{\uscore}of{\uscore}forces} and the result on forcing
+\isa{definition{\uscore}of{\uscore}forces} and the result on forcing
 values of a function.
@@
 a large project like the present one. For example, it is
-(mathematically) misleading when automatic tools (\isatt{simp},
-\isatt{auto}, etc) stop working just because of the sheer size of the
+(mathematically) misleading when automatic tools (\isa{simp},
+\isa{auto}, etc) stop working just because of the sheer size of the
 goal (v.g., the same statement with 7 variables succeeds but with 8

### 2022-09-12 Independence_CH/main_defs.tex
Bye isatt, bye slanted

@@
 cardinal arithmetic as implemented in
-\isatt{T{\kern0pt}r{\kern0pt}a{\kern0pt}n{\kern0pt}s{\kern0pt}i{\kern0pt}t{\kern0pt}i{\kern0pt}v{\kern0pt}e{\kern0pt}{\char`\_}{\kern0pt}M{\kern0pt}o{\kern0pt}d{\kern0pt}e{\kern0pt}l{\kern0pt}s{\kern0pt}}.
+\isa{T{\kern0pt}r{\kern0pt}a{\kern0pt}n{\kern0pt}s{\kern0pt}i{\kern0pt}t{\kern0pt}i{\kern0pt}v{\kern0pt}e{\kern0pt}{\char`\_}{\kern0pt}M{\kern0pt}o{\kern0pt}d{\kern0pt}e{\kern0pt}l{\kern0pt}s{\kern0pt}}.
 But in case one wants to dive deeper, the
@@
 “complete”. (Mind the
-\isatt{\isasymlbrakk P; Q\isasymrbrakk\ \isasymLongrightarrow\ R}
+\isa{\isasymlbrakk P; Q\isasymrbrakk\ \isasymLongrightarrow\ R}
 abbreviation for
-\isatt{P\ \isasymLongrightarrow\ Q\ \isasymLongrightarrow\ R}.)
+\isa{P\ \isasymLongrightarrow\ Q\ \isasymLongrightarrow\ R}.)
 
@@
 \end{isabelle}%
-The \isatt{arity} function strictly bounding the free de Bruijn
+The \isa{arity} function strictly bounding the free de Bruijn
 indices of a formula is defined below:
@@
 first-order formula \isa{{\isasymcdot}CH{\isasymcdot}} (since it
-actually is a sentence, it does not depend on \isatt{env}, which
+actually is a sentence, it does not depend on \isa{env}, which
 appears only because the definition of $\models$ requires that argument).%

### 2022-09-12 Independence_CH/main_results.tex
Bye isatt, bye slanted

@@
 The first 4 instances, collected in the subset
-\isatt{instances1{\uscore}fms} of \formula, consist of basic
+\isa{instances1{\uscore}fms} of \formula, consist of basic
 constructions:
@@
 
-The next 4 instances (gathered in \isatt{instances2{\uscore}fms})
+The next 4 instances (gathered in \isa{instances2{\uscore}fms})
 are needed to set up
@@
 In our development, the mapping $\phi\mapsto\psi$ defined above is given by the
-$\isatt{ground{\uscore}repl{\uscore}fm}$ function, and all ground replacement
+$\isa{ground{\uscore}repl{\uscore}fm}$ function, and all ground replacement
 instances appear in the locale \locale{M{\uscore}ZF3} and form the set
-\isatt{instances3{\uscore}fms}. These are expressed using
+\isa{instances3{\uscore}fms}. These are expressed using
 the
 \[
-  \isatt{ground{\uscore}replacement{\uscore}assm}(M,\mathit{env},\phi)
+  \isa{ground{\uscore}replacement{\uscore}assm}(M,\mathit{env},\phi)
 \]
@@
 obtained by replacing $\phi$ by
-$\isatt{ground{\uscore}repl{\uscore}fm}(\phi)$ in Eq.~(\ref{eq:replacement_assm_def}).
+$\isa{ground{\uscore}repl{\uscore}fm}(\phi)$ in Eq.~(\ref{eq:replacement_assm_def}).
 
@@
 require the following 3 instances, which form the set
-\isatt{instances{\uscore}ground{\uscore}fms}:
+\isa{instances{\uscore}ground{\uscore}fms}:
 %
@@
 The proof of the $\Delta$-System Lemma requires 2 instances which form the set
-\isatt{instances{\uscore}ground{\uscore}notCH{\uscore}fms}, that are
+\isa{instances{\uscore}ground{\uscore}notCH{\uscore}fms}, that are
 used for the recursive construction of sets using a choice function (as in the
@@
 The $21$ formulas up to this point are collected into the set
-\isatt{overhead{\uscore}notCH}, which is enough to
+\isa{overhead{\uscore}notCH}, which is enough to
 force $\neg\CH$. To force $\CH$, we required one further instance for
@@
 at least one extra instance
-% (\isatt{banach{\uscore}iterates{\uscore}fm})
+% (\isa{banach{\uscore}iterates{\uscore}fm})
 arising from an iteration. We also managed to avoid 12 further
@@
 require any extra replacement; actually, they only need the 7
-instances appearing in \isatt{instances1{\uscore}fms} and
-\isatt{instances{\uscore}ground{\uscore}fms}.  But this seems not be
+instances appearing in \isa{instances1{\uscore}fms} and
+\isa{instances{\uscore}ground{\uscore}fms}.  But this seems not be
 the case for Separation, at least by inspecting our formalization:
@@
 in the proof of Theorem~\ref{th:forcing-thms}(\ref{item:truth-lemma}),
-that appears as the \isatt{truth{\uscore}lemma} in our development; it
-depends on \isatt{truth{\uscore}lemma'} and
-\isatt{truth{\uscore}lemma{\uscore}Neg}, which explicitly invoke
-\isatt{separation{\uscore}ax}. In any case, our intended grounds
+that appears as the \isa{truth{\uscore}lemma} in our development; it
+depends on \isa{truth{\uscore}lemma'} and
+\isa{truth{\uscore}lemma{\uscore}Neg}, which explicitly invoke
+\isa{separation{\uscore}ax}. In any case, our intended grounds
 (v.g., the transitive collapse of countable elementary submodels of a
@@
 \end{isabelle}
-where $\approx$ is equipotence, and the predicate \isatt{Transset}
+where $\approx$ is equipotence, and the predicate \isa{Transset}
 holds for
@@
 
-Here, the 7-element set \isatt{overhead} is enough to construct a proper
+Here, the 7-element set \isa{overhead} is enough to construct a proper
 extension. It is  the union of
@@
 Also,
-\isatt{{\isasymcdot}Z{\isasymcdot}} denotes Zermelo set theory. In the
+\isa{{\isasymcdot}Z{\isasymcdot}} denotes Zermelo set theory. In the
 next theorem, the relevant set of formulas is
-\isatt{overhead{\isacharunderscore}{\kern0pt}notCH}, defined above in
-Section~\ref{sec:repl-instances}, and \isatt{ZC} denotes Zermelo set
+\isa{overhead{\isacharunderscore}{\kern0pt}notCH}, defined above in
+Section~\ref{sec:repl-instances}, and \isa{ZC} denotes Zermelo set
 theory plus Choice:
@@
 
-Finally, \isatt{overhead{\isacharunderscore}{\kern0pt}CH} is the union
-of \isatt{overhead{\isacharunderscore}{\kern0pt}CH} with the $\DC$
-instance \isatt{replacement{\isacharunderscore}{\kern0pt}dcwit{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}body{\isacharunderscore}{\kern0pt}fm}:
+Finally, \isa{overhead{\isacharunderscore}{\kern0pt}CH} is the union
+of \isa{overhead{\isacharunderscore}{\kern0pt}CH} with the $\DC$
+instance \isa{replacement{\isacharunderscore}{\kern0pt}dcwit{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}body{\isacharunderscore}{\kern0pt}fm}:
 \begin{isabelle}

### 2022-09-12 Independence_CH/non-absolute.tex
Bye isatt, bye slanted

@@
 could be minimized after proving the relevant absoluteness results:
-For example, the lemma \isatt{Union{\uscore}abs},
+For example, the lemma \isa{Union{\uscore}abs},
 \[
-  M(A) \implies M(z) \implies \isatt{big{\uscore}union}(M, A, z) \longleftrightarrow z = \union
+  M(A) \implies M(z) \implies \isa{big{\uscore}union}(M, A, z) \longleftrightarrow z = \union
   A
@@
 partenaire, and the internalized formulas. For instance, consider the 
-$\isatt{cardinal}::\tyi \fun \tyi$ function defined in
+$\isa{cardinal}::\tyi \fun \tyi$ function defined in
 \session{Isabelle/ZF}. Then the commands
@@
 define the relative cardinal function
-$\isatt{cardinal{\uscore}rel}::(\tyi \fun \tyo) \fun \tyi \fun\tyi$
+$\isa{cardinal{\uscore}rel}::(\tyi \fun \tyo) \fun \tyi \fun\tyi$
 (denoted  $|\cdot|^M$, as expected),
-the relational version $\isatt{is{\uscore}cardinal}$ of the latter, the
-internalized formula \isatt{is{\uscore}cardinal{\uscore}fm} whose
+the relational version $\isa{is{\uscore}cardinal}$ of the latter, the
+internalized formula \isa{is{\uscore}cardinal{\uscore}fm} whose
 satisfaction by a set is equivalent to the relational version, and
 prove the previous statement (analogous to (\ref{eq:sats_big_union_fm})).
-The proof that $\isatt{is{\uscore}cardinal}(M,x,z)$  encodes the
+The proof that $\isa{is{\uscore}cardinal}(M,x,z)$  encodes the
 statement $|x|^M = z$ must still be done by hand, since the definition
-of $\isatt{cardinal{\uscore}rel}$ already involves some tacit
+of $\isa{cardinal{\uscore}rel}$ already involves some tacit
 absoluteness results (“\textit{the least $z \in \Ord$ such that $z

### 2022-09-12 Independence_CH/proof_assists_isabelle.tex
Bye isatt, bye slanted

@@
 with a predicate for membership, constants for the empty set and an
-infinite set, and functions $\isatt{Pow}::\tyi\fun\tyi$,
-$\union::\tyi\fun\tyi$, and $\isatt{PrimReplace} :: \tyi \fun (\tyi
+infinite set, and functions $\isa{Pow}::\tyi\fun\tyi$,
+$\union::\tyi\fun\tyi$, and $\isa{PrimReplace} :: \tyi \fun (\tyi
 \fun \tyi \fun\tyo)\fun \tyi$ (for Replacement). The Axiom of
@@
   \longrightarrow y=z) \implies \\
-  b \in \isatt{PrimReplace}(A, P)
+  b \in \isa{PrimReplace}(A, P)
   \longleftrightarrow(\exists x \in A .\ P(x, b)) 
@@
 The development of relativization and absoluteness for classes $C::
-\isatt{i} \fun \isatt{o}$ follows the same pattern. Each particular
+\isa{i} \fun \isa{o}$ follows the same pattern. Each particular
 concept was manually written in a relational form and relativized.
@@
 %https://isabelle.in.tum.de/dist/library/ZF/ZF-Constructible/Relative.html#Relative.big_union_def_raw|axiom
-$\isatt{big{\uscore}union}:: (\isatt{i} \fun \isatt{o}) \fun \isatt{i}
-\fun \isatt{i} \fun \isatt{o}$ of the statement
+$\isa{big{\uscore}union}:: (\isa{i} \fun \isa{o}) \fun \isa{i}
+\fun \isa{i} \fun \isa{o}$ of the statement
 “$\union A = z$”:
 \[
- \isatt{big{\uscore}union}(M,A,z) \equiv \forall x[M].\ x \in z
+ \isa{big{\uscore}union}(M,A,z) \equiv \forall x[M].\ x \in z
  \longleftrightarrow (\exists y[M].\ y \in A \land x \in y)
@@
 (this is in constrast to general well-founded recursion, for which
-one has to work with the fundamental recursor $\isatt{wfrec}$).}  Its
+one has to work with the fundamental recursor $\isa{wfrec}$).}  Its
 atomic formulas have the form $\cdot x \in y\cdot$ and $\cdot x =
@@
 $z\in\omega$, $z$ is free in $\phi$ if it occurs under at most $z$
-quantifiers. The \isatt{arity} function on $\phi$ is one plus the
+quantifiers. The \isa{arity} function on $\phi$ is one plus the
 maximum free index occurring in $\phi$.
@@
 The satisfaction predicate
-$\isatt{sats}::\tyi\fun\tyi\fun\tyi\fun\tyo$ takes as arguments a set
-$M$, a list $\mathit{env}\in\isatt{list}(M)$ for the assignment of
+$\isa{sats}::\tyi\fun\tyi\fun\tyi\fun\tyo$ takes as arguments a set
+$M$, a list $\mathit{env}\in\isa{list}(M)$ for the assignment of
 free indices, and $\phi\in\formula$, and it is written
@@
 %%   $M,\mathit{env}\models\phi$ is only meaningful if
-%%   $\isatt{length}(\mathit{env}) \geqslant
-%% \isatt{arity}(\phi)$.\textbf{$\leftarrow$ necessary?}}
+%%   $\isa{length}(\mathit{env}) \geqslant
+%% \isa{arity}(\phi)$.\textbf{$\leftarrow$ necessary?}}
 This completes the
@@
 \begin{multline*}
-  \isatt{big{\uscore}union{\uscore}fm}(A,z) \equiv \\
+  \isa{big{\uscore}union{\uscore}fm}(A,z) \equiv \\
   (({\cdot}\forall{\cdot}\,{\cdot}0 \in \mathit{succ}(z){\cdot} \longleftrightarrow
@@
 \begin{multline}\label{eq:sats_big_union_fm}
-  A \in \omega \implies z \in \omega \implies \mathit{env} \in \isatt{list}(M)
+  A \in \omega \implies z \in \omega \implies \mathit{env} \in \isa{list}(M)
   \implies \\
-  \bigl(M, \mathit{env} \models \isatt{big{\uscore}union{\uscore}fm}(A,z)\bigr)%\cdot\union x \isatt{ is } y\cdot
+  \bigl(M, \mathit{env} \models \isa{big{\uscore}union{\uscore}fm}(A,z)\bigr)%\cdot\union x \isa{ is } y\cdot
   \longleftrightarrow \\
-  \isatt{big{\uscore}union}(\isatt{\#\#} M, \isatt{nth}(A,
-  \mathit{env}), \isatt{nth}(z, \mathit{env}))
+  \isa{big{\uscore}union}(\isa{\#\#} M, \isa{nth}(A,
+  \mathit{env}), \isa{nth}(z, \mathit{env}))
 \end{multline}
-Above, $\isatt{nth}(x,\mathit{env})$ is the $x$th element of $\mathit{env}$
-and $\isatt{\#\#}M::\tyi\fun \tyo$ is the class corresponding to the
+Above, $\isa{nth}(x,\mathit{env})$ is the $x$th element of $\mathit{env}$
+and $\isa{\#\#}M::\tyi\fun \tyo$ is the class corresponding to the
 set $M::\tyi$.

### 2022-09-12 Independence_CH/replacement_instances.tex
Bye isatt, bye slanted

@@
 The first 4 instances, collected in the subset
-\isatt{instances1{\uscore}fms} of \formula, consist of basic
+\isa{instances1{\uscore}fms} of \formula, consist of basic
 constructions.
@@
   \item
-    \sout{\isatt{eclose{\uscore}repl1{\uscore}intf{\uscore}fm}.} \isatt{eclose{\uscore}closed{\uscore}fm}.
+    \sout{\isa{eclose{\uscore}repl1{\uscore}intf{\uscore}fm}.} \isa{eclose{\uscore}closed{\uscore}fm}.
 
@@
   \item
-    \sout{\isatt{eclose{\uscore}repl2{\uscore}intf{\uscore}fm}} \isatt{eclose{\uscore}abs{\uscore}fm}.
+    \sout{\isa{eclose{\uscore}repl2{\uscore}intf{\uscore}fm}} \isa{eclose{\uscore}abs{\uscore}fm}.
 
@@
 %% \locale{M{\uscore}eclose}.
-\item \isatt{wfrec{\uscore}rank{\uscore}fm}.
+\item \isa{wfrec{\uscore}rank{\uscore}fm}.
 
@@
   %
-\item \sout{\isatt{trans{\uscore}repl{\uscore}HVFrom{\uscore}fm}.} \isatt{transrec{\uscore}VFrom{\uscore}fm}.
+\item \sout{\isa{trans{\uscore}repl{\uscore}HVFrom{\uscore}fm}.} \isa{transrec{\uscore}VFrom{\uscore}fm}.
 
@@
 %% they are definitions by well-founded recursion.
-The next 4 instances (gathered in \isatt{instances2{\uscore}fms})
+The next 4 instances (gathered in \isa{instances2{\uscore}fms})
 are needed to set up
@@
   \item
-    \sout{\isatt{wfrec{\uscore}replacement{\uscore}order{\uscore}pred{\uscore}fm}.}
-    \isatt{wfrec{\uscore}ordertype{\uscore}fm}.
+    \sout{\isa{wfrec{\uscore}replacement{\uscore}order{\uscore}pred{\uscore}fm}.}
+    \isa{wfrec{\uscore}ordertype{\uscore}fm}.
 
@@
   \item
-    \sout{\isatt{replacement{\uscore}is{\uscore}order{\uscore}eq{\uscore}map{\uscore}fm}.}
-    \isatt{omap{\uscore}replacement{\uscore}fm}.
+    \sout{\isa{replacement{\uscore}is{\uscore}order{\uscore}eq{\uscore}map{\uscore}fm}.}
+    \isa{omap{\uscore}replacement{\uscore}fm}.
 
@@
   \item
-    \sout{\isatt{replacement{\uscore}is{\uscore}order{\uscore}body{\uscore}fm}.}
-    \isatt{otype{\uscore}replacement{\uscore}fm}.
+    \sout{\isa{replacement{\uscore}is{\uscore}order{\uscore}body{\uscore}fm}.}
+    \isa{otype{\uscore}replacement{\uscore}fm}.
 
@@
   \item
-    \sout{\isatt{replacement{\uscore}HAleph{\uscore}wfrec{\uscore}repl{\uscore}body{\uscore}fm}.}
-    \isatt{wfrec{\uscore}Aleph{\uscore}fm}.
+    \sout{\isa{replacement{\uscore}HAleph{\uscore}wfrec{\uscore}repl{\uscore}body{\uscore}fm}.}
+    \isa{wfrec{\uscore}Aleph{\uscore}fm}.
 
@@
 In our development, the mapping $\phi\mapsto\psi$ defined above is given by the
-$\isatt{ground{\uscore}repl{\uscore}fm}$ function, and all ground replacement
+$\isa{ground{\uscore}repl{\uscore}fm}$ function, and all ground replacement
 instances appear in the locale \locale{M{\uscore}ZF3} and form the set
-\isatt{instances3{\uscore}fms}. These are expressed using
-the $\isatt{ground{\uscore}replacement{\uscore}assm}(M,\mathit{env},\phi)$ predicate
+\isa{instances3{\uscore}fms}. These are expressed using
+the $\isa{ground{\uscore}replacement{\uscore}assm}(M,\mathit{env},\phi)$ predicate
 obtained by replacing $\phi$ by
-$\isatt{ground{\uscore}repl{\uscore}fm}(\phi)$ in Eq.~(\ref{eq:replacement_assm_def}).
+$\isa{ground{\uscore}repl{\uscore}fm}(\phi)$ in Eq.~(\ref{eq:replacement_assm_def}).
 
@@
 require the following 3 instances, which form the set
-\isatt{instances{\uscore}ground{\uscore}fms}.
+\isa{instances{\uscore}ground{\uscore}fms}.
 
 \begin{itemize}
-\item \isatt{wfrec{\uscore}Hcheck{\uscore}fm}.
+\item \isa{wfrec{\uscore}Hcheck{\uscore}fm}.
 
@@
   %
-\item \isatt{wfrec{\uscore}Hfrc{\uscore}at{\uscore}fm}.
+\item \isa{wfrec{\uscore}Hfrc{\uscore}at{\uscore}fm}.
 
@@
 \item
-  \sout{\isatt{Lambda{\uscore}in{\uscore}M{\uscore}fm(check{\uscore}fm(2,0,1),1)}.}
-  \isatt{lam{\uscore}replacement{\uscore}check{\uscore}fm}.
+  \sout{\isa{Lambda{\uscore}in{\uscore}M{\uscore}fm(check{\uscore}fm(2,0,1),1)}.}
+  \isa{lam{\uscore}replacement{\uscore}check{\uscore}fm}.
 
@@
 The next two formulas, which form the set
-\isatt{instances{\uscore}ground{\uscore}notCH{\uscore}fms},
+\isa{instances{\uscore}ground{\uscore}notCH{\uscore}fms},
 are needed for the $\Delta$-System Lemma.
@@
 \item
-  \sout{\isatt{replacement{\uscore}is{\uscore}trans{\uscore}apply{\uscore}image{\uscore}fm}.}
-  \isatt{recursive{\uscore}construction{\uscore}fm}.
+  \sout{\isa{replacement{\uscore}is{\uscore}trans{\uscore}apply{\uscore}image{\uscore}fm}.}
+  \isa{recursive{\uscore}construction{\uscore}fm}.
 
@@
 \item
-  \sout{\isatt{replacement{\uscore}transrec{\uscore}apply{\uscore}image{\uscore}body{\uscore}fm}.}
-  \isatt{recursive{\uscore}construction{\uscore}abs{\uscore}fm}.
+  \sout{\isa{replacement{\uscore}transrec{\uscore}apply{\uscore}image{\uscore}body{\uscore}fm}.}
+  \isa{recursive{\uscore}construction{\uscore}abs{\uscore}fm}.
 
@@
 The $21$ formulas up to this point are collected into the set
-\isatt{overhead{\uscore}notCH}, which is enough to
+\isa{overhead{\uscore}notCH}, which is enough to
 force $\neg\CH$. To force $\CH$, we required one further instance:
@@
 \item
-  \sout{\isatt{replacement{\uscore}dcwit{\uscore}repl{\uscore}body{\uscore}fm}.}
-  \isatt{dc{\uscore}abs{\uscore}fm}.
+  \sout{\isa{replacement{\uscore}dcwit{\uscore}repl{\uscore}body{\uscore}fm}.}
+  \isa{dc{\uscore}abs{\uscore}fm}.
 

### 2022-09-13 Independence_CH/sample_proof.tex
Minor edits
Updated formal proof & snippets

@@
   $Q=(\mathcal{P}(\operatorname{dom}(\tau) \times
-  \mathbb{P}))^{M}$.}
+  \mathbb{P}))^{M}$. This is the set of all names $\vartheta \in
+  M^{\mathbb{P}}$} [\dots]%% ---it is pretty laborious to show that things
+%% are in $M$; we omit 17 lines of code to that effect,
+%% that also apply previously proved lemmas.
 \begin{isabelle}
 \ \ \isacommand{let}\isamarkupfalse%
-\ {\isacharquery}{\kern0pt}Q{\isacharequal}{\kern0pt}{\isachardoublequoteopen}Pow{\isacharparenleft}{\kern0pt}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}{\isasymtimes}\isasymbbbP{\isacharparenright}{\kern0pt}\ {\isasyminter}\ M{\isachardoublequoteclose}
+\ {\isacharquery}{\kern0pt}Q{\isacharequal}{\kern0pt}{\isachardoublequoteopen}Pow\isactrlbsup M\isactrlesup {\isacharparenleft}{\kern0pt}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}{\isasymtimes}{\isasymbbbP}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}
 \end{isabelle}
-\textit{This is the set of all names $\vartheta \in
-  M^{\mathbb{P}}$} [\dots]---it is pretty laborious to show that things
-are in $M$; we omit 17 lines of code to that effect,
-that also apply previously proved lemmas.
+\textit{ Let $\pi=Q \times\{\1\}$ and let
+  $b=\pi_{G}=$ $\left\{\vartheta_{G}: \vartheta \in Q\right\}$.}
+\begin{isabelle}
+\ \ \isacommand{let}\isamarkupfalse%
+\ {\isacharquery}{\kern0pt}{\isasympi}{\isacharequal}{\kern0pt}{\isachardoublequoteopen}{\isacharquery}{\kern0pt}Q{\isasymtimes}{\isacharbraceleft}{\kern0pt}{\isasymone}{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline
+\ \ \isacommand{let}\isamarkupfalse%
+\ {\isacharquery}{\kern0pt}b{\isacharequal}{\kern0pt}{\isachardoublequoteopen}val{\isacharparenleft}{\kern0pt}G{\isacharcomma}{\kern0pt}{\isacharquery}{\kern0pt}{\isasympi}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}
+\end{isabelle}
+(Recall: \textit{\dots there is a $b\in M[G]$ such that\dots})
 \begin{isabelle}
@@
 \ \ \isacommand{have}\isamarkupfalse%
-\ {\isachardoublequoteopen}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}{\isasymtimes}\isasymbbbP\ {\isasymin}\ M{\isachardoublequoteclose}\ {\isachardoublequoteopen}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}\ {\isasymin}\ M{\isachardoublequoteclose}\isanewline
+\ {\isachardoublequoteopen}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}{\isasymtimes}{\isasymbbbP}\ {\isasymin}\ M{\isachardoublequoteclose}\ {\isachardoublequoteopen}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}\ {\isasymin}\ M{\isachardoublequoteclose}\isanewline
 \ \ \ \ \isacommand{by}\isamarkupfalse%
@@
 \ \ \isacommand{have}\isamarkupfalse%
-\ {\isachardoublequoteopen}{\isacharquery}{\kern0pt}Q\ {\isasymin}\ M{\isachardoublequoteclose}
-\end{isabelle}
-[\dots]\textit{ Let $\pi=Q \times\{\1\}$ and let
-  $b=\pi_{G}=$ $\left\{\vartheta_{G}: \vartheta \in Q\right\}$.}
-\begin{isabelle}
-\ \ \isacommand{let}\isamarkupfalse%
-\ {\isacharquery}{\kern0pt}{\isasympi}{\isacharequal}{\kern0pt}{\isachardoublequoteopen}{\isacharquery}{\kern0pt}Q{\isasymtimes}{\isacharbraceleft}{\kern0pt}{\isasymone}{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline
-\ \ \isacommand{let}\isamarkupfalse%
-\ {\isacharquery}{\kern0pt}b{\isacharequal}{\kern0pt}{\isachardoublequoteopen}val{\isacharparenleft}{\kern0pt}G{\isacharcomma}{\kern0pt}{\isacharquery}{\kern0pt}{\isasympi}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}%% \isanewline
-%% \ \ \isacommand{from}\isamarkupfalse%
-%% \ {\isacartoucheopen}{\isacharquery}{\kern0pt}Q{\isasymin}M{\isacartoucheclose}\isanewline
-%% \ \ \isacommand{have}\isamarkupfalse%
-%% \ {\isachardoublequoteopen}{\isacharquery}{\kern0pt}{\isasympi}{\isasymin}M{\isachardoublequoteclose}\isanewline
-%% \ \ \ \ \isacommand{using}\isamarkupfalse%
-%% \ one{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}\isasymbbbP\ P{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}M\ transitivity\isanewline
-%% \ \ \ \ \isacommand{by}\isamarkupfalse%
-%% \ {\isacharparenleft}{\kern0pt}simp\ flip{\isacharcolon}{\kern0pt}\ setclass{\isacharunderscore}{\kern0pt}iff{\isacharparenright}{\kern0pt}\isanewline
-%% \ \ \isacommand{then}\isamarkupfalse%
-%% \isanewline
-%% \ \ \isacommand{have}\isamarkupfalse%
-%% \ {\isachardoublequoteopen}{\isacharquery}{\kern0pt}b\ {\isasymin}\ M{\isacharbrackleft}{\kern0pt}G{\isacharbrackright}{\kern0pt}{\isachardoublequoteclose}\isanewline
-%% \ \ \ \ \isacommand{using}\isamarkupfalse%
-%% \ GenExtI\ \isacommand{by}\isamarkupfalse%
-%% \ simp
+\ {\isachardoublequoteopen}{\isacharquery}{\kern0pt}b\ {\isasymin}\ M{\isacharbrackleft}{\kern0pt}G{\isacharbrackright}{\kern0pt}{\isachardoublequoteclose}\isanewline
+\ \ \ \ \isacommand{by}\isamarkupfalse%
+\ {\isacharparenleft}{\kern0pt}auto\ intro{\isacharbang}{\kern0pt}{\isacharcolon}{\kern0pt}GenExtI{\isacharparenright}{\kern0pt}
 \end{isabelle}
@@
 \ \ \ \ \isacommand{have}\isamarkupfalse%
-\ {\isachardoublequoteopen}arity{\isacharparenleft}{\kern0pt}forces{\isacharparenleft}{\kern0pt}Member{\isacharparenleft}{\kern0pt}{\isadigit{0}}{\isacharcomma}{\kern0pt}{\isadigit{1}}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ {\isadigit{6}}{\isachardoublequoteclose}\isanewline
+\ {\isachardoublequoteopen}arity{\isacharparenleft}{\kern0pt}forces{\isacharparenleft}{\kern0pt}\ {\isasymcdot}{\isadigit{0}}\ {\isasymin}\ {\isadigit{1}}{\isasymcdot}\ {\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ {\isadigit{6}}{\isachardoublequoteclose}\isanewline
 \ \ \ \ \ \ \isacommand{using}\isamarkupfalse%
@@
 \ \ \ \ \ \ \isacommand{by}\isamarkupfalse%
-\ simp
-\end{isabelle}
+\ simp\end{isabelle}
 \textit{Since
@@
 \begin{isabelle}
-\ \ \ \ \isacommand{then}\isamarkupfalse%
-\isanewline
+\ \ \ \ \isacommand{with}\isamarkupfalse%
+\ {\isacartoucheopen}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}{\isasymtimes}{\isasymbbbP}\ {\isasymin}\ M{\isacartoucheclose}\isanewline
 \ \ \ \ \isacommand{have}\isamarkupfalse%
-\ {\isachardoublequoteopen}{\isacharquery}{\kern0pt}{\isasymtheta}\ {\isasymin}\ {\isacharquery}{\kern0pt}Q{\isachardoublequoteclose}\ \isacommand{by}\isamarkupfalse%
-\ auto\isanewline
-\ \ \ \ \isacommand{then}\isamarkupfalse%
-\isanewline
-\ \ \ \ \isacommand{have}\isamarkupfalse%
-\ {\isachardoublequoteopen}val{\isacharparenleft}{\kern0pt}G{\isacharcomma}{\kern0pt}{\isacharquery}{\kern0pt}{\isasymtheta}{\isacharparenright}{\kern0pt}\ {\isasymin}\ {\isacharquery}{\kern0pt}b{\isachardoublequoteclose}\isanewline
+\ {\isachardoublequoteopen}{\isacharquery}{\kern0pt}{\isasymtheta}\ {\isasymin}\ {\isacharquery}{\kern0pt}Q{\isachardoublequoteclose}\isanewline
 \ \ \ \ \ \ \isacommand{using}\isamarkupfalse%
-\ one{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}G\ generic\ val{\isacharunderscore}{\kern0pt}of{\isacharunderscore}{\kern0pt}elem\ {\isacharbrackleft}{\kern0pt}of\ {\isacharquery}{\kern0pt}{\isasymtheta}\ {\isasymone}\ {\isacharquery}{\kern0pt}{\isasympi}\ G{\isacharbrackright}{\kern0pt}\isanewline
-\ \ \ \ \ \ \isacommand{by}\isamarkupfalse%
+\ Pow{\isacharunderscore}{\kern0pt}rel{\isacharunderscore}{\kern0pt}char\ \isacommand{by}\isamarkupfalse%
 \ auto
@@
 %
-\ \ \ \ \isacommand{proof}\isamarkupfalse%
-\ {\isacharminus}{\kern0pt}\isanewline
-\ \ \ \ \ \ \dots\isanewline
-%
-\ \ \ \ \isacommand{qed}\isamarkupfalse%
+\ \ \ \ \isacommand{proof} \ \mbox{$[\dots]$}
 \end{isabelle}

### 2022-09-15 Independence_CH/lessons.tex
Two fixes and one typo

@@
 are presented in the \session{Delta\_System\_Lemma} session. A similar
-problem is that we require the forcing poset to be an element of $M$,
+problem, which can be traced to our reading of Kunen's suggestion on
+how to formalize the forcing relation \cite[p.~260]{kunen2011set},
+is that we require the forcing poset to be an element of $M$,
 so the present infrastructure does not allow class forcing out of the
 box. (The latter change seems to be rather straightforward, but the
-former does not.)\\
-\fbox{Note that this is a consequence of following
-  Kunen's blueprint for the formalization}
+former does not.)
 

### 2022-09-15 Independence_CH/main_results.tex
New names for replacement instances
Two fixes and one typo

@@
   ordertypes;
-\item 2 instances for Aleph: Replacement through $x\mapsto
-  \otype(x)$ (for Hartogs' Theorem) and the well-founded recursion
+\item 2 instances for Aleph: Replacement through the ordertype function (for Hartogs' Theorem) and the well-founded recursion
   using it.
@@
 Finally, \isa{overhead{\isacharunderscore}{\kern0pt}CH} is the union
-of \isa{overhead{\isacharunderscore}{\kern0pt}CH} with the $\DC$
-instance \isa{replacement{\isacharunderscore}{\kern0pt}dcwit{\isacharunderscore}{\kern0pt}repl{\isacharunderscore}{\kern0pt}body{\isacharunderscore}{\kern0pt}fm}:
+of \isa{overhead{\isacharunderscore}{\kern0pt}notCH} with the $\DC$
+instance \isa{dc{\uscore}abs{\uscore}fm}:
 \begin{isabelle}

### 2022-09-15 Independence_CH/non-absolute.tex
Two fixes and one typo

@@
 \label{sec:tools-relativization}
-In order to cope with the inconveniences of the completely relational approach, \sout{we added the missing step from the
-literature consisting of relative versions of the various non-absolute
-functions and programmed in ML limited automatic facilities that
-define the needed concepts, and  state and
-prove the requisite lemmas} we
-provide automatic tools
-to ease the definitions of relativized versions, their fully relational
-partenaire, and the internalized formulas. For instance, consider the 
+The missing step, that naturally appears in the literature, consists
+of having relative \emph{functions} like $\Pow^M$, and the ability to
+translate between the different presentations discussed so far.
+
+To achieve this, we provide automatic tools to ease the definitions of
+such relative versions, their fully relational counteparts, and the
+internalized formulas. For instance, consider the
 $\isa{cardinal}::\tyi \fun \tyi$ function defined in

### 2022-09-15 Independence_CH/proof_assists_isabelle.tex
Reference on internal datatypes

@@
 \footnote{These, alongside with lists, are instances of
-Isabelle/ZF treatment of inductively defined (internal) datatypes;
+Isabelle/ZF treatment of inductively defined (internal) datatypes \cite[Sect.~4]{Paulson1995-wz};
 induction and recursion theorems for them are proved automatically

### 2022-09-15 Independence_CH/replacement_instances.tex
New names for replacement instances
Leaving only essential info in appendix on instances

@@
 
-We isolated 22 instances of Replacement that are sufficient to force
-$\CH$ or $\neg\CH$, which are enumerated below by the name of the
-corresponding internalized first order formula. Many of these were already present in
-relational form in the \session{ZF-Constructible} library.
-
-The first 4 instances, collected in the subset
-\isa{instances1{\uscore}fms} of \formula, consist of basic
-constructions.
+In \isa{instances1{\uscore}fms}:
 
 \begin{itemize}
-\item 2 instances for transitive closure.
+\item Transitive closure:
   \begin{itemize}
@@
 %% they are definitions by well-founded recursion.
-The next 4 instances (gathered in \isa{instances2{\uscore}fms})
-are needed to set up
-cardinal arithmetic in $M$.
+In \isa{instances2{\uscore}fms} (for cardinal arithmetic):
 \begin{itemize}
-\item 2 instances for ordertypes:
+\item Ordertypes:
   \begin{itemize}
@@
   \end{itemize}
-\item 2 instances for Aleph:
+\item Aleph:
   \begin{itemize}
@@
     \sout{\isa{replacement{\uscore}is{\uscore}order{\uscore}body{\uscore}fm}.}
-    \isa{otype{\uscore}replacement{\uscore}fm}.
+    \isa{ordtype{\uscore}replacement{\uscore}fm}.
 
-    Replacement through $x\mapsto \otype(x)$, for Hartogs' Theorem.
+    Replacement through the ordertype function, for Hartogs' Theorem.
     %
@@
 
-We also need a one extra replacement instance $\psi$ on $M$ for each
-$\phi$ of the
-previous ones to have them in $M[G]$:
-\[
-  \psi(x,\alpha,y_1,\dots,y_n) \defi \quine{\alpha = \min \bigl\{
-    \beta \mid \exists\tau\in V_\beta.\  \mathit{snd}(x) \forces
-    \phi\ [\mathit{fst}(x),\tau,y_1,\dots,y_n]\bigr\}}
-\]
-Here, $\mathit{fst}(\lb a,b\rb) = a$, $\mathit{snd}(\lb a,b\rb) = b$
-(with default value $0$ for non pairs).
-In our development, the mapping $\phi\mapsto\psi$ defined above is given by the
-$\isa{ground{\uscore}repl{\uscore}fm}$ function, and all ground replacement
-instances appear in the locale \locale{M{\uscore}ZF3} and form the set
-\isa{instances3{\uscore}fms}. These are expressed using
-the $\isa{ground{\uscore}replacement{\uscore}assm}(M,\mathit{env},\phi)$ predicate
-obtained by replacing $\phi$ by
-$\isa{ground{\uscore}repl{\uscore}fm}(\phi)$ in Eq.~(\ref{eq:replacement_assm_def}).
-
-That makes 16 instances up to now. For the setup of forcing, we
-require the following 3 instances, which form the set
-\isa{instances{\uscore}ground{\uscore}fms}.
+In \isa{instances{\uscore}ground{\uscore}fms}:
 
@@
 \end{itemize}
-The next two formulas, which form the set
-\isa{instances{\uscore}ground{\uscore}notCH{\uscore}fms},
-are needed for the $\Delta$-System Lemma.
+
+In 
+\isa{instances{\uscore}ground{\uscore}notCH{\uscore}fms}:
 \begin{itemize}
@@
 
-  Recursive construction of sets using a choice function (as in the
-  construction of a wellorder of $X$ given a choice function on $\Pow(X)$).
+  Recursive construction of sets using a choice function.
   %
@@
 \end{itemize}
-%
-The $21$ formulas up to this point are collected into the set
-\isa{overhead{\uscore}notCH}, which is enough to
-force $\neg\CH$. To force $\CH$, we required one further instance:
 %
@@
 
-The particular choice of some of the instances above arose from
-Paulson's architecture on which we based our development.
-This applies every time
-a locale from \session{ZF-Constructible} has to be
-interpreted (\locale{M{\uscore}eclose} and
-\locale{M{\uscore}ordertype}, respectively, for the “auxiliary” instances).
-%% For instance, the first
-%% instance required for the definition of relative ordertypes arises
-%% from Paulson's \session{ZF-Constructible}.
-% https://isabelle.in.tum.de/dist/library/ZF/ZF-Constructible/Rank.html#offset_1123..1139
-
-On the other hand, we replaced the original proof of the
-Schröder-Bernstein Theorem by Zermelo's one
-\cite[Exr. x4.27]{moschovakis1994notes}, because the former required
-at least one extra instance
-% (\isa{banach{\uscore}iterates{\uscore}fm})
-arising from an iteration. We also managed to avoid 12 further
-replacements by restructuring some of original theories in
-\session{ZF-Constructible}, so these modifications are included as
-part of our project.
-
-It is to be noted that the proofs of the Forcing Theorems do not
-require any extra replacement; actually, they only need the 7
-instances appearing in \isa{instances1{\uscore}fms} and
-\isa{instances{\uscore}ground{\uscore}fms}.  But this seems not be
-the case for Separation, at least by inspecting our formalization:
-More instances holding in $M$ are needed 
-as the complexity of $\phi$ grows. One point where this is apparent is
-in the proof of Theorem~\ref{th:forcing-thms}(\ref{item:truth-lemma}),
-that appears as the \isa{truth{\uscore}lemma} in our development; it
-depends on \isa{truth{\uscore}lemma'} and
-\isa{truth{\uscore}lemma{\uscore}Neg}, which explicitly invoke
-\isa{separation{\uscore}ax}. In any case, our intended grounds
-(v.g., the transitive collapse of countable elementary submodels of a
-rank initial segment $V_\alpha$ or an $H(\kappa)$) all satisfy full
-Separation.
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "independence_ch_isabelle"
+%%% ispell-local-dictionary: "american"
+%%% End:

### 2022-09-15 Independence_CH/forcing.tex
Section on set models revised

@@
 The description of set models of fragments of $\ZFC$ was performed
-using locales that fix a variable $M::\tyi$
-and pack assumptions stating that $\lb M, \in\rb$ satisfy some
-axioms.
-\begin{itemize}
-\item First introduce axioms as satisfactions;
-\item Comment on the convenience of relative forms;
-\item We retain satisfactions for the predicates for Separation and Replacement.
-\end{itemize}
-For example, the locale \locale{M{\uscore}Z{\uscore}basic}
-states that Zermelo set theory holds in $M$. The Union Axiom (\isa{Union{\uscore}ax}), for
+using locales that fix a variable $M::\tyi$ and pack assumptions
+stating that $\lb M, \in\rb$ satisfy some axioms; for example, the
+locale \locale{M{\uscore}Z{\uscore}basic} states that Zermelo set
+theory holds in $M$. It would be natural to state those assumptions
+directly as the corresponding satisfactions, as in
+\[
+  M, [\,]\models \isa{{\isasymcdot}Union\ Ax{\isasymcdot}}
+\]
+where \isa{{\isasymcdot}Union\ Ax{\isasymcdot}} is the $\formula$ code for the
+Union Axiom. We actually decided to express the axioms other than the
+infinite schemes in relational form, by using terms already
+available in \session{ZF-Constructible} and for which useful lemmas
+had already been proved (and, as it was mentioned in Section~\ref{sec:isabellezf},
+this third layer of the formalization has more tools at our
+disposal); the Union Axiom (\isa{Union{\uscore}ax}), for
 instance, is defined as follows:
@@
 %% \end{isabelle}%
-using relativized, relational versions of the axioms of Isabelle/ZF
-since the interaction with \session{ZF-Constructible} was smoother
-that way and, as it was mentioned in Section~\ref{sec:isabellezf},
-this third layer of the formalization has more tools at our
-disposal. Later, an equivalent statement “in the codes” (our fourth
-layer) is obtained,
+Both assumptions are then shown to be equivalent:
 \[
@@
 %% \end{isabelle}
-where \isa{{\isasymcdot}Union\ Ax{\isasymcdot}} is the $\formula$ code for the
-Union axiom. For the axiom schemes, \session{ZF-Constructible} defines
+
+For stating the axiom schemes, \session{ZF-Constructible} defines
 the expressions
@@
 \]
-%%%%%%%%%%
-\sout{that consume a class $N$ and predicates $Q::\tyi\fun\tyo$ and $R::\tyi\fun\tyi\fun\tyo$, and state that $N$
-satisfies the $Q$-instance of Separation and the $R$-instance of
-Replacement, respectively.
-In the statement of the Separation Axiom in
-\locale{M{\uscore}Z{\uscore}basic} and
-the many replacement instances, predicates $Q$ and $R$ are actually
-defined as the satisfaction of formulas of the appropriate arity; we can only show
-that this form of the axioms hold in generic extensions.}
-%%%%%%%%%%
 whose first argument $N$ is a class and their second arguments $Q$ and
-$R$ are higher-order predicates of types $R :: \tyi \fun \tyo$ and $Q ::
-\tyi \fun \tyi \fun \tyo$, resp.
-The Separation axiom is axiomatized like
-\[
-  \text{This}
-\]
-Particular replacement instances expressed by means of the
-$\isa{replacement{\uscore}assm}(M,\mathit{env},\phi)$ predicate ($@$ denotes
-list concatenation):
-\begin{multline}\label{eq:replacement_assm_def}
-\varphi \in \formula  \limp \mathit{env} \in \isa{list}(M) \limp \isa{arity}(\varphi) \leq 2+ \isa{length}(\mathit{env}) \limp \\
- \isa{strong{\uscore}replacement}(\isa{\#\#} M, \lambda x\, y.\ (M, [x,y]
-\mathbin{@} \mathit{env}  \models \varphi))
-\end{multline}
+$R$ are predicates of types $R :: \tyi \fun \tyo$ and $Q ::
+\tyi \fun \tyi \fun \tyo$, respectively. The Separation Axiom appears in
+\locale{M{\uscore}Z{\uscore}basic} as follows, where 
+\isa{\isasymphi} is free and \isa{\isacharat} denotes
+list concatenation:
+\begin{isabelle}
+separation{\isacharunderscore}{\kern0pt}ax{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}{\isasymphi}\ {\isasymin}\ formula\ {\isasymLongrightarrow}\ env\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}\ {\isasymLongrightarrow}\isanewline
+\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ arity{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isadigit{1}}\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ length{\isacharparenleft}{\kern0pt}env{\isacharparenright}{\kern0pt}\ {\isasymLongrightarrow}\isanewline
+\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ separation{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}M{\isacharcomma}{\kern0pt}{\isasymlambda}x{\isachardot}{\kern0pt}\ {\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}x{\isacharbrackright}{\kern0pt}\ {\isacharat}{\kern0pt}\ env\ {\isasymTurnstile}\ {\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}
+\end{isabelle}
+%% \sout{we can only show
+%% that this form of the axioms hold in generic extensions.}
+Note that the predicate $Q$ mentioned above corresponds to the
+satisfaction of \isa{\isasymphi}.
+
+In constrast to Separation, we stated each instance of the Replacement
+Axiom separately by means of the
+$\isa{replacement{\uscore}assm}(M,\mathit{env},\phi)$ predicate:
+\begin{isabelle}
+{\isasymphi}\ {\isasymin}\ formula\ {\isasymlongrightarrow}\ env\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\isanewline
+\ \ arity{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isadigit{2}}\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ length{\isacharparenleft}{\kern0pt}env{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\isanewline
+\ \ \ \ strong{\isacharunderscore}{\kern0pt}replacement{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}M{\isacharcomma}{\kern0pt}{\isasymlambda}x\ y{\isachardot}{\kern0pt}\ {\isacharparenleft}{\kern0pt}M\ {\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}x{\isacharcomma}{\kern0pt}y{\isacharbrackright}{\kern0pt}{\isacharat}{\kern0pt}env\ {\isasymTurnstile}\ {\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}
+\end{isabelle}
 %% % Same as above
-%% \begin{isabelle}
-%% {\isasymphi}\;{\isasymin}\;formula\ {\isasymlongrightarrow}\ env\;{\isasymin}\;list{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow} arity{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}\;{\isasymle}\;{\isadigit{2}}\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ length{\isacharparenleft}{\kern0pt}env{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\isanewline
-%% \ \ \ \ strong{\isacharunderscore}{\kern0pt}replacement{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}M{\isacharcomma}{\kern0pt}{\isasymlambda}x\ y{\isachardot}{\kern0pt}\ {\isacharparenleft}{\kern0pt}M\ {\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}x{\isacharcomma}{\kern0pt}y{\isacharbrackright}{\kern0pt}{\isacharat}{\kern0pt}env\ {\isasymTurnstile}\ {\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}
-%% \end{isabelle}
+%% \begin{multline}\label{eq:replacement_assm_def}
+%% \varphi \in \formula  \limp \mathit{env} \in \isa{list}(M) \limp \isa{arity}(\varphi) \leq 2+ \isa{length}(\mathit{env}) \limp \\
+%%  \isa{strong{\uscore}replacement}(\isa{\#\#} M, \lambda x\, y.\ (M, [x,y]
+%% \mathbin{@} \mathit{env}  \models \varphi))
+%% \end{multline}
 In turn, the \isa{{\isasymcdot}Replacement{\isasymcdot}} function
@@
 \end{isabelle}
+%%%%%%%
+%% \sout{In \session{ZF-Constructible} some results are proved assuming some instances of
+%% separation or replacement (mejor si ponemos ejemplo concreto). In contrast,
+%% our locale \isa{M{\uscore}Z{\uscore}basic} assumes the full scheme of separation in terms
+%% of satisfaction of some internalized formula but no instance of replacement. (Mostrar el enunciado de separation\_ax:
+%% es distinto a Reemplazo y se nombra después)
+%% 
+%% In order to have a finer control of the instances of replacement needed,
+%% we postulate further locales with the instances needed to interpret at
+%% the class\dots
+%% (tal vez no sea necesario mostrar la definición de .Replacement.)}
+%%%%%%%
 
-In \session{ZF-Constructible} some results are proved assuming some instances of
-separation or replacement (mejor si ponemos ejemplo concreto). In contrast,
-our locale \isa{M{\uscore}Z{\uscore}basic} assumes the full scheme of separation in terms
-of satisfaction of some internalized formula but no instance of replacement. (Mostrar el enunciado de separation\_ax:
-es distinto a Reemplazo y se nombra después)
-
-In order to have a finer control of the instances of replacement needed,
-we postulate further locales with the instances needed to interpret at
-the class\dots
-(tal vez no sea necesario mostrar la definición de .Replacement.)
-
-%
-%% Some of those instances, which we included for simplicity of design,
-%% are actually “fake” instances in that they can be obtained from
-%% Powerset, or from other replacement instances already available in the
-%% respective context. An example of this is the sole instance of the
-%% \locale{M{\uscore}basic} locale, originally from
-%% \session{ZF-Constructible}; we managed to eliminate it but in
-%% doing so we had to reprove many lemmas from
-%% that session in the weakened context.
-
-\sout{The locales gathering all 22 replacement instances necessary are named
-\locale{M{\uscore}ZF1} through \locale{M{\uscore}ZF3},
-\locale{M{\uscore}ZF{\uscore}ground},
-\locale{M{\uscore}ZF{\uscore}ground{\uscore}notCH}, and
-\locale{M{\uscore}ZF{\uscore}ground{\uscore}CH} (see
-Section~\ref{sec:repl-instances} for details).}
+Starting from \locale{M{\uscore}Z{\uscore}basic}, stronger locales are
+defined by assuming more replacement instances.
 These assumptions are then invoked to interpret at the class

### 2022-09-18 Independence_CH/sample_proof.tex
More details about Naproche

@@
 
-Undoubtedly, even for this cherry-picked example, the formalization
-looks a bit codish. It is therefore inevitable to compare this to the
-magnificent results obtained by P.~Koepke and his team by using
-Isabelle/Naproche \cite{10.1007/978-3-030-81097-9_2} (particularly,
-the proof of König's Theorem). The trick there consists in presenting the result
-being formalized as a restricted first-order problem, and then each
-proof step can be handled by an automatic theorem prover.
+This cherry-picked example shows that the formalization can be close
+to the mathematical exposition and might be useful to reconstruct the
+proof from the book; nonetheless, it also has significantly more
+details than the mathematical prose, even with some indications to
+direct the automatic tools.
+
+There has been some progress on assistants where one writes statements
+and proofs in natural language; recently P.~Koepke and his team
+achieved magnificient results by using Isabelle/Naproche
+\cite{10.1007/978-3-030-81097-9_2} to formalize proofs of several
+results (particularly, the proof of König's Theorem). The input
+language of Isabelle/Naproche is a \emph{controlled} natural language
+that presents the result being formalized as a deduction in
+first-order logic, where every assumption and the ``whole logical
+scenario'' are explicitly given. From the input language,
+Isabelle/Naproche builds ``proof tasks'' that are handled to automatic
+theorem provers. As far as we can tell, Isabelle/Naproche is
+promising but still unsuitable for a project of the magnitude of ours.
+
+% Undoubtedly, even for this cherry-picked example, the formalization
+% looks a bit codish. It is therefore inevitable to compare this to the
+% magnificent results obtained by P.~Koepke and his team by using
+% Isabelle/Naproche \cite{10.1007/978-3-030-81097-9_2} (particularly,
+% the proof of König's Theorem). The trick there consists in presenting the result
+% being formalized as a restricted first-order problem, and then each
+% proof step can be handled by an automatic theorem prover.
 

### 2022-09-19 Independence_CH/axioms_metatheory.tex
New appendix with the axioms of the meta-theory.

\section{Axioms of the meta-theory}
\label{appendix:axioms}

In this appendix we list the complete set of axioms of our meta-theory.

\subsection{Pure}
\begin{isabelle}
Pure.abstract\_rule: (⋀x. ?f(x) ≡ ?g(x)) ⟹ λx. ?f(x) ≡ λx. ?g(x)\isanewline
Pure.combination: ?f ≡ ?g ⟹ ?x ≡ ?y ⟹ ?f(?x) ≡ ?g(?y)\isanewline
Pure.equal\_elim: PROP ?A ≡ PROP ?B ⟹ PROP ?A ⟹ PROP ?B\isanewline
Pure.equal\_intr: (PROP ?A ⟹ PROP ?B) ⟹ (PROP ?B ⟹ PROP ?A) ⟹ \isanewline
\ \ \ \ \ \ \ \ \ \ PROP ?A ≡ PROP ?B\isanewline
Pure.reflexive: ?x ≡ ?x\isanewline
Pure.symmetric: ?x ≡ ?y ⟹ ?y ≡ ?x\isanewline
Pure.transitive: ?x ≡ ?y ⟹ ?y ≡ ?z ⟹ ?x ≡ ?z
\end{isabelle}

\subsection{IFOL and FOL}
In the axioms \isa{refl, subst, allI, spec, exE, exI,eq\_reflection} there is a constraint
for the type of the variable \isa{x : ?'a} to be in the class \isa{term\_class}.
\begin{isabelle}
IFOL.FalseE: ⋀P. False ⟹ P\isanewline
IFOL.refl:  (⋀a. a = a)\isanewline
IFOL.subst: (⋀a b P. a = b ⟹ P(a) ⟹ P(b))\isanewline
IFOL.allI:  (⋀P. (⋀x. P(x)) ⟹ ∀x. P(x))\isanewline
IFOL.spec:  (⋀P x. ∀x. P(x) ⟹ P(x))\isanewline
IFOL.exE:   (⋀P R. ∃x. P(x) ⟹ (⋀x. P(x) ⟹ R) ⟹ R)\isanewline
IFOL.exI:   (⋀P x. P(x) ⟹ ∃x. P(x))\isanewline
IFOL.conjI: ⋀P Q. P ⟹ Q ⟹ P ∧ Q\isanewline
IFOL.conjunct1: ⋀P Q. P ∧ Q ⟹ P\isanewline
IFOL.conjunct2: ⋀P Q. P ∧ Q ⟹ Q\isanewline
IFOL.disjE: ⋀P Q R. P ∨ Q ⟹ (P ⟹ R) ⟹ (Q ⟹ R) ⟹ R\isanewline
IFOL.disjI1: ⋀P Q. P ⟹ P ∨ Q\isanewline
IFOL.disjI2: ⋀P Q. Q ⟹ P ∨ Q\isanewline
IFOL.eq\_reflection: (⋀x y. x = y ⟹ x ≡ y)\isanewline
IFOL.iff\_reflection: ⋀P Q. P ⟷ Q ⟹ P ≡ Q\isanewline
IFOL.impI: ⋀P Q. (P ⟹ Q) ⟹ P ⟶ Q\isanewline
IFOL.mp: ⋀P Q. P ⟶ Q ⟹ P ⟹ Q\isanewline%\isanewline
FOL.classical: ⋀P. (¬ P ⟹ P) ⟹ P
\end{isabelle}

\subsection{ZF\_Base}
The following symbols are introduced in this theory:
\begin{isabelle}
axiomatization\isanewline
\ \ \ \      mem :: "[i, i] ⇒ o"  (infixl ‹∈› 50)  ― ‹membership relation›\isanewline
  and zero :: "i"  (‹0›)  ― ‹the empty set›\isanewline
  and Pow :: "i ⇒ i"  ― ‹power sets›\isanewline
  and Inf :: "i"  ― ‹infinite set›\isanewline
  and Union :: "i ⇒ i"  (‹⋃\_› [90] 90)\isanewline
  and PrimReplace :: "[i, [i, i] ⇒ o] ⇒ i"
\end{isabelle}
\noindent with these axioms
\begin{isabelle}
ZF\_Base.Pow\_iff: ⋀A B. A ∈ Pow(B) ⟷ A ⊆ B\isanewline
ZF\_Base.Union\_iff: ⋀A C. A ∈ ⋃C ⟷ (∃B∈C. A ∈ B)\isanewline
ZF\_Base.extension: ⋀A B. A = B ⟷ A ⊆ B ∧ B ⊆ A\isanewline
ZF\_Base.foundation: ⋀A. A = 0 ∨ (∃x∈A. ∀y∈x. y ∉ A)\isanewline
ZF\_Base.infinity: 0 ∈ Inf ∧ (∀y∈Inf. succ(y) ∈ Inf)\isanewline
ZF\_Base.replacement: ⋀A P b. ∀x∈A. ∀y z. P(x, y) ∧ P(x, z) ⟶ y = z ⟹ \isanewline
\ \ \ \ \ \ \ \ \ b ∈ PrimReplace(A, P) ⟷ (∃x∈A. P(x, b))
\end{isabelle}

\subsection{AC}
\begin{isabelle}
AC.AC: ⋀a A B. a ∈ A ⟹ (⋀x. x ∈ A ⟹ ∃y. y ∈ B(x)) ⟹\isanewline
  \ \ \ \ \ \ ∃z. z ∈ Pi(A, B)
\end{isabelle}

### 2022-09-19 Independence_CH/lessons.tex
Eliminating old fm names. microtype commented out temporarily. Edits after discussion with Miguel

@@
 prioritized trying to obtain formal proofs that mimicked standard
-prose (the highlight being the sample proof in
+prose (as can be seen in the sample proof in
 Section~\ref{sec:sample-formal-proof}). We feel that the Isar language

### 2022-09-19 Independence_CH/main_results.tex
Section on set models revised
Eliminating old fm names. microtype commented out temporarily. Edits after discussion with Miguel

@@
 (with default value $0$ for non pairs).
-In our development, the mapping $\phi\mapsto\psi$ defined above is given by the
-$\isa{ground{\uscore}repl{\uscore}fm}$ function, and all ground replacement
+All ground replacement
 instances appear in the locale \locale{M{\uscore}ZF3} and form the set
-\isa{instances3{\uscore}fms}. These are expressed using
-the
-\[
-  \isa{ground{\uscore}replacement{\uscore}assm}(M,\mathit{env},\phi)
-\]
-predicate
-obtained by replacing $\phi$ by
-$\isa{ground{\uscore}repl{\uscore}fm}(\phi)$ in Eq.~(\ref{eq:replacement_assm_def}).
+\isa{instances3{\uscore}fms}.
 
@@
 Also,
-\isa{{\isasymcdot}Z{\isasymcdot}} denotes Zermelo set theory. In the
+\isa{{\isasymcdot}Z{\isasymcdot}} denotes Zermelo set theory and one
+can use the parameter $\Phi$ to ensure those replacement instances in the extension.
+
+In the
 next theorem, the relevant set of formulas is

### 2022-09-19 Independence_CH/related.tex
Bye isatt, bye slanted
Eliminating old fm names. microtype commented out temporarily. Edits after discussion with Miguel

@@
 presentation of the mathematics is somewhat more elegant and cohesive,
-since they go for the Boolean valued approach and this is closer to
-obtain proofs of relative consistency; they actually set up the
+since they go for the Boolean valued approach; they also  set up the
 calculus of first-order logic, and \emph{en route} to forcing they
@@
 \cite{han_et_al:LIPIcs:2019:11074}. It is not clear to us if they are
-able to avoid such strength (at least $\Con(\ZFC)$ is provable in
-their context, a requisite to get either unprovability result).
+able to avoid such strength: At least, $\Con(\ZFC)$ is provable in
+their context. %, a requisite to get either unprovability result.
 % by Gödel's Second).
@@
 Flypitch requires choice in the metatheory
-\cite[p.~11]{han_et_al:LIPIcs:2019:11074}. Our formalization works
-entirely in $\ZF$, and only imports the Isabelle theory postulating
-$\AC$ for demostration purposes.
+\cite[p.~11]{han_et_al:LIPIcs:2019:11074}, while our formalization works
+entirely in $\ZF$. %%  and only imports the Isabelle theory postulating
+%% $\AC$ for demostration purposes.
 
@@
 \begin{itemize}
-\item $\isatt{Ord}(x)$,
-\item $\isatt{ordinal}(\isatt{\#\#}M,x)$ (the relational
-  definition relativized to $M$),
+\item $\isa{Ord}(x)$;
+\item $\isa{ordinal}(\isa{\#\#}M,x)$ (the relational
+  definition relativized to $M$);
 \item
-  $M,[x] \models {\cdot}0\isatt{ is ordinal}{\cdot}$,
-\item
-  $M[G], [x] \models {\cdot}0\isatt{ is ordinal}{\cdot}$ for every generic $G$,
+  $M,[x] \models {\cdot}0\isa{ is ordinal}{\cdot}$.
+%% \item
+%%   $M[G], [x] \models {\cdot}0\isa{ is ordinal}{\cdot}$ for every generic $G$,
 \end{itemize}
-where ${\cdot}n\isatt{ is ordinal}{\cdot}$ is the code for the
+where ${\cdot}n\isa{ is ordinal}{\cdot}$ is the code for the
 appropriate first-order formula  ($0$ is a de

### 2022-09-19 Independence_CH/replacement_instances.tex
Eliminating old fm names. microtype commented out temporarily. Edits after discussion with Miguel

@@
   \item
-    \sout{\isa{eclose{\uscore}repl1{\uscore}intf{\uscore}fm}.} \isa{eclose{\uscore}closed{\uscore}fm}.
+    \isa{eclose{\uscore}closed{\uscore}fm}.
 
@@
   \item
-    \sout{\isa{eclose{\uscore}repl2{\uscore}intf{\uscore}fm}} \isa{eclose{\uscore}abs{\uscore}fm}.
+    \isa{eclose{\uscore}abs{\uscore}fm}.
 
@@
   %
-\item \sout{\isa{trans{\uscore}repl{\uscore}HVFrom{\uscore}fm}.} \isa{transrec{\uscore}VFrom{\uscore}fm}.
+\item \isa{transrec{\uscore}VFrom{\uscore}fm}.
 
@@
   \item
-    \sout{\isa{wfrec{\uscore}replacement{\uscore}order{\uscore}pred{\uscore}fm}.}
     \isa{wfrec{\uscore}ordertype{\uscore}fm}.
@@
   \item
-    \sout{\isa{replacement{\uscore}is{\uscore}order{\uscore}eq{\uscore}map{\uscore}fm}.}
     \isa{omap{\uscore}replacement{\uscore}fm}.
@@
   \item
-    \sout{\isa{replacement{\uscore}is{\uscore}order{\uscore}body{\uscore}fm}.}
     \isa{ordtype{\uscore}replacement{\uscore}fm}.
@@
   \item
-    \sout{\isa{replacement{\uscore}HAleph{\uscore}wfrec{\uscore}repl{\uscore}body{\uscore}fm}.}
     \isa{wfrec{\uscore}Aleph{\uscore}fm}.
@@
 \item
-  \sout{\isa{Lambda{\uscore}in{\uscore}M{\uscore}fm(check{\uscore}fm(2,0,1),1)}.}
   \isa{lam{\uscore}replacement{\uscore}check{\uscore}fm}.
@@
 
+In our development, the $\isa{ground{\uscore}repl{\uscore}fm}$
+function maps instances to \dots. All ground replacement instances
+appear in the locale \locale{M{\uscore}ZF3} and form the set
+\isa{instances3{\uscore}fms}. These are expressed using the following
+predicate:
+\[
+  \isa{ground{\uscore}replacement{\uscore}assm}(M,\mathit{env},\phi)
+\]
+defined as
+$\isa{replacement{\uscore}assm}(M,\mathit{env},\isa{ground{\uscore}repl{\uscore}fm}(\phi))$.
+
 In 
@@
 \item
-  \sout{\isa{replacement{\uscore}is{\uscore}trans{\uscore}apply{\uscore}image{\uscore}fm}.}
   \isa{recursive{\uscore}construction{\uscore}fm}.
@@
 \item
-  \sout{\isa{replacement{\uscore}transrec{\uscore}apply{\uscore}image{\uscore}body{\uscore}fm}.}
   \isa{recursive{\uscore}construction{\uscore}abs{\uscore}fm}.
@@
 \item
-  \sout{\isa{replacement{\uscore}dcwit{\uscore}repl{\uscore}body{\uscore}fm}.}
   \isa{dc{\uscore}abs{\uscore}fm}.

### 2022-09-20 Independence_CH/replacement_instances.tex
Replacement instances in tabular format.

@@
 
-In \isa{instances1{\uscore}fms}:
-
-\begin{itemize}
-\item Transitive closure:
-  \begin{itemize}
-  \item
-    \isa{eclose{\uscore}closed{\uscore}fm}.
-
-    To prove closure under iteration of $X\mapsto\union X$.
-  \item
-    \isa{eclose{\uscore}abs{\uscore}fm}.
-
-    Auxiliary instance used to show absoluteness.
-  \end{itemize}
-%% The instances so far
-%% are needed to interpret locale
-%% \locale{M{\uscore}eclose}.
-\item \isa{wfrec{\uscore}rank{\uscore}fm}.
-
-  For $\in$-rank.
-  %
-\item \isa{transrec{\uscore}VFrom{\uscore}fm}.
-
-  For the cumulative hierarchy (rank initial segments).
-\end{itemize}
-
-%% The last two and next pair have the same syntactic structure, because
-%% they are definitions by well-founded recursion.
-In \isa{instances2{\uscore}fms} (for cardinal arithmetic):
-\begin{itemize}
-\item Ordertypes:
-  \begin{itemize}
-  \item
-    \isa{wfrec{\uscore}ordertype{\uscore}fm}.
-
-    Well-founded recursion for the construction of ordertypes.
-    %
-  \item
-    \isa{omap{\uscore}replacement{\uscore}fm}.
-
-    Auxiliary instance for the definition of ordertypes.
-  \end{itemize}
-\item Aleph:
-  \begin{itemize}
-    %
-  \item
-    \isa{ordtype{\uscore}replacement{\uscore}fm}.
-
-    Replacement through the ordertype function, for Hartogs' Theorem.
-    %
-  \item
-    \isa{wfrec{\uscore}Aleph{\uscore}fm}.
-
-    The well-founded recursion to define Aleph.
-  \end{itemize}
-\end{itemize}
-
-In \isa{instances{\uscore}ground{\uscore}fms}:
-
-\begin{itemize}
-\item \isa{wfrec{\uscore}Hcheck{\uscore}fm}.
-
-  Well-founded recursion to define check-names.
-  %
-\item \isa{wfrec{\uscore}Hfrc{\uscore}at{\uscore}fm}.
-
-  Well-founded recursion for the definition of forcing for atomic formulas.
-  %
-\item
-  \isa{lam{\uscore}replacement{\uscore}check{\uscore}fm}.
-
-  Replacement through $x\mapsto \lb x,\check{x}\rb$ (for the
-  definition of $\punto{G}$).
-  %
-\end{itemize}
+In Table~\ref{tab:instances1} we show the name of the fourteen formulas
+involved in the twenty two instances of replacement needed in our
+mechanization. The formulas marked with ($\dagger$) are needed twice:
+one by themselves and the other as the argument for $\isa{ground{\uscore}repl{\uscore}fm}$.
 
 In our development, the $\isa{ground{\uscore}repl{\uscore}fm}$
-function maps instances to \dots. All ground replacement instances
-appear in the locale \locale{M{\uscore}ZF3} and form the set
-\isa{instances3{\uscore}fms}. These are expressed using the following
-predicate:
-\[
-  \isa{ground{\uscore}replacement{\uscore}assm}(M,\mathit{env},\phi)
-\]
-defined as
-$\isa{replacement{\uscore}assm}(M,\mathit{env},\isa{ground{\uscore}repl{\uscore}fm}(\phi))$.
+function maps $\phi$ to $\psi$ as described in
+Sect.~\ref{sec:repl-instances}. These instances form the set
+\isa{instances3{\uscore}fms}.
 
-In 
-\isa{instances{\uscore}ground{\uscore}notCH{\uscore}fms}:
-\begin{itemize}
-\item
-  \isa{recursive{\uscore}construction{\uscore}fm}.
+% These are expressed using the following predicate:
+% \[
+%   \isa{ground{\uscore}replacement{\uscore}assm}(M,\mathit{env},\phi)
+% \]
+% defined as
+% $\isa{replacement{\uscore}assm}(M,\mathit{env},\isa{ground{\uscore}repl{\uscore}fm}(\phi))$.
+\newcommand{\groundRepl}{\ensuremath{{}^\dagger}}
+\newcommand{\replInstSet}[1]{\multicolumn{3}{@{}l}{\hspace{0.3em}{$\mathbf{\mathit{#1}}$}}}
 
-  Recursive construction of sets using a choice function.
-  %
-\item
-  \isa{recursive{\uscore}construction{\uscore}abs{\uscore}fm}.
-
-  Absoluteness of the previous construction.
-\end{itemize}
-%
-\begin{itemize}
-\item
-  \isa{dc{\uscore}abs{\uscore}fm}.
-
-  Absoluteness of the recursive construction in the proof of the
-  Dependent Choices from $\AC$.
-\end{itemize}
+\begin{table}[!h]
+\centering
+\begin{threeparttable}
+\begin{tabular}{r<{\stepcounter{replInstCount}\thereplInstCount.} >{\hspace{2pt}}l @{\hspace{1ex}} p{6cm}}
+  \toprule
+  \multicolumn{1}{r}{No.} & name & {purpose} \\
+  \midrule
+  \replInstSet{instances1{\uscore}fms}\\
+%  \addlinespace
+  & \isa{eclose{\uscore}closed{\uscore}fm} \groundRepl & closure under iteration of $X\mapsto\union X$. \\
+  & \isa{eclose{\uscore}abs{\uscore}fm} \groundRepl & Absoluteness of the previous construction.\\
+%  \addlinespace
+  & \isa{wfrec{\uscore}rank{\uscore}fm} \groundRepl &  For $\in$-rank.\\
+%  \addlinespace
+  & \isa{transrec{\uscore}VFrom{\uscore}fm} \groundRepl & For rank initial segments.\\
+  \midrule
+  \replInstSet{instances2{\uscore}fms}\\
+%  \addlinespace
+  & \isa{wfrec{\uscore}ordertype{\uscore}fm} \groundRepl & Well-founded recursion for the construction of ordertypes. \\
+  & \isa{omap{\uscore}replacement{\uscore}fm} \groundRepl & Auxiliary instance for the definition of ordertypes. \\
+%  \addlinespace
+  & \isa{ordtype{\uscore}replacement{\uscore}fm} \groundRepl& Replacement through the ordertype function, for Hartogs' Theorem.\\
+%  \addlinespace
+  & \isa{wfrec{\uscore}Aleph{\uscore}fm} \groundRepl& Well-founded recursion to define Aleph.\\
+  \midrule
+  \replInstSet{instances{\uscore}ground{\uscore}fms}\\
+%  \addlinespace
+  & \isa{wfrec{\uscore}Hcheck{\uscore}fm} & Well-founded recursion to define check.\\
+  & \isa{wfrec{\uscore}Hfrc{\uscore}at{\uscore}fm}. & Well-founded recursion for the definition of forcing for atomic formulas.\\
+  & \isa{lam{\uscore}replacement{\uscore}check{\uscore}fm} & Replacement through $x\mapsto \lb x,\check{x}\rb$ (for the
+  definition of $\punto{G}$).\\
+  \midrule
+  \replInstSet{instances{\uscore}ground{\uscore}notCH{\uscore}fms}\\
+%  \addlinespace
+  &   \isa{recursive{\uscore}construction{\uscore}fm} &
+  Recursive construction of sets using a choice function.\\
+  & \isa{recursive{\uscore}construction{\uscore}abs{\uscore}fm} &
+  Absoluteness of the previous construction \\
+  \midrule
+  \replInstSet{instances{\uscore}ground{\uscore}CH{\uscore}fms}\\
+%  \addlinespace
+  & \isa{dc{\uscore}abs{\uscore}fm} &  Absoluteness of the recursive construction in the proof of
+  Dependent Choice from $\AC$. \\
+  \midrule
+  \replInstSet{instances3{\uscore}fms}\\
+%  \addlinespace
+  \multicolumn{1}{r}{15-22.} & $\isa{ground{\uscore}repl{\uscore}fm}(\phi)$ & one for each formula $\phi$ marked with $\dagger$ \\
+  \bottomrule
+\end{tabular}
+\caption{Replacement Instances used in our mechanization}
+\label{tab:instances1}
+\end{threeparttable}
+\end{table}
 

### 2022-09-20 Independence_CH/axioms_metatheory.tex
Minor editing

@@
 In the axioms \isa{refl, subst, allI, spec, exE, exI,eq\_reflection} there is a constraint
-for the type of the variable \isa{x : ?'a} to be in the class \isa{term\_class}.
+for the \emph{type of} the variables \isa{a, b, x} to be in the class \isa{term\_class}.
 \begin{isabelle}
@@
 axiomatization\isanewline
-\ \ \ \      mem :: "[i, i] ⇒ o"  (infixl ‹∈› 50)  ― ‹membership relation›\isanewline
-  and zero :: "i"  (‹0›)  ― ‹the empty set›\isanewline
-  and Pow :: "i ⇒ i"  ― ‹power sets›\isanewline
-  and Inf :: "i"  ― ‹infinite set›\isanewline
+\ \ \ \      mem :: "[i, i] ⇒ o"  (infixl ‹∈› 50)  \isamarkupcmt{membership relation}\isanewline
+  and zero :: "i"  (‹0›)  \isamarkupcmt{the empty set}\isanewline
+  and Pow :: "i ⇒ i"  \isamarkupcmt{power sets}\isanewline
+  and Inf :: "i"  \isamarkupcmt{infinite set}\isanewline
   and Union :: "i ⇒ i"  (‹⋃\_› [90] 90)\isanewline
@@
 \end{isabelle}
+
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: "independence_ch_isabelle"
+%%% ispell-local-dictionary: "american"
+%%% End:

### 2022-09-20 Independence_CH/replacement_instances.tex
Minor editing

@@
 
-In Table~\ref{tab:instances1} we show the name of the fourteen formulas
-involved in the twenty two instances of replacement needed in our
-mechanization. The formulas marked with ($\dagger$) are needed twice:
-one by themselves and the other as the argument for $\isa{ground{\uscore}repl{\uscore}fm}$.
-
-In our development, the $\isa{ground{\uscore}repl{\uscore}fm}$
-function maps $\phi$ to $\psi$ as described in
-Sect.~\ref{sec:repl-instances}. These instances form the set
-\isa{instances3{\uscore}fms}.
+In Table~\ref{tab:instances1} we show the name of the fourteen
+formulas involved in the twenty two instances of replacement needed in
+our mechanization. The formulas marked with ($\dagger$) are needed
+twice: one by themselves and the other as the argument for
+$\isa{ground{\uscore}repl{\uscore}fm}$. The
+$\isa{ground{\uscore}repl{\uscore}fm}$ function maps $\phi$ to $\psi$
+as described in Sect.~\ref{sec:repl-instances}. These eight instances form
+the set \isa{instances3{\uscore}fms}.
 
@@
   \toprule
-  \multicolumn{1}{r}{No.} & name & {purpose} \\
+  \multicolumn{1}{r}{No.} & Formula's name & Comment \\
   \midrule
@@
 %  \addlinespace
-  & \isa{eclose{\uscore}closed{\uscore}fm} \groundRepl & closure under iteration of $X\mapsto\union X$. \\
+  & \isa{eclose{\uscore}closed{\uscore}fm} \groundRepl & Closure under iteration of $X\mapsto\union X$. \\
   & \isa{eclose{\uscore}abs{\uscore}fm} \groundRepl & Absoluteness of the previous construction.\\
@@
   & \isa{wfrec{\uscore}Hfrc{\uscore}at{\uscore}fm}. & Well-founded recursion for the definition of forcing for atomic formulas.\\
-  & \isa{lam{\uscore}replacement{\uscore}check{\uscore}fm} & Replacement through $x\mapsto \lb x,\check{x}\rb$ (for the
-  definition of $\punto{G}$).\\
+  & \isa{lam{\uscore}replacement{\uscore}check{\uscore}fm} & Replacement through $x\mapsto \lb x,\check{x}\rb$, for $\punto{G}$.\\
   \midrule
@@
   & \isa{recursive{\uscore}construction{\uscore}abs{\uscore}fm} &
-  Absoluteness of the previous construction \\
+  Absoluteness of the previous construction.\\
   \midrule

### 2022-09-22 Independence_CH/lam_replacement.tex
Table with assumptions of M_replacement; comments about AC.

@@
 enough to assume the lambda replacement
-$x\mapsto \lb x,H(\mathit{fst}(x),\mathit{snd}(x)))\rb$.
+$x\mapsto \lb x,H(\mathit{fst}(x),\mathit{snd}(x)))\rb$. We summarize
+the assumptions in Table~\ref{tab:m-repl-instances}.
+
+\newcommand{\lamRepl}[2][x]{#1 \mapsto\langle #1,#2\rangle}
+\begin{table}[!h]
+\centering
+\begin{threeparttable}
+\begin{tabular}{r<{\stepcounter{LamReplCount}\theLamReplCount.} >{\hspace{1ex}}l @{\hspace{1em}} l}
+  \toprule
+  \multicolumn{1}{r}{No.} & Name & Instance  \\
+  \midrule
+  \multicolumn{3}{@{}l}{\hspace{0.3em}{\textit{{Replacement Instances}}}}\\
+  & \isa{lam{\uscore}replacement{\uscore}fst} & $\lamRepl{\mathit{fst}(x)}$ \\
+  & \isa{lam{\uscore}replacement{\uscore}snd} & $\lamRepl{\mathit{snd}(x)}$ \\
+  & \isa{lam{\uscore}replacement{\uscore}Union} & $\lamRepl{\bigcup(x)}$ \\
+  & \isa{lam{\uscore}replacement{\uscore}Image} & $\lamRepl{\mathit{fst}(x)``\mathit{snd}(x)}$ \\
+  & \isa{lam{\uscore}replacement{\uscore}middle{\uscore}del} &
+        $\lamRepl{\langle \mathit{fst}(\mathit{fst}(x)),\mathit{snd}(\mathit{snd}(x)) \rangle}$ \\
+  & \isa{lam{\uscore}replacement{\uscore}prodRepl} &
+        $\lamRepl{\langle \mathit{snd}(\mathit{fst}(x)),\langle \mathit{fst}(\mathit{fst}(x)),\mathit{snd}(\mathit{snd}(x))\rangle \rangle}$\\
+  \midrule
+  \multicolumn{3}{@{}l}{\hspace{0.3em}{\textit{{Separation Instances}}}}\\
+  & \isa{middle{\uscore}separation} & $\mathit{snd}(\mathit{fst}(x))=\mathit{fst}(\mathit{snd}(x))$ \\
+  & \isa{separation{\uscore}fst{\uscore}in{\uscore}snd} & $\mathit{fst}(\mathit{snd}(x)) \in \mathit{snd}(\mathit{snd}(x))$\\
+  \bottomrule
+\end{tabular}
+\caption{Replacement and Separation instances of the locale \isa{M{\uscore}remplacement}}
+\label{tab:m-repl-instances}
+\end{threeparttable}
+\end{table}
 

### 2022-09-22 Independence_CH/lam_replacement.tex
Removed a repeated word.

@@
 Faced with this situation, we decided that we needed some sort of
-\emph{compositionality} in order to obtain instances new instances
-from the ones already proved: Having Replacement for class functions
-$F$ and $G$ does not entail immediately replacement under $F\circ G$
-(unless you use one further instance of Separation, and the net gain
-is zero). The solution was to postulate for the relevant $F$s, instead
-of replacement through $x\mapsto F(x)$, a \emph{lambda replacement}
+\emph{compositionality} in order to obtain new instances from the ones
+already proved: Having Replacement for class functions $F$ and $G$
+does not entail immediately replacement under $F\circ G$ (unless you
+use one further instance of Separation, and the net gain is zero). The
+solution was to postulate for the relevant $F$s, instead of
+replacement through $x\mapsto F(x)$, a \emph{lambda replacement}
 through $x\mapsto \lb x,F(x)\rb$. The name “lambda” corresponds to the

### 2022-09-26 Independence_CH/axioms_metatheory.tex
Table with assumptions of M_replacement; comments about AC.
Meta vs object logic. Appendix order. Metatheory now a word.

@@
-\section{Axioms of the meta-theory}
+\section{Axioms of Isabelle/ZF}
 \label{appendix:axioms}
 
-In this appendix we list the complete set of axioms of our meta-theory.
+In this appendix we list the complete set of axioms of Isabelle's
+metatheory and logic.
 
-\subsection{Pure}
+\subsection{The metatheory Pure}
 \begin{isabelle}
@@
 \subsection{AC}
+
+The theory \theory{AC} is only imported in the theory
+\isa{Absolute{\uscore}Versions}; none of the main results
+depends on $\AC$. The theory \theory{Absolute\_Versions}
+shows that some absolute results can be obtained from the
+relativized versions on $\mathcal{V}$.
+
 \begin{isabelle}

### 2022-09-26 Independence_CH/conclusion.tex
Meta vs object logic. Appendix order. Metatheory now a word.

@@
 our presentation of this material during the Latin American Congress
-of Mathematicians (CLAM 2021); we make this extensive to the rest of
-the Isabelle community, for its kind support through the Users mailing
+of Mathematicians (CLAM 2021); we extend our gratitude to the rest of
+the Isabelle community, for their kind support through the Users mailing
 list. We also want to warmly thank Larry Paulson for his work and

### 2022-09-26 Independence_CH/forcing.tex
Meta vs object logic. Appendix order. Metatheory now a word.

@@
 p.~260 of this book, the structural recursion to define the forcing
-relation was replaced by one  involving codes for formulas. Thus, the meta-theoretic formula
+relation was replaced by one  involving codes for formulas. Thus, the metatheoretic formula
 transformer $\phi\mapsto\mathit{Forces}_\phi$ was replaced by a

### 2022-09-26 Independence_CH/lam_replacement.tex
Meta vs object logic. Appendix order. Metatheory now a word.

@@
 process. As further properties of closure of the ground $M$ were
-required, we gather the relevant instances of Separation and
+required, we gathered the relevant instances of Separation and
 Replacement into a new locale (always assuming a class model, for

### 2022-09-26 Independence_CH/proof_assists_isabelle.tex
Comments on theories brought from ZF-Constructible. Shortening fm names

@@
 We recommend the survey \cite{2020arXiv200909541A} by
-J.~Avigad for details about the different logical foundations in which
+J.~Avigad for details about the different logical foundations on which
 assistants are based.}

### 2022-09-26 Independence_CH/replacement_instances.tex
Comments on theories brought from ZF-Constructible. Shortening fm names

@@
 %  \addlinespace
-  &   \isa{recursive{\uscore}construction{\uscore}fm} &
+  &   \isa{rec{\uscore}constr{\uscore}fm} &
   Recursive construction of sets using a choice function.\\
-  & \isa{recursive{\uscore}construction{\uscore}abs{\uscore}fm} &
+  & \isa{rec{\uscore}constr{\uscore}abs{\uscore}fm} &
   Absoluteness of the previous construction.\\

### 2022-09-30 Independence_CH/conclusion.tex
Joint editing with Miguel

@@
 
-In line with identifying (almost) minimal set of axioms, we would like
-to obtained formalized results concerning which instances of
-Separation are needed to use forcing. A necessary ingredient will
+As we did for Replacement, we would like to pinpoint an (almost)
+minimal set of instances of
+Separation needed to use forcing. A necessary ingredient will
 certainly be an implementation of Gödel operations
@@
 accomplishment after seeing one's writings validated beyond doubt is
-in some ways comparable that of finding a proof of an important
+in some ways comparable to that of finding a proof of an important
 lemma. It also allows subtly different ways of reasoning (with their

### 2022-09-30 Independence_CH/lessons.tex
Joint editing with Miguel

@@
   we only proved the Delta System Lemma for $\aleph_1$-sized families;
-  thus limiting us to $\kappa=\aleph_1$ case of the $\kappa$-chain condition, and dodging the relativization
+  thus limiting us to the case of the $\aleph_1$-chain condition, and avoiding the relativization
   of the material on cofinalities \cite{Delta_System_Lemma-AFP};
@@
 \end{itemize}
-In doing this we went against the tried and
-true advice that one should formalize the most general version of the
+In doing this we went against the conventional wisdom that one should formalize the most general version of the
 results available. Another shortcut we took was to
@@
 link in the chain is the laying down of definitions building up to the
-concepts needed to state the target results. (Another related aspect,
+concepts needed to state the target results.%
+%%%%%%%
+\footnote{%
+Another related aspect,
 concerning the way results are printed and parsed by assistant versus
-their internal meaning, was studied by Wiedijk \cite{zbMATH06319597}.)
+their internal meaning, was studied by Wiedijk \cite{zbMATH06319597}.}
+%%%%%%%
 
@@
 Frequently, we formalized material by directly typing the proof we
-knew by heart, and in doing so we assumed that some definitions
+knew by heart, and in so doing we assumed that some definitions
 accommodated some of our preconceptions.
@@
 
-A final aspect on this topic concerns automated methods. As it was
-mentioned in the Introduction, some proofs can actually be
-\emph{obscured} by automation, simply because one has to take
-complicated opaque computations “on faith.”
+A final aspect on this topic concerns automated methods. In the
+Introduction we hinted at the fact that a proof can actually be
+\emph{obscured} by automation. Specifically, proof steps that were solved automatically give no
+information for someone who wants to understand the details of the
+argument; by the same token, automatic methods might silently exploit
+inconsistencies in the definitions, and this will only be apparent in
+a later stage of the development.
+
+%, simply because one has to take
+%complicated opaque computations “on faith.”
 
@@
 
-Nevertheless, the experience of the CS members of the team (and
-community) is invaluable for the kind of engineering that is needed in
+Nevertheless, experience  in software engineering is invaluable in
 a large project like the present one. For example, it is
@@
 variables doesn't). Scale issues are very easily disregarded in the
-abstract but, as an example, the formula $\forceisa({\cdot}0\in 1{\cdot})$
+abstract but, as a colorful example, the formula $\forceisa({\cdot}0\in 1{\cdot})$
 can be explicitly printed by Isabelle2021-1 (it spans nearly 20k symbols), but
@@
 
-Another point where CS expertise was a prime asset was
+Another point where computer science expertise was a prime asset was
 the very definition of $\forceisa$. As a
 proof of concept, one of us tried to obtain its definition by
-using formula synthesis exclusively, which was supposed to be trivial,
-since it is mathematically (as in the case of
+using formula synthesis exclusively, which was supposed to be as
+trivial as in the usual
+mathematical development (similarly to the case of
 Equation~(\ref{eq:sats_big_union_fm})).

### 2022-09-30 Independence_CH/main_defs.tex
Meta vs object logic. Appendix order. Metatheory now a word.
Joint editing with Miguel

@@
 For the basic logic ZF we restrict ourselves to just a few
-concepts.
+concepts (for its axioms, consult Appendix~\ref{appendix:axioms}).
 %
@@
 i\ {\isacharless}{\kern0pt}\ j\ {\isasymequiv}\ i\ {\isasymin}\ j\ {\isasymand}\ Ord{\isacharparenleft}{\kern0pt}j{\isacharparenright}{\kern0pt}\isasep\isanewline%
-i\ {\isasymle}\ j\ {\isasymlongleftrightarrow}\ i\ {\isacharless}{\kern0pt}\ j\ {\isasymor}\ i\ {\isacharequal}{\kern0pt}\ j\ {\isasymand}\ Ord{\isacharparenleft}{\kern0pt}j{\isacharparenright}{\kern0pt}%
+i\ {\isasymle}\ j\ {\isasymlongleftrightarrow}\ i\ {\isacharless}{\kern0pt}\ j\ {\isasymor}\ {\isacharparenleft}i\ {\isacharequal}{\kern0pt}\ j\ {\isasymand}\ Ord{\isacharparenleft}{\kern0pt}j{\isacharparenright}{\isacharparenright}{\kern0pt}%
 \end{isabelle}%

### 2022-09-30 Independence_CH/forcing.tex
Line breaks

@@
 \end{theorem}
-The items in Theorem~\ref{th:forcing-thms} appear as three
-separated lemmas in the theory
-\theory{Forcing{\uscore}Theorems} of our
-\session{Independence\_CH} session \cite{Independence_CH-AFP}.
+The items in Theorem~\ref{th:forcing-thms} appear in our
+\session{Independence\_CH} session \cite{Independence_CH-AFP} as three
+separate lemmas (located in the theory
+\theory{Forcing{\uscore}Theorems}).
 For instance, the Truth Lemma is stated as

### 2022-09-30 Independence_CH/main_defs.tex
Line breaks

@@
 \begin{isabelle}%
-pair{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ a{\isacharcomma}{\kern0pt}\ b{\isacharcomma}{\kern0pt}\ z{\isacharparenright}{\kern0pt}\ {\isasymequiv}\isanewline
-{\isasymexists}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ upair{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ a{\isacharcomma}{\kern0pt}\ a{\isacharcomma}{\kern0pt}\ x{\isacharparenright}{\kern0pt}\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymexists}y{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ upair{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ a{\isacharcomma}{\kern0pt}\ b{\isacharcomma}{\kern0pt}\ y{\isacharparenright}{\kern0pt}\ {\isasymand}\ upair{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ x{\isacharcomma}{\kern0pt}\ y{\isacharcomma}{\kern0pt}\ z{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}%
+pair{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ a{\isacharcomma}{\kern0pt}\ b{\isacharcomma}{\kern0pt}\ z{\isacharparenright}{\kern0pt}\ {\isasymequiv}\ 
+{\isasymexists}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ upair{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ a{\isacharcomma}{\kern0pt}\ a{\isacharcomma}{\kern0pt}\ x{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
+\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ {\isacharparenleft}{\kern0pt}{\isasymexists}y{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ upair{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ a{\isacharcomma}{\kern0pt}\ b{\isacharcomma}{\kern0pt}\ y{\isacharparenright}{\kern0pt}\ {\isasymand}\ upair{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ x{\isacharcomma}{\kern0pt}\ y{\isacharcomma}{\kern0pt}\ z{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}%
 \end{isabelle}%
@@
 \begin{isabelle}%
-power{\isacharunderscore}{\kern0pt}ax{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}\ {\isasymequiv}\ {\isasymforall}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymexists}z{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymforall}xa{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ xa\ {\isasymin}\ z\ {\isasymlongleftrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymforall}xb{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ xb\ {\isasymin}\ xa\ {\isasymlongrightarrow}\ xb\ {\isasymin}\ x{\isacharparenright}{\kern0pt}%
+power{\isacharunderscore}{\kern0pt}ax{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}\ {\isasymequiv}\ {\isasymforall}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymexists}z{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymforall}xa{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ xa\ {\isasymin}\ z\ {\isasymlongleftrightarrow}\isanewline
+\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ {\isacharparenleft}{\kern0pt}{\isasymforall}xb{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ xb\ {\isasymin}\ xa\ {\isasymlongrightarrow}\ xb\ {\isasymin}\ x{\isacharparenright}{\kern0pt}%
 \end{isabelle}%
@@
 \begin{isabelle}%
-choice{\isacharunderscore}{\kern0pt}ax{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}\ {\isasymequiv}\ {\isasymforall}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymexists}a{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymexists}f{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ ordinal{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ a{\isacharparenright}{\kern0pt}\ {\isasymand}\ surjection{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ a{\isacharcomma}{\kern0pt}\ x{\isacharcomma}{\kern0pt}\ f{\isacharparenright}{\kern0pt}%
+choice{\isacharunderscore}{\kern0pt}ax{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}\ {\isasymequiv}\ {\isasymforall}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymexists}a{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymexists}f{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ ordinal{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ a{\isacharparenright}{\kern0pt}\ {\isasymand}\isanewline
+\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ surjection{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ a{\isacharcomma}{\kern0pt}\ x{\isacharcomma}{\kern0pt}\ f{\isacharparenright}{\kern0pt}%
 \end{isabelle}%
@@
 {\isasymforall}A{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\isanewline
-\isaindent{\ \ \ }univalent{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ A{\isacharcomma}{\kern0pt}\ P{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymexists}Y{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymforall}b{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ b\ {\isasymin}\ Y\ {\isasymlongleftrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymexists}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ x\ {\isasymin}\ A\ {\isasymand}\ P{\isacharparenleft}{\kern0pt}x{\isacharcomma}{\kern0pt}\ b{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}%
+\isaindent{\ \ \ }univalent{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ A{\isacharcomma}{\kern0pt}\ P{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymexists}Y{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymforall}b{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ b\ {\isasymin}\ Y\ {\isasymlongleftrightarrow}\isanewline
+\ \ \ \ \ \ {\isacharparenleft}{\kern0pt}{\isasymexists}x{\isacharbrackleft}{\kern0pt}M{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ x\ {\isasymin}\ A\ {\isasymand}
+P{\isacharparenleft}{\kern0pt}x{\isacharcomma}{\kern0pt}\ b{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}%
 \end{isabelle}%

### 2022-09-30 Independence_CH/arxiv.sh
Scripts for producing the arXiv version

@@ -1,0 +1,37 @@
+#!/bin/bash
+
+rootbase=independence_ch_isabelle
+root=$rootbase.tex
+arxiv_dir=__"$rootbase"_ArXiv_tmp
+arxiv="$rootbase"_arxiv.tex
+tgz="$rootbase"_arxiv.tar.gz
+extra_files="llncs.cls"
+
+pdflatex -draftmode \\nonstopmode \\input $root > /dev/null
+bibtex $rootbase
+makeindex $rootbase
+pdflatex  -draftmode \\nonstopmode \\input $root > /dev/null
+pdflatex \\nonstopmode \\input $root
+
+mkdir $arxiv_dir
+
+for x in *.tex
+do
+    ../scripts/stripcomments.pl $x > $arxiv_dir/$x
+done
+
+cp *sty *bbl $extra_files $arxiv_dir
+
+cd $arxiv_dir
+
+sed 's/\\bibliographystyle{[^}]*}//g' $root | \
+    sed 's/\\bibliography{'$rootbase'}/\\input{'$rootbase.bbl'}/g'\
+ > $arxiv
+rm -f $root
+
+tar --exclude=$root -zcvf $tgz *tex *sty *cls *bbl $extra_files
+mv $tgz ..
+
+cd ..
+
+rm -rf $arxiv_dir

### 2022-09-30 Independence_CH/independence_ch_isabelle.tex
Signaling edits after meeting with Miguel. Proposed new section on replacement instances, and new names in appendix
Added microtype package.
New appendix with the axioms of the meta-theory.
Eliminating old fm names. microtype commented out temporarily. Edits after discussion with Miguel
Replacement instances in tabular format.
Table with assumptions of M_replacement; comments about AC.
Meta vs object logic. Appendix order. Metatheory now a word.
Joint editing with Miguel
Updated abstact and keywords
Supporting projects

@@
 %
-\usepackage[utf8]{inputenc}
+%\usepackage[utf8]{inputenc}
+%\usepackage[final]{microtype}
 \usepackage{isabelle_indepCH,isabellesym_indepCH}
+\usepackage{booktabs,array,threeparttable}
 \input{header-indepCH}
@@
 % \renewcommand\UrlFont{\color{blue}\rmfamily}
+\input{unicode_support}
+\usepackage{array}
+\newcounter{replInstCount}
+\setcounter{replInstCount}{0}
 
+\newcounter{LamReplCount}
+\setcounter{LamReplCount}{0}
 \begin{document}
@@
 \title{The formal verification of the ctm approach to forcing%Some lessons after the formalization of the ctm approach to forcing%
-  \thanks{Supported by Secyt-UNC project 33620180100465CB and Conicet.}%
+  \thanks{Supported by Secyt-UNC projects 33620180100855CB and 33620180100465CB, and Conicet.}%
 }
@@
 
-  %% The vantage point of the talk will be that of a mathematician but
-  %% elements from the computer science perspective will be
-  %% present. Perhaps some myths regarding what can effectively be done
-  %% using  proof assistants/checkers will
-  %% be dispelled.
-
-  We also compare our formalization with the recent one by J.~M.~Han and F.~van Doorn in the proof assistant \emph{Lean}.
-
   \keywords{forcing \and Isabelle/ZF \and countable transitive models
-    \and absoluteness \and generic extension.}
+    \and continuum hypothesis
+    \and proof assistants \and interactive theorem provers \and generic extension}
 \end{abstract}
@@
 
-\input{lam_replacement}
+\input{main_defs}
 
@@
 
-\input{main_defs}
+\input{axioms_metatheory}
+
+\input{lam_replacement}
+
+\input{replacement_instances}
 

### 2022-09-30 Independence_CH/arxiv.sh
Eliminating repeated file for the tarball

@@ -29,7 +29,7 @@ sed 's/\\bibliographystyle{[^}]*}//g' $root | \
  > $arxiv
 rm -f $root
 
-tar --exclude=$root -zcvf $tgz *tex *sty *cls *bbl $extra_files
+tar --exclude=$root -zcvf $tgz *tex *sty *bbl $extra_files
 mv $tgz ..
 
 cd ..

### 2022-09-30 Independence_CH/forcing.tex
Spell-check

@@
 
-In constrast to Separation, we stated each instance of the Replacement
+In contrast to Separation, we stated each instance of the Replacement
 Axiom separately by means of the

### 2022-09-30 Independence_CH/intro.tex
Meta vs object logic. Appendix order. Metatheory now a word.
Spell-check

@@
 $\AC$), and we formalized the basic forcing notions required to obtain
-ctms of $\ZFC + \neg\CH$ and of $\ZFC + \CH$ (Sect.~\ref{sec:models-ch-negation}). No meta-theoretic issues
+ctms of $\ZFC + \neg\CH$ and of $\ZFC + \CH$ (Sect.~\ref{sec:models-ch-negation}). No metatheoretic issues
 (consistency, FOL calculi, etc) were formalized, so we were mainly
@@
 
-The lastest version of our formalization can be accessed at the
+The latest version of our formalization can be accessed at the
 development site of the AFP, via the following link:

### 2022-09-30 Independence_CH/lam_replacement.tex
Line breaks
Spell-check

--- 2022-09-22 Independence_CH/lam_replacement.tex
@@
 process. As further properties of closure of the ground $M$ were
-required, we gather the relevant instances of Separation and
+required, we gathered the relevant instances of Separation and
 Replacement into a new locale (always assuming a class model, for
-added generality), and proceded to apply them to those closure proofs.
+added generality), and proceeded to apply them to those closure proofs.
 
@@
 \begin{threeparttable}
-\begin{tabular}{r<{\stepcounter{LamReplCount}\theLamReplCount.} >{\hspace{1ex}}l @{\hspace{1em}} l}
+\begin{tabular}{r<{\stepcounter{LamReplCount}\theLamReplCount.} >{\hspace{1ex}}l @{\hspace{0.8em}} l}
   \toprule
@@
 \end{tabular}
-\caption{Replacement and Separation instances of the locale \isa{M{\uscore}remplacement}}
+\caption{Replacement and Separation instances of the locale \isa{M{\uscore}replacement}}
 \label{tab:m-repl-instances}

### 2022-09-30 Independence_CH/main_defs.tex
Spell-check

@@
 In order to state the defining property of the relative
-equipotence relation, we work under the assumptions of the
+equipotency relation, we work under the assumptions of the
 locale \isa{M{\isacharunderscore}{\kern0pt}cardinals}. They comprise a finite set

### 2022-09-30 Independence_CH/main_results.tex
Meta vs object logic. Appendix order. Metatheory now a word.
Spell-check

@@
 \item 2 instances for the definition of
-  ordertypes;
+  ordertypes: The relevant well-founded recursion and a technical
+  auxiliary instance.
 \item 2 instances for Aleph: Replacement through the ordertype function (for Hartogs' Theorem) and the well-founded recursion
@@
 the absoluteness of the recursive construction in the proof of
-Dependent Choices from $\AC$.
+Dependent Choices from $\AC$. A listing with the names of all the formulas
+can be found in Appendix~\ref{sec:repl-instances-appendix}.
 
@@
 \end{isabelle}
-where $\approx$ is equipotence, and the predicate \isa{Transset}
+where $\approx$ is equipotency, and the predicate \isa{Transset}
 holds for

### 2022-09-30 Independence_CH/non-absolute.tex
Spell-check

@@
 To achieve this, we provide automatic tools to ease the definitions of
-such relative versions, their fully relational counteparts, and the
+such relative versions, their fully relational counterparts, and the
 internalized formulas. For instance, consider the

### 2022-09-30 Independence_CH/proof_assists_isabelle.tex
Spell-check

@@
 constructs proofs in response to user scripts or other modules) was reduced to a small
-\emph{kernel}. Hence, by verfying the correctness of the kernel, one
+\emph{kernel}. Hence, by verifying the correctness of the kernel, one
 achieves confidence on the whole system.
@@
 induction and recursion theorems for them are proved automatically
-(this is in constrast to general well-founded recursion, for which
+(this is in contrast to general well-founded recursion, for which
 one has to work with the fundamental recursor $\isa{wfrec}$).}  Its

### 2022-09-30 Independence_CH/related.tex
Meta vs object logic. Appendix order. Metatheory now a word.
Spell-check

@@
 It should be emphasized, however, that the Flypitch project was
-carried out assuming a rather strong meta-theory. 
+carried out assuming a rather strong metatheory.
 Carneiro \cite{carneiro-ms-thesis} reports that Werner's results in
@@
 is based, and in Isabelle/ZF are rather cumbersome. Also, a
-typed discipline provides aid to write succintly and many assumptions
+typed discipline provides aid to write succinctly and many assumptions
 are satisfied by mere notation. To be clear, those benefits come from

### 2022-09-30 Independence_CH/sample_proof.tex
Spell-check

@@
 and proofs in natural language; recently P.~Koepke and his team
-achieved magnificient results by using Isabelle/Naproche
+achieved magnificent results by using Isabelle/Naproche
 \cite{10.1007/978-3-030-81097-9_2} to formalize proofs of several

### 2022-10-01 Independence_CH/arxiv.sh
Explicit bbl for arXiv due to overflows

@@ -2,6 +2,7 @@
 
 rootbase=independence_ch_isabelle
 root=$rootbase.tex
+bbl=$rootbase\_arxiv.bbl
 arxiv_dir=__"$rootbase"_ArXiv_tmp
 arxiv="$rootbase"_arxiv.tex
 tgz="$rootbase"_arxiv.tar.gz
@@ -20,18 +21,22 @@ do
     ../scripts/stripcomments.pl $x > $arxiv_dir/$x
 done
 
-cp *sty *bbl $extra_files $arxiv_dir
+cp *sty $bbl $extra_files $arxiv_dir
 
 cd $arxiv_dir
 
 sed 's/\\bibliographystyle{[^}]*}//g' $root | \
-    sed 's/\\bibliography{'$rootbase'}/\\input{'$rootbase.bbl'}/g'\
+    sed 's/\\bibliography{'$rootbase'}/\\input{'$bbl'}/g'\
  > $arxiv
 rm -f $root
 
-tar --exclude=$root -zcvf $tgz *tex *sty *bbl $extra_files
+tar --exclude=$root -zcvf $tgz *tex *sty $bbl $extra_files
 mv $tgz ..
 
 cd ..
 
 rm -rf $arxiv_dir
+
+echo
+echo "Beware:"
+echo "        Don't forget to update \""$bbl"\" !"

### 2022-10-01 Independence_CH/intro.tex
Style by Miguel

@@
 reader surely understands that, aside from trivial results, no current
-technology allows us to state a reasonably complex (and correct) theorem
-statement and expect to obtain a proof after hitting ``Enter'', at
+technology allows us to write a reasonably complex (and correct)
+theorem statement in a computer and expect to obtain a proof after hitting “Enter”, at
 least not after a humanly feasible wait. On the other hand, it is

### 2022-10-02 Independence_CH/conclusion.tex
Spell-check
Removed apostrophes in verbs and one minor change in style.

@@
 Another desirable goal is to construct transitive set models of $\ZFC$
-from a large cardinal. There's some work to be done for that: Even the
+from a large cardinal. There is some work to be done for that: Even the
 definition of inaccessibles, and of the transitive collapse (for that
@@
 daunting, but at the same time extremely rewarding: The feeling of
-accomplishment after seeing one's writings validated beyond doubt is
+accomplishment after seeing your own writings validated beyond doubt is
 in some ways comparable to that of finding a proof of an important
@@
 paper is once you are fully engaged in directing your assistant). We
-hope that at some point these experiencies are shared by our community
+hope that at some point these experiences are shared by our community
 at large.

### 2022-10-02 Independence_CH/intro.tex
Removed apostrophes in verbs and one minor change in style.

@@
 
-We'll continue our description of proof assistants in
+We will continue our description of proof assistants in
 Section~\ref{sec:proof-assist-isabelle}. We kindly invite the reader
@@
 minutiae of the formalism” on our part, as it was put by Cohen
-\cite{zbMATH02012060}, but we'd rather say that we were driven by
+\cite{zbMATH02012060}, but we would rather say that we were driven by
 curiosity.

### 2022-10-02 Independence_CH/lessons.tex
Line breaks
Removed apostrophes in verbs and one minor change in style.

@@
 Nevertheless, experience  in software engineering is invaluable in
-a large project like the present one. For example, it is
+large projects like the present one. For example, it is
 (mathematically) misleading when automatic tools (\isa{simp},
@@
 goal (v.g., the same statement with 7 variables succeeds but with 8
-variables doesn't). Scale issues are very easily disregarded in the
+variables does not). Scale issues are very easily disregarded in the
 abstract but, as a colorful example, the formula $\forceisa({\cdot}0\in 1{\cdot})$

### 2022-10-02 Independence_CH/axioms_metatheory.tex
Line breaks
Minor edits

@@
 ZF\_Base.infinity: 0 ∈ Inf ∧ (∀y∈Inf. succ(y) ∈ Inf)\isanewline
-ZF\_Base.replacement: ⋀A P b. ∀x∈A. ∀y z. P(x, y) ∧ P(x, z) ⟶ y = z ⟹ \isanewline
-\ \ \ \ \ \ \ \ \ b ∈ PrimReplace(A, P) ⟷ (∃x∈A. P(x, b))
+ZF\_Base.replacement: ⋀A P b. ∀x∈A. ∀y z. P(x, y) ∧ P(x, z) ⟶ y = z \isanewline
+\ \ \ \ \ \ \ \ \ ⟹ b ∈ PrimReplace(A, P) ⟷ (∃x∈A. P(x, b))
 \end{isabelle}
@@
 The theory \theory{AC} is only imported in the theory
-\isa{Absolute{\uscore}Versions}; none of the main results
-depends on $\AC$. The theory \theory{Absolute\_Versions}
+\theory{Absolute\_Versions}; none of the main results
+depends on $\AC$. The latter theory
 shows that some absolute results can be obtained from the

### 2022-10-02 Independence_CH/forcing.tex
Minor edits

@@
 analogous fashion. But, in contrast to the point made on
-p.~260 of this book, the structural recursion to define the forcing
+p.~260 of this book, the structural recursion used to define the forcing
 relation was replaced by one  involving codes for formulas. Thus, the metatheoretic formula

### 2022-10-02 Independence_CH/main_results.tex
Things in the abstract must appear somewhere

@@
 \]
-Here, $\mathit{fst}(\lb a,b\rb) = a$, $\mathit{snd}(\lb a,b\rb) = b$
-(with default value $0$ for non pairs).
-All ground replacement
+Here, $\mathit{fst}(\lb a,b\rb) = a$ and $\mathit{snd}(\lb a,b\rb) = b$.
+% (with default value $0$ for non pairs).
+The map $\phi\mapsto\psi$ is
+the function $\calF$ referred to in the abstract.
+All such “ground” replacement
 instances appear in the locale \locale{M{\uscore}ZF3} and form the set
@@
 The $21$ formulas up to this point are collected into the set
-\isa{overhead{\uscore}notCH}, which is enough to
+\isa{overhead{\uscore}notCH} (called $\Omega$ in the abstract), which is enough to
 force $\neg\CH$. To force $\CH$, we required one further instance for

### 2022-10-03 Independence_CH/main_results.tex
More minor edits

@@
 We isolated 22 instances of Replacement that are sufficient to force
-$\CH$ or $\neg\CH$, which are enumerated below by the name of the
-corresponding internalized first order formula. Many of these were already present in
+$\CH$ or $\neg\CH$. Many of these were already present in
 relational form in the \session{ZF-Constructible} library.
@@
 It is to be noted that the proofs of the Forcing Theorems do not
-require any extra replacement; actually, they only need the 7
+require any extra replacement on the ground model; actually, they only need the 7
 instances appearing in \isa{instances1{\uscore}fms} and

### 2022-10-03 Independence_CH/intro.tex
(Soon) freezed AFP entry

@@
 
-The latest version of our formalization can be accessed at the
-development site of the AFP, via the following link:
+The code of our formalization can be accessed at the
+AFP site, via the following link:
 \begin{center}
-  \url{https://devel.isa-afp.org/entries/Independence_CH.html}
+  \url{https://www.isa-afp.org/entries/Independence_CH.html}
 \end{center}
+
 %%% Local Variables: 

### 2022-10-13 notes/chow.tex
Some words stemming from comments by Tim Chow

Elaboration on comments by Timothy Y. Chow
===========================================

On the relative consistency of $\neg\CH$
----------------------------------------

It is perhaps important to make clear that we have not obtained the
relative consistency of $\neg\CH$, and we are
actually not aiming for that in the short term. In our context, going
for the plain consistency relative to $\ZF$ seems off the mark, since we
can not weaken our base theory (which is essentially equivalent to $\ZF$)%% ,
%% and we catch our tail.
Even if we intended to do so, the standard route to use ctms for a
relative consistency proof (through the Reflection Theorem) is
prohibitive for us, since our metatheory does not have the required
induction principles (on Isabelle/ZF formulas---in contrast to
formulas coded as sets, as in our presentation).

Work by Mathias
---------------

A summary is offered in \cite[Sect.~6]{kanamori-mathias}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In \cite[Sect.~1]{mathias:hal-01188043}, models $M$ of Zermelo Set
Theory are offered that each of the inclusions $M\subseteq M[G]$ and
$M\supseteq M[G]$ fail, where the poset $\pP$ is the trivial
$\{\1 \}$. In one of them, $K$, we have $\om\in K \sm K[G]$, hence the
ordinals do not coincide.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In \cite{mathias-provident}, a reasonably minimal fragment
$\mathsc{Prov}$ of $\ZF$ that allows to do set forcing is identified,
and transitive sets satisfying it are called
\emph{provident}. Existence of rank and of transitive closure are
among them; hence their appearance in our list seems
justified. Nevertheless, $\mathsc{Prov}$ is even a restriction of
Kripke-Platek Set Theory, and thus it does not include neither
Powerset nor full Separation. The detailed theory of provident sets is
developed in \cite{mathias-bowler-gentle}. % Its Section~0 is very
% readable.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2022-10-24 Independence_CH/conclusion.tex
Some comments inspired by Tim Chow added

@@
 \label{sec:acknowledgments}
-We thank Z.~Vidnyánszky for some discussions where the idea of
+We are indebted to Timothy Chow for insightful comments on a draft of
+this paper. We also thank Z.~Vidnyánszky for some discussions where the idea of
 building the summary in \theory{Definitions\_Main} took shape.

### 2022-10-24 Independence_CH/related.tex
Minor edits
Some comments inspired by Tim Chow added

@@
 
+This is perhaps an appropriate time to insist that we have \emph{not}
+formalized the relative consistency of $\neg\CH$, and we are actually
+not aiming for that in the short term. In our context, going for the
+plain consistency result seems off the mark, since we can not weaken
+our base theory (which is essentially equivalent to $\ZF$)%%,
+%% and we catch our tail.
+Even if we intended to do so, the standard route to use ctms for a
+relative consistency proof (through the Reflection Theorem) is
+prohibitive for us, since our metatheory does not have the required
+induction principles (on Isabelle/ZF formulas---in contrast to
+formulas coded as sets, as in our presentation).
+
 We believe that our alternative formalization using the ctm approach
@@
 are satisfied by mere notation. To be clear, those benefits come from
-doing set theory in a non set-theoretical meta-language. On the other
+doing set theory in a non set-theoretical language. On the other
 hand, Isar proofs, as the one shown in
@@
 
+Concerning the minimum amount of Replacement needed to construct
+forcing extensions, only recently we learned about Mathias' work on the
+subject (for which a summary is offered in
+\cite[Sect.~6]{kanamori-mathias}). In
+\cite[Sect.~1]{mathias:hal-01188043}, models $M$ of Zermelo Set Theory
+are offered for which each of the inclusions $M\subseteq M[G]$ and
+$M\supseteq M[G]$ fail, where the poset $\pP$ is the trivial $\{\1
+\}$. In one of them, $K$, we have $\om\in K \sm K[G]$, hence the
+ordinals do not coincide.
+
+Also, in the reference \cite{mathias-provident}, a reasonably minimal
+fragment \textsc{Prov} of $\ZF$ that allows to do set forcing is
+identified, and transitive sets satisfying it are called
+\emph{provident}. Existence of rank and of transitive closure are
+among them; hence their appearance in our list seems
+justified. Nevertheless, \textsc{Prov} is far weaker than the
+fragments of $\ZF$ considered here, since it is a restriction of
+Kripke-Platek Set Theory, and thus it does not include neither
+Powerset nor full Separation. The detailed theory of provident sets is
+developed in \cite{mathias-bowler-gentle}. % Its Section~0 is very
+% readable.
+
 %%% Local Variables: 

### 2022-10-24 notes/chow.tex
Some comments inspired by Tim Chow added

@@
 
-It is perhaps important to make clear that we have not obtained the
+It is perhaps important to make clear that we have not formalized the
 relative consistency of $\neg\CH$, and we are
 actually not aiming for that in the short term. In our context, going
-for the plain consistency relative to $\ZF$ seems off the mark, since we
+for the plain consistency result seems off the mark, since we
 can not weaken our base theory (which is essentially equivalent to $\ZF$)%% ,
@@
 In \cite[Sect.~1]{mathias:hal-01188043}, models $M$ of Zermelo Set
-Theory are offered that each of the inclusions $M\subseteq M[G]$ and
+Theory are offered for which each of the inclusions $M\subseteq M[G]$ and
 $M\supseteq M[G]$ fail, where the poset $\pP$ is the trivial

### 2022-10-24 Independence_CH/intro.tex
Clarifying defs/thms (q by Tim Chow)

@@
 the Archive of Formal Proofs (AFP) based on Isabelle; and formalizations of
-brand new and complex objects like the definition of perfectoid spaces \cite{10.1145/3372885.3373830}
-and the Liquid Tensor Experiment \cite{LTE2020,LTE2021} have been
-achieved using Lean.
+brand new mathematics like the Liquid Tensor Experiment
+\cite{LTE2020,LTE2021} and the definition of perfectoid spaces \cite{10.1145/3372885.3373830}
+have been achieved using Lean.
 

### 2022-10-25 Independence_CH/related.tex
Repeated words

@@
 \cite[Sect.~1]{mathias:hal-01188043}, models $M$ of Zermelo Set Theory
-are offered for which each of the inclusions $M\subseteq M[G]$ and
+are constructed for which each of the inclusions $M\subseteq M[G]$ and
 $M\supseteq M[G]$ fail, where the poset $\pP$ is the trivial $\{\1

### 2022-10-25 Independence_CH/related.tex
Minor

@@
 % by Gödel's Second).
-
 On a lesser note, in order to prove $\AC$ in the generic extension,
@@
 plain consistency result seems off the mark, since we can not weaken
-our base theory (which is essentially equivalent to $\ZF$)%%,
+our base theory (which is essentially equivalent to $\ZF$). %%,
 %% and we catch our tail.
@@
 
-We believe that our alternative formalization using the ctm approach
+We believe that our formalization using the ctm approach
 over Isabelle/ZF might be more appealing to set-theorists because of the
@@
 hand, Isar proofs, as the one shown in
-Sec.~\ref{sec:sample-formal-proof}, are easier to write and understand
+Sec.~\ref{sec:sample-formal-proof}, are easier to understand
 than the language of tactics of Lean.

### 2022-10-25 Independence_CH/intro.tex
Non trivial results & automation

@@
 A formalized proof is not the same as an \emph{automatic proof}. The
-reader surely understands that, aside from trivial results, no current
+reader surely understands that, aside from results of a very specific sort, no current
 technology allows us to write a reasonably complex (and correct)

### 2022-10-27 Independence_CH/lessons.tex
Non trivial results & automation
Some more words on countable closure

@@
   closed forcings (in fact, we formalized the bare minimum requirement
-  of being $(<\omega+1)$-closed, that is, under plain $\omega$-sequences).
+  of being $({<}\omega{+}1)$-closed, that is, closure under
+  $\omega$-sequences and not $\delta$-sequences for every countable $\delta$).
 \end{itemize}
@@
 results. The takeaway is that intuition may drive proofs
-even if you are not working on what you believe we are.
+even if you are not working on what you think you are.
 

### 2022-10-27 Independence_CH/arxiv.sh
Fails if biblio dates are wrong

@@ -8,6 +8,12 @@ arxiv="$rootbase"_arxiv.tex
 tgz="$rootbase"_arxiv.tar.gz
 extra_files="llncs.cls"
 
+if [ $rootbase.bbl -nt $bbl ]
+then
+   echo "Error: "$rootbase.bbl" newer than "$bbl"!"
+   exit 1
+fi
+
 pdflatex -draftmode \\nonstopmode \\input $root > /dev/null
 bibtex $rootbase
 makeindex $rootbase

### 2022-10-27 Independence_CH/independence_ch_isabelle.tex
Spell-check
Changed email for Miguel.
Things in the abstract must appear somewhere
Abstract and capitalization of Zst

@@
     Conicet. C\'ordoba. Argentina. \\
-    \email{\{gunther,pagano,sterraf\}@famaf.unc.edu.ar\\
-        matias.steinberg@mi.unc.edu.ar}
+    \email{\{gunther,sterraf\}@famaf.unc.edu.ar\\
+        \{miguel.pagano,matias.steinberg\}@unc.edu.ar}
 }
@@
   defined $\calF:\calR\to\calR$
-  such that for every finite $\Phi\sbq\calR$
-  we have
+  such that for every $\Phi\sbq\calR$ and $M$-generic $G$,
   $M\models \ZC \cup \calF\dimg\Phi \cup \Omega$ implies
-  $M[G]\models \ZC \cup \Phi \cup \{ \neg \CH \}$ for any $M$-generic $G$, where $\ZC$ is
+  $M[G]\models \ZC \cup \Phi \cup \{ \neg \CH \}$, where $\ZC$ is
   Zermelo set theory with Choice.
-  We also obtained the formulas yielded by the Forcing Definability Theorem
-  explicitly.
+  %% We also obtained the formulas yielded by the Forcing Definability Theorem
+  %% explicitly.
 
   To achieve this, we worked in the proof assistant \emph{Isabelle},
-  basing our development on the theory Isabelle/ZF by L.~Paulson and
+  basing our development on the Isabelle/ZF library by L.~Paulson and
   others.

### 2022-10-27 Independence_CH/related.tex
Abstract and capitalization of Zst

@@
 \cite[Sect.~6]{kanamori-mathias}). In
-\cite[Sect.~1]{mathias:hal-01188043}, models $M$ of Zermelo Set Theory
+\cite[Sect.~1]{mathias:hal-01188043}, models $M$ of Zermelo set theory
 are constructed for which each of the inclusions $M\subseteq M[G]$ and

### 2022-10-29 Independence_CH/related.tex
Minor corrections

@@
 \emph{provident}. Existence of rank and of transitive closure are
-among them; hence their appearance in our list seems
+implied by \textsc{Prov}; hence their appearance in our list seems
 justified. Nevertheless, \textsc{Prov} is far weaker than the
 fragments of $\ZF$ considered here, since it is a restriction of
-Kripke-Platek Set Theory, and thus it does not include neither
+Kripke-Platek set theory, and thus it does not include neither
 Powerset nor full Separation. The detailed theory of provident sets is

### 2022-10-30 Independence_CH/related.tex
Related & corrected bib. Pruned header

@@
 range of results in combinatorics and other set-theoretical material
-that was swiftly formalized in this setting: Nash-Williams theorem and
+that was swiftly formalized in this setting: Erdős-Milner
+partition theorem \cite{2021arXiv210411613P}, Nash-Williams theorem and
 Larson's $\forall k\ \omega^{\omega}\longrightarrow(\omega^\omega,k)$

### 2023-01-04 Independence_CH/axioms_metatheory.tex
Some comments inspired by Tim Chow added
Repeated words
Minor formatting edits

@@
 \subsection{IFOL and FOL}
-In the axioms \isa{refl, subst, allI, spec, exE, exI,eq\_reflection} there is a constraint
+In the axioms \isa{refl}, \isa{subst}, \isa{allI}, \isa{spec},
+\isa{exE}, \isa{exI}, and \isa{eq\_reflection} there is a constraint
 for the \emph{type of} the variables \isa{a, b, x} to be in the class \isa{term\_class}.
@@
 \end{isabelle}
-\noindent with these axioms
+\noindent After the definitions of $\notin$, $\subseteq$, $\isa{succ}$,
+and relative quantifications are presented, the following axioms are postulated:
 \begin{isabelle}

### 2023-01-04 Independence_CH/lam_replacement.tex
Minor edits
Minor formatting edits

@@
 
-Faced with this situation, we decided that we needed some sort of
+Facing this situation, we decided that we needed some sort of
 \emph{compositionality} in order to obtain new instances from the ones
@@
 \begin{threeparttable}
-\begin{tabular}{r<{\stepcounter{LamReplCount}\theLamReplCount.} >{\hspace{1ex}}l @{\hspace{0.8em}} l}
+\begin{tabular}{r<{\stepcounter{LamReplCount}\theLamReplCount.} >{\hspace{1ex}}l @{\hspace{0.8em}} p{16.3em}}
   \toprule
@@
   & \isa{lam{\uscore}replacement{\uscore}prodRepl} &
-        $\lamRepl{\langle \mathit{snd}(\mathit{fst}(x)),\langle \mathit{fst}(\mathit{fst}(x)),\mathit{snd}(\mathit{snd}(x))\rangle \rangle}$\\
+  $x \mapsto \bigl\langle \mathit{snd}(\mathit{fst}(x)),\bigr.$
+  \newline
+  $\bigl.\phantom{x \mapsto \langle \mathit{snd}}\langle \mathit{fst}(\mathit{fst}(x)),\mathit{snd}(\mathit{snd}(x))\rangle \bigr\rangle$\\
   \midrule

### 2023-09-27 Independence_CH/lessons.tex
Offending comment on LTE out

@@
 efficacy. Another crucial aspect to achieve this goal is the level of
-detail of the blueprint for the formalization. We must however confess
+detail of a blueprint for the formalization.
+
+We must however confess
 that we learned many of the subtleties of Isabelle in the making, and
@@
 precise way things would develop in the future.
-
-A similar experience, but on an opposite side of the formalization
-spectrum happened to the Liquid Tensor Experiment \cite{nature-lte} as described by Scholze
-in \cite{LTE2021}. People involved in the formalization simply pushed
-their way to reach the summit, formalizing lemma after lemma. They
-actually wrote the blueprint for that formalization \emph{afterwards}
-it was complete! From time to time, we were also frenziedly trying to
+From time to time, we were frenziedly trying to
 get the results formalized, going beyond what we had planned.

### 2023-09-27 Independence_CH/non-absolute.tex
Minimal explanation of the type of cardinal_rel

@@
 $\isa{cardinal{\uscore}rel}::(\tyi \fun \tyo) \fun \tyi \fun\tyi$
-(denoted  $|\cdot|^M$, as expected),
+(whose first argument is the class $M$, and is denoted $|\cdot|^M$ as expected),
 the relational version $\isa{is{\uscore}cardinal}$ of the latter, the

### 2023-09-27 Independence_CH/axioms_metatheory.tex
Explanation of Hessenberg and minor edits

@@
 \isa{exE}, \isa{exI}, and \isa{eq\_reflection} there is a constraint
-for the \emph{type of} the variables \isa{a, b, x} to be in the class \isa{term\_class}.
+for the \emph{type} of the variables \isa{a, b, x} to be in the class \isa{term\_class}.
 \begin{isabelle}

### 2023-09-28 Independence_CH/forcing.tex
First typo spotted after arXiv v1
Explanation of separation, strong_repl, blowup, ack Commelin

@@
 For stating the axiom schemes, \session{ZF-Constructible} defines
-the expressions
+$\isa{separation}(N,Q)$:
 \[
-  \isa{separation}(N,Q)
-  \text{ and }
-  \isa{strong{\uscore}replacement}(N,R)
+  \forall z[N].\ \exists y[N].\ \forall
+  x[N].\ x\in y \longleftrightarrow x\in z \land Q(x),
 \]
+and $\isa{strong{\uscore}replacement}(N,R)$:
+\begin{multline*}
+  \forall
+  A[N].\ \isa{univalent}(N,A,R) \longrightarrow \\
+  (\exists Y[N].\ \forall b[N].\ b\in Y\longleftrightarrow (\exists x[N].\ x\in
+  A \land R(x,b))),
+\end{multline*}
+%% \begin{isabelle}%
+%% separation{\isacharparenleft}{\kern0pt}N{\isacharcomma}{\kern0pt}\ Q{\isacharparenright}{\kern0pt}\ {\isasymequiv}\ {\isasymforall}z{\isacharbrackleft}{\kern0pt}N{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymexists}y{\isacharbrackleft}{\kern0pt}N{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymforall}x{\isacharbrackleft}{\kern0pt}N{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ x\ {\isasymin}\ y\ {\isasymlongleftrightarrow}\ x\ {\isasymin}\ z\ {\isasymand}\ Q{\isacharparenleft}{\kern0pt}x{\isacharparenright}{\kern0pt}%
+%% \end{isabelle}%
+%% \begin{isabelle}%
+%% strong{\isacharunderscore}{\kern0pt}replacement{\isacharparenleft}{\kern0pt}N{\isacharcomma}{\kern0pt}\ R{\isacharparenright}{\kern0pt}\ {\isasymequiv}\isanewline
+%% {\isasymforall}A{\isacharbrackleft}{\kern0pt}N{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\isanewline
+%% \isaindent{\ \ \ }univalent{\isacharparenleft}{\kern0pt}N{\isacharcomma}{\kern0pt}\ A{\isacharcomma}{\kern0pt}\ R{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymexists}Y{\isacharbrackleft}{\kern0pt}N{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymforall}b{\isacharbrackleft}{\kern0pt}N{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ b\ {\isasymin}\ Y\ {\isasymlongleftrightarrow}\isanewline
+%% \ \ \ \ \ \ {\isacharparenleft}{\kern0pt}{\isasymexists}x{\isacharbrackleft}{\kern0pt}N{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ x\ {\isasymin}\ A\ {\isasymand}
+%% R{\isacharparenleft}{\kern0pt}x{\isacharcomma}{\kern0pt}\ b{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}%
+%% \end{isabelle}%
 whose first argument $N$ is a class and their second arguments $Q$ and
-$R$ are predicates of types $R :: \tyi \fun \tyo$ and $Q ::
-\tyi \fun \tyi \fun \tyo$, respectively. The Separation Axiom appears in
+$R$ are predicates of types $Q :: \tyi \fun \tyo$ and $R ::
+\tyi \fun \tyi \fun \tyo$, respectively (the \isatt{univalent}
+predicate states that $R$ is functional in its first variable; see Appendix~\ref{sec:def-main-relative}). The Separation Axiom appears in
 \locale{M{\uscore}Z{\uscore}basic} as follows, where 
@@
 \ \ arity{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isadigit{2}}\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ length{\isacharparenleft}{\kern0pt}env{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\isanewline
-\ \ \ \ strong{\isacharunderscore}{\kern0pt}replacement{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}M{\isacharcomma}{\kern0pt}{\isasymlambda}x\ y{\isachardot}{\kern0pt}\ {\isacharparenleft}{\kern0pt}M\ {\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}x{\isacharcomma}{\kern0pt}y{\isacharbrackright}{\kern0pt}{\isacharat}{\kern0pt}env\ {\isasymTurnstile}\ {\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}
+\ \ \ \ strong{\isacharunderscore}{\kern0pt}replacement{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}M{\isacharcomma}{\kern0pt}{\isasymlambda}x\ y{\isachardot}{\kern0pt}\ {\isacharparenleft}{\kern0pt}M\ {\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}x{\isacharcomma}{\kern0pt}y{\isacharbrackright}{\kern0pt}{\isacharat}{\kern0pt}env\ {\isasymTurnstile}\ {\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose},
 \end{isabelle}
@@
 %% \end{multline}
+which encodes the instance of replacement for the model $M$
+corresponding to the predicate $R(x,y)$ given by
+$\phi^M(x,y,x_0,\dots,x_n)$, where $\isatt{env} = [x_0,\dots,x_n]$.
+
 In turn, the \isa{{\isasymcdot}Replacement{\isasymcdot}} function

### 2023-10-02 Independence_CH/cofinality.tex
Comments on DSL

@@
-\section{Recursions in cofinality}\label{sec:recursions-cofinality}
+\section{Recursions in cofinality and the Delta System Lemma}\label{sec:recursions-cofinality}
 
@@
 Lemma by “algebraic” reasoning only.
+We expect that the relativization of these
+results be straightforward, when time permits.
 
-We therefore expect that the relativization of these
-results be straightforward, when time permits.
+The AFP entry \cite{Delta_System_Lemma-AFP} also includes the
+formalization of the (absolute) Delta System Lemma (DSL). Formalizing its
+proof was rather straightforward, once the many prerequisites were
+taken care of. Some of those were really basic, for instance:
+\begin{itemize}
+\item $\omega$ injects into every infinite set;
+\item surjective images of countable sets are countable;
+\item the union over a countable index set $J$ of a family $X :: \tyi
+  \fun \tyi$ of countable sets is countable.
+\end{itemize}
+It was also convenient to isolate the relevant recursive construction
+principle (\isatt{bounded{\uscore}cardinal{\uscore}selection}) that
+appears in the proof of DSL, which was also useful for showing the
+first item:
+\begin{lemma}
+  Assume $F$ is nonempty,
+  $\gamma$ is a cardinal, and $Q$ is a binary predicate over $F$ satisfying 
+  \[
+    \forall X \sbq F.\ |X| < \gamma \implies \exists a\in F.\ \forall
+    x\in X.\ Q(s,a).
+  \]
+  Then there exists  $S:\gamma \to F$ such that
+  $Q(S(\alpha),S(\beta))$ for all $\alpha<\beta<\gamma$.
+\end{lemma}
+
+Concerning the relativization of the proof of DSL for its use in
+forcing, it required inserting proofs that all the relevant objects
+lied in the model $M$; this was only tedious. But the point where the
+last item above was used required some expansion due to an application
+of Replacement (the relevant $X$ was a class function).
 

### 2023-10-02 Independence_CH/related.tex
Comment on relative consistency

@@
 not aiming for that in the short term. In our context, going for the
-plain consistency result seems off the mark, since we can not weaken
-our base theory (which is essentially equivalent to $\ZF$). %%,
+plain consistency result (that is, proving the arithmetical statement
+$\Con(ZFC) \implies \Con(ZFC + \neg\CH)$ over a weak base theory)
+seems off the mark, since our base is fixed (and essentially equivalent to $\ZF$). %%,
 %% and we catch our tail.
-Even if we intended to do so, the standard route to use ctms for a
-relative consistency proof (through the Reflection Theorem) is
-prohibitive for us, since our metatheory does not have the required
-induction principles (on Isabelle/ZF formulas---in contrast to
-formulas coded as sets, as in our presentation).
+If we intended to do so, we would a need a development of a weak base with a
+new type of (external) formulas having the required
+induction principles, plus an interface from this new type to
+Isabelle/ZF. This might allow us to take the standard route to use ctms for a
+relative consistency proof (through a version of the Reflection
+Theorem proved by induction on the external formulas).
 
-We believe that our formalization using the ctm approach
+Back to the comparison, we believe that our formalization using the ctm approach
 over Isabelle/ZF might be more appealing to set-theorists because of the

### 2023-10-04 Independence_CH/axioms_metatheory.tex
Probably final revision
Co-authored-by: Miguel Pagano <miguel.pagano@gmail.com>

@@
 
-\subsection{The metatheory Pure}
+\subsection{The metatheory $\Meta$}
+Below, the \isatt{PROP} operator corresponds to the
+injection $[\cdot]$ of Section~\ref{sec:isabelle-metalogic-meta}.
 \begin{isabelle}
@@
 \isa{exE}, \isa{exI}, and \isa{eq\_reflection} there is a constraint
-for the \emph{type} of the variables \isa{a, b, x} to be in the class \isa{term\_class}.
+on the \emph{type} of the variables \isa{a, b, x}: It should be in the class \isa{term\_class}.
 \begin{isabelle}

### 2023-10-04 Independence_CH/cofinality.tex
Probably final revision
Co-authored-by: Miguel Pagano <miguel.pagano@gmail.com>

--- 2022-08-12 Independence_CH/cofinality.tex
@@
 Lemma by “algebraic” reasoning only.
+We expect the relativization of these
+results to be straightforward.
 
-We therefore expect that the relativization of these
-results be straightforward, when time permits.
+The AFP entry \cite{Delta_System_Lemma-AFP} also includes the
+formalization of the (absolute) Delta System Lemma (DSL). Formalizing its
+proof was rather straightforward, once the many prerequisites were
+taken care of. Some of those were really basic, for instance:
+\begin{enumerate}
+\item \label{item:1}$\omega$ injects into every infinite set;
+\item \label{item:2}surjective images of countable sets are countable;
+\item \label{item:3}the union over a countable index set $J$ of a family $X :: \tyi
+  \fun \tyi$ of countable sets is countable.
+\end{enumerate}
+It was also convenient to isolate the relevant recursive construction
+principle (\isatt{bounded{\uscore}cardinal{\uscore}selection}) that
+appears in the proof of DSL, which was also useful for showing
+Item~\ref{item:1}:
+\begin{lemma}\label{lem:bdd-card-selection}
+  Assume $F$ is nonempty,
+  $\gamma$ is a cardinal, and $Q$ is a binary predicate over $F$ satisfying 
+  \[
+    \forall Y \sbq F.\ |Y| < \gamma \implies \exists a\in F.\ \forall
+    y\in Y.\ Q(y,a).
+  \]
+  Then there exists  $S:\gamma \to F$ such that
+  $Q(S(\alpha),S(\beta))$ for all $\alpha<\beta<\gamma$.
+\end{lemma}
+
+Concerning the relativization of the proof of DSL for its use in
+forcing, it required inserting proofs that all the relevant objects
+lied in the model $M$; this was only tedious.
+A bit more effort was required at the point where Item~\ref{item:3}
+was used, because it involved an application of Replacement (being $X$
+a class function); relativizing
+Lemma~\ref{lem:bdd-card-selection} also required some work, because of the
+recursion.
 

### 2023-10-04 Independence_CH/lessons.tex
Explanation of separation, strong_repl, blowup, ack Commelin
Probably final revision
Co-authored-by: Miguel Pagano <miguel.pagano@gmail.com>

@@
 
-Nearly the final stage of the project, we decided to go for the minimal
+Upon reaching the final stage of the project, we decided to go for the minimal
 set of definitions and versions of lemmas that were needed to obtain
@@
 can not.
+(For this particular example, one reason for the blow-up in
+length comes from the encoding of negation using the “nand”
+connective, but otherwise the formulas grow bigger by the sheer
+complexity of the definition of the forcing relation.)
 

### 2023-10-04 Independence_CH/proof_assists_isabelle.tex
Explanation of Hessenberg and minor edits
Probably final revision
Co-authored-by: Miguel Pagano <miguel.pagano@gmail.com>

@@
 
-Isabelle/ZF reaches essentially Hessenberg's $|A|\cdot|A| = |A|$. Our decision
+The amount of set theory developed in “vanilla” Isabelle/ZF is quite
+modest, essentially going no further than Hessenberg's theorem on the
+cardinality of products $|A\times A|=|A|\cdot|A| = |A|$. Instead, our decision
 (during 2017) to

### 2023-10-04 Independence_CH/related.tex
Probably final revision
Co-authored-by: Miguel Pagano <miguel.pagano@gmail.com>

@@
 %% and we catch our tail.
-If we intended to do so, we would a need a development of a weak base with a
+If we intended to do so, we would need a development of a weak base with a
 new type of (external) formulas having the required

### 2023-10-04 Independence_CH/cover_letter.tex
Cover letter with replies to reviewer

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
%\usepackage[utf8]{inputenc}
%\usepackage[final]{microtype}
\usepackage{isabelle_indepCH,isabellesym_indepCH}
\usepackage{booktabs,array,threeparttable}
\input{header-indepCH}
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
\hypersetup{
  colorlinks,
  urlcolor={blue},
  linkcolor={blue!50!black},
  citecolor={blue!50!black},
}
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
% \renewcommand\UrlFont{\color{blue}\rmfamily}
\input{unicode_support}
\usepackage{array}
\newcounter{replInstCount}
\setcounter{replInstCount}{0}

\newcounter{LamReplCount}
\setcounter{LamReplCount}{0}
\begin{document}
%
\title{Cover Letter: Replies to Referee
}
%
\titlerunning{Formalization of ctm forcing}%Lessons after formalizing ctm forcing}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Emmanuel Gunther\inst{1} \and
Miguel Pagano\inst{1} \and \\
Pedro Sánchez Terraf\inst{1,2}%\orcidID{0000-0003-3928-6942}
\and
Matías Steinberg\inst{1}
}
%
\authorrunning{E.~Gunther, M.~Pagano, P.~Sánchez Terraf, M.~Steinberg}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Universidad Nacional de C\'ordoba. 
  \\  Facultad de Matem\'atica, Astronom\'{\i}a,  F\'{\i}sica y
  Computaci\'on.
  \and
    Centro de Investigaci\'on y Estudios de Matem\'atica (CIEM-FaMAF),
    Conicet. C\'ordoba. Argentina. \\
    \email{\{gunther,sterraf\}@famaf.unc.edu.ar\\
        \{miguel.pagano,matias.steinberg\}@unc.edu.ar}
}
%
\maketitle              % typeset the header of the contribution
%
First and foremost, we are very grateful
for your comments and pointers. After the comments in this cover
letter, you may find useful a version of the paper with highlighted
changes.

\begin{verbatim}
p 5 l 33: Isabelle Z/F reaches Hessenberg's |A|.|A| = |A|: I
 have no idea what this means.
\end{verbatim}

We expanded that section.

\begin{verbatim}
p 7 l 40: It would help to explain the type of cardinal_rel, in
 particular explain that the first argument is M.
\end{verbatim}

Same.

\begin{verbatim}
p 8 l 19: You haven't said what a locale is; it would help to
 explain that the text that follows the term is an explanation
 of that.
\end{verbatim}

We did, in the last paragraph of Sect.~2.2. Should we expand on
that?

\begin{verbatim}
p 8 bottom: it would help to show separation(N, Q) and
 strong_replacement(N, R) and explain the formula at the bottom
 of the page.
\end{verbatim}

Please indicate if the edition makes things clearer.

\begin{verbatim}
The example on page 11 is nice.
\end{verbatim}

Much appreciated!

\begin{verbatim}
p 13: it would be nice to say more about the Delta system lemma
 and its proof (and whether it was relatively hard or easy
 compared to other aspects of the formalization).
\end{verbatim}

We added comments to Appendix C.

\begin{verbatim}
p 16 last sentence of the first full paragraph ("Even if we
 intended ..."): I don't understand what is going on here. Do
 you mean relative consistency in terms of probability?  Is the
 difficult proving soundness and completeness of first-order
 logic?  You should clarify what this paragraph aims to get
 across.
\end{verbatim}

We meant a finitary, relative consistency proof. We hope now
that is better expressed on the manuscript.

\begin{verbatim}
p 17 toward the bottom: "They actually wrote the blueprint for
 the formalization *afterwards* it was complete": this is
 false. Commelin and Topaz have given many talks explaining how
 the blueprint was used, and how essential it was to guiding the
 collaboration. The project made key use of an online version of
 the blueprint that showed the status of the formalization and
 provided links from the informal text to the formal proof. The
 prior sentence sounds wrong too. Even Scholze's post describes
 the process of revising the proof and discovering new
 approaches as they proceeded. If the authors want to make
 claims about LTE, they should check the claims with a member of
 the project to make sure they are accurate.
\end{verbatim}

We checked with Commelin and he acknowledged that there was some
back and forth between the formalization effort and the
blueprint. But he did not confirm that they shared the same kind of
“feeling” we were trying to convey in that paragraph, so we deleted
the comment.

\begin{verbatim}
p 19 l -15: It would be helpful to say something about where the
 blowup in length comes from.
\end{verbatim}

Done.

\begin{verbatim}
  If length is an issue, the appendices could be shortened, but
 the data is interesting and I recommend keeping them if
 possible.
\end{verbatim}

Thanks again for your support.
\end{document}

%%% Local Variables: 
%%% mode: latex
%%% ispell-local-dictionary: "american"
%%% End:

### 2023-11-08 Independence_CH/lessons.tex
LTE at nature; style

@@
 (For this particular example, one reason for the blow-up in
-length comes from the encoding of negation using the “nand”
+length comes from the encoding of negation using the \isatt{Nand}
 connective, but otherwise the formulas grow bigger by the sheer

### 2024-02-24 Independence_CH/cover_letter.tex
Fragments of diff
Co-authored-by: Miguel Pagano <miguel.pagano@gmail.com>

@@
 %
-First and foremost, we are very grateful
-for your comments and pointers. After the comments in this cover
-letter, you may find useful a version of the paper with highlighted
-changes.
+First and foremost, we are very grateful for your comments and
+pointers. After the replies in this cover letter, we show the
+fragments of the paper with changes; additions and deletions are
+highlighted in context. We hope they are useful for you.
+
+Please give us notice if any of the corrections is insufficient.
 
